### 2017-10-30 submission/manuscript.Rmd
first commit of manuscript

---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1.0in
bibliography: references.bib
---

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../code/functions.R')

loadLibs(c("tidyverse", "scales"))

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)
opts_chunk$set("message" = FALSE)

```

```{r rdata_tables}

```

# Normalization of the microbiota in patients after treatment for colonic lesions

\begin{center}
\vspace{25mm}

Marc A Sze${^1}$ and Patrick D Schloss${^1}$${^\dagger}$

\vspace{20mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI

\end{center}

Co-author e-mails:

* marcsze@med.umich.edu

\newpage
\linenumbers

## Abstract

**Background.** Colorectal cancer is a worldwide health problem. Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma. This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon. We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`).

**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma. The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`). Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups. Although the adenoma and carcinoma models could reliably differentiate between the pre and post-treatment samples (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.001), the advanced-adenoma model could not (P-value = `r format(filter(treat_random_pvalues, study == "srn")[, "pvalue"], digits = 2, nsmall = 2)`). Furthermore, there was little overlap between the OTUs that were indicative of each treatment. To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas. We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples. Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.001, invisible("<"), invisible(">"))` 0.001). Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).

**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will eventually be validated that can be used to quantify the risk of recurrence and the likelihood of survival. Although it was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma, this was not the case for carcinomas. Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.

### Keywords

microbiota; colorectal cancer; polyps; treatment; risk factor.

\newpage

## Background

Colorectal cancer (CRC) is the third most common cause of cancer deaths in the United States [@siegel_cancer_2016; @haggar_colorectal_2009]. Disease mortality has significantly decreased, predominately due to improvements in screening [@haggar_colorectal_2009]. Despite these improvements, there are still approximately 50,000 CRC-related deaths per year in the United States [@siegel_cancer_2016]. Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 35% of all patients will die within five years [@hellinger_reoperation_2006; @ryuk_predictive_2014; @national_cancer_institute_seer_nodate]. Identification of methods to assess patients' risk of recurrence is of great importance to reduce mortality and healthcare costs.

There is growing evidence that the gut microbiota is involved in the progression of CRC. Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009]. Furthermore, studies that shift the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces have shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014]. Collectively, these studies support the hypothesis that the microbiota can alter the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].

Building upon this evidence, several human studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015]. One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012]. Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014]. These studies have associated *Fusobacterium nucleatum* and other oral pathogens with colonic lesions (adenoma, advanced adenoma, and carcinoma). They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions. This suggests that gut bacteria have a role in tumorigenesis with potential as useful biomarkers for aiding in the early detection of disease [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012;@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014].

Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota. If the microbial community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would not only remove the lesion, but also the microbiota that promoted the tumorigenesis and hence the risk of recurrence. To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner? If so, does the treatment alter the community to more closely resemble that of individuals with normal colons?

We answered these questions by sequencing the V4 region of 16S rRNA genes amplified from fecal samples of individuals with adenoma, advanced adenoma, and carcinomas pre and post-treatment. We used classical community analysis to compare the alpha and beta-diversity of communities pre and post-treatment. Next, we generated Random Forest models to identify bacterial populations that were indicative of treatment for each diagnosis group. Finally, we measured the predictive probabilities to assess whether treatment yielded bacterial communities similar to those individuals with normal colons.  We found that treatment alters the composition of the gut microbiota and that, for those with carcinomas, the gut microbiota shifted more towards that of a normal colon after treatment.  In the individuals with carcinomas, no difference was found by the type of treatment (surgery alone, surgery with chemotherapy, surgery with chemotherapy and radiation).  Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence and the likelihood of survival.

\newpage

## Results

***Treatment for colonic lesions alters the bacterial community structure***
Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1]. None of the individuals in this study had any recorded antibiotic usage that was not associated with surgical treatment of their respective lesion. The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A]. We found that the communities obtained pre and post-treatment among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001). There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05). Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D]. We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value = `r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value = `r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value = `r beta_summary["Carcinoma", "PERMANOVA"]`). Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].

***The treatment of lesions are not consistent across diagnosis groups.***
We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group. First, we sought to identify individual OTUs that could account for the change in overall community structure. However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) > 0.05, invisible(">"), invisible("<"))` 0.05). It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test. We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups. The adenoma and carcinoma models performed well (adenoma AUC range = `r format(1 - max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(1 - min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` and carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`); however, the model for patients treated for advanced adenomas was not able to reliably differentiate between the pre and post-treatment samples (advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`). Interestingly, the top 10 most important OTUs by MDA that were used for each model had little overlap with each other [Figure 2]. Although treatment had an impact on the overall community structure, the effect of treatment was not consistent across patients and diagnosis groups. Both the adenoma and carcioma treatment models had AUCs that were significantly higher than a random model permutation (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.0001).

***Post-treatment samples from patients with carcinoma more closely resemble those of a normal colon.***
Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons. To test this, we used an expanded cohort of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma [Table 2]. We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon. The models performed moderately with CRC being the best (adenoma AUC range = `r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S1). The OTUs that were in the top 10% of importance for the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A]. Among the OTUs that were shared across the three models were those populations generally considered beneficial to their host (e.g. *Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B]. Although many of important OTUs in the top 10% were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S2] with some individuals showing a marked decrease in relative abundance [Figure S3]. Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model. A decrease in the positive probability would indicate that the microbiota more closely resembled that of a patient with a normal colon. There was no significant change in the positive probability for the adenoma or advanced adenoma groups (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] > 0.05 & all_probs_paired_wilcox["SRN", "BH_correction"] > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4]. The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals (P-value = `r format(round(all_probs_paired_wilcox["carcinoma", "BH_correction"], digits = 3))`). Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit. These results indicate that, although there were changes in the microbiota associated with treatment, those experienced by patients with carcinoma after treatment yielded gut bacterial communities of greater similarity to that of a normal colon.

***Difficult to identify effects of specific treatments on the change in the microbiota.***
The type of treatment that the patients received varied across diagnosis groups. Those with adenomas and advanced adenomas received surgical resection (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "Y")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "Y")))`) or polyp removal during colonoscopy (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "N")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "N")))`) and those with carcinomas received surgical resection (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "no")))`), surgical resection with chemotherapy (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "no")))`), and surgical resection with chemotherapy and radiation (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "yes")))`). Regardless of treatment used there was no significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05). Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`). Finally, the change in the positive probability was not significantly different between any of the treatment groups (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`). Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota.

\newpage

## Discussion

Our study focused on comparing the microbiota of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment. For all three groups of patients, we observed changes in their microbiota. Some of these changes, specifically for adenoma, may be due to normal temporal variation, however, those with advanced adenoma and carcinoma clearly had large microbiota changes. After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups. This change resulted in communities that more closely resembled those of patients with a normal colon.  This may suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at reducing the associated bacterial communities.  Understanding the effect of treatment on the microbiota of those diagnosed with carcinomas may have important implications for reducing disease recurrence.  It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to also assess the risk of recurrence due to these changes in the microbiota.

Patients diagnosed with adenoma and advanced adenoma, however, did not experience a shift towards a community structure that resembled those with normal colons. This may be due to the fundamental differences between the features of adenomas and advanced adenomas and carcinoma. Specifically, carcinomas may create an inflammatory milieu that would impact the structure of the community and removal of that stimulus would alter said structure. It is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle. This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S1]. Given the irregular distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.

There was a subset of patients (`r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the `r length(rownames(filter(good_metaf, dx == "cancer")))` with carcinomas) who demonstrated an elevated probability of carcinoma after treatment. This may reflect an elevated risk of recurrence. The `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]). We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage I, `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage II, and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[3]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients. We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` (`r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"])["25%"], digits = 0)` - `r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"])["75%"], digits = 0)`) days) or decreased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` (`r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])["25%"], digits = 0)` - `r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])["75%"], digits = 0)`) days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`) (all days data displayed as median (IQR)). Finally, it is possible that these patients may not have responded to treatment as well as the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way. Again, further studies looking at the role of the microbiota in recurrence are needed to understand the dynamics following treatment.

Our final hypothesis was that the specific type of treatment altered the structure of the microbiome. The treatment to remove adenomas and advanced adenomas was either polyp removal or surgical resection whereas it was surgical resection alone or in combination with chemotherapy or with chemotherapy and radiation for individuals with carcinoma. Because chemotherapy and radiation target rapidly growing cells, these treatments would be more likely to cause a turnover of the colonic epithelium driving a more significant change in the structure of the microbiota. Although, we were able to test for an effect across these specific types of treatment, the number of patients in each treatment group was relatively small.  Finally, those undergoing surgery would have received antibiotics and this may be a potential confounder.  However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(median(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` (`r format(quantile(filter(good_metaf, dx == "cancer")[, "time"])["25%"], digits = 0)` - `r format(quantile(filter(good_metaf, dx == "cancer")[, "time"])["75%"], digits = 0)`) days). We also found no difference in the community structure of those that received surgery and those that did not as a treatment for adenoma or advanced adenoma.

## Conclusion

This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions. We were surprised by the lack of a consistent signal that was associated with treatment of patients with adenomas or advanced adenomas. The lack of a large effect size may be due to differences in the role of bacteria in the formation of adenomas and carcinomas or it could be due to differences in the behaviors and medications within these classes of patients. One of the most exciting of these future directions is the possibility that markers within the microbiota could be used to potentially evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma. If such an approach is effective, it might be possible to target the microbiota as part of adjuvant therapy, if the biomarkers identified play a key role in the disease process. Our data provides additional evidence on the importance of the microbiota in tumorigenesis by addressing the recovery of the microbiota after treatment and opens interesting avenues of research into how these changes may affect recurrence.

\newpage

## Methods

***Study Design and Patient Sampling.***
Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016]. Briefly, samples were stored on ice for at least 24h before freezing. Although we cannot exclude that this sampling protocol may have impacted the gut microbiota composition all samples were subjected to the same methodology. Study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis. Samples used to build the models for prediction were collected either prior to a colonoscopy or between one and two weeks after initial colonoscopy. The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013]. Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 1]. Patients were diagnosed by colonoscopic examination and histopathological review of any biopsies taken. Patients were classified as having advanced adenoma if they had an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology. This study was approved by the University of Michigan Institutional Review Board. All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration. The original protocol for the study did not provide for tracking patients after the follow up samples and so it was not possible for us to ascertain their diagnosis after the completion of the study.

***Treatment.***
For this study treatment refers specifically to the removal of a lesion with or without chemotherapy and radiation. The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surgically [Table 1] but rather via colonoscopy. All patients diagnosed with carcinomas were treated with at least surgery or a combination of surgery and chemotherapy or surgery, chemotherapy, and radiation. The type of chemotherapy used for patients with CRC included Oxaliplatin, Levicovorin, Folfox, Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin. These were used individually or in combination with others depending on the patient [Table 1]. If an individual was treated with radiation they were also always treated with chemotherapy. Radiation therapy generally used 18 mV photons for treatment.

***16S rRNA Gene Sequencing.***
Sequencing was completed as described by Kozich, et al. [@kozich_development_2013]. DNA extraction used the 96-well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf). The V4 variable region was amplified and the resulting product was split between four sequencing runs with normal, adenoma, and carcinoma evenly represented on each run. Each group was randomly assigned to avoid biases based on sample collection location. The pre and post-treatment samples were sequenced on the same run.

***Sequence Processing.***
The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences and has been previously described [@kozich_development_2013]. The general workflow using mothur included merging paired-end reads into contigs, filtering for low quality contigs, aligning to the SILVA database [@pruesse_silva_2007], screening for chimeras using UCHIME [@edgar_uchime_2011], classifying with a naive Bayesian classifier using the Ribosomal Database Project (RDP)[@wang_naive_2007], and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm [@schloss_assessing_2011]. The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` to minimize the impacts of uneven sampling.

***Model Building.*** The Random Forest [@breiman_random_2001] algorithm was used to create the three models used to classify pre and post-treatment samples by diagnosis (adenoma, advanced adenoma, or carcinoma) as well as to assess the probability that a sample was more similar to the patient's original diagnosis or that of a disease-free patient. All models included only OTU data obtained from 16S rRNA sequencing and were processed using the caret (v`r paste(packageVersion("caret"))`) R package. For each model we optimized the mtry hyper-parameter, which defines the number of OTUs to investigate at each split before a new division of the data was created with the Random Forest model [@breiman_random_2001]. To insure that our optimization did not result in over-fitting of the data, we made 100 different 80/20 (train/test) splits of the data where the same proportion was present within both the whole data set and the 80/20 split. For each of the 100 splits, 20 repeated 10-fold cross validation was performed on the 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic). The resulting model was then tested on the 20% of the data that were held out. A summary of the mtry hyperparameter values that were tried is available in Table S5. The reported P-values for each model relative to a random labeling was assessed by comparing the distribution of the 100 80/20 splits for the correctly labeled data to the distribution of randomly labeled data.

The three diagnosis models were constructed by using the data from Baxter et al. [@baxter_microbiota-based_2016], which was censored for the pre-treatment samples of the patients that we had post-treatment samples. The treatment models were then used to quantify the model probability that a patient with an initial diagnosis retained that diagnosis or a disease-free diagnosis.

***Statistical Analysis.***
The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis. Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`). Comparisons between probabilities as well as overall differences in the median relative abundance of each OTU between pre and post-treatment samples utilized a paired Wilcoxon ranked sum test. Comparisons between different treatment for lesions utilized a Kruskal Wallis test. Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant. The P-values reported are those that were BH corrected. Model rank importance was determined by obtaining the median MDA from the 100, 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.

***Reproducible Methods.***
A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_FollowUps_Microbiome_2017. Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.

\newpage

## Declarations

### Ethics approval and consent to participate

The University of Michigan Institutional Review Board approved this study, and all subjects provided informed consent. This study conformed to the guidelines of the Helsinki Declaration.

### Consent for publication

Not applicable.

### Availability of data and material

A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017. Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.

### Competing Interests

All authors declare that they do not have any relevant competing interests to report.

### Funding

This study was supported by funding from the National Institutes of Health to P. Schloss (R01GM099514, P30DK034933) and to the Early Detection Research Network (U01CA86400).

### Authors' contributions

All authors were involved in the conception and design of the study. MAS analyzed the data. NTB processed samples and analyzed the data. All authors interpreted the data. MAS and PDS wrote the manuscript. All authors reviewed and revised the manuscript. All authors read and approved the final manuscript.

### Acknowledgements

The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study. We would also like to thank Amanda Elmore for reviewing and correcting code error and providing feedback on manuscript drafts. We would also like to thank Nicholas Lesniak for providing feedback on manuscript drafts.

\newpage

## References

<div id="refs"></div>

\newpage

**Table 1: Demographic data of patients in the pre and post-treatment cohort**

```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time, Surgery, chemo_received, radiation_received, Abx)

t2 <- as.data.frame(matrix(nrow = 10, ncol = 3,
                           dimnames =
                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time",  
                                           "Surgery_only", "Surg_Chemo", "Surg_Chemo_Radiation", "antibiotics"),
                                  coln = c("Adenoma", "SRN", "Carcinoma"))))

data_filter1 <- c("adenoma", "adenoma", "cancer")
data_filter2 <- c("adenoma", "adv_adenoma", "cancer")

# Generate the summary stats for each group
for(i in 1:length(colnames(t2))){

  t2["Dx_Bin", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
  t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "female") %>% select(n)
  t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
  t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
  t2["post_time", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(time, na.rm = TRUE))
  t2["Surgery_only", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & Surgery == "Y" & chemo_received == "no") %>% select(Surgery) %>% count()
  t2["Surg_Chemo", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & chemo_received == "yes" & radiation_received == "no") %>% select(chemo_received) %>% count()
  t2["Surg_Chemo_Radiation", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & radiation_received == "yes") %>% select(radiation_received) %>% count()
  t2["antibiotics", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & Abx > 0) %>% select(Abx) %>% count()

}

# Generate SD for applicable metrics
age_sd <- c()
BMI_sd <- c()
post_time_sd <- c()

for(i in 1:length(colnames(t2))){

  age_sd <- c(age_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
  BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
  post_time_sd <- c(post_time_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(time, na.rm = TRUE)))[, "std"])
}

# Create combined table
# "u\00B1" is to add a +/- symbol where applicable

ttemptable <- as.data.frame(t(t2)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2),
                                                    label = "\u00B1",
                                                    format(age_sd, digits = 2, nmsall = 2)),
                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2),
                                        BMI = paste(format(BMI, digits = 2, nsmall = 2),
                                                    label = "\u00B1",
                                                    format(BMI_sd, digits = 2, nmsall = 2)),
                                        White = format(White, digits =2, nsmall = 2),
                                        post_time = paste(format(post_time, digits = 2, nsmall = 2),
                                                    label = "\u00B1",
                                                    format(post_time_sd, digits = 2, nmsall = 2)),
                                        Surgery_only = format(Surgery_only, digits = 2, nsmall = 0),
                                        Surg_Chemo = format(Surg_Chemo, digits = 2, nsmall = 0),
                                        Surg_Chemo_Radiation = format(Surg_Chemo_Radiation, digits = 2, nsmall = 0),
                                        antibiotics = format(antibiotics, digits = 2, nsmall = 0))

ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f,
             ttemptable$BMI, ttemptable$White, ttemptable$post_time, ttemptable$Surgery_only, ttemptable$Surg_Chemo, ttemptable$Surg_Chemo_Radiation)

age_label <- paste("Age (Mean ", label = "\u00B1", " SD)", sep = "")
bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")
post_time_label <- paste("Days Between Colonoscopy (Mean ", label = "\u00B1", " SD)", sep = "")

colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label,
                   "Surgery Only", "Surgery & Chemotherapy", "Surgery, Chemotherapy, & Radiation")

test <- t(ft1)
colnames(test) <- test["Group", ]
test <- test[-1, ]

kable(test, format = "markdown", align = 'c', padding = 2)
```

\newpage

**Table 2: Demographic data of training cohort**

```{r table2, echo=FALSE, message=FALSE, warning=FALSE}

metadata <- filter(metaI, !sample %in% c(good_metaf$initial, good_metaf$followUp)) %>%
  select(Dx_Bin, dx, Age, Gender, BMI, White) %>% mutate(Dx_Bin = gsub("adv Adenoma", "adv_adenoma", Dx_Bin), Dx_Bin = gsub("High Risk Normal", "Normal", Dx_Bin))

t1 <- as.data.frame(matrix(nrow = 5, ncol = 4,
                           dimnames =
                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White"),
                                  coln = c("Normal", "Adenoma", "SRN", "Carcinoma"))))

data_filter1 <- c("normal", "adenoma", "adenoma", "cancer")
data_filter2 <- c("Normal", "Adenoma", "adv_adenoma", "Cancer")

# Generate the summary stats for each group
for(i in 1:length(colnames(t1))){

  t1["Dx_Bin", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
  t1["Age", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
  t1["Gender_f", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "female") %>% select(n)
  t1["BMI", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(as.numeric(BMI), na.rm = TRUE))
  t1["White", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
}

# Generate SD for applicable metrics
age_sd <- c()
BMI_sd <- c()

for(i in 1:length(colnames(t1))){

  age_sd <- c(age_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
  BMI_sd <- c(BMI_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
}

# Create combined table
# "u\00B1" is to add a +/- symbol where applicable

ttemptable <- as.data.frame(t(t1)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2),
                                                    label = "\u00B1",
                                                    format(age_sd, digits = 2, nmsall = 2)),
                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2),
                                        BMI = paste(format(BMI, digits = 2, nsmall = 2),
                                                    label = "\u00B1",
                                                    format(BMI_sd, digits = 2, nmsall = 2)),
                                        White = format(White, digits =2, nsmall = 2))

ft1 <- cbind(Group = c("Normal", "Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f,
             ttemptable$BMI, ttemptable$White)

age_label <- paste("Age (Mean ", label = "\u00B1", " SD)", sep = "")
bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")

colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)")

test <- t(ft1)
colnames(test) <- test["Group", ]
test <- test[-1, ]

kable(test, format = "markdown", align = 'c', padding = 2)
```

\newpage

**Figure 1: General differences between adenoma, advanced adenoma, and carcinoma groups after treatment.** A) $\theta$~YC~ distances from pre versus post sample within each individual. A significant difference was found between the adenoma and carcinoma group (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`). Solid black points represent the median value for each diagnosis group. B) NMDS of the pre and post-treatment samples for the adenoma group. C) NMDS of the pre and post-treatment samples for the advanced adenoma group. D) NMDS of the pre and post-treatment samples for the carcinoma group.

**Figure 2: The top 10 most important OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs. B) Advanced Adenoma OTUs. C) Carcinoma OTUs. The darker circle highlights the median log10 MDA value obtained from 100 different 80/20 splits while the lighter colored circles represents the value obtained for a specific run.

**Figure 3: Top 10% most important OTUs common to those models used to differentiate between patients with normal colons and those with adenoma, advanced adenoma, and carcinoma.** A) Venn diagram showing the OTU overlap between each model.  B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.

**Figure 4: Treatment response based on models built for adenoma, advanced adenoma, or carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma. B) Positive probability change from initial to follow up sample in those with advanced adenoma. C) Positive probability change from initial to follow up sample in those with carcinoma.

\newpage

**Figure S1: ROC curves of the adenoma, advanced adenoma, and carcinoma models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification. B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification. C) Carcinoma ROC curve: The light red shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark red line represents the model using 100% of the data set and what was used for subsequent classification.

**Figure S2: Summary of top 10% of important OTUs for the adenoma, advanced adenoma, and carcinoma models.** A) MDA of the most important variables in the adenoma model. The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs. B) Summary of Important Variables in the advanced adenoma model. MDA of the most important variables in the SRN model. The dark yellow point represents the mean and the lighter yellow points are the value of each of the 100 different runs. C) MDA of the most important variables in the carcinoma model. The dark red point represents the mean and the lighter red points are the value of each of the 100 different runs.

**Figure S3: Pre and post-treatment relative abundance of CRC associated OTUs within the carcinoma model.**

\newpage

### 2017-10-30 submission/manuscript.Rmd
remove previous manuscript writing

@@
 
-**Background.** Colorectal cancer is a worldwide health problem.
-Despite growing evidence that members of the gut microbiota can drive tumorigenesis, little is known about what happens to it after treatment for an adenoma or carcinoma.
-This study tested the hypothesis that treatment for adenoma or carcinoma alters the abundance of bacterial populations associated with disease to those associated with a normal colon.
-We tested this hypothesis by sequencing the 16S rRNA genes in the feces of `r length(rownames(good_metaf))` individuals before and after treatment for adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), and carcinoma (N = `r change_summary["thetayc", "crc_n"]`).
+**Background.**
 
-**Results.** There were small changes to the bacterial community associated with adenoma or advanced adenoma and large changes associated with carcinoma.
-The communities from patients with carcinomas changed significantly more than those with adenoma following treatment (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"]<0.001,"< 0.001", format(change_summary["thetayc", "Pvalue_adnVcrc"], scientific = FALSE, digits = 2, nsmall = 2))`).
-Although treatment was associated with intrapersonal changes, the change in the abundance of individual OTUs in response to treatment was not consistent within diagnosis groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Because the distribution of OTUs across patients and diagnosis groups was irregular, we used the Random Forest machine learning algorithm to identify groups of OTUs that could be used to classify pre and post-treatment samples for each of the diagnosis groups.
-Although the adenoma and carcinoma models could reliably differentiate between the pre and post-treatment samples (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.001), the advanced-adenoma model could not (P-value = `r format(filter(treat_random_pvalues, study == "srn")[, "pvalue"], digits = 2, nsmall = 2)`).
-Furthermore, there was little overlap between the OTUs that were indicative of each treatment.
-To determine whether individuals who underwent treatment were more likely to have OTUs associated with normal colons we used a larger cohort that contained individuals with normal colons and those with adenomas, advanced adenomas, and carcinomas.
-We again built Random Forest models and measured the change in the positive probability of having one of the three diagnoses to assess whether the post-treatment samples received the same classification as the pre-treatment samples.
-Samples from patients who had carcinomas changed towards a microbial milieu that resembles the normal colon after treatment (P-value `r ifelse(all_probs_paired_wilcox["carcinoma", "BH_correction"] < 0.001, invisible("<"), invisible(">"))` 0.001).
-Finally, we were unable to detect any significant differences in the microbiota of individuals treated with surgery alone and those treated with chemotherapy or chemotherapy and radiation (P-value `r ifelse(min(chemo_rad_summary$bh, na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
+**Results.**
 
-**Conclusions.** By better understanding the response of the microbiota to treatment for adenomas and carcinomas, it is likely that biomarkers will eventually be validated that can be used to quantify the risk of recurrence and the likelihood of survival.
-Although it was difficult to identify significant differences between pre and post-treatment samples from patients with adenoma and advanced adenoma, this was not the case for carcinomas.
-Not only were there large changes in pre versus post-treatment samples for those with carcinoma, but these changes were towards a more normal microbiota.
+**Conclusions.**
 
@@
 
-Colorectal cancer (CRC) is the third most common cause of cancer deaths in the United States [@siegel_cancer_2016; @haggar_colorectal_2009].
-Disease mortality has significantly decreased, predominately due to improvements in screening [@haggar_colorectal_2009].
-Despite these improvements, there are still approximately 50,000 CRC-related deaths per year in the United States [@siegel_cancer_2016].
-Current estimates indicate that 20-30% of those who undergo treatment will experience recurrence and 35% of all patients will die within five years [@hellinger_reoperation_2006; @ryuk_predictive_2014; @national_cancer_institute_seer_nodate].
-Identification of methods to assess patients' risk of recurrence is of great importance to reduce mortality and healthcare costs.
-
-There is growing evidence that the gut microbiota is involved in the progression of CRC.
-Mouse-based studies have identified populations of *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* that alter disease progression [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_microbial_2014; @kostic_fusobacterium_2013; @wu_human_2009].
-Furthermore, studies that shift the structure of the microbiota through the use of antibiotics or inoculation of germ free mice with human feces have shown that varying community compositions can result in varied tumor burden [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014].
-Collectively, these studies support the hypothesis that the microbiota can alter the amount of inflammation in the colon and with it the rate of tumorigenesis [@flynn_metabolic_2016].
-
-Building upon this evidence, several human studies have identified unique signatures of colonic lesions [@wang_structural_2012; @chen_decreased_2013; @chen_human_2012; @shen_molecular_2010; @kostic_genomic_2012; @feng_gut_2015].
-One line of research has identified community-level differences between those bacteria that are found on and adjacent to colonic lesions and have supported a role for *Bacteroides fragilis*, *Escherichia coli*, and *Fusobacterium nucleatum* in tumorigenesis [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012].
-Others have proposed feces-based biomarkers that could be used to diagnose the presence of colonic adenomas and carcinomas [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014].
-These studies have associated *Fusobacterium nucleatum* and other oral pathogens with colonic lesions (adenoma, advanced adenoma, and carcinoma).
-They have also noted that the loss of bacteria generally thought to produce short chain fatty acids, which can suppress inflammation, is associated with colonic lesions.
-This suggests that gut bacteria have a role in tumorigenesis with potential as useful biomarkers for aiding in the early detection of disease [@dejea_microbiota_2014; @mima_fusobacterium_2015; @arthur_intestinal_2012;@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014].
-
-Despite advances in understanding the role between the gut microbiota and colonic tumorigenesis, we still do not understand how treatments including resection, chemotherapy, and/or radiation affect the composition of the gut microbiota.
-If the microbial community drives tumorigenesis then one would hypothesize that treatment to remove a lesion would not only remove the lesion, but also the microbiota that promoted the tumorigenesis and hence the risk of recurrence.
-To test this hypothesis, we addressed two related questions: Does treatment affect the colonic microbiota in a predictable manner?
-If so, does the treatment alter the community to more closely resemble that of individuals with normal colons?
-
-We answered these questions by sequencing the V4 region of 16S rRNA genes amplified from fecal samples of individuals with adenoma, advanced adenoma, and carcinomas pre and post-treatment.
-We used classical community analysis to compare the alpha and beta-diversity of communities pre and post-treatment.
-Next, we generated Random Forest models to identify bacterial populations that were indicative of treatment for each diagnosis group.
-Finally, we measured the predictive probabilities to assess whether treatment yielded bacterial communities similar to those individuals with normal colons.
-We found that treatment alters the composition of the gut microbiota and that, for those with carcinomas, the gut microbiota shifted more towards that of a normal colon after treatment.
-In the individuals with carcinomas, no difference was found by the type of treatment (surgery alone, surgery with chemotherapy, surgery with chemotherapy and radiation).
-Understanding how the community responds to these treatments could be a valuable tool for identifying biomarkers to quantify the risk of recurrence and the likelihood of survival.
-
 \newpage
@@
 
-***Treatment for colonic lesions alters the bacterial community structure***
-Within our `r length(rownames(good_metaf))`-person cohort we tested whether the microbiota of patients with adenoma (N = `r change_summary["thetayc", "adn_n"]`), advanced adenoma (N = `r change_summary["thetayc", "srn_n"]`), or carcinoma (N = `r change_summary["thetayc", "crc_n"]`) had any broad differences between pre and post-treatment samples [Table 1].
-None of the individuals in this study had any recorded antibiotic usage that was not associated with surgical treatment of their respective lesion.
-The structure of the microbial communities of the pre and post-treatment samples differed, as measured by the $\theta$~YC~ beta diversity metric [Figure 1A].
-We found that the communities obtained pre and post-treatment among the patients with carcinomas changed significantly more than those patients with adenoma (P-value `r ifelse(change_summary["thetayc", "Pvalue_adnVcrc"] < 0.001, invisible("<"), invisible(">"))` 0.001).
-There were no significant differences in the amount of change observed between the patients with adenoma and advanced adenoma or between the patients with advanced adenoma and carcinoma (P-value > 0.05).
-Next, we tested whether there was a consistent direction in the change in the community structure between the pre and post-treatment samples for each of the diagnosis groups [Figure 1B-D].
-We only observed a consistent shift in community structure for the patients with carcinoma when using a PERMANOVA test (adenoma P-value = `r beta_summary["Adenoma", "PERMANOVA"]`, advanced adenoma P-value = `r beta_summary["SRN", "PERMANOVA"]`, and carcinoma P-value = `r beta_summary["Carcinoma", "PERMANOVA"]`).
-Finally, we measured the number of observed OTUs, Shannon evenness, and Shannon diversity in the pre and post-treatment samples and did not observe a significant change for any of the diagnosis groups (P-value > 0.05) [Table S1].
+***Section 1***
 
-***The treatment of lesions are not consistent across diagnosis groups.***
-We used two approaches to identify those bacterial populations that change between the two samples for each diagnosis group.
-First, we sought to identify individual OTUs that could account for the change in overall community structure.
-However, using a paired Wilcoxon test we were unable to identify any OTUs that were significantly different in the pre and post-treatment groups (P-value `r ifelse(min(all_OTU_pair_wilcox$BH_corrected, na.rm = TRUE) > 0.05, invisible(">"), invisible("<"))` 0.05).
-It is likely that high inter-individual variation and the irregular distribution of OTUs across individuals limited the statistical power of the test.
-We attempted to overcome these problems by using Random Forest models to identify collections of OTUs that would allow us to differentiate between pre and post-treatment samples from each of the diagnosis groups.
-The adenoma and carcinoma models performed well (adenoma AUC range = `r format(1 - max(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(1 - min(adn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` and carcinoma AUC range = `r format(min(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`); however, the model for patients treated for advanced adenomas was not able to reliably differentiate between the pre and post-treatment samples (advanced adenoma AUC range = `r format(min(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_treat_AUC_summary$ROC), digits = 2, nsmall=2L)`).
-Interestingly, the top 10 most important OTUs by MDA that were used for each model had little overlap with each other [Figure 2].
-Although treatment had an impact on the overall community structure, the effect of treatment was not consistent across patients and diagnosis groups.
-Both the adenoma and carcioma treatment models had AUCs that were significantly higher than a random model permutation (P-value `r ifelse(filter(treat_random_pvalues, study == "adn")[, "pvalue"] > 0.0001 & filter(treat_random_pvalues, study == "crc")[, "pvalue"] > 0.0001, invisible(">"), invisible("<"))` 0.0001).
+***Section 2***
 
-***Post-treatment samples from patients with carcinoma more closely resemble those of a normal colon.***
-Next, we determined whether treatment changed the microbiota in a way that the post-treatment communities resembled that of patients with normal colons.
-To test this, we used an expanded cohort of `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals that were diagnosed under the same protocol as having normal colons or colons with adenoma, advanced adenoma, or carcinoma [Table 2].
-We then constructed Random Forest models to classify the study samples, with the 3 diagnosis groups (adenoma, advanced adenoma, or carcinoma), or having a normal colon.
-The models performed moderately with CRC being the best (adenoma AUC range = `r format(min(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(adn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, advanced adenoma AUC range = `r format(min(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(srn_AUC_run_summary$ROC), digits = 2, nsmall=2L)`, carcinoma AUC range = `r format(min(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)` - `r format(max(crc_AUC_run_summary$ROC), digits = 2, nsmall=2L)`; Figure S1).
-The OTUs that were in the top 10% of importance for the adenoma and advanced adenoma models largely overlapped and those OTUs that were used to classify the carcinoma samples were largely distinct from those of the other two models [Figure 3A].
-Among the OTUs that were shared across the three models were those populations generally considered beneficial to their host (e.g.
-*Faecalibacterium*, *Lachnospiraceae*, *Bacteroides*, *Dorea*, *Anaerostipes*, and *Roseburia*) [Figures 3B].
-Although many of important OTUs in the top 10% were also included in the model differentiating between patients with normal colons and those with carcinoma, this model also included OTUs affiliated with populations that have previously been associated with carcinoma (*Fusobacterium*, *Porphyromonas*, *Parvimonas*) [@baxter_microbiota-based_2016; @zeller_potential_2014; @zackular_human_2014] [Figure S2] with some individuals showing a marked decrease in relative abundance [Figure S3].
-Finally, we applied these three models to the pre and post-treatment samples for each diagnosis group and quantified the change in the positive probability of the model.
-A decrease in the positive probability would indicate that the microbiota more closely resembled that of a patient with a normal colon.
-There was no significant change in the positive probability for the adenoma or advanced adenoma groups (P-value `r ifelse(all_probs_paired_wilcox["adenoma", "BH_correction"] > 0.05 & all_probs_paired_wilcox["SRN", "BH_correction"] > 0.05, invisible(">"), invisible("<"))` 0.05) [Figure 4].
-The positive probability for the pre and post-treatment samples from patients diagnosed with carcinoma significantly decreased with treatment, suggesting a shift toward a normal microbiota for most individuals (P-value = `r format(round(all_probs_paired_wilcox["carcinoma", "BH_correction"], digits = 3))`).
-Only, `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the 26 patients (`r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`%) who were diagnosed with a carcinoma had a higher positive probability after treatment; one of those was re-diagnosed with carcinoma on the follow up visit.
-These results indicate that, although there were changes in the microbiota associated with treatment, those experienced by patients with carcinoma after treatment yielded gut bacterial communities of greater similarity to that of a normal colon.
+***Section 3***
 
-***Difficult to identify effects of specific treatments on the change in the microbiota.***
-The type of treatment that the patients received varied across diagnosis groups.
-Those with adenomas and advanced adenomas received surgical resection (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "Y")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "Y")))`) or polyp removal during colonoscopy (adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adenoma" & Surgery == "N")))`; advanced adenoma, N=`r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma" & Surgery == "N")))`) and those with carcinomas received surgical resection (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "no")))`), surgical resection with chemotherapy (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "no")))`), and surgical resection with chemotherapy and radiation (N=`r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Surgery == "Y" & chemo_received == "yes" & radiation_received == "yes")))`).
-Regardless of treatment used there was no significant difference in the effect of these treatments on the number of observed OTUs, Shannon diversity, or Shannon evenness (P-value `r ifelse(min(chemo_rad_summary[, "bh"], na.rm = TRUE) < 0.05, invisible("<"), invisible(">"))` 0.05).
-Furthermore, there was not a significant difference in the effect of the treatments on the amount of change in the community structure (P-value = `r ifelse(chemo_rad_summary["dist", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
-Finally, the change in the positive probability was not significantly different between any of the treatment groups (P-value = `r ifelse(chemo_rad_summary["red_crc", "bh"] == 1, invisible("0.999"), format(chemo_rad_summary["dist", "bh"], digits = 3, nsmall = 3))`).
-Due to the relatively small number of samples in each treatment group, it was difficult to make a definitive statement regarding the specific type of treatment on the amount of change in the structure of the microbiota.
+***Section 4***
 
@@
 
-Our study focused on comparing the microbiota of patients diagnosed with adenoma, advanced adenoma, and carcinoma before and after treatment.
-For all three groups of patients, we observed changes in their microbiota.
-Some of these changes, specifically for adenoma, may be due to normal temporal variation, however, those with advanced adenoma and carcinoma clearly had large microbiota changes.
-After treatment, the microbiota of patients with carcinoma changed significantly more than the other groups.
-This change resulted in communities that more closely resembled those of patients with a normal colon.
-This may suggest that treatment for carcinoma is not only successful for removing the carcinoma but also at reducing the associated bacterial communities.
-Understanding the effect of treatment on the microbiota of those diagnosed with carcinomas may have important implications for reducing disease recurrence.
-It is intriguing that it may be possible to use microbiome-based biomarkers to not only predict the presence of lesions but to also assess the risk of recurrence due to these changes in the microbiota.
-
-Patients diagnosed with adenoma and advanced adenoma, however, did not experience a shift towards a community structure that resembled those with normal colons.
-This may be due to the fundamental differences between the features of adenomas and advanced adenomas and carcinoma.
-Specifically, carcinomas may create an inflammatory milieu that would impact the structure of the community and removal of that stimulus would alter said structure.
-It is possible that the difference between the microbiota of patients with adenoma and advanced adenoma and those with normal colons is subtle.
-This is supported by the reduced ability of our models to correctly classify patients with adenomas and advanced adenomas relative to those diagnosed with carcinomas [Figure S1].
-Given the irregular distribution of microbiota across patients in the different diagnosis groups, it is possible that we lacked the statistical power to adequately characterize the change in the communities following treatment.
-
-There was a subset of patients (`r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` of the `r length(rownames(filter(good_metaf, dx == "cancer")))` with carcinomas) who demonstrated an elevated probability of carcinoma after treatment.
-This may reflect an elevated risk of recurrence.
-The `r round(table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]/26*100,digits = 2)`% prevalence of increased carcinoma probability from our study is within the expected rate of recurrence (20-30% [@hellinger_reoperation_2006; @ryuk_predictive_2014]).
-We hypothesized that these individuals may have had more severe tumors; however, the tumor severity of these `r table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` individuals (`r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[1]` with Stage I, `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[2]` with Stage II, and `r table(good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes) %>% select(stage))[3]` with Stage III) was similar to the distribution observed among the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients.
-We also hypothesized that we may have sampled these patients later than the rest and their communities may have reverted to a carcinoma-associated state; however, there was not a statistically significant difference in the length of time between sample collection among those whose probabilities increased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"]), digits = 0)` (`r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"])["25%"], digits = 0)` - `r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"])["75%"], digits = 0)`) days) or decreased (`r format(median((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"]), digits = 0)` (`r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])["25%"], digits = 0)` - `r format(quantile((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])["75%"], digits = 0)`) days) (Wilcoxon Test; P-value = `r format(wilcox.test((good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes > initial_yes))[, "time"], (good_metaf %>% filter(dx == "cancer") %>% mutate(initial_yes = filter(crc_probs, sampleType == "initial")[, "Yes"], follow_yes = filter(crc_probs, sampleType == "followup")[, "Yes"]) %>% filter(follow_yes < initial_yes))[, "time"])$p.value, digits = 2)`) (all days data displayed as median (IQR)).
-Finally, it is possible that these patients may not have responded to treatment as well as the other `r length(rownames(filter(good_metaf, dx == "cancer"))) - table(filter(crc_probs, sampleType == "initial")[, "Yes"] < filter(crc_probs, sampleType == "followup")[, "Yes"])["TRUE"]` patients diagnosed with carcinoma and so the microbiota may not have been impacted the same way.
-Again, further studies looking at the role of the microbiota in recurrence are needed to understand the dynamics following treatment.
-
-Our final hypothesis was that the specific type of treatment altered the structure of the microbiome.
-The treatment to remove adenomas and advanced adenomas was either polyp removal or surgical resection whereas it was surgical resection alone or in combination with chemotherapy or with chemotherapy and radiation for individuals with carcinoma.
-Because chemotherapy and radiation target rapidly growing cells, these treatments would be more likely to cause a turnover of the colonic epithelium driving a more significant change in the structure of the microbiota.
-Although, we were able to test for an effect across these specific types of treatment, the number of patients in each treatment group was relatively small.
-Finally, those undergoing surgery would have received antibiotics and this may be a potential confounder.
-However, our pre-treatment stool samples were obtained before the surgery and the post-treatment samples were obtained long after any effects due to antibiotic administration on the microbiome would be expected to occur (`r format(median(filter(good_metaf, dx == "cancer")[, "time"]), digits = 0)` (`r format(quantile(filter(good_metaf, dx == "cancer")[, "time"])["25%"], digits = 0)` - `r format(quantile(filter(good_metaf, dx == "cancer")[, "time"])["75%"], digits = 0)`) days).
-We also found no difference in the community structure of those that received surgery and those that did not as a treatment for adenoma or advanced adenoma.
-
 ## Conclusion
-
-This study expands upon existing research that has established a role for the microbiota in tumorigenesis and that demonstrated the utility of microbiome-based biomarkers to predict the presence of colonic lesions.
-We were surprised by the lack of a consistent signal that was associated with treatment of patients with adenomas or advanced adenomas.
-The lack of a large effect size may be due to differences in the role of bacteria in the formation of adenomas and carcinomas or it could be due to differences in the behaviors and medications within these classes of patients.
-One of the most exciting of these future directions is the possibility that markers within the microbiota could be used to potentially evaluate the effect of treatment and predict recurrence for those diagnosed with carcinoma.
-If such an approach is effective, it might be possible to target the microbiota as part of adjuvant therapy, if the biomarkers identified play a key role in the disease process.
-Our data provides additional evidence on the importance of the microbiota in tumorigenesis by addressing the recovery of the microbiota after treatment and opens interesting avenues of research into how these changes may affect recurrence.
 
@@
 
-***Study Design and Patient Sampling.***
-Sampling and design have been previously reported in Baxter, et al [@baxter_microbiota-based_2016].
-Briefly, samples were stored on ice for at least 24h before freezing.
-Although we cannot exclude that this sampling protocol may have impacted the gut microbiota composition all samples were subjected to the same methodology.
-Study exclusion involved those who had already undergone surgery, radiation, or chemotherapy, had colorectal cancer before a baseline fecal sample could be obtained, had IBD, a known hereditary non-polyposis colorectal cancer, or familial adenomatous polyposis.
-Samples used to build the models for prediction were collected either prior to a colonoscopy or between one and two weeks after initial colonoscopy.
-The bacterial community has been shown to normalize back to a pre-colonoscopy community within this time period [@obrien_impact_2013].
-Our study cohort consisted of `r length(rownames(good_metaf))` individuals with an initial sample as described and a follow up sample obtained between `r min(good_metaf$time)` - `r max(good_metaf$time)` days after treatment of lesion [Table 1].
-Patients were diagnosed by colonoscopic examination and histopathological review of any biopsies taken.
-Patients were classified as having advanced adenoma if they had an adenoma greater than 1 cm, more than three adenomas of any size, or an adenoma with villous histology.
-This study was approved by the University of Michigan Institutional Review Board.
-All study participants provided informed consent and the study itself conformed to the guidelines set out by the Helsinki Declaration.
-The original protocol for the study did not provide for tracking patients after the follow up samples and so it was not possible for us to ascertain their diagnosis after the completion of the study.
-
-***Treatment.***
-For this study treatment refers specifically to the removal of a lesion with or without chemotherapy and radiation.
-The majority of patients undergoing treatment for adenoma or advanced adenoma were not treated surgically [Table 1] but rather via colonoscopy.
-All patients diagnosed with carcinomas were treated with at least surgery or a combination of surgery and chemotherapy or surgery, chemotherapy, and radiation.
-The type of chemotherapy used for patients with CRC included Oxaliplatin, Levicovorin, Folfox, Xeloda, Capecitabine, Avastin, Fluorouracil, and Glucovorin.
-These were used individually or in combination with others depending on the patient [Table 1].
-If an individual was treated with radiation they were also always treated with chemotherapy.
-Radiation therapy generally used 18 mV photons for treatment.
+***Study Design***
 
 ***16S rRNA Gene Sequencing.***
-Sequencing was completed as described by Kozich, et al.
-[@kozich_development_2013].
-DNA extraction used the 96-well Soil DNA isolation kit (MO BIO Laboratories) and an epMotion 5075 automated pipetting system (Eppendorf).
-The V4 variable region was amplified and the resulting product was split between four sequencing runs with normal, adenoma, and carcinoma evenly represented on each run.
-Each group was randomly assigned to avoid biases based on sample collection location.
-The pre and post-treatment samples were sequenced on the same run.
 
 ***Sequence Processing.***
-The mothur software package (v1.37.5) was used to process the 16S rRNA gene sequences and has been previously described [@kozich_development_2013].
-The general workflow using mothur included merging paired-end reads into contigs, filtering for low quality contigs, aligning to the SILVA database [@pruesse_silva_2007], screening for chimeras using UCHIME [@edgar_uchime_2011], classifying with a naive Bayesian classifier using the Ribosomal Database Project (RDP)[@wang_naive_2007], and clustered into Operational Taxonomic Units (OTUs) using a 97% similarity cutoff with an average neighbor clustering algorithm [@schloss_assessing_2011].
-The number of sequences for each sample was rarefied to `r ifelse(sd(total_sub_Seqs) == 0, paste(mean(total_sub_Seqs)), paste("Error Check Subsampling"))` to minimize the impacts of uneven sampling.
-
-***Model Building.*** The Random Forest [@breiman_random_2001] algorithm was used to create the three models used to classify pre and post-treatment samples by diagnosis (adenoma, advanced adenoma, or carcinoma) as well as to assess the probability that a sample was more similar to the patient's original diagnosis or that of a disease-free patient.
-All models included only OTU data obtained from 16S rRNA sequencing and were processed using the caret (v`r paste(packageVersion("caret"))`) R package.
-For each model we optimized the mtry hyper-parameter, which defines the number of OTUs to investigate at each split before a new division of the data was created with the Random Forest model [@breiman_random_2001].
-To insure that our optimization did not result in over-fitting of the data, we made 100 different 80/20 (train/test) splits of the data where the same proportion was present within both the whole data set and the 80/20 split.
-For each of the 100 splits, 20 repeated 10-fold cross validation was performed on the 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic).
-The resulting model was then tested on the 20% of the data that were held out.
-A summary of the mtry hyperparameter values that were tried is available in Table S5.
-The reported P-values for each model relative to a random labeling was assessed by comparing the distribution of the 100 80/20 splits for the correctly labeled data to the distribution of randomly labeled data.
-
-The three diagnosis models were constructed by using the data from Baxter et al.
-[@baxter_microbiota-based_2016], which was censored for the pre-treatment samples of the patients that we had post-treatment samples.
-The treatment models were then used to quantify the model probability that a patient with an initial diagnosis retained that diagnosis or a disease-free diagnosis.
 
 ***Statistical Analysis.***
-The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical analysis.
-Comparisons between bacterial community structure utilized PERMANOVA [@anderson_permanova_2013] in the vegan package (v`r paste(packageVersion("vegan"))`).
-Comparisons between probabilities as well as overall differences in the median relative abundance of each OTU between pre and post-treatment samples utilized a paired Wilcoxon ranked sum test.
-Comparisons between different treatment for lesions utilized a Kruskal Wallis test.
-Where multiple comparison testing was appropriate, a Benjamini-Hochberg (BH) correction was applied [@benjamini_controlling_1995] and a corrected P-value of less than 0.05 was considered significant.
-The P-values reported are those that were BH corrected.
-Model rank importance was determined by obtaining the median MDA from the 100, 20 repeated 10-fold cross validation and then ranking from largest to smallest MDA.
+The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical.
 
 ***Reproducible Methods.***
-A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_FollowUps_Microbiome_2017.
-Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.
 
@@
 
-The University of Michigan Institutional Review Board approved this study, and all subjects provided informed consent. This study conformed to the guidelines of the Helsinki Declaration.
+Need to fill in.
 
@@
 
-A detailed and reproducible description of how the data were processed and analyzed can be found at https://github.com/SchlossLab/Sze_followUps_2017.
-Raw sequences have been deposited into the NCBI Sequence Read Archive (SRP062005 and SRP096978) and the necessary metadata can be found at https://www.ncbi.nlm.nih.gov/Traces/study/ and searching the respective SRA study accession.
+Need to fill in.
 
@@
 
-This study was supported by funding from the National Institutes of Health to P. Schloss (R01GM099514, P30DK034933) and to the Early Detection Research Network (U01CA86400).
+Need to fill in.
 
@@
 
-All authors were involved in the conception and design of the study.
-MAS analyzed the data.
-NTB processed samples and analyzed the data.
-All authors interpreted the data.
-MAS and PDS wrote the manuscript.
-All authors reviewed and revised the manuscript.
-All authors read and approved the final manuscript.
+Need to fill in.
 
@@
 
-The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
-We would also like to thank Amanda Elmore for reviewing and correcting code error and providing feedback on manuscript drafts.
-We would also like to thank Nicholas Lesniak for providing feedback on manuscript drafts.
+Need to fill in.
 
@@
 
-**Table 1: Demographic data of patients in the pre and post-treatment cohort**
+**Table 1: **
 
 ```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
-test_cohort <- good_metaf %>% select(Dx_Bin, dx, Age, Gender, BMI, White, time, Surgery, chemo_received, radiation_received, Abx)
 
-t2 <- as.data.frame(matrix(nrow = 10, ncol = 3,
-                           dimnames =
-                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White", "post_time",  
-                                           "Surgery_only", "Surg_Chemo", "Surg_Chemo_Radiation", "antibiotics"),
-                                  coln = c("Adenoma", "SRN", "Carcinoma"))))
-
-data_filter1 <- c("adenoma", "adenoma", "cancer")
-data_filter2 <- c("adenoma", "adv_adenoma", "cancer")
-
-# Generate the summary stats for each group
-for(i in 1:length(colnames(t2))){
-
-  t2["Dx_Bin", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
-  t2["Age", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
-  t2["Gender_f", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "female") %>% select(n)
-  t2["BMI", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(BMI, na.rm = TRUE))
-  t2["White", i] <- length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
-  t2["post_time", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(time, na.rm = TRUE))
-  t2["Surgery_only", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & Surgery == "Y" & chemo_received == "no") %>% select(Surgery) %>% count()
-  t2["Surg_Chemo", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & chemo_received == "yes" & radiation_received == "no") %>% select(chemo_received) %>% count()
-  t2["Surg_Chemo_Radiation", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & radiation_received == "yes") %>% select(radiation_received) %>% count()
-  t2["antibiotics", i] <- filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & Abx > 0) %>% select(Abx) %>% count()
-
-}
-
-# Generate SD for applicable metrics
-age_sd <- c()
-BMI_sd <- c()
-post_time_sd <- c()
-
-for(i in 1:length(colnames(t2))){
-
-  age_sd <- c(age_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
-  BMI_sd <- c(BMI_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
-  post_time_sd <- c(post_time_sd, (filter(test_cohort, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(time, na.rm = TRUE)))[, "std"])
-}
-
-# Create combined table
-# "u\00B1" is to add a +/- symbol where applicable
-
-ttemptable <- as.data.frame(t(t2)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2),
-                                                    label = "\u00B1",
-                                                    format(age_sd, digits = 2, nmsall = 2)),
-                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2),
-                                        BMI = paste(format(BMI, digits = 2, nsmall = 2),
-                                                    label = "\u00B1",
-                                                    format(BMI_sd, digits = 2, nmsall = 2)),
-                                        White = format(White, digits =2, nsmall = 2),
-                                        post_time = paste(format(post_time, digits = 2, nsmall = 2),
-                                                    label = "\u00B1",
-                                                    format(post_time_sd, digits = 2, nmsall = 2)),
-                                        Surgery_only = format(Surgery_only, digits = 2, nsmall = 0),
-                                        Surg_Chemo = format(Surg_Chemo, digits = 2, nsmall = 0),
-                                        Surg_Chemo_Radiation = format(Surg_Chemo_Radiation, digits = 2, nsmall = 0),
-                                        antibiotics = format(antibiotics, digits = 2, nsmall = 0))
-
-ft1 <- cbind(Group = c("Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f,
-             ttemptable$BMI, ttemptable$White, ttemptable$post_time, ttemptable$Surgery_only, ttemptable$Surg_Chemo, ttemptable$Surg_Chemo_Radiation)
-
-age_label <- paste("Age (Mean ", label = "\u00B1", " SD)", sep = "")
-bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")
-post_time_label <- paste("Days Between Colonoscopy (Mean ", label = "\u00B1", " SD)", sep = "")
-
-colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)", post_time_label,
-                   "Surgery Only", "Surgery & Chemotherapy", "Surgery, Chemotherapy, & Radiation")
-
-test <- t(ft1)
-colnames(test) <- test["Group", ]
-test <- test[-1, ]
-
-kable(test, format = "markdown", align = 'c', padding = 2)
 ```
@@
 
-**Table 2: Demographic data of training cohort**
+**Table 2: **
 
@@
 
-metadata <- filter(metaI, !sample %in% c(good_metaf$initial, good_metaf$followUp)) %>%
-  select(Dx_Bin, dx, Age, Gender, BMI, White) %>% mutate(Dx_Bin = gsub("adv Adenoma", "adv_adenoma", Dx_Bin), Dx_Bin = gsub("High Risk Normal", "Normal", Dx_Bin))
-
-t1 <- as.data.frame(matrix(nrow = 5, ncol = 4,
-                           dimnames =
-                             list(rown = c("Dx_Bin", "Age", "Gender_f", "BMI", "White"),
-                                  coln = c("Normal", "Adenoma", "SRN", "Carcinoma"))))
-
-data_filter1 <- c("normal", "adenoma", "adenoma", "cancer")
-data_filter2 <- c("Normal", "Adenoma", "adv_adenoma", "Cancer")
-
-# Generate the summary stats for each group
-for(i in 1:length(colnames(t1))){
-
-  t1["Dx_Bin", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i])))
-  t1["Age", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(Age))
-  t1["Gender_f", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% count(Gender) %>% mutate(n = n/sum(n) * 100) %>%  filter(Gender == "female") %>% select(n)
-  t1["BMI", i] <- filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(avg = mean(as.numeric(BMI), na.rm = TRUE))
-  t1["White", i] <- length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i] & White == 1)))/length(rownames(filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]))) * 100
-}
-
-# Generate SD for applicable metrics
-age_sd <- c()
-BMI_sd <- c()
-
-for(i in 1:length(colnames(t1))){
-
-  age_sd <- c(age_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(Age)))[, "std"])
-  BMI_sd <- c(BMI_sd, (filter(metadata, dx == data_filter1[i] & Dx_Bin == data_filter2[i]) %>% summarise(std = sd(BMI, na.rm = TRUE)))[, "std"])
-}
-
-# Create combined table
-# "u\00B1" is to add a +/- symbol where applicable
-
-ttemptable <- as.data.frame(t(t1)) %>% mutate(Age = paste(format(Age, digits = 2, nsmall = 2),
-                                                    label = "\u00B1",
-                                                    format(age_sd, digits = 2, nmsall = 2)),
-                                        Gender_f = format(Gender_f, digits = 2, nsmall = 2),
-                                        BMI = paste(format(BMI, digits = 2, nsmall = 2),
-                                                    label = "\u00B1",
-                                                    format(BMI_sd, digits = 2, nmsall = 2)),
-                                        White = format(White, digits =2, nsmall = 2))
-
-ft1 <- cbind(Group = c("Normal", "Adenoma", "Advanced Adenoma", "Carcinoma"), ttemptable$Dx_Bin, ttemptable$Age, ttemptable$Gender_f,
-             ttemptable$BMI, ttemptable$White)
-
-age_label <- paste("Age (Mean ", label = "\u00B1", " SD)", sep = "")
-bmi_label <- paste("BMI (Mean ", label = "\u00B1", " SD)", sep = "")
-
-colnames(ft1) <- c("Group", "n", age_label, "Sex (%F)", bmi_label, "Caucasian (%)")
-
-test <- t(ft1)
-colnames(test) <- test["Group", ]
-test <- test[-1, ]
-
-kable(test, format = "markdown", align = 'c', padding = 2)
 ```
@@
 
-**Figure 1: General differences between adenoma, advanced adenoma, and carcinoma groups after treatment.** A) $\theta$~YC~ distances from pre versus post sample within each individual.
-A significant difference was found between the adenoma and carcinoma group (P-value = `r format(change_summary['thetayc', 'Pvalue_adnVcrc'], scientific = TRUE, digits = 3)`).
-Solid black points represent the median value for each diagnosis group.
-B) NMDS of the pre and post-treatment samples for the adenoma group.
-C) NMDS of the pre and post-treatment samples for the advanced adenoma group.
-D) NMDS of the pre and post-treatment samples for the carcinoma group.
+**Figure 1: **
 
-**Figure 2: The top 10 most important OTUs used to classify treatment for adenoma, advanced adenoma, and carcinoma.** A) Adenoma OTUs.
-B) Advanced Adenoma OTUs.
-C) Carcinoma OTUs.
-The darker circle highlights the median log10 MDA value obtained from 100 different 80/20 splits while the lighter colored circles represents the value obtained for a specific run.
+**Figure 2: **
 
-**Figure 3: Top 10% most important OTUs common to those models used to differentiate between patients with normal colons and those with adenoma, advanced adenoma, and carcinoma.** A) Venn diagram showing the OTU overlap between each model.
-B) For each common OTU the lowest taxonomic identification and importance rank for each model run is shown.
+**Figure 3: **
 
-**Figure 4: Treatment response based on models built for adenoma, advanced adenoma, or carcinoma.** A) Positive probability change from initial to follow up sample in those with adenoma.
-B) Positive probability change from initial to follow up sample in those with advanced adenoma.
-C) Positive probability change from initial to follow up sample in those with carcinoma.
+**Figure 4: **
 
@@
 
-**Figure S1: ROC curves of the adenoma, advanced adenoma, and carcinoma models.** A) Adenoma ROC curve: The light green shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark green line represents the model using 100% of the data set and what was used for subsequent classification.
-B) Advanced Adenoma ROC curve: The light yellow shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark yellow line represents the model using 100% of the data set and what was used for subsequent classification.
-C) Carcinoma ROC curve: The light red shaded areas represent the range of values of a 100 different 80/20 splits of the test set data and the dark red line represents the model using 100% of the data set and what was used for subsequent classification.
+**Figure S1: **
 
-**Figure S2: Summary of top 10% of important OTUs for the adenoma, advanced adenoma, and carcinoma models.** A) MDA of the most important variables in the adenoma model.
-The dark green point represents the mean and the lighter green points are the value of each of the 100 different runs.
-B) Summary of Important Variables in the advanced adenoma model.
-MDA of the most important variables in the SRN model.
-The dark yellow point represents the mean and the lighter yellow points are the value of each of the 100 different runs.
-C) MDA of the most important variables in the carcinoma model.
-The dark red point represents the mean and the lighter red points are the value of each of the 100 different runs.
+**Figure S2: **
 
-**Figure S3: Pre and post-treatment relative abundance of CRC associated OTUs within the carcinoma model.**
+**Figure S3: **
 

### 2017-10-30 submission/manuscript.Rmd
add title
3/4 completed methods

@@
 
-opts_chunk$set("tidy" = TRUE)
-opts_chunk$set("echo" = FALSE)
-opts_chunk$set("eval" = TRUE)
-opts_chunk$set("warning" = FALSE)
-opts_chunk$set("cache" = FALSE)
-opts_chunk$set("message" = FALSE)
-
 ```
@@
 
-# Normalization of the microbiota in patients after treatment for colonic lesions
+# The Importance of the Microbiota and Individual Community Members in Colorectal Cancer
 
@@
 
-***Study Design***
+***Obtaining Data Sets:***
+Studies used for this meta-analysis were identified through the review articles written by Keku, et al.
+and Vogtmann, et al.
+[@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016].
+All studies were included that used tissue or feces as their sample source for 16S rRNA gene sequencing analysis.
+Studies using either 454 or Illumina sequencing technology were included.
+Only data sets that had the raw sequences available for analysis were included.
+Some studies did not have publically available raw sequences or did not have meta data in which the authors were able to share.
+After this filtering step the following studies remained: Ahn [@ahn_human_2013], Baxter [@baxter_microbiota-based_2016], Brim [@brim_microbiome_2013], Burns [@burns_virulence_2015], Chen [@chen_human_2012], Dejea [@dejea_microbiota_2014], Flemer [@flemer_tumour-associated_2017], Geng [@geng_diversified_2013], Hale [@hale_shifts_2017], Kostic [@kostic_genomic_2012], Lu [@lu_mucosal_2016], Sanapareddy [@sanapareddy_increased_2012], Wang [@wang_structural_2012], Weir [@weir_stool_2013], and Zeller [@zeller_potential_2014].
+The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study.
+The Kostic study was not used since after sequence processing all the case samples did not have more than 100 sequences remaining.
+This left a total of 13 studies in which complete analysis could be completed.
 
-***16S rRNA Gene Sequencing.***
+***Data Set Breakdown:***
+In total there were 7 studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), 5 studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and 2 studies with both fecal and tissue samples (Chen and Flemer).
+The total number of individuals initially run through the sequence processing for the fecal samples was 1899 and for the tissue samples was 462.
 
-***Sequence Processing.***
+***Sequence Processing:***
+For the majority of studies raw sequences were downloaded from the SRA (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the following website: http://www.ncbi.nlm.nih.gov/Traces/study/ by searching the respective accession number of the study.
+Of the studies that did not have sequences and meta data on the SRA one study had the data stored on DBGap [@ahn_human_2013] and four studies the data was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013].
+Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009].
+Where possible quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing.
+If it was not possible to use these defaults the author stated quality cut-offs were used instead.
+Chimeras were identifed and removed using the VSEARCH [@rognes_vsearch_2016] program and *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.
 
-***Statistical Analysis.***
-The R software package (v`r paste(version$major, ".", version$minor, sep = "")`) was used for all statistical.
+***Statistical Analysis:***
+All statistical analysis after sequence processing utilized the R software package (v`r paste(version$major, ".", version$minor, sep = "")`).
+For the alpha diversity analysis values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package.
+Testing for alpha diversity differences utilized both T-tests and ANOVA.
+Linear mixed-effect models were created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for both study and variable region effect in the diversity measures when analyzing colorectal cancer groups.
+Relative Risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages.
+Relative risk significance testing utilized the chi-squred test.
+Beta-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package.
+Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages.
+Random Forest testing of the obtained AUC versus a random model AUC utilized T-tests.
+Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages.
+All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages.
 
-***Reproducible Methods.***
+***Study Analysis Overview:***
+
+***Reproducible Methods:***
+The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_XXXX_2016.
+Unless mentioned otherwise the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file, on the GitHub repository or in the original manuscript.
 

### 2017-10-30 submission/manuscript.tex
3/4 completed methods

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{}
  \pretitle{\vspace{\droptitle}}
  \posttitle{}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

\begin{document}

\section{The Importance of the Microbiota and Individual Community
Members in Colorectal
Cancer}\label{the-importance-of-the-microbiota-and-individual-community-members-in-colorectal-cancer}

\begin{center}
\vspace{25mm}

Marc A Sze${^1}$ and Patrick D Schloss${^1}$${^\dagger}$

\vspace{20mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

$1$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor, MI

\end{center}

Co-author e-mails:

\begin{itemize}
\tightlist
\item
  \href{mailto:marcsze@med.umich.edu}{\nolinkurl{marcsze@med.umich.edu}}
\end{itemize}

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

\textbf{Background.}

\textbf{Results.}

\textbf{Conclusions.}

\subsubsection{Keywords}\label{keywords}

microbiota; colorectal cancer; polyps; treatment; risk factor.

\newpage

\subsection{Background}\label{background}

\newpage

\subsection{Results}\label{results}

\textbf{\emph{Section 1}}

\textbf{\emph{Section 2}}

\textbf{\emph{Section 3}}

\textbf{\emph{Section 4}}

\newpage

\subsection{Discussion}\label{discussion}

\subsection{Conclusion}\label{conclusion}

\newpage

\subsection{Methods}\label{methods}

\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
were identified through the review articles written by Keku, et al. and
Vogtmann, et al. {[}1,2{]}. All studies were included that used tissue
or feces as their sample source for 16S rRNA gene sequencing analysis.
Studies using either 454 or Illumina sequencing technology were
included. Only data sets that had the raw sequences available for
analysis were included. Some studies did not have publically available
raw sequences or did not have meta data in which the authors were able
to share. After this filtering step the following studies remained: Ahn
{[}3{]}, Baxter {[}4{]}, Brim {[}5{]}, Burns {[}6{]}, Chen {[}7{]},
Dejea {[}8{]}, Flemer {[}9{]}, Geng {[}10{]}, Hale {[}11{]}, Kostic
{[}12{]}, Lu {[}13{]}, Sanapareddy {[}14{]}, Wang {[}15{]}, Weir
{[}16{]}, and Zeller {[}17{]}. The Zackular {[}18{]} study was not
included becasue the 90 individuals analyzed within the study are
contained within the larger Baxter study. The Kostic study was not used
since after sequence processing all the case samples did not have more
than 100 sequences remaining. This left a total of 13 studies in which
complete analysis could be completed.

\textbf{\emph{Data Set Breakdown:}} In total there were 7 studies with
only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), 5
studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy),
and 2 studies with both fecal and tissue samples (Chen and Flemer). The
total number of individuals initially run through the sequence
processing for the fecal samples was 1899 and for the tissue samples was
462.

\textbf{\emph{Sequence Processing:}} For the majority of studies raw
sequences were downloaded from the SRA
(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
and metadata was obtained from the following website:
\url{http://www.ncbi.nlm.nih.gov/Traces/study/} by searching the
respective accession number of the study. Of the studies that did not
have sequences and meta data on the SRA one study had the data stored on
DBGap {[}3{]} and four studies the data was obtained directly from the
authors {[}9,11,14,16{]}. Each study was processed using the mothur
(v1.39.3) software program {[}19{]}. Where possible quality filtering
utilized the default methods used in mothur for either 454 or Illumina
based sequencing. If it was not possible to use these defaults the
author stated quality cut-offs were used instead. Chimeras were
identifed and removed using the VSEARCH {[}20{]} program and \emph{de
novo} OTU clustering at 97\% similarity using the OptiClust algorithm
{[}21{]} was utilized.

\textbf{\emph{Statistical Analysis.}} All statistical analysis after
sequence processing utilized the R software package (v3.4.2). For the
alpha diversity analysis values were power transformed using the
rcompanion (v1.10.1) package and then Z-score normalized using the car
(v2.1.5) package. Testing for alpha diversity differences utilized both
T-tests and ANOVA. Linear mixed-effect models were created using the
lme4 (v1.1.14) package to correct for both study and variable region
effect in the diversity measures when analyzing colorectal cancer
groups. Relative Risk was analyzed using both the epiR (v0.9.87) and
metafor (v2.0.0) packages. Relative risk significance testing utilized
the chi-squred test. Beta-diversity differences utilized a Bray-Curtis
distance matrix and PERMANOVA executed with the vegan (v2.4.4) package.
Random Forest models were built using both the caret (v6.0.77) and
randomForest (v4.6.12) packages. Random Forest testing of the obtained
AUC versus a random model AUC utilized T-tests. Power analysis and
estimations were made using the pwr (v1.2.1) and statmod (v1.4.30)
packages. All figures were created using both ggplot2 (v2.2.1) and
gridExtra (v2.3) packages.

\textbf{\emph{Reproducible Methods.}} The code and analysis can be found
here \url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_XXXX_2016}.
Unless mentioned otherwise the accession number for the raw sequences
for the studies used in this analysis can be found directly in the
respective batch file, on the GitHub repository or in the original
manuscript.

\newpage

\subsection{Declarations}\label{declarations}

\subsubsection{Ethics approval and consent to
participate}\label{ethics-approval-and-consent-to-participate}

Need to fill in.

\subsubsection{Consent for publication}\label{consent-for-publication}

Not applicable.

\subsubsection{Availability of data and
material}\label{availability-of-data-and-material}

Need to fill in.

\subsubsection{Competing Interests}\label{competing-interests}

All authors declare that they do not have any relevant competing
interests to report.

\subsubsection{Funding}\label{funding}

Need to fill in.

\subsubsection{Authors' contributions}\label{authors-contributions}

Need to fill in.

\subsubsection{Acknowledgements}\label{acknowledgements}

Need to fill in.

\newpage

\subsection{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-keku_gastrointestinal_2015}{}
1. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
microbiota and colorectal cancer. American Journal of Physiology -
Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
Oct 30{]};308:G351--63. Available from:
\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}

\hypertarget{ref-vogtmann_epidemiologic_2016}{}
2. Vogtmann E, Goedert JJ. Epidemiologic studies of the human microbiome
and cancer. British Journal of Cancer {[}Internet{]}. 2016 {[}cited 2017
Oct 30{]};114:237--42. Available from:
\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}

\hypertarget{ref-ahn_human_2013}{}
3. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
microbiome and risk for colorectal cancer. Journal of the National
Cancer Institute. 2013;105:1907--11.

\hypertarget{ref-baxter_microbiota-based_2016}{}
4. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
improves the sensitivity of fecal immunochemical test for detecting
colonic lesions. Genome Medicine. 2016;8:37.

\hypertarget{ref-brim_microbiome_2013}{}
5. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
al. Microbiome analysis of stool samples from African Americans with
colon polyps. PloS One. 2013;8:e81352.

\hypertarget{ref-burns_virulence_2015}{}
6. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
are a signature of the microbiome in the colorectal tumor
microenvironment. Genome Medicine. 2015;7:55.

\hypertarget{ref-chen_human_2012}{}
7. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
mucosa-associated microbiota in patients with colorectal cancer. PloS
One. 2012;7:e39743.

\hypertarget{ref-dejea_microbiota_2014}{}
8. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, et al. Microbiota organization is a distinct feature of
proximal colorectal cancers. Proceedings of the National Academy of
Sciences of the United States of America. 2014;111:18321--6.

\hypertarget{ref-flemer_tumour-associated_2017}{}
9. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
al. Tumour-associated and non-tumour-associated microbiota in colorectal
cancer. Gut. 2017;66:633--43.

\hypertarget{ref-geng_diversified_2013}{}
10. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.

\hypertarget{ref-hale_shifts_2017}{}
11. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
American Association for Cancer Research, Cosponsored by the American
Society of Preventive Oncology. 2017;26:85--94.

\hypertarget{ref-kostic_genomic_2012}{}
12. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
al. Genomic analysis identifies association of Fusobacterium with
colorectal carcinoma. Genome Research. 2012;22:292--8.

\hypertarget{ref-lu_mucosal_2016}{}
13. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
adherent bacterial dysbiosis in patients with colorectal adenomas.
Scientific Reports. 2016;6:26337.

\hypertarget{ref-sanapareddy_increased_2012}{}
14. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
et al. Increased rectal microbial richness is associated with the
presence of colorectal adenomas in humans. The ISME journal.
2012;6:1858--68.

\hypertarget{ref-wang_structural_2012}{}
15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
segregation of gut microbiota between colorectal cancer patients and
healthy volunteers. The ISME journal. 2012;6:320--9.

\hypertarget{ref-weir_stool_2013}{}
16. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
Stool microbiome and metabolome differences between colorectal cancer
patients and healthy adults. PloS One. 2013;8:e70803.

\hypertarget{ref-zeller_potential_2014}{}
17. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
Potential of fecal microbiota for early-stage detection of colorectal
cancer. Molecular Systems Biology. 2014;10:766.

\hypertarget{ref-zackular_human_2014}{}
18. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
microbiome as a screening tool for colorectal cancer. Cancer Prevention
Research (Philadelphia, Pa.). 2014;7:1112--21.

\hypertarget{ref-schloss_introducing_2009}{}
19. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, et al. Introducing mothur: Open-Source, Platform-Independent,
Community-Supported Software for Describing and Comparing Microbial
Communities. Appl.Environ.Microbiol. {[}Internet{]}. 2009 {[}cited 12AD
Jan 1{]};75:7537--41. Available from:
\url{http://aem.asm.org/cgi/content/abstract/75/23/7537}

\hypertarget{ref-rognes_vsearch_2016}{}
20. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.

\hypertarget{ref-westcott_opticlust_2017}{}
21. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
2017;2.

\newpage

\textbf{Table 1: }

\newpage

\textbf{Table 2: }

\newpage

\textbf{Figure 1: }

\textbf{Figure 2: }

\textbf{Figure 3: }

\textbf{Figure 4: }

\newpage

\textbf{Figure S1: }

\textbf{Figure S2: }

\textbf{Figure S3: }

\newpage

\end{document}

### 2017-10-31 code/dejea.R
fix dejea metadata problem

@@ -7,27 +7,28 @@
 # Load in needed functions and libraries
 source('code/functions.R')
 
-loadLibs(c("dplyr", "tidyr"))
+loadLibs(c("tidyverse"))
 
-shared <- read.table("data/process/dejea/combined.unique.good.filter.unique.precluster.pick.pick.opti_mcc.unique_list.shared", 
-                     header=T, stringsAsFactors = F)
+shared <- read_tsv("data/process/dejea/combined.unique.good.filter.unique.precluster.pick.pick.opti_mcc.unique_list.shared")
 
 
-metadata1 <- read.csv("data/process/dejea/dejea_metadata.csv", 
-                      header = T, stringsAsFactors=F)
+metadata1 <- read_csv("data/process/dejea/dejea_metadata.csv")
 
-metadata2 <- read.delim("data/process/dejea/SraRunTable.txt", 
-                        header = T, stringsAsFactors = F)
+metadata2 <- read_tsv("data/process/dejea/SraRunTable.txt")
 
 # Merge the metadata together and create needed columns
 temp_data <- metadata2 %>% separate(Sample_Name_s, 
                                into = c("patient_id", "disease", "location")) %>% 
-  select(Run_s, patient_id) %>% 
+  select(Run_s, patient_id, disease, location) %>% 
   inner_join(metadata1, by = "patient_id") %>% 
   mutate(white = ifelse(race == "caucasian" | race == "hispanic", 1, 0)) %>% 
-  rename(sample = Run_s, disease = patient_type, site = tumor_site) %>% 
-  select(-patient_id) %>% 
-  mutate(matched = rep("y", length(sample)))
+  rename(sample = Run_s, site = tumor_site) %>% 
+  select(-patient_type) %>% 
+  
+  mutate(disease = stringr::str_replace(disease, "Normal", "control"), 
+         disease = stringr::str_replace(disease, "Tumor", "cancer"), 
+         disease = stringr::str_replace(disease, "Polyp", "polyp"), 
+    matched = rep("y", length(sample)))
 
 
 # Align shared file with metadata
@@ -39,5 +40,5 @@ stopifnot(temp_shared$Group == temp_data$sample)
 
 write.table(temp_shared, file="data/process/dejea/dejea.shared", quote=F, sep='\t', row.names=F)
 
-write.table(temp_data, file="data/process/dejea/dejea.metadata", quote=F, sep='\t', row.names=F)
+write_tsv(temp_data, "data/process/dejea/dejea.metadata")
  

### 2017-10-31 code/make_est_power_graph.R

@@ -78,10 +78,45 @@ adn_sample_n <- adn_power %>%
         axis.text.y = element_text(size = 10))
 
 
+crc_power %>% 
+  mutate(study = factor(study, 
+                        levels = c("zeller", "weir", "wang", "sana", "hale", "geng", 
+                                   "flemer", "burns", "baxter", "ahn"), 
+                        labels = c("Zeller", "Weir", "Wang", "Sanapareddy", "Hale", "Geng", 
+                                   "Flemer", "Burns", "Baxter", "Ahn")), 
+         effect_size = factor(effect_size, 
+                              levels = c(0.01, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3), 
+                              labels = c("1%", "5%", "10%", "15%", "20%", "25%", "30%"))) %>% 
+  ggplot(aes(effect_size, study_power, color = study, group = study)) + 
+  geom_jitter(width = 0.3, size = 3) + 
+  coord_cartesian(ylim = c(0,1)) + 
+  geom_vline(xintercept = c(1.5, 2.5, 3.5, 4.5, 5.5, 6.5), color = "gray") + 
+  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") + 
+  labs(x = "Effect Size", y = "Study Power") + theme_bw() + ggtitle("A") + 
+  scale_color_manual(name = "Study", 
+                     values = c('#FDE725FF', '#24878EFF', '#97D83FFF', '#1F998AFF', 
+                                '#67CC5CFF', '#CBE11EFF', '#440154FF', '#453581FF', 
+                                '#481D6FFF', '#40BC72FF')) + 
+  theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank(), 
+        axis.text.y = element_text(size = 10))
 
 
-
-
+# flemer - #440154FF
+# lu - #FDE725FF
+# burns - #453581FF
+# chen - #3D4D8AFF
+# sana - #1F998AFF
+# dejea - #2B748EFF
+# geng - #CBE11EFF
+# brim - #34618DFF
+# zeller - #FDE725FF
+# baxter - #481D6FFF
+# hale - #67CC5CFF
+# wang - #97D83FFF
+# weir - #24878EFF
+# ahn - #40BC72FF
 
 
 ##############################################################################################

### 2017-10-31 code/run_power_analysis.R
add dejea

@@ -18,7 +18,7 @@ stool_sets <- c("brim", "wang", "weir", "ahn", "zeller", "baxter", "hale", "flem
 # Remove Lu since it only has polyps and no cancer cases
 # Have to remove dejea since there are no controls
 # Have to omit Chen since their is only 1 case
-tissue_sets <- c("lu", "geng", "sana", "burns")
+tissue_sets <- c("dejea", "lu", "geng", "sana", "burns")
 
 
 

### 2017-10-31 code/make_est_power_graph.R
add graph write out

@@ -43,7 +43,7 @@ adn_study_power <- adn_power %>%
                               levels = c(0.01, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3), 
                               labels = c("1%", "5%", "10%", "15%", "20%", "25%", "30%"))) %>% 
   ggplot(aes(effect_size, study_power, color = study, group = study)) + 
-  geom_jitter(width = 0.3, size = 3) + 
+  geom_jitter(width = 0.3, size = 3, show.legend = F) + 
   coord_cartesian(ylim = c(0,1)) + 
   geom_vline(xintercept = c(1.5, 2.5, 3.5, 4.5, 5.5, 6.5), color = "gray") + 
   geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") + 
@@ -51,7 +51,7 @@ adn_study_power <- adn_power %>%
   scale_color_manual(name = "Study", 
                      values = c('#FDE725FF', '#FDE725FF', '#67CC5CFF', '#440154FF', 
                                 '#34618DFF', '#481D6FFF')) + 
-  theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
+  theme(plot.title = element_text(face="bold", hjust = -0.15, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10))
@@ -72,53 +72,71 @@ adn_sample_n <- adn_power %>%
   scale_color_manual(name = "Study", 
                      values = c('#FDE725FF', '#FDE725FF', '#67CC5CFF', '#440154FF', 
                                 '#34618DFF', '#481D6FFF')) + 
-  theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
+  theme(plot.title = element_text(face="bold", hjust = -0.25, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10))
 
 
-crc_power %>% 
+crc_study_power <- crc_power %>% 
   mutate(study = factor(study, 
                         levels = c("zeller", "weir", "wang", "sana", "hale", "geng", 
-                                   "flemer", "burns", "baxter", "ahn"), 
+                                   "flemer", "dejea", "burns", "baxter", "ahn"), 
                         labels = c("Zeller", "Weir", "Wang", "Sanapareddy", "Hale", "Geng", 
[…]

### 2017-10-31 submission/manuscript.Rmd
update methods

@@
 ***Study Analysis Overview:***
+Alpha diversity was first assessed for differences between controls and adenoma versus cancer and controls versus adenoma.
+We analyzed the data using standard T-tests, ANOVA, linear mixed-effect models, and relative risk.
+Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk.
+We then built Random Forest models based on all genera or the select CRC-associated genera.
+The models were trained on one study then tested on the remaining studies for every study.
+The data was split between feces and tissue samples.
+Within the tissue groups the data was further divided between matched and unmatched tissue samples.
+Both prediction for adenoma and carcinoma were tested.
+This same approach was then applied at the OTU level with the exception that instead of testing on the other studies a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs.
+The power of each study was assessed for and effect size ranging from 1% to 30%.
+An estimated sample n for these effect sizes was also generated based on 80% power.
 

### 2017-10-31 submission/manuscript.tex
update methods

@@
 
-\textbf{\emph{Statistical Analysis.}} All statistical analysis after
+\textbf{\emph{Statistical Analysis:}} All statistical analysis after
 sequence processing utilized the R software package (v3.4.2). For the
@@
 
-\textbf{\emph{Reproducible Methods.}} The code and analysis can be found
+\textbf{\emph{Study Analysis Overview:}} Alpha diversity was first
+assessed for differences between controls and adenoma versus cancer and
+controls versus adenoma. We analyzed the data using standard T-tests,
+ANOVA, linear mixed-effect models, and relative risk. Next, four
+specific CRC-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
+differences in relative risk. We then built Random Forest models based
+on all genera or the select CRC-associated genera. The models were
+trained on one study then tested on the remaining studies for every
+study. The data was split between feces and tissue samples. Within the
+tissue groups the data was further divided between matched and unmatched
+tissue samples. Both prediction for adenoma and carcinoma were tested.
+This same approach was then applied at the OTU level with the exception
+that instead of testing on the other studies a 10-fold cross validation
+was utilized and 100 different models were created based on random 80/20
+splitting of the data to generate a range of expected AUCs. The power of
+each study was assessed for and effect size ranging from 1\% to 30\%. An
+estimated sample n for these effect sizes was also generated based on
+80\% power.
+
+\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
 here \url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_XXXX_2016}.

### 2017-10-31 code/make_est_power_graph.R
update sample n

@@ -12,7 +12,9 @@ loadLibs(c("tidyverse", "gridExtra", "viridis"))
 adn_power <- read_csv("data/process/tables/adn_predicted_pwr_and_n.csv")
 crc_power <- read_csv("data/process/tables/cancer_predicted_pwr_and_n.csv")
 
-
+combined_data <- adn_power %>% 
+  filter(study %in% c("lu", "brim")) %>% 
+  bind_rows(crc_power)
 
 ##############################################################################################
 ############################## List of functions to be used  #################################
@@ -51,28 +53,8 @@ adn_study_power <- adn_power %>%
   scale_color_manual(name = "Study", 
                      values = c('#FDE725FF', '#FDE725FF', '#67CC5CFF', '#440154FF', 
                                 '#34618DFF', '#481D6FFF')) + 
+  annotate("text", label = paste("Adenoma"), x = 0.78, y = 1.03, size = 2.5) +
   theme(plot.title = element_text(face="bold", hjust = -0.15, size = 20), 
-        panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank(), 
-        axis.text.y = element_text(size = 10))
-
-
-adn_sample_n <- adn_power %>% 
-  mutate(study = factor(study, 
-                        levels = c("zeller", "lu", "hale", "flemer_t", "brim", "baxter"), 
-                        labels = c("Zeller", "Lu", "Hale", "Flemer\n(Tissue)", "Brim", "Baxter")), 
-         effect_size = factor(effect_size, 
-                              levels = c(0.01, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3), 
-                              labels = c("1%", "5%", "10%", "15%", "20%", "25%", "30%"))) %>% 
-  ggplot(aes(effect_size, log(pwr80_needed_n), color = study, group = study)) + 
-  geom_jitter(width = 0.3, size = 3) + 
-  coord_cartesian(ylim = c(0,12)) + 
-  geom_vline(xintercept = c(1.5, 2.5, 3.5, 4.5, 5.5, 6.5), color = "gray") + 
-  labs(x = "Effect Size", y = expression(Log["10"]~Sample~Number)) + theme_bw() + ggtitle("B") + 
-  scale_color_manual(name = "Study", 
-                     values = c('#FDE725FF', '#FDE725FF', '#67CC5CFF', '#440154FF', 
-                                '#34618DFF', '#481D6FFF')) + 
[…]

### 2017-10-31 submission/manuscript.tex
change title

@@
 
-\section{The Importance of the Microbiota and Individual Community
-Members in Colorectal
-Cancer}\label{the-importance-of-the-microbiota-and-individual-community-members-in-colorectal-cancer}
+\section{The Microbiota and Individual Community Members in Colorectal
+Cancer: Is There a Common
+Theme?}\label{the-microbiota-and-individual-community-members-in-colorectal-cancer-is-there-a-common-theme}
 

### 2017-10-31 submission/manuscript.Rmd
change title
change hperlink

@@
 
-# The Importance of the Microbiota and Individual Community Members in Colorectal Cancer
+# The Microbiota and Individual Community Members in Colorectal Cancer: Is There a Common Theme?
 
@@
 ***Reproducible Methods:***
-The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_XXXX_2016.
+The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
 Unless mentioned otherwise the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file, on the GitHub repository or in the original manuscript.

### 2017-10-31 submission/manuscript.Rmd
simplify methods and add first part of section 1 results

@@
 
-```{r rdata_tables}
+```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+#Load in linear mixed-effect data tables
+stool_lme <- as.data.frame(read_csv("../data/process/tables/alpha_mixeffect_results.csv"))
+unmatch_tissue_lme <- as.data.frame(read_csv("../data/process/tables/alpha_unmatched_mixeffect_results.csv"))
+match_tissue_lme <- as.data.frame(read_csv("../data/process/tables/alpha_matched_mixeffect_results.csv"))
 
@@
 
-microbiota; colorectal cancer; polyps; treatment; risk factor.
+microbiota; colorectal cancer; polyps; adenoma; meta-analysis.
 
@@
 
-***Section 1***
+***Fecal Diversity is Lower in Those with Carcinoma:***
+Using power transformed and Z-score normalized alpha diversity metrics both evenness and the Shannon diversity metrics in feces was lower in those with carcinoma then in controls but not for tissue samples [Figure 1].
+Using linear mixed-effects to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`).
+This effect was not observed in tissue when additionally controlling for whether the sample came from the same individual (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
 
@@
 For the alpha diversity analysis values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package.
-Testing for alpha diversity differences utilized both T-tests and ANOVA.
-Linear mixed-effect models were created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for both study and variable region effect in the diversity measures when analyzing colorectal cancer groups.
+Testing for alpha diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for both study and variable region effect in the diversity measures when analyzing colorectal cancer groups.
 Relative Risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages.
@@
 Alpha diversity was first assessed for differences between controls and adenoma versus cancer and controls versus adenoma.
-We analyzed the data using standard T-tests, ANOVA, linear mixed-effect models, and relative risk.
+We analyzed the data using linear mixed-effect models, and relative risk.
+Beta-diversity was then assessed for each inidividual study.
 Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk.

### 2017-10-31 submission/manuscript.tex
change hperlink
simplify methods and add first part of section 1 results

@@
 
-\textbf{\emph{Section 1}}
+\textbf{\emph{Fecal Diversity is Lower in Those with Carcinoma:}} Using
+power transformed and Z-score normalized alpha diversity metrics both
+evenness and the Shannon diversity metrics in feces was lower in those
+with carcinoma then in controls but not for tissue samples {[}Figure
+1{]}. Using linear mixed-effects to control for study and variable
+region there was a significant decrease from control to adenoma to
+carcinoma for both evenness (P-value = 0.025) and Shannon diversity
+(P-value = 0.043). This effect was not observed in tissue when
+additionally controlling for whether the sample came from the same
+individual (P-value \textgreater{} 0.05).
 
@@
 rcompanion (v1.10.1) package and then Z-score normalized using the car
-(v2.1.5) package. Testing for alpha diversity differences utilized both
-T-tests and ANOVA. Linear mixed-effect models were created using the
-lme4 (v1.1.14) package to correct for both study and variable region
-effect in the diversity measures when analyzing colorectal cancer
-groups. Relative Risk was analyzed using both the epiR (v0.9.87) and
-metafor (v2.0.0) packages. Relative risk significance testing utilized
-the chi-squred test. Beta-diversity differences utilized a Bray-Curtis
-distance matrix and PERMANOVA executed with the vegan (v2.4.4) package.
-Random Forest models were built using both the caret (v6.0.77) and
-randomForest (v4.6.12) packages. Random Forest testing of the obtained
-AUC versus a random model AUC utilized T-tests. Power analysis and
-estimations were made using the pwr (v1.2.1) and statmod (v1.4.30)
-packages. All figures were created using both ggplot2 (v2.2.1) and
-gridExtra (v2.3) packages.
+(v2.1.5) package. Testing for alpha diversity differences utilized
+linear mixed-effect models created using the lme4 (v1.1.14) package to
+correct for both study and variable region effect in the diversity
+measures when analyzing colorectal cancer groups. Relative Risk was
+analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages.
+Relative risk significance testing utilized the chi-squred test.
+Beta-diversity differences utilized a Bray-Curtis distance matrix and
+PERMANOVA executed with the vegan (v2.4.4) package. Random Forest models
+were built using both the caret (v6.0.77) and randomForest (v4.6.12)
+packages. Random Forest testing of the obtained AUC versus a random
+model AUC utilized T-tests. Power analysis and estimations were made
+using the pwr (v1.2.1) and statmod (v1.4.30) packages. All figures were
+created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages.
 
@@
 assessed for differences between controls and adenoma versus cancer and
-controls versus adenoma. We analyzed the data using standard T-tests,
-ANOVA, linear mixed-effect models, and relative risk. Next, four
-specific CRC-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
-differences in relative risk. We then built Random Forest models based
-on all genera or the select CRC-associated genera. The models were
-trained on one study then tested on the remaining studies for every
-study. The data was split between feces and tissue samples. Within the
-tissue groups the data was further divided between matched and unmatched
-tissue samples. Both prediction for adenoma and carcinoma were tested.
-This same approach was then applied at the OTU level with the exception
-that instead of testing on the other studies a 10-fold cross validation
-was utilized and 100 different models were created based on random 80/20
-splitting of the data to generate a range of expected AUCs. The power of
-each study was assessed for and effect size ranging from 1\% to 30\%. An
-estimated sample n for these effect sizes was also generated based on
-80\% power.
+controls versus adenoma. We analyzed the data using linear mixed-effect
+models, and relative risk. Beta-diversity was then assessed for each
+inidividual study. Next, four specific CRC-associated genera
+(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, and
+\emph{Porphyromonas}) were assessed for differences in relative risk. We
+then built Random Forest models based on all genera or the select
+CRC-associated genera. The models were trained on one study then tested
+on the remaining studies for every study. The data was split between
+feces and tissue samples. Within the tissue groups the data was further
+divided between matched and unmatched tissue samples. Both prediction
+for adenoma and carcinoma were tested. This same approach was then
+applied at the OTU level with the exception that instead of testing on
+the other studies a 10-fold cross validation was utilized and 100
+different models were created based on random 80/20 splitting of the
+data to generate a range of expected AUCs. The power of each study was
+assessed for and effect size ranging from 1\% to 30\%. An estimated
+sample n for these effect sizes was also generated based on 80\% power.
 
 \textbf{\emph{Reproducible Methods:}} The code and analysis can be found
-here \url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_XXXX_2016}.
+here
+\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
 Unless mentioned otherwise the accession number for the raw sequences

### 2017-10-31 submission/manuscript.Rmd
complete first result section

@@
 
-#Load in linear mixed-effect data tables
+#Load in linear mixed-effect alpha diversity data tables 
 stool_lme <- as.data.frame(read_csv("../data/process/tables/alpha_mixeffect_results.csv"))
@@
 match_tissue_lme <- as.data.frame(read_csv("../data/process/tables/alpha_matched_mixeffect_results.csv"))
+
+#Load in relative risk alpha diversity data tables
+alpha_RR_crc_stool <- as.data.frame(read_csv("../data/process/tables/alpha_RR_composite.csv"))
+alpha_RR_adn_stool <- as.data.frame(read_csv("../data/process/tables/alpha_adn_RR_composite.csv"))
+
+alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_RR_unmatched_tissue_composite.csv"))
+alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_RR_matched_tissue_composite.csv"))
+
+alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_RR_tissue_composite.csv"))
 
@@
 
-***Fecal Diversity is Lower in Those with Carcinoma:***
-Using power transformed and Z-score normalized alpha diversity metrics both evenness and the Shannon diversity metrics in feces was lower in those with carcinoma then in controls but not for tissue samples [Figure 1].
+***Fecal Diversity is Lower in Those with Carcinoma and increases Relative Risk:***
+Using power transformed and Z-score normalized alpha diversity metrics both evenness and the Shannon diversity metrics in feces are lower in those with carcinoma then in controls but not for tissue samples [Figure 1].
 Using linear mixed-effects to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`).
 This effect was not observed in tissue when additionally controlling for whether the sample came from the same individual (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
+For fecal samples a decrease in Shannon diversity and evenness resulted in a significantly increased relative risk for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2].
+Although these values were significant the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2].
+There was no increased relative risk for these metrics for adenoma or for tissue in general [Figure S1-3]
 

### 2017-10-31 submission/manuscript.tex
complete first result section

@@
 
-\textbf{\emph{Fecal Diversity is Lower in Those with Carcinoma:}} Using
-power transformed and Z-score normalized alpha diversity metrics both
-evenness and the Shannon diversity metrics in feces was lower in those
-with carcinoma then in controls but not for tissue samples {[}Figure
-1{]}. Using linear mixed-effects to control for study and variable
-region there was a significant decrease from control to adenoma to
-carcinoma for both evenness (P-value = 0.025) and Shannon diversity
-(P-value = 0.043). This effect was not observed in tissue when
-additionally controlling for whether the sample came from the same
-individual (P-value \textgreater{} 0.05).
+\textbf{\emph{Fecal Diversity is Lower in Those with Carcinoma and
+increases Relative Risk:}} Using power transformed and Z-score
+normalized alpha diversity metrics both evenness and the Shannon
+diversity metrics in feces are lower in those with carcinoma then in
+controls but not for tissue samples {[}Figure 1{]}. Using linear
+mixed-effects to control for study and variable region there was a
+significant decrease from control to adenoma to carcinoma for both
+evenness (P-value = 0.025) and Shannon diversity (P-value = 0.043). This
+effect was not observed in tissue when additionally controlling for
+whether the sample came from the same individual (P-value \textgreater{}
+0.05). For fecal samples a decrease in Shannon diversity and evenness
+resulted in a significantly increased relative risk for carcinoma
+(P-value = 0.01 and P-value = 0.0011, respectively) {[}Figure 2{]}.
+Although these values were significant the effect size was relatively
+small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
+{[}Figure 2{]}. There was no increased relative risk for these metrics
+for adenoma or for tissue in general {[}Figure S1-3{]}
 

### 2017-10-31 submission/manuscript.Rmd
fix typo and add next section

@@
 
-***Fecal Diversity is Lower in Those with Carcinoma and increases Relative Risk:***
+***Fecal Diversity is Lower in Those with Carcinoma and Increases Relative Risk:***
 Using power transformed and Z-score normalized alpha diversity metrics both evenness and the Shannon diversity metrics in feces are lower in those with carcinoma then in controls but not for tissue samples [Figure 1].
@@
 
-***Section 2***
+***Genera Previously Associated with Carcinoma Increases Relative Risk More than Alpha Diversity:***
 

### 2017-10-31 submission/manuscript.Rmd
add results section 2

@@
 alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_RR_tissue_composite.csv"))
+
+#Load in relative risk select genera data tables
+select_RR_stool <- as.data.frame(read_csv("../data/process/tables/select_genus_RR_stool_composite.csv"))
+select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_RR_tissue_composite.csv"))
 
@@
 ***Genera Previously Associated with Carcinoma Increases Relative Risk More than Alpha Diversity:***
+Both fecal and tissue samples had a significantly increased RR for carcinoma but not for adenoma [Figure 3] which was greater than either evenness or Shannon diversity [Figure 2 & 3].
+The relative risk did not increase when considering the total abundance or increasing number of carcinoma associated genera [Figure 3].
+The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) .
+This decrease may be explained by the fact that tissue samples include matched samples.
 

### 2017-10-31 submission/manuscript.tex
add results section 2

@@
 \textbf{\emph{Fecal Diversity is Lower in Those with Carcinoma and
-increases Relative Risk:}} Using power transformed and Z-score
+Increases Relative Risk:}} Using power transformed and Z-score
 normalized alpha diversity metrics both evenness and the Shannon
@@
 
-\textbf{\emph{Section 2}}
+\textbf{\emph{Genera Previously Associated with Carcinoma Increases
+Relative Risk More than Alpha Diversity:}} Both fecal and tissue samples
+had a significantly increased RR for carcinoma but not for adenoma
+{[}Figure 3{]} which was greater than either evenness or Shannon
+diversity {[}Figure 2 \& 3{]}. The relative risk did not increase when
+considering the total abundance or increasing number of carcinoma
+associated genera {[}Figure 3{]}. The RR effect size was greater for
+stool (RR range = 1.78 - 2.64) then for tissue (RR range = 1.33 - 1.53)
+. This decrease may be explained by the fact that tissue samples include
+matched samples.
 

### 2017-10-31 submission/supplemental.Rmd
first commit

---
title: ''
csl: microbiome.csl
fontsize: 12pt
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
geometry: margin=1.0in
bibliography: null
---

# The Microbiota and Individual Community Members in Colorectal Cancer: Is There a Common Theme?

\vspace{10mm}

\begin{center}
Marc A Sze and Patrick D Schloss

\vspace{10mm}

Supplemental
\end{center}

\newpage

```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

source('../code/functions.R')

loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown"))

```

**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA Adenoma Stool Studies**

```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}

temp_t1 <- read_csv("../data/process/tables/beta_perm_adn_stool_summary.csv")

table_s1 <- temp_t1 %>% 
  mutate(fstat = round(fstat, digits = 2), 
         r2 = round(r2, digits = 3), 
         pvalue = round(pvalue, digits = 4), 
         study = stringr::str_replace(study, 
                                      c("brim", "zeller", "baxter", "hale"), 
                                      c("Brim", "Zeller", "Baxter", "Hale")))

colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Adenoma n", "Control n")

kable(table_s1, format = "markdown", align = 'c', padding = 2)
```

\newpage

```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}

```

\newpage

\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}

**Table S2: Summary of Common OTUs for All Diagnosis Groups**

\footnotesize

```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}

```
\normalsize
\newpage

**Table S3: Summary of Comparisons Based on Additive Treatment Received (Chemotherapy or Chemotherapy and Radiation)**
\footnotesize
```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}

```
\normalsize

\newpage

**Table S4: Summary of Adenoma based on Surgery Received**

```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}

```

\newpage

**Table S5: Summary of mtry ranges used for the treatment and diagnosis models**

\footnotesize

```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}

```
\normalsize

### 2017-10-31 submission/supplemental.Rmd
add stool bray st2

@@
 
-**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA Adenoma Stool Studies**
+**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Adenoma**
 
@@
 
+**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Carcinoma**
+
 ```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
 
+temp_t2 <- read_csv("../data/process/tables/beta_perm_stool_summary.csv")
+
+table_s2 <- temp_t2 %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = stringr::str_replace(study, 
+                                      c("wang", "weir", "ahn", "zeller", "baxter", "hale", "flemer"), 
+                                      c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")))
+
+colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma n", "Control n")
+
+kable(table_s2, format = "markdown", align = 'c', padding = 2)
 ```
+
+\newpage
+
+**Table S3: **
+\footnotesize
+```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
+
+```
+\normalsize
 
@@
 
-```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
+```{r supplemental_table, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 \normalsize
-\newpage
-
-**Table S3: Summary of Comparisons Based on Additive Treatment Received (Chemotherapy or Chemotherapy and Radiation)**
-\footnotesize
-```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
-
-```
-\normalsize
-
 \newpage

### 2017-10-31 submission/supplemental.Rmd
add table S3

@@
 
-**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Adenoma**
+**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Adenoma**
 
@@
 
-**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Carcinoma**
+**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Carcinoma**
 
@@
 
-**Table S3: **
-\footnotesize
+**Table S3: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue Adenoma**
+
 ```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
 
+temp_t3a <- read_csv("../data/process/tables/beta_perm_adn_unmatched_tissue_summary.csv")
+temp_t3b <- read_csv("../data/process/tables/beta_perm_adn_matched_tissue_summary.csv")
+
+table_s3 <- temp_t3a %>% 
+  bind_rows(temp_t3b) %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = c("Lu", "Flemer", "Lu (Matched)"))
+
+colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma n", "Control n")
+
+kable(table_s3, format = "markdown", align = 'c', padding = 2)
 ```
-\normalsize
 

### 2017-10-31 submission/supplemental.Rmd
add table s4

@@
 
-colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma n", "Control n")
+colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Adenoma n", "Control n")
 
 kable(table_s3, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S4: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue Carcinoma**
+
+```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
+
+temp_t4a <- read_csv("../data/process/tables/beta_perm_unmatched_tissue_summary.csv")
+temp_t4b <- read_csv("../data/process/tables/beta_perm_matched_tissue_summary.csv")
+
+table_s4 <- temp_t4a %>% 
+  bind_rows(temp_t4b) %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = c("Sanapareddy", "Burns", "Flemer", "Chen", 
+                   "Dejea (Matched)", "Geng (Matched)", "Burns (Matched)"))
+
+colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma n", "Control n")
+
+kable(table_s4, format = "markdown", align = 'c', padding = 2)
+
 ```
@@
 
-**Table S4: Summary of Adenoma based on Surgery Received**
-
-```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
-
-```
-
-\newpage
-
 **Table S5: Summary of mtry ranges used for the treatment and diagnosis models**

### 2017-10-31 submission/manuscript.Rmd
minor fixes
complete first result section

@@
 Although these values were significant the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2].
-There was no increased relative risk for these metrics for adenoma or for tissue in general [Figure S1-3]
+There was no increased relative risk for these metrics for adenoma or for tissue in general [Figure S1-3].
+
+Using the Bray-Curtis distance metric, the fecal microbiota did not have a different community diversity between adenoma and control but did for carcinoma across studies [Table S1 & S2].
+The majority of unmatched tissue samples had a significant difference for both adenoma and carcinoma versus controls [Table S3 & S4].
+All matched tissue samples accross studies had no difference between any of the compared groups [Table S3 & S4].
 
@@
 
-***Section 3***
+***Using the Whole Community Increases AUC:***
 
@@
 
-**Table 1: **
+**Table 1:**
 

### 2017-10-31 submission/manuscript.tex
complete first result section

@@
 {[}Figure 2{]}. There was no increased relative risk for these metrics
-for adenoma or for tissue in general {[}Figure S1-3{]}
+for adenoma or for tissue in general {[}Figure S1-3{]}.
+
+Using the Bray-Curtis distance metric, the fecal microbiota did not have
+a different community diversity between adenoma and control but did for
+carcinoma across studies {[}Table S1 \& S2{]}. The majority of unmatched
+tissue samples had a significant difference for both adenoma and
+carcinoma versus controls {[}Table S3 \& S4{]}. All matched tissue
+samples accross studies had no difference between any of the compared
+groups {[}Table S3 \& S4{]}.
 
@@
 
-\textbf{\emph{Section 3}}
+\textbf{\emph{Using the Whole Community Increases AUC:}}
 
@@
 
-\textbf{Table 1: }
+\textbf{Table 1:}
 

### 2017-11-01 submission/manuscript.tex
update result section name and add methods on CRC associated Genera OTU model

@@
 
-\textbf{\emph{Using the Whole Community Increases AUC:}}
+\textbf{\emph{Using the Whole Community Increases Model AUC:}}
 
@@
 different models were created based on random 80/20 splitting of the
-data to generate a range of expected AUCs. The power of each study was
-assessed for and effect size ranging from 1\% to 30\%. An estimated
-sample n for these effect sizes was also generated based on 80\% power.
+data to generate a range of expected AUCs. For OTU based models the CRC
+Associated Genera included all OTUs that had a taxonomic classification
+to \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus},
+or \emph{Porphyromonas}. The power of each study was assessed for and
+effect size ranging from 1\% to 30\%. An estimated sample n for these
+effect sizes was also generated based on 80\% power.
 

### 2017-11-01 submission/manuscript.Rmd
update result section name and add methods on CRC associated Genera OTU model
add results on whole community versus specific OTUs

@@
 
-***Using the Whole Community Increases AUC:***
+***Using the Whole Community Increases Model AUC:***
+For both fecal and tissue samples (matched and unmatched) there was a decrease in AUC when only OTUs from the CRC associated genera are used [Figure 4 & 5].
+This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5].
+The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S4-S4].
+Both genus models perform similarily in their ability to be able to predict lesion (adenoma or carcinoma) with carcinoma having a higher AUC then adenoma [Figure S6-S8].
+Matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S6A, S7B, & S8] then carcinoma models [Figure S6B & S7A].
 
@@
 This same approach was then applied at the OTU level with the exception that instead of testing on the other studies a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs.
+For OTU based models the CRC Associated Genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
 The power of each study was assessed for and effect size ranging from 1% to 30%.

### 2017-11-01 submission/manuscript.tex
add results on whole community versus specific OTUs

@@
 
-\textbf{\emph{Using the Whole Community Increases Model AUC:}}
+\textbf{\emph{Using the Whole Community Increases Model AUC:}} For both
+fecal and tissue samples (matched and unmatched) there was a decrease in
+AUC when only OTUs from the CRC associated genera are used {[}Figure 4
+\& 5{]}. This decrease is observed in both adenoma and carcinoma groups
+{[}Figure 4 \& 5{]}. The genus models generally had similar trends as
+observed for the OTU based models with the full genera models performing
+better then the CRC associated genera models {[}Figure S4-S4{]}. Both
+genus models perform similarily in their ability to be able to predict
+lesion (adenoma or carcinoma) with carcinoma having a higher AUC then
+adenoma {[}Figure S6-S8{]}. Matched tissue samples for those with
+carcinoma had an AUC that was more similar to the adenoma models
+{[}Figure S6A, S7B, \& S8{]} then carcinoma models {[}Figure S6B \&
+S7A{]}.
 

### 2017-11-01 code/make_est_power_graph.R
conver to log2

@@ -95,11 +95,11 @@ combined_sample_n <- combined_data %>%
          effect_size = factor(effect_size, 
                               levels = c(0.01, 0.05, 0.1, 0.15, 0.2, 0.25, 0.3), 
                               labels = c("1%", "5%", "10%", "15%", "20%", "25%", "30%"))) %>% 
-  ggplot(aes(effect_size, log(pwr80_needed_n), color = study, group = study)) + 
+  ggplot(aes(effect_size, log2(pwr80_needed_n), color = study, group = study)) + 
   geom_jitter(width = 0.3, size = 3) + 
-  coord_cartesian(ylim = c(0,12)) + 
+  coord_cartesian(ylim = c(0,16)) + 
   geom_vline(xintercept = c(1.5, 2.5, 3.5, 4.5, 5.5, 6.5), color = "gray") + 
-  labs(x = "Effect Size", y = expression(Log["10"]~Sample~Number)) + theme_bw() + ggtitle("C") + 
+  labs(x = "Effect Size", y = expression(Log["2"]~Sample~Number)) + theme_bw() + ggtitle("C") + 
   scale_color_manual(name = "Study", 
                      values = c('#FDE725FF', '#24878EFF', '#97D83FFF', '#1F998AFF', '#FDE725FF',  
                                 '#67CC5CFF', '#CBE11EFF', '#440154FF', '#2B748EFF', '#453581FF', 

### 2017-11-01 submission/manuscript.Rmd
complete addition of last result section

@@
 This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5].
-The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S4-S4].
+The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S4-S5].
 Both genus models perform similarily in their ability to be able to predict lesion (adenoma or carcinoma) with carcinoma having a higher AUC then adenoma [Figure S6-S8].
@@
 
-***Section 4***
+***Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
+When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B].
+No single study that was analyzed had the standard 80% power to detect a difference that was eqaul to or below 10% [Figure 6A & B].
+In order to achieve adequate power for small effect sizes it would be necessary to recruit over 1000 individuals for each arm of the study [Figure 6C].
 

### 2017-11-01 submission/manuscript.tex
complete addition of last result section

@@
 observed for the OTU based models with the full genera models performing
-better then the CRC associated genera models {[}Figure S4-S4{]}. Both
+better then the CRC associated genera models {[}Figure S4-S5{]}. Both
 genus models perform similarily in their ability to be able to predict
@@
 
-\textbf{\emph{Section 4}}
+\textbf{\emph{Majority of Studies are Underpowered for Detecting Small
+Effect Size Differences:}} When assessing the power of each study at
+different effect sizes the majority of studies for both adenoma and
+carcinoma have an 80\% power to detect a 30\% difference {[}Figure 6A \&
+B{]}. No single study that was analyzed had the standard 80\% power to
+detect a difference that was eqaul to or below 10\% {[}Figure 6A \&
+B{]}. In order to achieve adequate power for small effect sizes it would
+be necessary to recruit over 1000 individuals for each arm of the study
+{[}Figure 6C{]}.
 

### 2017-11-01 submission/manuscript.Rmd
fill in declarations section

@@
 
-Need to fill in.
+Ethics approval and informed consent for each of the studies used is mentioned in the respective manuscript used in this meta-analysis.
 
@@
 
-Need to fill in.
+A detailed and reporducible description of how the data were processed and analyzed for each study can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
+Raw sequences can be downloaded from the SRA in most cases and can be found in the respective studies batch file in the GitHub repo or within the original publication.
+When sequences were not publicly available contacting the corresponding author for raw sequences needs to be undertaken.
 
@@
 
-Need to fill in.
+MAS is supported by a CIHR fellowship and a University of Michigan PTSP fellowship grant.
 
@@
 
-Need to fill in.
+All authors helped to design and conceptualize the study.
+MAS identified and analyzed the data.
+MAS and PDS interpreted the data.
+MAS wrote the first draft of the manuscript and both he and PDS reviewed and revised updated versions.
+All authors approved the final manuscript.
 
@@
 
-Need to fill in.
+The authors would like to thank all the study participants who were apart of each of the individual studies uitlized.
+We would also like to thank each of the study authors for making their data available for use.
+Finally we would like to thank the members of the Schloss lab for valuable feed back and proof reading during the formulation of this manuscript.
 
@@
 
-**Table 2: **
-
-```{r table2, echo=FALSE, message=FALSE, warning=FALSE}
-
-```
-
-\newpage
-
 **Figure 1: **
@@
 **Figure 4: **
+
+**Figure 5: **
+
+**Figure 6: **
 
@@
 
+**Figure S4: **
+
+**Figure S5: **
+
+**Figure S6: **
+
+**Figure S7: **
+
+**Figure S8: **
+
 \newpage

### 2017-11-01 submission/manuscript.tex
fill in declarations section

@@
 
-Need to fill in.
+Ethics approval and informed consent for each of the studies used is
+mentioned in the respective manuscript used in this meta-analysis.
 
@@
 
-Need to fill in.
+A detailed and reporducible description of how the data were processed
+and analyzed for each study can be found at
+\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
+Raw sequences can be downloaded from the SRA in most cases and can be
+found in the respective studies batch file in the GitHub repo or within
+the original publication. When sequences were not publicly available
+contacting the corresponding author for raw sequences needs to be
+undertaken.
 
@@
 
-Need to fill in.
+MAS is supported by a CIHR fellowship and a University of Michigan PTSP
+fellowship grant.
 
@@
 
-Need to fill in.
+All authors helped to design and conceptualize the study. MAS identified
+and analyzed the data. MAS and PDS interpreted the data. MAS wrote the
+first draft of the manuscript and both he and PDS reviewed and revised
+updated versions. All authors approved the final manuscript.
 
@@
 
-Need to fill in.
+The authors would like to thank all the study participants who were
+apart of each of the individual studies uitlized. We would also like to
+thank each of the study authors for making their data available for use.
+Finally we would like to thank the members of the Schloss lab for
+valuable feed back and proof reading during the formulation of this
+manuscript.
 
@@
 
-\textbf{Table 2: }
-
-\newpage
-
 \textbf{Figure 1: }
@@
 \textbf{Figure 4: }
+
+\textbf{Figure 5: }
+
+\textbf{Figure 6: }
 
@@
 
+\textbf{Figure S4: }
+
+\textbf{Figure S5: }
+
+\textbf{Figure S6: }
+
+\textbf{Figure S7: }
+
+\textbf{Figure S8: }
+
 \newpage

### 2017-11-02 submission/manuscript.Rmd
add first paragraph of discussion

@@
 The relative risk did not increase when considering the total abundance or increasing number of carcinoma associated genera [Figure 3].
-The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) .
+The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`).
 This decrease may be explained by the fact that tissue samples include matched samples.
 
-***Using the Whole Community Increases Model AUC:***
+***Using the Whole Community Increases Model AUC over CRC Associated Genera:***
 For both fecal and tissue samples (matched and unmatched) there was a decrease in AUC when only OTUs from the CRC associated genera are used [Figure 4 & 5].
@@
 
+Our study identifies clear diversity changes both at the community level and within individual genera that are present in indivdiuals with carcinoma versus those without the disease.
+Although there was a step wise decrease in diversity from control to adenoma to carcinoma; this did not translate into large effect sizes for the relative risk of either of these two conditions.
+These clear changes were not easily recapitulated in those with adenoma.
+Even though CRC associated genera increase the relative risk of carcinoma they do not increase the relative risk of adenoma.
+This information suggests that these specific genera may not be the primary members of the microbial community that contributes to the formation of an adenoma but is for a carcinoma.
+Additionally, our data shows that by using the whole community our models perform better then when they only use the CRC associated genera.
+CRC associated genera are clearly important to carcinoma but the context or community in which these microbes are a part of can drastically increase the ability of models to make predictions.
+This data supports the concept that small localized changes within the community may be occuring that are important in the disease progression of colorectal cancer and that they may not directly involve CRC associated genera.
+
+- Flynn review and how data supports an idea of early specific localized community changes that becomes global in CRC
+
+- Builiding on this idea moving forward how CRC associated genera interact with the overall community rather than in isolation will be important since they do not increase the RR of polyps
+  - suggesting localized changes to community may be more important
+
+- In closing to get at these more difficult questions the Power calculations and effect size in future studies need to be accounted for especially for adenoma where the predicted effect sizes are going to be small
+
 ## Conclusion
+
+-aggregate studies done with feces and stool to investigate the role of the microbiota in adenoma and carcinoma.
+
+- Significantly adds to the existing research using this approach by expanding both the total individuals included and the scope.
+
+- leads to clear identifications of the challenges that lie ahead with respect to adenoma
 

### 2017-11-02 submission/manuscript.tex
add first paragraph of discussion

@@
 associated genera {[}Figure 3{]}. The RR effect size was greater for
-stool (RR range = 1.78 - 2.64) then for tissue (RR range = 1.33 - 1.53)
-. This decrease may be explained by the fact that tissue samples include
+stool (RR range = 1.78 - 2.64) then for tissue (RR range = 1.33 - 1.53).
+This decrease may be explained by the fact that tissue samples include
 matched samples.
 
-\textbf{\emph{Using the Whole Community Increases Model AUC:}} For both
-fecal and tissue samples (matched and unmatched) there was a decrease in
-AUC when only OTUs from the CRC associated genera are used {[}Figure 4
-\& 5{]}. This decrease is observed in both adenoma and carcinoma groups
-{[}Figure 4 \& 5{]}. The genus models generally had similar trends as
-observed for the OTU based models with the full genera models performing
-better then the CRC associated genera models {[}Figure S4-S5{]}. Both
-genus models perform similarily in their ability to be able to predict
-lesion (adenoma or carcinoma) with carcinoma having a higher AUC then
-adenoma {[}Figure S6-S8{]}. Matched tissue samples for those with
-carcinoma had an AUC that was more similar to the adenoma models
-{[}Figure S6A, S7B, \& S8{]} then carcinoma models {[}Figure S6B \&
-S7A{]}.
+\textbf{\emph{Using the Whole Community Increases Model AUC over CRC
+Associated Genera:}} For both fecal and tissue samples (matched and
+unmatched) there was a decrease in AUC when only OTUs from the CRC
+associated genera are used {[}Figure 4 \& 5{]}. This decrease is
+observed in both adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The
+genus models generally had similar trends as observed for the OTU based
+models with the full genera models performing better then the CRC
+associated genera models {[}Figure S4-S5{]}. Both genus models perform
+similarily in their ability to be able to predict lesion (adenoma or
+carcinoma) with carcinoma having a higher AUC then adenoma {[}Figure
+S6-S8{]}. Matched tissue samples for those with carcinoma had an AUC
+that was more similar to the adenoma models {[}Figure S6A, S7B, \& S8{]}
+then carcinoma models {[}Figure S6B \& S7A{]}.
 
@@
 
+Our study identifies clear diversity changes both at the community level
+and within individual genera that are present in indivdiuals with
+carcinoma versus those without the disease. Although there was a step
+wise decrease in diversity from control to adenoma to carcinoma; this
+did not translate into large effect sizes for the relative risk of
+either of these two conditions. These clear changes were not easily
+recapitulated in those with adenoma. Even though CRC associated genera
+increase the relative risk of carcinoma they do not increase the
+relative risk of adenoma. This information suggests that these specific
+genera may not be the primary members of the microbial community that
+contributes to the formation of an adenoma but is for a carcinoma.
+Additionally, our data shows that by using the whole community our
+models perform better then when they only use the CRC associated genera.
+CRC associated genera are clearly important to carcinoma but the context
+or community in which these microbes are a part of can drastically
+increase the ability of models to make predictions. This data supports
+the concept that small localized changes within the community may be
+occuring that are important in the disease progression of colorectal
+cancer and that they may not directly involve CRC associated genera.
+
+\begin{itemize}
+\item
+  Flynn review and how data supports an idea of early specific localized
+  community changes that becomes global in CRC
+\item
+  Builiding on this idea moving forward how CRC associated genera
+  interact with the overall community rather than in isolation will be
+  important since they do not increase the RR of polyps
+\item
+  suggesting localized changes to community may be more important
+\item
+  In closing to get at these more difficult questions the Power
+  calculations and effect size in future studies need to be accounted
+  for especially for adenoma where the predicted effect sizes are going
+  to be small
+\end{itemize}
+
 \subsection{Conclusion}\label{conclusion}
+
+-aggregate studies done with feces and stool to investigate the role of
+the microbiota in adenoma and carcinoma.
+
+\begin{itemize}
+\item
+  Significantly adds to the existing research using this approach by
+  expanding both the total individuals included and the scope.
+\item
+  leads to clear identifications of the challenges that lie ahead with
+  respect to adenoma
+\end{itemize}
 

### 2017-11-02 submission/manuscript.Rmd
add second discussion paragraph on driver-passenger

@@
 
-- Flynn review and how data supports an idea of early specific localized community changes that becomes global in CRC
+The driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], can be supported with this data for carcinoma but not necessarily for adenoma.
+The drasitically increased relative risk of disease when considering the CRC associated genera is highly supportive of this type of process.
+In a driver-passenger scenario it is possible that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring.
+This would account for the observation that there is no synergistic incrase in relative risk when accounting for either the total number or increasing abundance of these genera.
+The initial establishment of the driver within the system is also dependent on the community that is present and this could explain why adding the community context to our models increases the prediction.
+Although the driver-passenger model fits well with the transtion of adenoma to carcinoma for our carcinoma observations it does not fit well with our observed data on control to adenoma.
 

### 2017-11-02 submission/manuscript.tex
add second discussion paragraph on driver-passenger

@@
 
+The driver-passenger model of the microbial role in CRC, as summarized
+by Flynn {[}1{]}, can be supported with this data for carcinoma but not
+necessarily for adenoma. The drasitically increased relative risk of
+disease when considering the CRC associated genera is highly supportive
+of this type of process. In a driver-passenger scenario it is possible
+that simply having the driver present or only identifying the passenger
+is a good enough proxy that the event is occuring. This would account
+for the observation that there is no synergistic incrase in relative
+risk when accounting for either the total number or increasing abundance
+of these genera. The initial establishment of the driver within the
+system is also dependent on the community that is present and this could
+explain why adding the community context to our models increases the
+prediction. Although the driver-passenger model fits well with the
+transtion of adenoma to carcinoma for our carcinoma observations it does
+not fit well with our observed data on control to adenoma.
+
 \begin{itemize}
-\item
-  Flynn review and how data supports an idea of early specific localized
-  community changes that becomes global in CRC
 \item
@@
 were identified through the review articles written by Keku, et al. and
-Vogtmann, et al. {[}1,2{]}. All studies were included that used tissue
+Vogtmann, et al. {[}2,3{]}. All studies were included that used tissue
 or feces as their sample source for 16S rRNA gene sequencing analysis.
@@
 to share. After this filtering step the following studies remained: Ahn
-{[}3{]}, Baxter {[}4{]}, Brim {[}5{]}, Burns {[}6{]}, Chen {[}7{]},
-Dejea {[}8{]}, Flemer {[}9{]}, Geng {[}10{]}, Hale {[}11{]}, Kostic
-{[}12{]}, Lu {[}13{]}, Sanapareddy {[}14{]}, Wang {[}15{]}, Weir
-{[}16{]}, and Zeller {[}17{]}. The Zackular {[}18{]} study was not
+{[}4{]}, Baxter {[}5{]}, Brim {[}6{]}, Burns {[}7{]}, Chen {[}8{]},
+Dejea {[}9{]}, Flemer {[}10{]}, Geng {[}11{]}, Hale {[}12{]}, Kostic
+{[}13{]}, Lu {[}14{]}, Sanapareddy {[}15{]}, Wang {[}16{]}, Weir
+{[}17{]}, and Zeller {[}18{]}. The Zackular {[}19{]} study was not
 included becasue the 90 individuals analyzed within the study are
@@
 have sequences and meta data on the SRA one study had the data stored on
-DBGap {[}3{]} and four studies the data was obtained directly from the
-authors {[}9,11,14,16{]}. Each study was processed using the mothur
-(v1.39.3) software program {[}19{]}. Where possible quality filtering
+DBGap {[}4{]} and four studies the data was obtained directly from the
+authors {[}10,12,15,17{]}. Each study was processed using the mothur
+(v1.39.3) software program {[}20{]}. Where possible quality filtering
 utilized the default methods used in mothur for either 454 or Illumina
@@
 author stated quality cut-offs were used instead. Chimeras were
-identifed and removed using the VSEARCH {[}20{]} program and \emph{de
-novo} OTU clustering at 97\% similarity using the OptiClust algorithm
-{[}21{]} was utilized.
+identifed and removed using the VSEARCH {[}{\textbf{???}}{]} program and
+\emph{de novo} OTU clustering at 97\% similarity using the OptiClust
+algorithm {[}21{]} was utilized.
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-flynn_metabolic_2016}{}
+1. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
+Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
+
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-1. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
+2. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
 microbiota and colorectal cancer. American Journal of Physiology -
@@
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-2. Vogtmann E, Goedert JJ. Epidemiologic studies of the human microbiome
+3. Vogtmann E, Goedert JJ. Epidemiologic studies of the human microbiome
 and cancer. British Journal of Cancer {[}Internet{]}. 2016 {[}cited 2017
@@
 \hypertarget{ref-ahn_human_2013}{}
-3. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
+4. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
 microbiome and risk for colorectal cancer. Journal of the National
@@
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-4. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+5. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
 improves the sensitivity of fecal immunochemical test for detecting
@@
 \hypertarget{ref-brim_microbiome_2013}{}
-5. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
+6. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
 al. Microbiome analysis of stool samples from African Americans with
@@
 \hypertarget{ref-burns_virulence_2015}{}
-6. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
+7. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
 are a signature of the microbiome in the colorectal tumor
@@
 \hypertarget{ref-chen_human_2012}{}
-7. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
+8. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
 mucosa-associated microbiota in patients with colorectal cancer. PloS
@@
 \hypertarget{ref-dejea_microbiota_2014}{}
-8. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+9. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
 Rossetti BJ, et al. Microbiota organization is a distinct feature of
@@
 \hypertarget{ref-flemer_tumour-associated_2017}{}
-9. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
+10. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
 al. Tumour-associated and non-tumour-associated microbiota in colorectal
@@
 \hypertarget{ref-geng_diversified_2013}{}
-10. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
+11. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
 human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
@@
 \hypertarget{ref-hale_shifts_2017}{}
-11. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
+12. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
 Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
@@
 \hypertarget{ref-kostic_genomic_2012}{}
-12. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
+13. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
 al. Genomic analysis identifies association of Fusobacterium with
@@
 \hypertarget{ref-lu_mucosal_2016}{}
-13. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
+14. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
 adherent bacterial dysbiosis in patients with colorectal adenomas.
@@
 \hypertarget{ref-sanapareddy_increased_2012}{}
-14. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
+15. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
 et al. Increased rectal microbial richness is associated with the
@@
 \hypertarget{ref-wang_structural_2012}{}
-15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
+16. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
 segregation of gut microbiota between colorectal cancer patients and
@@
 \hypertarget{ref-weir_stool_2013}{}
-16. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
+17. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
 Stool microbiome and metabolome differences between colorectal cancer
@@
 \hypertarget{ref-zeller_potential_2014}{}
-17. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+18. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
 Potential of fecal microbiota for early-stage detection of colorectal
@@
 \hypertarget{ref-zackular_human_2014}{}
-18. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+19. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-19. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+20. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, et al. Introducing mothur: Open-Source, Platform-Independent,
@@
 \url{http://aem.asm.org/cgi/content/abstract/75/23/7537}
-
-\hypertarget{ref-rognes_vsearch_2016}{}
-20. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
-versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.
 

### 2017-11-02 submission/manuscript.Rmd
add thrid paragraph

@@
 The initial establishment of the driver within the system is also dependent on the community that is present and this could explain why adding the community context to our models increases the prediction.
-Although the driver-passenger model fits well with the transtion of adenoma to carcinoma for our carcinoma observations it does not fit well with our observed data on control to adenoma.
 
-- Builiding on this idea moving forward how CRC associated genera interact with the overall community rather than in isolation will be important since they do not increase the RR of polyps
-  - suggesting localized changes to community may be more important
+Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma.
+In contrast, the adenoma observations do not fit well with this model and suggests that the transition from control to adenoma do not fit this framework.
+The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle.
+Although there may be localized changes that do depend on the driver-passenger model, our observations show that they do not involve the CRC associated genera.
+It is possible to hypothesize that at early stages of the diease how the host interacts to these subtle changes could be the catlyst that causes adenoma formation.
+Subsequent transition to carcinoma could then fit into the proposed driver-passenger model framework.
 

### 2017-11-02 submission/manuscript.tex
add thrid paragraph

@@
 explain why adding the community context to our models increases the
-prediction. Although the driver-passenger model fits well with the
-transtion of adenoma to carcinoma for our carcinoma observations it does
-not fit well with our observed data on control to adenoma.
+prediction.
+
+Our carcinoma observations fit the driver-passenger model and support
+this concept within the framework of the transition from adenoma to
+carcinoma. In contrast, the adenoma observations do not fit well with
+this model and suggests that the transition from control to adenoma do
+not fit this framework. The stepwise decrease in diversity suggests that
+the adenoma community is not normal but this change is subtle. Although
+there may be localized changes that do depend on the driver-passenger
+model, our observations show that they do not involve the CRC associated
+genera. It is possible to hypothesize that at early stages of the diease
+how the host interacts to these subtle changes could be the catlyst that
+causes adenoma formation. Subsequent transition to carcinoma could then
+fit into the proposed driver-passenger model framework.
 
 \begin{itemize}
-\item
-  Builiding on this idea moving forward how CRC associated genera
-  interact with the overall community rather than in isolation will be
-  important since they do not increase the RR of polyps
-\item
-  suggesting localized changes to community may be more important
+\tightlist
 \item

### 2017-11-02 submission/manuscript.Rmd
add last paragraph

@@
 
-- In closing to get at these more difficult questions the Power calculations and effect size in future studies need to be accounted for especially for adenoma where the predicted effect sizes are going to be small
+Although there are still questions that need to be answered for the microbiota and carcinoma a clearer framework is beginning to develop as to how this occurs.
+The role of the microbiota in adenoma is still not clear and part of the reason may be because many studies are not powered effectely to observe the small changes reported here.
+It is realistic to suspect that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control.
+Most of the studies have sufficient power to detect these changes.
+In contrast, our data suggests that the adenoma effect size is relatively small.
+None of the studies analyzed  were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences occur for adenoma.
+Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma genesis.
 

### 2017-11-02 submission/manuscript.tex
add last paragraph

@@
 
-\begin{itemize}
-\tightlist
-\item
-  In closing to get at these more difficult questions the Power
-  calculations and effect size in future studies need to be accounted
-  for especially for adenoma where the predicted effect sizes are going
-  to be small
-\end{itemize}
+Although there are still questions that need to be answered for the
+microbiota and carcinoma a clearer framework is beginning to develop as
+to how this occurs. The role of the microbiota in adenoma is still not
+clear and part of the reason may be because many studies are not powered
+effectely to observe the small changes reported here. It is realistic to
+suspect that many changes in carcinoma could easily result in effect
+sizes that are 30\% or more between the case and control. Most of the
+studies have sufficient power to detect these changes. In contrast, our
+data suggests that the adenoma effect size is relatively small. None of
+the studies analyzed were properly powered to detect a 10\% or lower
+change between case and controls and this may well be the range in which
+differences occur for adenoma. Future studies investigating adenoma and
+the microbiota need to take these factors into consideration if we are
+to work out the role of the microbiota in adenoma genesis.
 
@@
 author stated quality cut-offs were used instead. Chimeras were
-identifed and removed using the VSEARCH {[}{\textbf{???}}{]} program and
-\emph{de novo} OTU clustering at 97\% similarity using the OptiClust
-algorithm {[}21{]} was utilized.
+identifed and removed using the VSEARCH {[}21{]} program and \emph{de
+novo} OTU clustering at 97\% similarity using the OptiClust algorithm
+{[}22{]} was utilized.
 
@@
 
+\hypertarget{ref-rognes_vsearch_2016}{}
+21. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
+versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.
+
 \hypertarget{ref-westcott_opticlust_2017}{}
-21. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
+22. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
 Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.

### 2017-11-02 submission/manuscript.Rmd
change discussion grammar and add conclusion

@@
 In a driver-passenger scenario it is possible that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring.
-This would account for the observation that there is no synergistic incrase in relative risk when accounting for either the total number or increasing abundance of these genera.
-The initial establishment of the driver within the system is also dependent on the community that is present and this could explain why adding the community context to our models increases the prediction.
+This would account for the observation that there is no synergistic increase in relative risk when accounting for either the total number or increasing abundance of these genera.
+The initial establishment of the driver within the system is also dependent on the community that is present and this is supported by the observation that when adding the community context to our models along with the CRC associated genera the model AUC increases.
 
@@
 Although there may be localized changes that do depend on the driver-passenger model, our observations show that they do not involve the CRC associated genera.
-It is possible to hypothesize that at early stages of the diease how the host interacts to these subtle changes could be the catlyst that causes adenoma formation.
+It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that causes adenoma formation.
 Subsequent transition to carcinoma could then fit into the proposed driver-passenger model framework.
@@
 It is realistic to suspect that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control.
-Most of the studies have sufficient power to detect these changes.
+Most of the studies analyzed have sufficient power to detect these changes.
 In contrast, our data suggests that the adenoma effect size is relatively small.
-None of the studies analyzed  were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences occur for adenoma.
+None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences occur in adenoma.
 Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma genesis.
@@
 
--aggregate studies done with feces and stool to investigate the role of the microbiota in adenoma and carcinoma.
-
-- Significantly adds to the existing research using this approach by expanding both the total individuals included and the scope.
-
-- leads to clear identifications of the challenges that lie ahead with respect to adenoma
+By aggregating together a large collection of studies from both feces and tissue we are able to provide information in support of the driver-passenger model in the context of carcinoma.
+However, within the context of adenoma it is less clear that this relationship exists.
+These observations highlight the importance of power and sample number considerations when considering investigations into the microbiota and adenoma due to the subtle changes in community.
+This study helps to identify the problems that have been solved and the challenges that lie ahead in the invesitgation of colorectal cancer and the microbiota.
 

### 2017-11-02 submission/manuscript.tex
change discussion grammar and add conclusion

@@
 is a good enough proxy that the event is occuring. This would account
-for the observation that there is no synergistic incrase in relative
+for the observation that there is no synergistic increase in relative
 risk when accounting for either the total number or increasing abundance
 of these genera. The initial establishment of the driver within the
-system is also dependent on the community that is present and this could
-explain why adding the community context to our models increases the
-prediction.
+system is also dependent on the community that is present and this is
+supported by the observation that when adding the community context to
+our models along with the CRC associated genera the model AUC increases.
 
@@
 model, our observations show that they do not involve the CRC associated
-genera. It is possible to hypothesize that at early stages of the diease
-how the host interacts to these subtle changes could be the catlyst that
-causes adenoma formation. Subsequent transition to carcinoma could then
-fit into the proposed driver-passenger model framework.
+genera. It is possible to hypothesize that at early stages of the
+diease, how the host interacts to these subtle changes could be the
+catlyst that causes adenoma formation. Subsequent transition to
+carcinoma could then fit into the proposed driver-passenger model
+framework.
 
@@
 sizes that are 30\% or more between the case and control. Most of the
-studies have sufficient power to detect these changes. In contrast, our
-data suggests that the adenoma effect size is relatively small. None of
-the studies analyzed were properly powered to detect a 10\% or lower
-change between case and controls and this may well be the range in which
-differences occur for adenoma. Future studies investigating adenoma and
-the microbiota need to take these factors into consideration if we are
-to work out the role of the microbiota in adenoma genesis.
+studies analyzed have sufficient power to detect these changes. In
+contrast, our data suggests that the adenoma effect size is relatively
+small. None of the studies analyzed were properly powered to detect a
+10\% or lower change between case and controls and this may well be the
+range in which differences occur in adenoma. Future studies
+investigating adenoma and the microbiota need to take these factors into
+consideration if we are to work out the role of the microbiota in
+adenoma genesis.
 
@@
 
--aggregate studies done with feces and stool to investigate the role of
-the microbiota in adenoma and carcinoma.
-
-\begin{itemize}
-\item
-  Significantly adds to the existing research using this approach by
-  expanding both the total individuals included and the scope.
-\item
-  leads to clear identifications of the challenges that lie ahead with
-  respect to adenoma
-\end{itemize}
+By aggregating together a large collection of studies from both feces
+and tissue we are able to provide information in support of the
+driver-passenger model in the context of carcinoma. However, within the
+context of adenoma it is less clear that this relationship exists. These
+observations highlight the importance of power and sample number
+considerations when considering investigations into the microbiota and
+adenoma due to the subtle changes in community. This study helps to
+identify the problems that have been solved and the challenges that lie
+ahead in the invesitgation of colorectal cancer and the microbiota.
 

### 2017-11-03 submission/manuscript.Rmd
add intro w/o citations

@@
 ## Background
+
+Colorectal cancer (CRC) is a growing world wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis.
+Numerous studies have shown the importance of both individual microbes () and the overall community () in polyp formation in mouse models.
+There has also been numerous case control studies investigating the microbiota in both adenoma and carcinoma.
+Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets ().
+Many of the studies along with the current meta-analysis focus on identifying biomarkers or individual microbes but do not critically investigate the community role in the disease.
+
+Using both fecal and tissue samples totalling over 2100 total individuals across 14 studies within our data analysis we exapnd both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
+We first assessed the diversity of controls, adenoma, and carcinoma individuals and tested whether they change and if it results in an increased relative risk of adenoma or carcinoma.
+Next, we assessed how this relative risk compared to CRC associated genera for both adenoma and carcinoma.
+Third, using Random Forest models we assessed whether the community context can increase the classification model area under the curve (AUC).
+Finally, we examine whether the studies that were used were adequately powered and if not what effect size they were powered for.
+
+Our analysis found a continuous decreae in Shannon diversity from control to adenoma to carcinoma and a significantly increased relative risk for carcinoma with lower diversity.
+Using the CRC associated genera only this relative risk was higher than Shannon diversity.
+However, adding the community context in which these CRC associated genera are present increases prediction models AUC.
+Although we analyze a data set with a large number of total individuals each individual study was underpowered for effect size differences of 10% or below between the case and control.
 

### 2017-11-03 submission/manuscript.tex
add intro w/o citations

@@
 \subsection{Background}\label{background}
+
+Colorectal cancer (CRC) is a growing world wide health problem in which
+the microbiota has been purported to play an active role in disease
+pathogenesis. Numerous studies have shown the importance of both
+individual microbes () and the overall community () in polyp formation
+in mouse models. There has also been numerous case control studies
+investigating the microbiota in both adenoma and carcinoma. Recently, a
+meta-analysis was published investigating whether specific biomarkers
+could be consistently identified using multiple data sets (). Many of
+the studies along with the current meta-analysis focus on identifying
+biomarkers or individual microbes but do not critically investigate the
+community role in the disease.
+
+Using both fecal and tissue samples totalling over 2100 total
+individuals across 14 studies within our data analysis we exapnd both
+the breadth and scope of the previous meta-analysis to investigate
+whether the bacterial community is an important risk factor for both
+adenoma and carcinoma. We first assessed the diversity of controls,
+adenoma, and carcinoma individuals and tested whether they change and if
+it results in an increased relative risk of adenoma or carcinoma. Next,
+we assessed how this relative risk compared to CRC associated genera for
+both adenoma and carcinoma. Third, using Random Forest models we
+assessed whether the community context can increase the classification
+model area under the curve (AUC). Finally, we examine whether the
+studies that were used were adequately powered and if not what effect
+size they were powered for.
+
+Our analysis found a continuous decreae in Shannon diversity from
+control to adenoma to carcinoma and a significantly increased relative
+risk for carcinoma with lower diversity. Using the CRC associated genera
+only this relative risk was higher than Shannon diversity. However,
+adding the community context in which these CRC associated genera are
+present increases prediction models AUC. Although we analyze a data set
+with a large number of total individuals each individual study was
+underpowered for effect size differences of 10\% or below between the
+case and control.
 

### 2017-11-03 submission/manuscript.Rmd
add refs to background

@@
 
-Colorectal cancer (CRC) is a growing world wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis.
-Numerous studies have shown the importance of both individual microbes () and the overall community () in polyp formation in mouse models.
+Colorectal cancer (CRC) is a growing world wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016].
+Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in polyp formation in mouse models.
 There has also been numerous case control studies investigating the microbiota in both adenoma and carcinoma.
-Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets ().
+Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
 Many of the studies along with the current meta-analysis focus on identifying biomarkers or individual microbes but do not critically investigate the community role in the disease.
 
-Using both fecal and tissue samples totalling over 2100 total individuals across 14 studies within our data analysis we exapnd both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
+Using both fecal and tissue samples totalling over 2100 total individuals across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] within our data analysis we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
 We first assessed the diversity of controls, adenoma, and carcinoma individuals and tested whether they change and if it results in an increased relative risk of adenoma or carcinoma.

### 2017-11-03 submission/manuscript.tex
add refs to background

@@
 
-Colorectal cancer (CRC) is a growing world wide health problem in which
-the microbiota has been purported to play an active role in disease
-pathogenesis. Numerous studies have shown the importance of both
-individual microbes () and the overall community () in polyp formation
-in mouse models. There has also been numerous case control studies
-investigating the microbiota in both adenoma and carcinoma. Recently, a
-meta-analysis was published investigating whether specific biomarkers
-could be consistently identified using multiple data sets (). Many of
-the studies along with the current meta-analysis focus on identifying
-biomarkers or individual microbes but do not critically investigate the
-community role in the disease.
+Colorectal cancer (CRC) is a growing world wide health problem {[}1{]}
+in which the microbiota has been purported to play an active role in
+disease pathogenesis {[}2{]}. Numerous studies have shown the importance
+of both individual microbes {[}3--7{]} and the overall community
+{[}8--10{]} in polyp formation in mouse models. There has also been
+numerous case control studies investigating the microbiota in both
+adenoma and carcinoma. Recently, a meta-analysis was published
+investigating whether specific biomarkers could be consistently
+identified using multiple data sets {[}11{]}. Many of the studies along
+with the current meta-analysis focus on identifying biomarkers or
+individual microbes but do not critically investigate the community role
+in the disease.
 
 Using both fecal and tissue samples totalling over 2100 total
-individuals across 14 studies within our data analysis we exapnd both
-the breadth and scope of the previous meta-analysis to investigate
-whether the bacterial community is an important risk factor for both
-adenoma and carcinoma. We first assessed the diversity of controls,
-adenoma, and carcinoma individuals and tested whether they change and if
-it results in an increased relative risk of adenoma or carcinoma. Next,
-we assessed how this relative risk compared to CRC associated genera for
-both adenoma and carcinoma. Third, using Random Forest models we
-assessed whether the community context can increase the classification
-model area under the curve (AUC). Finally, we examine whether the
-studies that were used were adequately powered and if not what effect
-size they were powered for.
+individuals across 14 studies {[}12--25{]} within our data analysis we
+expand both the breadth and scope of the previous meta-analysis to
+investigate whether the bacterial community is an important risk factor
+for both adenoma and carcinoma. We first assessed the diversity of
+controls, adenoma, and carcinoma individuals and tested whether they
+change and if it results in an increased relative risk of adenoma or
+carcinoma. Next, we assessed how this relative risk compared to CRC
+associated genera for both adenoma and carcinoma. Third, using Random
+Forest models we assessed whether the community context can increase the
+classification model area under the curve (AUC). Finally, we examine
+whether the studies that were used were adequately powered and if not
+what effect size they were powered for.
 
@@
 The driver-passenger model of the microbial role in CRC, as summarized
-by Flynn {[}1{]}, can be supported with this data for carcinoma but not
+by Flynn {[}2{]}, can be supported with this data for carcinoma but not
 necessarily for adenoma. The drasitically increased relative risk of
@@
 were identified through the review articles written by Keku, et al. and
-Vogtmann, et al. {[}2,3{]}. All studies were included that used tissue
+Vogtmann, et al. {[}26,27{]}. All studies were included that used tissue
 or feces as their sample source for 16S rRNA gene sequencing analysis.
@@
 to share. After this filtering step the following studies remained: Ahn
-{[}4{]}, Baxter {[}5{]}, Brim {[}6{]}, Burns {[}7{]}, Chen {[}8{]},
-Dejea {[}9{]}, Flemer {[}10{]}, Geng {[}11{]}, Hale {[}12{]}, Kostic
-{[}13{]}, Lu {[}14{]}, Sanapareddy {[}15{]}, Wang {[}16{]}, Weir
-{[}17{]}, and Zeller {[}18{]}. The Zackular {[}19{]} study was not
+{[}21{]}, Baxter {[}24{]}, Brim {[}17{]}, Burns {[}22{]}, Chen {[}14{]},
+Dejea {[}19{]}, Flemer {[}13{]}, Geng {[}25{]}, Hale {[}12{]}, Kostic
+{[}28{]}, Lu {[}16{]}, Sanapareddy {[}20{]}, Wang {[}15{]}, Weir
+{[}18{]}, and Zeller {[}23{]}. The Zackular {[}29{]} study was not
 included becasue the 90 individuals analyzed within the study are
@@
 have sequences and meta data on the SRA one study had the data stored on
-DBGap {[}4{]} and four studies the data was obtained directly from the
-authors {[}10,12,15,17{]}. Each study was processed using the mothur
-(v1.39.3) software program {[}20{]}. Where possible quality filtering
+DBGap {[}21{]} and four studies the data was obtained directly from the
+authors {[}12,13,18,20{]}. Each study was processed using the mothur
+(v1.39.3) software program {[}30{]}. Where possible quality filtering
 utilized the default methods used in mothur for either 454 or Illumina
@@
 author stated quality cut-offs were used instead. Chimeras were
-identifed and removed using the VSEARCH {[}21{]} program and \emph{de
+identifed and removed using the VSEARCH {[}31{]} program and \emph{de
 novo} OTU clustering at 97\% similarity using the OptiClust algorithm
-{[}22{]} was utilized.
+{[}32{]} was utilized.
 
@@
 \hypertarget{refs}{}
+\hypertarget{ref-siegel_cancer_2016}{}
+1. Siegel RL, Miller KD, Jemal A. Cancer statistics, 2016. CA: a cancer
+journal for clinicians. 2016;66:7--30.
+
 \hypertarget{ref-flynn_metabolic_2016}{}
-1. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
+2. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
 Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
 
-\hypertarget{ref-keku_gastrointestinal_2015}{}
-2. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
-microbiota and colorectal cancer. American Journal of Physiology -
-Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
-Oct 30{]};308:G351--63. Available from:
-\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}
+\hypertarget{ref-goodwin_polyamine_2011}{}
+3. Goodwin AC, Destefano Shields CE, Wu S, Huso DL, Wu X, Murray-Stewart
+TR, et al. Polyamine catabolism contributes to enterotoxigenic
+Bacteroides fragilis-induced colon tumorigenesis. Proceedings of the
+National Academy of Sciences of the United States of America.
+2011;108:15354--9.
 
-\hypertarget{ref-vogtmann_epidemiologic_2016}{}
-3. Vogtmann E, Goedert JJ. Epidemiologic studies of the human microbiome
-and cancer. British Journal of Cancer {[}Internet{]}. 2016 {[}cited 2017
-Oct 30{]};114:237--42. Available from:
-\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}
+\hypertarget{ref-abed_fap2_2016}{}
+4. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, et al.
+Fap2 Mediates Fusobacterium nucleatum Colorectal Adenocarcinoma
+Enrichment by Binding to Tumor-Expressed Gal-GalNAc. Cell Host \&
+Microbe. 2016;20:215--25.
 
-\hypertarget{ref-ahn_human_2013}{}
-4. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
-microbiome and risk for colorectal cancer. Journal of the National
-Cancer Institute. 2013;105:1907--11.
+\hypertarget{ref-arthur_intestinal_2012}{}
+5. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
+T-J, et al. Intestinal inflammation targets cancer-inducing activity of
+the microbiota. Science (New York, N.Y.). 2012;338:120--3.
 
-\hypertarget{ref-baxter_microbiota-based_2016}{}
-5. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
-improves the sensitivity of fecal immunochemical test for detecting
-colonic lesions. Genome Medicine. 2016;8:37.
+\hypertarget{ref-kostic_fusobacterium_2013}{}
+6. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
+et al. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
+modulates the tumor-immune microenvironment. Cell Host \& Microbe.
+2013;14:207--15.
 
-\hypertarget{ref-brim_microbiome_2013}{}
-6. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
-al. Microbiome analysis of stool samples from African Americans with
-colon polyps. PloS One. 2013;8:e81352.
+\hypertarget{ref-wu_human_2009}{}
+7. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, et al. A
+human colonic commensal promotes colon tumorigenesis via activation of T
+helper type 17 T cell responses. Nature Medicine. 2009;15:1016--22.
 
-\hypertarget{ref-burns_virulence_2015}{}
-7. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
-are a signature of the microbiome in the colorectal tumor
-microenvironment. Genome Medicine. 2015;7:55.
+\hypertarget{ref-zackular_manipulation_2016}{}
+8. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
+Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
 
-\hypertarget{ref-chen_human_2012}{}
-8. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
-mucosa-associated microbiota in patients with colorectal cancer. PloS
-One. 2012;7:e39743.
+\hypertarget{ref-zackular_gut_2013}{}
+9. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen GY,
+et al. The gut microbiome modulates colon tumorigenesis. mBio.
+2013;4:e00692--00613.
 
-\hypertarget{ref-dejea_microbiota_2014}{}
-9. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
-Rossetti BJ, et al. Microbiota organization is a distinct feature of
-proximal colorectal cancers. Proceedings of the National Academy of
-Sciences of the United States of America. 2014;111:18321--6.
+\hypertarget{ref-baxter_structure_2014}{}
+10. Baxter NT, Zackular JP, Chen GY, Schloss PD. Structure of the gut
+microbiome following colonization with human feces determines colonic
+tumor burden. Microbiome. 2014;2:20.
 
-\hypertarget{ref-flemer_tumour-associated_2017}{}
-10. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
-al. Tumour-associated and non-tumour-associated microbiota in colorectal
-cancer. Gut. 2017;66:633--43.
-
-\hypertarget{ref-geng_diversified_2013}{}
-11. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
-human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
+\hypertarget{ref-shah_leveraging_2017}{}
+11. Shah MS, DeSantis TZ, Weinmaier T, McMurdie PJ, Cope JL, Altrichter
+A, et al. Leveraging sequence-based faecal microbial community survey
+data to identify a composite biomarker for colorectal cancer. Gut. 2017;
 
@@
 
-\hypertarget{ref-kostic_genomic_2012}{}
-13. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
-al. Genomic analysis identifies association of Fusobacterium with
-colorectal carcinoma. Genome Research. 2012;22:292--8.
+\hypertarget{ref-flemer_tumour-associated_2017}{}
+13. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
+al. Tumour-associated and non-tumour-associated microbiota in colorectal
+cancer. Gut. 2017;66:633--43.
+
+\hypertarget{ref-chen_human_2012}{}
+14. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
+mucosa-associated microbiota in patients with colorectal cancer. PloS
+One. 2012;7:e39743.
+
+\hypertarget{ref-wang_structural_2012}{}
+15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
+segregation of gut microbiota between colorectal cancer patients and
+healthy volunteers. The ISME journal. 2012;6:320--9.
 
 \hypertarget{ref-lu_mucosal_2016}{}
-14. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
+16. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
 adherent bacterial dysbiosis in patients with colorectal adenomas.
@@
 
+\hypertarget{ref-brim_microbiome_2013}{}
+17. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
+al. Microbiome analysis of stool samples from African Americans with
+colon polyps. PloS One. 2013;8:e81352.
+
+\hypertarget{ref-weir_stool_2013}{}
+18. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
+Stool microbiome and metabolome differences between colorectal cancer
+patients and healthy adults. PloS One. 2013;8:e70803.
+
+\hypertarget{ref-dejea_microbiota_2014}{}
+19. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+Rossetti BJ, et al. Microbiota organization is a distinct feature of
+proximal colorectal cancers. Proceedings of the National Academy of
+Sciences of the United States of America. 2014;111:18321--6.
+
 \hypertarget{ref-sanapareddy_increased_2012}{}
-15. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
+20. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
 et al. Increased rectal microbial richness is associated with the
@@
 
-\hypertarget{ref-wang_structural_2012}{}
-16. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
-segregation of gut microbiota between colorectal cancer patients and
-healthy volunteers. The ISME journal. 2012;6:320--9.
+\hypertarget{ref-ahn_human_2013}{}
+21. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
+microbiome and risk for colorectal cancer. Journal of the National
+Cancer Institute. 2013;105:1907--11.
 
-\hypertarget{ref-weir_stool_2013}{}
-17. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
-Stool microbiome and metabolome differences between colorectal cancer
-patients and healthy adults. PloS One. 2013;8:e70803.
+\hypertarget{ref-burns_virulence_2015}{}
+22. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
+are a signature of the microbiome in the colorectal tumor
+microenvironment. Genome Medicine. 2015;7:55.
 
 \hypertarget{ref-zeller_potential_2014}{}
-18. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+23. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
 Potential of fecal microbiota for early-stage detection of colorectal
@@
 
+\hypertarget{ref-baxter_microbiota-based_2016}{}
+24. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+improves the sensitivity of fecal immunochemical test for detecting
+colonic lesions. Genome Medicine. 2016;8:37.
+
+\hypertarget{ref-geng_diversified_2013}{}
+25. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
+human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
+
+\hypertarget{ref-keku_gastrointestinal_2015}{}
+26. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
+microbiota and colorectal cancer. American Journal of Physiology -
+Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
+Oct 30{]};308:G351--63. Available from:
+\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}
+
+\hypertarget{ref-vogtmann_epidemiologic_2016}{}
+27. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
+microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
+{[}cited 2017 Oct 30{]};114:237--42. Available from:
+\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}
+
+\hypertarget{ref-kostic_genomic_2012}{}
+28. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
+al. Genomic analysis identifies association of Fusobacterium with
+colorectal carcinoma. Genome Research. 2012;22:292--8.
+
 \hypertarget{ref-zackular_human_2014}{}
-19. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+29. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-20. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+30. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, et al. Introducing mothur: Open-Source, Platform-Independent,
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-21. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
+31. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
 versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-22. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
+32. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
 Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.

### 2017-11-03 submission/manuscript.Rmd
update results

@@
 
+adn_select_RR_stool <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_RR_stool_composite.csv"))
+adn_select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_RR_tissue_composite.csv"))
+
 ```
@@
 
-**Background.**
+**Background.** An increasing body of literature suggests that there a role for the microbiota in colorectal cancer.
+Important players within this axis have ranged from individual microbes to the whole community.
+A recent meta-analysis investigated this but only focused on potential biomarkers.
+This study expands on this previous research and tests the hypothesis that detectable changes in the bacterial community are important both to increasing relative risk and model accuracy for adenoma and carcinoma.
+To test this hypothesis we examined both fecal and tissue samples from 2109 individuals across 14 different studies.
 
-**Results.**
+**Results.** There was a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness for fecal samples (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study and variable region.
+Lower Shannon diversity and evennes in fecal samples resulted in a significant increase in relative risk for carinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05).
+Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
 
@@
 ***Genera Previously Associated with Carcinoma Increases Relative Risk More than Alpha Diversity:***
-Both fecal and tissue samples had a significantly increased RR for carcinoma but not for adenoma [Figure 3] which was greater than either evenness or Shannon diversity [Figure 2 & 3].
+Both fecal and tissue samples had a significantly increased RR for carcinoma but not for adenoma [Figure 3].
+The relative risk for feces was greater than either evenness or Shannon diversity [Figure 2 & 3].
 The relative risk did not increase when considering the total abundance or increasing number of carcinoma associated genera [Figure 3].
@@
 This decrease may be explained by the fact that tissue samples include matched samples.
+
+Only two measures for adenoma were significant when investigating these CRC associated genera.
+The first was *Porphyromonas* for feces (RR range: `r round(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2)` - `r round(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2)`, P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and the second was the total relative abundance of all four CRC associated genera on tissue (RR range: `r round(filter(adn_select_RR_tissue, measure == "total_four")[, "ci_lb"], digits = 2)` - `r round(filter(adn_select_RR_tissue, measure == "total_four")[, "ci_ub"], digits = 2)`, P-value = `r format(filter(adn_select_RR_tissue, measure == "total_four")[, "pvalue"], digits = 3, nsmall = 3)`).
 

### 2017-11-03 submission/manuscript.tex
update results

@@
 
-\textbf{Background.}
+\textbf{Background.} An increasing body of literature suggests that
+there a role for the microbiota in colorectal cancer. Important players
+within this axis have ranged from individual microbes to the whole
+community. A recent meta-analysis investigated this but only focused on
+potential biomarkers. This study expands on this previous research and
+tests the hypothesis that detectable changes in the bacterial community
+are important both to increasing relative risk and model accuracy for
+adenoma and carcinoma. To test this hypothesis we examined both fecal
+and tissue samples from 2109 individuals across 14 different studies.
 
-\textbf{Results.}
+\textbf{Results.} There was a significant decrease from control to
+adenoma to carcinoma for both Shannon diversity and evenness for fecal
+samples (P-value \textless{} 0.05) after correcting for study and
+variable region. Lower Shannon diversity and evennes in fecal samples
+resulted in a significant increase in relative risk for carinoma
+(P-value \textless{} 0.05) but not adenoma (P-value \textgreater{}
+0.05). Previously associated colorectal cancer genera
+(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
+\emph{Porphyromonas}) followed a similar pattern with a significantly
+increased relative risk by their presence for carcinoma (P-value
+\textless{} 0.05) but not adenoma (P-value \textgreater{} 0.05) with the
+exception of \emph{Porphyromonas} (P-value \textless{} 0.05).
 
@@
 had a significantly increased RR for carcinoma but not for adenoma
-{[}Figure 3{]} which was greater than either evenness or Shannon
-diversity {[}Figure 2 \& 3{]}. The relative risk did not increase when
-considering the total abundance or increasing number of carcinoma
-associated genera {[}Figure 3{]}. The RR effect size was greater for
-stool (RR range = 1.78 - 2.64) then for tissue (RR range = 1.33 - 1.53).
-This decrease may be explained by the fact that tissue samples include
-matched samples.
+{[}Figure 3{]}. The relative risk for feces was greater than either
+evenness or Shannon diversity {[}Figure 2 \& 3{]}. The relative risk did
+not increase when considering the total abundance or increasing number
+of carcinoma associated genera {[}Figure 3{]}. The RR effect size was
+greater for stool (RR range = 1.78 - 2.64) then for tissue (RR range =
+1.33 - 1.53). This decrease may be explained by the fact that tissue
+samples include matched samples.
+
+Only two measures for adenoma were significant when investigating these
+CRC associated genera. The first was \emph{Porphyromonas} for feces (RR
+range: 1.06 - 2.29, P-value = 0.023) and the second was the total
+relative abundance of all four CRC associated genera on tissue (RR
+range: 1.21 - 3.89, P-value = 0.00933)
 

### 2017-11-03 submission/manuscript.Rmd
complete abstract

@@
 Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
+Using the whole community resulted in a higher classification model AUC for both adenoma and carcinoma for fecal and tissue samples.
+Most studies are adequately powered for large effect size differences which may be more amenable for carcinoma than for adenoma.
 
-**Conclusions.**
+**Conclusions.** This data provides support for the idea of the "driver-passenger" model in both adenoma and carinoma.
+The data is much stronger for carcinoma and this may in part be due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
 
@@
 Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma.
-In contrast, the adenoma observations do not fit well with this model and suggests that the transition from control to adenoma do not fit this framework.
+In contrast, with the present data we can only suggest that the adenoma observations might fit well with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma or specific location.
 The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle.
-Although there may be localized changes that do depend on the driver-passenger model, our observations show that they do not involve the CRC associated genera.
-It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that causes adenoma formation.
-Subsequent transition to carcinoma could then fit into the proposed driver-passenger model framework.
+Although there may be localized changes that do depend on the driver-passenger model, supported by the *Porphyromonas* finding in adenoma stool and the relative total abundance of all four CRC associated genera in tissue, there may be another process involved that ultimately exacerbates the condition from a subtle localized change to a global community one.
+It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that ultimately leads to this larger global dysfunctional community.
 
 Although there are still questions that need to be answered for the microbiota and carcinoma a clearer framework is beginning to develop as to how this occurs.
-The role of the microbiota in adenoma is still not clear and part of the reason may be because many studies are not powered effectely to observe the small changes reported here.
+The role of the microbiota in adenoma is still not clear and part of the reason may be because many studies are not powered effectively to observe the small changes reported here.
 It is realistic to suspect that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control.

### 2017-11-03 submission/manuscript.tex
complete abstract

@@
 \textless{} 0.05) but not adenoma (P-value \textgreater{} 0.05) with the
-exception of \emph{Porphyromonas} (P-value \textless{} 0.05).
+exception of \emph{Porphyromonas} (P-value \textless{} 0.05). Using the
+whole community resulted in a higher classification model AUC for both
+adenoma and carcinoma for fecal and tissue samples. Most studies are
+adequately powered for large effect size differences which may be more
+amenable for carcinoma than for adenoma.
 
-\textbf{Conclusions.}
+\textbf{Conclusions.} This data provides support for the idea of the
+``driver-passenger'' model in both adenoma and carinoma. The data is
+much stronger for carcinoma and this may in part be due to the low power
+to detect more subtle changes in the majority of studies that have been
+performed to date.
 
@@
 relative abundance of all four CRC associated genera on tissue (RR
-range: 1.21 - 3.89, P-value = 0.00933)
+range: 1.21 - 3.89, P-value = 0.00933).
 
@@
 this concept within the framework of the transition from adenoma to
-carcinoma. In contrast, the adenoma observations do not fit well with
-this model and suggests that the transition from control to adenoma do
-not fit this framework. The stepwise decrease in diversity suggests that
+carcinoma. In contrast, with the present data we can only suggest that
+the adenoma observations might fit well with this model but the changes
+that occur at this timepoint are small and possibly focal to the adenoma
+or specific location. The stepwise decrease in diversity suggests that
 the adenoma community is not normal but this change is subtle. Although
 there may be localized changes that do depend on the driver-passenger
-model, our observations show that they do not involve the CRC associated
-genera. It is possible to hypothesize that at early stages of the
+model, supported by the \emph{Porphyromonas} finding in adenoma stool
+and the relative total abundance of all four CRC associated genera in
+tissue, there may be another process involved that ultimately
+exacerbates the condition from a subtle localized change to a global
+community one. It is possible to hypothesize that at early stages of the
 diease, how the host interacts to these subtle changes could be the
-catlyst that causes adenoma formation. Subsequent transition to
-carcinoma could then fit into the proposed driver-passenger model
-framework.
+catlyst that ultimately leads to this larger global dysfunctional
+community.
 
@@
 clear and part of the reason may be because many studies are not powered
-effectely to observe the small changes reported here. It is realistic to
-suspect that many changes in carcinoma could easily result in effect
+effectively to observe the small changes reported here. It is realistic
+to suspect that many changes in carcinoma could easily result in effect
 sizes that are 30\% or more between the case and control. Most of the

### 2017-11-03 submission/manuscript.Rmd
amend abstract and discussion

@@
 
-loadLibs(c("tidyverse", "scales"))
+loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown"))
 
@@
 
-**Conclusions.** This data provides support for the idea of the "driver-passenger" model in both adenoma and carinoma.
+**Conclusions.** This data provides support for the idea of the "driver-passenger" model in carinoma but only partially for adenoma.
 The data is much stronger for carcinoma and this may in part be due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
@@
 Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma.
-In contrast, with the present data we can only suggest that the adenoma observations might fit well with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma or specific location.
+In contrast, with the present data we can only suggest that the adenoma observations might fit with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma or specific location.
 The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle.
-Although there may be localized changes that do depend on the driver-passenger model, supported by the *Porphyromonas* finding in adenoma stool and the relative total abundance of all four CRC associated genera in tissue, there may be another process involved that ultimately exacerbates the condition from a subtle localized change to a global community one.
+Although there may be localized changes that do depend on the driver-passenger model, supported by the *Porphyromonas* finding in adenoma stool and the relative total abundance of all four CRC associated genera in tissue [Figure 3], there may be another process involved that ultimately exacerbates the condition from a subtle localized change to a global community one.
+The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case.
 It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that ultimately leads to this larger global dysfunctional community.
@@
 
+studies <- c("Ahn", "Baxter", "Brim", "Burns", "Chen", "Dejea", "Flemer", "Geng", "Hale", "Lu", "Sanapareddy", "Wang", "Weir", "Zeller")
+
+kable(table1, format = "markdown", align = 'c', padding = 2)
 ```

### 2017-11-03 submission/manuscript.tex
amend abstract and discussion

@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{longtable,booktabs}
 \usepackage{graphicx,grffile}
@@
 \textbf{Conclusions.} This data provides support for the idea of the
-``driver-passenger'' model in both adenoma and carinoma. The data is
-much stronger for carcinoma and this may in part be due to the low power
-to detect more subtle changes in the majority of studies that have been
-performed to date.
+``driver-passenger'' model in carinoma but only partially for adenoma.
+The data is much stronger for carcinoma and this may in part be due to
+the low power to detect more subtle changes in the majority of studies
+that have been performed to date.
 
@@
 carcinoma. In contrast, with the present data we can only suggest that
-the adenoma observations might fit well with this model but the changes
-that occur at this timepoint are small and possibly focal to the adenoma
-or specific location. The stepwise decrease in diversity suggests that
-the adenoma community is not normal but this change is subtle. Although
+the adenoma observations might fit with this model but the changes that
+occur at this timepoint are small and possibly focal to the adenoma or
+specific location. The stepwise decrease in diversity suggests that the
+adenoma community is not normal but this change is subtle. Although
 there may be localized changes that do depend on the driver-passenger
@@
 and the relative total abundance of all four CRC associated genera in
-tissue, there may be another process involved that ultimately
-exacerbates the condition from a subtle localized change to a global
-community one. It is possible to hypothesize that at early stages of the
-diease, how the host interacts to these subtle changes could be the
+tissue {[}Figure 3{]}, there may be another process involved that
+ultimately exacerbates the condition from a subtle localized change to a
+global community one. The poor performance of the Random Forest models
+for classifying adenoma based only on the microbiota would suggest that
+this is the case. It is possible to hypothesize that at early stages of
+the diease, how the host interacts to these subtle changes could be the
 catlyst that ultimately leads to this larger global dysfunctional
@@
 
+\begin{longtable}[]{@{}cccc@{}}
+\toprule
+country & year & cases & population\tabularnewline
+\midrule
+\endhead
+Afghanistan & 1999 & 745 & 19987071\tabularnewline
+Afghanistan & 2000 & 2666 & 20595360\tabularnewline
+Brazil & 1999 & 37737 & 172006362\tabularnewline
+Brazil & 2000 & 80488 & 174504898\tabularnewline
+China & 1999 & 212258 & 1272915272\tabularnewline
+China & 2000 & 213766 & 1280428583\tabularnewline
+\bottomrule
+\end{longtable}
+
 \newpage

### 2017-11-03 submission/manuscript.Rmd
add data table

@@
 
-loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown"))
+loadLibs(c("tidyverse", "scales", "knitr", "rmarkdown", "stringr"))
 
@@
 ```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+#Read in meta data files used
+studies <- c("Ahn", "Baxter", "Brim", "Burns", "Chen", "Dejea", "Flemer", "Geng", "Hale", "Lu", "Sana", "Wang", "Weir", "Zeller")
+
+study_metadata <- sapply(studies, function(x) 
+  read_tsv(paste("../data/process/", str_to_lower(x), "/", str_to_lower(x), ".metadata", sep = "")), simplify = F)
 
@@
 
+\footnotesize
+
 ```{r table1, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
+sample_type <- c("Stool", "Stool", "Stool", "Tissue", "Stool + Tissue", "Tissue", "Stool + Tissue", "Tissue", "Stool", "Tissue", "Tissue", "Stool", "Stool", "Stool")
+
+individuals <- c("control, carcinoma", "control, adenoma, carcinoma", "control, adenoma", "carcinoma", "control, carcinoma", "adenoma, carcinoma", "control, adenoma, carcinoma", "carcinoma", "control, adenoma, carcinoma", "control, adenoma", "control, carcinoma", "control, carcinoma", "control, carcinoma", "control, adenoma, carcinoma")
+
+data_obtained <- c("DBGap", "SRA", "SRA", "SRA", "SRA", "SRA", "Author", "SRA", "Author", "SRA", "Author", 
+                   "SRA", "Author", "SRA")
+
+final_n <- c(length(unique(study_metadata[["Ahn"]]$sample)), length(unique(study_metadata[["Baxter"]]$sample)), length(unique(study_metadata[["Brim"]]$sample)), length(rownames(filter(study_metadata[["Burns"]], disease == "cancer"))), length(rownames(filter(study_metadata[["Chen"]], sample_type != "swab"))), length(unique(study_metadata[["Dejea"]]$patient_id)), length(unique(study_metadata[["Flemer"]]$sample)), length(rownames(filter(study_metadata[["Geng"]], disease == "cancer"))), length(unique(study_metadata[["Hale"]]$sample)), round((length(rownames(filter(study_metadata[["Lu"]], disease == "control"))) - length(rownames(filter(study_metadata[["Lu"]], disease == "polyp"))))/2) + length(rownames(filter(study_metadata[["Lu"]], matched == "n"))) + length(rownames(filter(study_metadata[["Lu"]], disease == "polyp"))) + 1, length(unique(study_metadata[["Sana"]]$sample)), length(unique(study_metadata[["Wang"]]$sample)), length(unique(study_metadata[["Weir"]]$sample)), length(unique(study_metadata[["Zeller"]]$sample)))
+
+table1 <- cbind(studies, sample_type, individuals, data_obtained, final_n)
+
+colnames(table1) <- c("Study", "Sample Type", "Individuals", "Data Stored", "mothur Processed n")
+
 kable(table1, format = "markdown", align = 'c', padding = 2)
@@
 
+\normalsize
 \newpage

### 2017-11-03 submission/manuscript.tex
add data table

@@
 
-\begin{longtable}[]{@{}cccc@{}}
+\footnotesize
+
+\begin{longtable}[]{@{}ccccc@{}}
 \toprule
-country & year & cases & population\tabularnewline
+Study & Sample Type & Individuals & Data Stored & mothur Processed
+n\tabularnewline
 \midrule
 \endhead
-Afghanistan & 1999 & 745 & 19987071\tabularnewline
-Afghanistan & 2000 & 2666 & 20595360\tabularnewline
-Brazil & 1999 & 37737 & 172006362\tabularnewline
-Brazil & 2000 & 80488 & 174504898\tabularnewline
-China & 1999 & 212258 & 1272915272\tabularnewline
-China & 2000 & 213766 & 1280428583\tabularnewline
+Ahn & Stool & control, carcinoma & DBGap & 240\tabularnewline
+Baxter & Stool & control, adenoma, carcinoma & SRA & 490\tabularnewline
+Brim & Stool & control, adenoma & SRA & 12\tabularnewline
+Burns & Tissue & carcinoma & SRA & 43\tabularnewline
+Chen & Stool + Tissue & control, carcinoma & SRA & 100\tabularnewline
+Dejea & Tissue & adenoma, carcinoma & SRA & 25\tabularnewline
+Flemer & Stool + Tissue & control, adenoma, carcinoma & Author &
+314\tabularnewline
+Geng & Tissue & carcinoma & SRA & 8\tabularnewline
+Hale & Stool & control, adenoma, carcinoma & Author & 770\tabularnewline
+Lu & Tissue & control, adenoma & SRA & 67\tabularnewline
+Sanapareddy & Tissue & control, carcinoma & Author & 71\tabularnewline
+Wang & Stool & control, carcinoma & SRA & 102\tabularnewline
+Weir & Stool & control, carcinoma & Author & 15\tabularnewline
+Zeller & Stool & control, adenoma, carcinoma & SRA & 224\tabularnewline
 \bottomrule

### 2017-11-03 code/make_study_count_table.R
add file read outs

@@ -9,27 +9,57 @@ source('code/functions.R')
 # Load needed libraries
 loadLibs(c("tidyverse"))
 
-# Read in needed data tables
-
+# Read in adenoma stool files
 adn_stool <- read_csv("data/process/tables/alpha_adn_group_counts_summary.csv") %>% 
   filter(measure == "shannon") %>% 
   mutate(control = high_N + low_N, adenoma = high_Y + low_Y) %>% 
   select(study, control, adenoma)
-  
-  
+
+# Read in the carcinoma stool files
 crc_stool <- read_csv("data/process/tables/alpha_group_counts_summary.csv") %>% 
   filter(measure == "shannon") %>% 
   mutate(control = high_N + low_N, cancer = high_Y + low_Y) %>% 
   select(study, control, cancer)
 
+# Combine and aggregate the total n together
+combine_stool <- adn_stool %>% full_join(crc_stool, by = "study") %>% 
+  mutate(final_control = ifelse(is.na(control.x), invisible(control.y), invisible(control.x))) %>% 
+  select(study, final_control, adenoma, cancer) %>% 
+  rename(control = final_control)
+
+# Read in the adenoma tissue files
 adn_tissue <- read.csv("data/process/tables/alpha_adn_group_counts_tissue_summary.csv") %>% 
-  filter(measure == "shannon")
+  filter(measure == "shannon") %>% 
+  mutate(control = high_N + low_N, adenoma = high_Y + low_Y) %>% 
+  select(study, control, adenoma)
 
+# Read in the crc matched tissue files
 crc_tissue_matched <- read_csv("data/process/tables/alpha_group_counts_matched_tissue_summary.csv") %>% 
-  filter(measure == "shannon")
+  filter(measure == "shannon") %>% 
+  mutate(control = high_N + low_N, cancer = high_Y + low_Y) %>% 
+  select(study, control, cancer)
 
+# Read in the crc unmatched tissue files
 crc_tissue_unmatched <- read_csv("data/process/tables/alpha_group_counts_tissue_summary.csv") %>% 
-  filter(measure == "shannon")
+  filter(measure == "shannon") %>% 
+  mutate(control = high_N + low_N, cancer = high_Y + low_Y) %>% 
+  select(study, control, cancer)
+
+# combine all the tissue samples together into a single file and aggregate
+combine_tissue <- adn_tissue %>% full_join(crc_tissue_unmatched, by = "study") %>% 
[…]

### 2017-11-04 submission/manuscript.Rmd
add study n tables

--- 2017-11-03 submission/manuscript.Rmd
@@
 ```{r rdata_tables, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
+
+#Read study data files
+stool_n <- read_csv("../data/process/tables/stool_study_n_analyzed.csv")
+tissue_n <- read_csv("../data/process/tables/tissue_study_n_analyzed.csv")
 
@@
 
-**Table 1:**
+**Table 1: Studies with Stool Samples Included in the Analysis**
+
+\footnotesize
 
@@
 
-studies <- c("Ahn", "Baxter", "Brim", "Burns", "Chen", "Dejea", "Flemer", "Geng", "Hale", "Lu", "Sanapareddy", "Wang", "Weir", "Zeller")
+stool_n <- stool_n %>% arrange(study)
+
+studies <- c("Ahn", "Baxter", "Brim", "Flemer", "Hale", "Wang", "Weir", "Zeller")
+
+data_obtained <- c("DBGap", "SRA", "SRA", "Author", "Author", "SRA", "Author", "SRA")
+
+variable_region <- c("V3-4", "V4", "V1-3", "V3-4", "V3-5", "V3", "V4", "V4")
+
+con_n <- gsub("NA", 0, str_replace_na(stool_n$control))
+adn_n <- gsub("NA", 0, str_replace_na(stool_n$adenoma))
+crc_n <- gsub("NA", 0, str_replace_na(stool_n$cancer))
+
+table1 <- cbind(studies, data_obtained, variable_region, con_n, adn_n, crc_n)
+
+colnames(table1) <- c("Study", "Data Stored", "16S Region", "Controls", "Adenoma", "Carcinoma")
 
@@
 
+\normalsize
+\newpage
+
+**Table 2: Studies with Tissue Samples Included in the Analysis**
+
+\footnotesize
+
+```{r table2, echo=FALSE, message=FALSE, warning=FALSE}
+
+tissue_n <- tissue_n %>% arrange(study)
+
+studies <- c("Burns", "Chen", "Dejea", "Flemer", "Geng", "Lu", "Sanapareddy")
+
+data_obtained <- c("SRA", "SRA", "SRA", "Author", "SRA", "SRA", "Author")
+
+variable_region <- c("V5-6", "V1-V3", "V3-5", "V3-4", "V1-2", "V3-4", "V1-2")
+
+con_n <- tissue_n$control
+adn_n <- gsub("NA", 0, str_replace_na(tissue_n$adenoma))
+crc_n <- gsub("NA", 0, str_replace_na(tissue_n$cancer))
+
+table2 <- cbind(studies, data_obtained, variable_region, con_n, adn_n, crc_n)
+
+colnames(table2) <- c("Study", "Data Stored", "16S Region", "Controls", "Adenoma", "Carcinoma")
+
+kable(table2, format = "markdown", align = 'c', padding = 2)
+```
+
+\normalsize
 \newpage

### 2017-11-04 submission/manuscript.tex
add study n tables

--- 2017-11-03 submission/manuscript.tex
@@
 resulted in a significant increase in relative risk for carinoma
-(P-value \textless{} 0.05) but not adenoma (P-value \textgreater{}
-0.05). Previously associated colorectal cancer genera
-(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
-\emph{Porphyromonas}) followed a similar pattern with a significantly
-increased relative risk by their presence for carcinoma (P-value
-\textless{} 0.05) but not adenoma (P-value \textgreater{} 0.05) with the
-exception of \emph{Porphyromonas} (P-value \textless{} 0.05). Using the
-whole community resulted in a higher classification model AUC for both
-adenoma and carcinoma for fecal and tissue samples. Most studies are
-adequately powered for large effect size differences which may be more
-amenable for carcinoma than for adenoma.
+(P-value \textless{} 0.05) but not adenoma (P-value check tables 0.05).
+Previously associated colorectal cancer genera (\emph{Fusobacterium},
+\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
+followed a similar pattern with a significantly increased relative risk
+by their presence for carcinoma (P-value \textless{} 0.05) but not
+adenoma (P-value \textgreater{} 0.05) with the exception of
+\emph{Porphyromonas} (P-value \textless{} 0.05). Using the whole
+community resulted in a higher classification model AUC for both adenoma
+and carcinoma for fecal and tissue samples. Most studies are adequately
+powered for large effect size differences which may be more amenable for
+carcinoma than for adenoma.
 
@@
 
-\textbf{Table 1:}
+\textbf{Table 1: Studies with Stool Samples Included in the Analysis}
 
-\begin{longtable}[]{@{}cccc@{}}
+\footnotesize
+
+\begin{longtable}[]{@{}cccccc@{}}
 \toprule
-country & year & cases & population\tabularnewline
+Study & Data Stored & 16S Region & Controls & Adenoma &
+Carcinoma\tabularnewline
 \midrule
 \endhead
-Afghanistan & 1999 & 745 & 19987071\tabularnewline
-Afghanistan & 2000 & 2666 & 20595360\tabularnewline
-Brazil & 1999 & 37737 & 172006362\tabularnewline
-Brazil & 2000 & 80488 & 174504898\tabularnewline
-China & 1999 & 212258 & 1272915272\tabularnewline
-China & 2000 & 213766 & 1280428583\tabularnewline
+Ahn & DBGap & V3-4 & 148 & 0 & 62\tabularnewline
+Baxter & SRA & V4 & 172 & 198 & 120\tabularnewline
+Brim & SRA & V1-3 & 6 & 6 & 0\tabularnewline
+Flemer & Author & V3-4 & 37 & 0 & 43\tabularnewline
+Hale & Author & V3-5 & 473 & 214 & 17\tabularnewline
+Wang & SRA & V3 & 56 & 0 & 46\tabularnewline
+Weir & Author & V4 & 4 & 0 & 7\tabularnewline
+Zeller & SRA & V4 & 50 & 37 & 41\tabularnewline
 \bottomrule
@@
 
+\normalsize
+\newpage
+
+\textbf{Table 2: Studies with Tissue Samples Included in the Analysis}
+
+\footnotesize
+
+\begin{longtable}[]{@{}cccccc@{}}
+\toprule
+Study & Data Stored & 16S Region & Controls & Adenoma &
+Carcinoma\tabularnewline
+\midrule
+\endhead
+Burns & SRA & V5-6 & 18 & 0 & 16\tabularnewline
+Chen & SRA & V1-V3 & 9 & 0 & 9\tabularnewline
+Dejea & SRA & V3-5 & 31 & 0 & 32\tabularnewline
+Flemer & Author & V3-4 & 103 & 37 & 94\tabularnewline
+Geng & SRA & V1-2 & 16 & 0 & 16\tabularnewline
+Lu & SRA & V3-4 & 20 & 20 & 0\tabularnewline
+Sanapareddy & Author & V1-2 & 38 & 0 & 33\tabularnewline
+\bottomrule
+\end{longtable}
+
+\normalsize
 \newpage

### 2017-11-04 submission/manuscript.Rmd
incorporate n into manuscript and incorporate tables into methods

@@
 This study expands on this previous research and tests the hypothesis that detectable changes in the bacterial community are important both to increasing relative risk and model accuracy for adenoma and carcinoma.
-To test this hypothesis we examined both fecal and tissue samples from 2109 individuals across 14 different studies.
+To test this hypothesis we examined both feces (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) across 14 different studies.
 
@@
 
-Our analysis found a continuous decreae in Shannon diversity from control to adenoma to carcinoma and a significantly increased relative risk for carcinoma with lower diversity.
+Our analysis found a continuous decrease in Shannon diversity from control to adenoma to carcinoma and a significantly increased relative risk for carcinoma with lower diversity.
 Using the CRC associated genera only this relative risk was higher than Shannon diversity.
@@
 In total there were 7 studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), 5 studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and 2 studies with both fecal and tissue samples (Chen and Flemer).
-The total number of individuals initially run through the sequence processing for the fecal samples was 1899 and for the tissue samples was 462.
+The total number of individuals that were analyzed after sequence processing for feces was `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` [Table 1].
+The total number of matched and unmatched tissue samples that were analyzed after sequence processing was `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` [Table 2].
 

### 2017-11-04 submission/manuscript.tex
incorporate n into manuscript and incorporate tables into methods

@@
 are important both to increasing relative risk and model accuracy for
-adenoma and carcinoma. To test this hypothesis we examined both fecal
-and tissue samples from 2109 individuals across 14 different studies.
+adenoma and carcinoma. To test this hypothesis we examined both feces
+(total individuals = 1737) and tissue (total samples = 492) across 14
+different studies.
 
@@
 
-Our analysis found a continuous decreae in Shannon diversity from
+Our analysis found a continuous decrease in Shannon diversity from
 control to adenoma to carcinoma and a significantly increased relative
@@
 and 2 studies with both fecal and tissue samples (Chen and Flemer). The
-total number of individuals initially run through the sequence
-processing for the fecal samples was 1899 and for the tissue samples was
-462.
+total number of individuals that were analyzed after sequence processing
+for feces was 1737 {[}Table 1{]}. The total number of matched and
+unmatched tissue samples that were analyzed after sequence processing
+was 492 {[}Table 2{]}.
 

### 2017-11-06 submission/manuscript.Rmd
add total numbers to background

@@
 
-Using both fecal and tissue samples totalling over 2100 total individuals across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] within our data analysis we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
+Using both fecal (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue samples (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] within our data analysis we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
 We first assessed the diversity of controls, adenoma, and carcinoma individuals and tested whether they change and if it results in an increased relative risk of adenoma or carcinoma.

### 2017-11-06 submission/manuscript.tex
add total numbers to background

@@
 
-Using both fecal and tissue samples totalling over 2100 total
-individuals across 14 studies {[}12--25{]} within our data analysis we
-expand both the breadth and scope of the previous meta-analysis to
-investigate whether the bacterial community is an important risk factor
-for both adenoma and carcinoma. We first assessed the diversity of
-controls, adenoma, and carcinoma individuals and tested whether they
-change and if it results in an increased relative risk of adenoma or
-carcinoma. Next, we assessed how this relative risk compared to CRC
-associated genera for both adenoma and carcinoma. Third, using Random
-Forest models we assessed whether the community context can increase the
-classification model area under the curve (AUC). Finally, we examine
-whether the studies that were used were adequately powered and if not
-what effect size they were powered for.
+Using both fecal (total individuals = 1737) and tissue samples (total
+samples = 492) totalling over 2229 total samples across 14 studies
+{[}12--25{]} within our data analysis we expand both the breadth and
+scope of the previous meta-analysis to investigate whether the bacterial
+community is an important risk factor for both adenoma and carcinoma. We
+first assessed the diversity of controls, adenoma, and carcinoma
+individuals and tested whether they change and if it results in an
+increased relative risk of adenoma or carcinoma. Next, we assessed how
+this relative risk compared to CRC associated genera for both adenoma
+and carcinoma. Third, using Random Forest models we assessed whether the
+community context can increase the classification model area under the
+curve (AUC). Finally, we examine whether the studies that were used were
+adequately powered and if not what effect size they were powered for.
 

### 2017-11-06 code/make_alpha_normalized_graphs.R
make figure 1

@@ -93,7 +93,7 @@ matched_tissue_graph <- matched_tissue_data %>%
 alpha_all_graph <- grid.arrange(stool_graph, unmatched_tissue_graph, matched_tissue_graph)
 
 
-ggsave("results/figures/zscore_alpha_graph.pdf", 
+ggsave("results/figures/Figure1.pdf", 
        alpha_all_graph, width = 6, height = 9, dpi = 300)
 
 

### 2017-11-06 code/make_alpha_normalized_graphs.R
change shannon evenness to evenness

@@ -36,7 +36,7 @@ stool_graph <- stool_data %>%
                           labels = c("Control", "Adenoma", "Cancer")), 
          alpha_metric = factor(alpha_metric, 
                                levels = c("sobs", "shannoneven", "shannon"), 
-                               labels = c("Observed OTUs", "Shannon Evenness", "Shannon Diversity"))) %>% 
+                               labels = c("Observed OTUs", "Evenness", "Shannon Diversity"))) %>% 
   ggplot(aes(disease, measurement, color = disease)) + 
   geom_boxplot(show.legend = F) + facet_grid(. ~ alpha_metric) + 
   scale_color_manual(values = c('#228B22', '#FFD700', '#DC143C')) + 
@@ -55,7 +55,7 @@ unmatched_tissue_graph <- unmatched_tissue_data %>%
                           labels = c("Control", "Adenoma", "Cancer")), 
          alpha_metric = factor(alpha_metric, 
                                levels = c("sobs", "shannoneven", "shannon"), 
-                               labels = c("Observed OTUs", "Shannon Evenness", "Shannon Diversity"))) %>% 
+                               labels = c("Observed OTUs", "Evenness", "Shannon Diversity"))) %>% 
   ggplot(aes(disease, measurement, color = disease)) + 
   geom_boxplot(show.legend = F) + facet_grid(. ~ alpha_metric) + 
   scale_color_manual(values = c('#228B22', '#FFD700', '#DC143C')) + 
@@ -73,7 +73,7 @@ matched_tissue_graph <- matched_tissue_data %>%
                           labels = c("Control", "Cancer")), 
          alpha_metric = factor(alpha_metric, 
                                levels = c("sobs", "shannoneven", "shannon"), 
-                               labels = c("Observed OTUs", "Shannon Evenness", "Shannon Diversity"))) %>% 
+                               labels = c("Observed OTUs", "Evenness", "Shannon Diversity"))) %>% 
   ggplot(aes(disease, measurement, color = disease)) + 
   geom_boxplot(show.legend = F) + facet_grid(. ~ alpha_metric) + 
   scale_color_manual(values = c('#228B22', '#DC143C')) + 

### 2017-11-06 code/get_tissue_combined_alpha.R
change mixed effect analysis to icnlude polyp

@@ -27,7 +27,8 @@ tissue_matched <- read.csv("data/process/tables/alpha_tissue_matched_data.csv",
   mutate(matchings = ifelse(disease == "cancer" | disease == "polyp", 1, 0), 
          v_region = ifelse(study == "burns", invisible("V5-6"), 
                            ifelse(study == "dejea", invisible("V3-5"), 
-                                  ifelse(study == "lu", invisible("V3-4"), invisible("V1-2")))))
+                                  ifelse(study == "lu", invisible("V3-4"), invisible("V1-2"))))) %>% 
+  mutate(disease = stringr::str_replace(disease, "Polyp", "polyp"))
 
 tissue_unmatched <- read.csv("data/process/tables/alpha_tissue_unmatched_data.csv", 
                              header = T, stringsAsFactors = F) %>% 
@@ -189,7 +190,7 @@ unmatched_mixeffect_results <- mapply(get_mixed_effect, c("sobs", "shannon", "sh
 
 
 matched_mixeffect_results <- mapply(get_mixed_effect, c("sobs", "shannon", "shannoneven"), 
-                                      individual = "id", data_set = "good_tissue_matched", 
+                                      individual = "id", data_set = "tissue_matched", 
                                     USE.NAMES = T, SIMPLIFY = F) %>% 
   bind_rows() %>% gather(key = alpha_metric, value = value) %>% 
   mutate(measure = rep(c("chi_sq", "pvalue"), length(value)/2)) %>% 
@@ -202,7 +203,7 @@ mapply(make_the_tables,
        c("unmatched_ttest_tissue", "matched_ttest_tissue", "unmatched_tukey_results", 
          "unmatched_mixeffect_results", "matched_mixeffect_results"))
 
-write_csv(good_tissue_matched, "data/process/tables/matched_tissue_normalized_alpha_all_data.csv")
+write_csv(tissue_matched, "data/process/tables/matched_tissue_normalized_alpha_all_data.csv")
 write_csv(tissue_unmatched, "data/process/tables/unmatched_tissue_normalized_alpha_all_data.csv")
 
 

### 2017-11-06 code/make_alpha_normalized_graphs.R
update figure 1

@@ -33,12 +33,13 @@ unmatched_tissue_data <- read_csv("data/process/tables/unmatched_tissue_normaliz
 stool_graph <- stool_data %>% 
   mutate(disease = factor(disease, 
                           levels = c("control", "polyp", "cancer"), 
-                          labels = c("Control", "Adenoma", "Cancer")), 
+                          labels = c("Control", "Adenoma", "Carcinoma")), 
          alpha_metric = factor(alpha_metric, 
                                levels = c("sobs", "shannoneven", "shannon"), 
                                labels = c("Observed OTUs", "Evenness", "Shannon Diversity"))) %>% 
   ggplot(aes(disease, measurement, color = disease)) + 
   geom_boxplot(show.legend = F) + facet_grid(. ~ alpha_metric) + 
+  geom_hline(yintercept = 0, linetype = "dashed", color = "black") + 
   scale_color_manual(values = c('#228B22', '#FFD700', '#DC143C')) + 
   labs(x = "", y = "Z-Score") + theme_bw() + ggtitle("A") + 
   annotate("text", label = paste("Stool"), x = 2, y = 3.5, size = 2.5) + 
@@ -52,12 +53,13 @@ stool_graph <- stool_data %>%
 unmatched_tissue_graph <- unmatched_tissue_data %>% 
   mutate(disease = factor(disease, 
                           levels = c("control", "polyp", "cancer"), 
-                          labels = c("Control", "Adenoma", "Cancer")), 
+                          labels = c("Control", "Adenoma", "Carcinoma")), 
          alpha_metric = factor(alpha_metric, 
                                levels = c("sobs", "shannoneven", "shannon"), 
                                labels = c("Observed OTUs", "Evenness", "Shannon Diversity"))) %>% 
   ggplot(aes(disease, measurement, color = disease)) + 
   geom_boxplot(show.legend = F) + facet_grid(. ~ alpha_metric) + 
+  geom_hline(yintercept = 0, linetype = "dashed", color = "black") + 
   scale_color_manual(values = c('#228B22', '#FFD700', '#DC143C')) + 
   labs(x = "", y = "Z-Score") + theme_bw() + ggtitle("B") + 
   annotate("text", label = paste("Unmatched Tissue"), x = 2, y = 3, size = 2.5) + 
@@ -69,16 +71,17 @@ unmatched_tissue_graph <- unmatched_tissue_data %>%
 
[…]

### 2017-11-06 code/get_adneoma_select_inc_genera_positivity_RR_stool.R
add code to aggregate pooled counts

@@ -1,0 +1,188 @@
+### Test each increasing positive amount against 0 for RR of Adenoma
+### Marc Sze
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+# Load needed libraries
+loadLibs(c("tidyverse", "epiR", "metafor"))
+
+
+# Stool Only polyp sets
+# Hale, Wang, Brim, Weir, Ahn, Zeller, Baxter
+stool_sets <- c("brim", "zeller", "baxter", "hale")
+
+# CRC genera of interest
+crc_genera <- c("Fusobacterium", "Peptostreptococcus", "Porphyromonas", "Parvimonas")
+
+
+##############################################################################################
+############### List of functions used to analyze the data ###################################
+##############################################################################################
+
+# Control function to get all the data, basically runs the above functions in a
+# contained location withouth having to repeat them
+get_data <- function(i){
+  # i is the study of interest
+  
+  # gets original sample names
+  sample_names <- (get_file(i, "data/process/", "_genera_shared.csv") %>% 
+                     mutate(sample_names = rownames(.)) %>% select(sample_names))[, "sample_names"]
+  # grabs subsampled data and assigns rownames from sample names to table
+  sub_genera_data <- get_file(i, "data/process/", "_subsample_genera.csv", rows_present = F, 
+                              vec_of_rownames = sample_names) %>% 
+    as.data.frame() %>% mutate(sample_ID = rownames(.)) %>% 
+    select(sample_ID, everything())
+  # grabs the meta data and transforms polyp to control (polyp/control vs cancer) 
+  study_meta <- get_file(i, "data/process/", ".metadata", rows_present = F,  
+                         "stool", metadata = T)
+  # conditional that checks for whether length of rows of meta data is smaller
+  if(length(rownames(study_meta)) < length(rownames(sub_genera_data))){
+    # grab only the samples in the meta data file for down stream analysis
+    sub_genera_data <- sub_genera_data %>% slice(match(study_meta$sampleID, sample_ID))
+    
+  } else{
+    # grab only files in the data file for analysis
+    study_meta <- study_meta %>% slice(match(sub_genera_data$sample_ID, sampleID))
+  }
+  # re assigns the rown names while removing the extra column used for sorting
+  sample_names <- sub_genera_data$sample_ID
+  # creates a list file with both data sets
+  dataList <- list(sub_genera_data = sub_genera_data, 
+                   study_meta = study_meta)
+  # returns the combined list file
+  return(dataList)
+  
+}
[…]

### 2017-11-06 code/get_adneoma_select_inc_genera_positivity_RR_tissue.R

@@ -1,0 +1,207 @@
+### Test each increasing positive amount against 0 for RR of Adenoma
+### Tissue
+### Marc Sze
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+# Load needed libraries
+loadLibs(c("tidyverse", "epiR", "metafor"))
+
+
+# Tissue Only sets
+# Lu, Dejea, Sana, Burns, Geng
+# Remove Lu since it only has polyps and no cancer cases
+tissue_sets <- c("lu")
+
+# Both Tissue and Stool
+# flemer sampletype = biopsy or stool
+# chen sample_type = tissue or stool
+# Flemer, Chen
+both_sets <- c("flemer")
+
+# CRC genera of interest
+crc_genera <- c("Fusobacterium", "Peptostreptococcus", "Porphyromonas", "Parvimonas")
+
+
+# Read in specific data tables to be used
+
+tissue_matched <- read.csv("data/process/tables/alpha_tissue_matched_data.csv", 
+                           header = T, stringsAsFactors = F) %>% 
+  mutate(matchings = ifelse(disease == "cancer" | disease == "polyp", 1, 0)) %>% 
+  filter(study %in% c(both_sets, tissue_sets))
+
+tissue_unmatched <- read.csv("data/process/tables/alpha_tissue_unmatched_data.csv", 
+                             header = T, stringsAsFactors = F) %>% 
+  filter(study %in% c(both_sets, tissue_sets), disease != "cancer")
+
+
+combined_tissue <- tissue_unmatched %>% 
+  select(one_of("group", "study", "disease")) %>% 
+  bind_rows(
+    select(tissue_matched, 
+           one_of("group", "study", "disease")))
+
+##############################################################################################
+############### List of functions used to analyze the data ###################################
+##############################################################################################
+
+# Control function to get all the data, basically runs the above functions in a
+# contained location withouth having to repeat them
+get_data <- function(i, metadata_table){
+  # i is the study of interest
+  
+  # gets original sample names
+  sample_names <- (get_file(i, "data/process/", "_genera_shared.csv") %>% 
+                     mutate(sample_names = rownames(.)) %>% select(sample_names))[, "sample_names"]
+  # grabs subsampled data and assigns rownames from sample names to table
+  sub_genera_data <- get_file(i, "data/process/", "_subsample_genera.csv", rows_present = F, 
+                              vec_of_rownames = sample_names) %>% 
[…]

### 2017-11-06 submission/manuscript.Rmd
update fig 3 graph and code

@@
 Although these values were significant the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2].
-There was no increased relative risk for these metrics for adenoma or for tissue in general [Figure S1-3].
+There was no increased relative risk for these metrics for adenoma or for tissue in general [Figure 2A & S1].
 
@@
 
-**Figure 1: **
+**Figure 1: Alpha Diversity Differences between Control, Adenoma, and Carcinoma Across Sampling Site.** A) Alpha diversity metric differences by group in stool samples.
+B) Alpha diversity metric differences by group in unmatched tissue samples.
+C) Alpha diversity metric differences by group in matched tissue samples.
+The dashed line represents a Z-score of 0 or no difference from the median.
 
-**Figure 2: **
+**Figure 2: Relative Risk for Adenoma or Carcinoma based on Alpha Diversity Metrics in Stool.** A) Alpha metric relative risk for adenoma.
+B) Alpha metric relative risk for carcinoma.
+Colors represent the different variable regions used within the respective study.
 
-**Figure 3: **
+**Figure 3: Colorectal Cancer Associated Genera Relative Risk for Adenoma and Carcinoma in Stool and Tissue.** A) Adenoma relative risk in stool.
+B) Carinoma relative risk in stool.
+C) Adenoma relative risk in tissue.
+D) Carcinoma relative risk in tissue.
+For all panels CRC Associated Genera Expansion is the total relative abundance of all 4 CRC associated genera above or below the median.
+Increased CRC Associated Genera Positivity
 
@@
 
-**Figure S1: **
+**Figure S1: Relative Risk for Adenoma or Carcinoma based on Alpha Diversity Metrics in Tissue.** A) Alpha metric relative risk for adenoma.
+B) Alpha metric relative risk for carcinoma.
+Colors represent the different variable regions used within the respective study.
 

### 2017-11-06 code/make_select_genus_RR_graph.R

@@ -11,24 +11,38 @@ loadLibs(c("tidyverse", "gridExtra", "viridis"))
 adn_all_stool <- read_csv("data/process/tables/adn_select_genus_RR_stool_composite.csv") %>% 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% 
-  bind_rows(read_csv("data/process/tables/adn_select_genus_RR_stool_ind_results.csv"))
+  bind_rows(read_csv("data/process/tables/adn_select_genus_RR_stool_ind_results.csv")) %>% 
+  bind_rows(read_csv("data/process/tables/adn_select_genus_inc_4_stool.csv") %>% 
+              rename(est = rr, lower = ci_lb, upper = ci_ub)) %>% 
+  mutate(study = ifelse(is.na(study), invisible("composite"), invisible(study)))
 
 adn_all_tissue <- read_csv("data/process/tables/adn_select_genus_RR_tissue_composite.csv") %>% 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% 
-  bind_rows(read_csv("data/process/tables/adn_select_genus_RR_tissue_ind_results.csv"))
+  bind_rows(read_csv("data/process/tables/adn_select_genus_RR_tissue_ind_results.csv")) %>% 
+  bind_rows(read_csv("data/process/tables/adn_select_genus_inc_4_tissue.csv") %>% 
+              rename(est = rr, lower = ci_lb, upper = ci_ub)) %>% 
+  mutate(study = ifelse(is.na(study), invisible("composite"), invisible(study)))
 
 # Load in needed data tables (carcinoma)
 crc_all_stool <- read_csv("data/process/tables/select_genus_RR_stool_composite.csv") %>% 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% 
-  bind_rows(read_csv("data/process/tables/select_genus_RR_stool_ind_results.csv"))
+  bind_rows(read_csv("data/process/tables/select_genus_RR_stool_ind_results.csv")) %>% 
+  bind_rows(read_csv("data/process/tables/select_genus_inc_4_stool.csv") %>% 
+              rename(est = rr, lower = ci_lb, upper = ci_ub)) %>% 
+  mutate(study = ifelse(is.na(study), invisible("composite"), invisible(study))) %>% 
[…]

### 2017-11-07 submission/manuscript.Rmd
update figure 3
update data table read in

@@
 adn_select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_RR_tissue_composite.csv"))
+
+select_inc_4_stool <- as.data.frame(read_csv("../data/process/tables/select_genus_inc_4_stool.csv"))
+select_inc_4_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_inc_4_tissue.csv"))
+
+adn_select_inc_4_RR_stool <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_inc_4_stool.csv"))
+adn_select_inc_4_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_inc_4_tissue.csv"))
 
@@
 D) Carcinoma relative risk in tissue.
-For all panels CRC Associated Genera Expansion is the total relative abundance of all 4 CRC associated genera above or below the median.
-Increased CRC Associated Genera Positivity
+For all panels the relative risk was also compared to whether one, two, three, or four of the CRC associated genera were present.
 

### 2017-11-07 submission/manuscript.Rmd
add changes based on fig 3

@@
 The relative risk for feces was greater than either evenness or Shannon diversity [Figure 2 & 3].
-The relative risk did not increase when considering the total abundance or increasing number of carcinoma associated genera [Figure 3].
+The relative risk of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D].
 The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`).
@@
 
-Only two measures for adenoma were significant when investigating these CRC associated genera.
-The first was *Porphyromonas* for feces (RR range: `r round(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2)` - `r round(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2)`, P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and the second was the total relative abundance of all four CRC associated genera on tissue (RR range: `r round(filter(adn_select_RR_tissue, measure == "total_four")[, "ci_lb"], digits = 2)` - `r round(filter(adn_select_RR_tissue, measure == "total_four")[, "ci_ub"], digits = 2)`, P-value = `r format(filter(adn_select_RR_tissue, measure == "total_four")[, "pvalue"], digits = 3, nsmall = 3)`).
+Two measures in stool for adenoma were significant when investigating these CRC associated genera.
+The first was *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and the second was being positive for three CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A].
+For tissue three measures for adenoma were significant; being positive for one CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), being positive for two CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].
 
@@
 Although there was a step wise decrease in diversity from control to adenoma to carcinoma; this did not translate into large effect sizes for the relative risk of either of these two conditions.
-These clear changes were not easily recapitulated in those with adenoma.
-Even though CRC associated genera increase the relative risk of carcinoma they do not increase the relative risk of adenoma.
-This information suggests that these specific genera may not be the primary members of the microbial community that contributes to the formation of an adenoma but is for a carcinoma.
+Even though CRC associated genera increases the relative risk of carcinoma it does not increase the relative risk of adenoma.
+This information suggests that these specific genera may not be the primary members of the microbial community that contributes to the formation of an adenoma but is for carcinoma.
 Additionally, our data shows that by using the whole community our models perform better then when they only use the CRC associated genera.
@@
 The driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], can be supported with this data for carcinoma but not necessarily for adenoma.
-The drasitically increased relative risk of disease when considering the CRC associated genera is highly supportive of this type of process.
-In a driver-passenger scenario it is possible that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring.
-This would account for the observation that there is no synergistic increase in relative risk when accounting for either the total number or increasing abundance of these genera.
-The initial establishment of the driver within the system is also dependent on the community that is present and this is supported by the observation that when adding the community context to our models along with the CRC associated genera the model AUC increases.
+The drasitically increased relative risk of disease when considering the CRC associated genera is highly supportive of this type of process, especially in the context of increasing relative risk with more CRC associated genera positivity.
+It is also possible that in a driver-passenger scenario it is possible that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring.
+This would account for the observation that there is no constant additive effect on relative risk for increasing positivity.
+Additionally, the initial establishment of the driver within the system is also dependent on the community that is present and this is supported by the observation that when adding the community context to our models along with the CRC associated genera the model AUC increases.
 
@@
 The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle.
-Although there may be localized changes that do depend on the driver-passenger model, supported by the *Porphyromonas* finding in adenoma stool and the relative total abundance of all four CRC associated genera in tissue [Figure 3], there may be another process involved that ultimately exacerbates the condition from a subtle localized change to a global community one.
+Although there may be localized changes that do depend on the driver-passenger model, supported by the one, two, and four positive CRC associated genera in tissue relative risk increases [Figure 3C], there may be another process involved that ultimately exacerbates the condition from a subtle localized change to a global community one.
 The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case.
@@
 
-Although there are still questions that need to be answered for the microbiota and carcinoma a clearer framework is beginning to develop as to how this occurs.
-The role of the microbiota in adenoma is still not clear and part of the reason may be because many studies are not powered effectively to observe the small changes reported here.
+Although there are still questions that need to be answered for the microbiota and carcinoma, a clearer framework is beginning to develop as to how this occurs.
+The role of the microbiota in adenoma is still not clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here.
 It is realistic to suspect that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control.
-Most of the studies analyzed have sufficient power to detect these changes.
+Most of the studies analyzed have sufficient power to detect these type of changes.
 In contrast, our data suggests that the adenoma effect size is relatively small.
-None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences occur in adenoma.
-Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma genesis.
+None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences consistently occur in adenoma.
+Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma formation.
 
@@
 However, within the context of adenoma it is less clear that this relationship exists.
-These observations highlight the importance of power and sample number considerations when considering investigations into the microbiota and adenoma due to the subtle changes in community.
-This study helps to identify the problems that have been solved and the challenges that lie ahead in the invesitgation of colorectal cancer and the microbiota.
+These observations highlight the importance of power and sample number considerations when considering investigations into the microbiota and adenoma due to the subtle changes in the community.
 

### 2017-11-07 submission/manuscript.tex
add changes based on fig 3

@@
 {[}Figure 2{]}. There was no increased relative risk for these metrics
-for adenoma or for tissue in general {[}Figure S1-3{]}.
+for adenoma or for tissue in general {[}Figure 2A \& S1{]}.
 
@@
 {[}Figure 3{]}. The relative risk for feces was greater than either
-evenness or Shannon diversity {[}Figure 2 \& 3{]}. The relative risk did
-not increase when considering the total abundance or increasing number
-of carcinoma associated genera {[}Figure 3{]}. The RR effect size was
-greater for stool (RR range = 1.78 - 2.64) then for tissue (RR range =
-1.33 - 1.53). This decrease may be explained by the fact that tissue
+evenness or Shannon diversity {[}Figure 2 \& 3{]}. The relative risk of
+carcinoma continuously increased as individuals tested positive for more
+CRC associated genera {[}Figure 3B \& 3D{]}. The RR effect size was
+greater for stool (RR range = 1.61 - 2.74) then for tissue (RR range =
+1.21 - 1.81). This decrease may be explained by the fact that tissue
 samples include matched samples.
 
-Only two measures for adenoma were significant when investigating these
-CRC associated genera. The first was \emph{Porphyromonas} for feces (RR
-range: 1.06 - 2.29, P-value = 0.023) and the second was the total
-relative abundance of all four CRC associated genera on tissue (RR
-range: 1.21 - 3.89, P-value = 0.00933).
+Two measures in stool for adenoma were significant when investigating
+these CRC associated genera. The first was \emph{Porphyromonas} (P-value
+= 0.023) and the second was being positive for three CRC associated
+genera (P-value = 0.022) {[}Figure 3A{]}. For tissue three measures for
+adenoma were significant; being positive for one CRC associated genera
+(P-value = 0.032), being positive for two CRC associated genera (P-value
+= 0.008), and being positive for four CRC associated genera (P-value =
+0.039) {[}Figure 3C{]}.
 
@@
 did not translate into large effect sizes for the relative risk of
-either of these two conditions. These clear changes were not easily
-recapitulated in those with adenoma. Even though CRC associated genera
-increase the relative risk of carcinoma they do not increase the
+either of these two conditions. Even though CRC associated genera
+increases the relative risk of carcinoma it does not increase the
 relative risk of adenoma. This information suggests that these specific
 genera may not be the primary members of the microbial community that
-contributes to the formation of an adenoma but is for a carcinoma.
+contributes to the formation of an adenoma but is for carcinoma.
 Additionally, our data shows that by using the whole community our
@@
 disease when considering the CRC associated genera is highly supportive
-of this type of process. In a driver-passenger scenario it is possible
-that simply having the driver present or only identifying the passenger
-is a good enough proxy that the event is occuring. This would account
-for the observation that there is no synergistic increase in relative
-risk when accounting for either the total number or increasing abundance
-of these genera. The initial establishment of the driver within the
-system is also dependent on the community that is present and this is
-supported by the observation that when adding the community context to
-our models along with the CRC associated genera the model AUC increases.
+of this type of process, especially in the context of increasing
+relative risk with more CRC associated genera positivity. It is also
+possible that in a driver-passenger scenario it is possible that simply
+having the driver present or only identifying the passenger is a good
+enough proxy that the event is occuring. This would account for the
+observation that there is no constant additive effect on relative risk
+for increasing positivity. Additionally, the initial establishment of
+the driver within the system is also dependent on the community that is
+present and this is supported by the observation that when adding the
+community context to our models along with the CRC associated genera the
+model AUC increases.
 
@@
 there may be localized changes that do depend on the driver-passenger
-model, supported by the \emph{Porphyromonas} finding in adenoma stool
-and the relative total abundance of all four CRC associated genera in
-tissue {[}Figure 3{]}, there may be another process involved that
-ultimately exacerbates the condition from a subtle localized change to a
-global community one. The poor performance of the Random Forest models
-for classifying adenoma based only on the microbiota would suggest that
-this is the case. It is possible to hypothesize that at early stages of
-the diease, how the host interacts to these subtle changes could be the
-catlyst that ultimately leads to this larger global dysfunctional
-community.
+model, supported by the one, two, and four positive CRC associated
+genera in tissue relative risk increases {[}Figure 3C{]}, there may be
+another process involved that ultimately exacerbates the condition from
+a subtle localized change to a global community one. The poor
+performance of the Random Forest models for classifying adenoma based
+only on the microbiota would suggest that this is the case. It is
+possible to hypothesize that at early stages of the diease, how the host
+interacts to these subtle changes could be the catlyst that ultimately
+leads to this larger global dysfunctional community.
 
 Although there are still questions that need to be answered for the
-microbiota and carcinoma a clearer framework is beginning to develop as
+microbiota and carcinoma, a clearer framework is beginning to develop as
 to how this occurs. The role of the microbiota in adenoma is still not
-clear and part of the reason may be because many studies are not powered
-effectively to observe the small changes reported here. It is realistic
-to suspect that many changes in carcinoma could easily result in effect
-sizes that are 30\% or more between the case and control. Most of the
-studies analyzed have sufficient power to detect these changes. In
-contrast, our data suggests that the adenoma effect size is relatively
-small. None of the studies analyzed were properly powered to detect a
-10\% or lower change between case and controls and this may well be the
-range in which differences occur in adenoma. Future studies
-investigating adenoma and the microbiota need to take these factors into
-consideration if we are to work out the role of the microbiota in
-adenoma genesis.
+clear and part of the reason this may be is because many studies are not
+powered effectively to observe the small changes reported here. It is
+realistic to suspect that many changes in carcinoma could easily result
+in effect sizes that are 30\% or more between the case and control. Most
+of the studies analyzed have sufficient power to detect these type of
+changes. In contrast, our data suggests that the adenoma effect size is
+relatively small. None of the studies analyzed were properly powered to
+detect a 10\% or lower change between case and controls and this may
+well be the range in which differences consistently occur in adenoma.
+Future studies investigating adenoma and the microbiota need to take
+these factors into consideration if we are to work out the role of the
+microbiota in adenoma formation.
 
@@
 considerations when considering investigations into the microbiota and
-adenoma due to the subtle changes in community. This study helps to
-identify the problems that have been solved and the challenges that lie
-ahead in the invesitgation of colorectal cancer and the microbiota.
+adenoma due to the subtle changes in the community.
 
@@
 
-\textbf{Figure 1: }
+\textbf{Figure 1: Alpha Diversity Differences between Control, Adenoma,
+and Carcinoma Across Sampling Site.} A) Alpha diversity metric
+differences by group in stool samples. B) Alpha diversity metric
+differences by group in unmatched tissue samples. C) Alpha diversity
+metric differences by group in matched tissue samples. The dashed line
+represents a Z-score of 0 or no difference from the median.
 
-\textbf{Figure 2: }
+\textbf{Figure 2: Relative Risk for Adenoma or Carcinoma based on Alpha
+Diversity Metrics in Stool.} A) Alpha metric relative risk for adenoma.
+B) Alpha metric relative risk for carcinoma. Colors represent the
+different variable regions used within the respective study.
 
-\textbf{Figure 3: }
+\textbf{Figure 3: Colorectal Cancer Associated Genera Relative Risk for
+Adenoma and Carcinoma in Stool and Tissue.} A) Adenoma relative risk in
+stool. B) Carinoma relative risk in stool. C) Adenoma relative risk in
+tissue. D) Carcinoma relative risk in tissue. For all panels the
+relative risk was also compared to whether one, two, three, or four of
+the CRC associated genera were present.
 
@@
 
-\textbf{Figure S1: }
+\textbf{Figure S1: Relative Risk for Adenoma or Carcinoma based on Alpha
+Diversity Metrics in Tissue.} A) Alpha metric relative risk for adenoma.
+B) Alpha metric relative risk for carcinoma. Colors represent the
+different variable regions used within the respective study.
 

### 2017-11-07 code/make_rf_auc_full_versus_specific_OTU_graph.R
add figure 4 and 5 save

@@ -211,10 +211,10 @@ tissue_auc_graph <- grid.arrange(adn_tissue_graph, crc_tissue_graph)
 
 stool_auc_graph <- grid.arrange(adn_stool_graph, crc_stool_graph)
 
-ggsave("results/figures/tissue_otu_all_v_specific.pdf", 
+ggsave("results/figures/Figure5.pdf", 
        tissue_auc_graph, width = 7, height = 6, dpi = 300)
 
-ggsave("results/figures/stool_otu_all_v_specific.pdf", 
+ggsave("results/figures/Figure4.pdf", 
        stool_auc_graph, width = 7, height = 6, dpi = 300)
 
 

### 2017-11-07 code/make_rf_auc_full_versus_specific_OTU_graph.R
Change figure titles

@@ -96,7 +96,7 @@ adn_tissue_graph <- adn_tissue %>%
                        labels = c("Unmatched Tissue", "Matched Tissue")), 
          model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("Full Community", "CRC Associated\nGenera Community Only")), 
          study = factor(study, 
                         levels = c("flemer", "lu"), 
                         labels = c("Flemer", "Lu\n(matched)"))) %>% 
@@ -121,7 +121,7 @@ adn_tissue_graph <- adn_tissue %>%
 adn_stool_graph <- adn_all_stool %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("Full Community", "CRC Associated\nGenera Community Only")), 
          study = factor(study, 
                         levels = c("baxter", "brim", "hale", "zeller"), 
                         labels = c("Baxter", "Brim", "Hale", "Zeller"))) %>% 
@@ -147,7 +147,7 @@ adn_stool_graph <- adn_all_stool %>%
 crc_tissue_graph <- crc_all_tissue %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("Full Community", "CRC Associated\nGenera Community Only")), 
          combined_study = factor(combined_study, 
                         levels = c("sana_unmatched", "geng_matched", "flemer_unmatched", 
                                    "dejea_matched", "chen_unmatched", "burns_unmatched",
@@ -177,7 +177,7 @@ crc_tissue_graph <- crc_all_tissue %>%
 crc_stool_graph <- crc_all_stool %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("Full Community", "CRC Associated\nGenera Community Only")), 
          study = factor(study, 
                         levels = c("zeller", "weir", "wang", "hale", "flemer", "baxter", "ahn"), 
                         labels = c("Zeller", "Weir", "Wang", "Hale", "Flemer", "Baxter", "Ahn"))) %>% 

### 2017-11-07 code/make_rf_auc_full_versus_specific_OTU_graph.R
minor tweaks

@@ -110,7 +110,7 @@ adn_tissue_graph <- adn_tissue %>%
   labs(x = "", y = "Model AUC") + theme_bw() + ggtitle("A") + 
   scale_color_manual(name = "Study", 
                      values = c('#440154FF', '#FDE725FF')) + 
-  annotate("text", label = paste("Adenoma"), x = 2.5, y = 0.06, size = 2.5) + 
+  annotate("text", label = paste("Adenoma"), x = 2.5, y = 0.05, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
         legend.text = element_text(size = 6), 
         panel.grid.major = element_blank(), 
@@ -165,7 +165,7 @@ crc_tissue_graph <- crc_all_tissue %>%
   scale_color_manual(name = "Study", 
                      values = c('#1F998AFF', '#CBE11EFF', '#440154FF', '#2B748EFF', 
                                 '#3D4D8AFF', '#453581FF', '#453581FF')) + 
-  annotate("text", label = paste("Cancer"), x = 7.35, y = 0.03, size = 2.5) + 
+  annotate("text", label = paste("Carcinoma"), x = 7.35, y = 0.06, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.15, size = 20), 
         legend.text = element_text(size = 6), 
         panel.grid.major = element_blank(), 
@@ -193,7 +193,7 @@ crc_stool_graph <- crc_all_stool %>%
   scale_color_manual(name = "Study", 
                      values = c('#FDE725FF', '#24878EFF', '#97D83FFF', '#67CC5CFF', 
                                 '#440154FF', '#481D6FFF', '#40BC72FF')) + 
-  annotate("text", label = paste("Cancer (Stool)"), x = 7.4, y = 0.08, size = 2.5) + 
+  annotate("text", label = paste("Carcinoma (Stool)"), x = 7.4, y = 0.1, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 

### 2017-11-07 submission/manuscript.Rmd
add fig legend 4 and 5 to manuscript

@@
 
-**Figure 4: **
+**Figure 4: OTU Random Forest Model of Stool Across Studies.** A) Adenoma random forest model between the full community and CRC associated genera OTUs only.
+B) Carcinoma random forest model between the full community and CRC associated genera OTUs only.
+The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.
 
-**Figure 5: **
+**Figure 5: OTU Random Forest Model of Tissue Across Studies.** A) Adenoma random forest model between the full community and CRC associated genera OTUs only.
+B) Carcinoma random forest model between the full community and CRC associated genera OTUs only.
+The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.
 

### 2017-11-07 submission/manuscript.tex
add fig legend 4 and 5 to manuscript

@@
 
-\textbf{Figure 4: }
+\textbf{Figure 4: OTU Random Forest Model of Stool Across Studies.} A)
+Adenoma random forest model between the full community and CRC
+associated genera OTUs only. B) Carcinoma random forest model between
+the full community and CRC associated genera OTUs only. The dotted line
+represents an AUC of 0.5 and the lines represent the range in which the
+AUC for the 100 different 80/20 runs fell between.
 
-\textbf{Figure 5: }
+\textbf{Figure 5: OTU Random Forest Model of Tissue Across Studies.} A)
+Adenoma random forest model between the full community and CRC
+associated genera OTUs only. B) Carcinoma random forest model between
+the full community and CRC associated genera OTUs only. The dotted line
+represents an AUC of 0.5 and the lines represent the range in which the
+AUC for the 100 different 80/20 runs fell between.
 

### 2017-11-07 code/make_rf_auc_full_versus_specific_graph.R
switch to carcinoma

@@ -145,7 +145,7 @@ crc_stool_graph <- crc_all_stool %>%
   labs(x = "", y = "Model AUC") + theme_bw() + ggtitle("B") + 
   scale_color_manual(name = "Study", 
                      values = c('#40BC72FF', '#481D6FFF', '#440154FF', '#67CC5CFF', '#97D83FFF', '#24878EFF', '#FDE725FF')) + 
-  annotate("text", label = paste("Cancer (Stool)"), x = 0.75, y = 1.09, size = 2.5) + 
+  annotate("text", label = paste("Carcinoma (Stool)"), x = 0.8, y = 1.09, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.25, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
@@ -159,9 +159,9 @@ crc_stool_graph <- crc_all_stool %>%
 tissue_auc_graph <- grid.arrange(adn_tissue_graph, crc_tissue_graph)
 stool_auc_graph <- grid.arrange(adn_stool_graph, crc_stool_graph, ncol = 2)
 
-ggsave("results/figures/tissue_all_vs_select_genera_AUC.pdf", 
+ggsave("results/figures/FigureS3.pdf", 
        tissue_auc_graph, width = 6, height = 6, dpi = 300)
 
-ggsave("results/figures/stool_all_vs_select_genera_AUC.pdf", 
+ggsave("results/figures/FigureS2.pdf", 
        stool_auc_graph, width = 6, height = 6, dpi = 300)
 

### 2017-11-07 code/make_rf_auc_full_versus_specific_graph.R
switch to carcinoma

@@ -122,7 +122,7 @@ crc_tissue_graph <- crc_all_tissue %>%
   labs(x = "", y = "Model AUC") + theme_bw() + ggtitle("B") + 
   scale_color_manual(name = "Study", 
                      values = c('#453581FF', '#3D4D8AFF', '#2B748EFF', '#440154FF', '#CBE11EFF', '#1F998AFF')) + 
-  annotate("text", label = paste("Cancer"), x = 0.55, y = 1.07, size = 2.5) + 
+  annotate("text", label = paste("Carcinoma"), x = 0.6, y = 1.07, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
         legend.text = element_text(size = 6), 
         panel.grid.major = element_blank(), 

### 2017-11-07 submission/manuscript.Rmd
add supplemental legend 2 and 3

@@
 
-**Figure S2: **
+**Figure S2: Random Forest Genus Model AUC for each Stool Study.** A) AUC of Adenoma models using all genera or CRC associated genera only.
+B) AUC of Carcinoma models using all genera or CRC associated genera only.
+The black line represents the median within each group.
 
-**Figure S3: **
+**Figure S3: Random Forest Genus Model AUC for each Tissue Study.** A) AUC of Adenoma models using all genera or CRC associated genera only divided between matched and unmatched tissue.
+B) AUC of Carcinoma models using all genera or CRC associated genera only.
+The black line represents the median within each group divided between matched and unmatched tissue.
 

### 2017-11-07 submission/manuscript.tex
add supplemental legend 2 and 3

@@
 
-\textbf{Figure S2: }
+\textbf{Figure S2: Random Forest Genus Model AUC for each Stool Study.}
+A) AUC of Adenoma models using all genera or CRC associated genera only.
+B) AUC of Carcinoma models using all genera or CRC associated genera
+only. The black line represents the median within each group.
 
-\textbf{Figure S3: }
+\textbf{Figure S3: Random Forest Genus Model AUC for each Tissue Study.}
+A) AUC of Adenoma models using all genera or CRC associated genera only
+divided between matched and unmatched tissue. B) AUC of Carcinoma models
+using all genera or CRC associated genera only. The black line
+represents the median within each group divided between matched and
+unmatched tissue.
 

### 2017-11-07 code/make_genus_rf_auc_against_study_graph.R
minor adj and swtich to supplement

@@ -114,10 +114,9 @@ adn_tissue_graph <- adn_all_tissue %>%
   geom_point(size = 3.5, show.legend = F) + facet_grid(. ~ model_type) + 
   geom_hline(yintercept = 0.5, linetype = "dashed") + 
   coord_cartesian(ylim = c(0, 1.05)) + 
-  labs(x = "", y = "AUC") + theme_bw() + ggtitle("A") + 
+  labs(x = "", y = "AUC") + theme_bw() + 
   scale_color_manual(name = "Study", 
                      values = c('#440154FF', '#FDE725FF')) + 
-  annotate("text", label = paste("Adenoma (Unmatched Tissue)"), x = 0.9, y = 1.07, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
         legend.text = element_text(size = 6), 
         panel.grid.major = element_blank(), 
@@ -166,7 +165,7 @@ crc_unmatched_tissue_graph <- crc_all_unmatched_tissue %>%
   labs(x = "", y = "AUC") + theme_bw() + ggtitle("A") + 
   scale_color_manual(name = "Study", 
                      values = c('#453581FF', '#3D4D8AFF', '#440154FF', '#1F998AFF')) + 
-  annotate("text", label = paste("Cancer (Unmatched Tissue)"), x = 1.4, y = 1.07, size = 2.5) + 
+  annotate("text", label = paste("Carcinoma (Unmatched Tissue)"), x = 1.6, y = 1.07, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
         legend.text = element_text(size = 6), 
         panel.grid.major = element_blank(), 
@@ -191,7 +190,7 @@ crc_matched_tissue_graph <- crc_all_matched_tissue %>%
   labs(x = "", y = "AUC") + theme_bw() + ggtitle("B") + 
   scale_color_manual(name = "Study", 
                      values = c('#453581FF', '#2B748EFF', '#CBE11EFF')) + 
-  annotate("text", label = paste("Cancer (Matched Tissue)"), x = 1.1, y = 1.07, size = 2.5) + 
+  annotate("text", label = paste("Carcinoma (Matched Tissue)"), x = 1.3, y = 1.07, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
         legend.text = element_text(size = 6), 
[…]

### 2017-11-07 code/make_genus_rf_auc_against_study_graph.R
add titles

@@ -142,7 +142,7 @@ adn_stool_graph <- adn_all_stool %>%
   scale_color_manual(name = "Study", 
                      values = c('#481D6FFF', '#34618DFF', '#67CC5CFF', '#FDE725FF')) + 
   annotate("text", label = paste("Adenoma (Stool)"), x = 1.0, y = 1.07, size = 2.5) + 
-  theme(plot.title = element_text(face="bold", hjust = -0.25, size = 20), 
+  theme(plot.title = element_text(face="bold", hjust = -0.09, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10))
@@ -217,7 +217,7 @@ crc_stool_graph <- crc_all_stool %>%
                      values = c('#40BC72FF', '#481D6FFF', '#440154FF', '#67CC5CFF', 
                                 '#97D83FFF', '#24878EFF', '#FDE725FF')) + 
   annotate("text", label = paste("Carcinoma (Stool)"), x = 1.4, y = 1.07, size = 2.5) + 
-  theme(plot.title = element_text(face="bold", hjust = -0.25, size = 20), 
+  theme(plot.title = element_text(face="bold", hjust = -0.09, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10))

### 2017-11-07 code/make_genus_rf_auc_against_study_graph.R
minor axis correction

@@ -103,13 +103,13 @@ crc_all_unmatched_tissue <- make_table(crc_unmatched_tissue_studies, "data/proce
 adn_tissue_graph <- adn_all_tissue %>% 
   mutate(model = factor(model, 
                         levels = c("flemer", "lu"), 
-                        labels = c("Flemer", "Lu")), 
+                        labels = c("Flemer", "Lu\n(Matched)")), 
          model_type = factor(model_type, 
                              levels = c("full", "select"), 
                              labels = c("All Genera", "CRC Associated\nGenera Only")), 
          study = factor(study, 
                         levels = c("flemer", "lu"), 
-                        labels = c("Flemer", "Lu"))) %>% 
+                        labels = c("Flemer", "Lu\n(Matched)"))) %>% 
   ggplot(aes(model, auc, color = study, group = model_type)) + 
   geom_point(size = 3.5, show.legend = F) + facet_grid(. ~ model_type) + 
   geom_hline(yintercept = 0.5, linetype = "dashed") + 

### 2017-11-07 code/make_est_power_graph.R
switch power to fig6

@@ -79,7 +79,7 @@ crc_study_power <- crc_power %>%
                      values = c('#FDE725FF', '#24878EFF', '#97D83FFF', '#1F998AFF', 
                                 '#67CC5CFF', '#CBE11EFF', '#440154FF', '#2B748EFF', 
                                 '#453581FF', '#481D6FFF', '#40BC72FF')) + 
-  annotate("text", label = paste("Cancer"), x = 0.7, y = 1.03, size = 2.5) + 
+  annotate("text", label = paste("Carcinoma"), x = 0.85, y = 1.03, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.15, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
@@ -104,7 +104,7 @@ combined_sample_n <- combined_data %>%
                      values = c('#FDE725FF', '#24878EFF', '#97D83FFF', '#1F998AFF', '#FDE725FF',  
                                 '#67CC5CFF', '#CBE11EFF', '#440154FF', '#2B748EFF', '#453581FF', 
                                 '#34618DFF', '#481D6FFF', '#40BC72FF')) + 
-  theme(plot.title = element_text(face="bold", hjust = -0.25, size = 20), 
+  theme(plot.title = element_text(face="bold", hjust = -0.18, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10))
@@ -117,7 +117,7 @@ combined_sample_n <- combined_data %>%
 power_graph <- grid.arrange(adn_study_power, crc_study_power, combined_sample_n, 
                             layout_matrix = rbind(c(1, 3), c(2, 3)))
 
-ggsave("results/figures/power_graph.pdf", 
+ggsave("results/figures/Figure6.pdf", 
        power_graph, width = 10, height = 8, dpi = 300)
 
 

### 2017-11-07 submission/manuscript.Rmd
complete supplemental fig legend

@@
 
-**Figure S4: **
+**Figure S4: Random Forest Prediction Success Using Genera for each Stool Study.** A) AUC for prediction in Adenoma using all genera or CRC associated genera only.
+B) AUC for prediction in Carcinoma using all genera or CRC associated genera only.
+The dotted line represents an AUC of 0.5.
+The x-axis is the data set in which the model was initially trained on.
 
-**Figure S5: **
+**Figure S5: Random Forest Prediction Success of Carcinoma Using Genera for each Tissue Study.** A) AUC for prediction in unmatched tissue for all genera or CRC associated genera only.
+B) AUC for prediction in matched tissue using all genera or CRC associated genera only.
+The dotted line represents an AUC of 0.5.
+The x-axis is the data set in which the model was initially trained on.
 
-**Figure S6: **
-
-**Figure S7: **
-
-**Figure S8: **
+**Figure S6: Random Forest Prediction Success of Adenoma Using Genera for each Tissue Study.**
 

### 2017-11-07 submission/manuscript.tex
complete supplemental fig legend

@@
 
-\textbf{Figure S4: }
+\textbf{Figure S4: Random Forest Prediction Success Using Genera for
+each Stool Study.} A) AUC for prediction in Adenoma using all genera or
+CRC associated genera only. B) AUC for prediction in Carcinoma using all
+genera or CRC associated genera only. The dotted line represents an AUC
+of 0.5. The x-axis is the data set in which the model was initially
+trained on.
 
-\textbf{Figure S5: }
+\textbf{Figure S5: Random Forest Prediction Success of Carcinoma Using
+Genera for each Tissue Study.} A) AUC for prediction in unmatched tissue
+for all genera or CRC associated genera only. B) AUC for prediction in
+matched tissue using all genera or CRC associated genera only. The
+dotted line represents an AUC of 0.5. The x-axis is the data set in
+which the model was initially trained on.
 
-\textbf{Figure S6: }
-
-\textbf{Figure S7: }
-
-\textbf{Figure S8: }
+\textbf{Figure S6: Random Forest Prediction Success of Adenoma Using
+Genera for each Tissue Study.}
 

### 2017-11-07 submission/manuscript.Rmd
complete adding figure legend

@@
 
-**Figure 6: **
+**Figure 6: Power and Effect Size Analysis of Studies Included.** A) Power based on effect size for studies with Adenoma individuals.
+B) Power based on effect size for studies with Carcinoma individuals.
+C) The estimated sample number needed for each arm of each study to detect aneffect size of 1-30%.
+The dotted red lines in A) and B) represent the generally used power of 0.8.
 

### 2017-11-07 submission/manuscript.tex
complete adding figure legend

@@
 
-\textbf{Figure 6: }
+\textbf{Figure 6: Power and Effect Size Analysis of Studies Included.}
+A) Power based on effect size for studies with Adenoma individuals. B)
+Power based on effect size for studies with Carcinoma individuals. C)
+The estimated sample number needed for each arm of each study to detect
+aneffect size of 1-30\%. The dotted red lines in A) and B) represent the
+generally used power of 0.8.
 

### 2017-11-07 submission/manuscript.Rmd
update supplemental refs in text

@@
 This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5].
-The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S4-S5].
-Both genus models perform similarily in their ability to be able to predict lesion (adenoma or carcinoma) with carcinoma having a higher AUC then adenoma [Figure S6-S8].
-Matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S6A, S7B, & S8] then carcinoma models [Figure S6B & S7A].
+The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S2-S3].
+Both genus models perform similarily in their ability to be able to predict lesion (adenoma or carcinoma) with carcinoma having a higher AUC then adenoma [Figure S4-S5].
+Matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S4A, S5B, & S6] then carcinoma models [Figure S4B & S5A].
 

### 2017-11-07 submission/manuscript.tex
update supplemental refs in text

@@
 models with the full genera models performing better then the CRC
-associated genera models {[}Figure S4-S5{]}. Both genus models perform
+associated genera models {[}Figure S2-S3{]}. Both genus models perform
 similarily in their ability to be able to predict lesion (adenoma or
 carcinoma) with carcinoma having a higher AUC then adenoma {[}Figure
-S6-S8{]}. Matched tissue samples for those with carcinoma had an AUC
-that was more similar to the adenoma models {[}Figure S6A, S7B, \& S8{]}
-then carcinoma models {[}Figure S6B \& S7A{]}.
+S4-S5{]}. Matched tissue samples for those with carcinoma had an AUC
+that was more similar to the adenoma models {[}Figure S4A, S5B, \& S6{]}
+then carcinoma models {[}Figure S4B \& S5A{]}.
 

### 2017-11-07 submission/manuscript.Rmd
abstract fixes

@@
 
-**Background.** An increasing body of literature suggests that there a role for the microbiota in colorectal cancer.
-Important players within this axis have ranged from individual microbes to the whole community.
-A recent meta-analysis investigated this but only focused on potential biomarkers.
-This study expands on this previous research and tests the hypothesis that detectable changes in the bacterial community are important both to increasing relative risk and model accuracy for adenoma and carcinoma.
+**Background.** An increasing body of literature suggests that there is a role for the microbiota in colorectal cancer.
+Important drivers within this axis have ranged from individual microbes to the whole community.
+A recent meta-analysis investigated whether consistent biomarkers could be identified across studies.
+This report expands on this previous research and tests the hypothesis that the bacterial community is an important driver of disease from control to adenoma to carcinoma.
 To test this hypothesis we examined both feces (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) across 14 different studies.
 
-**Results.** There was a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness for fecal samples (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study and variable region.
-Lower Shannon diversity and evennes in fecal samples resulted in a significant increase in relative risk for carinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05).
+**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced.
+Lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) while only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk.
 Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
-Using the whole community resulted in a higher classification model AUC for both adenoma and carcinoma for fecal and tissue samples.
-Most studies are adequately powered for large effect size differences which may be more amenable for carcinoma than for adenoma.
+Using the whole community versus only CRC associated genera to build a prediction model resulted in a higher classification AUC for both adenoma and carcinoma for fecal and tissue samples.
+For the studies that were analyzed, most were adequately powered for large effect size differences which may be more amenable for carcinoma than for adenoma microbiota research.
 
-**Conclusions.** This data provides support for the idea of the "driver-passenger" model in carinoma but only partially for adenoma.
-The data is much stronger for carcinoma and this may in part be due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
+**Conclusions.** This data provides support for the importance of the bacteral community to both adenoma and carcinoma genesis.
+The evidence that has been collated within this study is much stronger for carcinoma and this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
 

### 2017-11-07 submission/manuscript.tex
abstract fixes

@@
 \textbf{Background.} An increasing body of literature suggests that
-there a role for the microbiota in colorectal cancer. Important players
-within this axis have ranged from individual microbes to the whole
-community. A recent meta-analysis investigated this but only focused on
-potential biomarkers. This study expands on this previous research and
-tests the hypothesis that detectable changes in the bacterial community
-are important both to increasing relative risk and model accuracy for
-adenoma and carcinoma. To test this hypothesis we examined both feces
-(total individuals = 1737) and tissue (total samples = 492) across 14
-different studies.
+there is a role for the microbiota in colorectal cancer. Important
+drivers within this axis have ranged from individual microbes to the
+whole community. A recent meta-analysis investigated whether consistent
+biomarkers could be identified across studies. This report expands on
+this previous research and tests the hypothesis that the bacterial
+community is an important driver of disease from control to adenoma to
+carcinoma. To test this hypothesis we examined both feces (total
+individuals = 1737) and tissue (total samples = 492) across 14 different
+studies.
 
-\textbf{Results.} There was a significant decrease from control to
-adenoma to carcinoma for both Shannon diversity and evenness for fecal
-samples (P-value \textless{} 0.05) after correcting for study and
-variable region. Lower Shannon diversity and evennes in fecal samples
-resulted in a significant increase in relative risk for carinoma
-(P-value \textless{} 0.05) but not adenoma (P-value check tables 0.05).
+\textbf{Results.} Fecal samples had a significant decrease from control
+to adenoma to carcinoma for both Shannon diversity and evenness (P-value
+\textless{} 0.05) after correcting for study effect and variable region
+sequenced. Lower Shannon diversity and evenness in fecal samples
+resulted in a significant increase in relative risk for carcinoma
+(P-value \textless{} 0.05) while only evenness for adenoma (P-value
+\textless{} 0.05) resulted in a slightly increased relative risk.
 Previously associated colorectal cancer genera (\emph{Fusobacterium},
@@
 \emph{Porphyromonas} (P-value \textless{} 0.05). Using the whole
-community resulted in a higher classification model AUC for both adenoma
-and carcinoma for fecal and tissue samples. Most studies are adequately
-powered for large effect size differences which may be more amenable for
-carcinoma than for adenoma.
+community versus only CRC associated genera to build a prediction model
+resulted in a higher classification AUC for both adenoma and carcinoma
+for fecal and tissue samples. For the studies that were analyzed, most
+were adequately powered for large effect size differences which may be
+more amenable for carcinoma than for adenoma microbiota research.
 
-\textbf{Conclusions.} This data provides support for the idea of the
-``driver-passenger'' model in carinoma but only partially for adenoma.
-The data is much stronger for carcinoma and this may in part be due to
-the low power to detect more subtle changes in the majority of studies
-that have been performed to date.
+\textbf{Conclusions.} This data provides support for the importance of
+the bacteral community to both adenoma and carcinoma genesis. The
+evidence that has been collated within this study is much stronger for
+carcinoma and this may be in part due to the low power to detect more
+subtle changes in the majority of studies that have been performed to
+date.
 

### 2017-11-07 submission/manuscript.Rmd
bacground/intro edits

@@
 Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in polyp formation in mouse models.
-There has also been numerous case control studies investigating the microbiota in both adenoma and carcinoma.
+There has also been numerous case/control studies investigating the microbiota in both adenoma and carcinoma.
 Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
@@
 
-Using both fecal (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue samples (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] within our data analysis we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
-We first assessed the diversity of controls, adenoma, and carcinoma individuals and tested whether they change and if it results in an increased relative risk of adenoma or carcinoma.
-Next, we assessed how this relative risk compared to CRC associated genera for both adenoma and carcinoma.
-Third, using Random Forest models we assessed whether the community context can increase the classification model area under the curve (AUC).
-Finally, we examine whether the studies that were used were adequately powered and if not what effect size they were powered for.
+Using both fecal (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue samples (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] we expand both the breadth and scope of the previous meta-analysis; to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
+To accomplish this we first assessed whether the diversity changes across disease (control to adenoma to carcinoma) and if it results in an increased relative risk for adenoma or carcinoma.
+Next, we assessed how typical CRC associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma.
+Third, using Random Forest models we analyzed whether the full community or only the CRC associated genera resulted in better model classification area under the curve (AUC).
+Finally, we examined at what effect size and sample n the studies that were used were adequately powered for.
 
-Our analysis found a continuous decrease in Shannon diversity from control to adenoma to carcinoma and a significantly increased relative risk for carcinoma with lower diversity.
-Using the CRC associated genera only this relative risk was higher than Shannon diversity.
-However, adding the community context in which these CRC associated genera are present increases prediction models AUC.
+Our analysis found a continuous decrease in Shannon diversity from control to adenoma to carcinoma and a significantly increased relative risk for carcinoma with this lower diversity.
+Using only the CRC associated genera this relative risk was higher than what was observed for Shannon diversity.
+However, by using the full community instead of only the CRC associated genera the AUC of the classification models increased.
 Although we analyze a data set with a large number of total individuals each individual study was underpowered for effect size differences of 10% or below between the case and control.

### 2017-11-07 submission/manuscript.tex
bacground/intro edits

@@
 {[}8--10{]} in polyp formation in mouse models. There has also been
-numerous case control studies investigating the microbiota in both
+numerous case/control studies investigating the microbiota in both
 adenoma and carcinoma. Recently, a meta-analysis was published
@@
 samples = 492) totalling over 2229 total samples across 14 studies
-{[}12--25{]} within our data analysis we expand both the breadth and
-scope of the previous meta-analysis to investigate whether the bacterial
-community is an important risk factor for both adenoma and carcinoma. We
-first assessed the diversity of controls, adenoma, and carcinoma
-individuals and tested whether they change and if it results in an
-increased relative risk of adenoma or carcinoma. Next, we assessed how
-this relative risk compared to CRC associated genera for both adenoma
-and carcinoma. Third, using Random Forest models we assessed whether the
-community context can increase the classification model area under the
-curve (AUC). Finally, we examine whether the studies that were used were
-adequately powered and if not what effect size they were powered for.
+{[}12--25{]} we expand both the breadth and scope of the previous
+meta-analysis; to investigate whether the bacterial community is an
+important risk factor for both adenoma and carcinoma. To accomplish this
+we first assessed whether the diversity changes across disease (control
+to adenoma to carcinoma) and if it results in an increased relative risk
+for adenoma or carcinoma. Next, we assessed how typical CRC associated
+genera (\emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Peptostreptococcus}, or \emph{Porphyromonas}) affect the relative
+risk of adenoma or carcinoma. Third, using Random Forest models we
+analyzed whether the full community or only the CRC associated genera
+resulted in better model classification area under the curve (AUC).
+Finally, we examined at what effect size and sample n the studies that
+were used were adequately powered for.
 
@@
 control to adenoma to carcinoma and a significantly increased relative
-risk for carcinoma with lower diversity. Using the CRC associated genera
-only this relative risk was higher than Shannon diversity. However,
-adding the community context in which these CRC associated genera are
-present increases prediction models AUC. Although we analyze a data set
-with a large number of total individuals each individual study was
-underpowered for effect size differences of 10\% or below between the
-case and control.
+risk for carcinoma with this lower diversity. Using only the CRC
+associated genera this relative risk was higher than what was observed
+for Shannon diversity. However, by using the full community instead of
+only the CRC associated genera the AUC of the classification models
+increased. Although we analyze a data set with a large number of total
+individuals each individual study was underpowered for effect size
+differences of 10\% or below between the case and control.
 

### 2017-11-07 submission/manuscript.Rmd
add result edits

@@
 
-***Fecal Diversity is Lower in Those with Carcinoma and Increases Relative Risk:***
-Using power transformed and Z-score normalized alpha diversity metrics both evenness and the Shannon diversity metrics in feces are lower in those with carcinoma then in controls but not for tissue samples [Figure 1].
-Using linear mixed-effects to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`).
+***Diversity is Lower in Those with Carcinoma and Increases Relative Risk:***
+Using power transformed and Z-score normalized alpha diversity metrics, both evenness and Shannon diversity in feces and not tissue were  lower in those with carcinoma [Figure 1].
+Using linear mixed-effect models to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`).
 This effect was not observed in tissue when additionally controlling for whether the sample came from the same individual (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
 For fecal samples a decrease in Shannon diversity and evenness resulted in a significantly increased relative risk for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2].
-Although these values were significant the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2].
-There was no increased relative risk for these metrics for adenoma or for tissue in general [Figure 2A & S1].
+Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2].
+Only a decrease in evenness had an increased relative risk for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))`).
 
-Using the Bray-Curtis distance metric, the fecal microbiota did not have a different community diversity between adenoma and control but did for carcinoma across studies [Table S1 & S2].
-The majority of unmatched tissue samples had a significant difference for both adenoma and carcinoma versus controls [Table S3 & S4].
-All matched tissue samples accross studies had no difference between any of the compared groups [Table S3 & S4].
+Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls but not between adenoma and controls [Table S1 & S2].
+Many studies with unmatched tissue samples had a significant difference for both adenoma and carcinoma versus controls [Table S3 & S4] while studies with matched tissue samples had no differences [Table S3 & S4].
 
-***Genera Previously Associated with Carcinoma Increases Relative Risk More than Alpha Diversity:***
-Both fecal and tissue samples had a significantly increased RR for carcinoma but not for adenoma [Figure 3].
-The relative risk for feces was greater than either evenness or Shannon diversity [Figure 2 & 3].
-The relative risk of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D].
-The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`).
-This decrease may be explained by the fact that tissue samples include matched samples.
+***Genera Previously Associated with Carcinoma Predominately Increases Relative Risk for Carcinoma but not Adenoma:***
+The majority of CRC associated genera for both feces and tissue had a significantly increased relative risk for carcinoma but not for adenoma [Figure 3].
+In fecal samples the relative risk due to CRC associated genera was greater than either evenness or Shannon diversity [Figure 2 & 3].
+Additionally, the relative risk of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D].
+The relative risk effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`).
+This decrease may be explained by the fact that the tissue analysis include matched samples.
 
@@
 ***Using the Whole Community Increases Model AUC over CRC Associated Genera:***
-For both fecal and tissue samples (matched and unmatched) there was a decrease in AUC when only OTUs from the CRC associated genera are used [Figure 4 & 5].
+For both fecal and tissue samples (matched and unmatched) the AUC decreases when only OTUs from the CRC associated genera are used [Figure 4 & 5].
 This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5].
@@
 When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B].
-No single study that was analyzed had the standard 80% power to detect a difference that was eqaul to or below 10% [Figure 6A & B].
+No single study that was analyzed had the standard 80% power to detect an effect size difference that was eqaul to or below 10% [Figure 6A & B].
 In order to achieve adequate power for small effect sizes it would be necessary to recruit over 1000 individuals for each arm of the study [Figure 6C].

### 2017-11-07 submission/manuscript.tex
add result edits

@@
 
-\textbf{\emph{Fecal Diversity is Lower in Those with Carcinoma and
-Increases Relative Risk:}} Using power transformed and Z-score
-normalized alpha diversity metrics both evenness and the Shannon
-diversity metrics in feces are lower in those with carcinoma then in
-controls but not for tissue samples {[}Figure 1{]}. Using linear
-mixed-effects to control for study and variable region there was a
+\textbf{\emph{Diversity is Lower in Those with Carcinoma and Increases
+Relative Risk:}} Using power transformed and Z-score normalized alpha
+diversity metrics, both evenness and Shannon diversity in feces and not
+tissue were lower in those with carcinoma {[}Figure 1{]}. Using linear
+mixed-effect models to control for study and variable region there was a
 significant decrease from control to adenoma to carcinoma for both
@@
 (P-value = 0.01 and P-value = 0.0011, respectively) {[}Figure 2{]}.
-Although these values were significant the effect size was relatively
+Although these values were significant, the effect size was relatively
 small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
-{[}Figure 2{]}. There was no increased relative risk for these metrics
-for adenoma or for tissue in general {[}Figure 2A \& S1{]}.
+{[}Figure 2{]}. Only a decrease in evenness had an increased relative
+risk for adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect
+size was even smaller than what was observed for carcinoma (RR = 1.16).
 
-Using the Bray-Curtis distance metric, the fecal microbiota did not have
-a different community diversity between adenoma and control but did for
-carcinoma across studies {[}Table S1 \& S2{]}. The majority of unmatched
-tissue samples had a significant difference for both adenoma and
-carcinoma versus controls {[}Table S3 \& S4{]}. All matched tissue
-samples accross studies had no difference between any of the compared
-groups {[}Table S3 \& S4{]}.
+Using the Bray-Curtis distance metric, there was a significant
+difference across studies in the bacterial community of fecal samples
+between carcinoma and controls but not between adenoma and controls
+{[}Table S1 \& S2{]}. Many studies with unmatched tissue samples had a
+significant difference for both adenoma and carcinoma versus controls
+{[}Table S3 \& S4{]} while studies with matched tissue samples had no
+differences {[}Table S3 \& S4{]}.
 
-\textbf{\emph{Genera Previously Associated with Carcinoma Increases
-Relative Risk More than Alpha Diversity:}} Both fecal and tissue samples
-had a significantly increased RR for carcinoma but not for adenoma
-{[}Figure 3{]}. The relative risk for feces was greater than either
-evenness or Shannon diversity {[}Figure 2 \& 3{]}. The relative risk of
-carcinoma continuously increased as individuals tested positive for more
-CRC associated genera {[}Figure 3B \& 3D{]}. The RR effect size was
-greater for stool (RR range = 1.61 - 2.74) then for tissue (RR range =
-1.21 - 1.81). This decrease may be explained by the fact that tissue
-samples include matched samples.
+\textbf{\emph{Genera Previously Associated with Carcinoma Predominately
+Increases Relative Risk for Carcinoma but not Adenoma:}} The majority of
+CRC associated genera for both feces and tissue had a significantly
+increased relative risk for carcinoma but not for adenoma {[}Figure
+3{]}. In fecal samples the relative risk due to CRC associated genera
+was greater than either evenness or Shannon diversity {[}Figure 2 \&
+3{]}. Additionally, the relative risk of carcinoma continuously
+increased as individuals tested positive for more CRC associated genera
+{[}Figure 3B \& 3D{]}. The relative risk effect size was greater for
+stool (RR range = 1.61 - 2.74) then for tissue (RR range = 1.21 - 1.81).
+This decrease may be explained by the fact that the tissue analysis
+include matched samples.
 
@@
 Associated Genera:}} For both fecal and tissue samples (matched and
-unmatched) there was a decrease in AUC when only OTUs from the CRC
-associated genera are used {[}Figure 4 \& 5{]}. This decrease is
-observed in both adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The
-genus models generally had similar trends as observed for the OTU based
-models with the full genera models performing better then the CRC
-associated genera models {[}Figure S2-S3{]}. Both genus models perform
-similarily in their ability to be able to predict lesion (adenoma or
-carcinoma) with carcinoma having a higher AUC then adenoma {[}Figure
-S4-S5{]}. Matched tissue samples for those with carcinoma had an AUC
-that was more similar to the adenoma models {[}Figure S4A, S5B, \& S6{]}
-then carcinoma models {[}Figure S4B \& S5A{]}.
+unmatched) the AUC decreases when only OTUs from the CRC associated
+genera are used {[}Figure 4 \& 5{]}. This decrease is observed in both
+adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The genus models
+generally had similar trends as observed for the OTU based models with
+the full genera models performing better then the CRC associated genera
+models {[}Figure S2-S3{]}. Both genus models perform similarily in their
+ability to be able to predict lesion (adenoma or carcinoma) with
+carcinoma having a higher AUC then adenoma {[}Figure S4-S5{]}. Matched
+tissue samples for those with carcinoma had an AUC that was more similar
+to the adenoma models {[}Figure S4A, S5B, \& S6{]} then carcinoma models
+{[}Figure S4B \& S5A{]}.
 
@@
 B{]}. No single study that was analyzed had the standard 80\% power to
-detect a difference that was eqaul to or below 10\% {[}Figure 6A \&
-B{]}. In order to achieve adequate power for small effect sizes it would
-be necessary to recruit over 1000 individuals for each arm of the study
-{[}Figure 6C{]}.
+detect an effect size difference that was eqaul to or below 10\%
+{[}Figure 6A \& B{]}. In order to achieve adequate power for small
+effect sizes it would be necessary to recruit over 1000 individuals for
+each arm of the study {[}Figure 6C{]}.
 

### 2017-11-07 submission/manuscript.Rmd
discussion edits

@@
 
-Our study identifies clear diversity changes both at the community level and within individual genera that are present in indivdiuals with carcinoma versus those without the disease.
+Our study identifies clear diversity changes both at the community level and within individual genera that are present in individuals with carcinoma versus those without the disease.
 Although there was a step wise decrease in diversity from control to adenoma to carcinoma; this did not translate into large effect sizes for the relative risk of either of these two conditions.
-Even though CRC associated genera increases the relative risk of carcinoma it does not increase the relative risk of adenoma.
-This information suggests that these specific genera may not be the primary members of the microbial community that contributes to the formation of an adenoma but is for carcinoma.
+Even though CRC associated genera increases the relative risk of carcinoma it does not consistently increase the relative risk of adenoma.
+This information suggests that these specific genera are important in carcinoma genesis but may not be the primary members of the microbial community that contribute to the formation of an adenoma.
 Additionally, our data shows that by using the whole community our models perform better then when they only use the CRC associated genera.
@@
 The driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], can be supported with this data for carcinoma but not necessarily for adenoma.
-The drasitically increased relative risk of disease when considering the CRC associated genera is highly supportive of this type of process, especially in the context of increasing relative risk with more CRC associated genera positivity.
-It is also possible that in a driver-passenger scenario it is possible that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring.
+The drasitically increased relative risk of disease, when considering the CRC associated genera, is highly supportive of this type of process, especially in the context of increasing relative risk with more CRC associated genera positivity.
+It is also possible that in a driver-passenger scenario that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring.
 This would account for the observation that there is no constant additive effect on relative risk for increasing positivity.
@@
 Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma.
-In contrast, with the present data we can only suggest that the adenoma observations might fit with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma or specific location.
+In contrast, with the present data we can only suggest that the adenoma observations might fit with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma.
 The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle.
-Although there may be localized changes that do depend on the driver-passenger model, supported by the one, two, and four positive CRC associated genera in tissue relative risk increases [Figure 3C], there may be another process involved that ultimately exacerbates the condition from a subtle localized change to a global community one.
+Although there may be localized changes that do depend on the driver-passenger model, supported by an increased relative risk for one, two, and four positive CRC associated genera in tissue [Figure 3C], there may be other processes involved that ultimately exacerbates the condition from a subtle localized change to a global community one.
 The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case.
@@
 
-By aggregating together a large collection of studies from both feces and tissue we are able to provide information in support of the driver-passenger model in the context of carcinoma.
+By aggregating together a large collection of studies from both feces and tissue we are able to provide information in support of the importance of the bacterial community in both adenoma and carcinoma.
+We are also able to provide support for the driver-passenger model in the context of carcinoma.
 However, within the context of adenoma it is less clear that this relationship exists.

### 2017-11-07 submission/manuscript.tex
discussion edits

@@
 Our study identifies clear diversity changes both at the community level
-and within individual genera that are present in indivdiuals with
+and within individual genera that are present in individuals with
 carcinoma versus those without the disease. Although there was a step
@@
 either of these two conditions. Even though CRC associated genera
-increases the relative risk of carcinoma it does not increase the
-relative risk of adenoma. This information suggests that these specific
-genera may not be the primary members of the microbial community that
-contributes to the formation of an adenoma but is for carcinoma.
-Additionally, our data shows that by using the whole community our
-models perform better then when they only use the CRC associated genera.
-CRC associated genera are clearly important to carcinoma but the context
-or community in which these microbes are a part of can drastically
-increase the ability of models to make predictions. This data supports
-the concept that small localized changes within the community may be
-occuring that are important in the disease progression of colorectal
-cancer and that they may not directly involve CRC associated genera.
+increases the relative risk of carcinoma it does not consistently
+increase the relative risk of adenoma. This information suggests that
+these specific genera are important in carcinoma genesis but may not be
+the primary members of the microbial community that contribute to the
+formation of an adenoma. Additionally, our data shows that by using the
+whole community our models perform better then when they only use the
+CRC associated genera. CRC associated genera are clearly important to
+carcinoma but the context or community in which these microbes are a
+part of can drastically increase the ability of models to make
+predictions. This data supports the concept that small localized changes
+within the community may be occuring that are important in the disease
+progression of colorectal cancer and that they may not directly involve
+CRC associated genera.
 
@@
 necessarily for adenoma. The drasitically increased relative risk of
-disease when considering the CRC associated genera is highly supportive
-of this type of process, especially in the context of increasing
-relative risk with more CRC associated genera positivity. It is also
-possible that in a driver-passenger scenario it is possible that simply
-having the driver present or only identifying the passenger is a good
-enough proxy that the event is occuring. This would account for the
-observation that there is no constant additive effect on relative risk
-for increasing positivity. Additionally, the initial establishment of
-the driver within the system is also dependent on the community that is
+disease, when considering the CRC associated genera, is highly
+supportive of this type of process, especially in the context of
+increasing relative risk with more CRC associated genera positivity. It
+is also possible that in a driver-passenger scenario that simply having
+the driver present or only identifying the passenger is a good enough
+proxy that the event is occuring. This would account for the observation
+that there is no constant additive effect on relative risk for
+increasing positivity. Additionally, the initial establishment of the
+driver within the system is also dependent on the community that is
 present and this is supported by the observation that when adding the
@@
 the adenoma observations might fit with this model but the changes that
-occur at this timepoint are small and possibly focal to the adenoma or
-specific location. The stepwise decrease in diversity suggests that the
-adenoma community is not normal but this change is subtle. Although
-there may be localized changes that do depend on the driver-passenger
-model, supported by the one, two, and four positive CRC associated
-genera in tissue relative risk increases {[}Figure 3C{]}, there may be
-another process involved that ultimately exacerbates the condition from
-a subtle localized change to a global community one. The poor
-performance of the Random Forest models for classifying adenoma based
-only on the microbiota would suggest that this is the case. It is
-possible to hypothesize that at early stages of the diease, how the host
-interacts to these subtle changes could be the catlyst that ultimately
-leads to this larger global dysfunctional community.
+occur at this timepoint are small and possibly focal to the adenoma. The
+stepwise decrease in diversity suggests that the adenoma community is
+not normal but this change is subtle. Although there may be localized
+changes that do depend on the driver-passenger model, supported by an
+increased relative risk for one, two, and four positive CRC associated
+genera in tissue {[}Figure 3C{]}, there may be other processes involved
+that ultimately exacerbates the condition from a subtle localized change
+to a global community one. The poor performance of the Random Forest
+models for classifying adenoma based only on the microbiota would
+suggest that this is the case. It is possible to hypothesize that at
+early stages of the diease, how the host interacts to these subtle
+changes could be the catlyst that ultimately leads to this larger global
+dysfunctional community.
 
@@
 and tissue we are able to provide information in support of the
-driver-passenger model in the context of carcinoma. However, within the
-context of adenoma it is less clear that this relationship exists. These
-observations highlight the importance of power and sample number
-considerations when considering investigations into the microbiota and
-adenoma due to the subtle changes in the community.
+importance of the bacterial community in both adenoma and carcinoma. We
+are also able to provide support for the driver-passenger model in the
+context of carcinoma. However, within the context of adenoma it is less
+clear that this relationship exists. These observations highlight the
+importance of power and sample number considerations when considering
+investigations into the microbiota and adenoma due to the subtle changes
+in the community.
 

### 2017-11-07 submission/manuscript.Rmd
method edits

@@
 [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016].
+Additional studies not mentioned in the reviews were obtained from the authors knowledge of the literature.
 All studies were included that used tissue or feces as their sample source for 16S rRNA gene sequencing analysis.
 Studies using either 454 or Illumina sequencing technology were included.
-Only data sets that had the raw sequences available for analysis were included.
-Some studies did not have publically available raw sequences or did not have meta data in which the authors were able to share.
-After this filtering step the following studies remained: Ahn [@ahn_human_2013], Baxter [@baxter_microbiota-based_2016], Brim [@brim_microbiome_2013], Burns [@burns_virulence_2015], Chen [@chen_human_2012], Dejea [@dejea_microbiota_2014], Flemer [@flemer_tumour-associated_2017], Geng [@geng_diversified_2013], Hale [@hale_shifts_2017], Kostic [@kostic_genomic_2012], Lu [@lu_mucosal_2016], Sanapareddy [@sanapareddy_increased_2012], Wang [@wang_structural_2012], Weir [@weir_stool_2013], and Zeller [@zeller_potential_2014].
+Only data sets that had sequences available for analysis were included.
+Some studies did not have publically available sequences or did not have meta data in which the authors were able to share.
+After these filtering steps the following studies remained: Ahn [@ahn_human_2013], Baxter [@baxter_microbiota-based_2016], Brim [@brim_microbiome_2013], Burns [@burns_virulence_2015], Chen [@chen_human_2012], Dejea [@dejea_microbiota_2014], Flemer [@flemer_tumour-associated_2017], Geng [@geng_diversified_2013], Hale [@hale_shifts_2017], Kostic [@kostic_genomic_2012], Lu [@lu_mucosal_2016], Sanapareddy [@sanapareddy_increased_2012], Wang [@wang_structural_2012], Weir [@weir_stool_2013], and Zeller [@zeller_potential_2014].
 The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study.
 The Kostic study was not used since after sequence processing all the case samples did not have more than 100 sequences remaining.
-This left a total of 13 studies in which complete analysis could be completed.
+This left a total of 14 studies in which complete analysis could be completed.
 
@@
 For the majority of studies raw sequences were downloaded from the SRA (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the following website: http://www.ncbi.nlm.nih.gov/Traces/study/ by searching the respective accession number of the study.
-Of the studies that did not have sequences and meta data on the SRA one study had the data stored on DBGap [@ahn_human_2013] and four studies the data was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013].
+Of the studies that did not have sequences and meta data on the SRA one study had the data stored on DBGap [@ahn_human_2013] and for four studies the data was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013].
 Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009].
@@
 For the alpha diversity analysis values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package.
-Testing for alpha diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for both study and variable region effect in the diversity measures when analyzing colorectal cancer groups.
+Testing for alpha diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue.
 Relative Risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages.
@@
 ***Study Analysis Overview:***
-Alpha diversity was first assessed for differences between controls and adenoma versus cancer and controls versus adenoma.
-We analyzed the data using linear mixed-effect models, and relative risk.
+Alpha diversity was first assessed for differences between controls, adenoma, and carcinoma.
+We analyzed the data using linear mixed-effect models and relative risk.
 Beta-diversity was then assessed for each inidividual study.
@@
 Within the tissue groups the data was further divided between matched and unmatched tissue samples.
-Both prediction for adenoma and carcinoma were tested.
+Where applicable for each study predictions for adenoma and carcinoma were tested.
 This same approach was then applied at the OTU level with the exception that instead of testing on the other studies a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs.
-For OTU based models the CRC Associated Genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
-The power of each study was assessed for and effect size ranging from 1% to 30%.
+For OTU based models the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
+The power of each study was assessed for an effect size ranging from 1% to 30%.
 An estimated sample n for these effect sizes was also generated based on 80% power.

### 2017-11-07 submission/manuscript.tex
method edits

@@
 were identified through the review articles written by Keku, et al. and
-Vogtmann, et al. {[}26,27{]}. All studies were included that used tissue
-or feces as their sample source for 16S rRNA gene sequencing analysis.
-Studies using either 454 or Illumina sequencing technology were
-included. Only data sets that had the raw sequences available for
-analysis were included. Some studies did not have publically available
-raw sequences or did not have meta data in which the authors were able
-to share. After this filtering step the following studies remained: Ahn
-{[}21{]}, Baxter {[}24{]}, Brim {[}17{]}, Burns {[}22{]}, Chen {[}14{]},
-Dejea {[}19{]}, Flemer {[}13{]}, Geng {[}25{]}, Hale {[}12{]}, Kostic
-{[}28{]}, Lu {[}16{]}, Sanapareddy {[}20{]}, Wang {[}15{]}, Weir
-{[}18{]}, and Zeller {[}23{]}. The Zackular {[}29{]} study was not
-included becasue the 90 individuals analyzed within the study are
-contained within the larger Baxter study. The Kostic study was not used
-since after sequence processing all the case samples did not have more
-than 100 sequences remaining. This left a total of 13 studies in which
-complete analysis could be completed.
+Vogtmann, et al. {[}26,27{]}. Additional studies not mentioned in the
+reviews were obtained from the authors knowledge of the literature. All
+studies were included that used tissue or feces as their sample source
+for 16S rRNA gene sequencing analysis. Studies using either 454 or
+Illumina sequencing technology were included. Only data sets that had
+sequences available for analysis were included. Some studies did not
+have publically available sequences or did not have meta data in which
+the authors were able to share. After these filtering steps the
+following studies remained: Ahn {[}21{]}, Baxter {[}24{]}, Brim
+{[}17{]}, Burns {[}22{]}, Chen {[}14{]}, Dejea {[}19{]}, Flemer
+{[}13{]}, Geng {[}25{]}, Hale {[}12{]}, Kostic {[}28{]}, Lu {[}16{]},
+Sanapareddy {[}20{]}, Wang {[}15{]}, Weir {[}18{]}, and Zeller {[}23{]}.
+The Zackular {[}29{]} study was not included becasue the 90 individuals
+analyzed within the study are contained within the larger Baxter study.
+The Kostic study was not used since after sequence processing all the
+case samples did not have more than 100 sequences remaining. This left a
+total of 14 studies in which complete analysis could be completed.
 
@@
 have sequences and meta data on the SRA one study had the data stored on
-DBGap {[}21{]} and four studies the data was obtained directly from the
-authors {[}12,13,18,20{]}. Each study was processed using the mothur
+DBGap {[}21{]} and for four studies the data was obtained directly from
+the authors {[}12,13,18,20{]}. Each study was processed using the mothur
 (v1.39.3) software program {[}30{]}. Where possible quality filtering
@@
 linear mixed-effect models created using the lme4 (v1.1.14) package to
-correct for both study and variable region effect in the diversity
-measures when analyzing colorectal cancer groups. Relative Risk was
+correct for study and variable region effects in feces and study,
+variable region, and individual effects in tissue. Relative Risk was
 analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages.
@@
 \textbf{\emph{Study Analysis Overview:}} Alpha diversity was first
-assessed for differences between controls and adenoma versus cancer and
-controls versus adenoma. We analyzed the data using linear mixed-effect
-models, and relative risk. Beta-diversity was then assessed for each
-inidividual study. Next, four specific CRC-associated genera
-(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, and
-\emph{Porphyromonas}) were assessed for differences in relative risk. We
-then built Random Forest models based on all genera or the select
-CRC-associated genera. The models were trained on one study then tested
-on the remaining studies for every study. The data was split between
-feces and tissue samples. Within the tissue groups the data was further
-divided between matched and unmatched tissue samples. Both prediction
-for adenoma and carcinoma were tested. This same approach was then
-applied at the OTU level with the exception that instead of testing on
-the other studies a 10-fold cross validation was utilized and 100
-different models were created based on random 80/20 splitting of the
-data to generate a range of expected AUCs. For OTU based models the CRC
-Associated Genera included all OTUs that had a taxonomic classification
-to \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus},
-or \emph{Porphyromonas}. The power of each study was assessed for and
-effect size ranging from 1\% to 30\%. An estimated sample n for these
-effect sizes was also generated based on 80\% power.
+assessed for differences between controls, adenoma, and carcinoma. We
+analyzed the data using linear mixed-effect models and relative risk.
+Beta-diversity was then assessed for each inidividual study. Next, four
+specific CRC-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
+differences in relative risk. We then built Random Forest models based
+on all genera or the select CRC-associated genera. The models were
+trained on one study then tested on the remaining studies for every
+study. The data was split between feces and tissue samples. Within the
+tissue groups the data was further divided between matched and unmatched
+tissue samples. Where applicable for each study predictions for adenoma
+and carcinoma were tested. This same approach was then applied at the
+OTU level with the exception that instead of testing on the other
+studies a 10-fold cross validation was utilized and 100 different models
+were created based on random 80/20 splitting of the data to generate a
+range of expected AUCs. For OTU based models the CRC associated genera
+included all OTUs that had a taxonomic classification to
+\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
+\emph{Porphyromonas}. The power of each study was assessed for an effect
+size ranging from 1\% to 30\%. An estimated sample n for these effect
+sizes was also generated based on 80\% power.
 

### 2017-11-07 submission/manuscript.Rmd
small fix

@@
 However, by using the full community instead of only the CRC associated genera the AUC of the classification models increased.
-Although we analyze a data set with a large number of total individuals each individual study was underpowered for effect size differences of 10% or below between the case and control.
+Although we analyze a data set with a large number of total individuals each individual study was underpowered for effect size differences of 10% or below between the case and control group.
 

### 2017-11-07 submission/manuscript.tex
small fix

@@
 individuals each individual study was underpowered for effect size
-differences of 10\% or below between the case and control.
+differences of 10\% or below between the case and control group.
 

### 2017-11-14 submission/manuscript.Rmd
incorporate Will's method suggestions

@@
 ***Diversity is Lower in Those with Carcinoma and Increases Relative Risk:***
-Using power transformed and Z-score normalized alpha diversity metrics, both evenness and Shannon diversity in feces and not tissue were  lower in those with carcinoma [Figure 1].
+Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces and not tissue were  lower in those with carcinoma [Figure 1].
 Using linear mixed-effect models to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`).
@@
 ***Obtaining Data Sets:***
-Studies used for this meta-analysis were identified through the review articles written by Keku, et al.
-and Vogtmann, et al.
-[@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016].
-Additional studies not mentioned in the reviews were obtained from the authors knowledge of the literature.
-All studies were included that used tissue or feces as their sample source for 16S rRNA gene sequencing analysis.
+Studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016].
+Additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature.
+Studies that used tissue or feces as their sample source for 16S rRNA gene sequencing analysis were included.
 Studies using either 454 or Illumina sequencing technology were included.
 Only data sets that had sequences available for analysis were included.
-Some studies did not have publically available sequences or did not have meta data in which the authors were able to share.
-After these filtering steps the following studies remained: Ahn [@ahn_human_2013], Baxter [@baxter_microbiota-based_2016], Brim [@brim_microbiome_2013], Burns [@burns_virulence_2015], Chen [@chen_human_2012], Dejea [@dejea_microbiota_2014], Flemer [@flemer_tumour-associated_2017], Geng [@geng_diversified_2013], Hale [@hale_shifts_2017], Kostic [@kostic_genomic_2012], Lu [@lu_mucosal_2016], Sanapareddy [@sanapareddy_increased_2012], Wang [@wang_structural_2012], Weir [@weir_stool_2013], and Zeller [@zeller_potential_2014].
-The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study.
-The Kostic study was not used since after sequence processing all the case samples did not have more than 100 sequences remaining.
-This left a total of 14 studies in which complete analysis could be completed.
+Some studies did not have publically available sequences or did not have metadata in which the authors were able to share.
+After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014].
+The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016].
+Additionally, after sequence processing all the case samples for the Kostic study only had 100 sequences remaining and was not used.
+This left a total of 14 studies for which analysis could be completed.
 
 ***Data Set Breakdown:***
-In total there were 7 studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), 5 studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and 2 studies with both fecal and tissue samples (Chen and Flemer).
+In total, there were seven studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal and tissue samples (Chen and Flemer).
 The total number of individuals that were analyzed after sequence processing for feces was `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` [Table 1].
@@
 ***Sequence Processing:***
-For the majority of studies raw sequences were downloaded from the SRA (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the following website: http://www.ncbi.nlm.nih.gov/Traces/study/ by searching the respective accession number of the study.
-Of the studies that did not have sequences and meta data on the SRA one study had the data stored on DBGap [@ahn_human_2013] and for four studies the data was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013].
+For the majority of studies raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the by searching the respective accession number of the study following website: http://www.ncbi.nlm.nih.gov/Traces/study/.
+Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap for one study [@ahn_human_2013] and for four studies was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013].
 Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009].
-Where possible quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing.
-If it was not possible to use these defaults the author stated quality cut-offs were used instead.
-Chimeras were identifed and removed using the VSEARCH [@rognes_vsearch_2016] program and *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.
+Where possible, quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing.
+If it was not possible to use these defaults, the stated quality cut-offs were used instead.
+Chimeras were identifed and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.
 
@@
 All statistical analysis after sequence processing utilized the R software package (v`r paste(version$major, ".", version$minor, sep = "")`).
-For the alpha diversity analysis values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package.
-Testing for alpha diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue.
-Relative Risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages.
+For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package.
+Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue.
+Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages.
 Relative risk significance testing utilized the chi-squred test.
-Beta-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package.
+$\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package.
 Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages.
-Random Forest testing of the obtained AUC versus a random model AUC utilized T-tests.
+Differences between the obtained AUC versus a random model AUC was assessed using T-tests.
 Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages.
@@
 ***Study Analysis Overview:***
-Alpha diversity was first assessed for differences between controls, adenoma, and carcinoma.
+$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma.
 We analyzed the data using linear mixed-effect models and relative risk.
-Beta-diversity was then assessed for each inidividual study.
+$\beta$-diversity was then assessed for each inidividual study.
 Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk.
@@
 Within the tissue groups the data was further divided between matched and unmatched tissue samples.
-Where applicable for each study predictions for adenoma and carcinoma were tested.
-This same approach was then applied at the OTU level with the exception that instead of testing on the other studies a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs.
+Where applicable for each study, predictions for adenoma and carcinoma were tested.
+This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs.
 For OTU based models the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
@@
 The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
-Unless mentioned otherwise the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file, on the GitHub repository or in the original manuscript.
+Unless mentioned otherwise, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.
 
@@
 
-**Figure 1: Alpha Diversity Differences between Control, Adenoma, and Carcinoma Across Sampling Site.** A) Alpha diversity metric differences by group in stool samples.
-B) Alpha diversity metric differences by group in unmatched tissue samples.
-C) Alpha diversity metric differences by group in matched tissue samples.
+**Figure 1: $\alpha$-Diversity Differences between Control, Adenoma, and Carcinoma Across Sampling Site.** A) $\alpha$-diversity metric differences by group in stool samples.
+B) $\alpha$-diversity metric differences by group in unmatched tissue samples.
+C) $\alpha$-diversity metric differences by group in matched tissue samples.
 The dashed line represents a Z-score of 0 or no difference from the median.
 
-**Figure 2: Relative Risk for Adenoma or Carcinoma based on Alpha Diversity Metrics in Stool.** A) Alpha metric relative risk for adenoma.
-B) Alpha metric relative risk for carcinoma.
+**Figure 2: Relative Risk for Adenoma or Carcinoma based on $\alpha$-Diversity Metrics in Stool.** A) $\alpha$-metric relative risk for adenoma.
+B) $\alpha$-metric relative risk for carcinoma.
 Colors represent the different variable regions used within the respective study.
@@
 
-**Figure S1: Relative Risk for Adenoma or Carcinoma based on Alpha Diversity Metrics in Tissue.** A) Alpha metric relative risk for adenoma.
-B) Alpha metric relative risk for carcinoma.
+**Figure S1: Relative Risk for Adenoma or Carcinoma based on $\alpha$-Diversity Metrics in Tissue.** A) $\alpha$-metric relative risk for adenoma.
+B) $\alpha$-metric relative risk for carcinoma.
 Colors represent the different variable regions used within the respective study.

### 2017-11-14 submission/manuscript.tex
incorporate Will's method suggestions

@@
 \usepackage{pdfpages}
+
+\usepackage{amsmath}
+
+\usepackage{mathtools}
 
@@
 \textbf{\emph{Diversity is Lower in Those with Carcinoma and Increases
-Relative Risk:}} Using power transformed and Z-score normalized alpha
-diversity metrics, both evenness and Shannon diversity in feces and not
-tissue were lower in those with carcinoma {[}Figure 1{]}. Using linear
-mixed-effect models to control for study and variable region there was a
-significant decrease from control to adenoma to carcinoma for both
-evenness (P-value = 0.025) and Shannon diversity (P-value = 0.043). This
-effect was not observed in tissue when additionally controlling for
-whether the sample came from the same individual (P-value \textgreater{}
-0.05). For fecal samples a decrease in Shannon diversity and evenness
-resulted in a significantly increased relative risk for carcinoma
-(P-value = 0.01 and P-value = 0.0011, respectively) {[}Figure 2{]}.
-Although these values were significant, the effect size was relatively
-small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
-{[}Figure 2{]}. Only a decrease in evenness had an increased relative
-risk for adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect
-size was even smaller than what was observed for carcinoma (RR = 1.16).
+Relative Risk:}} Using power transformed and Z-score normalized
+\(\alpha\)-diversity metrics, both evenness and Shannon diversity in
+feces and not tissue were lower in those with carcinoma {[}Figure 1{]}.
+Using linear mixed-effect models to control for study and variable
+region there was a significant decrease from control to adenoma to
+carcinoma for both evenness (P-value = 0.025) and Shannon diversity
+(P-value = 0.043). This effect was not observed in tissue when
+additionally controlling for whether the sample came from the same
+individual (P-value \textgreater{} 0.05). For fecal samples a decrease
+in Shannon diversity and evenness resulted in a significantly increased
+relative risk for carcinoma (P-value = 0.01 and P-value = 0.0011,
+respectively) {[}Figure 2{]}. Although these values were significant,
+the effect size was relatively small for both metrics (Shannon RR = 1.31
+and evenness RR = 1.34) {[}Figure 2{]}. Only a decrease in evenness had
+an increased relative risk for adenoma (P-value = 0.032) {[}Figure 2A \&
+S1{]} but this effect size was even smaller than what was observed for
+carcinoma (RR = 1.16).
 
@@
 \textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
-were identified through the review articles written by Keku, et al. and
-Vogtmann, et al. {[}26,27{]}. Additional studies not mentioned in the
-reviews were obtained from the authors knowledge of the literature. All
-studies were included that used tissue or feces as their sample source
-for 16S rRNA gene sequencing analysis. Studies using either 454 or
-Illumina sequencing technology were included. Only data sets that had
-sequences available for analysis were included. Some studies did not
-have publically available sequences or did not have meta data in which
-the authors were able to share. After these filtering steps the
-following studies remained: Ahn {[}21{]}, Baxter {[}24{]}, Brim
-{[}17{]}, Burns {[}22{]}, Chen {[}14{]}, Dejea {[}19{]}, Flemer
-{[}13{]}, Geng {[}25{]}, Hale {[}12{]}, Kostic {[}28{]}, Lu {[}16{]},
-Sanapareddy {[}20{]}, Wang {[}15{]}, Weir {[}18{]}, and Zeller {[}23{]}.
-The Zackular {[}29{]} study was not included becasue the 90 individuals
-analyzed within the study are contained within the larger Baxter study.
-The Kostic study was not used since after sequence processing all the
-case samples did not have more than 100 sequences remaining. This left a
-total of 14 studies in which complete analysis could be completed.
+were identified through the review articles written by Keku, \emph{et
+al.} and Vogtmann, \emph{et al.} {[}26,27{]}. Additional studies not
+mentioned in the reviews were obtained based on the authors' knowledge
+of the literature. Studies that used tissue or feces as their sample
+source for 16S rRNA gene sequencing analysis were included. Studies
+using either 454 or Illumina sequencing technology were included. Only
+data sets that had sequences available for analysis were included. Some
+studies did not have publically available sequences or did not have
+metadata in which the authors were able to share. After these filtering
+steps, the following studies remained: Ahn, \emph{et al.} {[}21{]},
+Baxter, \emph{et al.} {[}24{]}, Brim, \emph{et al.} {[}17{]}, Burns,
+\emph{et al.} {[}22{]}, Chen, \emph{et al.} {[}14{]}, Dejea, \emph{et
+al.} {[}19{]}, Flemer, \emph{et al.} {[}13{]}, Geng, \emph{et al.}
+{[}25{]}, Hale, \emph{et al.} {[}12{]}, Kostic, \emph{et al.} {[}28{]},
+Lu, \emph{et al.} {[}16{]}, Sanapareddy, \emph{et al.} {[}20{]}, Wang,
+\emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}18{]}, and Zeller,
+\emph{et al.} {[}23{]}. The Zackular {[}29{]} study was not included
+becasue the 90 individuals analyzed within the study are contained
+within the larger Baxter study {[}24{]}. Additionally, after sequence
+processing all the case samples for the Kostic study only had 100
+sequences remaining and was not used. This left a total of 14 studies
+for which analysis could be completed.
 
-\textbf{\emph{Data Set Breakdown:}} In total there were 7 studies with
-only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), 5
-studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy),
-and 2 studies with both fecal and tissue samples (Chen and Flemer). The
-total number of individuals that were analyzed after sequence processing
-for feces was 1737 {[}Table 1{]}. The total number of matched and
-unmatched tissue samples that were analyzed after sequence processing
-was 492 {[}Table 2{]}.
+\textbf{\emph{Data Set Breakdown:}} In total, there were seven studies
+with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and
+Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu,
+Sanapareddy), and two studies with both fecal and tissue samples (Chen
+and Flemer). The total number of individuals that were analyzed after
+sequence processing for feces was 1737 {[}Table 1{]}. The total number
+of matched and unmatched tissue samples that were analyzed after
+sequence processing was 492 {[}Table 2{]}.
 
 \textbf{\emph{Sequence Processing:}} For the majority of studies raw
-sequences were downloaded from the SRA
+sequences were downloaded from the Sequence Read Archive (SRA)
 (\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
-and metadata was obtained from the following website:
-\url{http://www.ncbi.nlm.nih.gov/Traces/study/} by searching the
-respective accession number of the study. Of the studies that did not
-have sequences and meta data on the SRA one study had the data stored on
-DBGap {[}21{]} and for four studies the data was obtained directly from
+and metadata was obtained from the by searching the respective accession
+number of the study following website:
+\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
+not have sequences and metadata on the SRA, data was obtained from DBGap
+for one study {[}21{]} and for four studies was obtained directly from
 the authors {[}12,13,18,20{]}. Each study was processed using the mothur
-(v1.39.3) software program {[}30{]}. Where possible quality filtering
+(v1.39.3) software program {[}30{]}. Where possible, quality filtering
 utilized the default methods used in mothur for either 454 or Illumina
-based sequencing. If it was not possible to use these defaults the
-author stated quality cut-offs were used instead. Chimeras were
-identifed and removed using the VSEARCH {[}31{]} program and \emph{de
-novo} OTU clustering at 97\% similarity using the OptiClust algorithm
-{[}32{]} was utilized.
+based sequencing. If it was not possible to use these defaults, the
+stated quality cut-offs were used instead. Chimeras were identifed and
+removed using VSEARCH {[}31{]} before \emph{de novo} OTU clustering at
+97\% similarity using the OptiClust algorithm {[}32{]} was utilized.
 
@@
 sequence processing utilized the R software package (v3.4.2). For the
-alpha diversity analysis values were power transformed using the
+\(\alpha\)-diversity analysis, values were power transformed using the
 rcompanion (v1.10.1) package and then Z-score normalized using the car
-(v2.1.5) package. Testing for alpha diversity differences utilized
+(v2.1.5) package. Testing for \(\alpha\)-diversity differences utilized
 linear mixed-effect models created using the lme4 (v1.1.14) package to
 correct for study and variable region effects in feces and study,
-variable region, and individual effects in tissue. Relative Risk was
+variable region, and individual effects in tissue. Relative risk was
 analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages.
 Relative risk significance testing utilized the chi-squred test.
-Beta-diversity differences utilized a Bray-Curtis distance matrix and
-PERMANOVA executed with the vegan (v2.4.4) package. Random Forest models
-were built using both the caret (v6.0.77) and randomForest (v4.6.12)
-packages. Random Forest testing of the obtained AUC versus a random
-model AUC utilized T-tests. Power analysis and estimations were made
-using the pwr (v1.2.1) and statmod (v1.4.30) packages. All figures were
-created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages.
+\(\beta\)-diversity differences utilized a Bray-Curtis distance matrix
+and PERMANOVA executed with the vegan (v2.4.4) package. Random Forest
+models were built using both the caret (v6.0.77) and randomForest
+(v4.6.12) packages. Differences between the obtained AUC versus a random
+model AUC was assessed using T-tests. Power analysis and estimations
+were made using the pwr (v1.2.1) and statmod (v1.4.30) packages. All
+figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
+packages.
 
-\textbf{\emph{Study Analysis Overview:}} Alpha diversity was first
+\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
 assessed for differences between controls, adenoma, and carcinoma. We
 analyzed the data using linear mixed-effect models and relative risk.
-Beta-diversity was then assessed for each inidividual study. Next, four
-specific CRC-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
-differences in relative risk. We then built Random Forest models based
-on all genera or the select CRC-associated genera. The models were
-trained on one study then tested on the remaining studies for every
-study. The data was split between feces and tissue samples. Within the
-tissue groups the data was further divided between matched and unmatched
-tissue samples. Where applicable for each study predictions for adenoma
-and carcinoma were tested. This same approach was then applied at the
-OTU level with the exception that instead of testing on the other
-studies a 10-fold cross validation was utilized and 100 different models
-were created based on random 80/20 splitting of the data to generate a
-range of expected AUCs. For OTU based models the CRC associated genera
-included all OTUs that had a taxonomic classification to
-\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
-\emph{Porphyromonas}. The power of each study was assessed for an effect
-size ranging from 1\% to 30\%. An estimated sample n for these effect
-sizes was also generated based on 80\% power.
+\(\beta\)-diversity was then assessed for each inidividual study. Next,
+four specific CRC-associated genera (\emph{Fusobacterium},
+\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas})
+were assessed for differences in relative risk. We then built Random
+Forest models based on all genera or the select CRC-associated genera.
+The models were trained on one study then tested on the remaining
+studies for every study. The data was split between feces and tissue
+samples. Within the tissue groups the data was further divided between
+matched and unmatched tissue samples. Where applicable for each study,
+predictions for adenoma and carcinoma were tested. This same approach
+was then applied at the OTU level with the exception that instead of
+testing on the other studies, a 10-fold cross validation was utilized
+and 100 different models were created based on random 80/20 splitting of
+the data to generate a range of expected AUCs. For OTU based models the
+CRC associated genera included all OTUs that had a taxonomic
+classification to \emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Peptostreptococcus}, or \emph{Porphyromonas}. The power of each
+study was assessed for an effect size ranging from 1\% to 30\%. An
+estimated sample n for these effect sizes was also generated based on
+80\% power.
 
@@
 \url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
-Unless mentioned otherwise the accession number for the raw sequences
+Unless mentioned otherwise, the accession number for the raw sequences
 for the studies used in this analysis can be found directly in the
-respective batch file, on the GitHub repository or in the original
+respective batch file in the GitHub repository or in the original
 manuscript.
@@
 
-\textbf{Figure 1: Alpha Diversity Differences between Control, Adenoma,
-and Carcinoma Across Sampling Site.} A) Alpha diversity metric
-differences by group in stool samples. B) Alpha diversity metric
-differences by group in unmatched tissue samples. C) Alpha diversity
-metric differences by group in matched tissue samples. The dashed line
-represents a Z-score of 0 or no difference from the median.
+\textbf{Figure 1: \(\alpha\)-Diversity Differences between Control,
+Adenoma, and Carcinoma Across Sampling Site.} A) \(\alpha\)-diversity
+metric differences by group in stool samples. B) \(\alpha\)-diversity
+metric differences by group in unmatched tissue samples. C)
+\(\alpha\)-diversity metric differences by group in matched tissue
+samples. The dashed line represents a Z-score of 0 or no difference from
+the median.
 
-\textbf{Figure 2: Relative Risk for Adenoma or Carcinoma based on Alpha
-Diversity Metrics in Stool.} A) Alpha metric relative risk for adenoma.
-B) Alpha metric relative risk for carcinoma. Colors represent the
-different variable regions used within the respective study.
+\textbf{Figure 2: Relative Risk for Adenoma or Carcinoma based on
+\(\alpha\)-Diversity Metrics in Stool.} A) \(\alpha\)-metric relative
+risk for adenoma. B) \(\alpha\)-metric relative risk for carcinoma.
+Colors represent the different variable regions used within the
+respective study.
 
@@
 
-\textbf{Figure S1: Relative Risk for Adenoma or Carcinoma based on Alpha
-Diversity Metrics in Tissue.} A) Alpha metric relative risk for adenoma.
-B) Alpha metric relative risk for carcinoma. Colors represent the
-different variable regions used within the respective study.
+\textbf{Figure S1: Relative Risk for Adenoma or Carcinoma based on
+\(\alpha\)-Diversity Metrics in Tissue.} A) \(\alpha\)-metric relative
+risk for adenoma. B) \(\alpha\)-metric relative risk for carcinoma.
+Colors represent the different variable regions used within the
+respective study.
 

### 2017-11-14 submission/manuscript.Rmd
add Will's Declaration changes

@@
 
-Ethics approval and informed consent for each of the studies used is mentioned in the respective manuscript used in this meta-analysis.
+Ethics approval and informed consent for each of the studies used is mentioned in the respective manuscripts used in this meta-analysis.
 
@@
 
-A detailed and reporducible description of how the data were processed and analyzed for each study can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
-Raw sequences can be downloaded from the SRA in most cases and can be found in the respective studies batch file in the GitHub repo or within the original publication.
-When sequences were not publicly available contacting the corresponding author for raw sequences needs to be undertaken.
+A detailed and reproducible description of how the data were processed and analyzed for each study can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
+Raw sequences can be downloaded from the SRA in most cases and can be found in the respective studies batch file in the GitHub repository or within the original publication.
+For instances when sequences are not publicly available, they may be accessed by contacting the corresponding authors from whence the data came.
 
@@
 
-MAS is supported by a CIHR fellowship and a University of Michigan PTSP fellowship grant.
+MAS is supported by a Candian Institute of Health Research fellowship and a University of Michigan Postdoctoral Translational Scholar Program grant.
 
@@
 
-The authors would like to thank all the study participants who were apart of each of the individual studies uitlized.
+The authors would like to thank all the study participants who were a part of each of the individual studies uitlized.
 We would also like to thank each of the study authors for making their data available for use.

### 2017-11-14 submission/manuscript.tex
add Will's Declaration changes

@@
 Ethics approval and informed consent for each of the studies used is
-mentioned in the respective manuscript used in this meta-analysis.
+mentioned in the respective manuscripts used in this meta-analysis.
 
@@
 
-A detailed and reporducible description of how the data were processed
+A detailed and reproducible description of how the data were processed
 and analyzed for each study can be found at
@@
 Raw sequences can be downloaded from the SRA in most cases and can be
-found in the respective studies batch file in the GitHub repo or within
-the original publication. When sequences were not publicly available
-contacting the corresponding author for raw sequences needs to be
-undertaken.
+found in the respective studies batch file in the GitHub repository or
+within the original publication. For instances when sequences are not
+publicly available, they may be accessed by contacting the corresponding
+authors from whence the data came.
 
@@
 
-MAS is supported by a CIHR fellowship and a University of Michigan PTSP
-fellowship grant.
+MAS is supported by a Candian Institute of Health Research fellowship
+and a University of Michigan Postdoctoral Translational Scholar Program
+grant.
 
@@
 
-The authors would like to thank all the study participants who were
-apart of each of the individual studies uitlized. We would also like to
+The authors would like to thank all the study participants who were a
+part of each of the individual studies uitlized. We would also like to
 thank each of the study authors for making their data available for use.

### 2017-11-14 submission/manuscript.Rmd
add Will's comments on figure legends

@@
 
-**Figure 3: Colorectal Cancer Associated Genera Relative Risk for Adenoma and Carcinoma in Stool and Tissue.** A) Adenoma relative risk in stool.
-B) Carinoma relative risk in stool.
+**Figure 3: CRC-Associated Genera Relative Risk for Adenoma and Carcinoma in Stool and Tissue.** A) Adenoma relative risk in stool.
+B) Carcinoma relative risk in stool.
 C) Adenoma relative risk in tissue.
 D) Carcinoma relative risk in tissue.
-For all panels the relative risk was also compared to whether one, two, three, or four of the CRC associated genera were present.
+For all panels the relative risk was also compared to whether one, two, three, or four of the CRC-associated genera were present.
 
-**Figure 4: OTU Random Forest Model of Stool Across Studies.** A) Adenoma random forest model between the full community and CRC associated genera OTUs only.
-B) Carcinoma random forest model between the full community and CRC associated genera OTUs only.
+**Figure 4: OTU Random Forest Model of Stool Across Studies.** A) Adenoma random forest model between the full community and CRC-associated genera OTUs only.
+B) Carcinoma random forest model between the full community and CRC-associated genera OTUs only.
 The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.
 
-**Figure 5: OTU Random Forest Model of Tissue Across Studies.** A) Adenoma random forest model between the full community and CRC associated genera OTUs only.
-B) Carcinoma random forest model between the full community and CRC associated genera OTUs only.
+**Figure 5: OTU Random Forest Model of Tissue Across Studies.** A) Adenoma random forest model between the full community and CRC-associated genera OTUs only.
+B) Carcinoma random forest model between the full community and CRC-associated genera OTUs only.
 The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.
 
-**Figure 6: Power and Effect Size Analysis of Studies Included.** A) Power based on effect size for studies with Adenoma individuals.
-B) Power based on effect size for studies with Carcinoma individuals.
+**Figure 6: Power and Effect Size Analysis of Studies Included.** A) Power based on effect size for studies with adenoma individuals.
+B) Power based on effect size for studies with carcinoma individuals.
 C) The estimated sample number needed for each arm of each study to detect aneffect size of 1-30%.
-The dotted red lines in A) and B) represent the generally used power of 0.8.
+The dotted red lines in A) and B) represent a power of 0.8.
 
@@
 
-**Figure S2: Random Forest Genus Model AUC for each Stool Study.** A) AUC of Adenoma models using all genera or CRC associated genera only.
-B) AUC of Carcinoma models using all genera or CRC associated genera only.
+**Figure S2: Random Forest Genus Model AUC for each Stool Study.** A) AUC of adenoma models using all genera or CRC-associated genera only.
+B) AUC of carcinoma models using all genera or CRC-associated genera only.
 The black line represents the median within each group.
 
-**Figure S3: Random Forest Genus Model AUC for each Tissue Study.** A) AUC of Adenoma models using all genera or CRC associated genera only divided between matched and unmatched tissue.
-B) AUC of Carcinoma models using all genera or CRC associated genera only.
+**Figure S3: Random Forest Genus Model AUC for each Tissue Study.** A) AUC of adenoma models using all genera or only CRC-associated genera divided between matched and unmatched tissue.
+B) AUC of carcinoma models using all genera or CRC-associated genera only.
 The black line represents the median within each group divided between matched and unmatched tissue.
 
-**Figure S4: Random Forest Prediction Success Using Genera for each Stool Study.** A) AUC for prediction in Adenoma using all genera or CRC associated genera only.
-B) AUC for prediction in Carcinoma using all genera or CRC associated genera only.
+**Figure S4: Random Forest Prediction Success Using Genera for each Stool Study.** A) AUC for prediction in adenoma using all genera or CRC associated genera only.
+B) AUC for prediction in carcinoma using all genera or CRC-associated genera only.
 The dotted line represents an AUC of 0.5.
@@
 
-**Figure S5: Random Forest Prediction Success of Carcinoma Using Genera for each Tissue Study.** A) AUC for prediction in unmatched tissue for all genera or CRC associated genera only.
-B) AUC for prediction in matched tissue using all genera or CRC associated genera only.
+**Figure S5: Random Forest Prediction Success of Carcinoma Using Genera for each Tissue Study.** A) AUC for prediction in unmatched tissue for all genera or CRC-associated genera only.
+B) AUC for prediction in matched tissue using all genera or CRC-associated genera only.
 The dotted line represents an AUC of 0.5.

### 2017-11-14 submission/manuscript.tex
add Will's comments on figure legends

@@
 
-\textbf{Figure 3: Colorectal Cancer Associated Genera Relative Risk for
-Adenoma and Carcinoma in Stool and Tissue.} A) Adenoma relative risk in
-stool. B) Carinoma relative risk in stool. C) Adenoma relative risk in
-tissue. D) Carcinoma relative risk in tissue. For all panels the
-relative risk was also compared to whether one, two, three, or four of
-the CRC associated genera were present.
+\textbf{Figure 3: CRC-Associated Genera Relative Risk for Adenoma and
+Carcinoma in Stool and Tissue.} A) Adenoma relative risk in stool. B)
+Carcinoma relative risk in stool. C) Adenoma relative risk in tissue. D)
+Carcinoma relative risk in tissue. For all panels the relative risk was
+also compared to whether one, two, three, or four of the CRC-associated
+genera were present.
 
 \textbf{Figure 4: OTU Random Forest Model of Stool Across Studies.} A)
-Adenoma random forest model between the full community and CRC
-associated genera OTUs only. B) Carcinoma random forest model between
-the full community and CRC associated genera OTUs only. The dotted line
-represents an AUC of 0.5 and the lines represent the range in which the
-AUC for the 100 different 80/20 runs fell between.
+Adenoma random forest model between the full community and
+CRC-associated genera OTUs only. B) Carcinoma random forest model
+between the full community and CRC-associated genera OTUs only. The
+dotted line represents an AUC of 0.5 and the lines represent the range
+in which the AUC for the 100 different 80/20 runs fell between.
 
 \textbf{Figure 5: OTU Random Forest Model of Tissue Across Studies.} A)
-Adenoma random forest model between the full community and CRC
-associated genera OTUs only. B) Carcinoma random forest model between
-the full community and CRC associated genera OTUs only. The dotted line
-represents an AUC of 0.5 and the lines represent the range in which the
-AUC for the 100 different 80/20 runs fell between.
+Adenoma random forest model between the full community and
+CRC-associated genera OTUs only. B) Carcinoma random forest model
+between the full community and CRC-associated genera OTUs only. The
+dotted line represents an AUC of 0.5 and the lines represent the range
+in which the AUC for the 100 different 80/20 runs fell between.
 
 \textbf{Figure 6: Power and Effect Size Analysis of Studies Included.}
-A) Power based on effect size for studies with Adenoma individuals. B)
-Power based on effect size for studies with Carcinoma individuals. C)
+A) Power based on effect size for studies with adenoma individuals. B)
+Power based on effect size for studies with carcinoma individuals. C)
 The estimated sample number needed for each arm of each study to detect
-aneffect size of 1-30\%. The dotted red lines in A) and B) represent the
-generally used power of 0.8.
+aneffect size of 1-30\%. The dotted red lines in A) and B) represent a
+power of 0.8.
 
@@
 \textbf{Figure S2: Random Forest Genus Model AUC for each Stool Study.}
-A) AUC of Adenoma models using all genera or CRC associated genera only.
-B) AUC of Carcinoma models using all genera or CRC associated genera
+A) AUC of adenoma models using all genera or CRC-associated genera only.
+B) AUC of carcinoma models using all genera or CRC-associated genera
 only. The black line represents the median within each group.
@@
 \textbf{Figure S3: Random Forest Genus Model AUC for each Tissue Study.}
-A) AUC of Adenoma models using all genera or CRC associated genera only
-divided between matched and unmatched tissue. B) AUC of Carcinoma models
-using all genera or CRC associated genera only. The black line
+A) AUC of adenoma models using all genera or only CRC-associated genera
+divided between matched and unmatched tissue. B) AUC of carcinoma models
+using all genera or CRC-associated genera only. The black line
 represents the median within each group divided between matched and
@@
 \textbf{Figure S4: Random Forest Prediction Success Using Genera for
-each Stool Study.} A) AUC for prediction in Adenoma using all genera or
-CRC associated genera only. B) AUC for prediction in Carcinoma using all
-genera or CRC associated genera only. The dotted line represents an AUC
+each Stool Study.} A) AUC for prediction in adenoma using all genera or
+CRC associated genera only. B) AUC for prediction in carcinoma using all
+genera or CRC-associated genera only. The dotted line represents an AUC
 of 0.5. The x-axis is the data set in which the model was initially
@@
 Genera for each Tissue Study.} A) AUC for prediction in unmatched tissue
-for all genera or CRC associated genera only. B) AUC for prediction in
-matched tissue using all genera or CRC associated genera only. The
+for all genera or CRC-associated genera only. B) AUC for prediction in
+matched tissue using all genera or CRC-associated genera only. The
 dotted line represents an AUC of 0.5. The x-axis is the data set in

### 2017-11-14 submission/supplemental.Rmd
remove unused table legends
add Will's supplemental suggestions

@@
 
-colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Adenoma n", "Control n")
+colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Adenoma (n)", "Control (n)")
 
@@
 
-colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma n", "Control n")
+colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma (n)", "Control (n)")
 
@@
 
-colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Adenoma n", "Control n")
+colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Adenoma (n)", "Control (n)")
 
@@
 
-colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma n", "Control n")
+colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma (n)", "Control (n)")
 
@@
 ```
-
-\newpage
-
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
-
-**Table S2: Summary of Common OTUs for All Diagnosis Groups**
-
-\footnotesize
-
-```{r supplemental_table, echo=FALSE, message=FALSE, warning=FALSE}
-
-```
-\normalsize
-\newpage
-
-**Table S5: Summary of mtry ranges used for the treatment and diagnosis models**
-
-\footnotesize
-
-```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
-
-```
-\normalsize

### 2017-11-14 submission/manuscript.Rmd
add Will's abstract suggestions

@@
 
-**Background.** An increasing body of literature suggests that there is a role for the microbiota in colorectal cancer.
-Important drivers within this axis have ranged from individual microbes to the whole community.
-A recent meta-analysis investigated whether consistent biomarkers could be identified across studies.
-This report expands on this previous research and tests the hypothesis that the bacterial community is an important driver of disease from control to adenoma to carcinoma.
-To test this hypothesis we examined both feces (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) across 14 different studies.
+**Background.** An increasing body of literature suggests that there is a role for the microbiota in colorectal cancer (CRC).
+Important drivers within this context have ranged from individual microbes to the whole community.
+Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community is an important driver of both early (adenoma) and late (carcinoma) stage of disease.
+To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r sum(filter(tissue_n, study == "flemer")[, c("control", "adenoma", "cancer")]) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.
 
 **Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced.
-Lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) while only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk.
+Only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk while lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
 Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
-Using the whole community versus only CRC associated genera to build a prediction model resulted in a higher classification AUC for both adenoma and carcinoma for fecal and tissue samples.
-For the studies that were analyzed, most were adequately powered for large effect size differences which may be more amenable for carcinoma than for adenoma microbiota research.
+Using the whole community versus only CRC associated genera to build a prediction model resulted in a higher classification based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples.
+For the included studies, most were adequately powered for large effect size differences.
+This may be more amenable for carcinoma than for adenoma microbiota research due to the smaller community level changes observed in our results.
 
 **Conclusions.** This data provides support for the importance of the bacteral community to both adenoma and carcinoma genesis.
-The evidence that has been collated within this study is much stronger for carcinoma and this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
+The evidence collected within this study on the role of the microbiota in CRC pathogenesis is much stronger for carcinoma then adenoma.
+A strong reason for this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
 

### 2017-11-14 submission/manuscript.tex
add Will's abstract suggestions

@@
 \textbf{Background.} An increasing body of literature suggests that
-there is a role for the microbiota in colorectal cancer. Important
-drivers within this axis have ranged from individual microbes to the
-whole community. A recent meta-analysis investigated whether consistent
-biomarkers could be identified across studies. This report expands on
-this previous research and tests the hypothesis that the bacterial
-community is an important driver of disease from control to adenoma to
-carcinoma. To test this hypothesis we examined both feces (total
-individuals = 1737) and tissue (total samples = 492) across 14 different
-studies.
+there is a role for the microbiota in colorectal cancer (CRC). Important
+drivers within this context have ranged from individual microbes to the
+whole community. Our study expands on a recent meta-analysis
+investigating microbial biomarkers for CRC by testing the hypothesis
+that the bacterial community is an important driver of both early
+(adenoma) and late (carcinoma) stage of disease. To test this hypothesis
+we examined both feces (n = 1737) and tissue (492 total samples from 400
+individuals) across 14 different studies.
 
@@
 \textless{} 0.05) after correcting for study effect and variable region
-sequenced. Lower Shannon diversity and evenness in fecal samples
-resulted in a significant increase in relative risk for carcinoma
-(P-value \textless{} 0.05) while only evenness for adenoma (P-value
-\textless{} 0.05) resulted in a slightly increased relative risk.
-Previously associated colorectal cancer genera (\emph{Fusobacterium},
-\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
-followed a similar pattern with a significantly increased relative risk
-by their presence for carcinoma (P-value \textless{} 0.05) but not
-adenoma (P-value \textgreater{} 0.05) with the exception of
-\emph{Porphyromonas} (P-value \textless{} 0.05). Using the whole
-community versus only CRC associated genera to build a prediction model
-resulted in a higher classification AUC for both adenoma and carcinoma
-for fecal and tissue samples. For the studies that were analyzed, most
-were adequately powered for large effect size differences which may be
-more amenable for carcinoma than for adenoma microbiota research.
+sequenced. Only evenness for adenoma (P-value \textless{} 0.05) resulted
+in a slightly increased relative risk while lower Shannon diversity and
+evenness in fecal samples resulted in a significant increase in relative
+risk for carcinoma (P-value \textless{} 0.05). Previously associated
+colorectal cancer genera (\emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Peptostreptococcus}, or \emph{Porphyromonas}) followed a similar
+pattern with a significantly increased relative risk by their presence
+for carcinoma (P-value \textless{} 0.05) but not adenoma (P-value
+\textgreater{} 0.05) with the exception of \emph{Porphyromonas} (P-value
+\textless{} 0.05). Using the whole community versus only CRC associated
+genera to build a prediction model resulted in a higher classification
+based on Area Under the Curve (AUC) for both adenoma and carcinoma using
+fecal and tissue samples. For the included studies, most were adequately
+powered for large effect size differences. This may be more amenable for
+carcinoma than for adenoma microbiota research due to the smaller
+community level changes observed in our results.
 
@@
 the bacteral community to both adenoma and carcinoma genesis. The
-evidence that has been collated within this study is much stronger for
-carcinoma and this may be in part due to the low power to detect more
+evidence collected within this study on the role of the microbiota in
+CRC pathogenesis is much stronger for carcinoma then adenoma. A strong
+reason for this may be in part due to the low power to detect more
 subtle changes in the majority of studies that have been performed to

### 2017-11-14 submission/manuscript.Rmd
update tables with n

@@
 
-**Table 1: Studies with Stool Samples Included in the Analysis**
+**Table 1: Total Individuals in each Study Included in the Stool Analysis**
 
@@
 
-colnames(table1) <- c("Study", "Data Stored", "16S Region", "Controls", "Adenoma", "Carcinoma")
+colnames(table1) <- c("Study", "Data Stored", "16S Region", "Control (n)", "Adenoma (n)", "Carcinoma (n)")
 
@@
 
-colnames(table2) <- c("Study", "Data Stored", "16S Region", "Controls", "Adenoma", "Carcinoma")
+colnames(table2) <- c("Study", "Data Stored", "16S Region", "Control (n)", "Adenoma (n)", "Carcinoma (n)")
 

### 2017-11-14 submission/manuscript.tex
update tables with n

@@
 
-\textbf{Table 1: Studies with Stool Samples Included in the Analysis}
+\textbf{Table 1: Total Individuals in each Study Included in the Stool
+Analysis}
 
@@
 \toprule
-Study & Data Stored & 16S Region & Controls & Adenoma &
-Carcinoma\tabularnewline
+Study & Data Stored & 16S Region & Control (n) & Adenoma (n) & Carcinoma
+(n)\tabularnewline
 \midrule

### 2017-11-14 code/get_matched_tissue.R
corrent error of the matched samples of flemer

@@ -5,10 +5,9 @@
 ###### Notes to self ######################################################
 
 # Chen is unmatched - chen.metadata
-# Flemer is unmatched - flemer.metadata
 # Sanaparedy is unmatched - sana.metadata
 
-
+# Flemer is matched (want Individual column)- flemer.metadata
 # Burns is matched (want host_subject_id_s column) - burnsMetadata.csv
 # Lu data A matches with B - luData.csv
 # Dejea is matched (host_tissue_sampled_s column) - SraRunTable.txt
@@ -34,7 +33,7 @@ tissue_sets <- c("lu", "dejea", "sana", "burns", "geng")
 both_sets <- c("flemer", "chen")
 
 # Create vector with data that has matched samples
-matched_sets <- c("burns", "dejea", "lu", "geng")
+matched_sets <- c("burns", "dejea", "lu", "geng", "flemer")
 
 
 
@@ -243,7 +242,9 @@ data_to_match_list <- list(
   geng = tissue_metadata[["geng"]] %>% 
     select(Run_s, Sample_Name_s) %>% 
     mutate(Sample_Name_s = gsub("[A-Z]", "", Sample_Name_s)) %>% 
-    rename(id = Sample_Name_s, group = Run_s))
+    rename(id = Sample_Name_s, group = Run_s), 
+  flemer = tissue_metadata[["flemer"]] %>% select(sample, Individual) %>% 
+    rename(id = Individual, group = sample))
 
 alpha_to_match_list <- zscore_pwr_transform_data[matched_sets]
 
@@ -259,7 +260,7 @@ tissue_matched_data <- bind_rows(mapply(get_matched_set_data, matched_sets, "mat
                     ifelse(disease == "Tumor", invisible("cancer"), invisible(disease))))
 
 tissue_unmatched_data <- bind_rows(mapply(get_matched_set_data, matched_sets, "unmatched"), 
-                  zscore_pwr_transform_data[c("chen", "flemer", "sana")]) %>% 
+                  zscore_pwr_transform_data[c("chen", "sana")]) %>% 
   mutate(disease = ifelse(is.na(disease), invisible(disease.x), invisible(disease)))
 
 

### 2017-11-14 code/get_matched_tissue.R
revert changes

@@ -33,7 +33,7 @@ tissue_sets <- c("lu", "dejea", "sana", "burns", "geng")
 both_sets <- c("flemer", "chen")
 
 # Create vector with data that has matched samples
-matched_sets <- c("burns", "dejea", "lu", "geng", "flemer")
+matched_sets <- c("burns", "dejea", "lu", "geng")
 
 
 
@@ -242,9 +242,7 @@ data_to_match_list <- list(
   geng = tissue_metadata[["geng"]] %>% 
     select(Run_s, Sample_Name_s) %>% 
     mutate(Sample_Name_s = gsub("[A-Z]", "", Sample_Name_s)) %>% 
-    rename(id = Sample_Name_s, group = Run_s), 
-  flemer = tissue_metadata[["flemer"]] %>% select(sample, Individual) %>% 
-    rename(id = Individual, group = sample))
+    rename(id = Sample_Name_s, group = Run_s))
 
 alpha_to_match_list <- zscore_pwr_transform_data[matched_sets]
 
@@ -260,7 +258,7 @@ tissue_matched_data <- bind_rows(mapply(get_matched_set_data, matched_sets, "mat
                     ifelse(disease == "Tumor", invisible("cancer"), invisible(disease))))
 
 tissue_unmatched_data <- bind_rows(mapply(get_matched_set_data, matched_sets, "unmatched"), 
-                  zscore_pwr_transform_data[c("chen", "sana")]) %>% 
+                  zscore_pwr_transform_data[c("chen", "flemer", "sana")]) %>% 
   mutate(disease = ifelse(is.na(disease), invisible(disease.x), invisible(disease)))
 
 

### 2017-11-14 code/get_matched_tissue.R
add ids for the flemer study

@@ -242,12 +242,14 @@ data_to_match_list <- list(
   geng = tissue_metadata[["geng"]] %>% 
     select(Run_s, Sample_Name_s) %>% 
     mutate(Sample_Name_s = gsub("[A-Z]", "", Sample_Name_s)) %>% 
-    rename(id = Sample_Name_s, group = Run_s))
+    rename(id = Sample_Name_s, group = Run_s), 
+  flemer = tissue_metadata[["flemer"]] %>% select(sample, Individual) %>% 
+    rename(id = Individual, group = sample))
 
-alpha_to_match_list <- zscore_pwr_transform_data[matched_sets]
+alpha_to_match_list <- zscore_pwr_transform_data[c(matched_sets, "flemer")]
 
 # generate the combined data
-combination_data <- mapply(get_combinations, matched_sets, USE.NAMES = T, SIMPLIFY = F)
+combination_data <- mapply(get_combinations, c(matched_sets, "flemer"), USE.NAMES = T, SIMPLIFY = F)
 
 
 # Get the matched data
@@ -258,7 +260,9 @@ tissue_matched_data <- bind_rows(mapply(get_matched_set_data, matched_sets, "mat
                     ifelse(disease == "Tumor", invisible("cancer"), invisible(disease))))
 
 tissue_unmatched_data <- bind_rows(mapply(get_matched_set_data, matched_sets, "unmatched"), 
-                  zscore_pwr_transform_data[c("chen", "flemer", "sana")]) %>% 
+                                   get_matched_set_data("flemer", "matched"), 
+                                   get_matched_set_data("flemer", "unmatched"), 
+                  zscore_pwr_transform_data[c("chen", "sana")]) %>% 
   mutate(disease = ifelse(is.na(disease), invisible(disease.x), invisible(disease)))
 
 
@@ -271,7 +275,7 @@ raw_alpha <- mapply(get_data, c(tissue_sets, both_sets), "tissue", SIMPLIFY = F)
 tissue_matched_data <- tissue_matched_data %>% inner_join(raw_alpha, by = "group")
 tissue_unmatched_data <- tissue_unmatched_data %>% inner_join(raw_alpha, by = "group")
 
-
+test <- get_matched_set_data("flemer", "matched")
 
 
 

### 2017-11-14 code/get_matched_tissue.R
finalize the changes

@@ -275,8 +275,6 @@ raw_alpha <- mapply(get_data, c(tissue_sets, both_sets), "tissue", SIMPLIFY = F)
 tissue_matched_data <- tissue_matched_data %>% inner_join(raw_alpha, by = "group")
 tissue_unmatched_data <- tissue_unmatched_data %>% inner_join(raw_alpha, by = "group")
 
-test <- get_matched_set_data("flemer", "matched")
-
 
 
 # Write out the data tables for analysis

### 2017-11-14 code/get_tissue_combined_alpha.R
add correction for flemer multiple samples from same individual

@@ -36,7 +36,8 @@ tissue_unmatched <- read.csv("data/process/tables/alpha_tissue_unmatched_data.cs
          v_region = ifelse(study %in% c("lu", "flemer"), invisible("V3-4"), 
                            ifelse(study == "burns", invisible("V5-6"), 
                                   ifelse(study == "dejea", invisible("V3-5"), 
-                                         ifelse(study == "chen", invisible("V1-3"), invisible("V1-2"))))))
+                                         ifelse(study == "chen", invisible("V1-3"), invisible("V1-2")))))) %>% 
+  mutate(id = ifelse(study %in% c("chen", "sana"), invisible(group), invisible(id)))
 
 
 # Run tests for alpha metrics using t-tests
@@ -182,7 +183,7 @@ unmatched_tukey_results <- mapply(get_anova_comparisons, c("sobs", "shannon", "s
 
 # RUn linear mixed-effect models
 unmatched_mixeffect_results <- mapply(get_mixed_effect, c("sobs", "shannon", "shannoneven"), 
-                            data_set = "tissue_unmatched", USE.NAMES = T, SIMPLIFY = F) %>% 
+                                      individual = "id", data_set = "tissue_unmatched", USE.NAMES = T, SIMPLIFY = F) %>% 
   bind_rows() %>% gather(key = alpha_metric, value = value) %>% 
   mutate(measure = rep(c("chi_sq", "pvalue"), length(value)/2)) %>% 
   spread(measure, value) %>% 

### 2017-11-14 submission/manuscript.Rmd
add updated numbers

@@
 tissue_n <- read_csv("../data/process/tables/tissue_study_n_analyzed.csv")
+flemer_data <- read_tsv("../data/process/flemer/flemer.metadata")
 
@@
 Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community is an important driver of both early (adenoma) and late (carcinoma) stage of disease.
-To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r sum(filter(tissue_n, study == "flemer")[, c("control", "adenoma", "cancer")]) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.
+To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.
 
@@
 
-Colorectal cancer (CRC) is a growing world wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016].
-Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in polyp formation in mouse models.
-There has also been numerous case/control studies investigating the microbiota in both adenoma and carcinoma.
+Colorectal cancer (CRC) is a growing world-wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016].
+Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in polyp formation using mouse models of CRC.
+There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma.
 Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
-Many of the studies along with the current meta-analysis focus on identifying biomarkers or individual microbes but do not critically investigate the community role in the disease.
+Many of the studies, along with the previous meta-analysis, focus on identifying biomarkers or individual microbes but do not critically investigate the role the community has in CRC.
 
-Using both fecal (total individuals = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue samples (total samples = `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)`) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] we expand both the breadth and scope of the previous meta-analysis; to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
-To accomplish this we first assessed whether the diversity changes across disease (control to adenoma to carcinoma) and if it results in an increased relative risk for adenoma or carcinoma.
+Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
+To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma.
 Next, we assessed how typical CRC associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma.
-Third, using Random Forest models we analyzed whether the full community or only the CRC associated genera resulted in better model classification area under the curve (AUC).
-Finally, we examined at what effect size and sample n the studies that were used were adequately powered for.
+Third, using Random Forest models, we analyzed whether the full community or only the CRC associated genera resulted in better model classification based on the area under the curve (AUC).
+Since the changes in community were subtle for adenoma we also examined what effect size and sample size the studies that were used were adequately powered for.
 

### 2017-11-14 submission/manuscript.tex
add updated numbers

@@
 (adenoma) and late (carcinoma) stage of disease. To test this hypothesis
-we examined both feces (n = 1737) and tissue (492 total samples from 400
+we examined both feces (n = 1737) and tissue (492 total samples from 350
 individuals) across 14 different studies.
@@
 
-Colorectal cancer (CRC) is a growing world wide health problem {[}1{]}
+Colorectal cancer (CRC) is a growing world-wide health problem {[}1{]}
 in which the microbiota has been purported to play an active role in
@@
 of both individual microbes {[}3--7{]} and the overall community
-{[}8--10{]} in polyp formation in mouse models. There has also been
-numerous case/control studies investigating the microbiota in both
-adenoma and carcinoma. Recently, a meta-analysis was published
-investigating whether specific biomarkers could be consistently
-identified using multiple data sets {[}11{]}. Many of the studies along
-with the current meta-analysis focus on identifying biomarkers or
-individual microbes but do not critically investigate the community role
-in the disease.
+{[}8--10{]} in polyp formation using mouse models of CRC. There have
+also been numerous case/control studies investigating the microbiota in
+the formation of both adenoma and carcinoma. Recently, a meta-analysis
+was published investigating whether specific biomarkers could be
+consistently identified using multiple data sets {[}11{]}. Many of the
+studies, along with the previous meta-analysis, focus on identifying
+biomarkers or individual microbes but do not critically investigate the
+role the community has in CRC.
 
-Using both fecal (total individuals = 1737) and tissue samples (total
-samples = 492) totalling over 2229 total samples across 14 studies
-{[}12--25{]} we expand both the breadth and scope of the previous
-meta-analysis; to investigate whether the bacterial community is an
+Using both feces (n = 1737) and tissue (492 samples from 350
+individuals) totalling over 2229 total samples across 14 studies
+{[}12--25{]}, we expand both the breadth and scope of the previous
+meta-analysis to investigate whether the bacterial community is an
 important risk factor for both adenoma and carcinoma. To accomplish this
-we first assessed whether the diversity changes across disease (control
-to adenoma to carcinoma) and if it results in an increased relative risk
-for adenoma or carcinoma. Next, we assessed how typical CRC associated
-genera (\emph{Fusobacterium}, \emph{Parvimonas},
+we first assessed whether the diversity changes throughout disease
+(control to adenoma to carcinoma) and if it results in an increased
+relative risk (RR) for adenoma or carcinoma. Next, we assessed how
+typical CRC associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
 \emph{Peptostreptococcus}, or \emph{Porphyromonas}) affect the relative
-risk of adenoma or carcinoma. Third, using Random Forest models we
+risk of adenoma or carcinoma. Third, using Random Forest models, we
 analyzed whether the full community or only the CRC associated genera
-resulted in better model classification area under the curve (AUC).
-Finally, we examined at what effect size and sample n the studies that
-were used were adequately powered for.
+resulted in better model classification based on the area under the
+curve (AUC). Since the changes in community were subtle for adenoma we
+also examined what effect size and sample size the studies that were
+used were adequately powered for.
 

### 2017-11-14 submission/manuscript.Rmd
add WIll's Intro suggestions

@@
 
-Our analysis found a continuous decrease in Shannon diversity from control to adenoma to carcinoma and a significantly increased relative risk for carcinoma with this lower diversity.
-Using only the CRC associated genera this relative risk was higher than what was observed for Shannon diversity.
-However, by using the full community instead of only the CRC associated genera the AUC of the classification models increased.
-Although we analyze a data set with a large number of total individuals each individual study was underpowered for effect size differences of 10% or below between the case and control group.
+Our analysis found a continuous decrease in Shannon diversity as disease became more severe which correlated with a significantly increased RR for carcinoma.
+Using only CRC-associated genera, the RR for carcinoma was higher relative to what was observed for the Shannon diversity RR.
+Conversely, we demonstrate that the AUC of the classification models increased when the full community was incorporated as opposed to only CRC-associated genera.
+Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.
 

### 2017-11-14 submission/manuscript.tex
add WIll's Intro suggestions

@@
 
-Our analysis found a continuous decrease in Shannon diversity from
-control to adenoma to carcinoma and a significantly increased relative
-risk for carcinoma with this lower diversity. Using only the CRC
-associated genera this relative risk was higher than what was observed
-for Shannon diversity. However, by using the full community instead of
-only the CRC associated genera the AUC of the classification models
-increased. Although we analyze a data set with a large number of total
-individuals each individual study was underpowered for effect size
-differences of 10\% or below between the case and control group.
+Our analysis found a continuous decrease in Shannon diversity as disease
+became more severe which correlated with a significantly increased RR
+for carcinoma. Using only CRC-associated genera, the RR for carcinoma
+was higher relative to what was observed for the Shannon diversity RR.
+Conversely, we demonstrate that the AUC of the classification models
+increased when the full community was incorporated as opposed to only
+CRC-associated genera. Although we analyzed data sets which sampled
+large numbers of individuals, our results indicate the individual
+studies were underpowered for detecting effect size differences of 10\%
+or below between the case and control groups.
 

### 2017-11-14 submission/manuscript.Rmd
add Will's suggestions to the results

@@
 
-Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
+Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
 To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma.
@@
 
-***Diversity is Lower in Those with Carcinoma and Increases Relative Risk:***
-Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces and not tissue were  lower in those with carcinoma [Figure 1].
-Using linear mixed-effect models to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`).
-This effect was not observed in tissue when additionally controlling for whether the sample came from the same individual (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
-For fecal samples a decrease in Shannon diversity and evenness resulted in a significantly increased relative risk for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2].
+***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
+Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces, but not tissue, were  lower in those with carcinoma [Figure 1].
+Using linear mixed-effect models to control for study and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`).
+However, in tissue, this effect was not observed in tissue when resampling of the same individual was also controlled for (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
+Within fecal samples, a decrease in Shannon diversity and evenness resulted in a significantly increased RR for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2].
 Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2].
-Only a decrease in evenness had an increased relative risk for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))`).
+Only a decrease in evenness had an increased RR for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))`).
 
-Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls but not between adenoma and controls [Table S1 & S2].
-Many studies with unmatched tissue samples had a significant difference for both adenoma and carcinoma versus controls [Table S3 & S4] while studies with matched tissue samples had no differences [Table S3 & S4].
+Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S1 & S2].
+For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with matched tissue samples had no differences [Table S3 & S4].
 
-***Genera Previously Associated with Carcinoma Predominately Increases Relative Risk for Carcinoma but not Adenoma:***
-The majority of CRC associated genera for both feces and tissue had a significantly increased relative risk for carcinoma but not for adenoma [Figure 3].
-In fecal samples the relative risk due to CRC associated genera was greater than either evenness or Shannon diversity [Figure 2 & 3].
-Additionally, the relative risk of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D].
-The relative risk effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`).
-This decrease may be explained by the fact that the tissue analysis include matched samples.
+***Carcinoma-Associated Genera Minimally Impacts RR of Adenoma:***
+The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3].
+In fecal samples the RR due to CRC associated genera was greater than either the RR assoicated with evenness or Shannon diversity [Figure 2 & 3].
+Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D].
+The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) than for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`).
+This decrease may be explained by the fact that the tissue analysis included matched samples.
 
-Two measures in stool for adenoma were significant when investigating these CRC associated genera.
-The first was *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and the second was being positive for three CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A].
-For tissue three measures for adenoma were significant; being positive for one CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), being positive for two CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].
+There were two significant measures for increased RR of adenoma when investigating CRC-associated genera in stool: 1) Having a higher then median value of *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and 2) whether samples were positive for three CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A].
+With tissue, there were three significant measures for an increased RR of adenoma: 1) being positive for one CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), 2) being positive for two CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and 3) being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].
 
-***Using the Whole Community Increases Model AUC over CRC Associated Genera:***
-For both fecal and tissue samples (matched and unmatched) the AUC decreases when only OTUs from the CRC associated genera are used [Figure 4 & 5].
+***Using the Whole Community Instead of Only CRC-Associated Genera Increases Model AUC:***
+For both fecal and tissue (matched and unmatched) samples, the AUC decreases when only OTUs from the CRC-associated genera are used [Figure 4 & 5].
 This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5].
-The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S2-S3].
-Both genus models perform similarily in their ability to be able to predict lesion (adenoma or carcinoma) with carcinoma having a higher AUC then adenoma [Figure S4-S5].
-Matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S4A, S5B, & S6] then carcinoma models [Figure S4B & S5A].
+The genus models generally performed similar to the OTU-based models with the full genera models performing better than the CRC-associated genera models [Figure S2-S3].
+Both genus models were similarily predictive in their ability to detect adenomas or carcinomas, with carcinoma having a higher AUC then adenoma [Figure S4-S5].
+Of note, matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S4A, S5B, & S6] than carcinoma models [Figure S4B & S5A].
 
-***Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
+***A Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
 When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B].
-No single study that was analyzed had the standard 80% power to detect an effect size difference that was eqaul to or below 10% [Figure 6A & B].
-In order to achieve adequate power for small effect sizes it would be necessary to recruit over 1000 individuals for each arm of the study [Figure 6C].
+No single study that was analyzed had the standard 80% power to detect an effect size difference that was equal to or below 10% [Figure 6A & B].
+In order to achieve adequate power for small effect sizes, studies would need to recruit over 1000 individuals for each arm [Figure 6C].
 

### 2017-11-14 submission/manuscript.tex
add Will's suggestions to the results

@@
 individuals) totalling over 2229 total samples across 14 studies
-{[}12--25{]}, we expand both the breadth and scope of the previous
-meta-analysis to investigate whether the bacterial community is an
-important risk factor for both adenoma and carcinoma. To accomplish this
-we first assessed whether the diversity changes throughout disease
-(control to adenoma to carcinoma) and if it results in an increased
-relative risk (RR) for adenoma or carcinoma. Next, we assessed how
-typical CRC associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Peptostreptococcus}, or \emph{Porphyromonas}) affect the relative
-risk of adenoma or carcinoma. Third, using Random Forest models, we
-analyzed whether the full community or only the CRC associated genera
-resulted in better model classification based on the area under the
-curve (AUC). Since the changes in community were subtle for adenoma we
-also examined what effect size and sample size the studies that were
-used were adequately powered for.
+{[}12--25{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope of
+the previous meta-analysis to investigate whether the bacterial
+community is an important risk factor for both adenoma and carcinoma. To
+accomplish this we first assessed whether the diversity changes
+throughout disease (control to adenoma to carcinoma) and if it results
+in an increased relative risk (RR) for adenoma or carcinoma. Next, we
+assessed how typical CRC associated genera (\emph{Fusobacterium},
+\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
+affect the relative risk of adenoma or carcinoma. Third, using Random
+Forest models, we analyzed whether the full community or only the CRC
+associated genera resulted in better model classification based on the
+area under the curve (AUC). Since the changes in community were subtle
+for adenoma we also examined what effect size and sample size the
+studies that were used were adequately powered for.
 
@@
 
-\textbf{\emph{Diversity is Lower in Those with Carcinoma and Increases
-Relative Risk:}} Using power transformed and Z-score normalized
+\textbf{\emph{Lower Community Diversity is Associated with Increased RR
+of Carcinomas:}} Using power transformed and Z-score normalized
 \(\alpha\)-diversity metrics, both evenness and Shannon diversity in
-feces and not tissue were lower in those with carcinoma {[}Figure 1{]}.
-Using linear mixed-effect models to control for study and variable
-region there was a significant decrease from control to adenoma to
+feces, but not tissue, were lower in those with carcinoma {[}Figure
+1{]}. Using linear mixed-effect models to control for study and variable
+region, there was a significant decrease from control to adenoma to
 carcinoma for both evenness (P-value = 0.025) and Shannon diversity
-(P-value = 0.043). This effect was not observed in tissue when
-additionally controlling for whether the sample came from the same
-individual (P-value \textgreater{} 0.05). For fecal samples a decrease
-in Shannon diversity and evenness resulted in a significantly increased
-relative risk for carcinoma (P-value = 0.01 and P-value = 0.0011,
-respectively) {[}Figure 2{]}. Although these values were significant,
-the effect size was relatively small for both metrics (Shannon RR = 1.31
-and evenness RR = 1.34) {[}Figure 2{]}. Only a decrease in evenness had
-an increased relative risk for adenoma (P-value = 0.032) {[}Figure 2A \&
-S1{]} but this effect size was even smaller than what was observed for
-carcinoma (RR = 1.16).
+(P-value = 0.043). However, in tissue, this effect was not observed in
+tissue when resampling of the same individual was also controlled for
+(P-value \textgreater{} 0.05). Within fecal samples, a decrease in
+Shannon diversity and evenness resulted in a significantly increased RR
+for carcinoma (P-value = 0.01 and 0.0011, respectively) {[}Figure 2{]}.
+Although these values were significant, the effect size was relatively
+small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
+{[}Figure 2{]}. Only a decrease in evenness had an increased RR for
+adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect size was
+even smaller than what was observed for carcinoma (RR = 1.16).
 
@@
 difference across studies in the bacterial community of fecal samples
-between carcinoma and controls but not between adenoma and controls
-{[}Table S1 \& S2{]}. Many studies with unmatched tissue samples had a
-significant difference for both adenoma and carcinoma versus controls
-{[}Table S3 \& S4{]} while studies with matched tissue samples had no
-differences {[}Table S3 \& S4{]}.
+between carcinoma and controls, but not adenoma and controls {[}Table S1
+\& S2{]}. For studies with unmatched tissue samples a similar trend was
+observed {[}Table S3 \& S4{]} while studies with matched tissue samples
+had no differences {[}Table S3 \& S4{]}.
 
-\textbf{\emph{Genera Previously Associated with Carcinoma Predominately
-Increases Relative Risk for Carcinoma but not Adenoma:}} The majority of
-CRC associated genera for both feces and tissue had a significantly
-increased relative risk for carcinoma but not for adenoma {[}Figure
-3{]}. In fecal samples the relative risk due to CRC associated genera
-was greater than either evenness or Shannon diversity {[}Figure 2 \&
-3{]}. Additionally, the relative risk of carcinoma continuously
-increased as individuals tested positive for more CRC associated genera
-{[}Figure 3B \& 3D{]}. The relative risk effect size was greater for
-stool (RR range = 1.61 - 2.74) then for tissue (RR range = 1.21 - 1.81).
-This decrease may be explained by the fact that the tissue analysis
-include matched samples.
+\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
+Adenoma:}} The majority of CRC-associated genera for both feces and
+tissue had a significantly increased RR for carcinoma but not for
+adenoma {[}Figure 3{]}. In fecal samples the RR due to CRC associated
+genera was greater than either the RR assoicated with evenness or
+Shannon diversity {[}Figure 2 \& 3{]}. Additionally, the RR of carcinoma
+continuously increased as individuals tested positive for more CRC
+associated genera {[}Figure 3B \& 3D{]}. The RR effect size was greater
+for stool (RR range = 1.61 - 2.74) than for tissue (RR range = 1.21 -
+1.81). This decrease may be explained by the fact that the tissue
+analysis included matched samples.
 
-Two measures in stool for adenoma were significant when investigating
-these CRC associated genera. The first was \emph{Porphyromonas} (P-value
-= 0.023) and the second was being positive for three CRC associated
-genera (P-value = 0.022) {[}Figure 3A{]}. For tissue three measures for
-adenoma were significant; being positive for one CRC associated genera
-(P-value = 0.032), being positive for two CRC associated genera (P-value
-= 0.008), and being positive for four CRC associated genera (P-value =
-0.039) {[}Figure 3C{]}.
+There were two significant measures for increased RR of adenoma when
+investigating CRC-associated genera in stool: 1) Having a higher then
+median value of \emph{Porphyromonas} (P-value = 0.023) and 2) whether
+samples were positive for three CRC associated genera (P-value = 0.022)
+{[}Figure 3A{]}. With tissue, there were three significant measures for
+an increased RR of adenoma: 1) being positive for one CRC-associated
+genera (P-value = 0.032), 2) being positive for two CRC associated
+genera (P-value = 0.008), and 3) being positive for four CRC associated
+genera (P-value = 0.039) {[}Figure 3C{]}.
 
-\textbf{\emph{Using the Whole Community Increases Model AUC over CRC
-Associated Genera:}} For both fecal and tissue samples (matched and
-unmatched) the AUC decreases when only OTUs from the CRC associated
-genera are used {[}Figure 4 \& 5{]}. This decrease is observed in both
-adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The genus models
-generally had similar trends as observed for the OTU based models with
-the full genera models performing better then the CRC associated genera
-models {[}Figure S2-S3{]}. Both genus models perform similarily in their
-ability to be able to predict lesion (adenoma or carcinoma) with
-carcinoma having a higher AUC then adenoma {[}Figure S4-S5{]}. Matched
-tissue samples for those with carcinoma had an AUC that was more similar
-to the adenoma models {[}Figure S4A, S5B, \& S6{]} then carcinoma models
+\textbf{\emph{Using the Whole Community Instead of Only CRC-Associated
+Genera Increases Model AUC:}} For both fecal and tissue (matched and
+unmatched) samples, the AUC decreases when only OTUs from the
+CRC-associated genera are used {[}Figure 4 \& 5{]}. This decrease is
+observed in both adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The
+genus models generally performed similar to the OTU-based models with
+the full genera models performing better than the CRC-associated genera
+models {[}Figure S2-S3{]}. Both genus models were similarily predictive
+in their ability to detect adenomas or carcinomas, with carcinoma having
+a higher AUC then adenoma {[}Figure S4-S5{]}. Of note, matched tissue
+samples for those with carcinoma had an AUC that was more similar to the
+adenoma models {[}Figure S4A, S5B, \& S6{]} than carcinoma models
 {[}Figure S4B \& S5A{]}.
 
-\textbf{\emph{Majority of Studies are Underpowered for Detecting Small
+\textbf{\emph{A Majority of Studies are Underpowered for Detecting Small
 Effect Size Differences:}} When assessing the power of each study at
@@
 B{]}. No single study that was analyzed had the standard 80\% power to
-detect an effect size difference that was eqaul to or below 10\%
+detect an effect size difference that was equal to or below 10\%
 {[}Figure 6A \& B{]}. In order to achieve adequate power for small
-effect sizes it would be necessary to recruit over 1000 individuals for
-each arm of the study {[}Figure 6C{]}.
+effect sizes, studies would need to recruit over 1000 individuals for
+each arm {[}Figure 6C{]}.
 

### 2017-11-14 submission/manuscript.Rmd
add RR generation to methods section

@@
 Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue.
-Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages.
+Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages by assessing how many with and without disease were above and below the overall median value within the specific study.
 Relative risk significance testing utilized the chi-squred test.

### 2017-11-14 submission/manuscript.Rmd
add Will's comments discussion up to P2

@@
 
-Our study identifies clear diversity changes both at the community level and within individual genera that are present in individuals with carcinoma versus those without the disease.
-Although there was a step wise decrease in diversity from control to adenoma to carcinoma; this did not translate into large effect sizes for the relative risk of either of these two conditions.
-Even though CRC associated genera increases the relative risk of carcinoma it does not consistently increase the relative risk of adenoma.
-This information suggests that these specific genera are important in carcinoma genesis but may not be the primary members of the microbial community that contribute to the formation of an adenoma.
-Additionally, our data shows that by using the whole community our models perform better then when they only use the CRC associated genera.
-CRC associated genera are clearly important to carcinoma but the context or community in which these microbes are a part of can drastically increase the ability of models to make predictions.
-This data supports the concept that small localized changes within the community may be occuring that are important in the disease progression of colorectal cancer and that they may not directly involve CRC associated genera.
+Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease.
+Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the relative risk of either of these two conditions.
+Even though CRC-associated genera increase the RR of carcinoma, they do not consistently increase the relative risk of adenoma.
+This information suggests that these specific genera are important in carcinoma genesis but may not be the primary members of the microbial community contributing to the formation of an adenoma.
+Additionally, our data show that by using the whole community, our models perform better than when only the CRC-associated genera are included.
+CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions.
+These observations suggest that small localized changes within the community may be occuring that are important in early disease progression of CRC and that this process may not directly involve CRC-associated genera.
 
-The driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], can be supported with this data for carcinoma but not necessarily for adenoma.
-The drasitically increased relative risk of disease, when considering the CRC associated genera, is highly supportive of this type of process, especially in the context of increasing relative risk with more CRC associated genera positivity.
-It is also possible that in a driver-passenger scenario that simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring.
-This would account for the observation that there is no constant additive effect on relative risk for increasing positivity.
-Additionally, the initial establishment of the driver within the system is also dependent on the community that is present and this is supported by the observation that when adding the community context to our models along with the CRC associated genera the model AUC increases.
+The data presented herein supports the driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], when applied to carcinoma but not necessarily adenoma.
+Both the drastically increased RR of CRC-associated genera versus diversity for carcinoma and increasing RR with more CRC-associated generea positivity are highly supportive of this model.
+It is also possible that in a driver-passenger scenario, simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring.
+This would account for the observation that there is no constant additive effect on RR for increasing positivity.
+Additionally, the initial establishment of the driver within the system appears to be dependent on the current community.
+This is supported by our finding that when adding the community context to our models in addition to the CRC-associated genera, the model AUC increases.
 

### 2017-11-14 submission/manuscript.tex
add RR generation to methods section
add Will's comments discussion up to P2

@@
 
-Our study identifies clear diversity changes both at the community level
-and within individual genera that are present in individuals with
-carcinoma versus those without the disease. Although there was a step
-wise decrease in diversity from control to adenoma to carcinoma; this
-did not translate into large effect sizes for the relative risk of
-either of these two conditions. Even though CRC associated genera
-increases the relative risk of carcinoma it does not consistently
-increase the relative risk of adenoma. This information suggests that
-these specific genera are important in carcinoma genesis but may not be
-the primary members of the microbial community that contribute to the
-formation of an adenoma. Additionally, our data shows that by using the
-whole community our models perform better then when they only use the
-CRC associated genera. CRC associated genera are clearly important to
-carcinoma but the context or community in which these microbes are a
-part of can drastically increase the ability of models to make
-predictions. This data supports the concept that small localized changes
-within the community may be occuring that are important in the disease
-progression of colorectal cancer and that they may not directly involve
-CRC associated genera.
+Our study identifies clear differences in diversity both at the
+community level and within individual genera that are present in
+individuals with CRC versus those without the disease. Although there
+was a step-wise decrease in diversity from control to adenoma to
+carcinoma, this did not translate into large effect sizes for the
+relative risk of either of these two conditions. Even though
+CRC-associated genera increase the RR of carcinoma, they do not
+consistently increase the relative risk of adenoma. This information
+suggests that these specific genera are important in carcinoma genesis
+but may not be the primary members of the microbial community
+contributing to the formation of an adenoma. Additionally, our data show
+that by using the whole community, our models perform better than when
+only the CRC-associated genera are included. CRC-associated genera are
+clearly important to carcinoma pathogenesis but accounting for the
+community in which these microbes exist can drastically increase the
+ability of models to make predictions. These observations suggest that
+small localized changes within the community may be occuring that are
+important in early disease progression of CRC and that this process may
+not directly involve CRC-associated genera.
 
-The driver-passenger model of the microbial role in CRC, as summarized
-by Flynn {[}2{]}, can be supported with this data for carcinoma but not
-necessarily for adenoma. The drasitically increased relative risk of
-disease, when considering the CRC associated genera, is highly
-supportive of this type of process, especially in the context of
-increasing relative risk with more CRC associated genera positivity. It
-is also possible that in a driver-passenger scenario that simply having
-the driver present or only identifying the passenger is a good enough
-proxy that the event is occuring. This would account for the observation
-that there is no constant additive effect on relative risk for
+The data presented herein supports the driver-passenger model of the
+microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
+carcinoma but not necessarily adenoma. Both the drastically increased RR
+of CRC-associated genera versus diversity for carcinoma and increasing
+RR with more CRC-associated generea positivity are highly supportive of
+this model. It is also possible that in a driver-passenger scenario,
+simply having the driver present or only identifying the passenger is a
+good enough proxy that the event is occuring. This would account for the
+observation that there is no constant additive effect on RR for
 increasing positivity. Additionally, the initial establishment of the
-driver within the system is also dependent on the community that is
-present and this is supported by the observation that when adding the
-community context to our models along with the CRC associated genera the
-model AUC increases.
+driver within the system appears to be dependent on the current
+community. This is supported by our finding that when adding the
+community context to our models in addition to the CRC-associated
+genera, the model AUC increases.
 
@@
 variable region, and individual effects in tissue. Relative risk was
-analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages.
-Relative risk significance testing utilized the chi-squred test.
-\(\beta\)-diversity differences utilized a Bray-Curtis distance matrix
-and PERMANOVA executed with the vegan (v2.4.4) package. Random Forest
-models were built using both the caret (v6.0.77) and randomForest
-(v4.6.12) packages. Differences between the obtained AUC versus a random
-model AUC was assessed using T-tests. Power analysis and estimations
-were made using the pwr (v1.2.1) and statmod (v1.4.30) packages. All
-figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
-packages.
+analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages by
+assessing how many with and without disease were above and below the
+overall median value within the specific study. Relative risk
+significance testing utilized the chi-squred test. \(\beta\)-diversity
+differences utilized a Bray-Curtis distance matrix and PERMANOVA
+executed with the vegan (v2.4.4) package. Random Forest models were
+built using both the caret (v6.0.77) and randomForest (v4.6.12)
+packages. Differences between the obtained AUC versus a random model AUC
+was assessed using T-tests. Power analysis and estimations were made
+using the pwr (v1.2.1) and statmod (v1.4.30) packages. All figures were
+created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages.
 

### 2017-11-15 submission/manuscript.tex
change title

@@
 
-\section{The Microbiota and Individual Community Members in Colorectal
-Cancer: Is There a Common
-Theme?}\label{the-microbiota-and-individual-community-members-in-colorectal-cancer-is-there-a-common-theme}
+\section{Investigating the Microbiota and Colorectal Cancer: The
+Importance of
+Community}\label{investigating-the-microbiota-and-colorectal-cancer-the-importance-of-community}
 

### 2017-11-15 submission/manuscript.Rmd
change title
finish Will's suggestions for discussion and conclusion

@@
 
-# The Microbiota and Individual Community Members in Colorectal Cancer: Is There a Common Theme?
+# Investigating the Microbiota and Colorectal Cancer: The Importance of Community
 
@@
 
-Our carcinoma observations fit the driver-passenger model and support this concept within the framework of the transition from adenoma to carcinoma.
-In contrast, with the present data we can only suggest that the adenoma observations might fit with this model but the changes that occur at this timepoint are small and possibly focal to the adenoma.
-The stepwise decrease in diversity suggests that the adenoma community is not normal but this change is subtle.
-Although there may be localized changes that do depend on the driver-passenger model, supported by an increased relative risk for one, two, and four positive CRC associated genera in tissue [Figure 3C], there may be other processes involved that ultimately exacerbates the condition from a subtle localized change to a global community one.
+With regards to carcinoma, our results support the driver-passenger model within the framework of the transition from adenoma to carcinoma.
+Conversely, using the present data, we can only theorize observations related to adenoma development fit this model.
+The changes that occur at this timepoint are small and possibly focal to the adenoma itself.
+The stepwise decrease in diversity suggests that the adenoma community is not normal but has changed subtly.
+Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change to a change in the global community.
 The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case.
-It is possible to hypothesize that at early stages of the diease, how the host interacts to these subtle changes could be the catlyst that ultimately leads to this larger global dysfunctional community.
+It is possible to hypothesize that at early stages of the diease, the host interacts with these subtle changes and catalyzes what ultimately leads to a thoroughly dysfunctional community.
 
-Although there are still questions that need to be answered for the microbiota and carcinoma, a clearer framework is beginning to develop as to how this occurs.
-The role of the microbiota in adenoma is still not clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here.
-It is realistic to suspect that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control.
-Most of the studies analyzed have sufficient power to detect these type of changes.
-In contrast, our data suggests that the adenoma effect size is relatively small.
-None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls and this may well be the range in which differences consistently occur in adenoma.
+Although there are still questions that need to be answered regarding the microbiota and carcinoma, a clearer framework for their relationship is beginning to develop as to how this occurs.
+From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 6].
+Conversely, the role of the microbiota in adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here.
+None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls.
+This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma.
 Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma formation.
@@
 
-By aggregating together a large collection of studies from both feces and tissue we are able to provide information in support of the importance of the bacterial community in both adenoma and carcinoma.
+By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in both adenoma and carcinoma.
 We are also able to provide support for the driver-passenger model in the context of carcinoma.
-However, within the context of adenoma it is less clear that this relationship exists.
-These observations highlight the importance of power and sample number considerations when considering investigations into the microbiota and adenoma due to the subtle changes in the community.
+However, within the context of adenoma, it is less clear that this relationship exists.
+These observations highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community.
+Although there are power limitations associated with adenoma, this report highlights the strong influence the microbiota has on CRC development.
 

### 2017-11-15 submission/manuscript.tex
finish Will's suggestions for discussion and conclusion

@@
 
-Our carcinoma observations fit the driver-passenger model and support
-this concept within the framework of the transition from adenoma to
-carcinoma. In contrast, with the present data we can only suggest that
-the adenoma observations might fit with this model but the changes that
-occur at this timepoint are small and possibly focal to the adenoma. The
+With regards to carcinoma, our results support the driver-passenger
+model within the framework of the transition from adenoma to carcinoma.
+Conversely, using the present data, we can only theorize observations
+related to adenoma development fit this model. The changes that occur at
+this timepoint are small and possibly focal to the adenoma itself. The
 stepwise decrease in diversity suggests that the adenoma community is
-not normal but this change is subtle. Although there may be localized
-changes that do depend on the driver-passenger model, supported by an
-increased relative risk for one, two, and four positive CRC associated
-genera in tissue {[}Figure 3C{]}, there may be other processes involved
-that ultimately exacerbates the condition from a subtle localized change
-to a global community one. The poor performance of the Random Forest
-models for classifying adenoma based only on the microbiota would
-suggest that this is the case. It is possible to hypothesize that at
-early stages of the diease, how the host interacts to these subtle
-changes could be the catlyst that ultimately leads to this larger global
-dysfunctional community.
+not normal but has changed subtly. Although there appears to be
+localized changes that do depend on the driver-passenger model, as
+supported by an increased RR for one, two, and four positive
+CRC-associated genera in tissue {[}Figure 3C{]}, there may be other
+processes at play that ultimately exacerbate the condition from a subtle
+localized change to a change in the global community. The poor
+performance of the Random Forest models for classifying adenoma based
+only on the microbiota would suggest that this is the case. It is
+possible to hypothesize that at early stages of the diease, the host
+interacts with these subtle changes and catalyzes what ultimately leads
+to a thoroughly dysfunctional community.
 
-Although there are still questions that need to be answered for the
-microbiota and carcinoma, a clearer framework is beginning to develop as
-to how this occurs. The role of the microbiota in adenoma is still not
-clear and part of the reason this may be is because many studies are not
-powered effectively to observe the small changes reported here. It is
-realistic to suspect that many changes in carcinoma could easily result
-in effect sizes that are 30\% or more between the case and control. Most
-of the studies analyzed have sufficient power to detect these type of
-changes. In contrast, our data suggests that the adenoma effect size is
-relatively small. None of the studies analyzed were properly powered to
-detect a 10\% or lower change between case and controls and this may
-well be the range in which differences consistently occur in adenoma.
+Although there are still questions that need to be answered regarding
+the microbiota and carcinoma, a clearer framework for their relationship
+is beginning to develop as to how this occurs. From our observations
+many changes in carcinoma could easily result in effect sizes that are
+30\% or more between the case and control and most studies analyzed have
+sufficient power to detect these types of changes {[}Figure 6{]}.
+Conversely, the role of the microbiota in adenoma is less clear and part
+of the reason this may be is because many studies are not powered
+effectively to observe the small changes reported here. None of the
+studies analyzed were properly powered to detect a 10\% or lower change
+between case and controls. This small effect size range may well be the
+scope in which differences consistently occur in adenoma due to the
+subtle changes in community that occur between control and adenoma.
 Future studies investigating adenoma and the microbiota need to take
@@
 By aggregating together a large collection of studies from both feces
-and tissue we are able to provide information in support of the
-importance of the bacterial community in both adenoma and carcinoma. We
-are also able to provide support for the driver-passenger model in the
-context of carcinoma. However, within the context of adenoma it is less
-clear that this relationship exists. These observations highlight the
-importance of power and sample number considerations when considering
-investigations into the microbiota and adenoma due to the subtle changes
-in the community.
+and tissue, we are able to provide evidence in support of the importance
+of the bacterial community in both adenoma and carcinoma. We are also
+able to provide support for the driver-passenger model in the context of
+carcinoma. However, within the context of adenoma, it is less clear that
+this relationship exists. These observations highlight the importance of
+power and sample number considerations when undertaking investigations
+into the microbiota and adenoma due to the subtle changes in the
+community. Although there are power limitations associated with adenoma,
+this report highlights the strong influence the microbiota has on CRC
+development.
 

### 2017-11-15 submission/manuscript.Rmd
add R package references

@@
 ***Statistical Analysis:***
-All statistical analysis after sequence processing utilized the R software package (v`r paste(version$major, ".", version$minor, sep = "")`).
-For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package.
-Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue.
-Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages by assessing how many with and without disease were above and below the overall median value within the specific study.
+All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
+For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
+Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue.
+Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many with and without disease were above and below the overall median value within the specific study.
 Relative risk significance testing utilized the chi-squred test.
-$\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package.
-Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages.
+$\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017].
+Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002].
 Differences between the obtained AUC versus a random model AUC was assessed using T-tests.
-Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages.
-All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages.
+Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016].
+All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].
 

### 2017-11-15 submission/manuscript.tex
add R package references

@@
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R software package (v3.4.2). For the
-\(\alpha\)-diversity analysis, values were power transformed using the
-rcompanion (v1.10.1) package and then Z-score normalized using the car
-(v2.1.5) package. Testing for \(\alpha\)-diversity differences utilized
-linear mixed-effect models created using the lme4 (v1.1.14) package to
-correct for study and variable region effects in feces and study,
-variable region, and individual effects in tissue. Relative risk was
-analyzed using both the epiR (v0.9.87) and metafor (v2.0.0) packages by
-assessing how many with and without disease were above and below the
-overall median value within the specific study. Relative risk
-significance testing utilized the chi-squred test. \(\beta\)-diversity
-differences utilized a Bray-Curtis distance matrix and PERMANOVA
-executed with the vegan (v2.4.4) package. Random Forest models were
-built using both the caret (v6.0.77) and randomForest (v4.6.12)
-packages. Differences between the obtained AUC versus a random model AUC
-was assessed using T-tests. Power analysis and estimations were made
-using the pwr (v1.2.1) and statmod (v1.4.30) packages. All figures were
-created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages.
+sequence processing utilized the R (v3.4.2) software package {[}33{]}.
+For the \(\alpha\)-diversity analysis, values were power transformed
+using the rcompanion (v1.10.1) package {[}34{]} and then Z-score
+normalized using the car (v2.1.5) package {[}35{]}. Testing for
+\(\alpha\)-diversity differences utilized linear mixed-effect models
+created using the lme4 (v1.1.14) package {[}36{]} to correct for study
+and variable region effects in feces and study, variable region, and
+individual effects in tissue. Relative risk was analyzed using both the
+epiR (v0.9.87) and metafor (v2.0.0) packages {[}37,38{]} by assessing
+how many with and without disease were above and below the overall
+median value within the specific study. Relative risk significance
+testing utilized the chi-squred test. \(\beta\)-diversity differences
+utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
+vegan (v2.4.4) package {[}39{]}. Random Forest models were built using
+both the caret (v6.0.77) and randomForest (v4.6.12) packages
+{[}40,41{]}. Differences between the obtained AUC versus a random model
+AUC was assessed using T-tests. Power analysis and estimations were made
+using the pwr (v1.2.1) and statmod (v1.4.30) packages {[}42,43{]}. All
+figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
+packages {[}44,45{]}.
 
@@
 
+\hypertarget{ref-r_citation_2017}{}
+33. R Core Team. R: A language and environment for statistical computing
+{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
+2017. Available from: \url{https://www.R-project.org/}
+
+\hypertarget{ref-rcompanion_citation_2017}{}
+34. Mangiafico S. Rcompanion: Functions to support extension education
+program evaluation {[}Internet{]}. 2017. Available from:
+\url{https://CRAN.R-project.org/package=rcompanion}
+
+\hypertarget{ref-car_citation_2011}{}
+35. Fox J, Weisberg S. An R companion to applied regression
+{[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
+\url{http://socserv.socsci.mcmaster.ca/jfox/Books/Companion}
+
+\hypertarget{ref-lme4_citation_2015}{}
+36. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
+models using lme4. Journal of Statistical Software. 2015;67:1--48.
+
+\hypertarget{ref-epir_citation_2017}{}
+37. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
+J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
+epidemiological data {[}Internet{]}. 2017. Available from:
+\url{https://CRAN.R-project.org/package=epiR}
+
+\hypertarget{ref-metafor_citation_2010}{}
+38. Viechtbauer W. Conducting meta-analyses in R with the metafor
+package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
+Available from: \url{http://www.jstatsoft.org/v36/i03/}
+
+\hypertarget{ref-vegan_citation_2017}{}
+39. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
+from: \url{https://CRAN.R-project.org/package=vegan}
+
+\hypertarget{ref-caret_citation_2017}{}
+40. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
+Cooper T, et al. Caret: Classification and regression training
+{[}Internet{]}. 2017. Available from:
+\url{https://CRAN.R-project.org/package=caret}
+
+\hypertarget{ref-randomforest_citation_2002}{}
+41. Liaw A, Wiener M. Classification and regression by randomForest. R
+News {[}Internet{]}. 2002;2:18--22. Available from:
+\url{http://CRAN.R-project.org/doc/Rnews/}
+
+\hypertarget{ref-pwr_citation_2017}{}
+42. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
+2017. Available from: \url{https://CRAN.R-project.org/package=pwr}
+
+\hypertarget{ref-statmod_citation_2016}{}
+43. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
+gaussian distribution. R Journal. 2016;8:339--51.
+
+\hypertarget{ref-ggplot2_citation_2009}{}
+44. Wickham H. Ggplot2: Elegant graphics for data analysis
+{[}Internet{]}. Springer-Verlag New York; 2009. Available from:
+\url{http://ggplot2.org}
+
+\hypertarget{ref-gridextra_citation_2017}{}
+45. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
+{[}Internet{]}. 2017. Available from:
+\url{https://CRAN.R-project.org/package=gridExtra}
+
 \newpage

### 2017-11-15 code/make_select_genus_RR_graph.R
add Nick suggestion Fig3

@@ -65,7 +65,7 @@ adn_stool_graph <- adn_all_stool %>%
   geom_point(size = 3, show.legend = F) + 
   labs(x = expression(Log["2"]~Relative~Risk), y = "") + theme_bw() + ggtitle("A") + 
   scale_color_manual(values = c('#000000', '#000000', '#000000', '#000000', 
-                                '#8B5A00', '#FFA500', '#FFC125', '#E3CF57')) + 
+                                '#778899', '#778899', '#778899', '#778899')) + 
   scale_y_discrete(labels=expression(All~4, At~Least~3, At~Least~2, At~Least~1, italic(Fusobacterium), 
                                      italic(Porphyromonas), italic(Peptostreptococcus), italic(Parvimonas))) +
   annotate("text", label = paste("Adenoma (Stool)"), x = -2.9, y = 8.4, size = 2.5) + 
@@ -91,7 +91,7 @@ crc_stool_graph <- crc_all_stool %>%
   geom_point(size = 3, show.legend = F) + 
   labs(x = expression(Log["2"]~Relative~Risk), y = "") + theme_bw() + ggtitle("B") + 
   scale_color_manual(values = c('#000000', '#000000', '#000000', '#000000', 
-                                '#B0171F', '#FF4040', '#CD5C5C', '#FFC1C1')) + 
+                                '#778899', '#778899', '#778899', '#778899')) + 
   scale_y_discrete(labels=expression(All~4, At~Least~3, At~Least~2, At~Least~1, italic(Fusobacterium), 
                                      italic(Porphyromonas), italic(Peptostreptococcus), italic(Parvimonas))) +
   annotate("text", label = paste("Carcinoma (Stool)"), x = -2.8, y = 8.4, size = 2.5) + 
@@ -117,7 +117,7 @@ adn_tissue_graph <- adn_all_tissue %>%
   geom_point(size = 3, show.legend = F) + 
   labs(x = expression(Log["2"]~Relative~Risk), y = "") + theme_bw() + ggtitle("C") + 
   scale_color_manual(values = c('#000000', '#000000', '#000000', '#000000', 
-                                '#8B5A00', '#FFA500', '#FFC125', '#E3CF57')) + 
+                                '#778899', '#778899', '#778899', '#778899')) + 
   scale_y_discrete(labels=expression(All~4, At~Least~3, At~Least~2, At~Least~1, italic(Fusobacterium), 
[…]

### 2017-11-15 code/make_rf_auc_full_versus_specific_OTU_graph.R
incorporate medians to make interpretation easier --Nick comment

@@ -90,6 +90,15 @@ crc_all_stool <- read_csv("data/process/tables/stool_rf_otu_random_comparison_su
 # weir - #24878EFF
 # ahn - #40BC72FF
 
+# Creates a median secondary table for adenoma tissue
+adn_tissue_medians <- adn_tissue %>% 
+  filter(model == "act") %>% group_by(model_type) %>% 
+  summarise(auc_median = median(AUC)) %>% 
+  mutate(model_type = factor(model_type, 
+                             levels = c("full", "select"), 
+                             labels = c("Full Community", "CRC Associated\nGenera Community Only")))
+
+# Creates the adenoma tissue graph
 adn_tissue_graph <- adn_tissue %>% 
   mutate(type = factor(type, 
                        levels = c("unmatched", "matched"), 
@@ -102,6 +111,7 @@ adn_tissue_graph <- adn_tissue %>%
                         labels = c("Flemer", "Lu\n(matched)"))) %>% 
   filter(grepl("rand", model) != T) %>% 
   ggplot(aes(study, AUC, color = study, group = study)) + 
+  geom_hline(data = adn_tissue_medians, aes(yintercept = auc_median), linetype = "solid", color = "red", alpha = 0.7) + 
   geom_point(size = 3.5, show.legend = F) + 
   geom_errorbar(aes(ymin = AUC-deviation, ymax = AUC+deviation), 
                 width = 0.25, size = 0.7, show.legend = F) + 
@@ -117,7 +127,15 @@ adn_tissue_graph <- adn_tissue %>%
         panel.grid.minor = element_blank(), 
         axis.text.y = element_text(size = 10))
 
+# Creates a median secondary table for adenoma stool
+adn_stool_medians <- adn_all_stool %>% 
+  filter(model == "act") %>% group_by(model_type) %>% 
+  summarise(auc_median = median(AUC)) %>% 
+  mutate(model_type = factor(model_type, 
+                             levels = c("full", "select"), 
+                             labels = c("Full Community", "CRC Associated\nGenera Community Only")))
 
+# Creates the adenoma stool graph
 adn_stool_graph <- adn_all_stool %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
@@ -127,11 +145,12 @@ adn_stool_graph <- adn_all_stool %>%
                         labels = c("Baxter", "Brim", "Hale", "Zeller"))) %>% 
[…]

### 2017-11-15 submission/manuscript.Rmd
update fig4 and fig5 legend
nick's abstract suggestions

@@
 
-**Background.** An increasing body of literature suggests that there is a role for the microbiota in colorectal cancer (CRC).
+**Background.** An increasing body of literature suggests that there is a crucial role for the microbiota in colorectal cancer (CRC) pathogenesis.
 Important drivers within this context have ranged from individual microbes to the whole community.
@@
 Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
-Using the whole community versus only CRC associated genera to build a prediction model resulted in a higher classification based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples.
+Using the whole community versus only CRC associated genera to build a prediction model resulted in higher classification success based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples.
 For the included studies, most were adequately powered for large effect size differences.
@@
 The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.
+The solid red line represents the median AUC of all the studies for either the full community or CRC-associated genera OTUS only model.
 

### 2017-11-15 submission/manuscript.tex
update fig4 and fig5 legend
nick's abstract suggestions

@@
 \textbf{Background.} An increasing body of literature suggests that
-there is a role for the microbiota in colorectal cancer (CRC). Important
-drivers within this context have ranged from individual microbes to the
-whole community. Our study expands on a recent meta-analysis
-investigating microbial biomarkers for CRC by testing the hypothesis
-that the bacterial community is an important driver of both early
-(adenoma) and late (carcinoma) stage of disease. To test this hypothesis
-we examined both feces (n = 1737) and tissue (492 total samples from 350
-individuals) across 14 different studies.
+there is a crucial role for the microbiota in colorectal cancer (CRC)
+pathogenesis. Important drivers within this context have ranged from
+individual microbes to the whole community. Our study expands on a
+recent meta-analysis investigating microbial biomarkers for CRC by
+testing the hypothesis that the bacterial community is an important
+driver of both early (adenoma) and late (carcinoma) stage of disease. To
+test this hypothesis we examined both feces (n = 1737) and tissue (492
+total samples from 350 individuals) across 14 different studies.
 
@@
 \textless{} 0.05). Using the whole community versus only CRC associated
-genera to build a prediction model resulted in a higher classification
-based on Area Under the Curve (AUC) for both adenoma and carcinoma using
-fecal and tissue samples. For the included studies, most were adequately
-powered for large effect size differences. This may be more amenable for
-carcinoma than for adenoma microbiota research due to the smaller
-community level changes observed in our results.
+genera to build a prediction model resulted in higher classification
+success based on Area Under the Curve (AUC) for both adenoma and
+carcinoma using fecal and tissue samples. For the included studies, most
+were adequately powered for large effect size differences. This may be
+more amenable for carcinoma than for adenoma microbiota research due to
+the smaller community level changes observed in our results.
 
@@
 dotted line represents an AUC of 0.5 and the lines represent the range
-in which the AUC for the 100 different 80/20 runs fell between.
+in which the AUC for the 100 different 80/20 runs fell between. The
+solid red line represents the median AUC of all the studies for either
+the full community or CRC-associated genera OTUS only model.
 

### 2017-11-15 submission/manuscript.Rmd
Nick's background suggestions

@@
 To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma.
-Next, we assessed how typical CRC associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma.
+Next, we assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma.
 Third, using Random Forest models, we analyzed whether the full community or only the CRC associated genera resulted in better model classification based on the area under the curve (AUC).
-Since the changes in community were subtle for adenoma we also examined what effect size and sample size the studies that were used were adequately powered for.
+Since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for.
 

### 2017-11-15 submission/manuscript.tex
Nick's background suggestions

@@
 in an increased relative risk (RR) for adenoma or carcinoma. Next, we
-assessed how typical CRC associated genera (\emph{Fusobacterium},
+assessed how common CRC-associated genera (\emph{Fusobacterium},
 \emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
@@
 area under the curve (AUC). Since the changes in community were subtle
-for adenoma we also examined what effect size and sample size the
-studies that were used were adequately powered for.
+for adenoma we also examined what effect and sample size the studies
+that were used were adequately powered for.
 

### 2017-11-15 submission/manuscript.Rmd
add Nick's result suggestions

@@
 Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S1 & S2].
-For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with matched tissue samples had no differences [Table S3 & S4].
+For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with tissue samples from the same individual (matched) had no differences [Table S3 & S4].
 
@@
 The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) than for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`).
-This decrease may be explained by the fact that the tissue analysis included matched samples.
+This decrease may be explained by the fact that the tissue analysis included matched samples from the same individual.
 
@@
 The data was split between feces and tissue samples.
-Within the tissue groups the data was further divided between matched and unmatched tissue samples.
+Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched) tissue samples.
 Where applicable for each study, predictions for adenoma and carcinoma were tested.

### 2017-11-15 submission/manuscript.tex
add Nick's result suggestions

@@
 \& S2{]}. For studies with unmatched tissue samples a similar trend was
-observed {[}Table S3 \& S4{]} while studies with matched tissue samples
-had no differences {[}Table S3 \& S4{]}.
+observed {[}Table S3 \& S4{]} while studies with tissue samples from the
+same individual (matched) had no differences {[}Table S3 \& S4{]}.
 
@@
 1.81). This decrease may be explained by the fact that the tissue
-analysis included matched samples.
+analysis included matched samples from the same individual.
 
@@
 samples. Within the tissue groups the data was further divided between
-matched and unmatched tissue samples. Where applicable for each study,
+samples from the same individual (matched) and those from different
+individuals (unmatched) tissue samples. Where applicable for each study,
 predictions for adenoma and carcinoma were tested. This same approach

### 2017-11-15 submission/manuscript.Rmd
complete Nick's comments on discussion, methods, and Declarations

@@
 
-Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease.
+Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease [Figure 1-3].
 Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the relative risk of either of these two conditions.
@@
 CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions.
-These observations suggest that small localized changes within the community may be occuring that are important in early disease progression of CRC and that this process may not directly involve CRC-associated genera.
+These observations suggest that small localized changes within the community on tissue may be occuring that are important in early disease progression of CRC and that this process may not directly involve CRC-associated genera.
 
@@
 This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma.
-Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out the role of the microbiota in adenoma formation.
+Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out how the microbiota contributes to adenoma formation.
 
@@
 ***Sequence Processing:***
-For the majority of studies raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the by searching the respective accession number of the study following website: http://www.ncbi.nlm.nih.gov/Traces/study/.
+For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the by searching the respective accession number of the study following website: http://www.ncbi.nlm.nih.gov/Traces/study/.
 Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap for one study [@ahn_human_2013] and for four studies was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013].
@@
 This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs.
-For OTU based models the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
+For OTU based models, the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
 The power of each study was assessed for an effect size ranging from 1% to 30%.
@@
 A detailed and reproducible description of how the data were processed and analyzed for each study can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
-Raw sequences can be downloaded from the SRA in most cases and can be found in the respective studies batch file in the GitHub repository or within the original publication.
+Raw sequences can be downloaded from the SRA in most cases and can be found in the respective study batch file in the GitHub repository or within the original publication.
 For instances when sequences are not publicly available, they may be accessed by contacting the corresponding authors from whence the data came.
@@
 
-The authors would like to thank all the study participants who were a part of each of the individual studies uitlized.
+The authors would like to thank all the study participants who were a part of each of the individual studies utilized.
 We would also like to thank each of the study authors for making their data available for use.

### 2017-11-15 submission/manuscript.tex
complete Nick's comments on discussion, methods, and Declarations

@@
 community level and within individual genera that are present in
-individuals with CRC versus those without the disease. Although there
-was a step-wise decrease in diversity from control to adenoma to
-carcinoma, this did not translate into large effect sizes for the
-relative risk of either of these two conditions. Even though
+individuals with CRC versus those without the disease {[}Figure 1-3{]}.
+Although there was a step-wise decrease in diversity from control to
+adenoma to carcinoma, this did not translate into large effect sizes for
+the relative risk of either of these two conditions. Even though
 CRC-associated genera increase the RR of carcinoma, they do not
@@
 ability of models to make predictions. These observations suggest that
-small localized changes within the community may be occuring that are
-important in early disease progression of CRC and that this process may
-not directly involve CRC-associated genera.
+small localized changes within the community on tissue may be occuring
+that are important in early disease progression of CRC and that this
+process may not directly involve CRC-associated genera.
 
@@
 Future studies investigating adenoma and the microbiota need to take
-these factors into consideration if we are to work out the role of the
-microbiota in adenoma formation.
+these factors into consideration if we are to work out how the
+microbiota contributes to adenoma formation.
 
@@
 
-\textbf{\emph{Sequence Processing:}} For the majority of studies raw
+\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
 sequences were downloaded from the Sequence Read Archive (SRA)
@@
 and 100 different models were created based on random 80/20 splitting of
-the data to generate a range of expected AUCs. For OTU based models the
+the data to generate a range of expected AUCs. For OTU based models, the
 CRC associated genera included all OTUs that had a taxonomic
@@
 Raw sequences can be downloaded from the SRA in most cases and can be
-found in the respective studies batch file in the GitHub repository or
+found in the respective study batch file in the GitHub repository or
 within the original publication. For instances when sequences are not
@@
 The authors would like to thank all the study participants who were a
-part of each of the individual studies uitlized. We would also like to
+part of each of the individual studies utilized. We would also like to
 thank each of the study authors for making their data available for use.

### 2017-11-15 code/make_stool_alpha_RR_graph.R
add legend ala Nick suggestion

@@ -50,18 +50,19 @@ adn_stool_graph <- adn_all_stool %>%
                         levels = c("composite", "zeller", "hale", "brim", "baxter"), 
                         labels = c( "Pooled", "Zeller", "Hale", "Brim", "Baxter")), 
          region = factor(region, 
-                         levels = c("combined", "V4", "V3-V5", "V1-V3")), 
+                         levels = c("combined", "V4", "V3-V5", "V1-V3"), 
+                         labels = c("Combined", "V4", "V3-V5", "V1-V3")), 
          measure = factor(measure, 
                           levels = c("sobs", "shannoneven", "shannon"), 
                           labels = c("Observed OTUs", "Evenness", "Shannon Diversity"))) %>% 
   ggplot(aes(log2(est), study, xmax=log2(upper), xmin=log2(lower), colour=region)) + 
   coord_cartesian(xlim=c(-4.2, 4.2)) + 
   geom_vline(xintercept = 0.0, linetype=2, alpha=0.75) + 
-  geom_errorbarh(alpha=0.5, size = 1, height=0, show.legend = F) + 
-  geom_point(size = 3, show.legend = F) + 
+  geom_errorbarh(alpha=0.5, size = 1, height=0, show.legend = T) + 
+  geom_point(size = 3, show.legend = T) + 
   facet_grid(. ~ measure) + 
   labs(x = expression(Log["2"]~Relative~Risk), y = "") + theme_bw() + ggtitle("A") + 
-  scale_color_manual(values = c( '#000000', '#440154FF', '#443A83FF', '#31688EFF')) + 
+  scale_color_manual(name = "Variable Region", values = c( '#000000', '#440154FF', '#443A83FF', '#31688EFF')) + 
   annotate("text", label = paste("Adenoma (Stool)"), x = -2.75, y = 5.5, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
@@ -77,18 +78,20 @@ crc_stool_graph <- crc_all_stool %>%
                         labels = c( "Pooled", "Zeller", "Weir", "Wang",  "Hale", 
                                     "Flemer", "Baxter", "Ahn")), 
          region = factor(region, 
-                         levels = c("combined", "V3", "V4", "V3-V4", "V3-V5")), 
[…]

### 2017-11-15 code/make_tissue_alpha_RR_graph.R
add legend a la Nick suggestion

@@ -50,17 +50,20 @@ adn_tissue_graph <- adn_all_tissue %>%
   mutate(study = factor(study, 
                         levels = c("composite", "flemer", "lu"), 
                         labels = c( "Pooled", "Flemer", "Lu")), 
+         region = factor(region, 
+                         levels = c("combined", "V3-V4"), 
+                         labels = c("Combined", "V3-V4")), 
          measure = factor(measure, 
                           levels = c("sobs", "shannoneven", "shannon"), 
                           labels = c("Observed OTUs", "Evenness", "Shannon Diversity")))  %>% 
   ggplot(aes(log2(est), study, xmax=log2(upper), xmin=log2(lower), colour=region)) + 
   coord_cartesian(xlim=c(-4.2, 4.2)) + 
   geom_vline(xintercept = 0.0, linetype=2, alpha=0.75) + 
-  geom_errorbarh(alpha=0.5, size = 1, height=0, show.legend = F) + 
-  geom_point(size = 3, show.legend = F) + 
+  geom_errorbarh(alpha=0.5, size = 1, height=0, show.legend = T) + 
+  geom_point(size = 3, show.legend = T) + 
   facet_grid(. ~ measure) + 
   labs(x = expression(Log["2"]~Relative~Risk), y = "") + theme_bw() + ggtitle("A") + 
-  scale_color_manual(values = c('#000000', '#35B779FF')) + 
+  scale_color_manual(name = "Variable Region", values = c('#000000', '#35B779FF')) + 
   annotate("text", label = paste("Adenoma (Tissue)"), x = -2.70, y = 3.5, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
@@ -73,18 +76,19 @@ crc_tissue_graph <- crc_all_tissue %>%
                         levels = c("composite", "sana", "geng", "flemer", "dejea", "chen", "burns"), 
                         labels = c( "Pooled", "Sanapareddy", "Geng", "Flemer", "Dejea", "Chen", "Burns")), 
          region = factor(region, 
-                         levels = c("combined", "V1-V2", "V1-V3", "V3-V4", "V3-V5", "V5-V6")), 
[…]

### 2017-11-27 submission/manuscript.Rmd
fix a final typo
update with Pat's comments 3-6

@@
 ***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
+Using the combined data set we first assessed whether there were any broad scale community differences that could be detected as disease severity worsened.
 Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces, but not tissue, were  lower in those with carcinoma [Figure 1].
 Using linear mixed-effect models to control for study and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`).
-However, in tissue, this effect was not observed in tissue when resampling of the same individual was also controlled for (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
+However, in tissue, this effect was not observed when resampling of the same individual was also controlled for (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
+We next tested whether these detectable differences in the community resulted in a significant increase in RR.
 Within fecal samples, a decrease in Shannon diversity and evenness resulted in a significantly increased RR for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2].
-Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2].
-Only a decrease in evenness had an increased RR for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))`).
+Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`) and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)) [Figure 2].
+Only a decrease in evenness had an increased RR for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)).
+Interestingly, for both adenoma and carcinoma there was no increase in RR within tissue samples for any alpha diversity metric investigated [Table S1-S3].
 
-Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S1 & S2].
-For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with tissue samples from the same individual (matched) had no differences [Table S3 & S4].
+Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S4 & S5].
+For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with tissue samples from the same individual (matched) had no differences [Table S6 & S7].
 
 ***Carcinoma-Associated Genera Minimally Impacts RR of Adenoma:***
+Based on the small increase in RR using $\alpha$-diversity metrics, we wanted to know if the presence of specific genera resulted in a higher RR for both stool and tissue.
+To investigate this we analyzed the classically associated CRC genera, *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for and increase in RR.
 The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3].
@@
 ***Using the Whole Community Instead of Only CRC-Associated Genera Increases Model AUC:***
-For both fecal and tissue (matched and unmatched) samples, the AUC decreases when only OTUs from the CRC-associated genera are used [Figure 4 & 5].
-This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5].
-The genus models generally performed similar to the OTU-based models with the full genera models performing better than the CRC-associated genera models [Figure S2-S3].
-Both genus models were similarily predictive in their ability to detect adenomas or carcinomas, with carcinoma having a higher AUC then adenoma [Figure S4-S5].
-Of note, matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S4A, S5B, & S6] than carcinoma models [Figure S4B & S5A].
+We then tested whether the overall bacterial community was at all important to classifying disease or if the CRC-associated genera were sufficient alone.
+To test this we used two approaches.
+The first used genus level data and tested whether there were any differences in AUC when training on one study and testing on all the others when using either all genera present or only the CRC-associated genera.
+The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model across studies using either all OTUs or only OTUs that taxonomically classified to CRC-associated genera.
+
+The genus based models showed an AUC decrease in model classification on the training set for both stool and tissue studies [Figure S2-S3].
+With respect to the test sets, comprised of genera data from other studies, both the all genera model and CRC-associated models had a similar ability to detect adenomas or carcinomas [Figure S4-S6].
+The AUC for classification of adenomas was lower than carcionmas for both tissue and stool [Figure S4-S6].
+
+The OTU based models for both fecal and tissue (matched and unmatched) samples, showed an AUC decrease when only OTUs from the CRC-associated genera are used versus full community of OTUs [Figure 4 & 5].
+This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5] with the largest difference in median AUC for stool being in carcinoma classification [Figure 4B] and for tissue being in adenoma [Figure 5A].
 
 ***A Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
+Based on the previous observations we then assessed whether the studies that we included are realistically powered to identify small, medium, and large scale differences between case and control.
 When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B].
@@
 An estimated sample n for these effect sizes was also generated based on 80% power.
+For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study.
+Similarily, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study.
+
+***Obtaining CRC-Associated Genera:***
+For the CRC-associated genera analysis of the RR, the total average counts were collected for each respective OTU that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for 100 different subsamplings.
+The OTU based Random Forest Models that used CRC-associated genera used a similar approach except that the OTUs were not aggregated together by genus by kept as separate OTUs.
+So, OTU Random Forest models using the full community included all OTUs while those using CRC-associated genera included only those OTUs that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.
+
+***Matched versus Unmatched Tissue Samples:***
+In general, tissue samples that had control and lesion samples that did not belong to the same indivdiual were classified as unmatched while samples that belonged to the same individual were classified as matched.
+Studies with matched data included Burns, Dejea, Geng, and Lu.
+Studies that had unmatched data were Burns, Flemer, Chen, and Sanapareddy.
+For some studies samples became unmatched due to one of the corresponding matched samples not making it through sequence processing.
+For the linear mixed-effect models samples from the same individual were taken into account.
+For all other analysis matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.
 
@@
 B) Power based on effect size for studies with carcinoma individuals.
-C) The estimated sample number needed for each arm of each study to detect aneffect size of 1-30%.
+C) The estimated sample number needed for each arm of each study to detect an effect size of 1-30%.
 The dotted red lines in A) and B) represent a power of 0.8.
@@
 The x-axis is the data set in which the model was initially trained on.
+The red lines represent the median AUC using that specific study as the training set.
 
@@
 The x-axis is the data set in which the model was initially trained on.
+The red lines represent the median AUC using that specific study as the training set.
 
-**Figure S6: Random Forest Prediction Success of Adenoma Using Genera for each Tissue Study.**
+**Figure S6: Random Forest Prediction Success of Adenoma Using Genera for each Tissue Study.** The red lines represent the median AUC using that specific study as the training set.
 

### 2017-11-27 submission/manuscript.tex
fix a final typo
update with Pat's comments 3-6

@@
 \textbf{\emph{Lower Community Diversity is Associated with Increased RR
-of Carcinomas:}} Using power transformed and Z-score normalized
-\(\alpha\)-diversity metrics, both evenness and Shannon diversity in
-feces, but not tissue, were lower in those with carcinoma {[}Figure
-1{]}. Using linear mixed-effect models to control for study and variable
-region, there was a significant decrease from control to adenoma to
-carcinoma for both evenness (P-value = 0.025) and Shannon diversity
-(P-value = 0.043). However, in tissue, this effect was not observed in
-tissue when resampling of the same individual was also controlled for
-(P-value \textgreater{} 0.05). Within fecal samples, a decrease in
-Shannon diversity and evenness resulted in a significantly increased RR
-for carcinoma (P-value = 0.01 and 0.0011, respectively) {[}Figure 2{]}.
-Although these values were significant, the effect size was relatively
-small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
+of Carcinomas:}} Using the combined data set we first assessed whether
+there were any broad scale community differences that could be detected
+as disease severity worsened. Using power transformed and Z-score
+normalized \(\alpha\)-diversity metrics, both evenness and Shannon
+diversity in feces, but not tissue, were lower in those with carcinoma
+{[}Figure 1{]}. Using linear mixed-effect models to control for study
+and variable region, there was a significant decrease from control to
+adenoma to carcinoma for both evenness (P-value = 0.025) and Shannon
+diversity (P-value = 0.043). However, in tissue, this effect was not
+observed when resampling of the same individual was also controlled for
+(P-value \textgreater{} 0.05). We next tested whether these detectable
+differences in the community resulted in a significant increase in RR.
+Within fecal samples, a decrease in Shannon diversity and evenness
+resulted in a significantly increased RR for carcinoma (P-value = 0.0047
+and 0.00034, respectively) {[}Figure 2{]}. Although these values were
+significant, the effect size was relatively small for both metrics
+(Shannon RR = 1.33 (1.09 - 1.62) and evenness RR = 1.36 (1.15 - 1.61))
 {[}Figure 2{]}. Only a decrease in evenness had an increased RR for
 adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect size was
-even smaller than what was observed for carcinoma (RR = 1.16).
+even smaller than what was observed for carcinoma (RR = 1.16 (1.01 -
+1.34)). Interestingly, for both adenoma and carcinoma there was no
+increase in RR within tissue samples for any alpha diversity metric
+investigated {[}Table S1-S3{]}.
 
@@
 difference across studies in the bacterial community of fecal samples
-between carcinoma and controls, but not adenoma and controls {[}Table S1
-\& S2{]}. For studies with unmatched tissue samples a similar trend was
+between carcinoma and controls, but not adenoma and controls {[}Table S4
+\& S5{]}. For studies with unmatched tissue samples a similar trend was
 observed {[}Table S3 \& S4{]} while studies with tissue samples from the
-same individual (matched) had no differences {[}Table S3 \& S4{]}.
+same individual (matched) had no differences {[}Table S6 \& S7{]}.
 
 \textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
-Adenoma:}} The majority of CRC-associated genera for both feces and
-tissue had a significantly increased RR for carcinoma but not for
-adenoma {[}Figure 3{]}. In fecal samples the RR due to CRC associated
-genera was greater than either the RR assoicated with evenness or
-Shannon diversity {[}Figure 2 \& 3{]}. Additionally, the RR of carcinoma
-continuously increased as individuals tested positive for more CRC
-associated genera {[}Figure 3B \& 3D{]}. The RR effect size was greater
-for stool (RR range = 1.61 - 2.74) than for tissue (RR range = 1.21 -
-1.81). This decrease may be explained by the fact that the tissue
+Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
+metrics, we wanted to know if the presence of specific genera resulted
+in a higher RR for both stool and tissue. To investigate this we
+analyzed the classically associated CRC genera, \emph{Fusobacterium},
+\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}
+for and increase in RR. The majority of CRC-associated genera for both
+feces and tissue had a significantly increased RR for carcinoma but not
+for adenoma {[}Figure 3{]}. In fecal samples the RR due to CRC
+associated genera was greater than either the RR assoicated with
+evenness or Shannon diversity {[}Figure 2 \& 3{]}. Additionally, the RR
+of carcinoma continuously increased as individuals tested positive for
+more CRC associated genera {[}Figure 3B \& 3D{]}. The RR effect size was
+greater for stool (RR range = 1.62 - 2.37) than for tissue (RR range =
+1.21 - 1.81). This decrease may be explained by the fact that the tissue
 analysis included matched samples from the same individual.
@@
 \textbf{\emph{Using the Whole Community Instead of Only CRC-Associated
-Genera Increases Model AUC:}} For both fecal and tissue (matched and
-unmatched) samples, the AUC decreases when only OTUs from the
-CRC-associated genera are used {[}Figure 4 \& 5{]}. This decrease is
-observed in both adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The
-genus models generally performed similar to the OTU-based models with
-the full genera models performing better than the CRC-associated genera
-models {[}Figure S2-S3{]}. Both genus models were similarily predictive
-in their ability to detect adenomas or carcinomas, with carcinoma having
-a higher AUC then adenoma {[}Figure S4-S5{]}. Of note, matched tissue
-samples for those with carcinoma had an AUC that was more similar to the
-adenoma models {[}Figure S4A, S5B, \& S6{]} than carcinoma models
-{[}Figure S4B \& S5A{]}.
+Genera Increases Model AUC:}} We then tested whether the overall
+bacterial community was at all important to classifying disease or if
+the CRC-associated genera were sufficient alone. To test this we used
+two approaches. The first used genus level data and tested whether there
+were any differences in AUC when training on one study and testing on
+all the others when using either all genera present or only the
+CRC-associated genera. The second approach used OTU level data and
+tested whether there was a generalized decrease in the 10-fold cross
+validation (CV) model across studies using either all OTUs or only OTUs
+that taxonomically classified to CRC-associated genera.
+
+The genus based models showed an AUC decrease in model classification on
+the training set for both stool and tissue studies {[}Figure S2-S3{]}.
+With respect to the test sets, comprised of genera data from other
+studies, both the all genera model and CRC-associated models had a
+similar ability to detect adenomas or carcinomas {[}Figure S4-S6{]}. The
+AUC for classification of adenomas was lower than carcionmas for both
+tissue and stool {[}Figure S4-S6{]}.
+
+The OTU based models for both fecal and tissue (matched and unmatched)
+samples, showed an AUC decrease when only OTUs from the CRC-associated
+genera are used versus full community of OTUs {[}Figure 4 \& 5{]}. This
+decrease is observed in both adenoma and carcinoma groups {[}Figure 4 \&
+5{]} with the largest difference in median AUC for stool being in
+carcinoma classification {[}Figure 4B{]} and for tissue being in adenoma
+{[}Figure 5A{]}.
 
 \textbf{\emph{A Majority of Studies are Underpowered for Detecting Small
-Effect Size Differences:}} When assessing the power of each study at
-different effect sizes the majority of studies for both adenoma and
-carcinoma have an 80\% power to detect a 30\% difference {[}Figure 6A \&
-B{]}. No single study that was analyzed had the standard 80\% power to
-detect an effect size difference that was equal to or below 10\%
-{[}Figure 6A \& B{]}. In order to achieve adequate power for small
-effect sizes, studies would need to recruit over 1000 individuals for
-each arm {[}Figure 6C{]}.
+Effect Size Differences:}} Based on the previous observations we then
+assessed whether the studies that we included are realistically powered
+to identify small, medium, and large scale differences between case and
+control. When assessing the power of each study at different effect
+sizes the majority of studies for both adenoma and carcinoma have an
+80\% power to detect a 30\% difference {[}Figure 6A \& B{]}. No single
+study that was analyzed had the standard 80\% power to detect an effect
+size difference that was equal to or below 10\% {[}Figure 6A \& B{]}. In
+order to achieve adequate power for small effect sizes, studies would
+need to recruit over 1000 individuals for each arm {[}Figure 6C{]}.
 
@@
 estimated sample n for these effect sizes was also generated based on
-80\% power.
+80\% power. For comparisons in which normal versus adenoma were made the
+carcinoma samples were excluded from each respective study. Similarily,
+for comparisons in which normal versus carcinoma were made the adenoma
+samples were excluded from each respective study.
+
+\textbf{\emph{Obtaining CRC-Associated Genera:}} For the CRC-associated
+genera analysis of the RR, the total average counts were collected for
+each respective OTU that had a genus level taxonomic classification to
+\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, and
+\emph{Porphyromonas} for 100 different subsamplings. The OTU based
+Random Forest Models that used CRC-associated genera used a similar
+approach except that the OTUs were not aggregated together by genus by
+kept as separate OTUs. So, OTU Random Forest models using the full
+community included all OTUs while those using CRC-associated genera
+included only those OTUs that had a genus level taxonomic classification
+to \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus},
+and \emph{Porphyromonas}.
+
+\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
+tissue samples that had control and lesion samples that did not belong
+to the same indivdiual were classified as unmatched while samples that
+belonged to the same individual were classified as matched. Studies with
+matched data included Burns, Dejea, Geng, and Lu. Studies that had
+unmatched data were Burns, Flemer, Chen, and Sanapareddy. For some
+studies samples became unmatched due to one of the corresponding matched
+samples not making it through sequence processing. For the linear
+mixed-effect models samples from the same individual were taken into
+account. For all other analysis matched and unmatched samples were
+analyzed separately using the statistical approaches mentioned in the
+Statistical Analysis section.
 
@@
 The estimated sample number needed for each arm of each study to detect
-aneffect size of 1-30\%. The dotted red lines in A) and B) represent a
+an effect size of 1-30\%. The dotted red lines in A) and B) represent a
 power of 0.8.
@@
 of 0.5. The x-axis is the data set in which the model was initially
-trained on.
+trained on. The red lines represent the median AUC using that specific
+study as the training set.
 
@@
 dotted line represents an AUC of 0.5. The x-axis is the data set in
-which the model was initially trained on.
+which the model was initially trained on. The red lines represent the
+median AUC using that specific study as the training set.
 
 \textbf{Figure S6: Random Forest Prediction Success of Adenoma Using
-Genera for each Tissue Study.}
+Genera for each Tissue Study.} The red lines represent the median AUC
+using that specific study as the training set.
 

### 2017-11-27 submission/supplemental.Rmd
update title to match change in manuscript
update with Pat's comments 3-6

--- 2017-10-31 submission/supplemental.Rmd
@@
 
-**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Adenoma**
+**Table S1: Alpha Diversity Metrics RR in Tissue for Adenoma**
 
@@
 
+alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_RR_tissue_composite.csv")) %>% 
+  mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = round(pvalue, digits = 2), 
+         measure = c("OTU Richness", "Shannon Diversity", "Evenness"))
+
+colnames(alpha_RR_adn_tissue) <- c("Relative Risk", "Lower Bound", "Upper Bound", "P-value", "Measure")
+
+kable(alpha_RR_adn_tissue, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S2: Alpha Diversity Metrics RR in Tissue for Unmatched Carcinoma Non-Carcinoma Samples**
+
+```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
+
+alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_RR_unmatched_tissue_composite.csv")) %>%   mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = round(pvalue, digits = 2), 
+         measure = c("OTU Richness", "Shannon Diversity", "Evenness"))
+
+colnames(alpha_RR_crc_unmatch_tissue) <- c("Relative Risk", "Lower Bound", "Upper Bound", "P-value", "Measure")
+
+kable(alpha_RR_crc_unmatch_tissue, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S3: Alpha Diversity Metrics RR in Tissue for Matched Carcinoma Non-Carcinoma Samples**
+
+```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
+
+alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_RR_matched_tissue_composite.csv")) %>% 
+  mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = round(pvalue, digits = 2), 
+         measure = c("OTU Richness", "Shannon Diversity", "Evenness"))
+
+colnames(alpha_RR_crc_match_tissue) <- c("Relative Risk", "Lower Bound", "Upper Bound", "P-value", "Measure")
+
+kable(alpha_RR_crc_match_tissue, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S4: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Adenoma**
+
+```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
+
 temp_t1 <- read_csv("../data/process/tables/beta_perm_adn_stool_summary.csv")
 
-table_s1 <- temp_t1 %>% 
+table_s4 <- temp_t1 %>% 
   mutate(fstat = round(fstat, digits = 2), 
@@
 
-colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Adenoma n", "Control n")
+colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Adenoma (n)", "Control (n)")
 
-kable(table_s1, format = "markdown", align = 'c', padding = 2)
+kable(table_s4, format = "markdown", align = 'c', padding = 2)
 ```
@@
 
-**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Carcinoma**
+**Table S5: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Carcinoma**
 
-```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
+```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-table_s2 <- temp_t2 %>% 
+table_s5 <- temp_t2 %>% 
   mutate(fstat = round(fstat, digits = 2), 
@@
 
-colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma n", "Control n")
+colnames(table_s5) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma (n)", "Control (n)")
 
-kable(table_s2, format = "markdown", align = 'c', padding = 2)
+kable(table_s5, format = "markdown", align = 'c', padding = 2)
 ```
@@
 
-**Table S3: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue Adenoma**
+**Table S6: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue Adenoma**
 
-```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
+```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-table_s3 <- temp_t3a %>% 
+table_s6 <- temp_t3a %>% 
   bind_rows(temp_t3b) %>% 
@@
 
-colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma n", "Control n")
+colnames(table_s6) <- c("F statistic", "R2", "P-value", "Study", "Adenoma (n)", "Control (n)")
 
-kable(table_s3, format = "markdown", align = 'c', padding = 2)
+kable(table_s6, format = "markdown", align = 'c', padding = 2)
 ```
@@
 
-\newcommand{\blandscape}{\begin{landscape}}
-\newcommand{\elandscape}{\end{landscape}}
+**Table S7: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue Carcinoma**
 
-**Table S2: Summary of Common OTUs for All Diagnosis Groups**
+```{r supplemental_table7, echo=FALSE, message=FALSE, warning=FALSE}
 
-\footnotesize
+temp_t4a <- read_csv("../data/process/tables/beta_perm_unmatched_tissue_summary.csv")
+temp_t4b <- read_csv("../data/process/tables/beta_perm_matched_tissue_summary.csv")
 
-```{r supplemental_table, echo=FALSE, message=FALSE, warning=FALSE}
+table_s7 <- temp_t4a %>% 
+  bind_rows(temp_t4b) %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = c("Sanapareddy", "Burns", "Flemer", "Chen", 
+                   "Dejea (Matched)", "Geng (Matched)", "Burns (Matched)"))
+
+colnames(table_s7) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma (n)", "Control (n)")
+
+kable(table_s7, format = "markdown", align = 'c', padding = 2)
 
 ```
-\normalsize
-\newpage
-
-**Table S4: Summary of Adenoma based on Surgery Received**
-
-```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
-
-```
-
-\newpage
-
-**Table S5: Summary of mtry ranges used for the treatment and diagnosis models**
-
-\footnotesize
-
-```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
-
-```
-\normalsize

### 2017-11-27 code/run_comparison_RF_imp_otus_stool.R
first commit

@@ -1,0 +1,27 @@
+### Run Random Forest Analysis OTU Level - Tissue
+### Generate model and then test on remaining studies
+### Marc Sze
+
+## For tissue set CV to 5 since there are in general less samples per study
+
+# Load in needed functions and libraries
+source('code/functions.R')
+
+# Load needed libraries
+loadLibs(c("tidyverse", "caret", "pROC"))
+
+
+##############################################################################################
+########################## Group of Functions needed to run the analysis #####################
+##############################################################################################
+
+
+
+
+
+
+
+
+##############################################################################################
+########################## Code used to run the analysis  ####################################
+##############################################################################################

### 2017-11-28 submission/manuscript.Rmd
add imp otu and new figs

@@
 
+In stool the most common genera in the genus based models belonged mostly to resident genera [Figure 6A & B].
+This included genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia*.
+With respect to the CRC-associated genera *Fusobacterium* was the only genus present in adenoma while all four were present in carcinoma [Figure 6A & B].
+However, none of these CRC-associated genera were present in the majority of studies.
+For the adenoma OTU models, OTUs that classified as  *Ruminococcaceae* or *Roseburia* were present in the top 10 OTUs for the vast majority of studies [Figure 6C].
+Although for the carcinoma OTU models *Ruminococcaceae* was in the top 10 for many studies, *Bacteroides* was present in the overwhelming majority of the carcinoma OTU stool models [Figure 6D].
+
+Conversely to the stool models, both genera and OTU based models in tissue had the vast majority of their top 10 occur in a study specific manner [Figure S7].
+*Fusobacterium* and *Fusobacteriaceae* show up more often in the top 10 for matched tissue samples but are not present in the top 10 or much lower for unmatched tissue [Figure S7B-C & S7E-F].
+There appears to be very little overlap in the top 10 most important variables between stool and tissue for both adenoma and carcinoma [Figure 6 & S7].
+
 ***A Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
 Based on the previous observations we then assessed whether the studies that we included are realistically powered to identify small, medium, and large scale differences between case and control.
-When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B].
-No single study that was analyzed had the standard 80% power to detect an effect size difference that was equal to or below 10% [Figure 6A & B].
-In order to achieve adequate power for small effect sizes, studies would need to recruit over 1000 individuals for each arm [Figure 6C].
+When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 7A & B].
+No single study that was analyzed had the standard 80% power to detect an effect size difference that was equal to or below 10% [Figure 7A & B].
+In order to achieve adequate power for small effect sizes, studies would need to recruit over 1000 individuals for each arm [Figure 7C].
 
@@
 For all other analysis matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.
+
+***Assessing Important Random Forest Model Variables:***
+The genus based models collected the top 10 most important variables from each training set and assessed how many times that genera showed up in the top 10 across each study.
+The OTU based models recorded the medians for each OTU across 100 different 80/20 splits of the data for each study.
+The lowest classification for each OTU was obtained using the RDP database and the number of times the specific classification occured in the top 10 across studies was recorded.
+For the adenoma tissue genus and OTU models there was only one matched and unmatched study and these results were grouped together for the counting of the top 10.
 
@@
 
-**Figure 6: Power and Effect Size Analysis of Studies Included.** A) Power based on effect size for studies with adenoma individuals.
+**Figure 6: Most Common Genera Across Full Community Stool Study Models.** A) Common genera in the top 10 for adenoma Random Forest genus models.
+B) Common genera in the top 10 for carcinoma Random Forest genus models.
+C) Common genera in the top 10 for adenoma Random Forest OTU models.
+D) Common genera in the top 10 for carcinoma Random Forest OTU models.
+
+**Figure 7: Power and Effect Size Analysis of Studies Included.** A) Power based on effect size for studies with adenoma individuals.
 B) Power based on effect size for studies with carcinoma individuals.
@@
 
+**Figure S7: Most Common Genera Across Full Community Tissue Study Models.** A) Common genera in the top 10 for adenoma Random Forest genus models.
+B) Common genera in the top 10 for unmatched carcinoma Random Forest genus models.
+B) Common genera in the top 10 for matched carcinoma Random Forest genus models.
+D) Common genera in the top 10 for adenoma Random Forest OTU models.
+E) Common genera in the top 10 for unmatched carcinoma Random Forest OTU models.
+F) Common genera in the top 10 for matched carcinoma Random Forest OTU models.
+
 \newpage

### 2017-11-28 submission/manuscript.tex
add imp otu and new figs

@@
 
+In stool the most common genera in the genus based models belonged
+mostly to resident genera {[}Figure 6A \& B{]}. This included genera
+such as \emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia}.
+With respect to the CRC-associated genera \emph{Fusobacterium} was the
+only genus present in adenoma while all four were present in carcinoma
+{[}Figure 6A \& B{]}. However, none of these CRC-associated genera were
+present in the majority of studies. For the adenoma OTU models, OTUs
+that classified as \emph{Ruminococcaceae} or \emph{Roseburia} were
+present in the top 10 OTUs for the vast majority of studies {[}Figure
+6C{]}. Although for the carcinoma OTU models \emph{Ruminococcaceae} was
+in the top 10 for many studies, \emph{Bacteroides} was present in the
+overwhelming majority of the carcinoma OTU stool models {[}Figure 6D{]}.
+
+Conversely to the stool models, both genera and OTU based models in
+tissue had the vast majority of their top 10 occur in a study specific
+manner {[}Figure S7{]}. \emph{Fusobacterium} and \emph{Fusobacteriaceae}
+show up more often in the top 10 for matched tissue samples but are not
+present in the top 10 or much lower for unmatched tissue {[}Figure S7B-C
+\& S7E-F{]}. There appears to be very little overlap in the top 10 most
+important variables between stool and tissue for both adenoma and
+carcinoma {[}Figure 6 \& S7{]}.
+
 \textbf{\emph{A Majority of Studies are Underpowered for Detecting Small
@@
 sizes the majority of studies for both adenoma and carcinoma have an
-80\% power to detect a 30\% difference {[}Figure 6A \& B{]}. No single
+80\% power to detect a 30\% difference {[}Figure 7A \& B{]}. No single
 study that was analyzed had the standard 80\% power to detect an effect
-size difference that was equal to or below 10\% {[}Figure 6A \& B{]}. In
+size difference that was equal to or below 10\% {[}Figure 7A \& B{]}. In
 order to achieve adequate power for small effect sizes, studies would
-need to recruit over 1000 individuals for each arm {[}Figure 6C{]}.
+need to recruit over 1000 individuals for each arm {[}Figure 7C{]}.
 
@@
 
+\textbf{\emph{Assessing Important Random Forest Model Variables:}} The
+genus based models collected the top 10 most important variables from
+each training set and assessed how many times that genera showed up in
+the top 10 across each study. The OTU based models recorded the medians
+for each OTU across 100 different 80/20 splits of the data for each
+study. The lowest classification for each OTU was obtained using the RDP
+database and the number of times the specific classification occured in
+the top 10 across studies was recorded. For the adenoma tissue genus and
+OTU models there was only one matched and unmatched study and these
+results were grouped together for the counting of the top 10.
+
 \textbf{\emph{Reproducible Methods:}} The code and analysis can be found
@@
 
-\textbf{Figure 6: Power and Effect Size Analysis of Studies Included.}
+\textbf{Figure 6: Most Common Genera Across Full Community Stool Study
+Models.} A) Common genera in the top 10 for adenoma Random Forest genus
+models. B) Common genera in the top 10 for carcinoma Random Forest genus
+models. C) Common genera in the top 10 for adenoma Random Forest OTU
+models. D) Common genera in the top 10 for carcinoma Random Forest OTU
+models.
+
+\textbf{Figure 7: Power and Effect Size Analysis of Studies Included.}
 A) Power based on effect size for studies with adenoma individuals. B)
@@
 
+\textbf{Figure S7: Most Common Genera Across Full Community Tissue Study
+Models.} A) Common genera in the top 10 for adenoma Random Forest genus
+models. B) Common genera in the top 10 for unmatched carcinoma Random
+Forest genus models. B) Common genera in the top 10 for matched
+carcinoma Random Forest genus models. D) Common genera in the top 10 for
+adenoma Random Forest OTU models. E) Common genera in the top 10 for
+unmatched carcinoma Random Forest OTU models. F) Common genera in the
+top 10 for matched carcinoma Random Forest OTU models.
+
 \newpage

### 2017-11-29 submission/manuscript.Rmd
re write discussion and conclusion

@@
 The data presented herein supports the driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], when applied to carcinoma but not necessarily adenoma.
-Both the drastically increased RR of CRC-associated genera versus diversity for carcinoma and increasing RR with more CRC-associated generea positivity are highly supportive of this model.
-It is also possible that in a driver-passenger scenario, simply having the driver present or only identifying the passenger is a good enough proxy that the event is occuring.
-This would account for the observation that there is no constant additive effect on RR for increasing positivity.
-Additionally, the initial establishment of the driver within the system appears to be dependent on the current community.
-This is supported by our finding that when adding the community context to our models in addition to the CRC-associated genera, the model AUC increases.
+The central idea of the model is that a single bacterium initiates an environment in which other non-resident microbes may then be able to colonize, creating a vicious cycle that is conducive for CRC.
+Both the drastically increased RR of CRC-associated genera versus $\alpha$-diversity metrics for carcinoma and increasing RR with more CRC-associated generea positivity are highly supportive of this model.
+However, the initial establishment of the driver within the system appears to be dependent on the current community.
+This is supported by our finding that when adding the community context to our models in addition to the CRC-associated genera, the model AUC increases [Figure 4 & 5].
+Conversely, using the present data, it is less likely that adenoma development fits this model.
+The changes that occur at this timepoint are small and possibly focal to the adenoma itself.
+The stepwise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1].
+Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change, to a change in the global community.
+The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case.
+One potential hypothesis from these observations is that at early stages of the diease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.
 
-With regards to carcinoma, our results support the driver-passenger model within the framework of the transition from adenoma to carcinoma.
-Conversely, using the present data, we can only theorize observations related to adenoma development fit this model.
-The changes that occur at this timepoint are small and possibly focal to the adenoma itself.
-The stepwise decrease in diversity suggests that the adenoma community is not normal but has changed subtly.
-Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change to a change in the global community.
-The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case.
-It is possible to hypothesize that at early stages of the diease, the host interacts with these subtle changes and catalyzes what ultimately leads to a thoroughly dysfunctional community.
+Within stool, common resident microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6].
+Changes in *Bacteroides*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma models [Figure 6].
+This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community.
+These observations would also suggest that initial changes in the resident community, specifically to *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma.
+Based on these observations, the early changes in the community may give rise to initial polyp formation via interactions with the host.
+These changes then create new niches in which any one of the CRC-associated genera could gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma via a driver-passenger type mechanism.
+
+The tissue studies did not provide a clearer understanding of how the microbiota may drive the progression of disease severity.
+For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 across studies.
+Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F].
+The majority of the results seem to be study specific with many top 10 taxa being present only in a single study.
+One other potentially worrying sign is the presence of *Propionibacterium* within the top 10 for the genera and OTU models and could be a marker of contamination.
+The low biomass of these samples coupled with potential contamination might be a possible reason why the tissue results seem to be more sporadic then the stool results.
 
 Although there are still questions that need to be answered regarding the microbiota and carcinoma, a clearer framework for their relationship is beginning to develop as to how this occurs.
-From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 6].
+From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 7].
 Conversely, the role of the microbiota in adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here.
@@
 
+\newpage
+
 ## Conclusion
@@
 By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in both adenoma and carcinoma.
-We are also able to provide support for the driver-passenger model in the context of carcinoma.
-However, within the context of adenoma, it is less clear that this relationship exists.
-These observations highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community.
+Overall, our results support a framework by which early localized community changes give rise to polyp formation.
+With the host as a potential catalyst, new niches arise by which non-resident CRC-associated microbes can then gain a foothold and create an environment that allows more of these microbes to colonize.
+This exacerbates the existing community changes and creates a vicious cycle conducive of carcinoma formation.
+Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community.
 Although there are power limitations associated with adenoma, this report highlights the strong influence the microbiota has on CRC development.

### 2017-11-29 submission/manuscript.tex
re write discussion and conclusion

@@
 microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
-carcinoma but not necessarily adenoma. Both the drastically increased RR
-of CRC-associated genera versus diversity for carcinoma and increasing
-RR with more CRC-associated generea positivity are highly supportive of
-this model. It is also possible that in a driver-passenger scenario,
-simply having the driver present or only identifying the passenger is a
-good enough proxy that the event is occuring. This would account for the
-observation that there is no constant additive effect on RR for
-increasing positivity. Additionally, the initial establishment of the
+carcinoma but not necessarily adenoma. The central idea of the model is
+that a single bacterium initiates an environment in which other
+non-resident microbes may then be able to colonize, creating a vicious
+cycle that is conducive for CRC. Both the drastically increased RR of
+CRC-associated genera versus \(\alpha\)-diversity metrics for carcinoma
+and increasing RR with more CRC-associated generea positivity are highly
+supportive of this model. However, the initial establishment of the
 driver within the system appears to be dependent on the current
@@
 community context to our models in addition to the CRC-associated
-genera, the model AUC increases.
-
-With regards to carcinoma, our results support the driver-passenger
-model within the framework of the transition from adenoma to carcinoma.
-Conversely, using the present data, we can only theorize observations
-related to adenoma development fit this model. The changes that occur at
-this timepoint are small and possibly focal to the adenoma itself. The
-stepwise decrease in diversity suggests that the adenoma community is
-not normal but has changed subtly. Although there appears to be
-localized changes that do depend on the driver-passenger model, as
-supported by an increased RR for one, two, and four positive
-CRC-associated genera in tissue {[}Figure 3C{]}, there may be other
-processes at play that ultimately exacerbate the condition from a subtle
-localized change to a change in the global community. The poor
-performance of the Random Forest models for classifying adenoma based
-only on the microbiota would suggest that this is the case. It is
-possible to hypothesize that at early stages of the diease, the host
-interacts with these subtle changes and catalyzes what ultimately leads
-to a thoroughly dysfunctional community.
+genera, the model AUC increases {[}Figure 4 \& 5{]}. Conversely, using
+the present data, it is less likely that adenoma development fits this
+model. The changes that occur at this timepoint are small and possibly
+focal to the adenoma itself. The stepwise decrease in diversity suggests
+that the adenoma community is not normal but has changed subtly
+{[}Figure 1{]}. Although there appears to be localized changes that do
+depend on the driver-passenger model, as supported by an increased RR
+for one, two, and four positive CRC-associated genera in tissue
+{[}Figure 3C{]}, there may be other processes at play that ultimately
+exacerbate the condition from a subtle localized change, to a change in
+the global community. The poor performance of the Random Forest models
+for classifying adenoma based only on the microbiota would suggest that
+this is the case. One potential hypothesis from these observations is
+that at early stages of the diease, how the host interacts with these
+subtle changes is what ultimately leads to a thoroughly dysfunctional
+community that is supportive of CRC genesis.
+
+Within stool, common resident microbes were most consistently present in
+the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
+\emph{Bacteroides}, \emph{Ruminococcus}, and \emph{Roseburia} were
+consistently found to be disciminative across the different studies for
+both adenoma and carcinoma models {[}Figure 6{]}. This data would
+suggest that whether the non-resident bacterium is \emph{Fusobacteria}
+or \emph{Peptostreptococcus} is not as important as how these bacteria
+interact with the changing resident community. These observations would
+also suggest that initial changes in the resident community,
+specifically to \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
+\emph{Roseburia}, carry on from adenoma to carcinoma. Based on these
+observations, the early changes in the community may give rise to
+initial polyp formation via interactions with the host. These changes
+then create new niches in which any one of the CRC-associated genera
+could gain a foothold, exacerbating the initial changes in community and
+facilitating the transition from adenoma to carcinoma via a
+driver-passenger type mechanism.
+
+The tissue studies did not provide a clearer understanding of how the
+microbiota may drive the progression of disease severity. For the OTU
+models both the unmatched and matched {[}Figure S7E \& F{]} tissue
+samples had some concordance with the stool data, with resident bacteria
+being the most prevalent in the top 10 across studies. Unlike in stool,
+\emph{Fusobacterium} was the only CRC-associated bacteria consistently
+present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
+majority of the results seem to be study specific with many top 10 taxa
+being present only in a single study. One other potentially worrying
+sign is the presence of \emph{Propionibacterium} within the top 10 for
+the genera and OTU models and could be a marker of contamination. The
+low biomass of these samples coupled with potential contamination might
+be a possible reason why the tissue results seem to be more sporadic
+then the stool results.
 
@@
 30\% or more between the case and control and most studies analyzed have
-sufficient power to detect these types of changes {[}Figure 6{]}.
+sufficient power to detect these types of changes {[}Figure 7{]}.
 Conversely, the role of the microbiota in adenoma is less clear and part
@@
 
+\newpage
+
 \subsection{Conclusion}\label{conclusion}
@@
 and tissue, we are able to provide evidence in support of the importance
-of the bacterial community in both adenoma and carcinoma. We are also
-able to provide support for the driver-passenger model in the context of
-carcinoma. However, within the context of adenoma, it is less clear that
-this relationship exists. These observations highlight the importance of
-power and sample number considerations when undertaking investigations
-into the microbiota and adenoma due to the subtle changes in the
-community. Although there are power limitations associated with adenoma,
-this report highlights the strong influence the microbiota has on CRC
-development.
+of the bacterial community in both adenoma and carcinoma. Overall, our
+results support a framework by which early localized community changes
+give rise to polyp formation. With the host as a potential catalyst, new
+niches arise by which non-resident CRC-associated microbes can then gain
+a foothold and create an environment that allows more of these microbes
+to colonize. This exacerbates the existing community changes and creates
+a vicious cycle conducive of carcinoma formation. Our observations also
+highlight the importance of power and sample number considerations when
+undertaking investigations into the microbiota and adenoma due to the
+subtle changes in the community. Although there are power limitations
+associated with adenoma, this report highlights the strong influence the
+microbiota has on CRC development.
 

### 2017-11-29 submission/manuscript.Rmd
refine major questions and problems this manuscript answers in intro

@@
 Using the whole community versus only CRC associated genera to build a prediction model resulted in higher classification success based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples.
+The most important OTUs for these models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies.
 For the included studies, most were adequately powered for large effect size differences.
@@
 
+The focus on biomarkers within stool makes sense since the current gold standard for diagnosis is a colonoscopy which can be time-consuming and is not without risks.
+Stool offers an easy and cost-effective way in which to stratify risk within the population.
+Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to what is actually happening on the adenoma or carcinoma.
+Some studies have begun to try and assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
+Since the stool possibly represents an agregate community of the overall gastrointestinal (GI) tract, sampling the adenoma or carinoma directly might do a better job at identifying important microbes and could be drastically different then what is identified in stool.
+Sampling the tissue directly may provide clearer answers but it is not without problems.
+Due to bowel prep the communities left for sampling may not be the truly resident microbes but rather a collection of what is able to keep adhered to the mucosa.
+Further, these samples would contain much more host DNA then microbial DNA.
+Low biomass samples are very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014].
+Due to these variables, one question our meta-analysis aims to answer is whether it is possible to find consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.
+
+This intense focus on identifying biomarkers has identified mostly mouth-associated microbes as potential CRC-associated microbes in the GI tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014].
+The main bacteria of interest arising from this set of microbes has been those within the *Fusobacterium* genus.
+Yet, the question remains as to whether or not this is indeed the most important genera to be focusing on since many studies microbiota based studies identify a collection of oral microbes rather than single genera [@baxter_microbiota-based_2016; @zeller_potential_2014].
+Based on this discrepency, the second question we want to answer with this meta-analysis is if there really is one dominant CRC-associated genera that can be identified across studies.
+
+The heavy focus on the identification of biomarkers has mostly dominated the study of CRC and has had an unattended consequence of reducing the focus on changes that occur within the underlying resident community.
+This has been borne out by the majority of previous studies, both within stool and tissue, that have focused predominately on biomarkers as well as the only meta-analysis investigating this area [@shah_leveraging_2017].
+Since it has not been investigated in detail the final question that this meta-analysis aims to answer is whether there are consistent detectable community differences as disease severity increases.
+
 Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
@@
 Third, using Random Forest models, we analyzed whether the full community or only the CRC associated genera resulted in better model classification based on the area under the curve (AUC).
-Since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for.
-
-Our analysis found a continuous decrease in Shannon diversity as disease became more severe which correlated with a significantly increased RR for carcinoma.
-Using only CRC-associated genera, the RR for carcinoma was higher relative to what was observed for the Shannon diversity RR.
-Conversely, we demonstrate that the AUC of the classification models increased when the full community was incorporated as opposed to only CRC-associated genera.
+We observed that the community changes as disease severity worsens and that this community is important for disease classification.
+However, since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for.
 Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.

### 2017-11-29 submission/manuscript.tex
refine major questions and problems this manuscript answers in intro

@@
 success based on Area Under the Curve (AUC) for both adenoma and
-carcinoma using fecal and tissue samples. For the included studies, most
-were adequately powered for large effect size differences. This may be
-more amenable for carcinoma than for adenoma microbiota research due to
-the smaller community level changes observed in our results.
+carcinoma using fecal and tissue samples. The most important OTUs for
+these models consistently belonged to genera such as
+\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
+studies. For the included studies, most were adequately powered for
+large effect size differences. This may be more amenable for carcinoma
+than for adenoma microbiota research due to the smaller community level
+changes observed in our results.
 
@@
 
+The focus on biomarkers within stool makes sense since the current gold
+standard for diagnosis is a colonoscopy which can be time-consuming and
+is not without risks. Stool offers an easy and cost-effective way in
+which to stratify risk within the population. Although stool represents
+an easy and less invasive way to assess risk, it is not clear how
+reflective this sample is to what is actually happening on the adenoma
+or carcinoma. Some studies have begun to try and assess this in health
+and disease but are limited by their sample size {[}12,13{]}. Since the
+stool possibly represents an agregate community of the overall
+gastrointestinal (GI) tract, sampling the adenoma or carinoma directly
+might do a better job at identifying important microbes and could be
+drastically different then what is identified in stool. Sampling the
+tissue directly may provide clearer answers but it is not without
+problems. Due to bowel prep the communities left for sampling may not be
+the truly resident microbes but rather a collection of what is able to
+keep adhered to the mucosa. Further, these samples would contain much
+more host DNA then microbial DNA. Low biomass samples are very difficult
+to work with and results can end up being study dependent due to the
+randomness of contamination {[}14{]}. Due to these variables, one
+question our meta-analysis aims to answer is whether it is possible to
+find consistent patterns that emerge across studies regardless of
+whether they used stool or tissue samples.
+
+This intense focus on identifying biomarkers has identified mostly
+mouth-associated microbes as potential CRC-associated microbes in the GI
+tract {[}15--17{]}. The main bacteria of interest arising from this set
+of microbes has been those within the \emph{Fusobacterium} genus. Yet,
+the question remains as to whether or not this is indeed the most
+important genera to be focusing on since many studies microbiota based
+studies identify a collection of oral microbes rather than single genera
+{[}16,17{]}. Based on this discrepency, the second question we want to
+answer with this meta-analysis is if there really is one dominant
+CRC-associated genera that can be identified across studies.
+
+The heavy focus on the identification of biomarkers has mostly dominated
+the study of CRC and has had an unattended consequence of reducing the
+focus on changes that occur within the underlying resident community.
+This has been borne out by the majority of previous studies, both within
+stool and tissue, that have focused predominately on biomarkers as well
+as the only meta-analysis investigating this area {[}11{]}. Since it has
+not been investigated in detail the final question that this
+meta-analysis aims to answer is whether there are consistent detectable
+community differences as disease severity increases.
+
 Using both feces (n = 1737) and tissue (492 samples from 350
 individuals) totalling over 2229 total samples across 14 studies
-{[}12--25{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope of
-the previous meta-analysis to investigate whether the bacterial
+{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
+of the previous meta-analysis to investigate whether the bacterial
 community is an important risk factor for both adenoma and carcinoma. To
@@
 associated genera resulted in better model classification based on the
-area under the curve (AUC). Since the changes in community were subtle
-for adenoma we also examined what effect and sample size the studies
-that were used were adequately powered for.
-
-Our analysis found a continuous decrease in Shannon diversity as disease
-became more severe which correlated with a significantly increased RR
-for carcinoma. Using only CRC-associated genera, the RR for carcinoma
-was higher relative to what was observed for the Shannon diversity RR.
-Conversely, we demonstrate that the AUC of the classification models
-increased when the full community was incorporated as opposed to only
-CRC-associated genera. Although we analyzed data sets which sampled
-large numbers of individuals, our results indicate the individual
-studies were underpowered for detecting effect size differences of 10\%
-or below between the case and control groups.
+area under the curve (AUC). We observed that the community changes as
+disease severity worsens and that this community is important for
+disease classification. However, since the changes in community were
+subtle for adenoma we also examined what effect and sample size the
+studies that were used were adequately powered for. Although we analyzed
+data sets which sampled large numbers of individuals, our results
+indicate the individual studies were underpowered for detecting effect
+size differences of 10\% or below between the case and control groups.
 
@@
 were identified through the review articles written by Keku, \emph{et
-al.} and Vogtmann, \emph{et al.} {[}26,27{]}. Additional studies not
+al.} and Vogtmann, \emph{et al.} {[}29,30{]}. Additional studies not
 mentioned in the reviews were obtained based on the authors' knowledge
@@
 metadata in which the authors were able to share. After these filtering
-steps, the following studies remained: Ahn, \emph{et al.} {[}21{]},
-Baxter, \emph{et al.} {[}24{]}, Brim, \emph{et al.} {[}17{]}, Burns,
-\emph{et al.} {[}22{]}, Chen, \emph{et al.} {[}14{]}, Dejea, \emph{et
-al.} {[}19{]}, Flemer, \emph{et al.} {[}13{]}, Geng, \emph{et al.}
-{[}25{]}, Hale, \emph{et al.} {[}12{]}, Kostic, \emph{et al.} {[}28{]},
-Lu, \emph{et al.} {[}16{]}, Sanapareddy, \emph{et al.} {[}20{]}, Wang,
-\emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}18{]}, and Zeller,
-\emph{et al.} {[}23{]}. The Zackular {[}29{]} study was not included
+steps, the following studies remained: Ahn, \emph{et al.} {[}26{]},
+Baxter, \emph{et al.} {[}16{]}, Brim, \emph{et al.} {[}22{]}, Burns,
+\emph{et al.} {[}27{]}, Chen, \emph{et al.} {[}19{]}, Dejea, \emph{et
+al.} {[}24{]}, Flemer, \emph{et al.} {[}12{]}, Geng, \emph{et al.}
+{[}28{]}, Hale, \emph{et al.} {[}18{]}, Kostic, \emph{et al.} {[}15{]},
+Lu, \emph{et al.} {[}21{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang,
+\emph{et al.} {[}20{]}, Weir, \emph{et al.} {[}23{]}, and Zeller,
+\emph{et al.} {[}17{]}. The Zackular {[}31{]} study was not included
 becasue the 90 individuals analyzed within the study are contained
-within the larger Baxter study {[}24{]}. Additionally, after sequence
+within the larger Baxter study {[}16{]}. Additionally, after sequence
 processing all the case samples for the Kostic study only had 100
@@
 not have sequences and metadata on the SRA, data was obtained from DBGap
-for one study {[}21{]} and for four studies was obtained directly from
-the authors {[}12,13,18,20{]}. Each study was processed using the mothur
-(v1.39.3) software program {[}30{]}. Where possible, quality filtering
+for one study {[}26{]} and for four studies was obtained directly from
+the authors {[}12,18,23,25{]}. Each study was processed using the mothur
+(v1.39.3) software program {[}32{]}. Where possible, quality filtering
 utilized the default methods used in mothur for either 454 or Illumina
@@
 stated quality cut-offs were used instead. Chimeras were identifed and
-removed using VSEARCH {[}31{]} before \emph{de novo} OTU clustering at
-97\% similarity using the OptiClust algorithm {[}32{]} was utilized.
+removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
+97\% similarity using the OptiClust algorithm {[}34{]} was utilized.
 
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.2) software package {[}33{]}.
+sequence processing utilized the R (v3.4.2) software package {[}35{]}.
 For the \(\alpha\)-diversity analysis, values were power transformed
-using the rcompanion (v1.10.1) package {[}34{]} and then Z-score
-normalized using the car (v2.1.5) package {[}35{]}. Testing for
+using the rcompanion (v1.10.1) package {[}36{]} and then Z-score
+normalized using the car (v2.1.5) package {[}37{]}. Testing for
 \(\alpha\)-diversity differences utilized linear mixed-effect models
-created using the lme4 (v1.1.14) package {[}36{]} to correct for study
+created using the lme4 (v1.1.14) package {[}38{]} to correct for study
 and variable region effects in feces and study, variable region, and
 individual effects in tissue. Relative risk was analyzed using both the
-epiR (v0.9.87) and metafor (v2.0.0) packages {[}37,38{]} by assessing
+epiR (v0.9.87) and metafor (v2.0.0) packages {[}39,40{]} by assessing
 how many with and without disease were above and below the overall
@@
 utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
-vegan (v2.4.4) package {[}39{]}. Random Forest models were built using
+vegan (v2.4.4) package {[}41{]}. Random Forest models were built using
 both the caret (v6.0.77) and randomForest (v4.6.12) packages
-{[}40,41{]}. Differences between the obtained AUC versus a random model
+{[}42,43{]}. Differences between the obtained AUC versus a random model
 AUC was assessed using T-tests. Power analysis and estimations were made
-using the pwr (v1.2.1) and statmod (v1.4.30) packages {[}42,43{]}. All
+using the pwr (v1.2.1) and statmod (v1.4.30) packages {[}44,45{]}. All
 figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
-packages {[}44,45{]}.
+packages {[}46,47{]}.
 
@@
 
+\hypertarget{ref-flemer_tumour-associated_2017}{}
+12. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
+al. Tumour-associated and non-tumour-associated microbiota in colorectal
+cancer. Gut. 2017;66:633--43.
+
+\hypertarget{ref-Flynn_preprint_2017}{}
+13. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
+the native colon microbiota in healthy adults. Cold Spring Harbor
+Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}
+
+\hypertarget{ref-Salter_contamination_2014}{}
+14. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
+al. Reagent and laboratory contamination can critically impact
+sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
+Nature; 2014;12. Available from:
+\url{https://doi.org/10.1186/s12915-014-0087-z}
+
+\hypertarget{ref-kostic_genomic_2012}{}
+15. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
+al. Genomic analysis identifies association of Fusobacterium with
+colorectal carcinoma. Genome Research. 2012;22:292--8.
+
+\hypertarget{ref-baxter_microbiota-based_2016}{}
+16. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+improves the sensitivity of fecal immunochemical test for detecting
+colonic lesions. Genome Medicine. 2016;8:37.
+
+\hypertarget{ref-zeller_potential_2014}{}
+17. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+Potential of fecal microbiota for early-stage detection of colorectal
+cancer. Molecular Systems Biology. 2014;10:766.
+
 \hypertarget{ref-hale_shifts_2017}{}
-12. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
+18. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
 Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
@@
 
-\hypertarget{ref-flemer_tumour-associated_2017}{}
-13. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
-al. Tumour-associated and non-tumour-associated microbiota in colorectal
-cancer. Gut. 2017;66:633--43.
-
 \hypertarget{ref-chen_human_2012}{}
-14. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
+19. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
 mucosa-associated microbiota in patients with colorectal cancer. PloS
@@
 \hypertarget{ref-wang_structural_2012}{}
-15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
+20. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
 segregation of gut microbiota between colorectal cancer patients and
@@
 \hypertarget{ref-lu_mucosal_2016}{}
-16. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
+21. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
 adherent bacterial dysbiosis in patients with colorectal adenomas.
@@
 \hypertarget{ref-brim_microbiome_2013}{}
-17. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
+22. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
 al. Microbiome analysis of stool samples from African Americans with
@@
 \hypertarget{ref-weir_stool_2013}{}
-18. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
+23. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
 Stool microbiome and metabolome differences between colorectal cancer
@@
 \hypertarget{ref-dejea_microbiota_2014}{}
-19. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+24. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
 Rossetti BJ, et al. Microbiota organization is a distinct feature of
@@
 \hypertarget{ref-sanapareddy_increased_2012}{}
-20. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
+25. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
 et al. Increased rectal microbial richness is associated with the
@@
 \hypertarget{ref-ahn_human_2013}{}
-21. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
+26. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
 microbiome and risk for colorectal cancer. Journal of the National
@@
 \hypertarget{ref-burns_virulence_2015}{}
-22. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
+27. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
 are a signature of the microbiome in the colorectal tumor
@@
 
-\hypertarget{ref-zeller_potential_2014}{}
-23. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
-Potential of fecal microbiota for early-stage detection of colorectal
-cancer. Molecular Systems Biology. 2014;10:766.
-
-\hypertarget{ref-baxter_microbiota-based_2016}{}
-24. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
-improves the sensitivity of fecal immunochemical test for detecting
-colonic lesions. Genome Medicine. 2016;8:37.
-
 \hypertarget{ref-geng_diversified_2013}{}
-25. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
+28. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
 human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
@@
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-26. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
+29. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
 microbiota and colorectal cancer. American Journal of Physiology -
@@
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-27. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
+30. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
 microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
@@
 
-\hypertarget{ref-kostic_genomic_2012}{}
-28. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
-al. Genomic analysis identifies association of Fusobacterium with
-colorectal carcinoma. Genome Research. 2012;22:292--8.
-
 \hypertarget{ref-zackular_human_2014}{}
-29. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+31. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-30. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+32. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, et al. Introducing mothur: Open-Source, Platform-Independent,
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-31. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
+33. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
 versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-32. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
+34. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
 Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
@@
 \hypertarget{ref-r_citation_2017}{}
-33. R Core Team. R: A language and environment for statistical computing
+35. R Core Team. R: A language and environment for statistical computing
 {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-34. Mangiafico S. Rcompanion: Functions to support extension education
+36. Mangiafico S. Rcompanion: Functions to support extension education
 program evaluation {[}Internet{]}. 2017. Available from:
@@
 \hypertarget{ref-car_citation_2011}{}
-35. Fox J, Weisberg S. An R companion to applied regression
+37. Fox J, Weisberg S. An R companion to applied regression
 {[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
@@
 \hypertarget{ref-lme4_citation_2015}{}
-36. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
+38. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
 models using lme4. Journal of Statistical Software. 2015;67:1--48.
@@
 \hypertarget{ref-epir_citation_2017}{}
-37. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
+39. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
 J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
@@
 \hypertarget{ref-metafor_citation_2010}{}
-38. Viechtbauer W. Conducting meta-analyses in R with the metafor
+40. Viechtbauer W. Conducting meta-analyses in R with the metafor
 package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-39. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+41. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
@@
 \hypertarget{ref-caret_citation_2017}{}
-40. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
+42. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
 Cooper T, et al. Caret: Classification and regression training
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-41. Liaw A, Wiener M. Classification and regression by randomForest. R
+43. Liaw A, Wiener M. Classification and regression by randomForest. R
 News {[}Internet{]}. 2002;2:18--22. Available from:
@@
 \hypertarget{ref-pwr_citation_2017}{}
-42. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
+44. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
 2017. Available from: \url{https://CRAN.R-project.org/package=pwr}
@@
 \hypertarget{ref-statmod_citation_2016}{}
-43. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
+45. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
 gaussian distribution. R Journal. 2016;8:339--51.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-44. Wickham H. Ggplot2: Elegant graphics for data analysis
+46. Wickham H. Ggplot2: Elegant graphics for data analysis
 {[}Internet{]}. Springer-Verlag New York; 2009. Available from:
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-45. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
+47. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
 {[}Internet{]}. 2017. Available from:

### 2017-11-29 submission/manuscript.Rmd
grammar updates and flow corrections

--- 2017-11-29 submission/manuscript.Rmd
@@
 
+Targeting the identification of biomarkers within stool seems logical since the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free.
+Stool offers an easy and cost-effective way in which to stratify CRC risk within the population.
+Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to what is actually happening on the adenoma or carcinoma.
+Some studies have begun to try and assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
+Since the stool possibly represents an agregate community of the overall gastrointestinal (GI) tract, sampling the adenoma or carinoma directly could more accurately identify important microbes which could be drastically different then what is identified in stool.
+Sampling the tissue directly may provide clearer answers but it is not without problems.
+Due to bowel prep the communities left for sampling may not be reflective of the resident microbiota, but rather a collection of what is able to keep adhered to the mucosa.
+Additionally, these samples would contain much more host DNA then microbial DNA, potentially limiting the types of analysis that can be done.
+It has also been well published that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014].
+Due to these many differences, one question our meta-analysis aims to answer is whether it is possible to find consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.
+
+This intense focus on identifying biomarkers has identified mostly mouth-associated microbes as potential CRC-associated microbes in the GI tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014].
+The main bacteria of interest arising from this set of microbes has been those within the *Fusobacterium* genus.
+Yet, the question remains as to whether or not this is indeed the most important genera to be focusing on, since many microbiota-based studies typically have identified a collection of oral microbes rather than single species from a single genera [@baxter_microbiota-based_2016; @zeller_potential_2014].
+Based on this discrepency, the second question we want to answer with this meta-analysis is if there is one dominant CRC-associated genera that can be identified across studies.
+
+The heavy concentration on the identification of biomarkers has mostly dominated the study of CRC and has had an unattended consequence of reducing the focus on changes that occur within the underlying resident community.
+This has been borne out by the majority of previous studies, within stool, tissue, and the only meta-analysis investigating this area, that have focused predominately on biomarker identification.
+Since it has not been investigated in detail the final question that this meta-analysis aims to answer is whether there are consistent detectable community differences as disease severity increases.
+
 Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
 To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma.
-Next, we assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma.
-Third, using Random Forest models, we analyzed whether the full community or only the CRC associated genera resulted in better model classification based on the area under the curve (AUC).
-Since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for.
-
-Our analysis found a continuous decrease in Shannon diversity as disease became more severe which correlated with a significantly increased RR for carcinoma.
-Using only CRC-associated genera, the RR for carcinoma was higher relative to what was observed for the Shannon diversity RR.
-Conversely, we demonstrate that the AUC of the classification models increased when the full community was incorporated as opposed to only CRC-associated genera.
+We then assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma.
+Next, using Random Forest models, we analyzed whether the full community or only the CRC-associated genera resulted in better model classification based on the area under the curve (AUC).
+We observed that the community changes as disease severity worsens and that this community is important for disease classification.
+However, since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for.
 Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.
@@
 These observations would also suggest that initial changes in the resident community, specifically to *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma.
-Based on these observations, the early changes in the community may give rise to initial polyp formation via interactions with the host.
+Based on these observations, it is possible to hypothesize that the early changes in the community may give rise to initial polyp formation via interactions with the host and not necessarily via interactions with CRC-assciated genera.
 These changes then create new niches in which any one of the CRC-associated genera could gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma via a driver-passenger type mechanism.
@@
 The tissue studies did not provide a clearer understanding of how the microbiota may drive the progression of disease severity.
-For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 across studies.
+For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 important variables across studies.
 Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F].

### 2017-11-29 submission/manuscript.tex
grammar updates and flow corrections

@@
 
-The focus on biomarkers within stool makes sense since the current gold
-standard for diagnosis is a colonoscopy which can be time-consuming and
-is not without risks. Stool offers an easy and cost-effective way in
-which to stratify risk within the population. Although stool represents
-an easy and less invasive way to assess risk, it is not clear how
-reflective this sample is to what is actually happening on the adenoma
-or carcinoma. Some studies have begun to try and assess this in health
-and disease but are limited by their sample size {[}12,13{]}. Since the
-stool possibly represents an agregate community of the overall
-gastrointestinal (GI) tract, sampling the adenoma or carinoma directly
-might do a better job at identifying important microbes and could be
-drastically different then what is identified in stool. Sampling the
-tissue directly may provide clearer answers but it is not without
-problems. Due to bowel prep the communities left for sampling may not be
-the truly resident microbes but rather a collection of what is able to
-keep adhered to the mucosa. Further, these samples would contain much
-more host DNA then microbial DNA. Low biomass samples are very difficult
-to work with and results can end up being study dependent due to the
-randomness of contamination {[}14{]}. Due to these variables, one
+Targeting the identification of biomarkers within stool seems logical
+since the current gold standard for diagnosis, a colonoscopy, can be
+time-consuming and is not risk free. Stool offers an easy and
+cost-effective way in which to stratify CRC risk within the population.
+Although stool represents an easy and less invasive way to assess risk,
+it is not clear how reflective this sample is to what is actually
+happening on the adenoma or carcinoma. Some studies have begun to try
+and assess this in health and disease but are limited by their sample
+size {[}12,13{]}. Since the stool possibly represents an agregate
+community of the overall gastrointestinal (GI) tract, sampling the
+adenoma or carinoma directly could more accurately identify important
+microbes which could be drastically different then what is identified in
+stool. Sampling the tissue directly may provide clearer answers but it
+is not without problems. Due to bowel prep the communities left for
+sampling may not be reflective of the resident microbiota, but rather a
+collection of what is able to keep adhered to the mucosa. Additionally,
+these samples would contain much more host DNA then microbial DNA,
+potentially limiting the types of analysis that can be done. It has also
+been well published that low biomass samples can be very difficult to
+work with and results can end up being study dependent due to the
+randomness of contamination {[}14{]}. Due to these many differences, one
 question our meta-analysis aims to answer is whether it is possible to
@@
 the question remains as to whether or not this is indeed the most
-important genera to be focusing on since many studies microbiota based
-studies identify a collection of oral microbes rather than single genera
-{[}16,17{]}. Based on this discrepency, the second question we want to
-answer with this meta-analysis is if there really is one dominant
-CRC-associated genera that can be identified across studies.
-
-The heavy focus on the identification of biomarkers has mostly dominated
-the study of CRC and has had an unattended consequence of reducing the
-focus on changes that occur within the underlying resident community.
-This has been borne out by the majority of previous studies, both within
-stool and tissue, that have focused predominately on biomarkers as well
-as the only meta-analysis investigating this area {[}11{]}. Since it has
-not been investigated in detail the final question that this
+important genera to be focusing on, since many microbiota-based studies
+typically have identified a collection of oral microbes rather than
+single species from a single genera {[}16,17{]}. Based on this
+discrepency, the second question we want to answer with this
+meta-analysis is if there is one dominant CRC-associated genera that can
+be identified across studies.
+
+The heavy concentration on the identification of biomarkers has mostly
+dominated the study of CRC and has had an unattended consequence of
+reducing the focus on changes that occur within the underlying resident
+community. This has been borne out by the majority of previous studies,
+within stool, tissue, and the only meta-analysis investigating this
+area, that have focused predominately on biomarker identification. Since
+it has not been investigated in detail the final question that this
 meta-analysis aims to answer is whether there are consistent detectable
@@
 throughout disease (control to adenoma to carcinoma) and if it results
-in an increased relative risk (RR) for adenoma or carcinoma. Next, we
+in an increased relative risk (RR) for adenoma or carcinoma. We then
 assessed how common CRC-associated genera (\emph{Fusobacterium},
 \emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
-affect the relative risk of adenoma or carcinoma. Third, using Random
-Forest models, we analyzed whether the full community or only the CRC
-associated genera resulted in better model classification based on the
-area under the curve (AUC). We observed that the community changes as
-disease severity worsens and that this community is important for
+affect the relative risk of adenoma or carcinoma. Next, using Random
+Forest models, we analyzed whether the full community or only the
+CRC-associated genera resulted in better model classification based on
+the area under the curve (AUC). We observed that the community changes
+as disease severity worsens and that this community is important for
 disease classification. However, since the changes in community were
@@
 \emph{Roseburia}, carry on from adenoma to carcinoma. Based on these
-observations, the early changes in the community may give rise to
-initial polyp formation via interactions with the host. These changes
-then create new niches in which any one of the CRC-associated genera
-could gain a foothold, exacerbating the initial changes in community and
-facilitating the transition from adenoma to carcinoma via a
-driver-passenger type mechanism.
+observations, it is possible to hypothesize that the early changes in
+the community may give rise to initial polyp formation via interactions
+with the host and not necessarily via interactions with CRC-assciated
+genera. These changes then create new niches in which any one of the
+CRC-associated genera could gain a foothold, exacerbating the initial
+changes in community and facilitating the transition from adenoma to
+carcinoma via a driver-passenger type mechanism.
 
@@
 samples had some concordance with the stool data, with resident bacteria
-being the most prevalent in the top 10 across studies. Unlike in stool,
-\emph{Fusobacterium} was the only CRC-associated bacteria consistently
-present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
-majority of the results seem to be study specific with many top 10 taxa
-being present only in a single study. One other potentially worrying
-sign is the presence of \emph{Propionibacterium} within the top 10 for
-the genera and OTU models and could be a marker of contamination. The
-low biomass of these samples coupled with potential contamination might
-be a possible reason why the tissue results seem to be more sporadic
-then the stool results.
+being the most prevalent in the top 10 important variables across
+studies. Unlike in stool, \emph{Fusobacterium} was the only
+CRC-associated bacteria consistently present in the top 10 of the CRC
+models {[}Figure S7B-C \& E-F{]}. The majority of the results seem to be
+study specific with many top 10 taxa being present only in a single
+study. One other potentially worrying sign is the presence of
+\emph{Propionibacterium} within the top 10 for the genera and OTU models
+and could be a marker of contamination. The low biomass of these samples
+coupled with potential contamination might be a possible reason why the
+tissue results seem to be more sporadic then the stool results.
 

### 2017-11-29 submission/manuscript.Rmd
add missing result and correct a grammar error

@@
 The AUC for classification of adenomas was lower than carcionmas for both tissue and stool [Figure S4-S6].
+Interestingly, the AUC for the classification of carcinoma was consistently lower for the tissue models than the stool models [Figure S4-S6].
 
-The OTU based models for both fecal and tissue (matched and unmatched) samples, showed an AUC decrease when only OTUs from the CRC-associated genera are used versus full community of OTUs [Figure 4 & 5].
-This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5] with the largest difference in median AUC for stool being in carcinoma classification [Figure 4B] and for tissue being in adenoma [Figure 5A].
+The OTU based models for both fecal and tissue (matched and unmatched) samples, showed an AUC decrease when only OTUs from the CRC-associated genera are used versus the full community of OTUs [Figure 4 & 5].
+This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5] with the largest difference in median AUC for stool being in carcinoma classification [Figure 4B] and for tissue being an adenoma [Figure 5A].
 

### 2017-11-29 submission/manuscript.tex
add missing result and correct a grammar error

@@
 AUC for classification of adenomas was lower than carcionmas for both
-tissue and stool {[}Figure S4-S6{]}.
+tissue and stool {[}Figure S4-S6{]}. Interestingly, the AUC for the
+classification of carcinoma was consistently lower for the tissue models
+than the stool models {[}Figure S4-S6{]}.
 
@@
 samples, showed an AUC decrease when only OTUs from the CRC-associated
-genera are used versus full community of OTUs {[}Figure 4 \& 5{]}. This
-decrease is observed in both adenoma and carcinoma groups {[}Figure 4 \&
-5{]} with the largest difference in median AUC for stool being in
-carcinoma classification {[}Figure 4B{]} and for tissue being in adenoma
+genera are used versus the full community of OTUs {[}Figure 4 \& 5{]}.
+This decrease is observed in both adenoma and carcinoma groups {[}Figure
+4 \& 5{]} with the largest difference in median AUC for stool being in
+carcinoma classification {[}Figure 4B{]} and for tissue being an adenoma
 {[}Figure 5A{]}.

### 2017-12-11 submission/manuscript.Rmd
add intro update -- more explicit differences

@@
 Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.
+
+In comparison to the previous meta-analysis, this study  significantly increases the total stool samples investigated, examines differences between stool and tissue microbiota in the context of CRC, and takes a more community centric approach rather than a biomarker focused approach to investigating commonalities across study for the microbiota and CRC severity.
+Importantly, this community centric approach could provide valuable insights into the importance of accounting for the community in CRC disease not previously provided by earlier meta-analysis studies [@shah_leveraging_2017].
 

### 2017-12-11 submission/manuscript.tex
add intro update -- more explicit differences

@@
 
+In comparison to the previous meta-analysis, this study significantly
+increases the total stool samples investigated, examines differences
+between stool and tissue microbiota in the context of CRC, and takes a
+more community centric approach rather than a biomarker focused approach
+to investigating commonalities across study for the microbiota and CRC
+severity. Importantly, this community centric approach could provide
+valuable insights into the importance of accounting for the community in
+CRC disease not previously provided by earlier meta-analysis studies
+{[}11{]}.
+
 \newpage

### 2017-12-11 submission/manuscript.Rmd
change paragraph order
remove/replace role with associations where appropriate

@@
 **Conclusions.** This data provides support for the importance of the bacteral community to both adenoma and carcinoma genesis.
-The evidence collected within this study on the role of the microbiota in CRC pathogenesis is much stronger for carcinoma then adenoma.
+The evidence collected within this study on the associations between the microbiota and CRC are much stronger for carcinoma then adenoma.
 A strong reason for this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
@@
 Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
-Many of the studies, along with the previous meta-analysis, focus on identifying biomarkers or individual microbes but do not critically investigate the role the community has in CRC.
+Many of the studies, along with the previous meta-analysis, focus on identifying biomarkers or individual microbes but do not critically investigate how the community changes in CRC.
 
@@
 
+In comparison to the previous meta-analysis, this study  significantly increases the total stool samples investigated, examines differences between stool and tissue microbiota in the context of CRC, and takes a more community centric approach rather than a biomarker focused approach to investigating commonalities across study for the microbiota and CRC severity.
+Importantly, this community centric approach could provide valuable insights into the importance of accounting for the community in CRC disease not previously provided by earlier meta-analysis studies [@shah_leveraging_2017].
+
 Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
@@
 Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.
-
-In comparison to the previous meta-analysis, this study  significantly increases the total stool samples investigated, examines differences between stool and tissue microbiota in the context of CRC, and takes a more community centric approach rather than a biomarker focused approach to investigating commonalities across study for the microbiota and CRC severity.
-Importantly, this community centric approach could provide valuable insights into the importance of accounting for the community in CRC disease not previously provided by earlier meta-analysis studies [@shah_leveraging_2017].
 
@@
 From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 7].
-Conversely, the role of the microbiota in adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here.
+Conversely, the associations between the microbiota and adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here.
 None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls.

### 2017-12-11 submission/manuscript.tex
change paragraph order
remove/replace role with associations where appropriate

@@
 the bacteral community to both adenoma and carcinoma genesis. The
-evidence collected within this study on the role of the microbiota in
-CRC pathogenesis is much stronger for carcinoma then adenoma. A strong
-reason for this may be in part due to the low power to detect more
-subtle changes in the majority of studies that have been performed to
-date.
+evidence collected within this study on the associations between the
+microbiota and CRC are much stronger for carcinoma then adenoma. A
+strong reason for this may be in part due to the low power to detect
+more subtle changes in the majority of studies that have been performed
+to date.
 
@@
 studies, along with the previous meta-analysis, focus on identifying
-biomarkers or individual microbes but do not critically investigate the
-role the community has in CRC.
+biomarkers or individual microbes but do not critically investigate how
+the community changes in CRC.
 
@@
 
+In comparison to the previous meta-analysis, this study significantly
+increases the total stool samples investigated, examines differences
+between stool and tissue microbiota in the context of CRC, and takes a
+more community centric approach rather than a biomarker focused approach
+to investigating commonalities across study for the microbiota and CRC
+severity. Importantly, this community centric approach could provide
+valuable insights into the importance of accounting for the community in
+CRC disease not previously provided by earlier meta-analysis studies
+{[}11{]}.
+
 Using both feces (n = 1737) and tissue (492 samples from 350
@@
 size differences of 10\% or below between the case and control groups.
-
-In comparison to the previous meta-analysis, this study significantly
-increases the total stool samples investigated, examines differences
-between stool and tissue microbiota in the context of CRC, and takes a
-more community centric approach rather than a biomarker focused approach
-to investigating commonalities across study for the microbiota and CRC
-severity. Importantly, this community centric approach could provide
-valuable insights into the importance of accounting for the community in
-CRC disease not previously provided by earlier meta-analysis studies
-{[}11{]}.
 
@@
 sufficient power to detect these types of changes {[}Figure 7{]}.
-Conversely, the role of the microbiota in adenoma is less clear and part
-of the reason this may be is because many studies are not powered
-effectively to observe the small changes reported here. None of the
-studies analyzed were properly powered to detect a 10\% or lower change
-between case and controls. This small effect size range may well be the
-scope in which differences consistently occur in adenoma due to the
-subtle changes in community that occur between control and adenoma.
+Conversely, the associations between the microbiota and adenoma is less
+clear and part of the reason this may be is because many studies are not
+powered effectively to observe the small changes reported here. None of
+the studies analyzed were properly powered to detect a 10\% or lower
+change between case and controls. This small effect size range may well
+be the scope in which differences consistently occur in adenoma due to
+the subtle changes in community that occur between control and adenoma.
 Future studies investigating adenoma and the microbiota need to take

### 2017-12-11 submission/manuscript.Rmd
update abstract with more direct sentences

@@
 
-**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) after correcting for study effect and variable region sequenced.
-Only evenness for adenoma (P-value `r ifelse(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) resulted in a slightly increased relative risk while lower Shannon diversity and evenness in fecal samples resulted in a significant increase in relative risk for carcinoma (P-value `r ifelse(min(filter(alpha_RR_crc_stool, !(measure %in% "sobs"))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
-Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern with a significantly increased relative risk by their presence for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value `r ifelse(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
+**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
+This reduction in evenness resulted in small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only resulted in an increased relative risk for carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
+Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern, with their presence significantly increasing the relative risk for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value = `r format(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]), digits = 2)`).
 Using the whole community versus only CRC associated genera to build a prediction model resulted in higher classification success based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples.
 The most important OTUs for these models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies.
-For the included studies, most were adequately powered for large effect size differences.
-This may be more amenable for carcinoma than for adenoma microbiota research due to the smaller community level changes observed in our results.
+Overall, there were less associations between the microbiota and adenoma versus carcinoma and one reason why this may be is that most studies were only adequatly powered for large effect sizes.
 
 **Conclusions.** This data provides support for the importance of the bacteral community to both adenoma and carcinoma genesis.
-The evidence collected within this study on the associations between the microbiota and CRC are much stronger for carcinoma then adenoma.
+The evidence collected within this study on the role of the microbiota in CRC pathogenesis shows stronger associations between carcinoma then adenoma.
 A strong reason for this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

### 2017-12-11 submission/manuscript.tex
update abstract with more direct sentences

@@
 \textbf{Results.} Fecal samples had a significant decrease from control
-to adenoma to carcinoma for both Shannon diversity and evenness (P-value
-\textless{} 0.05) after correcting for study effect and variable region
-sequenced. Only evenness for adenoma (P-value \textless{} 0.05) resulted
-in a slightly increased relative risk while lower Shannon diversity and
-evenness in fecal samples resulted in a significant increase in relative
-risk for carcinoma (P-value \textless{} 0.05). Previously associated
-colorectal cancer genera (\emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Peptostreptococcus}, or \emph{Porphyromonas}) followed a similar
-pattern with a significantly increased relative risk by their presence
-for carcinoma (P-value \textless{} 0.05) but not adenoma (P-value
-\textgreater{} 0.05) with the exception of \emph{Porphyromonas} (P-value
-\textless{} 0.05). Using the whole community versus only CRC associated
-genera to build a prediction model resulted in higher classification
-success based on Area Under the Curve (AUC) for both adenoma and
-carcinoma using fecal and tissue samples. The most important OTUs for
-these models consistently belonged to genera such as
+to adenoma to carcinoma for both Shannon diversity and evenness after
+correcting for study effect and variable region sequenced (P-value
+\textless{} 0.05). This reduction in evenness resulted in small
+increases in relative risk for adenoma (P-value = 0.032) and carcinoma
+(P-value = 0.00034) while the reduction in Shannon diversity only
+resulted in an increased relative risk for carcinoma (P-value = 0.0047).
+Previously associated colorectal cancer genera (\emph{Fusobacterium},
+\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
+followed a similar pattern, with their presence significantly increasing
+the relative risk for carcinoma (P-value \textless{} 0.05) but not
+adenoma (P-value \textgreater{} 0.05) with the exception of
+\emph{Porphyromonas} (P-value = 0.023). Using the whole community versus
+only CRC associated genera to build a prediction model resulted in
+higher classification success based on Area Under the Curve (AUC) for
+both adenoma and carcinoma using fecal and tissue samples. The most
+important OTUs for these models consistently belonged to genera such as
 \emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
-studies. For the included studies, most were adequately powered for
-large effect size differences. This may be more amenable for carcinoma
-than for adenoma microbiota research due to the smaller community level
-changes observed in our results.
+studies. Overall, there were less associations between the microbiota
+and adenoma versus carcinoma and one reason why this may be is that most
+studies were only adequatly powered for large effect sizes.
 
@@
 the bacteral community to both adenoma and carcinoma genesis. The
-evidence collected within this study on the associations between the
-microbiota and CRC are much stronger for carcinoma then adenoma. A
-strong reason for this may be in part due to the low power to detect
-more subtle changes in the majority of studies that have been performed
-to date.
+evidence collected within this study on the role of the microbiota in
+CRC pathogenesis shows stronger associations between carcinoma then
+adenoma. A strong reason for this may be in part due to the low power to
+detect more subtle changes in the majority of studies that have been
+performed to date.
 

### 2017-12-11 submission/manuscript.Rmd
complete intro more specific sentence structure

@@
 Colorectal cancer (CRC) is a growing world-wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016].
-Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in polyp formation using mouse models of CRC.
+Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC.
 There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma.
-Recently, a meta-analysis was published investigating whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
-Many of the studies, along with the previous meta-analysis, focus on identifying biomarkers or individual microbes but do not critically investigate how the community changes in CRC.
+A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
+This meta-analysis, along with other studies, focused on identifying biomarkers or individual microbes but do not critically investigate how the community changes in CRC.
 
-Targeting the identification of biomarkers within stool seems logical since the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free.
-Stool offers an easy and cost-effective way in which to stratify CRC risk within the population.
-Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to what is actually happening on the adenoma or carcinoma.
-Some studies have begun to try and assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
-Since the stool possibly represents an agregate community of the overall gastrointestinal (GI) tract, sampling the adenoma or carinoma directly could more accurately identify important microbes which could be drastically different then what is identified in stool.
-Sampling the tissue directly may provide clearer answers but it is not without problems.
+Targeting the identification of biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free.
+Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to the community on the adenoma or carcinoma.
+Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
+One advantage of tissue is that stool possibly represents an agregate community of the overall gastrointestinal (GI) tract and sampling the adenoma or carinoma directly could more accurately identify important microbes for CRC.
+Sampling the tissue directly may provide clearer answers but is not without limitations.
 Due to bowel prep the communities left for sampling may not be reflective of the resident microbiota, but rather a collection of what is able to keep adhered to the mucosa.
-Additionally, these samples would contain much more host DNA then microbial DNA, potentially limiting the types of analysis that can be done.
-It has also been well published that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014].
-Due to these many differences, one question our meta-analysis aims to answer is whether it is possible to find consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.
+Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done.
+It is well known that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014].
+Due to these many differences that could arise between stool and tissue, one question our meta-analysis aims to answer is whether there are consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.
 
-This intense focus on identifying biomarkers has identified mostly mouth-associated microbes as potential CRC-associated microbes in the GI tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014].
-The main bacteria of interest arising from this set of microbes has been those within the *Fusobacterium* genus.
+This intense focus on identifying biomarkers has found strong associations with resident mouth microbes as potential CRC-associated microbes in the GI tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014].
+The main bacteria of interest, arising from this set of microbes, has been those within the *Fusobacterium* genus.
 Yet, the question remains as to whether or not this is indeed the most important genera to be focusing on, since many microbiota-based studies typically have identified a collection of oral microbes rather than single species from a single genera [@baxter_microbiota-based_2016; @zeller_potential_2014].
@@
 
-The heavy concentration on the identification of biomarkers has mostly dominated the study of CRC and has had an unattended consequence of reducing the focus on changes that occur within the underlying resident community.
-This has been borne out by the majority of previous studies, within stool, tissue, and the only meta-analysis investigating this area, that have focused predominately on biomarker identification.
-Since it has not been investigated in detail the final question that this meta-analysis aims to answer is whether there are consistent detectable community differences as disease severity increases.
+The identification of microbial biomarkers has mostly dominated the study of CRC and has had an unattended consequence of reducing the focus on changes that occur within the underlying resident community.
+This has been borne out by the majority of previous studies, within stool, tissue, and the only meta-analysis investigating this area to date, that focus predominately on biomarker identification.
+Yet, how these CRC-associated microbes interact with their community may be an important component to consider, and has not been investigated in great detail by other meta-analyses.
+In response to this gap, our study aims to answer whether there are consistent detectable community differences as disease severity increases.
 
@@
 Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
-To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma.
+To accomplish this we first assessed whether diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma.
 We then assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma.
 Next, using Random Forest models, we analyzed whether the full community or only the CRC-associated genera resulted in better model classification based on the area under the curve (AUC).
-We observed that the community changes as disease severity worsens and that this community is important for disease classification.
+Our results suggest that the community changes as disease severity worsens and that this community is important for disease classification.
 However, since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for.
-Although we analyzed data sets which sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.
+Although we analyzed data sets that sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.
 

### 2017-12-11 submission/manuscript.tex
complete intro more specific sentence structure

@@
 of both individual microbes {[}3--7{]} and the overall community
-{[}8--10{]} in polyp formation using mouse models of CRC. There have
-also been numerous case/control studies investigating the microbiota in
-the formation of both adenoma and carcinoma. Recently, a meta-analysis
-was published investigating whether specific biomarkers could be
-consistently identified using multiple data sets {[}11{]}. Many of the
-studies, along with the previous meta-analysis, focus on identifying
-biomarkers or individual microbes but do not critically investigate how
-the community changes in CRC.
+{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
+been numerous case/control studies investigating the microbiota in the
+formation of both adenoma and carcinoma. A recent meta-analysis
+investigated whether specific biomarkers could be consistently
+identified using multiple data sets {[}11{]}. This meta-analysis, along
+with other studies, focused on identifying biomarkers or individual
+microbes but do not critically investigate how the community changes in
+CRC.
 
 Targeting the identification of biomarkers within stool seems logical
-since the current gold standard for diagnosis, a colonoscopy, can be
-time-consuming and is not risk free. Stool offers an easy and
-cost-effective way in which to stratify CRC risk within the population.
-Although stool represents an easy and less invasive way to assess risk,
-it is not clear how reflective this sample is to what is actually
-happening on the adenoma or carcinoma. Some studies have begun to try
-and assess this in health and disease but are limited by their sample
-size {[}12,13{]}. Since the stool possibly represents an agregate
-community of the overall gastrointestinal (GI) tract, sampling the
-adenoma or carinoma directly could more accurately identify important
-microbes which could be drastically different then what is identified in
-stool. Sampling the tissue directly may provide clearer answers but it
-is not without problems. Due to bowel prep the communities left for
-sampling may not be reflective of the resident microbiota, but rather a
+since it offers an easy and cost-effective way to stratify risk and the
+current gold standard for diagnosis, a colonoscopy, can be
+time-consuming and is not risk free. Although stool represents an easy
+and less invasive way to assess risk, it is not clear how reflective
+this sample is to the community on the adenoma or carcinoma. Some
+studies have tried to assess this in health and disease but are limited
+by their sample size {[}12,13{]}. One advantage of tissue is that stool
+possibly represents an agregate community of the overall
+gastrointestinal (GI) tract and sampling the adenoma or carinoma
+directly could more accurately identify important microbes for CRC.
+Sampling the tissue directly may provide clearer answers but is not
+without limitations. Due to bowel prep the communities left for sampling
+may not be reflective of the resident microbiota, but rather a
 collection of what is able to keep adhered to the mucosa. Additionally,
-these samples would contain much more host DNA then microbial DNA,
-potentially limiting the types of analysis that can be done. It has also
-been well published that low biomass samples can be very difficult to
-work with and results can end up being study dependent due to the
-randomness of contamination {[}14{]}. Due to these many differences, one
-question our meta-analysis aims to answer is whether it is possible to
-find consistent patterns that emerge across studies regardless of
-whether they used stool or tissue samples.
-
-This intense focus on identifying biomarkers has identified mostly
-mouth-associated microbes as potential CRC-associated microbes in the GI
-tract {[}15--17{]}. The main bacteria of interest arising from this set
-of microbes has been those within the \emph{Fusobacterium} genus. Yet,
-the question remains as to whether or not this is indeed the most
-important genera to be focusing on, since many microbiota-based studies
-typically have identified a collection of oral microbes rather than
-single species from a single genera {[}16,17{]}. Based on this
-discrepency, the second question we want to answer with this
-meta-analysis is if there is one dominant CRC-associated genera that can
-be identified across studies.
-
-The heavy concentration on the identification of biomarkers has mostly
-dominated the study of CRC and has had an unattended consequence of
-reducing the focus on changes that occur within the underlying resident
-community. This has been borne out by the majority of previous studies,
-within stool, tissue, and the only meta-analysis investigating this
-area, that have focused predominately on biomarker identification. Since
-it has not been investigated in detail the final question that this
-meta-analysis aims to answer is whether there are consistent detectable
-community differences as disease severity increases.
+these samples contain more host DNA, potentially limiting the types of
+analysis that can be done. It is well known that low biomass samples can
+be very difficult to work with and results can end up being study
+dependent due to the randomness of contamination {[}14{]}. Due to these
+many differences that could arise between stool and tissue, one question
+our meta-analysis aims to answer is whether there are consistent
+patterns that emerge across studies regardless of whether they used
+stool or tissue samples.
+
+This intense focus on identifying biomarkers has found strong
+associations with resident mouth microbes as potential CRC-associated
+microbes in the GI tract {[}15--17{]}. The main bacteria of interest,
+arising from this set of microbes, has been those within the
+\emph{Fusobacterium} genus. Yet, the question remains as to whether or
+not this is indeed the most important genera to be focusing on, since
+many microbiota-based studies typically have identified a collection of
+oral microbes rather than single species from a single genera
+{[}16,17{]}. Based on this discrepency, the second question we want to
+answer with this meta-analysis is if there is one dominant
+CRC-associated genera that can be identified across studies.
+
+The identification of microbial biomarkers has mostly dominated the
+study of CRC and has had an unattended consequence of reducing the focus
+on changes that occur within the underlying resident community. This has
+been borne out by the majority of previous studies, within stool,
+tissue, and the only meta-analysis investigating this area to date, that
+focus predominately on biomarker identification. Yet, how these
+CRC-associated microbes interact with their community may be an
+important component to consider, and has not been investigated in great
+detail by other meta-analyses. In response to this gap, our study aims
+to answer whether there are consistent detectable community differences
+as disease severity increases.
 
@@
 community is an important risk factor for both adenoma and carcinoma. To
-accomplish this we first assessed whether the diversity changes
-throughout disease (control to adenoma to carcinoma) and if it results
-in an increased relative risk (RR) for adenoma or carcinoma. We then
-assessed how common CRC-associated genera (\emph{Fusobacterium},
+accomplish this we first assessed whether diversity changes throughout
+disease (control to adenoma to carcinoma) and if it results in an
+increased relative risk (RR) for adenoma or carcinoma. We then assessed
+how common CRC-associated genera (\emph{Fusobacterium},
 \emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
@@
 CRC-associated genera resulted in better model classification based on
-the area under the curve (AUC). We observed that the community changes
-as disease severity worsens and that this community is important for
-disease classification. However, since the changes in community were
+the area under the curve (AUC). Our results suggest that the community
+changes as disease severity worsens and that this community is important
+for disease classification. However, since the changes in community were
 subtle for adenoma we also examined what effect and sample size the
 studies that were used were adequately powered for. Although we analyzed
-data sets which sampled large numbers of individuals, our results
+data sets that sampled large numbers of individuals, our results
 indicate the individual studies were underpowered for detecting effect

### 2017-12-11 submission/manuscript.Rmd
incorporate edits to first results section -- adding clarity

@@
 ***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
-Using the combined data set we first assessed whether there were any broad scale community differences that could be detected as disease severity worsened.
-Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces, but not tissue, were  lower in those with carcinoma [Figure 1].
-Using linear mixed-effect models to control for study and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`).
-However, in tissue, this effect was not observed when resampling of the same individual was also controlled for (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
-We next tested whether these detectable differences in the community resulted in a significant increase in RR.
-Within fecal samples, a decrease in Shannon diversity and evenness resulted in a significantly increased RR for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2].
-Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`) and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)) [Figure 2].
-Only a decrease in evenness had an increased RR for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)).
+By power transforming and Z-score normalizing the $\alpha$-diversity metrics for the entire data set we assessed whether there were any broad scale community differences that could be detected as disease severity worsened.
+For feces and not tissue a decrease in evenness and Shannon diversity was associated with carcinoma [Figure 1].
+Using linear mixed-effect models to control for study, resampling of the same individual (tissue studies only), and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness and Shannon diversity in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) but not for tissue (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
+We next tested whether these detectable differences in community resulted in significant increases in RR.
+For fecal samples, a decrease in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))` and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
 Interestingly, for both adenoma and carcinoma there was no increase in RR within tissue samples for any alpha diversity metric investigated [Table S1-S3].

### 2017-12-11 submission/manuscript.tex
incorporate edits to first results section -- adding clarity

@@
 \textbf{\emph{Lower Community Diversity is Associated with Increased RR
-of Carcinomas:}} Using the combined data set we first assessed whether
+of Carcinomas:}} By power transforming and Z-score normalizing the
+\(\alpha\)-diversity metrics for the entire data set we assessed whether
 there were any broad scale community differences that could be detected
-as disease severity worsened. Using power transformed and Z-score
-normalized \(\alpha\)-diversity metrics, both evenness and Shannon
-diversity in feces, but not tissue, were lower in those with carcinoma
-{[}Figure 1{]}. Using linear mixed-effect models to control for study
-and variable region, there was a significant decrease from control to
-adenoma to carcinoma for both evenness (P-value = 0.025) and Shannon
-diversity (P-value = 0.043). However, in tissue, this effect was not
-observed when resampling of the same individual was also controlled for
-(P-value \textgreater{} 0.05). We next tested whether these detectable
-differences in the community resulted in a significant increase in RR.
-Within fecal samples, a decrease in Shannon diversity and evenness
-resulted in a significantly increased RR for carcinoma (P-value = 0.0047
-and 0.00034, respectively) {[}Figure 2{]}. Although these values were
-significant, the effect size was relatively small for both metrics
-(Shannon RR = 1.33 (1.09 - 1.62) and evenness RR = 1.36 (1.15 - 1.61))
-{[}Figure 2{]}. Only a decrease in evenness had an increased RR for
-adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect size was
-even smaller than what was observed for carcinoma (RR = 1.16 (1.01 -
-1.34)). Interestingly, for both adenoma and carcinoma there was no
-increase in RR within tissue samples for any alpha diversity metric
-investigated {[}Table S1-S3{]}.
+as disease severity worsened. For feces and not tissue a decrease in
+evenness and Shannon diversity was associated with carcinoma {[}Figure
+1{]}. Using linear mixed-effect models to control for study, resampling
+of the same individual (tissue studies only), and variable region, there
+was a significant decrease from control to adenoma to carcinoma for both
+evenness and Shannon diversity in stool (P-value = 0.025 and 0.043,
+respectively) but not for tissue (P-value \textgreater{} 0.05). We next
+tested whether these detectable differences in community resulted in
+significant increases in RR. For fecal samples, a decrease in evenness
+resulted in a significantly increased RR for carcinoma (RR = 1.36 (1.15
+- 1.61), P-value = 0.00034 and adenoma (RR = 1.16 (1.01 - 1.34), P-value
+= 0.032) while a decrease in Shannon diversity only increased the RR for
+carcinoma (RR = 1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}.
+Interestingly, for both adenoma and carcinoma there was no increase in
+RR within tissue samples for any alpha diversity metric investigated
+{[}Table S1-S3{]}.
 

### 2017-12-12 submission/manuscript.Rmd
update first and second result sections

@@
 By power transforming and Z-score normalizing the $\alpha$-diversity metrics for the entire data set we assessed whether there were any broad scale community differences that could be detected as disease severity worsened.
-For feces and not tissue a decrease in evenness and Shannon diversity was associated with carcinoma [Figure 1].
-Using linear mixed-effect models to control for study, resampling of the same individual (tissue studies only), and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness and Shannon diversity in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) but not for tissue (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
+Using linear mixed-effect models to control for study, resampling of the same individual (tissue studies only), and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness and Shannon diversity in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) but not for tissue (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1].
 We next tested whether these detectable differences in community resulted in significant increases in RR.
@@
 ***Carcinoma-Associated Genera Minimally Impacts RR of Adenoma:***
-Based on the small increase in RR using $\alpha$-diversity metrics, we wanted to know if the presence of specific genera resulted in a higher RR for both stool and tissue.
-To investigate this we analyzed the classically associated CRC genera, *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for and increase in RR.
+Based on the small increase in RR using $\alpha$-diversity metrics, we assessed if the presence of specific genera resulted in a higher RR for both stool and tissue.
+To investigate this we analyzed the classically associated CRC genera, *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for an increase in RR.
 The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3].
-In fecal samples the RR due to CRC associated genera was greater than either the RR assoicated with evenness or Shannon diversity [Figure 2 & 3].
-Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D].
 The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) than for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`).
 This decrease may be explained by the fact that the tissue analysis included matched samples from the same individual.
+In fecal samples, the RR for carcinoma due to the prescence of CRC-associated genera was greater than either the RR associated with evenness or Shannon diversity [Figure 2 & 3].
+Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC-associated genera [Figure 3B & 3D].
 

### 2017-12-12 submission/manuscript.tex
update first and second result sections

@@
 there were any broad scale community differences that could be detected
-as disease severity worsened. For feces and not tissue a decrease in
-evenness and Shannon diversity was associated with carcinoma {[}Figure
-1{]}. Using linear mixed-effect models to control for study, resampling
-of the same individual (tissue studies only), and variable region, there
-was a significant decrease from control to adenoma to carcinoma for both
-evenness and Shannon diversity in stool (P-value = 0.025 and 0.043,
-respectively) but not for tissue (P-value \textgreater{} 0.05). We next
-tested whether these detectable differences in community resulted in
-significant increases in RR. For fecal samples, a decrease in evenness
-resulted in a significantly increased RR for carcinoma (RR = 1.36 (1.15
-- 1.61), P-value = 0.00034 and adenoma (RR = 1.16 (1.01 - 1.34), P-value
-= 0.032) while a decrease in Shannon diversity only increased the RR for
+as disease severity worsened. Using linear mixed-effect models to
+control for study, resampling of the same individual (tissue studies
+only), and variable region, there was a significant decrease from
+control to adenoma to carcinoma for both evenness and Shannon diversity
+in stool (P-value = 0.025 and 0.043, respectively) but not for tissue
+(P-value \textgreater{} 0.05) {[}Figure 1{]}. We next tested whether
+these detectable differences in community resulted in significant
+increases in RR. For fecal samples, a decrease in evenness resulted in a
+significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
+P-value = 0.00034 and adenoma (RR = 1.16 (1.01 - 1.34), P-value = 0.032)
+while a decrease in Shannon diversity only increased the RR for
 carcinoma (RR = 1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}.
@@
 Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
-metrics, we wanted to know if the presence of specific genera resulted
-in a higher RR for both stool and tissue. To investigate this we
-analyzed the classically associated CRC genera, \emph{Fusobacterium},
+metrics, we assessed if the presence of specific genera resulted in a
+higher RR for both stool and tissue. To investigate this we analyzed the
+classically associated CRC genera, \emph{Fusobacterium},
 \emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}
-for and increase in RR. The majority of CRC-associated genera for both
+for an increase in RR. The majority of CRC-associated genera for both
 feces and tissue had a significantly increased RR for carcinoma but not
-for adenoma {[}Figure 3{]}. In fecal samples the RR due to CRC
-associated genera was greater than either the RR assoicated with
-evenness or Shannon diversity {[}Figure 2 \& 3{]}. Additionally, the RR
-of carcinoma continuously increased as individuals tested positive for
-more CRC associated genera {[}Figure 3B \& 3D{]}. The RR effect size was
-greater for stool (RR range = 1.62 - 2.37) than for tissue (RR range =
-1.21 - 1.81). This decrease may be explained by the fact that the tissue
-analysis included matched samples from the same individual.
+for adenoma {[}Figure 3{]}. The RR effect size was greater for stool (RR
+range = 1.62 - 2.37) than for tissue (RR range = 1.21 - 1.81). This
+decrease may be explained by the fact that the tissue analysis included
+matched samples from the same individual. In fecal samples, the RR for
+carcinoma due to the prescence of CRC-associated genera was greater than
+either the RR associated with evenness or Shannon diversity {[}Figure 2
+\& 3{]}. Additionally, the RR of carcinoma continuously increased as
+individuals tested positive for more CRC-associated genera {[}Figure 3B
+\& 3D{]}.
 

### 2017-12-12 submission/manuscript.Rmd
add results section models --better version

@@
 
-***Using the Whole Community Instead of Only CRC-Associated Genera Increases Model AUC:***
-We then tested whether the overall bacterial community was at all important to classifying disease or if the CRC-associated genera were sufficient alone.
+***Whole Community Models Add Important Community Context:***
+Since CRC-associated genera had increased RR for carcinoma over diversity metrics we wondered whether the overall bacterial community was at all important to classifying disease or if the CRC-associated genera were sufficient alone.
 To test this we used two approaches.
-The first used genus level data and tested whether there were any differences in AUC when training on one study and testing on all the others when using either all genera present or only the CRC-associated genera.
+The first approach used genus level data and tested whether there were any differences in AUC when training on one study and testing on all the others when using either all genera present or only the CRC-associated genera.
 The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model across studies using either all OTUs or only OTUs that taxonomically classified to CRC-associated genera.
 
-The genus based models showed an AUC decrease in model classification on the training set for both stool and tissue studies [Figure S2-S3].
+Our first approach using genus based models showed an AUC decrease in model classification on the training set for both stool and tissue studies [Figure S2-S3].
 With respect to the test sets, comprised of genera data from other studies, both the all genera model and CRC-associated models had a similar ability to detect adenomas or carcinomas [Figure S4-S6].
-The AUC for classification of adenomas was lower than carcionmas for both tissue and stool [Figure S4-S6].
-Interestingly, the AUC for the classification of carcinoma was consistently lower for the tissue models than the stool models [Figure S4-S6].
+Two interesting but seperate general observations from these models were that: 1) classification of adenomas was lower than carcinomas for both tissue and stool and 2) AUC for the classification of carcinoma was consistently lower for the tissue models than the stool models [Figure S4-S6].
 
-The OTU based models for both fecal and tissue (matched and unmatched) samples, showed an AUC decrease when only OTUs from the CRC-associated genera are used versus the full community of OTUs [Figure 4 & 5].
-This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5] with the largest difference in median AUC for stool being in carcinoma classification [Figure 4B] and for tissue being an adenoma [Figure 5A].
+A similar trend was observed for the OTU based models for both fecal and tissue (matched and unmatched) samples.
+There was a generalized decrease in AUC when only OTUs from the CRC-associated genera were used versus the full community of OTUs for both adenoma and carcinoma [Figure 4 & 5].
+The largest difference in median AUC for stool, between the full community of OTUs and only CRC-associated OTUs, was in carcinoma classification [Figure 4B] and for tissue it was in adenoma classification [Figure 5A].
 
-In stool the most common genera in the genus based models belonged mostly to resident genera [Figure 6A & B].
-This included genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia*.
-With respect to the CRC-associated genera *Fusobacterium* was the only genus present in adenoma while all four were present in carcinoma [Figure 6A & B].
-However, none of these CRC-associated genera were present in the majority of studies.
-For the adenoma OTU models, OTUs that classified as  *Ruminococcaceae* or *Roseburia* were present in the top 10 OTUs for the vast majority of studies [Figure 6C].
-Although for the carcinoma OTU models *Ruminococcaceae* was in the top 10 for many studies, *Bacteroides* was present in the overwhelming majority of the carcinoma OTU stool models [Figure 6D].
+In stool the most common genera in the genus based models belonged mostly to resident genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
+With respect to the CRC-associated genera, *Fusobacterium* was the only genus present in adenoma while all four were present in carcinoma [Figure 6A & B].
+Conversely, none of these CRC-associated genera were present in the majority of studies.
+When we move onto the OTU based models, the adenoma OTU models had OTUs that classified as  *Ruminococcaceae* or *Roseburia* present in the top 10 OTUs for the vast majority of studies [Figure 6C].
+*Ruminococcaceae* was also present in the top 10 in some studies for the carcinoma OTU models, but it was *Bacteroides* that was present in the overwhelming majority of the carcinoma OTU stool models [Figure 6D].
 
 Conversely to the stool models, both genera and OTU based models in tissue had the vast majority of their top 10 occur in a study specific manner [Figure S7].
-*Fusobacterium* and *Fusobacteriaceae* show up more often in the top 10 for matched tissue samples but are not present in the top 10 or much lower for unmatched tissue [Figure S7B-C & S7E-F].
+*Fusobacterium* and *Fusobacteriaceae* show up more often in the top 10 for matched tissue samples but were not present in the top 10 or were much lower ranked for unmatched tissue [Figure S7B-C & S7E-F].
 There appears to be very little overlap in the top 10 most important variables between stool and tissue for both adenoma and carcinoma [Figure 6 & S7].

### 2017-12-12 submission/manuscript.tex
add results section models --better version

@@
 
-\textbf{\emph{Using the Whole Community Instead of Only CRC-Associated
-Genera Increases Model AUC:}} We then tested whether the overall
-bacterial community was at all important to classifying disease or if
-the CRC-associated genera were sufficient alone. To test this we used
-two approaches. The first used genus level data and tested whether there
-were any differences in AUC when training on one study and testing on
-all the others when using either all genera present or only the
-CRC-associated genera. The second approach used OTU level data and
-tested whether there was a generalized decrease in the 10-fold cross
-validation (CV) model across studies using either all OTUs or only OTUs
-that taxonomically classified to CRC-associated genera.
-
-The genus based models showed an AUC decrease in model classification on
-the training set for both stool and tissue studies {[}Figure S2-S3{]}.
-With respect to the test sets, comprised of genera data from other
-studies, both the all genera model and CRC-associated models had a
-similar ability to detect adenomas or carcinomas {[}Figure S4-S6{]}. The
-AUC for classification of adenomas was lower than carcionmas for both
-tissue and stool {[}Figure S4-S6{]}. Interestingly, the AUC for the
+\textbf{\emph{Whole Community Models Add Important Community Context:}}
+Since CRC-associated genera had increased RR for carcinoma over
+diversity metrics we wondered whether the overall bacterial community
+was at all important to classifying disease or if the CRC-associated
+genera were sufficient alone. To test this we used two approaches. The
+first approach used genus level data and tested whether there were any
+differences in AUC when training on one study and testing on all the
+others when using either all genera present or only the CRC-associated
+genera. The second approach used OTU level data and tested whether there
+was a generalized decrease in the 10-fold cross validation (CV) model
+across studies using either all OTUs or only OTUs that taxonomically
+classified to CRC-associated genera.
+
+Our first approach using genus based models showed an AUC decrease in
+model classification on the training set for both stool and tissue
+studies {[}Figure S2-S3{]}. With respect to the test sets, comprised of
+genera data from other studies, both the all genera model and
+CRC-associated models had a similar ability to detect adenomas or
+carcinomas {[}Figure S4-S6{]}. Two interesting but seperate general
+observations from these models were that: 1) classification of adenomas
+was lower than carcinomas for both tissue and stool and 2) AUC for the
 classification of carcinoma was consistently lower for the tissue models
@@
 
-The OTU based models for both fecal and tissue (matched and unmatched)
-samples, showed an AUC decrease when only OTUs from the CRC-associated
-genera are used versus the full community of OTUs {[}Figure 4 \& 5{]}.
-This decrease is observed in both adenoma and carcinoma groups {[}Figure
-4 \& 5{]} with the largest difference in median AUC for stool being in
-carcinoma classification {[}Figure 4B{]} and for tissue being an adenoma
-{[}Figure 5A{]}.
+A similar trend was observed for the OTU based models for both fecal and
+tissue (matched and unmatched) samples. There was a generalized decrease
+in AUC when only OTUs from the CRC-associated genera were used versus
+the full community of OTUs for both adenoma and carcinoma {[}Figure 4 \&
+5{]}. The largest difference in median AUC for stool, between the full
+community of OTUs and only CRC-associated OTUs, was in carcinoma
+classification {[}Figure 4B{]} and for tissue it was in adenoma
+classification {[}Figure 5A{]}.
 
 In stool the most common genera in the genus based models belonged
-mostly to resident genera {[}Figure 6A \& B{]}. This included genera
-such as \emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia}.
-With respect to the CRC-associated genera \emph{Fusobacterium} was the
-only genus present in adenoma while all four were present in carcinoma
-{[}Figure 6A \& B{]}. However, none of these CRC-associated genera were
-present in the majority of studies. For the adenoma OTU models, OTUs
-that classified as \emph{Ruminococcaceae} or \emph{Roseburia} were
-present in the top 10 OTUs for the vast majority of studies {[}Figure
-6C{]}. Although for the carcinoma OTU models \emph{Ruminococcaceae} was
-in the top 10 for many studies, \emph{Bacteroides} was present in the
+mostly to resident genera such as \emph{Ruminococcus},
+\emph{Bacteroides}, and \emph{Roseburia} {[}Figure 6A \& B{]}. With
+respect to the CRC-associated genera, \emph{Fusobacterium} was the only
+genus present in adenoma while all four were present in carcinoma
+{[}Figure 6A \& B{]}. Conversely, none of these CRC-associated genera
+were present in the majority of studies. When we move onto the OTU based
+models, the adenoma OTU models had OTUs that classified as
+\emph{Ruminococcaceae} or \emph{Roseburia} present in the top 10 OTUs
+for the vast majority of studies {[}Figure 6C{]}. \emph{Ruminococcaceae}
+was also present in the top 10 in some studies for the carcinoma OTU
+models, but it was \emph{Bacteroides} that was present in the
 overwhelming majority of the carcinoma OTU stool models {[}Figure 6D{]}.
@@
 manner {[}Figure S7{]}. \emph{Fusobacterium} and \emph{Fusobacteriaceae}
-show up more often in the top 10 for matched tissue samples but are not
-present in the top 10 or much lower for unmatched tissue {[}Figure S7B-C
-\& S7E-F{]}. There appears to be very little overlap in the top 10 most
-important variables between stool and tissue for both adenoma and
-carcinoma {[}Figure 6 \& S7{]}.
+show up more often in the top 10 for matched tissue samples but were not
+present in the top 10 or were much lower ranked for unmatched tissue
+{[}Figure S7B-C \& S7E-F{]}. There appears to be very little overlap in
+the top 10 most important variables between stool and tissue for both
+adenoma and carcinoma {[}Figure 6 \& S7{]}.
 

### 2017-12-12 submission/manuscript.Rmd
complete imporvements to disussion -- reduce causation talk and improve flow

@@
 Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease [Figure 1-3].
-Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the relative risk of either of these two conditions.
+Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the relative risk of either of these two.
 Even though CRC-associated genera increase the RR of carcinoma, they do not consistently increase the relative risk of adenoma.
-This information suggests that these specific genera are important in carcinoma genesis but may not be the primary members of the microbial community contributing to the formation of an adenoma.
-Additionally, our data show that by using the whole community, our models perform better than when only the CRC-associated genera are included.
+This information suggests that these specific genera are strongly associated with carcinoma but may not be associated with adenoma.
+Although CRC-associated genera are asscoiated with carcinomas, our data show that by using the whole community, classification models perform better than when only the CRC-associated genera are included.
 CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions.
-These observations suggest that small localized changes within the community on tissue may be occuring that are important in early disease progression of CRC and that this process may not directly involve CRC-associated genera.
+These observations suggest that small localized changes within the community on tissue may be occuring, that are important in early disease progression of CRC, and that this process may not directly involve CRC-associated genera.
 
@@
 The central idea of the model is that a single bacterium initiates an environment in which other non-resident microbes may then be able to colonize, creating a vicious cycle that is conducive for CRC.
-Both the drastically increased RR of CRC-associated genera versus $\alpha$-diversity metrics for carcinoma and increasing RR with more CRC-associated generea positivity are highly supportive of this model.
+Both the drastically increased RR of CRC-associated genera versus $\alpha$-diversity metrics for carcinoma and increasing RR with more CRC-associated genera positivity are highly supportive of this model.
 However, the initial establishment of the driver within the system appears to be dependent on the current community.
@@
 The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case.
-One potential hypothesis from these observations is that at early stages of the diease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.
+One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.
 
@@
 This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community.
-These observations would also suggest that initial changes in the resident community, specifically to *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma.
+These observations would also suggest that initial changes in the resident community, specifically to *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma.
 Based on these observations, it is possible to hypothesize that the early changes in the community may give rise to initial polyp formation via interactions with the host and not necessarily via interactions with CRC-assciated genera.
@@
 
-The tissue studies did not provide a clearer understanding of how the microbiota may drive the progression of disease severity.
+The tissue studies did not provide a clearer understanding of how the microbiota may be associated with disease severity.
 For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 important variables across studies.
@@
 
-Although there are still questions that need to be answered regarding the microbiota and carcinoma, a clearer framework for their relationship is beginning to develop as to how this occurs.
-From our observations many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 7].
-Conversely, the associations between the microbiota and adenoma is less clear and part of the reason this may be is because many studies are not powered effectively to observe the small changes reported here.
+The associations between the microbiota and adenoma are less clear in part because many studies may not be powered effectively to observe the small changes reported here.
 None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls.
@@
 Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out how the microbiota contributes to adenoma formation.
+In contrast to adenoma, our observations suggest that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 7].
 

### 2017-12-12 submission/manuscript.tex
complete imporvements to disussion -- reduce causation talk and improve flow

@@
 adenoma to carcinoma, this did not translate into large effect sizes for
-the relative risk of either of these two conditions. Even though
-CRC-associated genera increase the RR of carcinoma, they do not
-consistently increase the relative risk of adenoma. This information
-suggests that these specific genera are important in carcinoma genesis
-but may not be the primary members of the microbial community
-contributing to the formation of an adenoma. Additionally, our data show
-that by using the whole community, our models perform better than when
-only the CRC-associated genera are included. CRC-associated genera are
-clearly important to carcinoma pathogenesis but accounting for the
-community in which these microbes exist can drastically increase the
-ability of models to make predictions. These observations suggest that
-small localized changes within the community on tissue may be occuring
-that are important in early disease progression of CRC and that this
-process may not directly involve CRC-associated genera.
+the relative risk of either of these two. Even though CRC-associated
+genera increase the RR of carcinoma, they do not consistently increase
+the relative risk of adenoma. This information suggests that these
+specific genera are strongly associated with carcinoma but may not be
+associated with adenoma. Although CRC-associated genera are asscoiated
+with carcinomas, our data show that by using the whole community,
+classification models perform better than when only the CRC-associated
+genera are included. CRC-associated genera are clearly important to
+carcinoma pathogenesis but accounting for the community in which these
+microbes exist can drastically increase the ability of models to make
+predictions. These observations suggest that small localized changes
+within the community on tissue may be occuring, that are important in
+early disease progression of CRC, and that this process may not directly
+involve CRC-associated genera.
 
@@
 CRC-associated genera versus \(\alpha\)-diversity metrics for carcinoma
-and increasing RR with more CRC-associated generea positivity are highly
+and increasing RR with more CRC-associated genera positivity are highly
 supportive of this model. However, the initial establishment of the
@@
 this is the case. One potential hypothesis from these observations is
-that at early stages of the diease, how the host interacts with these
+that at early stages of the disease, how the host interacts with these
 subtle changes is what ultimately leads to a thoroughly dysfunctional
@@
 also suggest that initial changes in the resident community,
-specifically to \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
-\emph{Roseburia}, carry on from adenoma to carcinoma. Based on these
-observations, it is possible to hypothesize that the early changes in
-the community may give rise to initial polyp formation via interactions
-with the host and not necessarily via interactions with CRC-assciated
-genera. These changes then create new niches in which any one of the
-CRC-associated genera could gain a foothold, exacerbating the initial
-changes in community and facilitating the transition from adenoma to
-carcinoma via a driver-passenger type mechanism.
+specifically to \emph{Bacteroides}, \emph{Ruminococcaceae},
+\emph{Ruminococcus}, and \emph{Roseburia}, carry on from adenoma to
+carcinoma. Based on these observations, it is possible to hypothesize
+that the early changes in the community may give rise to initial polyp
+formation via interactions with the host and not necessarily via
+interactions with CRC-assciated genera. These changes then create new
+niches in which any one of the CRC-associated genera could gain a
+foothold, exacerbating the initial changes in community and facilitating
+the transition from adenoma to carcinoma via a driver-passenger type
+mechanism.
 
 The tissue studies did not provide a clearer understanding of how the
-microbiota may drive the progression of disease severity. For the OTU
-models both the unmatched and matched {[}Figure S7E \& F{]} tissue
-samples had some concordance with the stool data, with resident bacteria
-being the most prevalent in the top 10 important variables across
-studies. Unlike in stool, \emph{Fusobacterium} was the only
-CRC-associated bacteria consistently present in the top 10 of the CRC
-models {[}Figure S7B-C \& E-F{]}. The majority of the results seem to be
-study specific with many top 10 taxa being present only in a single
-study. One other potentially worrying sign is the presence of
-\emph{Propionibacterium} within the top 10 for the genera and OTU models
-and could be a marker of contamination. The low biomass of these samples
-coupled with potential contamination might be a possible reason why the
-tissue results seem to be more sporadic then the stool results.
-
-Although there are still questions that need to be answered regarding
-the microbiota and carcinoma, a clearer framework for their relationship
-is beginning to develop as to how this occurs. From our observations
-many changes in carcinoma could easily result in effect sizes that are
-30\% or more between the case and control and most studies analyzed have
+microbiota may be associated with disease severity. For the OTU models
+both the unmatched and matched {[}Figure S7E \& F{]} tissue samples had
+some concordance with the stool data, with resident bacteria being the
+most prevalent in the top 10 important variables across studies. Unlike
+in stool, \emph{Fusobacterium} was the only CRC-associated bacteria
+consistently present in the top 10 of the CRC models {[}Figure S7B-C \&
+E-F{]}. The majority of the results seem to be study specific with many
+top 10 taxa being present only in a single study. One other potentially
+worrying sign is the presence of \emph{Propionibacterium} within the top
+10 for the genera and OTU models and could be a marker of contamination.
+The low biomass of these samples coupled with potential contamination
+might be a possible reason why the tissue results seem to be more
+sporadic then the stool results.
+
+The associations between the microbiota and adenoma are less clear in
+part because many studies may not be powered effectively to observe the
+small changes reported here. None of the studies analyzed were properly
+powered to detect a 10\% or lower change between case and controls. This
+small effect size range may well be the scope in which differences
+consistently occur in adenoma due to the subtle changes in community
+that occur between control and adenoma. Future studies investigating
+adenoma and the microbiota need to take these factors into consideration
+if we are to work out how the microbiota contributes to adenoma
+formation. In contrast to adenoma, our observations suggest that many
+changes in carcinoma could easily result in effect sizes that are 30\%
+or more between the case and control and most studies analyzed have
 sufficient power to detect these types of changes {[}Figure 7{]}.
-Conversely, the associations between the microbiota and adenoma is less
-clear and part of the reason this may be is because many studies are not
-powered effectively to observe the small changes reported here. None of
-the studies analyzed were properly powered to detect a 10\% or lower
-change between case and controls. This small effect size range may well
-be the scope in which differences consistently occur in adenoma due to
-the subtle changes in community that occur between control and adenoma.
-Future studies investigating adenoma and the microbiota need to take
-these factors into consideration if we are to work out how the
-microbiota contributes to adenoma formation.
 

### 2017-12-12 submission/manuscript.Rmd
tone down conclusions
fix grammar and structure of methods and declaration section

@@
 Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community.
-Although there are power limitations associated with adenoma, this report highlights the strong influence the microbiota has on CRC development.
+Although there are power limitations associated with adenoma, this report highlights the strong associations the microbiota has on CRC.
 
@@
 Additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature.
-Studies that used tissue or feces as their sample source for 16S rRNA gene sequencing analysis were included.
-Studies using either 454 or Illumina sequencing technology were included.
+Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis were included.
 Only data sets that had sequences available for analysis were included.
-Some studies did not have publically available sequences or did not have metadata in which the authors were able to share.
+Some studies did not have publically available sequences or did not have metadata in which the authors were able to share and were excluded.
 After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014].
 The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016].
-Additionally, after sequence processing all the case samples for the Kostic study only had 100 sequences remaining and was not used.
+Additionally, after sequence processing all the case samples for the Kostic study only had 100 or less sequences remaining and was not used.
 This left a total of 14 studies for which analysis could be completed.
@@
 ***Sequence Processing:***
-For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained from the by searching the respective accession number of the study following website: http://www.ncbi.nlm.nih.gov/Traces/study/.
+For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained by searching the respective accession number of the study at the following website: http://www.ncbi.nlm.nih.gov/Traces/study/.
 Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap for one study [@ahn_human_2013] and for four studies was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013].
@@
 If it was not possible to use these defaults, the stated quality cut-offs were used instead.
-Chimeras were identifed and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] was utilized.
+Chimeras were identifed and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].
 
@@
 For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
-Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study and variable region effects in feces and study, variable region, and individual effects in tissue.
+Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and variable region effects.
 Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many with and without disease were above and below the overall median value within the specific study.
@@
 Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002].
-Differences between the obtained AUC versus a random model AUC was assessed using T-tests.
 Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016].
@@
 We analyzed the data using linear mixed-effect models and relative risk.
-$\beta$-diversity was then assessed for each inidividual study.
+$\beta$-diversity was then assessed for each individual study.
 Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk.
@@
 The data was split between feces and tissue samples.
-Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched) tissue samples.
-Where applicable for each study, predictions for adenoma and carcinoma were tested.
+Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched).
+Where applicable for each study, predictions for adenoma and carcinoma were then tested for feces, matched tissue, and unmatched tissue.
 This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs.
 For OTU based models, the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
-The power of each study was assessed for an effect size ranging from 1% to 30%.
+Finally, the power of each study was assessed for an effect size ranging from 1% to 30%.
 An estimated sample n for these effect sizes was also generated based on 80% power.
@@
 ***Assessing Important Random Forest Model Variables:***
-The genus based models collected the top 10 most important variables from each training set and assessed how many times that genera showed up in the top 10 across each study.
+The genus based models collected the top 10 most important variables, measured by Mean Decrease in Accuracy (MDA), from each training set and assessed how many times that genera showed up in the top 10 across each study.
 The OTU based models recorded the medians for each OTU across 100 different 80/20 splits of the data for each study.
@@
 The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
-Unless mentioned otherwise, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.
+Unless otherwise mentioned, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.
 
@@
 We would also like to thank each of the study authors for making their data available for use.
-Finally we would like to thank the members of the Schloss lab for valuable feed back and proof reading during the formulation of this manuscript.
+Finally, we would like to thank the members of the Schloss lab for valuable feed back and proof reading during the formulation of this manuscript.
 

### 2017-12-12 submission/manuscript.tex
tone down conclusions
fix grammar and structure of methods and declaration section

@@
 subtle changes in the community. Although there are power limitations
-associated with adenoma, this report highlights the strong influence the
-microbiota has on CRC development.
+associated with adenoma, this report highlights the strong associations
+the microbiota has on CRC.
 
@@
 of the literature. Studies that used tissue or feces as their sample
-source for 16S rRNA gene sequencing analysis were included. Studies
-using either 454 or Illumina sequencing technology were included. Only
-data sets that had sequences available for analysis were included. Some
-studies did not have publically available sequences or did not have
-metadata in which the authors were able to share. After these filtering
-steps, the following studies remained: Ahn, \emph{et al.} {[}26{]},
-Baxter, \emph{et al.} {[}16{]}, Brim, \emph{et al.} {[}22{]}, Burns,
-\emph{et al.} {[}27{]}, Chen, \emph{et al.} {[}19{]}, Dejea, \emph{et
-al.} {[}24{]}, Flemer, \emph{et al.} {[}12{]}, Geng, \emph{et al.}
-{[}28{]}, Hale, \emph{et al.} {[}18{]}, Kostic, \emph{et al.} {[}15{]},
-Lu, \emph{et al.} {[}21{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang,
-\emph{et al.} {[}20{]}, Weir, \emph{et al.} {[}23{]}, and Zeller,
-\emph{et al.} {[}17{]}. The Zackular {[}31{]} study was not included
-becasue the 90 individuals analyzed within the study are contained
-within the larger Baxter study {[}16{]}. Additionally, after sequence
-processing all the case samples for the Kostic study only had 100
-sequences remaining and was not used. This left a total of 14 studies
-for which analysis could be completed.
+source for 454 or Illumina 16S rRNA gene sequencing analysis were
+included. Only data sets that had sequences available for analysis were
+included. Some studies did not have publically available sequences or
+did not have metadata in which the authors were able to share and were
+excluded. After these filtering steps, the following studies remained:
+Ahn, \emph{et al.} {[}26{]}, Baxter, \emph{et al.} {[}16{]}, Brim,
+\emph{et al.} {[}22{]}, Burns, \emph{et al.} {[}27{]}, Chen, \emph{et
+al.} {[}19{]}, Dejea, \emph{et al.} {[}24{]}, Flemer, \emph{et al.}
+{[}12{]}, Geng, \emph{et al.} {[}28{]}, Hale, \emph{et al.} {[}18{]},
+Kostic, \emph{et al.} {[}15{]}, Lu, \emph{et al.} {[}21{]}, Sanapareddy,
+\emph{et al.} {[}25{]}, Wang, \emph{et al.} {[}20{]}, Weir, \emph{et
+al.} {[}23{]}, and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]}
+study was not included becasue the 90 individuals analyzed within the
+study are contained within the larger Baxter study {[}16{]}.
+Additionally, after sequence processing all the case samples for the
+Kostic study only had 100 or less sequences remaining and was not used.
+This left a total of 14 studies for which analysis could be completed.
 
@@
 (\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
-and metadata was obtained from the by searching the respective accession
-number of the study following website:
+and metadata was obtained by searching the respective accession number
+of the study at the following website:
 \url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
@@
 removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
-97\% similarity using the OptiClust algorithm {[}34{]} was utilized.
+97\% similarity was completed using the OptiClust algorithm {[}34{]}.
 
@@
 \(\alpha\)-diversity differences utilized linear mixed-effect models
-created using the lme4 (v1.1.14) package {[}38{]} to correct for study
-and variable region effects in feces and study, variable region, and
-individual effects in tissue. Relative risk was analyzed using both the
-epiR (v0.9.87) and metafor (v2.0.0) packages {[}39,40{]} by assessing
-how many with and without disease were above and below the overall
-median value within the specific study. Relative risk significance
-testing utilized the chi-squred test. \(\beta\)-diversity differences
-utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
-vegan (v2.4.4) package {[}41{]}. Random Forest models were built using
-both the caret (v6.0.77) and randomForest (v4.6.12) packages
-{[}42,43{]}. Differences between the obtained AUC versus a random model
-AUC was assessed using T-tests. Power analysis and estimations were made
-using the pwr (v1.2.1) and statmod (v1.4.30) packages {[}44,45{]}. All
-figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
-packages {[}46,47{]}.
+created using the lme4 (v1.1.14) package {[}38{]} to correct for study,
+repeat sampling of individuals (tissue only), and variable region
+effects. Relative risk was analyzed using both the epiR (v0.9.87) and
+metafor (v2.0.0) packages {[}39,40{]} by assessing how many with and
+without disease were above and below the overall median value within the
+specific study. Relative risk significance testing utilized the
+chi-squred test. \(\beta\)-diversity differences utilized a Bray-Curtis
+distance matrix and PERMANOVA executed with the vegan (v2.4.4) package
+{[}41{]}. Random Forest models were built using both the caret (v6.0.77)
+and randomForest (v4.6.12) packages {[}42,43{]}. Power analysis and
+estimations were made using the pwr (v1.2.1) and statmod (v1.4.30)
+packages {[}44,45{]}. All figures were created using both ggplot2
+(v2.2.1) and gridExtra (v2.3) packages {[}46,47{]}.
 
@@
 analyzed the data using linear mixed-effect models and relative risk.
-\(\beta\)-diversity was then assessed for each inidividual study. Next,
+\(\beta\)-diversity was then assessed for each individual study. Next,
 four specific CRC-associated genera (\emph{Fusobacterium},
@@
 samples from the same individual (matched) and those from different
-individuals (unmatched) tissue samples. Where applicable for each study,
-predictions for adenoma and carcinoma were tested. This same approach
-was then applied at the OTU level with the exception that instead of
-testing on the other studies, a 10-fold cross validation was utilized
-and 100 different models were created based on random 80/20 splitting of
-the data to generate a range of expected AUCs. For OTU based models, the
-CRC associated genera included all OTUs that had a taxonomic
-classification to \emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Peptostreptococcus}, or \emph{Porphyromonas}. The power of each
-study was assessed for an effect size ranging from 1\% to 30\%. An
-estimated sample n for these effect sizes was also generated based on
-80\% power. For comparisons in which normal versus adenoma were made the
-carcinoma samples were excluded from each respective study. Similarily,
-for comparisons in which normal versus carcinoma were made the adenoma
-samples were excluded from each respective study.
+individuals (unmatched). Where applicable for each study, predictions
+for adenoma and carcinoma were then tested for feces, matched tissue,
+and unmatched tissue. This same approach was then applied at the OTU
+level with the exception that instead of testing on the other studies, a
+10-fold cross validation was utilized and 100 different models were
+created based on random 80/20 splitting of the data to generate a range
+of expected AUCs. For OTU based models, the CRC associated genera
+included all OTUs that had a taxonomic classification to
+\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
+\emph{Porphyromonas}. Finally, the power of each study was assessed for
+an effect size ranging from 1\% to 30\%. An estimated sample n for these
+effect sizes was also generated based on 80\% power. For comparisons in
+which normal versus adenoma were made the carcinoma samples were
+excluded from each respective study. Similarily, for comparisons in
+which normal versus carcinoma were made the adenoma samples were
+excluded from each respective study.
 
@@
 \textbf{\emph{Assessing Important Random Forest Model Variables:}} The
-genus based models collected the top 10 most important variables from
-each training set and assessed how many times that genera showed up in
-the top 10 across each study. The OTU based models recorded the medians
-for each OTU across 100 different 80/20 splits of the data for each
-study. The lowest classification for each OTU was obtained using the RDP
-database and the number of times the specific classification occured in
-the top 10 across studies was recorded. For the adenoma tissue genus and
-OTU models there was only one matched and unmatched study and these
-results were grouped together for the counting of the top 10.
+genus based models collected the top 10 most important variables,
+measured by Mean Decrease in Accuracy (MDA), from each training set and
+assessed how many times that genera showed up in the top 10 across each
+study. The OTU based models recorded the medians for each OTU across 100
+different 80/20 splits of the data for each study. The lowest
+classification for each OTU was obtained using the RDP database and the
+number of times the specific classification occured in the top 10 across
+studies was recorded. For the adenoma tissue genus and OTU models there
+was only one matched and unmatched study and these results were grouped
+together for the counting of the top 10.
 
@@
 \url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
-Unless mentioned otherwise, the accession number for the raw sequences
+Unless otherwise mentioned, the accession number for the raw sequences
 for the studies used in this analysis can be found directly in the
@@
 thank each of the study authors for making their data available for use.
-Finally we would like to thank the members of the Schloss lab for
+Finally, we would like to thank the members of the Schloss lab for
 valuable feed back and proof reading during the formulation of this

### 2017-12-12 submission/manuscript.Rmd
final edits of final read through after pat rd 2 suggestions

@@
 Important drivers within this context have ranged from individual microbes to the whole community.
-Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community is an important driver of both early (adenoma) and late (carcinoma) stage of disease.
+Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community has important associations to both early (adenoma) and late (carcinoma) stage disease.
 To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.
@@
 Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern, with their presence significantly increasing the relative risk for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value = `r format(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]), digits = 2)`).
-Using the whole community versus only CRC associated genera to build a prediction model resulted in higher classification success based on Area Under the Curve (AUC) for both adenoma and carcinoma using fecal and tissue samples.
+Using the whole community versus only CRC-associated genera to build a prediction model resulted in higher classification success, based on Area Under the Curve (AUC), for both adenoma and carcinoma using fecal and tissue samples.
 The most important OTUs for these models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies.
-Overall, there were less associations between the microbiota and adenoma versus carcinoma and one reason why this may be is that most studies were only adequatly powered for large effect sizes.
+Overall, there were less associations between the microbiota and adenoma and one reason why this may be is that most studies were only adequately powered for large effect sizes.
 
-**Conclusions.** This data provides support for the importance of the bacteral community to both adenoma and carcinoma genesis.
+**Conclusions.** This data provides support for the importance of the bacterial community to both adenoma and carcinoma genesis.
 The evidence collected within this study on the role of the microbiota in CRC pathogenesis shows stronger associations between carcinoma then adenoma.
-A strong reason for this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
+One reason for this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
 
@@
 A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
-This meta-analysis, along with other studies, focused on identifying biomarkers or individual microbes but do not critically investigate how the community changes in CRC.
+This meta-analysis focused on identifying biomarkers or individual microbes but did not critically investigate how the community changes in CRC.
 
@@
 Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
-One advantage of tissue is that stool possibly represents an agregate community of the overall gastrointestinal (GI) tract and sampling the adenoma or carinoma directly could more accurately identify important microbes for CRC.
 Sampling the tissue directly may provide clearer answers but is not without limitations.
@@
 
-This intense focus on identifying biomarkers has found strong associations with resident mouth microbes as potential CRC-associated microbes in the GI tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014].
+This intense focus on identifying biomarkers has found strong associations with resident mouth microbes as potential CRC-associated microbes in the gastrointestinal (GI) tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014].
 The main bacteria of interest, arising from this set of microbes, has been those within the *Fusobacterium* genus.
 Yet, the question remains as to whether or not this is indeed the most important genera to be focusing on, since many microbiota-based studies typically have identified a collection of oral microbes rather than single species from a single genera [@baxter_microbiota-based_2016; @zeller_potential_2014].
-Based on this discrepency, the second question we want to answer with this meta-analysis is if there is one dominant CRC-associated genera that can be identified across studies.
+Based on this discrepancy, the second question we can answer with this meta-analysis is if there is one dominant CRC-associated genera that can be identified across studies.
 
@@
 
-In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, examines differences between stool and tissue microbiota in the context of CRC, and takes a more community centric approach rather than a biomarker focused approach to investigating commonalities across study for the microbiota and CRC severity.
+In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, examines differences between stool and tissue microbiota in the context of CRC, and takes a more community centric approach rather than a biomarker focused approach to investigating commonalities across study for the microbiota and CRC.
 Importantly, this community centric approach could provide valuable insights into the importance of accounting for the community in CRC disease not previously provided by earlier meta-analysis studies [@shah_leveraging_2017].
 
-Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
+Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
 To accomplish this we first assessed whether diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma.
-We then assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma.
+We then assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the RR of adenoma or carcinoma.
 Next, using Random Forest models, we analyzed whether the full community or only the CRC-associated genera resulted in better model classification based on the area under the curve (AUC).
@@
 By power transforming and Z-score normalizing the $\alpha$-diversity metrics for the entire data set we assessed whether there were any broad scale community differences that could be detected as disease severity worsened.
-Using linear mixed-effect models to control for study, resampling of the same individual (tissue studies only), and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness and Shannon diversity in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) but not for tissue (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1].
+Using linear mixed-effect models to control for study, re-sampling of the same individual (tissue studies only), and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness and Shannon diversity in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) and no $\alpha$-diversity correlations in tissue (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1].
 We next tested whether these detectable differences in community resulted in significant increases in RR.
-For fecal samples, a decrease in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))` and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
+For fecal samples, a decrease in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
 Interestingly, for both adenoma and carcinoma there was no increase in RR within tissue samples for any alpha diversity metric investigated [Table S1-S3].
 
-Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S4 & S5].
+Using the Bray-Curtis distance metric, there was a significant difference across study in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S4 & S5].
 For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with tissue samples from the same individual (matched) had no differences [Table S6 & S7].
@@
 This decrease may be explained by the fact that the tissue analysis included matched samples from the same individual.
-In fecal samples, the RR for carcinoma due to the prescence of CRC-associated genera was greater than either the RR associated with evenness or Shannon diversity [Figure 2 & 3].
+In fecal samples, the RR for carcinoma due to the presence of CRC-associated genera was greater than either the RR associated with evenness or Shannon diversity [Figure 2 & 3].
 Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC-associated genera [Figure 3B & 3D].
 
-There were two significant measures for increased RR of adenoma when investigating CRC-associated genera in stool: 1) Having a higher then median value of *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and 2) whether samples were positive for three CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A].
-With tissue, there were three significant measures for an increased RR of adenoma: 1) being positive for one CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), 2) being positive for two CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and 3) being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].
+There were two significant measures for increased RR of adenoma when investigating CRC-associated genera in stool: 1) Having a higher then median value of *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and 2) whether samples were positive for three CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A].
+With tissue, there were three significant measures for an increased RR of adenoma: 1) being positive for one CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), 2) being positive for two CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and 3) being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].
 
@@
 With respect to the test sets, comprised of genera data from other studies, both the all genera model and CRC-associated models had a similar ability to detect adenomas or carcinomas [Figure S4-S6].
-Two interesting but seperate general observations from these models were that: 1) classification of adenomas was lower than carcinomas for both tissue and stool and 2) AUC for the classification of carcinoma was consistently lower for the tissue models than the stool models [Figure S4-S6].
+Two interesting but separate general observations from these models were that: 1) classification of adenomas was lower than carcinomas for both tissue and stool and 2) AUC for the classification of carcinoma was consistently lower for the tissue models than the stool models [Figure S4-S6].
 
@@
 Conversely, none of these CRC-associated genera were present in the majority of studies.
-When we move onto the OTU based models, the adenoma OTU models had OTUs that classified as  *Ruminococcaceae* or *Roseburia* present in the top 10 OTUs for the vast majority of studies [Figure 6C].
+When we move on to the OTU based models, the adenoma OTU models had OTUs that classified as  *Ruminococcaceae* or *Roseburia* present in the top 10 OTUs for the vast majority of studies [Figure 6C].
 *Ruminococcaceae* was also present in the top 10 in some studies for the carcinoma OTU models, but it was *Bacteroides* that was present in the overwhelming majority of the carcinoma OTU stool models [Figure 6D].
@@
 Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease [Figure 1-3].
-Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the relative risk of either of these two.
+Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of either of these two.
 Even though CRC-associated genera increase the RR of carcinoma, they do not consistently increase the relative risk of adenoma.
 This information suggests that these specific genera are strongly associated with carcinoma but may not be associated with adenoma.
-Although CRC-associated genera are asscoiated with carcinomas, our data show that by using the whole community, classification models perform better than when only the CRC-associated genera are included.
+Although CRC-associated genera are associated with carcinomas, our data show that by using the whole community, classification models perform better than when only the CRC-associated genera are included.
 CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions.
-These observations suggest that small localized changes within the community on tissue may be occuring, that are important in early disease progression of CRC, and that this process may not directly involve CRC-associated genera.
 
@@
 The central idea of the model is that a single bacterium initiates an environment in which other non-resident microbes may then be able to colonize, creating a vicious cycle that is conducive for CRC.
-Both the drastically increased RR of CRC-associated genera versus $\alpha$-diversity metrics for carcinoma and increasing RR with more CRC-associated genera positivity are highly supportive of this model.
+Both the drastically increased carcinoma RR of CRC-associated genera versus $\alpha$-diversity metrics and increasing RR with more CRC-associated genera positivity, are highly supportive of this model.
 However, the initial establishment of the driver within the system appears to be dependent on the current community.
@@
 The changes that occur at this timepoint are small and possibly focal to the adenoma itself.
-The stepwise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1].
+The step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1].
 Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change, to a change in the global community.
-The poor performance of the Random Forest models for classifying adenoma based only on the microbiota would suggest that this is the case.
+The poor performance of the Random Forest models for classifying adenoma, based only on the microbiota, would suggest that this is the case.
 One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.
@@
 These observations would also suggest that initial changes in the resident community, specifically to *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma.
-Based on these observations, it is possible to hypothesize that the early changes in the community may give rise to initial polyp formation via interactions with the host and not necessarily via interactions with CRC-assciated genera.
+Based on these observations, it is possible to hypothesize that the early changes in the community may give rise to initial polyp formation via interactions with the host and not necessarily via interactions with CRC-associated genera.
 These changes then create new niches in which any one of the CRC-associated genera could gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma via a driver-passenger type mechanism.
@@
 Only data sets that had sequences available for analysis were included.
-Some studies did not have publically available sequences or did not have metadata in which the authors were able to share and were excluded.
+Some studies did not have publicly available sequences or did not have metadata in which the authors were able to share and were excluded.
 After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014].
-The Zackular [@zackular_human_2014] study was not included becasue the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016].
+The Zackular [@zackular_human_2014] study was not included because the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016].
 Additionally, after sequence processing all the case samples for the Kostic study only had 100 or less sequences remaining and was not used.
@@
 If it was not possible to use these defaults, the stated quality cut-offs were used instead.
-Chimeras were identifed and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].
+Chimeras were identified and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].
 
@@
 Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many with and without disease were above and below the overall median value within the specific study.
-Relative risk significance testing utilized the chi-squred test.
+Relative risk significance testing utilized the chi-squared test.
 $\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017].
@@
 Where applicable for each study, predictions for adenoma and carcinoma were then tested for feces, matched tissue, and unmatched tissue.
-This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold cross validation was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs.
-For OTU based models, the CRC associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
+This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold CV was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs.
+For OTU based models, the CRC-associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
 Finally, the power of each study was assessed for an effect size ranging from 1% to 30%.
-An estimated sample n for these effect sizes was also generated based on 80% power.
+An estimated sample size for these effect sizes was also generated based on 80% power.
 For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study.
-Similarily, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study.
+Similarly, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study.
 
@@
 ***Matched versus Unmatched Tissue Samples:***
-In general, tissue samples that had control and lesion samples that did not belong to the same indivdiual were classified as unmatched while samples that belonged to the same individual were classified as matched.
+In general, tissue samples that had control and lesion samples that did not belong to the same individual were classified as unmatched while samples that belonged to the same individual were classified as matched.
 Studies with matched data included Burns, Dejea, Geng, and Lu.
@@
 The OTU based models recorded the medians for each OTU across 100 different 80/20 splits of the data for each study.
-The lowest classification for each OTU was obtained using the RDP database and the number of times the specific classification occured in the top 10 across studies was recorded.
+The lowest classification for each OTU was obtained using the RDP database and the number of times the specific classification occurred in the top 10 across studies was recorded.
 For the adenoma tissue genus and OTU models there was only one matched and unmatched study and these results were grouped together for the counting of the top 10.
@@
 
-MAS is supported by a Candian Institute of Health Research fellowship and a University of Michigan Postdoctoral Translational Scholar Program grant.
+MAS is supported by a Canadian Institute of Health Research fellowship and a University of Michigan Postdoctoral Translational Scholar Program grant.
 

### 2017-12-12 submission/manuscript.tex
final edits of final read through after pat rd 2 suggestions

@@
 recent meta-analysis investigating microbial biomarkers for CRC by
-testing the hypothesis that the bacterial community is an important
-driver of both early (adenoma) and late (carcinoma) stage of disease. To
-test this hypothesis we examined both feces (n = 1737) and tissue (492
-total samples from 350 individuals) across 14 different studies.
+testing the hypothesis that the bacterial community has important
+associations to both early (adenoma) and late (carcinoma) stage disease.
+To test this hypothesis we examined both feces (n = 1737) and tissue
+(492 total samples from 350 individuals) across 14 different studies.
 
@@
 \emph{Porphyromonas} (P-value = 0.023). Using the whole community versus
-only CRC associated genera to build a prediction model resulted in
-higher classification success based on Area Under the Curve (AUC) for
+only CRC-associated genera to build a prediction model resulted in
+higher classification success, based on Area Under the Curve (AUC), for
 both adenoma and carcinoma using fecal and tissue samples. The most
@@
 studies. Overall, there were less associations between the microbiota
-and adenoma versus carcinoma and one reason why this may be is that most
-studies were only adequatly powered for large effect sizes.
+and adenoma and one reason why this may be is that most studies were
+only adequately powered for large effect sizes.
 
 \textbf{Conclusions.} This data provides support for the importance of
-the bacteral community to both adenoma and carcinoma genesis. The
+the bacterial community to both adenoma and carcinoma genesis. The
 evidence collected within this study on the role of the microbiota in
 CRC pathogenesis shows stronger associations between carcinoma then
-adenoma. A strong reason for this may be in part due to the low power to
+adenoma. One reason for this may be in part due to the low power to
 detect more subtle changes in the majority of studies that have been
@@
 investigated whether specific biomarkers could be consistently
-identified using multiple data sets {[}11{]}. This meta-analysis, along
-with other studies, focused on identifying biomarkers or individual
-microbes but do not critically investigate how the community changes in
-CRC.
+identified using multiple data sets {[}11{]}. This meta-analysis focused
+on identifying biomarkers or individual microbes but did not critically
+investigate how the community changes in CRC.
 
@@
 studies have tried to assess this in health and disease but are limited
-by their sample size {[}12,13{]}. One advantage of tissue is that stool
-possibly represents an agregate community of the overall
-gastrointestinal (GI) tract and sampling the adenoma or carinoma
-directly could more accurately identify important microbes for CRC.
-Sampling the tissue directly may provide clearer answers but is not
-without limitations. Due to bowel prep the communities left for sampling
-may not be reflective of the resident microbiota, but rather a
-collection of what is able to keep adhered to the mucosa. Additionally,
-these samples contain more host DNA, potentially limiting the types of
-analysis that can be done. It is well known that low biomass samples can
-be very difficult to work with and results can end up being study
-dependent due to the randomness of contamination {[}14{]}. Due to these
-many differences that could arise between stool and tissue, one question
-our meta-analysis aims to answer is whether there are consistent
-patterns that emerge across studies regardless of whether they used
-stool or tissue samples.
+by their sample size {[}12,13{]}. Sampling the tissue directly may
+provide clearer answers but is not without limitations. Due to bowel
+prep the communities left for sampling may not be reflective of the
+resident microbiota, but rather a collection of what is able to keep
+adhered to the mucosa. Additionally, these samples contain more host
+DNA, potentially limiting the types of analysis that can be done. It is
+well known that low biomass samples can be very difficult to work with
+and results can end up being study dependent due to the randomness of
+contamination {[}14{]}. Due to these many differences that could arise
+between stool and tissue, one question our meta-analysis aims to answer
+is whether there are consistent patterns that emerge across studies
+regardless of whether they used stool or tissue samples.
 
@@
 associations with resident mouth microbes as potential CRC-associated
-microbes in the GI tract {[}15--17{]}. The main bacteria of interest,
-arising from this set of microbes, has been those within the
-\emph{Fusobacterium} genus. Yet, the question remains as to whether or
-not this is indeed the most important genera to be focusing on, since
-many microbiota-based studies typically have identified a collection of
-oral microbes rather than single species from a single genera
-{[}16,17{]}. Based on this discrepency, the second question we want to
-answer with this meta-analysis is if there is one dominant
+microbes in the gastrointestinal (GI) tract {[}15--17{]}. The main
+bacteria of interest, arising from this set of microbes, has been those
+within the \emph{Fusobacterium} genus. Yet, the question remains as to
+whether or not this is indeed the most important genera to be focusing
+on, since many microbiota-based studies typically have identified a
+collection of oral microbes rather than single species from a single
+genera {[}16,17{]}. Based on this discrepancy, the second question we
+can answer with this meta-analysis is if there is one dominant
 CRC-associated genera that can be identified across studies.
@@
 more community centric approach rather than a biomarker focused approach
-to investigating commonalities across study for the microbiota and CRC
-severity. Importantly, this community centric approach could provide
-valuable insights into the importance of accounting for the community in
-CRC disease not previously provided by earlier meta-analysis studies
+to investigating commonalities across study for the microbiota and CRC.
+Importantly, this community centric approach could provide valuable
+insights into the importance of accounting for the community in CRC
+disease not previously provided by earlier meta-analysis studies
 {[}11{]}.
@@
 Using both feces (n = 1737) and tissue (492 samples from 350
-individuals) totalling over 2229 total samples across 14 studies
+individuals) totaling over 2229 total samples across 14 studies
 {[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
@@
 \emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
-affect the relative risk of adenoma or carcinoma. Next, using Random
-Forest models, we analyzed whether the full community or only the
-CRC-associated genera resulted in better model classification based on
-the area under the curve (AUC). Our results suggest that the community
-changes as disease severity worsens and that this community is important
-for disease classification. However, since the changes in community were
-subtle for adenoma we also examined what effect and sample size the
-studies that were used were adequately powered for. Although we analyzed
-data sets that sampled large numbers of individuals, our results
-indicate the individual studies were underpowered for detecting effect
-size differences of 10\% or below between the case and control groups.
+affect the RR of adenoma or carcinoma. Next, using Random Forest models,
+we analyzed whether the full community or only the CRC-associated genera
+resulted in better model classification based on the area under the
+curve (AUC). Our results suggest that the community changes as disease
+severity worsens and that this community is important for disease
+classification. However, since the changes in community were subtle for
+adenoma we also examined what effect and sample size the studies that
+were used were adequately powered for. Although we analyzed data sets
+that sampled large numbers of individuals, our results indicate the
+individual studies were underpowered for detecting effect size
+differences of 10\% or below between the case and control groups.
 
@@
 as disease severity worsened. Using linear mixed-effect models to
-control for study, resampling of the same individual (tissue studies
+control for study, re-sampling of the same individual (tissue studies
 only), and variable region, there was a significant decrease from
 control to adenoma to carcinoma for both evenness and Shannon diversity
-in stool (P-value = 0.025 and 0.043, respectively) but not for tissue
-(P-value \textgreater{} 0.05) {[}Figure 1{]}. We next tested whether
-these detectable differences in community resulted in significant
-increases in RR. For fecal samples, a decrease in evenness resulted in a
-significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
-P-value = 0.00034 and adenoma (RR = 1.16 (1.01 - 1.34), P-value = 0.032)
-while a decrease in Shannon diversity only increased the RR for
-carcinoma (RR = 1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}.
-Interestingly, for both adenoma and carcinoma there was no increase in
-RR within tissue samples for any alpha diversity metric investigated
-{[}Table S1-S3{]}.
+in stool (P-value = 0.025 and 0.043, respectively) and no
+\(\alpha\)-diversity correlations in tissue (P-value \textgreater{}
+0.05) {[}Figure 1{]}. We next tested whether these detectable
+differences in community resulted in significant increases in RR. For
+fecal samples, a decrease in evenness resulted in a significantly
+increased RR for carcinoma (RR = 1.36 (1.15 - 1.61), P-value = 0.00034)
+and adenoma (RR = 1.16 (1.01 - 1.34), P-value = 0.032) while a decrease
+in Shannon diversity only increased the RR for carcinoma (RR = 1.33
+(1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}. Interestingly, for both
+adenoma and carcinoma there was no increase in RR within tissue samples
+for any alpha diversity metric investigated {[}Table S1-S3{]}.
 
 Using the Bray-Curtis distance metric, there was a significant
-difference across studies in the bacterial community of fecal samples
+difference across study in the bacterial community of fecal samples
 between carcinoma and controls, but not adenoma and controls {[}Table S4
@@
 matched samples from the same individual. In fecal samples, the RR for
-carcinoma due to the prescence of CRC-associated genera was greater than
+carcinoma due to the presence of CRC-associated genera was greater than
 either the RR associated with evenness or Shannon diversity {[}Figure 2
@@
 median value of \emph{Porphyromonas} (P-value = 0.023) and 2) whether
-samples were positive for three CRC associated genera (P-value = 0.022)
+samples were positive for three CRC-associated genera (P-value = 0.022)
 {[}Figure 3A{]}. With tissue, there were three significant measures for
 an increased RR of adenoma: 1) being positive for one CRC-associated
-genera (P-value = 0.032), 2) being positive for two CRC associated
+genera (P-value = 0.032), 2) being positive for two CRC-associated
 genera (P-value = 0.008), and 3) being positive for four CRC associated
@@
 CRC-associated models had a similar ability to detect adenomas or
-carcinomas {[}Figure S4-S6{]}. Two interesting but seperate general
+carcinomas {[}Figure S4-S6{]}. Two interesting but separate general
 observations from these models were that: 1) classification of adenomas
@@
 {[}Figure 6A \& B{]}. Conversely, none of these CRC-associated genera
-were present in the majority of studies. When we move onto the OTU based
-models, the adenoma OTU models had OTUs that classified as
+were present in the majority of studies. When we move on to the OTU
+based models, the adenoma OTU models had OTUs that classified as
 \emph{Ruminococcaceae} or \emph{Roseburia} present in the top 10 OTUs
@@
 adenoma to carcinoma, this did not translate into large effect sizes for
-the relative risk of either of these two. Even though CRC-associated
-genera increase the RR of carcinoma, they do not consistently increase
-the relative risk of adenoma. This information suggests that these
-specific genera are strongly associated with carcinoma but may not be
-associated with adenoma. Although CRC-associated genera are asscoiated
-with carcinomas, our data show that by using the whole community,
+the RR of either of these two. Even though CRC-associated genera
+increase the RR of carcinoma, they do not consistently increase the
+relative risk of adenoma. This information suggests that these specific
+genera are strongly associated with carcinoma but may not be associated
+with adenoma. Although CRC-associated genera are associated with
+carcinomas, our data show that by using the whole community,
 classification models perform better than when only the CRC-associated
@@
 microbes exist can drastically increase the ability of models to make
-predictions. These observations suggest that small localized changes
-within the community on tissue may be occuring, that are important in
-early disease progression of CRC, and that this process may not directly
-involve CRC-associated genera.
+predictions.
 
@@
 non-resident microbes may then be able to colonize, creating a vicious
-cycle that is conducive for CRC. Both the drastically increased RR of
-CRC-associated genera versus \(\alpha\)-diversity metrics for carcinoma
-and increasing RR with more CRC-associated genera positivity are highly
-supportive of this model. However, the initial establishment of the
-driver within the system appears to be dependent on the current
+cycle that is conducive for CRC. Both the drastically increased
+carcinoma RR of CRC-associated genera versus \(\alpha\)-diversity
+metrics and increasing RR with more CRC-associated genera positivity,
+are highly supportive of this model. However, the initial establishment
+of the driver within the system appears to be dependent on the current
 community. This is supported by our finding that when adding the
@@
 model. The changes that occur at this timepoint are small and possibly
-focal to the adenoma itself. The stepwise decrease in diversity suggests
-that the adenoma community is not normal but has changed subtly
+focal to the adenoma itself. The step-wise decrease in diversity
+suggests that the adenoma community is not normal but has changed subtly
 {[}Figure 1{]}. Although there appears to be localized changes that do
@@
 the global community. The poor performance of the Random Forest models
-for classifying adenoma based only on the microbiota would suggest that
-this is the case. One potential hypothesis from these observations is
-that at early stages of the disease, how the host interacts with these
-subtle changes is what ultimately leads to a thoroughly dysfunctional
-community that is supportive of CRC genesis.
+for classifying adenoma, based only on the microbiota, would suggest
+that this is the case. One potential hypothesis from these observations
+is that at early stages of the disease, how the host interacts with
+these subtle changes is what ultimately leads to a thoroughly
+dysfunctional community that is supportive of CRC genesis.
 
@@
 formation via interactions with the host and not necessarily via
-interactions with CRC-assciated genera. These changes then create new
+interactions with CRC-associated genera. These changes then create new
 niches in which any one of the CRC-associated genera could gain a
@@
 included. Only data sets that had sequences available for analysis were
-included. Some studies did not have publically available sequences or
-did not have metadata in which the authors were able to share and were
+included. Some studies did not have publicly available sequences or did
+not have metadata in which the authors were able to share and were
 excluded. After these filtering steps, the following studies remained:
@@
 al.} {[}23{]}, and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]}
-study was not included becasue the 90 individuals analyzed within the
+study was not included because the 90 individuals analyzed within the
 study are contained within the larger Baxter study {[}16{]}.
@@
 based sequencing. If it was not possible to use these defaults, the
-stated quality cut-offs were used instead. Chimeras were identifed and
+stated quality cut-offs were used instead. Chimeras were identified and
 removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
@@
 specific study. Relative risk significance testing utilized the
-chi-squred test. \(\beta\)-diversity differences utilized a Bray-Curtis
+chi-squared test. \(\beta\)-diversity differences utilized a Bray-Curtis
 distance matrix and PERMANOVA executed with the vegan (v2.4.4) package
@@
 level with the exception that instead of testing on the other studies, a
-10-fold cross validation was utilized and 100 different models were
-created based on random 80/20 splitting of the data to generate a range
-of expected AUCs. For OTU based models, the CRC associated genera
-included all OTUs that had a taxonomic classification to
-\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, or
-\emph{Porphyromonas}. Finally, the power of each study was assessed for
-an effect size ranging from 1\% to 30\%. An estimated sample n for these
-effect sizes was also generated based on 80\% power. For comparisons in
-which normal versus adenoma were made the carcinoma samples were
-excluded from each respective study. Similarily, for comparisons in
-which normal versus carcinoma were made the adenoma samples were
-excluded from each respective study.
+10-fold CV was utilized and 100 different models were created based on
+random 80/20 splitting of the data to generate a range of expected AUCs.
+For OTU based models, the CRC-associated genera included all OTUs that
+had a taxonomic classification to \emph{Fusobacterium},
+\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas}.
+Finally, the power of each study was assessed for an effect size ranging
+from 1\% to 30\%. An estimated sample size for these effect sizes was
+also generated based on 80\% power. For comparisons in which normal
+versus adenoma were made the carcinoma samples were excluded from each
+respective study. Similarly, for comparisons in which normal versus
+carcinoma were made the adenoma samples were excluded from each
+respective study.
 
@@
 tissue samples that had control and lesion samples that did not belong
-to the same indivdiual were classified as unmatched while samples that
+to the same individual were classified as unmatched while samples that
 belonged to the same individual were classified as matched. Studies with
@@
 classification for each OTU was obtained using the RDP database and the
-number of times the specific classification occured in the top 10 across
-studies was recorded. For the adenoma tissue genus and OTU models there
-was only one matched and unmatched study and these results were grouped
-together for the counting of the top 10.
+number of times the specific classification occurred in the top 10
+across studies was recorded. For the adenoma tissue genus and OTU models
+there was only one matched and unmatched study and these results were
+grouped together for the counting of the top 10.
 
@@
 
-MAS is supported by a Candian Institute of Health Research fellowship
+MAS is supported by a Canadian Institute of Health Research fellowship
 and a University of Michigan Postdoctoral Translational Scholar Program

### 2018-01-04 submission/manuscript.Rmd
methods re-write

@@
 ***Obtaining Data Sets:***
-Studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016].
-Additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature.
-Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis were included.
-Only data sets that had sequences available for analysis were included.
-Some studies did not have publicly available sequences or did not have metadata in which the authors were able to share and were excluded.
+The studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016] and additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature.
+Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis were included and only data sets that had sequences available for analysis were included.
+Some studies were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share and were excluded.
 After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014].
 The Zackular [@zackular_human_2014] study was not included because the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016].
-Additionally, after sequence processing all the case samples for the Kostic study only had 100 or less sequences remaining and was not used.
-This left a total of 14 studies for which analysis could be completed.
+After sequence processing all the case samples for the Kostic study had 100 or less sequences remaining and was not used which left a total of 14 studies that analysis could be completed on.
 
@@
 For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained by searching the respective accession number of the study at the following website: http://www.ncbi.nlm.nih.gov/Traces/study/.
-Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap for one study [@ahn_human_2013] and for four studies was obtained directly from the authors [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013].
-Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009].
-Where possible, quality filtering utilized the default methods used in mothur for either 454 or Illumina based sequencing.
-If it was not possible to use these defaults, the stated quality cut-offs were used instead.
+Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap (n = 1, [@ahn_human_2013]) and directly from the authors (n = 4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]).
+Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009] and quality filtering utilized the default methods, used in mothur, for either 454 or Illumina based sequencing.
+If it was not possible to use the defaults, the stated quality cut-offs, from the study itself, were used instead.
 Chimeras were identified and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].
@@
 For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
-Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and variable region effects.
-Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many with and without disease were above and below the overall median value within the specific study.
+Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hypervariable region used.
+Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study.
 Relative risk significance testing utilized the chi-squared test.
@@
 ***Study Analysis Overview:***
-$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma.
-We analyzed the data using linear mixed-effect models and relative risk.
-$\beta$-diversity was then assessed for each individual study.
-Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in relative risk.
-We then built Random Forest models based on all genera or the select CRC-associated genera.
-The models were trained on one study then tested on the remaining studies for every study.
+$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma using both linear mixed-effect models and RR.
+$\beta$-diversity was then assessed for each individual study for differences between control-adenoma and control-carcinoma.
+Next, four specific mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in RR for adenoma and carcinoma.
+We then built Random Forest models based on all or mouth-associated genera and these models were trained on one study then tested on the remaining studies, for every study.
+A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
+For these OTU based models, the mouth-associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
+Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power.
+For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study.
+Similarly, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study.
 The data was split between feces and tissue samples.
@@
 Where applicable for each study, predictions for adenoma and carcinoma were then tested for feces, matched tissue, and unmatched tissue.
-This same approach was then applied at the OTU level with the exception that instead of testing on the other studies, a 10-fold CV was utilized and 100 different models were created based on random 80/20 splitting of the data to generate a range of expected AUCs.
-For OTU based models, the CRC-associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
-Finally, the power of each study was assessed for an effect size ranging from 1% to 30%.
-An estimated sample size for these effect sizes was also generated based on 80% power.
-For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study.
-Similarly, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study.
 
-***Obtaining CRC-Associated Genera:***
-For the CRC-associated genera analysis of the RR, the total average counts were collected for each respective OTU that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for 100 different subsamplings.
-The OTU based Random Forest Models that used CRC-associated genera used a similar approach except that the OTUs were not aggregated together by genus by kept as separate OTUs.
-So, OTU Random Forest models using the full community included all OTUs while those using CRC-associated genera included only those OTUs that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.
+***Obtaining Mouth-Associated Genera:***
+For the mouth-associated genera analysis of the RR, the total average counts, for 100 different subsamplings, were collected for each respective OTU that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.
+The OTU based Random Forest Models using mouth-associated genera utilized a similar approach except that the OTUs were not aggregated together by genus but kept as separate OTUs.
+OTU Random Forest models using the full community included all OTUs while those using mouth-associated genera included only those OTUs that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.
 
 ***Matched versus Unmatched Tissue Samples:***
-In general, tissue samples that had control and lesion samples that did not belong to the same individual were classified as unmatched while samples that belonged to the same individual were classified as matched.
-Studies with matched data included Burns, Dejea, Geng, and Lu.
-Studies that had unmatched data were Burns, Flemer, Chen, and Sanapareddy.
+In general, tissue samples that had control and lesion samples, that did not belong to the same individual, were classified as unmatched while samples, that belonged to the same individual, were classified as matched.
+Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy.
 For some studies samples became unmatched due to one of the corresponding matched samples not making it through sequence processing.
-For the linear mixed-effect models samples from the same individual were taken into account.
-For all other analysis matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.
+All samples, from both tissue sample types, were analyzed together for the linear mixed-effect models with samples from the same individual corrected for.
+For all other analysis, not mentioned herein, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.
 
 ***Assessing Important Random Forest Model Variables:***
-The genus based models collected the top 10 most important variables, measured by Mean Decrease in Accuracy (MDA), from each training set and assessed how many times that genera showed up in the top 10 across each study.
-The OTU based models recorded the medians for each OTU across 100 different 80/20 splits of the data for each study.
-The lowest classification for each OTU was obtained using the RDP database and the number of times the specific classification occurred in the top 10 across studies was recorded.
-For the adenoma tissue genus and OTU models there was only one matched and unmatched study and these results were grouped together for the counting of the top 10.
+Using Mean Decrease in Accuracy (MDA) the top 10 most important variables were obtained in two different ways depending on whether the model used genera or OTU data.
+For the genus based models, the number of times that a genus showed up in the top 10 of the training set across each study was counted while, for the OTU based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study.
+Common taxa, for the OTU based models, were identified by using the lowest classification within the RDP database for each of the specific OTUs obtained from the previous counts and the number of times this classification occurred in the top 10, in each study, was recorded.
+The two studies that had adenoma tissue were equally divided between matched and unmatched groups and were grouped together for the counting of the top 10 genera and OTUs.
 
 ***Reproducible Methods:***
-The code and analysis can be found here https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
+The code and analysis can be found here at  https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
 Unless otherwise mentioned, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.

### 2018-01-04 submission/manuscript.tex
methods re-write

@@
 
-\textbf{\emph{Obtaining Data Sets:}} Studies used for this meta-analysis
-were identified through the review articles written by Keku, \emph{et
-al.} and Vogtmann, \emph{et al.} {[}29,30{]}. Additional studies not
-mentioned in the reviews were obtained based on the authors' knowledge
-of the literature. Studies that used tissue or feces as their sample
-source for 454 or Illumina 16S rRNA gene sequencing analysis were
-included. Only data sets that had sequences available for analysis were
-included. Some studies did not have publicly available sequences or did
-not have metadata in which the authors were able to share and were
-excluded. After these filtering steps, the following studies remained:
-Ahn, \emph{et al.} {[}26{]}, Baxter, \emph{et al.} {[}16{]}, Brim,
-\emph{et al.} {[}22{]}, Burns, \emph{et al.} {[}27{]}, Chen, \emph{et
-al.} {[}19{]}, Dejea, \emph{et al.} {[}24{]}, Flemer, \emph{et al.}
-{[}12{]}, Geng, \emph{et al.} {[}28{]}, Hale, \emph{et al.} {[}18{]},
-Kostic, \emph{et al.} {[}15{]}, Lu, \emph{et al.} {[}21{]}, Sanapareddy,
-\emph{et al.} {[}25{]}, Wang, \emph{et al.} {[}20{]}, Weir, \emph{et
-al.} {[}23{]}, and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]}
-study was not included because the 90 individuals analyzed within the
-study are contained within the larger Baxter study {[}16{]}.
-Additionally, after sequence processing all the case samples for the
-Kostic study only had 100 or less sequences remaining and was not used.
-This left a total of 14 studies for which analysis could be completed.
+\textbf{\emph{Obtaining Data Sets:}} The studies used for this
+meta-analysis were identified through the review articles written by
+Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}29,30{]} and
+additional studies not mentioned in the reviews were obtained based on
+the authors' knowledge of the literature. Studies that used tissue or
+feces as their sample source for 454 or Illumina 16S rRNA gene
+sequencing analysis were included and only data sets that had sequences
+available for analysis were included. Some studies were excluded because
+they did not have publicly available sequences or did not have metadata
+in which the authors were able to share and were excluded. After these
+filtering steps, the following studies remained: Ahn, \emph{et al.}
+{[}26{]}, Baxter, \emph{et al.} {[}16{]}, Brim, \emph{et al.} {[}22{]},
+Burns, \emph{et al.} {[}27{]}, Chen, \emph{et al.} {[}19{]}, Dejea,
+\emph{et al.} {[}24{]}, Flemer, \emph{et al.} {[}12{]}, Geng, \emph{et
+al.} {[}28{]}, Hale, \emph{et al.} {[}18{]}, Kostic, \emph{et al.}
+{[}15{]}, Lu, \emph{et al.} {[}21{]}, Sanapareddy, \emph{et al.}
+{[}25{]}, Wang, \emph{et al.} {[}20{]}, Weir, \emph{et al.} {[}23{]},
+and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]} study was not
+included because the 90 individuals analyzed within the study are
+contained within the larger Baxter study {[}16{]}. After sequence
+processing all the case samples for the Kostic study had 100 or less
+sequences remaining and was not used which left a total of 14 studies
+that analysis could be completed on.
 
@@
 not have sequences and metadata on the SRA, data was obtained from DBGap
-for one study {[}26{]} and for four studies was obtained directly from
-the authors {[}12,18,23,25{]}. Each study was processed using the mothur
-(v1.39.3) software program {[}32{]}. Where possible, quality filtering
-utilized the default methods used in mothur for either 454 or Illumina
-based sequencing. If it was not possible to use these defaults, the
-stated quality cut-offs were used instead. Chimeras were identified and
+(n = 1, {[}26{]}) and directly from the authors (n = 4,
+{[}12,18,23,25{]}). Each study was processed using the mothur (v1.39.3)
+software program {[}32{]} and quality filtering utilized the default
+methods, used in mothur, for either 454 or Illumina based sequencing. If
+it was not possible to use the defaults, the stated quality cut-offs,
+from the study itself, were used instead. Chimeras were identified and
 removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
@@
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.2) software package {[}35{]}.
+sequence processing utilized the R (v3.4.3) software package {[}35{]}.
 For the \(\alpha\)-diversity analysis, values were power transformed
@@
 created using the lme4 (v1.1.14) package {[}38{]} to correct for study,
-repeat sampling of individuals (tissue only), and variable region
-effects. Relative risk was analyzed using both the epiR (v0.9.87) and
-metafor (v2.0.0) packages {[}39,40{]} by assessing how many with and
-without disease were above and below the overall median value within the
-specific study. Relative risk significance testing utilized the
-chi-squared test. \(\beta\)-diversity differences utilized a Bray-Curtis
-distance matrix and PERMANOVA executed with the vegan (v2.4.4) package
-{[}41{]}. Random Forest models were built using both the caret (v6.0.77)
-and randomForest (v4.6.12) packages {[}42,43{]}. Power analysis and
-estimations were made using the pwr (v1.2.1) and statmod (v1.4.30)
-packages {[}44,45{]}. All figures were created using both ggplot2
-(v2.2.1) and gridExtra (v2.3) packages {[}46,47{]}.
+repeat sampling of individuals (tissue only), and 16S hypervariable
+region used. Relative risk was analyzed using both the epiR (v0.9.87)
+and metafor (v2.0.0) packages {[}39,40{]} by assessing how many
+individuals with and without disease were above and below the overall
+median value within each specific study. Relative risk significance
+testing utilized the chi-squared test. \(\beta\)-diversity differences
+utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
+vegan (v2.4.4) package {[}41{]}. Random Forest models were built using
+both the caret (v6.0.77) and randomForest (v4.6.12) packages
+{[}42,43{]}. Power analysis and estimations were made using the pwr
+(v1.2.1) and statmod (v1.4.30) packages {[}44,45{]}. All figures were
+created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
+{[}46,47{]}.
 
 \textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
-assessed for differences between controls, adenoma, and carcinoma. We
-analyzed the data using linear mixed-effect models and relative risk.
-\(\beta\)-diversity was then assessed for each individual study. Next,
-four specific CRC-associated genera (\emph{Fusobacterium},
-\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas})
-were assessed for differences in relative risk. We then built Random
-Forest models based on all genera or the select CRC-associated genera.
-The models were trained on one study then tested on the remaining
-studies for every study. The data was split between feces and tissue
-samples. Within the tissue groups the data was further divided between
-samples from the same individual (matched) and those from different
-individuals (unmatched). Where applicable for each study, predictions
-for adenoma and carcinoma were then tested for feces, matched tissue,
-and unmatched tissue. This same approach was then applied at the OTU
-level with the exception that instead of testing on the other studies, a
-10-fold CV was utilized and 100 different models were created based on
-random 80/20 splitting of the data to generate a range of expected AUCs.
-For OTU based models, the CRC-associated genera included all OTUs that
-had a taxonomic classification to \emph{Fusobacterium},
+assessed for differences between controls, adenoma, and carcinoma using
+both linear mixed-effect models and RR. \(\beta\)-diversity was then
+assessed for each individual study for differences between
+control-adenoma and control-carcinoma. Next, four specific
+mouth-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
+differences in RR for adenoma and carcinoma. We then built Random Forest
+models based on all or mouth-associated genera and these models were
+trained on one study then tested on the remaining studies, for every
+study. A similar approach was then applied at the OTU level with the
+exception that a 10-fold CV over 100 different models, based on random
+80/20 splitting of the data, was used to generate a range of expected
+AUCs. For these OTU based models, the mouth-associated genera included
+all OTUs that had a taxonomic classification to \emph{Fusobacterium},
 \emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas}.
 Finally, the power of each study was assessed for an effect size ranging
-from 1\% to 30\%. An estimated sample size for these effect sizes was
-also generated based on 80\% power. For comparisons in which normal
+from 1\% to 30\% and an estimated sample size, for these effect sizes,
+was generated based on 80\% power. For comparisons in which normal
 versus adenoma were made the carcinoma samples were excluded from each
@@
 carcinoma were made the adenoma samples were excluded from each
-respective study.
-
-\textbf{\emph{Obtaining CRC-Associated Genera:}} For the CRC-associated
-genera analysis of the RR, the total average counts were collected for
-each respective OTU that had a genus level taxonomic classification to
-\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus}, and
-\emph{Porphyromonas} for 100 different subsamplings. The OTU based
-Random Forest Models that used CRC-associated genera used a similar
-approach except that the OTUs were not aggregated together by genus by
-kept as separate OTUs. So, OTU Random Forest models using the full
-community included all OTUs while those using CRC-associated genera
-included only those OTUs that had a genus level taxonomic classification
-to \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus},
-and \emph{Porphyromonas}.
+respective study. The data was split between feces and tissue samples.
+Within the tissue groups the data was further divided between samples
+from the same individual (matched) and those from different individuals
+(unmatched). Where applicable for each study, predictions for adenoma
+and carcinoma were then tested for feces, matched tissue, and unmatched
+tissue.
+
+\textbf{\emph{Obtaining Mouth-Associated Genera:}} For the
+mouth-associated genera analysis of the RR, the total average counts,
+for 100 different subsamplings, were collected for each respective OTU
+that had a genus level taxonomic classification to \emph{Fusobacterium},
+\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}.
+The OTU based Random Forest Models using mouth-associated genera
+utilized a similar approach except that the OTUs were not aggregated
+together by genus but kept as separate OTUs. OTU Random Forest models
+using the full community included all OTUs while those using
+mouth-associated genera included only those OTUs that had a genus level
+taxonomic classification to \emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Peptostreptococcus}, and \emph{Porphyromonas}.
 
 \textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
-tissue samples that had control and lesion samples that did not belong
-to the same individual were classified as unmatched while samples that
-belonged to the same individual were classified as matched. Studies with
-matched data included Burns, Dejea, Geng, and Lu. Studies that had
-unmatched data were Burns, Flemer, Chen, and Sanapareddy. For some
+tissue samples that had control and lesion samples, that did not belong
+to the same individual, were classified as unmatched while samples, that
+belonged to the same individual, were classified as matched. Studies
+with matched data included Burns, Dejea, Geng, and Lu while those with
+unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some
 studies samples became unmatched due to one of the corresponding matched
-samples not making it through sequence processing. For the linear
-mixed-effect models samples from the same individual were taken into
-account. For all other analysis matched and unmatched samples were
-analyzed separately using the statistical approaches mentioned in the
-Statistical Analysis section.
-
-\textbf{\emph{Assessing Important Random Forest Model Variables:}} The
-genus based models collected the top 10 most important variables,
-measured by Mean Decrease in Accuracy (MDA), from each training set and
-assessed how many times that genera showed up in the top 10 across each
-study. The OTU based models recorded the medians for each OTU across 100
-different 80/20 splits of the data for each study. The lowest
-classification for each OTU was obtained using the RDP database and the
-number of times the specific classification occurred in the top 10
-across studies was recorded. For the adenoma tissue genus and OTU models
-there was only one matched and unmatched study and these results were
-grouped together for the counting of the top 10.
+samples not making it through sequence processing. All samples, from
+both tissue sample types, were analyzed together for the linear
+mixed-effect models with samples from the same individual corrected for.
+For all other analysis, not mentioned herein, matched and unmatched
+samples were analyzed separately using the statistical approaches
+mentioned in the Statistical Analysis section.
+
+\textbf{\emph{Assessing Important Random Forest Model Variables:}} Using
+Mean Decrease in Accuracy (MDA) the top 10 most important variables were
+obtained in two different ways depending on whether the model used
+genera or OTU data. For the genus based models, the number of times that
+a genus showed up in the top 10 of the training set across each study
+was counted while, for the OTU based models, the medians for each OTU
+across 100 different 80/20 splits of the data was generated and the top
+10 OTUs then counted for each study. Common taxa, for the OTU based
+models, were identified by using the lowest classification within the
+RDP database for each of the specific OTUs obtained from the previous
+counts and the number of times this classification occurred in the top
+10, in each study, was recorded. The two studies that had adenoma tissue
+were equally divided between matched and unmatched groups and were
+grouped together for the counting of the top 10 genera and OTUs.
 
 \textbf{\emph{Reproducible Methods:}} The code and analysis can be found
-here
+here at
 \url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.

### 2018-01-04 submission/manuscript.Rmd
conclusion re-write

@@
 
-By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in both adenoma and carcinoma.
-Overall, our results support a framework by which early localized community changes give rise to polyp formation.
-With the host as a potential catalyst, new niches arise by which non-resident CRC-associated microbes can then gain a foothold and create an environment that allows more of these microbes to colonize.
-This exacerbates the existing community changes and creates a vicious cycle conducive of carcinoma formation.
+By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in CRC.
+Further, the data presented here suggests that non-resident mouth-associated microbes can gain a foothold within the colon and the more of these microbes that are present the greater the RR of carcinoma.
+Conversely, no clear signal with these mouth-associated micobes could be detected for adenoma.
 Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community.
-Although there are power limitations associated with adenoma, this report highlights the strong associations the microbiota has on CRC.
+Overall, the microbiota associations with carcinoma are much stronger than the associations with adenoma.
 

### 2018-01-04 submission/manuscript.tex
conclusion re-write

@@
 and tissue, we are able to provide evidence in support of the importance
-of the bacterial community in both adenoma and carcinoma. Overall, our
-results support a framework by which early localized community changes
-give rise to polyp formation. With the host as a potential catalyst, new
-niches arise by which non-resident CRC-associated microbes can then gain
-a foothold and create an environment that allows more of these microbes
-to colonize. This exacerbates the existing community changes and creates
-a vicious cycle conducive of carcinoma formation. Our observations also
-highlight the importance of power and sample number considerations when
-undertaking investigations into the microbiota and adenoma due to the
-subtle changes in the community. Although there are power limitations
-associated with adenoma, this report highlights the strong associations
-the microbiota has on CRC.
+of the bacterial community in CRC. Further, the data presented here
+suggests that non-resident mouth-associated microbes can gain a foothold
+within the colon and the more of these microbes that are present the
+greater the RR of carcinoma. Conversely, no clear signal with these
+mouth-associated micobes could be detected for adenoma. Our observations
+also highlight the importance of power and sample number considerations
+when undertaking investigations into the microbiota and adenoma due to
+the subtle changes in the community. Overall, the microbiota
+associations with carcinoma are much stronger than the associations with
+adenoma.
 

### 2018-01-04 submission/manuscript.Rmd
re-write of discussion

@@
 
-Our study identifies clear differences in diversity both at the community level and within individual genera that are present in individuals with CRC versus those without the disease [Figure 1-3].
-Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of either of these two.
-Even though CRC-associated genera increase the RR of carcinoma, they do not consistently increase the relative risk of adenoma.
-This information suggests that these specific genera are strongly associated with carcinoma but may not be associated with adenoma.
-Although CRC-associated genera are associated with carcinomas, our data show that by using the whole community, classification models perform better than when only the CRC-associated genera are included.
-CRC-associated genera are clearly important to carcinoma pathogenesis but accounting for the community in which these microbes exist can drastically increase the ability of models to make predictions.
+Our study identifies clear differences in diversity, both at the community level and for individual genera, that are present in patients with and without CRC [Figure 1-3].
+Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of lesion.
+Even though mouth-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma.
+These mouth-associated genera are clearly important to carcinoma classification but ourobservations suggest that accounting for the community in which these microbes exist can increase the ability of models to make predictions.
 
-The data presented herein supports the driver-passenger model of the microbial role in CRC, as summarized by Flynn [@flynn_metabolic_2016], when applied to carcinoma but not necessarily adenoma.
-The central idea of the model is that a single bacterium initiates an environment in which other non-resident microbes may then be able to colonize, creating a vicious cycle that is conducive for CRC.
-Both the drastically increased carcinoma RR of CRC-associated genera versus $\alpha$-diversity metrics and increasing RR with more CRC-associated genera positivity, are highly supportive of this model.
-However, the initial establishment of the driver within the system appears to be dependent on the current community.
-This is supported by our finding that when adding the community context to our models in addition to the CRC-associated genera, the model AUC increases [Figure 4 & 5].
-Conversely, using the present data, it is less likely that adenoma development fits this model.
-The changes that occur at this timepoint are small and possibly focal to the adenoma itself.
+The data presented herein supports the importance of mouth-associated genera for carcinoma but not necessarily adenoma formation.
+A proposed mechanism by which these mouth-associated microbes cause CRC stems from the creation of a vicious cycle involving host inflammation and repair.
+Both the drastically increased carcinoma RR of mouth-associated genera versus $\alpha$-diversity metrics and increasing RR with more mouth-associated genera positivity [Figure 3C], are highly supportive of their importance.
+However, our observations show that when adding the community context to our Random Forest models along with the mouth-associated genera, the model AUC increases [Figure 4 & 5].
+This suggests the resident community cannot be ignored completely and needs to be accounted for when investigating mouth-associated microbes and carcinoma.
+Conversely, using the present data, it is less likely that mouth-associated microbes are correlated with adenoma development.
 The step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1].
-Although there appears to be localized changes that do depend on the driver-passenger model, as supported by an increased RR for one, two, and four positive CRC-associated genera in tissue [Figure 3C], there may be other processes at play that ultimately exacerbate the condition from a subtle localized change, to a change in the global community.
-The poor performance of the Random Forest models for classifying adenoma, based only on the microbiota, would suggest that this is the case.
+Yet, this change in diversity, at this early stage of disease, could be focal to the adenoma itself.
+The poor performance of the Random Forest models for classifying adenoma, based only on the stool microbiota, would suggest that this is the case.
 One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.
@@
 Within stool, common resident microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6].
-Changes in *Bacteroides*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma models [Figure 6].
+Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma [Figure 6].
 This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community.
-These observations would also suggest that initial changes in the resident community, specifically to *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia*, carry on from adenoma to carcinoma.
-Based on these observations, it is possible to hypothesize that the early changes in the community may give rise to initial polyp formation via interactions with the host and not necessarily via interactions with CRC-associated genera.
-These changes then create new niches in which any one of the CRC-associated genera could gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma via a driver-passenger type mechanism.
+Based on these observations, it is possible to hypothesize that tsmall changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma.
 
-The tissue studies did not provide a clearer understanding of how the microbiota may be associated with disease severity.
-For the OTU models both the unmatched and matched [Figure S7E & F] tissue samples had some concordance with the stool data, with resident bacteria being the most prevalent in the top 10 important variables across studies.
-Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F].
-The majority of the results seem to be study specific with many top 10 taxa being present only in a single study.
-One other potentially worrying sign is the presence of *Propionibacterium* within the top 10 for the genera and OTU models and could be a marker of contamination.
-The low biomass of these samples coupled with potential contamination might be a possible reason why the tissue results seem to be more sporadic then the stool results.
+The tissue studies did not provide a clearer understanding of how the microbiota may be associated with lesion.
+Generally, the OTU models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study.
+Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F].
+The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study.
+Additionally, the presence of *Propionibacterium*, within the top 10 most important variables for the genera and OTU models, is worrying because it is commonly a marker of contamination.
+The low bacterial biomass of tissue samples coupled with potential contaminationcould ultimately explain why the these results seem to be more sporadic than the stool results.
 
-The associations between the microbiota and adenoma are less clear in part because many studies may not be powered effectively to observe the small changes reported here.
-None of the studies analyzed were properly powered to detect a 10% or lower change between case and controls.
-This small effect size range may well be the scope in which differences consistently occur in adenoma due to the subtle changes in community that occur between control and adenoma.
-Future studies investigating adenoma and the microbiota need to take these factors into consideration if we are to work out how the microbiota contributes to adenoma formation.
-In contrast to adenoma, our observations suggest that many changes in carcinoma could easily result in effect sizes that are 30% or more between the case and control and most studies analyzed have sufficient power to detect these types of changes [Figure 7].
+The associations between the microbiota and adenoma are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes.
+None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls.
+A small effect size may well be the scope in which differences consistently occur in adenoma based on the results observed within our meta-analysis.
+Future studies investigating adenoma and the microbiota need to take power into consideration if we are to reproducibly study whether the microbiota contributes to poly formation.
+In contrast to adenoma, our observations suggest that most studies analyzed have sufficient power to detect many changes in carcinoma because of large effect size differences between cases and controls [Figure 7].
 

### 2018-01-04 submission/manuscript.tex
re-write of discussion

@@
 
-Our study identifies clear differences in diversity both at the
-community level and within individual genera that are present in
-individuals with CRC versus those without the disease {[}Figure 1-3{]}.
-Although there was a step-wise decrease in diversity from control to
-adenoma to carcinoma, this did not translate into large effect sizes for
-the RR of either of these two. Even though CRC-associated genera
-increase the RR of carcinoma, they do not consistently increase the
-relative risk of adenoma. This information suggests that these specific
-genera are strongly associated with carcinoma but may not be associated
-with adenoma. Although CRC-associated genera are associated with
-carcinomas, our data show that by using the whole community,
-classification models perform better than when only the CRC-associated
-genera are included. CRC-associated genera are clearly important to
-carcinoma pathogenesis but accounting for the community in which these
-microbes exist can drastically increase the ability of models to make
-predictions.
-
-The data presented herein supports the driver-passenger model of the
-microbial role in CRC, as summarized by Flynn {[}2{]}, when applied to
-carcinoma but not necessarily adenoma. The central idea of the model is
-that a single bacterium initiates an environment in which other
-non-resident microbes may then be able to colonize, creating a vicious
-cycle that is conducive for CRC. Both the drastically increased
-carcinoma RR of CRC-associated genera versus \(\alpha\)-diversity
-metrics and increasing RR with more CRC-associated genera positivity,
-are highly supportive of this model. However, the initial establishment
-of the driver within the system appears to be dependent on the current
-community. This is supported by our finding that when adding the
-community context to our models in addition to the CRC-associated
-genera, the model AUC increases {[}Figure 4 \& 5{]}. Conversely, using
-the present data, it is less likely that adenoma development fits this
-model. The changes that occur at this timepoint are small and possibly
-focal to the adenoma itself. The step-wise decrease in diversity
-suggests that the adenoma community is not normal but has changed subtly
-{[}Figure 1{]}. Although there appears to be localized changes that do
-depend on the driver-passenger model, as supported by an increased RR
-for one, two, and four positive CRC-associated genera in tissue
-{[}Figure 3C{]}, there may be other processes at play that ultimately
-exacerbate the condition from a subtle localized change, to a change in
-the global community. The poor performance of the Random Forest models
-for classifying adenoma, based only on the microbiota, would suggest
-that this is the case. One potential hypothesis from these observations
-is that at early stages of the disease, how the host interacts with
-these subtle changes is what ultimately leads to a thoroughly
-dysfunctional community that is supportive of CRC genesis.
+Our study identifies clear differences in diversity, both at the
+community level and for individual genera, that are present in patients
+with and without CRC {[}Figure 1-3{]}. Although there was a step-wise
+decrease in diversity from control to adenoma to carcinoma, this did not
+translate into large effect sizes for the RR of lesion. Even though
+mouth-associated genera increased the RR of carcinoma, they did not
+consistently increase the RR of adenoma. These mouth-associated genera
+are clearly important to carcinoma classification but ourobservations
+suggest that accounting for the community in which these microbes exist
+can increase the ability of models to make predictions.
+
+The data presented herein supports the importance of mouth-associated
+genera for carcinoma but not necessarily adenoma formation. A proposed
+mechanism by which these mouth-associated microbes cause CRC stems from
+the creation of a vicious cycle involving host inflammation and repair.
+Both the drastically increased carcinoma RR of mouth-associated genera
+versus \(\alpha\)-diversity metrics and increasing RR with more
+mouth-associated genera positivity {[}Figure 3C{]}, are highly
+supportive of their importance. However, our observations show that when
+adding the community context to our Random Forest models along with the
+mouth-associated genera, the model AUC increases {[}Figure 4 \& 5{]}.
+This suggests the resident community cannot be ignored completely and
+needs to be accounted for when investigating mouth-associated microbes
+and carcinoma. Conversely, using the present data, it is less likely
+that mouth-associated microbes are correlated with adenoma development.
+The step-wise decrease in diversity suggests that the adenoma community
+is not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
+diversity, at this early stage of disease, could be focal to the adenoma
+itself. The poor performance of the Random Forest models for classifying
+adenoma, based only on the stool microbiota, would suggest that this is
+the case. One potential hypothesis from these observations is that at
+early stages of the disease, how the host interacts with these subtle
+changes is what ultimately leads to a thoroughly dysfunctional community
+that is supportive of CRC genesis.
 
@@
 the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
-\emph{Bacteroides}, \emph{Ruminococcus}, and \emph{Roseburia} were
-consistently found to be disciminative across the different studies for
-both adenoma and carcinoma models {[}Figure 6{]}. This data would
-suggest that whether the non-resident bacterium is \emph{Fusobacteria}
-or \emph{Peptostreptococcus} is not as important as how these bacteria
-interact with the changing resident community. These observations would
-also suggest that initial changes in the resident community,
-specifically to \emph{Bacteroides}, \emph{Ruminococcaceae},
-\emph{Ruminococcus}, and \emph{Roseburia}, carry on from adenoma to
-carcinoma. Based on these observations, it is possible to hypothesize
-that the early changes in the community may give rise to initial polyp
-formation via interactions with the host and not necessarily via
-interactions with CRC-associated genera. These changes then create new
-niches in which any one of the CRC-associated genera could gain a
-foothold, exacerbating the initial changes in community and facilitating
-the transition from adenoma to carcinoma via a driver-passenger type
-mechanism.
+\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
+\emph{Roseburia} were consistently found to be disciminative across the
+different studies for both adenoma and carcinoma {[}Figure 6{]}. This
+data would suggest that whether the non-resident bacterium is
+\emph{Fusobacteria} or \emph{Peptostreptococcus} is not as important as
+how these bacteria interact with the changing resident community. Based
+on these observations, it is possible to hypothesize that tsmall changes
+in community structure lead to new niches in which any one of the
+mouth-associated genera can gain a foothold, exacerbating the initial
+changes in community and facilitating the transition from adenoma to
+carcinoma.
 
 The tissue studies did not provide a clearer understanding of how the
-microbiota may be associated with disease severity. For the OTU models
-both the unmatched and matched {[}Figure S7E \& F{]} tissue samples had
-some concordance with the stool data, with resident bacteria being the
-most prevalent in the top 10 important variables across studies. Unlike
-in stool, \emph{Fusobacterium} was the only CRC-associated bacteria
-consistently present in the top 10 of the CRC models {[}Figure S7B-C \&
-E-F{]}. The majority of the results seem to be study specific with many
-top 10 taxa being present only in a single study. One other potentially
-worrying sign is the presence of \emph{Propionibacterium} within the top
-10 for the genera and OTU models and could be a marker of contamination.
-The low biomass of these samples coupled with potential contamination
-might be a possible reason why the tissue results seem to be more
-sporadic then the stool results.
-
-The associations between the microbiota and adenoma are less clear in
-part because many studies may not be powered effectively to observe the
-small changes reported here. None of the studies analyzed were properly
-powered to detect a 10\% or lower change between case and controls. This
-small effect size range may well be the scope in which differences
-consistently occur in adenoma due to the subtle changes in community
-that occur between control and adenoma. Future studies investigating
-adenoma and the microbiota need to take these factors into consideration
-if we are to work out how the microbiota contributes to adenoma
-formation. In contrast to adenoma, our observations suggest that many
-changes in carcinoma could easily result in effect sizes that are 30\%
-or more between the case and control and most studies analyzed have
-sufficient power to detect these types of changes {[}Figure 7{]}.
+microbiota may be associated with lesion. Generally, the OTU models of
+unmatched and matched {[}Figure S7E \& F{]} tissue samples were
+concordant with stool showing that resident microbes were the most
+prevalent in the top 10 across study. Unlike in stool,
+\emph{Fusobacterium} was the only mouth-associated bacteria consistently
+present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
+majority of the tissue results seem to be study specific with many top
+10 taxa being present only in a single study. Additionally, the presence
+of \emph{Propionibacterium}, within the top 10 most important variables
+for the genera and OTU models, is worrying because it is commonly a
+marker of contamination. The low bacterial biomass of tissue samples
+coupled with potential contaminationcould ultimately explain why the
+these results seem to be more sporadic than the stool results.
+
+The associations between the microbiota and adenoma are inconclusive, in
+part, because many studies may not be powered effectively to observe
+small effect sizes. None of the studies analyzed were properly powered
+to detect a 10\% or lower change between cases and controls. A small
+effect size may well be the scope in which differences consistently
+occur in adenoma based on the results observed within our meta-analysis.
+Future studies investigating adenoma and the microbiota need to take
+power into consideration if we are to reproducibly study whether the
+microbiota contributes to poly formation. In contrast to adenoma, our
+observations suggest that most studies analyzed have sufficient power to
+detect many changes in carcinoma because of large effect size
+differences between cases and controls {[}Figure 7{]}.
 

### 2018-01-04 submission/manuscript.Rmd
re-write last two paragraphs of results

@@
 
-Conversely to the stool models, both genera and OTU based models in tissue had the vast majority of their top 10 occur in a study specific manner [Figure S7].
-*Fusobacterium* and *Fusobacteriaceae* show up more often in the top 10 for matched tissue samples but were not present in the top 10 or were much lower ranked for unmatched tissue [Figure S7B-C & S7E-F].
-There appears to be very little overlap in the top 10 most important variables between stool and tissue for both adenoma and carcinoma [Figure 6 & S7].
+Unlike stool, tissue based Random Forest models showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 across study [Figure S7].
+The vast majority of the top 10 genera and OTUs occured in a study specific manner for these tissue based Random Forest models.
+For both the genera and OTU Random Forest models of adenoma and carcinoma, there appears to be very little overlap in the top 10 most important variables between stool and tissue [Figure 6 & S7].
+This discordance between stool and tissue also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of Random Forest models using matched versus unmatched tissue samples [Figure S7B-C & S7E-F].
 
-***A Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
-Based on the previous observations we then assessed whether the studies that we included are realistically powered to identify small, medium, and large scale differences between case and control.
-When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 7A & B].
-No single study that was analyzed had the standard 80% power to detect an effect size difference that was equal to or below 10% [Figure 7A & B].
-In order to achieve adequate power for small effect sizes, studies would need to recruit over 1000 individuals for each arm [Figure 7C].
+***CRC Studies are Underpowered for Detecting Small Effect Sizes:***
+Based on our reported results we assessed whether the studies analyzed were realistically powered to identify small, medium, and large scale differences between cases and controls.
+When assessing the power of each study at different effect sizes the majority of studies, for both adenoma and carcinoma, acheived 80% power to detect a 30% or greater difference between groups [Figure 7A & B].
+No study that was analyzed had the standard 80% power to detect an effect size difference that was equal to or below 10% [Figure 7A & B].
+In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for each arm [Figure 7C].
 

### 2018-01-04 submission/manuscript.tex
re-write last two paragraphs of results

@@
 
-Conversely to the stool models, both genera and OTU based models in
-tissue had the vast majority of their top 10 occur in a study specific
-manner {[}Figure S7{]}. \emph{Fusobacterium} and \emph{Fusobacteriaceae}
-show up more often in the top 10 for matched tissue samples but were not
-present in the top 10 or were much lower ranked for unmatched tissue
-{[}Figure S7B-C \& S7E-F{]}. There appears to be very little overlap in
-the top 10 most important variables between stool and tissue for both
-adenoma and carcinoma {[}Figure 6 \& S7{]}.
-
-\textbf{\emph{A Majority of Studies are Underpowered for Detecting Small
-Effect Size Differences:}} Based on the previous observations we then
-assessed whether the studies that we included are realistically powered
-to identify small, medium, and large scale differences between case and
-control. When assessing the power of each study at different effect
-sizes the majority of studies for both adenoma and carcinoma have an
-80\% power to detect a 30\% difference {[}Figure 7A \& B{]}. No single
-study that was analyzed had the standard 80\% power to detect an effect
-size difference that was equal to or below 10\% {[}Figure 7A \& B{]}. In
-order to achieve adequate power for small effect sizes, studies would
-need to recruit over 1000 individuals for each arm {[}Figure 7C{]}.
+Unlike stool, tissue based Random Forest models showed no consistent
+representation of \emph{Ruminococcaceae}, \emph{Ruminococcus},
+\emph{Bacteroides}, and \emph{Roseburia} in the top 10 across study
+{[}Figure S7{]}. The vast majority of the top 10 genera and OTUs occured
+in a study specific manner for these tissue based Random Forest models.
+For both the genera and OTU Random Forest models of adenoma and
+carcinoma, there appears to be very little overlap in the top 10 most
+important variables between stool and tissue {[}Figure 6 \& S7{]}. This
+discordance between stool and tissue also applies to the
+mouth-associated genera with one noticeable skew being that
+\emph{Fusobacterium} and \emph{Fusobacteriaceae} occur more often in the
+top 10 of Random Forest models using matched versus unmatched tissue
+samples {[}Figure S7B-C \& S7E-F{]}.
+
+\textbf{\emph{CRC Studies are Underpowered for Detecting Small Effect
+Sizes:}} Based on our reported results we assessed whether the studies
+analyzed were realistically powered to identify small, medium, and large
+scale differences between cases and controls. When assessing the power
+of each study at different effect sizes the majority of studies, for
+both adenoma and carcinoma, acheived 80\% power to detect a 30\% or
+greater difference between groups {[}Figure 7A \& B{]}. No study that
+was analyzed had the standard 80\% power to detect an effect size
+difference that was equal to or below 10\% {[}Figure 7A \& B{]}. In
+order to achieve a power of 80\%, for small effect sizes, studies used
+in our meta-analysis would need to recruit over 1000 individuals for
+each arm {[}Figure 7C{]}.
 

### 2018-01-05 submission/manuscript.Rmd
results re-write upto diversity before CRC-associated re-analysis

@@
 Based on the small increase in RR using $\alpha$-diversity metrics, we assessed if the presence of specific genera resulted in a higher RR for both stool and tissue.
-To investigate this we analyzed the classically associated CRC genera, *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas* for an increase in RR.
+To investigate this we analyzed the classically associated CRC genera, *Bacteroides*, *Escherichia*, *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, *Porphyromonas*, and *Streptococcus* for an increase in RR.
 The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3].
@@
 This decrease may be explained by the fact that the tissue analysis included matched samples from the same individual.
-In fecal samples, the RR for carcinoma due to the presence of CRC-associated genera was greater than either the RR associated with evenness or Shannon diversity [Figure 2 & 3].
+In fecal samples, the RR for carcinoma when having CRC-associated genera at a higher than overal study median relative abundance was greater than the RR associated with evenness or Shannon diversity [Figure 2 & 3].
 Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC-associated genera [Figure 3B & 3D].
 
-There were two significant measures for increased RR of adenoma when investigating CRC-associated genera in stool: 1) Having a higher then median value of *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and 2) whether samples were positive for three CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A].
-With tissue, there were three significant measures for an increased RR of adenoma: 1) being positive for one CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), 2) being positive for two CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and 3) being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].
+In stool the significant measures that increased RR of adenoma for CRC-associated genera were having a higher than overall study median relative abundance of *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and being positive for three CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A].
+Within tissue the significant measures for an increased RR of adenoma were being positive for one CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), being positive for two CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].
 
-***Whole Community Models Add Important Community Context:***
-Since CRC-associated genera had increased RR for carcinoma over diversity metrics we wondered whether the overall bacterial community was at all important to classifying disease or if the CRC-associated genera were sufficient alone.
-To test this we used two approaches.
-The first approach used genus level data and tested whether there were any differences in AUC when training on one study and testing on all the others when using either all genera present or only the CRC-associated genera.
-The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model across studies using either all OTUs or only OTUs that taxonomically classified to CRC-associated genera.
+***Whole Community Models Add Important Context:***
+Since CRC-associated genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only CRC-associated genera and tested this using two approaches.
+The first approach used genus level data and tested whether there were any differences in AUC between all genera and CRC-associated genera only during training on one study or testing on all other studies.
+The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model AUC across studies when comparing all OTUs versus only OTUs that taxonomically classified to CRC-associated genera.
 
-Our first approach using genus based models showed an AUC decrease in model classification on the training set for both stool and tissue studies [Figure S2-S3].
-With respect to the test sets, comprised of genera data from other studies, both the all genera model and CRC-associated models had a similar ability to detect adenomas or carcinomas [Figure S4-S6].
-Two interesting but separate general observations from these models were that: 1) classification of adenomas was lower than carcinomas for both tissue and stool and 2) AUC for the classification of carcinoma was consistently lower for the tissue models than the stool models [Figure S4-S6].
+For the genera-based models both tissue and stool studies had a median AUC decrease in model classification when only CRC-associated genera versus all genera were included [Figure S2-S3].
+With respect to the test sets, comprised of genera data from other studies, both all and CRC-associated genera models had a similar ability to detect adenomas or carcinomas [Figure S4-S6].
+The OTU-based models showed a similar trend to the genera models where a generalized decrease in median AUC was observed when only OTUs from the CRC-associated genera were used versus the full community of OTUs for both adenoma and carcinoma, with the largest median change for stool in carcinoma models and for tissue in adenoma models [Figure 4 & 5].
 
-A similar trend was observed for the OTU based models for both fecal and tissue (matched and unmatched) samples.
-There was a generalized decrease in AUC when only OTUs from the CRC-associated genera were used versus the full community of OTUs for both adenoma and carcinoma [Figure 4 & 5].
-The largest difference in median AUC for stool, between the full community of OTUs and only CRC-associated OTUs, was in carcinoma classification [Figure 4B] and for tissue it was in adenoma classification [Figure 5A].
-
-In stool the most common genera in the genus based models belonged mostly to resident genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
-With respect to the CRC-associated genera, *Fusobacterium* was the only genus present in adenoma while all four were present in carcinoma [Figure 6A & B].
-Conversely, none of these CRC-associated genera were present in the majority of studies.
-When we move on to the OTU based models, the adenoma OTU models had OTUs that classified as  *Ruminococcaceae* or *Roseburia* present in the top 10 OTUs for the vast majority of studies [Figure 6C].
-*Ruminococcaceae* was also present in the top 10 in some studies for the carcinoma OTU models, but it was *Bacteroides* that was present in the overwhelming majority of the carcinoma OTU stool models [Figure 6D].
+In stool the most common genera in the genus based models belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
+Although all four CRC-associated genera were present in carcinoma, none were present in the majority of studies and *Fusobacterium* was the only genus present in adenoma [Figure 6A & B].
+For the OTU based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].
 
@@
 For both the genera and OTU Random Forest models of adenoma and carcinoma, there appears to be very little overlap in the top 10 most important variables between stool and tissue [Figure 6 & S7].
-This discordance between stool and tissue also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of Random Forest models using matched versus unmatched tissue samples [Figure S7B-C & S7E-F].
+This discordance between stool and tissue also applies to the CRC-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of Random Forest models using matched versus unmatched tissue samples [Figure S7B-C & S7E-F].
 
@@
 Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of lesion.
-Even though mouth-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma.
-These mouth-associated genera are clearly important to carcinoma classification but ourobservations suggest that accounting for the community in which these microbes exist can increase the ability of models to make predictions.
+Even though CRC-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma.
+These CRC-associated genera are clearly important to carcinoma classification but ourobservations suggest that accounting for the community in which these microbes exist can increase the ability of models to make predictions.
 
-The data presented herein supports the importance of mouth-associated genera for carcinoma but not necessarily adenoma formation.
-A proposed mechanism by which these mouth-associated microbes cause CRC stems from the creation of a vicious cycle involving host inflammation and repair.
-Both the drastically increased carcinoma RR of mouth-associated genera versus $\alpha$-diversity metrics and increasing RR with more mouth-associated genera positivity [Figure 3C], are highly supportive of their importance.
-However, our observations show that when adding the community context to our Random Forest models along with the mouth-associated genera, the model AUC increases [Figure 4 & 5].
-This suggests the resident community cannot be ignored completely and needs to be accounted for when investigating mouth-associated microbes and carcinoma.
-Conversely, using the present data, it is less likely that mouth-associated microbes are correlated with adenoma development.
+The data presented herein supports the importance of CRC-associated genera for carcinoma but not necessarily adenoma formation.
+A proposed mechanism by which these CRC-associated microbes cause CRC stems from the creation of a vicious cycle involving host inflammation and repair.
+Both the drastically increased carcinoma RR of CRC-associated genera versus $\alpha$-diversity metrics and increasing RR with more CRC-associated genera positivity [Figure 3C], are highly supportive of their importance.
+However, our observations show that when adding the community context to our Random Forest models along with the CRC-associated genera, the model AUC increases [Figure 4 & 5].
+This suggests the resident community cannot be ignored completely and needs to be accounted for when investigating CRC-associated microbes and carcinoma.
+Conversely, using the present data, it is less likely that CRC-associated microbes are correlated with adenoma development.
 The step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1].
@@
 This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community.
-Based on these observations, it is possible to hypothesize that tsmall changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma.
+Based on these observations, it is possible to hypothesize that tsmall changes in community structure lead to new niches in which any one of the CRC-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma.
 
@@
 Generally, the OTU models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study.
-Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F].
+Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F].
 The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study.
@@
 By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in CRC.
-Further, the data presented here suggests that non-resident mouth-associated microbes can gain a foothold within the colon and the more of these microbes that are present the greater the RR of carcinoma.
-Conversely, no clear signal with these mouth-associated micobes could be detected for adenoma.
+Further, the data presented here suggests that non-resident CRC-associated microbes can gain a foothold within the colon and the more of these microbes that are present the greater the RR of carcinoma.
+Conversely, no clear signal with these CRC-associated micobes could be detected for adenoma.
 Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community.
@@
 $\beta$-diversity was then assessed for each individual study for differences between control-adenoma and control-carcinoma.
-Next, four specific mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in RR for adenoma and carcinoma.
-We then built Random Forest models based on all or mouth-associated genera and these models were trained on one study then tested on the remaining studies, for every study.
+Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in RR for adenoma and carcinoma.
+We then built Random Forest models based on all or CRC-associated genera and these models were trained on one study then tested on the remaining studies, for every study.
 A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
-For these OTU based models, the mouth-associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
+For these OTU based models, the CRC-associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
 Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power.
@@
 
-***Obtaining Mouth-Associated Genera:***
-For the mouth-associated genera analysis of the RR, the total average counts, for 100 different subsamplings, were collected for each respective OTU that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.
-The OTU based Random Forest Models using mouth-associated genera utilized a similar approach except that the OTUs were not aggregated together by genus but kept as separate OTUs.
-OTU Random Forest models using the full community included all OTUs while those using mouth-associated genera included only those OTUs that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.
+***Obtaining CRC-associated Genera:***
+For the CRC-associated genera analysis of the RR, the total average counts, for 100 different subsamplings, were collected for each respective OTU that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.
+The OTU based Random Forest Models using CRC-associated genera utilized a similar approach except that the OTUs were not aggregated together by genus but kept as separate OTUs.
+OTU Random Forest models using the full community included all OTUs while those using CRC-associated genera included only those OTUs that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.
 

### 2018-01-05 submission/manuscript.tex
results re-write upto diversity before CRC-associated re-analysis

@@
 higher RR for both stool and tissue. To investigate this we analyzed the
-classically associated CRC genera, \emph{Fusobacterium},
-\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}
-for an increase in RR. The majority of CRC-associated genera for both
-feces and tissue had a significantly increased RR for carcinoma but not
-for adenoma {[}Figure 3{]}. The RR effect size was greater for stool (RR
-range = 1.62 - 2.37) than for tissue (RR range = 1.21 - 1.81). This
-decrease may be explained by the fact that the tissue analysis included
-matched samples from the same individual. In fecal samples, the RR for
-carcinoma due to the presence of CRC-associated genera was greater than
-either the RR associated with evenness or Shannon diversity {[}Figure 2
-\& 3{]}. Additionally, the RR of carcinoma continuously increased as
-individuals tested positive for more CRC-associated genera {[}Figure 3B
-\& 3D{]}.
-
-There were two significant measures for increased RR of adenoma when
-investigating CRC-associated genera in stool: 1) Having a higher then
-median value of \emph{Porphyromonas} (P-value = 0.023) and 2) whether
-samples were positive for three CRC-associated genera (P-value = 0.022)
-{[}Figure 3A{]}. With tissue, there were three significant measures for
-an increased RR of adenoma: 1) being positive for one CRC-associated
-genera (P-value = 0.032), 2) being positive for two CRC-associated
-genera (P-value = 0.008), and 3) being positive for four CRC associated
-genera (P-value = 0.039) {[}Figure 3C{]}.
-
-\textbf{\emph{Whole Community Models Add Important Community Context:}}
-Since CRC-associated genera had increased RR for carcinoma over
-diversity metrics we wondered whether the overall bacterial community
-was at all important to classifying disease or if the CRC-associated
-genera were sufficient alone. To test this we used two approaches. The
-first approach used genus level data and tested whether there were any
-differences in AUC when training on one study and testing on all the
-others when using either all genera present or only the CRC-associated
-genera. The second approach used OTU level data and tested whether there
-was a generalized decrease in the 10-fold cross validation (CV) model
-across studies using either all OTUs or only OTUs that taxonomically
-classified to CRC-associated genera.
-
-Our first approach using genus based models showed an AUC decrease in
-model classification on the training set for both stool and tissue
-studies {[}Figure S2-S3{]}. With respect to the test sets, comprised of
-genera data from other studies, both the all genera model and
-CRC-associated models had a similar ability to detect adenomas or
-carcinomas {[}Figure S4-S6{]}. Two interesting but separate general
-observations from these models were that: 1) classification of adenomas
-was lower than carcinomas for both tissue and stool and 2) AUC for the
-classification of carcinoma was consistently lower for the tissue models
-than the stool models {[}Figure S4-S6{]}.
-
-A similar trend was observed for the OTU based models for both fecal and
-tissue (matched and unmatched) samples. There was a generalized decrease
-in AUC when only OTUs from the CRC-associated genera were used versus
-the full community of OTUs for both adenoma and carcinoma {[}Figure 4 \&
-5{]}. The largest difference in median AUC for stool, between the full
-community of OTUs and only CRC-associated OTUs, was in carcinoma
-classification {[}Figure 4B{]} and for tissue it was in adenoma
-classification {[}Figure 5A{]}.
-
-In stool the most common genera in the genus based models belonged
-mostly to resident genera such as \emph{Ruminococcus},
-\emph{Bacteroides}, and \emph{Roseburia} {[}Figure 6A \& B{]}. With
-respect to the CRC-associated genera, \emph{Fusobacterium} was the only
-genus present in adenoma while all four were present in carcinoma
-{[}Figure 6A \& B{]}. Conversely, none of these CRC-associated genera
-were present in the majority of studies. When we move on to the OTU
-based models, the adenoma OTU models had OTUs that classified as
-\emph{Ruminococcaceae} or \emph{Roseburia} present in the top 10 OTUs
-for the vast majority of studies {[}Figure 6C{]}. \emph{Ruminococcaceae}
-was also present in the top 10 in some studies for the carcinoma OTU
-models, but it was \emph{Bacteroides} that was present in the
-overwhelming majority of the carcinoma OTU stool models {[}Figure 6D{]}.
+classically associated CRC genera, \emph{Bacteroides},
+\emph{Escherichia}, \emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Peptostreptococcus}, \emph{Porphyromonas}, and
+\emph{Streptococcus} for an increase in RR. The majority of
+CRC-associated genera for both feces and tissue had a significantly
+increased RR for carcinoma but not for adenoma {[}Figure 3{]}. The RR
+effect size was greater for stool (RR range = 1.62 - 2.37) than for
+tissue (RR range = 1.21 - 1.81). This decrease may be explained by the
+fact that the tissue analysis included matched samples from the same
+individual. In fecal samples, the RR for carcinoma when having
+CRC-associated genera at a higher than overal study median relative
+abundance was greater than the RR associated with evenness or Shannon
+diversity {[}Figure 2 \& 3{]}. Additionally, the RR of carcinoma
+continuously increased as individuals tested positive for more
+CRC-associated genera {[}Figure 3B \& 3D{]}.
+
+In stool the significant measures that increased RR of adenoma for
+CRC-associated genera were having a higher than overall study median
+relative abundance of \emph{Porphyromonas} (P-value = 0.023) and being
+positive for three CRC-associated genera (P-value = 0.022) {[}Figure
+3A{]}. Within tissue the significant measures for an increased RR of
+adenoma were being positive for one CRC-associated genera (P-value =
+0.032), being positive for two CRC-associated genera (P-value = 0.008),
+and being positive for four CRC associated genera (P-value = 0.039)
+{[}Figure 3C{]}.
+
+\textbf{\emph{Whole Community Models Add Important Context:}} Since
+CRC-associated genera increased RR for carcinoma over diversity metrics
+we assessed whether the bacterial community was better at classifying
+disease versus only CRC-associated genera and tested this using two
+approaches. The first approach used genus level data and tested whether
+there were any differences in AUC between all genera and CRC-associated
+genera only during training on one study or testing on all other
+studies. The second approach used OTU level data and tested whether
+there was a generalized decrease in the 10-fold cross validation (CV)
+model AUC across studies when comparing all OTUs versus only OTUs that
+taxonomically classified to CRC-associated genera.
+
+For the genera-based models both tissue and stool studies had a median
+AUC decrease in model classification when only CRC-associated genera
+versus all genera were included {[}Figure S2-S3{]}. With respect to the
+test sets, comprised of genera data from other studies, both all and
+CRC-associated genera models had a similar ability to detect adenomas or
+carcinomas {[}Figure S4-S6{]}. The OTU-based models showed a similar
+trend to the genera models where a generalized decrease in median AUC
+was observed when only OTUs from the CRC-associated genera were used
+versus the full community of OTUs for both adenoma and carcinoma, with
+the largest median change for stool in carcinoma models and for tissue
+in adenoma models {[}Figure 4 \& 5{]}.
+
+In stool the most common genera in the genus based models belonged to
+genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
+\emph{Roseburia} {[}Figure 6A \& B{]}. Although all four CRC-associated
+genera were present in carcinoma, none were present in the majority of
+studies and \emph{Fusobacterium} was the only genus present in adenoma
+{[}Figure 6A \& B{]}. For the OTU based models, \emph{Ruminococcaceae}
+was present in the top 10 consistently for both adenoma and carcinoma
+models while \emph{Roseburia} was only present in many adenoma models
+and \emph{Bacteroides} was present in the overwhelming majority of the
+carcinoma models {[}Figure 6C \& 6D{]}.
 
@@
 important variables between stool and tissue {[}Figure 6 \& S7{]}. This
-discordance between stool and tissue also applies to the
-mouth-associated genera with one noticeable skew being that
-\emph{Fusobacterium} and \emph{Fusobacteriaceae} occur more often in the
-top 10 of Random Forest models using matched versus unmatched tissue
-samples {[}Figure S7B-C \& S7E-F{]}.
+discordance between stool and tissue also applies to the CRC-associated
+genera with one noticeable skew being that \emph{Fusobacterium} and
+\emph{Fusobacteriaceae} occur more often in the top 10 of Random Forest
+models using matched versus unmatched tissue samples {[}Figure S7B-C \&
+S7E-F{]}.
 
@@
 translate into large effect sizes for the RR of lesion. Even though
-mouth-associated genera increased the RR of carcinoma, they did not
-consistently increase the RR of adenoma. These mouth-associated genera
-are clearly important to carcinoma classification but ourobservations
+CRC-associated genera increased the RR of carcinoma, they did not
+consistently increase the RR of adenoma. These CRC-associated genera are
+clearly important to carcinoma classification but ourobservations
 suggest that accounting for the community in which these microbes exist
@@
 
-The data presented herein supports the importance of mouth-associated
+The data presented herein supports the importance of CRC-associated
 genera for carcinoma but not necessarily adenoma formation. A proposed
-mechanism by which these mouth-associated microbes cause CRC stems from
+mechanism by which these CRC-associated microbes cause CRC stems from
 the creation of a vicious cycle involving host inflammation and repair.
-Both the drastically increased carcinoma RR of mouth-associated genera
+Both the drastically increased carcinoma RR of CRC-associated genera
 versus \(\alpha\)-diversity metrics and increasing RR with more
-mouth-associated genera positivity {[}Figure 3C{]}, are highly
-supportive of their importance. However, our observations show that when
-adding the community context to our Random Forest models along with the
-mouth-associated genera, the model AUC increases {[}Figure 4 \& 5{]}.
-This suggests the resident community cannot be ignored completely and
-needs to be accounted for when investigating mouth-associated microbes
-and carcinoma. Conversely, using the present data, it is less likely
-that mouth-associated microbes are correlated with adenoma development.
-The step-wise decrease in diversity suggests that the adenoma community
-is not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
+CRC-associated genera positivity {[}Figure 3C{]}, are highly supportive
+of their importance. However, our observations show that when adding the
+community context to our Random Forest models along with the
+CRC-associated genera, the model AUC increases {[}Figure 4 \& 5{]}. This
+suggests the resident community cannot be ignored completely and needs
+to be accounted for when investigating CRC-associated microbes and
+carcinoma. Conversely, using the present data, it is less likely that
+CRC-associated microbes are correlated with adenoma development. The
+step-wise decrease in diversity suggests that the adenoma community is
+not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
 diversity, at this early stage of disease, could be focal to the adenoma
@@
 in community structure lead to new niches in which any one of the
-mouth-associated genera can gain a foothold, exacerbating the initial
+CRC-associated genera can gain a foothold, exacerbating the initial
 changes in community and facilitating the transition from adenoma to
@@
 prevalent in the top 10 across study. Unlike in stool,
-\emph{Fusobacterium} was the only mouth-associated bacteria consistently
+\emph{Fusobacterium} was the only CRC-associated bacteria consistently
 present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
@@
 of the bacterial community in CRC. Further, the data presented here
-suggests that non-resident mouth-associated microbes can gain a foothold
+suggests that non-resident CRC-associated microbes can gain a foothold
 within the colon and the more of these microbes that are present the
 greater the RR of carcinoma. Conversely, no clear signal with these
-mouth-associated micobes could be detected for adenoma. Our observations
+CRC-associated micobes could be detected for adenoma. Our observations
 also highlight the importance of power and sample number considerations
@@
 control-adenoma and control-carcinoma. Next, four specific
-mouth-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
+CRC-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
 \emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
 differences in RR for adenoma and carcinoma. We then built Random Forest
-models based on all or mouth-associated genera and these models were
+models based on all or CRC-associated genera and these models were
 trained on one study then tested on the remaining studies, for every
@@
 80/20 splitting of the data, was used to generate a range of expected
-AUCs. For these OTU based models, the mouth-associated genera included
-all OTUs that had a taxonomic classification to \emph{Fusobacterium},
+AUCs. For these OTU based models, the CRC-associated genera included all
+OTUs that had a taxonomic classification to \emph{Fusobacterium},
 \emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas}.
@@
 
-\textbf{\emph{Obtaining Mouth-Associated Genera:}} For the
-mouth-associated genera analysis of the RR, the total average counts,
-for 100 different subsamplings, were collected for each respective OTU
-that had a genus level taxonomic classification to \emph{Fusobacterium},
+\textbf{\emph{Obtaining CRC-associated Genera:}} For the CRC-associated
+genera analysis of the RR, the total average counts, for 100 different
+subsamplings, were collected for each respective OTU that had a genus
+level taxonomic classification to \emph{Fusobacterium},
 \emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}.
-The OTU based Random Forest Models using mouth-associated genera
-utilized a similar approach except that the OTUs were not aggregated
-together by genus but kept as separate OTUs. OTU Random Forest models
-using the full community included all OTUs while those using
-mouth-associated genera included only those OTUs that had a genus level
-taxonomic classification to \emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Peptostreptococcus}, and \emph{Porphyromonas}.
+The OTU based Random Forest Models using CRC-associated genera utilized
+a similar approach except that the OTUs were not aggregated together by
+genus but kept as separate OTUs. OTU Random Forest models using the full
+community included all OTUs while those using CRC-associated genera
+included only those OTUs that had a genus level taxonomic classification
+to \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus},
+and \emph{Porphyromonas}.
 

### 2018-01-08 code/make_select_genus_RR_graph.R
remove combos and add top 5 genera RR

@@ -10,39 +10,58 @@ loadLibs(c("tidyverse", "gridExtra", "viridis"))
 # Load needed data tables (adenoma)
 adn_all_stool <- read_csv("data/process/tables/adn_select_genus_RR_stool_composite.csv") %>% 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
-  mutate(study = "composite") %>% 
-  bind_rows(read_csv("data/process/tables/adn_select_genus_RR_stool_ind_results.csv")) %>% 
-  bind_rows(read_csv("data/process/tables/adn_select_genus_inc_4_stool.csv") %>% 
-              rename(est = rr, lower = ci_lb, upper = ci_ub)) %>% 
-  mutate(study = ifelse(is.na(study), invisible("composite"), invisible(study)))
+  mutate(study = "composite") %>% arrange(pvalue, est) %>% 
+  filter(est > 1) %>% slice(1:5) %>% 
+  bind_rows(read_csv("data/process/tables/adn_select_genus_RR_stool_composite.csv") %>% 
+              rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
+              mutate(study = "composite") %>% arrange(pvalue, est) %>% 
+              filter(est < 1) %>% slice(1:5))
+
+adn_ind_stool <- read_csv("data/process/tables/adn_select_genus_RR_stool_ind_results.csv") %>% 
+  filter(measure %in% as.data.frame(adn_all_stool)[, "measure"])
 
 adn_all_tissue <- read_csv("data/process/tables/adn_select_genus_RR_tissue_composite.csv") %>% 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% 
-  bind_rows(read_csv("data/process/tables/adn_select_genus_RR_tissue_ind_results.csv")) %>% 
-  bind_rows(read_csv("data/process/tables/adn_select_genus_inc_4_tissue.csv") %>% 
-              rename(est = rr, lower = ci_lb, upper = ci_ub)) %>% 
-  mutate(study = ifelse(is.na(study), invisible("composite"), invisible(study)))
+  arrange(pvalue, est) %>% 
+  filter(est > 1) %>% slice(1:5) %>% 
+  bind_rows(read_csv("data/process/tables/adn_select_genus_RR_tissue_composite.csv") %>% 
+              rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
[…]

### 2018-01-08 submission/supplemental.Rmd
add tables with P-value < 0.05 for genera RR

@@
 ```
+
+\newpage
+
+**Table S8: RR of Adenoma in Stool for each Individual Genera with Initial P-Value under 0.05**
+
+```{r supplemental_table8, echo=FALSE, message=FALSE, warning=FALSE}
+
+temp_8 <- read_csv("../data/process/tables/adn_select_genus_RR_stool_composite.csv") %>% 
+  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
+  arrange(bh, pvalue, rr) %>% 
+  mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = round(pvalue, digits = 4), 
+         bh = round(bh, digits = 3), 
+         measure = str_replace_all(measure, "_", " ")) %>% 
+  select(rr, ci_lb, ci_ub, pvalue, bh, measure)
+
+colnames(temp_8) <- c("RR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+
+kable(temp_8, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\newpage
+
+**Table S9: RR of Carcinoma in Stool for each Individual Genera with Initial P-Value under 0.05**
+
+```{r supplemental_table9, echo=FALSE, message=FALSE, warning=FALSE}
+
+temp_9 <- read_csv("../data/process/tables/select_genus_RR_stool_composite.csv") %>% 
+  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
+  arrange(bh, pvalue, rr) %>% 
+  mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = format(pvalue, scientific = T, digits = 3), 
+         bh = format(bh, scientific = T, digits = 3), 
+         measure = str_replace_all(measure, "_", " ")) %>% 
+  select(rr, ci_lb, ci_ub, pvalue, bh, measure)
+
+colnames(temp_9) <- c("RR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+
+kable(temp_9, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\newpage
+
+**Table S10: RR of Adenoma in Tissue for each Individual Genera with Initial P-Value under 0.05**
+
+```{r supplemental_table10, echo=FALSE, message=FALSE, warning=FALSE}
+
+temp_10 <- read_csv("../data/process/tables/adn_select_genus_RR_tissue_composite.csv") %>% 
+  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
+  arrange(bh, pvalue, rr) %>% 
+  mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = round(pvalue, digits = 4), 
+         bh = round(bh, digits = 3), 
+         measure = str_replace_all(measure, "_", " ")) %>% 
+  select(rr, ci_lb, ci_ub, pvalue, bh, measure)
+
+colnames(temp_10) <- c("RR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+
+kable(temp_10, format = "markdown", align = 'c', padding = 2)
+
+```
+
+\newpage
+
+**Table S11: RR of Carcinoma in Tissue for each Individual Genera with Initial P-Value under 0.05**
+
+```{r supplemental_table11, echo=FALSE, message=FALSE, warning=FALSE}
+
+temp_11 <- read_csv("../data/process/tables/select_genus_RR_tissue_composite.csv") %>% 
+  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
+  arrange(bh, pvalue, rr) %>% 
+  mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = format(pvalue, scientific = T, digits = 3), 
+         bh = format(bh, scientific = T, digits = 3), 
+         measure = str_replace_all(measure, "_", " ")) %>% 
+  select(rr, ci_lb, ci_ub, pvalue, bh, measure)
+
+colnames(temp_11) <- c("RR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+
+kable(temp_11, format = "markdown", align = 'c', padding = 2)
+
+```

### 2018-01-08 submission/manuscript.Rmd
incorporate RR genera updates

@@
 adn_select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_RR_tissue_composite.csv"))
-
-select_inc_4_stool <- as.data.frame(read_csv("../data/process/tables/select_genus_inc_4_stool.csv"))
-select_inc_4_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_inc_4_tissue.csv"))
-
-adn_select_inc_4_RR_stool <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_inc_4_stool.csv"))
-adn_select_inc_4_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_inc_4_tissue.csv"))
 
@@
 
-***Carcinoma-Associated Genera Minimally Impacts RR of Adenoma:***
-Based on the small increase in RR using $\alpha$-diversity metrics, we assessed if the presence of specific genera resulted in a higher RR for both stool and tissue.
-To investigate this we analyzed the classically associated CRC genera, *Bacteroides*, *Escherichia*, *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, *Porphyromonas*, and *Streptococcus* for an increase in RR.
-The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3].
-The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) than for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`).
-This decrease may be explained by the fact that the tissue analysis included matched samples from the same individual.
-In fecal samples, the RR for carcinoma when having CRC-associated genera at a higher than overal study median relative abundance was greater than the RR associated with evenness or Shannon diversity [Figure 2 & 3].
-Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC-associated genera [Figure 3B & 3D].
+***Mouth-Associated Genera are Associated with an Increased RR of Lesion:***
+Based on the small increase in RR using $\alpha$-diversity metrics, we assessed if being higher than the median for any specific genera resulted in an altered RR for adenoma or carcinoma in stool and tissue.
+To investigate this we analyzed all common genera across each study, in tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR.
+For both stool and tissue none of the genera in the top 5 most significantly increased RR were the same between adenoma and carcinoma [Figure 3].
+Mouth-associated genera were in the top 5 genera associated with an increased RR of adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and tissue.
+Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and tissue [Figure 3].
+Overall, there was little direct overlap of the top 5 increased or decreased RR genera between stool and tissue.
 
-In stool the significant measures that increased RR of adenoma for CRC-associated genera were having a higher than overall study median relative abundance of *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and being positive for three CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A].
-Within tissue the significant measures for an increased RR of adenoma were being positive for one CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), being positive for two CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].
+For adenoma, there was almost no overlap between genera with a RR P-value of less than 0.05 and when they were similar the RR was in opposite directions (e.g.
+*Lactococcus*) [Table S8 & S9].
+Many of the genera that had a P-value under 0.05 for either an increased or dereased RR are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
+For carcinoma, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both tissue and stool [Table S10 & S11].
+Although the most significant increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].
 
@@
 Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of lesion.
-Even though CRC-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma.
-These CRC-associated genera are clearly important to carcinoma classification but ourobservations suggest that accounting for the community in which these microbes exist can increase the ability of models to make predictions.
+Even though mouth-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma.
+These mouth-associated genera are clearly important to carcinoma classification but our observations suggest that accounting for the community in which these microbes exist can increase the ability of models to make predictions.
 
-The data presented herein supports the importance of CRC-associated genera for carcinoma but not necessarily adenoma formation.
-A proposed mechanism by which these CRC-associated microbes cause CRC stems from the creation of a vicious cycle involving host inflammation and repair.
-Both the drastically increased carcinoma RR of CRC-associated genera versus $\alpha$-diversity metrics and increasing RR with more CRC-associated genera positivity [Figure 3C], are highly supportive of their importance.
-However, our observations show that when adding the community context to our Random Forest models along with the CRC-associated genera, the model AUC increases [Figure 4 & 5].
-This suggests the resident community cannot be ignored completely and needs to be accounted for when investigating CRC-associated microbes and carcinoma.
-Conversely, using the present data, it is less likely that CRC-associated microbes are correlated with adenoma development.
+The data presented herein supports the importance of mouth-associated genera for carcinoma but not necessarily adenoma formation.
+A proposed mechanism by which these mouth-associated microbes cause CRC stems from the creation of a vicious cycle involving host inflammation and repair.
+Both the drastically increased carcinoma RR of mouth-associated genera versus $\alpha$-diversity metrics and increasing RR with more mouth-associated genera positivity [Figure 3C], are highly supportive of their importance.
+However, our observations show that when adding the community context to our Random Forest models along with the mouth-associated genera, the model AUC increases [Figure 4 & 5].
+This suggests the resident community cannot be ignored completely and needs to be accounted for when investigating mouth-associated microbes and carcinoma.
+Conversely, using the present data, it is less likely that mouth-associated microbes are correlated with adenoma development.
 The step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1].
@@
 Generally, the OTU models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study.
-Unlike in stool, *Fusobacterium* was the only CRC-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F].
+Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F].
 The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study.
-Additionally, the presence of *Propionibacterium*, within the top 10 most important variables for the genera and OTU models, is worrying because it is commonly a marker of contamination.
+Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models, is worrying because it is commonly a marker of contamination.
 The low bacterial biomass of tissue samples coupled with potential contaminationcould ultimately explain why the these results seem to be more sporadic than the stool results.
@@
 By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in CRC.
-Further, the data presented here suggests that non-resident CRC-associated microbes can gain a foothold within the colon and the more of these microbes that are present the greater the RR of carcinoma.
-Conversely, no clear signal with these CRC-associated micobes could be detected for adenoma.
+Further, the data presented here suggests that non-resident mouth-associated microbes can gain a foothold within the colon and the more of these microbes that are present the greater the RR of carcinoma.
+Conversely, no clear signal with these mouth-associated micobes could be detected for adenoma.
 Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community.

### 2018-01-08 submission/manuscript.tex
incorporate RR genera updates

@@
 the relative risk for carcinoma (P-value \textless{} 0.05) but not
-adenoma (P-value \textgreater{} 0.05) with the exception of
+adenoma (P-value check tables 0.05) with the exception of
 \emph{Porphyromonas} (P-value = 0.023). Using the whole community versus
@@
 
-\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
-Adenoma:}} Based on the small increase in RR using \(\alpha\)-diversity
-metrics, we assessed if the presence of specific genera resulted in a
-higher RR for both stool and tissue. To investigate this we analyzed the
-classically associated CRC genera, \emph{Bacteroides},
-\emph{Escherichia}, \emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Peptostreptococcus}, \emph{Porphyromonas}, and
-\emph{Streptococcus} for an increase in RR. The majority of
-CRC-associated genera for both feces and tissue had a significantly
-increased RR for carcinoma but not for adenoma {[}Figure 3{]}. The RR
-effect size was greater for stool (RR range = 1.62 - 2.37) than for
-tissue (RR range = 1.21 - 1.81). This decrease may be explained by the
-fact that the tissue analysis included matched samples from the same
-individual. In fecal samples, the RR for carcinoma when having
-CRC-associated genera at a higher than overal study median relative
-abundance was greater than the RR associated with evenness or Shannon
-diversity {[}Figure 2 \& 3{]}. Additionally, the RR of carcinoma
-continuously increased as individuals tested positive for more
-CRC-associated genera {[}Figure 3B \& 3D{]}.
-
-In stool the significant measures that increased RR of adenoma for
-CRC-associated genera were having a higher than overall study median
-relative abundance of \emph{Porphyromonas} (P-value = 0.023) and being
-positive for three CRC-associated genera (P-value = 0.022) {[}Figure
-3A{]}. Within tissue the significant measures for an increased RR of
-adenoma were being positive for one CRC-associated genera (P-value =
-0.032), being positive for two CRC-associated genera (P-value = 0.008),
-and being positive for four CRC associated genera (P-value = 0.039)
-{[}Figure 3C{]}.
+\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
+RR of Lesion:}} Based on the small increase in RR using
+\(\alpha\)-diversity metrics, we assessed if being higher than the
+median for any specific genera resulted in an altered RR for adenoma or
+carcinoma in stool and tissue. To investigate this we analyzed all
+common genera across each study, in tissue or stool, and assessed
+whether a relative abundance higher than the median results in an
+increase or decrease in RR. For both stool and tissue none of the genera
+in the top 5 most significantly increased RR were the same between
+adenoma and carcinoma {[}Figure 3{]}. Mouth-associated genera were in
+the top 5 genera associated with an increased RR of adenoma
+(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
+3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
+\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and tissue.
+Conversely, genera commonly associated with a normal gastrointestinal
+tract were correlated with a decreased RR for both adenoma and carcinoma
+for both stool and tissue {[}Figure 3{]}. Overall, there was little
+direct overlap of the top 5 increased or decreased RR genera between
+stool and tissue.
+
+For adenoma, there was almost no overlap between genera with a RR
+P-value of less than 0.05 and when they were similar the RR was in
+opposite directions (e.g. \emph{Lactococcus}) {[}Table S8 \& S9{]}. Many
+of the genera that had a P-value under 0.05 for either an increased or
+dereased RR are also highly prevalent in contamination, specifically,
+\emph{Novosphingobium}, \emph{Selemonas}, and \emph{Achromobacter}
+{[}Figure 3 \& Table S8-S9{]}. For carcinoma, certain mouth-associated
+genera (\emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Peptostreptococcus}) had a high RR for both tissue and stool
+{[}Table S10 \& S11{]}. Although the most significant increased RR
+genera for tissue was \emph{Camplyobacter} while in stool it was
+\emph{Peptostreptococcus} {[}Table S10 \& S11{]}.
 
@@
 translate into large effect sizes for the RR of lesion. Even though
-CRC-associated genera increased the RR of carcinoma, they did not
-consistently increase the RR of adenoma. These CRC-associated genera are
-clearly important to carcinoma classification but ourobservations
+mouth-associated genera increased the RR of carcinoma, they did not
+consistently increase the RR of adenoma. These mouth-associated genera
+are clearly important to carcinoma classification but our observations
 suggest that accounting for the community in which these microbes exist
@@
 
-The data presented herein supports the importance of CRC-associated
+The data presented herein supports the importance of mouth-associated
 genera for carcinoma but not necessarily adenoma formation. A proposed
-mechanism by which these CRC-associated microbes cause CRC stems from
+mechanism by which these mouth-associated microbes cause CRC stems from
 the creation of a vicious cycle involving host inflammation and repair.
-Both the drastically increased carcinoma RR of CRC-associated genera
+Both the drastically increased carcinoma RR of mouth-associated genera
 versus \(\alpha\)-diversity metrics and increasing RR with more
-CRC-associated genera positivity {[}Figure 3C{]}, are highly supportive
-of their importance. However, our observations show that when adding the
-community context to our Random Forest models along with the
-CRC-associated genera, the model AUC increases {[}Figure 4 \& 5{]}. This
-suggests the resident community cannot be ignored completely and needs
-to be accounted for when investigating CRC-associated microbes and
-carcinoma. Conversely, using the present data, it is less likely that
-CRC-associated microbes are correlated with adenoma development. The
-step-wise decrease in diversity suggests that the adenoma community is
-not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
+mouth-associated genera positivity {[}Figure 3C{]}, are highly
+supportive of their importance. However, our observations show that when
+adding the community context to our Random Forest models along with the
+mouth-associated genera, the model AUC increases {[}Figure 4 \& 5{]}.
+This suggests the resident community cannot be ignored completely and
+needs to be accounted for when investigating mouth-associated microbes
+and carcinoma. Conversely, using the present data, it is less likely
+that mouth-associated microbes are correlated with adenoma development.
+The step-wise decrease in diversity suggests that the adenoma community
+is not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
 diversity, at this early stage of disease, could be focal to the adenoma
@@
 prevalent in the top 10 across study. Unlike in stool,
-\emph{Fusobacterium} was the only CRC-associated bacteria consistently
+\emph{Fusobacterium} was the only mouth-associated bacteria consistently
 present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
@@
 10 taxa being present only in a single study. Additionally, the presence
-of \emph{Propionibacterium}, within the top 10 most important variables
-for the genera and OTU models, is worrying because it is commonly a
-marker of contamination. The low bacterial biomass of tissue samples
-coupled with potential contaminationcould ultimately explain why the
-these results seem to be more sporadic than the stool results.
+of genera associated with contamination, within the top 10 most
+important variables for the genera and OTU models, is worrying because
+it is commonly a marker of contamination. The low bacterial biomass of
+tissue samples coupled with potential contaminationcould ultimately
+explain why the these results seem to be more sporadic than the stool
+results.
 
@@
 of the bacterial community in CRC. Further, the data presented here
-suggests that non-resident CRC-associated microbes can gain a foothold
+suggests that non-resident mouth-associated microbes can gain a foothold
 within the colon and the more of these microbes that are present the
 greater the RR of carcinoma. Conversely, no clear signal with these
-CRC-associated micobes could be detected for adenoma. Our observations
+mouth-associated micobes could be detected for adenoma. Our observations
 also highlight the importance of power and sample number considerations

### 2018-01-08 submission/supplemental.Rmd
add RR tables P<0.05

@@
 
-**Table S9: RR of Carcinoma in Stool for each Individual Genera with Initial P-Value under 0.05**
+**Table S9: RR of Adenoma in Tissue for each Individual Genera with Initial P-Value under 0.05**
 
@@
 
-temp_9 <- read_csv("../data/process/tables/select_genus_RR_stool_composite.csv") %>% 
+temp_9 <- read_csv("../data/process/tables/adn_select_genus_RR_tissue_composite.csv") %>% 
   mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
@@
          ci_ub = round(ci_ub, digits = 2), 
-         pvalue = format(pvalue, scientific = T, digits = 3), 
-         bh = format(bh, scientific = T, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         bh = round(bh, digits = 3), 
          measure = str_replace_all(measure, "_", " ")) %>% 
@@
 
-**Table S10: RR of Adenoma in Tissue for each Individual Genera with Initial P-Value under 0.05**
+**Table S10: RR of Carcinoma in Stool for each Individual Genera with Initial P-Value under 0.05**
 
@@
 
-temp_10 <- read_csv("../data/process/tables/adn_select_genus_RR_tissue_composite.csv") %>% 
+temp_10 <- read_csv("../data/process/tables/select_genus_RR_stool_composite.csv") %>% 
   mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
@@
          ci_ub = round(ci_ub, digits = 2), 
-         pvalue = round(pvalue, digits = 4), 
-         bh = round(bh, digits = 3), 
+         pvalue = format(pvalue, scientific = T, digits = 3), 
+         bh = format(bh, scientific = T, digits = 3), 
          measure = str_replace_all(measure, "_", " ")) %>% 

### 2018-01-09 submission/manuscript.Rmd
re-write second result paragraph

@@
 
-Using the Bray-Curtis distance metric, there was a significant difference across study in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S4 & S5].
-For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with tissue samples from the same individual (matched) had no differences [Table S6 & S7].
+Using the Bray-Curtis distance metric and PERMANOVA, it was possible to identify significant bacterial community changes, in specific studies, for both carcinoma versus control and adenoma versus control [Table S4 & S5].
+Significant differences in bacterial community, assessed by PERMANOVA, were also identified in unmatched tissue samples (control and adenoma or carcinoma tissue not from the same individual) for both those with adenoma and carcinoma [Table S6 & S7].
+For studies with matched samples (unaffected tissue and an adenoma or carcinoma from the same individual) no differences in bacterial community were observed when assessed with PERMANOVA [Table S6 & S7].
 

### 2018-01-09 submission/manuscript.tex
re-write second result paragraph

@@
 
-Using the Bray-Curtis distance metric, there was a significant
-difference across study in the bacterial community of fecal samples
-between carcinoma and controls, but not adenoma and controls {[}Table S4
-\& S5{]}. For studies with unmatched tissue samples a similar trend was
-observed {[}Table S3 \& S4{]} while studies with tissue samples from the
-same individual (matched) had no differences {[}Table S6 \& S7{]}.
+Using the Bray-Curtis distance metric and PERMANOVA, it was possible to
+identify significant bacterial community changes, in specific studies,
+for both carcinoma versus control and adenoma versus control {[}Table S4
+\& S5{]}. Significant differences in bacterial community, assessed by
+PERMANOVA, were also identified in unmatched tissue samples (control and
+adenoma or carcinoma tissue not from the same individual) for both those
+with adenoma and carcinoma {[}Table S6 \& S7{]}. For studies with
+matched samples (unaffected tissue and an adenoma or carcinoma from the
+same individual) no differences in bacterial community were observed
+when assessed with PERMANOVA {[}Table S6 \& S7{]}.
 

### 2018-01-09 submission/supplemental.Rmd
change table orders

@@
 
-**Table S1: Alpha Diversity Metrics RR in Tissue for Adenoma**
+**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Adenoma**
 
 ```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}
+
+temp_t1 <- read_csv("../data/process/tables/beta_perm_adn_stool_summary.csv")
+
+table_s1 <- temp_t1 %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = stringr::str_replace(study, 
+                                      c("brim", "zeller", "baxter", "hale"), 
+                                      c("Brim", "Zeller", "Baxter", "Hale")))
+
+colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Adenoma (n)", "Control (n)")
+
+kable(table_s1, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Carcinoma**
+
+```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
+
+temp_t2 <- read_csv("../data/process/tables/beta_perm_stool_summary.csv")
+
+table_s2 <- temp_t2 %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = stringr::str_replace(study, 
+                                      c("wang", "weir", "ahn", "zeller", "baxter", "hale", "flemer"), 
+                                      c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")))
+
+colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma (n)", "Control (n)")
+
+kable(table_s2, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S3: Alpha Diversity Metrics RR in Tissue for Adenoma**
+
+```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-**Table S2: Alpha Diversity Metrics RR in Tissue for Unmatched Carcinoma Non-Carcinoma Samples**
+**Table S4: Alpha Diversity Metrics RR in Tissue for Unmatched Carcinoma Non-Carcinoma Samples**
 
-```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
+```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-**Table S3: Alpha Diversity Metrics RR in Tissue for Matched Carcinoma Non-Carcinoma Samples**
+**Table S5: Alpha Diversity Metrics RR in Tissue for Matched Carcinoma Non-Carcinoma Samples**
 
-```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
+```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 kable(alpha_RR_crc_match_tissue, format = "markdown", align = 'c', padding = 2)
-```
-
-\newpage
-
-**Table S4: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Adenoma**
-
-```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
-
-temp_t1 <- read_csv("../data/process/tables/beta_perm_adn_stool_summary.csv")
-
-table_s4 <- temp_t1 %>% 
-  mutate(fstat = round(fstat, digits = 2), 
-         r2 = round(r2, digits = 3), 
-         pvalue = round(pvalue, digits = 4), 
-         study = stringr::str_replace(study, 
-                                      c("brim", "zeller", "baxter", "hale"), 
-                                      c("Brim", "Zeller", "Baxter", "Hale")))
-
-colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Adenoma (n)", "Control (n)")
-
-kable(table_s4, format = "markdown", align = 'c', padding = 2)
-```
-
-\newpage
-
-**Table S5: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Carcinoma**
-
-```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
-
-temp_t2 <- read_csv("../data/process/tables/beta_perm_stool_summary.csv")
-
-table_s5 <- temp_t2 %>% 
-  mutate(fstat = round(fstat, digits = 2), 
-         r2 = round(r2, digits = 3), 
-         pvalue = round(pvalue, digits = 4), 
-         study = stringr::str_replace(study, 
-                                      c("wang", "weir", "ahn", "zeller", "baxter", "hale", "flemer"), 
-                                      c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")))
-
-colnames(table_s5) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma (n)", "Control (n)")
-
-kable(table_s5, format = "markdown", align = 'c', padding = 2)
 ```

### 2018-01-09 submission/manuscript.Rmd
re write first result section

--- 2018-01-08 submission/manuscript.Rmd
@@
 ***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
-By power transforming and Z-score normalizing the $\alpha$-diversity metrics for the entire data set we assessed whether there were any broad scale community differences that could be detected as disease severity worsened.
-Using linear mixed-effect models to control for study, re-sampling of the same individual (tissue studies only), and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness and Shannon diversity in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) and no $\alpha$-diversity correlations in tissue (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1].
+Power transforming and Z-score normalizing OTU richness, evenness, and Shannon diversity for the entire data set allowed for the combination of data from all studies to assess for differences in these broad scale community metrics as disease severity worsens.
+Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased.
+There was a significant decrease in both evenness and Shannon diversity from control to adenoma to carcinoma in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A].
 We next tested whether these detectable differences in community resulted in significant increases in RR.
 For fecal samples, a decrease in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
-Interestingly, for both adenoma and carcinoma there was no increase in RR within tissue samples for any alpha diversity metric investigated [Table S1-S3].
+Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma versus control and adenoma versus control [Table S1 & S2].
 
-Using the Bray-Curtis distance metric, there was a significant difference across study in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S4 & S5].
-For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with tissue samples from the same individual (matched) had no differences [Table S6 & S7].
+Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased.
+For tissue there were no significant changes in OTU richness, evenness, and Shannon diversity from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1B & C].
+When analyzing the RR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) tissue samples, for both adenoma and carcinoma there was no significant change in RR based on lower than median values for OTU richness, evenness, and Shannon diversity [Table S3-S5].
+Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were also identified in unmatched tissue samples  for both those with adenoma and carcinoma [Table S6 & S7].
+For studies with matched samples no differences in bacterial community were observed when assessed with PERMANOVA [Table S6 & S7].
 

### 2018-01-09 submission/manuscript.tex
re write first result section

--- 2018-01-08 submission/manuscript.tex
@@
 \textbf{\emph{Lower Community Diversity is Associated with Increased RR
-of Carcinomas:}} By power transforming and Z-score normalizing the
-\(\alpha\)-diversity metrics for the entire data set we assessed whether
-there were any broad scale community differences that could be detected
-as disease severity worsened. Using linear mixed-effect models to
-control for study, re-sampling of the same individual (tissue studies
-only), and variable region, there was a significant decrease from
-control to adenoma to carcinoma for both evenness and Shannon diversity
-in stool (P-value = 0.025 and 0.043, respectively) and no
-\(\alpha\)-diversity correlations in tissue (P-value \textgreater{}
-0.05) {[}Figure 1{]}. We next tested whether these detectable
-differences in community resulted in significant increases in RR. For
-fecal samples, a decrease in evenness resulted in a significantly
-increased RR for carcinoma (RR = 1.36 (1.15 - 1.61), P-value = 0.00034)
-and adenoma (RR = 1.16 (1.01 - 1.34), P-value = 0.032) while a decrease
-in Shannon diversity only increased the RR for carcinoma (RR = 1.33
-(1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}. Interestingly, for both
-adenoma and carcinoma there was no increase in RR within tissue samples
-for any alpha diversity metric investigated {[}Table S1-S3{]}.
-
-Using the Bray-Curtis distance metric, there was a significant
-difference across study in the bacterial community of fecal samples
-between carcinoma and controls, but not adenoma and controls {[}Table S4
-\& S5{]}. For studies with unmatched tissue samples a similar trend was
-observed {[}Table S3 \& S4{]} while studies with tissue samples from the
-same individual (matched) had no differences {[}Table S6 \& S7{]}.
+of Carcinomas:}} Power transforming and Z-score normalizing OTU
+richness, evenness, and Shannon diversity for the entire data set
+allowed for the combination of data from all studies to assess for
+differences in these broad scale community metrics as disease severity
+worsens. Using linear mixed-effect models to control for study and
+variable region we assessed whether OTU richness, evenness, or Shannon
+diversity changed in a step-wise manner as disease severity increased.
+There was a significant decrease in both evenness and Shannon diversity
+from control to adenoma to carcinoma in stool (P-value = 0.025 and
+0.043, respectively) {[}Figure 1A{]}. We next tested whether these
+detectable differences in community resulted in significant increases in
+RR. For fecal samples, a decrease in evenness resulted in a
+significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
+P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
+0.032) while a decrease in Shannon diversity only increased the RR for
+carcinoma (RR = 1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}.
+Using the Bray-Curtis distance metric and PERMANOVA, it was also
+possible to identify significant bacterial community changes, in
+specific studies, for both carcinoma versus control and adenoma versus
+control {[}Table S1 \& S2{]}.
+
+Using similar transformations for tissue samples, linear mixed-effect
+models were used on the transformed combined data to control for study,
+re-sampling of the same individual, and variable region to test whether
+OTU richness, evenness, or Shannon diversity changed in a step-wise
+manner as disease severity increased. For tissue there were no
+significant changes in OTU richness, evenness, and Shannon diversity
+from control to adenoma to carcinoma (P-value \textgreater{} 0.05)
+{[}Figure 1B \& C{]}. When analyzing the RR, for matched (unaffected
+tissue and an adenoma or carcinoma from the same individual) and
+unmatched (control and adenoma or carcinoma tissue not from the same
+individual) tissue samples, for both adenoma and carcinoma there was no
+significant change in RR based on lower than median values for OTU
+richness, evenness, and Shannon diversity {[}Table S3-S5{]}. Similar to
+stool samples, significant differences in bacterial community, assessed
+by PERMANOVA, were also identified in unmatched tissue samples for both
+those with adenoma and carcinoma {[}Table S6 \& S7{]}. For studies with
+matched samples no differences in bacterial community were observed when
+assessed with PERMANOVA {[}Table S6 \& S7{]}.
 

### 2018-01-09 submission/manuscript.Rmd
fix minor grammar errors in first result section

@@
 
-***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
-Power transforming and Z-score normalizing OTU richness, evenness, and Shannon diversity for the entire data set allowed for the combination of data from all studies to assess for differences in these broad scale community metrics as disease severity worsens.
-Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased.
-There was a significant decrease in both evenness and Shannon diversity from control to adenoma to carcinoma in stool (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A].
-We next tested whether these detectable differences in community resulted in significant increases in RR.
-For fecal samples, a decrease in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
+***Lower Bacterial Diversity is Associated with Increased RR of Carcinomas:***
+Power transforming and Z-score normalizing OTU richness, evenness, and Shannon diversity for the entire data set allowed for the combination of data from all studies to assess differences in broad scale community metrics as disease severity worsens.
+Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity.
+In stool, there was a significant decrease in both evenness and Shannon diversity from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A].
+We next tested whether these detectable differences in community resulted in significant increases in RR of lesion (adenoma or carcinoma).
+For fecal samples, a decrease versus the overal median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
 Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma versus control and adenoma versus control [Table S1 & S2].
@@
 Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased.
-For tissue there were no significant changes in OTU richness, evenness, and Shannon diversity from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1B & C].
+For tissue, there were no significant changes in OTU richness, evenness, and Shannon diversity from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1B & C].
 When analyzing the RR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) tissue samples, for both adenoma and carcinoma there was no significant change in RR based on lower than median values for OTU richness, evenness, and Shannon diversity [Table S3-S5].

### 2018-01-09 submission/manuscript.tex
fix minor grammar errors in first result section

@@
 
-\textbf{\emph{Lower Community Diversity is Associated with Increased RR
+\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
 of Carcinomas:}} Power transforming and Z-score normalizing OTU
 richness, evenness, and Shannon diversity for the entire data set
-allowed for the combination of data from all studies to assess for
-differences in these broad scale community metrics as disease severity
+allowed for the combination of data from all studies to assess
+differences in broad scale community metrics as disease severity
 worsens. Using linear mixed-effect models to control for study and
 variable region we assessed whether OTU richness, evenness, or Shannon
-diversity changed in a step-wise manner as disease severity increased.
-There was a significant decrease in both evenness and Shannon diversity
-from control to adenoma to carcinoma in stool (P-value = 0.025 and
-0.043, respectively) {[}Figure 1A{]}. We next tested whether these
-detectable differences in community resulted in significant increases in
-RR. For fecal samples, a decrease in evenness resulted in a
-significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
-P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
-0.032) while a decrease in Shannon diversity only increased the RR for
-carcinoma (RR = 1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}.
-Using the Bray-Curtis distance metric and PERMANOVA, it was also
-possible to identify significant bacterial community changes, in
-specific studies, for both carcinoma versus control and adenoma versus
-control {[}Table S1 \& S2{]}.
+diversity changed in a step-wise manner with disease severity. In stool,
+there was a significant decrease in both evenness and Shannon diversity
+from control to adenoma to carcinoma (P-value = 0.025 and 0.043,
+respectively) {[}Figure 1A{]}. We next tested whether these detectable
+differences in community resulted in significant increases in RR of
+lesion (adenoma or carcinoma). For fecal samples, a decrease versus the
+overal median in evenness resulted in a significantly increased RR for
+carcinoma (RR = 1.36 (1.15 - 1.61), P-value = 0.00034) and adenoma (RR =
+1.16 (1.01 - 1.34), P-value = 0.032) while a decrease versus the overal
+median in Shannon diversity only increased the RR for carcinoma (RR =
+1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}. Using the
+Bray-Curtis distance metric and PERMANOVA, it was also possible to
+identify significant bacterial community changes, in specific studies,
+for both carcinoma versus control and adenoma versus control {[}Table S1
+\& S2{]}.
 
@@
 OTU richness, evenness, or Shannon diversity changed in a step-wise
-manner as disease severity increased. For tissue there were no
+manner as disease severity increased. For tissue, there were no
 significant changes in OTU richness, evenness, and Shannon diversity

### 2018-01-09 submission/manuscript.Rmd
change intro final paragrpah

@@
 
-Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma.
-To accomplish this we first assessed whether diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma.
-We then assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the RR of adenoma or carcinoma.
-Next, using Random Forest models, we analyzed whether the full community or only the CRC-associated genera resulted in better model classification based on the area under the curve (AUC).
+Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor or if select members can recapitulate the same information for both adenoma and carcinoma.
+To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma.
+We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma.
+Next, using Random Forest models, we analyzed whether the full community or only the top 5 increased and top 5 decreased RR genera combined resulted in better model classification, based on the area under the curve (AUC).
 Our results suggest that the community changes as disease severity worsens and that this community is important for disease classification.
-However, since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for.
-Although we analyzed data sets that sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.
+Finally, we also examined at what effect and sample size the studies that were used were powered for and the sample size needed to get to an 80% power.
+Our power analysis suggests that each individual study used was underpowered for detecting effect size differences of 10% or below between the case and control groups.
 

### 2018-01-09 submission/manuscript.tex
change intro final paragrpah

@@
 of the previous meta-analysis to investigate whether the bacterial
-community is an important risk factor for both adenoma and carcinoma. To
-accomplish this we first assessed whether diversity changes throughout
-disease (control to adenoma to carcinoma) and if it results in an
-increased relative risk (RR) for adenoma or carcinoma. We then assessed
-how common CRC-associated genera (\emph{Fusobacterium},
-\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
-affect the RR of adenoma or carcinoma. Next, using Random Forest models,
-we analyzed whether the full community or only the CRC-associated genera
-resulted in better model classification based on the area under the
-curve (AUC). Our results suggest that the community changes as disease
-severity worsens and that this community is important for disease
-classification. However, since the changes in community were subtle for
-adenoma we also examined what effect and sample size the studies that
-were used were adequately powered for. Although we analyzed data sets
-that sampled large numbers of individuals, our results indicate the
-individual studies were underpowered for detecting effect size
+community is an important risk factor or if select members can
+recapitulate the same information for both adenoma and carcinoma. To
+accomplish this we first assessed whether bacterial diversity changes
+throughout disease (control to adenoma to carcinoma) and if it results
+in an increased relative risk (RR) for adenoma or carcinoma. We then
+assessed what genera, if any, increase or decrease the RR of adenoma or
+carcinoma. Next, using Random Forest models, we analyzed whether the
+full community or only the top 5 increased and top 5 decreased RR genera
+combined resulted in better model classification, based on the area
+under the curve (AUC). Our results suggest that the community changes as
+disease severity worsens and that this community is important for
+disease classification. Finally, we also examined at what effect and
+sample size the studies that were used were powered for and the sample
+size needed to get to an 80\% power. Our power analysis suggests that
+each individual study used was underpowered for detecting effect size
 differences of 10\% or below between the case and control groups.

### 2018-01-10 submission/manuscript.Rmd
rewrite background - add clearer picture of the confusion in current understanding

@@
 
+Although there has been an intense focus on microbiota based biomarker discovery for CRC, the candidate genera seem to be numerous.
+Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
+Yet, even in some of these studies mouth-associated genera are far from the only microbes identified to be associated with CRC.
+These other genera include, but are not necessarily limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012].
+In fact, *Escherichia/Shigella* also has good *in vivo* evidence that it could be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012].
+Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential additional markers of CRC with *Bacteroides fragilis* having good mechanistic studies into how this may occur [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014].
+Still, other studies have identified genera like *Roseburia* as increased in CRC while others have reported it to be decreased or have found no difference in any genera at all [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
+
+Most of these studies have focused on carcinoma but the adenoma results are not any clearer at identifying candidate genera.
+Some groups have focused on more broad scale community metrics and have found that metrics such as richness had decreased in adenoma versus controls while others have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
+A recent meta-analysis confirmed the associations of mouth-associated genera and *Akkermansia muciniphila* in carcinoma [@shah_leveraging_2017].
+However, it only analyzed a total of 509 stool samples and with some recent studies, that were not included, matching that sample size alone it is hard to know how extrapolatable these findings truly are.
+This meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) that need to be investigated further since a number of these genera have been found to be enriched in controls and not lesion (adenoma or carcinoma) [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].
+
 Targeting the identification of biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free.
@@
 
-This intense focus on identifying biomarkers has found strong associations with resident mouth microbes as potential CRC-associated microbes in the gastrointestinal (GI) tract [@kostic_genomic_2012; @baxter_microbiota-based_2016; @zeller_potential_2014].
-The main bacteria of interest, arising from this set of microbes, has been those within the *Fusobacterium* genus.
-Yet, the question remains as to whether or not this is indeed the most important genera to be focusing on, since many microbiota-based studies typically have identified a collection of oral microbes rather than single species from a single genera [@baxter_microbiota-based_2016; @zeller_potential_2014].
-Based on this discrepancy, the second question we can answer with this meta-analysis is if there is one dominant CRC-associated genera that can be identified across studies.
-
-The identification of microbial biomarkers has mostly dominated the study of CRC and has had an unattended consequence of reducing the focus on changes that occur within the underlying resident community.
-This has been borne out by the majority of previous studies, within stool, tissue, and the only meta-analysis investigating this area to date, that focus predominately on biomarker identification.
-Yet, how these CRC-associated microbes interact with their community may be an important component to consider, and has not been investigated in great detail by other meta-analyses.
-In response to this gap, our study aims to answer whether there are consistent detectable community differences as disease severity increases.
-
-In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, examines differences between stool and tissue microbiota in the context of CRC, and takes a more community centric approach rather than a biomarker focused approach to investigating commonalities across study for the microbiota and CRC.
-Importantly, this community centric approach could provide valuable insights into the importance of accounting for the community in CRC disease not previously provided by earlier meta-analysis studies [@shah_leveraging_2017].
+In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of CRC.
+Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not previously provided by earlier meta-analysis studies [@shah_leveraging_2017].
 

### 2018-01-10 submission/manuscript.tex
rewrite background - add clearer picture of the confusion in current understanding

@@
 
+Although there has been an intense focus on microbiota based biomarker
+discovery for CRC, the candidate genera seem to be numerous. Some
+studies point towards mouth-associated genera such as
+\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
+\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
+some of these studies mouth-associated genera are far from the only
+microbes identified to be associated with CRC. These other genera
+include, but are not necessarily limited to, \emph{Providencia},
+\emph{Mogibacterium}, \emph{Enterococcus}, \emph{Escherichia/Shigella},
+\emph{Klebsiella}, and \emph{Streptococcus} {[}14--16{]}. In fact,
+\emph{Escherichia/Shigella} also has good \emph{in vivo} evidence that
+it could be important in the pathogenesis of CRC {[}5,19{]}. Other
+studies have also identified \emph{Akkermansia muciniphila} and
+\emph{Bacteroides fragilis} as potential additional markers of CRC with
+\emph{Bacteroides fragilis} having good mechanistic studies into how
+this may occur {[}15,20,21{]}. Still, other studies have identified
+genera like \emph{Roseburia} as increased in CRC while others have
+reported it to be decreased or have found no difference in any genera at
+all {[}15,18,22,23{]}.
+
+Most of these studies have focused on carcinoma but the adenoma results
+are not any clearer at identifying candidate genera. Some groups have
+focused on more broad scale community metrics and have found that
+metrics such as richness had decreased in adenoma versus controls while
+others have identified \emph{Lactococcus}, \emph{Pseudomonas},
+\emph{Acidovorax}, \emph{Cloacibacterium}, \emph{Helicobacter},
+\emph{Lactobacillus}, \emph{Bilophila}, \emph{Desulfovibrio}, and
+\emph{Mogibacterium} to be increased in adenoma {[}24--26{]}. A recent
+meta-analysis confirmed the associations of mouth-associated genera and
+\emph{Akkermansia muciniphila} in carcinoma {[}11{]}. However, it only
+analyzed a total of 509 stool samples and with some recent studies, that
+were not included, matching that sample size alone it is hard to know
+how extrapolatable these findings truly are. This meta-analysis also
+added more potential microbial associations to both carcinoma
+(\emph{Pantoea agglomerans} \emph{Ruminococcus}, \emph{Lactobacillus})
+and adenoma (\emph{Prevotella}, \emph{Methanosphaera},
+\emph{Succinovibrio}, \emph{Haemophilus parainfluenzae},
+\emph{Ruminococcus}, \emph{Lactobacillus}) that need to be investigated
+further since a number of these genera have been found to be enriched in
+controls and not lesion (adenoma or carcinoma) {[}13,16,17{]}.
+
 Targeting the identification of biomarkers within stool seems logical
@@
 studies have tried to assess this in health and disease but are limited
-by their sample size {[}12,13{]}. Sampling the tissue directly may
+by their sample size {[}18,27{]}. Sampling the tissue directly may
 provide clearer answers but is not without limitations. Due to bowel
@@
 and results can end up being study dependent due to the randomness of
-contamination {[}14{]}. Due to these many differences that could arise
+contamination {[}28{]}. Due to these many differences that could arise
 between stool and tissue, one question our meta-analysis aims to answer
@@
 
-This intense focus on identifying biomarkers has found strong
-associations with resident mouth microbes as potential CRC-associated
-microbes in the gastrointestinal (GI) tract {[}15--17{]}. The main
-bacteria of interest, arising from this set of microbes, has been those
-within the \emph{Fusobacterium} genus. Yet, the question remains as to
-whether or not this is indeed the most important genera to be focusing
-on, since many microbiota-based studies typically have identified a
-collection of oral microbes rather than single species from a single
-genera {[}16,17{]}. Based on this discrepancy, the second question we
-can answer with this meta-analysis is if there is one dominant
-CRC-associated genera that can be identified across studies.
-
-The identification of microbial biomarkers has mostly dominated the
-study of CRC and has had an unattended consequence of reducing the focus
-on changes that occur within the underlying resident community. This has
-been borne out by the majority of previous studies, within stool,
-tissue, and the only meta-analysis investigating this area to date, that
-focus predominately on biomarker identification. Yet, how these
-CRC-associated microbes interact with their community may be an
-important component to consider, and has not been investigated in great
-detail by other meta-analyses. In response to this gap, our study aims
-to answer whether there are consistent detectable community differences
-as disease severity increases.
-
 In comparison to the previous meta-analysis, this study significantly
-increases the total stool samples investigated, examines differences
-between stool and tissue microbiota in the context of CRC, and takes a
-more community centric approach rather than a biomarker focused approach
-to investigating commonalities across study for the microbiota and CRC.
-Importantly, this community centric approach could provide valuable
-insights into the importance of accounting for the community in CRC
-disease not previously provided by earlier meta-analysis studies
-{[}11{]}.
+increases the total stool samples investigated, re-examines important
+genera across adenoma and carcinoma across study, and examines
+differences and similarities between stool and tissue microbiota in the
+context of CRC. Importantly, this analysis and approach could provide
+valuable insights into the common genera that are both protective and
+detrimental in CRC and whether broad bacterial community measurements
+can account for these changes that were not previously provided by
+earlier meta-analysis studies {[}11{]}.
 
@@
 individuals) totaling over 2229 total samples across 14 studies
-{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
-of the previous meta-analysis to investigate whether the bacterial
-community is an important risk factor or if select members can
+{[}12--18,20,22--26,29{]} {[}Table 1 \& 2{]}, we expand both the breadth
+and scope of the previous meta-analysis to investigate whether the
+bacterial community is an important risk factor or if select members can
 recapitulate the same information for both adenoma and carcinoma. To
@@
 meta-analysis were identified through the review articles written by
-Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}29,30{]} and
+Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}30,31{]} and
 additional studies not mentioned in the reviews were obtained based on
@@
 filtering steps, the following studies remained: Ahn, \emph{et al.}
-{[}26{]}, Baxter, \emph{et al.} {[}16{]}, Brim, \emph{et al.} {[}22{]},
-Burns, \emph{et al.} {[}27{]}, Chen, \emph{et al.} {[}19{]}, Dejea,
-\emph{et al.} {[}24{]}, Flemer, \emph{et al.} {[}12{]}, Geng, \emph{et
-al.} {[}28{]}, Hale, \emph{et al.} {[}18{]}, Kostic, \emph{et al.}
-{[}15{]}, Lu, \emph{et al.} {[}21{]}, Sanapareddy, \emph{et al.}
-{[}25{]}, Wang, \emph{et al.} {[}20{]}, Weir, \emph{et al.} {[}23{]},
-and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}31{]} study was not
+{[}12{]}, Baxter, \emph{et al.} {[}13{]}, Brim, \emph{et al.} {[}29{]},
+Burns, \emph{et al.} {[}16{]}, Chen, \emph{et al.} {[}14{]}, Dejea,
+\emph{et al.} {[}23{]}, Flemer, \emph{et al.} {[}18{]}, Geng, \emph{et
+al.} {[}22{]}, Hale, \emph{et al.} {[}26{]}, Kostic, \emph{et al.}
+{[}32{]}, Lu, \emph{et al.} {[}25{]}, Sanapareddy, \emph{et al.}
+{[}24{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}20{]},
+and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}33{]} study was not
 included because the 90 individuals analyzed within the study are
-contained within the larger Baxter study {[}16{]}. After sequence
+contained within the larger Baxter study {[}13{]}. After sequence
 processing all the case samples for the Kostic study had 100 or less
@@
 not have sequences and metadata on the SRA, data was obtained from DBGap
-(n = 1, {[}26{]}) and directly from the authors (n = 4,
-{[}12,18,23,25{]}). Each study was processed using the mothur (v1.39.3)
-software program {[}32{]} and quality filtering utilized the default
+(n = 1, {[}12{]}) and directly from the authors (n = 4,
+{[}18,20,24,26{]}). Each study was processed using the mothur (v1.39.3)
+software program {[}34{]} and quality filtering utilized the default
 methods, used in mothur, for either 454 or Illumina based sequencing. If
@@
 from the study itself, were used instead. Chimeras were identified and
-removed using VSEARCH {[}33{]} before \emph{de novo} OTU clustering at
-97\% similarity was completed using the OptiClust algorithm {[}34{]}.
+removed using VSEARCH {[}35{]} before \emph{de novo} OTU clustering at
+97\% similarity was completed using the OptiClust algorithm {[}36{]}.
 
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package {[}35{]}.
+sequence processing utilized the R (v3.4.3) software package {[}37{]}.
 For the \(\alpha\)-diversity analysis, values were power transformed
-using the rcompanion (v1.10.1) package {[}36{]} and then Z-score
-normalized using the car (v2.1.5) package {[}37{]}. Testing for
+using the rcompanion (v1.10.1) package {[}38{]} and then Z-score
+normalized using the car (v2.1.5) package {[}39{]}. Testing for
 \(\alpha\)-diversity differences utilized linear mixed-effect models
-created using the lme4 (v1.1.14) package {[}38{]} to correct for study,
+created using the lme4 (v1.1.14) package {[}40{]} to correct for study,
 repeat sampling of individuals (tissue only), and 16S hypervariable
 region used. Relative risk was analyzed using both the epiR (v0.9.87)
-and metafor (v2.0.0) packages {[}39,40{]} by assessing how many
+and metafor (v2.0.0) packages {[}41,42{]} by assessing how many
 individuals with and without disease were above and below the overall
@@
 utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
-vegan (v2.4.4) package {[}41{]}. Random Forest models were built using
+vegan (v2.4.4) package {[}43{]}. Random Forest models were built using
 both the caret (v6.0.77) and randomForest (v4.6.12) packages
-{[}42,43{]}. Power analysis and estimations were made using the pwr
-(v1.2.1) and statmod (v1.4.30) packages {[}44,45{]}. All figures were
+{[}44,45{]}. Power analysis and estimations were made using the pwr
+(v1.2.1) and statmod (v1.4.30) packages {[}46,47{]}. All figures were
 created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
-{[}46,47{]}.
+{[}48,49{]}.
 
@@
 
+\hypertarget{ref-ahn_human_2013}{}
+12. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
+microbiome and risk for colorectal cancer. Journal of the National
+Cancer Institute. 2013;105:1907--11.
+
+\hypertarget{ref-baxter_microbiota-based_2016}{}
+13. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+improves the sensitivity of fecal immunochemical test for detecting
+colonic lesions. Genome Medicine. 2016;8:37.
+
+\hypertarget{ref-chen_human_2012}{}
+14. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
+mucosa-associated microbiota in patients with colorectal cancer. PloS
+One. 2012;7:e39743.
+
+\hypertarget{ref-wang_structural_2012}{}
+15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
+segregation of gut microbiota between colorectal cancer patients and
+healthy volunteers. The ISME journal. 2012;6:320--9.
+
+\hypertarget{ref-burns_virulence_2015}{}
+16. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
+are a signature of the microbiome in the colorectal tumor
+microenvironment. Genome Medicine. 2015;7:55.
+
+\hypertarget{ref-zeller_potential_2014}{}
+17. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+Potential of fecal microbiota for early-stage detection of colorectal
+cancer. Molecular Systems Biology. 2014;10:766.
+
 \hypertarget{ref-flemer_tumour-associated_2017}{}
-12. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
+18. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
 al. Tumour-associated and non-tumour-associated microbiota in colorectal
@@
 
+\hypertarget{ref-ecoli_Arthur_2014}{}
+19. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
+McCafferty J, et al. Microbial genomic analysis reveals the essential
+role of inflammation in bacteria-induced colorectal cancer. Nature
+Communications {[}Internet{]}. Springer Nature; 2014;5:4724. Available
+from: \url{https://doi.org/10.1038/ncomms5724}
+
+\hypertarget{ref-weir_stool_2013}{}
+20. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
+Stool microbiome and metabolome differences between colorectal cancer
+patients and healthy adults. PloS One. 2013;8:e70803.
+
+\hypertarget{ref-bfrag_Boleij_2014}{}
+21. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
+MG, et al. The bacteroides fragilis toxin gene is prevalent in the colon
+mucosa of colorectal cancer patients. Clinical Infectious Diseases
+{[}Internet{]}. Oxford University Press (OUP); 2014;60:208--15.
+Available from: \url{https://doi.org/10.1093/cid/ciu787}
+
+\hypertarget{ref-geng_diversified_2013}{}
+22. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
+human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
+
+\hypertarget{ref-dejea_microbiota_2014}{}
+23. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+Rossetti BJ, et al. Microbiota organization is a distinct feature of
+proximal colorectal cancers. Proceedings of the National Academy of
+Sciences of the United States of America. 2014;111:18321--6.
+
+\hypertarget{ref-sanapareddy_increased_2012}{}
+24. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
+et al. Increased rectal microbial richness is associated with the
+presence of colorectal adenomas in humans. The ISME journal.
+2012;6:1858--68.
+
+\hypertarget{ref-lu_mucosal_2016}{}
+25. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
+adherent bacterial dysbiosis in patients with colorectal adenomas.
+Scientific Reports. 2016;6:26337.
+
+\hypertarget{ref-hale_shifts_2017}{}
+26. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
+Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
+Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
+American Association for Cancer Research, Cosponsored by the American
+Society of Preventive Oncology. 2017;26:85--94.
+
 \hypertarget{ref-Flynn_preprint_2017}{}
-13. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
+27. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
 the native colon microbiota in healthy adults. Cold Spring Harbor
@@
 \hypertarget{ref-Salter_contamination_2014}{}
-14. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
+28. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
 al. Reagent and laboratory contamination can critically impact
@@
 
-\hypertarget{ref-kostic_genomic_2012}{}
-15. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
-al. Genomic analysis identifies association of Fusobacterium with
-colorectal carcinoma. Genome Research. 2012;22:292--8.
-
-\hypertarget{ref-baxter_microbiota-based_2016}{}
-16. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
-improves the sensitivity of fecal immunochemical test for detecting
-colonic lesions. Genome Medicine. 2016;8:37.
-
-\hypertarget{ref-zeller_potential_2014}{}
-17. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
-Potential of fecal microbiota for early-stage detection of colorectal
-cancer. Molecular Systems Biology. 2014;10:766.
-
-\hypertarget{ref-hale_shifts_2017}{}
-18. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
-Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
-Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
-American Association for Cancer Research, Cosponsored by the American
-Society of Preventive Oncology. 2017;26:85--94.
-
-\hypertarget{ref-chen_human_2012}{}
-19. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
-mucosa-associated microbiota in patients with colorectal cancer. PloS
-One. 2012;7:e39743.
-
-\hypertarget{ref-wang_structural_2012}{}
-20. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
-segregation of gut microbiota between colorectal cancer patients and
-healthy volunteers. The ISME journal. 2012;6:320--9.
-
-\hypertarget{ref-lu_mucosal_2016}{}
-21. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
-adherent bacterial dysbiosis in patients with colorectal adenomas.
-Scientific Reports. 2016;6:26337.
-
 \hypertarget{ref-brim_microbiome_2013}{}
-22. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
+29. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
 al. Microbiome analysis of stool samples from African Americans with
@@
 
-\hypertarget{ref-weir_stool_2013}{}
-23. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
-Stool microbiome and metabolome differences between colorectal cancer
-patients and healthy adults. PloS One. 2013;8:e70803.
-
-\hypertarget{ref-dejea_microbiota_2014}{}
-24. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
-Rossetti BJ, et al. Microbiota organization is a distinct feature of
-proximal colorectal cancers. Proceedings of the National Academy of
-Sciences of the United States of America. 2014;111:18321--6.
-
-\hypertarget{ref-sanapareddy_increased_2012}{}
-25. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
-et al. Increased rectal microbial richness is associated with the
-presence of colorectal adenomas in humans. The ISME journal.
-2012;6:1858--68.
-
-\hypertarget{ref-ahn_human_2013}{}
-26. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
-microbiome and risk for colorectal cancer. Journal of the National
-Cancer Institute. 2013;105:1907--11.
-
-\hypertarget{ref-burns_virulence_2015}{}
-27. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
-are a signature of the microbiome in the colorectal tumor
-microenvironment. Genome Medicine. 2015;7:55.
-
-\hypertarget{ref-geng_diversified_2013}{}
-28. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
-human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
-
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-29. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
+30. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
 microbiota and colorectal cancer. American Journal of Physiology -
@@
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-30. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
+31. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
 microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
@@
 
+\hypertarget{ref-kostic_genomic_2012}{}
+32. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
+al. Genomic analysis identifies association of Fusobacterium with
+colorectal carcinoma. Genome Research. 2012;22:292--8.
+
 \hypertarget{ref-zackular_human_2014}{}
-31. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+33. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-32. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+34. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, et al. Introducing mothur: Open-Source, Platform-Independent,
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-33. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
+35. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
 versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-34. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
+36. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
 Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
@@
 \hypertarget{ref-r_citation_2017}{}
-35. R Core Team. R: A language and environment for statistical computing
+37. R Core Team. R: A language and environment for statistical computing
 {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-36. Mangiafico S. Rcompanion: Functions to support extension education
+38. Mangiafico S. Rcompanion: Functions to support extension education
 program evaluation {[}Internet{]}. 2017. Available from:
@@
 \hypertarget{ref-car_citation_2011}{}
-37. Fox J, Weisberg S. An R companion to applied regression
+39. Fox J, Weisberg S. An R companion to applied regression
 {[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
@@
 \hypertarget{ref-lme4_citation_2015}{}
-38. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
+40. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
 models using lme4. Journal of Statistical Software. 2015;67:1--48.
@@
 \hypertarget{ref-epir_citation_2017}{}
-39. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
+41. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
 J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
@@
 \hypertarget{ref-metafor_citation_2010}{}
-40. Viechtbauer W. Conducting meta-analyses in R with the metafor
+42. Viechtbauer W. Conducting meta-analyses in R with the metafor
 package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-41. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+43. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
@@
 \hypertarget{ref-caret_citation_2017}{}
-42. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
+44. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
 Cooper T, et al. Caret: Classification and regression training
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-43. Liaw A, Wiener M. Classification and regression by randomForest. R
+45. Liaw A, Wiener M. Classification and regression by randomForest. R
 News {[}Internet{]}. 2002;2:18--22. Available from:
@@
 \hypertarget{ref-pwr_citation_2017}{}
-44. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
+46. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
 2017. Available from: \url{https://CRAN.R-project.org/package=pwr}
@@
 \hypertarget{ref-statmod_citation_2016}{}
-45. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
+47. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
 gaussian distribution. R Journal. 2016;8:339--51.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-46. Wickham H. Ggplot2: Elegant graphics for data analysis
+48. Wickham H. Ggplot2: Elegant graphics for data analysis
 {[}Internet{]}. Springer-Verlag New York; 2009. Available from:
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-47. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
+49. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
 {[}Internet{]}. 2017. Available from:

### 2018-01-10 submission/manuscript.Rmd
update discussion to include caveat on previously identified species

@@
 The low bacterial biomass of tissue samples coupled with potential contaminationcould ultimately explain why the these results seem to be more sporadic than the stool results.
+
+One important caveat to this study is that even though specific genera associated with species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp.
+*gallolyticus* were not identified, it does not necessarily mean these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017].
+Since we are limited in our aggregation of the data, across study, to the genus level, it is not possible to clearly delineate what specific species are contributing to overall disease progression.
+However, *Streptococcus gallolyticus* subsp.
+*gallolyticus* is a mouth-associated micobe and our study clearly shows that regardless of whether the sample is tissue or stool, mouth-associated genera are commonly associated with an increased RR of both adenoma and carcinoma.
 

### 2018-01-10 submission/manuscript.tex
update discussion to include caveat on previously identified species

@@
 
+One important caveat to this study is that even though specific genera
+associated with species such as \emph{Bacteroides fragilis} and
+\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
+identified, it does not necessarily mean these specific species are not
+important in human CRC {[}21,30{]}. Since we are limited in our
+aggregation of the data, across study, to the genus level, it is not
+possible to clearly delineate what specific species are contributing to
+overall disease progression. However, \emph{Streptococcus gallolyticus}
+subsp. \emph{gallolyticus} is a mouth-associated micobe and our study
+clearly shows that regardless of whether the sample is tissue or stool,
+mouth-associated genera are commonly associated with an increased RR of
+both adenoma and carcinoma.
+
 The associations between the microbiota and adenoma are inconclusive, in
@@
 meta-analysis were identified through the review articles written by
-Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}30,31{]} and
+Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}31,32{]} and
 additional studies not mentioned in the reviews were obtained based on
@@
 al.} {[}22{]}, Hale, \emph{et al.} {[}26{]}, Kostic, \emph{et al.}
-{[}32{]}, Lu, \emph{et al.} {[}25{]}, Sanapareddy, \emph{et al.}
+{[}33{]}, Lu, \emph{et al.} {[}25{]}, Sanapareddy, \emph{et al.}
 {[}24{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}20{]},
-and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}33{]} study was not
+and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}34{]} study was not
 included because the 90 individuals analyzed within the study are
@@
 {[}18,20,24,26{]}). Each study was processed using the mothur (v1.39.3)
-software program {[}34{]} and quality filtering utilized the default
+software program {[}35{]} and quality filtering utilized the default
 methods, used in mothur, for either 454 or Illumina based sequencing. If
@@
 from the study itself, were used instead. Chimeras were identified and
-removed using VSEARCH {[}35{]} before \emph{de novo} OTU clustering at
-97\% similarity was completed using the OptiClust algorithm {[}36{]}.
+removed using VSEARCH {[}36{]} before \emph{de novo} OTU clustering at
+97\% similarity was completed using the OptiClust algorithm {[}37{]}.
 
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package {[}37{]}.
+sequence processing utilized the R (v3.4.3) software package {[}38{]}.
 For the \(\alpha\)-diversity analysis, values were power transformed
-using the rcompanion (v1.10.1) package {[}38{]} and then Z-score
-normalized using the car (v2.1.5) package {[}39{]}. Testing for
+using the rcompanion (v1.10.1) package {[}39{]} and then Z-score
+normalized using the car (v2.1.5) package {[}40{]}. Testing for
 \(\alpha\)-diversity differences utilized linear mixed-effect models
-created using the lme4 (v1.1.14) package {[}40{]} to correct for study,
+created using the lme4 (v1.1.14) package {[}41{]} to correct for study,
 repeat sampling of individuals (tissue only), and 16S hypervariable
 region used. Relative risk was analyzed using both the epiR (v0.9.87)
-and metafor (v2.0.0) packages {[}41,42{]} by assessing how many
+and metafor (v2.0.0) packages {[}42,43{]} by assessing how many
 individuals with and without disease were above and below the overall
@@
 utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
-vegan (v2.4.4) package {[}43{]}. Random Forest models were built using
+vegan (v2.4.4) package {[}44{]}. Random Forest models were built using
 both the caret (v6.0.77) and randomForest (v4.6.12) packages
-{[}44,45{]}. Power analysis and estimations were made using the pwr
-(v1.2.1) and statmod (v1.4.30) packages {[}46,47{]}. All figures were
+{[}45,46{]}. Power analysis and estimations were made using the pwr
+(v1.2.1) and statmod (v1.4.30) packages {[}47,48{]}. All figures were
 created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
-{[}48,49{]}.
+{[}49,50{]}.
 
@@
 
+\hypertarget{ref-strep_Aymeric_2017}{}
+30. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
+et al. Colorectal cancer specific conditions promoteStreptococcus
+gallolyticusgut colonization. Proceedings of the National Academy of
+Sciences {[}Internet{]}. Proceedings of the National Academy of
+Sciences; 2017;115:E283--91. Available from:
+\url{https://doi.org/10.1073/pnas.1715112115}
+
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-30. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
+31. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
 microbiota and colorectal cancer. American Journal of Physiology -
@@
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-31. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
+32. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
 microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
@@
 \hypertarget{ref-kostic_genomic_2012}{}
-32. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
+33. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
 al. Genomic analysis identifies association of Fusobacterium with
@@
 \hypertarget{ref-zackular_human_2014}{}
-33. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+34. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-34. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+35. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, et al. Introducing mothur: Open-Source, Platform-Independent,
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-35. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
+36. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
 versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-36. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
+37. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
 Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
@@
 \hypertarget{ref-r_citation_2017}{}
-37. R Core Team. R: A language and environment for statistical computing
+38. R Core Team. R: A language and environment for statistical computing
 {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-38. Mangiafico S. Rcompanion: Functions to support extension education
+39. Mangiafico S. Rcompanion: Functions to support extension education
 program evaluation {[}Internet{]}. 2017. Available from:
@@
 \hypertarget{ref-car_citation_2011}{}
-39. Fox J, Weisberg S. An R companion to applied regression
+40. Fox J, Weisberg S. An R companion to applied regression
 {[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
@@
 \hypertarget{ref-lme4_citation_2015}{}
-40. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
+41. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
 models using lme4. Journal of Statistical Software. 2015;67:1--48.
@@
 \hypertarget{ref-epir_citation_2017}{}
-41. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
+42. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
 J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
@@
 \hypertarget{ref-metafor_citation_2010}{}
-42. Viechtbauer W. Conducting meta-analyses in R with the metafor
+43. Viechtbauer W. Conducting meta-analyses in R with the metafor
 package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-43. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+44. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
@@
 \hypertarget{ref-caret_citation_2017}{}
-44. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
+45. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
 Cooper T, et al. Caret: Classification and regression training
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-45. Liaw A, Wiener M. Classification and regression by randomForest. R
+46. Liaw A, Wiener M. Classification and regression by randomForest. R
 News {[}Internet{]}. 2002;2:18--22. Available from:
@@
 \hypertarget{ref-pwr_citation_2017}{}
-46. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
+47. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
 2017. Available from: \url{https://CRAN.R-project.org/package=pwr}
@@
 \hypertarget{ref-statmod_citation_2016}{}
-47. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
+48. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
 gaussian distribution. R Journal. 2016;8:339--51.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-48. Wickham H. Ggplot2: Elegant graphics for data analysis
+49. Wickham H. Ggplot2: Elegant graphics for data analysis
 {[}Internet{]}. Springer-Verlag New York; 2009. Available from:
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-49. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
+50. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
 {[}Internet{]}. 2017. Available from:

### 2018-01-10 submission/manuscript.Rmd
change conclusion to match better with result changes

@@
 By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in CRC.
-Further, the data presented here suggests that non-resident mouth-associated microbes can gain a foothold within the colon and the more of these microbes that are present the greater the RR of carcinoma.
-Conversely, no clear signal with these mouth-associated micobes could be detected for adenoma.
+Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest RR of carcinoma.
+Conversely, no conclusive signal with these mouth-associated micobes could be detected for adenoma.
 Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community.

### 2018-01-10 submission/manuscript.tex
change conclusion to match better with result changes

@@
 of the bacterial community in CRC. Further, the data presented here
-suggests that non-resident mouth-associated microbes can gain a foothold
-within the colon and the more of these microbes that are present the
-greater the RR of carcinoma. Conversely, no clear signal with these
-mouth-associated micobes could be detected for adenoma. Our observations
-also highlight the importance of power and sample number considerations
-when undertaking investigations into the microbiota and adenoma due to
-the subtle changes in the community. Overall, the microbiota
-associations with carcinoma are much stronger than the associations with
-adenoma.
+suggests that mouth-associated microbes can gain a foothold within the
+colon and are are commonly associated with the greatest RR of carcinoma.
+Conversely, no conclusive signal with these mouth-associated micobes
+could be detected for adenoma. Our observations also highlight the
+importance of power and sample number considerations when undertaking
+investigations into the microbiota and adenoma due to the subtle changes
+in the community. Overall, the microbiota associations with carcinoma
+are much stronger than the associations with adenoma.
 

### 2018-01-10 submission/manuscript.Rmd
update methods section to include top 5 component in workflow

@@
 $\beta$-diversity was then assessed for each individual study for differences between control-adenoma and control-carcinoma.
-Next, four specific CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*) were assessed for differences in RR for adenoma and carcinoma.
-We then built Random Forest models based on all or CRC-associated genera and these models were trained on one study then tested on the remaining studies, for every study.
+Next, all common genera were assessed for differences in RR for adenoma and carcinoma.
+We then built Random Forest models based on all or a selected model, based on the top 5 increased and top 5 decreased RR based on P-value, and these models were trained on one study then tested on the remaining studies, for every study.
 A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
-For these OTU based models, the CRC-associated genera included all OTUs that had a taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*.
+For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a variable in the top 5 increased and top 5 decreased RR based on P-value.
 Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power.
@@
 
-***Obtaining CRC-associated Genera:***
-For the CRC-associated genera analysis of the RR, the total average counts, for 100 different subsamplings, were collected for each respective OTU that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.
-The OTU based Random Forest Models using CRC-associated genera utilized a similar approach except that the OTUs were not aggregated together by genus but kept as separate OTUs.
-OTU Random Forest models using the full community included all OTUs while those using CRC-associated genera included only those OTUs that had a genus level taxonomic classification to *Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, and *Porphyromonas*.
+***Obtaining Genera Relative Abundance and Selected Models:***
+For the genera analysis of the RR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings, were collected.
+The OTU based Random Forest Models using selected OTUS utilized a similar approach except that the OTUs were not aggregated together by taxonomic identity but kept as separate OTUs.
+OTU Random Forest models using the full community included all OTUs while those for the selected model included only those OTUs that had a taxonomic classification to a variable in the top 5 increased of top 5 decreased RR based on P-value.
 

### 2018-01-10 submission/manuscript.tex
update methods section to include top 5 component in workflow

@@
 assessed for each individual study for differences between
-control-adenoma and control-carcinoma. Next, four specific
-CRC-associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Peptostreptococcus}, and \emph{Porphyromonas}) were assessed for
-differences in RR for adenoma and carcinoma. We then built Random Forest
-models based on all or CRC-associated genera and these models were
-trained on one study then tested on the remaining studies, for every
-study. A similar approach was then applied at the OTU level with the
-exception that a 10-fold CV over 100 different models, based on random
-80/20 splitting of the data, was used to generate a range of expected
-AUCs. For these OTU based models, the CRC-associated genera included all
-OTUs that had a taxonomic classification to \emph{Fusobacterium},
-\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas}.
-Finally, the power of each study was assessed for an effect size ranging
-from 1\% to 30\% and an estimated sample size, for these effect sizes,
-was generated based on 80\% power. For comparisons in which normal
-versus adenoma were made the carcinoma samples were excluded from each
-respective study. Similarly, for comparisons in which normal versus
-carcinoma were made the adenoma samples were excluded from each
-respective study. The data was split between feces and tissue samples.
-Within the tissue groups the data was further divided between samples
-from the same individual (matched) and those from different individuals
-(unmatched). Where applicable for each study, predictions for adenoma
-and carcinoma were then tested for feces, matched tissue, and unmatched
-tissue.
-
-\textbf{\emph{Obtaining CRC-associated Genera:}} For the CRC-associated
-genera analysis of the RR, the total average counts, for 100 different
-subsamplings, were collected for each respective OTU that had a genus
-level taxonomic classification to \emph{Fusobacterium},
-\emph{Parvimonas}, \emph{Peptostreptococcus}, and \emph{Porphyromonas}.
-The OTU based Random Forest Models using CRC-associated genera utilized
-a similar approach except that the OTUs were not aggregated together by
-genus but kept as separate OTUs. OTU Random Forest models using the full
-community included all OTUs while those using CRC-associated genera
-included only those OTUs that had a genus level taxonomic classification
-to \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Peptostreptococcus},
-and \emph{Porphyromonas}.
+control-adenoma and control-carcinoma. Next, all common genera were
+assessed for differences in RR for adenoma and carcinoma. We then built
+Random Forest models based on all or a selected model, based on the top
+5 increased and top 5 decreased RR based on P-value, and these models
+were trained on one study then tested on the remaining studies, for
+every study. A similar approach was then applied at the OTU level with
+the exception that a 10-fold CV over 100 different models, based on
+random 80/20 splitting of the data, was used to generate a range of
+expected AUCs. For these OTU-based models, the selected model included
+all OTUs that had a taxonomic classification to a variable in the top 5
+increased and top 5 decreased RR based on P-value. Finally, the power of
+each study was assessed for an effect size ranging from 1\% to 30\% and
+an estimated sample size, for these effect sizes, was generated based on
+80\% power. For comparisons in which normal versus adenoma were made the
+carcinoma samples were excluded from each respective study. Similarly,
+for comparisons in which normal versus carcinoma were made the adenoma
+samples were excluded from each respective study. The data was split
+between feces and tissue samples. Within the tissue groups the data was
+further divided between samples from the same individual (matched) and
+those from different individuals (unmatched). Where applicable for each
+study, predictions for adenoma and carcinoma were then tested for feces,
+matched tissue, and unmatched tissue.
+
+\textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
+For the genera analysis of the RR, OTUs were added together based on the
+genus or lowest available taxonomic classification level and the total
+average counts, for 100 different subsamplings, were collected. The OTU
+based Random Forest Models using selected OTUS utilized a similar
+approach except that the OTUs were not aggregated together by taxonomic
+identity but kept as separate OTUs. OTU Random Forest models using the
+full community included all OTUs while those for the selected model
+included only those OTUs that had a taxonomic classification to a
+variable in the top 5 increased of top 5 decreased RR based on P-value.
 

### 2018-01-10 submission/manuscript.Rmd
correct one small missing statement in methods
fix grammar errors in intro updates

@@
 These other genera include, but are not necessarily limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012].
-In fact, *Escherichia/Shigella* also has good *in vivo* evidence that it could be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012].
+In fact, *Escherichia/Shigella* and *Streptococcus* also have good *in vivo* evidence that individual species within the genus can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017].
 Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential additional markers of CRC with *Bacteroides fragilis* having good mechanistic studies into how this may occur [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014].
@@
 
-Most of these studies have focused on carcinoma but the adenoma results are not any clearer at identifying candidate genera.
-Some groups have focused on more broad scale community metrics and have found that metrics such as richness had decreased in adenoma versus controls while others have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
-A recent meta-analysis confirmed the associations of mouth-associated genera and *Akkermansia muciniphila* in carcinoma [@shah_leveraging_2017].
-However, it only analyzed a total of 509 stool samples and with some recent studies, that were not included, matching that sample size alone it is hard to know how extrapolatable these findings truly are.
-This meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) that need to be investigated further since a number of these genera have been found to be enriched in controls and not lesion (adenoma or carcinoma) [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].
+Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera.
+Some groups have focused on more broad scale community metrics and have found that metrics such as richness are decreased in adenoma versus controls while others have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
+A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017].
+However, it only analyzed a total of 509 stool samples and with some recent studies, that were not included, matching this sample size alone, it is hard to know how extrapolatable these findings truly are.
+That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) that need to be investigated further, since a number of these genera have been found to be enriched in controls and not lesion (adenoma or carcinoma) [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].
 
@@
 $\beta$-diversity was then assessed for each individual study for differences between control-adenoma and control-carcinoma.
-Next, all common genera were assessed for differences in RR for adenoma and carcinoma.
+Next, all common genera were assessed for differences in RR for adenoma and carcinoma and ranked based on P-value.
 We then built Random Forest models based on all or a selected model, based on the top 5 increased and top 5 decreased RR based on P-value, and these models were trained on one study then tested on the remaining studies, for every study.

### 2018-01-10 submission/manuscript.tex
correct one small missing statement in methods
fix grammar errors in intro updates

@@
 \emph{Klebsiella}, and \emph{Streptococcus} {[}14--16{]}. In fact,
-\emph{Escherichia/Shigella} also has good \emph{in vivo} evidence that
-it could be important in the pathogenesis of CRC {[}5,19{]}. Other
-studies have also identified \emph{Akkermansia muciniphila} and
-\emph{Bacteroides fragilis} as potential additional markers of CRC with
-\emph{Bacteroides fragilis} having good mechanistic studies into how
-this may occur {[}15,20,21{]}. Still, other studies have identified
-genera like \emph{Roseburia} as increased in CRC while others have
-reported it to be decreased or have found no difference in any genera at
-all {[}15,18,22,23{]}.
-
-Most of these studies have focused on carcinoma but the adenoma results
-are not any clearer at identifying candidate genera. Some groups have
-focused on more broad scale community metrics and have found that
-metrics such as richness had decreased in adenoma versus controls while
-others have identified \emph{Lactococcus}, \emph{Pseudomonas},
+\emph{Escherichia/Shigella} and \emph{Streptococcus} also have good
+\emph{in vivo} evidence that individual species within the genus can be
+important in the pathogenesis of CRC {[}5,19,20{]}. Other studies have
+also identified \emph{Akkermansia muciniphila} and \emph{Bacteroides
+fragilis} as potential additional markers of CRC with \emph{Bacteroides
+fragilis} having good mechanistic studies into how this may occur
+{[}15,21,22{]}. Still, other studies have identified genera like
+\emph{Roseburia} as increased in CRC while others have reported it to be
+decreased or have found no difference in any genera at all
+{[}15,18,23,24{]}.
+
+Most of these studies have focused on carcinoma but the adenoma
+observations are not any clearer at identifying candidate genera. Some
+groups have focused on more broad scale community metrics and have found
+that metrics such as richness are decreased in adenoma versus controls
+while others have identified \emph{Lactococcus}, \emph{Pseudomonas},
 \emph{Acidovorax}, \emph{Cloacibacterium}, \emph{Helicobacter},
 \emph{Lactobacillus}, \emph{Bilophila}, \emph{Desulfovibrio}, and
-\emph{Mogibacterium} to be increased in adenoma {[}24--26{]}. A recent
-meta-analysis confirmed the associations of mouth-associated genera and
-\emph{Akkermansia muciniphila} in carcinoma {[}11{]}. However, it only
-analyzed a total of 509 stool samples and with some recent studies, that
-were not included, matching that sample size alone it is hard to know
-how extrapolatable these findings truly are. This meta-analysis also
-added more potential microbial associations to both carcinoma
-(\emph{Pantoea agglomerans} \emph{Ruminococcus}, \emph{Lactobacillus})
-and adenoma (\emph{Prevotella}, \emph{Methanosphaera},
-\emph{Succinovibrio}, \emph{Haemophilus parainfluenzae},
-\emph{Ruminococcus}, \emph{Lactobacillus}) that need to be investigated
-further since a number of these genera have been found to be enriched in
-controls and not lesion (adenoma or carcinoma) {[}13,16,17{]}.
+\emph{Mogibacterium} to be increased in adenoma {[}25--27{]}. A recent
+meta-analysis confirmed the correlations of certain mouth-associated
+genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
+However, it only analyzed a total of 509 stool samples and with some
+recent studies, that were not included, matching this sample size alone,
+it is hard to know how extrapolatable these findings truly are. That
+particular meta-analysis also added more potential microbial
+associations to both carcinoma (\emph{Pantoea agglomerans}
+\emph{Ruminococcus}, \emph{Lactobacillus}) and adenoma
+(\emph{Prevotella}, \emph{Methanosphaera}, \emph{Succinovibrio},
+\emph{Haemophilus parainfluenzae}, \emph{Ruminococcus},
+\emph{Lactobacillus}) that need to be investigated further, since a
+number of these genera have been found to be enriched in controls and
+not lesion (adenoma or carcinoma) {[}13,16,17{]}.
 
@@
 studies have tried to assess this in health and disease but are limited
-by their sample size {[}18,27{]}. Sampling the tissue directly may
+by their sample size {[}18,28{]}. Sampling the tissue directly may
 provide clearer answers but is not without limitations. Due to bowel
@@
 and results can end up being study dependent due to the randomness of
-contamination {[}28{]}. Due to these many differences that could arise
+contamination {[}29{]}. Due to these many differences that could arise
 between stool and tissue, one question our meta-analysis aims to answer
@@
 individuals) totaling over 2229 total samples across 14 studies
-{[}12--18,20,22--26,29{]} {[}Table 1 \& 2{]}, we expand both the breadth
+{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
 and scope of the previous meta-analysis to investigate whether the
@@
 identified, it does not necessarily mean these specific species are not
-important in human CRC {[}21,30{]}. Since we are limited in our
+important in human CRC {[}20,22{]}. Since we are limited in our
 aggregation of the data, across study, to the genus level, it is not
@@
 filtering steps, the following studies remained: Ahn, \emph{et al.}
-{[}12{]}, Baxter, \emph{et al.} {[}13{]}, Brim, \emph{et al.} {[}29{]},
+{[}12{]}, Baxter, \emph{et al.} {[}13{]}, Brim, \emph{et al.} {[}30{]},
 Burns, \emph{et al.} {[}16{]}, Chen, \emph{et al.} {[}14{]}, Dejea,
-\emph{et al.} {[}23{]}, Flemer, \emph{et al.} {[}18{]}, Geng, \emph{et
-al.} {[}22{]}, Hale, \emph{et al.} {[}26{]}, Kostic, \emph{et al.}
-{[}33{]}, Lu, \emph{et al.} {[}25{]}, Sanapareddy, \emph{et al.}
-{[}24{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}20{]},
+\emph{et al.} {[}24{]}, Flemer, \emph{et al.} {[}18{]}, Geng, \emph{et
+al.} {[}23{]}, Hale, \emph{et al.} {[}27{]}, Kostic, \emph{et al.}
+{[}33{]}, Lu, \emph{et al.} {[}26{]}, Sanapareddy, \emph{et al.}
+{[}25{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}21{]},
 and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}34{]} study was not
@@
 (n = 1, {[}12{]}) and directly from the authors (n = 4,
-{[}18,20,24,26{]}). Each study was processed using the mothur (v1.39.3)
+{[}18,21,25,27{]}). Each study was processed using the mothur (v1.39.3)
 software program {[}35{]} and quality filtering utilized the default
@@
 control-adenoma and control-carcinoma. Next, all common genera were
-assessed for differences in RR for adenoma and carcinoma. We then built
-Random Forest models based on all or a selected model, based on the top
-5 increased and top 5 decreased RR based on P-value, and these models
-were trained on one study then tested on the remaining studies, for
-every study. A similar approach was then applied at the OTU level with
-the exception that a 10-fold CV over 100 different models, based on
-random 80/20 splitting of the data, was used to generate a range of
-expected AUCs. For these OTU-based models, the selected model included
-all OTUs that had a taxonomic classification to a variable in the top 5
-increased and top 5 decreased RR based on P-value. Finally, the power of
-each study was assessed for an effect size ranging from 1\% to 30\% and
-an estimated sample size, for these effect sizes, was generated based on
-80\% power. For comparisons in which normal versus adenoma were made the
-carcinoma samples were excluded from each respective study. Similarly,
-for comparisons in which normal versus carcinoma were made the adenoma
-samples were excluded from each respective study. The data was split
-between feces and tissue samples. Within the tissue groups the data was
-further divided between samples from the same individual (matched) and
-those from different individuals (unmatched). Where applicable for each
-study, predictions for adenoma and carcinoma were then tested for feces,
-matched tissue, and unmatched tissue.
+assessed for differences in RR for adenoma and carcinoma and ranked
+based on P-value. We then built Random Forest models based on all or a
+selected model, based on the top 5 increased and top 5 decreased RR
+based on P-value, and these models were trained on one study then tested
+on the remaining studies, for every study. A similar approach was then
+applied at the OTU level with the exception that a 10-fold CV over 100
+different models, based on random 80/20 splitting of the data, was used
+to generate a range of expected AUCs. For these OTU-based models, the
+selected model included all OTUs that had a taxonomic classification to
+a variable in the top 5 increased and top 5 decreased RR based on
+P-value. Finally, the power of each study was assessed for an effect
+size ranging from 1\% to 30\% and an estimated sample size, for these
+effect sizes, was generated based on 80\% power. For comparisons in
+which normal versus adenoma were made the carcinoma samples were
+excluded from each respective study. Similarly, for comparisons in which
+normal versus carcinoma were made the adenoma samples were excluded from
+each respective study. The data was split between feces and tissue
+samples. Within the tissue groups the data was further divided between
+samples from the same individual (matched) and those from different
+individuals (unmatched). Where applicable for each study, predictions
+for adenoma and carcinoma were then tested for feces, matched tissue,
+and unmatched tissue.
 
@@
 
-\hypertarget{ref-weir_stool_2013}{}
-20. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
-Stool microbiome and metabolome differences between colorectal cancer
-patients and healthy adults. PloS One. 2013;8:e70803.
-
-\hypertarget{ref-bfrag_Boleij_2014}{}
-21. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
-MG, et al. The bacteroides fragilis toxin gene is prevalent in the colon
-mucosa of colorectal cancer patients. Clinical Infectious Diseases
-{[}Internet{]}. Oxford University Press (OUP); 2014;60:208--15.
-Available from: \url{https://doi.org/10.1093/cid/ciu787}
-
-\hypertarget{ref-geng_diversified_2013}{}
-22. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
-human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
-
-\hypertarget{ref-dejea_microbiota_2014}{}
-23. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
-Rossetti BJ, et al. Microbiota organization is a distinct feature of
-proximal colorectal cancers. Proceedings of the National Academy of
-Sciences of the United States of America. 2014;111:18321--6.
-
-\hypertarget{ref-sanapareddy_increased_2012}{}
-24. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
-et al. Increased rectal microbial richness is associated with the
-presence of colorectal adenomas in humans. The ISME journal.
-2012;6:1858--68.
-
-\hypertarget{ref-lu_mucosal_2016}{}
-25. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
-adherent bacterial dysbiosis in patients with colorectal adenomas.
-Scientific Reports. 2016;6:26337.
-
-\hypertarget{ref-hale_shifts_2017}{}
-26. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
-Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
-Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
-American Association for Cancer Research, Cosponsored by the American
-Society of Preventive Oncology. 2017;26:85--94.
-
-\hypertarget{ref-Flynn_preprint_2017}{}
-27. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
-the native colon microbiota in healthy adults. Cold Spring Harbor
-Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}
-
-\hypertarget{ref-Salter_contamination_2014}{}
-28. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
-al. Reagent and laboratory contamination can critically impact
-sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
-Nature; 2014;12. Available from:
-\url{https://doi.org/10.1186/s12915-014-0087-z}
-
-\hypertarget{ref-brim_microbiome_2013}{}
-29. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
-al. Microbiome analysis of stool samples from African Americans with
-colon polyps. PloS One. 2013;8:e81352.
-
 \hypertarget{ref-strep_Aymeric_2017}{}
-30. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
+20. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
 et al. Colorectal cancer specific conditions promoteStreptococcus
@@
 \url{https://doi.org/10.1073/pnas.1715112115}
+
+\hypertarget{ref-weir_stool_2013}{}
+21. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
+Stool microbiome and metabolome differences between colorectal cancer
+patients and healthy adults. PloS One. 2013;8:e70803.
+
+\hypertarget{ref-bfrag_Boleij_2014}{}
+22. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
+MG, et al. The bacteroides fragilis toxin gene is prevalent in the colon
+mucosa of colorectal cancer patients. Clinical Infectious Diseases
+{[}Internet{]}. Oxford University Press (OUP); 2014;60:208--15.
+Available from: \url{https://doi.org/10.1093/cid/ciu787}
+
+\hypertarget{ref-geng_diversified_2013}{}
+23. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
+human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
+
+\hypertarget{ref-dejea_microbiota_2014}{}
+24. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+Rossetti BJ, et al. Microbiota organization is a distinct feature of
+proximal colorectal cancers. Proceedings of the National Academy of
+Sciences of the United States of America. 2014;111:18321--6.
+
+\hypertarget{ref-sanapareddy_increased_2012}{}
+25. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
+et al. Increased rectal microbial richness is associated with the
+presence of colorectal adenomas in humans. The ISME journal.
+2012;6:1858--68.
+
+\hypertarget{ref-lu_mucosal_2016}{}
+26. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
+adherent bacterial dysbiosis in patients with colorectal adenomas.
+Scientific Reports. 2016;6:26337.
+
+\hypertarget{ref-hale_shifts_2017}{}
+27. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
+Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
+Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
+American Association for Cancer Research, Cosponsored by the American
+Society of Preventive Oncology. 2017;26:85--94.
+
+\hypertarget{ref-Flynn_preprint_2017}{}
+28. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
+the native colon microbiota in healthy adults. Cold Spring Harbor
+Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}
+
+\hypertarget{ref-Salter_contamination_2014}{}
+29. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
+al. Reagent and laboratory contamination can critically impact
+sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
+Nature; 2014;12. Available from:
+\url{https://doi.org/10.1186/s12915-014-0087-z}
+
+\hypertarget{ref-brim_microbiome_2013}{}
+30. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
+al. Microbiome analysis of stool samples from African Americans with
+colon polyps. PloS One. 2013;8:e81352.
 

### 2018-01-11 submission/manuscript.Rmd
update RF section except one paragraph

@@
 ***Whole Community Models Add Important Context:***
-Since CRC-associated genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only CRC-associated genera and tested this using two approaches.
-The first approach used genus level data and tested whether there were any differences in AUC between all genera and CRC-associated genera only during training on one study or testing on all other studies.
-The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model AUC across studies when comparing all OTUs versus only OTUs that taxonomically classified to CRC-associated genera.
+Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera based on RR and P-value significance and tested this using two approaches.
+The first approach used genus level data and tested whether there were any differences in AUC between all genera and selected genera during training on one study or testing on all other studies.
+The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model AUC across studies when comparing all OTUs versus only OTUs that taxonomically classified to selected genera.
 
@@
 
-In stool the most common genera in the genus based models belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
+In stool the most common genera in the full genera-based models belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
 Although all four CRC-associated genera were present in carcinoma, none were present in the majority of studies and *Fusobacterium* was the only genus present in adenoma [Figure 6A & B].
-For the OTU based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].
+For the full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].
 
-Unlike stool, tissue based Random Forest models showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 across study [Figure S7].
+Unlike stool, the tissue full genera-based Random Forest models showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 across study [Figure S7].
 The vast majority of the top 10 genera and OTUs occured in a study specific manner for these tissue based Random Forest models.
-For both the genera and OTU Random Forest models of adenoma and carcinoma, there appears to be very little overlap in the top 10 most important variables between stool and tissue [Figure 6 & S7].
-This discordance between stool and tissue also applies to the CRC-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of Random Forest models using matched versus unmatched tissue samples [Figure S7B-C & S7E-F].
+For both the full genera and OTU Random Forest models of adenoma and carcinoma, there appears to be very little overlap in the top 10 most important variables between stool and tissue [Figure 6 & S7].
+This discordance between stool and tissue also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of Random Forest models using matched versus unmatched tissue samples [Figure S7B-C & S7E-F].
 

### 2018-01-11 submission/manuscript.tex
update RF section except one paragraph

@@
 \textbf{\emph{Whole Community Models Add Important Context:}} Since
-CRC-associated genera increased RR for carcinoma over diversity metrics
-we assessed whether the bacterial community was better at classifying
-disease versus only CRC-associated genera and tested this using two
-approaches. The first approach used genus level data and tested whether
-there were any differences in AUC between all genera and CRC-associated
-genera only during training on one study or testing on all other
-studies. The second approach used OTU level data and tested whether
-there was a generalized decrease in the 10-fold cross validation (CV)
-model AUC across studies when comparing all OTUs versus only OTUs that
-taxonomically classified to CRC-associated genera.
+specific genera increased RR for carcinoma over diversity metrics we
+assessed whether the bacterial community was better at classifying
+disease versus only a select group of genera based on RR and P-value
+significance and tested this using two approaches. The first approach
+used genus level data and tested whether there were any differences in
+AUC between all genera and selected genera during training on one study
+or testing on all other studies. The second approach used OTU level data
+and tested whether there was a generalized decrease in the 10-fold cross
+validation (CV) model AUC across studies when comparing all OTUs versus
+only OTUs that taxonomically classified to selected genera.
 
@@
 
-In stool the most common genera in the genus based models belonged to
-genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
+In stool the most common genera in the full genera-based models belonged
+to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
 \emph{Roseburia} {[}Figure 6A \& B{]}. Although all four CRC-associated
@@
 studies and \emph{Fusobacterium} was the only genus present in adenoma
-{[}Figure 6A \& B{]}. For the OTU based models, \emph{Ruminococcaceae}
-was present in the top 10 consistently for both adenoma and carcinoma
-models while \emph{Roseburia} was only present in many adenoma models
-and \emph{Bacteroides} was present in the overwhelming majority of the
-carcinoma models {[}Figure 6C \& 6D{]}.
-
-Unlike stool, tissue based Random Forest models showed no consistent
-representation of \emph{Ruminococcaceae}, \emph{Ruminococcus},
-\emph{Bacteroides}, and \emph{Roseburia} in the top 10 across study
-{[}Figure S7{]}. The vast majority of the top 10 genera and OTUs occured
-in a study specific manner for these tissue based Random Forest models.
-For both the genera and OTU Random Forest models of adenoma and
-carcinoma, there appears to be very little overlap in the top 10 most
-important variables between stool and tissue {[}Figure 6 \& S7{]}. This
-discordance between stool and tissue also applies to the CRC-associated
-genera with one noticeable skew being that \emph{Fusobacterium} and
-\emph{Fusobacteriaceae} occur more often in the top 10 of Random Forest
-models using matched versus unmatched tissue samples {[}Figure S7B-C \&
-S7E-F{]}.
+{[}Figure 6A \& B{]}. For the full OTU-based models,
+\emph{Ruminococcaceae} was present in the top 10 consistently for both
+adenoma and carcinoma models while \emph{Roseburia} was only present in
+many adenoma models and \emph{Bacteroides} was present in the
+overwhelming majority of the carcinoma models {[}Figure 6C \& 6D{]}.
+
+Unlike stool, the tissue full genera-based Random Forest models showed
+no consistent representation of \emph{Ruminococcaceae},
+\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} in the top
+10 across study {[}Figure S7{]}. The vast majority of the top 10 genera
+and OTUs occured in a study specific manner for these tissue based
+Random Forest models. For both the full genera and OTU Random Forest
+models of adenoma and carcinoma, there appears to be very little overlap
+in the top 10 most important variables between stool and tissue
+{[}Figure 6 \& S7{]}. This discordance between stool and tissue also
+applies to the mouth-associated genera with one noticeable skew being
+that \emph{Fusobacterium} and \emph{Fusobacteriaceae} occur more often
+in the top 10 of Random Forest models using matched versus unmatched
+tissue samples {[}Figure S7B-C \& S7E-F{]}.
 

### 2018-01-11 submission/manuscript.Rmd
update discussion to match better with new direction

@@
 
-Within stool, common resident microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6].
+Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6].
 Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma [Figure 6].
 This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community.
-Based on these observations, it is possible to hypothesize that tsmall changes in community structure lead to new niches in which any one of the CRC-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma.
+Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma.
 
 The tissue studies did not provide a clearer understanding of how the microbiota may be associated with lesion.
-Generally, the OTU models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study.
-Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the CRC models [Figure S7B-C & E-F].
+Generally, the full OTU-based models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study.
+Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the full CRC models [Figure S7B-C & E-F].
 The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study.

### 2018-01-11 submission/manuscript.tex
update discussion to match better with new direction

@@
 
-Within stool, common resident microbes were most consistently present in
-the top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
+Within stool, common GI microbes were most consistently present in the
+top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
 \emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
@@
 how these bacteria interact with the changing resident community. Based
-on these observations, it is possible to hypothesize that tsmall changes
+on these observations, it is possible to hypothesize that small changes
 in community structure lead to new niches in which any one of the
-CRC-associated genera can gain a foothold, exacerbating the initial
+mouth-associated genera can gain a foothold, exacerbating the initial
 changes in community and facilitating the transition from adenoma to
@@
 The tissue studies did not provide a clearer understanding of how the
-microbiota may be associated with lesion. Generally, the OTU models of
-unmatched and matched {[}Figure S7E \& F{]} tissue samples were
-concordant with stool showing that resident microbes were the most
+microbiota may be associated with lesion. Generally, the full OTU-based
+models of unmatched and matched {[}Figure S7E \& F{]} tissue samples
+were concordant with stool showing that resident microbes were the most
 prevalent in the top 10 across study. Unlike in stool,
 \emph{Fusobacterium} was the only mouth-associated bacteria consistently
-present in the top 10 of the CRC models {[}Figure S7B-C \& E-F{]}. The
-majority of the tissue results seem to be study specific with many top
-10 taxa being present only in a single study. Additionally, the presence
-of genera associated with contamination, within the top 10 most
+present in the top 10 of the full CRC models {[}Figure S7B-C \& E-F{]}.
+The majority of the tissue results seem to be study specific with many
+top 10 taxa being present only in a single study. Additionally, the
+presence of genera associated with contamination, within the top 10 most
 important variables for the genera and OTU models, is worrying because

### 2018-01-12 submission/manuscript.Rmd
update discussion paragraph on b frag and strep gallo

@@
 Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models, is worrying because it is commonly a marker of contamination.
-The low bacterial biomass of tissue samples coupled with potential contaminationcould ultimately explain why the these results seem to be more sporadic than the stool results.
+The low bacterial biomass of tissue samples coupled with potential contamination could ultimately explain why the these results seem to be more sporadic than the stool results.
 
@@
 Since we are limited in our aggregation of the data, across study, to the genus level, it is not possible to clearly delineate what specific species are contributing to overall disease progression.
-However, *Streptococcus gallolyticus* subsp.
+Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp.
+*gallolyticus*.
+As an example, for stool, the full RF models consistently identified the genus *Bacteroides* as well as OTUs that classified as *Bacteroides* to be important model component across studies.
+Suggesting that even though *Bacteroides* may not increase the RR of lesion and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.
+Additionaly, *Streptococcus gallolyticus* subsp.
 *gallolyticus* is a mouth-associated micobe and our study clearly shows that regardless of whether the sample is tissue or stool, mouth-associated genera are commonly associated with an increased RR of both adenoma and carcinoma.

### 2018-01-12 submission/manuscript.tex
update discussion paragraph on b frag and strep gallo

@@
 it is commonly a marker of contamination. The low bacterial biomass of
-tissue samples coupled with potential contaminationcould ultimately
+tissue samples coupled with potential contamination could ultimately
 explain why the these results seem to be more sporadic than the stool
@@
 possible to clearly delineate what specific species are contributing to
-overall disease progression. However, \emph{Streptococcus gallolyticus}
-subsp. \emph{gallolyticus} is a mouth-associated micobe and our study
-clearly shows that regardless of whether the sample is tissue or stool,
+overall disease progression. Our observations are not inconsistent with
+the previous literature on either \emph{Bacteroides fragilis} or
+\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus}. As an
+example, for stool, the full RF models consistently identified the genus
+\emph{Bacteroides} as well as OTUs that classified as \emph{Bacteroides}
+to be important model component across studies. Suggesting that even
+though \emph{Bacteroides} may not increase the RR of lesion and may not
+vary in relative abundance, like \emph{Fusobacterium}, it is still
+important in CRC. Additionaly, \emph{Streptococcus gallolyticus} subsp.
+\emph{gallolyticus} is a mouth-associated micobe and our study clearly
+shows that regardless of whether the sample is tissue or stool,
 mouth-associated genera are commonly associated with an increased RR of

### 2018-01-12 submission/manuscript.tex
change the title

@@
 
-\section{Investigating the Microbiota and Colorectal Cancer: The
-Importance of
-Community}\label{investigating-the-microbiota-and-colorectal-cancer-the-importance-of-community}
+\section{Making Sense of the Noise: Leveraging Existing 16S rRNA Gene
+Surveys to Identify Key Community Members in Colorectal
+Cancer}\label{making-sense-of-the-noise-leveraging-existing-16s-rrna-gene-surveys-to-identify-key-community-members-in-colorectal-cancer}
 

### 2018-01-23 code/make_rf_auc_full_versus_specific_OTU_graph.R
update to include new AUCs with new select models

@@ -96,7 +96,7 @@ adn_tissue_medians <- adn_tissue %>%
   summarise(auc_median = median(AUC)) %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "CRC Associated\nGenera Community Only")))
+                             labels = c("Full Community", "Select Genera Only")))
 
 # Creates the adenoma tissue graph
 adn_tissue_graph <- adn_tissue %>% 
@@ -105,7 +105,7 @@ adn_tissue_graph <- adn_tissue %>%
                        labels = c("Unmatched Tissue", "Matched Tissue")), 
          model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "CRC Associated\nGenera Community Only")), 
+                             labels = c("Full Community", "Select Genera Only")), 
          study = factor(study, 
                         levels = c("flemer", "lu"), 
                         labels = c("Flemer", "Lu\n(matched)"))) %>% 
@@ -133,13 +133,13 @@ adn_stool_medians <- adn_all_stool %>%
   summarise(auc_median = median(AUC)) %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "CRC Associated\nGenera Community Only")))
+                             labels = c("Full Community", "Select Genera Only")))
 
 # Creates the adenoma stool graph
 adn_stool_graph <- adn_all_stool %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "CRC Associated\nGenera Community Only")), 
+                             labels = c("Full Community", "Select Genera Only")), 
          study = factor(study, 
                         levels = c("baxter", "brim", "hale", "zeller"), 
                         labels = c("Baxter", "Brim", "Hale", "Zeller"))) %>% 
@@ -168,13 +168,13 @@ crc_tissue_medians <- crc_all_tissue %>%
   summarise(auc_median = median(AUC)) %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "CRC Associated\nGenera Community Only")))
+                             labels = c("Full Community", "Select Genera Only")))
 
 # Creates the crc tissue graph
 crc_tissue_graph <- crc_all_tissue %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
[…]

### 2018-01-23 code/make_rf_auc_full_versus_specific_graph.R
update to include new AUCs from new select models

@@ -63,7 +63,7 @@ adn_tissue_graph <- adn_all_tissue %>%
                        labels = c("Unmatched Tissue", "Matched Tissue")), 
          model_type = factor(model_type, 
                              levels = c("full_model", "select_model"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("All Genera", "Select Genera Only")), 
          study = factor(study, 
                         levels = c("flemer", "lu"), 
                         labels = c("Flemer", "Lu"))) %>% 
@@ -88,7 +88,7 @@ adn_tissue_graph <- adn_all_tissue %>%
 adn_stool_graph <- adn_all_stool %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full_model", "select_model"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("All Genera", "Select Genera Only")), 
          study = factor(study, 
                         levels = c("baxter", "brim", "hale", "zeller"), 
                         labels = c("Baxter", "Brim", "Hale", "Zeller"))) %>% 
@@ -116,7 +116,7 @@ crc_tissue_graph <- crc_all_tissue %>%
                        labels = c("Unmatched Tissue", "Matched Tissue")), 
          model_type = factor(model_type, 
                              levels = c("full_model", "select_model"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("All Genera", "Select Genera Only")), 
          study = factor(study, 
                         levels = c("burns", "chen", "dejea", "flemer", "geng", "sana"), 
                         labels = c("Burns", "Chen", "Dejea", "Flemer", "Geng", "Sanapareddy"))) %>% 
@@ -143,7 +143,7 @@ crc_tissue_graph <- crc_all_tissue %>%
 crc_stool_graph <- crc_all_stool %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full_model", "select_model"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("All Genera", "Select Genera")), 
          study = factor(study, 
                         levels = c("ahn", "baxter", "flemer", "hale", "wang", "weir", "zeller"), 
                         labels = c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller"))) %>% 

### 2018-01-23 code/make_genus_rf_auc_against_study_graph.R
update with new AUCs based on new select models

@@ -106,7 +106,7 @@ adn_tissue_graph <- adn_all_tissue %>%
                         labels = c("Flemer", "Lu\n(Matched)")), 
          model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("All Genera", "Select Genera Only")), 
          study = factor(study, 
                         levels = c("flemer", "lu"), 
                         labels = c("Flemer", "Lu\n(Matched)"))) %>% 
@@ -136,7 +136,7 @@ adn_stool_graph <- adn_all_stool %>%
                         labels = c("Baxter", "Brim", "Hale", "Zeller")), 
          model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("All Genera", "Select Genera Only")), 
          study = factor(study, 
                         levels = c("baxter", "brim", "hale", "zeller"), 
                         labels = c("Baxter", "Brim", "Hale", "Zeller"))) %>% 
@@ -167,7 +167,7 @@ crc_unmatched_tissue_graph <- crc_all_unmatched_tissue %>%
                         labels = c("Burns", "Chen", "Flemer", "Sanapareddy")), 
          model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("All Genera", "Select Genera Only")), 
          study = factor(study, 
                         levels = c("burns", "chen", "flemer", "sana"), 
                         labels = c("Burns", "Chen", "Flemer", "Sanapareddy"))) %>% 
@@ -198,7 +198,7 @@ crc_matched_tissue_graph <- crc_all_matched_tissue %>%
                         labels = c("Burns", "Dejea", "Geng")), 
          model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("All Genera", "CRC Associated\nGenera Only")), 
+                             labels = c("All Genera", "Select Genera Only")), 
          study = factor(study, 
                         levels = c("burns", "dejea", "geng"), 
                         labels = c("Burns", "Dejea", "Geng"))) %>% 
@@ -229,7 +229,7 @@ crc_stool_graph <- crc_all_stool %>%
                         labels = c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller")), 
[…]

### 2018-01-23 submission/manuscript.Rmd
change the title
update figure legends

@@
 
-# Investigating the Microbiota and Colorectal Cancer: The Importance of Community
+# Making Sense of the Noise: Leveraging Existing 16S rRNA Gene Surveys to Identify Key Community Members in Colorectal Cancer
 
@@
 
-**Figure 3: CRC-Associated Genera Relative Risk for Adenoma and Carcinoma in Stool and Tissue.** A) Adenoma relative risk in stool.
+**Figure 3: Top 5 Genera that Decrease and Increase Relative Risk for Lesion.** A) Adenoma relative risk in stool.
 B) Carcinoma relative risk in stool.
@@
 
-**Figure 4: OTU Random Forest Model of Stool Across Studies.** A) Adenoma random forest model between the full community and CRC-associated genera OTUs only.
-B) Carcinoma random forest model between the full community and CRC-associated genera OTUs only.
+**Figure 4: Stool OTU Random Forest Model Across Studies.** A) Adenoma random forest model between the full community and select genera OTUs only.
+B) Carcinoma random forest model between the full community and select genera OTUs only.
 The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.
-The solid red line represents the median AUC of all the studies for either the full community or CRC-associated genera OTUS only model.
+The solid red line represents the median AUC of all the studies for either the full community or select genera OTUS only model.
 
-**Figure 5: OTU Random Forest Model of Tissue Across Studies.** A) Adenoma random forest model between the full community and CRC-associated genera OTUs only.
-B) Carcinoma random forest model between the full community and CRC-associated genera OTUs only.
+**Figure 5: Tissue OTU Random Forest Model Across Studies.** A) Adenoma random forest model between the full community and select genera OTUs only.
+B) Carcinoma random forest model between the full community and select genera OTUs only.
 The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.
-The solid red line represents the median AUC of all the studies for either the full community or CRC-associated genera OTUS only model.
+The solid red line represents the median AUC of all the studies for either the full community or select genera OTUS only model.
 
@@
 
-**Figure S2: Random Forest Genus Model AUC for each Stool Study.** A) AUC of adenoma models using all genera or CRC-associated genera only.
-B) AUC of carcinoma models using all genera or CRC-associated genera only.
+**Figure S2: Stool Random Forest Genus Model AUC for each Study.** A) AUC of adenoma models using all genera or select genera only.
+B) AUC of carcinoma models using all genera or select genera only.
 The black line represents the median within each group.
 
-**Figure S3: Random Forest Genus Model AUC for each Tissue Study.** A) AUC of adenoma models using all genera or only CRC-associated genera divided between matched and unmatched tissue.
-B) AUC of carcinoma models using all genera or CRC-associated genera only.
+**Figure S3: Tissue Random Forest Genus Model AUC for each Study.** A) AUC of adenoma models using all genera or only select genera divided between matched and unmatched tissue.
+B) AUC of carcinoma models using all genera or select genera only.
 The black line represents the median within each group divided between matched and unmatched tissue.
 
-**Figure S4: Random Forest Prediction Success Using Genera for each Stool Study.** A) AUC for prediction in adenoma using all genera or CRC associated genera only.
-B) AUC for prediction in carcinoma using all genera or CRC-associated genera only.
+**Figure S4: Stool Random Forest Prediction Success Using Genera Across Studies.** A) AUC for prediction in adenoma using all genera or select genera only.
+B) AUC for prediction in carcinoma using all genera or select genera only.
 The dotted line represents an AUC of 0.5.
@@
 
-**Figure S5: Random Forest Prediction Success of Carcinoma Using Genera for each Tissue Study.** A) AUC for prediction in unmatched tissue for all genera or CRC-associated genera only.
-B) AUC for prediction in matched tissue using all genera or CRC-associated genera only.
+**Figure S5: Tissue Random Forest Prediction Success of Carcinoma Using Genera Across Studies.** A) AUC for prediction in unmatched tissue for all genera or select genera only.
+B) AUC for prediction in matched tissue using all genera or select genera only.
 The dotted line represents an AUC of 0.5.
@@
 
-**Figure S6: Random Forest Prediction Success of Adenoma Using Genera for each Tissue Study.** The red lines represent the median AUC using that specific study as the training set.
+**Figure S6: Tissue Random Forest Prediction Success of Adenoma Using Genera Across Studies.** The red lines represent the median AUC using that specific study as the training set.
 

### 2018-01-23 submission/manuscript.tex
update figure legends

@@
 For the \(\alpha\)-diversity analysis, values were power transformed
-using the rcompanion (v1.10.1) package {[}39{]} and then Z-score
-normalized using the car (v2.1.5) package {[}40{]}. Testing for
+using the rcompanion (v1.11.1) package {[}39{]} and then Z-score
+normalized using the car (v2.1.6) package {[}40{]}. Testing for
 \(\alpha\)-diversity differences utilized linear mixed-effect models
-created using the lme4 (v1.1.14) package {[}41{]} to correct for study,
+created using the lme4 (v1.1.15) package {[}41{]} to correct for study,
 repeat sampling of individuals (tissue only), and 16S hypervariable
-region used. Relative risk was analyzed using both the epiR (v0.9.87)
+region used. Relative risk was analyzed using both the epiR (v0.9.93)
 and metafor (v2.0.0) packages {[}42,43{]} by assessing how many
@@
 utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
-vegan (v2.4.4) package {[}44{]}. Random Forest models were built using
-both the caret (v6.0.77) and randomForest (v4.6.12) packages
+vegan (v2.4.5) package {[}44{]}. Random Forest models were built using
+both the caret (v6.0.78) and randomForest (v4.6.12) packages
 {[}45,46{]}. Power analysis and estimations were made using the pwr
@@
 
-\textbf{Figure 3: CRC-Associated Genera Relative Risk for Adenoma and
-Carcinoma in Stool and Tissue.} A) Adenoma relative risk in stool. B)
-Carcinoma relative risk in stool. C) Adenoma relative risk in tissue. D)
-Carcinoma relative risk in tissue. For all panels the relative risk was
-also compared to whether one, two, three, or four of the CRC-associated
-genera were present.
-
-\textbf{Figure 4: OTU Random Forest Model of Stool Across Studies.} A)
-Adenoma random forest model between the full community and
-CRC-associated genera OTUs only. B) Carcinoma random forest model
-between the full community and CRC-associated genera OTUs only. The
-dotted line represents an AUC of 0.5 and the lines represent the range
-in which the AUC for the 100 different 80/20 runs fell between. The
-solid red line represents the median AUC of all the studies for either
-the full community or CRC-associated genera OTUS only model.
-
-\textbf{Figure 5: OTU Random Forest Model of Tissue Across Studies.} A)
-Adenoma random forest model between the full community and
-CRC-associated genera OTUs only. B) Carcinoma random forest model
-between the full community and CRC-associated genera OTUs only. The
-dotted line represents an AUC of 0.5 and the lines represent the range
-in which the AUC for the 100 different 80/20 runs fell between. The
-solid red line represents the median AUC of all the studies for either
-the full community or CRC-associated genera OTUS only model.
+\textbf{Figure 3: Top 5 Genera that Decrease and Increase Relative Risk
+for Lesion.} A) Adenoma relative risk in stool. B) Carcinoma relative
+risk in stool. C) Adenoma relative risk in tissue. D) Carcinoma relative
+risk in tissue. For all panels the relative risk was also compared to
+whether one, two, three, or four of the CRC-associated genera were
+present.
+
+\textbf{Figure 4: Stool OTU Random Forest Model Across Studies.} A)
+Adenoma random forest model between the full community and select genera
+OTUs only. B) Carcinoma random forest model between the full community
+and select genera OTUs only. The dotted line represents an AUC of 0.5
+and the lines represent the range in which the AUC for the 100 different
+80/20 runs fell between. The solid red line represents the median AUC of
+all the studies for either the full community or select genera OTUS only
+model.
+
+\textbf{Figure 5: Tissue OTU Random Forest Model Across Studies.} A)
+Adenoma random forest model between the full community and select genera
+OTUs only. B) Carcinoma random forest model between the full community
+and select genera OTUs only. The dotted line represents an AUC of 0.5
+and the lines represent the range in which the AUC for the 100 different
+80/20 runs fell between. The solid red line represents the median AUC of
+all the studies for either the full community or select genera OTUS only
+model.
 
@@
 
-\textbf{Figure S2: Random Forest Genus Model AUC for each Stool Study.}
-A) AUC of adenoma models using all genera or CRC-associated genera only.
-B) AUC of carcinoma models using all genera or CRC-associated genera
-only. The black line represents the median within each group.
-
-\textbf{Figure S3: Random Forest Genus Model AUC for each Tissue Study.}
-A) AUC of adenoma models using all genera or only CRC-associated genera
-divided between matched and unmatched tissue. B) AUC of carcinoma models
-using all genera or CRC-associated genera only. The black line
-represents the median within each group divided between matched and
-unmatched tissue.
-
-\textbf{Figure S4: Random Forest Prediction Success Using Genera for
-each Stool Study.} A) AUC for prediction in adenoma using all genera or
-CRC associated genera only. B) AUC for prediction in carcinoma using all
-genera or CRC-associated genera only. The dotted line represents an AUC
-of 0.5. The x-axis is the data set in which the model was initially
-trained on. The red lines represent the median AUC using that specific
-study as the training set.
-
-\textbf{Figure S5: Random Forest Prediction Success of Carcinoma Using
-Genera for each Tissue Study.} A) AUC for prediction in unmatched tissue
-for all genera or CRC-associated genera only. B) AUC for prediction in
-matched tissue using all genera or CRC-associated genera only. The
-dotted line represents an AUC of 0.5. The x-axis is the data set in
-which the model was initially trained on. The red lines represent the
-median AUC using that specific study as the training set.
-
-\textbf{Figure S6: Random Forest Prediction Success of Adenoma Using
-Genera for each Tissue Study.} The red lines represent the median AUC
+\textbf{Figure S2: Stool Random Forest Genus Model AUC for each Study.}
+A) AUC of adenoma models using all genera or select genera only. B) AUC
+of carcinoma models using all genera or select genera only. The black
+line represents the median within each group.
+
+\textbf{Figure S3: Tissue Random Forest Genus Model AUC for each Study.}
+A) AUC of adenoma models using all genera or only select genera divided
+between matched and unmatched tissue. B) AUC of carcinoma models using
+all genera or select genera only. The black line represents the median
+within each group divided between matched and unmatched tissue.
+
+\textbf{Figure S4: Stool Random Forest Prediction Success Using Genera
+Across Studies.} A) AUC for prediction in adenoma using all genera or
+select genera only. B) AUC for prediction in carcinoma using all genera
+or select genera only. The dotted line represents an AUC of 0.5. The
+x-axis is the data set in which the model was initially trained on. The
+red lines represent the median AUC using that specific study as the
+training set.
+
+\textbf{Figure S5: Tissue Random Forest Prediction Success of Carcinoma
+Using Genera Across Studies.} A) AUC for prediction in unmatched tissue
+for all genera or select genera only. B) AUC for prediction in matched
+tissue using all genera or select genera only. The dotted line
+represents an AUC of 0.5. The x-axis is the data set in which the model
+was initially trained on. The red lines represent the median AUC using
+that specific study as the training set.
+
+\textbf{Figure S6: Tissue Random Forest Prediction Success of Adenoma
+Using Genera Across Studies.} The red lines represent the median AUC
 using that specific study as the training set.

### 2018-01-23 submission/manuscript.Rmd
add changes to results and discussion based on updated selected models

@@
 
-For the genera-based models both tissue and stool studies had a median AUC decrease in model classification when only CRC-associated genera versus all genera were included [Figure S2-S3].
-With respect to the test sets, comprised of genera data from other studies, both all and CRC-associated genera models had a similar ability to detect adenomas or carcinomas [Figure S4-S6].
-The OTU-based models showed a similar trend to the genera models where a generalized decrease in median AUC was observed when only OTUs from the CRC-associated genera were used versus the full community of OTUs for both adenoma and carcinoma, with the largest median change for stool in carcinoma models and for tissue in adenoma models [Figure 4 & 5].
+For the genera-based models the training set median AUC for model classification was similar for select genera and full genera models, for both tissue and stool studies [Figure S2-S3].
+With respect to the test sets, comprised of genera data from other studies, both all and select genera models had a similar ability to detect adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6].
+Conversely, the OTU-based models showed a slight decrease in median AUC between the full and select models, with one exception being carcinoma models for matched tissue [Figure 4 & 5].
 
@@
 
-The data presented herein supports the importance of mouth-associated genera for carcinoma but not necessarily adenoma formation.
-A proposed mechanism by which these mouth-associated microbes cause CRC stems from the creation of a vicious cycle involving host inflammation and repair.
-Both the drastically increased carcinoma RR of mouth-associated genera versus $\alpha$-diversity metrics and increasing RR with more mouth-associated genera positivity [Figure 3C], are highly supportive of their importance.
-However, our observations show that when adding the community context to our Random Forest models along with the mouth-associated genera, the model AUC increases [Figure 4 & 5].
-This suggests the resident community cannot be ignored completely and needs to be accounted for when investigating mouth-associated microbes and carcinoma.
-Conversely, using the present data, it is less likely that mouth-associated microbes are correlated with adenoma development.
-The step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1].
-Yet, this change in diversity, at this early stage of disease, could be focal to the adenoma itself.
-The poor performance of the Random Forest models for classifying adenoma, based only on the stool microbiota, would suggest that this is the case.
+The data presented herein supports the importance of select genera for carcinoma but not necessarily adenoma formation.
+Our observations show that when only using the select genera to create models for carcinoma the AUC are similar to using all genera or OTUs [Figure 4 & 5].
+This suggests that an interplay between a select number of potentially protective and exacerbating microbes, within the GI community, are crucial for carcinoma formation.
+Importantly, it suggests that there may be key members of the community that might be studied further to potentially reduce the risk of carcinoma.
+Conversely, using the present data, it is clear that new approaches may need to be used to identify members of the community that are associated with adenoma.
+Regardless of sample type and whether a full or select model was used our Random Forest models consistently performed poorly.
+Yet, the step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1].
+This change in diversity, at this early stage of disease, could be focal to the adenoma itself.
 One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

### 2018-01-23 submission/manuscript.tex
add changes to results and discussion based on updated selected models

@@
 
-For the genera-based models both tissue and stool studies had a median
-AUC decrease in model classification when only CRC-associated genera
-versus all genera were included {[}Figure S2-S3{]}. With respect to the
+For the genera-based models the training set median AUC for model
+classification was similar for select genera and full genera models, for
+both tissue and stool studies {[}Figure S2-S3{]}. With respect to the
 test sets, comprised of genera data from other studies, both all and
-CRC-associated genera models had a similar ability to detect adenomas or
-carcinomas {[}Figure S4-S6{]}. The OTU-based models showed a similar
-trend to the genera models where a generalized decrease in median AUC
-was observed when only OTUs from the CRC-associated genera were used
-versus the full community of OTUs for both adenoma and carcinoma, with
-the largest median change for stool in carcinoma models and for tissue
-in adenoma models {[}Figure 4 \& 5{]}.
+select genera models had a similar ability to detect adenomas or
+carcinomas, with the select genera models performing better in some
+instances {[}Figure S4-S6{]}. Conversely, the OTU-based models showed a
+slight decrease in median AUC between the full and select models, with
+one exception being carcinoma models for matched tissue {[}Figure 4 \&
+5{]}.
 
@@
 
-The data presented herein supports the importance of mouth-associated
-genera for carcinoma but not necessarily adenoma formation. A proposed
-mechanism by which these mouth-associated microbes cause CRC stems from
-the creation of a vicious cycle involving host inflammation and repair.
-Both the drastically increased carcinoma RR of mouth-associated genera
-versus \(\alpha\)-diversity metrics and increasing RR with more
-mouth-associated genera positivity {[}Figure 3C{]}, are highly
-supportive of their importance. However, our observations show that when
-adding the community context to our Random Forest models along with the
-mouth-associated genera, the model AUC increases {[}Figure 4 \& 5{]}.
-This suggests the resident community cannot be ignored completely and
-needs to be accounted for when investigating mouth-associated microbes
-and carcinoma. Conversely, using the present data, it is less likely
-that mouth-associated microbes are correlated with adenoma development.
-The step-wise decrease in diversity suggests that the adenoma community
-is not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
-diversity, at this early stage of disease, could be focal to the adenoma
-itself. The poor performance of the Random Forest models for classifying
-adenoma, based only on the stool microbiota, would suggest that this is
-the case. One potential hypothesis from these observations is that at
-early stages of the disease, how the host interacts with these subtle
-changes is what ultimately leads to a thoroughly dysfunctional community
-that is supportive of CRC genesis.
+The data presented herein supports the importance of select genera for
+carcinoma but not necessarily adenoma formation. Our observations show
+that when only using the select genera to create models for carcinoma
+the AUC are similar to using all genera or OTUs {[}Figure 4 \& 5{]}.
+This suggests that an interplay between a select number of potentially
+protective and exacerbating microbes, within the GI community, are
+crucial for carcinoma formation. Importantly, it suggests that there may
+be key members of the community that might be studied further to
+potentially reduce the risk of carcinoma. Conversely, using the present
+data, it is clear that new approaches may need to be used to identify
+members of the community that are associated with adenoma. Regardless of
+sample type and whether a full or select model was used our Random
+Forest models consistently performed poorly. Yet, the step-wise decrease
+in diversity suggests that the adenoma community is not normal but has
+changed subtly {[}Figure 1{]}. This change in diversity, at this early
+stage of disease, could be focal to the adenoma itself. One potential
+hypothesis from these observations is that at early stages of the
+disease, how the host interacts with these subtle changes is what
+ultimately leads to a thoroughly dysfunctional community that is
+supportive of CRC genesis.
 

### 2018-01-23 submission/manuscript.Rmd
update abstract

@@
 This reduction in evenness resulted in small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only resulted in an increased relative risk for carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
-Previously associated colorectal cancer genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) followed a similar pattern, with their presence significantly increasing the relative risk for carcinoma (P-value `r ifelse(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "pvalue"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05) but not adenoma (P-value `r ifelse(min(filter(adn_select_RR_stool, !(measure %in% c("all_four", "total_four", "Porphyromonas")))[, "pvalue"]) > 0.05, invisible(">"), invisible("check tables"))` 0.05) with the exception of *Porphyromonas* (P-value = `r format(min(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"]), digits = 2)`).
-Using the whole community versus only CRC-associated genera to build a prediction model resulted in higher classification success, based on Area Under the Curve (AUC), for both adenoma and carcinoma using fecal and tissue samples.
-The most important OTUs for these models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies.
+Mouth-associated genera were consistently in the top 5 OTUs or genera with the highest relative risk of adenoma and carcinoma for both stool and tissue.
+Using the whole community versus selected generea with the highest and lowest relative risk, to build a prediction model, resulted in similar classification success, based on Area Under the Curve (AUC), for both adenoma and carcinoma using fecal and tissue samples.
+The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies.
 Overall, there were less associations between the microbiota and adenoma and one reason why this may be is that most studies were only adequately powered for large effect sizes.

### 2018-01-23 submission/manuscript.tex
update abstract

@@
 resulted in an increased relative risk for carcinoma (P-value = 0.0047).
-Previously associated colorectal cancer genera (\emph{Fusobacterium},
-\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
-followed a similar pattern, with their presence significantly increasing
-the relative risk for carcinoma (P-value \textless{} 0.05) but not
-adenoma (P-value check tables 0.05) with the exception of
-\emph{Porphyromonas} (P-value = 0.023). Using the whole community versus
-only CRC-associated genera to build a prediction model resulted in
-higher classification success, based on Area Under the Curve (AUC), for
-both adenoma and carcinoma using fecal and tissue samples. The most
-important OTUs for these models consistently belonged to genera such as
-\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
-studies. Overall, there were less associations between the microbiota
-and adenoma and one reason why this may be is that most studies were
-only adequately powered for large effect sizes.
+Mouth-associated genera were consistently in the top 5 OTUs or genera
+with the highest relative risk of adenoma and carcinoma for both stool
+and tissue. Using the whole community versus selected generea with the
+highest and lowest relative risk, to build a prediction model, resulted
+in similar classification success, based on Area Under the Curve (AUC),
+for both adenoma and carcinoma using fecal and tissue samples. The most
+important groups within the full community models consistently belonged
+to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
+\emph{Roseburia} across studies. Overall, there were less associations
+between the microbiota and adenoma and one reason why this may be is
+that most studies were only adequately powered for large effect sizes.
 

### 2018-01-23 submission/manuscript.Rmd
fix abstract and other grammar errors

@@
 **Conclusions.** This data provides support for the importance of the bacterial community to both adenoma and carcinoma genesis.
-The evidence collected within this study on the role of the microbiota in CRC pathogenesis shows stronger associations between carcinoma then adenoma.
-One reason for this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
+The evidence collected within this study on the role of the microbiota in CRC pathogenesis shows stronger associations between carcinoma then adenoma and this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
 
@@
 
-Colorectal cancer (CRC) is a growing world-wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016].
+Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis [@siegel_cancer_2016; @flynn_metabolic_2016].
 Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC.
@@
 A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
-This meta-analysis focused on identifying biomarkers or individual microbes but did not critically investigate how the community changes in CRC.
+This meta-analysis focused on identifying biomarkers or individual microbes but did so on a small total number of individuals and only investigated stool.
 
-Although there has been an intense focus on microbiota based biomarker discovery for CRC, the candidate genera seem to be numerous.
+Although there has been an intense focus on microbiota-based biomarker discovery for CRC, the candidate genera seem to be numerous.
 Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
@@
 It is well known that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014].
-Due to these many differences that could arise between stool and tissue, one question our meta-analysis aims to answer is whether there are consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.
+Due to these many differences that could arise between stool and tissue, an additional question our meta-analysis aims to answer is whether there are consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.
 
@@
 Finally, we also examined at what effect and sample size the studies that were used were powered for and the sample size needed to get to an 80% power.
-Our power analysis suggests that each individual study used was underpowered for detecting effect size differences of 10% or below between the case and control groups.
 
@@
 ***Lower Bacterial Diversity is Associated with Increased RR of Carcinomas:***
-Power transforming and Z-score normalizing OTU richness, evenness, and Shannon diversity for the entire data set allowed for the combination of data from all studies to assess differences in broad scale community metrics as disease severity worsens.
+Power transforming and Z-score normalizing Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity for the entire data set allowed for the combination of data from all studies to assess differences in broad scale community metrics as disease severity worsens.
 Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity.
@@
 ***Mouth-Associated Genera are Associated with an Increased RR of Lesion:***
-Based on the small increase in RR using $\alpha$-diversity metrics, we assessed if being higher than the median for any specific genera resulted in an altered RR for adenoma or carcinoma in stool and tissue.
+Based on the small increase in RR using OTU richness and Shannon diversity, we assessed if being higher than the median for any specific genera resulted in an altered RR for adenoma or carcinoma in stool and tissue.
 To investigate this we analyzed all common genera across each study, in tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR.
@@
 
-***Whole Community Models Add Important Context:***
+***Select Genera Models can Recapitulate Whole Community Models:***
 Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera based on RR and P-value significance and tested this using two approaches.
@@
 
-In stool the most common genera in the full genera-based models belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
-Although all four CRC-associated genera were present in carcinoma, none were present in the majority of studies and *Fusobacterium* was the only genus present in adenoma [Figure 6A & B].
+In stool, the most common genera in the full genera-based models belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
+Although all mouth-associated genera were present in carcinoma, none were present in the majority of studies and *Fusobacterium* was the only genus present in adenoma [Figure 6A & B].
 For the full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].
@@
 Even though mouth-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma.
-These mouth-associated genera are clearly important to carcinoma classification but our observations suggest that accounting for the community in which these microbes exist can increase the ability of models to make predictions.
+These mouth-associated genera are clearly important to carcinoma classification but our observations suggest that accounting for other community members can increase the ability of models to make predictions.
 
@@
 This suggests that an interplay between a select number of potentially protective and exacerbating microbes, within the GI community, are crucial for carcinoma formation.
-Importantly, it suggests that there may be key members of the community that might be studied further to potentially reduce the risk of carcinoma.
+Importantly, it suggests that there may be key members of the GI community that might be studied further to potentially reduce the risk of carcinoma.
 Conversely, using the present data, it is clear that new approaches may need to be used to identify members of the community that are associated with adenoma.
@@
 The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study.
-Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models, is worrying because it is commonly a marker of contamination.
+Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying.
 The low bacterial biomass of tissue samples coupled with potential contamination could ultimately explain why the these results seem to be more sporadic than the stool results.
@@
 *gallolyticus*.
-As an example, for stool, the full RF models consistently identified the genus *Bacteroides* as well as OTUs that classified as *Bacteroides* to be important model component across studies.
+As an example, for stool, the full RF models consistently identified the genus *Bacteroides* as well as OTUs that classified as *Bacteroides* to be important model components across studies.
 Suggesting that even though *Bacteroides* may not increase the RR of lesion and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.
@@
 A small effect size may well be the scope in which differences consistently occur in adenoma based on the results observed within our meta-analysis.
-Future studies investigating adenoma and the microbiota need to take power into consideration if we are to reproducibly study whether the microbiota contributes to poly formation.
+Future studies investigating adenoma and the microbiota need to take power into consideration if we are to reproducibly study whether the microbiota contributes to polyp formation.
 In contrast to adenoma, our observations suggest that most studies analyzed have sufficient power to detect many changes in carcinoma because of large effect size differences between cases and controls [Figure 7].

### 2018-01-23 submission/manuscript.tex
fix abstract and other grammar errors

@@
 CRC pathogenesis shows stronger associations between carcinoma then
-adenoma. One reason for this may be in part due to the low power to
-detect more subtle changes in the majority of studies that have been
-performed to date.
+adenoma and this may be in part due to the low power to detect more
+subtle changes in the majority of studies that have been performed to
+date.
 
@@
 
-Colorectal cancer (CRC) is a growing world-wide health problem {[}1{]}
-in which the microbiota has been purported to play an active role in
-disease pathogenesis {[}2{]}. Numerous studies have shown the importance
-of both individual microbes {[}3--7{]} and the overall community
+Colorectal cancer (CRC) is a growing world-wide health problem in which
+the microbiota has been purported to play an active role in disease
+pathogenesis {[}1,2{]}. Numerous studies have shown the importance of
+both individual microbes {[}3--7{]} and the overall community
 {[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
@@
 identified using multiple data sets {[}11{]}. This meta-analysis focused
-on identifying biomarkers or individual microbes but did not critically
-investigate how the community changes in CRC.
-
-Although there has been an intense focus on microbiota based biomarker
+on identifying biomarkers or individual microbes but did so on a small
+total number of individuals and only investigated stool.
+
+Although there has been an intense focus on microbiota-based biomarker
 discovery for CRC, the candidate genera seem to be numerous. Some
@@
 contamination {[}29{]}. Due to these many differences that could arise
-between stool and tissue, one question our meta-analysis aims to answer
-is whether there are consistent patterns that emerge across studies
-regardless of whether they used stool or tissue samples.
+between stool and tissue, an additional question our meta-analysis aims
+to answer is whether there are consistent patterns that emerge across
+studies regardless of whether they used stool or tissue samples.
 
@@
 sample size the studies that were used were powered for and the sample
-size needed to get to an 80\% power. Our power analysis suggests that
-each individual study used was underpowered for detecting effect size
-differences of 10\% or below between the case and control groups.
+size needed to get to an 80\% power.
 
@@
 \textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
-of Carcinomas:}} Power transforming and Z-score normalizing OTU
-richness, evenness, and Shannon diversity for the entire data set
-allowed for the combination of data from all studies to assess
-differences in broad scale community metrics as disease severity
+of Carcinomas:}} Power transforming and Z-score normalizing Operational
+Taxonomic Unit (OTU) richness, evenness, and Shannon diversity for the
+entire data set allowed for the combination of data from all studies to
+assess differences in broad scale community metrics as disease severity
 worsens. Using linear mixed-effect models to control for study and
@@
 \textbf{\emph{Mouth-Associated Genera are Associated with an Increased
-RR of Lesion:}} Based on the small increase in RR using
-\(\alpha\)-diversity metrics, we assessed if being higher than the
-median for any specific genera resulted in an altered RR for adenoma or
-carcinoma in stool and tissue. To investigate this we analyzed all
-common genera across each study, in tissue or stool, and assessed
-whether a relative abundance higher than the median results in an
-increase or decrease in RR. For both stool and tissue none of the genera
-in the top 5 most significantly increased RR were the same between
-adenoma and carcinoma {[}Figure 3{]}. Mouth-associated genera were in
-the top 5 genera associated with an increased RR of adenoma
-(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
-3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
-\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and tissue.
-Conversely, genera commonly associated with a normal gastrointestinal
-tract were correlated with a decreased RR for both adenoma and carcinoma
-for both stool and tissue {[}Figure 3{]}. Overall, there was little
-direct overlap of the top 5 increased or decreased RR genera between
-stool and tissue.
+RR of Lesion:}} Based on the small increase in RR using OTU richness and
+Shannon diversity, we assessed if being higher than the median for any
+specific genera resulted in an altered RR for adenoma or carcinoma in
+stool and tissue. To investigate this we analyzed all common genera
+across each study, in tissue or stool, and assessed whether a relative
+abundance higher than the median results in an increase or decrease in
+RR. For both stool and tissue none of the genera in the top 5 most
+significantly increased RR were the same between adenoma and carcinoma
+{[}Figure 3{]}. Mouth-associated genera were in the top 5 genera
+associated with an increased RR of adenoma (\emph{Pyramidobacter}
+{[}Figure 3A{]} and \emph{Rothia} {[}Figure 3C{]}) and carcinoma
+(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Porphyromonas}, and
+\emph{Peptostreptococcus} {[}Figure 3B{]} and \emph{Fusobacterium}
+{[}Figure 3D{]}) for both stool and tissue. Conversely, genera commonly
+associated with a normal gastrointestinal tract were correlated with a
+decreased RR for both adenoma and carcinoma for both stool and tissue
+{[}Figure 3{]}. Overall, there was little direct overlap of the top 5
+increased or decreased RR genera between stool and tissue.
 
@@
 
-\textbf{\emph{Whole Community Models Add Important Context:}} Since
-specific genera increased RR for carcinoma over diversity metrics we
-assessed whether the bacterial community was better at classifying
-disease versus only a select group of genera based on RR and P-value
-significance and tested this using two approaches. The first approach
-used genus level data and tested whether there were any differences in
-AUC between all genera and selected genera during training on one study
-or testing on all other studies. The second approach used OTU level data
-and tested whether there was a generalized decrease in the 10-fold cross
-validation (CV) model AUC across studies when comparing all OTUs versus
-only OTUs that taxonomically classified to selected genera.
+\textbf{\emph{Select Genera Models can Recapitulate Whole Community
+Models:}} Since specific genera increased RR for carcinoma over
+diversity metrics we assessed whether the bacterial community was better
+at classifying disease versus only a select group of genera based on RR
+and P-value significance and tested this using two approaches. The first
+approach used genus level data and tested whether there were any
+differences in AUC between all genera and selected genera during
+training on one study or testing on all other studies. The second
+approach used OTU level data and tested whether there was a generalized
+decrease in the 10-fold cross validation (CV) model AUC across studies
+when comparing all OTUs versus only OTUs that taxonomically classified
+to selected genera.
 
@@
 
-In stool the most common genera in the full genera-based models belonged
-to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
-\emph{Roseburia} {[}Figure 6A \& B{]}. Although all four CRC-associated
+In stool, the most common genera in the full genera-based models
+belonged to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
+\emph{Roseburia} {[}Figure 6A \& B{]}. Although all mouth-associated
 genera were present in carcinoma, none were present in the majority of
@@
 are clearly important to carcinoma classification but our observations
-suggest that accounting for the community in which these microbes exist
-can increase the ability of models to make predictions.
+suggest that accounting for other community members can increase the
+ability of models to make predictions.
 
@@
 crucial for carcinoma formation. Importantly, it suggests that there may
-be key members of the community that might be studied further to
+be key members of the GI community that might be studied further to
 potentially reduce the risk of carcinoma. Conversely, using the present
@@
 presence of genera associated with contamination, within the top 10 most
-important variables for the genera and OTU models, is worrying because
-it is commonly a marker of contamination. The low bacterial biomass of
-tissue samples coupled with potential contamination could ultimately
-explain why the these results seem to be more sporadic than the stool
-results.
+important variables for the genera and OTU models is worrying. The low
+bacterial biomass of tissue samples coupled with potential contamination
+could ultimately explain why the these results seem to be more sporadic
+than the stool results.
 
@@
 \emph{Bacteroides} as well as OTUs that classified as \emph{Bacteroides}
-to be important model component across studies. Suggesting that even
+to be important model components across studies. Suggesting that even
 though \emph{Bacteroides} may not increase the RR of lesion and may not
@@
 power into consideration if we are to reproducibly study whether the
-microbiota contributes to poly formation. In contrast to adenoma, our
+microbiota contributes to polyp formation. In contrast to adenoma, our
 observations suggest that most studies analyzed have sufficient power to

### 2018-01-25 submission/manuscript.Rmd
ada edits on abstract

@@
 Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community has important associations to both early (adenoma) and late (carcinoma) stage disease.
-To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 different studies.
+To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 previously published 16S rRNA gene sequencing studies on CRC and the microbiota.
 
-**Results.** Fecal samples had a significant decrease from control to adenoma to carcinoma for both Shannon diversity and evenness after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
-This reduction in evenness resulted in small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only resulted in an increased relative risk for carcinoma (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
-Mouth-associated genera were consistently in the top 5 OTUs or genera with the highest relative risk of adenoma and carcinoma for both stool and tissue.
-Using the whole community versus selected generea with the highest and lowest relative risk, to build a prediction model, resulted in similar classification success, based on Area Under the Curve (AUC), for both adenoma and carcinoma using fecal and tissue samples.
+**Results.** Fecal samples had a significant decrease from those without disease to adenoma (less severe disease) to carcinoma (severe disease) for both Shannon diversity and evenness after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
+This reduction in evenness translated into small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased relative risk for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
+Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples.
+A prediction model for adenoma and carcinoma was built using either the whole community or selected genera with highest and lowest relative risk from fecal and tissue samples.
+Both approaches resulted in similar classification success according to Area Under the Curve (AUC) regardless of whether genera or OTUs were used to build the model.
 The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies.
-Overall, there were less associations between the microbiota and adenoma and one reason why this may be is that most studies were only adequately powered for large effect sizes.
+Although a number of associations between the microbiota and CRC were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.
 
-**Conclusions.** This data provides support for the importance of the bacterial community to both adenoma and carcinoma genesis.
-The evidence collected within this study on the role of the microbiota in CRC pathogenesis shows stronger associations between carcinoma then adenoma and this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.
+**Conclusions.** These data provide support for the importance of the bacterial community to both adenoma and carcinoma genesis.
+The evidence collected within this study on the role of the microbiota in CRC identifies a number of correlations that may not have been detected because of the low power associated with the majority of studies that have been performed to date.
 

### 2018-01-25 submission/manuscript.tex
ada edits on abstract

@@
 associations to both early (adenoma) and late (carcinoma) stage disease.
-To test this hypothesis we examined both feces (n = 1737) and tissue
-(492 total samples from 350 individuals) across 14 different studies.
-
-\textbf{Results.} Fecal samples had a significant decrease from control
-to adenoma to carcinoma for both Shannon diversity and evenness after
-correcting for study effect and variable region sequenced (P-value
-\textless{} 0.05). This reduction in evenness resulted in small
-increases in relative risk for adenoma (P-value = 0.032) and carcinoma
-(P-value = 0.00034) while the reduction in Shannon diversity only
-resulted in an increased relative risk for carcinoma (P-value = 0.0047).
-Mouth-associated genera were consistently in the top 5 OTUs or genera
-with the highest relative risk of adenoma and carcinoma for both stool
-and tissue. Using the whole community versus selected generea with the
-highest and lowest relative risk, to build a prediction model, resulted
-in similar classification success, based on Area Under the Curve (AUC),
-for both adenoma and carcinoma using fecal and tissue samples. The most
-important groups within the full community models consistently belonged
-to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
-\emph{Roseburia} across studies. Overall, there were less associations
-between the microbiota and adenoma and one reason why this may be is
-that most studies were only adequately powered for large effect sizes.
-
-\textbf{Conclusions.} This data provides support for the importance of
+To test this hypothesis we examined both feces (n = 1737) and colon
+tissue (492 total samples from 350 individuals) across 14 previously
+published 16S rRNA gene sequencing studies on CRC and the microbiota.
+
+\textbf{Results.} Fecal samples had a significant decrease from those
+without disease to adenoma (less severe disease) to carcinoma (severe
+disease) for both Shannon diversity and evenness after correcting for
+study effect and variable region sequenced (P-value \textless{} 0.05).
+This reduction in evenness translated into small increases in relative
+risk for adenoma (P-value = 0.032) and carcinoma stages of CRC (P-value
+= 0.00034) while the reduction in Shannon diversity only translated into
+an increased relative risk for developing carcinomas (P-value = 0.0047).
+Increases in mouth-associated microbes were commonly in the top 5 most
+significantly increased relative risk of adenoma and carcinoma for both
+stool and tissue samples. A prediction model for adenoma and carcinoma
+was built using either the whole community or selected genera with
+highest and lowest relative risk from fecal and tissue samples. Both
+approaches resulted in similar classification success according to Area
+Under the Curve (AUC) regardless of whether genera or OTUs were used to
+build the model. The most important groups within the full community
+models consistently belonged to genera such as \emph{Ruminococcus},
+\emph{Bacteroides}, and \emph{Roseburia} across studies. Although a
+number of associations between the microbiota and CRC were identified,
+the majority of studies that we used in this meta-analysis were only
+individually adequately powered for large effect sizes.
+
+\textbf{Conclusions.} These data provide support for the importance of
 the bacterial community to both adenoma and carcinoma genesis. The
 evidence collected within this study on the role of the microbiota in
-CRC pathogenesis shows stronger associations between carcinoma then
-adenoma and this may be in part due to the low power to detect more
-subtle changes in the majority of studies that have been performed to
-date.
+CRC identifies a number of correlations that may not have been detected
+because of the low power associated with the majority of studies that
+have been performed to date.
 

### 2018-01-25 submission/manuscript.Rmd
ada background edits

@@
 A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
-This meta-analysis focused on identifying biomarkers or individual microbes but did so on a small total number of individuals and only investigated stool.
+This meta-analysis focused on identifying microbial signatures of CRC (biomarkers) but did so on a small total number of individuals and only investigated stool.
 
-Although there has been an intense focus on microbiota-based biomarker discovery for CRC, the candidate genera seem to be numerous.
+Although there has been an intense focus on microbiota-based biomarker discovery for CRC, the number of candidate genera seem to be endless.
 Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
-Yet, even in some of these studies mouth-associated genera are far from the only microbes identified to be associated with CRC.
-These other genera include, but are not necessarily limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012].
-In fact, *Escherichia/Shigella* and *Streptococcus* also have good *in vivo* evidence that individual species within the genus can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017].
-Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential additional markers of CRC with *Bacteroides fragilis* having good mechanistic studies into how this may occur [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014].
-Still, other studies have identified genera like *Roseburia* as increased in CRC while others have reported it to be decreased or have found no difference in any genera at all [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
+Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with CRC.
+These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012].
+In fact, there is good *in vivo* evidence that *Escherichia/Shigella* and *Streptococcus* can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017].
+Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential markers of CRC with good mechanistic studies for the latter [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014].
+A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017].
+However, the sample size (n = 509) is equal to or less than some more recent individual studies investigating the microbiota and CRC, making it hard to know how extrapolatable these findings are.
+That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of CRC that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].
+There are also genera like *Roseburia* that some studies find to be increased in CRC but other studies report either decreased or no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
 
 Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera.
-Some groups have focused on more broad scale community metrics and have found that metrics such as richness are decreased in adenoma versus controls while others have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
-A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017].
-However, it only analyzed a total of 509 stool samples and with some recent studies, that were not included, matching this sample size alone, it is hard to know how extrapolatable these findings truly are.
-That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) that need to be investigated further, since a number of these genera have been found to be enriched in controls and not lesion (adenoma or carcinoma) [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].
+Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of CRC versus controls.
+Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
+Based on the studies mentioned, there seems to be very little overlap between the genera identified to be associated with adenoma and carcinoma, with *Lactobacillus* being one of the few commonalities.
 
-Targeting the identification of biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free.
-Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to the community on the adenoma or carcinoma.
+Targeting the identification of CRC microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications.
+Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities.
 Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
-Sampling the tissue directly may provide clearer answers but is not without limitations.
-Due to bowel prep the communities left for sampling may not be reflective of the resident microbiota, but rather a collection of what is able to keep adhered to the mucosa.
+Sampling the microbiota directly associated with colon tissue directly may provide clearer answers but is not without limitations.
+The community present for sampling following the colonoscopy bowel prep may reflect the better adhered microbiota versus the resident community.
 Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done.
-It is well known that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014].
-Due to these many differences that could arise between stool and tissue, an additional question our meta-analysis aims to answer is whether there are consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.
+It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].
 
 In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of CRC.
-Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not previously provided by earlier meta-analysis studies [@shah_leveraging_2017].
+Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].
 
-Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor or if select members can recapitulate the same information for both adenoma and carcinoma.
-To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma.
-We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma.
-Next, using Random Forest models, we analyzed whether the full community or only the top 5 increased and top 5 decreased RR genera combined resulted in better model classification, based on the area under the curve (AUC).
-Our results suggest that the community changes as disease severity worsens and that this community is important for disease classification.
-Finally, we also examined at what effect and sample size the studies that were used were powered for and the sample size needed to get to an 80% power.
+Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of CRC.
+To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma stages of CRC.
+We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma stages of CRC.
+Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased RR genera resulted in better model classification, based on the area under the curve (AUC).
+Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionaly accepted 80% power.
+Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the studies are underpowered for assessing small effect sizes.
 

### 2018-01-25 submission/manuscript.tex
ada background edits

@@
 identified using multiple data sets {[}11{]}. This meta-analysis focused
-on identifying biomarkers or individual microbes but did so on a small
-total number of individuals and only investigated stool.
+on identifying microbial signatures of CRC (biomarkers) but did so on a
+small total number of individuals and only investigated stool.
 
 Although there has been an intense focus on microbiota-based biomarker
-discovery for CRC, the candidate genera seem to be numerous. Some
-studies point towards mouth-associated genera such as
+discovery for CRC, the number of candidate genera seem to be endless.
+Some studies point towards mouth-associated genera such as
 \emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
 \emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
-some of these studies mouth-associated genera are far from the only
-microbes identified to be associated with CRC. These other genera
-include, but are not necessarily limited to, \emph{Providencia},
-\emph{Mogibacterium}, \emph{Enterococcus}, \emph{Escherichia/Shigella},
-\emph{Klebsiella}, and \emph{Streptococcus} {[}14--16{]}. In fact,
-\emph{Escherichia/Shigella} and \emph{Streptococcus} also have good
-\emph{in vivo} evidence that individual species within the genus can be
-important in the pathogenesis of CRC {[}5,19,20{]}. Other studies have
-also identified \emph{Akkermansia muciniphila} and \emph{Bacteroides
-fragilis} as potential additional markers of CRC with \emph{Bacteroides
-fragilis} having good mechanistic studies into how this may occur
-{[}15,21,22{]}. Still, other studies have identified genera like
-\emph{Roseburia} as increased in CRC while others have reported it to be
-decreased or have found no difference in any genera at all
-{[}15,18,23,24{]}.
+these studies, mouth-associated genera are far from the only microbes
+identified to be associated with CRC. These other genera include, but
+are not limited to, \emph{Providencia}, \emph{Mogibacterium},
+\emph{Enterococcus}, \emph{Escherichia/Shigella}, \emph{Klebsiella}, and
+\emph{Streptococcus} {[}14--16{]}. In fact, there is good \emph{in vivo}
+evidence that \emph{Escherichia/Shigella} and \emph{Streptococcus} can
+be important in the pathogenesis of CRC {[}5,19,20{]}. Other studies
+have also identified \emph{Akkermansia muciniphila} and
+\emph{Bacteroides fragilis} as potential markers of CRC with good
+mechanistic studies for the latter {[}15,21,22{]}. A recent
+meta-analysis confirmed the correlations of certain mouth-associated
+genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
+However, the sample size (n = 509) is equal to or less than some more
+recent individual studies investigating the microbiota and CRC, making
+it hard to know how extrapolatable these findings are. That particular
+meta-analysis also added more potential microbial associations to both
+carcinoma (\emph{Pantoea agglomerans} \emph{Ruminococcus},
+\emph{Lactobacillus}) and adenoma (\emph{Prevotella},
+\emph{Methanosphaera}, \emph{Succinovibrio}, \emph{Haemophilus
+parainfluenzae}, \emph{Ruminococcus}, \emph{Lactobacillus}) stages of
+CRC that need to be investigated further, since a number of these genera
+have been found to be enriched in controls and not disease
+{[}13,16,17{]}. There are also genera like \emph{Roseburia} that some
+studies find to be increased in CRC but other studies report either
+decreased or no difference {[}15,18,23,24{]}.
 
 Most of these studies have focused on carcinoma but the adenoma
-observations are not any clearer at identifying candidate genera. Some
-groups have focused on more broad scale community metrics and have found
-that metrics such as richness are decreased in adenoma versus controls
-while others have identified \emph{Lactococcus}, \emph{Pseudomonas},
+observations are not any clearer at identifying candidate genera. Groups
+focusing on broad scale community metrics have found that metrics such
+as richness are decreased in the adenoma stage of CRC versus controls.
+Other studies have identified \emph{Lactococcus}, \emph{Pseudomonas},
 \emph{Acidovorax}, \emph{Cloacibacterium}, \emph{Helicobacter},
 \emph{Lactobacillus}, \emph{Bilophila}, \emph{Desulfovibrio}, and
-\emph{Mogibacterium} to be increased in adenoma {[}25--27{]}. A recent
-meta-analysis confirmed the correlations of certain mouth-associated
-genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
-However, it only analyzed a total of 509 stool samples and with some
-recent studies, that were not included, matching this sample size alone,
-it is hard to know how extrapolatable these findings truly are. That
-particular meta-analysis also added more potential microbial
-associations to both carcinoma (\emph{Pantoea agglomerans}
-\emph{Ruminococcus}, \emph{Lactobacillus}) and adenoma
-(\emph{Prevotella}, \emph{Methanosphaera}, \emph{Succinovibrio},
-\emph{Haemophilus parainfluenzae}, \emph{Ruminococcus},
-\emph{Lactobacillus}) that need to be investigated further, since a
-number of these genera have been found to be enriched in controls and
-not lesion (adenoma or carcinoma) {[}13,16,17{]}.
-
-Targeting the identification of biomarkers within stool seems logical
-since it offers an easy and cost-effective way to stratify risk and the
-current gold standard for diagnosis, a colonoscopy, can be
-time-consuming and is not risk free. Although stool represents an easy
-and less invasive way to assess risk, it is not clear how reflective
-this sample is to the community on the adenoma or carcinoma. Some
-studies have tried to assess this in health and disease but are limited
-by their sample size {[}18,28{]}. Sampling the tissue directly may
-provide clearer answers but is not without limitations. Due to bowel
-prep the communities left for sampling may not be reflective of the
-resident microbiota, but rather a collection of what is able to keep
-adhered to the mucosa. Additionally, these samples contain more host
-DNA, potentially limiting the types of analysis that can be done. It is
-well known that low biomass samples can be very difficult to work with
-and results can end up being study dependent due to the randomness of
-contamination {[}29{]}. Due to these many differences that could arise
-between stool and tissue, an additional question our meta-analysis aims
-to answer is whether there are consistent patterns that emerge across
-studies regardless of whether they used stool or tissue samples.
+\emph{Mogibacterium} to be increased in adenoma {[}25--27{]}. Based on
+the studies mentioned, there seems to be very little overlap between the
+genera identified to be associated with adenoma and carcinoma, with
+\emph{Lactobacillus} being one of the few commonalities.
+
+Targeting the identification of CRC microbial biomarkers within stool
+seems logical since it offers an easy and cost-effective way to stratify
+risk and the current gold standard for diagnosis, a colonoscopy, can be
+time-consuming and is not without risk of complications. Although stool
+represents an easy and less invasive way to assess risk, it is not clear
+how well this sample reflects adenoma- and carcinoma- associated
+microbial communities. Some studies have tried to assess this in health
+and disease but are limited by their sample size {[}18,28{]}. Sampling
+the microbiota directly associated with colon tissue directly may
+provide clearer answers but is not without limitations. The community
+present for sampling following the colonoscopy bowel prep may reflect
+the better adhered microbiota versus the resident community.
+Additionally, these samples contain more host DNA, potentially limiting
+the types of analysis that can be done. It is well known that low
+biomass samples can be very difficult to work with and results can be
+study dependent due to the randomness of contamination {[}29{]}.
 
@@
 detrimental in CRC and whether broad bacterial community measurements
-can account for these changes that were not previously provided by
-earlier meta-analysis studies {[}11{]}.
-
-Using both feces (n = 1737) and tissue (492 samples from 350
+can account for these changes that were not provided by earlier
+meta-analysis studies {[}11{]}.
+
+Using both feces (n = 1737) and colon tissues (492 samples from 350
 individuals) totaling over 2229 total samples across 14 studies
@@
 and scope of the previous meta-analysis to investigate whether the
-bacterial community is an important risk factor or if select members can
-recapitulate the same information for both adenoma and carcinoma. To
-accomplish this we first assessed whether bacterial diversity changes
-throughout disease (control to adenoma to carcinoma) and if it results
-in an increased relative risk (RR) for adenoma or carcinoma. We then
+bacterial community or specific members are more important risk factors
+for both adenoma and carcinoma stages of CRC. To accomplish this we
+first assessed whether bacterial diversity changes throughout disease
+(control to adenoma to carcinoma) and if it results in an increased
+relative risk (RR) for adenoma or carcinoma stages of CRC. We then
 assessed what genera, if any, increase or decrease the RR of adenoma or
-carcinoma. Next, using Random Forest models, we analyzed whether the
-full community or only the top 5 increased and top 5 decreased RR genera
-combined resulted in better model classification, based on the area
-under the curve (AUC). Our results suggest that the community changes as
-disease severity worsens and that this community is important for
-disease classification. Finally, we also examined at what effect and
-sample size the studies that were used were powered for and the sample
-size needed to get to an 80\% power.
+carcinoma stages of CRC. Next, using Random Forest models, we analyzed
+whether the full community or only the combined top 5 increased and top
+5 decreased RR genera resulted in better model classification, based on
+the area under the curve (AUC). Finally, we also examined at what effect
+and sample size the studies used were powered for and the sample size
+needed to get to the traditionaly accepted 80\% power. Our results from
+these analyses suggests that the bacterial community changes as disease
+severity worsens, that specific members are important for disease
+classification, and that many of the studies are underpowered for
+assessing small effect sizes.
 

### 2018-01-25 submission/manuscript.Rmd
added first two sections of results ada suggestions

@@
 
-**Results.** Fecal samples had a significant decrease from those without disease to adenoma (less severe disease) to carcinoma (severe disease) for both Shannon diversity and evenness after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
+**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness after correcting for study effect and variable region sequenced with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
 This reduction in evenness translated into small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased relative risk for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
@@
 ***Lower Bacterial Diversity is Associated with Increased RR of Carcinomas:***
-Power transforming and Z-score normalizing Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity for the entire data set allowed for the combination of data from all studies to assess differences in broad scale community metrics as disease severity worsens.
+To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized.
+These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively.
 Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity.
-In stool, there was a significant decrease in both evenness and Shannon diversity from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A].
-We next tested whether these detectable differences in community resulted in significant increases in RR of lesion (adenoma or carcinoma).
+In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A].
+We next tested whether the detectable differences in community significantly increased in RR of having an adenoma or carcinoma.
 For fecal samples, a decrease versus the overal median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
-Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma versus control and adenoma versus control [Table S1 & S2].
+Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].
 
-Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased.
-For tissue, there were no significant changes in OTU richness, evenness, and Shannon diversity from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1B & C].
-When analyzing the RR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) tissue samples, for both adenoma and carcinoma there was no significant change in RR based on lower than median values for OTU richness, evenness, and Shannon diversity [Table S3-S5].
-Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were also identified in unmatched tissue samples  for both those with adenoma and carcinoma [Table S6 & S7].
+Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and 16S variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased.
+For colon tissue, there were no significant changes in OTU richness, evenness, or Shannon diversity as disease severity progressed from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1B & C].
+We next analyzed the RR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) colon tissue samples.
+For individuals at either an adenoma or carcinoma stage of disease there was no significant change in RR based on lower than median values for OTU richness, evenness, and Shannon diversity [Table S3-S5].
+Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were identified in unmatched tissue samples, for those at either adenoma or carcinoma stage of CRC [Table S6 & S7].
 For studies with matched samples no differences in bacterial community were observed when assessed with PERMANOVA [Table S6 & S7].
+These tissue results suggest that the microbiota within an individual are similar to each other regardless of disease status.
 
-***Mouth-Associated Genera are Associated with an Increased RR of Lesion:***
-Based on the small increase in RR using OTU richness and Shannon diversity, we assessed if being higher than the median for any specific genera resulted in an altered RR for adenoma or carcinoma in stool and tissue.
-To investigate this we analyzed all common genera across each study, in tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR.
-For both stool and tissue none of the genera in the top 5 most significantly increased RR were the same between adenoma and carcinoma [Figure 3].
-Mouth-associated genera were in the top 5 genera associated with an increased RR of adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and tissue.
-Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and tissue [Figure 3].
-Overall, there was little direct overlap of the top 5 increased or decreased RR genera between stool and tissue.
+***Mouth-Associated Genera are Associated with an Increased RR of CRC:***
+Next we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity.
+To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR.
+The top 5 most significantly increased RR were the same between adenoma and carcinoma for both stool and colon tissue [Figure 3].
+Mouth-associated genera were commonly found in the top 5 genera associated with an increased RR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples.
+Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3].
+Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased RR genera between both stages and sample site.
 
-For adenoma, there was almost no overlap between genera with a RR P-value of less than 0.05 and when they were similar the RR was in opposite directions (e.g.
+When observing RRs with a P-value less than 0.05 there was almost no overlap between genera from stool or colon tissue and when they were similar the RR was in opposite directions (e.g.
 *Lactococcus*) [Table S8 & S9].
-Many of the genera that had a P-value under 0.05 for either an increased or dereased RR are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
-For carcinoma, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both tissue and stool [Table S10 & S11].
-Although the most significant increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].
+Many of the genera that had RRs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
+For carcinoma stage of CRC, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both colon tissue and stool samples [Table S10 & S11].
+Finally, these data suggest that the most significantly increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].
 

### 2018-01-25 submission/manuscript.tex
added first two sections of results ada suggestions

@@
 
-\textbf{Results.} Fecal samples had a significant decrease from those
-without disease to adenoma (less severe disease) to carcinoma (severe
-disease) for both Shannon diversity and evenness after correcting for
-study effect and variable region sequenced (P-value \textless{} 0.05).
-This reduction in evenness translated into small increases in relative
-risk for adenoma (P-value = 0.032) and carcinoma stages of CRC (P-value
-= 0.00034) while the reduction in Shannon diversity only translated into
-an increased relative risk for developing carcinomas (P-value = 0.0047).
-Increases in mouth-associated microbes were commonly in the top 5 most
-significantly increased relative risk of adenoma and carcinoma for both
-stool and tissue samples. A prediction model for adenoma and carcinoma
-was built using either the whole community or selected genera with
-highest and lowest relative risk from fecal and tissue samples. Both
-approaches resulted in similar classification success according to Area
-Under the Curve (AUC) regardless of whether genera or OTUs were used to
-build the model. The most important groups within the full community
-models consistently belonged to genera such as \emph{Ruminococcus},
-\emph{Bacteroides}, and \emph{Roseburia} across studies. Although a
-number of associations between the microbiota and CRC were identified,
-the majority of studies that we used in this meta-analysis were only
-individually adequately powered for large effect sizes.
+\textbf{Results.} Fecal samples had a significant decrease for both
+Shannon diversity and evenness after correcting for study effect and
+variable region sequenced with more severe disease (P-value \textless{}
+0.05). This reduction in evenness translated into small increases in
+relative risk for adenoma (P-value = 0.032) and carcinoma stages of CRC
+(P-value = 0.00034) while the reduction in Shannon diversity only
+translated into an increased relative risk for developing carcinomas
+(P-value = 0.0047). Increases in mouth-associated microbes were commonly
+in the top 5 most significantly increased relative risk of adenoma and
+carcinoma for both stool and tissue samples. A prediction model for
+adenoma and carcinoma was built using either the whole community or
+selected genera with highest and lowest relative risk from fecal and
+tissue samples. Both approaches resulted in similar classification
+success according to Area Under the Curve (AUC) regardless of whether
+genera or OTUs were used to build the model. The most important groups
+within the full community models consistently belonged to genera such as
+\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
+studies. Although a number of associations between the microbiota and
+CRC were identified, the majority of studies that we used in this
+meta-analysis were only individually adequately powered for large effect
+sizes.
 
@@
 \textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
-of Carcinomas:}} Power transforming and Z-score normalizing Operational
-Taxonomic Unit (OTU) richness, evenness, and Shannon diversity for the
-entire data set allowed for the combination of data from all studies to
-assess differences in broad scale community metrics as disease severity
-worsens. Using linear mixed-effect models to control for study and
-variable region we assessed whether OTU richness, evenness, or Shannon
-diversity changed in a step-wise manner with disease severity. In stool,
-there was a significant decrease in both evenness and Shannon diversity
-from control to adenoma to carcinoma (P-value = 0.025 and 0.043,
-respectively) {[}Figure 1A{]}. We next tested whether these detectable
-differences in community resulted in significant increases in RR of
-lesion (adenoma or carcinoma). For fecal samples, a decrease versus the
-overal median in evenness resulted in a significantly increased RR for
-carcinoma (RR = 1.36 (1.15 - 1.61), P-value = 0.00034) and adenoma (RR =
-1.16 (1.01 - 1.34), P-value = 0.032) while a decrease versus the overal
-median in Shannon diversity only increased the RR for carcinoma (RR =
-1.33 (1.09 - 1.62), P-value = 0.0047) {[}Figure 2{]}. Using the
-Bray-Curtis distance metric and PERMANOVA, it was also possible to
-identify significant bacterial community changes, in specific studies,
-for both carcinoma versus control and adenoma versus control {[}Table S1
-\& S2{]}.
+of Carcinomas:}} To assess differences in broad scale community metrics
+as disease severity worsens Operational Taxonomic Unit (OTU) richness,
+evenness, and Shannon diversity measurements were power transformed and
+Z-score normalized. These metrics are commonly used to assess the total
+number of OTUs, the equality of their abundance, and the overall
+diversity, respectively. Using linear mixed-effect models to control for
+study and variable region we assessed whether OTU richness, evenness, or
+Shannon diversity changed in a step-wise manner with disease severity.
+In stool, there was a significant decrease in both evenness and Shannon
+diversity as disease severity moved from control to adenoma to carcinoma
+(P-value = 0.025 and 0.043, respectively) {[}Figure 1A{]}. We next
+tested whether the detectable differences in community significantly
+increased in RR of having an adenoma or carcinoma. For fecal samples, a
+decrease versus the overal median in evenness resulted in a
+significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
+P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
+0.032) while a decrease versus the overal median in Shannon diversity
+only increased the RR for carcinoma (RR = 1.33 (1.09 - 1.62), P-value =
+0.0047) {[}Figure 2{]}. Using the Bray-Curtis distance metric and
+PERMANOVA, it was also possible to identify significant bacterial
+community changes, in specific studies, for both carcinoma-associated
+and adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.
 
@@
 models were used on the transformed combined data to control for study,
-re-sampling of the same individual, and variable region to test whether
-OTU richness, evenness, or Shannon diversity changed in a step-wise
-manner as disease severity increased. For tissue, there were no
-significant changes in OTU richness, evenness, and Shannon diversity
-from control to adenoma to carcinoma (P-value \textgreater{} 0.05)
-{[}Figure 1B \& C{]}. When analyzing the RR, for matched (unaffected
-tissue and an adenoma or carcinoma from the same individual) and
-unmatched (control and adenoma or carcinoma tissue not from the same
-individual) tissue samples, for both adenoma and carcinoma there was no
-significant change in RR based on lower than median values for OTU
-richness, evenness, and Shannon diversity {[}Table S3-S5{]}. Similar to
-stool samples, significant differences in bacterial community, assessed
-by PERMANOVA, were also identified in unmatched tissue samples for both
-those with adenoma and carcinoma {[}Table S6 \& S7{]}. For studies with
-matched samples no differences in bacterial community were observed when
-assessed with PERMANOVA {[}Table S6 \& S7{]}.
+re-sampling of the same individual, and 16S variable region to test
+whether OTU richness, evenness, or Shannon diversity changed in a
+step-wise manner as disease severity increased. For colon tissue, there
+were no significant changes in OTU richness, evenness, or Shannon
+diversity as disease severity progressed from control to adenoma to
+carcinoma (P-value \textgreater{} 0.05) {[}Figure 1B \& C{]}. We next
+analyzed the RR, for matched (unaffected tissue and an adenoma or
+carcinoma from the same individual) and unmatched (control and adenoma
+or carcinoma tissue not from the same individual) colon tissue samples.
+For individuals at either an adenoma or carcinoma stage of disease there
+was no significant change in RR based on lower than median values for
+OTU richness, evenness, and Shannon diversity {[}Table S3-S5{]}. Similar
+to stool samples, significant differences in bacterial community,
+assessed by PERMANOVA, were identified in unmatched tissue samples, for
+those at either adenoma or carcinoma stage of CRC {[}Table S6 \& S7{]}.
+For studies with matched samples no differences in bacterial community
+were observed when assessed with PERMANOVA {[}Table S6 \& S7{]}. These
+tissue results suggest that the microbiota within an individual are
+similar to each other regardless of disease status.
 
 \textbf{\emph{Mouth-Associated Genera are Associated with an Increased
-RR of Lesion:}} Based on the small increase in RR using OTU richness and
-Shannon diversity, we assessed if being higher than the median for any
-specific genera resulted in an altered RR for adenoma or carcinoma in
-stool and tissue. To investigate this we analyzed all common genera
-across each study, in tissue or stool, and assessed whether a relative
+RR of CRC:}} Next we asked if being higher than the median relative
+abundance, for any specific genera, resulted in an altered RR for
+adenoma or carcinoma, in stool and colon tissue, due to our previous
+observations of small increases in RR using OTU richness and Shannon
+diversity. To investigate this we analyzed all common genera across each
+study, in colon tissue or stool, and assessed whether a relative
 abundance higher than the median results in an increase or decrease in
-RR. For both stool and tissue none of the genera in the top 5 most
-significantly increased RR were the same between adenoma and carcinoma
-{[}Figure 3{]}. Mouth-associated genera were in the top 5 genera
-associated with an increased RR of adenoma (\emph{Pyramidobacter}
-{[}Figure 3A{]} and \emph{Rothia} {[}Figure 3C{]}) and carcinoma
-(\emph{Fusobacterium}, \emph{Parvimonas}, \emph{Porphyromonas}, and
-\emph{Peptostreptococcus} {[}Figure 3B{]} and \emph{Fusobacterium}
-{[}Figure 3D{]}) for both stool and tissue. Conversely, genera commonly
-associated with a normal gastrointestinal tract were correlated with a
-decreased RR for both adenoma and carcinoma for both stool and tissue
-{[}Figure 3{]}. Overall, there was little direct overlap of the top 5
-increased or decreased RR genera between stool and tissue.
-
-For adenoma, there was almost no overlap between genera with a RR
-P-value of less than 0.05 and when they were similar the RR was in
-opposite directions (e.g. \emph{Lactococcus}) {[}Table S8 \& S9{]}. Many
-of the genera that had a P-value under 0.05 for either an increased or
-dereased RR are also highly prevalent in contamination, specifically,
-\emph{Novosphingobium}, \emph{Selemonas}, and \emph{Achromobacter}
-{[}Figure 3 \& Table S8-S9{]}. For carcinoma, certain mouth-associated
-genera (\emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Peptostreptococcus}) had a high RR for both tissue and stool
-{[}Table S10 \& S11{]}. Although the most significant increased RR
-genera for tissue was \emph{Camplyobacter} while in stool it was
-\emph{Peptostreptococcus} {[}Table S10 \& S11{]}.
+RR. The top 5 most significantly increased RR were the same between
+adenoma and carcinoma for both stool and colon tissue {[}Figure 3{]}.
+Mouth-associated genera were commonly found in the top 5 genera
+associated with an increased RR of having an adenoma
+(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
+3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
+\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and colon tissue
+samples. Conversely, genera commonly associated with a normal
+gastrointestinal tract were correlated with a decreased RR for both
+adenoma and carcinoma for both stool and colon tissue samples {[}Figure
+3{]}. Even though mouth-associated genera were identified across disease
+stage, there was little direct overlap of the top 5 increased or
+decreased RR genera between both stages and sample site.
+
+When observing RRs with a P-value less than 0.05 there was almost no
+overlap between genera from stool or colon tissue and when they were
+similar the RR was in opposite directions (e.g. \emph{Lactococcus})
+{[}Table S8 \& S9{]}. Many of the genera that had RRs with a P-value
+under 0.05 for colon tissue are also highly prevalent in contamination,
+specifically, \emph{Novosphingobium}, \emph{Selemonas}, and
+\emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma stage
+of CRC, certain mouth-associated genera (\emph{Fusobacterium},
+\emph{Parvimonas}, \emph{Peptostreptococcus}) had a high RR for both
+colon tissue and stool samples {[}Table S10 \& S11{]}. Finally, these
+data suggest that the most significantly increased RR genera for tissue
+was \emph{Camplyobacter} while in stool it was \emph{Peptostreptococcus}
+{[}Table S10 \& S11{]}.
 

### 2018-01-26 submission/manuscript.Rmd
discussion and conclusion -- ada suggestions

@@
 ***Select Genera Models can Recapitulate Whole Community Models:***
-Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera based on RR and P-value significance and tested this using two approaches.
-The first approach used genus level data and tested whether there were any differences in AUC between all genera and selected genera during training on one study or testing on all other studies.
-The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model AUC across studies when comparing all OTUs versus only OTUs that taxonomically classified to selected genera.
+Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera.
+We selected these genera based on their RR and P-value significance and used two approaches to test this question.
+The first approach used genus level data and tested for differences in AUC between all genera and selected genera.
+A single study was used for training the model prior to testing on all other studies and this was repeated for every study in the meta-analysis.
+The second approach used OTU level data and tested for a generalized decrease in the 10-fold cross validation (CV) model AUC which is a common approach used to guard against over-fitting.
+This was applied across study and the AUC of the all OTUs model was compared against the model that used only OTUs that taxonomically classified to selected genera.
 
-For the genera-based models the training set median AUC for model classification was similar for select genera and full genera models, for both tissue and stool studies [Figure S2-S3].
-With respect to the test sets, comprised of genera data from other studies, both all and select genera models had a similar ability to detect adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6].
-Conversely, the OTU-based models showed a slight decrease in median AUC between the full and select models, with one exception being carcinoma models for matched tissue [Figure 4 & 5].
+For the first approach using the genera-based models, the training set median AUC for model classification was similar for both the select genera and full genera models, for both tissue and stool studies [Figure S2-S3].
+When analyzing the tests sets that were comprised of genera data from other studies, both models had a similar ability to detect individuals with adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6].
+Conversely, the second approach that used OTU-based models showed a slight decrease in median AUC between the full and select models, with one exception to this generalization being the carcinoma models for matched colon tissue [Figure 4 & 5].
 
-In stool, the most common genera in the full genera-based models belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
-Although all mouth-associated genera were present in carcinoma, none were present in the majority of studies and *Fusobacterium* was the only genus present in adenoma [Figure 6A & B].
-For the full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].
+In stool, the most common genera in the top 10 most imporant variables, in models for the first approach, that used full genera-based models were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
+Regardless of sample type, mouth-associated genera were present in models for the carcinoma stage of CRC [Figure 6A & B].
+Yet, none were present in the majority of studies and *Fusobacterium* was the only genus present in the adenoma stage of CRC [Figure 6A & B].
+For the second approach that utilized full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].
 
-Unlike stool, the tissue full genera-based Random Forest models showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 across study [Figure S7].
-The vast majority of the top 10 genera and OTUs occured in a study specific manner for these tissue based Random Forest models.
-For both the full genera and OTU Random Forest models of adenoma and carcinoma, there appears to be very little overlap in the top 10 most important variables between stool and tissue [Figure 6 & S7].
-This discordance between stool and tissue also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of Random Forest models using matched versus unmatched tissue samples [Figure S7B-C & S7E-F].
+Unlike the stool-based Random Forest models, the tissue-based models, for the full genera from the first approach, showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 most important model variables across study [Figure S7].
+The vast majority of the top 10 model variables for the genera- and OTU-based models using colon tissue tended to be study specific.
+Further, there was very little overlap in the top 10 important variables between adenoma and carcinoma stage models, regardless of whether colon tissue or stool was used  [Figure 6 & S7].
+This discordance between stool and colon tissue samples also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of matched versus unmatched colon tissue Random Forest models [Figure S7B-C & S7E-F].
+This suggests that either the colon tissue microbiota is study and person dependent or that kit and/or other types of contamination associated with low biomass samples may be skewing the results.
 
 ***CRC Studies are Underpowered for Detecting Small Effect Sizes:***
-Based on our reported results we assessed whether the studies analyzed were realistically powered to identify small, medium, and large scale differences between cases and controls.
-When assessing the power of each study at different effect sizes the majority of studies, for both adenoma and carcinoma, acheived 80% power to detect a 30% or greater difference between groups [Figure 7A & B].
-No study that was analyzed had the standard 80% power to detect an effect size difference that was equal to or below 10% [Figure 7A & B].
-In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for each arm [Figure 7C].
+Next, we assessed how much confidence should be placed in the reported outcomes from each individual study by calculating the ability to detect a difference (power) and sample size needed for small, medium, and large effect size differences between cases and controls.
+When assessing the power of each study at different effect sizes the majority of studies acheived 80% power to detect a 30% or greater difference between groups [Figure 7A & B].
+No study that we analyzed had the standard 80% power to detect an effect size difference equal to or below 10% [Figure 7A & B].
+In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for both the case and control arms [Figure 7C]
 
@@
 
-Our study identifies clear differences in diversity, both at the community level and for individual genera, that are present in patients with and without CRC [Figure 1-3].
-Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of lesion.
-Even though mouth-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma.
-These mouth-associated genera are clearly important to carcinoma classification but our observations suggest that accounting for other community members can increase the ability of models to make predictions.
+Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with and without CRC [Figure 1-3].
+Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of CRC.
+Even though mouth-associated genera increased the RR of having a carcinoma, they did not consistently increase the RR of having an adenoma.
+Additionally, our observations suggest that by combining mouth-associated and CRC protective microbes we can classify either adenoma or carcinoma stage of disease as well as models that use the full community.
 
-The data presented herein supports the importance of select genera for carcinoma but not necessarily adenoma formation.
-Our observations show that when only using the select genera to create models for carcinoma the AUC are similar to using all genera or OTUs [Figure 4 & 5].
-This suggests that an interplay between a select number of potentially protective and exacerbating microbes, within the GI community, are crucial for carcinoma formation.
+The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, formation.
+The results that we have presented show that the both genera and OTU select and full models, for the carcinoma stage of CRC, had similar AUCs [Figure 4 & 5].
+This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community is crucial for carcinoma formation.
 Importantly, it suggests that there may be key members of the GI community that might be studied further to potentially reduce the risk of carcinoma.
-Conversely, using the present data, it is clear that new approaches may need to be used to identify members of the community that are associated with adenoma.
-Regardless of sample type and whether a full or select model was used our Random Forest models consistently performed poorly.
-Yet, the step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1].
+Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma stage of disease.
+Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly.
+Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly [Figure 1].
 This change in diversity, at this early stage of disease, could be focal to the adenoma itself.
-One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.
+One possible explanation is that how the host interacts with these subtle changes at early stages of the disease is what leads to a thoroughly dysfunctional community that is supportive of CRC genesis.
 
-Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across study [Figure 6].
-Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be disciminative across the different studies for both adenoma and carcinoma [Figure 6].
-This data would suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community.
-Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma.
+Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 6].
+Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both adenoma and carcinoma [Figure 6].
+These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community.
+Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.
 
-The tissue studies did not provide a clearer understanding of how the microbiota may be associated with lesion.
-Generally, the full OTU-based models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study.
-Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the full CRC models [Figure S7B-C & E-F].
-The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study.
+The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with CRC Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F].
+Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S7B-C & E-F].
+The majority of the colon tissue-based results seem to be study specific with many of top 10 taxa being present only in a single study.
 Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying.
-The low bacterial biomass of tissue samples coupled with potential contamination could ultimately explain why the these results seem to be more sporadic than the stool results.
+The low bacterial biomass of tissue samples coupled with potential contamination could explain why the these results seem to be more sporadic than the stool results.
 
-One important caveat to this study is that even though specific genera associated with species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp.
-*gallolyticus* were not identified, it does not necessarily mean these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017].
-Since we are limited in our aggregation of the data, across study, to the genus level, it is not possible to clearly delineate what specific species are contributing to overall disease progression.
+One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp.
+*gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017].
+Since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression.
 Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp.
 *gallolyticus*.
-As an example, for stool, the full RF models consistently identified the genus *Bacteroides* as well as OTUs that classified as *Bacteroides* to be important model components across studies.
-Suggesting that even though *Bacteroides* may not increase the RR of lesion and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.
-Additionaly, *Streptococcus gallolyticus* subsp.
-*gallolyticus* is a mouth-associated micobe and our study clearly shows that regardless of whether the sample is tissue or stool, mouth-associated genera are commonly associated with an increased RR of both adenoma and carcinoma.
+As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies.
+This suggests that even though *Bacteroides* may not increase the RR of CRC and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.
+Additionally, *Streptococcus gallolyticus* subsp.
+*gallolyticus* is a mouth-associated micobe, and the results from this study suggest that regardless sample type, mouth-associated genera are commonly associated with an increased RR for both adenoma and carcinoma stage of disease.
 
-The associations between the microbiota and adenoma are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes.
+The associations between the microbiota and adenoma stage of disease are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes.
 None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls.
-A small effect size may well be the scope in which differences consistently occur in adenoma based on the results observed within our meta-analysis.
-Future studies investigating adenoma and the microbiota need to take power into consideration if we are to reproducibly study whether the microbiota contributes to polyp formation.
-In contrast to adenoma, our observations suggest that most studies analyzed have sufficient power to detect many changes in carcinoma because of large effect size differences between cases and controls [Figure 7].
+The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and adenoma stage of disease.
+Future studies investigating adenoma stage and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation.
+In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 7].
 
@@
 
-By aggregating together a large collection of studies from both feces and tissue, we are able to provide evidence in support of the importance of the bacterial community in CRC.
-Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest RR of carcinoma.
-Conversely, no conclusive signal with these mouth-associated micobes could be detected for adenoma.
-Our observations also highlight the importance of power and sample number considerations when undertaking investigations into the microbiota and adenoma due to the subtle changes in the community.
-Overall, the microbiota associations with carcinoma are much stronger than the associations with adenoma.
+By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in CRC.
+Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest RR of having a carcinoma.
+No conclusive signal with these mouth-associated micobes could be detected for adenoma stage of disease.
+Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to the subtle changes in the community.
+Overall, associations of microbiota with the carcinoma stage of CRC are much stronger than those with the adenoma stage.
 

### 2018-01-26 submission/manuscript.tex
discussion and conclusion -- ada suggestions

@@
 diversity metrics we assessed whether the bacterial community was better
-at classifying disease versus only a select group of genera based on RR
-and P-value significance and tested this using two approaches. The first
-approach used genus level data and tested whether there were any
-differences in AUC between all genera and selected genera during
-training on one study or testing on all other studies. The second
-approach used OTU level data and tested whether there was a generalized
-decrease in the 10-fold cross validation (CV) model AUC across studies
-when comparing all OTUs versus only OTUs that taxonomically classified
-to selected genera.
-
-For the genera-based models the training set median AUC for model
-classification was similar for select genera and full genera models, for
-both tissue and stool studies {[}Figure S2-S3{]}. With respect to the
-test sets, comprised of genera data from other studies, both all and
-select genera models had a similar ability to detect adenomas or
-carcinomas, with the select genera models performing better in some
-instances {[}Figure S4-S6{]}. Conversely, the OTU-based models showed a
+at classifying disease versus only a select group of genera. We selected
+these genera based on their RR and P-value significance and used two
+approaches to test this question. The first approach used genus level
+data and tested for differences in AUC between all genera and selected
+genera. A single study was used for training the model prior to testing
+on all other studies and this was repeated for every study in the
+meta-analysis. The second approach used OTU level data and tested for a
+generalized decrease in the 10-fold cross validation (CV) model AUC
+which is a common approach used to guard against over-fitting. This was
+applied across study and the AUC of the all OTUs model was compared
+against the model that used only OTUs that taxonomically classified to
+selected genera.
+
+For the first approach using the genera-based models, the training set
+median AUC for model classification was similar for both the select
+genera and full genera models, for both tissue and stool studies
+{[}Figure S2-S3{]}. When analyzing the tests sets that were comprised of
+genera data from other studies, both models had a similar ability to
+detect individuals with adenomas or carcinomas, with the select genera
+models performing better in some instances {[}Figure S4-S6{]}.
+Conversely, the second approach that used OTU-based models showed a
 slight decrease in median AUC between the full and select models, with
-one exception being carcinoma models for matched tissue {[}Figure 4 \&
-5{]}.
-
-In stool, the most common genera in the full genera-based models
-belonged to genera such as \emph{Ruminococcus}, \emph{Bacteroides}, and
-\emph{Roseburia} {[}Figure 6A \& B{]}. Although all mouth-associated
-genera were present in carcinoma, none were present in the majority of
-studies and \emph{Fusobacterium} was the only genus present in adenoma
-{[}Figure 6A \& B{]}. For the full OTU-based models,
-\emph{Ruminococcaceae} was present in the top 10 consistently for both
-adenoma and carcinoma models while \emph{Roseburia} was only present in
-many adenoma models and \emph{Bacteroides} was present in the
-overwhelming majority of the carcinoma models {[}Figure 6C \& 6D{]}.
-
-Unlike stool, the tissue full genera-based Random Forest models showed
-no consistent representation of \emph{Ruminococcaceae},
-\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} in the top
-10 across study {[}Figure S7{]}. The vast majority of the top 10 genera
-and OTUs occured in a study specific manner for these tissue based
-Random Forest models. For both the full genera and OTU Random Forest
-models of adenoma and carcinoma, there appears to be very little overlap
-in the top 10 most important variables between stool and tissue
-{[}Figure 6 \& S7{]}. This discordance between stool and tissue also
-applies to the mouth-associated genera with one noticeable skew being
-that \emph{Fusobacterium} and \emph{Fusobacteriaceae} occur more often
-in the top 10 of Random Forest models using matched versus unmatched
-tissue samples {[}Figure S7B-C \& S7E-F{]}.
+one exception to this generalization being the carcinoma models for
+matched colon tissue {[}Figure 4 \& 5{]}.
+
+In stool, the most common genera in the top 10 most imporant variables,
+in models for the first approach, that used full genera-based models
+were \emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia}
+{[}Figure 6A \& B{]}. Regardless of sample type, mouth-associated genera
+were present in models for the carcinoma stage of CRC {[}Figure 6A \&
+B{]}. Yet, none were present in the majority of studies and
+\emph{Fusobacterium} was the only genus present in the adenoma stage of
+CRC {[}Figure 6A \& B{]}. For the second approach that utilized full
+OTU-based models, \emph{Ruminococcaceae} was present in the top 10
+consistently for both adenoma and carcinoma models while
+\emph{Roseburia} was only present in many adenoma models and
+\emph{Bacteroides} was present in the overwhelming majority of the
+carcinoma models {[}Figure 6C \& 6D{]}.
+
+Unlike the stool-based Random Forest models, the tissue-based models,
+for the full genera from the first approach, showed no consistent
+representation of \emph{Ruminococcaceae}, \emph{Ruminococcus},
+\emph{Bacteroides}, and \emph{Roseburia} in the top 10 most important
+model variables across study {[}Figure S7{]}. The vast majority of the
+top 10 model variables for the genera- and OTU-based models using colon
+tissue tended to be study specific. Further, there was very little
+overlap in the top 10 important variables between adenoma and carcinoma
+stage models, regardless of whether colon tissue or stool was used
+{[}Figure 6 \& S7{]}. This discordance between stool and colon tissue
+samples also applies to the mouth-associated genera with one noticeable
+skew being that \emph{Fusobacterium} and \emph{Fusobacteriaceae} occur
+more often in the top 10 of matched versus unmatched colon tissue Random
+Forest models {[}Figure S7B-C \& S7E-F{]}. This suggests that either the
+colon tissue microbiota is study and person dependent or that kit and/or
+other types of contamination associated with low biomass samples may be
+skewing the results.
 
 \textbf{\emph{CRC Studies are Underpowered for Detecting Small Effect
-Sizes:}} Based on our reported results we assessed whether the studies
-analyzed were realistically powered to identify small, medium, and large
-scale differences between cases and controls. When assessing the power
-of each study at different effect sizes the majority of studies, for
-both adenoma and carcinoma, acheived 80\% power to detect a 30\% or
-greater difference between groups {[}Figure 7A \& B{]}. No study that
-was analyzed had the standard 80\% power to detect an effect size
-difference that was equal to or below 10\% {[}Figure 7A \& B{]}. In
-order to achieve a power of 80\%, for small effect sizes, studies used
-in our meta-analysis would need to recruit over 1000 individuals for
-each arm {[}Figure 7C{]}.
+Sizes:}} Next, we assessed how much confidence should be placed in the
+reported outcomes from each individual study by calculating the ability
+to detect a difference (power) and sample size needed for small, medium,
+and large effect size differences between cases and controls. When
+assessing the power of each study at different effect sizes the majority
+of studies acheived 80\% power to detect a 30\% or greater difference
+between groups {[}Figure 7A \& B{]}. No study that we analyzed had the
+standard 80\% power to detect an effect size difference equal to or
+below 10\% {[}Figure 7A \& B{]}. In order to achieve a power of 80\%,
+for small effect sizes, studies used in our meta-analysis would need to
+recruit over 1000 individuals for both the case and control arms
+{[}Figure 7C{]}
 
@@
 Our study identifies clear differences in diversity, both at the
-community level and for individual genera, that are present in patients
-with and without CRC {[}Figure 1-3{]}. Although there was a step-wise
-decrease in diversity from control to adenoma to carcinoma, this did not
-translate into large effect sizes for the RR of lesion. Even though
-mouth-associated genera increased the RR of carcinoma, they did not
-consistently increase the RR of adenoma. These mouth-associated genera
-are clearly important to carcinoma classification but our observations
-suggest that accounting for other community members can increase the
-ability of models to make predictions.
-
-The data presented herein supports the importance of select genera for
-carcinoma but not necessarily adenoma formation. Our observations show
-that when only using the select genera to create models for carcinoma
-the AUC are similar to using all genera or OTUs {[}Figure 4 \& 5{]}.
+community level and for individual genera, present in patients with and
+without CRC {[}Figure 1-3{]}. Although there was a step-wise decrease in
+diversity as disease progressed from control to adenoma to carcinoma,
+this did not translate into large effect sizes for the RR of CRC. Even
+though mouth-associated genera increased the RR of having a carcinoma,
+they did not consistently increase the RR of having an adenoma.
+Additionally, our observations suggest that by combining
+mouth-associated and CRC protective microbes we can classify either
+adenoma or carcinoma stage of disease as well as models that use the
+full community.
+
+The data presented herein support the importance of select genera for
+carcinoma, but not necessarily adenoma, formation. The results that we
+have presented show that the both genera and OTU select and full models,
+for the carcinoma stage of CRC, had similar AUCs {[}Figure 4 \& 5{]}.
 This suggests that an interplay between a select number of potentially
-protective and exacerbating microbes, within the GI community, are
-crucial for carcinoma formation. Importantly, it suggests that there may
-be key members of the GI community that might be studied further to
-potentially reduce the risk of carcinoma. Conversely, using the present
-data, it is clear that new approaches may need to be used to identify
-members of the community that are associated with adenoma. Regardless of
-sample type and whether a full or select model was used our Random
-Forest models consistently performed poorly. Yet, the step-wise decrease
-in diversity suggests that the adenoma community is not normal but has
-changed subtly {[}Figure 1{]}. This change in diversity, at this early
-stage of disease, could be focal to the adenoma itself. One potential
-hypothesis from these observations is that at early stages of the
-disease, how the host interacts with these subtle changes is what
-ultimately leads to a thoroughly dysfunctional community that is
-supportive of CRC genesis.
+protective and exacerbating microbes within the GI community is crucial
+for carcinoma formation. Importantly, it suggests that there may be key
+members of the GI community that might be studied further to potentially
+reduce the risk of carcinoma. Conversely, using the present data, it is
+clear that new approaches may be needed to identify members of the
+community associated with adenoma stage of disease. Regardless of sample
+type and whether a full or select model was used, our Random Forest
+models consistently performed poorly. Yet, the step-wise decrease in
+diversity suggests that the adenoma-associated community is not normal
+but has changed subtly {[}Figure 1{]}. This change in diversity, at this
+early stage of disease, could be focal to the adenoma itself. One
+possible explanation is that how the host interacts with these subtle
+changes at early stages of the disease is what leads to a thoroughly
+dysfunctional community that is supportive of CRC genesis.
 
 Within stool, common GI microbes were most consistently present in the
-top 10 genera or OTUs across study {[}Figure 6{]}. Changes in
+top 10 genera or OTUs across studies {[}Figure 6{]}. Changes in
 \emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
-\emph{Roseburia} were consistently found to be disciminative across the
-different studies for both adenoma and carcinoma {[}Figure 6{]}. This
-data would suggest that whether the non-resident bacterium is
-\emph{Fusobacteria} or \emph{Peptostreptococcus} is not as important as
-how these bacteria interact with the changing resident community. Based
-on these observations, it is possible to hypothesize that small changes
-in community structure lead to new niches in which any one of the
+\emph{Roseburia} were consistently found to be in the top 10 most
+important variables across the different studies for both adenoma and
+carcinoma {[}Figure 6{]}. These data suggest that whether the
+non-resident bacterium is \emph{Fusobacteria} or
+\emph{Peptostreptococcus} is not as important as how these bacteria
+interact with the changing resident community. Based on these
+observations, it is possible to hypothesize that small changes in
+community structure lead to new niches in which any one of the
 mouth-associated genera can gain a foothold, exacerbating the initial
 changes in community and facilitating the transition from adenoma to
-carcinoma.
-
-The tissue studies did not provide a clearer understanding of how the
-microbiota may be associated with lesion. Generally, the full OTU-based
-models of unmatched and matched {[}Figure S7E \& F{]} tissue samples
-were concordant with stool showing that resident microbes were the most
-prevalent in the top 10 across study. Unlike in stool,
-\emph{Fusobacterium} was the only mouth-associated bacteria consistently
-present in the top 10 of the full CRC models {[}Figure S7B-C \& E-F{]}.
-The majority of the tissue results seem to be study specific with many
-top 10 taxa being present only in a single study. Additionally, the
-presence of genera associated with contamination, within the top 10 most
-important variables for the genera and OTU models is worrying. The low
-bacterial biomass of tissue samples coupled with potential contamination
-could ultimately explain why the these results seem to be more sporadic
-than the stool results.
-
-One important caveat to this study is that even though specific genera
-associated with species such as \emph{Bacteroides fragilis} and
+carcinoma stage of disease.
+
+The colon tissue-based studies did not provide a clearer understanding
+of how the microbiota may be associated with CRC Generally, the full
+OTU-based models of unmatched and matched colon tissue samples were
+concordant with stool samples showing that GI resident microbes were the
+most prevalent in the top 10 most important variables across study
+{[}Figure S7E \& F{]}. Unlike in stool, \emph{Fusobacterium} was the
+only mouth-associated bacteria consistently present in the top 10 most
+important variables of the full carcinoma stage models {[}Figure S7B-C
+\& E-F{]}. The majority of the colon tissue-based results seem to be
+study specific with many of top 10 taxa being present only in a single
+study. Additionally, the presence of genera associated with
+contamination, within the top 10 most important variables for the genera
+and OTU models is worrying. The low bacterial biomass of tissue samples
+coupled with potential contamination could explain why the these results
+seem to be more sporadic than the stool results.
+
+One important caveat to this study is that even though genera associated
+with certain species such as \emph{Bacteroides fragilis} and
 \emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
-identified, it does not necessarily mean these specific species are not
-important in human CRC {[}20,22{]}. Since we are limited in our
-aggregation of the data, across study, to the genus level, it is not
-possible to clearly delineate what specific species are contributing to
-overall disease progression. Our observations are not inconsistent with
-the previous literature on either \emph{Bacteroides fragilis} or
-\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus}. As an
-example, for stool, the full RF models consistently identified the genus
-\emph{Bacteroides} as well as OTUs that classified as \emph{Bacteroides}
-to be important model components across studies. Suggesting that even
-though \emph{Bacteroides} may not increase the RR of lesion and may not
-vary in relative abundance, like \emph{Fusobacterium}, it is still
-important in CRC. Additionaly, \emph{Streptococcus gallolyticus} subsp.
-\emph{gallolyticus} is a mouth-associated micobe and our study clearly
-shows that regardless of whether the sample is tissue or stool,
-mouth-associated genera are commonly associated with an increased RR of
-both adenoma and carcinoma.
-
-The associations between the microbiota and adenoma are inconclusive, in
-part, because many studies may not be powered effectively to observe
-small effect sizes. None of the studies analyzed were properly powered
-to detect a 10\% or lower change between cases and controls. A small
+identified, it does not necessarily mean that these specific species are
+not important in human CRC {[}20,22{]}. Since we are limited in our
+aggregation of the data to the genus level, it is not possible to
+clearly delineate which species are contributing to overall disease
+progression. Our observations are not inconsistent with the previous
+literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
+gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
+full community models consistently identified the genus
+\emph{Bacteroides}, as well as OTUs that classified as
+\emph{Bacteroides}, to be important model components across studies.
+This suggests that even though \emph{Bacteroides} may not increase the
+RR of CRC and may not vary in relative abundance, like
+\emph{Fusobacterium}, it is still important in CRC. Additionally,
+\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
+mouth-associated micobe, and the results from this study suggest that
+regardless sample type, mouth-associated genera are commonly associated
+with an increased RR for both adenoma and carcinoma stage of disease.
+
+The associations between the microbiota and adenoma stage of disease are
+inconclusive, in part, because many studies may not be powered
+effectively to observe small effect sizes. None of the studies analyzed
+were properly powered to detect a 10\% or lower change between cases and
+controls. The results within our meta-analysis suggest that a small
 effect size may well be the scope in which differences consistently
-occur in adenoma based on the results observed within our meta-analysis.
-Future studies investigating adenoma and the microbiota need to take
-power into consideration if we are to reproducibly study whether the
-microbiota contributes to polyp formation. In contrast to adenoma, our
+occur between controls and adenoma stage of disease. Future studies
+investigating adenoma stage and the microbiota need to take power into
+consideration to reproducibly study whether the microbiota contributes
+to polyp formation. In contrast to adenoma stage of disease, our
 observations suggest that most studies analyzed have sufficient power to
-detect many changes in carcinoma because of large effect size
-differences between cases and controls {[}Figure 7{]}.
+detect many changes in the carcinoma-associated microbiota because of
+large effect size differences between cases and controls {[}Figure 7{]}.
 
@@
 
-By aggregating together a large collection of studies from both feces
-and tissue, we are able to provide evidence in support of the importance
-of the bacterial community in CRC. Further, the data presented here
-suggests that mouth-associated microbes can gain a foothold within the
-colon and are are commonly associated with the greatest RR of carcinoma.
-Conversely, no conclusive signal with these mouth-associated micobes
-could be detected for adenoma. Our observations also highlight the
-importance of power and sample number considerations when undertaking
-investigations into the microbiota and adenoma due to the subtle changes
-in the community. Overall, the microbiota associations with carcinoma
-are much stronger than the associations with adenoma.
+By aggregating together a large collection of studies analyzing both
+fecal and colon tissue samples, we are able to provide evidence
+supporting the importance of the bacterial community in CRC. Further,
+the data presented here suggests that mouth-associated microbes can gain
+a foothold within the colon and are are commonly associated with the
+greatest RR of having a carcinoma. No conclusive signal with these
+mouth-associated micobes could be detected for adenoma stage of disease.
+Our observations also highlight the importance of power and sample
+number considerations when investigating the microbiota and adenoma
+stage of disease due to the subtle changes in the community. Overall,
+associations of microbiota with the carcinoma stage of CRC are much
+stronger than those with the adenoma stage.
 

### 2018-01-26 submission/manuscript.Rmd
add ada suggestions methods and discussion

@@
 The studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016] and additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature.
-Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis were included and only data sets that had sequences available for analysis were included.
-Some studies were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share and were excluded.
+Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis and had data sets with sequences available for analysis were included.
+Some studies were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share.
 After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014].
 The Zackular [@zackular_human_2014] study was not included because the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016].
-After sequence processing all the case samples for the Kostic study had 100 or less sequences remaining and was not used which left a total of 14 studies that analysis could be completed on.
+After sequence processing, all the case samples for the Kostic study had 100 or less sequences remaining and was excluded, leaving a total of 14 studies that analysis could be completed on.
 
@@
 In total, there were seven studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal and tissue samples (Chen and Flemer).
-The total number of individuals that were analyzed after sequence processing for feces was `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` [Table 1].
+The total number of individuals analyzed after sequence processing for feces was `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` [Table 1].
 The total number of matched and unmatched tissue samples that were analyzed after sequence processing was `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` [Table 2].
@@
 ***Sequence Processing:***
-For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata was obtained by searching the respective accession number of the study at the following website: http://www.ncbi.nlm.nih.gov/Traces/study/.
+For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata were obtained by searching the respective accession number of the study at the following website: http://www.ncbi.nlm.nih.gov/Traces/study/.
 Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap (n = 1, [@ahn_human_2013]) and directly from the authors (n = 4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]).
-Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009] and quality filtering utilized the default methods, used in mothur, for either 454 or Illumina based sequencing.
+Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009] and quality filtering utilized the default methods for both 454 and Illumina based sequencing.
 If it was not possible to use the defaults, the stated quality cut-offs, from the study itself, were used instead.
-Chimeras were identified and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].
+Sequences that were made up of an artificial combination of two or more different sequences and commonly known as chimeras were identified and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].
 
@@
 ***Study Analysis Overview:***
-$\alpha$-diversity was first assessed for differences between controls, adenoma, and carcinoma using both linear mixed-effect models and RR.
-$\beta$-diversity was then assessed for each individual study for differences between control-adenoma and control-carcinoma.
-Next, all common genera were assessed for differences in RR for adenoma and carcinoma and ranked based on P-value.
-We then built Random Forest models based on all or a selected model, based on the top 5 increased and top 5 decreased RR based on P-value, and these models were trained on one study then tested on the remaining studies, for every study.
+OTU richness, evenness, and Shannon diversity was first assessed for differences between controls, adenoma stage, and carcinoma stage using both linear mixed-effect models and RR.
+The Bray-Curtis index was used to assess, for each individual study, differences between control-adenoma and control-carcinoma.
+Next, all common genera were assessed for differences in RR for having an adenoma or carcinoma and ranked based on P-value.
+We then built Random Forest models based on full or  selected community, based on the top 5 increased and top 5 decreased RR based on P-value, and these models were trained on one study then tested on the remaining studies.
+This process was repeated for every study in the meta-analysis.
 A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
-For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a variable in the top 5 increased and top 5 decreased RR based on P-value.
+For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased RR based on P-value.
 Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power.
-For comparisons in which normal versus adenoma were made the carcinoma samples were excluded from each respective study.
-Similarly, for comparisons in which normal versus carcinoma were made the adenoma samples were excluded from each respective study.
-The data was split between feces and tissue samples.
-Within the tissue groups the data was further divided between samples from the same individual (matched) and those from different individuals (unmatched).
-Where applicable for each study, predictions for adenoma and carcinoma were then tested for feces, matched tissue, and unmatched tissue.
+For comparisons in which only control versus adenoma stage were made, the carcinoma samples were excluded from each respective study.
+Similarly, for comparisons in which control versus carcinoma stage were made the adenoma samples were excluded from each respective study.
+The data were split between feces and tissue samples.
+Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched).
+Where applicable for each study, predictions for adenoma and carcinoma stage of disease were then tested for feces, matched tissue, and unmatched tissue.
 
@@
 For the genera analysis of the RR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings, were collected.
-The OTU based Random Forest Models using selected OTUS utilized a similar approach except that the OTUs were not aggregated together by taxonomic identity but kept as separate OTUs.
+The OTU based Random Forest Models using selected OTUs utilized a similar approach except that the OTUs were not aggregated together by taxonomic identity but kept as separate OTUs.
 OTU Random Forest models using the full community included all OTUs while those for the selected model included only those OTUs that had a taxonomic classification to a variable in the top 5 increased of top 5 decreased RR based on P-value.
@@
 ***Matched versus Unmatched Tissue Samples:***
-In general, tissue samples that had control and lesion samples, that did not belong to the same individual, were classified as unmatched while samples, that belonged to the same individual, were classified as matched.
+In general, tissue samples with control and CRC samples from different indivdiuals were classified as unmatched while samples that belonged to the same individual were classified as matched.
 Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy.
-For some studies samples became unmatched due to one of the corresponding matched samples not making it through sequence processing.
+For some studies samples became unmatched when a corresponding matched sample did not make it through sequence processing.
 All samples, from both tissue sample types, were analyzed together for the linear mixed-effect models with samples from the same individual corrected for.
@@
 ***Assessing Important Random Forest Model Variables:***
-Using Mean Decrease in Accuracy (MDA) the top 10 most important variables were obtained in two different ways depending on whether the model used genera or OTU data.
+Using Mean Decrease in Accuracy (MDA) the top 10 most important variables to the Random Forest model were obtained in two different ways depending on whether the model used genera or OTU data.
 For the genus based models, the number of times that a genus showed up in the top 10 of the training set across each study was counted while, for the OTU based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study.
@@
 ***Reproducible Methods:***
-The code and analysis can be found here at  https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
-Unless otherwise mentioned, the accession number for the raw sequences for the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.
+The code and analysis can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
+Unless otherwise mentioned, the accession number of raw sequences from the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.
 

### 2018-01-26 submission/manuscript.tex
add ada suggestions methods and discussion

@@
 feces as their sample source for 454 or Illumina 16S rRNA gene
-sequencing analysis were included and only data sets that had sequences
-available for analysis were included. Some studies were excluded because
-they did not have publicly available sequences or did not have metadata
-in which the authors were able to share and were excluded. After these
-filtering steps, the following studies remained: Ahn, \emph{et al.}
-{[}12{]}, Baxter, \emph{et al.} {[}13{]}, Brim, \emph{et al.} {[}30{]},
-Burns, \emph{et al.} {[}16{]}, Chen, \emph{et al.} {[}14{]}, Dejea,
-\emph{et al.} {[}24{]}, Flemer, \emph{et al.} {[}18{]}, Geng, \emph{et
-al.} {[}23{]}, Hale, \emph{et al.} {[}27{]}, Kostic, \emph{et al.}
-{[}33{]}, Lu, \emph{et al.} {[}26{]}, Sanapareddy, \emph{et al.}
-{[}25{]}, Wang, \emph{et al.} {[}15{]}, Weir, \emph{et al.} {[}21{]},
-and Zeller, \emph{et al.} {[}17{]}. The Zackular {[}34{]} study was not
-included because the 90 individuals analyzed within the study are
-contained within the larger Baxter study {[}13{]}. After sequence
-processing all the case samples for the Kostic study had 100 or less
-sequences remaining and was not used which left a total of 14 studies
-that analysis could be completed on.
+sequencing analysis and had data sets with sequences available for
+analysis were included. Some studies were excluded because they did not
+have publicly available sequences or did not have metadata in which the
+authors were able to share. After these filtering steps, the following
+studies remained: Ahn, \emph{et al.} {[}12{]}, Baxter, \emph{et al.}
+{[}13{]}, Brim, \emph{et al.} {[}30{]}, Burns, \emph{et al.} {[}16{]},
+Chen, \emph{et al.} {[}14{]}, Dejea, \emph{et al.} {[}24{]}, Flemer,
+\emph{et al.} {[}18{]}, Geng, \emph{et al.} {[}23{]}, Hale, \emph{et
+al.} {[}27{]}, Kostic, \emph{et al.} {[}33{]}, Lu, \emph{et al.}
+{[}26{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang, \emph{et al.}
+{[}15{]}, Weir, \emph{et al.} {[}21{]}, and Zeller, \emph{et al.}
+{[}17{]}. The Zackular {[}34{]} study was not included because the 90
+individuals analyzed within the study are contained within the larger
+Baxter study {[}13{]}. After sequence processing, all the case samples
+for the Kostic study had 100 or less sequences remaining and was
+excluded, leaving a total of 14 studies that analysis could be completed
+on.
 
@@
 Sanapareddy), and two studies with both fecal and tissue samples (Chen
-and Flemer). The total number of individuals that were analyzed after
-sequence processing for feces was 1737 {[}Table 1{]}. The total number
-of matched and unmatched tissue samples that were analyzed after
-sequence processing was 492 {[}Table 2{]}.
+and Flemer). The total number of individuals analyzed after sequence
+processing for feces was 1737 {[}Table 1{]}. The total number of matched
+and unmatched tissue samples that were analyzed after sequence
+processing was 492 {[}Table 2{]}.
 
@@
 (\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
-and metadata was obtained by searching the respective accession number
+and metadata were obtained by searching the respective accession number
 of the study at the following website:
@@
 software program {[}35{]} and quality filtering utilized the default
-methods, used in mothur, for either 454 or Illumina based sequencing. If
-it was not possible to use the defaults, the stated quality cut-offs,
-from the study itself, were used instead. Chimeras were identified and
-removed using VSEARCH {[}36{]} before \emph{de novo} OTU clustering at
-97\% similarity was completed using the OptiClust algorithm {[}37{]}.
+methods for both 454 and Illumina based sequencing. If it was not
+possible to use the defaults, the stated quality cut-offs, from the
+study itself, were used instead. Sequences that were made up of an
+artificial combination of two or more different sequences and commonly
+known as chimeras were identified and removed using VSEARCH {[}36{]}
+before \emph{de novo} OTU clustering at 97\% similarity was completed
+using the OptiClust algorithm {[}37{]}.
 
@@
 
-\textbf{\emph{Study Analysis Overview:}} \(\alpha\)-diversity was first
-assessed for differences between controls, adenoma, and carcinoma using
-both linear mixed-effect models and RR. \(\beta\)-diversity was then
-assessed for each individual study for differences between
-control-adenoma and control-carcinoma. Next, all common genera were
-assessed for differences in RR for adenoma and carcinoma and ranked
-based on P-value. We then built Random Forest models based on all or a
-selected model, based on the top 5 increased and top 5 decreased RR
-based on P-value, and these models were trained on one study then tested
-on the remaining studies, for every study. A similar approach was then
-applied at the OTU level with the exception that a 10-fold CV over 100
-different models, based on random 80/20 splitting of the data, was used
-to generate a range of expected AUCs. For these OTU-based models, the
-selected model included all OTUs that had a taxonomic classification to
-a variable in the top 5 increased and top 5 decreased RR based on
-P-value. Finally, the power of each study was assessed for an effect
-size ranging from 1\% to 30\% and an estimated sample size, for these
-effect sizes, was generated based on 80\% power. For comparisons in
-which normal versus adenoma were made the carcinoma samples were
-excluded from each respective study. Similarly, for comparisons in which
-normal versus carcinoma were made the adenoma samples were excluded from
-each respective study. The data was split between feces and tissue
-samples. Within the tissue groups the data was further divided between
-samples from the same individual (matched) and those from different
-individuals (unmatched). Where applicable for each study, predictions
-for adenoma and carcinoma were then tested for feces, matched tissue,
-and unmatched tissue.
+\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
+Shannon diversity was first assessed for differences between controls,
+adenoma stage, and carcinoma stage using both linear mixed-effect models
+and RR. The Bray-Curtis index was used to assess, for each individual
+study, differences between control-adenoma and control-carcinoma. Next,
+all common genera were assessed for differences in RR for having an
+adenoma or carcinoma and ranked based on P-value. We then built Random
+Forest models based on full or selected community, based on the top 5
+increased and top 5 decreased RR based on P-value, and these models were
+trained on one study then tested on the remaining studies. This process
+was repeated for every study in the meta-analysis. A similar approach
+was then applied at the OTU level with the exception that a 10-fold CV
+over 100 different models, based on random 80/20 splitting of the data,
+was used to generate a range of expected AUCs. For these OTU-based
+models, the selected model included all OTUs that had a taxonomic
+classification to a taxa in the top 5 increased and top 5 decreased RR
+based on P-value. Finally, the power of each study was assessed for an
+effect size ranging from 1\% to 30\% and an estimated sample size, for
+these effect sizes, was generated based on 80\% power. For comparisons
+in which only control versus adenoma stage were made, the carcinoma
+samples were excluded from each respective study. Similarly, for
+comparisons in which control versus carcinoma stage were made the
+adenoma samples were excluded from each respective study. The data were
+split between feces and tissue samples. Within the tissue groups the
+data were further divided between samples from the same individual
+(matched) and those from different individuals (unmatched). Where
+applicable for each study, predictions for adenoma and carcinoma stage
+of disease were then tested for feces, matched tissue, and unmatched
+tissue.
 
@@
 average counts, for 100 different subsamplings, were collected. The OTU
-based Random Forest Models using selected OTUS utilized a similar
+based Random Forest Models using selected OTUs utilized a similar
 approach except that the OTUs were not aggregated together by taxonomic
@@
 \textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
-tissue samples that had control and lesion samples, that did not belong
-to the same individual, were classified as unmatched while samples, that
-belonged to the same individual, were classified as matched. Studies
-with matched data included Burns, Dejea, Geng, and Lu while those with
-unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some
-studies samples became unmatched due to one of the corresponding matched
-samples not making it through sequence processing. All samples, from
-both tissue sample types, were analyzed together for the linear
-mixed-effect models with samples from the same individual corrected for.
-For all other analysis, not mentioned herein, matched and unmatched
-samples were analyzed separately using the statistical approaches
-mentioned in the Statistical Analysis section.
+tissue samples with control and CRC samples from different indivdiuals
+were classified as unmatched while samples that belonged to the same
+individual were classified as matched. Studies with matched data
+included Burns, Dejea, Geng, and Lu while those with unmatched data were
+from Burns, Flemer, Chen, and Sanapareddy. For some studies samples
+became unmatched when a corresponding matched sample did not make it
+through sequence processing. All samples, from both tissue sample types,
+were analyzed together for the linear mixed-effect models with samples
+from the same individual corrected for. For all other analysis, not
+mentioned herein, matched and unmatched samples were analyzed separately
+using the statistical approaches mentioned in the Statistical Analysis
+section.
 
 \textbf{\emph{Assessing Important Random Forest Model Variables:}} Using
-Mean Decrease in Accuracy (MDA) the top 10 most important variables were
-obtained in two different ways depending on whether the model used
-genera or OTU data. For the genus based models, the number of times that
-a genus showed up in the top 10 of the training set across each study
-was counted while, for the OTU based models, the medians for each OTU
-across 100 different 80/20 splits of the data was generated and the top
-10 OTUs then counted for each study. Common taxa, for the OTU based
-models, were identified by using the lowest classification within the
-RDP database for each of the specific OTUs obtained from the previous
-counts and the number of times this classification occurred in the top
-10, in each study, was recorded. The two studies that had adenoma tissue
-were equally divided between matched and unmatched groups and were
-grouped together for the counting of the top 10 genera and OTUs.
+Mean Decrease in Accuracy (MDA) the top 10 most important variables to
+the Random Forest model were obtained in two different ways depending on
+whether the model used genera or OTU data. For the genus based models,
+the number of times that a genus showed up in the top 10 of the training
+set across each study was counted while, for the OTU based models, the
+medians for each OTU across 100 different 80/20 splits of the data was
+generated and the top 10 OTUs then counted for each study. Common taxa,
+for the OTU based models, were identified by using the lowest
+classification within the RDP database for each of the specific OTUs
+obtained from the previous counts and the number of times this
+classification occurred in the top 10, in each study, was recorded. The
+two studies that had adenoma tissue were equally divided between matched
+and unmatched groups and were grouped together for the counting of the
+top 10 genera and OTUs.
 
 \textbf{\emph{Reproducible Methods:}} The code and analysis can be found
-here at
+at
 \url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
-Unless otherwise mentioned, the accession number for the raw sequences
-for the studies used in this analysis can be found directly in the
+Unless otherwise mentioned, the accession number of raw sequences from
+the studies used in this analysis can be found directly in the
 respective batch file in the GitHub repository or in the original

### 2018-01-27 submission/manuscript.Rmd
Look over edits after incorporating Ada suggestions

@@
 
-**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness after correcting for study effect and variable region sequenced with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
+**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness, after correcting for study effect and variable region sequenced, with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
 This reduction in evenness translated into small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased relative risk for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
@@
 That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of CRC that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].
-There are also genera like *Roseburia* that some studies find to be increased in CRC but other studies report either decreased or no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
+Additionally, genera like *Roseburia* have been found in some studies to be increased in CRC but in others to either be decreased or have no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
 
-Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera.
+Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera of disease.
 Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of CRC versus controls.
@@
 Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
-Sampling the microbiota directly associated with colon tissue directly may provide clearer answers but is not without limitations.
+Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without limitations.
 The community present for sampling following the colonoscopy bowel prep may reflect the better adhered microbiota versus the resident community.
@@
 
-***Lower Bacterial Diversity is Associated with Increased RR of Carcinomas:***
+***Lower Bacterial Diversity is Associated with Increased RR of CRC:***
 To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized.
@@
 We next tested whether the detectable differences in community significantly increased in RR of having an adenoma or carcinoma.
-For fecal samples, a decrease versus the overal median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
+For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
 Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].
@@
 ***Mouth-Associated Genera are Associated with an Increased RR of CRC:***
-Next we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity.
+Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity.
 To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR.
-The top 5 most significantly increased RR were the same between adenoma and carcinoma for both stool and colon tissue [Figure 3].
 Mouth-associated genera were commonly found in the top 5 genera associated with an increased RR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples.
@@
 
-In stool, the most common genera in the top 10 most imporant variables, in models for the first approach, that used full genera-based models were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
+In stool, the most common genera in the top 10 most imporant variables, in the full community models using the first approach, were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
 Regardless of sample type, mouth-associated genera were present in models for the carcinoma stage of CRC [Figure 6A & B].
@@
 The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, formation.
-The results that we have presented show that the both genera and OTU select and full models, for the carcinoma stage of CRC, had similar AUCs [Figure 4 & 5].
+The results that we have presented show that both the genera and OTU select and full models, for the carcinoma stage of CRC, had similar AUCs [Figure 4 & 5].
 This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community is crucial for carcinoma formation.
@@
 Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both adenoma and carcinoma [Figure 6].
-These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* is not as important as how these bacteria interact with the changing resident community.
+These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community.
 Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.
 
-The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with CRC Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F].
+The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with CRC.
+Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F].
 Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S7B-C & E-F].
-The majority of the colon tissue-based results seem to be study specific with many of top 10 taxa being present only in a single study.
+The majority of the colon tissue-based results seem to be study specific with many of the top 10 taxa being present only in a single study.
 Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying.
-The low bacterial biomass of tissue samples coupled with potential contamination could explain why the these results seem to be more sporadic than the stool results.
+The low bacterial biomass of tissue samples coupled with potential contamination could explain why these results seem to be more sporadic than the stool results.
 
@@
 Additionally, *Streptococcus gallolyticus* subsp.
-*gallolyticus* is a mouth-associated micobe, and the results from this study suggest that regardless sample type, mouth-associated genera are commonly associated with an increased RR for both adenoma and carcinoma stage of disease.
+*gallolyticus* is a mouth-associated micobe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased RR for both adenoma and carcinoma stage of disease.
 
@@
 All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
-For the $\alpha$-diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
+For OTU richness, evenness, and Shannon diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
 Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hypervariable region used.

### 2018-01-27 submission/manuscript.tex
Look over edits after incorporating Ada suggestions

@@
 \textbf{Results.} Fecal samples had a significant decrease for both
-Shannon diversity and evenness after correcting for study effect and
-variable region sequenced with more severe disease (P-value \textless{}
+Shannon diversity and evenness, after correcting for study effect and
+variable region sequenced, with more severe disease (P-value \textless{}
 0.05). This reduction in evenness translated into small increases in
@@
 have been found to be enriched in controls and not disease
-{[}13,16,17{]}. There are also genera like \emph{Roseburia} that some
-studies find to be increased in CRC but other studies report either
-decreased or no difference {[}15,18,23,24{]}.
+{[}13,16,17{]}. Additionally, genera like \emph{Roseburia} have been
+found in some studies to be increased in CRC but in others to either be
+decreased or have no difference {[}15,18,23,24{]}.
 
 Most of these studies have focused on carcinoma but the adenoma
-observations are not any clearer at identifying candidate genera. Groups
-focusing on broad scale community metrics have found that metrics such
-as richness are decreased in the adenoma stage of CRC versus controls.
-Other studies have identified \emph{Lactococcus}, \emph{Pseudomonas},
-\emph{Acidovorax}, \emph{Cloacibacterium}, \emph{Helicobacter},
-\emph{Lactobacillus}, \emph{Bilophila}, \emph{Desulfovibrio}, and
-\emph{Mogibacterium} to be increased in adenoma {[}25--27{]}. Based on
-the studies mentioned, there seems to be very little overlap between the
-genera identified to be associated with adenoma and carcinoma, with
-\emph{Lactobacillus} being one of the few commonalities.
+observations are not any clearer at identifying candidate genera of
+disease. Groups focusing on broad scale community metrics have found
+that metrics such as richness are decreased in the adenoma stage of CRC
+versus controls. Other studies have identified \emph{Lactococcus},
+\emph{Pseudomonas}, \emph{Acidovorax}, \emph{Cloacibacterium},
+\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Bilophila},
+\emph{Desulfovibrio}, and \emph{Mogibacterium} to be increased in
+adenoma {[}25--27{]}. Based on the studies mentioned, there seems to be
+very little overlap between the genera identified to be associated with
+adenoma and carcinoma, with \emph{Lactobacillus} being one of the few
+commonalities.
 
@@
 and disease but are limited by their sample size {[}18,28{]}. Sampling
-the microbiota directly associated with colon tissue directly may
-provide clearer answers but is not without limitations. The community
-present for sampling following the colonoscopy bowel prep may reflect
-the better adhered microbiota versus the resident community.
-Additionally, these samples contain more host DNA, potentially limiting
-the types of analysis that can be done. It is well known that low
-biomass samples can be very difficult to work with and results can be
-study dependent due to the randomness of contamination {[}29{]}.
+the microbiota directly associated with colon tissue may provide clearer
+answers but is not without limitations. The community present for
+sampling following the colonoscopy bowel prep may reflect the better
+adhered microbiota versus the resident community. Additionally, these
+samples contain more host DNA, potentially limiting the types of
+analysis that can be done. It is well known that low biomass samples can
+be very difficult to work with and results can be study dependent due to
+the randomness of contamination {[}29{]}.
 
@@
 \textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
-of Carcinomas:}} To assess differences in broad scale community metrics
-as disease severity worsens Operational Taxonomic Unit (OTU) richness,
+of CRC:}} To assess differences in broad scale community metrics as
+disease severity worsens Operational Taxonomic Unit (OTU) richness,
 evenness, and Shannon diversity measurements were power transformed and
@@
 increased in RR of having an adenoma or carcinoma. For fecal samples, a
-decrease versus the overal median in evenness resulted in a
+decrease versus the overall median in evenness resulted in a
 significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
@@
 \textbf{\emph{Mouth-Associated Genera are Associated with an Increased
-RR of CRC:}} Next we asked if being higher than the median relative
+RR of CRC:}} Next, we asked if being higher than the median relative
 abundance, for any specific genera, resulted in an altered RR for
@@
 abundance higher than the median results in an increase or decrease in
-RR. The top 5 most significantly increased RR were the same between
-adenoma and carcinoma for both stool and colon tissue {[}Figure 3{]}.
-Mouth-associated genera were commonly found in the top 5 genera
+RR. Mouth-associated genera were commonly found in the top 5 genera
 associated with an increased RR of having an adenoma
@@
 In stool, the most common genera in the top 10 most imporant variables,
-in models for the first approach, that used full genera-based models
-were \emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia}
-{[}Figure 6A \& B{]}. Regardless of sample type, mouth-associated genera
-were present in models for the carcinoma stage of CRC {[}Figure 6A \&
-B{]}. Yet, none were present in the majority of studies and
+in the full community models using the first approach, were
+\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} {[}Figure
+6A \& B{]}. Regardless of sample type, mouth-associated genera were
+present in models for the carcinoma stage of CRC {[}Figure 6A \& B{]}.
+Yet, none were present in the majority of studies and
 \emph{Fusobacterium} was the only genus present in the adenoma stage of
@@
 carcinoma, but not necessarily adenoma, formation. The results that we
-have presented show that the both genera and OTU select and full models,
+have presented show that both the genera and OTU select and full models,
 for the carcinoma stage of CRC, had similar AUCs {[}Figure 4 \& 5{]}.
@@
 non-resident bacterium is \emph{Fusobacteria} or
-\emph{Peptostreptococcus} is not as important as how these bacteria
+\emph{Peptostreptococcus} may not be as important as how these bacteria
 interact with the changing resident community. Based on these
@@
 The colon tissue-based studies did not provide a clearer understanding
-of how the microbiota may be associated with CRC Generally, the full
+of how the microbiota may be associated with CRC. Generally, the full
 OTU-based models of unmatched and matched colon tissue samples were
@@
 \& E-F{]}. The majority of the colon tissue-based results seem to be
-study specific with many of top 10 taxa being present only in a single
-study. Additionally, the presence of genera associated with
+study specific with many of the top 10 taxa being present only in a
+single study. Additionally, the presence of genera associated with
 contamination, within the top 10 most important variables for the genera
 and OTU models is worrying. The low bacterial biomass of tissue samples
-coupled with potential contamination could explain why the these results
+coupled with potential contamination could explain why these results
 seem to be more sporadic than the stool results.
@@
 mouth-associated micobe, and the results from this study suggest that
-regardless sample type, mouth-associated genera are commonly associated
-with an increased RR for both adenoma and carcinoma stage of disease.
+regardless of sample type, mouth-associated genera are commonly
+associated with an increased RR for both adenoma and carcinoma stage of
+disease.
 
@@
 sequence processing utilized the R (v3.4.3) software package {[}38{]}.
-For the \(\alpha\)-diversity analysis, values were power transformed
-using the rcompanion (v1.11.1) package {[}39{]} and then Z-score
-normalized using the car (v2.1.6) package {[}40{]}. Testing for
-\(\alpha\)-diversity differences utilized linear mixed-effect models
+For OTU richness, evenness, and Shannon diversity analysis, values were
+power transformed using the rcompanion (v1.11.1) package {[}39{]} and
+then Z-score normalized using the car (v2.1.6) package {[}40{]}. Testing
+for \(\alpha\)-diversity differences utilized linear mixed-effect models
 created using the lme4 (v1.1.15) package {[}41{]} to correct for study,

### 2018-01-30 submission/manuscript.Rmd
update figure legends

@@
 
-**Figure 1: $\alpha$-Diversity Differences between Control, Adenoma, and Carcinoma Across Sampling Site.** A) $\alpha$-diversity metric differences by group in stool samples.
-B) $\alpha$-diversity metric differences by group in unmatched tissue samples.
-C) $\alpha$-diversity metric differences by group in matched tissue samples.
+**Figure 1: Community Differences between Control, Adenoma, and Carcinoma Across Sampling Site.** A) Stool sample community differences by disease group.
+B) Unmatched tissue samples differences by disease group.
+C) Matched tissue sample differences by group disease group.
 The dashed line represents a Z-score of 0 or no difference from the median.
 
-**Figure 2: Relative Risk for Adenoma or Carcinoma based on $\alpha$-Diversity Metrics in Stool.** A) $\alpha$-metric relative risk for adenoma.
-B) $\alpha$-metric relative risk for carcinoma.
+**Figure 2: Relative Risk for Adenoma or Carcinoma based on Bacterial Community Metrics in Stool.** A) Community-based relative risk for adenoma.
+B) Community-based relative risk for carcinoma.
 Colors represent the different variable regions used within the respective study.
@@
 
-**Figure S1: Relative Risk for Adenoma or Carcinoma based on $\alpha$-Diversity Metrics in Tissue.** A) $\alpha$-metric relative risk for adenoma.
-B) $\alpha$-metric relative risk for carcinoma.
+**Figure S1: Relative Risk for Adenoma or Carcinoma based on Bacterial Community Metrics in Tissue.** A) Community-based relative risk for adenoma.
+B) Community-based relative risk for carcinoma.
 Colors represent the different variable regions used within the respective study.

### 2018-01-30 submission/manuscript.tex
update figure legends

@@
 T-J, et al. Intestinal inflammation targets cancer-inducing activity of
-the microbiota. Science (New York, N.Y.). 2012;338:120--3.
+the microbiota. Science (New York, NY). 2012;338:120--3.
 
@@
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
-Research (Philadelphia, Pa.). 2014;7:1112--21.
+Research (Philadelphia, Pa). 2014;7:1112--21.
 
@@
 Community-Supported Software for Describing and Comparing Microbial
-Communities. Appl.Environ.Microbiol. {[}Internet{]}. 2009 {[}cited 12AD
-Jan 1{]};75:7537--41. Available from:
+Communities. ApplEnvironMicrobiol {[}Internet{]}. 2009 {[}cited 12AD Jan
+1{]};75:7537--41. Available from:
 \url{http://aem.asm.org/cgi/content/abstract/75/23/7537}
@@
 
-\textbf{Figure 1: \(\alpha\)-Diversity Differences between Control,
-Adenoma, and Carcinoma Across Sampling Site.} A) \(\alpha\)-diversity
-metric differences by group in stool samples. B) \(\alpha\)-diversity
-metric differences by group in unmatched tissue samples. C)
-\(\alpha\)-diversity metric differences by group in matched tissue
-samples. The dashed line represents a Z-score of 0 or no difference from
-the median.
+\textbf{Figure 1: Community Differences between Control, Adenoma, and
+Carcinoma Across Sampling Site.} A) Stool sample community differences
+by disease group. B) Unmatched tissue samples differences by disease
+group. C) Matched tissue sample differences by group disease group. The
+dashed line represents a Z-score of 0 or no difference from the median.
 
 \textbf{Figure 2: Relative Risk for Adenoma or Carcinoma based on
-\(\alpha\)-Diversity Metrics in Stool.} A) \(\alpha\)-metric relative
-risk for adenoma. B) \(\alpha\)-metric relative risk for carcinoma.
-Colors represent the different variable regions used within the
-respective study.
+Bacterial Community Metrics in Stool.} A) Community-based relative risk
+for adenoma. B) Community-based relative risk for carcinoma. Colors
+represent the different variable regions used within the respective
+study.
 
@@
 \textbf{Figure S1: Relative Risk for Adenoma or Carcinoma based on
-\(\alpha\)-Diversity Metrics in Tissue.} A) \(\alpha\)-metric relative
-risk for adenoma. B) \(\alpha\)-metric relative risk for carcinoma.
-Colors represent the different variable regions used within the
-respective study.
+Bacterial Community Metrics in Tissue.} A) Community-based relative risk
+for adenoma. B) Community-based relative risk for carcinoma. Colors
+represent the different variable regions used within the respective
+study.
 

### 2018-02-01 submission/manuscript.Rmd
run spell-check

@@
 Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased RR genera resulted in better model classification, based on the area under the curve (AUC).
-Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionaly accepted 80% power.
+Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power.
 Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the studies are underpowered for assessing small effect sizes.
@@
 We next tested whether the detectable differences in community significantly increased in RR of having an adenoma or carcinoma.
-For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
+For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
 Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].
@@
 
-In stool, the most common genera in the top 10 most imporant variables, in the full community models using the first approach, were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
+In stool, the most common genera in the top 10 most important variables, in the full community models using the first approach, were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
 Regardless of sample type, mouth-associated genera were present in models for the carcinoma stage of CRC [Figure 6A & B].
@@
 Next, we assessed how much confidence should be placed in the reported outcomes from each individual study by calculating the ability to detect a difference (power) and sample size needed for small, medium, and large effect size differences between cases and controls.
-When assessing the power of each study at different effect sizes the majority of studies acheived 80% power to detect a 30% or greater difference between groups [Figure 7A & B].
+When assessing the power of each study at different effect sizes the majority of studies achieved 80% power to detect a 30% or greater difference between groups [Figure 7A & B].
 No study that we analyzed had the standard 80% power to detect an effect size difference equal to or below 10% [Figure 7A & B].
@@
 Additionally, *Streptococcus gallolyticus* subsp.
-*gallolyticus* is a mouth-associated micobe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased RR for both adenoma and carcinoma stage of disease.
+*gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased RR for both adenoma and carcinoma stage of disease.
 
@@
 Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest RR of having a carcinoma.
-No conclusive signal with these mouth-associated micobes could be detected for adenoma stage of disease.
+No conclusive signal with these mouth-associated microbes could be detected for adenoma stage of disease.
 Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to the subtle changes in the community.
@@
 For OTU richness, evenness, and Shannon diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
-Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hypervariable region used.
+Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hyper-variable region used.
 Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study.
@@
 ***Matched versus Unmatched Tissue Samples:***
-In general, tissue samples with control and CRC samples from different indivdiuals were classified as unmatched while samples that belonged to the same individual were classified as matched.
+In general, tissue samples with control and CRC samples from different individuals were classified as unmatched while samples that belonged to the same individual were classified as matched.
 Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy.

### 2018-02-01 submission/manuscript.tex
run spell-check

@@
 and sample size the studies used were powered for and the sample size
-needed to get to the traditionaly accepted 80\% power. Our results from
+needed to get to the traditionally accepted 80\% power. Our results from
 these analyses suggests that the bacterial community changes as disease
@@
 P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
-0.032) while a decrease versus the overal median in Shannon diversity
+0.032) while a decrease versus the overall median in Shannon diversity
 only increased the RR for carcinoma (RR = 1.33 (1.09 - 1.62), P-value =
@@
 
-In stool, the most common genera in the top 10 most imporant variables,
+In stool, the most common genera in the top 10 most important variables,
 in the full community models using the first approach, were
@@
 assessing the power of each study at different effect sizes the majority
-of studies acheived 80\% power to detect a 30\% or greater difference
+of studies achieved 80\% power to detect a 30\% or greater difference
 between groups {[}Figure 7A \& B{]}. No study that we analyzed had the
@@
 \emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
-mouth-associated micobe, and the results from this study suggest that
+mouth-associated microbe, and the results from this study suggest that
 regardless of sample type, mouth-associated genera are commonly
@@
 greatest RR of having a carcinoma. No conclusive signal with these
-mouth-associated micobes could be detected for adenoma stage of disease.
-Our observations also highlight the importance of power and sample
-number considerations when investigating the microbiota and adenoma
-stage of disease due to the subtle changes in the community. Overall,
-associations of microbiota with the carcinoma stage of CRC are much
-stronger than those with the adenoma stage.
+mouth-associated microbes could be detected for adenoma stage of
+disease. Our observations also highlight the importance of power and
+sample number considerations when investigating the microbiota and
+adenoma stage of disease due to the subtle changes in the community.
+Overall, associations of microbiota with the carcinoma stage of CRC are
+much stronger than those with the adenoma stage.
 
@@
 created using the lme4 (v1.1.15) package {[}41{]} to correct for study,
-repeat sampling of individuals (tissue only), and 16S hypervariable
+repeat sampling of individuals (tissue only), and 16S hyper-variable
 region used. Relative risk was analyzed using both the epiR (v0.9.93)
@@
 \textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
-tissue samples with control and CRC samples from different indivdiuals
+tissue samples with control and CRC samples from different individuals
 were classified as unmatched while samples that belonged to the same

### 2018-02-02 code/make_rf_auc_full_versus_specific_OTU_graph.R
change to select community

@@ -96,7 +96,7 @@ adn_tissue_medians <- adn_tissue %>%
   summarise(auc_median = median(AUC)) %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "Select Genera Only")))
+                             labels = c("Full Community", "Select Community")))
 
 # Creates the adenoma tissue graph
 adn_tissue_graph <- adn_tissue %>% 
@@ -105,7 +105,7 @@ adn_tissue_graph <- adn_tissue %>%
                        labels = c("Unmatched Tissue", "Matched Tissue")), 
          model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "Select Genera Only")), 
+                             labels = c("Full Community", "Select Community")), 
          study = factor(study, 
                         levels = c("flemer", "lu"), 
                         labels = c("Flemer", "Lu\n(matched)"))) %>% 
@@ -133,13 +133,13 @@ adn_stool_medians <- adn_all_stool %>%
   summarise(auc_median = median(AUC)) %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "Select Genera Only")))
+                             labels = c("Full Community", "Select Community")))
 
 # Creates the adenoma stool graph
 adn_stool_graph <- adn_all_stool %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "Select Genera Only")), 
+                             labels = c("Full Community", "Select Community")), 
          study = factor(study, 
                         levels = c("baxter", "brim", "hale", "zeller"), 
                         labels = c("Baxter", "Brim", "Hale", "Zeller"))) %>% 
@@ -168,13 +168,13 @@ crc_tissue_medians <- crc_all_tissue %>%
   summarise(auc_median = median(AUC)) %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "Select Genera Only")))
+                             labels = c("Full Community", "Select Community")))
 
 # Creates the crc tissue graph
 crc_tissue_graph <- crc_all_tissue %>% 
   mutate(model_type = factor(model_type, 
                              levels = c("full", "select"), 
-                             labels = c("Full Community", "Select Genera Only")), 
+                             labels = c("Full Community", "Select Community")), 
[…]

### 2018-02-02 submission/manuscript.Rmd
update fig legends to select community

@@
 
-**Figure 4: Stool OTU Random Forest Model Across Studies.** A) Adenoma random forest model between the full community and select genera OTUs only.
-B) Carcinoma random forest model between the full community and select genera OTUs only.
+**Figure 4: Stool OTU Random Forest Model Across Studies.** A) Adenoma random forest model between the full and select community OTUs only.
+B) Carcinoma random forest model between the full and select community OTUs only.
 The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.
-The solid red line represents the median AUC of all the studies for either the full community or select genera OTUS only model.
+The solid red line represents the median AUC of all the studies for either the full or select community OTUS only model.
 
-**Figure 5: Tissue OTU Random Forest Model Across Studies.** A) Adenoma random forest model between the full community and select genera OTUs only.
-B) Carcinoma random forest model between the full community and select genera OTUs only.
+**Figure 5: Tissue OTU Random Forest Model Across Studies.** A) Adenoma random forest model between the full and select community OTUs only.
+B) Carcinoma random forest model between the full and select community OTUs only.
 The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.

### 2018-02-02 submission/manuscript.tex
update fig legends to select community

@@
 \textbf{Figure 4: Stool OTU Random Forest Model Across Studies.} A)
-Adenoma random forest model between the full community and select genera
-OTUs only. B) Carcinoma random forest model between the full community
-and select genera OTUs only. The dotted line represents an AUC of 0.5
-and the lines represent the range in which the AUC for the 100 different
-80/20 runs fell between. The solid red line represents the median AUC of
-all the studies for either the full community or select genera OTUS only
-model.
+Adenoma random forest model between the full and select community OTUs
+only. B) Carcinoma random forest model between the full and select
+community OTUs only. The dotted line represents an AUC of 0.5 and the
+lines represent the range in which the AUC for the 100 different 80/20
+runs fell between. The solid red line represents the median AUC of all
+the studies for either the full or select community OTUS only model.
 
 \textbf{Figure 5: Tissue OTU Random Forest Model Across Studies.} A)
-Adenoma random forest model between the full community and select genera
-OTUs only. B) Carcinoma random forest model between the full community
-and select genera OTUs only. The dotted line represents an AUC of 0.5
-and the lines represent the range in which the AUC for the 100 different
-80/20 runs fell between. The solid red line represents the median AUC of
-all the studies for either the full community or select genera OTUS only
+Adenoma random forest model between the full and select community OTUs
+only. B) Carcinoma random forest model between the full and select
+community OTUs only. The dotted line represents an AUC of 0.5 and the
+lines represent the range in which the AUC for the 100 different 80/20
+runs fell between. The solid red line represents the median AUC of all
+the studies for either the full community or select genera OTUS only
 model.

### 2018-02-03 submission/manuscript.Rmd
add one extra sentence to intro 1st paragraph
fix title

@@
 
-# Making Sense of the Noise: Leveraging Existing 16S rRNA Gene Surveys to Identify Key Community Members in Colorectal Cancer
+# Making Sense of the Noise: Leveraging Existing 16S rRNA Gene Surveys to Identify Key Community Members in Colorectal Tumors
 
@@
 This meta-analysis focused on identifying microbial signatures of CRC (biomarkers) but did so on a small total number of individuals and only investigated stool.
+This present meta-analysis addresses some of these major shortcomings.
 

### 2018-02-03 submission/manuscript.tex
add one extra sentence to intro 1st paragraph
fix title

@@
 Surveys to Identify Key Community Members in Colorectal
-Cancer}\label{making-sense-of-the-noise-leveraging-existing-16s-rrna-gene-surveys-to-identify-key-community-members-in-colorectal-cancer}
+Tumors}\label{making-sense-of-the-noise-leveraging-existing-16s-rrna-gene-surveys-to-identify-key-community-members-in-colorectal-tumors}
 
@@
 on identifying microbial signatures of CRC (biomarkers) but did so on a
-small total number of individuals and only investigated stool.
+small total number of individuals and only investigated stool. This
+present meta-analysis addresses some of these major shortcomings.
 

### 2018-02-03 submission/manuscript.Rmd
update to change CRC in specific places

@@
 Important drivers within this context have ranged from individual microbes to the whole community.
-Our study expands on a recent meta-analysis investigating microbial biomarkers for CRC by testing the hypothesis that the bacterial community has important associations to both early (adenoma) and late (carcinoma) stage disease.
+Our study expands on a recent meta-analysis investigating microbial biomarkers for tumors by testing the hypothesis that the bacterial community has important associations to both early (adenoma) and late (carcinoma) stage disease.
 To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 previously published 16S rRNA gene sequencing studies on CRC and the microbiota.
@@
 A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
-This meta-analysis focused on identifying microbial signatures of CRC (biomarkers) but did so on a small total number of individuals and only investigated stool.
+This meta-analysis focused on identifying microbial signatures of tumors (biomarkers) but did so on a small total number of individuals and only investigated stool.
 This present meta-analysis addresses some of these major shortcomings.
 
-Although there has been an intense focus on microbiota-based biomarker discovery for CRC, the number of candidate genera seem to be endless.
+Although there has been an intense focus on microbiota-based biomarker discovery for tumors, the number of candidate genera seem to be endless.
 Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
@@
 A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017].
-However, the sample size (n = 509) is equal to or less than some more recent individual studies investigating the microbiota and CRC, making it hard to know how extrapolatable these findings are.
-That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of CRC that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].
-Additionally, genera like *Roseburia* have been found in some studies to be increased in CRC but in others to either be decreased or have no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
+However, the sample size (n = 509) is equal to or less than some more recent individual studies investigating the microbiota and colorectal tumors, making it hard to know how extrapolatable these findings are.
+That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of disease that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].
+Additionally, genera like *Roseburia* have been found in some studies to be increased in tumors but in others to either be decreased or have no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
 
 Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera of disease.
-Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of CRC versus controls.
+Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of disease versus controls.
 Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
@@
 
-Targeting the identification of CRC microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications.
+Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications.
 Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities.
@@
 
-In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of CRC.
+In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of colorectal tumors.
 Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].
 
-Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of CRC.
-To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma stages of CRC.
-We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma stages of CRC.
-Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased RR genera resulted in better model classification, based on the area under the curve (AUC).
+Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of disease.
+To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma stages of disease We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma stages of disease Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased RR genera resulted in better model classification, based on the area under the curve (AUC).
 Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power.
@@
 
-***Lower Bacterial Diversity is Associated with Increased RR of CRC:***
+***Lower Bacterial Diversity is Associated with Increased RR of Tumors:***
 To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized.
@@
 
-***Mouth-Associated Genera are Associated with an Increased RR of CRC:***
+***Mouth-Associated Genera are Associated with an Increased RR of Tumor:***
 Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity.
@@
 Many of the genera that had RRs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
-For carcinoma stage of CRC, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both colon tissue and stool samples [Table S10 & S11].
+For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both colon tissue and stool samples [Table S10 & S11].
 Finally, these data suggest that the most significantly increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].
 
-***Select Genera Models can Recapitulate Whole Community Models:***
+***Select Community Models can Recapitulate Whole Community Models:***
 Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera.
@@
 Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with and without CRC [Figure 1-3].
-Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of CRC.
+Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of tumors.
 Even though mouth-associated genera increased the RR of having a carcinoma, they did not consistently increase the RR of having an adenoma.
@@
 
-The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with CRC.
+The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors.
 Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F].
@@
 No conclusive signal with these mouth-associated microbes could be detected for adenoma stage of disease.
+Overall, associations between the microbiota and carcinoma stage of disease were much stronger than with the adenoma stage.
 Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to the subtle changes in the community.
-Overall, associations of microbiota with the carcinoma stage of CRC are much stronger than those with the adenoma stage.
 
@@
 ***Matched versus Unmatched Tissue Samples:***
-In general, tissue samples with control and CRC samples from different individuals were classified as unmatched while samples that belonged to the same individual were classified as matched.
+In general, tissue samples with control and tumor samples from different individuals were classified as unmatched while samples that belonged to the same individual were classified as matched.
 Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy.

### 2018-02-03 submission/manuscript.tex
update to change CRC in specific places

@@
 individual microbes to the whole community. Our study expands on a
-recent meta-analysis investigating microbial biomarkers for CRC by
+recent meta-analysis investigating microbial biomarkers for tumors by
 testing the hypothesis that the bacterial community has important
@@
 identified using multiple data sets {[}11{]}. This meta-analysis focused
-on identifying microbial signatures of CRC (biomarkers) but did so on a
-small total number of individuals and only investigated stool. This
+on identifying microbial signatures of tumors (biomarkers) but did so on
+a small total number of individuals and only investigated stool. This
 present meta-analysis addresses some of these major shortcomings.
@@
 Although there has been an intense focus on microbiota-based biomarker
-discovery for CRC, the number of candidate genera seem to be endless.
+discovery for tumors, the number of candidate genera seem to be endless.
 Some studies point towards mouth-associated genera such as
@@
 However, the sample size (n = 509) is equal to or less than some more
-recent individual studies investigating the microbiota and CRC, making
-it hard to know how extrapolatable these findings are. That particular
-meta-analysis also added more potential microbial associations to both
-carcinoma (\emph{Pantoea agglomerans} \emph{Ruminococcus},
-\emph{Lactobacillus}) and adenoma (\emph{Prevotella},
-\emph{Methanosphaera}, \emph{Succinovibrio}, \emph{Haemophilus
-parainfluenzae}, \emph{Ruminococcus}, \emph{Lactobacillus}) stages of
-CRC that need to be investigated further, since a number of these genera
-have been found to be enriched in controls and not disease
-{[}13,16,17{]}. Additionally, genera like \emph{Roseburia} have been
-found in some studies to be increased in CRC but in others to either be
-decreased or have no difference {[}15,18,23,24{]}.
+recent individual studies investigating the microbiota and colorectal
+tumors, making it hard to know how extrapolatable these findings are.
+That particular meta-analysis also added more potential microbial
+associations to both carcinoma (\emph{Pantoea agglomerans}
+\emph{Ruminococcus}, \emph{Lactobacillus}) and adenoma
+(\emph{Prevotella}, \emph{Methanosphaera}, \emph{Succinovibrio},
+\emph{Haemophilus parainfluenzae}, \emph{Ruminococcus},
+\emph{Lactobacillus}) stages of disease that need to be investigated
+further, since a number of these genera have been found to be enriched
+in controls and not disease {[}13,16,17{]}. Additionally, genera like
+\emph{Roseburia} have been found in some studies to be increased in
+tumors but in others to either be decreased or have no difference
+{[}15,18,23,24{]}.
 
@@
 disease. Groups focusing on broad scale community metrics have found
-that metrics such as richness are decreased in the adenoma stage of CRC
-versus controls. Other studies have identified \emph{Lactococcus},
-\emph{Pseudomonas}, \emph{Acidovorax}, \emph{Cloacibacterium},
-\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Bilophila},
-\emph{Desulfovibrio}, and \emph{Mogibacterium} to be increased in
-adenoma {[}25--27{]}. Based on the studies mentioned, there seems to be
-very little overlap between the genera identified to be associated with
-adenoma and carcinoma, with \emph{Lactobacillus} being one of the few
-commonalities.
-
-Targeting the identification of CRC microbial biomarkers within stool
+that metrics such as richness are decreased in the adenoma stage of
+disease versus controls. Other studies have identified
+\emph{Lactococcus}, \emph{Pseudomonas}, \emph{Acidovorax},
+\emph{Cloacibacterium}, \emph{Helicobacter}, \emph{Lactobacillus},
+\emph{Bilophila}, \emph{Desulfovibrio}, and \emph{Mogibacterium} to be
+increased in adenoma {[}25--27{]}. Based on the studies mentioned, there
+seems to be very little overlap between the genera identified to be
+associated with adenoma and carcinoma, with \emph{Lactobacillus} being
+one of the few commonalities.
+
+Targeting the identification of tumor microbial biomarkers within stool
 seems logical since it offers an easy and cost-effective way to stratify
@@
 differences and similarities between stool and tissue microbiota in the
-context of CRC. Importantly, this analysis and approach could provide
-valuable insights into the common genera that are both protective and
-detrimental in CRC and whether broad bacterial community measurements
-can account for these changes that were not provided by earlier
-meta-analysis studies {[}11{]}.
+context of colorectal tumors. Importantly, this analysis and approach
+could provide valuable insights into the common genera that are both
+protective and detrimental in CRC and whether broad bacterial community
+measurements can account for these changes that were not provided by
+earlier meta-analysis studies {[}11{]}.
 
@@
 bacterial community or specific members are more important risk factors
-for both adenoma and carcinoma stages of CRC. To accomplish this we
+for both adenoma and carcinoma stages of disease. To accomplish this we
 first assessed whether bacterial diversity changes throughout disease
 (control to adenoma to carcinoma) and if it results in an increased
-relative risk (RR) for adenoma or carcinoma stages of CRC. We then
+relative risk (RR) for adenoma or carcinoma stages of disease We then
 assessed what genera, if any, increase or decrease the RR of adenoma or
-carcinoma stages of CRC. Next, using Random Forest models, we analyzed
-whether the full community or only the combined top 5 increased and top
-5 decreased RR genera resulted in better model classification, based on
-the area under the curve (AUC). Finally, we also examined at what effect
-and sample size the studies used were powered for and the sample size
-needed to get to the traditionally accepted 80\% power. Our results from
-these analyses suggests that the bacterial community changes as disease
-severity worsens, that specific members are important for disease
-classification, and that many of the studies are underpowered for
-assessing small effect sizes.
+carcinoma stages of disease Next, using Random Forest models, we
+analyzed whether the full community or only the combined top 5 increased
+and top 5 decreased RR genera resulted in better model classification,
+based on the area under the curve (AUC). Finally, we also examined at
+what effect and sample size the studies used were powered for and the
+sample size needed to get to the traditionally accepted 80\% power. Our
+results from these analyses suggests that the bacterial community
+changes as disease severity worsens, that specific members are important
+for disease classification, and that many of the studies are
+underpowered for assessing small effect sizes.
 
@@
 \textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
-of CRC:}} To assess differences in broad scale community metrics as
+of Tumors:}} To assess differences in broad scale community metrics as
 disease severity worsens Operational Taxonomic Unit (OTU) richness,
@@
 \textbf{\emph{Mouth-Associated Genera are Associated with an Increased
-RR of CRC:}} Next, we asked if being higher than the median relative
+RR of Tumor:}} Next, we asked if being higher than the median relative
 abundance, for any specific genera, resulted in an altered RR for
@@
 \emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma stage
-of CRC, certain mouth-associated genera (\emph{Fusobacterium},
+of disease, certain mouth-associated genera (\emph{Fusobacterium},
 \emph{Parvimonas}, \emph{Peptostreptococcus}) had a high RR for both
@@
 
-\textbf{\emph{Select Genera Models can Recapitulate Whole Community
+\textbf{\emph{Select Community Models can Recapitulate Whole Community
 Models:}} Since specific genera increased RR for carcinoma over
@@
 diversity as disease progressed from control to adenoma to carcinoma,
-this did not translate into large effect sizes for the RR of CRC. Even
-though mouth-associated genera increased the RR of having a carcinoma,
-they did not consistently increase the RR of having an adenoma.
-Additionally, our observations suggest that by combining
+this did not translate into large effect sizes for the RR of tumors.
+Even though mouth-associated genera increased the RR of having a
+carcinoma, they did not consistently increase the RR of having an
+adenoma. Additionally, our observations suggest that by combining
 mouth-associated and CRC protective microbes we can classify either
@@
 The colon tissue-based studies did not provide a clearer understanding
-of how the microbiota may be associated with CRC. Generally, the full
+of how the microbiota may be associated with tumors. Generally, the full
 OTU-based models of unmatched and matched colon tissue samples were
@@
 mouth-associated microbes could be detected for adenoma stage of
-disease. Our observations also highlight the importance of power and
-sample number considerations when investigating the microbiota and
-adenoma stage of disease due to the subtle changes in the community.
-Overall, associations of microbiota with the carcinoma stage of CRC are
-much stronger than those with the adenoma stage.
+disease. Overall, associations between the microbiota and carcinoma
+stage of disease were much stronger than with the adenoma stage. Our
+observations also highlight the importance of power and sample number
+considerations when investigating the microbiota and adenoma stage of
+disease due to the subtle changes in the community.
 
@@
 \textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
-tissue samples with control and CRC samples from different individuals
+tissue samples with control and tumor samples from different individuals
 were classified as unmatched while samples that belonged to the same

### 2018-02-03 code/make_stool_imp_otu_graph.R
correct spelling error

@@ -30,7 +30,7 @@ adn_genera <- adn_genera_occurances %>% arrange(occurance) %>%
   ggplot(aes(otu, occurance)) + 
   geom_bar(stat = "identity", fill = '#8B8B83') + 
   theme_bw() + 
-  labs(x = "", y = "Occurance Across Studies") + 
+  labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, 1)) + 
   ggtitle("A") + 
   scale_y_continuous(labels = scales::percent, expand = c(0,0.03)) + 
@@ -50,7 +50,7 @@ crc_genera <- crc_genera_occurances %>% arrange(occurance) %>%
   ggplot(aes(otu, occurance)) + 
   geom_bar(stat = "identity", fill = "black") + 
   theme_bw() + 
-  labs(x = "", y = "Occurance Across Studies") + 
+  labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, 1)) + 
   ggtitle("B") + 
   scale_y_continuous(labels = scales::percent, expand = c(0,0.03)) + 
@@ -69,7 +69,7 @@ adn_otu <- adn_otu_occurances %>% arrange(occurance) %>%
   ggplot(aes(genus, occurance)) + 
   geom_bar(stat = "identity", fill = '#8B8B83') + 
   theme_bw() + 
-  labs(x = "", y = "Occurance Across Studies") + 
+  labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, 1)) + 
   ggtitle("C") + 
   scale_y_continuous(labels = scales::percent, expand = c(0,0.03)) + 
@@ -88,7 +88,7 @@ crc_otu <- crc_otu_occurances %>% arrange(occurance) %>%
   ggplot(aes(genus, occurance)) + 
   geom_bar(stat = "identity", fill = 'black') + 
   theme_bw() + 
-  labs(x = "", y = "Occurance Across Studies") + 
+  labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, 1)) + 
   ggtitle("D") + 
   scale_y_continuous(labels = scales::percent, 

### 2018-02-03 code/make_tissue_imp_otu_graph.R

@@ -35,7 +35,7 @@ adn_genera <- adn_genera_occurances %>% arrange(occurance) %>%
   ggplot(aes(otu, occurance)) + 
   geom_bar(stat = "identity", fill = '#8B8B83') + 
   theme_bw() + 
-  labs(x = "", y = "Occurance Across Studies") + 
+  labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, 1)) + 
   ggtitle("A") + 
   scale_y_continuous(labels = scales::percent, expand = c(0,0.03)) + 
@@ -56,7 +56,7 @@ crc_unmatched_genera <- crc_unmatched_genera_occurances %>% arrange(occurance) %
   ggplot(aes(otu, occurance)) + 
   geom_bar(stat = "identity", fill = "black") + 
   theme_bw() + 
-  labs(x = "", y = "Occurance Across Studies") + 
+  labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, 1)) + 
   ggtitle("B") + 
   scale_y_continuous(labels = scales::percent, expand = c(0,0.03)) + 
@@ -75,7 +75,7 @@ crc_matched_genera <- crc_matched_genera_occurances %>% arrange(occurance) %>%
   ggplot(aes(otu, occurance)) + 
   geom_bar(stat = "identity", fill = "black") + 
   theme_bw() + 
-  labs(x = "", y = "Occurance Across Studies") + 
+  labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, 1)) + 
   ggtitle("C") + 
   scale_y_continuous(labels = scales::percent, expand = c(0,0.03)) + 
@@ -97,7 +97,7 @@ adn_otu <- adn_otu_occurances %>% arrange(occurance) %>%
   ggplot(aes(genus, occurance)) + 
   geom_bar(stat = "identity", fill = '#8B8B83') + 
   theme_bw() + 
-  labs(x = "", y = "Occurance Across Studies") + 
+  labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, 1)) + 
   ggtitle("D") + 
   scale_y_continuous(labels = scales::percent, expand = c(0,0.03)) + 
@@ -118,7 +118,7 @@ crc_unmatched_otu <- crc_unmatched_otu_occurances %>% arrange(occurance) %>%
   ggplot(aes(genus, occurance)) + 
[…]

### 2018-02-07 code/get_adenoma_tissue_RRisk.R
change to OR

@@ -101,7 +101,7 @@ run_rr <- function(high_low_vector, disease_vector){
   # runs the RR test based on the obtained 2x2 table
   test <- epi.2by2(contingency, method="cohort.count")
   # Pull only specific information from the stored list in "test"
-  test_values <- cbind(test$massoc$RR.strata.score, 
+  test_values <- cbind(test$massoc$OR.strata.score, 
                        pvalue = test$massoc$chisq.strata$p.value)
   # store both the obtained raw counts and the resulting RR with pvalue
   combined_data <- list(data_tbl = contingency, test_values = test_values)
@@ -149,7 +149,7 @@ run_pooled <- function(alpha_d, dataset = ind_counts_data){
   # Run the actual pooled test
   rr_pooled_test <- rma(ai = low_Y, bi = low_N, 
                         ci = high_Y, di = high_N, data = test_data, 
-                        measure = "RR", method = "REML")
+                        measure = "OR", method = "REML")
   # Store a vector of the important results of interest
   results <- c(exp(c(rr = rr_pooled_test$b[[1, 1]], ci_lb = rr_pooled_test$ci.lb, 
                      ci_ub=rr_pooled_test$ci.ub)), pvalue = rr_pooled_test$pval, 
@@ -188,8 +188,8 @@ pooled_results <- t(mapply(run_pooled, c("sobs", "shannon", "shannoneven"), USE.
 
 # Write out the important tables
 write.csv(ind_counts_data, "data/process/tables/alpha_adn_group_counts_tissue_summary.csv", row.names = F)
-write.csv(ind_RR_data, "data/process/tables/alpha_adn_RR_ind_tissue_results.csv", row.names = F)
-write.csv(pooled_results, "data/process/tables/alpha_adn_RR_tissue_composite.csv", row.names = F)
+write.csv(ind_RR_data, "data/process/tables/alpha_adn_OR_ind_tissue_results.csv", row.names = F)
+write.csv(pooled_results, "data/process/tables/alpha_adn_OR_tissue_composite.csv", row.names = F)
 
 
 

### 2018-02-07 code/get_specifc_genus_analysis_stool.R
switch to OR

@@ -196,7 +196,7 @@ run_rr <- function(high_low_vector, disease_vector){
   # runs the RR test based on the obtained 2x2 table
   test <- try(epi.2by2(contingency, method="cohort.count"))
   # Pull only specific information from the stored list in "test"
-  test_values <- try(cbind(test$massoc$RR.strata.score, 
+  test_values <- try(cbind(test$massoc$OR.strata.score, 
                        pvalue = test$massoc$chisq.strata$p.value))
   # store both the obtained raw counts and the resulting RR with pvalue
   combined_data <- try(list(data_tbl = contingency, test_values = test_values))
@@ -271,7 +271,7 @@ run_pooled <- function(alpha_d, dataset = ind_counts_data){
   # Run the actual pooled test
   rr_pooled_test <- rma(ai = high_Y, bi = high_N, 
                         ci = low_Y, di = low_N, data = test_data, 
-                        measure = "RR", method = "REML")
+                        measure = "OR", method = "REML")
   # Store a vector of the important results of interest
   results <- c(exp(c(rr = rr_pooled_test$b[[1, 1]], ci_lb = rr_pooled_test$ci.lb, 
                      ci_ub=rr_pooled_test$ci.ub)), pvalue = rr_pooled_test$pval, 
@@ -414,8 +414,8 @@ pooled_results <- t(mapply(run_pooled, all_genera, USE.NAMES = F)) %>%
 
 # Write out the important tables
 write.csv(ind_counts_data, "data/process/tables/select_genus_group_counts_summary.csv", row.names = F)
-write.csv(ind_RR_data, "data/process/tables/select_genus_RR_stool_ind_results.csv", row.names = F)
-write.csv(pooled_results, "data/process/tables/select_genus_RR_stool_composite.csv", row.names = F)
+write.csv(ind_RR_data, "data/process/tables/select_genus_OR_stool_ind_results.csv", row.names = F)
+write.csv(pooled_results, "data/process/tables/select_genus_OR_stool_composite.csv", row.names = F)
 
 
 

### 2018-02-07 code/get_specifc_genus_analysis_tissue.R
convert to OR and use emperical bayes method

@@ -180,7 +180,7 @@ run_rr <- function(high_low_vector, disease_vector){
   # runs the RR test based on the obtained 2x2 table
   test <- try(epi.2by2(contingency, method="cohort.count"))
   # Pull only specific information from the stored list in "test"
-  test_values <- try(cbind(test$massoc$RR.strata.score, 
+  test_values <- try(cbind(test$massoc$OR.strata.score, 
                            pvalue = test$massoc$chisq.strata$p.value))
   # store both the obtained raw counts and the resulting RR with pvalue
   combined_data <- try(list(data_tbl = contingency, test_values = test_values))
@@ -255,7 +255,7 @@ run_pooled <- function(alpha_d, dataset = ind_counts_data){
   # Run the actual pooled test
   rr_pooled_test <- rma(ai = high_Y, bi = high_N, 
                         ci = low_Y, di = low_N, data = test_data, 
-                        measure = "RR", method = "REML")
+                        measure = "OR", method = "EB")
   # Store a vector of the important results of interest
   results <- c(exp(c(rr = rr_pooled_test$b[[1, 1]], ci_lb = rr_pooled_test$ci.lb, 
                      ci_ub=rr_pooled_test$ci.ub)), pvalue = rr_pooled_test$pval, 
@@ -491,28 +491,28 @@ pooled_results <- t(sapply(all_genera,
 write.csv(matched_counts_data, 
           "data/process/tables/select_genus_matched_tissue_group_counts_summary.csv", 
           row.names = F)
-write.csv(matched_RR_data, "data/process/tables/select_genus_RR_matched_tissue_ind_results.csv", 
+write.csv(matched_RR_data, "data/process/tables/select_genus_OR_matched_tissue_ind_results.csv", 
           row.names = F)
-write.csv(matched_pooled_results, "data/process/tables/select_genus_RR_matched_tissue_composite.csv", 
+write.csv(matched_pooled_results, "data/process/tables/select_genus_OR_matched_tissue_composite.csv", 
           row.names = F)
 
 
 write.csv(unmatched_counts_data, 
           "data/process/tables/select_genus_unmatched_tissue_group_counts_summary.csv", 
           row.names = F)
-write.csv(unmatched_RR_data, "data/process/tables/select_genus_RR_unmatched_tissue_ind_results.csv", 
+write.csv(unmatched_RR_data, "data/process/tables/select_genus_OR_unmatched_tissue_ind_results.csv", 
           row.names = F)
[…]

### 2018-02-07 code/get_adenoma_specifc_genus_analysis_stool.R
change to OR

@@ -168,7 +168,7 @@ run_rr <- function(high_low_vector, disease_vector){
   # runs the RR test based on the obtained 2x2 table
   test <- try(epi.2by2(contingency, method="cohort.count"))
   # Pull only specific information from the stored list in "test"
-  test_values <- try(cbind(test$massoc$RR.strata.score, 
+  test_values <- try(cbind(test$massoc$OR.strata.score, 
                        pvalue = test$massoc$chisq.strata$p.value))
   # store both the obtained raw counts and the resulting RR with pvalue
   combined_data <- try(list(data_tbl = contingency, test_values = test_values))
@@ -242,7 +242,7 @@ run_pooled <- function(alpha_d, dataset = ind_counts_data){
   # Run the actual pooled test
   rr_pooled_test <- rma(ai = high_Y, bi = high_N, 
                         ci = low_Y, di = low_N, data = test_data, 
-                        measure = "RR", method = "REML")
+                        measure = "OR", method = "REML")
   # Store a vector of the important results of interest
   results <- c(exp(c(rr = rr_pooled_test$b[[1, 1]], ci_lb = rr_pooled_test$ci.lb, 
                      ci_ub=rr_pooled_test$ci.ub)), pvalue = rr_pooled_test$pval, 
@@ -353,8 +353,8 @@ pooled_results <- t(mapply(run_pooled, all_genera, USE.NAMES = F)) %>%
 
 # Write out the important tables
 write.csv(ind_counts_data, "data/process/tables/adn_select_genus_group_counts_summary.csv", row.names = F)
-write.csv(ind_RR_data, "data/process/tables/adn_select_genus_RR_stool_ind_results.csv", row.names = F)
-write.csv(pooled_results, "data/process/tables/adn_select_genus_RR_stool_composite.csv", row.names = F)
+write.csv(ind_RR_data, "data/process/tables/adn_select_genus_OR_stool_ind_results.csv", row.names = F)
+write.csv(pooled_results, "data/process/tables/adn_select_genus_OR_stool_composite.csv", row.names = F)
 
 
 

### 2018-02-07 code/get_adenoma_specific_genus_analysis_tissue.R
change to OR

@@ -190,7 +190,7 @@ run_rr <- function(high_low_vector, disease_vector){
   # runs the RR test based on the obtained 2x2 table
   test <- try(epi.2by2(contingency, method="cohort.count"))
   # Pull only specific information from the stored list in "test"
-  test_values <- try(cbind(test$massoc$RR.strata.score, 
+  test_values <- try(cbind(test$massoc$OR.strata.score, 
                            pvalue = test$massoc$chisq.strata$p.value))
   # store both the obtained raw counts and the resulting RR with pvalue
   combined_data <- try(list(data_tbl = contingency, test_values = test_values))
@@ -265,7 +265,7 @@ run_pooled <- function(alpha_d, dataset){
   # Run the actual pooled test
   rr_pooled_test <- rma(ai = high_Y, bi = high_N, 
                         ci = low_Y, di = low_N, data = test_data, 
-                        measure = "RR", method = "REML")
+                        measure = "OR", method = "REML")
   # Store a vector of the important results of interest
   results <- c(exp(c(rr = rr_pooled_test$b[[1, 1]], ci_lb = rr_pooled_test$ci.lb, 
                      ci_ub=rr_pooled_test$ci.ub)), pvalue = rr_pooled_test$pval, 
@@ -387,9 +387,9 @@ pooled_results <- t(sapply(all_genera,
 write.csv(counts_data, 
           "data/process/tables/adn_select_genus_tissue_group_counts_summary.csv", 
           row.names = F)
-write.csv(RR_data, "data/process/tables/adn_select_genus_RR_tissue_ind_results.csv", 
+write.csv(RR_data, "data/process/tables/adn_select_genus_OR_tissue_ind_results.csv", 
           row.names = F)
-write.csv(pooled_results, "data/process/tables/adn_select_genus_RR_tissue_composite.csv", 
+write.csv(pooled_results, "data/process/tables/adn_select_genus_OR_tissue_composite.csv", 
           row.names = F)
 
 

### 2018-02-07 code/get_stool_RRisk.R
change to OR

@@ -89,7 +89,7 @@ run_rr <- function(high_low_vector, disease_vector){
   # runs the RR test based on the obtained 2x2 table
   test <- epi.2by2(contingency, method="cohort.count")
   # Pull only specific information from the stored list in "test"
-  test_values <- cbind(test$massoc$RR.strata.score, 
+  test_values <- cbind(test$massoc$OR.strata.score, 
                        pvalue = test$massoc$chisq.strata$p.value)
   # store both the obtained raw counts and the resulting RR with pvalue
   combined_data <- list(data_tbl = contingency, test_values = test_values)
@@ -139,7 +139,7 @@ run_pooled <- function(alpha_d, dataset = ind_counts_data){
   # Run the actual pooled test
   rr_pooled_test <- rma(ai = low_Y, bi = low_N, 
                         ci = high_Y, di = high_N, data = test_data, 
-                        measure = "RR", method = "REML")
+                        measure = "OR", method = "REML")
   # Store a vector of the important results of interest
   results <- c(exp(c(rr = rr_pooled_test$b[[1, 1]], ci_lb = rr_pooled_test$ci.lb, 
                ci_ub=rr_pooled_test$ci.ub)), pvalue = rr_pooled_test$pval, 
@@ -178,7 +178,7 @@ pooled_results <- t(mapply(run_pooled, c("sobs", "shannon", "shannoneven"), USE.
 
 # Write out the important tables
 write.csv(ind_counts_data, "data/process/tables/alpha_group_counts_summary.csv", row.names = F)
-write.csv(ind_RR_data, "data/process/tables/alpha_RR_ind_results.csv", row.names = F)
-write.csv(pooled_results, "data/process/tables/alpha_RR_composite.csv", row.names = F)
+write.csv(ind_RR_data, "data/process/tables/alpha_OR_ind_results.csv", row.names = F)
+write.csv(pooled_results, "data/process/tables/alpha_OR_composite.csv", row.names = F)
 
 

### 2018-02-07 code/get_tissue_RRisk.R
change to OR

@@ -103,7 +103,7 @@ run_rr <- function(high_low_vector, disease_vector){
   # runs the RR test based on the obtained 2x2 table
   test <- epi.2by2(contingency, method="cohort.count")
   # Pull only specific information from the stored list in "test"
-  test_values <- cbind(test$massoc$RR.strata.score, 
+  test_values <- cbind(test$massoc$OR.strata.score, 
                        pvalue = test$massoc$chisq.strata$p.value)
   # store both the obtained raw counts and the resulting RR with pvalue
   combined_data <- list(data_tbl = contingency, test_values = test_values)
@@ -154,7 +154,7 @@ run_pooled <- function(alpha_d, dataset){
   # Run the actual pooled test
   rr_pooled_test <- rma(ai = low_Y, bi = low_N, 
                         ci = high_Y, di = high_N, data = test_data, 
-                        measure = "RR", method = "REML")
+                        measure = "OR", method = "REML")
   # Store a vector of the important results of interest
   results <- c(exp(c(rr = rr_pooled_test$b[[1, 1]], ci_lb = rr_pooled_test$ci.lb, 
                      ci_ub=rr_pooled_test$ci.ub)), pvalue = rr_pooled_test$pval, 
@@ -204,8 +204,8 @@ unmatched_pooled_results <- t(sapply(c("sobs", "shannon", "shannoneven"),
 # Write out the important tables
 write_csv(unmatched_ind_counts_data, 
           "data/process/tables/alpha_group_counts_unmatched_tissue_summary.csv")
-write_csv(unmatched_ind_RR_data, "data/process/tables/alpha_RR_ind_unmatched_tissue_results.csv")
-write_csv(unmatched_pooled_results, "data/process/tables/alpha_RR_unmatched_tissue_composite.csv")
+write_csv(unmatched_ind_RR_data, "data/process/tables/alpha_OR_ind_unmatched_tissue_results.csv")
+write_csv(unmatched_pooled_results, "data/process/tables/alpha_OR_unmatched_tissue_composite.csv")
 
 ##############################################################################################
 ############### Run the actual programs to get the data (combined) ###########################
@@ -238,8 +238,8 @@ pooled_results <- t(sapply(c("sobs", "shannon", "shannoneven"),
 
 # Write out the important tables
 write.csv(ind_counts_data, "data/process/tables/alpha_group_counts_tissue_summary.csv", row.names = F)
[…]

### 2018-02-07 submission/supplemental.Rmd
change the title
fix title
change to OR instead of RR

@@
 
-# Investigating the Microbiota and Colorectal Cancer: The Importance of Community
+# Making Sense of the Noise: Leveraging Existing 16S rRNA Gene Surveys to Identify Key Community Members in Colorectal Tumors
 
@@
 
-**Table S3: Alpha Diversity Metrics RR in Tissue for Adenoma**
+**Table S3: Alpha Diversity Metrics OR in Tissue for Adenoma**
 
@@
 
-alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_RR_tissue_composite.csv")) %>% 
+alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_OR_tissue_composite.csv")) %>% 
   mutate(rr = round(rr, digits = 2), 
@@
 
-colnames(alpha_RR_adn_tissue) <- c("Relative Risk", "Lower Bound", "Upper Bound", "P-value", "Measure")
+colnames(alpha_RR_adn_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure")
 
@@
 
-**Table S4: Alpha Diversity Metrics RR in Tissue for Unmatched Carcinoma Non-Carcinoma Samples**
+**Table S4: Alpha Diversity Metrics OR in Tissue for Unmatched Carcinoma Non-Carcinoma Samples**
 
@@
 
-alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_RR_unmatched_tissue_composite.csv")) %>%   mutate(rr = round(rr, digits = 2), 
+alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_unmatched_tissue_composite.csv")) %>%   mutate(rr = round(rr, digits = 2), 
          ci_lb = round(ci_lb, digits = 2), 
@@
 
-colnames(alpha_RR_crc_unmatch_tissue) <- c("Relative Risk", "Lower Bound", "Upper Bound", "P-value", "Measure")
+colnames(alpha_RR_crc_unmatch_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure")
 
@@
 
-**Table S5: Alpha Diversity Metrics RR in Tissue for Matched Carcinoma Non-Carcinoma Samples**
+**Table S5: Alpha Diversity Metrics OR in Tissue for Matched Carcinoma Non-Carcinoma Samples**
 
@@
 
-alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_RR_matched_tissue_composite.csv")) %>% 
+alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_matched_tissue_composite.csv")) %>% 
   mutate(rr = round(rr, digits = 2), 
@@
 
-colnames(alpha_RR_crc_match_tissue) <- c("Relative Risk", "Lower Bound", "Upper Bound", "P-value", "Measure")
+colnames(alpha_RR_crc_match_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure")
 
@@
 
-**Table S8: RR of Adenoma in Stool for each Individual Genera with Initial P-Value under 0.05**
+**Table S8: OR of Adenoma in Stool for each Individual Genera with Initial P-Value under 0.05**
 
@@
 
-temp_8 <- read_csv("../data/process/tables/adn_select_genus_RR_stool_composite.csv") %>% 
+temp_8 <- read_csv("../data/process/tables/adn_select_genus_OR_stool_composite.csv") %>% 
   mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
@@
 
-colnames(temp_8) <- c("RR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+colnames(temp_8) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
 
@@
 
-**Table S9: RR of Adenoma in Tissue for each Individual Genera with Initial P-Value under 0.05**
+**Table S9: OR of Adenoma in Tissue for each Individual Genera with Initial P-Value under 0.05**
 
@@
 
-temp_9 <- read_csv("../data/process/tables/adn_select_genus_RR_tissue_composite.csv") %>% 
+temp_9 <- read_csv("../data/process/tables/adn_select_genus_OR_tissue_composite.csv") %>% 
   mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
@@
 
-colnames(temp_9) <- c("RR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+colnames(temp_9) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
 
@@
 
-**Table S10: RR of Carcinoma in Stool for each Individual Genera with Initial P-Value under 0.05**
+**Table S10: OR of Carcinoma in Stool for each Individual Genera with Initial P-Value under 0.05**
 
@@
 
-temp_10 <- read_csv("../data/process/tables/select_genus_RR_stool_composite.csv") %>% 
+temp_10 <- read_csv("../data/process/tables/select_genus_OR_stool_composite.csv") %>% 
   mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
@@
 
-colnames(temp_10) <- c("RR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+colnames(temp_10) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
 
@@
 
-**Table S11: RR of Carcinoma in Tissue for each Individual Genera with Initial P-Value under 0.05**
+**Table S11: OR of Carcinoma in Tissue for each Individual Genera with Initial P-Value under 0.05**
 
@@
 
-temp_11 <- read_csv("../data/process/tables/select_genus_RR_tissue_composite.csv") %>% 
+temp_11 <- read_csv("../data/process/tables/select_genus_OR_tissue_composite.csv") %>% 
   mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
@@
 
-colnames(temp_11) <- c("RR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+colnames(temp_11) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
 

### 2018-02-07 submission/manuscript.Rmd
change to OR

@@
 
-#Load in relative risk alpha diversity data tables
-alpha_RR_crc_stool <- as.data.frame(read_csv("../data/process/tables/alpha_RR_composite.csv"))
-alpha_RR_adn_stool <- as.data.frame(read_csv("../data/process/tables/alpha_adn_RR_composite.csv"))
+#Load in odds ratio alpha diversity data tables
+alpha_RR_crc_stool <- as.data.frame(read_csv("../data/process/tables/alpha_OR_composite.csv"))
+alpha_RR_adn_stool <- as.data.frame(read_csv("../data/process/tables/alpha_adn_OR_composite.csv"))
 
-alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_RR_unmatched_tissue_composite.csv"))
-alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_RR_matched_tissue_composite.csv"))
+alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_unmatched_tissue_composite.csv"))
+alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_matched_tissue_composite.csv"))
 
-alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_RR_tissue_composite.csv"))
+alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_OR_tissue_composite.csv"))
 
-#Load in relative risk select genera data tables
-select_RR_stool <- as.data.frame(read_csv("../data/process/tables/select_genus_RR_stool_composite.csv"))
-select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_RR_tissue_composite.csv"))
+#Load in odds ratio select genera data tables
+select_RR_stool <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_stool_composite.csv"))
+select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_tissue_composite.csv"))
 
-adn_select_RR_stool <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_RR_stool_composite.csv"))
-adn_select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_RR_tissue_composite.csv"))
+adn_select_RR_stool <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_OR_stool_composite.csv"))
+adn_select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_OR_tissue_composite.csv"))
 
@@
 **Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness, after correcting for study effect and variable region sequenced, with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
-This reduction in evenness translated into small increases in relative risk for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased relative risk for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
+This reduction in evenness translated into small increases in the odds ratio for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
 Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples.
@@
 Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of disease.
-To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma stages of disease We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma stages of disease Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased RR genera resulted in better model classification, based on the area under the curve (AUC).
+To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased odds ratio (OR) for adenoma or carcinoma stages of disease We then assessed what genera, if any, increase or decrease the OR of adenoma or carcinoma stages of disease Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased OR genera resulted in better model classification, based on the area under the curve (AUC).
 Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power.
@@
 
-***Lower Bacterial Diversity is Associated with Increased RR of Tumors:***
+***Lower Bacterial Diversity is Associated with Increased OR of Tumors:***
 To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized.
@@
 In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A].
-We next tested whether the detectable differences in community significantly increased in RR of having an adenoma or carcinoma.
-For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
+We next tested whether the detectable differences in community significantly increased in OR of having an adenoma or carcinoma.
+For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the OR for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
 Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].
@@
 
-***Mouth-Associated Genera are Associated with an Increased RR of Tumor:***
-Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered RR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in RR using OTU richness and Shannon diversity.
+***Mouth-Associated Genera are Associated with an Increased OR of Tumor:***
+Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered OR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in OR using OTU richness and Shannon diversity.
 To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR.
-Mouth-associated genera were commonly found in the top 5 genera associated with an increased RR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples.
-Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3].
-Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased RR genera between both stages and sample site.
+Mouth-associated genera were commonly found in the top 5 genera associated with an increased OR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples.
+Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased OR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3].
+Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased OR genera between both stages and sample site.
 
-When observing RRs with a P-value less than 0.05 there was almost no overlap between genera from stool or colon tissue and when they were similar the RR was in opposite directions (e.g.
+When observing ORs with a P-value less than 0.05 there was almost no overlap between genera from stool or colon tissue and when they were similar the RR was in opposite directions (e.g.
 *Lactococcus*) [Table S8 & S9].
-Many of the genera that had RRs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
-For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high RR for both colon tissue and stool samples [Table S10 & S11].
+Many of the genera that had ORs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
+For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high OR for both colon tissue and stool samples [Table S10 & S11].
 Finally, these data suggest that the most significantly increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].
@@
 ***Select Community Models can Recapitulate Whole Community Models:***
-Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera.
-We selected these genera based on their RR and P-value significance and used two approaches to test this question.
+Since specific genera increased OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera.
+We selected these genera based on their OR and P-value significance and used two approaches to test this question.
 The first approach used genus level data and tested for differences in AUC between all genera and selected genera.
@@
 Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with and without CRC [Figure 1-3].
-Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of tumors.
-Even though mouth-associated genera increased the RR of having a carcinoma, they did not consistently increase the RR of having an adenoma.
+Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the OR of tumors.
+Even though mouth-associated genera increased the OR of having a carcinoma, they did not consistently increase the RR of having an adenoma.
 Additionally, our observations suggest that by combining mouth-associated and CRC protective microbes we can classify either adenoma or carcinoma stage of disease as well as models that use the full community.
@@
 By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in CRC.
-Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest RR of having a carcinoma.
+Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest OR of having a carcinoma.
 No conclusive signal with these mouth-associated microbes could be detected for adenoma stage of disease.
@@
 The OTU based Random Forest Models using selected OTUs utilized a similar approach except that the OTUs were not aggregated together by taxonomic identity but kept as separate OTUs.
-OTU Random Forest models using the full community included all OTUs while those for the selected model included only those OTUs that had a taxonomic classification to a variable in the top 5 increased of top 5 decreased RR based on P-value.
+OTU Random Forest models using the full community included all OTUs while those for the selected model included only those OTUs that had a taxonomic classification to a variable in the top 5 increased of top 5 decreased OR based on P-value.
 

### 2018-02-07 submission/manuscript.tex
change to OR

@@
 variable region sequenced, with more severe disease (P-value \textless{}
-0.05). This reduction in evenness translated into small increases in
-relative risk for adenoma (P-value = 0.032) and carcinoma stages of CRC
-(P-value = 0.00034) while the reduction in Shannon diversity only
-translated into an increased relative risk for developing carcinomas
-(P-value = 0.0047). Increases in mouth-associated microbes were commonly
+0.05). This reduction in evenness translated into small increases in the
+odds ratio for adenoma (P-value = 0.035) and carcinoma stages of CRC
+(P-value = 0.0021) while the reduction in Shannon diversity only
+translated into an increased odds ratio for developing carcinomas
+(P-value = 0.0069). Increases in mouth-associated microbes were commonly
 in the top 5 most significantly increased relative risk of adenoma and
@@
 first assessed whether bacterial diversity changes throughout disease
-(control to adenoma to carcinoma) and if it results in an increased
-relative risk (RR) for adenoma or carcinoma stages of disease We then
-assessed what genera, if any, increase or decrease the RR of adenoma or
-carcinoma stages of disease Next, using Random Forest models, we
-analyzed whether the full community or only the combined top 5 increased
-and top 5 decreased RR genera resulted in better model classification,
-based on the area under the curve (AUC). Finally, we also examined at
-what effect and sample size the studies used were powered for and the
-sample size needed to get to the traditionally accepted 80\% power. Our
-results from these analyses suggests that the bacterial community
-changes as disease severity worsens, that specific members are important
-for disease classification, and that many of the studies are
-underpowered for assessing small effect sizes.
+(control to adenoma to carcinoma) and if it results in an increased odds
+ratio (OR) for adenoma or carcinoma stages of disease We then assessed
+what genera, if any, increase or decrease the OR of adenoma or carcinoma
+stages of disease Next, using Random Forest models, we analyzed whether
+the full community or only the combined top 5 increased and top 5
+decreased OR genera resulted in better model classification, based on
+the area under the curve (AUC). Finally, we also examined at what effect
+and sample size the studies used were powered for and the sample size
+needed to get to the traditionally accepted 80\% power. Our results from
+these analyses suggests that the bacterial community changes as disease
+severity worsens, that specific members are important for disease
+classification, and that many of the studies are underpowered for
+assessing small effect sizes.
 
@@
 
-\textbf{\emph{Lower Bacterial Diversity is Associated with Increased RR
+\textbf{\emph{Lower Bacterial Diversity is Associated with Increased OR
 of Tumors:}} To assess differences in broad scale community metrics as
@@
 tested whether the detectable differences in community significantly
-increased in RR of having an adenoma or carcinoma. For fecal samples, a
+increased in OR of having an adenoma or carcinoma. For fecal samples, a
 decrease versus the overall median in evenness resulted in a
-significantly increased RR for carcinoma (RR = 1.36 (1.15 - 1.61),
-P-value = 0.00034) and adenoma (RR = 1.16 (1.01 - 1.34), P-value =
-0.032) while a decrease versus the overall median in Shannon diversity
-only increased the RR for carcinoma (RR = 1.33 (1.09 - 1.62), P-value =
-0.0047) {[}Figure 2{]}. Using the Bray-Curtis distance metric and
-PERMANOVA, it was also possible to identify significant bacterial
-community changes, in specific studies, for both carcinoma-associated
-and adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.
+significantly increased RR for carcinoma (OR = 1.66 (1.2 - 2.3), P-value
+= 0.0021) and adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) while a
+decrease versus the overall median in Shannon diversity only increased
+the OR for carcinoma (OR = 1.61 (1.14 - 2.28), P-value = 0.0069)
+{[}Figure 2{]}. Using the Bray-Curtis distance metric and PERMANOVA, it
+was also possible to identify significant bacterial community changes,
+in specific studies, for both carcinoma-associated and
+adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.
 
@@
 \textbf{\emph{Mouth-Associated Genera are Associated with an Increased
-RR of Tumor:}} Next, we asked if being higher than the median relative
-abundance, for any specific genera, resulted in an altered RR for
+OR of Tumor:}} Next, we asked if being higher than the median relative
+abundance, for any specific genera, resulted in an altered OR for
 adenoma or carcinoma, in stool and colon tissue, due to our previous
-observations of small increases in RR using OTU richness and Shannon
+observations of small increases in OR using OTU richness and Shannon
 diversity. To investigate this we analyzed all common genera across each
@@
 RR. Mouth-associated genera were commonly found in the top 5 genera
-associated with an increased RR of having an adenoma
+associated with an increased OR of having an adenoma
 (\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
@@
 samples. Conversely, genera commonly associated with a normal
-gastrointestinal tract were correlated with a decreased RR for both
+gastrointestinal tract were correlated with a decreased OR for both
 adenoma and carcinoma for both stool and colon tissue samples {[}Figure
@@
 stage, there was little direct overlap of the top 5 increased or
-decreased RR genera between both stages and sample site.
-
-When observing RRs with a P-value less than 0.05 there was almost no
+decreased OR genera between both stages and sample site.
+
+When observing ORs with a P-value less than 0.05 there was almost no
 overlap between genera from stool or colon tissue and when they were
 similar the RR was in opposite directions (e.g. \emph{Lactococcus})
-{[}Table S8 \& S9{]}. Many of the genera that had RRs with a P-value
+{[}Table S8 \& S9{]}. Many of the genera that had ORs with a P-value
 under 0.05 for colon tissue are also highly prevalent in contamination,
@@
 of disease, certain mouth-associated genera (\emph{Fusobacterium},
-\emph{Parvimonas}, \emph{Peptostreptococcus}) had a high RR for both
+\emph{Parvimonas}, \emph{Peptostreptococcus}) had a high OR for both
 colon tissue and stool samples {[}Table S10 \& S11{]}. Finally, these
@@
 \textbf{\emph{Select Community Models can Recapitulate Whole Community
-Models:}} Since specific genera increased RR for carcinoma over
+Models:}} Since specific genera increased OR for carcinoma over
 diversity metrics we assessed whether the bacterial community was better
 at classifying disease versus only a select group of genera. We selected
-these genera based on their RR and P-value significance and used two
+these genera based on their OR and P-value significance and used two
 approaches to test this question. The first approach used genus level
@@
 diversity as disease progressed from control to adenoma to carcinoma,
-this did not translate into large effect sizes for the RR of tumors.
-Even though mouth-associated genera increased the RR of having a
+this did not translate into large effect sizes for the OR of tumors.
+Even though mouth-associated genera increased the OR of having a
 carcinoma, they did not consistently increase the RR of having an
@@
 a foothold within the colon and are are commonly associated with the
-greatest RR of having a carcinoma. No conclusive signal with these
+greatest OR of having a carcinoma. No conclusive signal with these
 mouth-associated microbes could be detected for adenoma stage of
@@
 included only those OTUs that had a taxonomic classification to a
-variable in the top 5 increased of top 5 decreased RR based on P-value.
+variable in the top 5 increased of top 5 decreased OR based on P-value.
 

### 2018-02-07 code/make_stool_alpha_RR_graph.R
switch to OR

@@ -10,17 +10,17 @@ loadLibs(c("tidyverse", "gridExtra", "viridis"))
 
 
 # Load in needed data tables
-adn_all_stool <- read_csv("data/process/tables/alpha_adn_RR_composite.csv") %>% 
+adn_all_stool <- read_csv("data/process/tables/alpha_adn_OR_composite.csv") %>% 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% 
-  bind_rows(read_csv("data/process/tables/alpha_adn_RR_ind_results.csv")) %>% 
+  bind_rows(read_csv("data/process/tables/alpha_adn_OR_ind_results.csv")) %>% 
   mutate(region = c(rep("combined", 3), rep("V1-V3", 3), rep("V4", 6), rep("V3-V5", 3)))
 
 
-crc_all_stool <- read_csv("data/process/tables/alpha_RR_composite.csv") %>% 
+crc_all_stool <- read_csv("data/process/tables/alpha_OR_composite.csv") %>% 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% 
-  bind_rows(read_csv("data/process/tables/alpha_RR_ind_results.csv")) %>% 
+  bind_rows(read_csv("data/process/tables/alpha_OR_ind_results.csv")) %>% 
   mutate(region = c(rep("combined", 3), rep("V3", 3), rep("V4", 3), rep("V3-V4", 3), 
                     rep("V4", 6), rep("V3-V5", 3), rep("V3-V4", 3)))
 
@@ -58,12 +58,12 @@ adn_stool_graph <- adn_all_stool %>%
                           levels = c("sobs", "shannoneven", "shannon"), 
                           labels = c("Observed OTUs", "Evenness", "Shannon Diversity"))) %>% 
   ggplot(aes(log2(est), study, xmax=log2(upper), xmin=log2(lower), colour=region)) + 
-  coord_cartesian(xlim=c(-4.2, 4.2)) + 
+  coord_cartesian(xlim=c(-5.2, 5.2)) + 
   geom_vline(xintercept = 0.0, linetype=2, alpha=0.75) + 
   geom_errorbarh(alpha=0.5, size = 1, height=0, show.legend = T) + 
   geom_point(size = 3, show.legend = T) + 
   facet_grid(. ~ measure) + 
[…]

### 2018-02-07 submission/manuscript.Rmd
change legends to OR

@@
 
-**Figure 2: Relative Risk for Adenoma or Carcinoma based on Bacterial Community Metrics in Stool.** A) Community-based relative risk for adenoma.
-B) Community-based relative risk for carcinoma.
+**Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Stool.** A) Community-based odds ratio for adenoma.
+B) Community-based odds ratio for carcinoma.
 Colors represent the different variable regions used within the respective study.
 
-**Figure 3: Top 5 Genera that Decrease and Increase Relative Risk for Lesion.** A) Adenoma relative risk in stool.
-B) Carcinoma relative risk in stool.
-C) Adenoma relative risk in tissue.
-D) Carcinoma relative risk in tissue.
-For all panels the relative risk was also compared to whether one, two, three, or four of the CRC-associated genera were present.
+**Figure 3: Top 5 Genera that Decrease and Increase Odds Ratio for Lesion.** A) Adenoma odds ratio in stool.
+B) Carcinoma odds ratio in stool.
+C) Adenoma odds ratio in tissue.
+D) Carcinoma odds ratio in tissue.
+For all panels the odds ratio was also compared to whether one, two, three, or four of the CRC-associated genera were present.
 

### 2018-02-07 submission/manuscript.tex
change legends to OR

@@
 
-\textbf{Figure 2: Relative Risk for Adenoma or Carcinoma based on
-Bacterial Community Metrics in Stool.} A) Community-based relative risk
-for adenoma. B) Community-based relative risk for carcinoma. Colors
-represent the different variable regions used within the respective
-study.
-
-\textbf{Figure 3: Top 5 Genera that Decrease and Increase Relative Risk
-for Lesion.} A) Adenoma relative risk in stool. B) Carcinoma relative
-risk in stool. C) Adenoma relative risk in tissue. D) Carcinoma relative
-risk in tissue. For all panels the relative risk was also compared to
-whether one, two, three, or four of the CRC-associated genera were
-present.
+\textbf{Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial
+Community Metrics in Stool.} A) Community-based odds ratio for adenoma.
+B) Community-based odds ratio for carcinoma. Colors represent the
+different variable regions used within the respective study.
+
+\textbf{Figure 3: Top 5 Genera that Decrease and Increase Odds Ratio for
+Lesion.} A) Adenoma odds ratio in stool. B) Carcinoma odds ratio in
+stool. C) Adenoma odds ratio in tissue. D) Carcinoma odds ratio in
+tissue. For all panels the odds ratio was also compared to whether one,
+two, three, or four of the CRC-associated genera were present.
 

### 2018-02-07 code/make_select_genus_RR_graph.R
change to OR

@@ -8,80 +8,80 @@ source("code/functions.R")
 loadLibs(c("tidyverse", "gridExtra", "viridis"))
 
 # Load needed data tables (adenoma)
-adn_all_stool <- read_csv("data/process/tables/adn_select_genus_RR_stool_composite.csv") %>% 
+adn_all_stool <- read_csv("data/process/tables/adn_select_genus_OR_stool_composite.csv") %>% 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% arrange(pvalue, est) %>% 
   filter(est > 1) %>% slice(1:5) %>% 
-  bind_rows(read_csv("data/process/tables/adn_select_genus_RR_stool_composite.csv") %>% 
+  bind_rows(read_csv("data/process/tables/adn_select_genus_OR_stool_composite.csv") %>% 
               rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
               mutate(study = "composite") %>% arrange(pvalue, est) %>% 
               filter(est < 1) %>% slice(1:5)) %>% 
   mutate(high_low = ifelse(est > 1, invisible("high"), invisible("low")))
 
-adn_ind_stool <- read_csv("data/process/tables/adn_select_genus_RR_stool_ind_results.csv") %>% 
+adn_ind_stool <- read_csv("data/process/tables/adn_select_genus_OR_stool_ind_results.csv") %>% 
   filter(measure %in% as.data.frame(adn_all_stool)[, "measure"]) %>% 
   bind_rows(adn_all_stool) %>% 
-  mutate(high_low = ifelse(measure %in% c("Pyramidobacter", "Clostridium_XlVb", 
-                                          "Candidatus_Saccharibacteria_unclassified", 
-                                          "Novosphingobium", "Bacteroidales_unclassified"), 
+  mutate(high_low = ifelse(measure %in% c("Clostridium_XlVb", "Porphyromonas", 
+                                          "Novosphingobium", "Bacteroidales_unclassified", 
+                                          "Catenibacterium"), 
                            invisible("high"), invisible("low")))
 
 
-adn_all_tissue <- read_csv("data/process/tables/adn_select_genus_RR_tissue_composite.csv") %>% 
+adn_all_tissue <- read_csv("data/process/tables/adn_select_genus_OR_tissue_composite.csv") %>% 
[…]

### 2018-02-07 submission/manuscript.Rmd
update genera based on OR

@@
 To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR.
-Mouth-associated genera were commonly found in the top 5 genera associated with an increased OR of having an adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples.
+Mouth-associated genera were commonly found in the top 5 genera associated with an increased OR of having an adenoma (*Porphyromonas* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples.
 Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased OR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3].
@@
 *Lactococcus*) [Table S8 & S9].
-Many of the genera that had ORs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Selemonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
-For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*) had a high OR for both colon tissue and stool samples [Table S10 & S11].
-Finally, these data suggest that the most significantly increased RR genera for tissue was *Camplyobacter* while in stool it was *Peptostreptococcus* [Table S10 & S11].
+Many of the genera that had ORs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Pseudomonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
+For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*) had an increased OR for both colon tissue and stool samples [Table S10 & S11].
+Finally, these data suggest that the most significantly increased OR genera for tissue was *Lepttrichia* while in stool it was *Peptostreptococcus* [Table S10 & S11].
 
@@
 For all panels the odds ratio was also compared to whether one, two, three, or four of the CRC-associated genera were present.
+Points represented as only half on the graph have an OR of infinity in the positive or negative direction.
 

### 2018-02-07 submission/manuscript.tex
update genera based on OR

@@
 associated with an increased OR of having an adenoma
-(\emph{Pyramidobacter} {[}Figure 3A{]} and \emph{Rothia} {[}Figure
-3C{]}) and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
+(\emph{Porphyromonas} {[}Figure 3A{]} and \emph{Rothia} {[}Figure 3C{]})
+and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
 \emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
@@
 under 0.05 for colon tissue are also highly prevalent in contamination,
-specifically, \emph{Novosphingobium}, \emph{Selemonas}, and
+specifically, \emph{Novosphingobium}, \emph{Pseudomonas}, and
 \emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma stage
 of disease, certain mouth-associated genera (\emph{Fusobacterium},
-\emph{Parvimonas}, \emph{Peptostreptococcus}) had a high OR for both
-colon tissue and stool samples {[}Table S10 \& S11{]}. Finally, these
-data suggest that the most significantly increased RR genera for tissue
-was \emph{Camplyobacter} while in stool it was \emph{Peptostreptococcus}
-{[}Table S10 \& S11{]}.
+\emph{Parvimonas}) had an increased OR for both colon tissue and stool
+samples {[}Table S10 \& S11{]}. Finally, these data suggest that the
+most significantly increased OR genera for tissue was \emph{Lepttrichia}
+while in stool it was \emph{Peptostreptococcus} {[}Table S10 \& S11{]}.
 
@@
 tissue. For all panels the odds ratio was also compared to whether one,
-two, three, or four of the CRC-associated genera were present.
+two, three, or four of the CRC-associated genera were present. Points
+represented as only half on the graph have an OR of infinity in the
+positive or negative direction.
 

### 2018-02-07 code/run_adn_random_forest_genus_stool.R
update top 5 after change to OR

@@ -396,7 +396,7 @@ all_roc_values <- sapply(names(stool_final_data),
 ############### Run the actual programs to get the data (CRC Specific Genera) ################
 ##############################################################################################
 
-rr_data <- read_csv("data/process/tables/adn_select_genus_RR_stool_composite.csv") %>% arrange(pvalue, rr)
+rr_data <- read_csv("data/process/tables/adn_select_genus_OR_stool_composite.csv") %>% arrange(pvalue, rr)
 
 top5_pos_RR <- as.data.frame(rr_data %>% filter(rr > 1) %>% slice(1:5) %>% select(measure))[, "measure"]
 top5_neg_RR <- as.data.frame(rr_data %>% filter(rr < 1) %>% slice(1:5) %>% select(measure))[, "measure"]

### 2018-02-07 code/run_random_forest_genus_stool.R
update to use OR

@@ -444,7 +444,7 @@ all_roc_values <- sapply(names(stool_final_data),
 ############### Run the actual programs to get the data (CRC Specific Genera) ################
 ##############################################################################################
 
-rr_data <- read_csv("data/process/tables/select_genus_RR_stool_composite.csv") %>% arrange(pvalue, rr)
+rr_data <- read_csv("data/process/tables/select_genus_OR_stool_composite.csv") %>% arrange(pvalue, rr)
 
 top5_pos_RR <- as.data.frame(rr_data %>% filter(rr > 1) %>% slice(1:5) %>% select(measure))[, "measure"]
 top5_neg_RR <- as.data.frame(rr_data %>% filter(rr < 1) %>% slice(1:5) %>% select(measure))[, "measure"]

### 2018-02-07 code/run_adn_random_forest_genus_tissue.R
update to include OR based select models

@@ -445,24 +445,24 @@ matched_rf_datasets <- sapply(c(tissue_sets),
 
 # Generate data for each test (study) set
 # Definitely overfit since the classification is 100%
-matched_stool_model <- randomForest(disease ~ ., data = matched_rf_datasets[["lu"]], 
-                                    mtry = round(sqrt(ncol(matched_rf_datasets[["lu"]]))), 
-                                    importance = TRUE)
+#matched_stool_model <- randomForest(disease ~ ., data = matched_rf_datasets[["lu"]], 
+#                                    mtry = round(sqrt(ncol(matched_rf_datasets[["lu"]]))), 
+#                                    importance = TRUE)
 
-matched_model_rocs <- roc(matched_matched_genera_list$lu$disease ~ 
-                            matched_stool_model$votes[, "polyp"])
+#matched_model_rocs <- roc(matched_matched_genera_list$lu$disease ~ 
+#                            matched_stool_model$votes[, "polyp"])
 
 # Get important OTUs to the model, are they relevant 
-matched_model_importance_table <- matched_stool_model$importance %>% as.data.frame() %>% 
-  mutate(genera = rownames(.)) %>% 
-  arrange(desc(abs(MeanDecreaseAccuracy)))
+#matched_model_importance_table <- matched_stool_model$importance %>% as.data.frame() %>% 
+#  mutate(genera = rownames(.)) %>% 
+#  arrange(desc(abs(MeanDecreaseAccuracy)))
 
 
 ##############################################################################################
 ############### Run the actual programs to get the data (CRC Specific Genera - unmatched) ####
 ##############################################################################################
 
-rr_data <- read_csv("data/process/tables/adn_select_genus_RR_tissue_composite.csv") %>% arrange(pvalue, rr)
+rr_data <- read_csv("data/process/tables/adn_select_genus_OR_tissue_composite.csv") %>% arrange(pvalue, rr)
 
 top5_pos_RR <- as.data.frame(rr_data %>% filter(rr > 1) %>% slice(1:5) %>% select(measure))[, "measure"]
 top5_neg_RR <- as.data.frame(rr_data %>% filter(rr < 1) %>% slice(1:5) %>% select(measure))[, "measure"]

### 2018-02-09 code/make_genus_rf_auc_against_study_graph.R
update code to include ALL genera

@@ -40,15 +40,15 @@ make_table <- function(studies, path_to_file, first_data_part_name,
 adn_stool_studies <- c("baxter", "brim", "hale", "zeller")
 
 adn_all_stool <- make_table(adn_stool_studies, "data/process/tables/", 
-                            "adn_genus_stool_RF_full_", "adn_genus_stool_RF_select_", 
+                            "adn_ALL_genus_stool_RF_full_", "adn_genus_stool_RF_select_", 
                             "_pvalue_summary.csv")
 
 # Load needed data (adenoma -- tissue)
 adn_tissue_studies <- c("flemer", "lu")
 
 adn_all_tissue <- make_table(adn_tissue_studies, "data/process/tables/", 
-                            "adn_genus_unmatched_tissue_RF_full_", 
-                            "adn_genus_unmatched_tissue_RF_select_", 
+                            "adn_ALL_genus_unmatched_tissue_RF_full_", 
+                            "adn_ALL_genus_unmatched_tissue_RF_select_", 
                             "_pvalue_summary.csv")
 
 
@@ -56,24 +56,24 @@ adn_all_tissue <- make_table(adn_tissue_studies, "data/process/tables/",
 crc_stool_studies <- c("ahn", "baxter", "flemer", "hale", "wang", "weir", "zeller")
 
 crc_all_stool <- make_table(crc_stool_studies, "data/process/tables/", 
-                             "genus_stool_RF_full_", 
-                             "genus_stool_RF_select_", 
+                             "ALL_genus_stool_RF_full_", 
+                             "ALL_genus_stool_RF_select_", 
                              "_pvalue_summary.csv")
 
 # Load needed data (carcinoma -- tissue)
 crc_matched_tissue_studies <- c("burns", "dejea", "geng")
 
 crc_all_matched_tissue <- make_table(crc_matched_tissue_studies, "data/process/tables/", 
-                            "genus_matched_tissue_RF_", 
-                            "genus_matched_tissue_RF_select_", 
+                            "ALL_genus_matched_tissue_RF_", 
+                            "ALL_genus_matched_tissue_RF_select_", 
                             "_pvalue_summary.csv")
 
 
 crc_unmatched_tissue_studies <- c("burns", "chen", "flemer", "sana")
 
 crc_all_unmatched_tissue <- make_table(crc_unmatched_tissue_studies, "data/process/tables/", 
-                                     "genus_unmatched_tissue_RF_", 
-                                     "genus_unmatched_tissue_RF_select_", 
[…]

### 2018-02-09 code/make_tissue_imp_otu_graph.R
add all genera data to model occurences

@@ -9,12 +9,24 @@ loadLibs(c("tidyverse", "gridExtra", "viridis", "stringr"))
 
 
 # Read in Needed data
-crc_unmatched_genera_occurances <- read_csv("data/process/tables/crc_RF_genera_unmatched_tissue_top10.csv")
-crc_matched_genera_occurances <- read_csv("data/process/tables/adn_RF_genera_matched_tissue_top10.csv")
-adn_genera_occurances <- read_csv("data/process/tables/adn_RF_genera_tissue_top10.csv")
-crc_unmatched_otu_occurances <- read_csv("data/process/tables/crc_RF_otu_unmatched_tissue_top10.csv")
-crc_matched_otu_occurances <- read_csv("data/process/tables/adn_RF_otu_matched_tissue_top10.csv")
-adn_otu_occurances <- read_csv("data/process/tables/adn_RF_otu_tissue_top10.csv")
+crc_unmatched_genera_occurances <- read_csv("data/process/tables/crc_RF_genera_unmatched_tissue_top10.csv") %>% 
+  mutate(otu = str_replace_all(otu, "_unclassified", ""), 
+         otu = str_replace_all(otu, "\\.", "/"))
+crc_matched_genera_occurances <- read_csv("data/process/tables/adn_RF_genera_matched_tissue_top10.csv") %>% 
+  mutate(otu = str_replace_all(otu, "_unclassified", ""), 
+         otu = str_replace_all(otu, "\\.", "/"))
+adn_genera_occurances <- read_csv("data/process/tables/adn_RF_genera_tissue_top10.csv") %>% 
+  mutate(otu = str_replace_all(otu, "_unclassified", ""), 
+         otu = str_replace_all(otu, "\\.", "/"))
+crc_unmatched_otu_occurances <- read_csv("data/process/tables/crc_RF_otu_unmatched_tissue_top10.csv") %>% 
+  mutate(genus = str_replace_all(genus, "_unclassified", ""), 
+         genus = str_replace_all(genus, "\\.", "/"))
+crc_matched_otu_occurances <- read_csv("data/process/tables/adn_RF_otu_matched_tissue_top10.csv") %>% 
+  mutate(genus = str_replace_all(genus, "_unclassified", ""), 
+         genus = str_replace_all(genus, "\\.", "/"))
+adn_otu_occurances <- read_csv("data/process/tables/adn_RF_otu_tissue_top10.csv") %>% 
[…]

### 2018-02-09 code/make_tissue_alpha_RR_graph.R
update to use OR values

@@ -9,17 +9,17 @@ source("code/functions.R")
 loadLibs(c("tidyverse", "gridExtra", "viridis"))
 
 # Load in needed data tables
-adn_all_tissue <- read_csv("data/process/tables/alpha_adn_RR_tissue_composite.csv") %>% 
+adn_all_tissue <- read_csv("data/process/tables/alpha_adn_OR_tissue_composite.csv") %>% 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% 
-  bind_rows(read_csv("data/process/tables/alpha_adn_RR_ind_tissue_results.csv")) %>% 
+  bind_rows(read_csv("data/process/tables/alpha_adn_OR_ind_tissue_results.csv")) %>% 
   mutate(region = c(rep("combined", 3), rep("V3-V4", 6)))
 
 
-crc_all_tissue <- read_csv("data/process/tables/alpha_RR_tissue_composite.csv") %>% 
+crc_all_tissue <- read_csv("data/process/tables/alpha_OR_tissue_composite.csv") %>% 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% 
-  bind_rows(read_csv("data/process/tables/alpha_RR_ind_tissue_results.csv")) %>% 
+  bind_rows(read_csv("data/process/tables/alpha_OR_ind_tissue_results.csv")) %>% 
   mutate(region = c(rep("combined", 3), rep("V3-V5", 3), rep("V1-V2", 6), rep("V5-V6", 3), 
                     rep("V3-V4", 3), rep("V1-V3", 3)))
 
@@ -57,14 +57,14 @@ adn_tissue_graph <- adn_all_tissue %>%
                           levels = c("sobs", "shannoneven", "shannon"), 
                           labels = c("Observed OTUs", "Evenness", "Shannon Diversity")))  %>% 
   ggplot(aes(log2(est), study, xmax=log2(upper), xmin=log2(lower), colour=region)) + 
-  coord_cartesian(xlim=c(-4.2, 4.2)) + 
+  coord_cartesian(xlim=c(-6.2, 6.2)) + 
   geom_vline(xintercept = 0.0, linetype=2, alpha=0.75) + 
   geom_errorbarh(alpha=0.5, size = 1, height=0, show.legend = T) + 
   geom_point(size = 3, show.legend = T) + 
   facet_grid(. ~ measure) + 
[…]

### 2018-02-09 code/make_tissue_alpha_RR_graph.R
update x-axis

@@ -62,7 +62,7 @@ adn_tissue_graph <- adn_all_tissue %>%
   geom_errorbarh(alpha=0.5, size = 1, height=0, show.legend = T) + 
   geom_point(size = 3, show.legend = T) + 
   facet_grid(. ~ measure) + 
-  labs(x = expression(Log["2"]~Relative~Risk), y = "") + theme_bw() + ggtitle("A") + 
+  labs(x = expression(Log["2"]~Odds~Ratio), y = "") + theme_bw() + ggtitle("A") + 
   scale_color_manual(name = "Variable Region", values = c('#000000', '#35B779FF')) + 
   annotate("text", label = paste("Adenoma\n(Tissue)"), x = -4.70, y = 3.3, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
@@ -87,7 +87,7 @@ crc_tissue_graph <- crc_all_tissue %>%
   geom_errorbarh(alpha=0.5, size = 1, height=0, show.legend = T) + 
   geom_point(size = 3, show.legend = T) + 
   facet_grid(. ~ measure) + 
-  labs(x = expression(Log["2"]~Relative~Risk), y = "") + theme_bw() + ggtitle("B") + 
+  labs(x = expression(Log["2"]~Odds~Ratio), y = "") + theme_bw() + ggtitle("B") + 
   scale_color_manual(name = "Variable Region", values = c('#000000', '#8FD744FF', '#31688EFF', 
                                 '#35B779FF', '#443A83FF', '#FDE725FF')) + 
   annotate("text", label = paste("Carcinoma\n(Tissue)"), x = -4.4, y = 7, size = 2.5) + 

### 2018-02-09 submission/manuscript.Rmd
update with ALL genera model and OR

@@
 Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered OR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in OR using OTU richness and Shannon diversity.
-To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR.
-Mouth-associated genera were commonly found in the top 5 genera associated with an increased OR of having an adenoma (*Porphyromonas* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and colon tissue samples.
-Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased OR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3].
+To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in OR.
+Mouth-associated genera were commonly found in the top 5 genera associated with an increased OR of having an adenoma (*Porphyromonas* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* and *Parvimonas* [Figure 3D]) for both stool and colon tissue samples.
+Conversely, genera commonly associated with the gastrointestinal tract were correlated with a decreased OR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3].
 Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased OR genera between both stages and sample site.
 
-When observing ORs with a P-value less than 0.05 there was almost no overlap between genera from stool or colon tissue and when they were similar the RR was in opposite directions (e.g.
+When observing ORs for adenoma between genera from stool or colon tissue with a P-value less than 0.05 there was almost no overlap and when they were similar the OR was in opposite directions (e.g.
 *Lactococcus*) [Table S8 & S9].
-Many of the genera that had ORs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Pseudomonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
+Many of the adenoma associated genera ORs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Pseudomonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
 For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*) had an increased OR for both colon tissue and stool samples [Table S10 & S11].
-Finally, these data suggest that the most significantly increased OR genera for tissue was *Lepttrichia* while in stool it was *Peptostreptococcus* [Table S10 & S11].
+The genera with the highest increased OR for carcinoma in tissue was *Lepttrichia* while in stool it was *Peptostreptococcus* [Table S10 & S11].
 
 ***Select Community Models can Recapitulate Whole Community Models:***
-Since specific genera increased OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera.
+Since specific genera increased the OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera.
 We selected these genera based on their OR and P-value significance and used two approaches to test this question.
@@
 
-For the first approach using the genera-based models, the training set median AUC for model classification was similar for both the select genera and full genera models, for both tissue and stool studies [Figure S2-S3].
+For the first approach using the genera-based models, the training set median AUC for model classification was similar for both the full and select genera models, for both tissue and stool studies [Figure S2-S3].
 When analyzing the tests sets that were comprised of genera data from other studies, both models had a similar ability to detect individuals with adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6].
-Conversely, the second approach that used OTU-based models showed a slight decrease in median AUC between the full and select models, with one exception to this generalization being the carcinoma models for matched colon tissue [Figure 4 & 5].
+Conversely, the second approach that used OTU-based models showed a slight decrease in median AUC between the full and select models [Figure 4 & 5].
 
@@
 
-**Figure S1: Relative Risk for Adenoma or Carcinoma based on Bacterial Community Metrics in Tissue.** A) Community-based relative risk for adenoma.
-B) Community-based relative risk for carcinoma.
+**Figure S1: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Tissue.** A) Community-based odds ratio for adenoma.
+B) Community-based odds ratio for carcinoma.
 Colors represent the different variable regions used within the respective study.

### 2018-02-09 submission/manuscript.tex
update with ALL genera model and OR

@@
 abundance higher than the median results in an increase or decrease in
-RR. Mouth-associated genera were commonly found in the top 5 genera
+OR. Mouth-associated genera were commonly found in the top 5 genera
 associated with an increased OR of having an adenoma
@@
 \emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
-\emph{Fusobacterium} {[}Figure 3D{]}) for both stool and colon tissue
-samples. Conversely, genera commonly associated with a normal
-gastrointestinal tract were correlated with a decreased OR for both
-adenoma and carcinoma for both stool and colon tissue samples {[}Figure
-3{]}. Even though mouth-associated genera were identified across disease
-stage, there was little direct overlap of the top 5 increased or
-decreased OR genera between both stages and sample site.
-
-When observing ORs with a P-value less than 0.05 there was almost no
-overlap between genera from stool or colon tissue and when they were
-similar the RR was in opposite directions (e.g. \emph{Lactococcus})
-{[}Table S8 \& S9{]}. Many of the genera that had ORs with a P-value
-under 0.05 for colon tissue are also highly prevalent in contamination,
-specifically, \emph{Novosphingobium}, \emph{Pseudomonas}, and
-\emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma stage
-of disease, certain mouth-associated genera (\emph{Fusobacterium},
+\emph{Fusobacterium} and \emph{Parvimonas} {[}Figure 3D{]}) for both
+stool and colon tissue samples. Conversely, genera commonly associated
+with the gastrointestinal tract were correlated with a decreased OR for
+both adenoma and carcinoma for both stool and colon tissue samples
+{[}Figure 3{]}. Even though mouth-associated genera were identified
+across disease stage, there was little direct overlap of the top 5
+increased or decreased OR genera between both stages and sample site.
+
+When observing ORs for adenoma between genera from stool or colon tissue
+with a P-value less than 0.05 there was almost no overlap and when they
+were similar the OR was in opposite directions (e.g. \emph{Lactococcus})
+{[}Table S8 \& S9{]}. Many of the adenoma associated genera ORs with a
+P-value under 0.05 for colon tissue are also highly prevalent in
+contamination, specifically, \emph{Novosphingobium}, \emph{Pseudomonas},
+and \emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma
+stage of disease, certain mouth-associated genera (\emph{Fusobacterium},
 \emph{Parvimonas}) had an increased OR for both colon tissue and stool
-samples {[}Table S10 \& S11{]}. Finally, these data suggest that the
-most significantly increased OR genera for tissue was \emph{Lepttrichia}
-while in stool it was \emph{Peptostreptococcus} {[}Table S10 \& S11{]}.
+samples {[}Table S10 \& S11{]}. The genera with the highest increased OR
+for carcinoma in tissue was \emph{Lepttrichia} while in stool it was
+\emph{Peptostreptococcus} {[}Table S10 \& S11{]}.
 
 \textbf{\emph{Select Community Models can Recapitulate Whole Community
-Models:}} Since specific genera increased OR for carcinoma over
+Models:}} Since specific genera increased the OR for carcinoma over
 diversity metrics we assessed whether the bacterial community was better
@@
 For the first approach using the genera-based models, the training set
-median AUC for model classification was similar for both the select
-genera and full genera models, for both tissue and stool studies
-{[}Figure S2-S3{]}. When analyzing the tests sets that were comprised of
-genera data from other studies, both models had a similar ability to
-detect individuals with adenomas or carcinomas, with the select genera
-models performing better in some instances {[}Figure S4-S6{]}.
-Conversely, the second approach that used OTU-based models showed a
-slight decrease in median AUC between the full and select models, with
-one exception to this generalization being the carcinoma models for
-matched colon tissue {[}Figure 4 \& 5{]}.
+median AUC for model classification was similar for both the full and
+select genera models, for both tissue and stool studies {[}Figure
+S2-S3{]}. When analyzing the tests sets that were comprised of genera
+data from other studies, both models had a similar ability to detect
+individuals with adenomas or carcinomas, with the select genera models
+performing better in some instances {[}Figure S4-S6{]}. Conversely, the
+second approach that used OTU-based models showed a slight decrease in
+median AUC between the full and select models {[}Figure 4 \& 5{]}.
 
@@
 
-\textbf{Figure S1: Relative Risk for Adenoma or Carcinoma based on
-Bacterial Community Metrics in Tissue.} A) Community-based relative risk
-for adenoma. B) Community-based relative risk for carcinoma. Colors
+\textbf{Figure S1: Odds Ratio for Adenoma or Carcinoma based on
+Bacterial Community Metrics in Tissue.} A) Community-based odds ratio
+for adenoma. B) Community-based odds ratio for carcinoma. Colors
 represent the different variable regions used within the respective

### 2018-02-09 submission/manuscript.Rmd
update abstract

@@
 
-**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness, after correcting for study effect and variable region sequenced, with more severe disease (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
-This reduction in evenness translated into small increases in the odds ratio for adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma stages of CRC (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for developing carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
-Increases in mouth-associated microbes were commonly in the top 5 most significantly increased relative risk of adenoma and carcinoma for both stool and tissue samples.
-A prediction model for adenoma and carcinoma was built using either the whole community or selected genera with highest and lowest relative risk from fecal and tissue samples.
-Both approaches resulted in similar classification success according to Area Under the Curve (AUC) regardless of whether genera or OTUs were used to build the model.
+**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness as tumor severity increased, after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
+This reduction in evenness translated into small increases in the odds ratio for individuals to have both adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma tumors (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for individuals to have carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
+Increases in mouth-associated microbes were commonly in the top 5 most significantly increased odds ratios for individuals to have either adenoma or carcinoma tumors regardless of sample type.
+Prediction models built to classify either individuals with adenoma or carcinoma were trained on the whole community or selected genera (top 5 highest and lowest odds ratios) from either fecal or tissue samples.
+Both the full and select models for either adenoma or carcinoma resulted in similar classification success according to Area Under the Curve (AUC).
+Importantly, all models had a generalized decrease in AUC between train and test data sets.
 The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies.
-Although a number of associations between the microbiota and CRC were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.
+Although a number of associations between the microbiota and tumor were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.
 
 **Conclusions.** These data provide support for the importance of the bacterial community to both adenoma and carcinoma genesis.
-The evidence collected within this study on the role of the microbiota in CRC identifies a number of correlations that may not have been detected because of the low power associated with the majority of studies that have been performed to date.
+The evidence collected within this study on the role of the microbiota in those with tumors identifies a number of correlations that may not have been detected because of the low power associated with the majority of studies that have been performed to date.
 
@@
 
-microbiota; colorectal cancer; polyps; adenoma; meta-analysis.
+microbiota; colorectal cancer; polyps; adenoma; tumor; meta-analysis.
 

### 2018-02-09 submission/manuscript.tex
update abstract

@@
 \textbf{Results.} Fecal samples had a significant decrease for both
-Shannon diversity and evenness, after correcting for study effect and
-variable region sequenced, with more severe disease (P-value \textless{}
-0.05). This reduction in evenness translated into small increases in the
-odds ratio for adenoma (P-value = 0.035) and carcinoma stages of CRC
-(P-value = 0.0021) while the reduction in Shannon diversity only
-translated into an increased odds ratio for developing carcinomas
-(P-value = 0.0069). Increases in mouth-associated microbes were commonly
-in the top 5 most significantly increased relative risk of adenoma and
-carcinoma for both stool and tissue samples. A prediction model for
-adenoma and carcinoma was built using either the whole community or
-selected genera with highest and lowest relative risk from fecal and
-tissue samples. Both approaches resulted in similar classification
-success according to Area Under the Curve (AUC) regardless of whether
-genera or OTUs were used to build the model. The most important groups
-within the full community models consistently belonged to genera such as
-\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
-studies. Although a number of associations between the microbiota and
-CRC were identified, the majority of studies that we used in this
-meta-analysis were only individually adequately powered for large effect
-sizes.
+Shannon diversity and evenness as tumor severity increased, after
+correcting for study effect and variable region sequenced (P-value
+\textless{} 0.05). This reduction in evenness translated into small
+increases in the odds ratio for individuals to have both adenoma
+(P-value = 0.035) and carcinoma tumors (P-value = 0.0021) while the
+reduction in Shannon diversity only translated into an increased odds
+ratio for individuals to have carcinomas (P-value = 0.0069). Increases
+in mouth-associated microbes were commonly in the top 5 most
+significantly increased odds ratios for individuals to have either
+adenoma or carcinoma tumors regardless of sample type. Prediction models
+built to classify either individuals with adenoma or carcinoma were
+trained on the whole community or selected genera (top 5 highest and
+lowest odds ratios) from either fecal or tissue samples. Both the full
+and select models for either adenoma or carcinoma resulted in similar
+classification success according to Area Under the Curve (AUC).
+Importantly, all models had a generalized decrease in AUC between train
+and test data sets. The most important groups within the full community
+models consistently belonged to genera such as \emph{Ruminococcus},
+\emph{Bacteroides}, and \emph{Roseburia} across studies. Although a
+number of associations between the microbiota and tumor were identified,
+the majority of studies that we used in this meta-analysis were only
+individually adequately powered for large effect sizes.
 
@@
 evidence collected within this study on the role of the microbiota in
-CRC identifies a number of correlations that may not have been detected
-because of the low power associated with the majority of studies that
-have been performed to date.
+those with tumors identifies a number of correlations that may not have
+been detected because of the low power associated with the majority of
+studies that have been performed to date.
 

### 2018-02-09 submission/manuscript.Rmd
fix study analysis overview

@@
 
-**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness as tumor severity increased, after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
+**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness as tumor severity increased after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
 This reduction in evenness translated into small increases in the odds ratio for individuals to have both adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma tumors (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for individuals to have carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
-Increases in mouth-associated microbes were commonly in the top 5 most significantly increased odds ratios for individuals to have either adenoma or carcinoma tumors regardless of sample type.
+Increases in mouth-associated microbes were commonly in the top 5 most increased odds ratios for individuals to have either adenoma or carcinoma tumors regardless of sample type.
 Prediction models built to classify either individuals with adenoma or carcinoma were trained on the whole community or selected genera (top 5 highest and lowest odds ratios) from either fecal or tissue samples.
 Both the full and select models for either adenoma or carcinoma resulted in similar classification success according to Area Under the Curve (AUC).
-Importantly, all models had a generalized decrease in AUC between train and test data sets.
 The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies.
@@
 ***Study Analysis Overview:***
-OTU richness, evenness, and Shannon diversity was first assessed for differences between controls, adenoma stage, and carcinoma stage using both linear mixed-effect models and RR.
-The Bray-Curtis index was used to assess, for each individual study, differences between control-adenoma and control-carcinoma.
-Next, all common genera were assessed for differences in RR for having an adenoma or carcinoma and ranked based on P-value.
-We then built Random Forest models based on full or  selected community, based on the top 5 increased and top 5 decreased RR based on P-value, and these models were trained on one study then tested on the remaining studies.
+OTU richness, evenness, and Shannon diversity was first assessed for differences between controls, adenoma stage, and carcinoma stage using both linear mixed-effect models and OR.
+For each individual study the Bray-Curtis index was used to assess differences between control-adenoma and control-carcinoma.
+Next, all common genera were assessed for differences in OR for having an adenoma or carcinoma and ranked based on P-value.
+We then built Random Forest models based on the full or selected community (the top 5 increased and top 5 decreased OR based on P-value).
+Comparison between the full and selected models took two different approaches.
+In the first approach, models were trained on one study then tested on the remaining studies.
 This process was repeated for every study in the meta-analysis.
-A similar approach was then applied at the OTU level with the exception that a 10-fold CV over 100 different models, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
-For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased RR based on P-value.
+In the second approach models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
+For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased OR based on P-value.
 Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power.
-For comparisons in which only control versus adenoma stage were made, the carcinoma samples were excluded from each respective study.
-Similarly, for comparisons in which control versus carcinoma stage were made the adenoma samples were excluded from each respective study.
-The data were split between feces and tissue samples.
+For comparisons in which only control versus adenoma indivdiuals were made, the carcinoma individuals were excluded from each respective study.
+Similarly, for comparisons in which control versus carcinoma individuals were made the adenoma individuals were excluded from each respective study.
+For all analysis completed feces and tissue samples were kept seperate.
 Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched).
-Where applicable for each study, predictions for adenoma and carcinoma stage of disease were then tested for feces, matched tissue, and unmatched tissue.
 

### 2018-02-09 submission/manuscript.tex
fix study analysis overview

@@
 \textbf{Results.} Fecal samples had a significant decrease for both
-Shannon diversity and evenness as tumor severity increased, after
+Shannon diversity and evenness as tumor severity increased after
 correcting for study effect and variable region sequenced (P-value
@@
 ratio for individuals to have carcinomas (P-value = 0.0069). Increases
-in mouth-associated microbes were commonly in the top 5 most
-significantly increased odds ratios for individuals to have either
-adenoma or carcinoma tumors regardless of sample type. Prediction models
-built to classify either individuals with adenoma or carcinoma were
-trained on the whole community or selected genera (top 5 highest and
-lowest odds ratios) from either fecal or tissue samples. Both the full
-and select models for either adenoma or carcinoma resulted in similar
-classification success according to Area Under the Curve (AUC).
-Importantly, all models had a generalized decrease in AUC between train
-and test data sets. The most important groups within the full community
-models consistently belonged to genera such as \emph{Ruminococcus},
-\emph{Bacteroides}, and \emph{Roseburia} across studies. Although a
-number of associations between the microbiota and tumor were identified,
-the majority of studies that we used in this meta-analysis were only
-individually adequately powered for large effect sizes.
+in mouth-associated microbes were commonly in the top 5 most increased
+odds ratios for individuals to have either adenoma or carcinoma tumors
+regardless of sample type. Prediction models built to classify either
+individuals with adenoma or carcinoma were trained on the whole
+community or selected genera (top 5 highest and lowest odds ratios) from
+either fecal or tissue samples. Both the full and select models for
+either adenoma or carcinoma resulted in similar classification success
+according to Area Under the Curve (AUC). The most important groups
+within the full community models consistently belonged to genera such as
+\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
+studies. Although a number of associations between the microbiota and
+tumor were identified, the majority of studies that we used in this
+meta-analysis were only individually adequately powered for large effect
+sizes.
 
@@
 adenoma stage, and carcinoma stage using both linear mixed-effect models
-and RR. The Bray-Curtis index was used to assess, for each individual
-study, differences between control-adenoma and control-carcinoma. Next,
-all common genera were assessed for differences in RR for having an
+and OR. For each individual study the Bray-Curtis index was used to
+assess differences between control-adenoma and control-carcinoma. Next,
+all common genera were assessed for differences in OR for having an
 adenoma or carcinoma and ranked based on P-value. We then built Random
-Forest models based on full or selected community, based on the top 5
-increased and top 5 decreased RR based on P-value, and these models were
-trained on one study then tested on the remaining studies. This process
-was repeated for every study in the meta-analysis. A similar approach
-was then applied at the OTU level with the exception that a 10-fold CV
-over 100 different models, based on random 80/20 splitting of the data,
-was used to generate a range of expected AUCs. For these OTU-based
-models, the selected model included all OTUs that had a taxonomic
-classification to a taxa in the top 5 increased and top 5 decreased RR
-based on P-value. Finally, the power of each study was assessed for an
-effect size ranging from 1\% to 30\% and an estimated sample size, for
-these effect sizes, was generated based on 80\% power. For comparisons
-in which only control versus adenoma stage were made, the carcinoma
-samples were excluded from each respective study. Similarly, for
-comparisons in which control versus carcinoma stage were made the
-adenoma samples were excluded from each respective study. The data were
-split between feces and tissue samples. Within the tissue groups the
-data were further divided between samples from the same individual
-(matched) and those from different individuals (unmatched). Where
-applicable for each study, predictions for adenoma and carcinoma stage
-of disease were then tested for feces, matched tissue, and unmatched
-tissue.
+Forest models based on the full or selected community (the top 5
+increased and top 5 decreased OR based on P-value). Comparison between
+the full and selected models took two different approaches. In the first
+approach, models were trained on one study then tested on the remaining
+studies. This process was repeated for every study in the meta-analysis.
+In the second approach models were built using OTU level data and a
+10-fold CV over 100 different iterations, based on random 80/20
+splitting of the data, was used to generate a range of expected AUCs.
+For these OTU-based models, the selected model included all OTUs that
+had a taxonomic classification to a taxa in the top 5 increased and top
+5 decreased OR based on P-value. Finally, the power of each study was
+assessed for an effect size ranging from 1\% to 30\% and an estimated
+sample size, for these effect sizes, was generated based on 80\% power.
+For comparisons in which only control versus adenoma indivdiuals were
+made, the carcinoma individuals were excluded from each respective
+study. Similarly, for comparisons in which control versus carcinoma
+individuals were made the adenoma individuals were excluded from each
+respective study. For all analysis completed feces and tissue samples
+were kept seperate. Within the tissue groups the data were further
+divided between samples from the same individual (matched) and those
+from different individuals (unmatched).
 

### 2018-02-09 submission/manuscript.Rmd
update methods

@@
 ***Obtaining Genera Relative Abundance and Selected Models:***
-For the genera analysis of the RR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings, were collected.
-The OTU based Random Forest Models using selected OTUs utilized a similar approach except that the OTUs were not aggregated together by taxonomic identity but kept as separate OTUs.
-OTU Random Forest models using the full community included all OTUs while those for the selected model included only those OTUs that had a taxonomic classification to a variable in the top 5 increased of top 5 decreased OR based on P-value.
+For the genera analysis of the OR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings.
+The select models for the first approach utilized the top 5 most significantly increased and decreased ORs.
+For this approach, the full community models ufor each study were built by utilizing all genera and lowest taxnomic groups identified within that particular study.
+The select models for the second approach, that used OTU-based Random Forest models, utilized a similar method as the first approach.
+The main difference was that any OTU that taxonomically classified to one of the genera in the top 5 increased or decreased OR were included in the select model.
+OTU Random Forest models using the full community included all OTUs.
 
@@
 ***Assessing Important Random Forest Model Variables:***
-Using Mean Decrease in Accuracy (MDA) the top 10 most important variables to the Random Forest model were obtained in two different ways depending on whether the model used genera or OTU data.
-For the genus based models, the number of times that a genus showed up in the top 10 of the training set across each study was counted while, for the OTU based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study.
+Using Mean Decrease in Accuracy (MDA) the top 10 most important variables to the Random Forest model were obtained for the full models of the two different approaches used.
+For the first approach utilizing genus-based models, the number of times that a genus showed up in the top 10 of the training set across each study was counted.
+For the second approach, that utilized the OTU-based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study.
 Common taxa, for the OTU based models, were identified by using the lowest classification within the RDP database for each of the specific OTUs obtained from the previous counts and the number of times this classification occurred in the top 10, in each study, was recorded.
-The two studies that had adenoma tissue were equally divided between matched and unmatched groups and were grouped together for the counting of the top 10 genera and OTUs.
+Finally, for the two studies that had adenoma tissue (Lu and Flemer) were equally divided between matched and unmatched studies and were grouped together for the counting of the top 10 genera and OTUs.
 

### 2018-02-09 submission/manuscript.tex
update methods

@@
 \textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
-For the genera analysis of the RR, OTUs were added together based on the
+For the genera analysis of the OR, OTUs were added together based on the
 genus or lowest available taxonomic classification level and the total
-average counts, for 100 different subsamplings, were collected. The OTU
-based Random Forest Models using selected OTUs utilized a similar
-approach except that the OTUs were not aggregated together by taxonomic
-identity but kept as separate OTUs. OTU Random Forest models using the
-full community included all OTUs while those for the selected model
-included only those OTUs that had a taxonomic classification to a
-variable in the top 5 increased of top 5 decreased OR based on P-value.
+average counts, for 100 different subsamplings. The select models for
+the first approach utilized the top 5 most significantly increased and
+decreased ORs. For this approach, the full community models ufor each
+study were built by utilizing all genera and lowest taxnomic groups
+identified within that particular study. The select models for the
+second approach, that used OTU-based Random Forest models, utilized a
+similar method as the first approach. The main difference was that any
+OTU that taxonomically classified to one of the genera in the top 5
+increased or decreased OR were included in the select model. OTU Random
+Forest models using the full community included all OTUs.
 
@@
 Mean Decrease in Accuracy (MDA) the top 10 most important variables to
-the Random Forest model were obtained in two different ways depending on
-whether the model used genera or OTU data. For the genus based models,
-the number of times that a genus showed up in the top 10 of the training
-set across each study was counted while, for the OTU based models, the
-medians for each OTU across 100 different 80/20 splits of the data was
-generated and the top 10 OTUs then counted for each study. Common taxa,
-for the OTU based models, were identified by using the lowest
-classification within the RDP database for each of the specific OTUs
-obtained from the previous counts and the number of times this
-classification occurred in the top 10, in each study, was recorded. The
-two studies that had adenoma tissue were equally divided between matched
-and unmatched groups and were grouped together for the counting of the
-top 10 genera and OTUs.
+the Random Forest model were obtained for the full models of the two
+different approaches used. For the first approach utilizing genus-based
+models, the number of times that a genus showed up in the top 10 of the
+training set across each study was counted. For the second approach,
+that utilized the OTU-based models, the medians for each OTU across 100
+different 80/20 splits of the data was generated and the top 10 OTUs
+then counted for each study. Common taxa, for the OTU based models, were
+identified by using the lowest classification within the RDP database
+for each of the specific OTUs obtained from the previous counts and the
+number of times this classification occurred in the top 10, in each
+study, was recorded. Finally, for the two studies that had adenoma
+tissue (Lu and Flemer) were equally divided between matched and
+unmatched studies and were grouped together for the counting of the top
+10 genera and OTUs.
 

### 2018-02-09 submission/manuscript.Rmd
fix conclusion

@@
 
-By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in CRC.
-Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are are commonly associated with the greatest OR of having a carcinoma.
+By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in colorectal tumors.
+Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are commonly associated with the greatest OR of individuals having a carcinoma.
 No conclusive signal with these mouth-associated microbes could be detected for adenoma stage of disease.
-Overall, associations between the microbiota and carcinoma stage of disease were much stronger than with the adenoma stage.
-Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to the subtle changes in the community.
+Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to possible subtle changes in the community.
+Overall, associations between the microbiota and individuals with carcinoma stage of disease were much stronger than with to those with adenoma stage of disease.
 

### 2018-02-09 submission/manuscript.tex
fix conclusion

@@
 fecal and colon tissue samples, we are able to provide evidence
-supporting the importance of the bacterial community in CRC. Further,
-the data presented here suggests that mouth-associated microbes can gain
-a foothold within the colon and are are commonly associated with the
-greatest OR of having a carcinoma. No conclusive signal with these
-mouth-associated microbes could be detected for adenoma stage of
-disease. Overall, associations between the microbiota and carcinoma
-stage of disease were much stronger than with the adenoma stage. Our
-observations also highlight the importance of power and sample number
-considerations when investigating the microbiota and adenoma stage of
-disease due to the subtle changes in the community.
+supporting the importance of the bacterial community in colorectal
+tumors. Further, the data presented here suggests that mouth-associated
+microbes can gain a foothold within the colon and are commonly
+associated with the greatest OR of individuals having a carcinoma. No
+conclusive signal with these mouth-associated microbes could be detected
+for adenoma stage of disease. Our observations also highlight the
+importance of power and sample number considerations when investigating
+the microbiota and adenoma stage of disease due to possible subtle
+changes in the community. Overall, associations between the microbiota
+and individuals with carcinoma stage of disease were much stronger than
+with to those with adenoma stage of disease.
 

### 2018-02-10 submission/manuscript.Rmd
update changes in intro

@@
 Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC.
-There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma.
+There have also been numerous case-control studies investigating the microbiota in the formation of both adenoma and carcinoma.
 A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
@@
 Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
-Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with CRC.
+Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with tumors.
 These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012].
@@
 A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017].
-However, the sample size (n = 509) is equal to or less than some more recent individual studies investigating the microbiota and colorectal tumors, making it hard to know how extrapolatable these findings are.
+However, the sample size (n = 509) is equal to or less than some of the more recent individual studies investigating the microbiota and colorectal tumors, making it hard to know how extrapolatable these findings are.
 That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of disease that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].
@@
 
-Most of these studies have focused on carcinoma but the adenoma observations are not any clearer at identifying candidate genera of disease.
-Groups focusing on broad scale community metrics have found that metrics such as richness are decreased in the adenoma stage of disease versus controls.
-Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in adenoma [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
-Based on the studies mentioned, there seems to be very little overlap between the genera identified to be associated with adenoma and carcinoma, with *Lactobacillus* being one of the few commonalities.
+Most of these studies have focused on individuals with carcinomas but associations with the adenoma stage of disease are not any clearer at identifying candidate genera correlated with these earlier tumors.
+Groups focusing on broad scale community metrics have found that measures such as richness are decreased in those with adenomas versus controls.
+Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in those with adenoma tumors [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
+Additionally, based on these studies mentioned, there seems to be very few common genera that are associated with both adenoma and carcinoma tumors, with *Lactobacillus* being one of the few commonalities.
 
-Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications.
+Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk of disease.
+The current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications.
 Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities.
 Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
-Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without limitations.
-The community present for sampling following the colonoscopy bowel prep may reflect the better adhered microbiota versus the resident community.
+Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without their own limitations.
+After the colonoscopy bowel prep the bacterial community sampled may reflect the better adhered microbiota versus the resident community.
 Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done.
@@
 In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of colorectal tumors.
-Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in CRC and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].
+Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in individuals with adenoma or carcinoma and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].
 
 Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of disease.
-To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased odds ratio (OR) for adenoma or carcinoma stages of disease We then assessed what genera, if any, increase or decrease the OR of adenoma or carcinoma stages of disease Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased OR genera resulted in better model classification, based on the area under the curve (AUC).
+To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased odds ratio (OR) for individuals to have either an adenoma or carcinoma.
+We then assessed what genera, if any, increase or decrease the OR of an individual to have an adenoma or carcinoma.
+Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased OR genera resulted in better model classification, based on the area under the curve (AUC).
 Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power.
-Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the studies are underpowered for assessing small effect sizes.
+Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the individual studies are underpowered for assessing small effect sizes.
 
@@
 Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power.
-For comparisons in which only control versus adenoma indivdiuals were made, the carcinoma individuals were excluded from each respective study.
+For comparisons in which only control versus adenoma individuals were made, the carcinoma individuals were excluded from each respective study.
 Similarly, for comparisons in which control versus carcinoma individuals were made the adenoma individuals were excluded from each respective study.
-For all analysis completed feces and tissue samples were kept seperate.
+For all analysis completed feces and tissue samples were kept separate.
 Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched).
@@
 The select models for the first approach utilized the top 5 most significantly increased and decreased ORs.
-For this approach, the full community models ufor each study were built by utilizing all genera and lowest taxnomic groups identified within that particular study.
+For this approach, the full community models for each study were built by utilizing all genera and lowest taxonomic groups identified within that particular study.
 The select models for the second approach, that used OTU-based Random Forest models, utilized a similar method as the first approach.

### 2018-02-10 submission/manuscript.tex
update changes in intro

@@
 {[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
-been numerous case/control studies investigating the microbiota in the
+been numerous case-control studies investigating the microbiota in the
 formation of both adenoma and carcinoma. A recent meta-analysis
@@
 these studies, mouth-associated genera are far from the only microbes
-identified to be associated with CRC. These other genera include, but
+identified to be associated with tumors. These other genera include, but
 are not limited to, \emph{Providencia}, \emph{Mogibacterium},
@@
 genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
-However, the sample size (n = 509) is equal to or less than some more
-recent individual studies investigating the microbiota and colorectal
-tumors, making it hard to know how extrapolatable these findings are.
-That particular meta-analysis also added more potential microbial
-associations to both carcinoma (\emph{Pantoea agglomerans}
+However, the sample size (n = 509) is equal to or less than some of the
+more recent individual studies investigating the microbiota and
+colorectal tumors, making it hard to know how extrapolatable these
+findings are. That particular meta-analysis also added more potential
+microbial associations to both carcinoma (\emph{Pantoea agglomerans}
 \emph{Ruminococcus}, \emph{Lactobacillus}) and adenoma
@@
 
-Most of these studies have focused on carcinoma but the adenoma
-observations are not any clearer at identifying candidate genera of
-disease. Groups focusing on broad scale community metrics have found
-that metrics such as richness are decreased in the adenoma stage of
-disease versus controls. Other studies have identified
-\emph{Lactococcus}, \emph{Pseudomonas}, \emph{Acidovorax},
-\emph{Cloacibacterium}, \emph{Helicobacter}, \emph{Lactobacillus},
-\emph{Bilophila}, \emph{Desulfovibrio}, and \emph{Mogibacterium} to be
-increased in adenoma {[}25--27{]}. Based on the studies mentioned, there
-seems to be very little overlap between the genera identified to be
-associated with adenoma and carcinoma, with \emph{Lactobacillus} being
+Most of these studies have focused on individuals with carcinomas but
+associations with the adenoma stage of disease are not any clearer at
+identifying candidate genera correlated with these earlier tumors.
+Groups focusing on broad scale community metrics have found that
+measures such as richness are decreased in those with adenomas versus
+controls. Other studies have identified \emph{Lactococcus},
+\emph{Pseudomonas}, \emph{Acidovorax}, \emph{Cloacibacterium},
+\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Bilophila},
+\emph{Desulfovibrio}, and \emph{Mogibacterium} to be increased in those
+with adenoma tumors {[}25--27{]}. Additionally, based on these studies
+mentioned, there seems to be very few common genera that are associated
+with both adenoma and carcinoma tumors, with \emph{Lactobacillus} being
 one of the few commonalities.
@@
 seems logical since it offers an easy and cost-effective way to stratify
-risk and the current gold standard for diagnosis, a colonoscopy, can be
-time-consuming and is not without risk of complications. Although stool
-represents an easy and less invasive way to assess risk, it is not clear
-how well this sample reflects adenoma- and carcinoma- associated
+risk of disease. The current gold standard for diagnosis, a colonoscopy,
+can be time-consuming and is not without risk of complications. Although
+stool represents an easy and less invasive way to assess risk, it is not
+clear how well this sample reflects adenoma- and carcinoma- associated
 microbial communities. Some studies have tried to assess this in health
@@
 the microbiota directly associated with colon tissue may provide clearer
-answers but is not without limitations. The community present for
-sampling following the colonoscopy bowel prep may reflect the better
+answers but is not without their own limitations. After the colonoscopy
+bowel prep the bacterial community sampled may reflect the better
 adhered microbiota versus the resident community. Additionally, these
@@
 could provide valuable insights into the common genera that are both
-protective and detrimental in CRC and whether broad bacterial community
-measurements can account for these changes that were not provided by
-earlier meta-analysis studies {[}11{]}.
+protective and detrimental in individuals with adenoma or carcinoma and
+whether broad bacterial community measurements can account for these
+changes that were not provided by earlier meta-analysis studies
+{[}11{]}.
 
@@
 (control to adenoma to carcinoma) and if it results in an increased odds
-ratio (OR) for adenoma or carcinoma stages of disease We then assessed
-what genera, if any, increase or decrease the OR of adenoma or carcinoma
-stages of disease Next, using Random Forest models, we analyzed whether
-the full community or only the combined top 5 increased and top 5
-decreased OR genera resulted in better model classification, based on
-the area under the curve (AUC). Finally, we also examined at what effect
-and sample size the studies used were powered for and the sample size
-needed to get to the traditionally accepted 80\% power. Our results from
-these analyses suggests that the bacterial community changes as disease
-severity worsens, that specific members are important for disease
-classification, and that many of the studies are underpowered for
-assessing small effect sizes.
+ratio (OR) for individuals to have either an adenoma or carcinoma. We
+then assessed what genera, if any, increase or decrease the OR of an
+individual to have an adenoma or carcinoma. Next, using Random Forest
+models, we analyzed whether the full community or only the combined top
+5 increased and top 5 decreased OR genera resulted in better model
+classification, based on the area under the curve (AUC). Finally, we
+also examined at what effect and sample size the studies used were
+powered for and the sample size needed to get to the traditionally
+accepted 80\% power. Our results from these analyses suggests that the
+bacterial community changes as disease severity worsens, that specific
+members are important for disease classification, and that many of the
+individual studies are underpowered for assessing small effect sizes.
 
@@
 sample size, for these effect sizes, was generated based on 80\% power.
-For comparisons in which only control versus adenoma indivdiuals were
+For comparisons in which only control versus adenoma individuals were
 made, the carcinoma individuals were excluded from each respective
@@
 respective study. For all analysis completed feces and tissue samples
-were kept seperate. Within the tissue groups the data were further
+were kept separate. Within the tissue groups the data were further
 divided between samples from the same individual (matched) and those
@@
 the first approach utilized the top 5 most significantly increased and
-decreased ORs. For this approach, the full community models ufor each
-study were built by utilizing all genera and lowest taxnomic groups
+decreased ORs. For this approach, the full community models for each
+study were built by utilizing all genera and lowest taxonomic groups
 identified within that particular study. The select models for the

### 2018-02-10 submission/manuscript.Rmd
complete more abstract changes

@@
 Important drivers within this context have ranged from individual microbes to the whole community.
-Our study expands on a recent meta-analysis investigating microbial biomarkers for tumors by testing the hypothesis that the bacterial community has important associations to both early (adenoma) and late (carcinoma) stage disease.
-To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 previously published 16S rRNA gene sequencing studies on CRC and the microbiota.
+Our study expands on a recent meta-analysis investigating microbial biomarkers for tumors by testing the hypothesis that the bacterial community has important associations to both adenoma and carcinoma tumors.
+To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 previously published 16S rRNA gene sequencing studies on colorectal tumors and the microbiota.
 
-**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness as tumor severity increased after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
+**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness as tumor severity increased, after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
 This reduction in evenness translated into small increases in the odds ratio for individuals to have both adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma tumors (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for individuals to have carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
-Increases in mouth-associated microbes were commonly in the top 5 most increased odds ratios for individuals to have either adenoma or carcinoma tumors regardless of sample type.
+Increases in mouth-associated microbes were commonly in the top 5 most increased odds ratios for individuals to have either adenoma or carcinoma tumors, regardless of sample type.
 Prediction models built to classify either individuals with adenoma or carcinoma were trained on the whole community or selected genera (top 5 highest and lowest odds ratios) from either fecal or tissue samples.
@@
 
-**Conclusions.** These data provide support for the importance of the bacterial community to both adenoma and carcinoma genesis.
+**Conclusions.** These data provide support for the importance of the bacterial community to both adenoma and carcinoma tumorgenesis.
 The evidence collected within this study on the role of the microbiota in those with tumors identifies a number of correlations that may not have been detected because of the low power associated with the majority of studies that have been performed to date.

### 2018-02-10 submission/manuscript.tex
complete more abstract changes

@@
 testing the hypothesis that the bacterial community has important
-associations to both early (adenoma) and late (carcinoma) stage disease.
-To test this hypothesis we examined both feces (n = 1737) and colon
-tissue (492 total samples from 350 individuals) across 14 previously
-published 16S rRNA gene sequencing studies on CRC and the microbiota.
+associations to both adenoma and carcinoma tumors. To test this
+hypothesis we examined both feces (n = 1737) and colon tissue (492 total
+samples from 350 individuals) across 14 previously published 16S rRNA
+gene sequencing studies on colorectal tumors and the microbiota.
 
 \textbf{Results.} Fecal samples had a significant decrease for both
-Shannon diversity and evenness as tumor severity increased after
+Shannon diversity and evenness as tumor severity increased, after
 correcting for study effect and variable region sequenced (P-value
@@
 in mouth-associated microbes were commonly in the top 5 most increased
-odds ratios for individuals to have either adenoma or carcinoma tumors
+odds ratios for individuals to have either adenoma or carcinoma tumors,
 regardless of sample type. Prediction models built to classify either
@@
 \textbf{Conclusions.} These data provide support for the importance of
-the bacterial community to both adenoma and carcinoma genesis. The
+the bacterial community to both adenoma and carcinoma tumorgenesis. The
 evidence collected within this study on the role of the microbiota in

### 2018-02-10 submission/manuscript.Rmd
update and improve discussion

@@
 
-Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with and without CRC [Figure 1-3].
-Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect sizes for the OR of tumors.
-Even though mouth-associated genera increased the OR of having a carcinoma, they did not consistently increase the RR of having an adenoma.
-Additionally, our observations suggest that by combining mouth-associated and CRC protective microbes we can classify either adenoma or carcinoma stage of disease as well as models that use the full community.
+Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with tumors versus controls [Figure 1-3].
+Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect size increases in OR for either adenoma or carcinoma tumors.
+Even though mouth-associated genera increased individaul's OR of having a carcinoma, they did not consistently increase the OR of having an adenoma.
+Additionally, our observations suggest that by combining the top 5 most protective and detrimental microbes we can classify indviduals with either adenoma or carcinoma as well as models that use the full community.
 
-The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, formation.
-The results that we have presented show that both the genera and OTU select and full models, for the carcinoma stage of CRC, had similar AUCs [Figure 4 & 5].
-This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community is crucial for carcinoma formation.
-Importantly, it suggests that there may be key members of the GI community that might be studied further to potentially reduce the risk of carcinoma.
-Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma stage of disease.
+The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, tumor formation.
+The results that we have presented show that both the genera- and OTU-based select and full models, for indviduals with carcinoma, had similar AUCs [Figure 4 & 5].
+This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community could be crucial for carcinoma formation.
+Importantly, it suggests that there may be key members of the GI community that should be studied further to potentially help reduce the risk of carcinoma tumor formation.
+Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma turmors.
 Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly.
@@
 This change in diversity, at this early stage of disease, could be focal to the adenoma itself.
-One possible explanation is that how the host interacts with these subtle changes at early stages of the disease is what leads to a thoroughly dysfunctional community that is supportive of CRC genesis.
+Additionally, how the host interacts with these subtle changes at early stages of the disease could be what leads to a thoroughly dysfunctional community that is supportive of tumorgenesis.
 
 Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 6].
-Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both adenoma and carcinoma [Figure 6].
+Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 6].
 These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community.
@@
 As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies.
-This suggests that even though *Bacteroides* may not increase the RR of CRC and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.
+This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.
 Additionally, *Streptococcus gallolyticus* subsp.
-*gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased RR for both adenoma and carcinoma stage of disease.
+*gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased OR for individuals to have either an adenoma or carcinoma.
 
-The associations between the microbiota and adenoma stage of disease are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes.
+The associations between the microbiota and indivdials with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes.
 None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls.
-The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and adenoma stage of disease.
-Future studies investigating adenoma stage and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation.
+The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas.
+Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation.
 In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 7].

### 2018-02-10 submission/manuscript.tex
update and improve discussion

@@
 Our study identifies clear differences in diversity, both at the
-community level and for individual genera, present in patients with and
-without CRC {[}Figure 1-3{]}. Although there was a step-wise decrease in
-diversity as disease progressed from control to adenoma to carcinoma,
-this did not translate into large effect sizes for the OR of tumors.
-Even though mouth-associated genera increased the OR of having a
-carcinoma, they did not consistently increase the RR of having an
-adenoma. Additionally, our observations suggest that by combining
-mouth-associated and CRC protective microbes we can classify either
-adenoma or carcinoma stage of disease as well as models that use the
-full community.
+community level and for individual genera, present in patients with
+tumors versus controls {[}Figure 1-3{]}. Although there was a step-wise
+decrease in diversity as disease progressed from control to adenoma to
+carcinoma, this did not translate into large effect size increases in OR
+for either adenoma or carcinoma tumors. Even though mouth-associated
+genera increased individaul's OR of having a carcinoma, they did not
+consistently increase the OR of having an adenoma. Additionally, our
+observations suggest that by combining the top 5 most protective and
+detrimental microbes we can classify indviduals with either adenoma or
+carcinoma as well as models that use the full community.
 
 The data presented herein support the importance of select genera for
-carcinoma, but not necessarily adenoma, formation. The results that we
-have presented show that both the genera and OTU select and full models,
-for the carcinoma stage of CRC, had similar AUCs {[}Figure 4 \& 5{]}.
-This suggests that an interplay between a select number of potentially
-protective and exacerbating microbes within the GI community is crucial
-for carcinoma formation. Importantly, it suggests that there may be key
-members of the GI community that might be studied further to potentially
-reduce the risk of carcinoma. Conversely, using the present data, it is
-clear that new approaches may be needed to identify members of the
-community associated with adenoma stage of disease. Regardless of sample
-type and whether a full or select model was used, our Random Forest
-models consistently performed poorly. Yet, the step-wise decrease in
-diversity suggests that the adenoma-associated community is not normal
-but has changed subtly {[}Figure 1{]}. This change in diversity, at this
-early stage of disease, could be focal to the adenoma itself. One
-possible explanation is that how the host interacts with these subtle
-changes at early stages of the disease is what leads to a thoroughly
-dysfunctional community that is supportive of CRC genesis.
+carcinoma, but not necessarily adenoma, tumor formation. The results
+that we have presented show that both the genera- and OTU-based select
+and full models, for indviduals with carcinoma, had similar AUCs
+{[}Figure 4 \& 5{]}. This suggests that an interplay between a select
+number of potentially protective and exacerbating microbes within the GI
+community could be crucial for carcinoma formation. Importantly, it
+suggests that there may be key members of the GI community that should
+be studied further to potentially help reduce the risk of carcinoma
+tumor formation. Conversely, using the present data, it is clear that
+new approaches may be needed to identify members of the community
+associated with adenoma turmors. Regardless of sample type and whether a
+full or select model was used, our Random Forest models consistently
+performed poorly. Yet, the step-wise decrease in diversity suggests that
+the adenoma-associated community is not normal but has changed subtly
+{[}Figure 1{]}. This change in diversity, at this early stage of
+disease, could be focal to the adenoma itself. Additionally, how the
+host interacts with these subtle changes at early stages of the disease
+could be what leads to a thoroughly dysfunctional community that is
+supportive of tumorgenesis.
 
@@
 \emph{Roseburia} were consistently found to be in the top 10 most
-important variables across the different studies for both adenoma and
-carcinoma {[}Figure 6{]}. These data suggest that whether the
-non-resident bacterium is \emph{Fusobacteria} or
+important variables across the different studies for both individuals
+with adenoma and carcinoma {[}Figure 6{]}. These data suggest that
+whether the non-resident bacterium is \emph{Fusobacteria} or
 \emph{Peptostreptococcus} may not be as important as how these bacteria
@@
 This suggests that even though \emph{Bacteroides} may not increase the
-RR of CRC and may not vary in relative abundance, like
-\emph{Fusobacterium}, it is still important in CRC. Additionally,
-\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
-mouth-associated microbe, and the results from this study suggest that
-regardless of sample type, mouth-associated genera are commonly
-associated with an increased RR for both adenoma and carcinoma stage of
-disease.
-
-The associations between the microbiota and adenoma stage of disease are
-inconclusive, in part, because many studies may not be powered
-effectively to observe small effect sizes. None of the studies analyzed
-were properly powered to detect a 10\% or lower change between cases and
-controls. The results within our meta-analysis suggest that a small
-effect size may well be the scope in which differences consistently
-occur between controls and adenoma stage of disease. Future studies
-investigating adenoma stage and the microbiota need to take power into
-consideration to reproducibly study whether the microbiota contributes
-to polyp formation. In contrast to adenoma stage of disease, our
-observations suggest that most studies analyzed have sufficient power to
-detect many changes in the carcinoma-associated microbiota because of
-large effect size differences between cases and controls {[}Figure 7{]}.
+OR of individuals having an adenoma or carcinoma and may not vary in
+relative abundance, like \emph{Fusobacterium}, it is still important in
+CRC. Additionally, \emph{Streptococcus gallolyticus} subsp.
+\emph{gallolyticus} is a mouth-associated microbe, and the results from
+this study suggest that regardless of sample type, mouth-associated
+genera are commonly associated with an increased OR for individuals to
+have either an adenoma or carcinoma.
+
+The associations between the microbiota and indivdials with adenoma
+tumors are inconclusive, in part, because many studies may not be
+powered effectively to observe small effect sizes. None of the studies
+analyzed were properly powered to detect a 10\% or lower change between
+cases and controls. The results within our meta-analysis suggest that a
+small effect size may well be the scope in which differences
+consistently occur between controls and those with adenomas. Future
+studies investigating adenoma tumors and the microbiota need to take
+power into consideration to reproducibly study whether the microbiota
+contributes to polyp formation. In contrast to adenoma stage of disease,
+our observations suggest that most studies analyzed have sufficient
+power to detect many changes in the carcinoma-associated microbiota
+because of large effect size differences between cases and controls
+{[}Figure 7{]}.
 

### 2018-02-11 submission/manuscript.Rmd
spiffy up the conclusion section

@@
 By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in colorectal tumors.
-Further, the data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are commonly associated with the greatest OR of individuals having a carcinoma.
-No conclusive signal with these mouth-associated microbes could be detected for adenoma stage of disease.
-Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma stage of disease due to possible subtle changes in the community.
-Overall, associations between the microbiota and individuals with carcinoma stage of disease were much stronger than with to those with adenoma stage of disease.
+The data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are commonly associated with the greatest OR of individuals having a carcinoma.
+Conversely, no conclusive signal with these mouth-associated microbes could be detected for individuals with an adenoma.
+Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma tumors due to possible subtle changes in the community.
+Overall, associations between the microbiota and individuals with carcinomas were much stronger than with those with adenomas.
 

### 2018-02-11 submission/manuscript.tex
spiffy up the conclusion section

@@
 supporting the importance of the bacterial community in colorectal
-tumors. Further, the data presented here suggests that mouth-associated
-microbes can gain a foothold within the colon and are commonly
-associated with the greatest OR of individuals having a carcinoma. No
+tumors. The data presented here suggests that mouth-associated microbes
+can gain a foothold within the colon and are commonly associated with
+the greatest OR of individuals having a carcinoma. Conversely, no
 conclusive signal with these mouth-associated microbes could be detected
-for adenoma stage of disease. Our observations also highlight the
+for individuals with an adenoma. Our observations also highlight the
 importance of power and sample number considerations when investigating
-the microbiota and adenoma stage of disease due to possible subtle
-changes in the community. Overall, associations between the microbiota
-and individuals with carcinoma stage of disease were much stronger than
-with to those with adenoma stage of disease.
+the microbiota and adenoma tumors due to possible subtle changes in the
+community. Overall, associations between the microbiota and individuals
+with carcinomas were much stronger than with those with adenomas.
 

### 2018-02-11 submission/manuscript.Rmd
update methods section to make clearer, hopefully

@@
 For OTU richness, evenness, and Shannon diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
-Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hyper-variable region used.
-Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study.
-Relative risk significance testing utilized the chi-squared test.
-$\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017].
+Testing for OTU richness, evenness, and Shannon diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hyper-variable region used.
+Odds ratios (OR) were analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study.
+OR significance testing utilized the chi-squared test.
+Diversity differences measured by the Bray-Curtis index utilized the creation of distance matrix and testing with PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017].
 Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002].
@@
 ***Study Analysis Overview:***
-OTU richness, evenness, and Shannon diversity was first assessed for differences between controls, adenoma stage, and carcinoma stage using both linear mixed-effect models and OR.
-For each individual study the Bray-Curtis index was used to assess differences between control-adenoma and control-carcinoma.
-Next, all common genera were assessed for differences in OR for having an adenoma or carcinoma and ranked based on P-value.
-We then built Random Forest models based on the full or selected community (the top 5 increased and top 5 decreased OR based on P-value).
+OTU richness, evenness, and Shannon diversity were first assessed for differences between controls, adenoma tumors, and carcinoma turmors using both linear mixed-effect models and ORs.
+For each individual study the Bray-Curtis index was used to assess differences between control-adenoma and control-carcinoma individuals.
+Next, all common genera were assessed for differences in ORs for individuals having an adenoma or carcinoma and ranked based on P-value.
+We then built Random Forest models based on the full or selected community (the top 5 increased and top 5 decreased ORs based on P-value).
 Comparison between the full and selected models took two different approaches.
-In the first approach, models were trained on one study then tested on the remaining studies.
+In the first approach, models were trained on one study then tested on the remaining studies using genera-based relative abundances.
 This process was repeated for every study in the meta-analysis.
-In the second approach models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
-For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased OR based on P-value.
+In the second approach, models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
+For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased OR analysis based on P-value.
 Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power.
@@
 Similarly, for comparisons in which control versus carcinoma individuals were made the adenoma individuals were excluded from each respective study.
-For all analysis completed feces and tissue samples were kept separate.
+For all analysis completed fecal and tissue samples were kept separate.
 Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched).
@@
 ***Obtaining Genera Relative Abundance and Selected Models:***
-For the genera analysis of the OR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings.
-The select models for the first approach utilized the top 5 most significantly increased and decreased ORs.
+For the genera analysis of the ORs, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings was obtained.
+The select models for the first Random Forest model approach utilized the top 5 most significantly increased and decreased ORs.
 For this approach, the full community models for each study were built by utilizing all genera and lowest taxonomic groups identified within that particular study.
 The select models for the second approach, that used OTU-based Random Forest models, utilized a similar method as the first approach.
-The main difference was that any OTU that taxonomically classified to one of the genera in the top 5 increased or decreased OR were included in the select model.
-OTU Random Forest models using the full community included all OTUs.
+The main difference was that any OTU that taxonomically classified to one of the taxa in the top 5 increased or decreased OR analysis were included in the select model.
+OTU Random Forest models, using the full community, included all OTUs that did not have near zero variance within that specific study.
 
@@
 For some studies samples became unmatched when a corresponding matched sample did not make it through sequence processing.
-All samples, from both tissue sample types, were analyzed together for the linear mixed-effect models with samples from the same individual corrected for.
-For all other analysis, not mentioned herein, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.
+All samples, from both matched and unmatched tissue samples, were analyzed together for the linear mixed-effect models with samples from the same individual being corrected for.
+All other analysis, where it is not specified explicitly, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.
 
@@
 Using Mean Decrease in Accuracy (MDA) the top 10 most important variables to the Random Forest model were obtained for the full models of the two different approaches used.
-For the first approach utilizing genus-based models, the number of times that a genus showed up in the top 10 of the training set across each study was counted.
+For the first approach utilizing genus-based models, the number of times that a specific taxa showed up in the top 10 of the training set across each study was counted.
 For the second approach, that utilized the OTU-based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study.
-Common taxa, for the OTU based models, were identified by using the lowest classification within the RDP database for each of the specific OTUs obtained from the previous counts and the number of times this classification occurred in the top 10, in each study, was recorded.
-Finally, for the two studies that had adenoma tissue (Lu and Flemer) were equally divided between matched and unmatched studies and were grouped together for the counting of the top 10 genera and OTUs.
+Common taxa were then identified by using the lowest classification for each of the specific OTUs obtained from these counts and the number of times this classification occurred across the top 10 of each study was recorded.
+Finally, the two studies that had adenoma tumor tissue (Lu and Flemer) were equally divided between matched and unmatched studies and were grouped together for the counting of the top 10 genera and OTUs for both Random Forest approaches.
 

### 2018-02-11 submission/manuscript.tex
update methods section to make clearer, hopefully

@@
 then Z-score normalized using the car (v2.1.6) package {[}40{]}. Testing
-for \(\alpha\)-diversity differences utilized linear mixed-effect models
-created using the lme4 (v1.1.15) package {[}41{]} to correct for study,
-repeat sampling of individuals (tissue only), and 16S hyper-variable
-region used. Relative risk was analyzed using both the epiR (v0.9.93)
-and metafor (v2.0.0) packages {[}42,43{]} by assessing how many
-individuals with and without disease were above and below the overall
-median value within each specific study. Relative risk significance
-testing utilized the chi-squared test. \(\beta\)-diversity differences
-utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
-vegan (v2.4.5) package {[}44{]}. Random Forest models were built using
-both the caret (v6.0.78) and randomForest (v4.6.12) packages
-{[}45,46{]}. Power analysis and estimations were made using the pwr
-(v1.2.1) and statmod (v1.4.30) packages {[}47,48{]}. All figures were
-created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
-{[}49,50{]}.
+for OTU richness, evenness, and Shannon diversity differences utilized
+linear mixed-effect models created using the lme4 (v1.1.15) package
+{[}41{]} to correct for study, repeat sampling of individuals (tissue
+only), and 16S hyper-variable region used. Odds ratios (OR) were
+analyzed using both the epiR (v0.9.93) and metafor (v2.0.0) packages
+{[}42,43{]} by assessing how many individuals with and without disease
+were above and below the overall median value within each specific
+study. OR significance testing utilized the chi-squared test. Diversity
+differences measured by the Bray-Curtis index utilized the creation of
+distance matrix and testing with PERMANOVA executed with the vegan
+(v2.4.5) package {[}44{]}. Random Forest models were built using both
+the caret (v6.0.78) and randomForest (v4.6.12) packages {[}45,46{]}.
+Power analysis and estimations were made using the pwr (v1.2.1) and
+statmod (v1.4.30) packages {[}47,48{]}. All figures were created using
+both ggplot2 (v2.2.1) and gridExtra (v2.3) packages {[}49,50{]}.
 
 \textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
-Shannon diversity was first assessed for differences between controls,
-adenoma stage, and carcinoma stage using both linear mixed-effect models
-and OR. For each individual study the Bray-Curtis index was used to
-assess differences between control-adenoma and control-carcinoma. Next,
-all common genera were assessed for differences in OR for having an
-adenoma or carcinoma and ranked based on P-value. We then built Random
-Forest models based on the full or selected community (the top 5
-increased and top 5 decreased OR based on P-value). Comparison between
-the full and selected models took two different approaches. In the first
-approach, models were trained on one study then tested on the remaining
-studies. This process was repeated for every study in the meta-analysis.
-In the second approach models were built using OTU level data and a
-10-fold CV over 100 different iterations, based on random 80/20
-splitting of the data, was used to generate a range of expected AUCs.
-For these OTU-based models, the selected model included all OTUs that
-had a taxonomic classification to a taxa in the top 5 increased and top
-5 decreased OR based on P-value. Finally, the power of each study was
-assessed for an effect size ranging from 1\% to 30\% and an estimated
-sample size, for these effect sizes, was generated based on 80\% power.
-For comparisons in which only control versus adenoma individuals were
-made, the carcinoma individuals were excluded from each respective
-study. Similarly, for comparisons in which control versus carcinoma
-individuals were made the adenoma individuals were excluded from each
-respective study. For all analysis completed feces and tissue samples
-were kept separate. Within the tissue groups the data were further
-divided between samples from the same individual (matched) and those
-from different individuals (unmatched).
+Shannon diversity were first assessed for differences between controls,
+adenoma tumors, and carcinoma turmors using both linear mixed-effect
+models and ORs. For each individual study the Bray-Curtis index was used
+to assess differences between control-adenoma and control-carcinoma
+individuals. Next, all common genera were assessed for differences in
+ORs for individuals having an adenoma or carcinoma and ranked based on
+P-value. We then built Random Forest models based on the full or
+selected community (the top 5 increased and top 5 decreased ORs based on
+P-value). Comparison between the full and selected models took two
+different approaches. In the first approach, models were trained on one
+study then tested on the remaining studies using genera-based relative
+abundances. This process was repeated for every study in the
+meta-analysis. In the second approach, models were built using OTU level
+data and a 10-fold CV over 100 different iterations, based on random
+80/20 splitting of the data, was used to generate a range of expected
+AUCs. For these OTU-based models, the selected model included all OTUs
+that had a taxonomic classification to a taxa in the top 5 increased and
+top 5 decreased OR analysis based on P-value. Finally, the power of each
+study was assessed for an effect size ranging from 1\% to 30\% and an
+estimated sample size, for these effect sizes, was generated based on
+80\% power. For comparisons in which only control versus adenoma
+individuals were made, the carcinoma individuals were excluded from each
+respective study. Similarly, for comparisons in which control versus
+carcinoma individuals were made the adenoma individuals were excluded
+from each respective study. For all analysis completed fecal and tissue
+samples were kept separate. Within the tissue groups the data were
+further divided between samples from the same individual (matched) and
+those from different individuals (unmatched).
 
 \textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
-For the genera analysis of the OR, OTUs were added together based on the
-genus or lowest available taxonomic classification level and the total
-average counts, for 100 different subsamplings. The select models for
-the first approach utilized the top 5 most significantly increased and
-decreased ORs. For this approach, the full community models for each
-study were built by utilizing all genera and lowest taxonomic groups
-identified within that particular study. The select models for the
-second approach, that used OTU-based Random Forest models, utilized a
-similar method as the first approach. The main difference was that any
-OTU that taxonomically classified to one of the genera in the top 5
-increased or decreased OR were included in the select model. OTU Random
-Forest models using the full community included all OTUs.
+For the genera analysis of the ORs, OTUs were added together based on
+the genus or lowest available taxonomic classification level and the
+total average counts, for 100 different subsamplings was obtained. The
+select models for the first Random Forest model approach utilized the
+top 5 most significantly increased and decreased ORs. For this approach,
+the full community models for each study were built by utilizing all
+genera and lowest taxonomic groups identified within that particular
+study. The select models for the second approach, that used OTU-based
+Random Forest models, utilized a similar method as the first approach.
+The main difference was that any OTU that taxonomically classified to
+one of the taxa in the top 5 increased or decreased OR analysis were
+included in the select model. OTU Random Forest models, using the full
+community, included all OTUs that did not have near zero variance within
+that specific study.
 
@@
 became unmatched when a corresponding matched sample did not make it
-through sequence processing. All samples, from both tissue sample types,
-were analyzed together for the linear mixed-effect models with samples
-from the same individual corrected for. For all other analysis, not
-mentioned herein, matched and unmatched samples were analyzed separately
-using the statistical approaches mentioned in the Statistical Analysis
-section.
+through sequence processing. All samples, from both matched and
+unmatched tissue samples, were analyzed together for the linear
+mixed-effect models with samples from the same individual being
+corrected for. All other analysis, where it is not specified explicitly,
+matched and unmatched samples were analyzed separately using the
+statistical approaches mentioned in the Statistical Analysis section.
 
@@
 different approaches used. For the first approach utilizing genus-based
-models, the number of times that a genus showed up in the top 10 of the
-training set across each study was counted. For the second approach,
-that utilized the OTU-based models, the medians for each OTU across 100
-different 80/20 splits of the data was generated and the top 10 OTUs
-then counted for each study. Common taxa, for the OTU based models, were
-identified by using the lowest classification within the RDP database
-for each of the specific OTUs obtained from the previous counts and the
-number of times this classification occurred in the top 10, in each
-study, was recorded. Finally, for the two studies that had adenoma
-tissue (Lu and Flemer) were equally divided between matched and
-unmatched studies and were grouped together for the counting of the top
-10 genera and OTUs.
+models, the number of times that a specific taxa showed up in the top 10
+of the training set across each study was counted. For the second
+approach, that utilized the OTU-based models, the medians for each OTU
+across 100 different 80/20 splits of the data was generated and the top
+10 OTUs then counted for each study. Common taxa were then identified by
+using the lowest classification for each of the specific OTUs obtained
+from these counts and the number of times this classification occurred
+across the top 10 of each study was recorded. Finally, the two studies
+that had adenoma tumor tissue (Lu and Flemer) were equally divided
+between matched and unmatched studies and were grouped together for the
+counting of the top 10 genera and OTUs for both Random Forest
+approaches.
 

### 2018-02-13 submission/manuscript.Rmd
add docx

@@
 output:
+  word_document: default
   pdf_document:
@@
     keep_tex: yes
-  word_document: default
 geometry: margin=1.0in

### 2018-02-14 submission/supplemental.Rmd
first pass at combining tables

@@
 
-**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Adenoma**
+**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool**
 
@@
                                       c("brim", "zeller", "baxter", "hale"), 
-                                      c("Brim", "Zeller", "Baxter", "Hale")))
+                                      c("Brim", "Zeller", "Baxter", "Hale")), 
+         comparison = "Control vs. Adenoma") %>% 
+  select(fstat, r2, pvalue, study, comparison)
 
-colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Adenoma (n)", "Control (n)")
-
-kable(table_s1, format = "markdown", align = 'c', padding = 2)
-```
-
-\newpage
-
-**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool Carcinoma**
-
-```{r supplemental_table2, echo=FALSE, message=FALSE, warning=FALSE}
+colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Comparison")
 
@@
                                       c("wang", "weir", "ahn", "zeller", "baxter", "hale", "flemer"), 
-                                      c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")))
+                                      c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")), 
+         comparison = "Control vs. Carcinoma") %>% 
+  select(fstat, r2, pvalue, study, comparison)
 
-colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma (n)", "Control (n)")
+colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Comparison")
 
-kable(table_s2, format = "markdown", align = 'c', padding = 2)
+combined_table_s1 <- table_s1 %>% bind_rows(table_s2)
+
+kable(combined_table_s1, format = "markdown", align = 'c', padding = 2)
 ```
@@
 
-**Table S3: Alpha Diversity Metrics OR in Tissue for Adenoma**
+**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue**
+
+```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
+
+temp_t3a <- read_csv("../data/process/tables/beta_perm_adn_unmatched_tissue_summary.csv")
+temp_t3b <- read_csv("../data/process/tables/beta_perm_adn_matched_tissue_summary.csv")
+
+table_s3 <- temp_t3a %>% 
+  bind_rows(temp_t3b) %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = c("Lu", "Flemer", "Lu (Matched)"), 
+         comparison = "Control vs. Adenoma") %>% 
+  select(fstat, r2, pvalue, study, comparison)
+
+colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Comparison")
+
+temp_t4a <- read_csv("../data/process/tables/beta_perm_unmatched_tissue_summary.csv")
+temp_t4b <- read_csv("../data/process/tables/beta_perm_matched_tissue_summary.csv")
+
+table_s4 <- temp_t4a %>% 
+  bind_rows(temp_t4b) %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = c("Sanapareddy", "Burns", "Flemer", "Chen", 
+                   "Dejea (Matched)", "Geng (Matched)", "Burns (Matched)"), 
+         comparison = "Control vs. Carcinoma") %>% 
+  select(fstat, r2, pvalue, study, comparison)
+
+colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Comparison")
+
+combined_table_s2 <- table_s3 %>% bind_rows(table_s4)
+
+kable(combined_table_s2, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S3: Alpha Diversity Metrics OR in Tissue**
 
@@
          pvalue = round(pvalue, digits = 2), 
-         measure = c("OTU Richness", "Shannon Diversity", "Evenness"))
+         measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
+         comparison = "Control vs. Adenoma", 
+         tissue_type = "Combined")
 
-colnames(alpha_RR_adn_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure")
-
-kable(alpha_RR_adn_tissue, format = "markdown", align = 'c', padding = 2)
-```
-
-\newpage
-
-**Table S4: Alpha Diversity Metrics OR in Tissue for Unmatched Carcinoma Non-Carcinoma Samples**
-
-```{r supplemental_table4, echo=FALSE, message=FALSE, warning=FALSE}
+colnames(alpha_RR_adn_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure", 
+                                   "Comparison", "Tissue Group")
 
@@
          pvalue = round(pvalue, digits = 2), 
-         measure = c("OTU Richness", "Shannon Diversity", "Evenness"))
+         measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
+         comparison = "Control vs. Carcinoma", 
+         tissue_type = "Unmatched")
 
-colnames(alpha_RR_crc_unmatch_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure")
-
-kable(alpha_RR_crc_unmatch_tissue, format = "markdown", align = 'c', padding = 2)
-```
-
-\newpage
-
-**Table S5: Alpha Diversity Metrics OR in Tissue for Matched Carcinoma Non-Carcinoma Samples**
-
-```{r supplemental_table5, echo=FALSE, message=FALSE, warning=FALSE}
+colnames(alpha_RR_crc_unmatch_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure", 
+                                           "Comparison", "Tissue Group")
 
@@
          pvalue = round(pvalue, digits = 2), 
-         measure = c("OTU Richness", "Shannon Diversity", "Evenness"))
+         measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
+         comparison = "Control vs. Carcinoma", 
+         tissue_type = "Matched")
 
-colnames(alpha_RR_crc_match_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure")
+colnames(alpha_RR_crc_match_tissue) <-  c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure", 
+                                          "Comparison", "Tissue Group")
 
-kable(alpha_RR_crc_match_tissue, format = "markdown", align = 'c', padding = 2)
+combined_table_S3 <- alpha_RR_adn_tissue %>% bind_rows(alpha_RR_crc_unmatch_tissue, alpha_RR_crc_match_tissue)
+
+kable(combined_table_S3, format = "markdown", align = 'c', padding = 2)
 ```
@@
 
-**Table S6: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue Adenoma**
-
-```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
-
-temp_t3a <- read_csv("../data/process/tables/beta_perm_adn_unmatched_tissue_summary.csv")
-temp_t3b <- read_csv("../data/process/tables/beta_perm_adn_matched_tissue_summary.csv")
-
-table_s6 <- temp_t3a %>% 
-  bind_rows(temp_t3b) %>% 
-  mutate(fstat = round(fstat, digits = 2), 
-         r2 = round(r2, digits = 3), 
-         pvalue = round(pvalue, digits = 4), 
-         study = c("Lu", "Flemer", "Lu (Matched)"))
-
-colnames(table_s6) <- c("F statistic", "R2", "P-value", "Study", "Adenoma (n)", "Control (n)")
-
-kable(table_s6, format = "markdown", align = 'c', padding = 2)
-```
-
-\newpage
-
-**Table S7: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue Carcinoma**
-
-```{r supplemental_table7, echo=FALSE, message=FALSE, warning=FALSE}
-
-temp_t4a <- read_csv("../data/process/tables/beta_perm_unmatched_tissue_summary.csv")
-temp_t4b <- read_csv("../data/process/tables/beta_perm_matched_tissue_summary.csv")
-
-table_s7 <- temp_t4a %>% 
-  bind_rows(temp_t4b) %>% 
-  mutate(fstat = round(fstat, digits = 2), 
-         r2 = round(r2, digits = 3), 
-         pvalue = round(pvalue, digits = 4), 
-         study = c("Sanapareddy", "Burns", "Flemer", "Chen", 
-                   "Dejea (Matched)", "Geng (Matched)", "Burns (Matched)"))
-
-colnames(table_s7) <- c("F statistic", "R2", "P-value", "Study", "Carcinoma (n)", "Control (n)")
-
-kable(table_s7, format = "markdown", align = 'c', padding = 2)
-
-```
-
-\newpage
-
-**Table S8: OR of Adenoma in Stool for each Individual Genera with Initial P-Value under 0.05**
+**Table S4: OR of Tumor in Stool for each Individual Genera with Initial P-Value under 0.05**
 
@@
          ci_ub = round(ci_ub, digits = 2), 
-         pvalue = round(pvalue, digits = 4), 
-         bh = round(bh, digits = 3), 
-         measure = str_replace_all(measure, "_", " ")) %>% 
-  select(rr, ci_lb, ci_ub, pvalue, bh, measure)
+         pvalue = format(pvalue, scientific = T, digits = 3), 
+         bh = format(bh, scientific = T, digits = 3), 
+         measure = str_replace_all(measure, "_", " "), 
+         comparison = "Control vs. Adenoma") %>% 
+  select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison)
 
-colnames(temp_8) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+colnames(temp_8) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID", "Comparison")
 
-kable(temp_8, format = "markdown", align = 'c', padding = 2)
+temp_10 <- read_csv("../data/process/tables/select_genus_OR_stool_composite.csv") %>% 
+  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
+  arrange(bh, pvalue, rr) %>% 
+  mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = format(pvalue, scientific = T, digits = 3), 
+         bh = format(bh, scientific = T, digits = 3), 
+         measure = str_replace_all(measure, "_", " "), 
+         comparison = "Control vs. Carcinoma") %>% 
+  select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison)
+
+colnames(temp_10) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID", "Comparison")
+
+combined_table_S4 <- temp_8 %>% bind_rows(temp_10)
+
+kable(combined_table_S4, format = "markdown", align = 'c', padding = 2)
 
@@
          ci_ub = round(ci_ub, digits = 2), 
-         pvalue = round(pvalue, digits = 4), 
-         bh = round(bh, digits = 3), 
-         measure = str_replace_all(measure, "_", " ")) %>% 
-  select(rr, ci_lb, ci_ub, pvalue, bh, measure)
+         pvalue = format(pvalue, scientific = T, digits = 3), 
+         bh = format(bh, scientific = T, digits = 3), 
+         measure = str_replace_all(measure, "_", " "), 
+         comparison = "Control vs. Adenoma", 
+         tissue_type = "Combined") %>% 
+  select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison, tissue_type)
 
-colnames(temp_9) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+colnames(temp_9) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID", 
+                      "Comparison", "Tissue Group")
 
-kable(temp_9, format = "markdown", align = 'c', padding = 2)
-
-```
-
-\newpage
-
-**Table S10: OR of Carcinoma in Stool for each Individual Genera with Initial P-Value under 0.05**
-
-```{r supplemental_table10, echo=FALSE, message=FALSE, warning=FALSE}
-
-temp_10 <- read_csv("../data/process/tables/select_genus_OR_stool_composite.csv") %>% 
+temp_11 <- read_csv("../data/process/tables/select_genus_OR_unmatched_tissue_composite.csv") %>% 
   mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
@@
          bh = format(bh, scientific = T, digits = 3), 
-         measure = str_replace_all(measure, "_", " ")) %>% 
-  select(rr, ci_lb, ci_ub, pvalue, bh, measure)
+         measure = str_replace_all(measure, "_", " "), 
+         comparison = "Control vs. Carcinoma", 
+         tissue_type = "Unmatched") %>% 
+  select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison, tissue_type)
 
-colnames(temp_10) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+colnames(temp_11) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID", 
+                      "Comparison", "Tissue Group")
 
-kable(temp_10, format = "markdown", align = 'c', padding = 2)
-
-```
-
-\newpage
-
-**Table S11: OR of Carcinoma in Tissue for each Individual Genera with Initial P-Value under 0.05**
-
-```{r supplemental_table11, echo=FALSE, message=FALSE, warning=FALSE}
-
-temp_11 <- read_csv("../data/process/tables/select_genus_OR_tissue_composite.csv") %>% 
+temp_12 <- read_csv("../data/process/tables/select_genus_OR_matched_tissue_composite.csv") %>% 
   mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(pvalue < 0.05) %>% 
@@
          bh = format(bh, scientific = T, digits = 3), 
-         measure = str_replace_all(measure, "_", " ")) %>% 
-  select(rr, ci_lb, ci_ub, pvalue, bh, measure)
+         measure = str_replace_all(measure, "_", " "), 
+         comparison = "Control vs. Carcinoma", 
+         tissue_type = "Matched") %>% 
+  select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison, tissue_type)
 
-colnames(temp_11) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID")
+colnames(temp_12) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID", 
+                      "Comparison", "Tissue Group")
 
-kable(temp_11, format = "markdown", align = 'c', padding = 2)
+combined_table_S5 <- temp_9 %>% bind_rows(temp_11, temp_12)
+
+kable(combined_table_S5, format = "markdown", align = 'c', padding = 2)
 
 ```
+
+\newpage

### 2018-02-14 submission/supplemental.Rmd
complete table combination

@@
 
-**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool**
+**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stooland Tissue**
 
@@
                                       c("Brim", "Zeller", "Baxter", "Hale")), 
-         comparison = "Control vs. Adenoma") %>% 
-  select(fstat, r2, pvalue, study, comparison)
+         comparison = "Adenoma", 
+         sample_type = "Stool") %>% 
+  select(fstat, r2, pvalue, study, comparison, sample_type)
 
-colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Comparison")
+colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
 
@@
                                       c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")), 
-         comparison = "Control vs. Carcinoma") %>% 
-  select(fstat, r2, pvalue, study, comparison)
+         comparison = "Carcinoma", 
+         sample_type = "Stool") %>% 
+  select(fstat, r2, pvalue, study, comparison, sample_type)
 
-colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Comparison")
+colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
 
 combined_table_s1 <- table_s1 %>% bind_rows(table_s2)
-
-kable(combined_table_s1, format = "markdown", align = 'c', padding = 2)
-```
-
-\newpage
-
-**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Tissue**
-
-```{r supplemental_table6, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
          study = c("Lu", "Flemer", "Lu (Matched)"), 
-         comparison = "Control vs. Adenoma") %>% 
-  select(fstat, r2, pvalue, study, comparison)
+         comparison = "Adenoma", 
+         sample_type = "Tissue") %>% 
+  select(fstat, r2, pvalue, study, comparison, sample_type)
 
-colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Comparison")
+colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
 
@@
                    "Dejea (Matched)", "Geng (Matched)", "Burns (Matched)"), 
-         comparison = "Control vs. Carcinoma") %>% 
-  select(fstat, r2, pvalue, study, comparison)
+         comparison = "Carcinoma", 
+         sample_type = "Tissue") %>% 
+  select(fstat, r2, pvalue, study, comparison, sample_type)
 
-colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Comparison")
+colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
 
@@
 
-kable(combined_table_s2, format = "markdown", align = 'c', padding = 2)
+combined_table_s1 <- combined_table_s1 %>% bind_rows(combined_table_s2)
+
+kable(combined_table_s1, format = "markdown", align = 'c', padding = 2)
 ```
@@
 
-**Table S3: Alpha Diversity Metrics OR in Tissue**
+**Table S2: Alpha Diversity Metrics OR in Tissue**
+
+\footnotesize
 
@@
          measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
-         comparison = "Control vs. Adenoma", 
+         comparison = "Adenoma", 
          tissue_type = "Combined")
@@
 colnames(alpha_RR_adn_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure", 
-                                   "Comparison", "Tissue Group")
+                                   "Tumor", "Tissue Group")
 
@@
          measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
-         comparison = "Control vs. Carcinoma", 
+         comparison = "Carcinoma", 
          tissue_type = "Unmatched")
@@
 colnames(alpha_RR_crc_unmatch_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure", 
-                                           "Comparison", "Tissue Group")
+                                           "Tumor", "Tissue Group")
 
@@
          measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
-         comparison = "Control vs. Carcinoma", 
+         comparison = "Carcinoma", 
          tissue_type = "Matched")
@@
 colnames(alpha_RR_crc_match_tissue) <-  c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure", 
-                                          "Comparison", "Tissue Group")
+                                          "Tumor", "Tissue Group")
 
@@
 
+\normalsize
+
 \newpage
 
-**Table S4: OR of Tumor in Stool for each Individual Genera with Initial P-Value under 0.05**
+**Table S3: OR of Tumor in Stool by Individual Genera with Initial P-Value under 0.05**
+
+\footnotesize
 
@@
          bh = format(bh, scientific = T, digits = 3), 
-         measure = str_replace_all(measure, "_", " "), 
-         comparison = "Control vs. Adenoma") %>% 
+         measure = str_replace_all(measure, "_unclassified", ""), 
+         comparison = "Adenoma") %>% 
   select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison)
 
-colnames(temp_8) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID", "Comparison")
+colnames(temp_8) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Taxa ID", "Tumor")
 
@@
          bh = format(bh, scientific = T, digits = 3), 
-         measure = str_replace_all(measure, "_", " "), 
-         comparison = "Control vs. Carcinoma") %>% 
+         measure = str_replace_all(measure, "_unclassified", ""), 
+         comparison = "Carcinoma") %>% 
   select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison)
 
-colnames(temp_10) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID", "Comparison")
+colnames(temp_10) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Taxa ID", "Tumor")
 
@@
 
+\normalsize
+
 \newpage
 
-**Table S9: OR of Adenoma in Tissue for each Individual Genera with Initial P-Value under 0.05**
+**Table S4: OR of Tumor in Tissue by Individual Genera with Initial P-Value under 0.05**
+
+\footnotesize
 
@@
          bh = format(bh, scientific = T, digits = 3), 
-         measure = str_replace_all(measure, "_", " "), 
-         comparison = "Control vs. Adenoma", 
+         measure = str_replace_all(measure, "_unclassified", ""), 
+         comparison = "Adenoma", 
          tissue_type = "Combined") %>% 
@@
 
-colnames(temp_9) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID", 
-                      "Comparison", "Tissue Group")
+colnames(temp_9) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Taxa ID", 
+                      "Tumor", "Tissue Group")
 
@@
          bh = format(bh, scientific = T, digits = 3), 
-         measure = str_replace_all(measure, "_", " "), 
-         comparison = "Control vs. Carcinoma", 
+         measure = str_replace_all(measure, "_unclassified", ""), 
+         comparison = "Carcinoma", 
          tissue_type = "Unmatched") %>% 
@@
 
-colnames(temp_11) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID", 
-                      "Comparison", "Tissue Group")
+colnames(temp_11) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Taxa ID", 
+                      "Tumor", "Tissue Group")
 
@@
          bh = format(bh, scientific = T, digits = 3), 
-         measure = str_replace_all(measure, "_", " "), 
-         comparison = "Control vs. Carcinoma", 
+         measure = str_replace_all(measure, "_unclassified", ""), 
+         comparison = "Carcinoma", 
          tissue_type = "Matched") %>% 
@@
 
-colnames(temp_12) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Genera ID", 
-                      "Comparison", "Tissue Group")
+colnames(temp_12) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Taxa ID", 
+                      "Tumor", "Tissue Group")
 
@@
 
+\normalsize
+
 \newpage

### 2018-02-14 code/make_tissue_imp_otu_graph.R
change fig number

@@ -170,5 +170,5 @@ tissue_graph <- grid.arrange(adn_genera, crc_unmatched_genera, crc_matched_gener
                             layout_matrix = rbind(c(1, 2, 3), c(4, 5, 6)))
 
 
-ggsave("results/figures/FigureS5.pdf", 
+ggsave("results/figures/FigureS4.pdf", 
        tissue_graph, width = 13, height = 8, dpi = 300)

### 2018-02-14 code/make_genus_rf_auc_against_study_graph.R
change fig layouts

@@ -13,19 +13,30 @@ loadLibs(c("tidyverse", "gridExtra", "viridis"))
 
 # Function to generate the tables to be used
 make_table <- function(studies, path_to_file, first_data_part_name, 
-                       second_data_part_name, ending){
+                       second_data_part_name, ending, select_included = T){
+  
   
   tempFull <- sapply(studies, 
                      function(x) read_csv(paste(path_to_file, first_data_part_name, x, 
                                                 ending, sep = "")) %>% 
                        mutate(model = x, model_type = "full"), simplify = F) %>% bind_rows()
   
-  tempSelect <- sapply(studies, 
-                       function(x) read_csv(paste(path_to_file, second_data_part_name, x, 
-                                                  ending, sep = "")) %>% 
-                         mutate(model = x, model_type = "select"), simplify = F) %>% bind_rows()
+  if(select_included == T){
+    
+    tempSelect <- sapply(studies, 
+                         function(x) read_csv(paste(path_to_file, second_data_part_name, x, 
+                                                    ending, sep = "")) %>% 
+                           mutate(model = x, model_type = "select"), simplify = F) %>% bind_rows()
+    
+    tempALL <- tempFull %>% bind_rows(tempSelect)
+  } else{
+    
+    tempALL <- tempFull
+  }
   
-  tempALL <- tempFull %>% bind_rows(tempSelect)
+  
+  
+  
   
   return(tempALL)
   
@@ -40,16 +51,16 @@ make_table <- function(studies, path_to_file, first_data_part_name,
 adn_stool_studies <- c("baxter", "brim", "hale", "zeller")
 
 adn_all_stool <- make_table(adn_stool_studies, "data/process/tables/", 
-                            "adn_ALL_genus_stool_RF_full_", "adn_genus_stool_RF_select_", 
-                            "_pvalue_summary.csv")
+                            "adn_ALL_genus_stool_RF_full_", "adn_ALL_genus_stool_RF_full_", 
+                            "_pvalue_summary.csv",select_included = F)
 
 # Load needed data (adenoma -- tissue)
 adn_tissue_studies <- c("flemer", "lu")
 
 adn_all_tissue <- make_table(adn_tissue_studies, "data/process/tables/", 
                             "adn_ALL_genus_unmatched_tissue_RF_full_", 
[…]

### 2018-02-14 submission/manuscript.Rmd
updated figure legends

--- 2018-02-11 submission/manuscript.Rmd
@@
 
-**Figure 1: Community Differences between Control, Adenoma, and Carcinoma Across Sampling Site.** A) Stool sample community differences by disease group.
-B) Unmatched tissue samples differences by disease group.
-C) Matched tissue sample differences by group disease group.
-The dashed line represents a Z-score of 0 or no difference from the median.
-
-**Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Stool.** A) Community-based odds ratio for adenoma.
+**Figure 1: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Stool.** A) Community-based odds ratio for adenoma.
 B) Community-based odds ratio for carcinoma.
@@
 
-**Figure 3: Top 5 Genera that Decrease and Increase Odds Ratio for Lesion.** A) Adenoma odds ratio in stool.
-B) Carcinoma odds ratio in stool.
-C) Adenoma odds ratio in tissue.
-D) Carcinoma odds ratio in tissue.
-For all panels the odds ratio was also compared to whether one, two, three, or four of the CRC-associated genera were present.
-Points represented as only half on the graph have an OR of infinity in the positive or negative direction.
+**Figure 2: Stool Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
+B) Carcinoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
+The black line represents the median AUC for the respective group.
+If no values are present in the singificant OR taxa group then there were no significant taxa identified and no model was tested.
 
-**Figure 4: Stool OTU Random Forest Model Across Studies.** A) Adenoma random forest model between the full and select community OTUs only.
-B) Carcinoma random forest model between the full and select community OTUs only.
-The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.
-The solid red line represents the median AUC of all the studies for either the full or select community OTUS only model.
+**Figure 3: Stool Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
+B) Test AUCs of carcinoma models using all genera or significant OR taxa only.
+The black line represents the AUC at 0.5.
+The red lines represent the median AUC of all test AUCs for a specific study.
 
-**Figure 5: Tissue OTU Random Forest Model Across Studies.** A) Adenoma random forest model between the full and select community OTUs only.
-B) Carcinoma random forest model between the full and select community OTUs only.
-The dotted line represents an AUC of 0.5 and the lines represent the range in which the AUC for the 100 different 80/20 runs fell between.
-The solid red line represents the median AUC of all the studies for either the full community or select genera OTUS only model.
+**Figure 4: Most Common Taxa Across Full Community Stool Study Models.** A) Common taxa in the top 10 for adenoma Random Forest all genera-based models.
+B) Common taxa in the top 10 for carcinoma Random Forest all genera-based models.
+C) Common taxa in the top 10 for adenoma Random Forest all OTU-based models.
+D) Common genera in the top 10 for carcinoma Random Forest all OTU-based models.
 
-**Figure 6: Most Common Genera Across Full Community Stool Study Models.** A) Common genera in the top 10 for adenoma Random Forest genus models.
-B) Common genera in the top 10 for carcinoma Random Forest genus models.
-C) Common genera in the top 10 for adenoma Random Forest OTU models.
-D) Common genera in the top 10 for carcinoma Random Forest OTU models.
-
-**Figure 7: Power and Effect Size Analysis of Studies Included.** A) Power based on effect size for studies with adenoma individuals.
+**Figure 5: Power and Effect Size Analysis of Studies Included.** A) Power based on effect size for studies with adenoma individuals.
 B) Power based on effect size for studies with carcinoma individuals.
@@
 
-**Figure S2: Stool Random Forest Genus Model AUC for each Study.** A) AUC of adenoma models using all genera or select genera only.
-B) AUC of carcinoma models using all genera or select genera only.
-The black line represents the median within each group.
+**Figure S2: Tissue Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
+B) Carcinoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
+The black line represents the median AUC for the respective group.
+If no values are present in the singificant OR taxa group then there were no significant taxa identified and no model was tested.
 
-**Figure S3: Tissue Random Forest Genus Model AUC for each Study.** A) AUC of adenoma models using all genera or only select genera divided between matched and unmatched tissue.
-B) AUC of carcinoma models using all genera or select genera only.
-The black line represents the median within each group divided between matched and unmatched tissue.
+**Figure S3: Tissue Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
+B) Test AUCs of carcinoma models using all genera for matched tissue studies.
+C) Test AUCs of carcinoma models using all genera or significant OR taxa only for unmatched tissue studies The black line represents the AUC at 0.5.
+The red lines represent the median AUC of all test AUCs for a specific study.
 
-**Figure S4: Stool Random Forest Prediction Success Using Genera Across Studies.** A) AUC for prediction in adenoma using all genera or select genera only.
-B) AUC for prediction in carcinoma using all genera or select genera only.
-The dotted line represents an AUC of 0.5.
-The x-axis is the data set in which the model was initially trained on.
-The red lines represent the median AUC using that specific study as the training set.
-
-**Figure S5: Tissue Random Forest Prediction Success of Carcinoma Using Genera Across Studies.** A) AUC for prediction in unmatched tissue for all genera or select genera only.
-B) AUC for prediction in matched tissue using all genera or select genera only.
-The dotted line represents an AUC of 0.5.
-The x-axis is the data set in which the model was initially trained on.
-The red lines represent the median AUC using that specific study as the training set.
-
-**Figure S6: Tissue Random Forest Prediction Success of Adenoma Using Genera Across Studies.** The red lines represent the median AUC using that specific study as the training set.
-
-**Figure S7: Most Common Genera Across Full Community Tissue Study Models.** A) Common genera in the top 10 for adenoma Random Forest genus models.
-B) Common genera in the top 10 for unmatched carcinoma Random Forest genus models.
-B) Common genera in the top 10 for matched carcinoma Random Forest genus models.
-D) Common genera in the top 10 for adenoma Random Forest OTU models.
-E) Common genera in the top 10 for unmatched carcinoma Random Forest OTU models.
-F) Common genera in the top 10 for matched carcinoma Random Forest OTU models.
+**Figure S4: Most Common Genera Across Full Community Tissue Study Models.** A) Common genera in the top 10 for adenoma Random Forest for all genera-based models.
+B) Common genera in the top 10 for unmatched carcinoma Random Forest for all genera-based models.
+B) Common genera in the top 10 for matched carcinoma Random Forest for all genera-based models.
+D) Common genera in the top 10 for adenoma Random Forest for all OTU-based models.
+E) Common genera in the top 10 for unmatched carcinoma Random Forest for all OTU-based models.
+F) Common genera in the top 10 for matched carcinoma Random Forest for all OTU-based models.
 

### 2018-02-14 submission/manuscript.tex
updated figure legends

@@
 
-\textbf{Figure 1: Community Differences between Control, Adenoma, and
-Carcinoma Across Sampling Site.} A) Stool sample community differences
-by disease group. B) Unmatched tissue samples differences by disease
-group. C) Matched tissue sample differences by group disease group. The
-dashed line represents a Z-score of 0 or no difference from the median.
-
-\textbf{Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial
+\textbf{Figure 1: Odds Ratio for Adenoma or Carcinoma based on Bacterial
 Community Metrics in Stool.} A) Community-based odds ratio for adenoma.
@@
 
-\textbf{Figure 3: Top 5 Genera that Decrease and Increase Odds Ratio for
-Lesion.} A) Adenoma odds ratio in stool. B) Carcinoma odds ratio in
-stool. C) Adenoma odds ratio in tissue. D) Carcinoma odds ratio in
-tissue. For all panels the odds ratio was also compared to whether one,
-two, three, or four of the CRC-associated genera were present. Points
-represented as only half on the graph have an OR of infinity in the
-positive or negative direction.
-
-\textbf{Figure 4: Stool OTU Random Forest Model Across Studies.} A)
-Adenoma random forest model between the full and select community OTUs
-only. B) Carcinoma random forest model between the full and select
-community OTUs only. The dotted line represents an AUC of 0.5 and the
-lines represent the range in which the AUC for the 100 different 80/20
-runs fell between. The solid red line represents the median AUC of all
-the studies for either the full or select community OTUS only model.
-
-\textbf{Figure 5: Tissue OTU Random Forest Model Across Studies.} A)
-Adenoma random forest model between the full and select community OTUs
-only. B) Carcinoma random forest model between the full and select
-community OTUs only. The dotted line represents an AUC of 0.5 and the
-lines represent the range in which the AUC for the 100 different 80/20
-runs fell between. The solid red line represents the median AUC of all
-the studies for either the full community or select genera OTUS only
-model.
-
-\textbf{Figure 6: Most Common Genera Across Full Community Stool Study
-Models.} A) Common genera in the top 10 for adenoma Random Forest genus
-models. B) Common genera in the top 10 for carcinoma Random Forest genus
-models. C) Common genera in the top 10 for adenoma Random Forest OTU
-models. D) Common genera in the top 10 for carcinoma Random Forest OTU
-models.
-
-\textbf{Figure 7: Power and Effect Size Analysis of Studies Included.}
+\textbf{Figure 2: Stool Random Forest Model Train AUCs.} A) Adenoma
+random forest model AUCs between all genera, all OTU, and select model
+based on significant OR taxa. B) Carcinoma random forest model AUCs
+between all genera, all OTU, and select model based on significant OR
+taxa. The black line represents the median AUC for the respective group.
+If no values are present in the singificant OR taxa group then there
+were no significant taxa identified and no model was tested.
+
+\textbf{Figure 3: Stool Random Forest Genus-Based Model Test AUCs.} A)
+Test AUCs of adenoma models using all genera across study. B) Test AUCs
+of carcinoma models using all genera or significant OR taxa only. The
+black line represents the AUC at 0.5. The red lines represent the median
+AUC of all test AUCs for a specific study.
+
+\textbf{Figure 4: Most Common Taxa Across Full Community Stool Study
+Models.} A) Common taxa in the top 10 for adenoma Random Forest all
+genera-based models. B) Common taxa in the top 10 for carcinoma Random
+Forest all genera-based models. C) Common taxa in the top 10 for adenoma
+Random Forest all OTU-based models. D) Common genera in the top 10 for
+carcinoma Random Forest all OTU-based models.
+
+\textbf{Figure 5: Power and Effect Size Analysis of Studies Included.}
 A) Power based on effect size for studies with adenoma individuals. B)
@@
 
-\textbf{Figure S2: Stool Random Forest Genus Model AUC for each Study.}
-A) AUC of adenoma models using all genera or select genera only. B) AUC
-of carcinoma models using all genera or select genera only. The black
-line represents the median within each group.
-
-\textbf{Figure S3: Tissue Random Forest Genus Model AUC for each Study.}
-A) AUC of adenoma models using all genera or only select genera divided
-between matched and unmatched tissue. B) AUC of carcinoma models using
-all genera or select genera only. The black line represents the median
-within each group divided between matched and unmatched tissue.
-
-\textbf{Figure S4: Stool Random Forest Prediction Success Using Genera
-Across Studies.} A) AUC for prediction in adenoma using all genera or
-select genera only. B) AUC for prediction in carcinoma using all genera
-or select genera only. The dotted line represents an AUC of 0.5. The
-x-axis is the data set in which the model was initially trained on. The
-red lines represent the median AUC using that specific study as the
-training set.
-
-\textbf{Figure S5: Tissue Random Forest Prediction Success of Carcinoma
-Using Genera Across Studies.} A) AUC for prediction in unmatched tissue
-for all genera or select genera only. B) AUC for prediction in matched
-tissue using all genera or select genera only. The dotted line
-represents an AUC of 0.5. The x-axis is the data set in which the model
-was initially trained on. The red lines represent the median AUC using
-that specific study as the training set.
-
-\textbf{Figure S6: Tissue Random Forest Prediction Success of Adenoma
-Using Genera Across Studies.} The red lines represent the median AUC
-using that specific study as the training set.
-
-\textbf{Figure S7: Most Common Genera Across Full Community Tissue Study
-Models.} A) Common genera in the top 10 for adenoma Random Forest genus
-models. B) Common genera in the top 10 for unmatched carcinoma Random
-Forest genus models. B) Common genera in the top 10 for matched
-carcinoma Random Forest genus models. D) Common genera in the top 10 for
-adenoma Random Forest OTU models. E) Common genera in the top 10 for
-unmatched carcinoma Random Forest OTU models. F) Common genera in the
-top 10 for matched carcinoma Random Forest OTU models.
+\textbf{Figure S2: Tissue Random Forest Model Train AUCs.} A) Adenoma
+random forest model AUCs between all genera, all OTU, and select model
+based on significant OR taxa in unmatched and matched tissue. B)
+Carcinoma random forest model AUCs between all genera, all OTU, and
+select model based on significant OR taxa in unmatched and matched
+tissue. The black line represents the median AUC for the respective
+group. If no values are present in the singificant OR taxa group then
+there were no significant taxa identified and no model was tested.
+
+\textbf{Figure S3: Tissue Random Forest Genus-Based Model Test AUCs.} A)
+Test AUCs of adenoma models using all genera across study. B) Test AUCs
+of carcinoma models using all genera for matched tissue studies. C) Test
+AUCs of carcinoma models using all genera or significant OR taxa only
+for unmatched tissue studies The black line represents the AUC at 0.5.
+The red lines represent the median AUC of all test AUCs for a specific
+study.
+
+\textbf{Figure S4: Most Common Genera Across Full Community Tissue Study
+Models.} A) Common genera in the top 10 for adenoma Random Forest for
+all genera-based models. B) Common genera in the top 10 for unmatched
+carcinoma Random Forest for all genera-based models. B) Common genera in
+the top 10 for matched carcinoma Random Forest for all genera-based
+models. D) Common genera in the top 10 for adenoma Random Forest for all
+OTU-based models. E) Common genera in the top 10 for unmatched carcinoma
+Random Forest for all OTU-based models. F) Common genera in the top 10
+for matched carcinoma Random Forest for all OTU-based models.
 

### 2018-02-14 submission/manuscript.Rmd
update results to bring in line with upated fig legend

@@
 Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity.
-In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A].
+In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively).
 We next tested whether the detectable differences in community significantly increased in OR of having an adenoma or carcinoma.
-For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased RR for carcinoma  (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the OR for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2].
-Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1 & S2].
+For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased OR for carcinoma  (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the OR for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 1].
+Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1].
 
 Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and 16S variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased.
-For colon tissue, there were no significant changes in OTU richness, evenness, or Shannon diversity as disease severity progressed from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure 1B & C].
-We next analyzed the RR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) colon tissue samples.
-For individuals at either an adenoma or carcinoma stage of disease there was no significant change in RR based on lower than median values for OTU richness, evenness, and Shannon diversity [Table S3-S5].
-Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were identified in unmatched tissue samples, for those at either adenoma or carcinoma stage of CRC [Table S6 & S7].
-For studies with matched samples no differences in bacterial community were observed when assessed with PERMANOVA [Table S6 & S7].
+For colon tissue, there were no significant changes in OTU richness, evenness, or Shannon diversity as disease severity progressed from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
+We next analyzed the OR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) colon tissue samples.
+For individuals at either an adenoma or carcinoma stage of disease there was no significant change in OR based on lower than median values for OTU richness, evenness, and Shannon diversity [Figure S1 & Table S2].
+Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were identified in unmatched tissue samples, for those at either adenoma or carcinoma stage of CRC [Table S1].
+For studies with matched samples no differences in bacterial community were observed when assessed with PERMANOVA [Table S1].
 These tissue results suggest that the microbiota within an individual are similar to each other regardless of disease status.
@@
 To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in OR.
-Mouth-associated genera were commonly found in the top 5 genera associated with an increased OR of having an adenoma (*Porphyromonas* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* and *Parvimonas* [Figure 3D]) for both stool and colon tissue samples.
-Conversely, genera commonly associated with the gastrointestinal tract were correlated with a decreased OR for both adenoma and carcinoma for both stool and colon tissue samples [Figure 3].
-Even though mouth-associated genera were identified across disease stage, there was little direct overlap of the top 5 increased or decreased OR genera between both stages and sample site.
-
-When observing ORs for adenoma between genera from stool or colon tissue with a P-value less than 0.05 there was almost no overlap and when they were similar the OR was in opposite directions (e.g.
-*Lactococcus*) [Table S8 & S9].
-Many of the adenoma associated genera ORs with a P-value under 0.05 for colon tissue are also highly prevalent in contamination, specifically, *Novosphingobium*, *Pseudomonas*, and *Achromobacter* [Figure 3 & Table S8-S9].
-For carcinoma stage of disease, certain mouth-associated genera (*Fusobacterium*, *Parvimonas*) had an increased OR for both colon tissue and stool samples [Table S10 & S11].
-The genera with the highest increased OR for carcinoma in tissue was *Lepttrichia* while in stool it was *Peptostreptococcus* [Table S10 & S11].
+For both tissue and stool samples only ORs associated with an increase or decrease in carcioma tumors were significant after multiple comparison correction [Table S3 & S4].
+Mouth-associated genera significantly increased the ORs of carcinoma for stool samples and included *Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Table S3].
+Conversely, mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4].
+The significant OR taxa in tissue were *Dorea* (decreased OR) and *Weissella* (increased OR).
+Overall, there was little direct overlap in increased or decreased OR genera between both tumor type and sample site.
 
@@
 Since specific genera increased the OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera.
-We selected these genera based on their OR and P-value significance and used two approaches to test this question.
-The first approach used genus level data and tested for differences in AUC between all genera and selected genera.
-A single study was used for training the model prior to testing on all other studies and this was repeated for every study in the meta-analysis.
-The second approach used OTU level data and tested for a generalized decrease in the 10-fold cross validation (CV) model AUC which is a common approach used to guard against over-fitting.
-This was applied across study and the AUC of the all OTUs model was compared against the model that used only OTUs that taxonomically classified to selected genera.
+We selected these genera based on significance after multiple comparison correction.
+If no taxa were significant after multiple comparison correction then no model for that specific grouping (i.e.
+adenoma stool) was analyzed.
+We first tested three model AUCs.
+These models were created using Random Forest and where either all genera, all OTUs, or significant OR taxa only.
+Next, the all genera models and any significant OR taxa models were tested across all studies that were not used to train the model.
+For stool, all models used had similar AUCs [Figure 2].
+Although for adenoma and unmatched carcinoma this trend held, there were large differences in matched tissue based on whether all genera or OTUs were utilized [Figure S2].
+When analyzing the tests sets that were comprised of genera data from other studies, both the all genera and singificant OR taxa only models had a similar ability to detect individuals with carcinomas, for both stool and tissue samples [Figure 3-S3].
 
-For the first approach using the genera-based models, the training set median AUC for model classification was similar for both the full and select genera models, for both tissue and stool studies [Figure S2-S3].
-When analyzing the tests sets that were comprised of genera data from other studies, both models had a similar ability to detect individuals with adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6].
-Conversely, the second approach that used OTU-based models showed a slight decrease in median AUC between the full and select models [Figure 4 & 5].
+In stool, the most common genera in the top 10 most important variables, in the full community models using all genera-based models, were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 4].
+Regardless of sample type, mouth-associated genera were present in models for carcinomas [Figure 4B & D].
+Yet, none were present in the majority of studies and none were present in the adenoma models [Figure 4A & B].
+For the full community OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 4C & 4D].
 
-In stool, the most common genera in the top 10 most important variables, in the full community models using the first approach, were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B].
-Regardless of sample type, mouth-associated genera were present in models for the carcinoma stage of CRC [Figure 6A & B].
-Yet, none were present in the majority of studies and *Fusobacterium* was the only genus present in the adenoma stage of CRC [Figure 6A & B].
-For the second approach that utilized full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].
-
-Unlike the stool-based Random Forest models, the tissue-based models, for the full genera from the first approach, showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 most important model variables across study [Figure S7].
-The vast majority of the top 10 model variables for the genera- and OTU-based models using colon tissue tended to be study specific.
-Further, there was very little overlap in the top 10 important variables between adenoma and carcinoma stage models, regardless of whether colon tissue or stool was used  [Figure 6 & S7].
-This discordance between stool and colon tissue samples also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of matched versus unmatched colon tissue Random Forest models [Figure S7B-C & S7E-F].
-This suggests that either the colon tissue microbiota is study and person dependent or that kit and/or other types of contamination associated with low biomass samples may be skewing the results.
+Unlike the stool-based Random Forest models, the tissue-based models, for the full genera from the first approach, showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 most important model variables across study [Figure S4].
+The vast majority of the top 10 model variables for the full community genera- and OTU-based models using colon tissue tended to be study specific.
+Further, there was very little overlap in the top 10 important variables between adenoma and carcinoma stage models, regardless of whether colon tissue or stool was used  [Figure S4].
+This discordance between stool and colon tissue samples also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of matched versus unmatched colon tissue Random Forest models [Figure S4B-C & S4E-F].
+This suggests that either the colon tissue microbiota is study or person dependent, that kit and/or other types of contamination associated with low biomass samples may be skewing the results, or that multiple microbes could act as the inflammatory stimulus needed to exacerbate mutations.
 
@@
 Next, we assessed how much confidence should be placed in the reported outcomes from each individual study by calculating the ability to detect a difference (power) and sample size needed for small, medium, and large effect size differences between cases and controls.
-When assessing the power of each study at different effect sizes the majority of studies achieved 80% power to detect a 30% or greater difference between groups [Figure 7A & B].
-No study that we analyzed had the standard 80% power to detect an effect size difference equal to or below 10% [Figure 7A & B].
-In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for both the case and control arms [Figure 7C]
+When assessing the power of each study at different effect sizes the majority of studies achieved 80% power to detect a 30% or greater difference between groups [Figure 5A & B].
+No study that we analyzed had the standard 80% power to detect an effect size difference equal to or below 10% [Figure 5A & B].
+In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for both the case and control arms [Figure 5C]
 

### 2018-02-14 submission/manuscript.tex
update results to bring in line with upated fig legend

@@
 diversity as disease severity moved from control to adenoma to carcinoma
-(P-value = 0.025 and 0.043, respectively) {[}Figure 1A{]}. We next
-tested whether the detectable differences in community significantly
-increased in OR of having an adenoma or carcinoma. For fecal samples, a
-decrease versus the overall median in evenness resulted in a
-significantly increased RR for carcinoma (OR = 1.66 (1.2 - 2.3), P-value
-= 0.0021) and adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) while a
-decrease versus the overall median in Shannon diversity only increased
-the OR for carcinoma (OR = 1.61 (1.14 - 2.28), P-value = 0.0069)
-{[}Figure 2{]}. Using the Bray-Curtis distance metric and PERMANOVA, it
-was also possible to identify significant bacterial community changes,
-in specific studies, for both carcinoma-associated and
-adenoma-associated microbiota versus control {[}Table S1 \& S2{]}.
+(P-value = 0.025 and 0.043, respectively). We next tested whether the
+detectable differences in community significantly increased in OR of
+having an adenoma or carcinoma. For fecal samples, a decrease versus the
+overall median in evenness resulted in a significantly increased OR for
+carcinoma (OR = 1.66 (1.2 - 2.3), P-value = 0.0021) and adenoma (OR =
+1.3 (1.02 - 1.65), P-value = 0.035) while a decrease versus the overall
+median in Shannon diversity only increased the OR for carcinoma (OR =
+1.61 (1.14 - 2.28), P-value = 0.0069) {[}Figure 1{]}. Using the
+Bray-Curtis distance metric and PERMANOVA, it was also possible to
+identify significant bacterial community changes, in specific studies,
+for both carcinoma-associated and adenoma-associated microbiota versus
+control {[}Table S1{]}.
 
@@
 diversity as disease severity progressed from control to adenoma to
-carcinoma (P-value \textgreater{} 0.05) {[}Figure 1B \& C{]}. We next
-analyzed the RR, for matched (unaffected tissue and an adenoma or
-carcinoma from the same individual) and unmatched (control and adenoma
-or carcinoma tissue not from the same individual) colon tissue samples.
-For individuals at either an adenoma or carcinoma stage of disease there
-was no significant change in RR based on lower than median values for
-OTU richness, evenness, and Shannon diversity {[}Table S3-S5{]}. Similar
-to stool samples, significant differences in bacterial community,
-assessed by PERMANOVA, were identified in unmatched tissue samples, for
-those at either adenoma or carcinoma stage of CRC {[}Table S6 \& S7{]}.
-For studies with matched samples no differences in bacterial community
-were observed when assessed with PERMANOVA {[}Table S6 \& S7{]}. These
-tissue results suggest that the microbiota within an individual are
-similar to each other regardless of disease status.
+carcinoma (P-value \textgreater{} 0.05). We next analyzed the OR, for
+matched (unaffected tissue and an adenoma or carcinoma from the same
+individual) and unmatched (control and adenoma or carcinoma tissue not
+from the same individual) colon tissue samples. For individuals at
+either an adenoma or carcinoma stage of disease there was no significant
+change in OR based on lower than median values for OTU richness,
+evenness, and Shannon diversity {[}Figure S1 \& Table S2{]}. Similar to
+stool samples, significant differences in bacterial community, assessed
+by PERMANOVA, were identified in unmatched tissue samples, for those at
+either adenoma or carcinoma stage of CRC {[}Table S1{]}. For studies
+with matched samples no differences in bacterial community were observed
+when assessed with PERMANOVA {[}Table S1{]}. These tissue results
+suggest that the microbiota within an individual are similar to each
+other regardless of disease status.
 
@@
 abundance higher than the median results in an increase or decrease in
-OR. Mouth-associated genera were commonly found in the top 5 genera
-associated with an increased OR of having an adenoma
-(\emph{Porphyromonas} {[}Figure 3A{]} and \emph{Rothia} {[}Figure 3C{]})
-and carcinoma (\emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Figure 3B{]} and
-\emph{Fusobacterium} and \emph{Parvimonas} {[}Figure 3D{]}) for both
-stool and colon tissue samples. Conversely, genera commonly associated
-with the gastrointestinal tract were correlated with a decreased OR for
-both adenoma and carcinoma for both stool and colon tissue samples
-{[}Figure 3{]}. Even though mouth-associated genera were identified
-across disease stage, there was little direct overlap of the top 5
-increased or decreased OR genera between both stages and sample site.
-
-When observing ORs for adenoma between genera from stool or colon tissue
-with a P-value less than 0.05 there was almost no overlap and when they
-were similar the OR was in opposite directions (e.g. \emph{Lactococcus})
-{[}Table S8 \& S9{]}. Many of the adenoma associated genera ORs with a
-P-value under 0.05 for colon tissue are also highly prevalent in
-contamination, specifically, \emph{Novosphingobium}, \emph{Pseudomonas},
-and \emph{Achromobacter} {[}Figure 3 \& Table S8-S9{]}. For carcinoma
-stage of disease, certain mouth-associated genera (\emph{Fusobacterium},
-\emph{Parvimonas}) had an increased OR for both colon tissue and stool
-samples {[}Table S10 \& S11{]}. The genera with the highest increased OR
-for carcinoma in tissue was \emph{Lepttrichia} while in stool it was
-\emph{Peptostreptococcus} {[}Table S10 \& S11{]}.
+OR. For both tissue and stool samples only ORs associated with an
+increase or decrease in carcioma tumors were significant after multiple
+comparison correction {[}Table S3 \& S4{]}. Mouth-associated genera
+significantly increased the ORs of carcinoma for stool samples and
+included \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Porphyromonas},
+and \emph{Peptostreptococcus} {[}Table S3{]}. Conversely,
+mouth-associated genera were not significantly associated with an
+increased OR for carcinoma in tissue samples {[}Table S4{]}. The
+significant OR taxa in tissue were \emph{Dorea} (decreased OR) and
+\emph{Weissella} (increased OR). Overall, there was little direct
+overlap in increased or decreased OR genera between both tumor type and
+sample site.
 
@@
 at classifying disease versus only a select group of genera. We selected
-these genera based on their OR and P-value significance and used two
-approaches to test this question. The first approach used genus level
-data and tested for differences in AUC between all genera and selected
-genera. A single study was used for training the model prior to testing
-on all other studies and this was repeated for every study in the
-meta-analysis. The second approach used OTU level data and tested for a
-generalized decrease in the 10-fold cross validation (CV) model AUC
-which is a common approach used to guard against over-fitting. This was
-applied across study and the AUC of the all OTUs model was compared
-against the model that used only OTUs that taxonomically classified to
-selected genera.
-
-For the first approach using the genera-based models, the training set
-median AUC for model classification was similar for both the full and
-select genera models, for both tissue and stool studies {[}Figure
-S2-S3{]}. When analyzing the tests sets that were comprised of genera
-data from other studies, both models had a similar ability to detect
-individuals with adenomas or carcinomas, with the select genera models
-performing better in some instances {[}Figure S4-S6{]}. Conversely, the
-second approach that used OTU-based models showed a slight decrease in
-median AUC between the full and select models {[}Figure 4 \& 5{]}.
+these genera based on significance after multiple comparison correction.
+If no taxa were significant after multiple comparison correction then no
+model for that specific grouping (i.e.~adenoma stool) was analyzed. We
+first tested three model AUCs. These models were created using Random
+Forest and where either all genera, all OTUs, or significant OR taxa
+only. Next, the all genera models and any significant OR taxa models
+were tested across all studies that were not used to train the model.
+For stool, all models used had similar AUCs {[}Figure 2{]}. Although for
+adenoma and unmatched carcinoma this trend held, there were large
+differences in matched tissue based on whether all genera or OTUs were
+utilized {[}Figure S2{]}. When analyzing the tests sets that were
+comprised of genera data from other studies, both the all genera and
+singificant OR taxa only models had a similar ability to detect
+individuals with carcinomas, for both stool and tissue samples {[}Figure
+3-S3{]}.
 
 In stool, the most common genera in the top 10 most important variables,
-in the full community models using the first approach, were
+in the full community models using all genera-based models, were
 \emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} {[}Figure
-6A \& B{]}. Regardless of sample type, mouth-associated genera were
-present in models for the carcinoma stage of CRC {[}Figure 6A \& B{]}.
-Yet, none were present in the majority of studies and
-\emph{Fusobacterium} was the only genus present in the adenoma stage of
-CRC {[}Figure 6A \& B{]}. For the second approach that utilized full
-OTU-based models, \emph{Ruminococcaceae} was present in the top 10
-consistently for both adenoma and carcinoma models while
-\emph{Roseburia} was only present in many adenoma models and
-\emph{Bacteroides} was present in the overwhelming majority of the
-carcinoma models {[}Figure 6C \& 6D{]}.
+4{]}. Regardless of sample type, mouth-associated genera were present in
+models for carcinomas {[}Figure 4B \& D{]}. Yet, none were present in
+the majority of studies and none were present in the adenoma models
+{[}Figure 4A \& B{]}. For the full community OTU-based models,
+\emph{Ruminococcaceae} was present in the top 10 consistently for both
+adenoma and carcinoma models while \emph{Roseburia} was only present in
+many adenoma models and \emph{Bacteroides} was present in the
+overwhelming majority of the carcinoma models {[}Figure 4C \& 4D{]}.
 
@@
 \emph{Bacteroides}, and \emph{Roseburia} in the top 10 most important
-model variables across study {[}Figure S7{]}. The vast majority of the
-top 10 model variables for the genera- and OTU-based models using colon
-tissue tended to be study specific. Further, there was very little
-overlap in the top 10 important variables between adenoma and carcinoma
-stage models, regardless of whether colon tissue or stool was used
-{[}Figure 6 \& S7{]}. This discordance between stool and colon tissue
-samples also applies to the mouth-associated genera with one noticeable
-skew being that \emph{Fusobacterium} and \emph{Fusobacteriaceae} occur
-more often in the top 10 of matched versus unmatched colon tissue Random
-Forest models {[}Figure S7B-C \& S7E-F{]}. This suggests that either the
-colon tissue microbiota is study and person dependent or that kit and/or
-other types of contamination associated with low biomass samples may be
-skewing the results.
+model variables across study {[}Figure S4{]}. The vast majority of the
+top 10 model variables for the full community genera- and OTU-based
+models using colon tissue tended to be study specific. Further, there
+was very little overlap in the top 10 important variables between
+adenoma and carcinoma stage models, regardless of whether colon tissue
+or stool was used {[}Figure S4{]}. This discordance between stool and
+colon tissue samples also applies to the mouth-associated genera with
+one noticeable skew being that \emph{Fusobacterium} and
+\emph{Fusobacteriaceae} occur more often in the top 10 of matched versus
+unmatched colon tissue Random Forest models {[}Figure S4B-C \& S4E-F{]}.
+This suggests that either the colon tissue microbiota is study or person
+dependent, that kit and/or other types of contamination associated with
+low biomass samples may be skewing the results, or that multiple
+microbes could act as the inflammatory stimulus needed to exacerbate
+mutations.
 
@@
 of studies achieved 80\% power to detect a 30\% or greater difference
-between groups {[}Figure 7A \& B{]}. No study that we analyzed had the
+between groups {[}Figure 5A \& B{]}. No study that we analyzed had the
 standard 80\% power to detect an effect size difference equal to or
-below 10\% {[}Figure 7A \& B{]}. In order to achieve a power of 80\%,
+below 10\% {[}Figure 5A \& B{]}. In order to achieve a power of 80\%,
 for small effect sizes, studies used in our meta-analysis would need to
 recruit over 1000 individuals for both the case and control arms
-{[}Figure 7C{]}
+{[}Figure 5C{]}
 

### 2018-02-14 submission/supplemental.Rmd
update results to bring in line with upated fig legend

@@
 
-**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stooland Tissue**
+**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool and Tissue**
 
@@
 
-**Table S3: OR of Tumor in Stool by Individual Genera with Initial P-Value under 0.05**
+**Table S3: OR in Stool of Tumor by Individual Genera with Initial P-Value under 0.05**
 
@@
 
-**Table S4: OR of Tumor in Tissue by Individual Genera with Initial P-Value under 0.05**
+**Table S4: OR in Tissue of Tumor by Individual Genera with Initial P-Value under 0.05**
 

### 2018-02-14 submission/manuscript.Rmd
update to include relevent OR numbers

@@
 #Load in odds ratio select genera data tables
-select_RR_stool <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_stool_composite.csv"))
-select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_tissue_composite.csv"))
+crc_OR_stool <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_stool_composite.csv") %>% 
+  mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(bh < 0.05))
 
-adn_select_RR_stool <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_OR_stool_composite.csv"))
-adn_select_RR_tissue <- as.data.frame(read_csv("../data/process/tables/adn_select_genus_OR_tissue_composite.csv"))
+crc_unmatched_OR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_unmatched_tissue_composite.csv") %>% mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(bh < 0.05))
 
@@
 For both tissue and stool samples only ORs associated with an increase or decrease in carcioma tumors were significant after multiple comparison correction [Table S3 & S4].
-Mouth-associated genera significantly increased the ORs of carcinoma for stool samples and included *Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Table S3].
+Out of the `r length(crc_OR_stool$measure)` taxa that had significant ORs in stool samples `r length(filter(crc_OR_stool, measure %in% c("Fusobacterium", "Parvimonas", "Porphyromonas", "Peptostreptococcus"))[, "measure"])` were mouth-associated microbes.
+These mouth-associated genera significantly increased the ORs of carcinoma for stool samples and included *Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Table S3].
 Conversely, mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4].
-The significant OR taxa in tissue were *Dorea* (decreased OR) and *Weissella* (increased OR).
+Only unmatched tissue samples had significant OR taxa and these were *Dorea* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2)`)) and *Weissella* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2)`)).
 Overall, there was little direct overlap in increased or decreased OR genera between both tumor type and sample site.

### 2018-02-14 submission/manuscript.tex
update to include relevent OR numbers

@@
 increase or decrease in carcioma tumors were significant after multiple
-comparison correction {[}Table S3 \& S4{]}. Mouth-associated genera
-significantly increased the ORs of carcinoma for stool samples and
-included \emph{Fusobacterium}, \emph{Parvimonas}, \emph{Porphyromonas},
-and \emph{Peptostreptococcus} {[}Table S3{]}. Conversely,
-mouth-associated genera were not significantly associated with an
-increased OR for carcinoma in tissue samples {[}Table S4{]}. The
-significant OR taxa in tissue were \emph{Dorea} (decreased OR) and
-\emph{Weissella} (increased OR). Overall, there was little direct
-overlap in increased or decreased OR genera between both tumor type and
-sample site.
+comparison correction {[}Table S3 \& S4{]}. Out of the 8 taxa that had
+significant ORs in stool samples 4 were mouth-associated microbes. These
+mouth-associated genera significantly increased the ORs of carcinoma for
+stool samples and included \emph{Fusobacterium}, \emph{Parvimonas},
+\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Table S3{]}.
+Conversely, mouth-associated genera were not significantly associated
+with an increased OR for carcinoma in tissue samples {[}Table S4{]}.
+Only unmatched tissue samples had significant OR taxa and these were
+\emph{Dorea} (OR = 0.35 (0.22 - 0.55)) and \emph{Weissella} (OR = 5.15
+(2.02 - 13.14)). Overall, there was little direct overlap in increased
+or decreased OR genera between both tumor type and sample site.
 

### 2018-02-14 submission/manuscript.Rmd
update fig calls in discussion

@@
 The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, tumor formation.
-The results that we have presented show that both the genera- and OTU-based select and full models, for indviduals with carcinoma, had similar AUCs [Figure 4 & 5].
+The results that we have presented show that both the genera- and OTU-based select and full models, for indviduals with carcinoma, had similar AUCs [Figure 2 & 3].
 This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community could be crucial for carcinoma formation.
@@
 Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly.
-Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly [Figure 1].
+Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly.
 This change in diversity, at this early stage of disease, could be focal to the adenoma itself.
@@
 
-Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 6].
-Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 6].
+Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 4].
+Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 4].
 These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community.
@@
 The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors.
-Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S7E & F].
-Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S7B-C & E-F].
+Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S4E & F].
+Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S4B-C & E-F].
 The majority of the colon tissue-based results seem to be study specific with many of the top 10 taxa being present only in a single study.
@@
 Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation.
-In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 7].
+In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 5].
 

### 2018-02-14 submission/manuscript.tex
update fig calls in discussion

@@
 and full models, for indviduals with carcinoma, had similar AUCs
-{[}Figure 4 \& 5{]}. This suggests that an interplay between a select
+{[}Figure 2 \& 3{]}. This suggests that an interplay between a select
 number of potentially protective and exacerbating microbes within the GI
@@
 performed poorly. Yet, the step-wise decrease in diversity suggests that
-the adenoma-associated community is not normal but has changed subtly
-{[}Figure 1{]}. This change in diversity, at this early stage of
-disease, could be focal to the adenoma itself. Additionally, how the
-host interacts with these subtle changes at early stages of the disease
-could be what leads to a thoroughly dysfunctional community that is
-supportive of tumorgenesis.
+the adenoma-associated community is not normal but has changed subtly.
+This change in diversity, at this early stage of disease, could be focal
+to the adenoma itself. Additionally, how the host interacts with these
+subtle changes at early stages of the disease could be what leads to a
+thoroughly dysfunctional community that is supportive of tumorgenesis.
 
 Within stool, common GI microbes were most consistently present in the
-top 10 genera or OTUs across studies {[}Figure 6{]}. Changes in
+top 10 genera or OTUs across studies {[}Figure 4{]}. Changes in
 \emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
@@
 important variables across the different studies for both individuals
-with adenoma and carcinoma {[}Figure 6{]}. These data suggest that
+with adenoma and carcinoma {[}Figure 4{]}. These data suggest that
 whether the non-resident bacterium is \emph{Fusobacteria} or
@@
 most prevalent in the top 10 most important variables across study
-{[}Figure S7E \& F{]}. Unlike in stool, \emph{Fusobacterium} was the
+{[}Figure S4E \& F{]}. Unlike in stool, \emph{Fusobacterium} was the
 only mouth-associated bacteria consistently present in the top 10 most
-important variables of the full carcinoma stage models {[}Figure S7B-C
+important variables of the full carcinoma stage models {[}Figure S4B-C
 \& E-F{]}. The majority of the colon tissue-based results seem to be
@@
 because of large effect size differences between cases and controls
-{[}Figure 7{]}.
+{[}Figure 5{]}.
 

### 2018-02-15 submission/manuscript.Rmd
update discussion section

@@
 
-Our study identifies clear differences in diversity, both at the community level and for individual genera, present in patients with tumors versus controls [Figure 1-3].
+Our study identifies clear but small differences in diversity at the community level and larger differences for individual genera, present in patients with tumors versus controls [Figure 1-3].
 Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect size increases in OR for either adenoma or carcinoma tumors.
-Even though mouth-associated genera increased individaul's OR of having a carcinoma, they did not consistently increase the OR of having an adenoma.
-Additionally, our observations suggest that by combining the top 5 most protective and detrimental microbes we can classify indviduals with either adenoma or carcinoma as well as models that use the full community.
+Even though mouth-associated genera increased individaul's OR of having a carcinoma for certain sample types, they did not consistently increase the OR of having an adenoma.
+By using these taxa that had significant ORs after multiple comparison correction we found that we could classify indviduals with either adenoma or carcinoma as well as models that use either all genera or all OTUs.
+Finally, many studies were individually under powered to be able to reject the null hypothesis and this could one reason only the comparison between control and carcinoma individuals for stool samples had relible detectable differences.
 
-The data presented herein support the importance of select genera for carcinoma, but not necessarily adenoma, tumor formation.
-The results that we have presented show that both the genera- and OTU-based select and full models, for indviduals with carcinoma, had similar AUCs [Figure 2 & 3].
+The data presented herein support the importance of specific taxa for carcinoma, but not necessarily adenoma, tumor formation.
+The results that we have presented show that the significant OR taxa model and both the full genera and OTU models, for indviduals with carcinoma, had similar AUCs [Figure 2 & 3].
 This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community could be crucial for carcinoma formation.
 Importantly, it suggests that there may be key members of the GI community that should be studied further to potentially help reduce the risk of carcinoma tumor formation.
-Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma turmors.
-Regardless of sample type and whether a full or select model was used, our Random Forest models consistently performed poorly.
+Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma tumors.
+Regardless of sample type and whether a full genera- or OTU-based model was used, our Random Forest models consistently performed poorly.
 Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly.
 This change in diversity, at this early stage of disease, could be focal to the adenoma itself.
-Additionally, how the host interacts with these subtle changes at early stages of the disease could be what leads to a thoroughly dysfunctional community that is supportive of tumorgenesis.
+How the host interacts with these subtle changes at early stages of the disease could be what leads to a thoroughly dysfunctional community that is supportive of tumorgenesis.
 
-Within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 4].
+For the full genera- and OTU-based models within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 4].
 Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 4].
 These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community.
-Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.
+Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated or general inflammatory genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.
 
@@
 Additionally, *Streptococcus gallolyticus* subsp.
-*gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased OR for individuals to have either an adenoma or carcinoma.
+*gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased OR for individuals to have a carcinoma tumor.
 
-The associations between the microbiota and indivdials with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes.
+The associations between the microbiota and individuals with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes.
 None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls.
 The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas.
-Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to polyp formation.
+Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to adenoma formation.
 In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 5].
@@
 Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma tumors due to possible subtle changes in the community.
-Overall, associations between the microbiota and individuals with carcinomas were much stronger than with those with adenomas.
+Overall, the associations between the microbiota and individuals with carcinomas were much stronger than with those with adenomas.
 

### 2018-02-15 submission/manuscript.tex
update discussion section

@@
 
-Our study identifies clear differences in diversity, both at the
-community level and for individual genera, present in patients with
-tumors versus controls {[}Figure 1-3{]}. Although there was a step-wise
-decrease in diversity as disease progressed from control to adenoma to
-carcinoma, this did not translate into large effect size increases in OR
-for either adenoma or carcinoma tumors. Even though mouth-associated
-genera increased individaul's OR of having a carcinoma, they did not
-consistently increase the OR of having an adenoma. Additionally, our
-observations suggest that by combining the top 5 most protective and
-detrimental microbes we can classify indviduals with either adenoma or
-carcinoma as well as models that use the full community.
-
-The data presented herein support the importance of select genera for
+Our study identifies clear but small differences in diversity at the
+community level and larger differences for individual genera, present in
+patients with tumors versus controls {[}Figure 1-3{]}. Although there
+was a step-wise decrease in diversity as disease progressed from control
+to adenoma to carcinoma, this did not translate into large effect size
+increases in OR for either adenoma or carcinoma tumors. Even though
+mouth-associated genera increased individaul's OR of having a carcinoma
+for certain sample types, they did not consistently increase the OR of
+having an adenoma. By using these taxa that had significant ORs after
+multiple comparison correction we found that we could classify
+indviduals with either adenoma or carcinoma as well as models that use
+either all genera or all OTUs. Finally, many studies were individually
+under powered to be able to reject the null hypothesis and this could
+one reason only the comparison between control and carcinoma individuals
+for stool samples had relible detectable differences.
+
+The data presented herein support the importance of specific taxa for
 carcinoma, but not necessarily adenoma, tumor formation. The results
-that we have presented show that both the genera- and OTU-based select
-and full models, for indviduals with carcinoma, had similar AUCs
-{[}Figure 2 \& 3{]}. This suggests that an interplay between a select
-number of potentially protective and exacerbating microbes within the GI
-community could be crucial for carcinoma formation. Importantly, it
-suggests that there may be key members of the GI community that should
-be studied further to potentially help reduce the risk of carcinoma
-tumor formation. Conversely, using the present data, it is clear that
-new approaches may be needed to identify members of the community
-associated with adenoma turmors. Regardless of sample type and whether a
-full or select model was used, our Random Forest models consistently
-performed poorly. Yet, the step-wise decrease in diversity suggests that
-the adenoma-associated community is not normal but has changed subtly.
-This change in diversity, at this early stage of disease, could be focal
-to the adenoma itself. Additionally, how the host interacts with these
-subtle changes at early stages of the disease could be what leads to a
-thoroughly dysfunctional community that is supportive of tumorgenesis.
-
-Within stool, common GI microbes were most consistently present in the
-top 10 genera or OTUs across studies {[}Figure 4{]}. Changes in
-\emph{Bacteroides}, \emph{Ruminococcaceae}, \emph{Ruminococcus}, and
-\emph{Roseburia} were consistently found to be in the top 10 most
-important variables across the different studies for both individuals
-with adenoma and carcinoma {[}Figure 4{]}. These data suggest that
-whether the non-resident bacterium is \emph{Fusobacteria} or
-\emph{Peptostreptococcus} may not be as important as how these bacteria
-interact with the changing resident community. Based on these
-observations, it is possible to hypothesize that small changes in
-community structure lead to new niches in which any one of the
-mouth-associated genera can gain a foothold, exacerbating the initial
-changes in community and facilitating the transition from adenoma to
-carcinoma stage of disease.
+that we have presented show that the significant OR taxa model and both
+the full genera and OTU models, for indviduals with carcinoma, had
+similar AUCs {[}Figure 2 \& 3{]}. This suggests that an interplay
+between a select number of potentially protective and exacerbating
+microbes within the GI community could be crucial for carcinoma
+formation. Importantly, it suggests that there may be key members of the
+GI community that should be studied further to potentially help reduce
+the risk of carcinoma tumor formation. Conversely, using the present
+data, it is clear that new approaches may be needed to identify members
+of the community associated with adenoma tumors. Regardless of sample
+type and whether a full genera- or OTU-based model was used, our Random
+Forest models consistently performed poorly. Yet, the step-wise decrease
+in diversity suggests that the adenoma-associated community is not
+normal but has changed subtly. This change in diversity, at this early
+stage of disease, could be focal to the adenoma itself. How the host
+interacts with these subtle changes at early stages of the disease could
+be what leads to a thoroughly dysfunctional community that is supportive
+of tumorgenesis.
+
+For the full genera- and OTU-based models within stool, common GI
+microbes were most consistently present in the top 10 genera or OTUs
+across studies {[}Figure 4{]}. Changes in \emph{Bacteroides},
+\emph{Ruminococcaceae}, \emph{Ruminococcus}, and \emph{Roseburia} were
+consistently found to be in the top 10 most important variables across
+the different studies for both individuals with adenoma and carcinoma
+{[}Figure 4{]}. These data suggest that whether the non-resident
+bacterium is \emph{Fusobacteria} or \emph{Peptostreptococcus} may not be
+as important as how these bacteria interact with the changing resident
+community. Based on these observations, it is possible to hypothesize
+that small changes in community structure lead to new niches in which
+any one of the mouth-associated or general inflammatory genera can gain
+a foothold, exacerbating the initial changes in community and
+facilitating the transition from adenoma to carcinoma stage of disease.
 
@@
 genera are commonly associated with an increased OR for individuals to
-have either an adenoma or carcinoma.
-
-The associations between the microbiota and indivdials with adenoma
+have a carcinoma tumor.
+
+The associations between the microbiota and individuals with adenoma
 tumors are inconclusive, in part, because many studies may not be
@@
 power into consideration to reproducibly study whether the microbiota
-contributes to polyp formation. In contrast to adenoma stage of disease,
-our observations suggest that most studies analyzed have sufficient
-power to detect many changes in the carcinoma-associated microbiota
-because of large effect size differences between cases and controls
-{[}Figure 5{]}.
+contributes to adenoma formation. In contrast to adenoma stage of
+disease, our observations suggest that most studies analyzed have
+sufficient power to detect many changes in the carcinoma-associated
+microbiota because of large effect size differences between cases and
+controls {[}Figure 5{]}.
 
@@
 the microbiota and adenoma tumors due to possible subtle changes in the
-community. Overall, associations between the microbiota and individuals
-with carcinomas were much stronger than with those with adenomas.
+community. Overall, the associations between the microbiota and
+individuals with carcinomas were much stronger than with those with
+adenomas.
 

### 2018-02-15 submission/manuscript.Rmd
update methods section

@@
 For each individual study the Bray-Curtis index was used to assess differences between control-adenoma and control-carcinoma individuals.
-Next, all common genera were assessed for differences in ORs for individuals having an adenoma or carcinoma and ranked based on P-value.
-We then built Random Forest models based on the full or selected community (the top 5 increased and top 5 decreased ORs based on P-value).
-Comparison between the full and selected models took two different approaches.
-In the first approach, models were trained on one study then tested on the remaining studies using genera-based relative abundances.
+Next, all common genera were assessed for differences in ORs for individuals having an adenoma or carcinoma and corrected for multiple comparisons using the Benjamini-Hochberg method [@benjamini_controlling_1995].
+We then built Random Forest models based on all genera, all OTUs, or significant OR taxa (if any were present after multiple comparison correction).
+For both the full genera and significant OR taxa, models were trained on one study then tested on the remaining studies using genera-based relative abundances.
+The OTU-based models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
 This process was repeated for every study in the meta-analysis.
-In the second approach, models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
-For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased OR analysis based on P-value.
+Comparisons of the initial trained model AUCs for the full genera and significant OR taxa were made to the mean AUC generated from the 100 different 10-fold CV runs of the respective OTU-based model.
 Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power.
@@
 
-***Obtaining Genera Relative Abundance and Selected Models:***
+***Obtaining Genera Relative Abundance and Significant OR Taxa Models:***
 For the genera analysis of the ORs, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings was obtained.
-The select models for the first Random Forest model approach utilized the top 5 most significantly increased and decreased ORs.
-For this approach, the full community models for each study were built by utilizing all genera and lowest taxonomic groups identified within that particular study.
-The select models for the second approach, that used OTU-based Random Forest models, utilized a similar method as the first approach.
-The main difference was that any OTU that taxonomically classified to one of the taxa in the top 5 increased or decreased OR analysis were included in the select model.
-OTU Random Forest models, using the full community, included all OTUs that did not have near zero variance within that specific study.
+The significant OR taxa models for the  Random Forest models utilized all taxa that had significant ORs after multiple comparison correction.
+This meant only models for the carcinoma stool (`r length(crc_OR_stool$measure)` variables) and carcinoma unmatched (`r length(crc_unmatched_OR_tissue$measure)` variables) samples were possible to be created and analyzed.
 

### 2018-02-15 submission/manuscript.tex
update methods section

@@
 individuals. Next, all common genera were assessed for differences in
-ORs for individuals having an adenoma or carcinoma and ranked based on
-P-value. We then built Random Forest models based on the full or
-selected community (the top 5 increased and top 5 decreased ORs based on
-P-value). Comparison between the full and selected models took two
-different approaches. In the first approach, models were trained on one
-study then tested on the remaining studies using genera-based relative
-abundances. This process was repeated for every study in the
-meta-analysis. In the second approach, models were built using OTU level
-data and a 10-fold CV over 100 different iterations, based on random
-80/20 splitting of the data, was used to generate a range of expected
-AUCs. For these OTU-based models, the selected model included all OTUs
-that had a taxonomic classification to a taxa in the top 5 increased and
-top 5 decreased OR analysis based on P-value. Finally, the power of each
-study was assessed for an effect size ranging from 1\% to 30\% and an
-estimated sample size, for these effect sizes, was generated based on
-80\% power. For comparisons in which only control versus adenoma
-individuals were made, the carcinoma individuals were excluded from each
-respective study. Similarly, for comparisons in which control versus
-carcinoma individuals were made the adenoma individuals were excluded
-from each respective study. For all analysis completed fecal and tissue
-samples were kept separate. Within the tissue groups the data were
-further divided between samples from the same individual (matched) and
-those from different individuals (unmatched).
-
-\textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
-For the genera analysis of the ORs, OTUs were added together based on
-the genus or lowest available taxonomic classification level and the
-total average counts, for 100 different subsamplings was obtained. The
-select models for the first Random Forest model approach utilized the
-top 5 most significantly increased and decreased ORs. For this approach,
-the full community models for each study were built by utilizing all
-genera and lowest taxonomic groups identified within that particular
-study. The select models for the second approach, that used OTU-based
-Random Forest models, utilized a similar method as the first approach.
-The main difference was that any OTU that taxonomically classified to
-one of the taxa in the top 5 increased or decreased OR analysis were
-included in the select model. OTU Random Forest models, using the full
-community, included all OTUs that did not have near zero variance within
-that specific study.
+ORs for individuals having an adenoma or carcinoma and corrected for
+multiple comparisons using the Benjamini-Hochberg method {[}51{]}. We
+then built Random Forest models based on all genera, all OTUs, or
+significant OR taxa (if any were present after multiple comparison
+correction). For both the full genera and significant OR taxa, models
+were trained on one study then tested on the remaining studies using
+genera-based relative abundances. The OTU-based models were built using
+OTU level data and a 10-fold CV over 100 different iterations, based on
+random 80/20 splitting of the data, was used to generate a range of
+expected AUCs. This process was repeated for every study in the
+meta-analysis. Comparisons of the initial trained model AUCs for the
+full genera and significant OR taxa were made to the mean AUC generated
+from the 100 different 10-fold CV runs of the respective OTU-based
+model. Finally, the power of each study was assessed for an effect size
+ranging from 1\% to 30\% and an estimated sample size, for these effect
+sizes, was generated based on 80\% power. For comparisons in which only
+control versus adenoma individuals were made, the carcinoma individuals
+were excluded from each respective study. Similarly, for comparisons in
+which control versus carcinoma individuals were made the adenoma
+individuals were excluded from each respective study. For all analysis
+completed fecal and tissue samples were kept separate. Within the tissue
+groups the data were further divided between samples from the same
+individual (matched) and those from different individuals (unmatched).
+
+\textbf{\emph{Obtaining Genera Relative Abundance and Significant OR
+Taxa Models:}} For the genera analysis of the ORs, OTUs were added
+together based on the genus or lowest available taxonomic classification
+level and the total average counts, for 100 different subsamplings was
+obtained. The significant OR taxa models for the Random Forest models
+utilized all taxa that had significant ORs after multiple comparison
+correction. This meant only models for the carcinoma stool (8 variables)
+and carcinoma unmatched (3 variables) samples were possible to be
+created and analyzed.
 
@@
 
+\hypertarget{ref-benjamini_controlling_1995}{}
+51. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+practical and powerful approach to multiple testing. Journal of the
+Royal Statistical Society Series B (Methodological). 1995;57:289--300.
+
 \newpage

### 2018-02-15 submission/manuscript.Rmd
update sig OR for unmatched tissue

@@
 Conversely, mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4].
-Only unmatched tissue samples had significant OR taxa and these were *Dorea* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2)`)) and *Weissella* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2)`)).
-Overall, there was little direct overlap in increased or decreased OR genera between both tumor type and sample site.
+Only unmatched tissue samples had significant OR taxa and these were *Dorea* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2)`)), *Blautia* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2)`)), and *Weissella* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2)`)).
+Overall, there was little direct overlap in increased or decreased OR taxa between both tumor type and sample site.
 

### 2018-02-15 submission/manuscript.tex
update sig OR for unmatched tissue

@@
 Only unmatched tissue samples had significant OR taxa and these were
-\emph{Dorea} (OR = 0.35 (0.22 - 0.55)) and \emph{Weissella} (OR = 5.15
-(2.02 - 13.14)). Overall, there was little direct overlap in increased
-or decreased OR genera between both tumor type and sample site.
+\emph{Dorea} (OR = 0.35 (0.22 - 0.55)), \emph{Blautia} (OR = 0.47 (0.3 -
+0.73)), and \emph{Weissella} (OR = 5.15 (2.02 - 13.14)). Overall, there
+was little direct overlap in increased or decreased OR taxa between both
+tumor type and sample site.
 

### 2018-02-19 submission/manuscript.Rmd
update abstract with edits

@@
 
-**Background.** An increasing body of literature suggests that there is a crucial role for the microbiota in colorectal cancer (CRC) pathogenesis.
-Important drivers within this context have ranged from individual microbes to the whole community.
-Our study expands on a recent meta-analysis investigating microbial biomarkers for tumors by testing the hypothesis that the bacterial community has important associations to both adenoma and carcinoma tumors.
-To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 previously published 16S rRNA gene sequencing studies on colorectal tumors and the microbiota.
+**Background.** An increasing body of literature suggests that both individual and collections of bacteria are associated with the progression of colorectal cancer.
+As the number of studies investigating these associations increases and the number of subjects in each study increases, a meta-analysis to identify the associations that are the most predictive of disease progression is warranted.
+For our meta-analysis, we analyzed previously published 16S rRNA gene sequencing data collected from feces (`r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` individuals from `r length(stool_n$study)` studies) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals from `r length(tissue_n$study)` studies).
 
-**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness as tumor severity increased, after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05).
-This reduction in evenness translated into small increases in the odds ratio for individuals to have both adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma tumors (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for individuals to have carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`).
-Increases in mouth-associated microbes were commonly in the top 5 most increased odds ratios for individuals to have either adenoma or carcinoma tumors, regardless of sample type.
-Prediction models built to classify either individuals with adenoma or carcinoma were trained on the whole community or selected genera (top 5 highest and lowest odds ratios) from either fecal or tissue samples.
-Both the full and select models for either adenoma or carcinoma resulted in similar classification success according to Area Under the Curve (AUC).
-The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies.
-Although a number of associations between the microbiota and tumor were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.
+**Results.** We quantified the odds ratios for individual bacterial genera that were associated with an individual having tumors relative to a normal colon.
+Among the stool samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_OR_stool$measure)` genera with significant odds ratios associated with carcinoma.
+Similarly, among the tissue samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` genera with significant odds ratios associated with carcinoma.
+Among the significant odds ratios, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`.
+Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using the genera with the five highest and lowest odds ratios, using the entire collection of genera found in each study, and using operational taxonomic units defined based on a 97% similarity threshold.
+All training approaches yielded similar classification success as measured using the  Area Under the Curve.
+The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from stool or tissue.
 
-**Conclusions.** These data provide support for the importance of the bacterial community to both adenoma and carcinoma tumorgenesis.
-The evidence collected within this study on the role of the microbiota in those with tumors identifies a number of correlations that may not have been detected because of the low power associated with the majority of studies that have been performed to date.
+**Conclusions.** This meta-analysis confirms previous results indicating that individuals with adenomas cannot be readily classified based on their bacterial community, but that those with carcinomas can.
+Regardless of the dataset, we found a subset of the fecal community that was associated with carcinomas was as predictive as the full community.
 

### 2018-02-19 submission/manuscript.tex
update abstract with edits

@@
 
-\textbf{Background.} An increasing body of literature suggests that
-there is a crucial role for the microbiota in colorectal cancer (CRC)
-pathogenesis. Important drivers within this context have ranged from
-individual microbes to the whole community. Our study expands on a
-recent meta-analysis investigating microbial biomarkers for tumors by
-testing the hypothesis that the bacterial community has important
-associations to both adenoma and carcinoma tumors. To test this
-hypothesis we examined both feces (n = 1737) and colon tissue (492 total
-samples from 350 individuals) across 14 previously published 16S rRNA
-gene sequencing studies on colorectal tumors and the microbiota.
-
-\textbf{Results.} Fecal samples had a significant decrease for both
-Shannon diversity and evenness as tumor severity increased, after
-correcting for study effect and variable region sequenced (P-value
-\textless{} 0.05). This reduction in evenness translated into small
-increases in the odds ratio for individuals to have both adenoma
-(P-value = 0.035) and carcinoma tumors (P-value = 0.0021) while the
-reduction in Shannon diversity only translated into an increased odds
-ratio for individuals to have carcinomas (P-value = 0.0069). Increases
-in mouth-associated microbes were commonly in the top 5 most increased
-odds ratios for individuals to have either adenoma or carcinoma tumors,
-regardless of sample type. Prediction models built to classify either
-individuals with adenoma or carcinoma were trained on the whole
-community or selected genera (top 5 highest and lowest odds ratios) from
-either fecal or tissue samples. Both the full and select models for
-either adenoma or carcinoma resulted in similar classification success
-according to Area Under the Curve (AUC). The most important groups
-within the full community models consistently belonged to genera such as
-\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
-studies. Although a number of associations between the microbiota and
-tumor were identified, the majority of studies that we used in this
-meta-analysis were only individually adequately powered for large effect
-sizes.
-
-\textbf{Conclusions.} These data provide support for the importance of
-the bacterial community to both adenoma and carcinoma tumorgenesis. The
-evidence collected within this study on the role of the microbiota in
-those with tumors identifies a number of correlations that may not have
-been detected because of the low power associated with the majority of
-studies that have been performed to date.
+\textbf{Background.} An increasing body of literature suggests that both
+individual and collections of bacteria are associated with the
+progression of colorectal cancer. As the number of studies investigating
+these associations increases and the number of subjects in each study
+increases, a meta-analysis to identify the associations that are the
+most predictive of disease progression is warranted. For our
+meta-analysis, we analyzed previously published 16S rRNA gene sequencing
+data collected from feces (1737 individuals from 8 studies) and colon
+tissue (492 total samples from 350 individuals from 7 studies).
+
+\textbf{Results.} We quantified the odds ratios for individual bacterial
+genera that were associated with an individual having tumors relative to
+a normal colon. Among the stool samples, there were no genera that had a
+significant odds ratio associated with adenoma and there were 8 genera
+with significant odds ratios associated with carcinoma. Similarly, among
+the tissue samples, there were no genera that had a significant odds
+ratio associated with adenoma and there were 3 genera with significant
+odds ratios associated with carcinoma. Among the significant odds
+ratios, the association between individual taxa and tumor diagnosis was
+equal or below 7.11. Because individual taxa had limited association
+with tumor diagnosis, we trained Random Forest classification models
+using the genera with the five highest and lowest odds ratios, using the
+entire collection of genera found in each study, and using operational
+taxonomic units defined based on a 97\% similarity threshold. All
+training approaches yielded similar classification success as measured
+using the Area Under the Curve. The ability to correctly classify
+individuals with adenomas was poor and the ability to classify
+individuals with carcinomas was considerably better using sequences from
+stool or tissue.
+
+\textbf{Conclusions.} This meta-analysis confirms previous results
+indicating that individuals with adenomas cannot be readily classified
+based on their bacterial community, but that those with carcinomas can.
+Regardless of the dataset, we found a subset of the fecal community that
+was associated with carcinomas was as predictive as the full community.
 

### 2018-02-19 submission/manuscript.Rmd
update background

@@
 
-Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis [@siegel_cancer_2016; @flynn_metabolic_2016].
-Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC.
-There have also been numerous case-control studies investigating the microbiota in the formation of both adenoma and carcinoma.
-A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017].
-This meta-analysis focused on identifying microbial signatures of tumors (biomarkers) but did so on a small total number of individuals and only investigated stool.
-This present meta-analysis addresses some of these major shortcomings.
+Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been hypothesized to have a role in disease progression [@siegel_cancer_2016; @flynn_metabolic_2016].
+Numerous studies using murine models of CRC have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorigenesis.
+Numerous case-control studies have characterized the microbiota of individuals with colonic adenomas and carcinomas in an attempt to identify biomarkers of disease progression [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
+Because current CRC screening recommendations are poorly adhered to due to socioeconomic status, test invasiveness, and frequency of tests, development and validation of microbiome-associated biomarkers for CRC progression could further attempts to develop non-invasive diagnostics [@GimenoGarca2012].
 
-Although there has been an intense focus on microbiota-based biomarker discovery for tumors, the number of candidate genera seem to be endless.
-Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
-Yet, even in these studies, mouth-associated genera are far from the only microbes identified to be associated with tumors.
-These other genera include, but are not limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012].
-In fact, there is good *in vivo* evidence that *Escherichia/Shigella* and *Streptococcus* can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017].
-Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential markers of CRC with good mechanistic studies for the latter [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014].
-A recent meta-analysis confirmed the correlations of certain mouth-associated genera and *Akkermansia muciniphila* with carcinoma [@shah_leveraging_2017].
-However, the sample size (n = 509) is equal to or less than some of the more recent individual studies investigating the microbiota and colorectal tumors, making it hard to know how extrapolatable these findings are.
-That particular meta-analysis also added more potential microbial associations to both carcinoma (*Pantoea agglomerans* *Ruminococcus*, *Lactobacillus*) and adenoma (*Prevotella*, *Methanosphaera*, *Succinovibrio*, *Haemophilus parainfluenzae*, *Ruminococcus*, *Lactobacillus*) stages of disease that need to be investigated further, since a number of these genera have been found to be enriched in controls and not disease [@baxter_microbiota-based_2016; @zeller_potential_2014; @burns_virulence_2015].
-Additionally, genera like *Roseburia* have been found in some studies to be increased in tumors but in others to either be decreased or have no difference [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
+Recently, there has been an intense focus on identifying microbiota-based biomarker yielding a seemingly endless number of candidate taxa.
+Some studies point towards mouth-associated genera such as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that are enriched in people with carcinomas [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
+Other studies have identified members of *Akkermansia*, *Bacteroides*, *Enterococcus*, *Escherichia*, *Klebsiella*, *Mogibacterium*, *Streptococcus*, and *Providencia* are also associated with carciomas  [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012].
+Additionally, *Roseburia* has been found in some studies to be more abundant in people with tumors but in other studies it has been found to be either less abundant or no different than what is found in subjects with normal colons  [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
+There are strong results from tissue culture and murine models that Fusobacterium nucleatum, pks-positive strains of Escherichia coli, Streptococcus gallolyticus, and an entertoxin-producing strain of Bacteroides fragilis are important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017; @weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014].
+These results point to a causative role for the microbiota in CRC pathogenesis as well as their potential as diagnostic biomarkers.
 
-Most of these studies have focused on individuals with carcinomas but associations with the adenoma stage of disease are not any clearer at identifying candidate genera correlated with these earlier tumors.
-Groups focusing on broad scale community metrics have found that measures such as richness are decreased in those with adenomas versus controls.
-Other studies have identified *Lactococcus*, *Pseudomonas*, *Acidovorax*, *Cloacibacterium*, *Helicobacter*, *Lactobacillus*, *Bilophila*, *Desulfovibrio*, and *Mogibacterium* to be increased in those with adenoma tumors [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
-Additionally, based on these studies mentioned, there seems to be very few common genera that are associated with both adenoma and carcinoma tumors, with *Lactobacillus* being one of the few commonalities.
+Most studies have focused on identifying biomarkers in patients with carcinomas but there is a greater clinical need to identify biomarkers associated with adenomas.
+Studies focusing on broad scale community metrics have found that measures such as the total number of Operational Taxonomic Units (OTUs) are decreased in those with adenomas versus controls [@sanapareddy_increased_2012].
+Other studies have identified *Acidovorax*, *Bilophila*, *Cloacibacterium*, *Desulfovibrio*, *Helicobacter*, *Lactobacillus*, *Lactococcus*, *Mogibacterium*, and *Pseudomonas* to be enriched in those with adenomas [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
+There are few genera that are enriched in patients with adenoma or carcinoma tumors.
 
-Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk of disease.
-The current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications.
-Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities.
-Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
-Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without their own limitations.
-After the colonoscopy bowel prep the bacterial community sampled may reflect the better adhered microbiota versus the resident community.
-Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done.
-It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].
+Confirming some of these previous findings, a recent meta-analysis found that 16S rRNA gene sequences from members of the *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017].
+Contrary to previous studies they found sequences similar to members of *Lactobacillus* and *Ruminococcus* to be enriched in patients with adenoma or carcinoma relative to those with normal colons [@baxter_microbiota-based_2016; @burns_virulence_2015; @zeller_potential_2014].
+In addition, they found 16S rRNA gene sequences from members of *Haemophilus*, *Methanosphaera*, *Prevotella*, *Succinovibrio* were enriched in patients with adenoma and *Pantoea* were enriched in patients with carcinomas.
+Although this meta-analysis was helpful for distilling a large number of possible biomarkers, the aggregate number of samples included in the analysis (n = 509) was smaller than several larger case-control studies that have since been published [@baxter_microbiota-based_2016; @hale_shifts_2017]
 
-In comparison to the previous meta-analysis, this study significantly increases the total stool samples investigated, re-examines important genera across adenoma and carcinoma across study, and examines differences and similarities between stool and tissue microbiota in the context of colorectal tumors.
-Importantly, this analysis and approach could provide valuable insights into the common genera that are both protective and detrimental in individuals with adenoma or carcinoma and whether broad bacterial community measurements can account for these changes that were not provided by earlier meta-analysis studies [@shah_leveraging_2017].
-
-Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissues (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community or specific members are more important risk factors for both adenoma and carcinoma stages of disease.
-To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased odds ratio (OR) for individuals to have either an adenoma or carcinoma.
-We then assessed what genera, if any, increase or decrease the OR of an individual to have an adenoma or carcinoma.
-Next, using Random Forest models, we analyzed whether the full community or only the combined top 5 increased and top 5 decreased OR genera resulted in better model classification, based on the area under the curve (AUC).
-Finally, we also examined at what effect and sample size the studies used were powered for and the sample size needed to get to the traditionally accepted 80% power.
-Our results from these analyses suggests that the bacterial community changes as disease severity worsens, that specific members are important for disease classification, and that many of the individual studies are underpowered for assessing small effect sizes.
+Here we provide an updated meta-analysis using 16S rRNA gene sequence data from both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) from 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2].
+We expand both the breadth and scope of the previous meta-analysis to investigate whether biomarkers describing the bacterial community or specific members of the community can more accurately classify patients as having adenoma or carcinoma.
+Our results suggest that the bacterial community changes as disease severity worsens and that that a subset of the microbial community can be used to diagnose the presence of carcinoma.
 
@@
 ## Discussion
+
+*Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk of disease.
+The current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications.
+Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities.
+Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
+Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without their own limitations.
+After the colonoscopy bowel prep the bacterial community sampled may reflect the better adhered microbiota versus the resident community.
+Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done.
+It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].*
 

### 2018-02-19 submission/manuscript.tex
update background

@@
 Colorectal cancer (CRC) is a growing world-wide health problem in which
-the microbiota has been purported to play an active role in disease
-pathogenesis {[}1,2{]}. Numerous studies have shown the importance of
-both individual microbes {[}3--7{]} and the overall community
-{[}8--10{]} in tumorgenesis using mouse models of CRC. There have also
-been numerous case-control studies investigating the microbiota in the
-formation of both adenoma and carcinoma. A recent meta-analysis
-investigated whether specific biomarkers could be consistently
-identified using multiple data sets {[}11{]}. This meta-analysis focused
-on identifying microbial signatures of tumors (biomarkers) but did so on
-a small total number of individuals and only investigated stool. This
-present meta-analysis addresses some of these major shortcomings.
-
-Although there has been an intense focus on microbiota-based biomarker
-discovery for tumors, the number of candidate genera seem to be endless.
-Some studies point towards mouth-associated genera such as
-\emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas}, and
-\emph{Porphyromonas} as key enriched genera {[}6,12--18{]}. Yet, even in
-these studies, mouth-associated genera are far from the only microbes
-identified to be associated with tumors. These other genera include, but
-are not limited to, \emph{Providencia}, \emph{Mogibacterium},
-\emph{Enterococcus}, \emph{Escherichia/Shigella}, \emph{Klebsiella}, and
-\emph{Streptococcus} {[}14--16{]}. In fact, there is good \emph{in vivo}
-evidence that \emph{Escherichia/Shigella} and \emph{Streptococcus} can
-be important in the pathogenesis of CRC {[}5,19,20{]}. Other studies
-have also identified \emph{Akkermansia muciniphila} and
-\emph{Bacteroides fragilis} as potential markers of CRC with good
-mechanistic studies for the latter {[}15,21,22{]}. A recent
-meta-analysis confirmed the correlations of certain mouth-associated
-genera and \emph{Akkermansia muciniphila} with carcinoma {[}11{]}.
-However, the sample size (n = 509) is equal to or less than some of the
-more recent individual studies investigating the microbiota and
-colorectal tumors, making it hard to know how extrapolatable these
-findings are. That particular meta-analysis also added more potential
-microbial associations to both carcinoma (\emph{Pantoea agglomerans}
-\emph{Ruminococcus}, \emph{Lactobacillus}) and adenoma
-(\emph{Prevotella}, \emph{Methanosphaera}, \emph{Succinovibrio},
-\emph{Haemophilus parainfluenzae}, \emph{Ruminococcus},
-\emph{Lactobacillus}) stages of disease that need to be investigated
-further, since a number of these genera have been found to be enriched
-in controls and not disease {[}13,16,17{]}. Additionally, genera like
-\emph{Roseburia} have been found in some studies to be increased in
-tumors but in others to either be decreased or have no difference
-{[}15,18,23,24{]}.
-
-Most of these studies have focused on individuals with carcinomas but
-associations with the adenoma stage of disease are not any clearer at
-identifying candidate genera correlated with these earlier tumors.
-Groups focusing on broad scale community metrics have found that
-measures such as richness are decreased in those with adenomas versus
-controls. Other studies have identified \emph{Lactococcus},
-\emph{Pseudomonas}, \emph{Acidovorax}, \emph{Cloacibacterium},
-\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Bilophila},
-\emph{Desulfovibrio}, and \emph{Mogibacterium} to be increased in those
-with adenoma tumors {[}25--27{]}. Additionally, based on these studies
-mentioned, there seems to be very few common genera that are associated
-with both adenoma and carcinoma tumors, with \emph{Lactobacillus} being
-one of the few commonalities.
-
-Targeting the identification of tumor microbial biomarkers within stool
-seems logical since it offers an easy and cost-effective way to stratify
-risk of disease. The current gold standard for diagnosis, a colonoscopy,
-can be time-consuming and is not without risk of complications. Although
-stool represents an easy and less invasive way to assess risk, it is not
-clear how well this sample reflects adenoma- and carcinoma- associated
-microbial communities. Some studies have tried to assess this in health
-and disease but are limited by their sample size {[}18,28{]}. Sampling
-the microbiota directly associated with colon tissue may provide clearer
-answers but is not without their own limitations. After the colonoscopy
-bowel prep the bacterial community sampled may reflect the better
-adhered microbiota versus the resident community. Additionally, these
-samples contain more host DNA, potentially limiting the types of
-analysis that can be done. It is well known that low biomass samples can
-be very difficult to work with and results can be study dependent due to
-the randomness of contamination {[}29{]}.
-
-In comparison to the previous meta-analysis, this study significantly
-increases the total stool samples investigated, re-examines important
-genera across adenoma and carcinoma across study, and examines
-differences and similarities between stool and tissue microbiota in the
-context of colorectal tumors. Importantly, this analysis and approach
-could provide valuable insights into the common genera that are both
-protective and detrimental in individuals with adenoma or carcinoma and
-whether broad bacterial community measurements can account for these
-changes that were not provided by earlier meta-analysis studies
-{[}11{]}.
-
-Using both feces (n = 1737) and colon tissues (492 samples from 350
-individuals) totaling over 2229 total samples across 14 studies
-{[}12--18,21,23--27,30{]} {[}Table 1 \& 2{]}, we expand both the breadth
-and scope of the previous meta-analysis to investigate whether the
-bacterial community or specific members are more important risk factors
-for both adenoma and carcinoma stages of disease. To accomplish this we
-first assessed whether bacterial diversity changes throughout disease
-(control to adenoma to carcinoma) and if it results in an increased odds
-ratio (OR) for individuals to have either an adenoma or carcinoma. We
-then assessed what genera, if any, increase or decrease the OR of an
-individual to have an adenoma or carcinoma. Next, using Random Forest
-models, we analyzed whether the full community or only the combined top
-5 increased and top 5 decreased OR genera resulted in better model
-classification, based on the area under the curve (AUC). Finally, we
-also examined at what effect and sample size the studies used were
-powered for and the sample size needed to get to the traditionally
-accepted 80\% power. Our results from these analyses suggests that the
-bacterial community changes as disease severity worsens, that specific
-members are important for disease classification, and that many of the
-individual studies are underpowered for assessing small effect sizes.
+the microbiota has been hypothesized to have a role in disease
+progression {[}1,2{]}. Numerous studies using murine models of CRC have
+shown the importance of both individual microbes {[}3--7{]} and the
+overall community {[}8--10{]} in tumorigenesis. Numerous case-control
+studies have characterized the microbiota of individuals with colonic
+adenomas and carcinomas in an attempt to identify biomarkers of disease
+progression {[}6,11--17{]}. Because current CRC screening
+recommendations are poorly adhered to due to socioeconomic status, test
+invasiveness, and frequency of tests, development and validation of
+microbiome-associated biomarkers for CRC progression could further
+attempts to develop non-invasive diagnostics {[}18{]}.
+
+Recently, there has been an intense focus on identifying
+microbiota-based biomarker yielding a seemingly endless number of
+candidate taxa. Some studies point towards mouth-associated genera such
+as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that
+are enriched in people with carcinomas {[}6,11--17{]}. Other studies
+have identified members of \emph{Akkermansia}, \emph{Bacteroides},
+\emph{Enterococcus}, \emph{Escherichia}, \emph{Klebsiella},
+\emph{Mogibacterium}, \emph{Streptococcus}, and \emph{Providencia} are
+also associated with carciomas {[}13--15{]}. Additionally,
+\emph{Roseburia} has been found in some studies to be more abundant in
+people with tumors but in other studies it has been found to be either
+less abundant or no different than what is found in subjects with normal
+colons {[}14,17,19,20{]}. There are strong results from tissue culture
+and murine models that Fusobacterium nucleatum, pks-positive strains of
+Escherichia coli, Streptococcus gallolyticus, and an
+entertoxin-producing strain of Bacteroides fragilis are important in the
+pathogenesis of CRC {[}5,14,21--24{]}. These results point to a
+causative role for the microbiota in CRC pathogenesis as well as their
+potential as diagnostic biomarkers.
+
+Most studies have focused on identifying biomarkers in patients with
+carcinomas but there is a greater clinical need to identify biomarkers
+associated with adenomas. Studies focusing on broad scale community
+metrics have found that measures such as the total number of Operational
+Taxonomic Units (OTUs) are decreased in those with adenomas versus
+controls {[}25{]}. Other studies have identified \emph{Acidovorax},
+\emph{Bilophila}, \emph{Cloacibacterium}, \emph{Desulfovibrio},
+\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Lactococcus},
+\emph{Mogibacterium}, and \emph{Pseudomonas} to be enriched in those
+with adenomas {[}25--27{]}. There are few genera that are enriched in
+patients with adenoma or carcinoma tumors.
+
+Confirming some of these previous findings, a recent meta-analysis found
+that 16S rRNA gene sequences from members of the \emph{Akkermansia},
+\emph{Fusobacterium}, and \emph{Parvimonas} were fecal biomarkers for
+the presence of carcinomas {[}28{]}. Contrary to previous studies they
+found sequences similar to members of \emph{Lactobacillus} and
+\emph{Ruminococcus} to be enriched in patients with adenoma or carcinoma
+relative to those with normal colons {[}12,15,16{]}. In addition, they
+found 16S rRNA gene sequences from members of \emph{Haemophilus},
+\emph{Methanosphaera}, \emph{Prevotella}, \emph{Succinovibrio} were
+enriched in patients with adenoma and \emph{Pantoea} were enriched in
+patients with carcinomas. Although this meta-analysis was helpful for
+distilling a large number of possible biomarkers, the aggregate number
+of samples included in the analysis (n = 509) was smaller than several
+larger case-control studies that have since been published {[}12,27{]}
+
+Here we provide an updated meta-analysis using 16S rRNA gene sequence
+data from both feces (n = 1737) and colon tissue (492 samples from 350
+individuals) from 14 studies {[}11--17,19,20,23,25--27,29{]} {[}Table 1
+\& 2{]}. We expand both the breadth and scope of the previous
+meta-analysis to investigate whether biomarkers describing the bacterial
+community or specific members of the community can more accurately
+classify patients as having adenoma or carcinoma. Our results suggest
+that the bacterial community changes as disease severity worsens and
+that that a subset of the microbial community can be used to diagnose
+the presence of carcinoma.
 
@@
 
+\emph{Targeting the identification of tumor microbial biomarkers within
+stool seems logical since it offers an easy and cost-effective way to
+stratify risk of disease. The current gold standard for diagnosis, a
+colonoscopy, can be time-consuming and is not without risk of
+complications. Although stool represents an easy and less invasive way
+to assess risk, it is not clear how well this sample reflects adenoma-
+and carcinoma- associated microbial communities. Some studies have tried
+to assess this in health and disease but are limited by their sample
+size {[}17,30{]}. Sampling the microbiota directly associated with colon
+tissue may provide clearer answers but is not without their own
+limitations. After the colonoscopy bowel prep the bacterial community
+sampled may reflect the better adhered microbiota versus the resident
+community. Additionally, these samples contain more host DNA,
+potentially limiting the types of analysis that can be done. It is well
+known that low biomass samples can be very difficult to work with and
+results can be study dependent due to the randomness of contamination
+{[}31{]}.}
+
 Our study identifies clear but small differences in diversity at the
@@
 identified, it does not necessarily mean that these specific species are
-not important in human CRC {[}20,22{]}. Since we are limited in our
+not important in human CRC {[}22,24{]}. Since we are limited in our
 aggregation of the data to the genus level, it is not possible to
@@
 meta-analysis were identified through the review articles written by
-Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}31,32{]} and
+Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}32,33{]} and
 additional studies not mentioned in the reviews were obtained based on
@@
 authors were able to share. After these filtering steps, the following
-studies remained: Ahn, \emph{et al.} {[}12{]}, Baxter, \emph{et al.}
-{[}13{]}, Brim, \emph{et al.} {[}30{]}, Burns, \emph{et al.} {[}16{]},
-Chen, \emph{et al.} {[}14{]}, Dejea, \emph{et al.} {[}24{]}, Flemer,
-\emph{et al.} {[}18{]}, Geng, \emph{et al.} {[}23{]}, Hale, \emph{et
-al.} {[}27{]}, Kostic, \emph{et al.} {[}33{]}, Lu, \emph{et al.}
+studies remained: Ahn, \emph{et al.} {[}11{]}, Baxter, \emph{et al.}
+{[}12{]}, Brim, \emph{et al.} {[}29{]}, Burns, \emph{et al.} {[}15{]},
+Chen, \emph{et al.} {[}13{]}, Dejea, \emph{et al.} {[}20{]}, Flemer,
+\emph{et al.} {[}17{]}, Geng, \emph{et al.} {[}19{]}, Hale, \emph{et
+al.} {[}27{]}, Kostic, \emph{et al.} {[}34{]}, Lu, \emph{et al.}
 {[}26{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang, \emph{et al.}
-{[}15{]}, Weir, \emph{et al.} {[}21{]}, and Zeller, \emph{et al.}
-{[}17{]}. The Zackular {[}34{]} study was not included because the 90
+{[}14{]}, Weir, \emph{et al.} {[}23{]}, and Zeller, \emph{et al.}
+{[}16{]}. The Zackular {[}35{]} study was not included because the 90
 individuals analyzed within the study are contained within the larger
-Baxter study {[}13{]}. After sequence processing, all the case samples
+Baxter study {[}12{]}. After sequence processing, all the case samples
 for the Kostic study had 100 or less sequences remaining and was
@@
 not have sequences and metadata on the SRA, data was obtained from DBGap
-(n = 1, {[}12{]}) and directly from the authors (n = 4,
-{[}18,21,25,27{]}). Each study was processed using the mothur (v1.39.3)
-software program {[}35{]} and quality filtering utilized the default
+(n = 1, {[}11{]}) and directly from the authors (n = 4,
+{[}17,23,25,27{]}). Each study was processed using the mothur (v1.39.3)
+software program {[}36{]} and quality filtering utilized the default
 methods for both 454 and Illumina based sequencing. If it was not
@@
 artificial combination of two or more different sequences and commonly
-known as chimeras were identified and removed using VSEARCH {[}36{]}
+known as chimeras were identified and removed using VSEARCH {[}37{]}
 before \emph{de novo} OTU clustering at 97\% similarity was completed
-using the OptiClust algorithm {[}37{]}.
+using the OptiClust algorithm {[}38{]}.
 
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package {[}38{]}.
+sequence processing utilized the R (v3.4.3) software package {[}39{]}.
 For OTU richness, evenness, and Shannon diversity analysis, values were
-power transformed using the rcompanion (v1.11.1) package {[}39{]} and
-then Z-score normalized using the car (v2.1.6) package {[}40{]}. Testing
+power transformed using the rcompanion (v1.11.1) package {[}40{]} and
+then Z-score normalized using the car (v2.1.6) package {[}41{]}. Testing
 for OTU richness, evenness, and Shannon diversity differences utilized
 linear mixed-effect models created using the lme4 (v1.1.15) package
-{[}41{]} to correct for study, repeat sampling of individuals (tissue
+{[}42{]} to correct for study, repeat sampling of individuals (tissue
 only), and 16S hyper-variable region used. Odds ratios (OR) were
 analyzed using both the epiR (v0.9.93) and metafor (v2.0.0) packages
-{[}42,43{]} by assessing how many individuals with and without disease
+{[}43,44{]} by assessing how many individuals with and without disease
 were above and below the overall median value within each specific
@@
 distance matrix and testing with PERMANOVA executed with the vegan
-(v2.4.5) package {[}44{]}. Random Forest models were built using both
-the caret (v6.0.78) and randomForest (v4.6.12) packages {[}45,46{]}.
+(v2.4.5) package {[}45{]}. Random Forest models were built using both
+the caret (v6.0.78) and randomForest (v4.6.12) packages {[}46,47{]}.
 Power analysis and estimations were made using the pwr (v1.2.1) and
-statmod (v1.4.30) packages {[}47,48{]}. All figures were created using
-both ggplot2 (v2.2.1) and gridExtra (v2.3) packages {[}49,50{]}.
+statmod (v1.4.30) packages {[}48,49{]}. All figures were created using
+both ggplot2 (v2.2.1) and gridExtra (v2.3) packages {[}50,51{]}.
 
@@
 ORs for individuals having an adenoma or carcinoma and corrected for
-multiple comparisons using the Benjamini-Hochberg method {[}51{]}. We
+multiple comparisons using the Benjamini-Hochberg method {[}52{]}. We
 then built Random Forest models based on all genera, all OTUs, or
@@
 
-\hypertarget{ref-shah_leveraging_2017}{}
-11. Shah MS, DeSantis TZ, Weinmaier T, McMurdie PJ, Cope JL, Altrichter
-A, et al. Leveraging sequence-based faecal microbial community survey
-data to identify a composite biomarker for colorectal cancer. Gut. 2017;
-
 \hypertarget{ref-ahn_human_2013}{}
-12. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
+11. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
 microbiome and risk for colorectal cancer. Journal of the National
@@
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-13. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+12. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
 improves the sensitivity of fecal immunochemical test for detecting
@@
 \hypertarget{ref-chen_human_2012}{}
-14. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
+13. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
 mucosa-associated microbiota in patients with colorectal cancer. PloS
@@
 \hypertarget{ref-wang_structural_2012}{}
-15. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
+14. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
 segregation of gut microbiota between colorectal cancer patients and
@@
 \hypertarget{ref-burns_virulence_2015}{}
-16. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
+15. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
 are a signature of the microbiome in the colorectal tumor
@@
 \hypertarget{ref-zeller_potential_2014}{}
-17. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
+16. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
 Potential of fecal microbiota for early-stage detection of colorectal
@@
 \hypertarget{ref-flemer_tumour-associated_2017}{}
-18. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
+17. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
 al. Tumour-associated and non-tumour-associated microbiota in colorectal
@@
 
+\hypertarget{ref-GimenoGarca2012}{}
+18. García AZG. Factors influencing colorectal cancer screening
+participation. Gastroenterology Research and Practice {[}Internet{]}.
+Hindawi Limited; 2012;2012:1--8. Available from:
+\url{https://doi.org/10.1155/2012/483417}
+
+\hypertarget{ref-geng_diversified_2013}{}
+19. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
+human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
+
+\hypertarget{ref-dejea_microbiota_2014}{}
+20. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
+Rossetti BJ, et al. Microbiota organization is a distinct feature of
+proximal colorectal cancers. Proceedings of the National Academy of
+Sciences of the United States of America. 2014;111:18321--6.
+
 \hypertarget{ref-ecoli_Arthur_2014}{}
-19. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
+21. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
 McCafferty J, et al. Microbial genomic analysis reveals the essential
@@
 \hypertarget{ref-strep_Aymeric_2017}{}
-20. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
+22. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
 et al. Colorectal cancer specific conditions promoteStreptococcus
@@
 \hypertarget{ref-weir_stool_2013}{}
-21. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
+23. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
 Stool microbiome and metabolome differences between colorectal cancer
@@
 \hypertarget{ref-bfrag_Boleij_2014}{}
-22. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
+24. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
 MG, et al. The bacteroides fragilis toxin gene is prevalent in the colon
@@
 Available from: \url{https://doi.org/10.1093/cid/ciu787}
-
-\hypertarget{ref-geng_diversified_2013}{}
-23. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
-human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
-
-\hypertarget{ref-dejea_microbiota_2014}{}
-24. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
-Rossetti BJ, et al. Microbiota organization is a distinct feature of
-proximal colorectal cancers. Proceedings of the National Academy of
-Sciences of the United States of America. 2014;111:18321--6.
 
@@
 
+\hypertarget{ref-shah_leveraging_2017}{}
+28. Shah MS, DeSantis TZ, Weinmaier T, McMurdie PJ, Cope JL, Altrichter
+A, et al. Leveraging sequence-based faecal microbial community survey
+data to identify a composite biomarker for colorectal cancer. Gut. 2017;
+
+\hypertarget{ref-brim_microbiome_2013}{}
+29. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
+al. Microbiome analysis of stool samples from African Americans with
+colon polyps. PloS One. 2013;8:e81352.
+
 \hypertarget{ref-Flynn_preprint_2017}{}
-28. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
+30. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
 the native colon microbiota in healthy adults. Cold Spring Harbor
@@
 \hypertarget{ref-Salter_contamination_2014}{}
-29. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
+31. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
 al. Reagent and laboratory contamination can critically impact
@@
 
-\hypertarget{ref-brim_microbiome_2013}{}
-30. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
-al. Microbiome analysis of stool samples from African Americans with
-colon polyps. PloS One. 2013;8:e81352.
-
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-31. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
+32. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
 microbiota and colorectal cancer. American Journal of Physiology -
@@
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-32. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
+33. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
 microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
@@
 \hypertarget{ref-kostic_genomic_2012}{}
-33. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
+34. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
 al. Genomic analysis identifies association of Fusobacterium with
@@
 \hypertarget{ref-zackular_human_2014}{}
-34. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
+35. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
 microbiome as a screening tool for colorectal cancer. Cancer Prevention
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-35. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+36. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, et al. Introducing mothur: Open-Source, Platform-Independent,
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-36. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
+37. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
 versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-37. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
+38. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
 Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
@@
 \hypertarget{ref-r_citation_2017}{}
-38. R Core Team. R: A language and environment for statistical computing
+39. R Core Team. R: A language and environment for statistical computing
 {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-39. Mangiafico S. Rcompanion: Functions to support extension education
+40. Mangiafico S. Rcompanion: Functions to support extension education
 program evaluation {[}Internet{]}. 2017. Available from:
@@
 \hypertarget{ref-car_citation_2011}{}
-40. Fox J, Weisberg S. An R companion to applied regression
+41. Fox J, Weisberg S. An R companion to applied regression
 {[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
@@
 \hypertarget{ref-lme4_citation_2015}{}
-41. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
+42. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
 models using lme4. Journal of Statistical Software. 2015;67:1--48.
@@
 \hypertarget{ref-epir_citation_2017}{}
-42. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
+43. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
 J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
@@
 \hypertarget{ref-metafor_citation_2010}{}
-43. Viechtbauer W. Conducting meta-analyses in R with the metafor
+44. Viechtbauer W. Conducting meta-analyses in R with the metafor
 package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-44. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+45. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
@@
 \hypertarget{ref-caret_citation_2017}{}
-45. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
+46. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
 Cooper T, et al. Caret: Classification and regression training
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-46. Liaw A, Wiener M. Classification and regression by randomForest. R
+47. Liaw A, Wiener M. Classification and regression by randomForest. R
 News {[}Internet{]}. 2002;2:18--22. Available from:
@@
 \hypertarget{ref-pwr_citation_2017}{}
-47. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
+48. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
 2017. Available from: \url{https://CRAN.R-project.org/package=pwr}
@@
 \hypertarget{ref-statmod_citation_2016}{}
-48. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
+49. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
 gaussian distribution. R Journal. 2016;8:339--51.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-49. Wickham H. Ggplot2: Elegant graphics for data analysis
+50. Wickham H. Ggplot2: Elegant graphics for data analysis
 {[}Internet{]}. Springer-Verlag New York; 2009. Available from:
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-50. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
+51. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
 {[}Internet{]}. 2017. Available from:
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-51. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+52. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
 practical and powerful approach to multiple testing. Journal of the

### 2018-02-22 submission/manuscript.Rmd
pat updates

@@
 alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_OR_tissue_composite.csv"))
+
+# Load in PERMANOVA data
+adn_beta_stool <- read_csv("../data/process/tables/beta_perm_adn_stool_summary.csv")
+crc_beta_stool <- read_csv("../data/process/tables/beta_perm_stool_summary.csv")
+
+adn_unmatched <- read_csv("../data/process/tables/beta_perm_adn_unmatched_tissue_summary.csv")
+adn_matched <- read_csv("../data/process/tables/beta_perm_adn_matched_tissue_summary.csv")
+
+crc_unmatched <- read_csv("../data/process/tables/beta_perm_unmatched_tissue_summary.csv")
+crc_matched <- read_csv("../data/process/tables/beta_perm_matched_tissue_summary.csv")
 
@@
 
-***Lower Bacterial Diversity is Associated with Increased OR of Tumors:***
-To assess differences in broad scale community metrics as disease severity worsens Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity measurements were power transformed and Z-score normalized.
-These metrics are commonly used to assess the total number of OTUs, the equality of their abundance, and the overall diversity, respectively.
-Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity.
-In stool, there was a significant decrease in both evenness and Shannon diversity as disease severity moved from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively).
-We next tested whether the detectable differences in community significantly increased in OR of having an adenoma or carcinoma.
-For fecal samples, a decrease versus the overall median in evenness resulted in a significantly increased OR for carcinoma  (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overall median in Shannon diversity only increased the OR for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 1].
-Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma-associated and adenoma-associated microbiota versus control [Table S1].
+***Lower Bacterial Diversity is Associated with Increased Odds Ratio (OR) of Tumors:***
+We first assessed whether variation in broad community metrics like total number of operational taxonomic units (OTUs) (i.e.
+richness), the evenness of their abundance, and the overall diversity was associated with disease stage after controlling for study and variable region differences.
+In stool, there was a significant decrease in both evenness and diversity as disease severity progressed from normal to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1]; there was not a significant difference for richness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`).
+We next tested whether the decrease in these community metrics translated into significant ORs for having an adenoma or carcinoma.
+For fecal samples, the ORs for richness were not significantly greater than 1.0 for adenoma or carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "sobs")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A].
+The ORs for evenness were significantly higher than 1.0 for adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and carcinoma (OR =  `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2B].
+The ORs for diversity were only significantly greater than 1.0 for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`), but not for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2C].
+Although these OR are significantly greater than 1.0, it is doubtful that these are clinically meaningful values.
 
-Using similar transformations for tissue samples, linear mixed-effect models were used on the transformed combined data to control for study, re-sampling of the same individual, and 16S variable region to test whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner as disease severity increased.
-For colon tissue, there were no significant changes in OTU richness, evenness, or Shannon diversity as disease severity progressed from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
-We next analyzed the OR, for matched (unaffected tissue and an adenoma or carcinoma from the same individual) and unmatched (control and adenoma or carcinoma tissue not from the same individual) colon tissue samples.
-For individuals at either an adenoma or carcinoma stage of disease there was no significant change in OR based on lower than median values for OTU richness, evenness, and Shannon diversity [Figure S1 & Table S2].
-Similar to stool samples, significant differences in bacterial community, assessed by PERMANOVA, were identified in unmatched tissue samples, for those at either adenoma or carcinoma stage of CRC [Table S1].
-For studies with matched samples no differences in bacterial community were observed when assessed with PERMANOVA [Table S1].
-These tissue results suggest that the microbiota within an individual are similar to each other regardless of disease status.
+Similar to our analysis of sequences obtained from stool samples, we repeated the analysis using sequences obtained from colon tissue.
+There were no significant changes in richness, evenness, or diversity as disease severity progressed from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
+We next analyzed the OR, for matched (i.e.
+where unaffected tissue and tumors were obtained from the same individual) and unmatched (i.e.
+where unaffected tissue and tumor tissue were not obtained from the same individual) tissue samples.
+The ORs for adenoma and carcinoma by any measure were not significantly different from 1.0 (P-value `r ifelse(min(alpha_RR_crc_unmatch_tissue$pvalue) < 0.05 & min(alpha_RR_crc_match_tissue$pvalue) < 0.05 & min(alpha_RR_adn_tissue$pvalue) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure S1 & Table S1].
+This is likely due to the combination of a small effect size, as suggested from the results using stool, and the relatively small number of studies and size of studies used in the analysis.
+
+***Disease Progression is Associated with Community-Wide Changes in Composition and Abundance:*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages.
+We identified significant bacterial community differences in the stool of patients with adenomas relative to those with normal colons in `r length(as.data.frame(filter(adn_beta_stool, pvalue < 0.05))[, "study"])` of `r length(as.data.frame(adn_beta_stool)[, "study"])` studies and in patients with carcinomas relative to those with normal colons in 6 of 7 studies (PERMANOVA; P-value < 0.05) [Table S2].
+Similar to the analyses using stool samples, there were significant differences in bacterial community structure between subjects with normal colons and those with adenoma (1 of 2 studies) and carcinoma (1 of 3 studies) [Table S2].
+For studies that used matched samples no differences in bacterial community structures were observed [Table S2].
+Combined, these results indicate that there consistent and significant community-wide changes in the fecal community structure of subjects with carcinomas.
+However, the signal observed in subjects with adenomas or when using tissue samples was not as consistent.
+This is likely due to a smaller effect size or the relatively small sample sizes among the studies that characterized the tissue microbiota.
+
+***Individual Taxa are Associated with Significant ORs for Carcinomas:*** 
+Next we identified those taxa were associated with ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas.
+No taxa had a significant OR for the presence of adenomas when we used data collected from stool or tissue samples (Table S3 & S4).
+In contrast, 8 taxa had significant ORs for the presence of carcinomas using data from stool samples.
+Of these, 4 are commonly associated with the oral cavity: Fusobacterium (OR = 0.XX (0.XX - 0.XX)), Parvimonas (OR = 0.XX (0.XX - 0.XX)), Porphyromonas (OR = 0.XX (0.XX - 0.XX)), and Peptostreptococcus (OR = 0.XX (0.XX - 0.XX)) [Table S3].
+The other 4 were A  (OR = 0.XX (0.XX - 0.XX)), B (OR = 0.XX (0.XX - 0.XX)), C (OR = 0.XX (0.XX - 0.XX)), and D (OR = 0.XX (0.XX - 0.XX)).
+Among the data collected from tissue samples, only unmatched carcinoma samples had taxa with a significant OR.
+Those included Dorea (OR = 0.35 (0.22 - 0.55)), Blautia (OR = 0.47 (0.3 - 0.73)), and Weissella (OR = 5.15 (2.02 - 13.14)).
+Mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4].
+For example, Fusobacterium had an OR of 3.98 (1.19-13.24); however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg-corrected P-value was 0.93 [Table S4].
+It is interesting to note that Ruminococcus and members of Clostridium group XI in stool and Dorea and Blautia in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas.
+Overall, there was no overlap in the taxa with significant OR between stool and tissue samples.
 

### 2018-02-23 submission/manuscript.Rmd
update overall results

@@
 crc_unmatched_OR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_unmatched_tissue_composite.csv") %>% mutate(bh = p.adjust(pvalue, method = "BH")) %>% filter(bh < 0.05))
+
+crc_matched_OR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_matched_tissue_composite.csv") %>% mutate(bh = p.adjust(pvalue, method = "BH")))
 
@@
 No taxa had a significant OR for the presence of adenomas when we used data collected from stool or tissue samples (Table S3 & S4).
-In contrast, 8 taxa had significant ORs for the presence of carcinomas using data from stool samples.
-Of these, 4 are commonly associated with the oral cavity: Fusobacterium (OR = 0.XX (0.XX - 0.XX)), Parvimonas (OR = 0.XX (0.XX - 0.XX)), Porphyromonas (OR = 0.XX (0.XX - 0.XX)), and Peptostreptococcus (OR = 0.XX (0.XX - 0.XX)) [Table S3].
-The other 4 were A  (OR = 0.XX (0.XX - 0.XX)), B (OR = 0.XX (0.XX - 0.XX)), C (OR = 0.XX (0.XX - 0.XX)), and D (OR = 0.XX (0.XX - 0.XX)).
+In contrast, `r length(crc_OR_stool$measure)` taxa had significant ORs for the presence of carcinomas using data from stool samples.
+Of these, 4 are commonly associated with the oral cavity: Fusobacterium (OR = `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`)), Parvimonas (OR = `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_ub"], digits = 2))`)), Porphyromonas (OR = `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2))`)), and Peptostreptococcus (OR = `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_ub"], digits = 2))`)) [Table S3].
+The other 4 were Clostridium XI (OR = `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_ub"], digits = 2))`)), Enterobacteriaceae  (OR = `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_ub"], digits = 2))`)), Escherichia (OR = `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_ub"], digits = 2))`)), and Ruminococcus (OR = `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_ub"], digits = 2))`)).
 Among the data collected from tissue samples, only unmatched carcinoma samples had taxa with a significant OR.
-Those included Dorea (OR = 0.35 (0.22 - 0.55)), Blautia (OR = 0.47 (0.3 - 0.73)), and Weissella (OR = 5.15 (2.02 - 13.14)).
+Those included Dorea (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2))`)), Blautia (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2))`)), and Weissella (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2))`)).
 Mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4].
-For example, Fusobacterium had an OR of 3.98 (1.19-13.24); however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg-corrected P-value was 0.93 [Table S4].
+For example, Fusobacterium had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`; however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg-corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4].
 It is interesting to note that Ruminococcus and members of Clostridium group XI in stool and Dorea and Blautia in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas.
@@
 
-***Mouth-Associated Genera are Associated with an Increased OR of Tumor:***
-Next, we asked if being higher than the median relative abundance, for any specific genera, resulted in an altered OR for adenoma or carcinoma, in stool and colon tissue, due to our previous observations of small increases in OR using OTU richness and Shannon diversity.
-To investigate this we analyzed all common genera across each study, in colon tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in OR.
-For both tissue and stool samples only ORs associated with an increase or decrease in carcioma tumors were significant after multiple comparison correction [Table S3 & S4].
-Out of the `r length(crc_OR_stool$measure)` taxa that had significant ORs in stool samples `r length(filter(crc_OR_stool, measure %in% c("Fusobacterium", "Parvimonas", "Porphyromonas", "Peptostreptococcus"))[, "measure"])` were mouth-associated microbes.
-These mouth-associated genera significantly increased the ORs of carcinoma for stool samples and included *Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Table S3].
-Conversely, mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4].
-Only unmatched tissue samples had significant OR taxa and these were *Dorea* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2)`)), *Blautia* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2)`)), and *Weissella* (OR = `r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2)` (`r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2)` - `r round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2)`)).
-Overall, there was little direct overlap in increased or decreased OR taxa between both tumor type and sample site.
+***Individual Significant OR Taxa Classify Carcinoma Poorly:***
+Since specific taxa were associated with a significant OR for carcinoma we tested whether they would also be good classifiers of carcinoma.
+For stool, the `r length(crc_OR_stool$measure)` significant taxa did no better at classifying those with normal colons versus those with carcinomas than chance [Figure 3A].
+Likewise in unmathced tissue samples the `r length(crc_unmatched_OR_tissue$measure)` significant taxa were no better than an AUC of 0.5 [Figure 3B].
+These results suggest that although these taxa are significantly associated with a decreased or increased OR of carcinoma, individually they are poor classifiers of disease.
 
 ***Select Community Models can Recapitulate Whole Community Models:***
-Since specific genera increased the OR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera.
-We selected these genera based on significance after multiple comparison correction.
+Since specific taxa increased or decreased the OR for carcinoma but performed poorly as an individual classifier of carcinoma and assessing ORs for adenoma we assessed whether the overall bacterial community was better at classifying disease.
+Three models were tested and included a full taxa model, full OTU model, and significant OR taxa model.
 If no taxa were significant after multiple comparison correction then no model for that specific grouping (i.e.
@@
 We first tested three model AUCs.
-These models were created using Random Forest and where either all genera, all OTUs, or significant OR taxa only.
 Next, the all genera models and any significant OR taxa models were tested across all studies that were not used to train the model.
-For stool, all models used had similar AUCs [Figure 2].
-Although for adenoma and unmatched carcinoma this trend held, there were large differences in matched tissue based on whether all genera or OTUs were utilized [Figure S2].
-When analyzing the tests sets that were comprised of genera data from other studies, both the all genera and singificant OR taxa only models had a similar ability to detect individuals with carcinomas, for both stool and tissue samples [Figure 3-S3].
 
-In stool, the most common genera in the top 10 most important variables, in the full community models using all genera-based models, were *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 4].
-Regardless of sample type, mouth-associated genera were present in models for carcinomas [Figure 4B & D].
-Yet, none were present in the majority of studies and none were present in the adenoma models [Figure 4A & B].
-For the full community OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 4C & 4D].
+For stool, all models used had similar AUCs for both adenoma and carcinoma [Figure 4].
+However, the adenoma AUCs were barely better than chance [Figure 4A].
+Similarily, when analyzing the tests sets that were comprised of genera data from other studies the adenoma models performed poorly [Figure 5A].
+The carcinoma models performed much better, with both the all genera and singificant OR taxa only models having a similar ability to detect individuals with carcinomas [Figure 5B].
+The most common genera in the top 10 most important variables, in the full genera models used to classify carcinomas, were *Bacteroides* and *Lachnospiraceae* [Figure 6A].
+Regardless of sample type, mouth-associated genera were present in models for carcinomas but not consistently across studies [Figure 6].
+For the full community OTU-based models, both *Bacteroides*, *Blautia*, *Lachnospiraceae*, and *Ruminococcaceae* were present in the top 10 consistently across studies [Figure 6B].
+Overall, these results suggest that multiple microbes could act as the inflammatory stimulus needed to exacerbate mutations leading each individual microbe as a poor individual classifier but much better in aggregate with others.
 
-Unlike the stool-based Random Forest models, the tissue-based models, for the full genera from the first approach, showed no consistent representation of *Ruminococcaceae*, *Ruminococcus*, *Bacteroides*, and *Roseburia* in the top 10 most important model variables across study [Figure S4].
-The vast majority of the top 10 model variables for the full community genera- and OTU-based models using colon tissue tended to be study specific.
-Further, there was very little overlap in the top 10 important variables between adenoma and carcinoma stage models, regardless of whether colon tissue or stool was used  [Figure S4].
-This discordance between stool and colon tissue samples also applies to the mouth-associated genera with one noticeable skew being that *Fusobacterium* and *Fusobacteriaceae* occur more often in the top 10 of matched versus unmatched colon tissue Random Forest models [Figure S4B-C & S4E-F].
-This suggests that either the colon tissue microbiota is study or person dependent, that kit and/or other types of contamination associated with low biomass samples may be skewing the results, or that multiple microbes could act as the inflammatory stimulus needed to exacerbate mutations.
-
-***CRC Studies are Underpowered for Detecting Small Effect Sizes:***
-Next, we assessed how much confidence should be placed in the reported outcomes from each individual study by calculating the ability to detect a difference (power) and sample size needed for small, medium, and large effect size differences between cases and controls.
-When assessing the power of each study at different effect sizes the majority of studies achieved 80% power to detect a 30% or greater difference between groups [Figure 5A & B].
-No study that we analyzed had the standard 80% power to detect an effect size difference equal to or below 10% [Figure 5A & B].
-In order to achieve a power of 80%, for small effect sizes, studies used in our meta-analysis would need to recruit over 1000 individuals for both the case and control arms [Figure 5C]
+The tissue-based models had results that were dependent on genera or OTU level data and matched or unmatched samples [Figure S2].
+Similarily to stool, the significant OR taxa performed as well as both the full taxa and OTU models [Figure S3B].
+When analyzing the tests sets that were comprised of genera data from other studies, all models performed poorly with one exception being carcinoma classification in unmatched tissue [Figure S4].
+Unlike stool, the test sets for the significant OR taxa did not perform as well as the all genera-based models [Figure S4C].
+For the full genera models built using either matched or unmatched samples, showed no consistent taxa representation in the top 10 most important model variables across study [Figure S4A & S4B].
+Conversely, the OTU-based models built using either matched or unmatched samples showed more commonalities [Figure S4C & S4D].
+In models built using matched samples *Lachnospiraceae*, *Fusobacteriaceae*, *Comamonadaceae*, and *Bacteroidaceae* appeared in the top 10 percent in the majority of studies [Figure S4C].
+Although models built with unmatched samples also had *Lachnospiraceae* and *Bacteroidaceae*, a major difference was the presence of *Ruminococcaceae* in the top 10 of every study [Figure S4D].
+This results suggests that either the colon tissue microbiota is study or person dependent or that that kit and/or other types of contamination associated with low biomass samples may be skewing the results.
 

### 2018-02-23 submission/manuscript.tex
update overall results

@@
 
-\textbf{\emph{Lower Bacterial Diversity is Associated with Increased OR
-of Tumors:}} To assess differences in broad scale community metrics as
-disease severity worsens Operational Taxonomic Unit (OTU) richness,
-evenness, and Shannon diversity measurements were power transformed and
-Z-score normalized. These metrics are commonly used to assess the total
-number of OTUs, the equality of their abundance, and the overall
-diversity, respectively. Using linear mixed-effect models to control for
-study and variable region we assessed whether OTU richness, evenness, or
-Shannon diversity changed in a step-wise manner with disease severity.
-In stool, there was a significant decrease in both evenness and Shannon
-diversity as disease severity moved from control to adenoma to carcinoma
-(P-value = 0.025 and 0.043, respectively). We next tested whether the
-detectable differences in community significantly increased in OR of
-having an adenoma or carcinoma. For fecal samples, a decrease versus the
-overall median in evenness resulted in a significantly increased OR for
-carcinoma (OR = 1.66 (1.2 - 2.3), P-value = 0.0021) and adenoma (OR =
-1.3 (1.02 - 1.65), P-value = 0.035) while a decrease versus the overall
-median in Shannon diversity only increased the OR for carcinoma (OR =
-1.61 (1.14 - 2.28), P-value = 0.0069) {[}Figure 1{]}. Using the
-Bray-Curtis distance metric and PERMANOVA, it was also possible to
-identify significant bacterial community changes, in specific studies,
-for both carcinoma-associated and adenoma-associated microbiota versus
-control {[}Table S1{]}.
-
-Using similar transformations for tissue samples, linear mixed-effect
-models were used on the transformed combined data to control for study,
-re-sampling of the same individual, and 16S variable region to test
-whether OTU richness, evenness, or Shannon diversity changed in a
-step-wise manner as disease severity increased. For colon tissue, there
-were no significant changes in OTU richness, evenness, or Shannon
-diversity as disease severity progressed from control to adenoma to
-carcinoma (P-value \textgreater{} 0.05). We next analyzed the OR, for
-matched (unaffected tissue and an adenoma or carcinoma from the same
-individual) and unmatched (control and adenoma or carcinoma tissue not
-from the same individual) colon tissue samples. For individuals at
-either an adenoma or carcinoma stage of disease there was no significant
-change in OR based on lower than median values for OTU richness,
-evenness, and Shannon diversity {[}Figure S1 \& Table S2{]}. Similar to
-stool samples, significant differences in bacterial community, assessed
-by PERMANOVA, were identified in unmatched tissue samples, for those at
-either adenoma or carcinoma stage of CRC {[}Table S1{]}. For studies
-with matched samples no differences in bacterial community were observed
-when assessed with PERMANOVA {[}Table S1{]}. These tissue results
-suggest that the microbiota within an individual are similar to each
-other regardless of disease status.
-
-\textbf{\emph{Mouth-Associated Genera are Associated with an Increased
-OR of Tumor:}} Next, we asked if being higher than the median relative
-abundance, for any specific genera, resulted in an altered OR for
-adenoma or carcinoma, in stool and colon tissue, due to our previous
-observations of small increases in OR using OTU richness and Shannon
-diversity. To investigate this we analyzed all common genera across each
-study, in colon tissue or stool, and assessed whether a relative
-abundance higher than the median results in an increase or decrease in
-OR. For both tissue and stool samples only ORs associated with an
-increase or decrease in carcioma tumors were significant after multiple
-comparison correction {[}Table S3 \& S4{]}. Out of the 8 taxa that had
-significant ORs in stool samples 4 were mouth-associated microbes. These
-mouth-associated genera significantly increased the ORs of carcinoma for
-stool samples and included \emph{Fusobacterium}, \emph{Parvimonas},
-\emph{Porphyromonas}, and \emph{Peptostreptococcus} {[}Table S3{]}.
-Conversely, mouth-associated genera were not significantly associated
-with an increased OR for carcinoma in tissue samples {[}Table S4{]}.
-Only unmatched tissue samples had significant OR taxa and these were
-\emph{Dorea} (OR = 0.35 (0.22 - 0.55)), \emph{Blautia} (OR = 0.47 (0.3 -
-0.73)), and \emph{Weissella} (OR = 5.15 (2.02 - 13.14)). Overall, there
-was little direct overlap in increased or decreased OR taxa between both
-tumor type and sample site.
+\textbf{\emph{Lower Bacterial Diversity is Associated with Increased
+Odds Ratio (OR) of Tumors:}} We first assessed whether variation in
+broad community metrics like total number of operational taxonomic units
+(OTUs) (i.e.~richness), the evenness of their abundance, and the overall
+diversity was associated with disease stage after controlling for study
+and variable region differences. In stool, there was a significant
+decrease in both evenness and diversity as disease severity progressed
+from normal to adenoma to carcinoma (P-value = 0.025 and 0.043,
+respectively) {[}Figure 1{]}; there was not a significant difference for
+richness (P-value = 0.21). We next tested whether the decrease in these
+community metrics translated into significant ORs for having an adenoma
+or carcinoma. For fecal samples, the ORs for richness were not
+significantly greater than 1.0 for adenoma or carcinoma (P-value = 0.40)
+{[}Figure 2A{]}. The ORs for evenness were significantly higher than 1.0
+for adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) and carcinoma (OR
+= 1.66 (1.2 - 2.3), P-value = 0.0021) {[}Figure 2B{]}. The ORs for
+diversity were only significantly greater than 1.0 for carcinoma (OR =
+1.61 (1.14 - 2.28), P-value = 0.0069), but not for adenoma (P-value =
+0.11) {[}Figure 2C{]}. Although these OR are significantly greater than
+1.0, it is doubtful that these are clinically meaningful values.
+
+Similar to our analysis of sequences obtained from stool samples, we
+repeated the analysis using sequences obtained from colon tissue. There
+were no significant changes in richness, evenness, or diversity as
+disease severity progressed from control to adenoma to carcinoma
+(P-value \textgreater{} 0.05). We next analyzed the OR, for matched
+(i.e.~where unaffected tissue and tumors were obtained from the same
+individual) and unmatched (i.e.~where unaffected tissue and tumor tissue
+were not obtained from the same individual) tissue samples. The ORs for
+adenoma and carcinoma by any measure were not significantly different
+from 1.0 (P-value \textgreater{} 0.05) {[}Figure S1 \& Table S1{]}. This
+is likely due to the combination of a small effect size, as suggested
+from the results using stool, and the relatively small number of studies
+and size of studies used in the analysis.
+
+\textbf{\emph{Disease Progression is Associated with Community-Wide
+Changes in Composition and Abundance:}} Based on the differences in
+evenness and diversity, we next asked whether there were community-wide
+differences in the structure of the communities associated with
+different disease stages. We identified significant bacterial community
+differences in the stool of patients with adenomas relative to those
+with normal colons in 1 of 4 studies and in patients with carcinomas
+relative to those with normal colons in 6 of 7 studies (PERMANOVA;
+P-value \textless{} 0.05) {[}Table S2{]}. Similar to the analyses using
+stool samples, there were significant differences in bacterial community
+structure between subjects with normal colons and those with adenoma (1
+of 2 studies) and carcinoma (1 of 3 studies) {[}Table S2{]}. For studies
+that used matched samples no differences in bacterial community
+structures were observed {[}Table S2{]}. Combined, these results
+indicate that there consistent and significant community-wide changes in
+the fecal community structure of subjects with carcinomas. However, the
+signal observed in subjects with adenomas or when using tissue samples
+was not as consistent. This is likely due to a smaller effect size or
+the relatively small sample sizes among the studies that characterized
+the tissue microbiota.
+
+\textbf{\emph{Individual Taxa are Associated with Significant ORs for
+Carcinomas:}} Next we identified those taxa were associated with ORs
+that were significantly associated with having a normal colon or the
+presence of adenomas or carcinomas. No taxa had a significant OR for the
+presence of adenomas when we used data collected from stool or tissue
+samples (Table S3 \& S4). In contrast, 8 taxa had significant ORs for
+the presence of carcinomas using data from stool samples. Of these, 4
+are commonly associated with the oral cavity: Fusobacterium (OR = 2.74
+(1.95 - 3.85)), Parvimonas (OR = 3.07 (2.11 - 4.46)), Porphyromonas (OR
+= 3.2 (2.26 - 4.54)), and Peptostreptococcus (OR = 7.11 (3.84 - 13.17))
+{[}Table S3{]}. The other 4 were Clostridium XI (OR = 0.65 (0.49 -
+0.86)), Enterobacteriaceae (OR = 1.79 (1.33 - 2.41)), Escherichia (OR =
+2.15 (1.57 - 2.95)), and Ruminococcus (OR = 0.63 (0.48 - 0.83)). Among
+the data collected from tissue samples, only unmatched carcinoma samples
+had taxa with a significant OR. Those included Dorea (OR = 0.35 (0.22 -
+0.55)), Blautia (OR = 0.47 (0.3 - 0.73)), and Weissella (OR = 5.15 (2.02
+- 13.14)). Mouth-associated genera were not significantly associated
+with an increased OR for carcinoma in tissue samples {[}Table S4{]}. For
+example, Fusobacterium had an OR of 3.98 (1.19 - 13.24; however, due to
+the small number of studies and considerable variation in the data, the
+Benjimani-Hochberg-corrected P-value was 0.93 {[}Table S4{]}. It is
+interesting to note that Ruminococcus and members of Clostridium group
+XI in stool and Dorea and Blautia in tissue had ORs that were
+significantly less than 1.0, which suggests that these populations are
+protective against the development of carcinomas. Overall, there was no
+overlap in the taxa with significant OR between stool and tissue
+samples.
+
+\textbf{\emph{Individual Significant OR Taxa Classify Carcinoma
+Poorly:}} Since specific taxa were associated with a significant OR for
+carcinoma we tested whether they would also be good classifiers of
+carcinoma. For stool, the 8 significant taxa did no better at
+classifying those with normal colons versus those with carcinomas than
+chance {[}Figure 3A{]}. Likewise in unmathced tissue samples the 3
+significant taxa were no better than an AUC of 0.5 {[}Figure 3B{]}.
+These results suggest that although these taxa are significantly
+associated with a decreased or increased OR of carcinoma, individually
+they are poor classifiers of disease.
 
 \textbf{\emph{Select Community Models can Recapitulate Whole Community
-Models:}} Since specific genera increased the OR for carcinoma over
-diversity metrics we assessed whether the bacterial community was better
-at classifying disease versus only a select group of genera. We selected
-these genera based on significance after multiple comparison correction.
-If no taxa were significant after multiple comparison correction then no
-model for that specific grouping (i.e.~adenoma stool) was analyzed. We
-first tested three model AUCs. These models were created using Random
-Forest and where either all genera, all OTUs, or significant OR taxa
-only. Next, the all genera models and any significant OR taxa models
-were tested across all studies that were not used to train the model.
-For stool, all models used had similar AUCs {[}Figure 2{]}. Although for
-adenoma and unmatched carcinoma this trend held, there were large
-differences in matched tissue based on whether all genera or OTUs were
-utilized {[}Figure S2{]}. When analyzing the tests sets that were
-comprised of genera data from other studies, both the all genera and
-singificant OR taxa only models had a similar ability to detect
-individuals with carcinomas, for both stool and tissue samples {[}Figure
-3-S3{]}.
-
-In stool, the most common genera in the top 10 most important variables,
-in the full community models using all genera-based models, were
-\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} {[}Figure
-4{]}. Regardless of sample type, mouth-associated genera were present in
-models for carcinomas {[}Figure 4B \& D{]}. Yet, none were present in
-the majority of studies and none were present in the adenoma models
-{[}Figure 4A \& B{]}. For the full community OTU-based models,
-\emph{Ruminococcaceae} was present in the top 10 consistently for both
-adenoma and carcinoma models while \emph{Roseburia} was only present in
-many adenoma models and \emph{Bacteroides} was present in the
-overwhelming majority of the carcinoma models {[}Figure 4C \& 4D{]}.
-
-Unlike the stool-based Random Forest models, the tissue-based models,
-for the full genera from the first approach, showed no consistent
-representation of \emph{Ruminococcaceae}, \emph{Ruminococcus},
-\emph{Bacteroides}, and \emph{Roseburia} in the top 10 most important
-model variables across study {[}Figure S4{]}. The vast majority of the
-top 10 model variables for the full community genera- and OTU-based
-models using colon tissue tended to be study specific. Further, there
-was very little overlap in the top 10 important variables between
-adenoma and carcinoma stage models, regardless of whether colon tissue
-or stool was used {[}Figure S4{]}. This discordance between stool and
-colon tissue samples also applies to the mouth-associated genera with
-one noticeable skew being that \emph{Fusobacterium} and
-\emph{Fusobacteriaceae} occur more often in the top 10 of matched versus
-unmatched colon tissue Random Forest models {[}Figure S4B-C \& S4E-F{]}.
-This suggests that either the colon tissue microbiota is study or person
-dependent, that kit and/or other types of contamination associated with
-low biomass samples may be skewing the results, or that multiple
-microbes could act as the inflammatory stimulus needed to exacerbate
-mutations.
-
-\textbf{\emph{CRC Studies are Underpowered for Detecting Small Effect
-Sizes:}} Next, we assessed how much confidence should be placed in the
-reported outcomes from each individual study by calculating the ability
-to detect a difference (power) and sample size needed for small, medium,
-and large effect size differences between cases and controls. When
-assessing the power of each study at different effect sizes the majority
-of studies achieved 80\% power to detect a 30\% or greater difference
-between groups {[}Figure 5A \& B{]}. No study that we analyzed had the
-standard 80\% power to detect an effect size difference equal to or
-below 10\% {[}Figure 5A \& B{]}. In order to achieve a power of 80\%,
-for small effect sizes, studies used in our meta-analysis would need to
-recruit over 1000 individuals for both the case and control arms
-{[}Figure 5C{]}
+Models:}} Since specific taxa increased or decreased the OR for
+carcinoma but performed poorly as an individual classifier of carcinoma
+and assessing ORs for adenoma we assessed whether the overall bacterial
+community was better at classifying disease. Three models were tested
+and included a full taxa model, full OTU model, and significant OR taxa
+model. If no taxa were significant after multiple comparison correction
+then no model for that specific grouping (i.e.~adenoma stool) was
+analyzed. We first tested three model AUCs. Next, the all genera models
+and any significant OR taxa models were tested across all studies that
+were not used to train the model.
+
+For stool, all models used had similar AUCs for both adenoma and
+carcinoma {[}Figure 4{]}. However, the adenoma AUCs were barely better
+than chance {[}Figure 4A{]}. Similarily, when analyzing the tests sets
+that were comprised of genera data from other studies the adenoma models
+performed poorly {[}Figure 5A{]}. The carcinoma models performed much
+better, with both the all genera and singificant OR taxa only models
+having a similar ability to detect individuals with carcinomas {[}Figure
+5B{]}. The most common genera in the top 10 most important variables, in
+the full genera models used to classify carcinomas, were
+\emph{Bacteroides} and \emph{Lachnospiraceae} {[}Figure 6A{]}.
+Regardless of sample type, mouth-associated genera were present in
+models for carcinomas but not consistently across studies {[}Figure
+6{]}. For the full community OTU-based models, both \emph{Bacteroides},
+\emph{Blautia}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} were
+present in the top 10 consistently across studies {[}Figure 6B{]}.
+Overall, these results suggest that multiple microbes could act as the
+inflammatory stimulus needed to exacerbate mutations leading each
+individual microbe as a poor individual classifier but much better in
+aggregate with others.
+
+The tissue-based models had results that were dependent on genera or OTU
+level data and matched or unmatched samples {[}Figure S2{]}. Similarily
+to stool, the significant OR taxa performed as well as both the full
+taxa and OTU models {[}Figure S3B{]}. When analyzing the tests sets that
+were comprised of genera data from other studies, all models performed
+poorly with one exception being carcinoma classification in unmatched
+tissue {[}Figure S4{]}. Unlike stool, the test sets for the significant
+OR taxa did not perform as well as the all genera-based models {[}Figure
+S4C{]}. For the full genera models built using either matched or
+unmatched samples, showed no consistent taxa representation in the top
+10 most important model variables across study {[}Figure S4A \& S4B{]}.
+Conversely, the OTU-based models built using either matched or unmatched
+samples showed more commonalities {[}Figure S4C \& S4D{]}. In models
+built using matched samples \emph{Lachnospiraceae},
+\emph{Fusobacteriaceae}, \emph{Comamonadaceae}, and
+\emph{Bacteroidaceae} appeared in the top 10 percent in the majority of
+studies {[}Figure S4C{]}. Although models built with unmatched samples
+also had \emph{Lachnospiraceae} and \emph{Bacteroidaceae}, a major
+difference was the presence of \emph{Ruminococcaceae} in the top 10 of
+every study {[}Figure S4D{]}. This results suggests that either the
+colon tissue microbiota is study or person dependent or that that kit
+and/or other types of contamination associated with low biomass samples
+may be skewing the results.
 

### 2018-02-23 code/make_tissue_imp_otu_graph.R
update fig order

@@ -42,7 +42,7 @@ crc_unmatched_genera <- crc_unmatched_genera_occurances %>% arrange(occurance) %
   theme_bw() + 
   labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, crc_unmatched_sets)) + 
-  ggtitle("A") + 
+  ggtitle("B") + 
   scale_y_continuous(expand = c(0,0.03)) + 
   theme(axis.text.y = element_text(face = "italic"), 
         plot.title = element_text(face="bold", hjust = -0.35, size = 20), 
@@ -61,7 +61,7 @@ crc_matched_genera <- crc_matched_genera_occurances %>% arrange(occurance) %>%
   theme_bw() + 
   labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, crc_matched_sets)) + 
-  ggtitle("B") + 
+  ggtitle("A") + 
   scale_y_continuous(expand = c(0,0.03)) + 
   theme(axis.text.y = element_text(face = "italic"), 
         plot.title = element_text(face="bold", hjust = -0.20, size = 20), 
@@ -80,7 +80,7 @@ crc_unmatched_otu <- crc_unmatched_otu_occurances %>% arrange(occurance) %>%
   theme_bw() + 
   labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, crc_unmatched_sets)) + 
-  ggtitle("C") + 
+  ggtitle("D") + 
   scale_y_continuous(expand = c(0,0.03)) + 
   theme(axis.text.y = element_text(face = "italic"), 
         plot.title = element_text(face="bold", hjust = -0.35, size = 20), 
@@ -99,7 +99,7 @@ crc_matched_otu <- crc_matched_otu_occurances %>% arrange(occurance) %>%
   theme_bw() + 
   labs(x = "", y = "Occurrence Across Studies") + 
   coord_flip(ylim = c(0, crc_matched_sets)) + 
-  ggtitle("D") + 
+  ggtitle("C") + 
   scale_y_continuous(expand = c(0,0.03)) + 
   theme(axis.text.y = element_text(face = "italic"), 
         plot.title = element_text(face="bold", hjust = -0.35, size = 20), 
@@ -113,8 +113,8 @@ crc_matched_otu <- crc_matched_otu_occurances %>% arrange(occurance) %>%
 ##############################################################################################
 
 
[…]

### 2018-02-23 submission/manuscript.Rmd
fix figure legends

@@
 
-**Figure 1: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Stool.** A) Community-based odds ratio for adenoma.
+**Figure 1: Significant Bacterial Community Metrics for Adenoma or Carcinoma in Stool.** A) Adenoma evenness.
+B) Carcinoma evenness.
+C) Carcinoma Shannon diversity.
+Blue represents controls and red represents either adenoma (panel A) or carcinoma (panel B and C).
+The black lines represent the median value for each repsective group.
+
+**Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Stool.** A) Community-based odds ratio for adenoma.
 B) Community-based odds ratio for carcinoma.
@@
 
-**Figure 2: Stool Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
+**Figure 3: The AUC of Indivdiual Significant OR Taxa to classify Carcinoma.** A) Stool samples.
+B) Unmatched tissue samples.
+The larger circle represents the median AUC of all studies and the smaller circles represent the individual AUC for a particular study.
+The dotted line denotes an AUC of 0.5.
+
+**Figure 4: Stool Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
 B) Carcinoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
@@
 
-**Figure 3: Stool Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
+**Figure 5: Stool Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
 B) Test AUCs of carcinoma models using all genera or significant OR taxa only.
@@
 
-**Figure 4: Most Common Taxa Across Full Community Stool Study Models.** A) Common taxa in the top 10 for adenoma Random Forest all genera-based models.
-B) Common taxa in the top 10 for carcinoma Random Forest all genera-based models.
-C) Common taxa in the top 10 for adenoma Random Forest all OTU-based models.
-D) Common genera in the top 10 for carcinoma Random Forest all OTU-based models.
-
-**Figure 5: Power and Effect Size Analysis of Studies Included.** A) Power based on effect size for studies with adenoma individuals.
-B) Power based on effect size for studies with carcinoma individuals.
-C) The estimated sample number needed for each arm of each study to detect an effect size of 1-30%.
-The dotted red lines in A) and B) represent a power of 0.8.
+**Figure 6: Most Common Taxa Across Carcinoma Full Community Stool Study Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest all taxa-based models.
+B) Common taxa in the top 10 percent for carcinoma Random Forest all OTU-based models.
 
@@
 
-**Figure S4: Most Common Genera Across Full Community Tissue Study Models.** A) Common genera in the top 10 for adenoma Random Forest for all genera-based models.
-B) Common genera in the top 10 for unmatched carcinoma Random Forest for all genera-based models.
-B) Common genera in the top 10 for matched carcinoma Random Forest for all genera-based models.
-D) Common genera in the top 10 for adenoma Random Forest for all OTU-based models.
-E) Common genera in the top 10 for unmatched carcinoma Random Forest for all OTU-based models.
-F) Common genera in the top 10 for matched carcinoma Random Forest for all OTU-based models.
+**Figure S4: Most Common Genera Across Full Community Tissue Study Models.** A) Common genera in the top 10 percent for matched carcinoma Random Forest all genera-based models.
+B) Common genera in the top 10 percent for unmatched carcinoma Random Forest all genera-based models.
+C) Common genera in the top 10 percent for matched carcinoma Random Forest all OTU-based models.
+D) Common genera in the top 10 percent for unmatched carcinoma Random Forest all OTU-based models.
 

### 2018-02-23 submission/manuscript.tex
fix figure legends

@@
 
-\textbf{Figure 1: Odds Ratio for Adenoma or Carcinoma based on Bacterial
+\textbf{Figure 1: Significant Bacterial Community Metrics for Adenoma or
+Carcinoma in Stool.} A) Adenoma evenness. B) Carcinoma evenness. C)
+Carcinoma Shannon diversity. Blue represents controls and red represents
+either adenoma (panel A) or carcinoma (panel B and C). The black lines
+represent the median value for each repsective group.
+
+\textbf{Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial
 Community Metrics in Stool.} A) Community-based odds ratio for adenoma.
@@
 
-\textbf{Figure 2: Stool Random Forest Model Train AUCs.} A) Adenoma
+\textbf{Figure 3: The AUC of Indivdiual Significant OR Taxa to classify
+Carcinoma.} A) Stool samples. B) Unmatched tissue samples. The larger
+circle represents the median AUC of all studies and the smaller circles
+represent the individual AUC for a particular study. The dotted line
+denotes an AUC of 0.5.
+
+\textbf{Figure 4: Stool Random Forest Model Train AUCs.} A) Adenoma
 random forest model AUCs between all genera, all OTU, and select model
@@
 
-\textbf{Figure 3: Stool Random Forest Genus-Based Model Test AUCs.} A)
+\textbf{Figure 5: Stool Random Forest Genus-Based Model Test AUCs.} A)
 Test AUCs of adenoma models using all genera across study. B) Test AUCs
@@
 
-\textbf{Figure 4: Most Common Taxa Across Full Community Stool Study
-Models.} A) Common taxa in the top 10 for adenoma Random Forest all
-genera-based models. B) Common taxa in the top 10 for carcinoma Random
-Forest all genera-based models. C) Common taxa in the top 10 for adenoma
-Random Forest all OTU-based models. D) Common genera in the top 10 for
+\textbf{Figure 6: Most Common Taxa Across Carcinoma Full Community Stool
+Study Models.} A) Common taxa in the top 10 percent for carcinoma Random
+Forest all taxa-based models. B) Common taxa in the top 10 percent for
 carcinoma Random Forest all OTU-based models.
-
-\textbf{Figure 5: Power and Effect Size Analysis of Studies Included.}
-A) Power based on effect size for studies with adenoma individuals. B)
-Power based on effect size for studies with carcinoma individuals. C)
-The estimated sample number needed for each arm of each study to detect
-an effect size of 1-30\%. The dotted red lines in A) and B) represent a
-power of 0.8.
 
@@
 \textbf{Figure S4: Most Common Genera Across Full Community Tissue Study
-Models.} A) Common genera in the top 10 for adenoma Random Forest for
-all genera-based models. B) Common genera in the top 10 for unmatched
-carcinoma Random Forest for all genera-based models. B) Common genera in
-the top 10 for matched carcinoma Random Forest for all genera-based
-models. D) Common genera in the top 10 for adenoma Random Forest for all
-OTU-based models. E) Common genera in the top 10 for unmatched carcinoma
-Random Forest for all OTU-based models. F) Common genera in the top 10
-for matched carcinoma Random Forest for all OTU-based models.
+Models.} A) Common genera in the top 10 percent for matched carcinoma
+Random Forest all genera-based models. B) Common genera in the top 10
+percent for unmatched carcinoma Random Forest all genera-based models.
+C) Common genera in the top 10 percent for matched carcinoma Random
+Forest all OTU-based models. D) Common genera in the top 10 percent for
+unmatched carcinoma Random Forest all OTU-based models.
 

### 2018-02-23 submission/manuscript.Rmd
update methods

@@
 
-***Statistical Analysis:***
-All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
-For OTU richness, evenness, and Shannon diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
-Testing for OTU richness, evenness, and Shannon diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hyper-variable region used.
-Odds ratios (OR) were analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study.
-OR significance testing utilized the chi-squared test.
-Diversity differences measured by the Bray-Curtis index utilized the creation of distance matrix and testing with PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017].
-Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002].
-Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016].
-All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].
-
 ***Study Analysis Overview:***
@@
 Next, all common genera were assessed for differences in ORs for individuals having an adenoma or carcinoma and corrected for multiple comparisons using the Benjamini-Hochberg method [@benjamini_controlling_1995].
-We then built Random Forest models based on all genera, all OTUs, or significant OR taxa (if any were present after multiple comparison correction).
+We then built Random Forest models based on all genera, all OTUs, or significant OR taxa (only using taxa still significant after multiple comparison correction).
 For both the full genera and significant OR taxa, models were trained on one study then tested on the remaining studies using genera-based relative abundances.
@@
 Comparisons of the initial trained model AUCs for the full genera and significant OR taxa were made to the mean AUC generated from the 100 different 10-fold CV runs of the respective OTU-based model.
-Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power.
 For comparisons in which only control versus adenoma individuals were made, the carcinoma individuals were excluded from each respective study.
@@
 Finally, the two studies that had adenoma tumor tissue (Lu and Flemer) were equally divided between matched and unmatched studies and were grouped together for the counting of the top 10 genera and OTUs for both Random Forest approaches.
+
+***Statistical Analysis:***
+All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
+For OTU richness, evenness, and Shannon diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
+Testing for OTU richness, evenness, and Shannon diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hyper-variable region used.
+Odds ratios (OR) were analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study.
+OR significance testing utilized the chi-squared test.
+Diversity differences measured by the Bray-Curtis index utilized the creation of distance matrix and testing with PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017].
+Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002].
+All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].
 

### 2018-02-23 submission/manuscript.tex
update methods

@@
 
-\textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package {[}39{]}.
-For OTU richness, evenness, and Shannon diversity analysis, values were
-power transformed using the rcompanion (v1.11.1) package {[}40{]} and
-then Z-score normalized using the car (v2.1.6) package {[}41{]}. Testing
-for OTU richness, evenness, and Shannon diversity differences utilized
-linear mixed-effect models created using the lme4 (v1.1.15) package
-{[}42{]} to correct for study, repeat sampling of individuals (tissue
-only), and 16S hyper-variable region used. Odds ratios (OR) were
-analyzed using both the epiR (v0.9.93) and metafor (v2.0.0) packages
-{[}43,44{]} by assessing how many individuals with and without disease
-were above and below the overall median value within each specific
-study. OR significance testing utilized the chi-squared test. Diversity
-differences measured by the Bray-Curtis index utilized the creation of
-distance matrix and testing with PERMANOVA executed with the vegan
-(v2.4.5) package {[}45{]}. Random Forest models were built using both
-the caret (v6.0.78) and randomForest (v4.6.12) packages {[}46,47{]}.
-Power analysis and estimations were made using the pwr (v1.2.1) and
-statmod (v1.4.30) packages {[}48,49{]}. All figures were created using
-both ggplot2 (v2.2.1) and gridExtra (v2.3) packages {[}50,51{]}.
-
 \textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
@@
 ORs for individuals having an adenoma or carcinoma and corrected for
-multiple comparisons using the Benjamini-Hochberg method {[}52{]}. We
+multiple comparisons using the Benjamini-Hochberg method {[}39{]}. We
 then built Random Forest models based on all genera, all OTUs, or
-significant OR taxa (if any were present after multiple comparison
-correction). For both the full genera and significant OR taxa, models
-were trained on one study then tested on the remaining studies using
-genera-based relative abundances. The OTU-based models were built using
-OTU level data and a 10-fold CV over 100 different iterations, based on
-random 80/20 splitting of the data, was used to generate a range of
-expected AUCs. This process was repeated for every study in the
-meta-analysis. Comparisons of the initial trained model AUCs for the
-full genera and significant OR taxa were made to the mean AUC generated
-from the 100 different 10-fold CV runs of the respective OTU-based
-model. Finally, the power of each study was assessed for an effect size
-ranging from 1\% to 30\% and an estimated sample size, for these effect
-sizes, was generated based on 80\% power. For comparisons in which only
-control versus adenoma individuals were made, the carcinoma individuals
-were excluded from each respective study. Similarly, for comparisons in
-which control versus carcinoma individuals were made the adenoma
-individuals were excluded from each respective study. For all analysis
-completed fecal and tissue samples were kept separate. Within the tissue
-groups the data were further divided between samples from the same
-individual (matched) and those from different individuals (unmatched).
+significant OR taxa (only using taxa still significant after multiple
+comparison correction). For both the full genera and significant OR
+taxa, models were trained on one study then tested on the remaining
+studies using genera-based relative abundances. The OTU-based models
+were built using OTU level data and a 10-fold CV over 100 different
+iterations, based on random 80/20 splitting of the data, was used to
+generate a range of expected AUCs. This process was repeated for every
+study in the meta-analysis. Comparisons of the initial trained model
+AUCs for the full genera and significant OR taxa were made to the mean
+AUC generated from the 100 different 10-fold CV runs of the respective
+OTU-based model. For comparisons in which only control versus adenoma
+individuals were made, the carcinoma individuals were excluded from each
+respective study. Similarly, for comparisons in which control versus
+carcinoma individuals were made the adenoma individuals were excluded
+from each respective study. For all analysis completed fecal and tissue
+samples were kept separate. Within the tissue groups the data were
+further divided between samples from the same individual (matched) and
+those from different individuals (unmatched).
 
@@
 
+\textbf{\emph{Statistical Analysis:}} All statistical analysis after
+sequence processing utilized the R (v3.4.3) software package {[}40{]}.
+For OTU richness, evenness, and Shannon diversity analysis, values were
+power transformed using the rcompanion (v1.11.1) package {[}41{]} and
+then Z-score normalized using the car (v2.1.6) package {[}42{]}. Testing
+for OTU richness, evenness, and Shannon diversity differences utilized
+linear mixed-effect models created using the lme4 (v1.1.15) package
+{[}43{]} to correct for study, repeat sampling of individuals (tissue
+only), and 16S hyper-variable region used. Odds ratios (OR) were
+analyzed using both the epiR (v0.9.93) and metafor (v2.0.0) packages
+{[}44,45{]} by assessing how many individuals with and without disease
+were above and below the overall median value within each specific
+study. OR significance testing utilized the chi-squared test. Diversity
+differences measured by the Bray-Curtis index utilized the creation of
+distance matrix and testing with PERMANOVA executed with the vegan
+(v2.4.5) package {[}46{]}. Random Forest models were built using both
+the caret (v6.0.78) and randomForest (v4.6.12) packages {[}47,48{]}. All
+figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
+packages {[}49,50{]}.
+
 \textbf{\emph{Reproducible Methods:}} The code and analysis can be found
@@
 
+\hypertarget{ref-benjamini_controlling_1995}{}
+39. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
+practical and powerful approach to multiple testing. Journal of the
+Royal Statistical Society Series B (Methodological). 1995;57:289--300.
+
 \hypertarget{ref-r_citation_2017}{}
-39. R Core Team. R: A language and environment for statistical computing
+40. R Core Team. R: A language and environment for statistical computing
 {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-40. Mangiafico S. Rcompanion: Functions to support extension education
+41. Mangiafico S. Rcompanion: Functions to support extension education
 program evaluation {[}Internet{]}. 2017. Available from:
@@
 \hypertarget{ref-car_citation_2011}{}
-41. Fox J, Weisberg S. An R companion to applied regression
+42. Fox J, Weisberg S. An R companion to applied regression
 {[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
@@
 \hypertarget{ref-lme4_citation_2015}{}
-42. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
+43. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
 models using lme4. Journal of Statistical Software. 2015;67:1--48.
@@
 \hypertarget{ref-epir_citation_2017}{}
-43. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
+44. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
 J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
@@
 \hypertarget{ref-metafor_citation_2010}{}
-44. Viechtbauer W. Conducting meta-analyses in R with the metafor
+45. Viechtbauer W. Conducting meta-analyses in R with the metafor
 package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-45. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+46. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
@@
 \hypertarget{ref-caret_citation_2017}{}
-46. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
+47. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
 Cooper T, et al. Caret: Classification and regression training
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-47. Liaw A, Wiener M. Classification and regression by randomForest. R
+48. Liaw A, Wiener M. Classification and regression by randomForest. R
 News {[}Internet{]}. 2002;2:18--22. Available from:
@@
 
-\hypertarget{ref-pwr_citation_2017}{}
-48. Champely S. Pwr: Basic functions for power analysis {[}Internet{]}.
-2017. Available from: \url{https://CRAN.R-project.org/package=pwr}
-
-\hypertarget{ref-statmod_citation_2016}{}
-49. Giner G, Smyth GK. Statmod: Probability calculations for the inverse
-gaussian distribution. R Journal. 2016;8:339--51.
-
 \hypertarget{ref-ggplot2_citation_2009}{}
-50. Wickham H. Ggplot2: Elegant graphics for data analysis
+49. Wickham H. Ggplot2: Elegant graphics for data analysis
 {[}Internet{]}. Springer-Verlag New York; 2009. Available from:
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-51. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
+50. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
 {[}Internet{]}. 2017. Available from:
 \url{https://CRAN.R-project.org/package=gridExtra}
-
-\hypertarget{ref-benjamini_controlling_1995}{}
-52. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
-practical and powerful approach to multiple testing. Journal of the
-Royal Statistical Society Series B (Methodological). 1995;57:289--300.
 

### 2018-02-28 code/make_rf_auc_full_versus_specific_graph.R
fix order and jitter

@@ -111,19 +111,19 @@ adn_tissue_graph <- adn_tissue %>%
                        levels = c("unmatched", "matched"), 
                        labels = c("Unmatched Tissue", "Matched Tissue")), 
          model_type = factor(model_type, 
-                             levels = c("full_model", "full_otu", "select_model"), 
-                             labels = c("All Genera", "All OTUs", "Significant\nOR Taxa")), 
+                             levels = c("select_model", "full_model", "full_otu"), 
+                             labels = c("Significant\nOR Taxa", "All Taxa", "All OTUs")), 
          study = factor(study, 
                         levels = c("flemer", "lu"), 
                         labels = c("Flemer", "Lu"))) %>% 
   ggplot(aes(model_type, as.numeric(AUC), color = study, group = type)) + 
   stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 
                colour = "black", geom = "crossbar", size = 0.5, width = 0.5) +
-  geom_point(size = 3.5, show.legend = T) + coord_cartesian(ylim = c(0, 1.05)) + 
+  geom_point(size = 3.5, position = position_dodge(width = 0.3), show.legend = T) + 
+  coord_cartesian(ylim = c(0, 1.05)) + 
   facet_grid(. ~ type) + 
   labs(x = "", y = "Model AUC") + theme_bw() + ggtitle("A") + 
-  scale_color_manual(name = "Study", 
-                     values = c('#ED9121', '#8B7500')) + 
+  scale_color_manual(values = c('#ED9121', '#8B7500')) + 
   annotate("text", label = paste("Adenoma"), x = 0.7, y = 1.07, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
         legend.position = "bottom", 
@@ -138,18 +138,18 @@ adn_tissue_graph <- adn_tissue %>%
 adn_stool_graph <- adn_stool %>% 
   filter(study == train_model) %>% 
   mutate(model_type = factor(model_type, 
-                             levels = c("full_model", "full_otu", "select_model"), 
-                             labels = c("All Genera", "All OTUs", "Significant\nOR Taxa")), 
+                             levels = c("select_model", "full_model", "full_otu"), 
+                             labels = c("Significant\nOR Taxa", "All Taxa", "All OTUs")), 
[…]

### 2018-02-28 code/make_genus_rf_auc_against_study_graph.R
update order of groups

@@ -117,8 +117,8 @@ adn_tissue_graph <- adn_all_tissue %>%
                         levels = c("flemer", "lu"), 
                         labels = c("Flemer", "Lu\n(Matched)")), 
          model_type = factor(model_type, 
-                             levels = c("full", "select"), 
-                             labels = c("All Genera", "Significant\nOR Taxa")), 
+                             levels = c("select", "full"), 
+                             labels = c("Significant\nOR Taxa", "All Taxa")), 
          study = factor(study, 
                         levels = c("flemer", "lu"), 
                         labels = c("Flemer", "Lu\n(Matched)"))) %>% 
@@ -149,8 +149,8 @@ adn_stool_graph <- adn_all_stool %>%
                         levels = c("baxter", "brim", "hale", "zeller"), 
                         labels = c("Baxter", "Brim", "Hale", "Zeller")), 
          model_type = factor(model_type, 
-                             levels = c("full", "select"), 
-                             labels = c("All Genera", "Significant\nOR Taxa")), 
+                             levels = c("select", "full"), 
+                             labels = c("Significant\nOR Taxa", "All Taxa")), 
          study = factor(study, 
                         levels = c("baxter", "brim", "hale", "zeller"), 
                         labels = c("Baxter", "Brim", "Hale", "Zeller"))) %>% 
@@ -181,8 +181,8 @@ crc_unmatched_tissue_graph <- crc_all_unmatched_tissue %>%
                         levels = c("burns", "chen", "flemer", "sana"), 
                         labels = c("Burns", "Chen", "Flemer", "Sanapareddy")), 
          model_type = factor(model_type, 
-                             levels = c("full", "select"), 
-                             labels = c("All Genera", "Significant\nOR Taxa")), 
+                             levels = c("select", "full"), 
+                             labels = c("Significant\nOR Taxa", "All Taxa")), 
          study = factor(study, 
                         levels = c("burns", "chen", "flemer", "sana"), 
                         labels = c("Burns", "Chen", "Flemer", "Sanapareddy"))) %>% 
@@ -213,8 +213,8 @@ crc_matched_tissue_graph <- crc_all_matched_tissue %>%
                         levels = c("burns", "dejea", "geng"), 
                         labels = c("Burns", "Dejea", "Geng")), 
          model_type = factor(model_type, 
-                             levels = c("full", "select"), 
-                             labels = c("All Genera", "Significant\nOR Taxa")), 
[…]

### 2018-02-28 code/run_comparison_RF_imp_otus_stool.R
add and update data tables for taxa in full models

@@ -17,7 +17,7 @@ tissue_sets <- c("dejea", "sana", "burns", "geng")
 # Hale, Wang, Brim, Weir, Ahn, Zeller, Baxter
 # Ignore brim since it only has polyps
 # Ignore weir since none have nzv
-stool_sets <- c("wang", "ahn", "zeller", "baxter", "hale")
+stool_sets <- c("wang", "ahn", "zeller", "baxter", "hale", "weir")
 adn_stool_sets <- c("brim", "zeller", "baxter", "hale")
 
 # Both Tissue and Stool
@@ -86,6 +86,20 @@ get_occurances <- function(study_vector, dataList, lowest_var, var_of_int){
 }
 
 
+# Function to generate the top 10% of variables
+get_top10_percent <- function(study, dataList){
+  
+  tempData <- dataList[[study]]
+  
+  total_vars_to_select <- round(length(rownames(tempData)), digits = 0)
+  
+  selected_tempData <- tempData %>% 
+    slice(1:total_vars_to_select) %>% 
+    mutate(study = study)
+  
+  return(selected_tempData)
+}
+
 
 ##############################################################################################
 ########################## Code used to run the analysis  ####################################
@@ -117,11 +131,30 @@ adn_genera_occurances <- get_occurances(adn_stool_sets, adn_imp_genera, lowest_v
 crc_otu_occurances <- get_occurances(c(stool_sets, "flemer"), crc_imp_otu_data, lowest_var = 10, "genus")
 adn_otu_occurances <- get_occurances(adn_stool_sets, adn_imp_otu_data, lowest_var = 10, "genus")
 
-# Read out data tables
+# Generate the combined top 10% tables
+top10_crc_genera <- sapply(c(stool_sets, "flemer"), 
+                           function(x) get_top10_percent(x, crc_imp_genera), simplify = F) %>% bind_rows()
+
+top10_adn_genera <- sapply(adn_stool_sets, 
+                           function(x) get_top10_percent(x, adn_imp_genera), simplify = F) %>% bind_rows()
+
+top10_crc_otu <- sapply(c(stool_sets, "flemer"), 
+                           function(x) get_top10_percent(x, crc_imp_otu_data), simplify = F) %>% bind_rows()
+
+top10_adn_otu <- sapply(adn_stool_sets, 
[…]

### 2018-02-28 code/run_comparison_RF_imp_otus_tissue.R
update tables

@@ -101,6 +101,20 @@ get_occurances <- function(study_vector, dataList, lowest_var, var_of_int){
 }
 
 
+# Function to generate the top 10% of variables
+get_top10_percent <- function(study, dataList){
+  
+  tempData <- dataList[[study]]
+  
+  total_vars_to_select <- round(length(rownames(tempData)), digits = 0)
+  
+  selected_tempData <- tempData %>% 
+    slice(1:total_vars_to_select) %>% 
+    mutate(study = study)
+  
+  return(selected_tempData)
+}
+
 
 ##############################################################################################
 ########################## Code used to run the analysis  ####################################
@@ -143,6 +157,29 @@ crc_unmatched_otu_occurances <- get_occurances(unmatched_studies, crc_unmatched_
 crc_matched_otu_occurances <- get_occurances(matched_studies, crc_matched_imp_otu_data, lowest_var = 10, "genus")
 adn_otu_occurances <- get_occurances(adn_tissue, adn_imp_otu_data, lowest_var = 10, "genus")
 
+
+# Generate the combined top 10% tables
+top10_crc_unmatched_genera <- sapply(unmatched_studies, 
+                           function(x) get_top10_percent(x, crc_unmatched_imp_genera), simplify = F) %>% bind_rows()
+
+top10_crc_matched_genera <- sapply(matched_studies, 
+                                     function(x) get_top10_percent(x, crc_matched_imp_genera), simplify = F) %>% bind_rows()
+
+top10_adn_genera <- sapply(adn_tissue, 
+                           function(x) get_top10_percent(x, adn_imp_genera), simplify = F) %>% bind_rows()
+
+
+top10_crc_unmatched_otu <- sapply(unmatched_studies, 
+                                     function(x) get_top10_percent(x, crc_unmatched_imp_otu_data), 
+                                  simplify = F) %>% bind_rows()
+
+top10_crc_matched_otu <- sapply(matched_studies, 
+                                   function(x) get_top10_percent(x, crc_matched_imp_otu_data), 
+                                simplify = F) %>% bind_rows()
+
+top10_adn_otu <- sapply(adn_tissue, 
+                           function(x) get_top10_percent(x, adn_imp_otu_data), simplify = F) %>% bind_rows()
+
 # Read out data tables
[…]

### 2018-02-28 code/run_comparison_RF_imp_otus_tissue.R
complete update

@@ -180,7 +180,7 @@ top10_crc_matched_otu <- sapply(matched_studies,
 top10_adn_otu <- sapply(adn_tissue, 
                            function(x) get_top10_percent(x, adn_imp_otu_data), simplify = F) %>% bind_rows()
 
-# Read out data tables
+# Read out data tables for occurences
 write_csv(crc_unmatched_genera_occurances, "data/process/tables/crc_RF_genera_unmatched_tissue_top10.csv")
 write_csv(crc_matched_genera_occurances, "data/process/tables/adn_RF_genera_matched_tissue_top10.csv")
 write_csv(adn_genera_occurances, "data/process/tables/adn_RF_genera_tissue_top10.csv")
@@ -189,7 +189,12 @@ write_csv(crc_matched_otu_occurances, "data/process/tables/adn_RF_otu_matched_ti
 write_csv(adn_otu_occurances, "data/process/tables/adn_RF_otu_tissue_top10.csv")
 
 # Read out tables for heat maps
-
+write_csv(top10_crc_unmatched_genera, "data/process/tables/crc_RF_genera_unmatched_tissue_top10_mda.csv")
+write_csv(top10_crc_matched_genera, "data/process/tables/adn_RF_genera_matched_tissue_top10_mda.csv")
+write_csv(top10_adn_genera, "data/process/tables/adn_RF_genera_tissue_top10_mda.csv")
+write_csv(top10_crc_unmatched_otu, "data/process/tables/crc_RF_otu_unmatched_tissue_top10_mda.csv")
+write_csv(top10_crc_matched_otu, "data/process/tables/adn_RF_otu_matched_tissue_top10_mda.csv")
+write_csv(top10_adn_otu, "data/process/tables/adn_RF_otu_tissue_top10_mda.csv")
 
 
 

### 2018-02-28 code/run_non_common_RF_genera_tissue.R
update select mda data

@@ -361,7 +361,8 @@ get_imp_otu_data <- function(i, a_modelList){
     summarise(mda_median = median(Overall), 
               iqr25 = quantile(Overall)["25%"], 
               iqr75 = quantile(Overall)["75%"]) %>% 
-    arrange(desc(mda_median))
+    arrange(desc(mda_median)) %>% 
+    mutate(study = i)
   
   return(tempData)
 }
@@ -477,6 +478,12 @@ select_unmatched_matched_genera_list <- sapply(names(unmatched_stool_study_data)
                                                  unmatched_rf_datasets[[x]] %>% 
                                                  select(c("disease", combined_genera)), simplify = F)
 
+# Generate important vars in select model
+selected_training_imp_model_vars <- sapply(unmatched_studies, 
+                                           function(x) get_imp_otu_data(x, select_unmatched_matched_genera_list), 
+                                           simplify = F) %>% bind_rows()
+
+
 # Generate training data
 select_unmatched_rf_training_data <- sapply(
   unmatched_studies, function(x) create_training_data(x, select_unmatched_matched_genera_list), simplify = F)
@@ -556,6 +563,9 @@ sapply(matched_studies,
        function(x) write.csv(matched_training_imp_model_vars[[x]], 
                              paste("data/process/tables/ALL_genus_matched_tissue_RF_", 
                                    x, "_imp_vars.csv", sep = ""), row.names = F))
+
+write_csv(selected_training_imp_model_vars, 
+          "data/process/tables/ALL_genus_unmatched_tissue_RF_select_imp_vars.csv")
 
 #sapply(unmatched_studies, 
 #       function(x) write.csv(select_unmatched_train_test_pvalues[[x]], 

### 2018-02-28 submission/manuscript.Rmd
fix table errors

@@
 
-colnames(table1) <- c("Study", "Data Stored", "16S Region", "Control (n)", "Adenoma (n)", "Carcinoma (n)")
+colnames(table1) <- c("Study", "Data Stored", "Region", "Control (n)", "Adenoma (n)", "Carcinoma (n)")
 
@@
 
-variable_region <- c("V5-6", "V1-V3", "V3-5", "V3-4", "V1-2", "V3-4", "V1-2")
+variable_region <- c("V5-6", "V1-3", "V3-5", "V3-4", "V1-2", "V3-4", "V1-2")
 
@@
 
-colnames(table2) <- c("Study", "Data Stored", "16S Region", "Control (n)", "Adenoma (n)", "Carcinoma (n)")
+colnames(table2) <- c("Study", "Data Stored", "Region", "Control (n)", "Adenoma (n)", "Carcinoma (n)")
 

### 2018-02-28 submission/manuscript.tex
fix table errors

@@
 \toprule
-Study & Data Stored & 16S Region & Control (n) & Adenoma (n) & Carcinoma
+Study & Data Stored & Region & Control (n) & Adenoma (n) & Carcinoma
 (n)\tabularnewline
@@
 Burns & SRA & V5-6 & 18 & 0 & 16\tabularnewline
-Chen & SRA & V1-V3 & 9 & 0 & 9\tabularnewline
+Chen & SRA & V1-3 & 9 & 0 & 9\tabularnewline
 Dejea & SRA & V3-5 & 31 & 0 & 32\tabularnewline

### 2018-03-01 submission/manuscript.Rmd
update results with pat edits

@@
 crc_matched_OR_tissue <- as.data.frame(read_csv("../data/process/tables/select_genus_OR_matched_tissue_composite.csv") %>% mutate(bh = p.adjust(pvalue, method = "BH")))
+
+#Load in AUC data for individual genera
+ind_crc_stool_auc <- read_csv("../data/process/tables/ind_genera_auc_stool.csv") %>%  mutate(auc = ifelse(taxa == "Ruminococcus" | taxa == "Clostridium_XI", invisible(1-auc), invisible(auc)))
+ind_crc_combined_stool_auc <- ind_crc_stool_auc %>% group_by(taxa) %>% summarise(auc = median(auc)) %>% mutate(study = "median") %>% ungroup()
+
+ind_unmatched_tissue_crc_auc <- read_csv("../data/process/tables/ind_genera_auc_unmatched_tissue.csv") %>% mutate(auc = ifelse(taxa == "Dorea" | taxa == "Blautia", invisible(1-auc), invisible(auc)))
+ind_crc_combined_unmatched_auc <- ind_unmatched_tissue_crc_auc %>% group_by(taxa) %>% summarise(auc = median(auc)) %>% 
+  mutate(study = "median") %>% ungroup()
+
+#Load in AUC data between select and full taxa models
+adn_tissue_model_aucs <- 
+  as.data.frame(read_csv("../data/process/tables/adn_ALL_genus_unmatched_tissue_RF_fullvsselect_pvalue_summary.csv") %>% 
+  mutate(type = rep(c("unmatched", "matched"), 2), 
+         pvalue = as.numeric(pvalue)) %>% 
+  gather(key = model_type, value = AUC, full_model, select_model) %>% 
+  mutate(AUC = as.numeric(AUC)) %>% 
+  bind_rows(
+    read_csv("../data/process/tables/adn_tissue_rf_otu_random_comparison_summary.csv") %>% 
+      mutate(type = c("matched", "unmatched"), 
+             model_type = "full_otu", 
+             train_model = study, 
+             act_mean_auc = as.numeric(act_mean_auc)) %>% 
+      rename(AUC = act_mean_auc) %>% 
+      select(pvalue, study, train_model, type, model_type, AUC)))
+
+adn_stool_model_aucs <- as.data.frame(read_csv("../data/process/tables/adn_ALL_genus_stool_RF_fullvsselect_pvalue_summary.csv") %>% gather(key = model_type, value = AUC, full_model, select_model) %>% 
+  mutate(AUC = as.numeric(AUC), 
+         pvalue = as.numeric(pvalue)) %>% 
+  bind_rows(read_csv("../data/process/tables/adn_stool_rf_otu_random_comparison_summary.csv") %>% 
+              mutate(model_type = "full_otu", 
+                     train_model = study, 
+                     act_mean_auc = as.numeric(act_mean_auc)) %>% 
+              rename(AUC = act_mean_auc) %>% 
+              select(pvalue, study, train_model, model_type, AUC)))
+
+crc_tissue_model_aucs <- as.data.frame(read_csv("../data/process/tables/ALL_genus_unmatched_tissue_RF_fullvsselect_pvalue_summary.csv") %>% mutate(type = "unmatched") %>% 
+  bind_rows(read_csv("../data/process/tables/ALL_genus_matched_tissue_RF_fullvsselect_pvalue_summary.csv") %>% 
+  mutate(type = "matched")) %>% 
+  gather(key = model_type, value = AUC, full_model, select_model) %>% 
+  mutate(AUC = as.numeric(AUC)) %>% 
+  bind_rows(
+    read_csv("../data/process/tables/matched_tissue_rf_otu_random_comparison_summary.csv") %>% 
+      mutate(type = "matched", 
+             model_type = "full_otu", 
+             train_model = study, 
+             act_mean_auc = as.numeric(act_mean_auc)) %>% 
+      rename(AUC = act_mean_auc) %>% 
+      select(pvalue, study, train_model, type, model_type, AUC) %>% 
+      bind_rows(
+        read_csv("../data/process/tables/unmatched_tissue_rf_otu_random_comparison_summary.csv") %>% 
+          mutate(type = "unmatched", 
+                 model_type = "full_otu", 
+                 train_model = study, 
+                 act_mean_auc = as.numeric(act_mean_auc)) %>% 
+          rename(AUC = act_mean_auc) %>% 
+          select(pvalue, study, train_model, type, model_type, AUC))))
+
+crc_stool_model_aucs <- as.data.frame(read_csv("../data/process/tables/ALL_genus_stool_RF_fullvsselect_pvalue_summary.csv") %>% gather(key = model_type, value = AUC, full_model, select_model) %>% 
+  mutate(AUC = as.numeric(AUC), 
+         pvalue = as.numeric(pvalue)) %>% 
+  bind_rows(read_csv("../data/process/tables/stool_rf_otu_random_comparison_summary.csv") %>% 
+              mutate(model_type = "full_otu", 
+                     train_model = study, 
+                     act_mean_auc = as.numeric(act_mean_auc)) %>% 
+              rename(AUC = act_mean_auc) %>% 
+              select(pvalue, study, train_model, model_type, AUC)))
+
+# Pvalue comparisons tables of the different individual taxa and full taxa models
+stool_ind_vs_full_taxa <- read_csv("../data/process/tables/stool_ind_vs_full_taxa_pvalue_summary.csv")
+
+unmatched_tissue_ind_vs_full_taxa <- read_csv("../data/process/tables/unmatched_tissue_ind_vs_full_taxa_pvalue_summary.csv")
 
@@
 
-***Individual Significant OR Taxa Classify Carcinoma Poorly:***
-Since specific taxa were associated with a significant OR for carcinoma we tested whether they would also be good classifiers of carcinoma.
-For stool, the `r length(crc_OR_stool$measure)` significant taxa did no better at classifying those with normal colons versus those with carcinomas than chance [Figure 3A].
-Likewise in unmathced tissue samples the `r length(crc_unmatched_OR_tissue$measure)` significant taxa were no better than an AUC of 0.5 [Figure 3B].
-These results suggest that although these taxa are significantly associated with a decreased or increased OR of carcinoma, individually they are poor classifiers of disease.
+***Individual taxa with a significant OR do a poor job of differentiating subjects with normal colons and those with carcinoma:***
+ We next asked whether those taxa that had a significant OR associated with having a normal colon or carcinomas could be used individually, to classify subjects as having a normal colon or carcinomas.
+Whereas the OR was defined based on whether the relative abundance for a taxon in a subject was above or below the median relative abundance for that taxon across all subjects in a study, we generated receiver operator characteristic (ROC) curves for each taxon in each study and calculated the area under the curve (AUC).
+This allowed us to use a more fluid relative abundance threshold for defining disease status.
+Using data from stool samples, the `r length(crc_OR_stool$measure)` taxa did no better at classifying the subjects than one would expect by chance (i.e.
+AUC=0.50) [Figure 3A].
+The taxa that performed the best included Clostridium XI, Ruminococcus, and Escherichia and even these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)`.
+Likewise, in unmatched tissue samples the `r length(crc_OR_stool$measure)` taxa with significant ORs taxa were marginally better than one would expect by chance [Figure 3B].
+The relative abundance of Dorea was the best predictor of carcinomas and its median AUC was only `r round(max(filter(ind_crc_combined_unmatched_auc, taxa == "Dorea")[, "auc"]), digits = 3)`.
+These results suggest that although these taxa are associated with a decreased or increased OR for the presences of carcinomas, individually, they do a poor job of classifying a subject’s disease status.
 
-***Select Community Models can Recapitulate Whole Community Models:***
-Since specific taxa increased or decreased the OR for carcinoma but performed poorly as an individual classifier of carcinoma and assessing ORs for adenoma we assessed whether the overall bacterial community was better at classifying disease.
-Three models were tested and included a full taxa model, full OTU model, and significant OR taxa model.
-If no taxa were significant after multiple comparison correction then no model for that specific grouping (i.e.
-adenoma stool) was analyzed.
-We first tested three model AUCs.
-Next, the all genera models and any significant OR taxa models were tested across all studies that were not used to train the model.
+***Combined taxa model classifies subjects better than using individual taxa:*** 
+Instead of attempting to classify subjects based on individual taxa, next we generated Random Forest models that combined the individual taxa and evaluated the ability to classify as subject’s disease status.
+For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`).
+For the full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies.
+Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for Blautia and Weissella (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).
+For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies.
+Clearly, pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.
 
-For stool, all models used had similar AUCs for both adenoma and carcinoma [Figure 4].
-However, the adenoma AUCs were barely better than chance [Figure 4A].
-Similarily, when analyzing the tests sets that were comprised of genera data from other studies the adenoma models performed poorly [Figure 5A].
-The carcinoma models performed much better, with both the all genera and singificant OR taxa only models having a similar ability to detect individuals with carcinomas [Figure 5B].
-The most common genera in the top 10 most important variables, in the full genera models used to classify carcinomas, were *Bacteroides* and *Lachnospiraceae* [Figure 6A].
-Regardless of sample type, mouth-associated genera were present in models for carcinomas but not consistently across studies [Figure 6].
-For the full community OTU-based models, both *Bacteroides*, *Blautia*, *Lachnospiraceae*, and *Ruminococcaceae* were present in the top 10 consistently across studies [Figure 6B].
-Overall, these results suggest that multiple microbes could act as the inflammatory stimulus needed to exacerbate mutations leading each individual microbe as a poor individual classifier but much better in aggregate with others.
+***Performance of models based on taxa relative abundance in full community are not significantly better than those based on taxa with significant ORs:*** 
+Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR.
+Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were marginally better than 0.5 for any study[Figure 4A & S3A].
+In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 4B & S3B-C].
+When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed.
+Using the data from stool samples we found that although the AUC for 6 of 7 studies increased (mean decrease = `r round(mean((filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] - filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"])*100), digits = 2)`%), the more expansive models performed worse for 1 of the studies (decrease = `r round(abs(mean((filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] - filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"])*100)), digits = 2)`%).
+The overall improvement in performance was statistically significant (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
+Of the `r length(crc_OR_stool$measure)` taxa with significant ORs, all `r length(crc_OR_stool$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study .
+Similarly, using the data from unmatched tissue samples we found that the AUC for 4 out of 4 studies decreased between full versus select OR models (mean decrease = `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
+Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study.
+These results were surprising because it demonstrated that the ability to classify subjects could be done based on a limited characterization of the communities.
 
-The tissue-based models had results that were dependent on genera or OTU level data and matched or unmatched samples [Figure S2].
-Similarily to stool, the significant OR taxa performed as well as both the full taxa and OTU models [Figure S3B].
-When analyzing the tests sets that were comprised of genera data from other studies, all models performed poorly with one exception being carcinoma classification in unmatched tissue [Figure S4].
-Unlike stool, the test sets for the significant OR taxa did not perform as well as the all genera-based models [Figure S4C].
-For the full genera models built using either matched or unmatched samples, showed no consistent taxa representation in the top 10 most important model variables across study [Figure S4A & S4B].
-Conversely, the OTU-based models built using either matched or unmatched samples showed more commonalities [Figure S4C & S4D].
-In models built using matched samples *Lachnospiraceae*, *Fusobacteriaceae*, *Comamonadaceae*, and *Bacteroidaceae* appeared in the top 10 percent in the majority of studies [Figure S4C].
-Although models built with unmatched samples also had *Lachnospiraceae* and *Bacteroidaceae*, a major difference was the presence of *Ruminococcaceae* in the top 10 of every study [Figure S4D].
-This results suggests that either the colon tissue microbiota is study or person dependent or that that kit and/or other types of contamination associated with low biomass samples may be skewing the results.
+***Performance of models based on OTU relative abundance in full community are not significantly better than those based on taxa with significant ORs:*** 
+The previous models were based on relative abundance data where sequences were assigned to coarse taxonomic assignments (i.e.
+typically genus or family level).
+To determine whether model performance improved with a more fine scale classification, we assigned sequences to operational taxonomic units (OTUs) where the similarity among sequences within an OTU was more than 97%.
+We again found that classification models built using all of the sequence data for a community did a poor job of differentiating between subjects with normal colons and those with adenomas (median AUC: `r round(median(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]), but did a good job of differentiating between subjects with normal colons and those with carcinomas (median AUC: `r round(median(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]).
+The OTU-based models performed similarly to those constructed using the taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) and those using all taxa (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
+Among the OTUs that had the highest mean decrease in accuracy for the OTU-based models, we found that OTUs that affiliated with all of the `r length(crc_OR_stool$measure)` taxa that had a significant OR were within the top 10% for at least one study.
+Again, this result was surprising as it indicated that a finer scale classification of the sequence data and thus a larger number of features to select from, did not yield improved classification of the subjects.
+
+***Generalizability of taxon-based models trained on one dataset to the other datasets:*** 
+Considering the good performance of the Random Forest models using taxa with a significant OR and using all of the taxa, we next asked how well the models would perform when given data from a different subject cohort.
+For instance, if a model was trained using data from the Ahn study, we wanted to know how well it would perform using the data from the Baxter study.
+We found the models trained using the taxa with a significant OR all had a higher median AUC than the models trained using all of the taxa when tested on the other datasets [Figure 5].
+As might be expected, the difference between the performance of the modelling approaches appeared to vary with the size of the training cohort.
+These data suggest that given a sufficient number of subjects with normal colons and carcinomas, Random Forest models trained using a small number of taxa can accurately classify individuals from a different cohort.
 
@@
 ## Discussion
+
+Select subset community you can do as well as OTUs or genera when using all of the information and that these models perform well when trained on one cohort and tested on another.
+
+Why didn’t the whole community work better? Communities are patchy and higher level taxonomic information pools some of that patchiness
+There isn’t one single bug associated with disease – why ? possibly because multiple bugs can carry out the same function (e.g. inflammation). Tell some examples (Esch Fuso, pepto)
+Also, interesting that the loss of bacteria can be associated with disease. These appear to be commonly thought as beneficial bacteria that ferment to produce SCFAs
+
+Adenomas suck. 
+Which is kind of at odds with Baxter, but Baxter looked at lesions and also included FIT
+Also, variation in how adenomas are defined – small polyps vs larger neopplasias (e.g.
+baxter), but we couldn’t do this for the current study because we didn’t have that level of detail and most studies are small
+This isn’t such a novel result (Nich Chia)
+Stool may not be the best place to look for early stage adenomas
+
+Tissue vs. Stool
+Not much overlap between stool and tissue biomarkers
+Stool for diagnostic – does as well as tissue for classifying carcinomas
+Carcinoma result surprising (e.g. lack of Fuso) – largely because small studies?
+
+Missing microbes
+ETBfrag, Strepto, etc.
+B – is a lot of things that we might lose – good and bad; mucosal, right sided tumors? We might not see with stool
+
+Meta analysis are important (eg. obesity, crc), we need more. We learned that these are hard because…
+Inaccessible datasets
+Missing/vague metadata
+Varying sequence quality
+Small datasets
+
+Looking forward
+Potential for using microbiome as a diagnostic tool
+We’re excited that we were able to validate a set of biomarkes across multiple studies
+Found a phenotype we can relate to the microbiome, it’s replicable and it’s strong
 

### 2018-03-01 submission/manuscript.tex
update results with pat edits

@@
 
-\textbf{\emph{Individual Significant OR Taxa Classify Carcinoma
-Poorly:}} Since specific taxa were associated with a significant OR for
-carcinoma we tested whether they would also be good classifiers of
-carcinoma. For stool, the 8 significant taxa did no better at
-classifying those with normal colons versus those with carcinomas than
-chance {[}Figure 3A{]}. Likewise in unmathced tissue samples the 3
-significant taxa were no better than an AUC of 0.5 {[}Figure 3B{]}.
-These results suggest that although these taxa are significantly
-associated with a decreased or increased OR of carcinoma, individually
-they are poor classifiers of disease.
-
-\textbf{\emph{Select Community Models can Recapitulate Whole Community
-Models:}} Since specific taxa increased or decreased the OR for
-carcinoma but performed poorly as an individual classifier of carcinoma
-and assessing ORs for adenoma we assessed whether the overall bacterial
-community was better at classifying disease. Three models were tested
-and included a full taxa model, full OTU model, and significant OR taxa
-model. If no taxa were significant after multiple comparison correction
-then no model for that specific grouping (i.e.~adenoma stool) was
-analyzed. We first tested three model AUCs. Next, the all genera models
-and any significant OR taxa models were tested across all studies that
-were not used to train the model.
-
-For stool, all models used had similar AUCs for both adenoma and
-carcinoma {[}Figure 4{]}. However, the adenoma AUCs were barely better
-than chance {[}Figure 4A{]}. Similarily, when analyzing the tests sets
-that were comprised of genera data from other studies the adenoma models
-performed poorly {[}Figure 5A{]}. The carcinoma models performed much
-better, with both the all genera and singificant OR taxa only models
-having a similar ability to detect individuals with carcinomas {[}Figure
-5B{]}. The most common genera in the top 10 most important variables, in
-the full genera models used to classify carcinomas, were
-\emph{Bacteroides} and \emph{Lachnospiraceae} {[}Figure 6A{]}.
-Regardless of sample type, mouth-associated genera were present in
-models for carcinomas but not consistently across studies {[}Figure
-6{]}. For the full community OTU-based models, both \emph{Bacteroides},
-\emph{Blautia}, \emph{Lachnospiraceae}, and \emph{Ruminococcaceae} were
-present in the top 10 consistently across studies {[}Figure 6B{]}.
-Overall, these results suggest that multiple microbes could act as the
-inflammatory stimulus needed to exacerbate mutations leading each
-individual microbe as a poor individual classifier but much better in
-aggregate with others.
-
-The tissue-based models had results that were dependent on genera or OTU
-level data and matched or unmatched samples {[}Figure S2{]}. Similarily
-to stool, the significant OR taxa performed as well as both the full
-taxa and OTU models {[}Figure S3B{]}. When analyzing the tests sets that
-were comprised of genera data from other studies, all models performed
-poorly with one exception being carcinoma classification in unmatched
-tissue {[}Figure S4{]}. Unlike stool, the test sets for the significant
-OR taxa did not perform as well as the all genera-based models {[}Figure
-S4C{]}. For the full genera models built using either matched or
-unmatched samples, showed no consistent taxa representation in the top
-10 most important model variables across study {[}Figure S4A \& S4B{]}.
-Conversely, the OTU-based models built using either matched or unmatched
-samples showed more commonalities {[}Figure S4C \& S4D{]}. In models
-built using matched samples \emph{Lachnospiraceae},
-\emph{Fusobacteriaceae}, \emph{Comamonadaceae}, and
-\emph{Bacteroidaceae} appeared in the top 10 percent in the majority of
-studies {[}Figure S4C{]}. Although models built with unmatched samples
-also had \emph{Lachnospiraceae} and \emph{Bacteroidaceae}, a major
-difference was the presence of \emph{Ruminococcaceae} in the top 10 of
-every study {[}Figure S4D{]}. This results suggests that either the
-colon tissue microbiota is study or person dependent or that that kit
-and/or other types of contamination associated with low biomass samples
-may be skewing the results.
+\textbf{\emph{Individual taxa with a significant OR do a poor job of
+differentiating subjects with normal colons and those with carcinoma:}}
+We next asked whether those taxa that had a significant OR associated
+with having a normal colon or carcinomas could be used individually, to
+classify subjects as having a normal colon or carcinomas. Whereas the OR
+was defined based on whether the relative abundance for a taxon in a
+subject was above or below the median relative abundance for that taxon
+across all subjects in a study, we generated receiver operator
+characteristic (ROC) curves for each taxon in each study and calculated
+the area under the curve (AUC). This allowed us to use a more fluid
+relative abundance threshold for defining disease status. Using data
+from stool samples, the 8 taxa did no better at classifying the subjects
+than one would expect by chance (i.e.~AUC=0.50) {[}Figure 3A{]}. The
+taxa that performed the best included Clostridium XI, Ruminococcus, and
+Escherichia and even these had median AUC values less than 0.588.
+Likewise, in unmatched tissue samples the 8 taxa with significant ORs
+taxa were marginally better than one would expect by chance {[}Figure
+3B{]}. The relative abundance of Dorea was the best predictor of
+carcinomas and its median AUC was only 0.62. These results suggest that
+although these taxa are associated with a decreased or increased OR for
+the presences of carcinomas, individually, they do a poor job of
+classifying a subject's disease status.
+
+\textbf{\emph{Combined taxa model classifies subjects better than using
+individual taxa:}} Instead of attempting to classify subjects based on
+individual taxa, next we generated Random Forest models that combined
+the individual taxa and evaluated the ability to classify as subject's
+disease status. For data from stool samples, the combined model had an
+AUC of 0.75, which was significantly higher than any of the AUC values
+for the individual taxa (P-value \textless{} 0.033). For the full taxa
+models using stool, \emph{Bacteroides} and \emph{Lachnospiraceae} were
+the most common taxa in the top 10\% mean decrease in accuracy (MDA)
+across studies. Similarly, using data from the unmatched tissue samples,
+the combined model had an AUC of 0.77, which was significantly higher
+than the AUC values for Blautia and Weissella (P-value \textless{}
+0.037). For the full taxa models using unmatched tissue,
+\emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
+\emph{Ruminococcaceae} were the most common taxa in the top 10\% mean
+decrease in accuracy across studies. Clearly, pooling the information
+from the taxa with significant ORs results in a model that outperforms
+classifications made using individual taxa.
+
+\textbf{\emph{Performance of models based on taxa relative abundance in
+full community are not significantly better than those based on taxa
+with significant ORs:}} Next, we asked whether a Random Forest
+classification model built using all of the taxa found in the
+communities would outperform the models generated using those taxa with
+a significant OR. Similar to our inability to identify taxa associated
+with a significant OR for the presence of adenomas, the median AUCs to
+classify subjects as having normal colons or having adenomas using data
+from stool or tissue samples were marginally better than 0.5 for any
+study{[}Figure 4A \& S3A{]}. In contrast, the models for classifying
+subjects as having normal colons or having carcinomas using data from
+stool or tissue samples yielded AUC values meaningfully higher than 0.5
+{[}Figure 4B \& S3B-C{]}. When we compared the models based on all of
+the taxa in a community to models based on the taxa with significant
+ORs, the results were mixed. Using the data from stool samples we found
+that although the AUC for 6 of 7 studies increased (mean decrease =
+9.53\%), the more expansive models performed worse for 1 of the studies
+(decrease = 0.38\%). The overall improvement in performance was
+statistically significant (one-tailed paired T-test; P-value = 0.005).
+Of the 8 taxa with significant ORs, all 8 were among the top 10\% most
+important taxa as measured by mean decrease in accuracy, in at least one
+study . Similarly, using the data from unmatched tissue samples we found
+that the AUC for 4 out of 4 studies decreased between full versus select
+OR models (mean decrease = 19.11\%, one-tailed paired T-test; P-value =
+0.03). Of the 3 taxa with significant ORs, all 3 were among the top 10\%
+most important taxa as measured by mean decrease in accuracy, in at
+least one study. These results were surprising because it demonstrated
+that the ability to classify subjects could be done based on a limited
+characterization of the communities.
+
+\textbf{\emph{Performance of models based on OTU relative abundance in
+full community are not significantly better than those based on taxa
+with significant ORs:}} The previous models were based on relative
+abundance data where sequences were assigned to coarse taxonomic
+assignments (i.e.~typically genus or family level). To determine whether
+model performance improved with a more fine scale classification, we
+assigned sequences to operational taxonomic units (OTUs) where the
+similarity among sequences within an OTU was more than 97\%. We again
+found that classification models built using all of the sequence data
+for a community did a poor job of differentiating between subjects with
+normal colons and those with adenomas (median AUC: 0.53 {[}0.37-
+0.56{]}), but did a good job of differentiating between subjects with
+normal colons and those with carcinomas (median AUC: 0.71 {[}0.5-
+0.9{]}). The OTU-based models performed similarly to those constructed
+using the taxa with significant ORs (one-tailed paired T-test; P-value =
+0.966) and those using all taxa (one-tailed paired T-test; P-value =
+0.146). Among the OTUs that had the highest mean decrease in accuracy
+for the OTU-based models, we found that OTUs that affiliated with all of
+the 8 taxa that had a significant OR were within the top 10\% for at
+least one study. Again, this result was surprising as it indicated that
+a finer scale classification of the sequence data and thus a larger
+number of features to select from, did not yield improved classification
+of the subjects.
+
+\textbf{\emph{Generalizability of taxon-based models trained on one
+dataset to the other datasets:}} Considering the good performance of the
+Random Forest models using taxa with a significant OR and using all of
+the taxa, we next asked how well the models would perform when given
+data from a different subject cohort. For instance, if a model was
+trained using data from the Ahn study, we wanted to know how well it
+would perform using the data from the Baxter study. We found the models
+trained using the taxa with a significant OR all had a higher median AUC
+than the models trained using all of the taxa when tested on the other
+datasets {[}Figure 5{]}. As might be expected, the difference between
+the performance of the modelling approaches appeared to vary with the
+size of the training cohort. These data suggest that given a sufficient
+number of subjects with normal colons and carcinomas, Random Forest
+models trained using a small number of taxa can accurately classify
+individuals from a different cohort.
 
@@
 \subsection{Discussion}\label{discussion}
+
+Select subset community you can do as well as OTUs or genera when using
+all of the information and that these models perform well when trained
+on one cohort and tested on another.
+
+Why didn't the whole community work better? Communities are patchy and
+higher level taxonomic information pools some of that patchiness There
+isn't one single bug associated with disease -- why ? possibly because
+multiple bugs can carry out the same function (e.g.~inflammation). Tell
+some examples (Esch Fuso, pepto) Also, interesting that the loss of
+bacteria can be associated with disease. These appear to be commonly
+thought as beneficial bacteria that ferment to produce SCFAs
+
+Adenomas suck. Which is kind of at odds with Baxter, but Baxter looked
+at lesions and also included FIT Also, variation in how adenomas are
+defined -- small polyps vs larger neopplasias (e.g.~baxter), but we
+couldn't do this for the current study because we didn't have that level
+of detail and most studies are small This isn't such a novel result
+(Nich Chia) Stool may not be the best place to look for early stage
+adenomas
+
+Tissue vs.~Stool Not much overlap between stool and tissue biomarkers
+Stool for diagnostic -- does as well as tissue for classifying
+carcinomas Carcinoma result surprising (e.g.~lack of Fuso) -- largely
+because small studies?
+
+Missing microbes ETBfrag, Strepto, etc. B -- is a lot of things that we
+might lose -- good and bad; mucosal, right sided tumors? We might not
+see with stool
+
+Meta analysis are important (eg. obesity, crc), we need more. We learned
+that these are hard because\ldots{} Inaccessible datasets Missing/vague
+metadata Varying sequence quality Small datasets
+
+Looking forward Potential for using microbiome as a diagnostic tool
+We're excited that we were able to validate a set of biomarkes across
+multiple studies Found a phenotype we can relate to the microbiome, it's
+replicable and it's strong
 

### 2018-03-01 code/make_select_imp_otu_graph.R
first commits

@@ -1,0 +1,78 @@
+### Code for the crc most imp taxa in select models
+### Marc Sze
+
+# Load in needed generic functions
+source("code/functions.R")
+
+# Load in needed libraries
+loadLibs(c("tidyverse", "gridExtra", "viridis", "stringr"))
+
+# Read in Needed data
+stool_mda <- read_csv("data/process/tables/ALL_genus_stool_RF_select_imp_vars.csv") %>% 
+  mutate(otu = str_replace_all(otu, "_unclassified", ""), 
+         otu = str_replace_all(otu, "\\.Shigella", ""), 
+         otu = str_replace_all(otu, "_", " ")) %>% 
+  group_by(study) %>% mutate(zscore = scale(mda_median)) %>% 
+  ungroup()
+
+unmatched_tissue_mda <- read_csv("data/process/tables/ALL_genus_unmatched_tissue_RF_select_imp_vars.csv") %>% 
+  mutate(otu = str_replace_all(otu, "_unclassified", ""), 
+         otu = str_replace_all(otu, "\\.Shigella", ""), 
+         otu = str_replace_all(otu, "_", " ")) %>% 
+  group_by(study, otu) %>% filter(mda_median == max(mda_median)) %>% 
+  ungroup() %>% 
+  group_by(study) %>% mutate(zscore = scale(mda_median)) %>% 
+  ungroup()
+
+
+##############################################################################################
+############################## List of code to make figures ##################################
+##############################################################################################
+
+
+imp_stool_model <- stool_mda %>% 
+  mutate(study = factor(study, 
+                        levels = c("ahn", "baxter", "flemer", "hale", "wang", "weir", "zeller"), 
+                        labels= c(c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller")))) %>% 
+  ggplot(aes(study, otu, fill = zscore)) + 
+  geom_tile(color = "white") + 
+  scale_fill_gradient2(name = "Z-Score Median MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
+  theme_bw() + ggtitle("A") + 
+  labs(x = "", y = "") + 
+  theme(panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank(), 
+        legend.position = "bottom", 
+        axis.text.y = element_text(face = "italic", size = 10), 
+        plot.title = element_text(face="bold", hjust = -0.3, size = 20))
+
+
[…]

### 2018-03-01 code/run_pvalue_auc_ind_vs_full_taxa.R

@@ -1,0 +1,121 @@
+### Code for the crc most imp taxa in select models
+### Marc Sze
+
+# Load in needed generic functions
+source("code/functions.R")
+
+# Load in needed libraries
+loadLibs(c("tidyverse"))
+
+
+ind_crc_stool_auc <- read_csv("data/process/tables/ind_genera_auc_stool.csv") %>%  
+  mutate(auc = ifelse(taxa == "Ruminococcus" | taxa == "Clostridium_XI", invisible(1-auc), invisible(auc)))
+
+ind_unmatched_tissue_crc_auc <- read_csv("data/process/tables/ind_genera_auc_unmatched_tissue.csv") %>% 
+  mutate(auc = ifelse(taxa == "Dorea" | taxa == "Blautia", invisible(1-auc), invisible(auc)))
+
+
+crc_tissue_model_aucs <- read_csv("data/process/tables/ALL_genus_unmatched_tissue_RF_fullvsselect_pvalue_summary.csv") %>% 
+  mutate(type = "unmatched") %>% 
+  bind_rows(read_csv("data/process/tables/ALL_genus_matched_tissue_RF_fullvsselect_pvalue_summary.csv") %>% 
+              mutate(type = "matched")) %>% 
+  gather(key = model_type, value = AUC, full_model, select_model) %>% 
+  mutate(AUC = as.numeric(AUC)) %>% 
+  bind_rows(read_csv("data/process/tables/matched_tissue_rf_otu_random_comparison_summary.csv") %>% 
+              mutate(type = "matched", 
+                     model_type = "full_otu", 
+                     train_model = study, 
+                     act_mean_auc = as.numeric(act_mean_auc)) %>% 
+              rename(AUC = act_mean_auc) %>% 
+              select(pvalue, study, train_model, type, model_type, AUC) %>% 
+              bind_rows(read_csv("data/process/tables/unmatched_tissue_rf_otu_random_comparison_summary.csv") %>% 
+                          mutate(type = "unmatched", 
+                                 model_type = "full_otu", 
+                                 train_model = study, 
+                                 act_mean_auc = as.numeric(act_mean_auc)) %>% 
+                          rename(AUC = act_mean_auc) %>% 
+                          select(pvalue, study, train_model, type, model_type, AUC)))
+
+
+
[…]

### 2018-03-01 submission/manuscript.Rmd
update with new orders after pats edits

@@
 For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`).
-For the full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies.
+For the full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies [Figure S2].
 Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for Blautia and Weissella (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).
-For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies.
+For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S3].
 Clearly, pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.
 
-***Performance of models based on taxa relative abundance in full community are not significantly better than those based on taxa with significant ORs:*** 
+***Performance of models based on taxa relative abundance in full community are better than those based on taxa with significant ORs:*** 
 Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR.
-Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were marginally better than 0.5 for any study[Figure 4A & S3A].
-In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 4B & S3B-C].
+Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were marginally better than 0.5 for any study[Figure 4A & S4A].
+In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 4B & S4B-C].
 When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed.
@@
 Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study.
+The most important taxa across study within the significant OR taxa only models for stool were *Ruminococcus* and *Clostridium XI* [Figure 5A].
+For the singificant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the important based on mean decrease in accuracy [Figure 5B].
 These results were surprising because it demonstrated that the ability to classify subjects could be done based on a limited characterization of the communities.
@@
 For instance, if a model was trained using data from the Ahn study, we wanted to know how well it would perform using the data from the Baxter study.
-We found the models trained using the taxa with a significant OR all had a higher median AUC than the models trained using all of the taxa when tested on the other datasets [Figure 5].
-As might be expected, the difference between the performance of the modelling approaches appeared to vary with the size of the training cohort.
+We found the models trained using the taxa with a significant OR all had a higher median AUC than the models trained using all of the taxa when tested on the other datasets [Figure 6 & S5].
+As might be expected, the difference between the performance of the modelling approaches appeared to vary with the size of the training cohort [Figure 6].
 These data suggest that given a sufficient number of subjects with normal colons and carcinomas, Random Forest models trained using a small number of taxa can accurately classify individuals from a different cohort.
@@
 
-**Figure 5: Stool Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
+**Figure 5: Most Important Members in Significant OR Taxa Carcinoma Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest stool-based models.
+B) Common taxa in the top 10 percent for carcinoma Random Forest unmatched tissue-based models.
+Blue represents less important and red represents more important to the Random Forest Model.
+White means that particular taxa was not in the top 10%.
+
+**Figure 6: Stool Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
 B) Test AUCs of carcinoma models using all genera or significant OR taxa only.
@@
 The red lines represent the median AUC of all test AUCs for a specific study.
-
-**Figure 6: Most Common Taxa Across Carcinoma Full Community Stool Study Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest all taxa-based models.
-B) Common taxa in the top 10 percent for carcinoma Random Forest all OTU-based models.
 
@@
 
-**Figure S2: Tissue Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
+**Figure S2: Most Common Taxa Across Carcinoma Full Community Stool Study Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest all taxa-based models.
+B) Common taxa in the top 10 percent for carcinoma Random Forest all OTU-based models.
+Blue represents less important and red represents more important to the Random Forest Model.
+White means that particular taxa was not in the top 10%.
+
+**Figure S3: Most Common Genera Across Full Community Tissue Study Models.** A) Common genera in the top 10 percent for matched carcinoma Random Forest all genera-based models.
+B) Common genera in the top 10 percent for unmatched carcinoma Random Forest all genera-based models.
+C) Common genera in the top 10 percent for matched carcinoma Random Forest all OTU-based models.
+D) Common genera in the top 10 percent for unmatched carcinoma Random Forest all OTU-based models.
+Blue represents less important and red represents more important to the Random Forest Model.
+White means that particular taxa was not in the top 10%.
+
+**Figure S4: Tissue Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
 B) Carcinoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
@@
 
-**Figure S3: Tissue Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
+**Figure S5: Tissue Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
 B) Test AUCs of carcinoma models using all genera for matched tissue studies.
@@
 
-**Figure S4: Most Common Genera Across Full Community Tissue Study Models.** A) Common genera in the top 10 percent for matched carcinoma Random Forest all genera-based models.
-B) Common genera in the top 10 percent for unmatched carcinoma Random Forest all genera-based models.
-C) Common genera in the top 10 percent for matched carcinoma Random Forest all OTU-based models.
-D) Common genera in the top 10 percent for unmatched carcinoma Random Forest all OTU-based models.
-
 \newpage

### 2018-03-01 submission/manuscript.tex
update with new orders after pats edits

@@
 the most common taxa in the top 10\% mean decrease in accuracy (MDA)
-across studies. Similarly, using data from the unmatched tissue samples,
-the combined model had an AUC of 0.77, which was significantly higher
-than the AUC values for Blautia and Weissella (P-value \textless{}
-0.037). For the full taxa models using unmatched tissue,
-\emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
+across studies {[}Figure S2{]}. Similarly, using data from the unmatched
+tissue samples, the combined model had an AUC of 0.77, which was
+significantly higher than the AUC values for Blautia and Weissella
+(P-value \textless{} 0.037). For the full taxa models using unmatched
+tissue, \emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
 \emph{Ruminococcaceae} were the most common taxa in the top 10\% mean
-decrease in accuracy across studies. Clearly, pooling the information
-from the taxa with significant ORs results in a model that outperforms
-classifications made using individual taxa.
+decrease in accuracy across studies {[}Figure S3{]}. Clearly, pooling
+the information from the taxa with significant ORs results in a model
+that outperforms classifications made using individual taxa.
 
 \textbf{\emph{Performance of models based on taxa relative abundance in
-full community are not significantly better than those based on taxa
-with significant ORs:}} Next, we asked whether a Random Forest
-classification model built using all of the taxa found in the
-communities would outperform the models generated using those taxa with
-a significant OR. Similar to our inability to identify taxa associated
-with a significant OR for the presence of adenomas, the median AUCs to
-classify subjects as having normal colons or having adenomas using data
-from stool or tissue samples were marginally better than 0.5 for any
-study{[}Figure 4A \& S3A{]}. In contrast, the models for classifying
-subjects as having normal colons or having carcinomas using data from
-stool or tissue samples yielded AUC values meaningfully higher than 0.5
-{[}Figure 4B \& S3B-C{]}. When we compared the models based on all of
-the taxa in a community to models based on the taxa with significant
-ORs, the results were mixed. Using the data from stool samples we found
-that although the AUC for 6 of 7 studies increased (mean decrease =
-9.53\%), the more expansive models performed worse for 1 of the studies
-(decrease = 0.38\%). The overall improvement in performance was
-statistically significant (one-tailed paired T-test; P-value = 0.005).
-Of the 8 taxa with significant ORs, all 8 were among the top 10\% most
-important taxa as measured by mean decrease in accuracy, in at least one
-study . Similarly, using the data from unmatched tissue samples we found
-that the AUC for 4 out of 4 studies decreased between full versus select
-OR models (mean decrease = 19.11\%, one-tailed paired T-test; P-value =
-0.03). Of the 3 taxa with significant ORs, all 3 were among the top 10\%
-most important taxa as measured by mean decrease in accuracy, in at
-least one study. These results were surprising because it demonstrated
-that the ability to classify subjects could be done based on a limited
-characterization of the communities.
+full community are better than those based on taxa with significant
+ORs:}} Next, we asked whether a Random Forest classification model built
+using all of the taxa found in the communities would outperform the
+models generated using those taxa with a significant OR. Similar to our
+inability to identify taxa associated with a significant OR for the
+presence of adenomas, the median AUCs to classify subjects as having
+normal colons or having adenomas using data from stool or tissue samples
+were marginally better than 0.5 for any study{[}Figure 4A \& S4A{]}. In
+contrast, the models for classifying subjects as having normal colons or
+having carcinomas using data from stool or tissue samples yielded AUC
+values meaningfully higher than 0.5 {[}Figure 4B \& S4B-C{]}. When we
+compared the models based on all of the taxa in a community to models
+based on the taxa with significant ORs, the results were mixed. Using
+the data from stool samples we found that although the AUC for 6 of 7
+studies increased (mean decrease = 9.53\%), the more expansive models
+performed worse for 1 of the studies (decrease = 0.38\%). The overall
+improvement in performance was statistically significant (one-tailed
+paired T-test; P-value = 0.005). Of the 8 taxa with significant ORs, all
+8 were among the top 10\% most important taxa as measured by mean
+decrease in accuracy, in at least one study . Similarly, using the data
+from unmatched tissue samples we found that the AUC for 4 out of 4
+studies decreased between full versus select OR models (mean decrease =
+19.11\%, one-tailed paired T-test; P-value = 0.03). Of the 3 taxa with
+significant ORs, all 3 were among the top 10\% most important taxa as
+measured by mean decrease in accuracy, in at least one study. The most
+important taxa across study within the significant OR taxa only models
+for stool were \emph{Ruminococcus} and \emph{Clostridium XI} {[}Figure
+5A{]}. For the singificant OR taxa unmatched tissue models both
+\emph{Dorea} and \emph{Blautia} were the important based on mean
+decrease in accuracy {[}Figure 5B{]}. These results were surprising
+because it demonstrated that the ability to classify subjects could be
+done based on a limited characterization of the communities.
 
@@
 than the models trained using all of the taxa when tested on the other
-datasets {[}Figure 5{]}. As might be expected, the difference between
-the performance of the modelling approaches appeared to vary with the
-size of the training cohort. These data suggest that given a sufficient
-number of subjects with normal colons and carcinomas, Random Forest
-models trained using a small number of taxa can accurately classify
-individuals from a different cohort.
+datasets {[}Figure 6 \& S5{]}. As might be expected, the difference
+between the performance of the modelling approaches appeared to vary
+with the size of the training cohort {[}Figure 6{]}. These data suggest
+that given a sufficient number of subjects with normal colons and
+carcinomas, Random Forest models trained using a small number of taxa
+can accurately classify individuals from a different cohort.
 
@@
 
-\textbf{Figure 5: Stool Random Forest Genus-Based Model Test AUCs.} A)
+\textbf{Figure 5: Most Important Members in Significant OR Taxa
+Carcinoma Models.} A) Common taxa in the top 10 percent for carcinoma
+Random Forest stool-based models. B) Common taxa in the top 10 percent
+for carcinoma Random Forest unmatched tissue-based models. Blue
+represents less important and red represents more important to the
+Random Forest Model. White means that particular taxa was not in the top
+10\%.
+
+\textbf{Figure 6: Stool Random Forest Genus-Based Model Test AUCs.} A)
 Test AUCs of adenoma models using all genera across study. B) Test AUCs
@@
 AUC of all test AUCs for a specific study.
-
-\textbf{Figure 6: Most Common Taxa Across Carcinoma Full Community Stool
-Study Models.} A) Common taxa in the top 10 percent for carcinoma Random
-Forest all taxa-based models. B) Common taxa in the top 10 percent for
-carcinoma Random Forest all OTU-based models.
 
@@
 
-\textbf{Figure S2: Tissue Random Forest Model Train AUCs.} A) Adenoma
+\textbf{Figure S2: Most Common Taxa Across Carcinoma Full Community
+Stool Study Models.} A) Common taxa in the top 10 percent for carcinoma
+Random Forest all taxa-based models. B) Common taxa in the top 10
+percent for carcinoma Random Forest all OTU-based models. Blue
+represents less important and red represents more important to the
+Random Forest Model. White means that particular taxa was not in the top
+10\%.
+
+\textbf{Figure S3: Most Common Genera Across Full Community Tissue Study
+Models.} A) Common genera in the top 10 percent for matched carcinoma
+Random Forest all genera-based models. B) Common genera in the top 10
+percent for unmatched carcinoma Random Forest all genera-based models.
+C) Common genera in the top 10 percent for matched carcinoma Random
+Forest all OTU-based models. D) Common genera in the top 10 percent for
+unmatched carcinoma Random Forest all OTU-based models. Blue represents
+less important and red represents more important to the Random Forest
+Model. White means that particular taxa was not in the top 10\%.
+
+\textbf{Figure S4: Tissue Random Forest Model Train AUCs.} A) Adenoma
 random forest model AUCs between all genera, all OTU, and select model
@@
 
-\textbf{Figure S3: Tissue Random Forest Genus-Based Model Test AUCs.} A)
+\textbf{Figure S5: Tissue Random Forest Genus-Based Model Test AUCs.} A)
 Test AUCs of adenoma models using all genera across study. B) Test AUCs
@@
 
-\textbf{Figure S4: Most Common Genera Across Full Community Tissue Study
-Models.} A) Common genera in the top 10 percent for matched carcinoma
-Random Forest all genera-based models. B) Common genera in the top 10
-percent for unmatched carcinoma Random Forest all genera-based models.
-C) Common genera in the top 10 percent for matched carcinoma Random
-Forest all OTU-based models. D) Common genera in the top 10 percent for
-unmatched carcinoma Random Forest all OTU-based models.
-
 \newpage

### 2018-03-01 code/make_genus_rf_auc_against_study_graph.R
update fig orders

@@ -280,10 +280,10 @@ crc_tissue_auc_graph <- grid.arrange(adn_tissue_graph, crc_matched_tissue_graph,
                                      layout_matrix = rbind(c(1, 2), c(3, 3)))
 stool_auc_graph <- grid.arrange(adn_stool_graph, crc_stool_graph)
 
-ggsave("results/figures/FigureS3.pdf", 
+ggsave("results/figures/FigureS5.pdf", 
        crc_tissue_auc_graph, width = 6, height = 8, dpi = 300)
 
-ggsave("results/figures/Figure5.pdf", 
+ggsave("results/figures/Figure6.pdf", 
        stool_auc_graph, width = 7, height = 8, dpi = 300)
 
 

### 2018-03-01 code/make_rf_auc_full_versus_specific_graph.R

@@ -227,7 +227,7 @@ crc_stool_graph <- crc_stool %>%
 tissue_auc_graph <- grid.arrange(adn_tissue_graph, crc_tissue_graph)
 stool_auc_graph <- grid.arrange(adn_stool_graph, crc_stool_graph)
 
-ggsave("results/figures/FigureS2.pdf", 
+ggsave("results/figures/FigureS4.pdf", 
        tissue_auc_graph, width = 6, height = 8, dpi = 300)
 
 ggsave("results/figures/Figure4.pdf", 

### 2018-03-01 code/make_stool_imp_otu_graph.R

@@ -13,8 +13,7 @@ stool_mda <- read_csv("data/process/tables/crc_RF_genera_stool_top10_mda.csv") %
          otu = str_replace_all(otu, "\\.Shigella", ""), 
          otu = str_replace_all(otu, "_", " ")) %>% 
   group_by(study) %>% mutate(zscore = scale(mda_median)) %>% 
-  ungroup() %>% 
-  mutate(zscore = ifelse(zscore < 0, invisible(-zscore), invisible(zscore)))
+  ungroup()
 
 stool_otu_mda <- read_csv("data/process/tables/crc_RF_otu_stool_top10_mda.csv") %>% 
   mutate(genus = str_replace_all(genus, "_unclassified", ""), 
@@ -23,11 +22,8 @@ stool_otu_mda <- read_csv("data/process/tables/crc_RF_otu_stool_top10_mda.csv") 
   group_by(study, genus) %>% filter(mda_median == max(mda_median)) %>% 
   ungroup() %>% 
   group_by(study) %>% mutate(zscore = scale(mda_median)) %>% 
-  ungroup() %>% 
-  mutate(zscore = ifelse(zscore < 0, invisible(-zscore), invisible(zscore)))
+  ungroup()
 
-
-unmatched_tissue_mda <- read.csv("data/process/tables/crc_RF_genera_unmatched_tissue_top10_mda.csv")
 
 crc_stool_sets <- 7
 
@@ -43,7 +39,7 @@ crc_genera <- stool_mda %>%
                         labels= c(c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller")))) %>% 
   ggplot(aes(study, otu, fill = zscore)) + 
   geom_tile(color = "white") + 
-  scale_fill_gradient2(name = "Z-Score Median MDA", low = "white", high = "blue") + 
+  scale_fill_gradient2(name = "Z-Score Median MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
   theme_bw() + ggtitle("A") + 
   labs(x = "", y = "") + 
   theme(panel.grid.major = element_blank(), 
@@ -59,7 +55,7 @@ crc_otu <- stool_otu_mda %>%
                         labels= c(c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller")))) %>% 
   ggplot(aes(study, genus, fill = zscore)) + 
   geom_tile(color = "white") + 
[…]

### 2018-03-01 code/make_tissue_imp_otu_graph.R

@@ -16,8 +16,7 @@ crc_unmatched_genera <- read_csv("data/process/tables/crc_RF_genera_unmatched_ti
          otu = str_replace_all(otu, "Gp", "Acidobacteria Gp"), 
          otu = str_replace_all(otu, "Acidobacteria Acidobacteria Gp", "Acidobacteria Gp")) %>% 
   group_by(study) %>% mutate(zscore = scale(mda_median)) %>% 
-  ungroup() %>% 
-  mutate(zscore = ifelse(zscore < 0, invisible(-zscore), invisible(zscore)))
+  ungroup()
 
 crc_matched_genera <- read_csv("data/process/tables/adn_RF_genera_matched_tissue_top10_mda.csv") %>% 
   mutate(otu = str_replace_all(otu, "_unclassified", ""), 
@@ -26,31 +25,30 @@ crc_matched_genera <- read_csv("data/process/tables/adn_RF_genera_matched_tissue
          otu = str_replace_all(otu, "Gp", "Acidobacteria Gp"), 
          otu = str_replace_all(otu, "Acidobacteria Acidobacteria Gp", "Acidobacteria Gp")) %>% 
   group_by(study) %>% mutate(zscore = scale(mda_median)) %>% 
-  ungroup() %>% 
-  mutate(zscore = ifelse(zscore < 0, invisible(-zscore), invisible(zscore))) %>% 
-  filter(otu != "NA/")
+  ungroup() %>% filter(otu != "NA/")
 
-test <- crc_matched_otu %>% filter(grepl("Gp", genus) == TRUE)
 
 crc_unmatched_otu <- read_csv("data/process/tables/crc_RF_otu_unmatched_tissue_top10_mda.csv") %>% 
   mutate(genus = str_replace_all(genus, "_unclassified", ""), 
          genus = str_replace_all(genus, "\\.", "/"), 
-         genus = str_replace_all(genus, "_", " ")) %>% 
+         genus = str_replace_all(genus, "_", " "), 
+         genus = str_replace_all(genus, "Gp", "Acidobacteria Gp"), 
+         genus = str_replace_all(genus, "Acidobacteria Acidobacteria Gp", "Acidobacteria Gp")) %>% 
   group_by(study, genus) %>% filter(mda_median == max(mda_median)) %>% 
   ungroup() %>% 
   group_by(study) %>% mutate(zscore = scale(mda_median)) %>% 
-  ungroup() %>% 
[…]

### 2018-03-02 submission/manuscript.Rmd
cracking the discussion

@@
 Adenomas suck. 
+The associations between the microbiota and individuals with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes.
+None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls.
+The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas.
+Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to adenoma formation.
+In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 5].
+
 Which is kind of at odds with Baxter, but Baxter looked at lesions and also included FIT
@@
 
-Tissue vs. Stool
-Not much overlap between stool and tissue biomarkers
-Stool for diagnostic – does as well as tissue for classifying carcinomas
-Carcinoma result surprising (e.g. lack of Fuso) – largely because small studies?
+Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities.
+Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
+Sampling the microbiota directly associated with colon tissue may provide clearer answers but the colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors.
+Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S3].
+*Fusobacterium* was not identified consistently across studies and this could be due to both a small number of studies and a small sample size within these studies.
+Additionally, the majority of the colon tissue-based results were study specific with many of the top 10 taxa being present only in a single study.
+The presence of genera associated with contamination [@Salter_contamination_2014], within the top 10 most important variables for the genera and OTU models is worrying (i.e.
+*Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.).
+The low bacterial biomass of tissue samples coupled with potential contamination and small sample sizes could explain why these results seem to be more sporadic than the stool results.
 
-Missing microbes
-ETBfrag, Strepto, etc.
-B – is a lot of things that we might lose – good and bad; mucosal, right sided tumors? We might not see with stool
+One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp.
+*gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017].
+There are reports that *Bacteroides fragilis*, positive for the enterotoxigenic gene, are found at specific locations along the colon but the samples we were able to use in this meta-analysis could not identify these types of differences [@Purcell2017].
+Additionally, since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression.
+Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp.
+*gallolyticus*.
+As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies.
+This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.
+Additionally, *Streptococcus gallolyticus* subsp.
+*gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased OR for individuals to have a carcinoma tumor.
 
@@
 Found a phenotype we can relate to the microbiome, it’s replicable and it’s strong
-
-*Targeting the identification of tumor microbial biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk of disease.
-The current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not without risk of complications.
-Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities.
-Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
-Sampling the microbiota directly associated with colon tissue may provide clearer answers but is not without their own limitations.
-After the colonoscopy bowel prep the bacterial community sampled may reflect the better adhered microbiota versus the resident community.
-Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done.
-It is well known that low biomass samples can be very difficult to work with and results can be study dependent due to the randomness of contamination [@Salter_contamination_2014].*
 
@@
 Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated or general inflammatory genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.
-
-The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors.
-Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S4E & F].
-Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 most important variables of the full carcinoma stage models [Figure S4B-C & E-F].
-The majority of the colon tissue-based results seem to be study specific with many of the top 10 taxa being present only in a single study.
-Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying.
-The low bacterial biomass of tissue samples coupled with potential contamination could explain why these results seem to be more sporadic than the stool results.
-
-One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp.
-*gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017].
-Since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression.
-Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp.
-*gallolyticus*.
-As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies.
-This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.
-Additionally, *Streptococcus gallolyticus* subsp.
-*gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased OR for individuals to have a carcinoma tumor.
-
-The associations between the microbiota and individuals with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes.
-None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls.
-The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas.
-Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to adenoma formation.
-In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 5].
 

### 2018-03-02 submission/manuscript.Rmd
discussion and journal csl update

@@
 title: ''
-csl: microbiome.csl
+csl: mbio.csl
 fontsize: 12pt
@@
 
-Select subset community you can do as well as OTUs or genera when using all of the information and that these models perform well when trained on one cohort and tested on another.
+Although we expected that the full OTU models would perform the best at classifying individuals with and without carcinomas, our observations suggest that both the full and significant OR taxa models performed equally well.
+These results suggest that lower level classification to species and strain may not add extra useful information with respect to prediction models.
+This has been suggested in previous literature where metagenomics did not perform better than 16S rRNA gene sequencing data at classifying individuals with normal colons and those with carcinomas [@Hannigan2017].
+One possible reason as to why lower level classification may not result in better models is that the communities are patchy and higher level taxonomic information pools some of this patchiness allowing for better prediction models.
+There may also be a fair bit of data redundancy within models that utilize more of the community.
+An example of this redundancy would be when we trained models on one study and tested it on the other studies and the AUCs of the models created with the select OR taxa performed as well as full taxa models [Figure 6B].
 
-Why didn’t the whole community work better? Communities are patchy and higher level taxonomic information pools some of that patchiness
-There isn’t one single bug associated with disease – why ? possibly because multiple bugs can carry out the same function (e.g. inflammation). Tell some examples (Esch Fuso, pepto)
-Also, interesting that the loss of bacteria can be associated with disease. These appear to be commonly thought as beneficial bacteria that ferment to produce SCFAs
+Our observations also suggest that a small collection of taxa can classify diease as well as full OTU-based models but that these taxa individually perform quite poorly [Figure 3].
+This result supports the contention that there might be redundancy of function even amongst the taxa included in the significant OR models.
+As an example multiple different microbes could be similarily stimulating the activation of inflammatory pathways and by doing so exacerbate disease progression.
+Multiple reports within the literature have found that different bacteria, such as *Escherichia coli* and *Fusobacterium nucleatum*, can worsen similarily worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013].
+Although the inflammatory taxa were patchy in their importance and presence across studies those that were not typically associated with inflammation were consistently important for every study [Figure 6].
+The loss of these taxa (*Ruminococcus* and *Clostridium XI* in stool and *Dorea* and *Blautia* in unmatched tissue) is particularly interesting because many are commonly thought to be beneficial due to their involvement in production of short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].
 
-Adenomas suck. 
-The associations between the microbiota and individuals with adenoma tumors are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes.
-None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls.
-The results within our meta-analysis suggest that a small effect size may well be the scope in which differences consistently occur between controls and those with adenomas.
-Future studies investigating adenoma tumors and the microbiota need to take power into consideration to reproducibly study whether the microbiota contributes to adenoma formation.
-In contrast to adenoma stage of disease, our observations suggest that most studies analyzed have sufficient power to detect many changes in the carcinoma-associated microbiota because of large effect size differences between cases and controls [Figure 5].
-
-Which is kind of at odds with Baxter, but Baxter looked at lesions and also included FIT
-Also, variation in how adenomas are defined – small polyps vs larger neopplasias (e.g.
-baxter), but we couldn’t do this for the current study because we didn’t have that level of detail and most studies are small
-This isn’t such a novel result (Nich Chia)
-Stool may not be the best place to look for early stage adenomas
+The adenoma models as a whole performed poorly in classifying individuals with and without adenomas.
+This outcome is not inconsistent with what has been published previously [@hale_shifts_2017; @Sze2017].
+However, the modeling results are at odds with results obtained in Baxter, *et al.* [@baxter_microbiota-based_2016].
+There are some major differences between the models generated in this meta-analysis and what was used in this previous report.
+First, the prevoius report's models investigated the classification of lesions (individuals with adenoma or carcinoma) and not adenoma alone.
+The Baxter, *et al.* models also contained Fecal Immunoglobulin Test data while our meta-analysis models only contained 16S rRNA gene sequencing data.
+Although being able to classify individuals with adenomas is important, the classification of advanced adenomas is a more clinically meaningful diagnostic tool (i.e.
+those that are at high risk of progressing to a carcinoma).
+It is possible that we might have been able to detect differences in the bacterial community if advanced adenomas were separated from adenomas but that data was not available for the majority of studies analyzed.
+It is also possible that the initial changes to the bacterial community are focal to where the initial adenoma develops and would not be easily assessed with a fecal sample.
 
@@
 Additionally, the majority of the colon tissue-based results were study specific with many of the top 10 taxa being present only in a single study.
-The presence of genera associated with contamination [@Salter_contamination_2014], within the top 10 most important variables for the genera and OTU models is worrying (i.e.
+The presence of genera associated with contamination [@Salter_contamination_2014], within the top 10 most important variables for the genera and OTU models is worrying (e.g.
 *Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.).
@@
 
-Looking forward
-Potential for using microbiome as a diagnostic tool
-We’re excited that we were able to validate a set of biomarkes across multiple studies
-Found a phenotype we can relate to the microbiome, it’s replicable and it’s strong
-
-Our study identifies clear but small differences in diversity at the community level and larger differences for individual genera, present in patients with tumors versus controls [Figure 1-3].
-Although there was a step-wise decrease in diversity as disease progressed from control to adenoma to carcinoma, this did not translate into large effect size increases in OR for either adenoma or carcinoma tumors.
-Even though mouth-associated genera increased individaul's OR of having a carcinoma for certain sample types, they did not consistently increase the OR of having an adenoma.
-By using these taxa that had significant ORs after multiple comparison correction we found that we could classify indviduals with either adenoma or carcinoma as well as models that use either all genera or all OTUs.
-Finally, many studies were individually under powered to be able to reject the null hypothesis and this could one reason only the comparison between control and carcinoma individuals for stool samples had relible detectable differences.
-
-The data presented herein support the importance of specific taxa for carcinoma, but not necessarily adenoma, tumor formation.
-The results that we have presented show that the significant OR taxa model and both the full genera and OTU models, for indviduals with carcinoma, had similar AUCs [Figure 2 & 3].
-This suggests that an interplay between a select number of potentially protective and exacerbating microbes within the GI community could be crucial for carcinoma formation.
-Importantly, it suggests that there may be key members of the GI community that should be studied further to potentially help reduce the risk of carcinoma tumor formation.
-Conversely, using the present data, it is clear that new approaches may be needed to identify members of the community associated with adenoma tumors.
-Regardless of sample type and whether a full genera- or OTU-based model was used, our Random Forest models consistently performed poorly.
-Yet, the step-wise decrease in diversity suggests that the adenoma-associated community is not normal but has changed subtly.
-This change in diversity, at this early stage of disease, could be focal to the adenoma itself.
-How the host interacts with these subtle changes at early stages of the disease could be what leads to a thoroughly dysfunctional community that is supportive of tumorgenesis.
-
-For the full genera- and OTU-based models within stool, common GI microbes were most consistently present in the top 10 genera or OTUs across studies [Figure 4].
-Changes in *Bacteroides*, *Ruminococcaceae*, *Ruminococcus*, and *Roseburia* were consistently found to be in the top 10 most important variables across the different studies for both individuals with adenoma and carcinoma [Figure 4].
-These data suggest that whether the non-resident bacterium is *Fusobacteria* or *Peptostreptococcus* may not be as important as how these bacteria interact with the changing resident community.
-Based on these observations, it is possible to hypothesize that small changes in community structure lead to new niches in which any one of the mouth-associated or general inflammatory genera can gain a foothold, exacerbating the initial changes in community and facilitating the transition from adenoma to carcinoma stage of disease.
-
-\newpage
-
-## Conclusion
-
-By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in colorectal tumors.
-The data presented here suggests that mouth-associated microbes can gain a foothold within the colon and are commonly associated with the greatest OR of individuals having a carcinoma.
-Conversely, no conclusive signal with these mouth-associated microbes could be detected for individuals with an adenoma.
-Our observations also highlight the importance of power and sample number considerations when investigating the microbiota and adenoma tumors due to possible subtle changes in the community.
-Overall, the associations between the microbiota and individuals with carcinomas were much stronger than with those with adenomas.
+By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in carcinoma tumors.
+Although further validation of the biomarkers presented here need to be undertaken, the replciability of the AUC of a specific collection of taxa across multiple studies suggests a strong potential for the use of the microbiota as a risk stratification tool for individuals with carcinomas.
 

### 2018-03-02 submission/manuscript.tex
discussion and journal csl update

@@
 the microbiota has been hypothesized to have a role in disease
-progression {[}1,2{]}. Numerous studies using murine models of CRC have
-shown the importance of both individual microbes {[}3--7{]} and the
-overall community {[}8--10{]} in tumorigenesis. Numerous case-control
-studies have characterized the microbiota of individuals with colonic
-adenomas and carcinomas in an attempt to identify biomarkers of disease
-progression {[}6,11--17{]}. Because current CRC screening
-recommendations are poorly adhered to due to socioeconomic status, test
-invasiveness, and frequency of tests, development and validation of
-microbiome-associated biomarkers for CRC progression could further
-attempts to develop non-invasive diagnostics {[}18{]}.
+progression (1, 2). Numerous studies using murine models of CRC have
+shown the importance of both individual microbes (3--7) and the overall
+community (8--10) in tumorigenesis. Numerous case-control studies have
+characterized the microbiota of individuals with colonic adenomas and
+carcinomas in an attempt to identify biomarkers of disease progression
+(6, 11--17). Because current CRC screening recommendations are poorly
+adhered to due to socioeconomic status, test invasiveness, and frequency
+of tests, development and validation of microbiome-associated biomarkers
+for CRC progression could further attempts to develop non-invasive
+diagnostics (18).
 
@@
 as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that
-are enriched in people with carcinomas {[}6,11--17{]}. Other studies
-have identified members of \emph{Akkermansia}, \emph{Bacteroides},
+are enriched in people with carcinomas (6, 11--17). Other studies have
+identified members of \emph{Akkermansia}, \emph{Bacteroides},
 \emph{Enterococcus}, \emph{Escherichia}, \emph{Klebsiella},
 \emph{Mogibacterium}, \emph{Streptococcus}, and \emph{Providencia} are
-also associated with carciomas {[}13--15{]}. Additionally,
-\emph{Roseburia} has been found in some studies to be more abundant in
-people with tumors but in other studies it has been found to be either
-less abundant or no different than what is found in subjects with normal
-colons {[}14,17,19,20{]}. There are strong results from tissue culture
-and murine models that Fusobacterium nucleatum, pks-positive strains of
-Escherichia coli, Streptococcus gallolyticus, and an
-entertoxin-producing strain of Bacteroides fragilis are important in the
-pathogenesis of CRC {[}5,14,21--24{]}. These results point to a
-causative role for the microbiota in CRC pathogenesis as well as their
-potential as diagnostic biomarkers.
+also associated with carciomas (13--15). Additionally, \emph{Roseburia}
+has been found in some studies to be more abundant in people with tumors
+but in other studies it has been found to be either less abundant or no
+different than what is found in subjects with normal colons (14, 17, 19,
+20). There are strong results from tissue culture and murine models that
+Fusobacterium nucleatum, pks-positive strains of Escherichia coli,
+Streptococcus gallolyticus, and an entertoxin-producing strain of
+Bacteroides fragilis are important in the pathogenesis of CRC (5, 14,
+21--24). These results point to a causative role for the microbiota in
+CRC pathogenesis as well as their potential as diagnostic biomarkers.
 
@@
 Taxonomic Units (OTUs) are decreased in those with adenomas versus
-controls {[}25{]}. Other studies have identified \emph{Acidovorax},
+controls (25). Other studies have identified \emph{Acidovorax},
 \emph{Bilophila}, \emph{Cloacibacterium}, \emph{Desulfovibrio},
@@
 \emph{Mogibacterium}, and \emph{Pseudomonas} to be enriched in those
-with adenomas {[}25--27{]}. There are few genera that are enriched in
+with adenomas (25--27). There are few genera that are enriched in
 patients with adenoma or carcinoma tumors.
@@
 \emph{Fusobacterium}, and \emph{Parvimonas} were fecal biomarkers for
-the presence of carcinomas {[}28{]}. Contrary to previous studies they
-found sequences similar to members of \emph{Lactobacillus} and
+the presence of carcinomas (28). Contrary to previous studies they found
+sequences similar to members of \emph{Lactobacillus} and
 \emph{Ruminococcus} to be enriched in patients with adenoma or carcinoma
-relative to those with normal colons {[}12,15,16{]}. In addition, they
+relative to those with normal colons (12, 15, 16). In addition, they
 found 16S rRNA gene sequences from members of \emph{Haemophilus},
@@
 of samples included in the analysis (n = 509) was smaller than several
-larger case-control studies that have since been published {[}12,27{]}
+larger case-control studies that have since been published (12, 27)
 
@@
 data from both feces (n = 1737) and colon tissue (492 samples from 350
-individuals) from 14 studies {[}11--17,19,20,23,25--27,29{]} {[}Table 1
+individuals) from 14 studies (11--17, 19, 20, 23, 25--27, 29) {[}Table 1
 \& 2{]}. We expand both the breadth and scope of the previous
@@
 
-Select subset community you can do as well as OTUs or genera when using
-all of the information and that these models perform well when trained
-on one cohort and tested on another.
-
-Why didn't the whole community work better? Communities are patchy and
-higher level taxonomic information pools some of that patchiness There
-isn't one single bug associated with disease -- why ? possibly because
-multiple bugs can carry out the same function (e.g.~inflammation). Tell
-some examples (Esch Fuso, pepto) Also, interesting that the loss of
-bacteria can be associated with disease. These appear to be commonly
-thought as beneficial bacteria that ferment to produce SCFAs
-
-Adenomas suck. Which is kind of at odds with Baxter, but Baxter looked
-at lesions and also included FIT Also, variation in how adenomas are
-defined -- small polyps vs larger neopplasias (e.g.~baxter), but we
-couldn't do this for the current study because we didn't have that level
-of detail and most studies are small This isn't such a novel result
-(Nich Chia) Stool may not be the best place to look for early stage
-adenomas
-
-Tissue vs.~Stool Not much overlap between stool and tissue biomarkers
-Stool for diagnostic -- does as well as tissue for classifying
-carcinomas Carcinoma result surprising (e.g.~lack of Fuso) -- largely
-because small studies?
-
-Missing microbes ETBfrag, Strepto, etc. B -- is a lot of things that we
-might lose -- good and bad; mucosal, right sided tumors? We might not
-see with stool
-
-Meta analysis are important (eg. obesity, crc), we need more. We learned
-that these are hard because\ldots{} Inaccessible datasets Missing/vague
-metadata Varying sequence quality Small datasets
-
-Looking forward Potential for using microbiome as a diagnostic tool
-We're excited that we were able to validate a set of biomarkes across
-multiple studies Found a phenotype we can relate to the microbiome, it's
-replicable and it's strong
-
-\emph{Targeting the identification of tumor microbial biomarkers within
-stool seems logical since it offers an easy and cost-effective way to
-stratify risk of disease. The current gold standard for diagnosis, a
-colonoscopy, can be time-consuming and is not without risk of
-complications. Although stool represents an easy and less invasive way
-to assess risk, it is not clear how well this sample reflects adenoma-
-and carcinoma- associated microbial communities. Some studies have tried
-to assess this in health and disease but are limited by their sample
-size {[}17,30{]}. Sampling the microbiota directly associated with colon
-tissue may provide clearer answers but is not without their own
-limitations. After the colonoscopy bowel prep the bacterial community
-sampled may reflect the better adhered microbiota versus the resident
-community. Additionally, these samples contain more host DNA,
-potentially limiting the types of analysis that can be done. It is well
-known that low biomass samples can be very difficult to work with and
-results can be study dependent due to the randomness of contamination
-{[}31{]}.}
-
-Our study identifies clear but small differences in diversity at the
-community level and larger differences for individual genera, present in
-patients with tumors versus controls {[}Figure 1-3{]}. Although there
-was a step-wise decrease in diversity as disease progressed from control
-to adenoma to carcinoma, this did not translate into large effect size
-increases in OR for either adenoma or carcinoma tumors. Even though
-mouth-associated genera increased individaul's OR of having a carcinoma
-for certain sample types, they did not consistently increase the OR of
-having an adenoma. By using these taxa that had significant ORs after
-multiple comparison correction we found that we could classify
-indviduals with either adenoma or carcinoma as well as models that use
-either all genera or all OTUs. Finally, many studies were individually
-under powered to be able to reject the null hypothesis and this could
-one reason only the comparison between control and carcinoma individuals
-for stool samples had relible detectable differences.
-
-The data presented herein support the importance of specific taxa for
-carcinoma, but not necessarily adenoma, tumor formation. The results
-that we have presented show that the significant OR taxa model and both
-the full genera and OTU models, for indviduals with carcinoma, had
-similar AUCs {[}Figure 2 \& 3{]}. This suggests that an interplay
-between a select number of potentially protective and exacerbating
-microbes within the GI community could be crucial for carcinoma
-formation. Importantly, it suggests that there may be key members of the
-GI community that should be studied further to potentially help reduce
-the risk of carcinoma tumor formation. Conversely, using the present
-data, it is clear that new approaches may be needed to identify members
-of the community associated with adenoma tumors. Regardless of sample
-type and whether a full genera- or OTU-based model was used, our Random
-Forest models consistently performed poorly. Yet, the step-wise decrease
-in diversity suggests that the adenoma-associated community is not
-normal but has changed subtly. This change in diversity, at this early
-stage of disease, could be focal to the adenoma itself. How the host
-interacts with these subtle changes at early stages of the disease could
-be what leads to a thoroughly dysfunctional community that is supportive
-of tumorgenesis.
-
-For the full genera- and OTU-based models within stool, common GI
-microbes were most consistently present in the top 10 genera or OTUs
-across studies {[}Figure 4{]}. Changes in \emph{Bacteroides},
-\emph{Ruminococcaceae}, \emph{Ruminococcus}, and \emph{Roseburia} were
-consistently found to be in the top 10 most important variables across
-the different studies for both individuals with adenoma and carcinoma
-{[}Figure 4{]}. These data suggest that whether the non-resident
-bacterium is \emph{Fusobacteria} or \emph{Peptostreptococcus} may not be
-as important as how these bacteria interact with the changing resident
-community. Based on these observations, it is possible to hypothesize
-that small changes in community structure lead to new niches in which
-any one of the mouth-associated or general inflammatory genera can gain
-a foothold, exacerbating the initial changes in community and
-facilitating the transition from adenoma to carcinoma stage of disease.
-
-The colon tissue-based studies did not provide a clearer understanding
-of how the microbiota may be associated with tumors. Generally, the full
-OTU-based models of unmatched and matched colon tissue samples were
-concordant with stool samples showing that GI resident microbes were the
-most prevalent in the top 10 most important variables across study
-{[}Figure S4E \& F{]}. Unlike in stool, \emph{Fusobacterium} was the
-only mouth-associated bacteria consistently present in the top 10 most
-important variables of the full carcinoma stage models {[}Figure S4B-C
-\& E-F{]}. The majority of the colon tissue-based results seem to be
-study specific with many of the top 10 taxa being present only in a
-single study. Additionally, the presence of genera associated with
-contamination, within the top 10 most important variables for the genera
-and OTU models is worrying. The low bacterial biomass of tissue samples
-coupled with potential contamination could explain why these results
-seem to be more sporadic than the stool results.
+Although we expected that the full OTU models would perform the best at
+classifying individuals with and without carcinomas, our observations
+suggest that both the full and significant OR taxa models performed
+equally well. These results suggest that lower level classification to
+species and strain may not add extra useful information with respect to
+prediction models. This has been suggested in previous literature where
+metagenomics did not perform better than 16S rRNA gene sequencing data
+at classifying individuals with normal colons and those with carcinomas
+(30). One possible reason as to why lower level classification may not
+result in better models is that the communities are patchy and higher
+level taxonomic information pools some of this patchiness allowing for
+better prediction models. There may also be a fair bit of data
+redundancy within models that utilize more of the community. An example
+of this redundancy would be when we trained models on one study and
+tested it on the other studies and the AUCs of the models created with
+the select OR taxa performed as well as full taxa models {[}Figure
+6B{]}.
+
+Our observations also suggest that a small collection of taxa can
+classify diease as well as full OTU-based models but that these taxa
+individually perform quite poorly {[}Figure 3{]}. This result supports
+the contention that there might be redundancy of function even amongst
+the taxa included in the significant OR models. As an example multiple
+different microbes could be similarily stimulating the activation of
+inflammatory pathways and by doing so exacerbate disease progression.
+Multiple reports within the literature have found that different
+bacteria, such as \emph{Escherichia coli} and \emph{Fusobacterium
+nucleatum}, can worsen similarily worsen inflammation in mouse models of
+tumorigenesis (5, 6, 21). Although the inflammatory taxa were patchy in
+their importance and presence across studies those that were not
+typically associated with inflammation were consistently important for
+every study {[}Figure 6{]}. The loss of these taxa (\emph{Ruminococcus}
+and \emph{Clostridium XI} in stool and \emph{Dorea} and \emph{Blautia}
+in unmatched tissue) is particularly interesting because many are
+commonly thought to be beneficial due to their involvement in production
+of short chain fatty acids (31--33).
+
+The adenoma models as a whole performed poorly in classifying
+individuals with and without adenomas. This outcome is not inconsistent
+with what has been published previously (27, 34). However, the modeling
+results are at odds with results obtained in Baxter, \emph{et al.} (12).
+There are some major differences between the models generated in this
+meta-analysis and what was used in this previous report. First, the
+prevoius report's models investigated the classification of lesions
+(individuals with adenoma or carcinoma) and not adenoma alone. The
+Baxter, \emph{et al.} models also contained Fecal Immunoglobulin Test
+data while our meta-analysis models only contained 16S rRNA gene
+sequencing data. Although being able to classify individuals with
+adenomas is important, the classification of advanced adenomas is a more
+clinically meaningful diagnostic tool (i.e.~those that are at high risk
+of progressing to a carcinoma). It is possible that we might have been
+able to detect differences in the bacterial community if advanced
+adenomas were separated from adenomas but that data was not available
+for the majority of studies analyzed. It is also possible that the
+initial changes to the bacterial community are focal to where the
+initial adenoma develops and would not be easily assessed with a fecal
+sample.
+
+Although stool represents an easy and less invasive way to assess risk,
+it is not clear how well this sample reflects adenoma- and carcinoma-
+associated microbial communities. Some studies have tried to assess this
+in health and disease but are limited by their sample size (17, 35).
+Sampling the microbiota directly associated with colon tissue may
+provide clearer answers but the colon tissue-based studies did not
+provide a clearer understanding of how the microbiota may be associated
+with tumors. Generally, the full OTU-based models of unmatched and
+matched colon tissue samples were concordant with stool samples showing
+that GI resident microbes were the most prevalent in the top 10 most
+important variables across study {[}Figure S3{]}. \emph{Fusobacterium}
+was not identified consistently across studies and this could be due to
+both a small number of studies and a small sample size within these
+studies. Additionally, the majority of the colon tissue-based results
+were study specific with many of the top 10 taxa being present only in a
+single study. The presence of genera associated with contamination (36),
+within the top 10 most important variables for the genera and OTU models
+is worrying (e.g. \emph{Novosphingobium}, \emph{Acidobacteria Gp2},
+\emph{Sphingomonas}, etc.). The low bacterial biomass of tissue samples
+coupled with potential contamination and small sample sizes could
+explain why these results seem to be more sporadic than the stool
+results.
 
@@
 identified, it does not necessarily mean that these specific species are
-not important in human CRC {[}22,24{]}. Since we are limited in our
-aggregation of the data to the genus level, it is not possible to
-clearly delineate which species are contributing to overall disease
-progression. Our observations are not inconsistent with the previous
-literature on either \emph{Bacteroides fragilis} or \emph{Streptococcus
-gallolyticus} subsp. \emph{gallolyticus}. As an example, the stool-based
-full community models consistently identified the genus
-\emph{Bacteroides}, as well as OTUs that classified as
-\emph{Bacteroides}, to be important model components across studies.
-This suggests that even though \emph{Bacteroides} may not increase the
-OR of individuals having an adenoma or carcinoma and may not vary in
-relative abundance, like \emph{Fusobacterium}, it is still important in
-CRC. Additionally, \emph{Streptococcus gallolyticus} subsp.
-\emph{gallolyticus} is a mouth-associated microbe, and the results from
-this study suggest that regardless of sample type, mouth-associated
-genera are commonly associated with an increased OR for individuals to
-have a carcinoma tumor.
-
-The associations between the microbiota and individuals with adenoma
-tumors are inconclusive, in part, because many studies may not be
-powered effectively to observe small effect sizes. None of the studies
-analyzed were properly powered to detect a 10\% or lower change between
-cases and controls. The results within our meta-analysis suggest that a
-small effect size may well be the scope in which differences
-consistently occur between controls and those with adenomas. Future
-studies investigating adenoma tumors and the microbiota need to take
-power into consideration to reproducibly study whether the microbiota
-contributes to adenoma formation. In contrast to adenoma stage of
-disease, our observations suggest that most studies analyzed have
-sufficient power to detect many changes in the carcinoma-associated
-microbiota because of large effect size differences between cases and
-controls {[}Figure 5{]}.
-
-\newpage
-
-\subsection{Conclusion}\label{conclusion}
+not important in human CRC (22, 24). There are reports that
+\emph{Bacteroides fragilis}, positive for the enterotoxigenic gene, are
+found at specific locations along the colon but the samples we were able
+to use in this meta-analysis could not identify these types of
+differences (37). Additionally, since we are limited in our aggregation
+of the data to the genus level, it is not possible to clearly delineate
+which species are contributing to overall disease progression. Our
+observations are not inconsistent with the previous literature on either
+\emph{Bacteroides fragilis} or \emph{Streptococcus gallolyticus} subsp.
+\emph{gallolyticus}. As an example, the stool-based full community
+models consistently identified the genus \emph{Bacteroides}, as well as
+OTUs that classified as \emph{Bacteroides}, to be important model
+components across studies. This suggests that even though
+\emph{Bacteroides} may not increase the OR of individuals having an
+adenoma or carcinoma and may not vary in relative abundance, like
+\emph{Fusobacterium}, it is still important in CRC. Additionally,
+\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
+mouth-associated microbe, and the results from this study suggest that
+regardless of sample type, mouth-associated genera are commonly
+associated with an increased OR for individuals to have a carcinoma
+tumor.
+
+Meta analysis are important (eg. obesity, crc), we need more. We learned
+that these are hard because\ldots{} Inaccessible datasets Missing/vague
+metadata Varying sequence quality Small datasets
 
@@
 fecal and colon tissue samples, we are able to provide evidence
-supporting the importance of the bacterial community in colorectal
-tumors. The data presented here suggests that mouth-associated microbes
-can gain a foothold within the colon and are commonly associated with
-the greatest OR of individuals having a carcinoma. Conversely, no
-conclusive signal with these mouth-associated microbes could be detected
-for individuals with an adenoma. Our observations also highlight the
-importance of power and sample number considerations when investigating
-the microbiota and adenoma tumors due to possible subtle changes in the
-community. Overall, the associations between the microbiota and
-individuals with carcinomas were much stronger than with those with
-adenomas.
+supporting the importance of the bacterial community in carcinoma
+tumors. Although further validation of the biomarkers presented here
+need to be undertaken, the replciability of the AUC of a specific
+collection of taxa across multiple studies suggests a strong potential
+for the use of the microbiota as a risk stratification tool for
+individuals with carcinomas.
 
@@
 meta-analysis were identified through the review articles written by
-Keku, \emph{et al.} and Vogtmann, \emph{et al.} {[}32,33{]} and
-additional studies not mentioned in the reviews were obtained based on
-the authors' knowledge of the literature. Studies that used tissue or
-feces as their sample source for 454 or Illumina 16S rRNA gene
-sequencing analysis and had data sets with sequences available for
-analysis were included. Some studies were excluded because they did not
-have publicly available sequences or did not have metadata in which the
-authors were able to share. After these filtering steps, the following
-studies remained: Ahn, \emph{et al.} {[}11{]}, Baxter, \emph{et al.}
-{[}12{]}, Brim, \emph{et al.} {[}29{]}, Burns, \emph{et al.} {[}15{]},
-Chen, \emph{et al.} {[}13{]}, Dejea, \emph{et al.} {[}20{]}, Flemer,
-\emph{et al.} {[}17{]}, Geng, \emph{et al.} {[}19{]}, Hale, \emph{et
-al.} {[}27{]}, Kostic, \emph{et al.} {[}34{]}, Lu, \emph{et al.}
-{[}26{]}, Sanapareddy, \emph{et al.} {[}25{]}, Wang, \emph{et al.}
-{[}14{]}, Weir, \emph{et al.} {[}23{]}, and Zeller, \emph{et al.}
-{[}16{]}. The Zackular {[}35{]} study was not included because the 90
-individuals analyzed within the study are contained within the larger
-Baxter study {[}12{]}. After sequence processing, all the case samples
-for the Kostic study had 100 or less sequences remaining and was
-excluded, leaving a total of 14 studies that analysis could be completed
-on.
+Keku, \emph{et al.} and Vogtmann, \emph{et al.} (38, 39) and additional
+studies not mentioned in the reviews were obtained based on the authors'
+knowledge of the literature. Studies that used tissue or feces as their
+sample source for 454 or Illumina 16S rRNA gene sequencing analysis and
+had data sets with sequences available for analysis were included. Some
+studies were excluded because they did not have publicly available
+sequences or did not have metadata in which the authors were able to
+share. After these filtering steps, the following studies remained: Ahn,
+\emph{et al.} (11), Baxter, \emph{et al.} (12), Brim, \emph{et al.}
+(29), Burns, \emph{et al.} (15), Chen, \emph{et al.} (13), Dejea,
+\emph{et al.} (20), Flemer, \emph{et al.} (17), Geng, \emph{et al.}
+(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (40), Lu, \emph{et
+al.} (26), Sanapareddy, \emph{et al.} (25), Wang, \emph{et al.} (14),
+Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
+(41) study was not included because the 90 individuals analyzed within
+the study are contained within the larger Baxter study (12). After
+sequence processing, all the case samples for the Kostic study had 100
+or less sequences remaining and was excluded, leaving a total of 14
+studies that analysis could be completed on.
 
@@
 not have sequences and metadata on the SRA, data was obtained from DBGap
-(n = 1, {[}11{]}) and directly from the authors (n = 4,
-{[}17,23,25,27{]}). Each study was processed using the mothur (v1.39.3)
-software program {[}36{]} and quality filtering utilized the default
-methods for both 454 and Illumina based sequencing. If it was not
-possible to use the defaults, the stated quality cut-offs, from the
-study itself, were used instead. Sequences that were made up of an
-artificial combination of two or more different sequences and commonly
-known as chimeras were identified and removed using VSEARCH {[}37{]}
-before \emph{de novo} OTU clustering at 97\% similarity was completed
-using the OptiClust algorithm {[}38{]}.
+(n = 1, (11)) and directly from the authors (n = 4, (17, 23, 25, 27)).
+Each study was processed using the mothur (v1.39.3) software program
+(42) and quality filtering utilized the default methods for both 454 and
+Illumina based sequencing. If it was not possible to use the defaults,
+the stated quality cut-offs, from the study itself, were used instead.
+Sequences that were made up of an artificial combination of two or more
+different sequences and commonly known as chimeras were identified and
+removed using VSEARCH (43) before \emph{de novo} OTU clustering at 97\%
+similarity was completed using the OptiClust algorithm (44).
 
@@
 ORs for individuals having an adenoma or carcinoma and corrected for
-multiple comparisons using the Benjamini-Hochberg method {[}39{]}. We
-then built Random Forest models based on all genera, all OTUs, or
-significant OR taxa (only using taxa still significant after multiple
-comparison correction). For both the full genera and significant OR
-taxa, models were trained on one study then tested on the remaining
-studies using genera-based relative abundances. The OTU-based models
-were built using OTU level data and a 10-fold CV over 100 different
-iterations, based on random 80/20 splitting of the data, was used to
-generate a range of expected AUCs. This process was repeated for every
-study in the meta-analysis. Comparisons of the initial trained model
-AUCs for the full genera and significant OR taxa were made to the mean
-AUC generated from the 100 different 10-fold CV runs of the respective
-OTU-based model. For comparisons in which only control versus adenoma
-individuals were made, the carcinoma individuals were excluded from each
-respective study. Similarly, for comparisons in which control versus
-carcinoma individuals were made the adenoma individuals were excluded
-from each respective study. For all analysis completed fecal and tissue
-samples were kept separate. Within the tissue groups the data were
-further divided between samples from the same individual (matched) and
-those from different individuals (unmatched).
+multiple comparisons using the Benjamini-Hochberg method (45). We then
+built Random Forest models based on all genera, all OTUs, or significant
+OR taxa (only using taxa still significant after multiple comparison
+correction). For both the full genera and significant OR taxa, models
+were trained on one study then tested on the remaining studies using
+genera-based relative abundances. The OTU-based models were built using
+OTU level data and a 10-fold CV over 100 different iterations, based on
+random 80/20 splitting of the data, was used to generate a range of
+expected AUCs. This process was repeated for every study in the
+meta-analysis. Comparisons of the initial trained model AUCs for the
+full genera and significant OR taxa were made to the mean AUC generated
+from the 100 different 10-fold CV runs of the respective OTU-based
+model. For comparisons in which only control versus adenoma individuals
+were made, the carcinoma individuals were excluded from each respective
+study. Similarly, for comparisons in which control versus carcinoma
+individuals were made the adenoma individuals were excluded from each
+respective study. For all analysis completed fecal and tissue samples
+were kept separate. Within the tissue groups the data were further
+divided between samples from the same individual (matched) and those
+from different individuals (unmatched).
 
@@
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package {[}40{]}.
-For OTU richness, evenness, and Shannon diversity analysis, values were
-power transformed using the rcompanion (v1.11.1) package {[}41{]} and
-then Z-score normalized using the car (v2.1.6) package {[}42{]}. Testing
-for OTU richness, evenness, and Shannon diversity differences utilized
-linear mixed-effect models created using the lme4 (v1.1.15) package
-{[}43{]} to correct for study, repeat sampling of individuals (tissue
-only), and 16S hyper-variable region used. Odds ratios (OR) were
-analyzed using both the epiR (v0.9.93) and metafor (v2.0.0) packages
-{[}44,45{]} by assessing how many individuals with and without disease
-were above and below the overall median value within each specific
-study. OR significance testing utilized the chi-squared test. Diversity
-differences measured by the Bray-Curtis index utilized the creation of
-distance matrix and testing with PERMANOVA executed with the vegan
-(v2.4.5) package {[}46{]}. Random Forest models were built using both
-the caret (v6.0.78) and randomForest (v4.6.12) packages {[}47,48{]}. All
-figures were created using both ggplot2 (v2.2.1) and gridExtra (v2.3)
-packages {[}49,50{]}.
+sequence processing utilized the R (v3.4.3) software package (46). For
+OTU richness, evenness, and Shannon diversity analysis, values were
+power transformed using the rcompanion (v1.11.1) package (47) and then
+Z-score normalized using the car (v2.1.6) package (48). Testing for OTU
+richness, evenness, and Shannon diversity differences utilized linear
+mixed-effect models created using the lme4 (v1.1.15) package (49) to
+correct for study, repeat sampling of individuals (tissue only), and 16S
+hyper-variable region used. Odds ratios (OR) were analyzed using both
+the epiR (v0.9.93) and metafor (v2.0.0) packages (50, 51) by assessing
+how many individuals with and without disease were above and below the
+overall median value within each specific study. OR significance testing
+utilized the chi-squared test. Diversity differences measured by the
+Bray-Curtis index utilized the creation of distance matrix and testing
+with PERMANOVA executed with the vegan (v2.4.5) package (52). Random
+Forest models were built using both the caret (v6.0.78) and randomForest
+(v4.6.12) packages (53, 54). All figures were created using both ggplot2
+(v2.2.1) and gridExtra (v2.3) packages (55, 56).
 
@@
 \hypertarget{ref-siegel_cancer_2016}{}
-1. Siegel RL, Miller KD, Jemal A. Cancer statistics, 2016. CA: a cancer
-journal for clinicians. 2016;66:7--30.
+1. \textbf{Siegel, R. L.}, \textbf{K. D. Miller}, and \textbf{A. Jemal}.
+2016. Cancer statistics, 2016. CA: a cancer journal for clinicians
+\textbf{66}:7--30.
 
 \hypertarget{ref-flynn_metabolic_2016}{}
-2. Flynn KJ, Baxter NT, Schloss PD. Metabolic and Community Synergy of
-Oral Bacteria in Colorectal Cancer. mSphere. 2016;1.
+2. \textbf{Flynn, K. J.}, \textbf{N. T. Baxter}, and \textbf{P. D.
+Schloss}. 2016. Metabolic and Community Synergy of Oral Bacteria in
+Colorectal Cancer. mSphere \textbf{1}.
 
 \hypertarget{ref-goodwin_polyamine_2011}{}
-3. Goodwin AC, Destefano Shields CE, Wu S, Huso DL, Wu X, Murray-Stewart
-TR, et al. Polyamine catabolism contributes to enterotoxigenic
-Bacteroides fragilis-induced colon tumorigenesis. Proceedings of the
-National Academy of Sciences of the United States of America.
-2011;108:15354--9.
+3. \textbf{Goodwin, A. C.}, \textbf{C. E. Destefano Shields}, \textbf{S.
+Wu}, \textbf{D. L. Huso}, \textbf{X. Wu}, \textbf{T. R. Murray-Stewart},
+\textbf{A. Hacker-Prietz}, \textbf{S. Rabizadeh}, \textbf{P. M. Woster},
+\textbf{C. L. Sears}, and \textbf{R. A. Casero}. 2011. Polyamine
+catabolism contributes to enterotoxigenic Bacteroides fragilis-induced
+colon tumorigenesis. Proceedings of the National Academy of Sciences of
+the United States of America \textbf{108}:15354--15359.
 
 \hypertarget{ref-abed_fap2_2016}{}
-4. Abed J, Emgård JEM, Zamir G, Faroja M, Almogy G, Grenov A, et al.
-Fap2 Mediates Fusobacterium nucleatum Colorectal Adenocarcinoma
-Enrichment by Binding to Tumor-Expressed Gal-GalNAc. Cell Host \&
-Microbe. 2016;20:215--25.
+4. \textbf{Abed, J.}, \textbf{J. E. M. Emgård}, \textbf{G. Zamir},
+\textbf{M. Faroja}, \textbf{G. Almogy}, \textbf{A. Grenov}, \textbf{A.
+Sol}, \textbf{R. Naor}, \textbf{E. Pikarsky}, \textbf{K. A. Atlan},
+\textbf{A. Mellul}, \textbf{S. Chaushu}, \textbf{A. L. Manson},
+\textbf{A. M. Earl}, \textbf{N. Ou}, \textbf{C. A. Brennan}, \textbf{W.
+S. Garrett}, and \textbf{G. Bachrach}. 2016. Fap2 Mediates Fusobacterium
+nucleatum Colorectal Adenocarcinoma Enrichment by Binding to
+Tumor-Expressed Gal-GalNAc. Cell Host \& Microbe \textbf{20}:215--225.
 
 \hypertarget{ref-arthur_intestinal_2012}{}
-5. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
-T-J, et al. Intestinal inflammation targets cancer-inducing activity of
-the microbiota. Science (New York, NY). 2012;338:120--3.
+5. \textbf{Arthur, J. C.}, \textbf{E. Perez-Chanona}, \textbf{M.
+Mühlbauer}, \textbf{S. Tomkovich}, \textbf{J. M. Uronis}, \textbf{T.-J.
+Fan}, \textbf{B. J. Campbell}, \textbf{T. Abujamel}, \textbf{B. Dogan},
+\textbf{A. B. Rogers}, \textbf{J. M. Rhodes}, \textbf{A. Stintzi},
+\textbf{K. W. Simpson}, \textbf{J. J. Hansen}, \textbf{T. O. Keku},
+\textbf{A. A. Fodor}, and \textbf{C. Jobin}. 2012. Intestinal
+inflammation targets cancer-inducing activity of the microbiota. Science
+(New York, N.Y.) \textbf{338}:120--123.
 
 \hypertarget{ref-kostic_fusobacterium_2013}{}
-6. Kostic AD, Chun E, Robertson L, Glickman JN, Gallini CA, Michaud M,
-et al. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
-modulates the tumor-immune microenvironment. Cell Host \& Microbe.
-2013;14:207--15.
+6. \textbf{Kostic, A. D.}, \textbf{E. Chun}, \textbf{L. Robertson},
+\textbf{J. N. Glickman}, \textbf{C. A. Gallini}, \textbf{M. Michaud},
+\textbf{T. E. Clancy}, \textbf{D. C. Chung}, \textbf{P. Lochhead},
+\textbf{G. L. Hold}, \textbf{E. M. El-Omar}, \textbf{D. Brenner},
+\textbf{C. S. Fuchs}, \textbf{M. Meyerson}, and \textbf{W. S. Garrett}.
+2013. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
+modulates the tumor-immune microenvironment. Cell Host \& Microbe
+\textbf{14}:207--215.
 
 \hypertarget{ref-wu_human_2009}{}
-7. Wu S, Rhee K-J, Albesiano E, Rabizadeh S, Wu X, Yen H-R, et al. A
-human colonic commensal promotes colon tumorigenesis via activation of T
-helper type 17 T cell responses. Nature Medicine. 2009;15:1016--22.
+7. \textbf{Wu, S.}, \textbf{K.-J. Rhee}, \textbf{E. Albesiano},
+\textbf{S. Rabizadeh}, \textbf{X. Wu}, \textbf{H.-R. Yen}, \textbf{D. L.
+Huso}, \textbf{F. L. Brancati}, \textbf{E. Wick}, \textbf{F.
+McAllister}, \textbf{F. Housseau}, \textbf{D. M. Pardoll}, and
+\textbf{C. L. Sears}. 2009. A human colonic commensal promotes colon
+tumorigenesis via activation of T helper type 17 T cell responses.
+Nature Medicine \textbf{15}:1016--1022.
 
 \hypertarget{ref-zackular_manipulation_2016}{}
-8. Zackular JP, Baxter NT, Chen GY, Schloss PD. Manipulation of the Gut
-Microbiota Reveals Role in Colon Tumorigenesis. mSphere. 2016;1.
+8. \textbf{Zackular, J. P.}, \textbf{N. T. Baxter}, \textbf{G. Y. Chen},
+and \textbf{P. D. Schloss}. 2016. Manipulation of the Gut Microbiota
+Reveals Role in Colon Tumorigenesis. mSphere \textbf{1}.
 
 \hypertarget{ref-zackular_gut_2013}{}
-9. Zackular JP, Baxter NT, Iverson KD, Sadler WD, Petrosino JF, Chen GY,
-et al. The gut microbiome modulates colon tumorigenesis. mBio.
-2013;4:e00692--00613.
+9. \textbf{Zackular, J. P.}, \textbf{N. T. Baxter}, \textbf{K. D.
+Iverson}, \textbf{W. D. Sadler}, \textbf{J. F. Petrosino}, \textbf{G. Y.
+Chen}, and \textbf{P. D. Schloss}. 2013. The gut microbiome modulates
+colon tumorigenesis. mBio \textbf{4}:e00692--00613.
 
 \hypertarget{ref-baxter_structure_2014}{}
-10. Baxter NT, Zackular JP, Chen GY, Schloss PD. Structure of the gut
-microbiome following colonization with human feces determines colonic
-tumor burden. Microbiome. 2014;2:20.
+10. \textbf{Baxter, N. T.}, \textbf{J. P. Zackular}, \textbf{G. Y.
+Chen}, and \textbf{P. D. Schloss}. 2014. Structure of the gut microbiome
+following colonization with human feces determines colonic tumor burden.
+Microbiome \textbf{2}:20.
 
 \hypertarget{ref-ahn_human_2013}{}
-11. Ahn J, Sinha R, Pei Z, Dominianni C, Wu J, Shi J, et al. Human gut
-microbiome and risk for colorectal cancer. Journal of the National
-Cancer Institute. 2013;105:1907--11.
+11. \textbf{Ahn, J.}, \textbf{R. Sinha}, \textbf{Z. Pei}, \textbf{C.
+Dominianni}, \textbf{J. Wu}, \textbf{J. Shi}, \textbf{J. J. Goedert},
+\textbf{R. B. Hayes}, and \textbf{L. Yang}. 2013. Human gut microbiome
+and risk for colorectal cancer. Journal of the National Cancer Institute
+\textbf{105}:1907--1911.
 
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-12. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. Microbiota-based model
+12. \textbf{Baxter, N. T.}, \textbf{M. T. Ruffin}, \textbf{M. A. M.
+Rogers}, and \textbf{P. D. Schloss}. 2016. Microbiota-based model
 improves the sensitivity of fecal immunochemical test for detecting
-colonic lesions. Genome Medicine. 2016;8:37.
+colonic lesions. Genome Medicine \textbf{8}:37.
 
 \hypertarget{ref-chen_human_2012}{}
-13. Chen W, Liu F, Ling Z, Tong X, Xiang C. Human intestinal lumen and
+13. \textbf{Chen, W.}, \textbf{F. Liu}, \textbf{Z. Ling}, \textbf{X.
+Tong}, and \textbf{C. Xiang}. 2012. Human intestinal lumen and
 mucosa-associated microbiota in patients with colorectal cancer. PloS
-One. 2012;7:e39743.
+One \textbf{7}:e39743.
 
 \hypertarget{ref-wang_structural_2012}{}
-14. Wang T, Cai G, Qiu Y, Fei N, Zhang M, Pang X, et al. Structural
-segregation of gut microbiota between colorectal cancer patients and
-healthy volunteers. The ISME journal. 2012;6:320--9.
+14. \textbf{Wang, T.}, \textbf{G. Cai}, \textbf{Y. Qiu}, \textbf{N.
+Fei}, \textbf{M. Zhang}, \textbf{X. Pang}, \textbf{W. Jia}, \textbf{S.
+Cai}, and \textbf{L. Zhao}. 2012. Structural segregation of gut
+microbiota between colorectal cancer patients and healthy volunteers.
+The ISME journal \textbf{6}:320--329.
 
 \hypertarget{ref-burns_virulence_2015}{}
-15. Burns MB, Lynch J, Starr TK, Knights D, Blekhman R. Virulence genes
-are a signature of the microbiome in the colorectal tumor
-microenvironment. Genome Medicine. 2015;7:55.
+15. \textbf{Burns, M. B.}, \textbf{J. Lynch}, \textbf{T. K. Starr},
+\textbf{D. Knights}, and \textbf{R. Blekhman}. 2015. Virulence genes are
+a signature of the microbiome in the colorectal tumor microenvironment.
+Genome Medicine \textbf{7}:55.
 
 \hypertarget{ref-zeller_potential_2014}{}
-16. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, et al.
-Potential of fecal microbiota for early-stage detection of colorectal
-cancer. Molecular Systems Biology. 2014;10:766.
+16. \textbf{Zeller, G.}, \textbf{J. Tap}, \textbf{A. Y. Voigt},
+\textbf{S. Sunagawa}, \textbf{J. R. Kultima}, \textbf{P. I. Costea},
+\textbf{A. Amiot}, \textbf{J. Böhm}, \textbf{F. Brunetti}, \textbf{N.
+Habermann}, \textbf{R. Hercog}, \textbf{M. Koch}, \textbf{A. Luciani},
+\textbf{D. R. Mende}, \textbf{M. A. Schneider}, \textbf{P.
+Schrotz-King}, \textbf{C. Tournigand}, \textbf{J. Tran Van Nhieu},
+\textbf{T. Yamada}, \textbf{J. Zimmermann}, \textbf{V. Benes},
+\textbf{M. Kloor}, \textbf{C. M. Ulrich}, \textbf{M. von Knebel
+Doeberitz}, \textbf{I. Sobhani}, and \textbf{P. Bork}. 2014. Potential
+of fecal microbiota for early-stage detection of colorectal cancer.
+Molecular Systems Biology \textbf{10}:766.
 
 \hypertarget{ref-flemer_tumour-associated_2017}{}
-17. Flemer B, Lynch DB, Brown JMR, Jeffery IB, Ryan FJ, Claesson MJ, et
-al. Tumour-associated and non-tumour-associated microbiota in colorectal
-cancer. Gut. 2017;66:633--43.
+17. \textbf{Flemer, B.}, \textbf{D. B. Lynch}, \textbf{J. M. R. Brown},
+\textbf{I. B. Jeffery}, \textbf{F. J. Ryan}, \textbf{M. J. Claesson},
+\textbf{M. O'Riordain}, \textbf{F. Shanahan}, and \textbf{P. W.
+O'Toole}. 2017. Tumour-associated and non-tumour-associated microbiota
+in colorectal cancer. Gut \textbf{66}:633--643.
 
 \hypertarget{ref-GimenoGarca2012}{}
-18. García AZG. Factors influencing colorectal cancer screening
-participation. Gastroenterology Research and Practice {[}Internet{]}.
-Hindawi Limited; 2012;2012:1--8. Available from:
-\url{https://doi.org/10.1155/2012/483417}
+18. \textbf{García, A. Z. G.} 2012. Factors influencing colorectal
+cancer screening participation. Gastroenterology Research and Practice.
+Hindawi Limited \textbf{2012}:1--8.
 
 \hypertarget{ref-geng_diversified_2013}{}
-19. Geng J, Fan H, Tang X, Zhai H, Zhang Z. Diversified pattern of the
-human colorectal cancer microbiome. Gut Pathogens. 2013;5:2.
+19. \textbf{Geng, J.}, \textbf{H. Fan}, \textbf{X. Tang}, \textbf{H.
+Zhai}, and \textbf{Z. Zhang}. 2013. Diversified pattern of the human
+colorectal cancer microbiome. Gut Pathogens \textbf{5}:2.
 
 \hypertarget{ref-dejea_microbiota_2014}{}
-20. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
-Rossetti BJ, et al. Microbiota organization is a distinct feature of
-proximal colorectal cancers. Proceedings of the National Academy of
-Sciences of the United States of America. 2014;111:18321--6.
+20. \textbf{Dejea, C. M.}, \textbf{E. C. Wick}, \textbf{E. M.
+Hechenbleikner}, \textbf{J. R. White}, \textbf{J. L. Mark Welch},
+\textbf{B. J. Rossetti}, \textbf{S. N. Peterson}, \textbf{E. C.
+Snesrud}, \textbf{G. G. Borisy}, \textbf{M. Lazarev}, \textbf{E. Stein},
+\textbf{J. Vadivelu}, \textbf{A. C. Roslani}, \textbf{A. A. Malik},
+\textbf{J. W. Wanyiri}, \textbf{K. L. Goh}, \textbf{I. Thevambiga},
+\textbf{K. Fu}, \textbf{F. Wan}, \textbf{N. Llosa}, \textbf{F.
+Housseau}, \textbf{K. Romans}, \textbf{X. Wu}, \textbf{F. M.
+McAllister}, \textbf{S. Wu}, \textbf{B. Vogelstein}, \textbf{K. W.
+Kinzler}, \textbf{D. M. Pardoll}, and \textbf{C. L. Sears}. 2014.
+Microbiota organization is a distinct feature of proximal colorectal
+cancers. Proceedings of the National Academy of Sciences of the United
+States of America \textbf{111}:18321--18326.
 
 \hypertarget{ref-ecoli_Arthur_2014}{}
-21. Arthur JC, Gharaibeh RZ, Mühlbauer M, Perez-Chanona E, Uronis JM,
-McCafferty J, et al. Microbial genomic analysis reveals the essential
-role of inflammation in bacteria-induced colorectal cancer. Nature
-Communications {[}Internet{]}. Springer Nature; 2014;5:4724. Available
-from: \url{https://doi.org/10.1038/ncomms5724}
+21. \textbf{Arthur, J. C.}, \textbf{R. Z. Gharaibeh}, \textbf{M.
+Mühlbauer}, \textbf{E. Perez-Chanona}, \textbf{J. M. Uronis}, \textbf{J.
+McCafferty}, \textbf{A. A. Fodor}, and \textbf{C. Jobin}. 2014.
+Microbial genomic analysis reveals the essential role of inflammation in
+bacteria-induced colorectal cancer. Nature Communications. Springer
+Nature \textbf{5}:4724.
 
 \hypertarget{ref-strep_Aymeric_2017}{}
-22. Aymeric L, Donnadieu F, Mulet C, Merle L du, Nigro G, Saffarian A,
-et al. Colorectal cancer specific conditions promoteStreptococcus
-gallolyticusgut colonization. Proceedings of the National Academy of
-Sciences {[}Internet{]}. Proceedings of the National Academy of
-Sciences; 2017;115:E283--91. Available from:
-\url{https://doi.org/10.1073/pnas.1715112115}
+22. \textbf{Aymeric, L.}, \textbf{F. Donnadieu}, \textbf{C. Mulet},
+\textbf{L. du Merle}, \textbf{G. Nigro}, \textbf{A. Saffarian},
+\textbf{M. Bérard}, \textbf{C. Poyart}, \textbf{S. Robine}, \textbf{B.
+Regnault}, \textbf{P. Trieu-Cuot}, \textbf{P. J. Sansonetti}, and
+\textbf{S. Dramsi}. 2017. Colorectal cancer specific conditions
+promoteStreptococcus gallolyticusgut colonization. Proceedings of the
+National Academy of Sciences. Proceedings of the National Academy of
+Sciences \textbf{115}:E283--E291.
 
 \hypertarget{ref-weir_stool_2013}{}
-23. Weir TL, Manter DK, Sheflin AM, Barnett BA, Heuberger AL, Ryan EP.
-Stool microbiome and metabolome differences between colorectal cancer
-patients and healthy adults. PloS One. 2013;8:e70803.
+23. \textbf{Weir, T. L.}, \textbf{D. K. Manter}, \textbf{A. M. Sheflin},
+\textbf{B. A. Barnett}, \textbf{A. L. Heuberger}, and \textbf{E. P.
+Ryan}. 2013. Stool microbiome and metabolome differences between
+colorectal cancer patients and healthy adults. PloS One
+\textbf{8}:e70803.
 
 \hypertarget{ref-bfrag_Boleij_2014}{}
-24. Boleij A, Hechenbleikner EM, Goodwin AC, Badani R, Stein EM, Lazarev
-MG, et al. The bacteroides fragilis toxin gene is prevalent in the colon
-mucosa of colorectal cancer patients. Clinical Infectious Diseases
-{[}Internet{]}. Oxford University Press (OUP); 2014;60:208--15.
-Available from: \url{https://doi.org/10.1093/cid/ciu787}
+24. \textbf{Boleij, A.}, \textbf{E. M. Hechenbleikner}, \textbf{A. C.
+Goodwin}, \textbf{R. Badani}, \textbf{E. M. Stein}, \textbf{M. G.
+Lazarev}, \textbf{B. Ellis}, \textbf{K. C. Carroll}, \textbf{E.
+Albesiano}, \textbf{E. C. Wick}, \textbf{E. A. Platz}, \textbf{D. M.
+Pardoll}, and \textbf{C. L. Sears}. 2014. The bacteroides fragilis toxin
+gene is prevalent in the colon mucosa of colorectal cancer patients.
+Clinical Infectious Diseases. Oxford University Press (OUP)
+\textbf{60}:208--215.
 
 \hypertarget{ref-sanapareddy_increased_2012}{}
-25. Sanapareddy N, Legge RM, Jovov B, McCoy A, Burcal L, Araujo-Perez F,
-et al. Increased rectal microbial richness is associated with the
-presence of colorectal adenomas in humans. The ISME journal.
-2012;6:1858--68.
+25. \textbf{Sanapareddy, N.}, \textbf{R. M. Legge}, \textbf{B. Jovov},
+\textbf{A. McCoy}, \textbf{L. Burcal}, \textbf{F. Araujo-Perez},
+\textbf{T. A. Randall}, \textbf{J. Galanko}, \textbf{A. Benson},
+\textbf{R. S. Sandler}, \textbf{J. F. Rawls}, \textbf{Z. Abdo},
+\textbf{A. A. Fodor}, and \textbf{T. O. Keku}. 2012. Increased rectal
+microbial richness is associated with the presence of colorectal
+adenomas in humans. The ISME journal \textbf{6}:1858--1868.
 
 \hypertarget{ref-lu_mucosal_2016}{}
-26. Lu Y, Chen J, Zheng J, Hu G, Wang J, Huang C, et al. Mucosal
-adherent bacterial dysbiosis in patients with colorectal adenomas.
-Scientific Reports. 2016;6:26337.
+26. \textbf{Lu, Y.}, \textbf{J. Chen}, \textbf{J. Zheng}, \textbf{G.
+Hu}, \textbf{J. Wang}, \textbf{C. Huang}, \textbf{L. Lou}, \textbf{X.
+Wang}, and \textbf{Y. Zeng}. 2016. Mucosal adherent bacterial dysbiosis
+in patients with colorectal adenomas. Scientific Reports
+\textbf{6}:26337.
 
 \hypertarget{ref-hale_shifts_2017}{}
-27. Hale VL, Chen J, Johnson S, Harrington SC, Yab TC, Smyrk TC, et al.
+27. \textbf{Hale, V. L.}, \textbf{J. Chen}, \textbf{S. Johnson},
+\textbf{S. C. Harrington}, \textbf{T. C. Yab}, \textbf{T. C. Smyrk},
+\textbf{H. Nelson}, \textbf{L. A. Boardman}, \textbf{B. R. Druliner},
+\textbf{T. R. Levin}, \textbf{D. K. Rex}, \textbf{D. J. Ahnen},
+\textbf{P. Lance}, \textbf{D. A. Ahlquist}, and \textbf{N. Chia}. 2017.
 Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
@@
 American Association for Cancer Research, Cosponsored by the American
-Society of Preventive Oncology. 2017;26:85--94.
+Society of Preventive Oncology \textbf{26}:85--94.
 
 \hypertarget{ref-shah_leveraging_2017}{}
-28. Shah MS, DeSantis TZ, Weinmaier T, McMurdie PJ, Cope JL, Altrichter
-A, et al. Leveraging sequence-based faecal microbial community survey
-data to identify a composite biomarker for colorectal cancer. Gut. 2017;
+28. \textbf{Shah, M. S.}, \textbf{T. Z. DeSantis}, \textbf{T.
+Weinmaier}, \textbf{P. J. McMurdie}, \textbf{J. L. Cope}, \textbf{A.
+Altrichter}, \textbf{J.-M. Yamal}, and \textbf{E. B. Hollister}. 2017.
+Leveraging sequence-based faecal microbial community survey data to
+identify a composite biomarker for colorectal cancer. Gut.
 
 \hypertarget{ref-brim_microbiome_2013}{}
-29. Brim H, Yooseph S, Zoetendal EG, Lee E, Torralbo M, Laiyemo AO, et
-al. Microbiome analysis of stool samples from African Americans with
-colon polyps. PloS One. 2013;8:e81352.
+29. \textbf{Brim, H.}, \textbf{S. Yooseph}, \textbf{E. G. Zoetendal},
+\textbf{E. Lee}, \textbf{M. Torralbo}, \textbf{A. O. Laiyemo},
+\textbf{B. Shokrani}, \textbf{K. Nelson}, and \textbf{H. Ashktorab}.
+2013. Microbiome analysis of stool samples from African Americans with
+colon polyps. PloS One \textbf{8}:e81352.
+
+\hypertarget{ref-Hannigan2017}{}
+30. \textbf{Hannigan, G. D.}, \textbf{M. B. Duhaime}, \textbf{M. T.
+Ruffin}, \textbf{C. C. Koumpouras}, and \textbf{P. D. Schloss}. 2017.
+Diagnostic potential \& the interactive dynamics of the colorectal
+cancer virome. Cold Spring Harbor Laboratory.
+
+\hypertarget{ref-Venkataraman2016}{}
+31. \textbf{Venkataraman, A.}, \textbf{J. R. Sieber}, \textbf{A. W.
+Schmidt}, \textbf{C. Waldron}, \textbf{K. R. Theis}, and \textbf{T. M.
+Schmidt}. 2016. Variable responses of human microbiomes to dietary
+supplementation with resistant starch. Microbiome. Springer Nature
+\textbf{4}.
+
+\hypertarget{ref-Herrmann2018}{}
+32. \textbf{Herrmann, E.}, \textbf{W. Young}, \textbf{V.
+Reichert-Grimm}, \textbf{S. Weis}, \textbf{C. Riedel}, \textbf{D.
+Rosendale}, \textbf{H. Stoklosinski}, \textbf{M. Hunt}, and \textbf{M.
+Egert}. 2018. In vivo assessment of resistant starch degradation by the
+caecal microbiota of mice using RNA-based stable isotope probingA
+proof-of-principle study. Nutrients. MDPI AG \textbf{10}:179.
+
+\hypertarget{ref-Reichardt2017}{}
+33. \textbf{Reichardt, N.}, \textbf{M. Vollmer}, \textbf{G. Holtrop},
+\textbf{F. M. Farquharson}, \textbf{D. Wefers}, \textbf{M. Bunzel},
+\textbf{S. H. Duncan}, \textbf{J. E. Drew}, \textbf{L. M. Williams},
+\textbf{G. Milligan}, \textbf{T. Preston}, \textbf{D. Morrison},
+\textbf{H. J. Flint}, and \textbf{P. Louis}. 2017. Specific
+substrate-driven changes in human faecal microbiota composition contrast
+with functional redundancy in short-chain fatty acid production. The
+ISME Journal. Springer Nature \textbf{12}:610--622.
+
+\hypertarget{ref-Sze2017}{}
+34. \textbf{Sze, M. A.}, \textbf{N. T. Baxter}, \textbf{M. T. Ruffin},
+\textbf{M. A. M. Rogers}, and \textbf{P. D. Schloss}. 2017.
+Normalization of the microbiota in patients after treatment for colonic
+lesions. Microbiome. Springer Nature \textbf{5}.
 
 \hypertarget{ref-Flynn_preprint_2017}{}
-30. Flynn KJ, Ruffin MT, Turgeon DK, Schloss PD. Spatial variation of
-the native colon microbiota in healthy adults. Cold Spring Harbor
-Laboratory; 2017; Available from: \url{https://doi.org/10.1101/189886}
+35. \textbf{Flynn, K. J.}, \textbf{M. T. Ruffin}, \textbf{D. K.
+Turgeon}, and \textbf{P. D. Schloss}. 2017. Spatial variation of the
+native colon microbiota in healthy adults. Cold Spring Harbor
+Laboratory.
 
 \hypertarget{ref-Salter_contamination_2014}{}
-31. Salter SJ, Cox MJ, Turek EM, Calus ST, Cookson WO, Moffatt MF, et
-al. Reagent and laboratory contamination can critically impact
-sequence-based microbiome analyses. BMC Biology {[}Internet{]}. Springer
-Nature; 2014;12. Available from:
-\url{https://doi.org/10.1186/s12915-014-0087-z}
+36. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
+\textbf{S. T. Calus}, \textbf{W. O. Cookson}, \textbf{M. F. Moffatt},
+\textbf{P. Turner}, \textbf{J. Parkhill}, \textbf{N. J. Loman}, and
+\textbf{A. W. Walker}. 2014. Reagent and laboratory contamination can
+critically impact sequence-based microbiome analyses. BMC Biology.
+Springer Nature \textbf{12}.
+
+\hypertarget{ref-Purcell2017}{}
+37. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
+\textbf{L. Dixon}, \textbf{F. A. Frizelle}, and \textbf{J. I. Keenan}.
+2017. Colonization with enterotoxigenic bacteroides fragilis is
+associated with early-stage colorectal neoplasia. PLOS ONE. Public
+Library of Science (PLoS) \textbf{12}:e0171602.
 
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-32. Keku TO, Dulal S, Deveaux A, Jovov B, Han X. The gastrointestinal
+38. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
+\textbf{B. Jovov}, and \textbf{X. Han}. 2015. The gastrointestinal
 microbiota and colorectal cancer. American Journal of Physiology -
-Gastrointestinal and Liver Physiology {[}Internet{]}. 2015 {[}cited 2017
-Oct 30{]};308:G351--63. Available from:
-\url{http://ajpgi.physiology.org/lookup/doi/10.1152/ajpgi.00360.2012}
+Gastrointestinal and Liver Physiology \textbf{308}:G351--G363.
 
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-33. Vogtmann E, Goedert JJ. Epidemiologic studies of the human
-microbiome and cancer. British Journal of Cancer {[}Internet{]}. 2016
-{[}cited 2017 Oct 30{]};114:237--42. Available from:
-\url{http://www.nature.com/doifinder/10.1038/bjc.2015.465}
+39. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
+Epidemiologic studies of the human microbiome and cancer. British
+Journal of Cancer \textbf{114}:237--242.
 
 \hypertarget{ref-kostic_genomic_2012}{}
-34. Kostic AD, Gevers D, Pedamallu CS, Michaud M, Duke F, Earl AM, et
-al. Genomic analysis identifies association of Fusobacterium with
-colorectal carcinoma. Genome Research. 2012;22:292--8.
+40. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
+Pedamallu}, \textbf{M. Michaud}, \textbf{F. Duke}, \textbf{A. M. Earl},
+\textbf{A. I. Ojesina}, \textbf{J. Jung}, \textbf{A. J. Bass},
+\textbf{J. Tabernero}, \textbf{J. Baselga}, \textbf{C. Liu}, \textbf{R.
+A. Shivdasani}, \textbf{S. Ogino}, \textbf{B. W. Birren}, \textbf{C.
+Huttenhower}, \textbf{W. S. Garrett}, and \textbf{M. Meyerson}. 2012.
+Genomic analysis identifies association of Fusobacterium with colorectal
+carcinoma. Genome Research \textbf{22}:292--298.
 
 \hypertarget{ref-zackular_human_2014}{}
-35. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. The human gut
-microbiome as a screening tool for colorectal cancer. Cancer Prevention
-Research (Philadelphia, Pa). 2014;7:1112--21.
+41. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
+Ruffin}, and \textbf{P. D. Schloss}. 2014. The human gut microbiome as a
+screening tool for colorectal cancer. Cancer Prevention Research
+(Philadelphia, Pa.) \textbf{7}:1112--1121.
 
 \hypertarget{ref-schloss_introducing_2009}{}
-36. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
-EB, et al. Introducing mothur: Open-Source, Platform-Independent,
-Community-Supported Software for Describing and Comparing Microbial
-Communities. ApplEnvironMicrobiol {[}Internet{]}. 2009 {[}cited 12AD Jan
-1{]};75:7537--41. Available from:
-\url{http://aem.asm.org/cgi/content/abstract/75/23/7537}
+42. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
+Ryabin}, \textbf{J. R. Hall}, \textbf{M. Hartmann}, \textbf{E. B.
+Hollister}, \textbf{R. A. Lesniewski}, \textbf{B. B. Oakley}, \textbf{D.
+H. Parks}, \textbf{C. J. Robinson}, \textbf{J. W. Sahl}, \textbf{B.
+Stres}, \textbf{G. G. Thallinger}, \textbf{D. J. Van Horn}, and
+\textbf{C. F. Weber}. 2009. Introducing mothur: Open-Source,
+Platform-Independent, Community-Supported Software for Describing and
+Comparing Microbial Communities. Appl.Environ.Microbiol.
+\textbf{75}:7537--7541.
 
 \hypertarget{ref-rognes_vsearch_2016}{}
-37. Rognes T, Flouri T, Nichols B, Quince C, Mahé F. VSEARCH: A
-versatile open source tool for metagenomics. PeerJ. 2016;4:e2584.
+43. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
+\textbf{C. Quince}, and \textbf{F. Mahé}. 2016. VSEARCH: A versatile
+open source tool for metagenomics. PeerJ \textbf{4}:e2584.
 
 \hypertarget{ref-westcott_opticlust_2017}{}
-38. Westcott SL, Schloss PD. OptiClust, an Improved Method for Assigning
-Amplicon-Based Sequence Data to Operational Taxonomic Units. mSphere.
-2017;2.
+44. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
+OptiClust, an Improved Method for Assigning Amplicon-Based Sequence Data
+to Operational Taxonomic Units. mSphere \textbf{2}.
 
 \hypertarget{ref-benjamini_controlling_1995}{}
-39. Benjamini Y, Hochberg Y. Controlling the false discovery rate: A
-practical and powerful approach to multiple testing. Journal of the
-Royal Statistical Society Series B (Methodological). 1995;57:289--300.
+45. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
+the false discovery rate: A practical and powerful approach to multiple
+testing. Journal of the Royal Statistical Society. Series B
+(Methodological) \textbf{57}:289--300.
 
 \hypertarget{ref-r_citation_2017}{}
-40. R Core Team. R: A language and environment for statistical computing
-{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
-2017. Available from: \url{https://www.R-project.org/}
+46. \textbf{R Core Team}. 2017. R: A language and environment for
+statistical computing. R Foundation for Statistical Computing, Vienna,
+Austria.
 
 \hypertarget{ref-rcompanion_citation_2017}{}
-41. Mangiafico S. Rcompanion: Functions to support extension education
-program evaluation {[}Internet{]}. 2017. Available from:
-\url{https://CRAN.R-project.org/package=rcompanion}
+47. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
+extension education program evaluation.
 
 \hypertarget{ref-car_citation_2011}{}
-42. Fox J, Weisberg S. An R companion to applied regression
-{[}Internet{]}. Second. Thousand Oaks CA: Sage; 2011. Available from:
-\url{http://socserv.socsci.mcmaster.ca/jfox/Books/Companion}
+48. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
+applied regressionSecond. Sage, Thousand Oaks CA.
 
 \hypertarget{ref-lme4_citation_2015}{}
-43. Bates D, Mächler M, Bolker B, Walker S. Fitting linear mixed-effects
-models using lme4. Journal of Statistical Software. 2015;67:1--48.
+49. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
+\textbf{S. Walker}. 2015. Fitting linear mixed-effects models using
+lme4. Journal of Statistical Software \textbf{67}:1--48.
 
 \hypertarget{ref-epir_citation_2017}{}
-44. Telmo Nunes MS with contributions from, Heuer C, Marshall J, Sanchez
-J, Thornton R, Reiczigel J, et al. EpiR: Tools for the analysis of
-epidemiological data {[}Internet{]}. 2017. Available from:
-\url{https://CRAN.R-project.org/package=epiR}
+50. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
+Heuer}, \textbf{J. Marshall}, \textbf{J. Sanchez}, \textbf{R. Thornton},
+\textbf{J. Reiczigel}, \textbf{J. Robison-Cox}, \textbf{P. Sebastiani},
+\textbf{P. Solymos}, \textbf{K. Yoshida}, \textbf{G. Jones}, \textbf{S.
+Pirikahu}, \textbf{S. Firestone}, and \textbf{R. Kyle.} 2017. EpiR:
+Tools for the analysis of epidemiological data.
 
 \hypertarget{ref-metafor_citation_2010}{}
-45. Viechtbauer W. Conducting meta-analyses in R with the metafor
-package. Journal of Statistical Software {[}Internet{]}. 2010;36:1--48.
-Available from: \url{http://www.jstatsoft.org/v36/i03/}
+51. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
+the metafor package. Journal of Statistical Software \textbf{36}:1--48.
 
 \hypertarget{ref-vegan_citation_2017}{}
-46. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
-et al. Vegan: Community ecology package {[}Internet{]}. 2017. Available
-from: \url{https://CRAN.R-project.org/package=vegan}
+52. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
+\textbf{R. Kindt}, \textbf{P. Legendre}, \textbf{D. McGlinn}, \textbf{P.
+R. Minchin}, \textbf{R. B. O'Hara}, \textbf{G. L. Simpson}, \textbf{P.
+Solymos}, \textbf{M. H. H. Stevens}, \textbf{E. Szoecs}, and \textbf{H.
+Wagner}. 2017. Vegan: Community ecology package.
 
 \hypertarget{ref-caret_citation_2017}{}
-47. Jed Wing MKC from, Weston S, Williams A, Keefer C, Engelhardt A,
-Cooper T, et al. Caret: Classification and regression training
-{[}Internet{]}. 2017. Available from:
-\url{https://CRAN.R-project.org/package=caret}
+53. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
+Williams}, \textbf{C. Keefer}, \textbf{A. Engelhardt}, \textbf{T.
+Cooper}, \textbf{Z. Mayer}, \textbf{B. Kenkel}, \textbf{the R Core
+Team}, \textbf{M. Benesty}, \textbf{R. Lescarbeau}, \textbf{A. Ziem},
+\textbf{L. Scrucca}, \textbf{Y. Tang}, \textbf{C. Candan}, and
+\textbf{T. Hunt.} 2017. Caret: Classification and regression training.
 
 \hypertarget{ref-randomforest_citation_2002}{}
-48. Liaw A, Wiener M. Classification and regression by randomForest. R
-News {[}Internet{]}. 2002;2:18--22. Available from:
-\url{http://CRAN.R-project.org/doc/Rnews/}
+54. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
+regression by randomForest. R News \textbf{2}:18--22.
 
 \hypertarget{ref-ggplot2_citation_2009}{}
-49. Wickham H. Ggplot2: Elegant graphics for data analysis
-{[}Internet{]}. Springer-Verlag New York; 2009. Available from:
-\url{http://ggplot2.org}
+55. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
+analysis. Springer-Verlag New York.
 
 \hypertarget{ref-gridextra_citation_2017}{}
-50. Auguie B. GridExtra: Miscellaneous functions for ``grid'' graphics
-{[}Internet{]}. 2017. Available from:
-\url{https://CRAN.R-project.org/package=gridExtra}
+56. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
+``grid'' graphics.
 

### 2018-03-02 submission/manuscript.Rmd
complete discussion section with pat suggestions

@@
 
-Meta analysis are important (eg. obesity, crc), we need more. We learned that these are hard because…
-Inaccessible datasets
-Missing/vague metadata
-Varying sequence quality
-Small datasets
+Despite these limitations the findings that we present here would not be possible without performing a meta-analysis.
+These types of studies can be a useful tool in microbiota research because they can both validate existing research and make new discoveries by pooling many independent investigations together.
+Yet, it is still difficult to perform these studies because of inaccessible 16S sequencing data, missing or vague metadata (e.g.
+which samples are carcioma and which are not), varying sequence quality, and multiple small data sets.
+Better attention to these specific problems could help to increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses.
+Moving forward, meta-analyses will be important tools to help aggregate and find commonalities across studies when investigating the microbiota in the context of a specific disease and more are needed [@Sze2016; @Walters2014; @Finucane2014; @shah_leveraging_2017].
 

### 2018-03-02 submission/manuscript.tex
complete discussion section with pat suggestions

@@
 
-Meta analysis are important (eg. obesity, crc), we need more. We learned
-that these are hard because\ldots{} Inaccessible datasets Missing/vague
-metadata Varying sequence quality Small datasets
+Despite these limitations the findings that we present here would not be
+possible without performing a meta-analysis. These types of studies can
+be a useful tool in microbiota research because they can both validate
+existing research and make new discoveries by pooling many independent
+investigations together. Yet, it is still difficult to perform these
+studies because of inaccessible 16S sequencing data, missing or vague
+metadata (e.g.~which samples are carcioma and which are not), varying
+sequence quality, and multiple small data sets. Better attention to
+these specific problems could help to increase the reproducibility and
+replicability of microbiota studies and make it easier to perform these
+crucial meta-analyses. Moving forward, meta-analyses will be important
+tools to help aggregate and find commonalities across studies when
+investigating the microbiota in the context of a specific disease and
+more are needed (28, 38--40).
 
@@
 meta-analysis were identified through the review articles written by
-Keku, \emph{et al.} and Vogtmann, \emph{et al.} (38, 39) and additional
+Keku, \emph{et al.} and Vogtmann, \emph{et al.} (41, 42) and additional
 studies not mentioned in the reviews were obtained based on the authors'
@@
 \emph{et al.} (20), Flemer, \emph{et al.} (17), Geng, \emph{et al.}
-(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (40), Lu, \emph{et
+(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (43), Lu, \emph{et
 al.} (26), Sanapareddy, \emph{et al.} (25), Wang, \emph{et al.} (14),
 Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
-(41) study was not included because the 90 individuals analyzed within
+(44) study was not included because the 90 individuals analyzed within
 the study are contained within the larger Baxter study (12). After
@@
 Each study was processed using the mothur (v1.39.3) software program
-(42) and quality filtering utilized the default methods for both 454 and
+(45) and quality filtering utilized the default methods for both 454 and
 Illumina based sequencing. If it was not possible to use the defaults,
@@
 different sequences and commonly known as chimeras were identified and
-removed using VSEARCH (43) before \emph{de novo} OTU clustering at 97\%
-similarity was completed using the OptiClust algorithm (44).
+removed using VSEARCH (46) before \emph{de novo} OTU clustering at 97\%
+similarity was completed using the OptiClust algorithm (47).
 
@@
 ORs for individuals having an adenoma or carcinoma and corrected for
-multiple comparisons using the Benjamini-Hochberg method (45). We then
+multiple comparisons using the Benjamini-Hochberg method (48). We then
 built Random Forest models based on all genera, all OTUs, or significant
@@
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package (46). For
+sequence processing utilized the R (v3.4.3) software package (49). For
 OTU richness, evenness, and Shannon diversity analysis, values were
-power transformed using the rcompanion (v1.11.1) package (47) and then
-Z-score normalized using the car (v2.1.6) package (48). Testing for OTU
+power transformed using the rcompanion (v1.11.1) package (50) and then
+Z-score normalized using the car (v2.1.6) package (51). Testing for OTU
 richness, evenness, and Shannon diversity differences utilized linear
-mixed-effect models created using the lme4 (v1.1.15) package (49) to
+mixed-effect models created using the lme4 (v1.1.15) package (52) to
 correct for study, repeat sampling of individuals (tissue only), and 16S
 hyper-variable region used. Odds ratios (OR) were analyzed using both
-the epiR (v0.9.93) and metafor (v2.0.0) packages (50, 51) by assessing
+the epiR (v0.9.93) and metafor (v2.0.0) packages (53, 54) by assessing
 how many individuals with and without disease were above and below the
@@
 Bray-Curtis index utilized the creation of distance matrix and testing
-with PERMANOVA executed with the vegan (v2.4.5) package (52). Random
+with PERMANOVA executed with the vegan (v2.4.5) package (55). Random
 Forest models were built using both the caret (v6.0.78) and randomForest
-(v4.6.12) packages (53, 54). All figures were created using both ggplot2
-(v2.2.1) and gridExtra (v2.3) packages (55, 56).
+(v4.6.12) packages (56, 57). All figures were created using both ggplot2
+(v2.2.1) and gridExtra (v2.3) packages (58, 59).
 
@@
 
+\hypertarget{ref-Sze2016}{}
+38. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
+signal in the noise: Revisiting obesity and the microbiome. mBio.
+American Society for Microbiology \textbf{7}:e01018--16.
+
+\hypertarget{ref-Walters2014}{}
+39. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
+2014. Meta-analyses of human gut microbes associated with obesity and
+IBD. FEBS Letters. Wiley-Blackwell \textbf{588}:4223--4233.
+
+\hypertarget{ref-Finucane2014}{}
+40. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
+Laurent}, and \textbf{K. S. Pollard}. 2014. A taxonomic signature of
+obesity in the microbiome? Getting to the guts of the matter. PLoS ONE.
+Public Library of Science (PLoS) \textbf{9}:e84689.
+
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-38. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
+41. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
 \textbf{B. Jovov}, and \textbf{X. Han}. 2015. The gastrointestinal
@@
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-39. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
+42. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
 Epidemiologic studies of the human microbiome and cancer. British
@@
 \hypertarget{ref-kostic_genomic_2012}{}
-40. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
+43. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
 Pedamallu}, \textbf{M. Michaud}, \textbf{F. Duke}, \textbf{A. M. Earl},
@@
 \hypertarget{ref-zackular_human_2014}{}
-41. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
+44. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
 Ruffin}, and \textbf{P. D. Schloss}. 2014. The human gut microbiome as a
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-42. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
+45. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
 Ryabin}, \textbf{J. R. Hall}, \textbf{M. Hartmann}, \textbf{E. B.
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-43. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
+46. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
 \textbf{C. Quince}, and \textbf{F. Mahé}. 2016. VSEARCH: A versatile
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-44. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
+47. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
 OptiClust, an Improved Method for Assigning Amplicon-Based Sequence Data
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-45. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
+48. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
 the false discovery rate: A practical and powerful approach to multiple
@@
 \hypertarget{ref-r_citation_2017}{}
-46. \textbf{R Core Team}. 2017. R: A language and environment for
+49. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-47. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
+50. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
 extension education program evaluation.
@@
 \hypertarget{ref-car_citation_2011}{}
-48. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
+51. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
 applied regressionSecond. Sage, Thousand Oaks CA.
@@
 \hypertarget{ref-lme4_citation_2015}{}
-49. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
+52. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
 \textbf{S. Walker}. 2015. Fitting linear mixed-effects models using
@@
 \hypertarget{ref-epir_citation_2017}{}
-50. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
+53. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
 Heuer}, \textbf{J. Marshall}, \textbf{J. Sanchez}, \textbf{R. Thornton},
@@
 \hypertarget{ref-metafor_citation_2010}{}
-51. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
+54. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
 the metafor package. Journal of Statistical Software \textbf{36}:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-52. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
+55. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
 \textbf{R. Kindt}, \textbf{P. Legendre}, \textbf{D. McGlinn}, \textbf{P.
@@
 \hypertarget{ref-caret_citation_2017}{}
-53. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
+56. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
 Williams}, \textbf{C. Keefer}, \textbf{A. Engelhardt}, \textbf{T.
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-54. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
+57. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-55. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
+58. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
 analysis. Springer-Verlag New York.
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-56. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
+59. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
 ``grid'' graphics.

### 2018-03-02 submission/manuscript.Rmd
flow update

@@
 This has been suggested in previous literature where metagenomics did not perform better than 16S rRNA gene sequencing data at classifying individuals with normal colons and those with carcinomas [@Hannigan2017].
-One possible reason as to why lower level classification may not result in better models is that the communities are patchy and higher level taxonomic information pools some of this patchiness allowing for better prediction models.
+One possible reason as to why lower level classification may not result in better models is that the communities are patchy and higher level taxonomic information pools some of this patchiness, allowing for better prediction models.
 There may also be a fair bit of data redundancy within models that utilize more of the community.
@@
 
-Our observations also suggest that a small collection of taxa can classify diease as well as full OTU-based models but that these taxa individually perform quite poorly [Figure 3].
-This result supports the contention that there might be redundancy of function even amongst the taxa included in the significant OR models.
-As an example multiple different microbes could be similarily stimulating the activation of inflammatory pathways and by doing so exacerbate disease progression.
-Multiple reports within the literature have found that different bacteria, such as *Escherichia coli* and *Fusobacterium nucleatum*, can worsen similarily worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013].
+Our observations also suggest that a small collection of taxa can classify disease as well as full OTU-based models but that these taxa individually perform quite poorly [Figure 3].
+This result supports the contention that there might be redundancy of function amongst the taxa included in the significant OR models.
+As an example, multiple different microbes could be similarily stimulating the activation of inflammatory pathways and by doing so exacerbate disease progression.
+Multiple reports within the literature have found that different bacteria, such as *Escherichia coli* and *Fusobacterium nucleatum*, can similarily worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013].
 Although the inflammatory taxa were patchy in their importance and presence across studies those that were not typically associated with inflammation were consistently important for every study [Figure 6].
@@
 However, the modeling results are at odds with results obtained in Baxter, *et al.* [@baxter_microbiota-based_2016].
-There are some major differences between the models generated in this meta-analysis and what was used in this previous report.
+There are some major differences between the models generated in this meta-analysis and what was used in Baxter, *et al*.
 First, the prevoius report's models investigated the classification of lesions (individuals with adenoma or carcinoma) and not adenoma alone.
@@
 It is possible that we might have been able to detect differences in the bacterial community if advanced adenomas were separated from adenomas but that data was not available for the majority of studies analyzed.
-It is also possible that the initial changes to the bacterial community are focal to where the initial adenoma develops and would not be easily assessed with a fecal sample.
+Additionally, the initial changes to the bacterial community could be focal to where the initial adenoma develops and would not be easily assessed with a fecal sample.
 
-Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- associated microbial communities.
-Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017].
-Sampling the microbiota directly associated with colon tissue may provide clearer answers but the colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors.
+Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- tissue associated microbial communities.
+The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors.
 Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S3].
-*Fusobacterium* was not identified consistently across studies and this could be due to both a small number of studies and a small sample size within these studies.
-Additionally, the majority of the colon tissue-based results were study specific with many of the top 10 taxa being present only in a single study.
-The presence of genera associated with contamination [@Salter_contamination_2014], within the top 10 most important variables for the genera and OTU models is worrying (e.g.
+*Fusobacterium* was not consistently identified across studies and this could be due to both a small number of studies and a small sample size within these studies.
+The majority of the colon tissue-based results were also study specific with many of the top 10 taxa being present only in a single study.
+Additionally, the presence of genera associated with contamination [@Salter_contamination_2014], within the top 10 most important variables for the genera and OTU models is worrying (e.g.
 *Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.).
@@
 *gallolyticus*.
-As an example, the stool-based full community models consistently identified the genus *Bacteroides*, as well as OTUs that classified as *Bacteroides*, to be important model components across studies.
+As an example, the stool-based full community models consistently identified the *Bacteroides* taxa to be important model components across studies.
 This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.
-Additionally, *Streptococcus gallolyticus* subsp.
-*gallolyticus* is a mouth-associated microbe, and the results from this study suggest that regardless of sample type, mouth-associated genera are commonly associated with an increased OR for individuals to have a carcinoma tumor.
 
-Despite these limitations the findings that we present here would not be possible without performing a meta-analysis.
-These types of studies can be a useful tool in microbiota research because they can both validate existing research and make new discoveries by pooling many independent investigations together.
+Despite these limitations to the findings, meta-analyses are a useful tool in microbiota research because they can both validate existing research and make new discoveries by pooling many independent investigations together.
 Yet, it is still difficult to perform these studies because of inaccessible 16S sequencing data, missing or vague metadata (e.g.
@@
 By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in carcinoma tumors.
-Although further validation of the biomarkers presented here need to be undertaken, the replciability of the AUC of a specific collection of taxa across multiple studies suggests a strong potential for the use of the microbiota as a risk stratification tool for individuals with carcinomas.
+Although further validation of the biomarkers presented here need to be undertaken, the replicability of the AUC of a specific collection of taxa across multiple studies suggests a strong potential for the use of the microbiota as a risk stratification tool for individuals with carcinomas.
 

### 2018-03-02 submission/manuscript.Rmd
final re-read edits

@@
 Although being able to classify individuals with adenomas is important, the classification of advanced adenomas is a more clinically meaningful diagnostic tool (i.e.
-those that are at high risk of progressing to a carcinoma).
+those that are at highest risk of progressing to a carcinoma).
 It is possible that we might have been able to detect differences in the bacterial community if advanced adenomas were separated from adenomas but that data was not available for the majority of studies analyzed.
@@
 The majority of the colon tissue-based results were also study specific with many of the top 10 taxa being present only in a single study.
-Additionally, the presence of genera associated with contamination [@Salter_contamination_2014], within the top 10 most important variables for the genera and OTU models is worrying (e.g.
-*Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.).
+Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying (e.g.
+*Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.
+[@Salter_contamination_2014]).
 The low bacterial biomass of tissue samples coupled with potential contamination and small sample sizes could explain why these results seem to be more sporadic than the stool results.

### 2018-03-02 submission/manuscript.tex
final re-read edits

@@
 result in better models is that the communities are patchy and higher
-level taxonomic information pools some of this patchiness allowing for
+level taxonomic information pools some of this patchiness, allowing for
 better prediction models. There may also be a fair bit of data
@@
 Our observations also suggest that a small collection of taxa can
-classify diease as well as full OTU-based models but that these taxa
+classify disease as well as full OTU-based models but that these taxa
 individually perform quite poorly {[}Figure 3{]}. This result supports
-the contention that there might be redundancy of function even amongst
-the taxa included in the significant OR models. As an example multiple
+the contention that there might be redundancy of function amongst the
+taxa included in the significant OR models. As an example, multiple
 different microbes could be similarily stimulating the activation of
@@
 bacteria, such as \emph{Escherichia coli} and \emph{Fusobacterium
-nucleatum}, can worsen similarily worsen inflammation in mouse models of
+nucleatum}, can similarily worsen inflammation in mouse models of
 tumorigenesis (5, 6, 21). Although the inflammatory taxa were patchy in
@@
 There are some major differences between the models generated in this
-meta-analysis and what was used in this previous report. First, the
+meta-analysis and what was used in Baxter, \emph{et al}. First, the
 prevoius report's models investigated the classification of lesions
@@
 adenomas is important, the classification of advanced adenomas is a more
-clinically meaningful diagnostic tool (i.e.~those that are at high risk
-of progressing to a carcinoma). It is possible that we might have been
-able to detect differences in the bacterial community if advanced
+clinically meaningful diagnostic tool (i.e.~those that are at highest
+risk of progressing to a carcinoma). It is possible that we might have
+been able to detect differences in the bacterial community if advanced
 adenomas were separated from adenomas but that data was not available
-for the majority of studies analyzed. It is also possible that the
-initial changes to the bacterial community are focal to where the
-initial adenoma develops and would not be easily assessed with a fecal
-sample.
+for the majority of studies analyzed. Additionally, the initial changes
+to the bacterial community could be focal to where the initial adenoma
+develops and would not be easily assessed with a fecal sample.
 
@@
 it is not clear how well this sample reflects adenoma- and carcinoma-
-associated microbial communities. Some studies have tried to assess this
-in health and disease but are limited by their sample size (17, 35).
-Sampling the microbiota directly associated with colon tissue may
-provide clearer answers but the colon tissue-based studies did not
-provide a clearer understanding of how the microbiota may be associated
-with tumors. Generally, the full OTU-based models of unmatched and
-matched colon tissue samples were concordant with stool samples showing
-that GI resident microbes were the most prevalent in the top 10 most
-important variables across study {[}Figure S3{]}. \emph{Fusobacterium}
-was not identified consistently across studies and this could be due to
-both a small number of studies and a small sample size within these
-studies. Additionally, the majority of the colon tissue-based results
-were study specific with many of the top 10 taxa being present only in a
-single study. The presence of genera associated with contamination (36),
-within the top 10 most important variables for the genera and OTU models
-is worrying (e.g. \emph{Novosphingobium}, \emph{Acidobacteria Gp2},
-\emph{Sphingomonas}, etc.). The low bacterial biomass of tissue samples
-coupled with potential contamination and small sample sizes could
-explain why these results seem to be more sporadic than the stool
-results.
+tissue associated microbial communities. The colon tissue-based studies
+did not provide a clearer understanding of how the microbiota may be
+associated with tumors. Generally, the full OTU-based models of
+unmatched and matched colon tissue samples were concordant with stool
+samples showing that GI resident microbes were the most prevalent in the
+top 10 most important variables across study {[}Figure S3{]}.
+\emph{Fusobacterium} was not consistently identified across studies and
+this could be due to both a small number of studies and a small sample
+size within these studies. The majority of the colon tissue-based
+results were also study specific with many of the top 10 taxa being
+present only in a single study. Additionally, the presence of genera
+associated with contamination, within the top 10 most important
+variables for the genera and OTU models is worrying (e.g.
+\emph{Novosphingobium}, \emph{Acidobacteria Gp2}, \emph{Sphingomonas},
+etc. (35)). The low bacterial biomass of tissue samples coupled with
+potential contamination and small sample sizes could explain why these
+results seem to be more sporadic than the stool results.
 
@@
 to use in this meta-analysis could not identify these types of
-differences (37). Additionally, since we are limited in our aggregation
+differences (36). Additionally, since we are limited in our aggregation
 of the data to the genus level, it is not possible to clearly delineate
@@
 \emph{gallolyticus}. As an example, the stool-based full community
-models consistently identified the genus \emph{Bacteroides}, as well as
-OTUs that classified as \emph{Bacteroides}, to be important model
-components across studies. This suggests that even though
-\emph{Bacteroides} may not increase the OR of individuals having an
-adenoma or carcinoma and may not vary in relative abundance, like
-\emph{Fusobacterium}, it is still important in CRC. Additionally,
-\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} is a
-mouth-associated microbe, and the results from this study suggest that
-regardless of sample type, mouth-associated genera are commonly
-associated with an increased OR for individuals to have a carcinoma
-tumor.
-
-Despite these limitations the findings that we present here would not be
-possible without performing a meta-analysis. These types of studies can
-be a useful tool in microbiota research because they can both validate
-existing research and make new discoveries by pooling many independent
+models consistently identified the \emph{Bacteroides} taxa to be
+important model components across studies. This suggests that even
+though \emph{Bacteroides} may not increase the OR of individuals having
+an adenoma or carcinoma and may not vary in relative abundance, like
+\emph{Fusobacterium}, it is still important in CRC.
+
+Despite these limitations to the findings, meta-analyses are a useful
+tool in microbiota research because they can both validate existing
+research and make new discoveries by pooling many independent
 investigations together. Yet, it is still difficult to perform these
@@
 investigating the microbiota in the context of a specific disease and
-more are needed (28, 38--40).
+more are needed (28, 37--39).
 
@@
 tumors. Although further validation of the biomarkers presented here
-need to be undertaken, the replciability of the AUC of a specific
+need to be undertaken, the replicability of the AUC of a specific
 collection of taxa across multiple studies suggests a strong potential
@@
 meta-analysis were identified through the review articles written by
-Keku, \emph{et al.} and Vogtmann, \emph{et al.} (41, 42) and additional
+Keku, \emph{et al.} and Vogtmann, \emph{et al.} (40, 41) and additional
 studies not mentioned in the reviews were obtained based on the authors'
@@
 \emph{et al.} (20), Flemer, \emph{et al.} (17), Geng, \emph{et al.}
-(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (43), Lu, \emph{et
+(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (42), Lu, \emph{et
 al.} (26), Sanapareddy, \emph{et al.} (25), Wang, \emph{et al.} (14),
 Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
-(44) study was not included because the 90 individuals analyzed within
+(43) study was not included because the 90 individuals analyzed within
 the study are contained within the larger Baxter study (12). After
@@
 Each study was processed using the mothur (v1.39.3) software program
-(45) and quality filtering utilized the default methods for both 454 and
+(44) and quality filtering utilized the default methods for both 454 and
 Illumina based sequencing. If it was not possible to use the defaults,
@@
 different sequences and commonly known as chimeras were identified and
-removed using VSEARCH (46) before \emph{de novo} OTU clustering at 97\%
-similarity was completed using the OptiClust algorithm (47).
+removed using VSEARCH (45) before \emph{de novo} OTU clustering at 97\%
+similarity was completed using the OptiClust algorithm (46).
 
@@
 ORs for individuals having an adenoma or carcinoma and corrected for
-multiple comparisons using the Benjamini-Hochberg method (48). We then
+multiple comparisons using the Benjamini-Hochberg method (47). We then
 built Random Forest models based on all genera, all OTUs, or significant
@@
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package (49). For
+sequence processing utilized the R (v3.4.3) software package (48). For
 OTU richness, evenness, and Shannon diversity analysis, values were
-power transformed using the rcompanion (v1.11.1) package (50) and then
-Z-score normalized using the car (v2.1.6) package (51). Testing for OTU
+power transformed using the rcompanion (v1.11.1) package (49) and then
+Z-score normalized using the car (v2.1.6) package (50). Testing for OTU
 richness, evenness, and Shannon diversity differences utilized linear
-mixed-effect models created using the lme4 (v1.1.15) package (52) to
+mixed-effect models created using the lme4 (v1.1.15) package (51) to
 correct for study, repeat sampling of individuals (tissue only), and 16S
 hyper-variable region used. Odds ratios (OR) were analyzed using both
-the epiR (v0.9.93) and metafor (v2.0.0) packages (53, 54) by assessing
+the epiR (v0.9.93) and metafor (v2.0.0) packages (52, 53) by assessing
 how many individuals with and without disease were above and below the
@@
 Bray-Curtis index utilized the creation of distance matrix and testing
-with PERMANOVA executed with the vegan (v2.4.5) package (55). Random
+with PERMANOVA executed with the vegan (v2.4.5) package (54). Random
 Forest models were built using both the caret (v6.0.78) and randomForest
-(v4.6.12) packages (56, 57). All figures were created using both ggplot2
-(v2.2.1) and gridExtra (v2.3) packages (58, 59).
+(v4.6.12) packages (55, 56). All figures were created using both ggplot2
+(v2.2.1) and gridExtra (v2.3) packages (57, 58).
 
@@
 
-\hypertarget{ref-Flynn_preprint_2017}{}
-35. \textbf{Flynn, K. J.}, \textbf{M. T. Ruffin}, \textbf{D. K.
-Turgeon}, and \textbf{P. D. Schloss}. 2017. Spatial variation of the
-native colon microbiota in healthy adults. Cold Spring Harbor
-Laboratory.
-
 \hypertarget{ref-Salter_contamination_2014}{}
-36. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
+35. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
 \textbf{S. T. Calus}, \textbf{W. O. Cookson}, \textbf{M. F. Moffatt},
@@
 \hypertarget{ref-Purcell2017}{}
-37. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
+36. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
 \textbf{L. Dixon}, \textbf{F. A. Frizelle}, and \textbf{J. I. Keenan}.
@@
 \hypertarget{ref-Sze2016}{}
-38. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
+37. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
 signal in the noise: Revisiting obesity and the microbiome. mBio.
@@
 \hypertarget{ref-Walters2014}{}
-39. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
+38. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
 2014. Meta-analyses of human gut microbes associated with obesity and
@@
 \hypertarget{ref-Finucane2014}{}
-40. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
+39. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
 Laurent}, and \textbf{K. S. Pollard}. 2014. A taxonomic signature of
@@
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-41. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
+40. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
 \textbf{B. Jovov}, and \textbf{X. Han}. 2015. The gastrointestinal
@@
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-42. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
+41. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
 Epidemiologic studies of the human microbiome and cancer. British
@@
 \hypertarget{ref-kostic_genomic_2012}{}
-43. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
+42. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
 Pedamallu}, \textbf{M. Michaud}, \textbf{F. Duke}, \textbf{A. M. Earl},
@@
 \hypertarget{ref-zackular_human_2014}{}
-44. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
+43. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
 Ruffin}, and \textbf{P. D. Schloss}. 2014. The human gut microbiome as a
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-45. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
+44. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
 Ryabin}, \textbf{J. R. Hall}, \textbf{M. Hartmann}, \textbf{E. B.
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-46. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
+45. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
 \textbf{C. Quince}, and \textbf{F. Mahé}. 2016. VSEARCH: A versatile
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-47. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
+46. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
 OptiClust, an Improved Method for Assigning Amplicon-Based Sequence Data
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-48. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
+47. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
 the false discovery rate: A practical and powerful approach to multiple
@@
 \hypertarget{ref-r_citation_2017}{}
-49. \textbf{R Core Team}. 2017. R: A language and environment for
+48. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-50. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
+49. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
 extension education program evaluation.
@@
 \hypertarget{ref-car_citation_2011}{}
-51. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
+50. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
 applied regressionSecond. Sage, Thousand Oaks CA.
@@
 \hypertarget{ref-lme4_citation_2015}{}
-52. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
+51. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
 \textbf{S. Walker}. 2015. Fitting linear mixed-effects models using
@@
 \hypertarget{ref-epir_citation_2017}{}
-53. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
+52. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
 Heuer}, \textbf{J. Marshall}, \textbf{J. Sanchez}, \textbf{R. Thornton},
@@
 \hypertarget{ref-metafor_citation_2010}{}
-54. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
+53. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
 the metafor package. Journal of Statistical Software \textbf{36}:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-55. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
+54. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
 \textbf{R. Kindt}, \textbf{P. Legendre}, \textbf{D. McGlinn}, \textbf{P.
@@
 \hypertarget{ref-caret_citation_2017}{}
-56. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
+55. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
 Williams}, \textbf{C. Keefer}, \textbf{A. Engelhardt}, \textbf{T.
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-57. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
+56. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-58. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
+57. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
 analysis. Springer-Verlag New York.
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-59. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
+58. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
 ``grid'' graphics.

### 2018-03-03 submission/manuscript.Rmd
add importance, rm conclusion, change to sig OR only model

@@
 **Results.** We quantified the odds ratios for individual bacterial genera that were associated with an individual having tumors relative to a normal colon.
-Among the stool samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_OR_stool$measure)` genera with significant odds ratios associated with carcinoma.
+Among the stool samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_OR_stool$measure)` genera with significant odds ratios (ORs) associated with carcinoma.
 Similarly, among the tissue samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` genera with significant odds ratios associated with carcinoma.
 Among the significant odds ratios, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`.
-Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using the genera with the five highest and lowest odds ratios, using the entire collection of genera found in each study, and using operational taxonomic units defined based on a 97% similarity threshold.
+Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using only the taxa that had significant ORs, using the entire collection of taxa found in each study, and using operational taxonomic units defined based on a 97% similarity threshold.
 All training approaches yielded similar classification success as measured using the Area Under the Curve.
@@
 
-**Conclusions.** This meta-analysis confirms previous results indicating that individuals with adenomas cannot be readily classified based on their bacterial community, but that those with carcinomas can.
+\newpage
+
+**Importance.** This meta-analysis confirms previous results indicating that individuals with adenomas cannot be readily classified based on their bacterial community, but that those with carcinomas can.
 Regardless of the dataset, we found a subset of the fecal community that was associated with carcinomas was as predictive as the full community.

### 2018-03-03 submission/manuscript.tex
add importance, rm conclusion, change to sig OR only model

@@
 significant odds ratio associated with adenoma and there were 8 genera
-with significant odds ratios associated with carcinoma. Similarly, among
-the tissue samples, there were no genera that had a significant odds
-ratio associated with adenoma and there were 3 genera with significant
-odds ratios associated with carcinoma. Among the significant odds
-ratios, the association between individual taxa and tumor diagnosis was
-equal or below 7.11. Because individual taxa had limited association
+with significant odds ratios (ORs) associated with carcinoma. Similarly,
+among the tissue samples, there were no genera that had a significant
+odds ratio associated with adenoma and there were 3 genera with
+significant odds ratios associated with carcinoma. Among the significant
+odds ratios, the association between individual taxa and tumor diagnosis
+was equal or below 7.11. Because individual taxa had limited association
 with tumor diagnosis, we trained Random Forest classification models
-using the genera with the five highest and lowest odds ratios, using the
-entire collection of genera found in each study, and using operational
-taxonomic units defined based on a 97\% similarity threshold. All
-training approaches yielded similar classification success as measured
-using the Area Under the Curve. The ability to correctly classify
-individuals with adenomas was poor and the ability to classify
-individuals with carcinomas was considerably better using sequences from
-stool or tissue.
-
-\textbf{Conclusions.} This meta-analysis confirms previous results
+using only the taxa that had significant ORs, using the entire
+collection of taxa found in each study, and using operational taxonomic
+units defined based on a 97\% similarity threshold. All training
+approaches yielded similar classification success as measured using the
+Area Under the Curve. The ability to correctly classify individuals with
+adenomas was poor and the ability to classify individuals with
+carcinomas was considerably better using sequences from stool or tissue.
+
+\newpage
+
+\textbf{Importance.} This meta-analysis confirms previous results
 indicating that individuals with adenomas cannot be readily classified

### 2018-03-07 submission/manuscript.Rmd
incorporate comments from Pat on results section

@@
 
+# Create a correlation table for size versus median test AUC
+correlation_data <- as.data.frame(crc_stool_model_aucs %>% filter(model_type == "full_model", train_model != study) %>% group_by(study) %>% summarise(auc_median = median(AUC))) %>% left_join(stool_n) %>%
+  mutate(totals = control + cancer) %>% ungroup()
+
 ```
@@
 
-***Lower Bacterial Diversity is Associated with Increased Odds Ratio (OR) of Tumors:***
+***Lower bacterial diversity is associated with increased odds ratio (OR) of tumors:***
 We first assessed whether variation in broad community metrics like total number of operational taxonomic units (OTUs) (i.e.
@@
 
-***Disease Progression is Associated with Community-Wide Changes in Composition and Abundance:*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages.
+***Disease progression is associated with community-wide changes in composition and abundance:*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages.
 We identified significant bacterial community differences in the stool of patients with adenomas relative to those with normal colons in `r length(as.data.frame(filter(adn_beta_stool, pvalue < 0.05))[, "study"])` of `r length(as.data.frame(adn_beta_stool)[, "study"])` studies and in patients with carcinomas relative to those with normal colons in 6 of 7 studies (PERMANOVA; P-value < 0.05) [Table S2].
@@
 
-***Individual Taxa are Associated with Significant ORs for Carcinomas:*** 
+***Individual taxa are associated with significant ORs for carcinomas:*** 
 Next we identified those taxa were associated with ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas.
@@
 In contrast, `r length(crc_OR_stool$measure)` taxa had significant ORs for the presence of carcinomas using data from stool samples.
-Of these, 4 are commonly associated with the oral cavity: Fusobacterium (OR = `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`)), Parvimonas (OR = `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_ub"], digits = 2))`)), Porphyromonas (OR = `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2))`)), and Peptostreptococcus (OR = `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_ub"], digits = 2))`)) [Table S3].
-The other 4 were Clostridium XI (OR = `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_ub"], digits = 2))`)), Enterobacteriaceae  (OR = `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_ub"], digits = 2))`)), Escherichia (OR = `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_ub"], digits = 2))`)), and Ruminococcus (OR = `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_ub"], digits = 2))`)).
+Of these, 4 are commonly associated with the oral cavity: *Fusobacterium* (OR = `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`)), *Parvimonas* (OR = `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_ub"], digits = 2))`)), *Porphyromonas* (OR = `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2))`)), and *Peptostreptococcus* (OR = `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_ub"], digits = 2))`)) [Table S3].
+The other 4 were *Clostridium XI* (OR = `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_ub"], digits = 2))`)), *Enterobacteriaceae*  (OR = `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_ub"], digits = 2))`)), *Escherichia* (OR = `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_ub"], digits = 2))`)), and *Ruminococcus* (OR = `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_ub"], digits = 2))`)).
 Among the data collected from tissue samples, only unmatched carcinoma samples had taxa with a significant OR.
-Those included Dorea (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2))`)), Blautia (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2))`)), and Weissella (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2))`)).
+Those included *Dorea* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2))`)), *Blautia* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2))`)), and *Weissella* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2))`)).
 Mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4].
-For example, Fusobacterium had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`; however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg-corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4].
-It is interesting to note that Ruminococcus and members of Clostridium group XI in stool and Dorea and Blautia in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas.
+For example, *Fusobacterium* had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`; however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg-corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4].
+It is interesting to note that *Ruminococcus* and members of *Clostridium group XI* in stool and *Dorea* and *Blautia* in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas.
 Overall, there was no overlap in the taxa with significant OR between stool and tissue samples.
@@
  We next asked whether those taxa that had a significant OR associated with having a normal colon or carcinomas could be used individually, to classify subjects as having a normal colon or carcinomas.
-Whereas the OR was defined based on whether the relative abundance for a taxon in a subject was above or below the median relative abundance for that taxon across all subjects in a study, we generated receiver operator characteristic (ROC) curves for each taxon in each study and calculated the area under the curve (AUC).
+OR values were caluclated based on whether the relative abundance for a taxon in a subject was above or below the median relative abundance for that taxon across all subjects in a study.
+To measure the ability of these taxa to classify individuals we instead generated receiver operator characteristic (ROC) curves for each taxon in each study and calculated the area under the curve (AUC).
 This allowed us to use a more fluid relative abundance threshold for defining disease status.
@@
 AUC=0.50) [Figure 3A].
-The taxa that performed the best included Clostridium XI, Ruminococcus, and Escherichia and even these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)`.
+The taxa that performed the best included *Clostridium XI*, *Ruminococcus*, and *Escherichia*, however, these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)`.
 Likewise, in unmatched tissue samples the `r length(crc_OR_stool$measure)` taxa with significant ORs taxa were marginally better than one would expect by chance [Figure 3B].
@@
 ***Combined taxa model classifies subjects better than using individual taxa:*** 
-Instead of attempting to classify subjects based on individual taxa, next we generated Random Forest models that combined the individual taxa and evaluated the ability to classify as subject’s disease status.
+Instead of attempting to classify subjects based on individual taxa, next we combined information from the individual taxa and evaluated the ability to classify a subject’s disease status using Random Forest models.
 For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`).
 For the full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies [Figure S2].
+Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study.
+The most important taxa across study within the significant OR taxa only models for stool were *Ruminococcus* and *Clostridium XI* [Figure 5A].
 Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for Blautia and Weissella (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).
 For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S3].
-Clearly, pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.
+For the singificant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the important based on mean decrease in accuracy [Figure 5B].
+Pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.
 
@@
 Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR.
-Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were marginally better than 0.5 for any study[Figure 4A & S4A].
+Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were only marginally better than 0.5 for any study (median AUC = `r median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)` [`r min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)` - `r max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)`]) [Figure 4A & S4A].
 In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 4B & S4B-C].
 When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed.
-Using the data from stool samples we found that although the AUC for 6 of 7 studies increased (mean decrease = `r round(mean((filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] - filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"])*100), digits = 2)`%), the more expansive models performed worse for 1 of the studies (decrease = `r round(abs(mean((filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] - filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"])*100)), digits = 2)`%).
-The overall improvement in performance was statistically significant (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
-Of the `r length(crc_OR_stool$measure)` taxa with significant ORs, all `r length(crc_OR_stool$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study .
-Similarly, using the data from unmatched tissue samples we found that the AUC for 4 out of 4 studies decreased between full versus select OR models (mean decrease = `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
-Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study.
-The most important taxa across study within the significant OR taxa only models for stool were *Ruminococcus* and *Clostridium XI* [Figure 5A].
-For the singificant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the important based on mean decrease in accuracy [Figure 5B].
-These results were surprising because it demonstrated that the ability to classify subjects could be done based on a limited characterization of the communities.
+Using the data from stool samples we found that the AUC for 6 of 7 studies increased by an average of `r round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"]))*100), digits = 2)`%) and AUC for the Flemer study decreased by `r round(abs(mean(((1 - filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"]))*100)), digits = 2)`%).
+The overall improvement in performance was statistically significant (mean = `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"]))*100), digits = 2))`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
+Similarly, using the data from unmatched tissue samples we found that the AUC of studies increased by an average of `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`% when we used all of the taxa rather than the 3 taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
+Although the significant OR taxa models can classify indidivuals with and without carcinoma tumors, they are still missing taxa from the full community models that can increase the model accuracy.
 
 ***Performance of models based on OTU relative abundance in full community are not significantly better than those based on taxa with significant ORs:*** 
-The previous models were based on relative abundance data where sequences were assigned to coarse taxonomic assignments (i.e.
+The previous models were based on relative abundance data where sequences were classified to coarse taxonomic assignments (i.e.
 typically genus or family level).
-To determine whether model performance improved with a more fine scale classification, we assigned sequences to operational taxonomic units (OTUs) where the similarity among sequences within an OTU was more than 97%.
+To determine whether model performance improved with finer scale classification, we assigned sequences to operational taxonomic units (OTUs) where the similarity among sequences within an OTU was more than 97%.
 We again found that classification models built using all of the sequence data for a community did a poor job of differentiating between subjects with normal colons and those with adenomas (median AUC: `r round(median(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]), but did a good job of differentiating between subjects with normal colons and those with carcinomas (median AUC: `r round(median(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]).
-The OTU-based models performed similarly to those constructed using the taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) and those using all taxa (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
+The OTU-based models performed similarly to those constructed using the taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) and those using all taxa (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) [Figure 4].
 Among the OTUs that had the highest mean decrease in accuracy for the OTU-based models, we found that OTUs that affiliated with all of the `r length(crc_OR_stool$measure)` taxa that had a significant OR were within the top 10% for at least one study.
-Again, this result was surprising as it indicated that a finer scale classification of the sequence data and thus a larger number of features to select from, did not yield improved classification of the subjects.
+This result was surprising as it indicated that a finer scale classification of the sequences and thus a larger number of features to select from, did not yield improved classification of the subjects.
 
@@
 We found the models trained using the taxa with a significant OR all had a higher median AUC than the models trained using all of the taxa when tested on the other datasets [Figure 6 & S5].
-As might be expected, the difference between the performance of the modelling approaches appeared to vary with the size of the training cohort [Figure 6].
+As might be expected, the difference between the performance of the modelling approaches appeared to vary with the size of the training cohort (R^2^ = `r round(lm(auc_median ~ totals, data = correlation_data)$coefficients[1], digits = 2)`) [Figure 6].
 These data suggest that given a sufficient number of subjects with normal colons and carcinomas, Random Forest models trained using a small number of taxa can accurately classify individuals from a different cohort.

### 2018-03-07 submission/manuscript.tex
incorporate comments from Pat on results section

@@
 
-\textbf{\emph{Lower Bacterial Diversity is Associated with Increased
-Odds Ratio (OR) of Tumors:}} We first assessed whether variation in
+\textbf{\emph{Lower bacterial diversity is associated with increased
+odds ratio (OR) of tumors:}} We first assessed whether variation in
 broad community metrics like total number of operational taxonomic units
@@
 
-\textbf{\emph{Disease Progression is Associated with Community-Wide
-Changes in Composition and Abundance:}} Based on the differences in
+\textbf{\emph{Disease progression is associated with community-wide
+changes in composition and abundance:}} Based on the differences in
 evenness and diversity, we next asked whether there were community-wide
@@
 
-\textbf{\emph{Individual Taxa are Associated with Significant ORs for
-Carcinomas:}} Next we identified those taxa were associated with ORs
+\textbf{\emph{Individual taxa are associated with significant ORs for
+carcinomas:}} Next we identified those taxa were associated with ORs
 that were significantly associated with having a normal colon or the
@@
 the presence of carcinomas using data from stool samples. Of these, 4
-are commonly associated with the oral cavity: Fusobacterium (OR = 2.74
-(1.95 - 3.85)), Parvimonas (OR = 3.07 (2.11 - 4.46)), Porphyromonas (OR
-= 3.2 (2.26 - 4.54)), and Peptostreptococcus (OR = 7.11 (3.84 - 13.17))
-{[}Table S3{]}. The other 4 were Clostridium XI (OR = 0.65 (0.49 -
-0.86)), Enterobacteriaceae (OR = 1.79 (1.33 - 2.41)), Escherichia (OR =
-2.15 (1.57 - 2.95)), and Ruminococcus (OR = 0.63 (0.48 - 0.83)). Among
-the data collected from tissue samples, only unmatched carcinoma samples
-had taxa with a significant OR. Those included Dorea (OR = 0.35 (0.22 -
-0.55)), Blautia (OR = 0.47 (0.3 - 0.73)), and Weissella (OR = 5.15 (2.02
-- 13.14)). Mouth-associated genera were not significantly associated
-with an increased OR for carcinoma in tissue samples {[}Table S4{]}. For
-example, Fusobacterium had an OR of 3.98 (1.19 - 13.24; however, due to
-the small number of studies and considerable variation in the data, the
-Benjimani-Hochberg-corrected P-value was 0.93 {[}Table S4{]}. It is
-interesting to note that Ruminococcus and members of Clostridium group
-XI in stool and Dorea and Blautia in tissue had ORs that were
-significantly less than 1.0, which suggests that these populations are
-protective against the development of carcinomas. Overall, there was no
-overlap in the taxa with significant OR between stool and tissue
-samples.
+are commonly associated with the oral cavity: \emph{Fusobacterium} (OR =
+2.74 (1.95 - 3.85)), \emph{Parvimonas} (OR = 3.07 (2.11 - 4.46)),
+\emph{Porphyromonas} (OR = 3.2 (2.26 - 4.54)), and
+\emph{Peptostreptococcus} (OR = 7.11 (3.84 - 13.17)) {[}Table S3{]}. The
+other 4 were \emph{Clostridium XI} (OR = 0.65 (0.49 - 0.86)),
+\emph{Enterobacteriaceae} (OR = 1.79 (1.33 - 2.41)), \emph{Escherichia}
+(OR = 2.15 (1.57 - 2.95)), and \emph{Ruminococcus} (OR = 0.63 (0.48 -
+0.83)). Among the data collected from tissue samples, only unmatched
+carcinoma samples had taxa with a significant OR. Those included
+\emph{Dorea} (OR = 0.35 (0.22 - 0.55)), \emph{Blautia} (OR = 0.47 (0.3 -
+0.73)), and \emph{Weissella} (OR = 5.15 (2.02 - 13.14)).
+Mouth-associated genera were not significantly associated with an
+increased OR for carcinoma in tissue samples {[}Table S4{]}. For
+example, \emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24; however,
+due to the small number of studies and considerable variation in the
+data, the Benjimani-Hochberg-corrected P-value was 0.93 {[}Table S4{]}.
+It is interesting to note that \emph{Ruminococcus} and members of
+\emph{Clostridium group XI} in stool and \emph{Dorea} and \emph{Blautia}
+in tissue had ORs that were significantly less than 1.0, which suggests
+that these populations are protective against the development of
+carcinomas. Overall, there was no overlap in the taxa with significant
+OR between stool and tissue samples.
 
@@
 with having a normal colon or carcinomas could be used individually, to
-classify subjects as having a normal colon or carcinomas. Whereas the OR
-was defined based on whether the relative abundance for a taxon in a
+classify subjects as having a normal colon or carcinomas. OR values were
+caluclated based on whether the relative abundance for a taxon in a
 subject was above or below the median relative abundance for that taxon
-across all subjects in a study, we generated receiver operator
+across all subjects in a study. To measure the ability of these taxa to
+classify individuals we instead generated receiver operator
 characteristic (ROC) curves for each taxon in each study and calculated
@@
 than one would expect by chance (i.e.~AUC=0.50) {[}Figure 3A{]}. The
-taxa that performed the best included Clostridium XI, Ruminococcus, and
-Escherichia and even these had median AUC values less than 0.588.
-Likewise, in unmatched tissue samples the 8 taxa with significant ORs
-taxa were marginally better than one would expect by chance {[}Figure
-3B{]}. The relative abundance of Dorea was the best predictor of
-carcinomas and its median AUC was only 0.62. These results suggest that
-although these taxa are associated with a decreased or increased OR for
-the presences of carcinomas, individually, they do a poor job of
-classifying a subject's disease status.
+taxa that performed the best included \emph{Clostridium XI},
+\emph{Ruminococcus}, and \emph{Escherichia}, however, these had median
+AUC values less than 0.588. Likewise, in unmatched tissue samples the 8
+taxa with significant ORs taxa were marginally better than one would
+expect by chance {[}Figure 3B{]}. The relative abundance of Dorea was
+the best predictor of carcinomas and its median AUC was only 0.62. These
+results suggest that although these taxa are associated with a decreased
+or increased OR for the presences of carcinomas, individually, they do a
+poor job of classifying a subject's disease status.
 
@@
 individual taxa:}} Instead of attempting to classify subjects based on
-individual taxa, next we generated Random Forest models that combined
-the individual taxa and evaluated the ability to classify as subject's
-disease status. For data from stool samples, the combined model had an
-AUC of 0.75, which was significantly higher than any of the AUC values
-for the individual taxa (P-value \textless{} 0.033). For the full taxa
-models using stool, \emph{Bacteroides} and \emph{Lachnospiraceae} were
-the most common taxa in the top 10\% mean decrease in accuracy (MDA)
-across studies {[}Figure S2{]}. Similarly, using data from the unmatched
-tissue samples, the combined model had an AUC of 0.77, which was
-significantly higher than the AUC values for Blautia and Weissella
-(P-value \textless{} 0.037). For the full taxa models using unmatched
-tissue, \emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
-\emph{Ruminococcaceae} were the most common taxa in the top 10\% mean
-decrease in accuracy across studies {[}Figure S3{]}. Clearly, pooling
-the information from the taxa with significant ORs results in a model
-that outperforms classifications made using individual taxa.
+individual taxa, next we combined information from the individual taxa
+and evaluated the ability to classify a subject's disease status using
+Random Forest models. For data from stool samples, the combined model
+had an AUC of 0.75, which was significantly higher than any of the AUC
+values for the individual taxa (P-value \textless{} 0.033). For the full
+taxa models using stool, \emph{Bacteroides} and \emph{Lachnospiraceae}
+were the most common taxa in the top 10\% mean decrease in accuracy
+(MDA) across studies {[}Figure S2{]}. Of the 3 taxa with significant
+ORs, all 3 were among the top 10\% most important taxa as measured by
+mean decrease in accuracy, in at least one study. The most important
+taxa across study within the significant OR taxa only models for stool
+were \emph{Ruminococcus} and \emph{Clostridium XI} {[}Figure 5A{]}.
+Similarly, using data from the unmatched tissue samples, the combined
+model had an AUC of 0.77, which was significantly higher than the AUC
+values for Blautia and Weissella (P-value \textless{} 0.037). For the
+full taxa models using unmatched tissue, \emph{Lachnospiraceae},
+\emph{Bacteroidaceae}, and \emph{Ruminococcaceae} were the most common
+taxa in the top 10\% mean decrease in accuracy across studies {[}Figure
+S3{]}. For the singificant OR taxa unmatched tissue models both
+\emph{Dorea} and \emph{Blautia} were the important based on mean
+decrease in accuracy {[}Figure 5B{]}. Pooling the information from the
+taxa with significant ORs results in a model that outperforms
+classifications made using individual taxa.
 
@@
 normal colons or having adenomas using data from stool or tissue samples
-were marginally better than 0.5 for any study{[}Figure 4A \& S4A{]}. In
-contrast, the models for classifying subjects as having normal colons or
-having carcinomas using data from stool or tissue samples yielded AUC
-values meaningfully higher than 0.5 {[}Figure 4B \& S4B-C{]}. When we
-compared the models based on all of the taxa in a community to models
-based on the taxa with significant ORs, the results were mixed. Using
-the data from stool samples we found that although the AUC for 6 of 7
-studies increased (mean decrease = 9.53\%), the more expansive models
-performed worse for 1 of the studies (decrease = 0.38\%). The overall
-improvement in performance was statistically significant (one-tailed
-paired T-test; P-value = 0.005). Of the 8 taxa with significant ORs, all
-8 were among the top 10\% most important taxa as measured by mean
-decrease in accuracy, in at least one study . Similarly, using the data
-from unmatched tissue samples we found that the AUC for 4 out of 4
-studies decreased between full versus select OR models (mean decrease =
-19.11\%, one-tailed paired T-test; P-value = 0.03). Of the 3 taxa with
-significant ORs, all 3 were among the top 10\% most important taxa as
-measured by mean decrease in accuracy, in at least one study. The most
-important taxa across study within the significant OR taxa only models
-for stool were \emph{Ruminococcus} and \emph{Clostridium XI} {[}Figure
-5A{]}. For the singificant OR taxa unmatched tissue models both
-\emph{Dorea} and \emph{Blautia} were the important based on mean
-decrease in accuracy {[}Figure 5B{]}. These results were surprising
-because it demonstrated that the ability to classify subjects could be
-done based on a limited characterization of the communities.
+were only marginally better than 0.5 for any study (median AUC =
+0.5486298 {[}0.3671667 - 0.971{]}) {[}Figure 4A \& S4A{]}. In contrast,
+the models for classifying subjects as having normal colons or having
+carcinomas using data from stool or tissue samples yielded AUC values
+meaningfully higher than 0.5 {[}Figure 4B \& S4B-C{]}. When we compared
+the models based on all of the taxa in a community to models based on
+the taxa with significant ORs, the results were mixed. Using the data
+from stool samples we found that the AUC for 6 of 7 studies increased by
+an average of -14.8\%) and AUC for the Flemer study decreased by
+0.54\%). The overall improvement in performance was statistically
+significant (mean = 12.61\%, one-tailed paired T-test; P-value = 0.005).
+Similarly, using the data from unmatched tissue samples we found that
+the AUC of studies increased by an average of 19.11\% when we used all
+of the taxa rather than the 3 taxa with significant ORs (one-tailed
+paired T-test; P-value = 0.03). Although the significant OR taxa models
+can classify indidivuals with and without carcinoma tumors, they are
+still missing taxa from the full community models that can increase the
+model accuracy.
 
@@
 with significant ORs:}} The previous models were based on relative
-abundance data where sequences were assigned to coarse taxonomic
+abundance data where sequences were classified to coarse taxonomic
 assignments (i.e.~typically genus or family level). To determine whether
-model performance improved with a more fine scale classification, we
-assigned sequences to operational taxonomic units (OTUs) where the
-similarity among sequences within an OTU was more than 97\%. We again
-found that classification models built using all of the sequence data
-for a community did a poor job of differentiating between subjects with
-normal colons and those with adenomas (median AUC: 0.53 {[}0.37-
-0.56{]}), but did a good job of differentiating between subjects with
-normal colons and those with carcinomas (median AUC: 0.71 {[}0.5-
-0.9{]}). The OTU-based models performed similarly to those constructed
-using the taxa with significant ORs (one-tailed paired T-test; P-value =
-0.966) and those using all taxa (one-tailed paired T-test; P-value =
-0.146). Among the OTUs that had the highest mean decrease in accuracy
-for the OTU-based models, we found that OTUs that affiliated with all of
-the 8 taxa that had a significant OR were within the top 10\% for at
-least one study. Again, this result was surprising as it indicated that
-a finer scale classification of the sequence data and thus a larger
+model performance improved with finer scale classification, we assigned
+sequences to operational taxonomic units (OTUs) where the similarity
+among sequences within an OTU was more than 97\%. We again found that
+classification models built using all of the sequence data for a
+community did a poor job of differentiating between subjects with normal
+colons and those with adenomas (median AUC: 0.53 {[}0.37- 0.56{]}), but
+did a good job of differentiating between subjects with normal colons
+and those with carcinomas (median AUC: 0.71 {[}0.5- 0.9{]}). The
+OTU-based models performed similarly to those constructed using the taxa
+with significant ORs (one-tailed paired T-test; P-value = 0.966) and
+those using all taxa (one-tailed paired T-test; P-value = 0.146)
+{[}Figure 4{]}. Among the OTUs that had the highest mean decrease in
+accuracy for the OTU-based models, we found that OTUs that affiliated
+with all of the 8 taxa that had a significant OR were within the top
+10\% for at least one study. This result was surprising as it indicated
+that a finer scale classification of the sequences and thus a larger
 number of features to select from, did not yield improved classification
@@
 between the performance of the modelling approaches appeared to vary
-with the size of the training cohort {[}Figure 6{]}. These data suggest
-that given a sufficient number of subjects with normal colons and
-carcinomas, Random Forest models trained using a small number of taxa
-can accurately classify individuals from a different cohort.
+with the size of the training cohort (R\textsuperscript{2} = 0.66)
+{[}Figure 6{]}. These data suggest that given a sufficient number of
+subjects with normal colons and carcinomas, Random Forest models trained
+using a small number of taxa can accurately classify individuals from a
+different cohort.
 

### 2018-03-08 submission/manuscript.Rmd
add pat edits to the methods

@@
 
-***Obtaining Data Sets:***
-The studies used for this meta-analysis were identified through the review articles written by Keku, *et al.* and Vogtmann, *et al.* [@keku_gastrointestinal_2015; @vogtmann_epidemiologic_2016] and additional studies not mentioned in the reviews were obtained based on the authors' knowledge of the literature.
-Studies that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing analysis and had data sets with sequences available for analysis were included.
-Some studies were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share.
-After these filtering steps, the following studies remained: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014].
-The Zackular [@zackular_human_2014] study was not included because the 90 individuals analyzed within the study are contained within the larger Baxter study [@baxter_microbiota-based_2016].
-After sequence processing, all the case samples for the Kostic study had 100 or less sequences remaining and was excluded, leaving a total of 14 studies that analysis could be completed on.
-
-***Data Set Breakdown:***
-In total, there were seven studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal and tissue samples (Chen and Flemer).
-The total number of individuals analyzed after sequence processing for feces was `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` [Table 1].
-The total number of matched and unmatched tissue samples that were analyzed after sequence processing was `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` [Table 2].
+***Datasets:***
+The studies used for this meta-analysis were identified through the review articles written by Keku, et al.
+[@keku_gastrointestinal_2015] and Vogtmann, et al.
+[@vogtmann_epidemiologic_2016].
+Additional studies, not mentioned in those reviews were obtained based on the authors’ knowledge of the literature.
+Studies were included that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing.
+Some studies (N = 12) were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share.
+We were able to obtain sequence data and metadata from the following studies: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014].
+The Zackular [@zackular_human_2014] study was excluded because their 90 individuals were contained within the larger Baxter study [@baxter_microbiota-based_2016].
+The Kostic study was excluded because after we processed the sequences, all of the case samples had 100 or fewer sequences.
+The final analysis included 14 studies (Tables 1 and 2).
+There were seven studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal and tissue samples (Chen and Flemer).
+After curating the sequences, `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` stool samples and `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` tissue samples remained in the analysis [Tables 1 and 2].
 
 ***Sequence Processing:***
-For the majority of studies, raw sequences were downloaded from the Sequence Read Archive (SRA) (ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/) and metadata were obtained by searching the respective accession number of the study at the following website: http://www.ncbi.nlm.nih.gov/Traces/study/.
-Of the studies that did not have sequences and metadata on the SRA, data was obtained from DBGap (n = 1, [@ahn_human_2013]) and directly from the authors (n = 4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]).
-Each study was processed using the mothur (v1.39.3) software program [@schloss_introducing_2009] and quality filtering utilized the default methods for both 454 and Illumina based sequencing.
-If it was not possible to use the defaults, the stated quality cut-offs, from the study itself, were used instead.
-Sequences that were made up of an artificial combination of two or more different sequences and commonly known as chimeras were identified and removed using VSEARCH [@rognes_vsearch_2016] before *de novo* OTU clustering at 97% similarity was completed using the OptiClust algorithm [@westcott_opticlust_2017].
+Raw sequence data and metadata were primarily obtained from the Sequence Read Archive (SRA) and dbGaP.
+Other sequence and metadata were obtained directly from the authors (n = 4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]).
+Each dataset was processed separately using mothur (v1.39.3) [@schloss_introducing_2009] using the default quality filtering methods for both 454 and Illumina sequence data.
+If it was not possible to use the defaults because the sequences were trimmed too much, then the stated quality cut-offs from the original study were used.
+Chimeric sequences were identified and removed using VSEARCH [@rognes_vsearch_2016].
+The curated sequences were assigned to OTUs at 97% similarity using the OptiClust algorithm [@westcott_opticlust_2017] and classified to the deepest taxonomic level that had 80% support using the naïve Bayesian classifier trained on the RDP taxonomy outline (version 14, [@rdp_Wang2007]).
 
-***Study Analysis Overview:***
-OTU richness, evenness, and Shannon diversity were first assessed for differences between controls, adenoma tumors, and carcinoma turmors using both linear mixed-effect models and ORs.
-For each individual study the Bray-Curtis index was used to assess differences between control-adenoma and control-carcinoma individuals.
-Next, all common genera were assessed for differences in ORs for individuals having an adenoma or carcinoma and corrected for multiple comparisons using the Benjamini-Hochberg method [@benjamini_controlling_1995].
-We then built Random Forest models based on all genera, all OTUs, or significant OR taxa (only using taxa still significant after multiple comparison correction).
-For both the full genera and significant OR taxa, models were trained on one study then tested on the remaining studies using genera-based relative abundances.
-The OTU-based models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs.
-This process was repeated for every study in the meta-analysis.
-Comparisons of the initial trained model AUCs for the full genera and significant OR taxa were made to the mean AUC generated from the 100 different 10-fold CV runs of the respective OTU-based model.
-For comparisons in which only control versus adenoma individuals were made, the carcinoma individuals were excluded from each respective study.
-Similarly, for comparisons in which control versus carcinoma individuals were made the adenoma individuals were excluded from each respective study.
-For all analysis completed fecal and tissue samples were kept separate.
-Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched).
+***Community analysis:*** 
+We calculated alpha diversity metrics (i.e.
+OTU richness, evenness, and Shannon diversity) for each sample.
+Within each dataset, we ensured that the data followed a normal distribution using power transformations.
+Using the transformed data, we tested the hypothesis that individuals with normal colons, adenomas, and carcinomas had significantly different alpha diversity metrics using linear mixed-effect models.
+We also calculated the OR for each study and metric by considering any value above the median alpha diversity value to be positive.
+We measured the dissimilarity between individuals by calculating the pairwise Bray-Curtis index and used PERMANOVA [@permanova_Anderson2013] to test whether individuals with normal colons were significantly different from those with adenomas or carcinomas.
+Finally, after binning sequences into the deepest taxa that the naïve Bayesian classifier could calssify the sequences, we quantified the ORs for individuals having an adenoma or carcinoma and corrected for multiple comparisons using the Benjamini-Hochberg method [@benjamini_controlling_1995].
+Again, for each taxon, if the relative abundance was greater than the median relative abundance for that taxon in the study, the individual was considered to be positive.
 
-***Obtaining Genera Relative Abundance and Significant OR Taxa Models:***
-For the genera analysis of the ORs, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings was obtained.
-The significant OR taxa models for the  Random Forest models utilized all taxa that had significant ORs after multiple comparison correction.
-This meant only models for the carcinoma stool (`r length(crc_OR_stool$measure)` variables) and carcinoma unmatched (`r length(crc_unmatched_OR_tissue$measure)` variables) samples were possible to be created and analyzed.
-
-***Matched versus Unmatched Tissue Samples:***
-In general, tissue samples with control and tumor samples from different individuals were classified as unmatched while samples that belonged to the same individual were classified as matched.
-Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy.
-For some studies samples became unmatched when a corresponding matched sample did not make it through sequence processing.
-All samples, from both matched and unmatched tissue samples, were analyzed together for the linear mixed-effect models with samples from the same individual being corrected for.
-All other analysis, where it is not specified explicitly, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.
-
-***Assessing Important Random Forest Model Variables:***
-Using Mean Decrease in Accuracy (MDA) the top 10 most important variables to the Random Forest model were obtained for the full models of the two different approaches used.
-For the first approach utilizing genus-based models, the number of times that a specific taxa showed up in the top 10 of the training set across each study was counted.
-For the second approach, that utilized the OTU-based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study.
-Common taxa were then identified by using the lowest classification for each of the specific OTUs obtained from these counts and the number of times this classification occurred across the top 10 of each study was recorded.
-Finally, the two studies that had adenoma tumor tissue (Lu and Flemer) were equally divided between matched and unmatched studies and were grouped together for the counting of the top 10 genera and OTUs for both Random Forest approaches.
+***Random Forest classification analysis:*** 
+To classify individuals as having normal colons or tumors, we built Random Forest classification models for each dataset and comparison using taxa with significant ORs (after multiple comparison correction), all taxa, or OTUs.
+Because no taxa were identified as having a significant OR associated with adenomas using stool samples or tissue samples, classification models based on OR data were not constructed to classify individuals as having normal colons or adenomas.
+Within the training dataset, 10-fold cross validation (5-fold cross validation for small datasets) was used to build a model using the default mtry setting that was then evaluated on the testing set.
+For the models constructed using the taxa with significant ORs, the default mtry setting was used to train the model and this model was tested on the other datasets in the meta-analysis.
+The reported AUC values are the average AUCs for the application of the model on the test sets.
+For the OTU-based models, hyperparameters were fit splitting the dataset into training (80% of samples) and testing (20%) sets.
+Within the training dataset, 10-fold cross validation was used to identify the best hyperparameters, which were then used to build a model that was then evaluated on the testing set.
+The original 80/20 split and fitting was repeated 100 times.
+The Mean Decrease in Accuracy (MDA), a measure of the importance of each taxon to the overall model was used to rank the taxa used in each model.
 

### 2018-03-08 submission/manuscript.tex
add pat edits to the methods

@@
 
-\textbf{\emph{Obtaining Data Sets:}} The studies used for this
-meta-analysis were identified through the review articles written by
-Keku, \emph{et al.} and Vogtmann, \emph{et al.} (40, 41) and additional
-studies not mentioned in the reviews were obtained based on the authors'
-knowledge of the literature. Studies that used tissue or feces as their
-sample source for 454 or Illumina 16S rRNA gene sequencing analysis and
-had data sets with sequences available for analysis were included. Some
-studies were excluded because they did not have publicly available
-sequences or did not have metadata in which the authors were able to
-share. After these filtering steps, the following studies remained: Ahn,
+\textbf{\emph{Datasets:}} The studies used for this meta-analysis were
+identified through the review articles written by Keku, et al. (40) and
+Vogtmann, et al. (41). Additional studies, not mentioned in those
+reviews were obtained based on the authors' knowledge of the literature.
+Studies were included that used tissue or feces as their sample source
+for 454 or Illumina 16S rRNA gene sequencing. Some studies (N = 12) were
+excluded because they did not have publicly available sequences or did
+not have metadata in which the authors were able to share. We were able
+to obtain sequence data and metadata from the following studies: Ahn,
 \emph{et al.} (11), Baxter, \emph{et al.} (12), Brim, \emph{et al.}
@@
 Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
-(43) study was not included because the 90 individuals analyzed within
-the study are contained within the larger Baxter study (12). After
-sequence processing, all the case samples for the Kostic study had 100
-or less sequences remaining and was excluded, leaving a total of 14
-studies that analysis could be completed on.
-
-\textbf{\emph{Data Set Breakdown:}} In total, there were seven studies
-with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and
-Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu,
-Sanapareddy), and two studies with both fecal and tissue samples (Chen
-and Flemer). The total number of individuals analyzed after sequence
-processing for feces was 1737 {[}Table 1{]}. The total number of matched
-and unmatched tissue samples that were analyzed after sequence
-processing was 492 {[}Table 2{]}.
-
-\textbf{\emph{Sequence Processing:}} For the majority of studies, raw
-sequences were downloaded from the Sequence Read Archive (SRA)
-(\url{ftp://ftp-trace.ncbi.nih.gov/sra/sra-instant/reads/ByStudy/sra/SRP/})
-and metadata were obtained by searching the respective accession number
-of the study at the following website:
-\url{http://www.ncbi.nlm.nih.gov/Traces/study/}. Of the studies that did
-not have sequences and metadata on the SRA, data was obtained from DBGap
-(n = 1, (11)) and directly from the authors (n = 4, (17, 23, 25, 27)).
-Each study was processed using the mothur (v1.39.3) software program
-(44) and quality filtering utilized the default methods for both 454 and
-Illumina based sequencing. If it was not possible to use the defaults,
-the stated quality cut-offs, from the study itself, were used instead.
-Sequences that were made up of an artificial combination of two or more
-different sequences and commonly known as chimeras were identified and
-removed using VSEARCH (45) before \emph{de novo} OTU clustering at 97\%
-similarity was completed using the OptiClust algorithm (46).
-
-\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
-Shannon diversity were first assessed for differences between controls,
-adenoma tumors, and carcinoma turmors using both linear mixed-effect
-models and ORs. For each individual study the Bray-Curtis index was used
-to assess differences between control-adenoma and control-carcinoma
-individuals. Next, all common genera were assessed for differences in
-ORs for individuals having an adenoma or carcinoma and corrected for
-multiple comparisons using the Benjamini-Hochberg method (47). We then
-built Random Forest models based on all genera, all OTUs, or significant
-OR taxa (only using taxa still significant after multiple comparison
-correction). For both the full genera and significant OR taxa, models
-were trained on one study then tested on the remaining studies using
-genera-based relative abundances. The OTU-based models were built using
-OTU level data and a 10-fold CV over 100 different iterations, based on
-random 80/20 splitting of the data, was used to generate a range of
-expected AUCs. This process was repeated for every study in the
-meta-analysis. Comparisons of the initial trained model AUCs for the
-full genera and significant OR taxa were made to the mean AUC generated
-from the 100 different 10-fold CV runs of the respective OTU-based
-model. For comparisons in which only control versus adenoma individuals
-were made, the carcinoma individuals were excluded from each respective
-study. Similarly, for comparisons in which control versus carcinoma
-individuals were made the adenoma individuals were excluded from each
-respective study. For all analysis completed fecal and tissue samples
-were kept separate. Within the tissue groups the data were further
-divided between samples from the same individual (matched) and those
-from different individuals (unmatched).
-
-\textbf{\emph{Obtaining Genera Relative Abundance and Significant OR
-Taxa Models:}} For the genera analysis of the ORs, OTUs were added
-together based on the genus or lowest available taxonomic classification
-level and the total average counts, for 100 different subsamplings was
-obtained. The significant OR taxa models for the Random Forest models
-utilized all taxa that had significant ORs after multiple comparison
-correction. This meant only models for the carcinoma stool (8 variables)
-and carcinoma unmatched (3 variables) samples were possible to be
-created and analyzed.
-
-\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
-tissue samples with control and tumor samples from different individuals
-were classified as unmatched while samples that belonged to the same
-individual were classified as matched. Studies with matched data
-included Burns, Dejea, Geng, and Lu while those with unmatched data were
-from Burns, Flemer, Chen, and Sanapareddy. For some studies samples
-became unmatched when a corresponding matched sample did not make it
-through sequence processing. All samples, from both matched and
-unmatched tissue samples, were analyzed together for the linear
-mixed-effect models with samples from the same individual being
-corrected for. All other analysis, where it is not specified explicitly,
-matched and unmatched samples were analyzed separately using the
-statistical approaches mentioned in the Statistical Analysis section.
-
-\textbf{\emph{Assessing Important Random Forest Model Variables:}} Using
-Mean Decrease in Accuracy (MDA) the top 10 most important variables to
-the Random Forest model were obtained for the full models of the two
-different approaches used. For the first approach utilizing genus-based
-models, the number of times that a specific taxa showed up in the top 10
-of the training set across each study was counted. For the second
-approach, that utilized the OTU-based models, the medians for each OTU
-across 100 different 80/20 splits of the data was generated and the top
-10 OTUs then counted for each study. Common taxa were then identified by
-using the lowest classification for each of the specific OTUs obtained
-from these counts and the number of times this classification occurred
-across the top 10 of each study was recorded. Finally, the two studies
-that had adenoma tumor tissue (Lu and Flemer) were equally divided
-between matched and unmatched studies and were grouped together for the
-counting of the top 10 genera and OTUs for both Random Forest
-approaches.
+(43) study was excluded because their 90 individuals were contained
+within the larger Baxter study (12). The Kostic study was excluded
+because after we processed the sequences, all of the case samples had
+100 or fewer sequences. The final analysis included 14 studies (Tables 1
+and 2). There were seven studies with only fecal samples (Ahn, Baxter,
+Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue
+samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both
+fecal and tissue samples (Chen and Flemer). After curating the
+sequences, 1737 stool samples and 492 tissue samples remained in the
+analysis {[}Tables 1 and 2{]}.
+
+\textbf{\emph{Sequence Processing:}} Raw sequence data and metadata were
+primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other
+sequence and metadata were obtained directly from the authors (n = 4,
+(17, 23, 25, 27)). Each dataset was processed separately using mothur
+(v1.39.3) (44) using the default quality filtering methods for both 454
+and Illumina sequence data. If it was not possible to use the defaults
+because the sequences were trimmed too much, then the stated quality
+cut-offs from the original study were used. Chimeric sequences were
+identified and removed using VSEARCH (45). The curated sequences were
+assigned to OTUs at 97\% similarity using the OptiClust algorithm (46)
+and classified to the deepest taxonomic level that had 80\% support
+using the naïve Bayesian classifier trained on the RDP taxonomy outline
+(version 14, (47)).
+
+\textbf{\emph{Community analysis:}} We calculated alpha diversity
+metrics (i.e.~OTU richness, evenness, and Shannon diversity) for each
+sample. Within each dataset, we ensured that the data followed a normal
+distribution using power transformations. Using the transformed data, we
+tested the hypothesis that individuals with normal colons, adenomas, and
+carcinomas had significantly different alpha diversity metrics using
+linear mixed-effect models. We also calculated the OR for each study and
+metric by considering any value above the median alpha diversity value
+to be positive. We measured the dissimilarity between individuals by
+calculating the pairwise Bray-Curtis index and used PERMANOVA (48) to
+test whether individuals with normal colons were significantly different
+from those with adenomas or carcinomas. Finally, after binning sequences
+into the deepest taxa that the naïve Bayesian classifier could calssify
+the sequences, we quantified the ORs for individuals having an adenoma
+or carcinoma and corrected for multiple comparisons using the
+Benjamini-Hochberg method (49). Again, for each taxon, if the relative
+abundance was greater than the median relative abundance for that taxon
+in the study, the individual was considered to be positive.
+
+\textbf{\emph{Random Forest classification analysis:}} To classify
+individuals as having normal colons or tumors, we built Random Forest
+classification models for each dataset and comparison using taxa with
+significant ORs (after multiple comparison correction), all taxa, or
+OTUs. Because no taxa were identified as having a significant OR
+associated with adenomas using stool samples or tissue samples,
+classification models based on OR data were not constructed to classify
+individuals as having normal colons or adenomas. Within the training
+dataset, 10-fold cross validation (5-fold cross validation for small
+datasets) was used to build a model using the default mtry setting that
+was then evaluated on the testing set. For the models constructed using
+the taxa with significant ORs, the default mtry setting was used to
+train the model and this model was tested on the other datasets in the
+meta-analysis. The reported AUC values are the average AUCs for the
+application of the model on the test sets. For the OTU-based models,
+hyperparameters were fit splitting the dataset into training (80\% of
+samples) and testing (20\%) sets. Within the training dataset, 10-fold
+cross validation was used to identify the best hyperparameters, which
+were then used to build a model that was then evaluated on the testing
+set. The original 80/20 split and fitting was repeated 100 times. The
+Mean Decrease in Accuracy (MDA), a measure of the importance of each
+taxon to the overall model was used to rank the taxa used in each model.
 
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package (48). For
+sequence processing utilized the R (v3.4.3) software package (50). For
 OTU richness, evenness, and Shannon diversity analysis, values were
-power transformed using the rcompanion (v1.11.1) package (49) and then
-Z-score normalized using the car (v2.1.6) package (50). Testing for OTU
+power transformed using the rcompanion (v1.11.1) package (51) and then
+Z-score normalized using the car (v2.1.6) package (52). Testing for OTU
 richness, evenness, and Shannon diversity differences utilized linear
-mixed-effect models created using the lme4 (v1.1.15) package (51) to
+mixed-effect models created using the lme4 (v1.1.15) package (53) to
 correct for study, repeat sampling of individuals (tissue only), and 16S
 hyper-variable region used. Odds ratios (OR) were analyzed using both
-the epiR (v0.9.93) and metafor (v2.0.0) packages (52, 53) by assessing
+the epiR (v0.9.93) and metafor (v2.0.0) packages (54, 55) by assessing
 how many individuals with and without disease were above and below the
@@
 Bray-Curtis index utilized the creation of distance matrix and testing
-with PERMANOVA executed with the vegan (v2.4.5) package (54). Random
+with PERMANOVA executed with the vegan (v2.4.5) package (56). Random
 Forest models were built using both the caret (v6.0.78) and randomForest
-(v4.6.12) packages (55, 56). All figures were created using both ggplot2
-(v2.2.1) and gridExtra (v2.3) packages (57, 58).
+(v4.6.12) packages (57, 58). All figures were created using both ggplot2
+(v2.2.1) and gridExtra (v2.3) packages (59, 60).
 
@@
 
+\hypertarget{ref-rdp_Wang2007}{}
+47. \textbf{Wang, Q.}, \textbf{G. M. Garrity}, \textbf{J. M. Tiedje},
+and \textbf{J. R. Cole}. 2007. Naive bayesian classifier for rapid
+assignment of rRNA sequences into the new bacterial taxonomy. Applied
+and Environmental Microbiology. American Society for Microbiology
+\textbf{73}:5261--5267.
+
+\hypertarget{ref-permanova_Anderson2013}{}
+48. \textbf{Anderson, M. J.}, and \textbf{D. C. I. Walsh}. 2013.
+PERMANOVA, ANOSIM, and the mantel test in the face of heterogeneous
+dispersions: What null hypothesis are you testing? Ecological
+Monographs. Wiley-Blackwell \textbf{83}:557--574.
+
 \hypertarget{ref-benjamini_controlling_1995}{}
-47. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
+49. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
 the false discovery rate: A practical and powerful approach to multiple
@@
 \hypertarget{ref-r_citation_2017}{}
-48. \textbf{R Core Team}. 2017. R: A language and environment for
+50. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-49. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
+51. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
 extension education program evaluation.
@@
 \hypertarget{ref-car_citation_2011}{}
-50. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
+52. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
 applied regressionSecond. Sage, Thousand Oaks CA.
@@
 \hypertarget{ref-lme4_citation_2015}{}
-51. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
+53. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
 \textbf{S. Walker}. 2015. Fitting linear mixed-effects models using
@@
 \hypertarget{ref-epir_citation_2017}{}
-52. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
+54. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
 Heuer}, \textbf{J. Marshall}, \textbf{J. Sanchez}, \textbf{R. Thornton},
@@
 \hypertarget{ref-metafor_citation_2010}{}
-53. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
+55. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
 the metafor package. Journal of Statistical Software \textbf{36}:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-54. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
+56. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
 \textbf{R. Kindt}, \textbf{P. Legendre}, \textbf{D. McGlinn}, \textbf{P.
@@
 \hypertarget{ref-caret_citation_2017}{}
-55. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
+57. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
 Williams}, \textbf{C. Keefer}, \textbf{A. Engelhardt}, \textbf{T.
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-56. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
+58. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-57. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
+59. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
 analysis. Springer-Verlag New York.
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-58. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
+60. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
 ``grid'' graphics.

### 2018-03-08 submission/manuscript.Rmd
edit RF section of methods

@@
 For the models constructed using the taxa with significant ORs, the default mtry setting was used to train the model and this model was tested on the other datasets in the meta-analysis.
-The reported AUC values are the average AUCs for the application of the model on the test sets.
+The reported AUC values are the AUCs for the application of the model on the test sets.
 For the OTU-based models, hyperparameters were fit splitting the dataset into training (80% of samples) and testing (20%) sets.
-Within the training dataset, 10-fold cross validation was used to identify the best hyperparameters, which were then used to build a model that was then evaluated on the testing set.
-The original 80/20 split and fitting was repeated 100 times.
+Within the OTU dataset, 10-fold cross validation (5-fold cross validation for small datasets) was used to identify the best hyperparameters.
+The original 80/20 split and fitting was repeated 100 times and the average AUC from these 100 repeats was reported.
 The Mean Decrease in Accuracy (MDA), a measure of the importance of each taxon to the overall model was used to rank the taxa used in each model.

### 2018-03-08 submission/manuscript.tex
edit RF section of methods

@@
 train the model and this model was tested on the other datasets in the
-meta-analysis. The reported AUC values are the average AUCs for the
-application of the model on the test sets. For the OTU-based models,
-hyperparameters were fit splitting the dataset into training (80\% of
-samples) and testing (20\%) sets. Within the training dataset, 10-fold
-cross validation was used to identify the best hyperparameters, which
-were then used to build a model that was then evaluated on the testing
-set. The original 80/20 split and fitting was repeated 100 times. The
+meta-analysis. The reported AUC values are the AUCs for the application
+of the model on the test sets. For the OTU-based models, hyperparameters
+were fit splitting the dataset into training (80\% of samples) and
+testing (20\%) sets. Within the OTU dataset, 10-fold cross validation
+(5-fold cross validation for small datasets) was used to identify the
+best hyperparameters. The original 80/20 split and fitting was repeated
+100 times and the average AUC from these 100 repeats was reported. The
 Mean Decrease in Accuracy (MDA), a measure of the importance of each

### 2018-03-09 submission/manuscript.Rmd
update RF section to what was done

@@
 Because no taxa were identified as having a significant OR associated with adenomas using stool samples or tissue samples, classification models based on OR data were not constructed to classify individuals as having normal colons or adenomas.
-Within the training dataset, 10-fold cross validation (5-fold cross validation for small datasets) was used to build a model using the default mtry setting that was then evaluated on the testing set.
+Within the training dataset, 10-fold cross validation (5-fold cross validation for small datasets) was used to build a model that was then evaluated on the testing set.
 For the models constructed using the taxa with significant ORs, the default mtry setting was used to train the model and this model was tested on the other datasets in the meta-analysis.
 The reported AUC values are the AUCs for the application of the model on the test sets.
-For the OTU-based models, hyperparameters were fit splitting the dataset into training (80% of samples) and testing (20%) sets.
-Within the OTU dataset, 10-fold cross validation (5-fold cross validation for small datasets) was used to identify the best hyperparameters.
+For the OTU-based models, the dataset was split into training (80% of samples) and testing (20%) sets and 10-fold cross validation (5-fold cross validation for small datasets) on the training set was used to generate the model for the testing set.
 The original 80/20 split and fitting was repeated 100 times and the average AUC from these 100 repeats was reported.
 The Mean Decrease in Accuracy (MDA), a measure of the importance of each taxon to the overall model was used to rank the taxa used in each model.
+For all models, the default setting used was $\sqrt{p}$, where p is all the variables used in the respective model.
 

### 2018-03-09 submission/manuscript.tex
update RF section to what was done

@@
 dataset, 10-fold cross validation (5-fold cross validation for small
-datasets) was used to build a model using the default mtry setting that
-was then evaluated on the testing set. For the models constructed using
-the taxa with significant ORs, the default mtry setting was used to
-train the model and this model was tested on the other datasets in the
-meta-analysis. The reported AUC values are the AUCs for the application
-of the model on the test sets. For the OTU-based models, hyperparameters
-were fit splitting the dataset into training (80\% of samples) and
-testing (20\%) sets. Within the OTU dataset, 10-fold cross validation
-(5-fold cross validation for small datasets) was used to identify the
-best hyperparameters. The original 80/20 split and fitting was repeated
-100 times and the average AUC from these 100 repeats was reported. The
-Mean Decrease in Accuracy (MDA), a measure of the importance of each
-taxon to the overall model was used to rank the taxa used in each model.
+datasets) was used to build a model that was then evaluated on the
+testing set. For the models constructed using the taxa with significant
+ORs, the default mtry setting was used to train the model and this model
+was tested on the other datasets in the meta-analysis. The reported AUC
+values are the AUCs for the application of the model on the test sets.
+For the OTU-based models, the dataset was split into training (80\% of
+samples) and testing (20\%) sets and 10-fold cross validation (5-fold
+cross validation for small datasets) on the training set was used to
+generate the model for the testing set. The original 80/20 split and
+fitting was repeated 100 times and the average AUC from these 100
+repeats was reported. The Mean Decrease in Accuracy (MDA), a measure of
+the importance of each taxon to the overall model was used to rank the
+taxa used in each model. For all models, the default setting used was
+\(\sqrt{p}\), where p is all the variables used in the respective model.
 

### 2018-03-09 submission/manuscript.Rmd
discussion edits with pat comments in mind

@@
 
-Although we expected that the full OTU models would perform the best at classifying individuals with and without carcinomas, our observations suggest that both the full and significant OR taxa models performed equally well.
-These results suggest that lower level classification to species and strain may not add extra useful information with respect to prediction models.
-This has been suggested in previous literature where metagenomics did not perform better than 16S rRNA gene sequencing data at classifying individuals with normal colons and those with carcinomas [@Hannigan2017].
-One possible reason as to why lower level classification may not result in better models is that the communities are patchy and higher level taxonomic information pools some of this patchiness, allowing for better prediction models.
-There may also be a fair bit of data redundancy within models that utilize more of the community.
-An example of this redundancy would be when we trained models on one study and tested it on the other studies and the AUCs of the models created with the select OR taxa performed as well as full taxa models [Figure 6B].
+We performed a meta-analysis to identify and validate microbiome-base biomarkers that could be used to classify individuals as having normal colons or colonic tumors using fecal or tissue samples.
+To our surprise, Random Forest classification models constructed to differentiate individuals with normal colons from those with carcinomas using a subset of the community performed well relative to models constructed using the full communities.
+When we applied the models trained on each dataset to the other datasets in our study, we found that the models trained using the subset of the communities performed better than those using the full communities.
+These models were trained using data in which sequences were assigned to bacterial taxa using a classifier that typically assigned sequences to the family or genus level.
+When we attempted to improve the specificity of the classification by using an OTU-based approach the resulting models performed as well as those constructed using coarse taxonomic assignments.
+These results are significant because they strengthen the growing literature indicating a role of the microbiome in tumorigenesis [@zackular_gut_2013] and as a potential tool as a non-invasive diagnostic and for assessing risk of disease and recurrence [@baxter_microbiota-based_2016; @Sze2017].
 
-Our observations also suggest that a small collection of taxa can classify disease as well as full OTU-based models but that these taxa individually perform quite poorly [Figure 3].
-This result supports the contention that there might be redundancy of function amongst the taxa included in the significant OR models.
-As an example, multiple different microbes could be similarily stimulating the activation of inflammatory pathways and by doing so exacerbate disease progression.
-Multiple reports within the literature have found that different bacteria, such as *Escherichia coli* and *Fusobacterium nucleatum*, can similarily worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013].
-Although the inflammatory taxa were patchy in their importance and presence across studies those that were not typically associated with inflammation were consistently important for every study [Figure 6].
-The loss of these taxa (*Ruminococcus* and *Clostridium XI* in stool and *Dorea* and *Blautia* in unmatched tissue) is particularly interesting because many are commonly thought to be beneficial due to their involvement in production of short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].
+These results suggested that fine scale classification of sequences into OTUs does not improve our classification models.
+This has been suggested in previous literature where shotgun metagenomic data did not perform better than 16S rRNA gene sequencing data in classifying individuals with normal colons and those with carcinomas [@Hannigan2017].
+We hypothesize that fine scale classification may not result in better classification because distribution of microbiota between individuals is patchy.
+In contrast, models using coarser taxonomic assignments will pool the fine scale diversity and resulting in less patchiness and better classification.
+Furthermore, the ability of models trained using a subset of the community to outperform those using the full community when testing the models on the other datasets may also be a product of the patchiness of human-associated microbiota.
+The models based on the 8 taxa that had significant ORs used taxa that were found in every study and tended to have higher relative abundances.
+Similar to the OTU-based models, those models based on the full community taxonomy assignments were still sensitive to the patchy distribution of taxa.
+Regardless, it is encouraging that a collection of  `r length(crc_OR_stool$measure)` taxa could reliably classify individuals as having carcinomas considering the differences in cohorts, DNA extraction procedures, regions of the 16S rRNA gene, and sequencing methods.
 
-The adenoma models as a whole performed poorly in classifying individuals with and without adenomas.
-This outcome is not inconsistent with what has been published previously [@hale_shifts_2017; @Sze2017].
-However, the modeling results are at odds with results obtained in Baxter, *et al.* [@baxter_microbiota-based_2016].
-There are some major differences between the models generated in this meta-analysis and what was used in Baxter, *et al*.
-First, the prevoius report's models investigated the classification of lesions (individuals with adenoma or carcinoma) and not adenoma alone.
-The Baxter, *et al.* models also contained Fecal Immunoglobulin Test data while our meta-analysis models only contained 16S rRNA gene sequencing data.
-Although being able to classify individuals with adenomas is important, the classification of advanced adenomas is a more clinically meaningful diagnostic tool (i.e.
-those that are at highest risk of progressing to a carcinoma).
-It is possible that we might have been able to detect differences in the bacterial community if advanced adenomas were separated from adenomas but that data was not available for the majority of studies analyzed.
-Additionally, the initial changes to the bacterial community could be focal to where the initial adenoma develops and would not be easily assessed with a fecal sample.
+When used separately to classify individuals with carcinomas, the taxa with significant ORs could not reliably classify individuals [Figure 3].
+This result further supports the hypothesis that carcinoma-associated microbiota have a patchy distribution.
+Two individuals may have had the same classification, based on the relative abundance of different populations within this group of `r length(crc_OR_stool$measure)` taxa.
+Although these results only reflect associations with disease, it is tempting to hypothesize that the patchiness represents distinct mechanisms of exacerbating tumorigenesis or that multiple taxa have the same mechanism of exacerbating tumorigenesis.
+For example, strains of *Escherichia coli* and *Fusobacterium nucleatum* have been shown to worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013].
+In contrast to the patchiness of the taxa that were positively associated with carcinomas, potentially beneficial taxa had a more consistent association [Figure 6].
+This result was particularly interesting because members of these taxa (i.e.
+*Ruminococcus* and *Clostridium XI* in stool and *Dorea* and *Blautia* in tissue) are thought to be beneficial due to their involvement in production of anti-inflammatory short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].
 
-Although stool represents an easy and less invasive way to assess risk, it is not clear how well this sample reflects adenoma- and carcinoma- tissue associated microbial communities.
-The colon tissue-based studies did not provide a clearer understanding of how the microbiota may be associated with tumors.
-Generally, the full OTU-based models of unmatched and matched colon tissue samples were concordant with stool samples showing that GI resident microbes were the most prevalent in the top 10 most important variables across study [Figure S3].
-*Fusobacterium* was not consistently identified across studies and this could be due to both a small number of studies and a small sample size within these studies.
-The majority of the colon tissue-based results were also study specific with many of the top 10 taxa being present only in a single study.
-Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying (e.g.
-*Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.
-[@Salter_contamination_2014]).
-The low bacterial biomass of tissue samples coupled with potential contamination and small sample sizes could explain why these results seem to be more sporadic than the stool results.
+All of the adenoma classification models performed poorly.
+This result is not inconsistent with previous studies [@hale_shifts_2017; @Sze2017].
+However, the classification results are at odds with results of the multitarget microbiota test (MMT) from Baxter, *et al.* [@baxter_microbiota-based_2016] who observed an AUC of 0.755 when applied to individuals with adenomas.
+There are two major differences between the models generated in this meta-analysis and that analysis.
+The MMT attempted to classify individuals as having a normal colon or having colonic lesions (i.e.
+adenomas or carcinomas) and not adenomas alone.
+Further, the MMT incorporated fecal immunoglobulin test (FIT) data while our models only used 16S rRNA gene sequencing data.
+Because FIT data were not available for the other studies in our meta-analysis, it was not possible to validate the MMT approach.
+The ability to differentiate between individuals with and without adenomas is an important problem since early detection of tumors is critical.
+However, it is possible that we might have been able to detect differences in the bacterial community if individuals with non-advanced and advanced adenomas were separated.
+This is a clinically relevant distinction since advanced adenomas are at highest risk of progressing to a carcinoma.
+The initial changes of the microbiota during tumorigenesis could be focal to where the initial adenoma develops and would not be easily assessed using fecal samples from an individual with non-advanced adenomas.
+Unfortunately, distinguishing between individuals with advanced and non-advanced adenomas was not possible in our meta-analysis since the studies did not provide the clinical data needed to make that distinction.
 
-One important caveat to this study is that even though genera associated with certain species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp.
-*gallolyticus* were not identified, it does not necessarily mean that these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017].
-There are reports that *Bacteroides fragilis*, positive for the enterotoxigenic gene, are found at specific locations along the colon but the samples we were able to use in this meta-analysis could not identify these types of differences [@Purcell2017].
-Additionally, since we are limited in our aggregation of the data to the genus level, it is not possible to clearly delineate which species are contributing to overall disease progression.
-Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp.
-*gallolyticus*.
-As an example, the stool-based full community models consistently identified the *Bacteroides* taxa to be important model components across studies.
-This suggests that even though *Bacteroides* may not increase the OR of individuals having an adenoma or carcinoma and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC.
+Stool samples represent a non-invasive approach to assessing the structure of the gut microbiota and are potentially useful for diagnosing individuals as having colonic tumors.
+However, they do not reflect the structure of the mucosal microbiota [@Flynn_preprint_2017].
+Regardless, the taxa that were the most important in the stool-based models overlapped with those from the models trained using the data from unmatched and matched colon tissue samples [Figure S3].
+Mucosal biopsies are preferred for focused mechanistic studies and have offered researchers the opportunity to sample healthy and diseased tissue from the same individuals (i.e.
+matched) using each individual as their own control or in a cross sectional design (i.e.
+unmatched).
+Because obtaining these samples is invasive, carries risks to the individual, and is expensive, studies investigating the structure of the mucosal microbiota generally have a limited number of participants.
+Thus, it was not surprising that tissue-based studies did not provide clearer associations between the mucosal microbiota and the presence of tumors.
+Interestingly, Fusobacterium, which has received increased recent attention for its potential role in tumorigenesis [@kostic_fusobacterium_2013] was not consistently identified across the studies in our meta-analysis.
+This could be due to the relatively small number of individuals in the limited number of studies.
+The classification models trained using the tissue-based data performed well when tested with the training data (Figure S4), but performed poorly when tested on the other tissue-associated datasets (Figure S5).
+Disturbingly, taxa that are commonly associated with reagent contamination (e.g.
+*Novosphingobium*, *Acidobacteria Gp2*, *Sphingomonas*, etc.) were detected within the tissue datasets.
+Such contamination is common in studies where there is relatively low bacterial biomass [@Salter_contamination_2014].
+The lack of replication among the tissue-based biomarkers may be a product of the relatively small number of studies and individuals per study and possible reagent contamination.
 
-Despite these limitations to the findings, meta-analyses are a useful tool in microbiota research because they can both validate existing research and make new discoveries by pooling many independent investigations together.
-Yet, it is still difficult to perform these studies because of inaccessible 16S sequencing data, missing or vague metadata (e.g.
-which samples are carcioma and which are not), varying sequence quality, and multiple small data sets.
-Better attention to these specific problems could help to increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses.
-Moving forward, meta-analyses will be important tools to help aggregate and find commonalities across studies when investigating the microbiota in the context of a specific disease and more are needed [@Sze2016; @Walters2014; @Finucane2014; @shah_leveraging_2017].
+Among our stool samples, we failed to identify several notable populations that are commonly associated with carcinomas including an enterotoxigenic strain of *Bacteroides fragilis* (ETBF) and *Streptococcus gallolyticus* subsp.
+*gallolyticus* [@bfrag_Boleij_2014; @strep_Aymeric_2017].
+ETBF have been found in tumors in the proximal colon where they tend to form biofilms [@Purcell2017; @dejea_microbiota_2014].
+Considering DNA from bacteria that are more prevalent in the proximal colon may be degraded by the time it leaves the body, it is not surprising that we failed to identify a significant OR for Bacteroides with carcinomas.
+In addition, since our approach could only classify sequences to the genus level and there are likely multiple Bacteroides populations in the colon, it is possible that sequences from ETBF and non-oncogenic Bacteroides were pooled.
+This would then reduce the OR between Bacterioides and whether an individual had carcinomas.
+It is also necessary to distinguish between populations that are biomarkers for a disease and those that are known to cause disease.
+The former may also have a causative role in the disease.
+Although the latter have been shown to have a causative role, they may appear at low relative abundance, be found in specific locations, or may have a highly patchy distribution among affected individuals.
 
-By aggregating together a large collection of studies analyzing both fecal and colon tissue samples, we are able to provide evidence supporting the importance of the bacterial community in carcinoma tumors.
-Although further validation of the biomarkers presented here need to be undertaken, the replicability of the AUC of a specific collection of taxa across multiple studies suggests a strong potential for the use of the microbiota as a risk stratification tool for individuals with carcinomas.
+Meta-analyses are a useful tool in microbiome research because they can demonstrate whether a result can be replicated and facilitate new discoveries by pooling multiple independent investigations.
+There have been several meta-analyses similar to this study that have sought biomarkers for obesity [@Sze2016; @Walters2014; @Finucane2014], inflammatory bowel disease [@Walters2014], and colorectal cancer [@shah_leveraging_2017].
+Considering microbiome research is particularly prone to hype and overgeneralization of results, these analyses are critical.
+For example, previous meta-analyses have demonstrated that there are no clear fecal biomarkers for obesity [Sze2016; @Walters2014; @Finucane2014].
+Such meta-analyses are difficult to perform because the underlying 16S rRNA gene sequence data are not publicly available, metadata are missing, incomplete, or vague, sequence data are of poor quality or derived by non-standard approaches, and the original studies were significantly underpowered.
+Reluctance to publish negative results (i.e.
+the “file drawer effect”) is also likely to skew our understanding of the relationship between microbiota and disease.
+Better attention to these specifics will increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses.
+Moving forward, meta-analyses will be important tools to help aggregate and find commonalities across studies when investigating the microbiota in the context of a specific disease [@Sze2016; @Walters2014; @Finucane2014; @shah_leveraging_2017].
+
+Our meta-analysis suggests a strong association between the gut microbiota and colon tumorigenesis.
+By aggregating the results from studies that sequenced the 16S rRNA gene from fecal and tissue samples, we are able to provide evidence supporting the use of microbial biomarkers to diagnose the presence of colonic tumors.
+Further development of microbial biomarkers should focus on including other biomarkers (e.g.
+FIT), better categorizing of people with adenomas, and expanding datasets to include larger numbers of individuals.
+Based on prior research into the physiology of the biomarkers we identified, it is likely that they have a causative role in tumorigenesis.
+Their patchy distribution across individuals suggests that there are either multiple mechanisms causing disease or a single mechanism (e.g.
+inflammation) that can be mediated by multiple, diverse bacteria.
 
@@
 For all models, the default setting used was $\sqrt{p}$, where p is all the variables used in the respective model.
+Normally, $\sqrt{p}$ has been found to be what is chosen as the ideal mtry [@Breiman2001].
 

### 2018-03-09 submission/manuscript.tex
discussion edits with pat comments in mind

@@
 
-Although we expected that the full OTU models would perform the best at
-classifying individuals with and without carcinomas, our observations
-suggest that both the full and significant OR taxa models performed
-equally well. These results suggest that lower level classification to
-species and strain may not add extra useful information with respect to
-prediction models. This has been suggested in previous literature where
-metagenomics did not perform better than 16S rRNA gene sequencing data
-at classifying individuals with normal colons and those with carcinomas
-(30). One possible reason as to why lower level classification may not
-result in better models is that the communities are patchy and higher
-level taxonomic information pools some of this patchiness, allowing for
-better prediction models. There may also be a fair bit of data
-redundancy within models that utilize more of the community. An example
-of this redundancy would be when we trained models on one study and
-tested it on the other studies and the AUCs of the models created with
-the select OR taxa performed as well as full taxa models {[}Figure
-6B{]}.
-
-Our observations also suggest that a small collection of taxa can
-classify disease as well as full OTU-based models but that these taxa
-individually perform quite poorly {[}Figure 3{]}. This result supports
-the contention that there might be redundancy of function amongst the
-taxa included in the significant OR models. As an example, multiple
-different microbes could be similarily stimulating the activation of
-inflammatory pathways and by doing so exacerbate disease progression.
-Multiple reports within the literature have found that different
-bacteria, such as \emph{Escherichia coli} and \emph{Fusobacterium
-nucleatum}, can similarily worsen inflammation in mouse models of
-tumorigenesis (5, 6, 21). Although the inflammatory taxa were patchy in
-their importance and presence across studies those that were not
-typically associated with inflammation were consistently important for
-every study {[}Figure 6{]}. The loss of these taxa (\emph{Ruminococcus}
-and \emph{Clostridium XI} in stool and \emph{Dorea} and \emph{Blautia}
-in unmatched tissue) is particularly interesting because many are
-commonly thought to be beneficial due to their involvement in production
-of short chain fatty acids (31--33).
-
-The adenoma models as a whole performed poorly in classifying
-individuals with and without adenomas. This outcome is not inconsistent
-with what has been published previously (27, 34). However, the modeling
-results are at odds with results obtained in Baxter, \emph{et al.} (12).
-There are some major differences between the models generated in this
-meta-analysis and what was used in Baxter, \emph{et al}. First, the
-prevoius report's models investigated the classification of lesions
-(individuals with adenoma or carcinoma) and not adenoma alone. The
-Baxter, \emph{et al.} models also contained Fecal Immunoglobulin Test
-data while our meta-analysis models only contained 16S rRNA gene
-sequencing data. Although being able to classify individuals with
-adenomas is important, the classification of advanced adenomas is a more
-clinically meaningful diagnostic tool (i.e.~those that are at highest
-risk of progressing to a carcinoma). It is possible that we might have
-been able to detect differences in the bacterial community if advanced
-adenomas were separated from adenomas but that data was not available
-for the majority of studies analyzed. Additionally, the initial changes
-to the bacterial community could be focal to where the initial adenoma
-develops and would not be easily assessed with a fecal sample.
-
-Although stool represents an easy and less invasive way to assess risk,
-it is not clear how well this sample reflects adenoma- and carcinoma-
-tissue associated microbial communities. The colon tissue-based studies
-did not provide a clearer understanding of how the microbiota may be
-associated with tumors. Generally, the full OTU-based models of
-unmatched and matched colon tissue samples were concordant with stool
-samples showing that GI resident microbes were the most prevalent in the
-top 10 most important variables across study {[}Figure S3{]}.
-\emph{Fusobacterium} was not consistently identified across studies and
-this could be due to both a small number of studies and a small sample
-size within these studies. The majority of the colon tissue-based
-results were also study specific with many of the top 10 taxa being
-present only in a single study. Additionally, the presence of genera
-associated with contamination, within the top 10 most important
-variables for the genera and OTU models is worrying (e.g.
-\emph{Novosphingobium}, \emph{Acidobacteria Gp2}, \emph{Sphingomonas},
-etc. (35)). The low bacterial biomass of tissue samples coupled with
-potential contamination and small sample sizes could explain why these
-results seem to be more sporadic than the stool results.
-
-One important caveat to this study is that even though genera associated
-with certain species such as \emph{Bacteroides fragilis} and
-\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} were not
-identified, it does not necessarily mean that these specific species are
-not important in human CRC (22, 24). There are reports that
-\emph{Bacteroides fragilis}, positive for the enterotoxigenic gene, are
-found at specific locations along the colon but the samples we were able
-to use in this meta-analysis could not identify these types of
-differences (36). Additionally, since we are limited in our aggregation
-of the data to the genus level, it is not possible to clearly delineate
-which species are contributing to overall disease progression. Our
-observations are not inconsistent with the previous literature on either
-\emph{Bacteroides fragilis} or \emph{Streptococcus gallolyticus} subsp.
-\emph{gallolyticus}. As an example, the stool-based full community
-models consistently identified the \emph{Bacteroides} taxa to be
-important model components across studies. This suggests that even
-though \emph{Bacteroides} may not increase the OR of individuals having
-an adenoma or carcinoma and may not vary in relative abundance, like
-\emph{Fusobacterium}, it is still important in CRC.
-
-Despite these limitations to the findings, meta-analyses are a useful
-tool in microbiota research because they can both validate existing
-research and make new discoveries by pooling many independent
-investigations together. Yet, it is still difficult to perform these
-studies because of inaccessible 16S sequencing data, missing or vague
-metadata (e.g.~which samples are carcioma and which are not), varying
-sequence quality, and multiple small data sets. Better attention to
-these specific problems could help to increase the reproducibility and
-replicability of microbiota studies and make it easier to perform these
-crucial meta-analyses. Moving forward, meta-analyses will be important
-tools to help aggregate and find commonalities across studies when
-investigating the microbiota in the context of a specific disease and
-more are needed (28, 37--39).
-
-By aggregating together a large collection of studies analyzing both
-fecal and colon tissue samples, we are able to provide evidence
-supporting the importance of the bacterial community in carcinoma
-tumors. Although further validation of the biomarkers presented here
-need to be undertaken, the replicability of the AUC of a specific
-collection of taxa across multiple studies suggests a strong potential
-for the use of the microbiota as a risk stratification tool for
-individuals with carcinomas.
+We performed a meta-analysis to identify and validate microbiome-base
+biomarkers that could be used to classify individuals as having normal
+colons or colonic tumors using fecal or tissue samples. To our surprise,
+Random Forest classification models constructed to differentiate
+individuals with normal colons from those with carcinomas using a subset
+of the community performed well relative to models constructed using the
+full communities. When we applied the models trained on each dataset to
+the other datasets in our study, we found that the models trained using
+the subset of the communities performed better than those using the full
+communities. These models were trained using data in which sequences
+were assigned to bacterial taxa using a classifier that typically
+assigned sequences to the family or genus level. When we attempted to
+improve the specificity of the classification by using an OTU-based
+approach the resulting models performed as well as those constructed
+using coarse taxonomic assignments. These results are significant
+because they strengthen the growing literature indicating a role of the
+microbiome in tumorigenesis (9) and as a potential tool as a
+non-invasive diagnostic and for assessing risk of disease and recurrence
+(12, 30).
+
+These results suggested that fine scale classification of sequences into
+OTUs does not improve our classification models. This has been suggested
+in previous literature where shotgun metagenomic data did not perform
+better than 16S rRNA gene sequencing data in classifying individuals
+with normal colons and those with carcinomas (31). We hypothesize that
+fine scale classification may not result in better classification
+because distribution of microbiota between individuals is patchy. In
+contrast, models using coarser taxonomic assignments will pool the fine
+scale diversity and resulting in less patchiness and better
+classification. Furthermore, the ability of models trained using a
+subset of the community to outperform those using the full community
+when testing the models on the other datasets may also be a product of
+the patchiness of human-associated microbiota. The models based on the 8
+taxa that had significant ORs used taxa that were found in every study
+and tended to have higher relative abundances. Similar to the OTU-based
+models, those models based on the full community taxonomy assignments
+were still sensitive to the patchy distribution of taxa. Regardless, it
+is encouraging that a collection of 8 taxa could reliably classify
+individuals as having carcinomas considering the differences in cohorts,
+DNA extraction procedures, regions of the 16S rRNA gene, and sequencing
+methods.
+
+When used separately to classify individuals with carcinomas, the taxa
+with significant ORs could not reliably classify individuals {[}Figure
+3{]}. This result further supports the hypothesis that
+carcinoma-associated microbiota have a patchy distribution. Two
+individuals may have had the same classification, based on the relative
+abundance of different populations within this group of 8 taxa. Although
+these results only reflect associations with disease, it is tempting to
+hypothesize that the patchiness represents distinct mechanisms of
+exacerbating tumorigenesis or that multiple taxa have the same mechanism
+of exacerbating tumorigenesis. For example, strains of \emph{Escherichia
+coli} and \emph{Fusobacterium nucleatum} have been shown to worsen
+inflammation in mouse models of tumorigenesis (5, 6, 21). In contrast to
+the patchiness of the taxa that were positively associated with
+carcinomas, potentially beneficial taxa had a more consistent
+association {[}Figure 6{]}. This result was particularly interesting
+because members of these taxa (i.e. \emph{Ruminococcus} and
+\emph{Clostridium XI} in stool and \emph{Dorea} and \emph{Blautia} in
+tissue) are thought to be beneficial due to their involvement in
+production of anti-inflammatory short chain fatty acids (32--34).
+
+All of the adenoma classification models performed poorly. This result
+is not inconsistent with previous studies (27, 30). However, the
+classification results are at odds with results of the multitarget
+microbiota test (MMT) from Baxter, \emph{et al.} (12) who observed an
+AUC of 0.755 when applied to individuals with adenomas. There are two
+major differences between the models generated in this meta-analysis and
+that analysis. The MMT attempted to classify individuals as having a
+normal colon or having colonic lesions (i.e.~adenomas or carcinomas) and
+not adenomas alone. Further, the MMT incorporated fecal immunoglobulin
+test (FIT) data while our models only used 16S rRNA gene sequencing
+data. Because FIT data were not available for the other studies in our
+meta-analysis, it was not possible to validate the MMT approach. The
+ability to differentiate between individuals with and without adenomas
+is an important problem since early detection of tumors is critical.
+However, it is possible that we might have been able to detect
+differences in the bacterial community if individuals with non-advanced
+and advanced adenomas were separated. This is a clinically relevant
+distinction since advanced adenomas are at highest risk of progressing
+to a carcinoma. The initial changes of the microbiota during
+tumorigenesis could be focal to where the initial adenoma develops and
+would not be easily assessed using fecal samples from an individual with
+non-advanced adenomas. Unfortunately, distinguishing between individuals
+with advanced and non-advanced adenomas was not possible in our
+meta-analysis since the studies did not provide the clinical data needed
+to make that distinction.
+
+Stool samples represent a non-invasive approach to assessing the
+structure of the gut microbiota and are potentially useful for
+diagnosing individuals as having colonic tumors. However, they do not
+reflect the structure of the mucosal microbiota (35). Regardless, the
+taxa that were the most important in the stool-based models overlapped
+with those from the models trained using the data from unmatched and
+matched colon tissue samples {[}Figure S3{]}. Mucosal biopsies are
+preferred for focused mechanistic studies and have offered researchers
+the opportunity to sample healthy and diseased tissue from the same
+individuals (i.e.~matched) using each individual as their own control or
+in a cross sectional design (i.e.~unmatched). Because obtaining these
+samples is invasive, carries risks to the individual, and is expensive,
+studies investigating the structure of the mucosal microbiota generally
+have a limited number of participants. Thus, it was not surprising that
+tissue-based studies did not provide clearer associations between the
+mucosal microbiota and the presence of tumors. Interestingly,
+Fusobacterium, which has received increased recent attention for its
+potential role in tumorigenesis (6) was not consistently identified
+across the studies in our meta-analysis. This could be due to the
+relatively small number of individuals in the limited number of studies.
+The classification models trained using the tissue-based data performed
+well when tested with the training data (Figure S4), but performed
+poorly when tested on the other tissue-associated datasets (Figure S5).
+Disturbingly, taxa that are commonly associated with reagent
+contamination (e.g. \emph{Novosphingobium}, \emph{Acidobacteria Gp2},
+\emph{Sphingomonas}, etc.) were detected within the tissue datasets.
+Such contamination is common in studies where there is relatively low
+bacterial biomass (36). The lack of replication among the tissue-based
+biomarkers may be a product of the relatively small number of studies
+and individuals per study and possible reagent contamination.
+
+Among our stool samples, we failed to identify several notable
+populations that are commonly associated with carcinomas including an
+enterotoxigenic strain of \emph{Bacteroides fragilis} (ETBF) and
+\emph{Streptococcus gallolyticus} subsp. \emph{gallolyticus} (22, 24).
+ETBF have been found in tumors in the proximal colon where they tend to
+form biofilms (20, 37). Considering DNA from bacteria that are more
+prevalent in the proximal colon may be degraded by the time it leaves
+the body, it is not surprising that we failed to identify a significant
+OR for Bacteroides with carcinomas. In addition, since our approach
+could only classify sequences to the genus level and there are likely
+multiple Bacteroides populations in the colon, it is possible that
+sequences from ETBF and non-oncogenic Bacteroides were pooled. This
+would then reduce the OR between Bacterioides and whether an individual
+had carcinomas. It is also necessary to distinguish between populations
+that are biomarkers for a disease and those that are known to cause
+disease. The former may also have a causative role in the disease.
+Although the latter have been shown to have a causative role, they may
+appear at low relative abundance, be found in specific locations, or may
+have a highly patchy distribution among affected individuals.
+
+Meta-analyses are a useful tool in microbiome research because they can
+demonstrate whether a result can be replicated and facilitate new
+discoveries by pooling multiple independent investigations. There have
+been several meta-analyses similar to this study that have sought
+biomarkers for obesity (38--40), inflammatory bowel disease (39), and
+colorectal cancer (28). Considering microbiome research is particularly
+prone to hype and overgeneralization of results, these analyses are
+critical. For example, previous meta-analyses have demonstrated that
+there are no clear fecal biomarkers for obesity (39, 40). Such
+meta-analyses are difficult to perform because the underlying 16S rRNA
+gene sequence data are not publicly available, metadata are missing,
+incomplete, or vague, sequence data are of poor quality or derived by
+non-standard approaches, and the original studies were significantly
+underpowered. Reluctance to publish negative results (i.e.~the ``file
+drawer effect'') is also likely to skew our understanding of the
+relationship between microbiota and disease. Better attention to these
+specifics will increase the reproducibility and replicability of
+microbiota studies and make it easier to perform these crucial
+meta-analyses. Moving forward, meta-analyses will be important tools to
+help aggregate and find commonalities across studies when investigating
+the microbiota in the context of a specific disease (28, 38--40).
+
+Our meta-analysis suggests a strong association between the gut
+microbiota and colon tumorigenesis. By aggregating the results from
+studies that sequenced the 16S rRNA gene from fecal and tissue samples,
+we are able to provide evidence supporting the use of microbial
+biomarkers to diagnose the presence of colonic tumors. Further
+development of microbial biomarkers should focus on including other
+biomarkers (e.g.~FIT), better categorizing of people with adenomas, and
+expanding datasets to include larger numbers of individuals. Based on
+prior research into the physiology of the biomarkers we identified, it
+is likely that they have a causative role in tumorigenesis. Their patchy
+distribution across individuals suggests that there are either multiple
+mechanisms causing disease or a single mechanism (e.g.~inflammation)
+that can be mediated by multiple, diverse bacteria.
 
@@
 \textbf{\emph{Datasets:}} The studies used for this meta-analysis were
-identified through the review articles written by Keku, et al. (40) and
-Vogtmann, et al. (41). Additional studies, not mentioned in those
+identified through the review articles written by Keku, et al. (41) and
+Vogtmann, et al. (42). Additional studies, not mentioned in those
 reviews were obtained based on the authors' knowledge of the literature.
@@
 \emph{et al.} (20), Flemer, \emph{et al.} (17), Geng, \emph{et al.}
-(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (42), Lu, \emph{et
+(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (43), Lu, \emph{et
 al.} (26), Sanapareddy, \emph{et al.} (25), Wang, \emph{et al.} (14),
 Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
-(43) study was excluded because their 90 individuals were contained
+(44) study was excluded because their 90 individuals were contained
 within the larger Baxter study (12). The Kostic study was excluded
@@
 (17, 23, 25, 27)). Each dataset was processed separately using mothur
-(v1.39.3) (44) using the default quality filtering methods for both 454
+(v1.39.3) (45) using the default quality filtering methods for both 454
 and Illumina sequence data. If it was not possible to use the defaults
@@
 cut-offs from the original study were used. Chimeric sequences were
-identified and removed using VSEARCH (45). The curated sequences were
-assigned to OTUs at 97\% similarity using the OptiClust algorithm (46)
+identified and removed using VSEARCH (46). The curated sequences were
+assigned to OTUs at 97\% similarity using the OptiClust algorithm (47)
 and classified to the deepest taxonomic level that had 80\% support
 using the naïve Bayesian classifier trained on the RDP taxonomy outline
-(version 14, (47)).
+(version 14, (48)).
 
@@
 to be positive. We measured the dissimilarity between individuals by
-calculating the pairwise Bray-Curtis index and used PERMANOVA (48) to
+calculating the pairwise Bray-Curtis index and used PERMANOVA (49) to
 test whether individuals with normal colons were significantly different
@@
 or carcinoma and corrected for multiple comparisons using the
-Benjamini-Hochberg method (49). Again, for each taxon, if the relative
+Benjamini-Hochberg method (50). Again, for each taxon, if the relative
 abundance was greater than the median relative abundance for that taxon
@@
 \(\sqrt{p}\), where p is all the variables used in the respective model.
+Normally, \(\sqrt{p}\) has been found to be what is chosen as the ideal
+mtry (51).
 
 \textbf{\emph{Statistical Analysis:}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package (50). For
+sequence processing utilized the R (v3.4.3) software package (52). For
 OTU richness, evenness, and Shannon diversity analysis, values were
-power transformed using the rcompanion (v1.11.1) package (51) and then
-Z-score normalized using the car (v2.1.6) package (52). Testing for OTU
+power transformed using the rcompanion (v1.11.1) package (53) and then
+Z-score normalized using the car (v2.1.6) package (54). Testing for OTU
 richness, evenness, and Shannon diversity differences utilized linear
-mixed-effect models created using the lme4 (v1.1.15) package (53) to
+mixed-effect models created using the lme4 (v1.1.15) package (55) to
 correct for study, repeat sampling of individuals (tissue only), and 16S
 hyper-variable region used. Odds ratios (OR) were analyzed using both
-the epiR (v0.9.93) and metafor (v2.0.0) packages (54, 55) by assessing
+the epiR (v0.9.93) and metafor (v2.0.0) packages (56, 57) by assessing
 how many individuals with and without disease were above and below the
@@
 Bray-Curtis index utilized the creation of distance matrix and testing
-with PERMANOVA executed with the vegan (v2.4.5) package (56). Random
+with PERMANOVA executed with the vegan (v2.4.5) package (58). Random
 Forest models were built using both the caret (v6.0.78) and randomForest
-(v4.6.12) packages (57, 58). All figures were created using both ggplot2
-(v2.2.1) and gridExtra (v2.3) packages (59, 60).
+(v4.6.12) packages (59, 60). All figures were created using both ggplot2
+(v2.2.1) and gridExtra (v2.3) packages (61, 62).
 
@@
 
+\hypertarget{ref-Sze2017}{}
+30. \textbf{Sze, M. A.}, \textbf{N. T. Baxter}, \textbf{M. T. Ruffin},
+\textbf{M. A. M. Rogers}, and \textbf{P. D. Schloss}. 2017.
+Normalization of the microbiota in patients after treatment for colonic
+lesions. Microbiome. Springer Nature \textbf{5}.
+
 \hypertarget{ref-Hannigan2017}{}
-30. \textbf{Hannigan, G. D.}, \textbf{M. B. Duhaime}, \textbf{M. T.
+31. \textbf{Hannigan, G. D.}, \textbf{M. B. Duhaime}, \textbf{M. T.
 Ruffin}, \textbf{C. C. Koumpouras}, and \textbf{P. D. Schloss}. 2017.
@@
 \hypertarget{ref-Venkataraman2016}{}
-31. \textbf{Venkataraman, A.}, \textbf{J. R. Sieber}, \textbf{A. W.
+32. \textbf{Venkataraman, A.}, \textbf{J. R. Sieber}, \textbf{A. W.
 Schmidt}, \textbf{C. Waldron}, \textbf{K. R. Theis}, and \textbf{T. M.
@@
 \hypertarget{ref-Herrmann2018}{}
-32. \textbf{Herrmann, E.}, \textbf{W. Young}, \textbf{V.
+33. \textbf{Herrmann, E.}, \textbf{W. Young}, \textbf{V.
 Reichert-Grimm}, \textbf{S. Weis}, \textbf{C. Riedel}, \textbf{D.
@@
 \hypertarget{ref-Reichardt2017}{}
-33. \textbf{Reichardt, N.}, \textbf{M. Vollmer}, \textbf{G. Holtrop},
+34. \textbf{Reichardt, N.}, \textbf{M. Vollmer}, \textbf{G. Holtrop},
 \textbf{F. M. Farquharson}, \textbf{D. Wefers}, \textbf{M. Bunzel},
@@
 
-\hypertarget{ref-Sze2017}{}
-34. \textbf{Sze, M. A.}, \textbf{N. T. Baxter}, \textbf{M. T. Ruffin},
-\textbf{M. A. M. Rogers}, and \textbf{P. D. Schloss}. 2017.
-Normalization of the microbiota in patients after treatment for colonic
-lesions. Microbiome. Springer Nature \textbf{5}.
+\hypertarget{ref-Flynn_preprint_2017}{}
+35. \textbf{Flynn, K. J.}, \textbf{M. T. Ruffin}, \textbf{D. K.
+Turgeon}, and \textbf{P. D. Schloss}. 2017. Spatial variation of the
+native colon microbiota in healthy adults. Cold Spring Harbor
+Laboratory.
 
 \hypertarget{ref-Salter_contamination_2014}{}
-35. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
+36. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
 \textbf{S. T. Calus}, \textbf{W. O. Cookson}, \textbf{M. F. Moffatt},
@@
 \hypertarget{ref-Purcell2017}{}
-36. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
+37. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
 \textbf{L. Dixon}, \textbf{F. A. Frizelle}, and \textbf{J. I. Keenan}.
@@
 \hypertarget{ref-Sze2016}{}
-37. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
+38. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
 signal in the noise: Revisiting obesity and the microbiome. mBio.
@@
 \hypertarget{ref-Walters2014}{}
-38. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
+39. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
 2014. Meta-analyses of human gut microbes associated with obesity and
@@
 \hypertarget{ref-Finucane2014}{}
-39. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
+40. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
 Laurent}, and \textbf{K. S. Pollard}. 2014. A taxonomic signature of
@@
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-40. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
+41. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
 \textbf{B. Jovov}, and \textbf{X. Han}. 2015. The gastrointestinal
@@
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-41. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
+42. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
 Epidemiologic studies of the human microbiome and cancer. British
@@
 \hypertarget{ref-kostic_genomic_2012}{}
-42. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
+43. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
 Pedamallu}, \textbf{M. Michaud}, \textbf{F. Duke}, \textbf{A. M. Earl},
@@
 \hypertarget{ref-zackular_human_2014}{}
-43. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
+44. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
 Ruffin}, and \textbf{P. D. Schloss}. 2014. The human gut microbiome as a
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-44. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
+45. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
 Ryabin}, \textbf{J. R. Hall}, \textbf{M. Hartmann}, \textbf{E. B.
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-45. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
+46. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
 \textbf{C. Quince}, and \textbf{F. Mahé}. 2016. VSEARCH: A versatile
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-46. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
+47. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
 OptiClust, an Improved Method for Assigning Amplicon-Based Sequence Data
@@
 \hypertarget{ref-rdp_Wang2007}{}
-47. \textbf{Wang, Q.}, \textbf{G. M. Garrity}, \textbf{J. M. Tiedje},
+48. \textbf{Wang, Q.}, \textbf{G. M. Garrity}, \textbf{J. M. Tiedje},
 and \textbf{J. R. Cole}. 2007. Naive bayesian classifier for rapid
@@
 \hypertarget{ref-permanova_Anderson2013}{}
-48. \textbf{Anderson, M. J.}, and \textbf{D. C. I. Walsh}. 2013.
+49. \textbf{Anderson, M. J.}, and \textbf{D. C. I. Walsh}. 2013.
 PERMANOVA, ANOSIM, and the mantel test in the face of heterogeneous
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-49. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
+50. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
 the false discovery rate: A practical and powerful approach to multiple
@@
 
+\hypertarget{ref-Breiman2001}{}
+51. \textbf{Breiman, L.} 2001. Machine Learning. Springer Nature
+\textbf{45}:5--32.
+
 \hypertarget{ref-r_citation_2017}{}
-50. \textbf{R Core Team}. 2017. R: A language and environment for
+52. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-51. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
+53. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
 extension education program evaluation.
@@
 \hypertarget{ref-car_citation_2011}{}
-52. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
+54. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
 applied regressionSecond. Sage, Thousand Oaks CA.
@@
 \hypertarget{ref-lme4_citation_2015}{}
-53. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
+55. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
 \textbf{S. Walker}. 2015. Fitting linear mixed-effects models using
@@
 \hypertarget{ref-epir_citation_2017}{}
-54. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
+56. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
 Heuer}, \textbf{J. Marshall}, \textbf{J. Sanchez}, \textbf{R. Thornton},
@@
 \hypertarget{ref-metafor_citation_2010}{}
-55. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
+57. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
 the metafor package. Journal of Statistical Software \textbf{36}:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-56. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
+58. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
 \textbf{R. Kindt}, \textbf{P. Legendre}, \textbf{D. McGlinn}, \textbf{P.
@@
 \hypertarget{ref-caret_citation_2017}{}
-57. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
+59. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
 Williams}, \textbf{C. Keefer}, \textbf{A. Engelhardt}, \textbf{T.
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-58. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
+60. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-59. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
+61. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
 analysis. Springer-Verlag New York.
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-60. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
+62. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
 ``grid'' graphics.

### 2018-03-09 submission/manuscript.Rmd
remove unneeded extras leftover from microbiome formatting

@@
 
-## Declarations
-
-### Ethics approval and consent to participate
-
-Ethics approval and informed consent for each of the studies used is mentioned in the respective manuscripts used in this meta-analysis.
-
-### Consent for publication
-
-Not applicable.
-
-### Availability of data and material
-
-A detailed and reproducible description of how the data were processed and analyzed for each study can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
-Raw sequences can be downloaded from the SRA in most cases and can be found in the respective study batch file in the GitHub repository or within the original publication.
-For instances when sequences are not publicly available, they may be accessed by contacting the corresponding authors from whence the data came.
-
-### Competing Interests
-
-All authors declare that they do not have any relevant competing interests to report.
-
-### Funding
-
-MAS is supported by a Canadian Institute of Health Research fellowship and a University of Michigan Postdoctoral Translational Scholar Program grant.
-
-### Authors' contributions
-
-All authors helped to design and conceptualize the study.
-MAS identified and analyzed the data.
-MAS and PDS interpreted the data.
-MAS wrote the first draft of the manuscript and both he and PDS reviewed and revised updated versions.
-All authors approved the final manuscript.
-
 ### Acknowledgements
 
-The authors would like to thank all the study participants who were a part of each of the individual studies utilized.
-We would also like to thank each of the study authors for making their data available for use.
-Finally, we would like to thank the members of the Schloss lab for valuable feed back and proof reading during the formulation of this manuscript.
+The authors would like to thank all the study participants who were a part of each of the individual studies analyzed.
+We would also like to thank each of the study authors for making their sequencing reads and metadata available for use.
+Finally, we would like to thank the members of the Schloss lab for their valuable feedback and proofreading during the formulation of this manuscript.
 

### 2018-03-09 submission/manuscript.tex
remove unneeded extras leftover from microbiome formatting

@@
 
-\subsection{Declarations}\label{declarations}
-
-\subsubsection{Ethics approval and consent to
-participate}\label{ethics-approval-and-consent-to-participate}
-
-Ethics approval and informed consent for each of the studies used is
-mentioned in the respective manuscripts used in this meta-analysis.
-
-\subsubsection{Consent for publication}\label{consent-for-publication}
-
-Not applicable.
-
-\subsubsection{Availability of data and
-material}\label{availability-of-data-and-material}
-
-A detailed and reproducible description of how the data were processed
-and analyzed for each study can be found at
-\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
-Raw sequences can be downloaded from the SRA in most cases and can be
-found in the respective study batch file in the GitHub repository or
-within the original publication. For instances when sequences are not
-publicly available, they may be accessed by contacting the corresponding
-authors from whence the data came.
-
-\subsubsection{Competing Interests}\label{competing-interests}
-
-All authors declare that they do not have any relevant competing
-interests to report.
-
-\subsubsection{Funding}\label{funding}
-
-MAS is supported by a Canadian Institute of Health Research fellowship
-and a University of Michigan Postdoctoral Translational Scholar Program
-grant.
-
-\subsubsection{Authors' contributions}\label{authors-contributions}
-
-All authors helped to design and conceptualize the study. MAS identified
-and analyzed the data. MAS and PDS interpreted the data. MAS wrote the
-first draft of the manuscript and both he and PDS reviewed and revised
-updated versions. All authors approved the final manuscript.
-
 \subsubsection{Acknowledgements}\label{acknowledgements}
@@
 The authors would like to thank all the study participants who were a
-part of each of the individual studies utilized. We would also like to
-thank each of the study authors for making their data available for use.
-Finally, we would like to thank the members of the Schloss lab for
-valuable feed back and proof reading during the formulation of this
-manuscript.
+part of each of the individual studies analyzed. We would also like to
+thank each of the study authors for making their sequencing reads and
+metadata available for use. Finally, we would like to thank the members
+of the Schloss lab for their valuable feedback and proofreading during
+the formulation of this manuscript.
 

### 2018-03-09 submission/manuscript.Rmd
change : to . for titles

@@
 
-***Lower bacterial diversity is associated with increased odds ratio (OR) of tumors:***
+***Lower bacterial diversity is associated with increased odds ratio (OR) of tumors.***
 We first assessed whether variation in broad community metrics like total number of operational taxonomic units (OTUs) (i.e.
@@
 
-***Disease progression is associated with community-wide changes in composition and abundance:*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages.
+***Disease progression is associated with community-wide changes in composition and abundance.*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages.
 We identified significant bacterial community differences in the stool of patients with adenomas relative to those with normal colons in `r length(as.data.frame(filter(adn_beta_stool, pvalue < 0.05))[, "study"])` of `r length(as.data.frame(adn_beta_stool)[, "study"])` studies and in patients with carcinomas relative to those with normal colons in 6 of 7 studies (PERMANOVA; P-value < 0.05) [Table S2].
@@
 
-***Individual taxa are associated with significant ORs for carcinomas:*** 
+***Individual taxa are associated with significant ORs for carcinomas.*** 
 Next we identified those taxa were associated with ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas.
@@
 
-***Individual taxa with a significant OR do a poor job of differentiating subjects with normal colons and those with carcinoma:***
+***Individual taxa with a significant OR do a poor job of differentiating subjects with normal colons and those with carcinoma.***
  We next asked whether those taxa that had a significant OR associated with having a normal colon or carcinomas could be used individually, to classify subjects as having a normal colon or carcinomas.
@@
 
-***Combined taxa model classifies subjects better than using individual taxa:*** 
+***Combined taxa model classifies subjects better than using individual taxa.*** 
 Instead of attempting to classify subjects based on individual taxa, next we combined information from the individual taxa and evaluated the ability to classify a subject’s disease status using Random Forest models.
@@
 
-***Performance of models based on taxa relative abundance in full community are better than those based on taxa with significant ORs:*** 
+***Performance of models based on taxa relative abundance in full community are better than those based on taxa with significant ORs.*** 
 Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR.
@@
 
-***Performance of models based on OTU relative abundance in full community are not significantly better than those based on taxa with significant ORs:*** 
+***Performance of models based on OTU relative abundance in full community are not significantly better than those based on taxa with significant ORs.*** 
 The previous models were based on relative abundance data where sequences were classified to coarse taxonomic assignments (i.e.
@@
 
-***Generalizability of taxon-based models trained on one dataset to the other datasets:*** 
-Considering the good performance of the Random Forest models using taxa with a significant OR and using all of the taxa, we next asked how well the models would perform when given data from a different subject cohort.
+***Generalizability of taxon-based models trained on one dataset to the other datasets.*** 
+Considering the good performance of the Random Forest models using taxa with a significant R and using all of the taxa, we next asked how well the models would perform when given data from a different subject cohort.
 For instance, if a model was trained using data from the Ahn study, we wanted to know how well it would perform using the data from the Baxter study.
@@
 
-***Datasets:***
+***Datasets.***
 The studies used for this meta-analysis were identified through the review articles written by Keku, et al.
@@
 
-***Sequence Processing:***
+***Sequence Processing.***
 Raw sequence data and metadata were primarily obtained from the Sequence Read Archive (SRA) and dbGaP.
@@
 
-***Community analysis:*** 
+***Community analysis.*** 
 We calculated alpha diversity metrics (i.e.
@@
 
-***Random Forest classification analysis:*** 
+***Random Forest classification analysis.*** 
 To classify individuals as having normal colons or tumors, we built Random Forest classification models for each dataset and comparison using taxa with significant ORs (after multiple comparison correction), all taxa, or OTUs.
@@
 
-***Statistical Analysis:***
+***Statistical Analysis.***
 All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
@@
 
-***Reproducible Methods:***
+***Reproducible Methods.***
 The code and analysis can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.

### 2018-03-09 submission/manuscript.tex
change : to . for titles

@@
 \textbf{\emph{Lower bacterial diversity is associated with increased
-odds ratio (OR) of tumors:}} We first assessed whether variation in
+odds ratio (OR) of tumors.}} We first assessed whether variation in
 broad community metrics like total number of operational taxonomic units
@@
 \textbf{\emph{Disease progression is associated with community-wide
-changes in composition and abundance:}} Based on the differences in
+changes in composition and abundance.}} Based on the differences in
 evenness and diversity, we next asked whether there were community-wide
@@
 \textbf{\emph{Individual taxa are associated with significant ORs for
-carcinomas:}} Next we identified those taxa were associated with ORs
+carcinomas.}} Next we identified those taxa were associated with ORs
 that were significantly associated with having a normal colon or the
@@
 \textbf{\emph{Individual taxa with a significant OR do a poor job of
-differentiating subjects with normal colons and those with carcinoma:}}
+differentiating subjects with normal colons and those with carcinoma.}}
 We next asked whether those taxa that had a significant OR associated
@@
 \textbf{\emph{Combined taxa model classifies subjects better than using
-individual taxa:}} Instead of attempting to classify subjects based on
+individual taxa.}} Instead of attempting to classify subjects based on
 individual taxa, next we combined information from the individual taxa
@@
 full community are better than those based on taxa with significant
-ORs:}} Next, we asked whether a Random Forest classification model built
+ORs.}} Next, we asked whether a Random Forest classification model built
 using all of the taxa found in the communities would outperform the
@@
 full community are not significantly better than those based on taxa
-with significant ORs:}} The previous models were based on relative
+with significant ORs.}} The previous models were based on relative
 abundance data where sequences were classified to coarse taxonomic
@@
 \textbf{\emph{Generalizability of taxon-based models trained on one
-dataset to the other datasets:}} Considering the good performance of the
-Random Forest models using taxa with a significant OR and using all of
+dataset to the other datasets.}} Considering the good performance of the
+Random Forest models using taxa with a significant R and using all of
 the taxa, we next asked how well the models would perform when given
@@
 
-\textbf{\emph{Datasets:}} The studies used for this meta-analysis were
+\textbf{\emph{Datasets.}} The studies used for this meta-analysis were
 identified through the review articles written by Keku, et al. (41) and
@@
 
-\textbf{\emph{Sequence Processing:}} Raw sequence data and metadata were
+\textbf{\emph{Sequence Processing.}} Raw sequence data and metadata were
 primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other
@@
 
-\textbf{\emph{Community analysis:}} We calculated alpha diversity
+\textbf{\emph{Community analysis.}} We calculated alpha diversity
 metrics (i.e.~OTU richness, evenness, and Shannon diversity) for each
@@
 
-\textbf{\emph{Random Forest classification analysis:}} To classify
+\textbf{\emph{Random Forest classification analysis.}} To classify
 individuals as having normal colons or tumors, we built Random Forest
@@
 
-\textbf{\emph{Statistical Analysis:}} All statistical analysis after
+\textbf{\emph{Statistical Analysis.}} All statistical analysis after
 sequence processing utilized the R (v3.4.3) software package (52). For
@@
 
-\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
+\textbf{\emph{Reproducible Methods.}} The code and analysis can be found
 at

### 2018-03-12 code/make_rf_auc_full_versus_specific_graph.R
change figure order

@@ -234,6 +234,6 @@ stool_auc_graph <- grid.arrange(adn_stool_graph, crc_stool_graph)
 ggsave("results/figures/FigureS4.pdf", 
        tissue_auc_graph, width = 6, height = 8, dpi = 300)
 
-ggsave("results/figures/Figure4.pdf", 
+ggsave("results/figures/Figure5.pdf", 
        stool_auc_graph, width = 6, height = 8, dpi = 300)
 

### 2018-03-12 code/make_select_imp_otu_graph.R

@@ -70,7 +70,7 @@ imp_unmatched_tissue_model <- unmatched_tissue_mda %>%
 stool_graph <- grid.arrange(imp_stool_model, imp_unmatched_tissue_model, nrow = 1, ncol = 2)
 
 
-ggsave("results/figures/Figure5.pdf", 
+ggsave("results/figures/Figure4.pdf", 
        stool_graph, width = 11, height = 11, dpi = 300)
 
 

### 2018-03-12 code/make_rf_auc_full_versus_specific_graph.R
remove unneeded width call

@@ -146,7 +146,7 @@ adn_stool_graph <- adn_stool %>%
                         labels = c("Baxter", "Brim", "Hale", "Zeller"))) %>% 
   ggplot(aes(model_type, as.numeric(AUC), color = study)) + 
   geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") + 
-  geom_point(width = 0.2, size = 3.5, position = position_dodge(width = 0.3), show.legend = T) + 
+  geom_point(size = 3.5, position = position_dodge(width = 0.3), show.legend = T) + 
   coord_cartesian(ylim = c(0, 1.05)) + 
   stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 
                colour = "black", geom = "crossbar", size = 0.5, width = 0.5) + 
@@ -177,7 +177,7 @@ crc_tissue_graph <- crc_tissue %>%
                         labels = c("Burns", "Chen", "Dejea", "Flemer", "Geng", "Sanapareddy"))) %>% 
   ggplot(aes(model_type, AUC, color = study, group = type)) + 
   geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") + 
-  geom_point(width = 0.2, size = 3.5, position = position_dodge(width = 0.3), show.legend = T) + 
+  geom_point(size = 3.5, position = position_dodge(width = 0.3), show.legend = T) + 
   stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 
                colour = "black", geom = "crossbar", size = 0.5, width = 0.5) + 
   facet_grid(. ~ type) + coord_cartesian(ylim = c(0, 1.05)) + 
@@ -205,7 +205,7 @@ crc_stool_graph <- crc_stool %>%
                         labels = c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller"))) %>% 
   ggplot(aes(model_type, AUC, color = study)) + 
   geom_hline(yintercept = 0.5, color = "gray", linetype = "dashed") + 
-  geom_point(width = 0.2, size = 3.5, position = position_dodge(width = 0.3), show.legend = T) + 
+  geom_point(size = 3.5, position = position_dodge(width = 0.3), show.legend = T) + 
   stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 
                colour = "black", geom = "crossbar", size = 0.5, width = 0.5) + 
   coord_cartesian(ylim = c(0, 1.05)) + 

### 2018-03-12 submission/manuscript.Rmd
good read through upto results

@@
 
-**Results.** We quantified the odds ratios for individual bacterial genera that were associated with an individual having tumors relative to a normal colon.
-Among the stool samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_OR_stool$measure)` genera with significant odds ratios (ORs) associated with carcinoma.
-Similarly, among the tissue samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` genera with significant odds ratios associated with carcinoma.
+**Results.** We quantified the odds ratios for individual bacterial taxa that were associated with an individual having tumors relative to a normal colon.
+Among the stool samples, there were no taxa that had a significant odds ratio associated with adenoma and there were `r length(crc_OR_stool$measure)` taxa with significant odds ratios (ORs) associated with carcinoma.
+Similarly, among the tissue samples, there were no taxa that had a significant odds ratio associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` taxa with significant odds ratios associated with carcinoma.
 Among the significant odds ratios, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`.
@@
 
-Recently, there has been an intense focus on identifying microbiota-based biomarker yielding a seemingly endless number of candidate taxa.
+Recently, there has been an intense focus on identifying microbiota-based biomarkers yielding a seemingly endless number of candidate taxa.
 Some studies point towards mouth-associated genera such as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that are enriched in people with carcinomas [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
@@
 Additionally, *Roseburia* has been found in some studies to be more abundant in people with tumors but in other studies it has been found to be either less abundant or no different than what is found in subjects with normal colons  [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
-There are strong results from tissue culture and murine models that Fusobacterium nucleatum, pks-positive strains of Escherichia coli, Streptococcus gallolyticus, and an entertoxin-producing strain of Bacteroides fragilis are important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017; @weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014].
+There are strong results from tissue culture and murine models that *Fusobacterium nucleatum*, pks-positive strains of *Escherichia coli*, *Streptococcus gallolyticus*, and an entertoxin-producing strain of *Bacteroides fragilis* are important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017; @weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014].
 These results point to a causative role for the microbiota in CRC pathogenesis as well as their potential as diagnostic biomarkers.
@@
 
-Confirming some of these previous findings, a recent meta-analysis found that 16S rRNA gene sequences from members of the *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017].
+Confirming some of these previous findings, a recent meta-analysis found that 16S rRNA gene sequences from members of  *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017].
 Contrary to previous studies they found sequences similar to members of *Lactobacillus* and *Ruminococcus* to be enriched in patients with adenoma or carcinoma relative to those with normal colons [@baxter_microbiota-based_2016; @burns_virulence_2015; @zeller_potential_2014].
@@
 We expand both the breadth and scope of the previous meta-analysis to investigate whether biomarkers describing the bacterial community or specific members of the community can more accurately classify patients as having adenoma or carcinoma.
-Our results suggest that the bacterial community changes as disease severity worsens and that that a subset of the microbial community can be used to diagnose the presence of carcinoma.
+Our results suggest that the bacterial community changes as disease severity worsens and that a subset of the microbial community can be used to diagnose the presence of carcinoma.
 
@@
 richness), the evenness of their abundance, and the overall diversity was associated with disease stage after controlling for study and variable region differences.
-In stool, there was a significant decrease in both evenness and diversity as disease severity progressed from normal to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1]; there was not a significant difference for richness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`).
+In stool, there was a significant decrease in both evenness and diversity as disease severity progressed from normal to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1]; there was no significant difference for richness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`).
 We next tested whether the decrease in these community metrics translated into significant ORs for having an adenoma or carcinoma.
 For fecal samples, the ORs for richness were not significantly greater than 1.0 for adenoma or carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "sobs")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A].
-The ORs for evenness were significantly higher than 1.0 for adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and carcinoma (OR =  `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2B].
+The ORs for evenness were significantly higher than 1.0 for adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)(95% Confidence Interval), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and carcinoma (OR =  `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2B].
 The ORs for diversity were only significantly greater than 1.0 for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`), but not for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2C].
@@
 For studies that used matched samples no differences in bacterial community structures were observed [Table S2].
-Combined, these results indicate that there consistent and significant community-wide changes in the fecal community structure of subjects with carcinomas.
+Combined, these results indicate that there are consistent and significant community-wide changes in the fecal community structure of subjects with carcinomas.
 However, the signal observed in subjects with adenomas or when using tissue samples was not as consistent.
@@
 ***Individual taxa are associated with significant ORs for carcinomas.*** 
-Next we identified those taxa were associated with ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas.
+Next we identified those taxa that were associated with ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas.
 No taxa had a significant OR for the presence of adenomas when we used data collected from stool or tissue samples (Table S3 & S4).
@@
 Mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4].
-For example, *Fusobacterium* had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`; however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg-corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4].
-It is interesting to note that *Ruminococcus* and members of *Clostridium group XI* in stool and *Dorea* and *Blautia* in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas.
+For example, *Fusobacterium* had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`; however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4].
+It is interesting to note that *Ruminococcus* and members of *Clostridium XI* in stool and *Dorea* and *Blautia* in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas.
 Overall, there was no overlap in the taxa with significant OR between stool and tissue samples.
@@
 The taxa that performed the best included *Clostridium XI*, *Ruminococcus*, and *Escherichia*, however, these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)`.
-Likewise, in unmatched tissue samples the `r length(crc_OR_stool$measure)` taxa with significant ORs taxa were marginally better than one would expect by chance [Figure 3B].
-The relative abundance of Dorea was the best predictor of carcinomas and its median AUC was only `r round(max(filter(ind_crc_combined_unmatched_auc, taxa == "Dorea")[, "auc"]), digits = 3)`.
+Likewise, in unmatched tissue samples the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant OR taxa were marginally better than one would expect by chance [Figure 3B].
+The relative abundance of *Dorea* was the best predictor of carcinomas and its median AUC was only `r round(max(filter(ind_crc_combined_unmatched_auc, taxa == "Dorea")[, "auc"]), digits = 3)`.
 These results suggest that although these taxa are associated with a decreased or increased OR for the presences of carcinomas, individually, they do a poor job of classifying a subject’s disease status.
@@
 Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study.
-The most important taxa across study within the significant OR taxa only models for stool were *Ruminococcus* and *Clostridium XI* [Figure 5A].
-Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for Blautia and Weissella (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).
+Within the significant OR taxa only models the most important taxa across study for stool were *Ruminococcus* and *Clostridium XI* based on mean decrease in accuracy [Figure 4A].
+Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for *Blautia* and *Weissella* (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).
 For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S3].
-For the singificant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the important based on mean decrease in accuracy [Figure 5B].
+For the significant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the most important based on mean decrease in accuracy [Figure 4B].
 Pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.
@@
 Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR.
-Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were only marginally better than 0.5 for any study (median AUC = `r median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)` [`r min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)` - `r max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T)`]) [Figure 4A & S4A].
-In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 4B & S4B-C].
+Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were only marginally better than 0.5 for any study (median AUC = `r round(median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` (`r round(min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` - `r round(max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)`)(min-max)) [Figure 5A & S4A].
+In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 5B & S4B-C].
 When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed.
-Using the data from stool samples we found that the AUC for 6 of 7 studies increased by an average of `r round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"]))*100), digits = 2)`%) and AUC for the Flemer study decreased by `r round(abs(mean(((1 - filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"]))*100)), digits = 2)`%).
+Using the data from stool samples we found that the AUC for 6 of 7 studies increased by an average of `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"]))*100), digits = 2))`%) and AUC for the Flemer study decreased by `r round(abs(mean(((1 - filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"]))*100)), digits = 2)`%).
 The overall improvement in performance was statistically significant (mean = `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"]))*100), digits = 2))`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
 Similarly, using the data from unmatched tissue samples we found that the AUC of studies increased by an average of `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`% when we used all of the taxa rather than the 3 taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
-Although the significant OR taxa models can classify indidivuals with and without carcinoma tumors, they are still missing taxa from the full community models that can increase the model accuracy.
+Although the significant OR taxa models can classify individuals with and without carcinoma tumors, they are still missing taxa from the full community models that can increase the model accuracy.
 
@@
 To determine whether model performance improved with finer scale classification, we assigned sequences to operational taxonomic units (OTUs) where the similarity among sequences within an OTU was more than 97%.
-We again found that classification models built using all of the sequence data for a community did a poor job of differentiating between subjects with normal colons and those with adenomas (median AUC: `r round(median(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]), but did a good job of differentiating between subjects with normal colons and those with carcinomas (median AUC: `r round(median(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` [`r round(min(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`]).
+We again found that classification models built using all of the sequence data for a community did a poor job of differentiating between subjects with normal colons and those with adenomas (median AUC: `r round(median(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` (`r round(min(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`)), but did a good job of differentiating between subjects with normal colons and those with carcinomas (median AUC: `r round(median(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` (`r round(min(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`)).
 The OTU-based models performed similarly to those constructed using the taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) and those using all taxa (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) [Figure 4].
@@
 ***Generalizability of taxon-based models trained on one dataset to the other datasets.*** 
-Considering the good performance of the Random Forest models using taxa with a significant R and using all of the taxa, we next asked how well the models would perform when given data from a different subject cohort.
+Considering the good performance of the Random Forest models using taxa with a significant OR and using all of the taxa, we next asked how well the models would perform when given data from a different subject cohort.
 For instance, if a model was trained using data from the Ahn study, we wanted to know how well it would perform using the data from the Baxter study.

### 2018-03-12 submission/manuscript.tex
good read through upto results

@@
 \textbf{Results.} We quantified the odds ratios for individual bacterial
-genera that were associated with an individual having tumors relative to
-a normal colon. Among the stool samples, there were no genera that had a
-significant odds ratio associated with adenoma and there were 8 genera
+taxa that were associated with an individual having tumors relative to a
+normal colon. Among the stool samples, there were no taxa that had a
+significant odds ratio associated with adenoma and there were 8 taxa
 with significant odds ratios (ORs) associated with carcinoma. Similarly,
-among the tissue samples, there were no genera that had a significant
-odds ratio associated with adenoma and there were 3 genera with
-significant odds ratios associated with carcinoma. Among the significant
-odds ratios, the association between individual taxa and tumor diagnosis
-was equal or below 7.11. Because individual taxa had limited association
+among the tissue samples, there were no taxa that had a significant odds
+ratio associated with adenoma and there were 3 taxa with significant
+odds ratios associated with carcinoma. Among the significant odds
+ratios, the association between individual taxa and tumor diagnosis was
+equal or below 7.11. Because individual taxa had limited association
 with tumor diagnosis, we trained Random Forest classification models
@@
 Recently, there has been an intense focus on identifying
-microbiota-based biomarker yielding a seemingly endless number of
+microbiota-based biomarkers yielding a seemingly endless number of
 candidate taxa. Some studies point towards mouth-associated genera such
@@
 20). There are strong results from tissue culture and murine models that
-Fusobacterium nucleatum, pks-positive strains of Escherichia coli,
-Streptococcus gallolyticus, and an entertoxin-producing strain of
-Bacteroides fragilis are important in the pathogenesis of CRC (5, 14,
-21--24). These results point to a causative role for the microbiota in
-CRC pathogenesis as well as their potential as diagnostic biomarkers.
+\emph{Fusobacterium nucleatum}, pks-positive strains of
+\emph{Escherichia coli}, \emph{Streptococcus gallolyticus}, and an
+entertoxin-producing strain of \emph{Bacteroides fragilis} are important
+in the pathogenesis of CRC (5, 14, 21--24). These results point to a
+causative role for the microbiota in CRC pathogenesis as well as their
+potential as diagnostic biomarkers.
 
@@
 Confirming some of these previous findings, a recent meta-analysis found
-that 16S rRNA gene sequences from members of the \emph{Akkermansia},
+that 16S rRNA gene sequences from members of \emph{Akkermansia},
 \emph{Fusobacterium}, and \emph{Parvimonas} were fecal biomarkers for
@@
 that the bacterial community changes as disease severity worsens and
-that that a subset of the microbial community can be used to diagnose
-the presence of carcinoma.
+that a subset of the microbial community can be used to diagnose the
+presence of carcinoma.
 
@@
 from normal to adenoma to carcinoma (P-value = 0.025 and 0.043,
-respectively) {[}Figure 1{]}; there was not a significant difference for
+respectively) {[}Figure 1{]}; there was no significant difference for
 richness (P-value = 0.21). We next tested whether the decrease in these
@@
 {[}Figure 2A{]}. The ORs for evenness were significantly higher than 1.0
-for adenoma (OR = 1.3 (1.02 - 1.65), P-value = 0.035) and carcinoma (OR
-= 1.66 (1.2 - 2.3), P-value = 0.0021) {[}Figure 2B{]}. The ORs for
-diversity were only significantly greater than 1.0 for carcinoma (OR =
-1.61 (1.14 - 2.28), P-value = 0.0069), but not for adenoma (P-value =
-0.11) {[}Figure 2C{]}. Although these OR are significantly greater than
-1.0, it is doubtful that these are clinically meaningful values.
+for adenoma (OR = 1.3 (1.02 - 1.65)(95\% Confidence Interval), P-value =
+0.035) and carcinoma (OR = 1.66 (1.2 - 2.3), P-value = 0.0021) {[}Figure
+2B{]}. The ORs for diversity were only significantly greater than 1.0
+for carcinoma (OR = 1.61 (1.14 - 2.28), P-value = 0.0069), but not for
+adenoma (P-value = 0.11) {[}Figure 2C{]}. Although these OR are
+significantly greater than 1.0, it is doubtful that these are clinically
+meaningful values.
 
@@
 structures were observed {[}Table S2{]}. Combined, these results
-indicate that there consistent and significant community-wide changes in
-the fecal community structure of subjects with carcinomas. However, the
-signal observed in subjects with adenomas or when using tissue samples
-was not as consistent. This is likely due to a smaller effect size or
-the relatively small sample sizes among the studies that characterized
-the tissue microbiota.
+indicate that there are consistent and significant community-wide
+changes in the fecal community structure of subjects with carcinomas.
+However, the signal observed in subjects with adenomas or when using
+tissue samples was not as consistent. This is likely due to a smaller
+effect size or the relatively small sample sizes among the studies that
+characterized the tissue microbiota.
 
 \textbf{\emph{Individual taxa are associated with significant ORs for
-carcinomas.}} Next we identified those taxa were associated with ORs
-that were significantly associated with having a normal colon or the
+carcinomas.}} Next we identified those taxa that were associated with
+ORs that were significantly associated with having a normal colon or the
 presence of adenomas or carcinomas. No taxa had a significant OR for the
@@
 due to the small number of studies and considerable variation in the
-data, the Benjimani-Hochberg-corrected P-value was 0.93 {[}Table S4{]}.
+data, the Benjimani-Hochberg corrected P-value was 0.93 {[}Table S4{]}.
 It is interesting to note that \emph{Ruminococcus} and members of
-\emph{Clostridium group XI} in stool and \emph{Dorea} and \emph{Blautia}
-in tissue had ORs that were significantly less than 1.0, which suggests
+\emph{Clostridium XI} in stool and \emph{Dorea} and \emph{Blautia} in
+tissue had ORs that were significantly less than 1.0, which suggests
 that these populations are protective against the development of
@@
 \emph{Ruminococcus}, and \emph{Escherichia}, however, these had median
-AUC values less than 0.588. Likewise, in unmatched tissue samples the 8
-taxa with significant ORs taxa were marginally better than one would
-expect by chance {[}Figure 3B{]}. The relative abundance of Dorea was
-the best predictor of carcinomas and its median AUC was only 0.62. These
-results suggest that although these taxa are associated with a decreased
-or increased OR for the presences of carcinomas, individually, they do a
-poor job of classifying a subject's disease status.
+AUC values less than 0.588. Likewise, in unmatched tissue samples the 3
+taxa with significant OR taxa were marginally better than one would
+expect by chance {[}Figure 3B{]}. The relative abundance of \emph{Dorea}
+was the best predictor of carcinomas and its median AUC was only 0.62.
+These results suggest that although these taxa are associated with a
+decreased or increased OR for the presences of carcinomas, individually,
+they do a poor job of classifying a subject's disease status.
 
@@
 ORs, all 3 were among the top 10\% most important taxa as measured by
-mean decrease in accuracy, in at least one study. The most important
-taxa across study within the significant OR taxa only models for stool
-were \emph{Ruminococcus} and \emph{Clostridium XI} {[}Figure 5A{]}.
-Similarly, using data from the unmatched tissue samples, the combined
-model had an AUC of 0.77, which was significantly higher than the AUC
-values for Blautia and Weissella (P-value \textless{} 0.037). For the
-full taxa models using unmatched tissue, \emph{Lachnospiraceae},
-\emph{Bacteroidaceae}, and \emph{Ruminococcaceae} were the most common
-taxa in the top 10\% mean decrease in accuracy across studies {[}Figure
-S3{]}. For the singificant OR taxa unmatched tissue models both
-\emph{Dorea} and \emph{Blautia} were the important based on mean
-decrease in accuracy {[}Figure 5B{]}. Pooling the information from the
-taxa with significant ORs results in a model that outperforms
-classifications made using individual taxa.
+mean decrease in accuracy, in at least one study. Within the significant
+OR taxa only models the most important taxa across study for stool were
+\emph{Ruminococcus} and \emph{Clostridium XI} based on mean decrease in
+accuracy {[}Figure 4A{]}. Similarly, using data from the unmatched
+tissue samples, the combined model had an AUC of 0.77, which was
+significantly higher than the AUC values for \emph{Blautia} and
+\emph{Weissella} (P-value \textless{} 0.037). For the full taxa models
+using unmatched tissue, \emph{Lachnospiraceae}, \emph{Bacteroidaceae},
+and \emph{Ruminococcaceae} were the most common taxa in the top 10\%
+mean decrease in accuracy across studies {[}Figure S3{]}. For the
+significant OR taxa unmatched tissue models both \emph{Dorea} and
+\emph{Blautia} were the most important based on mean decrease in
+accuracy {[}Figure 4B{]}. Pooling the information from the taxa with
+significant ORs results in a model that outperforms classifications made
+using individual taxa.
 
@@
 normal colons or having adenomas using data from stool or tissue samples
-were only marginally better than 0.5 for any study (median AUC =
-0.5486298 {[}0.3671667 - 0.971{]}) {[}Figure 4A \& S4A{]}. In contrast,
-the models for classifying subjects as having normal colons or having
+were only marginally better than 0.5 for any study (median AUC = 0.549
+(0.367 - 0.971)(min-max)) {[}Figure 5A \& S4A{]}. In contrast, the
+models for classifying subjects as having normal colons or having
 carcinomas using data from stool or tissue samples yielded AUC values
-meaningfully higher than 0.5 {[}Figure 4B \& S4B-C{]}. When we compared
+meaningfully higher than 0.5 {[}Figure 5B \& S4B-C{]}. When we compared
 the models based on all of the taxa in a community to models based on
@@
 from stool samples we found that the AUC for 6 of 7 studies increased by
-an average of -14.8\%) and AUC for the Flemer study decreased by
-0.54\%). The overall improvement in performance was statistically
-significant (mean = 12.61\%, one-tailed paired T-test; P-value = 0.005).
-Similarly, using the data from unmatched tissue samples we found that
-the AUC of studies increased by an average of 19.11\% when we used all
-of the taxa rather than the 3 taxa with significant ORs (one-tailed
-paired T-test; P-value = 0.03). Although the significant OR taxa models
-can classify indidivuals with and without carcinoma tumors, they are
-still missing taxa from the full community models that can increase the
-model accuracy.
+an average of 14.8\%) and AUC for the Flemer study decreased by 0.54\%).
+The overall improvement in performance was statistically significant
+(mean = 12.61\%, one-tailed paired T-test; P-value = 0.005). Similarly,
+using the data from unmatched tissue samples we found that the AUC of
+studies increased by an average of 19.11\% when we used all of the taxa
+rather than the 3 taxa with significant ORs (one-tailed paired T-test;
+P-value = 0.03). Although the significant OR taxa models can classify
+individuals with and without carcinoma tumors, they are still missing
+taxa from the full community models that can increase the model
+accuracy.
 
@@
 community did a poor job of differentiating between subjects with normal
-colons and those with adenomas (median AUC: 0.53 {[}0.37- 0.56{]}), but
-did a good job of differentiating between subjects with normal colons
-and those with carcinomas (median AUC: 0.71 {[}0.5- 0.9{]}). The
-OTU-based models performed similarly to those constructed using the taxa
-with significant ORs (one-tailed paired T-test; P-value = 0.966) and
-those using all taxa (one-tailed paired T-test; P-value = 0.146)
-{[}Figure 4{]}. Among the OTUs that had the highest mean decrease in
-accuracy for the OTU-based models, we found that OTUs that affiliated
-with all of the 8 taxa that had a significant OR were within the top
-10\% for at least one study. This result was surprising as it indicated
-that a finer scale classification of the sequences and thus a larger
-number of features to select from, did not yield improved classification
-of the subjects.
+colons and those with adenomas (median AUC: 0.53 (0.37- 0.56)), but did
+a good job of differentiating between subjects with normal colons and
+those with carcinomas (median AUC: 0.71 (0.5- 0.9)). The OTU-based
+models performed similarly to those constructed using the taxa with
+significant ORs (one-tailed paired T-test; P-value = 0.966) and those
+using all taxa (one-tailed paired T-test; P-value = 0.146) {[}Figure
+4{]}. Among the OTUs that had the highest mean decrease in accuracy for
+the OTU-based models, we found that OTUs that affiliated with all of the
+8 taxa that had a significant OR were within the top 10\% for at least
+one study. This result was surprising as it indicated that a finer scale
+classification of the sequences and thus a larger number of features to
+select from, did not yield improved classification of the subjects.
 
@@
 dataset to the other datasets.}} Considering the good performance of the
-Random Forest models using taxa with a significant R and using all of
+Random Forest models using taxa with a significant OR and using all of
 the taxa, we next asked how well the models would perform when given

### 2018-03-12 submission/manuscript.Rmd
disucssion read through

@@
 We hypothesize that fine scale classification may not result in better classification because distribution of microbiota between individuals is patchy.
-In contrast, models using coarser taxonomic assignments will pool the fine scale diversity and resulting in less patchiness and better classification.
-Furthermore, the ability of models trained using a subset of the community to outperform those using the full community when testing the models on the other datasets may also be a product of the patchiness of human-associated microbiota.
+In contrast, models using coarser taxonomic assignments will pool the fine scale diversity, resulting in less patchiness and better classification.
+Furthermore, the ability of models trained using a subset of the community to outperform those using the full community when testing the models on the other datasets may also be a product of the patchiness of the human-associated microbiota.
 The models based on the 8 taxa that had significant ORs used taxa that were found in every study and tended to have higher relative abundances.
@@
 
-All of the adenoma classification models performed poorly.
-This result is not inconsistent with previous studies [@hale_shifts_2017; @Sze2017].
+All of the adenoma classification models performed poorly and is not inconsistent with previous studies [@hale_shifts_2017; @Sze2017].
 However, the classification results are at odds with results of the multitarget microbiota test (MMT) from Baxter, *et al.* [@baxter_microbiota-based_2016] who observed an AUC of 0.755 when applied to individuals with adenomas.
@@
 Studies were included that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing.
-Some studies (N = 12) were excluded because they did not have publicly available sequences or did not have metadata in which the authors were able to share.
+Some studies (N = 12) were excluded because they did not have publicly available sequences, did not use 454 or Illumina machines, or did not have metadata in which the authors were able to share.
 We were able to obtain sequence data and metadata from the following studies: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014].
@@
 For all models, the default setting used was $\sqrt{p}$, where p is all the variables used in the respective model.
-Normally, $\sqrt{p}$ has been found to be what is chosen as the ideal mtry [@Breiman2001].
+Normally, $\sqrt{p}$ has been found to be what is chosen as the ideal mtry after model tuneing [@Breiman2001].
 
@@
 
-**Figure 4: Stool Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
+**Figure 4: Most Important Members in Significant OR Taxa Carcinoma Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest stool-based models.
+B) Common taxa in the top 10 percent for carcinoma Random Forest unmatched tissue-based models.
+Blue represents less important and red represents more important to the Random Forest Model.
+White means that particular taxa was not in the top 10%.
+
+**Figure 5: Stool Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
 B) Carcinoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
@@
 If no values are present in the singificant OR taxa group then there were no significant taxa identified and no model was tested.
-
-**Figure 5: Most Important Members in Significant OR Taxa Carcinoma Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest stool-based models.
-B) Common taxa in the top 10 percent for carcinoma Random Forest unmatched tissue-based models.
-Blue represents less important and red represents more important to the Random Forest Model.
-White means that particular taxa was not in the top 10%.
 

### 2018-03-12 submission/manuscript.tex
disucssion read through

@@
 contrast, models using coarser taxonomic assignments will pool the fine
-scale diversity and resulting in less patchiness and better
-classification. Furthermore, the ability of models trained using a
-subset of the community to outperform those using the full community
-when testing the models on the other datasets may also be a product of
-the patchiness of human-associated microbiota. The models based on the 8
-taxa that had significant ORs used taxa that were found in every study
-and tended to have higher relative abundances. Similar to the OTU-based
-models, those models based on the full community taxonomy assignments
-were still sensitive to the patchy distribution of taxa. Regardless, it
-is encouraging that a collection of 8 taxa could reliably classify
+scale diversity, resulting in less patchiness and better classification.
+Furthermore, the ability of models trained using a subset of the
+community to outperform those using the full community when testing the
+models on the other datasets may also be a product of the patchiness of
+the human-associated microbiota. The models based on the 8 taxa that had
+significant ORs used taxa that were found in every study and tended to
+have higher relative abundances. Similar to the OTU-based models, those
+models based on the full community taxonomy assignments were still
+sensitive to the patchy distribution of taxa. Regardless, it is
+encouraging that a collection of 8 taxa could reliably classify
 individuals as having carcinomas considering the differences in cohorts,
@@
 
-All of the adenoma classification models performed poorly. This result
-is not inconsistent with previous studies (27, 30). However, the
-classification results are at odds with results of the multitarget
-microbiota test (MMT) from Baxter, \emph{et al.} (12) who observed an
-AUC of 0.755 when applied to individuals with adenomas. There are two
-major differences between the models generated in this meta-analysis and
-that analysis. The MMT attempted to classify individuals as having a
-normal colon or having colonic lesions (i.e.~adenomas or carcinomas) and
-not adenomas alone. Further, the MMT incorporated fecal immunoglobulin
-test (FIT) data while our models only used 16S rRNA gene sequencing
-data. Because FIT data were not available for the other studies in our
-meta-analysis, it was not possible to validate the MMT approach. The
-ability to differentiate between individuals with and without adenomas
-is an important problem since early detection of tumors is critical.
-However, it is possible that we might have been able to detect
-differences in the bacterial community if individuals with non-advanced
-and advanced adenomas were separated. This is a clinically relevant
-distinction since advanced adenomas are at highest risk of progressing
-to a carcinoma. The initial changes of the microbiota during
-tumorigenesis could be focal to where the initial adenoma develops and
-would not be easily assessed using fecal samples from an individual with
-non-advanced adenomas. Unfortunately, distinguishing between individuals
-with advanced and non-advanced adenomas was not possible in our
-meta-analysis since the studies did not provide the clinical data needed
-to make that distinction.
+All of the adenoma classification models performed poorly and is not
+inconsistent with previous studies (27, 30). However, the classification
+results are at odds with results of the multitarget microbiota test
+(MMT) from Baxter, \emph{et al.} (12) who observed an AUC of 0.755 when
+applied to individuals with adenomas. There are two major differences
+between the models generated in this meta-analysis and that analysis.
+The MMT attempted to classify individuals as having a normal colon or
+having colonic lesions (i.e.~adenomas or carcinomas) and not adenomas
+alone. Further, the MMT incorporated fecal immunoglobulin test (FIT)
+data while our models only used 16S rRNA gene sequencing data. Because
+FIT data were not available for the other studies in our meta-analysis,
+it was not possible to validate the MMT approach. The ability to
+differentiate between individuals with and without adenomas is an
+important problem since early detection of tumors is critical. However,
+it is possible that we might have been able to detect differences in the
+bacterial community if individuals with non-advanced and advanced
+adenomas were separated. This is a clinically relevant distinction since
+advanced adenomas are at highest risk of progressing to a carcinoma. The
+initial changes of the microbiota during tumorigenesis could be focal to
+where the initial adenoma develops and would not be easily assessed
+using fecal samples from an individual with non-advanced adenomas.
+Unfortunately, distinguishing between individuals with advanced and
+non-advanced adenomas was not possible in our meta-analysis since the
+studies did not provide the clinical data needed to make that
+distinction.
 
@@
 for 454 or Illumina 16S rRNA gene sequencing. Some studies (N = 12) were
-excluded because they did not have publicly available sequences or did
-not have metadata in which the authors were able to share. We were able
-to obtain sequence data and metadata from the following studies: Ahn,
-\emph{et al.} (11), Baxter, \emph{et al.} (12), Brim, \emph{et al.}
-(29), Burns, \emph{et al.} (15), Chen, \emph{et al.} (13), Dejea,
-\emph{et al.} (20), Flemer, \emph{et al.} (17), Geng, \emph{et al.}
-(19), Hale, \emph{et al.} (27), Kostic, \emph{et al.} (43), Lu, \emph{et
-al.} (26), Sanapareddy, \emph{et al.} (25), Wang, \emph{et al.} (14),
-Weir, \emph{et al.} (23), and Zeller, \emph{et al.} (16). The Zackular
-(44) study was excluded because their 90 individuals were contained
-within the larger Baxter study (12). The Kostic study was excluded
-because after we processed the sequences, all of the case samples had
-100 or fewer sequences. The final analysis included 14 studies (Tables 1
-and 2). There were seven studies with only fecal samples (Ahn, Baxter,
-Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue
-samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both
-fecal and tissue samples (Chen and Flemer). After curating the
-sequences, 1737 stool samples and 492 tissue samples remained in the
-analysis {[}Tables 1 and 2{]}.
+excluded because they did not have publicly available sequences, did not
+use 454 or Illumina machines, or did not have metadata in which the
+authors were able to share. We were able to obtain sequence data and
+metadata from the following studies: Ahn, \emph{et al.} (11), Baxter,
+\emph{et al.} (12), Brim, \emph{et al.} (29), Burns, \emph{et al.} (15),
+Chen, \emph{et al.} (13), Dejea, \emph{et al.} (20), Flemer, \emph{et
+al.} (17), Geng, \emph{et al.} (19), Hale, \emph{et al.} (27), Kostic,
+\emph{et al.} (43), Lu, \emph{et al.} (26), Sanapareddy, \emph{et al.}
+(25), Wang, \emph{et al.} (14), Weir, \emph{et al.} (23), and Zeller,
+\emph{et al.} (16). The Zackular (44) study was excluded because their
+90 individuals were contained within the larger Baxter study (12). The
+Kostic study was excluded because after we processed the sequences, all
+of the case samples had 100 or fewer sequences. The final analysis
+included 14 studies (Tables 1 and 2). There were seven studies with only
+fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five
+studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy),
+and two studies with both fecal and tissue samples (Chen and Flemer).
+After curating the sequences, 1737 stool samples and 492 tissue samples
+remained in the analysis {[}Tables 1 and 2{]}.
 
@@
 Normally, \(\sqrt{p}\) has been found to be what is chosen as the ideal
-mtry (51).
+mtry after model tuneing (51).
 
@@
 
-\textbf{Figure 4: Stool Random Forest Model Train AUCs.} A) Adenoma
+\textbf{Figure 4: Most Important Members in Significant OR Taxa
+Carcinoma Models.} A) Common taxa in the top 10 percent for carcinoma
+Random Forest stool-based models. B) Common taxa in the top 10 percent
+for carcinoma Random Forest unmatched tissue-based models. Blue
+represents less important and red represents more important to the
+Random Forest Model. White means that particular taxa was not in the top
+10\%.
+
+\textbf{Figure 5: Stool Random Forest Model Train AUCs.} A) Adenoma
 random forest model AUCs between all genera, all OTU, and select model
@@
 were no significant taxa identified and no model was tested.
-
-\textbf{Figure 5: Most Important Members in Significant OR Taxa
-Carcinoma Models.} A) Common taxa in the top 10 percent for carcinoma
-Random Forest stool-based models. B) Common taxa in the top 10 percent
-for carcinoma Random Forest unmatched tissue-based models. Blue
-represents less important and red represents more important to the
-Random Forest Model. White means that particular taxa was not in the top
-10\%.
 

### 2018-03-13 code/make_rf_auc_full_versus_specific_graph.R
surface changes

@@ -57,7 +57,7 @@ crc_tissue <- crc_tissue_unmatched %>% bind_rows(crc_tissue_matched) %>%
       mutate(type = "matched", 
              model_type = "full_otu", 
              train_model = study, 
-             act_mean_auc = as.numeric(act_mean_auc)) %>% 
+             act_mean_auc = 1 - as.numeric(act_mean_auc)) %>% 
       rename(AUC = act_mean_auc) %>% 
       select(pvalue, study, train_model, type, model_type, AUC) %>% 
       bind_rows(

### 2018-03-13 code/make_select_imp_otu_graph.R

@@ -71,7 +71,7 @@ stool_graph <- grid.arrange(imp_stool_model, imp_unmatched_tissue_model, nrow = 
 
 
 ggsave("results/figures/Figure4.pdf", 
-       stool_graph, width = 11, height = 11, dpi = 300)
+       stool_graph, width = 11, height = 6, dpi = 300)
 
 
 

### 2018-03-13 code/make_rf_auc_full_versus_specific_graph.R
order changes of figs

@@ -231,7 +231,7 @@ crc_stool_graph <- crc_stool %>%
 tissue_auc_graph <- grid.arrange(adn_tissue_graph, crc_tissue_graph)
 stool_auc_graph <- grid.arrange(adn_stool_graph, crc_stool_graph)
 
-ggsave("results/figures/FigureS4.pdf", 
+ggsave("results/figures/FigureS2.pdf", 
        tissue_auc_graph, width = 6, height = 8, dpi = 300)
 
 ggsave("results/figures/Figure5.pdf", 

### 2018-03-13 code/make_stool_imp_otu_graph.R

@@ -73,7 +73,7 @@ crc_otu <- stool_otu_mda %>%
 stool_graph <- grid.arrange(crc_genera, crc_otu, nrow = 1, ncol = 2)
 
 
-ggsave("results/figures/FigureS2.pdf", 
+ggsave("results/figures/FigureS3.pdf", 
        stool_graph, width = 10, height = 11, dpi = 300)
 
 

### 2018-03-13 code/make_tissue_imp_otu_graph.R

@@ -133,5 +133,5 @@ tissue_graph <- grid.arrange(crc_matched_genera_graph, crc_unmatched_genera_grap
                             layout_matrix = rbind(c(1, 2), c(3, 4)))
 
 
-ggsave("results/figures/FigureS3.pdf", 
+ggsave("results/figures/FigureS4.pdf", 
        tissue_graph, width = 10, height = 13, dpi = 300)

### 2018-03-13 submission/manuscript.Rmd
move problem section to full taxa vs combined taxa section

@@
 For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`).
-For the full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies [Figure S2].
-Of the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs, all `r length(crc_unmatched_OR_tissue$measure)` were among the top 10% most important taxa as measured by mean decrease in accuracy, in at least one study.
 Within the significant OR taxa only models the most important taxa across study for stool were *Ruminococcus* and *Clostridium XI* based on mean decrease in accuracy [Figure 4A].
 Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for *Blautia* and *Weissella* (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).
-For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S3].
 For the significant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the most important based on mean decrease in accuracy [Figure 4B].
@@
 Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR.
-Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were only marginally better than 0.5 for any study (median AUC = `r round(median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` (`r round(min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` - `r round(max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)`)(min-max)) [Figure 5A & S4A].
-In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 5B & S4B-C].
+Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were only marginally better than 0.5 for any study (median AUC = `r round(median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` (`r round(min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` - `r round(max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)`)(min-max)) [Figure 5A & S2A].
+In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 5B & S2B-C].
 When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed.
@@
 The overall improvement in performance was statistically significant (mean = `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"]))*100), digits = 2))`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
+For these full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies [Figure S3].
 Similarly, using the data from unmatched tissue samples we found that the AUC of studies increased by an average of `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`% when we used all of the taxa rather than the 3 taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
+For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S4].
 Although the significant OR taxa models can classify individuals with and without carcinoma tumors, they are still missing taxa from the full community models that can increase the model accuracy.

### 2018-03-13 submission/manuscript.tex
move problem section to full taxa vs combined taxa section

@@
 had an AUC of 0.75, which was significantly higher than any of the AUC
-values for the individual taxa (P-value \textless{} 0.033). For the full
-taxa models using stool, \emph{Bacteroides} and \emph{Lachnospiraceae}
-were the most common taxa in the top 10\% mean decrease in accuracy
-(MDA) across studies {[}Figure S2{]}. Of the 3 taxa with significant
-ORs, all 3 were among the top 10\% most important taxa as measured by
-mean decrease in accuracy, in at least one study. Within the significant
-OR taxa only models the most important taxa across study for stool were
-\emph{Ruminococcus} and \emph{Clostridium XI} based on mean decrease in
-accuracy {[}Figure 4A{]}. Similarly, using data from the unmatched
-tissue samples, the combined model had an AUC of 0.77, which was
-significantly higher than the AUC values for \emph{Blautia} and
-\emph{Weissella} (P-value \textless{} 0.037). For the full taxa models
-using unmatched tissue, \emph{Lachnospiraceae}, \emph{Bacteroidaceae},
-and \emph{Ruminococcaceae} were the most common taxa in the top 10\%
-mean decrease in accuracy across studies {[}Figure S3{]}. For the
-significant OR taxa unmatched tissue models both \emph{Dorea} and
-\emph{Blautia} were the most important based on mean decrease in
-accuracy {[}Figure 4B{]}. Pooling the information from the taxa with
-significant ORs results in a model that outperforms classifications made
-using individual taxa.
+values for the individual taxa (P-value \textless{} 0.033). Within the
+significant OR taxa only models the most important taxa across study for
+stool were \emph{Ruminococcus} and \emph{Clostridium XI} based on mean
+decrease in accuracy {[}Figure 4A{]}. Similarly, using data from the
+unmatched tissue samples, the combined model had an AUC of 0.77, which
+was significantly higher than the AUC values for \emph{Blautia} and
+\emph{Weissella} (P-value \textless{} 0.037). For the significant OR
+taxa unmatched tissue models both \emph{Dorea} and \emph{Blautia} were
+the most important based on mean decrease in accuracy {[}Figure 4B{]}.
+Pooling the information from the taxa with significant ORs results in a
+model that outperforms classifications made using individual taxa.
 
@@
 were only marginally better than 0.5 for any study (median AUC = 0.549
-(0.367 - 0.971)(min-max)) {[}Figure 5A \& S4A{]}. In contrast, the
+(0.367 - 0.971)(min-max)) {[}Figure 5A \& S2A{]}. In contrast, the
 models for classifying subjects as having normal colons or having
 carcinomas using data from stool or tissue samples yielded AUC values
-meaningfully higher than 0.5 {[}Figure 5B \& S4B-C{]}. When we compared
+meaningfully higher than 0.5 {[}Figure 5B \& S2B-C{]}. When we compared
 the models based on all of the taxa in a community to models based on
@@
 The overall improvement in performance was statistically significant
-(mean = 12.61\%, one-tailed paired T-test; P-value = 0.005). Similarly,
+(mean = 12.61\%, one-tailed paired T-test; P-value = 0.005). For these
+full taxa models using stool, \emph{Bacteroides} and
+\emph{Lachnospiraceae} were the most common taxa in the top 10\% mean
+decrease in accuracy (MDA) across studies {[}Figure S3{]}. Similarly,
 using the data from unmatched tissue samples we found that the AUC of
@@
 rather than the 3 taxa with significant ORs (one-tailed paired T-test;
-P-value = 0.03). Although the significant OR taxa models can classify
-individuals with and without carcinoma tumors, they are still missing
-taxa from the full community models that can increase the model
-accuracy.
+P-value = 0.03). For the full taxa models using unmatched tissue,
+\emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
+\emph{Ruminococcaceae} were the most common taxa in the top 10\% mean
+decrease in accuracy across studies {[}Figure S4{]}. Although the
+significant OR taxa models can classify individuals with and without
+carcinoma tumors, they are still missing taxa from the full community
+models that can increase the model accuracy.
 

### 2018-03-13 submission/manuscript.Rmd
update supplemental fig order

@@
 
-**Figure S2: Most Common Taxa Across Carcinoma Full Community Stool Study Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest all taxa-based models.
+**Figure S2: Tissue Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
+B) Carcinoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
+The black line represents the median AUC for the respective group.
+If no values are present in the singificant OR taxa group then there were no significant taxa identified and no model was tested.
+
+**Figure S3: Most Common Taxa Across Carcinoma Full Community Stool Study Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest all taxa-based models.
 B) Common taxa in the top 10 percent for carcinoma Random Forest all OTU-based models.
@@
 
-**Figure S3: Most Common Genera Across Full Community Tissue Study Models.** A) Common genera in the top 10 percent for matched carcinoma Random Forest all genera-based models.
+**Figure S4: Most Common Genera Across Full Community Tissue Study Models.** A) Common genera in the top 10 percent for matched carcinoma Random Forest all genera-based models.
 B) Common genera in the top 10 percent for unmatched carcinoma Random Forest all genera-based models.
@@
 White means that particular taxa was not in the top 10%.
-
-**Figure S4: Tissue Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
-B) Carcinoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
-The black line represents the median AUC for the respective group.
-If no values are present in the singificant OR taxa group then there were no significant taxa identified and no model was tested.
 

### 2018-03-13 submission/manuscript.tex
update supplemental fig order

@@
 
-\textbf{Figure S2: Most Common Taxa Across Carcinoma Full Community
+\textbf{Figure S2: Tissue Random Forest Model Train AUCs.} A) Adenoma
+random forest model AUCs between all genera, all OTU, and select model
+based on significant OR taxa in unmatched and matched tissue. B)
+Carcinoma random forest model AUCs between all genera, all OTU, and
+select model based on significant OR taxa in unmatched and matched
+tissue. The black line represents the median AUC for the respective
+group. If no values are present in the singificant OR taxa group then
+there were no significant taxa identified and no model was tested.
+
+\textbf{Figure S3: Most Common Taxa Across Carcinoma Full Community
 Stool Study Models.} A) Common taxa in the top 10 percent for carcinoma
@@
 
-\textbf{Figure S3: Most Common Genera Across Full Community Tissue Study
+\textbf{Figure S4: Most Common Genera Across Full Community Tissue Study
 Models.} A) Common genera in the top 10 percent for matched carcinoma
@@
 
-\textbf{Figure S4: Tissue Random Forest Model Train AUCs.} A) Adenoma
-random forest model AUCs between all genera, all OTU, and select model
-based on significant OR taxa in unmatched and matched tissue. B)
-Carcinoma random forest model AUCs between all genera, all OTU, and
-select model based on significant OR taxa in unmatched and matched
-tissue. The black line represents the median AUC for the respective
-group. If no values are present in the singificant OR taxa group then
-there were no significant taxa identified and no model was tested.
-
 \textbf{Figure S5: Tissue Random Forest Genus-Based Model Test AUCs.} A)

### 2018-03-13 submission/manuscript.Rmd
add pat's comments on abstract

@@
 
-**Background.** An increasing body of literature suggests that both individual and collections of bacteria are associated with the progression of colorectal cancer.
+An increasing body of literature suggests that both individual and collections of bacteria are associated with the progression of colorectal cancer.
 As the number of studies investigating these associations increases and the number of subjects in each study increases, a meta-analysis to identify the associations that are the most predictive of disease progression is warranted.
-For our meta-analysis, we analyzed previously published 16S rRNA gene sequencing data collected from feces (`r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` individuals from `r length(stool_n$study)` studies) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals from `r length(tissue_n$study)` studies).
+We analyzed previously published 16S rRNA gene sequencing data collected from feces and colon tissue.
 
-**Results.** We quantified the odds ratios for individual bacterial taxa that were associated with an individual having tumors relative to a normal colon.
-Among the stool samples, there were no taxa that had a significant odds ratio associated with adenoma and there were `r length(crc_OR_stool$measure)` taxa with significant odds ratios (ORs) associated with carcinoma.
-Similarly, among the tissue samples, there were no taxa that had a significant odds ratio associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` taxa with significant odds ratios associated with carcinoma.
-Among the significant odds ratios, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`.
+We quantified the odds ratios (ORs) for individual bacterial taxa that were associated with an individual having tumors relative to a normal colon.
+Among the fecal samples, there were no taxa that had a significant ORs associated with adenoma and there were `r length(crc_OR_stool$measure)` taxa with significant ORs associated with carcinoma.
+Similarly, among the tissue samples, there were no taxa that had a significant OR associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs associated with carcinoma.
+Among the significant ORs, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`.
 Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using only the taxa that had significant ORs, using the entire collection of taxa found in each study, and using operational taxonomic units defined based on a 97% similarity threshold.
 All training approaches yielded similar classification success as measured using the Area Under the Curve.
-The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from stool or tissue.
+The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from fecal or tissue.
 
@@
 
-**Importance.** This meta-analysis confirms previous results indicating that individuals with adenomas cannot be readily classified based on their bacterial community, but that those with carcinomas can.
-Regardless of the dataset, we found a subset of the fecal community that was associated with carcinomas was as predictive as the full community.
+## Importance
+
+Colorectal cancer is a significant and growing health problem in which animal models and epidemiological data suggest that the colonic microbiota have a role in tumorigenesis.
+These observations indicate that the colonic microbiota is a reservoir of biomarkers that may improve our ability to detect colonic tumors using non-invasive approaches.
+This meta-analysis identifies and validates a set of 8 bacterial taxa that can be used within a Random Forest modeling framework to differentiate individuals as having normal colons or carcinomas.
+When models trained using one dataset were tested on other datasets, the models performed well.
+These results lend support to the use of fecal biomarkers for the detection of tumors.
+Furthermore, these biomarkers are plausible candidates for further mechanistic studies into the role of the gut microbiota in tumorigenesis.
 

### 2018-03-13 submission/manuscript.Rmd
update with pat's suggestions for intro

@@
 Numerous case-control studies have characterized the microbiota of individuals with colonic adenomas and carcinomas in an attempt to identify biomarkers of disease progression [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
-Because current CRC screening recommendations are poorly adhered to due to socioeconomic status, test invasiveness, and frequency of tests, development and validation of microbiome-associated biomarkers for CRC progression could further attempts to develop non-invasive diagnostics [@GimenoGarca2012].
+Because current CRC screening recommendations are poorly adhered to due to person's socioeconomic status, test invasiveness, and frequency of tests, development and validation of microbiota-associated biomarkers for CRC progression could further attempts to develop non-invasive diagnostics [@GimenoGarca2012].
 
 Recently, there has been an intense focus on identifying microbiota-based biomarkers yielding a seemingly endless number of candidate taxa.
-Some studies point towards mouth-associated genera such as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that are enriched in people with carcinomas [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
-Other studies have identified members of *Akkermansia*, *Bacteroides*, *Enterococcus*, *Escherichia*, *Klebsiella*, *Mogibacterium*, *Streptococcus*, and *Providencia* are also associated with carciomas  [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012].
-Additionally, *Roseburia* has been found in some studies to be more abundant in people with tumors but in other studies it has been found to be either less abundant or no different than what is found in subjects with normal colons  [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
-There are strong results from tissue culture and murine models that *Fusobacterium nucleatum*, pks-positive strains of *Escherichia coli*, *Streptococcus gallolyticus*, and an entertoxin-producing strain of *Bacteroides fragilis* are important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017; @weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014].
-These results point to a causative role for the microbiota in CRC pathogenesis as well as their potential as diagnostic biomarkers.
+Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* that are enriched in people with carcinomas [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
+Other studies have identified members of *Akkermansia*, *Bacteroides*, *Enterococcus*, *Escherichia*, *Klebsiella*, *Mogibacterium*, *Streptococcus*, and *Providencia* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012].
+Additionally, *Roseburia* has been found in some studies to be more abundant in people with tumors but in other studies it has been found to be less abundant than what is found in subjects with normal colons  [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].
+There is support from mechanistic studies using tissue culture and murine models that *Fusobacterium nucleatum*, pks-positive strains of *Escherichia coli*, *Streptococcus gallolyticus*, and an entertoxin-producing strain of *Bacteroides fragilis* are important in tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017; @weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014].
+These results point to a causative role for the microbiota in tumorigenesis as well as their potential as diagnostic biomarkers.
 
-Most studies have focused on identifying biomarkers in patients with carcinomas but there is a greater clinical need to identify biomarkers associated with adenomas.
-Studies focusing on broad scale community metrics have found that measures such as the total number of Operational Taxonomic Units (OTUs) are decreased in those with adenomas versus controls [@sanapareddy_increased_2012].
+Most studies have focused on identifying biomarkers in patients with carcinomas but there is a clinical need to identify biomarkers associated with adenomas to facilitate early detection of the tumors.
+Studies focusing on broad scale community metrics have found that measures such as the total number of taxa (i.e.
+richness) are lower in those with adenomas versus controls [@sanapareddy_increased_2012].
 Other studies have identified *Acidovorax*, *Bilophila*, *Cloacibacterium*, *Desulfovibrio*, *Helicobacter*, *Lactobacillus*, *Lactococcus*, *Mogibacterium*, and *Pseudomonas* to be enriched in those with adenomas [@sanapareddy_increased_2012; @lu_mucosal_2016; @hale_shifts_2017].
-There are few genera that are enriched in patients with adenoma or carcinoma tumors.
+The ability to classify individuals as having normal colons or adenomas based solely on the taxa within fecal samples has been limited.
+However, when 16S rRNA gene sequence data was combined with the results of a fecal immunochemical test (FIT), the ability to diagnose individuals with adenomas was improved relative to using the FIT results alone [@baxter_microbiota-based_2016].
 
-Confirming some of these previous findings, a recent meta-analysis found that 16S rRNA gene sequences from members of  *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017].
+A recent meta-analysis found that 16S rRNA gene sequences from members of  *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017].
 Contrary to previous studies they found sequences similar to members of *Lactobacillus* and *Ruminococcus* to be enriched in patients with adenoma or carcinoma relative to those with normal colons [@baxter_microbiota-based_2016; @burns_virulence_2015; @zeller_potential_2014].
-In addition, they found 16S rRNA gene sequences from members of *Haemophilus*, *Methanosphaera*, *Prevotella*, *Succinovibrio* were enriched in patients with adenoma and *Pantoea* were enriched in patients with carcinomas.
-Although this meta-analysis was helpful for distilling a large number of possible biomarkers, the aggregate number of samples included in the analysis (n = 509) was smaller than several larger case-control studies that have since been published [@baxter_microbiota-based_2016; @hale_shifts_2017]
+In addition, they found that 16S rRNA gene sequences from members of *Haemophilus*, *Methanosphaera*, *Prevotella*, and *Succinovibrio* were enriched in patients with adenomas and *Pantoea* were enriched in patients with carcinomas.
+Although this meta-analysis was helpful for distilling a large number of possible biomarkers, the aggregate number of samples included in the analysis (n=509) was smaller than several larger case-control studies that have since been published [@baxter_microbiota-based_2016; @hale_shifts_2017]
 
-Here we provide an updated meta-analysis using 16S rRNA gene sequence data from both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) from 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2].
+Here we provide an updated meta-analysis using 16S rRNA gene sequence data from both feces (n=`r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) from 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2].
 We expand both the breadth and scope of the previous meta-analysis to investigate whether biomarkers describing the bacterial community or specific members of the community can more accurately classify patients as having adenoma or carcinoma.

### 2018-03-13 submission/manuscript.Rmd
add pat's result edits

@@
 
-***Lower bacterial diversity is associated with increased odds ratio (OR) of tumors.***
+***Lower bacterial diversity is associated with higher odds ratio (OR) of tumors.***
 We first assessed whether variation in broad community metrics like total number of operational taxonomic units (OTUs) (i.e.
-richness), the evenness of their abundance, and the overall diversity was associated with disease stage after controlling for study and variable region differences.
-In stool, there was a significant decrease in both evenness and diversity as disease severity progressed from normal to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1]; there was no significant difference for richness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`).
-We next tested whether the decrease in these community metrics translated into significant ORs for having an adenoma or carcinoma.
-For fecal samples, the ORs for richness were not significantly greater than 1.0 for adenoma or carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "sobs")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A].
-The ORs for evenness were significantly higher than 1.0 for adenoma (OR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`)(95% Confidence Interval), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and carcinoma (OR =  `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2B].
-The ORs for diversity were only significantly greater than 1.0 for carcinoma (OR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`), but not for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2C].
-Although these OR are significantly greater than 1.0, it is doubtful that these are clinically meaningful values.
+richness), the evenness of their abundance, and the overall diversity of the communities were associated with disease stage after controlling for study and variable region differences.
+In fecal samples, both evenness and diversity were significantly lower in successive disease severity categories (evenness P-value=`r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and diversity P-value=`r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`) [Figure 1]; there was no significant difference for richness (P-value=`r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`).
+We next tested whether the lower value of these community metrics translated into significant ORs for having an adenoma or carcinoma.
+For fecal samples, the ORs for richness were not significantly greater than 1.0 for adenoma or carcinoma (P-value=`r paste(format(filter(alpha_RR_crc_stool, measure == "sobs")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A].
+The ORs for evenness were significantly higher than 1.0 for adenoma (OR=`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (95% Confidence Interval: `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value=`r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and carcinoma (OR=`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value=`r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2B].
+The ORs for diversity were only significantly greater than 1.0 for carcinoma (OR=`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value=`r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`), but not for adenoma (P-value=`r paste(format(filter(alpha_RR_adn_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2C].
+Although these ORs are significantly greater than 1.0, it is doubtful that they are clinically meaningful.
 
-Similar to our analysis of sequences obtained from stool samples, we repeated the analysis using sequences obtained from colon tissue.
-There were no significant changes in richness, evenness, or diversity as disease severity progressed from control to adenoma to carcinoma (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
-We next analyzed the OR, for matched (i.e.
+Similar to our analysis of sequences obtained from fecal samples, we repeated the analysis using sequences obtained from colon tissue.
+There were no significant differences in richness, evenness, or diversity as disease severity progressed from control to adenoma to carcinoma (P-values `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05).
+We next analyzed the ORs, for matched (i.e.
 where unaffected tissue and tumors were obtained from the same individual) and unmatched (i.e.
 where unaffected tissue and tumor tissue were not obtained from the same individual) tissue samples.
-The ORs for adenoma and carcinoma by any measure were not significantly different from 1.0 (P-value `r ifelse(min(alpha_RR_crc_unmatch_tissue$pvalue) < 0.05 & min(alpha_RR_crc_match_tissue$pvalue) < 0.05 & min(alpha_RR_adn_tissue$pvalue) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure S1 & Table S1].
-This is likely due to the combination of a small effect size, as suggested from the results using stool, and the relatively small number of studies and size of studies used in the analysis.
+The ORs for adenoma and carcinoma were not significantly different from 1.0 for any measure (P-values `r ifelse(min(alpha_RR_crc_unmatch_tissue$pvalue) < 0.05 & min(alpha_RR_crc_match_tissue$pvalue) < 0.05 & min(alpha_RR_adn_tissue$pvalue) < 0.05, invisible("check tables"), invisible(">"))` 0.05) [Figure S1 & Table S1].
+This is likely due to the combination of a small effect size and the relatively small number of studies and size of studies used in the analysis.
 
-***Disease progression is associated with community-wide changes in composition and abundance.*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages.
-We identified significant bacterial community differences in the stool of patients with adenomas relative to those with normal colons in `r length(as.data.frame(filter(adn_beta_stool, pvalue < 0.05))[, "study"])` of `r length(as.data.frame(adn_beta_stool)[, "study"])` studies and in patients with carcinomas relative to those with normal colons in 6 of 7 studies (PERMANOVA; P-value < 0.05) [Table S2].
-Similar to the analyses using stool samples, there were significant differences in bacterial community structure between subjects with normal colons and those with adenoma (1 of 2 studies) and carcinoma (1 of 3 studies) [Table S2].
-For studies that used matched samples no differences in bacterial community structures were observed [Table S2].
-Combined, these results indicate that there are consistent and significant community-wide changes in the fecal community structure of subjects with carcinomas.
+***Disease progression is associated with changes in community structure.*** Based on the differences in evenness and diversity, we next asked whether there were community-wide differences in the structure of the communities associated with different disease stages.
+We identified significant bacterial community differences in the feces of patients with adenomas relative to those with normal colons in `r length(as.data.frame(filter(adn_beta_stool, pvalue < 0.05))[, "study"])` of `r length(as.data.frame(adn_beta_stool)[, "study"])` studies and in patients with carcinomas relative to those with normal colons in 6 of 7 studies (PERMANOVA; P-value < 0.05) [Table S2].
+Similar to the analyses using fecal samples, there were significant differences in the bacterial community structure of subjects with normal colons and those with adenomas (1 of 2 studies) and carcinomas (1 of 3 studies) [Table S2].
+For studies that used matched samples, we did not observe any differences in bacterial community structures [Table S2].
+Combined, these results indicate that there were consistent and significant community-wide changes in the fecal community structure of subjects with carcinomas.
 However, the signal observed in subjects with adenomas or when using tissue samples was not as consistent.
@@
 ***Individual taxa are associated with significant ORs for carcinomas.*** 
-Next we identified those taxa that were associated with ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas.
-No taxa had a significant OR for the presence of adenomas when we used data collected from stool or tissue samples (Table S3 & S4).
-In contrast, `r length(crc_OR_stool$measure)` taxa had significant ORs for the presence of carcinomas using data from stool samples.
-Of these, 4 are commonly associated with the oral cavity: *Fusobacterium* (OR = `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`)), *Parvimonas* (OR = `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_ub"], digits = 2))`)), *Porphyromonas* (OR = `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2))`)), and *Peptostreptococcus* (OR = `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_ub"], digits = 2))`)) [Table S3].
-The other 4 were *Clostridium XI* (OR = `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_ub"], digits = 2))`)), *Enterobacteriaceae*  (OR = `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_ub"], digits = 2))`)), *Escherichia* (OR = `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_ub"], digits = 2))`)), and *Ruminococcus* (OR = `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_ub"], digits = 2))`)).
+Next we identified those taxa that had ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas.
+No taxa had a significant OR for the presence of adenomas when we used data collected from fecal or tissue samples (Table S3 & S4).
+In contrast, `r length(crc_OR_stool$measure)` taxa had significant ORs for the presence of carcinomas using data from fecal samples.
+Of these, 4 are commonly associated with the oral cavity: *Fusobacterium* (OR=`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`)), *Parvimonas* (OR=`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_ub"], digits = 2))`)), *Porphyromonas* (OR=`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2))`)), and *Peptostreptococcus* (OR=`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_ub"], digits = 2))`)) [Table S3].
+The other 4 were *Clostridium XI* (OR=`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_ub"], digits = 2))`)), *Enterobacteriaceae*  (OR=`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_ub"], digits = 2))`)), *Escherichia* (OR = `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_ub"], digits = 2))`)), and *Ruminococcus* (OR=`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_ub"], digits = 2))`)).
 Among the data collected from tissue samples, only unmatched carcinoma samples had taxa with a significant OR.
-Those included *Dorea* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2))`)), *Blautia* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2))`)), and *Weissella* (OR = `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2))`)).
-Mouth-associated genera were not significantly associated with an increased OR for carcinoma in tissue samples [Table S4].
+Those included *Dorea* (OR=`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Dorea")[, "ci_ub"], digits = 2))`)), *Blautia* (OR=`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Blautia")[, "ci_ub"], digits = 2))`)), and *Weissella* (OR=`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_unmatched_OR_tissue, measure == "Weissella")[, "ci_ub"], digits = 2))`)).
+Mouth-associated genera were not significantly associated with a higher OR for carcinoma in tissue samples [Table S4].
 For example, *Fusobacterium* had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`; however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4].
-It is interesting to note that *Ruminococcus* and members of *Clostridium XI* in stool and *Dorea* and *Blautia* in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas.
-Overall, there was no overlap in the taxa with significant OR between stool and tissue samples.
+It is interesting to note that *Ruminococcus* and members of *Clostridium XI* in fecal samples and *Dorea* and *Blautia* in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas.
+Overall, there was no overlap in the taxa with significant OR between fecal and tissue samples.
 
@@
 To measure the ability of these taxa to classify individuals we instead generated receiver operator characteristic (ROC) curves for each taxon in each study and calculated the area under the curve (AUC).
-This allowed us to use a more fluid relative abundance threshold for defining disease status.
-Using data from stool samples, the `r length(crc_OR_stool$measure)` taxa did no better at classifying the subjects than one would expect by chance (i.e.
+This allowed us to use a more fluid relative abundance threshold for classifying individuals by their disease status.
+Using data from fecal samples, the `r length(crc_OR_stool$measure)` taxa did no better at classifying the subjects than one would expect by chance (i.e.
 AUC=0.50) [Figure 3A].
-The taxa that performed the best included *Clostridium XI*, *Ruminococcus*, and *Escherichia*, however, these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)`.
-Likewise, in unmatched tissue samples the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant OR taxa were marginally better than one would expect by chance [Figure 3B].
+The taxa that performed the best included *Clostridium XI*, *Ruminococcus*, and *Escherichia*.
+However, these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)` ndicating their limited value as biomarkers when used individually.
+Likewise, in unmatched tissue samples the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant OR taxa had AUC values that were marginally better than one would expect by chance [Figure 3B].
 The relative abundance of *Dorea* was the best predictor of carcinomas and its median AUC was only `r round(max(filter(ind_crc_combined_unmatched_auc, taxa == "Dorea")[, "auc"]), digits = 3)`.
-These results suggest that although these taxa are associated with a decreased or increased OR for the presences of carcinomas, individually, they do a poor job of classifying a subject’s disease status.
+These results suggest that although these taxa are associated with a significant OR for the presences of carcinomas, they do a poor job of classifying a subject’s disease status when used individually.
 
@@
 Instead of attempting to classify subjects based on individual taxa, next we combined information from the individual taxa and evaluated the ability to classify a subject’s disease status using Random Forest models.
-For data from stool samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`).
-Within the significant OR taxa only models the most important taxa across study for stool were *Ruminococcus* and *Clostridium XI* based on mean decrease in accuracy [Figure 4A].
-Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for *Blautia* and *Weissella* (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).
-For the significant OR taxa unmatched tissue models both *Dorea* and *Blautia* were the most important based on mean decrease in accuracy [Figure 4B].
-Pooling the information from the taxa with significant ORs results in a model that outperforms classifications made using individual taxa.
+For data from fecal samples, the combined model had an AUC of `r round(median(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"]), digits = 2)`, which was significantly higher than any of the AUC values for the individual taxa (P-value < `r round(max(stool_ind_vs_full_taxa$pvalue), digits = 3)`).
+When this approach was used to train models using data from each study, the most important taxa were *Ruminococcus* and *Clostridium XI* [Figure 4A].
+Similarly, using data from the unmatched tissue samples, the combined model had an AUC of `r round(median(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"]), digits = 2)`, which was significantly higher than the AUC values for classifying based on the relative abundances of *Blautia* and *Weissella* individually (P-value < `r round(max(filter(unmatched_tissue_ind_vs_full_taxa, taxa != "Dorea")[, "pvalue"]), digits = 3)`).
+Both *Dorea* and *Blautia* were the most important taxa in the tissue-based models [Figure 4B].
+Pooling the information from the taxa with significant ORs resulted in models that outperformed classifications made using the same taxa individually.
 
-***Performance of models based on taxa relative abundance in full community are better than those based on taxa with significant ORs.*** 
+***Performance of models based on taxa relative abundance in full community is better than that of models based on taxa with significant ORs.*** 
 Next, we asked whether a Random Forest classification model built using all of the taxa found in the communities would outperform the models generated using those taxa with a significant OR.
-Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from stool or tissue samples were only marginally better than 0.5 for any study (median AUC = `r round(median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` (`r round(min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` - `r round(max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)`)(min-max)) [Figure 5A & S2A].
-In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from stool or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 5B & S2B-C].
+Similar to our inability to identify taxa associated with a significant OR for the presence of adenomas, the median AUCs to classify subjects as having normal colons or having adenomas using data from fecal or tissue samples were only marginally better than 0.5 for any study (median AUC=`r round(median(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` (range: `r round(min(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)` - `r round(max(c(filter(adn_stool_model_aucs, study == train_model)[, "AUC"], filter(adn_tissue_model_aucs, study == train_model)[, "AUC"]), na.rm = T), digits = 3)`)) [Figure 5A & S2A].
+In contrast, the models for classifying subjects as having normal colons or having carcinomas using data from fecal or tissue samples yielded AUC values meaningfully higher than 0.5 [Figure 5B & S2B-C].
 When we compared the models based on all of the taxa in a community to models based on the taxa with significant ORs, the results were mixed.
-Using the data from stool samples we found that the AUC for 6 of 7 studies increased by an average of `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"]))*100), digits = 2))`%) and AUC for the Flemer study decreased by `r round(abs(mean(((1 - filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"]))*100)), digits = 2)`%).
-The overall improvement in performance was statistically significant (mean = `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"]))*100), digits = 2))`%, one-tailed paired T-test; P-value = `r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
-For these full taxa models using stool, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy (MDA) across studies [Figure S3].
-Similarly, using the data from unmatched tissue samples we found that the AUC of studies increased by an average of `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`% when we used all of the taxa rather than the 3 taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
-For the full taxa models using unmatched tissue, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S4].
-Although the significant OR taxa models can classify individuals with and without carcinoma tumors, they are still missing taxa from the full community models that can increase the model accuracy.
+Using the data from fecal samples, we found that the AUC for 6 of 7 studies were an average of `r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study != "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study != "flemer")[, "AUC"]))*100), digits = 2))`% higher and AUC for the Flemer study was `r round(abs(mean(((1 - filter(crc_stool_model_aucs, study == train_model, model_type == "full_model", study == "flemer")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model", study == "flemer")[, "AUC"]))*100)), digits = 2)`% lower when using the relative abundance data from all taxa relative to using the relative abundance of only the taxa with significant ORs.
+The overall improvement in performance was statistically significant (mean=`r abs(round(mean((1 - (filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"] / filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"]))*100), digits = 2))`%, one-tailed paired T-test; P-value=`r round(t.test(filter(crc_stool_model_aucs, study == train_model, model_type == "full_model")[, "AUC"], filter(crc_stool_model_aucs, study == train_model, model_type == "select_model")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
+Among the models trained using data from fecal samples, *Bacteroides* and *Lachnospiraceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S3].
+Using data from unmatched tissue samples to train classification models, we found that the AUC of studies was an average `r round(mean((filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"] - filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"])*100), digits = 2)`% higher when we used all of the taxa rather than the 3 taxa with significant ORs (one-tailed paired T-test; P-value=`r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`).
+For the models trained using data from unmatched tissue samples, *Lachnospiraceae*, *Bacteroidaceae*, and *Ruminococcaceae* were the most common taxa in the top 10% mean decrease in accuracy across studies [Figure S4].
+Although the models trained using those taxa with a significant OR perform well for classifying individuals with and without carcinomas, models trained using data from the full community perform better.
 
-***Performance of models based on OTU relative abundance in full community are not significantly better than those based on taxa with significant ORs.*** 
+***Performance of models based on OTU relative abundances are not significantly better than those based on taxa with significant ORs.*** 
 The previous models were based on relative abundance data where sequences were classified to coarse taxonomic assignments (i.e.
@@
 To determine whether model performance improved with finer scale classification, we assigned sequences to operational taxonomic units (OTUs) where the similarity among sequences within an OTU was more than 97%.
-We again found that classification models built using all of the sequence data for a community did a poor job of differentiating between subjects with normal colons and those with adenomas (median AUC: `r round(median(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` (`r round(min(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`)), but did a good job of differentiating between subjects with normal colons and those with carcinomas (median AUC: `r round(median(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` (`r round(min(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`)).
-The OTU-based models performed similarly to those constructed using the taxa with significant ORs (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) and those using all taxa (one-tailed paired T-test; P-value = `r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) [Figure 4].
+We again found that classification models built using all of the sequence data for a community did a poor job of differentiating between subjects with normal colons and those with adenomas (median AUC: `r round(median(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` (`r round(min(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`- `r round(max(filter(adn_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)`)).
+However, they did a good job of differentiating between subjects with normal colons and those with carcinomas (median AUC: `r round(median(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), digits = 2)` (`r format(min(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), nsmall = 2, digits = 2)`- `r format(max(filter(crc_stool_model_aucs, model_type == "full_otu")[, "AUC"]), nsmall = 2, digits = 2)`)).
+The OTU-based models performed similarly to those constructed using the taxa with significant ORs (one-tailed paired T-test; P-value=`r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "select_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) and those using all taxa (one-tailed paired T-test; P-value=`r round(t.test(filter(crc_tissue_model_aucs, study == train_model, model_type == "full_model", type == "unmatched")[, "AUC"], filter(crc_tissue_model_aucs, study == train_model, model_type == "full_otu", type == "unmatched")[, "AUC"], alternative = "greater", paired = TRUE)$p.value, digits = 3)`) [Figure 4].
 Among the OTUs that had the highest mean decrease in accuracy for the OTU-based models, we found that OTUs that affiliated with all of the `r length(crc_OR_stool$measure)` taxa that had a significant OR were within the top 10% for at least one study.
@@
 ***Generalizability of taxon-based models trained on one dataset to the other datasets.*** 
-Considering the good performance of the Random Forest models using taxa with a significant OR and using all of the taxa, we next asked how well the models would perform when given data from a different subject cohort.
+Considering the good performance of the Random Forest models trained using the relative abundance of taxa with significant ORs and models trained using the relative abundance of all taxa, we next asked how well the models would perform when given data from a different cohort.
 For instance, if a model was trained using data from the Ahn study, we wanted to know how well it would perform using the data from the Baxter study.
-We found the models trained using the taxa with a significant OR all had a higher median AUC than the models trained using all of the taxa when tested on the other datasets [Figure 6 & S5].
-As might be expected, the difference between the performance of the modelling approaches appeared to vary with the size of the training cohort (R^2^ = `r round(lm(auc_median ~ totals, data = correlation_data)$coefficients[1], digits = 2)`) [Figure 6].
+The models trained using the taxa with significant ORs all had a higher median AUC than the models trained using all of the taxa when tested on the other datasets [Figure 6 & S5].
+As might be expected, the difference between the performance of the modeling approaches appeared to vary with the size of the training cohort (R^2^=`r round(lm(auc_median ~ totals, data = correlation_data)$coefficients[1], digits = 2)`) [Figure 6].
 These data suggest that given a sufficient number of subjects with normal colons and carcinomas, Random Forest models trained using a small number of taxa can accurately classify individuals from a different cohort.

### 2018-03-13 submission/manuscript.tex
add pat's result edits

@@
 
-\textbf{Background.} An increasing body of literature suggests that both
-individual and collections of bacteria are associated with the
-progression of colorectal cancer. As the number of studies investigating
-these associations increases and the number of subjects in each study
+An increasing body of literature suggests that both individual and
+collections of bacteria are associated with the progression of
+colorectal cancer. As the number of studies investigating these
+associations increases and the number of subjects in each study
 increases, a meta-analysis to identify the associations that are the
-most predictive of disease progression is warranted. For our
-meta-analysis, we analyzed previously published 16S rRNA gene sequencing
-data collected from feces (1737 individuals from 8 studies) and colon
-tissue (492 total samples from 350 individuals from 7 studies).
-
-\textbf{Results.} We quantified the odds ratios for individual bacterial
-taxa that were associated with an individual having tumors relative to a
-normal colon. Among the stool samples, there were no taxa that had a
-significant odds ratio associated with adenoma and there were 8 taxa
-with significant odds ratios (ORs) associated with carcinoma. Similarly,
-among the tissue samples, there were no taxa that had a significant odds
-ratio associated with adenoma and there were 3 taxa with significant
-odds ratios associated with carcinoma. Among the significant odds
-ratios, the association between individual taxa and tumor diagnosis was
-equal or below 7.11. Because individual taxa had limited association
-with tumor diagnosis, we trained Random Forest classification models
-using only the taxa that had significant ORs, using the entire
-collection of taxa found in each study, and using operational taxonomic
-units defined based on a 97\% similarity threshold. All training
-approaches yielded similar classification success as measured using the
-Area Under the Curve. The ability to correctly classify individuals with
-adenomas was poor and the ability to classify individuals with
-carcinomas was considerably better using sequences from stool or tissue.
+most predictive of disease progression is warranted. We analyzed
+previously published 16S rRNA gene sequencing data collected from feces
+and colon tissue.
+
+We quantified the odds ratios (ORs) for individual bacterial taxa that
+were associated with an individual having tumors relative to a normal
+colon. Among the fecal samples, there were no taxa that had a
+significant ORs associated with adenoma and there were 8 taxa with
+significant ORs associated with carcinoma. Similarly, among the tissue
+samples, there were no taxa that had a significant OR associated with
+adenoma and there were 3 taxa with significant ORs associated with
+carcinoma. Among the significant ORs, the association between individual
+taxa and tumor diagnosis was equal or below 7.11. Because individual
+taxa had limited association with tumor diagnosis, we trained Random
+Forest classification models using only the taxa that had significant
+ORs, using the entire collection of taxa found in each study, and using
+operational taxonomic units defined based on a 97\% similarity
+threshold. All training approaches yielded similar classification
+success as measured using the Area Under the Curve. The ability to
+correctly classify individuals with adenomas was poor and the ability to
+classify individuals with carcinomas was considerably better using
+sequences from fecal or tissue.
 
@@
 
-\textbf{Importance.} This meta-analysis confirms previous results
-indicating that individuals with adenomas cannot be readily classified
-based on their bacterial community, but that those with carcinomas can.
-Regardless of the dataset, we found a subset of the fecal community that
-was associated with carcinomas was as predictive as the full community.
+\subsection{Importance}\label{importance}
+
+Colorectal cancer is a significant and growing health problem in which
+animal models and epidemiological data suggest that the colonic
+microbiota have a role in tumorigenesis. These observations indicate
+that the colonic microbiota is a reservoir of biomarkers that may
+improve our ability to detect colonic tumors using non-invasive
+approaches. This meta-analysis identifies and validates a set of 8
+bacterial taxa that can be used within a Random Forest modeling
+framework to differentiate individuals as having normal colons or
+carcinomas. When models trained using one dataset were tested on other
+datasets, the models performed well. These results lend support to the
+use of fecal biomarkers for the detection of tumors. Furthermore, these
+biomarkers are plausible candidates for further mechanistic studies into
+the role of the gut microbiota in tumorigenesis.
 
@@
 (6, 11--17). Because current CRC screening recommendations are poorly
-adhered to due to socioeconomic status, test invasiveness, and frequency
-of tests, development and validation of microbiome-associated biomarkers
-for CRC progression could further attempts to develop non-invasive
-diagnostics (18).
+adhered to due to person's socioeconomic status, test invasiveness, and
+frequency of tests, development and validation of microbiota-associated
+biomarkers for CRC progression could further attempts to develop
+non-invasive diagnostics (18).
 
@@
 candidate taxa. Some studies point towards mouth-associated genera such
-as Fusobacterium, Peptostreptococcus, Parvimonas, and Porphyromonas that
-are enriched in people with carcinomas (6, 11--17). Other studies have
-identified members of \emph{Akkermansia}, \emph{Bacteroides},
-\emph{Enterococcus}, \emph{Escherichia}, \emph{Klebsiella},
-\emph{Mogibacterium}, \emph{Streptococcus}, and \emph{Providencia} are
-also associated with carciomas (13--15). Additionally, \emph{Roseburia}
-has been found in some studies to be more abundant in people with tumors
-but in other studies it has been found to be either less abundant or no
-different than what is found in subjects with normal colons (14, 17, 19,
-20). There are strong results from tissue culture and murine models that
+as \emph{Fusobacterium}, \emph{Peptostreptococcus}, \emph{Parvimonas},
+and \emph{Porphyromonas} that are enriched in people with carcinomas (6,
+11--17). Other studies have identified members of \emph{Akkermansia},
+\emph{Bacteroides}, \emph{Enterococcus}, \emph{Escherichia},
+\emph{Klebsiella}, \emph{Mogibacterium}, \emph{Streptococcus}, and
+\emph{Providencia} (13--15). Additionally, \emph{Roseburia} has been
+found in some studies to be more abundant in people with tumors but in
+other studies it has been found to be less abundant than what is found
+in subjects with normal colons (14, 17, 19, 20). There is support from
+mechanistic studies using tissue culture and murine models that
 \emph{Fusobacterium nucleatum}, pks-positive strains of
@@
 entertoxin-producing strain of \emph{Bacteroides fragilis} are important
-in the pathogenesis of CRC (5, 14, 21--24). These results point to a
-causative role for the microbiota in CRC pathogenesis as well as their
-potential as diagnostic biomarkers.
+in tumorigenesis (5, 14, 21--24). These results point to a causative
+role for the microbiota in tumorigenesis as well as their potential as
+diagnostic biomarkers.
 
 Most studies have focused on identifying biomarkers in patients with
-carcinomas but there is a greater clinical need to identify biomarkers
-associated with adenomas. Studies focusing on broad scale community
-metrics have found that measures such as the total number of Operational
-Taxonomic Units (OTUs) are decreased in those with adenomas versus
-controls (25). Other studies have identified \emph{Acidovorax},
-\emph{Bilophila}, \emph{Cloacibacterium}, \emph{Desulfovibrio},
-\emph{Helicobacter}, \emph{Lactobacillus}, \emph{Lactococcus},
-\emph{Mogibacterium}, and \emph{Pseudomonas} to be enriched in those
-with adenomas (25--27). There are few genera that are enriched in
-patients with adenoma or carcinoma tumors.
-
-Confirming some of these previous findings, a recent meta-analysis found
-that 16S rRNA gene sequences from members of \emph{Akkermansia},
-\emph{Fusobacterium}, and \emph{Parvimonas} were fecal biomarkers for
-the presence of carcinomas (28). Contrary to previous studies they found
-sequences similar to members of \emph{Lactobacillus} and
-\emph{Ruminococcus} to be enriched in patients with adenoma or carcinoma
-relative to those with normal colons (12, 15, 16). In addition, they
-found 16S rRNA gene sequences from members of \emph{Haemophilus},
-\emph{Methanosphaera}, \emph{Prevotella}, \emph{Succinovibrio} were
-enriched in patients with adenoma and \emph{Pantoea} were enriched in
-patients with carcinomas. Although this meta-analysis was helpful for
-distilling a large number of possible biomarkers, the aggregate number
-of samples included in the analysis (n = 509) was smaller than several
-larger case-control studies that have since been published (12, 27)
+carcinomas but there is a clinical need to identify biomarkers
+associated with adenomas to facilitate early detection of the tumors.
+Studies focusing on broad scale community metrics have found that
+measures such as the total number of taxa (i.e.~richness) are lower in
+those with adenomas versus controls (25). Other studies have identified
+\emph{Acidovorax}, \emph{Bilophila}, \emph{Cloacibacterium},
+\emph{Desulfovibrio}, \emph{Helicobacter}, \emph{Lactobacillus},
+\emph{Lactococcus}, \emph{Mogibacterium}, and \emph{Pseudomonas} to be
+enriched in those with adenomas (25--27). The ability to classify
+individuals as having normal colons or adenomas based solely on the taxa
+within fecal samples has been limited. However, when 16S rRNA gene
+sequence data was combined with the results of a fecal immunochemical
+test (FIT), the ability to diagnose individuals with adenomas was
+improved relative to using the FIT results alone (12).
+
+A recent meta-analysis found that 16S rRNA gene sequences from members
+of \emph{Akkermansia}, \emph{Fusobacterium}, and \emph{Parvimonas} were
+fecal biomarkers for the presence of carcinomas (28). Contrary to
+previous studies they found sequences similar to members of
+\emph{Lactobacillus} and \emph{Ruminococcus} to be enriched in patients
+with adenoma or carcinoma relative to those with normal colons (12, 15,
+16). In addition, they found that 16S rRNA gene sequences from members
+of \emph{Haemophilus}, \emph{Methanosphaera}, \emph{Prevotella}, and
+\emph{Succinovibrio} were enriched in patients with adenomas and
+\emph{Pantoea} were enriched in patients with carcinomas. Although this
+meta-analysis was helpful for distilling a large number of possible
+biomarkers, the aggregate number of samples included in the analysis
+(n=509) was smaller than several larger case-control studies that have
+since been published (12, 27)
 
 Here we provide an updated meta-analysis using 16S rRNA gene sequence
-data from both feces (n = 1737) and colon tissue (492 samples from 350
+data from both feces (n=1737) and colon tissue (492 samples from 350
 individuals) from 14 studies (11--17, 19, 20, 23, 25--27, 29) {[}Table 1
@@
 
-\textbf{\emph{Lower bacterial diversity is associated with increased
-odds ratio (OR) of tumors.}} We first assessed whether variation in
-broad community metrics like total number of operational taxonomic units
+\textbf{\emph{Lower bacterial diversity is associated with higher odds
+ratio (OR) of tumors.}} We first assessed whether variation in broad
+community metrics like total number of operational taxonomic units
 (OTUs) (i.e.~richness), the evenness of their abundance, and the overall
-diversity was associated with disease stage after controlling for study
-and variable region differences. In stool, there was a significant
-decrease in both evenness and diversity as disease severity progressed
-from normal to adenoma to carcinoma (P-value = 0.025 and 0.043,
-respectively) {[}Figure 1{]}; there was no significant difference for
-richness (P-value = 0.21). We next tested whether the decrease in these
+diversity of the communities were associated with disease stage after
+controlling for study and variable region differences. In fecal samples,
+both evenness and diversity were significantly lower in successive
+disease severity categories (evenness P-value=0.025 and diversity
+P-value=0.043) {[}Figure 1{]}; there was no significant difference for
+richness (P-value=0.21). We next tested whether the lower value of these
 community metrics translated into significant ORs for having an adenoma
 or carcinoma. For fecal samples, the ORs for richness were not
-significantly greater than 1.0 for adenoma or carcinoma (P-value = 0.40)
+significantly greater than 1.0 for adenoma or carcinoma (P-value=0.40)
 {[}Figure 2A{]}. The ORs for evenness were significantly higher than 1.0
-for adenoma (OR = 1.3 (1.02 - 1.65)(95\% Confidence Interval), P-value =
-0.035) and carcinoma (OR = 1.66 (1.2 - 2.3), P-value = 0.0021) {[}Figure
-2B{]}. The ORs for diversity were only significantly greater than 1.0
-for carcinoma (OR = 1.61 (1.14 - 2.28), P-value = 0.0069), but not for
-adenoma (P-value = 0.11) {[}Figure 2C{]}. Although these OR are
-significantly greater than 1.0, it is doubtful that these are clinically
-meaningful values.
-
-Similar to our analysis of sequences obtained from stool samples, we
+for adenoma (OR=1.3 (95\% Confidence Interval: 1.02 - 1.65),
+P-value=0.035) and carcinoma (OR=1.66 (1.2 - 2.3), P-value=0.0021)
+{[}Figure 2B{]}. The ORs for diversity were only significantly greater
+than 1.0 for carcinoma (OR=1.61 (1.14 - 2.28), P-value=0.0069), but not
+for adenoma (P-value=0.11) {[}Figure 2C{]}. Although these ORs are
+significantly greater than 1.0, it is doubtful that they are clinically
+meaningful.
+
+Similar to our analysis of sequences obtained from fecal samples, we
 repeated the analysis using sequences obtained from colon tissue. There
-were no significant changes in richness, evenness, or diversity as
+were no significant differences in richness, evenness, or diversity as
 disease severity progressed from control to adenoma to carcinoma
-(P-value \textgreater{} 0.05). We next analyzed the OR, for matched
+(P-values \textgreater{} 0.05). We next analyzed the ORs, for matched
 (i.e.~where unaffected tissue and tumors were obtained from the same
@@
 were not obtained from the same individual) tissue samples. The ORs for
-adenoma and carcinoma by any measure were not significantly different
-from 1.0 (P-value \textgreater{} 0.05) {[}Figure S1 \& Table S1{]}. This
-is likely due to the combination of a small effect size, as suggested
-from the results using stool, and the relatively small number of studies
-and size of studies used in the analysis.
-
-\textbf{\emph{Disease progression is associated with community-wide
-changes in composition and abundance.}} Based on the differences in
-evenness and diversity, we next asked whether there were community-wide
-differences in the structure of the communities associated with
-different disease stages. We identified significant bacterial community
-differences in the stool of patients with adenomas relative to those
-with normal colons in 1 of 4 studies and in patients with carcinomas
-relative to those with normal colons in 6 of 7 studies (PERMANOVA;
-P-value \textless{} 0.05) {[}Table S2{]}. Similar to the analyses using
-stool samples, there were significant differences in bacterial community
-structure between subjects with normal colons and those with adenoma (1
-of 2 studies) and carcinoma (1 of 3 studies) {[}Table S2{]}. For studies
-that used matched samples no differences in bacterial community
-structures were observed {[}Table S2{]}. Combined, these results
-indicate that there are consistent and significant community-wide
-changes in the fecal community structure of subjects with carcinomas.
-However, the signal observed in subjects with adenomas or when using
-tissue samples was not as consistent. This is likely due to a smaller
-effect size or the relatively small sample sizes among the studies that
-characterized the tissue microbiota.
+adenoma and carcinoma were not significantly different from 1.0 for any
+measure (P-values \textgreater{} 0.05) {[}Figure S1 \& Table S1{]}. This
+is likely due to the combination of a small effect size and the
+relatively small number of studies and size of studies used in the
+analysis.
+
+\textbf{\emph{Disease progression is associated with changes in
+community structure.}} Based on the differences in evenness and
+diversity, we next asked whether there were community-wide differences
+in the structure of the communities associated with different disease
+stages. We identified significant bacterial community differences in the
+feces of patients with adenomas relative to those with normal colons in
+1 of 4 studies and in patients with carcinomas relative to those with
+normal colons in 6 of 7 studies (PERMANOVA; P-value \textless{} 0.05)
+{[}Table S2{]}. Similar to the analyses using fecal samples, there were
+significant differences in the bacterial community structure of subjects
+with normal colons and those with adenomas (1 of 2 studies) and
+carcinomas (1 of 3 studies) {[}Table S2{]}. For studies that used
+matched samples, we did not observe any differences in bacterial
+community structures {[}Table S2{]}. Combined, these results indicate
+that there were consistent and significant community-wide changes in the
+fecal community structure of subjects with carcinomas. However, the
+signal observed in subjects with adenomas or when using tissue samples
+was not as consistent. This is likely due to a smaller effect size or
+the relatively small sample sizes among the studies that characterized
+the tissue microbiota.
 
 \textbf{\emph{Individual taxa are associated with significant ORs for
-carcinomas.}} Next we identified those taxa that were associated with
-ORs that were significantly associated with having a normal colon or the
-presence of adenomas or carcinomas. No taxa had a significant OR for the
-presence of adenomas when we used data collected from stool or tissue
-samples (Table S3 \& S4). In contrast, 8 taxa had significant ORs for
-the presence of carcinomas using data from stool samples. Of these, 4
-are commonly associated with the oral cavity: \emph{Fusobacterium} (OR =
-2.74 (1.95 - 3.85)), \emph{Parvimonas} (OR = 3.07 (2.11 - 4.46)),
-\emph{Porphyromonas} (OR = 3.2 (2.26 - 4.54)), and
-\emph{Peptostreptococcus} (OR = 7.11 (3.84 - 13.17)) {[}Table S3{]}. The
-other 4 were \emph{Clostridium XI} (OR = 0.65 (0.49 - 0.86)),
-\emph{Enterobacteriaceae} (OR = 1.79 (1.33 - 2.41)), \emph{Escherichia}
-(OR = 2.15 (1.57 - 2.95)), and \emph{Ruminococcus} (OR = 0.63 (0.48 -
-0.83)). Among the data collected from tissue samples, only unmatched
-carcinoma samples had taxa with a significant OR. Those included
-\emph{Dorea} (OR = 0.35 (0.22 - 0.55)), \emph{Blautia} (OR = 0.47 (0.3 -
-0.73)), and \emph{Weissella} (OR = 5.15 (2.02 - 13.14)).
-Mouth-associated genera were not significantly associated with an
-increased OR for carcinoma in tissue samples {[}Table S4{]}. For
-example, \emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24; however,
-due to the small number of studies and considerable variation in the
-data, the Benjimani-Hochberg corrected P-value was 0.93 {[}Table S4{]}.
-It is interesting to note that \emph{Ruminococcus} and members of
-\emph{Clostridium XI} in stool and \emph{Dorea} and \emph{Blautia} in
-tissue had ORs that were significantly less than 1.0, which suggests
-that these populations are protective against the development of
-carcinomas. Overall, there was no overlap in the taxa with significant
-OR between stool and tissue samples.
+carcinomas.}} Next we identified those taxa that had ORs that were
+significantly associated with having a normal colon or the presence of
+adenomas or carcinomas. No taxa had a significant OR for the presence of
+adenomas when we used data collected from fecal or tissue samples (Table
+S3 \& S4). In contrast, 8 taxa had significant ORs for the presence of
+carcinomas using data from fecal samples. Of these, 4 are commonly
+associated with the oral cavity: \emph{Fusobacterium} (OR=2.74 (1.95 -
+3.85)), \emph{Parvimonas} (OR=3.07 (2.11 - 4.46)), \emph{Porphyromonas}
+(OR=3.2 (2.26 - 4.54)), and \emph{Peptostreptococcus} (OR=7.11 (3.84 -
+13.17)) {[}Table S3{]}. The other 4 were \emph{Clostridium XI} (OR=0.65
+(0.49 - 0.86)), \emph{Enterobacteriaceae} (OR=1.79 (1.33 - 2.41)),
+\emph{Escherichia} (OR = 2.15 (1.57 - 2.95)), and \emph{Ruminococcus}
+(OR=0.63 (0.48 - 0.83)). Among the data collected from tissue samples,
+only unmatched carcinoma samples had taxa with a significant OR. Those
+included \emph{Dorea} (OR=0.35 (0.22 - 0.55)), \emph{Blautia} (OR=0.47
+(0.3 - 0.73)), and \emph{Weissella} (OR=5.15 (2.02 - 13.14)).
+Mouth-associated genera were not significantly associated with a higher
+OR for carcinoma in tissue samples {[}Table S4{]}. For example,
+\emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24; however, due to
+the small number of studies and considerable variation in the data, the
+Benjimani-Hochberg corrected P-value was 0.93 {[}Table S4{]}. It is
+interesting to note that \emph{Ruminococcus} and members of
+\emph{Clostridium XI} in fecal samples and \emph{Dorea} and
+\emph{Blautia} in tissue had ORs that were significantly less than 1.0,
+which suggests that these populations are protective against the
+development of carcinomas. Overall, there was no overlap in the taxa
+with significant OR between fecal and tissue samples.
 
@@
 the area under the curve (AUC). This allowed us to use a more fluid
-relative abundance threshold for defining disease status. Using data
-from stool samples, the 8 taxa did no better at classifying the subjects
-than one would expect by chance (i.e.~AUC=0.50) {[}Figure 3A{]}. The
-taxa that performed the best included \emph{Clostridium XI},
-\emph{Ruminococcus}, and \emph{Escherichia}, however, these had median
-AUC values less than 0.588. Likewise, in unmatched tissue samples the 3
-taxa with significant OR taxa were marginally better than one would
-expect by chance {[}Figure 3B{]}. The relative abundance of \emph{Dorea}
-was the best predictor of carcinomas and its median AUC was only 0.62.
-These results suggest that although these taxa are associated with a
-decreased or increased OR for the presences of carcinomas, individually,
-they do a poor job of classifying a subject's disease status.
+relative abundance threshold for classifying individuals by their
+disease status. Using data from fecal samples, the 8 taxa did no better
+at classifying the subjects than one would expect by chance
+(i.e.~AUC=0.50) {[}Figure 3A{]}. The taxa that performed the best
+included \emph{Clostridium XI}, \emph{Ruminococcus}, and
+\emph{Escherichia}. However, these had median AUC values less than 0.588
+ndicating their limited value as biomarkers when used individually.
+Likewise, in unmatched tissue samples the 3 taxa with significant OR
+taxa had AUC values that were marginally better than one would expect by
+chance {[}Figure 3B{]}. The relative abundance of \emph{Dorea} was the
+best predictor of carcinomas and its median AUC was only 0.62. These
+results suggest that although these taxa are associated with a
+significant OR for the presences of carcinomas, they do a poor job of
+classifying a subject's disease status when used individually.
 
@@
 and evaluated the ability to classify a subject's disease status using
-Random Forest models. For data from stool samples, the combined model
+Random Forest models. For data from fecal samples, the combined model
 had an AUC of 0.75, which was significantly higher than any of the AUC
-values for the individual taxa (P-value \textless{} 0.033). Within the
-significant OR taxa only models the most important taxa across study for
-stool were \emph{Ruminococcus} and \emph{Clostridium XI} based on mean
-decrease in accuracy {[}Figure 4A{]}. Similarly, using data from the
-unmatched tissue samples, the combined model had an AUC of 0.77, which
-was significantly higher than the AUC values for \emph{Blautia} and
-\emph{Weissella} (P-value \textless{} 0.037). For the significant OR
-taxa unmatched tissue models both \emph{Dorea} and \emph{Blautia} were
-the most important based on mean decrease in accuracy {[}Figure 4B{]}.
-Pooling the information from the taxa with significant ORs results in a
-model that outperforms classifications made using individual taxa.
+values for the individual taxa (P-value \textless{} 0.033). When this
+approach was used to train models using data from each study, the most
+important taxa were \emph{Ruminococcus} and \emph{Clostridium XI}
+{[}Figure 4A{]}. Similarly, using data from the unmatched tissue
+samples, the combined model had an AUC of 0.77, which was significantly
+higher than the AUC values for classifying based on the relative
+abundances of \emph{Blautia} and \emph{Weissella} individually (P-value
+\textless{} 0.037). Both \emph{Dorea} and \emph{Blautia} were the most
+important taxa in the tissue-based models {[}Figure 4B{]}. Pooling the
+information from the taxa with significant ORs resulted in models that
+outperformed classifications made using the same taxa individually.
 
 \textbf{\emph{Performance of models based on taxa relative abundance in
-full community are better than those based on taxa with significant
-ORs.}} Next, we asked whether a Random Forest classification model built
-using all of the taxa found in the communities would outperform the
-models generated using those taxa with a significant OR. Similar to our
-inability to identify taxa associated with a significant OR for the
-presence of adenomas, the median AUCs to classify subjects as having
-normal colons or having adenomas using data from stool or tissue samples
-were only marginally better than 0.5 for any study (median AUC = 0.549
-(0.367 - 0.971)(min-max)) {[}Figure 5A \& S2A{]}. In contrast, the
-models for classifying subjects as having normal colons or having
-carcinomas using data from stool or tissue samples yielded AUC values
+full community is better than that of models based on taxa with
+significant ORs.}} Next, we asked whether a Random Forest classification
+model built using all of the taxa found in the communities would
+outperform the models generated using those taxa with a significant OR.
+Similar to our inability to identify taxa associated with a significant
+OR for the presence of adenomas, the median AUCs to classify subjects as
+having normal colons or having adenomas using data from fecal or tissue
+samples were only marginally better than 0.5 for any study (median
+AUC=0.549 (range: 0.367 - 0.971)) {[}Figure 5A \& S2A{]}. In contrast,
+the models for classifying subjects as having normal colons or having
+carcinomas using data from fecal or tissue samples yielded AUC values
 meaningfully higher than 0.5 {[}Figure 5B \& S2B-C{]}. When we compared
@@
 the taxa with significant ORs, the results were mixed. Using the data
-from stool samples we found that the AUC for 6 of 7 studies increased by
-an average of 14.8\%) and AUC for the Flemer study decreased by 0.54\%).
-The overall improvement in performance was statistically significant
-(mean = 12.61\%, one-tailed paired T-test; P-value = 0.005). For these
-full taxa models using stool, \emph{Bacteroides} and
+from fecal samples, we found that the AUC for 6 of 7 studies were an
+average of 14.8\% higher and AUC for the Flemer study was 0.54\% lower
+when using the relative abundance data from all taxa relative to using
+the relative abundance of only the taxa with significant ORs. The
+overall improvement in performance was statistically significant
+(mean=12.61\%, one-tailed paired T-test; P-value=0.005). Among the
+models trained using data from fecal samples, \emph{Bacteroides} and
 \emph{Lachnospiraceae} were the most common taxa in the top 10\% mean
-decrease in accuracy (MDA) across studies {[}Figure S3{]}. Similarly,
-using the data from unmatched tissue samples we found that the AUC of
-studies increased by an average of 19.11\% when we used all of the taxa
-rather than the 3 taxa with significant ORs (one-tailed paired T-test;
-P-value = 0.03). For the full taxa models using unmatched tissue,
-\emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
+decrease in accuracy across studies {[}Figure S3{]}. Using data from
+unmatched tissue samples to train classification models, we found that
+the AUC of studies was an average 19.11\% higher when we used all of the
+taxa rather than the 3 taxa with significant ORs (one-tailed paired
+T-test; P-value=0.03). For the models trained using data from unmatched
+tissue samples, \emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
 \emph{Ruminococcaceae} were the most common taxa in the top 10\% mean
-decrease in accuracy across studies {[}Figure S4{]}. Although the
-significant OR taxa models can classify individuals with and without
-carcinoma tumors, they are still missing taxa from the full community
-models that can increase the model accuracy.
-
-\textbf{\emph{Performance of models based on OTU relative abundance in
-full community are not significantly better than those based on taxa
-with significant ORs.}} The previous models were based on relative
-abundance data where sequences were classified to coarse taxonomic
-assignments (i.e.~typically genus or family level). To determine whether
-model performance improved with finer scale classification, we assigned
+decrease in accuracy across studies {[}Figure S4{]}. Although the models
+trained using those taxa with a significant OR perform well for
+classifying individuals with and without carcinomas, models trained
+using data from the full community perform better.
+
+\textbf{\emph{Performance of models based on OTU relative abundances are
+not significantly better than those based on taxa with significant
+ORs.}} The previous models were based on relative abundance data where
+sequences were classified to coarse taxonomic assignments
+(i.e.~typically genus or family level). To determine whether model
+performance improved with finer scale classification, we assigned
 sequences to operational taxonomic units (OTUs) where the similarity
@@
 community did a poor job of differentiating between subjects with normal
-colons and those with adenomas (median AUC: 0.53 (0.37- 0.56)), but did
-a good job of differentiating between subjects with normal colons and
-those with carcinomas (median AUC: 0.71 (0.5- 0.9)). The OTU-based
-models performed similarly to those constructed using the taxa with
-significant ORs (one-tailed paired T-test; P-value = 0.966) and those
-using all taxa (one-tailed paired T-test; P-value = 0.146) {[}Figure
-4{]}. Among the OTUs that had the highest mean decrease in accuracy for
-the OTU-based models, we found that OTUs that affiliated with all of the
-8 taxa that had a significant OR were within the top 10\% for at least
-one study. This result was surprising as it indicated that a finer scale
-classification of the sequences and thus a larger number of features to
-select from, did not yield improved classification of the subjects.
+colons and those with adenomas (median AUC: 0.53 (0.37- 0.56)). However,
+they did a good job of differentiating between subjects with normal
+colons and those with carcinomas (median AUC: 0.71 (0.50- 0.904)). The
+OTU-based models performed similarly to those constructed using the taxa
+with significant ORs (one-tailed paired T-test; P-value = 0.966) and
+those using all taxa (one-tailed paired T-test; P-value = 0.146)
+{[}Figure 4{]}. Among the OTUs that had the highest mean decrease in
+accuracy for the OTU-based models, we found that OTUs that affiliated
+with all of the 8 taxa that had a significant OR were within the top
+10\% for at least one study. This result was surprising as it indicated
+that a finer scale classification of the sequences and thus a larger
+number of features to select from, did not yield improved classification
+of the subjects.
 

### 2018-03-13 submission/manuscript.Rmd
add pat's discussion suggestions

@@
 
-We performed a meta-analysis to identify and validate microbiome-base biomarkers that could be used to classify individuals as having normal colons or colonic tumors using fecal or tissue samples.
+We performed a meta-analysis to identify and validate microbiota-base biomarkers that could be used to classify individuals as having normal colons or colonic tumors using fecal or tissue samples.
 To our surprise, Random Forest classification models constructed to differentiate individuals with normal colons from those with carcinomas using a subset of the community performed well relative to models constructed using the full communities.
@@
 When we attempted to improve the specificity of the classification by using an OTU-based approach the resulting models performed as well as those constructed using coarse taxonomic assignments.
-These results are significant because they strengthen the growing literature indicating a role of the microbiome in tumorigenesis [@zackular_gut_2013] and as a potential tool as a non-invasive diagnostic and for assessing risk of disease and recurrence [@baxter_microbiota-based_2016; @Sze2017].
+These results are significant because they strengthen the growing literature indicating a role for the colonic microbiota in tumorigenesis, as a potential tool as a non-invasive diagnostic, and for assessing risk of disease and recurrence [@zackular_gut_2013; @baxter_microbiota-based_2016; @Sze2017].
 
-These results suggested that fine scale classification of sequences into OTUs does not improve our classification models.
-This has been suggested in previous literature where shotgun metagenomic data did not perform better than 16S rRNA gene sequencing data in classifying individuals with normal colons and those with carcinomas [@Hannigan2017].
+Fine scale classification of sequences into OTUs did not improve our classification models.
+This was also tested in earlier efforts to use shotgun metagenomic data to classify individuals as having normal colons or tumors; however, it was shown that analyses performed using shotgun metagenomic data did not perform better than using 16S rRNA gene sequencing data [@Hannigan2017].
 We hypothesize that fine scale classification may not result in better classification because distribution of microbiota between individuals is patchy.
@@
 
-When used separately to classify individuals with carcinomas, the taxa with significant ORs could not reliably classify individuals [Figure 3].
+When used to classify individuals with carcinomas, the taxa with significant ORs could not reliably classify individuals on their own [Figure 3].
 This result further supports the hypothesis that carcinoma-associated microbiota have a patchy distribution.
 Two individuals may have had the same classification, based on the relative abundance of different populations within this group of `r length(crc_OR_stool$measure)` taxa.
-Although these results only reflect associations with disease, it is tempting to hypothesize that the patchiness represents distinct mechanisms of exacerbating tumorigenesis or that multiple taxa have the same mechanism of exacerbating tumorigenesis.
+Although these results only reflect associations with disease, it is tempting to hypothesize that the patchiness is indicative of distinct mechanisms of exacerbating tumorigenesis or that multiple taxa have the same mechanism of exacerbating tumorigenesis.
 For example, strains of *Escherichia coli* and *Fusobacterium nucleatum* have been shown to worsen inflammation in mouse models of tumorigenesis [@ecoli_Arthur_2014; @arthur_intestinal_2012; @kostic_fusobacterium_2013].
@@
 This result was particularly interesting because members of these taxa (i.e.
-*Ruminococcus* and *Clostridium XI* in stool and *Dorea* and *Blautia* in tissue) are thought to be beneficial due to their involvement in production of anti-inflammatory short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].
+*Ruminococcus* and *Clostridium XI* in fecal samples and *Dorea* and *Blautia* in tissue) are thought to be beneficial due to their involvement in production of anti-inflammatory short chain fatty acids [@Venkataraman2016; @Herrmann2018; @Reichardt2017].
 
-All of the adenoma classification models performed poorly and is not inconsistent with previous studies [@hale_shifts_2017; @Sze2017].
-However, the classification results are at odds with results of the multitarget microbiota test (MMT) from Baxter, *et al.* [@baxter_microbiota-based_2016] who observed an AUC of 0.755 when applied to individuals with adenomas.
+All of the adenoma classification models performed poorly, which is consistent with previous studies [@hale_shifts_2017; @Sze2017].
+However, the classification results are at odds with results of the multitarget microbiota test (MMT) from Baxter, et al.
+[@baxter_microbiota-based_2016] who observed an AUC of 0.755 when the test was applied to individuals with adenomas.
 There are two major differences between the models generated in this meta-analysis and that analysis.
@@
 Because FIT data were not available for the other studies in our meta-analysis, it was not possible to validate the MMT approach.
-The ability to differentiate between individuals with and without adenomas is an important problem since early detection of tumors is critical.
+The ability to differentiate between individuals with and without adenomas is an important problem since early detection of tumors is critical to patient survivorship.
 However, it is possible that we might have been able to detect differences in the bacterial community if individuals with non-advanced and advanced adenomas were separated.
-This is a clinically relevant distinction since advanced adenomas are at highest risk of progressing to a carcinoma.
+This is a clinically relevant distinction since advanced adenomas are at highest risk of progressing to carcinomas.
 The initial changes of the microbiota during tumorigenesis could be focal to where the initial adenoma develops and would not be easily assessed using fecal samples from an individual with non-advanced adenomas.
@@
 
-Stool samples represent a non-invasive approach to assessing the structure of the gut microbiota and are potentially useful for diagnosing individuals as having colonic tumors.
+Fecal samples represent a non-invasive approach to assess the structure of the gut microbiota and are potentially useful for diagnosing individuals as having colonic tumors.
 However, they do not reflect the structure of the mucosal microbiota [@Flynn_preprint_2017].
-Regardless, the taxa that were the most important in the stool-based models overlapped with those from the models trained using the data from unmatched and matched colon tissue samples [Figure S3].
+Regardless, the taxa that were the most important in the feces-based models overlapped with those from the models trained using the data from unmatched and matched colon tissue samples [Figure S3].
 Mucosal biopsies are preferred for focused mechanistic studies and have offered researchers the opportunity to sample healthy and diseased tissue from the same individuals (i.e.
-matched) using each individual as their own control or in a cross sectional design (i.e.
+matched) using each individual as their own control or in a cross-sectional design (i.e.
 unmatched).
@@
 Thus, it was not surprising that tissue-based studies did not provide clearer associations between the mucosal microbiota and the presence of tumors.
-Interestingly, Fusobacterium, which has received increased recent attention for its potential role in tumorigenesis [@kostic_fusobacterium_2013] was not consistently identified across the studies in our meta-analysis.
+Interestingly, *Fusobacterium*, which has received increased attention for its potential role in tumorigenesis [@kostic_fusobacterium_2013] was not consistently identified across the studies in our meta-analysis.
 This could be due to the relatively small number of individuals in the limited number of studies.
@@
 
-Among our stool samples, we failed to identify several notable populations that are commonly associated with carcinomas including an enterotoxigenic strain of *Bacteroides fragilis* (ETBF) and *Streptococcus gallolyticus* subsp.
+Among the fecal sample data, we failed to identify several notable populations that are commonly associated with carcinomas including an enterotoxigenic strain of *Bacteroides fragilis* (ETBF) and *Streptococcus gallolyticus* subsp.
 *gallolyticus* [@bfrag_Boleij_2014; @strep_Aymeric_2017].
 ETBF have been found in tumors in the proximal colon where they tend to form biofilms [@Purcell2017; @dejea_microbiota_2014].
-Considering DNA from bacteria that are more prevalent in the proximal colon may be degraded by the time it leaves the body, it is not surprising that we failed to identify a significant OR for Bacteroides with carcinomas.
-In addition, since our approach could only classify sequences to the genus level and there are likely multiple Bacteroides populations in the colon, it is possible that sequences from ETBF and non-oncogenic Bacteroides were pooled.
-This would then reduce the OR between Bacterioides and whether an individual had carcinomas.
+Considering DNA from bacteria that are more prevalent in the proximal colon may be degraded by the time it leaves the body, it is not surprising that we failed to identify a significant OR for *Bacteroides* with carcinomas.
+In addition, since our approach could only classify sequences to the genus level and there are likely multiple *Bacteroides* populations in the colon, it is possible that sequences from ETBF and non-oncogenic *Bacteroides* were pooled.
+This would then reduce the OR between *Bacterioides* and whether an individual had carcinomas.
 It is also necessary to distinguish between populations that are biomarkers for a disease and those that are known to cause disease.
-The former may also have a causative role in the disease.
 Although the latter have been shown to have a causative role, they may appear at low relative abundance, be found in specific locations, or may have a highly patchy distribution among affected individuals.
@@
 There have been several meta-analyses similar to this study that have sought biomarkers for obesity [@Sze2016; @Walters2014; @Finucane2014], inflammatory bowel disease [@Walters2014], and colorectal cancer [@shah_leveraging_2017].
-Considering microbiome research is particularly prone to hype and overgeneralization of results, these analyses are critical.
-For example, previous meta-analyses have demonstrated that there are no clear fecal biomarkers for obesity [Sze2016; @Walters2014; @Finucane2014].
-Such meta-analyses are difficult to perform because the underlying 16S rRNA gene sequence data are not publicly available, metadata are missing, incomplete, or vague, sequence data are of poor quality or derived by non-standard approaches, and the original studies were significantly underpowered.
+Considering microbiome research is particularly prone to hype and overgeneralization of results [@Hanage2014], these analyses are critical.
+Meta-analyses are difficult to perform because the underlying 16S rRNA gene sequence data are not publicly available, metadata are missing, incomplete, or vague, sequence data are of poor quality or derived by non-standard approaches, and the original studies may be significantly underpowered.
 Reluctance to publish negative results (i.e.
 the “file drawer effect”) is also likely to skew our understanding of the relationship between microbiota and disease.
-Better attention to these specifics will increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses.
+Better attention to these specific issues will increase the reproducibility and replicability of microbiota studies and make it easier to perform these crucial meta-analyses.
 Moving forward, meta-analyses will be important tools to help aggregate and find commonalities across studies when investigating the microbiota in the context of a specific disease [@Sze2016; @Walters2014; @Finucane2014; @shah_leveraging_2017].

### 2018-03-13 submission/manuscript.Rmd
method changes except RF section

@@
 Studies were included that used tissue or feces as their sample source for 454 or Illumina 16S rRNA gene sequencing.
-Some studies (N = 12) were excluded because they did not have publicly available sequences, did not use 454 or Illumina machines, or did not have metadata in which the authors were able to share.
-We were able to obtain sequence data and metadata from the following studies: Ahn, *et al.* [@ahn_human_2013], Baxter, *et al.* [@baxter_microbiota-based_2016], Brim, *et al.* [@brim_microbiome_2013], Burns, *et al.* [@burns_virulence_2015], Chen, *et al.*  [@chen_human_2012], Dejea, *et al.* [@dejea_microbiota_2014], Flemer, *et al.* [@flemer_tumour-associated_2017], Geng, *et al.* [@geng_diversified_2013], Hale, *et al.* [@hale_shifts_2017], Kostic, *et al.* [@kostic_genomic_2012], Lu, *et al.* [@lu_mucosal_2016], Sanapareddy, *et al.* [@sanapareddy_increased_2012], Wang, *et al.* [@wang_structural_2012], Weir, *et al.* [@weir_stool_2013], and Zeller, *et al.*  [@zeller_potential_2014].
-The Zackular [@zackular_human_2014] study was excluded because their 90 individuals were contained within the larger Baxter study [@baxter_microbiota-based_2016].
+A significant number of studies (N=12) were excluded from the meta-analysis because they did not have publicly available sequences, did not use 454 or Illumina sequencing platforms, or did not have metadata that the authors were able to share.
+We were able to obtain sequence data and metadata from the following studies: Ahn, et al.
+[@ahn_human_2013], Baxter, et al.
+[@baxter_microbiota-based_2016], Brim, et al.
+[@brim_microbiome_2013], Burns, et al.
+[@burns_virulence_2015], Chen, et al.
+[@chen_human_2012], Dejea, et al.
+[@dejea_microbiota_2014], Flemer, et al.
+[@flemer_tumour-associated_2017], Geng, et al.
+[@geng_diversified_2013], Hale, et al.
+[@hale_shifts_2017], Kostic, et al.
+[@kostic_genomic_2012], Lu, et al.
+[@lu_mucosal_2016], Sanapareddy, et al.
+[@sanapareddy_increased_2012], Wang, et al.
+[@wang_structural_2012], Weir, et al.
+[@weir_stool_2013], and Zeller, et al.
+[@zeller_potential_2014].
+The Zackular [@zackular_human_2014] study was excluded because the individuals studied were included within the larger Baxter study [@baxter_microbiota-based_2016].
 The Kostic study was excluded because after we processed the sequences, all of the case samples had 100 or fewer sequences.
@@
 There were seven studies with only fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal and tissue samples (Chen and Flemer).
-After curating the sequences, `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` stool samples and `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` tissue samples remained in the analysis [Tables 1 and 2].
+After curating the sequences, `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` fecal samples and `r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` tissue samples remained in the analysis [Tables 1 and 2].
 
@@
 Raw sequence data and metadata were primarily obtained from the Sequence Read Archive (SRA) and dbGaP.
-Other sequence and metadata were obtained directly from the authors (n = 4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]).
-Each dataset was processed separately using mothur (v1.39.3) [@schloss_introducing_2009] using the default quality filtering methods for both 454 and Illumina sequence data.
-If it was not possible to use the defaults because the sequences were trimmed too much, then the stated quality cut-offs from the original study were used.
+Other sequence and metadata were obtained directly from the authors (n=4, [@flemer_tumour-associated_2017; @hale_shifts_2017; @sanapareddy_increased_2012; @weir_stool_2013]).
+Each dataset was processed separately using mothur (v1.39.3) using the default quality filtering methods for both 454 and Illumina sequence data [@schloss_introducing_2009].
+If it was not possible to use the defaults because the trimmed sequences were too short, then the stated quality cut-offs from the original study were used.
 Chimeric sequences were identified and removed using VSEARCH [@rognes_vsearch_2016].
@@
 All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017].
-For OTU richness, evenness, and Shannon diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
-Testing for OTU richness, evenness, and Shannon diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hyper-variable region used.
-Odds ratios (OR) were analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study.
+For OTU richness, evenness, and Shannon diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011].
+Testing for OTU richness, evenness, and Shannon diversity differences utilized linear mixed-effect models to correct for study, repeat sampling of individuals (tissue only), and 16S rRNA gene sequence region used using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015].
+ORs were analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study.
 OR significance testing utilized the chi-squared test.
-Diversity differences measured by the Bray-Curtis index utilized the creation of distance matrix and testing with PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017].
+Community structure differences were calculated using the Bray-Curtis dissimilarity index and PERMANOVA was used to test for tumor-associated differences in structure with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017].
 Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002].
@@
 ***Reproducible Methods.***
-The code and analysis can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017.
+The analysis code can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2017.
 Unless otherwise mentioned, the accession number of raw sequences from the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.

### 2018-03-13 submission/manuscript.Rmd
all pat's changes except fig adjustment and RF section

@@
 
-**Table 1: Total Individuals in each Study Included in the Stool Analysis**
+**Table 1: Total Individuals in each Study Included in the Fecal Analysis**
 
@@
 
-**Figure 1: Significant Bacterial Community Metrics for Adenoma or Carcinoma in Stool.** A) Adenoma evenness.
+**Figure 1: Significant Bacterial Community Metrics for Adenoma or Carcinoma in Feces** A) Adenoma evenness.
 B) Carcinoma evenness.
@@
 
-**Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Stool.** A) Community-based odds ratio for adenoma.
+**Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Feces.** A) Community-based odds ratio for adenoma.
 B) Community-based odds ratio for carcinoma.
@@
 
-**Figure 3: The AUC of Indivdiual Significant OR Taxa to classify Carcinoma.** A) Stool samples.
+**Figure 3: The AUC of Indivdiual Significant OR Taxa to classify Carcinoma.** A) Fecal samples.
 B) Unmatched tissue samples.
@@
 
-**Figure 4: Most Important Members in Significant OR Taxa Carcinoma Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest stool-based models.
+**Figure 4: Most Important Members in Significant OR Taxa Carcinoma Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest fecal-based models.
 B) Common taxa in the top 10 percent for carcinoma Random Forest unmatched tissue-based models.
@@
 
-**Figure 5: Stool Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
+**Figure 5: Fecal Sample Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
 B) Carcinoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
@@
 
-**Figure 6: Stool Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
+**Figure 6: Fecal Sample Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
 B) Test AUCs of carcinoma models using all genera or significant OR taxa only.
@@
 
-**Figure S3: Most Common Taxa Across Carcinoma Full Community Stool Study Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest all taxa-based models.
+**Figure S3: Most Common Taxa Across Carcinoma Full Community Fecal Sample Study Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest all taxa-based models.
 B) Common taxa in the top 10 percent for carcinoma Random Forest all OTU-based models.

### 2018-03-13 submission/manuscript.tex
all pat's changes except fig adjustment and RF section

@@
 they did a good job of differentiating between subjects with normal
-colons and those with carcinomas (median AUC: 0.71 (0.50- 0.904)). The
+colons and those with carcinomas (median AUC: 0.71 (0.50- 0.90)). The
 OTU-based models performed similarly to those constructed using the taxa
-with significant ORs (one-tailed paired T-test; P-value = 0.966) and
-those using all taxa (one-tailed paired T-test; P-value = 0.146)
-{[}Figure 4{]}. Among the OTUs that had the highest mean decrease in
-accuracy for the OTU-based models, we found that OTUs that affiliated
-with all of the 8 taxa that had a significant OR were within the top
-10\% for at least one study. This result was surprising as it indicated
-that a finer scale classification of the sequences and thus a larger
-number of features to select from, did not yield improved classification
-of the subjects.
+with significant ORs (one-tailed paired T-test; P-value=0.966) and those
+using all taxa (one-tailed paired T-test; P-value=0.146) {[}Figure 4{]}.
+Among the OTUs that had the highest mean decrease in accuracy for the
+OTU-based models, we found that OTUs that affiliated with all of the 8
+taxa that had a significant OR were within the top 10\% for at least one
+study. This result was surprising as it indicated that a finer scale
+classification of the sequences and thus a larger number of features to
+select from, did not yield improved classification of the subjects.
 
@@
 dataset to the other datasets.}} Considering the good performance of the
-Random Forest models using taxa with a significant OR and using all of
-the taxa, we next asked how well the models would perform when given
-data from a different subject cohort. For instance, if a model was
-trained using data from the Ahn study, we wanted to know how well it
-would perform using the data from the Baxter study. We found the models
-trained using the taxa with a significant OR all had a higher median AUC
-than the models trained using all of the taxa when tested on the other
-datasets {[}Figure 6 \& S5{]}. As might be expected, the difference
-between the performance of the modelling approaches appeared to vary
-with the size of the training cohort (R\textsuperscript{2} = 0.66)
-{[}Figure 6{]}. These data suggest that given a sufficient number of
-subjects with normal colons and carcinomas, Random Forest models trained
-using a small number of taxa can accurately classify individuals from a
-different cohort.
+Random Forest models trained using the relative abundance of taxa with
+significant ORs and models trained using the relative abundance of all
+taxa, we next asked how well the models would perform when given data
+from a different cohort. For instance, if a model was trained using data
+from the Ahn study, we wanted to know how well it would perform using
+the data from the Baxter study. The models trained using the taxa with
+significant ORs all had a higher median AUC than the models trained
+using all of the taxa when tested on the other datasets {[}Figure 6 \&
+S5{]}. As might be expected, the difference between the performance of
+the modeling approaches appeared to vary with the size of the training
+cohort (R\textsuperscript{2}=0.66) {[}Figure 6{]}. These data suggest
+that given a sufficient number of subjects with normal colons and
+carcinomas, Random Forest models trained using a small number of taxa
+can accurately classify individuals from a different cohort.
 
@@
 
-We performed a meta-analysis to identify and validate microbiome-base
+We performed a meta-analysis to identify and validate microbiota-base
 biomarkers that could be used to classify individuals as having normal
@@
 using coarse taxonomic assignments. These results are significant
-because they strengthen the growing literature indicating a role of the
-microbiome in tumorigenesis (9) and as a potential tool as a
-non-invasive diagnostic and for assessing risk of disease and recurrence
-(12, 30).
-
-These results suggested that fine scale classification of sequences into
-OTUs does not improve our classification models. This has been suggested
-in previous literature where shotgun metagenomic data did not perform
-better than 16S rRNA gene sequencing data in classifying individuals
-with normal colons and those with carcinomas (31). We hypothesize that
-fine scale classification may not result in better classification
-because distribution of microbiota between individuals is patchy. In
-contrast, models using coarser taxonomic assignments will pool the fine
-scale diversity, resulting in less patchiness and better classification.
-Furthermore, the ability of models trained using a subset of the
-community to outperform those using the full community when testing the
-models on the other datasets may also be a product of the patchiness of
-the human-associated microbiota. The models based on the 8 taxa that had
-significant ORs used taxa that were found in every study and tended to
-have higher relative abundances. Similar to the OTU-based models, those
-models based on the full community taxonomy assignments were still
-sensitive to the patchy distribution of taxa. Regardless, it is
-encouraging that a collection of 8 taxa could reliably classify
-individuals as having carcinomas considering the differences in cohorts,
-DNA extraction procedures, regions of the 16S rRNA gene, and sequencing
-methods.
-
-When used separately to classify individuals with carcinomas, the taxa
-with significant ORs could not reliably classify individuals {[}Figure
-3{]}. This result further supports the hypothesis that
+because they strengthen the growing literature indicating a role for the
+colonic microbiota in tumorigenesis, as a potential tool as a
+non-invasive diagnostic, and for assessing risk of disease and
+recurrence (9, 12, 30).
+
+Fine scale classification of sequences into OTUs did not improve our
+classification models. This was also tested in earlier efforts to use
+shotgun metagenomic data to classify individuals as having normal colons
+or tumors; however, it was shown that analyses performed using shotgun
+metagenomic data did not perform better than using 16S rRNA gene
+sequencing data (31). We hypothesize that fine scale classification may
+not result in better classification because distribution of microbiota
+between individuals is patchy. In contrast, models using coarser
+taxonomic assignments will pool the fine scale diversity, resulting in
+less patchiness and better classification. Furthermore, the ability of
+models trained using a subset of the community to outperform those using
+the full community when testing the models on the other datasets may
+also be a product of the patchiness of the human-associated microbiota.
+The models based on the 8 taxa that had significant ORs used taxa that
+were found in every study and tended to have higher relative abundances.
+Similar to the OTU-based models, those models based on the full
+community taxonomy assignments were still sensitive to the patchy
+distribution of taxa. Regardless, it is encouraging that a collection of
+8 taxa could reliably classify individuals as having carcinomas
+considering the differences in cohorts, DNA extraction procedures,
+regions of the 16S rRNA gene, and sequencing methods.
+
+When used to classify individuals with carcinomas, the taxa with
+significant ORs could not reliably classify individuals on their own
+{[}Figure 3{]}. This result further supports the hypothesis that
 carcinoma-associated microbiota have a patchy distribution. Two
@@
 these results only reflect associations with disease, it is tempting to
-hypothesize that the patchiness represents distinct mechanisms of
+hypothesize that the patchiness is indicative of distinct mechanisms of
 exacerbating tumorigenesis or that multiple taxa have the same mechanism
@@
 because members of these taxa (i.e. \emph{Ruminococcus} and
-\emph{Clostridium XI} in stool and \emph{Dorea} and \emph{Blautia} in
-tissue) are thought to be beneficial due to their involvement in
-production of anti-inflammatory short chain fatty acids (32--34).
-
-All of the adenoma classification models performed poorly and is not
-inconsistent with previous studies (27, 30). However, the classification
+\emph{Clostridium XI} in fecal samples and \emph{Dorea} and
+\emph{Blautia} in tissue) are thought to be beneficial due to their
+involvement in production of anti-inflammatory short chain fatty acids
+(32--34).
+
+All of the adenoma classification models performed poorly, which is
+consistent with previous studies (27, 30). However, the classification
 results are at odds with results of the multitarget microbiota test
-(MMT) from Baxter, \emph{et al.} (12) who observed an AUC of 0.755 when
-applied to individuals with adenomas. There are two major differences
-between the models generated in this meta-analysis and that analysis.
-The MMT attempted to classify individuals as having a normal colon or
-having colonic lesions (i.e.~adenomas or carcinomas) and not adenomas
-alone. Further, the MMT incorporated fecal immunoglobulin test (FIT)
-data while our models only used 16S rRNA gene sequencing data. Because
-FIT data were not available for the other studies in our meta-analysis,
-it was not possible to validate the MMT approach. The ability to
-differentiate between individuals with and without adenomas is an
-important problem since early detection of tumors is critical. However,
-it is possible that we might have been able to detect differences in the
-bacterial community if individuals with non-advanced and advanced
-adenomas were separated. This is a clinically relevant distinction since
-advanced adenomas are at highest risk of progressing to a carcinoma. The
-initial changes of the microbiota during tumorigenesis could be focal to
-where the initial adenoma develops and would not be easily assessed
-using fecal samples from an individual with non-advanced adenomas.
-Unfortunately, distinguishing between individuals with advanced and
-non-advanced adenomas was not possible in our meta-analysis since the
-studies did not provide the clinical data needed to make that
-distinction.
-
-Stool samples represent a non-invasive approach to assessing the
-structure of the gut microbiota and are potentially useful for
-diagnosing individuals as having colonic tumors. However, they do not
-reflect the structure of the mucosal microbiota (35). Regardless, the
-taxa that were the most important in the stool-based models overlapped
-with those from the models trained using the data from unmatched and
-matched colon tissue samples {[}Figure S3{]}. Mucosal biopsies are
-preferred for focused mechanistic studies and have offered researchers
-the opportunity to sample healthy and diseased tissue from the same
-individuals (i.e.~matched) using each individual as their own control or
-in a cross sectional design (i.e.~unmatched). Because obtaining these
-samples is invasive, carries risks to the individual, and is expensive,
-studies investigating the structure of the mucosal microbiota generally
-have a limited number of participants. Thus, it was not surprising that
+(MMT) from Baxter, et al. (12) who observed an AUC of 0.755 when the
+test was applied to individuals with adenomas. There are two major
+differences between the models generated in this meta-analysis and that
+analysis. The MMT attempted to classify individuals as having a normal
+colon or having colonic lesions (i.e.~adenomas or carcinomas) and not
+adenomas alone. Further, the MMT incorporated fecal immunoglobulin test
+(FIT) data while our models only used 16S rRNA gene sequencing data.
+Because FIT data were not available for the other studies in our
+meta-analysis, it was not possible to validate the MMT approach. The
+ability to differentiate between individuals with and without adenomas
+is an important problem since early detection of tumors is critical to
+patient survivorship. However, it is possible that we might have been
+able to detect differences in the bacterial community if individuals
+with non-advanced and advanced adenomas were separated. This is a
+clinically relevant distinction since advanced adenomas are at highest
+risk of progressing to carcinomas. The initial changes of the microbiota
+during tumorigenesis could be focal to where the initial adenoma
+develops and would not be easily assessed using fecal samples from an
+individual with non-advanced adenomas. Unfortunately, distinguishing
+between individuals with advanced and non-advanced adenomas was not
+possible in our meta-analysis since the studies did not provide the
+clinical data needed to make that distinction.
+
+Fecal samples represent a non-invasive approach to assess the structure
+of the gut microbiota and are potentially useful for diagnosing
+individuals as having colonic tumors. However, they do not reflect the
+structure of the mucosal microbiota (35). Regardless, the taxa that were
+the most important in the feces-based models overlapped with those from
+the models trained using the data from unmatched and matched colon
+tissue samples {[}Figure S3{]}. Mucosal biopsies are preferred for
+focused mechanistic studies and have offered researchers the opportunity
+to sample healthy and diseased tissue from the same individuals
+(i.e.~matched) using each individual as their own control or in a
+cross-sectional design (i.e.~unmatched). Because obtaining these samples
+is invasive, carries risks to the individual, and is expensive, studies
+investigating the structure of the mucosal microbiota generally have a
+limited number of participants. Thus, it was not surprising that
 tissue-based studies did not provide clearer associations between the
 mucosal microbiota and the presence of tumors. Interestingly,
-Fusobacterium, which has received increased recent attention for its
+\emph{Fusobacterium}, which has received increased attention for its
 potential role in tumorigenesis (6) was not consistently identified
@@
 
-Among our stool samples, we failed to identify several notable
+Among the fecal sample data, we failed to identify several notable
 populations that are commonly associated with carcinomas including an
@@
 the body, it is not surprising that we failed to identify a significant
-OR for Bacteroides with carcinomas. In addition, since our approach
-could only classify sequences to the genus level and there are likely
-multiple Bacteroides populations in the colon, it is possible that
-sequences from ETBF and non-oncogenic Bacteroides were pooled. This
-would then reduce the OR between Bacterioides and whether an individual
-had carcinomas. It is also necessary to distinguish between populations
-that are biomarkers for a disease and those that are known to cause
-disease. The former may also have a causative role in the disease.
-Although the latter have been shown to have a causative role, they may
-appear at low relative abundance, be found in specific locations, or may
-have a highly patchy distribution among affected individuals.
+OR for \emph{Bacteroides} with carcinomas. In addition, since our
+approach could only classify sequences to the genus level and there are
+likely multiple \emph{Bacteroides} populations in the colon, it is
+possible that sequences from ETBF and non-oncogenic \emph{Bacteroides}
+were pooled. This would then reduce the OR between \emph{Bacterioides}
+and whether an individual had carcinomas. It is also necessary to
+distinguish between populations that are biomarkers for a disease and
+those that are known to cause disease. Although the latter have been
+shown to have a causative role, they may appear at low relative
+abundance, be found in specific locations, or may have a highly patchy
+distribution among affected individuals.
 
@@
 colorectal cancer (28). Considering microbiome research is particularly
-prone to hype and overgeneralization of results, these analyses are
-critical. For example, previous meta-analyses have demonstrated that
-there are no clear fecal biomarkers for obesity (39, 40). Such
-meta-analyses are difficult to perform because the underlying 16S rRNA
-gene sequence data are not publicly available, metadata are missing,
-incomplete, or vague, sequence data are of poor quality or derived by
-non-standard approaches, and the original studies were significantly
-underpowered. Reluctance to publish negative results (i.e.~the ``file
-drawer effect'') is also likely to skew our understanding of the
-relationship between microbiota and disease. Better attention to these
-specifics will increase the reproducibility and replicability of
-microbiota studies and make it easier to perform these crucial
-meta-analyses. Moving forward, meta-analyses will be important tools to
-help aggregate and find commonalities across studies when investigating
-the microbiota in the context of a specific disease (28, 38--40).
+prone to hype and overgeneralization of results (41), these analyses are
+critical. Meta-analyses are difficult to perform because the underlying
+16S rRNA gene sequence data are not publicly available, metadata are
+missing, incomplete, or vague, sequence data are of poor quality or
+derived by non-standard approaches, and the original studies may be
+significantly underpowered. Reluctance to publish negative results
+(i.e.~the ``file drawer effect'') is also likely to skew our
+understanding of the relationship between microbiota and disease. Better
+attention to these specific issues will increase the reproducibility and
+replicability of microbiota studies and make it easier to perform these
+crucial meta-analyses. Moving forward, meta-analyses will be important
+tools to help aggregate and find commonalities across studies when
+investigating the microbiota in the context of a specific disease (28,
+38--40).
 
@@
 \textbf{\emph{Datasets.}} The studies used for this meta-analysis were
-identified through the review articles written by Keku, et al. (41) and
-Vogtmann, et al. (42). Additional studies, not mentioned in those
+identified through the review articles written by Keku, et al. (42) and
+Vogtmann, et al. (43). Additional studies, not mentioned in those
 reviews were obtained based on the authors' knowledge of the literature.
 Studies were included that used tissue or feces as their sample source
-for 454 or Illumina 16S rRNA gene sequencing. Some studies (N = 12) were
-excluded because they did not have publicly available sequences, did not
-use 454 or Illumina machines, or did not have metadata in which the
-authors were able to share. We were able to obtain sequence data and
-metadata from the following studies: Ahn, \emph{et al.} (11), Baxter,
-\emph{et al.} (12), Brim, \emph{et al.} (29), Burns, \emph{et al.} (15),
-Chen, \emph{et al.} (13), Dejea, \emph{et al.} (20), Flemer, \emph{et
-al.} (17), Geng, \emph{et al.} (19), Hale, \emph{et al.} (27), Kostic,
-\emph{et al.} (43), Lu, \emph{et al.} (26), Sanapareddy, \emph{et al.}
-(25), Wang, \emph{et al.} (14), Weir, \emph{et al.} (23), and Zeller,
-\emph{et al.} (16). The Zackular (44) study was excluded because their
-90 individuals were contained within the larger Baxter study (12). The
-Kostic study was excluded because after we processed the sequences, all
-of the case samples had 100 or fewer sequences. The final analysis
-included 14 studies (Tables 1 and 2). There were seven studies with only
-fecal samples (Ahn, Baxter, Brim, Hale, Wang, Weir, and Zeller), five
-studies with only tissue samples (Burns, Dejea, Geng, Lu, Sanapareddy),
-and two studies with both fecal and tissue samples (Chen and Flemer).
-After curating the sequences, 1737 stool samples and 492 tissue samples
-remained in the analysis {[}Tables 1 and 2{]}.
+for 454 or Illumina 16S rRNA gene sequencing. A significant number of
+studies (N=12) were excluded from the meta-analysis because they did not
+have publicly available sequences, did not use 454 or Illumina
+sequencing platforms, or did not have metadata that the authors were
+able to share. We were able to obtain sequence data and metadata from
+the following studies: Ahn, et al. (11), Baxter, et al. (12), Brim, et
+al. (29), Burns, et al. (15), Chen, et al. (13), Dejea, et al. (20),
+Flemer, et al. (17), Geng, et al. (19), Hale, et al. (27), Kostic, et
+al. (44), Lu, et al. (26), Sanapareddy, et al. (25), Wang, et al. (14),
+Weir, et al. (23), and Zeller, et al. (16). The Zackular (45) study was
+excluded because the individuals studied were included within the larger
+Baxter study (12). The Kostic study was excluded because after we
+processed the sequences, all of the case samples had 100 or fewer
+sequences. The final analysis included 14 studies (Tables 1 and 2).
+There were seven studies with only fecal samples (Ahn, Baxter, Brim,
+Hale, Wang, Weir, and Zeller), five studies with only tissue samples
+(Burns, Dejea, Geng, Lu, Sanapareddy), and two studies with both fecal
+and tissue samples (Chen and Flemer). After curating the sequences, 1737
+fecal samples and 492 tissue samples remained in the analysis {[}Tables
+1 and 2{]}.
 
@@
 primarily obtained from the Sequence Read Archive (SRA) and dbGaP. Other
-sequence and metadata were obtained directly from the authors (n = 4,
-(17, 23, 25, 27)). Each dataset was processed separately using mothur
-(v1.39.3) (45) using the default quality filtering methods for both 454
-and Illumina sequence data. If it was not possible to use the defaults
-because the sequences were trimmed too much, then the stated quality
+sequence and metadata were obtained directly from the authors (n=4, (17,
+23, 25, 27)). Each dataset was processed separately using mothur
+(v1.39.3) using the default quality filtering methods for both 454 and
+Illumina sequence data (46). If it was not possible to use the defaults
+because the trimmed sequences were too short, then the stated quality
 cut-offs from the original study were used. Chimeric sequences were
-identified and removed using VSEARCH (46). The curated sequences were
-assigned to OTUs at 97\% similarity using the OptiClust algorithm (47)
+identified and removed using VSEARCH (47). The curated sequences were
+assigned to OTUs at 97\% similarity using the OptiClust algorithm (48)
 and classified to the deepest taxonomic level that had 80\% support
 using the naïve Bayesian classifier trained on the RDP taxonomy outline
-(version 14, (48)).
+(version 14, (49)).
 
@@
 to be positive. We measured the dissimilarity between individuals by
-calculating the pairwise Bray-Curtis index and used PERMANOVA (49) to
+calculating the pairwise Bray-Curtis index and used PERMANOVA (50) to
 test whether individuals with normal colons were significantly different
@@
 or carcinoma and corrected for multiple comparisons using the
-Benjamini-Hochberg method (50). Again, for each taxon, if the relative
+Benjamini-Hochberg method (51). Again, for each taxon, if the relative
 abundance was greater than the median relative abundance for that taxon
@@
 Normally, \(\sqrt{p}\) has been found to be what is chosen as the ideal
-mtry after model tuneing (51).
+mtry after model tuneing (52).
 
 \textbf{\emph{Statistical Analysis.}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package (52). For
+sequence processing utilized the R (v3.4.3) software package (53). For
 OTU richness, evenness, and Shannon diversity analysis, values were
-power transformed using the rcompanion (v1.11.1) package (53) and then
-Z-score normalized using the car (v2.1.6) package (54). Testing for OTU
+power transformed using the rcompanion (v1.11.1) package (54) and
+Z-score normalized using the car (v2.1.6) package (55). Testing for OTU
 richness, evenness, and Shannon diversity differences utilized linear
-mixed-effect models created using the lme4 (v1.1.15) package (55) to
-correct for study, repeat sampling of individuals (tissue only), and 16S
-hyper-variable region used. Odds ratios (OR) were analyzed using both
-the epiR (v0.9.93) and metafor (v2.0.0) packages (56, 57) by assessing
-how many individuals with and without disease were above and below the
-overall median value within each specific study. OR significance testing
-utilized the chi-squared test. Diversity differences measured by the
-Bray-Curtis index utilized the creation of distance matrix and testing
-with PERMANOVA executed with the vegan (v2.4.5) package (58). Random
-Forest models were built using both the caret (v6.0.78) and randomForest
-(v4.6.12) packages (59, 60). All figures were created using both ggplot2
-(v2.2.1) and gridExtra (v2.3) packages (61, 62).
-
-\textbf{\emph{Reproducible Methods.}} The code and analysis can be found
-at
-\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_Microbiome_2017}.
+mixed-effect models to correct for study, repeat sampling of individuals
+(tissue only), and 16S rRNA gene sequence region used using the lme4
+(v1.1.15) package (56). ORs were analyzed using both the epiR (v0.9.93)
+and metafor (v2.0.0) packages (57, 58) by assessing how many individuals
+with and without disease were above and below the overall median value
+within each specific study. OR significance testing utilized the
+chi-squared test. Community structure differences were calculated using
+the Bray-Curtis dissimilarity index and PERMANOVA was used to test for
+tumor-associated differences in structure with the vegan (v2.4.5)
+package (59). Random Forest models were built using both the caret
+(v6.0.78) and randomForest (v4.6.12) packages (60, 61). All figures were
+created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages (62,
+63).
+
+\textbf{\emph{Reproducible Methods.}} The analysis code can be found at
+\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2017}.
 Unless otherwise mentioned, the accession number of raw sequences from
@@
 
+\hypertarget{ref-Hanage2014}{}
+41. \textbf{Hanage, W. P.} 2014. Microbiology: Microbiome science needs
+a healthy dose of scepticism. Nature. Springer Nature
+\textbf{512}:247--248.
+
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-41. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
+42. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
 \textbf{B. Jovov}, and \textbf{X. Han}. 2015. The gastrointestinal
@@
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-42. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
+43. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
 Epidemiologic studies of the human microbiome and cancer. British
@@
 \hypertarget{ref-kostic_genomic_2012}{}
-43. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
+44. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
 Pedamallu}, \textbf{M. Michaud}, \textbf{F. Duke}, \textbf{A. M. Earl},
@@
 \hypertarget{ref-zackular_human_2014}{}
-44. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
+45. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
 Ruffin}, and \textbf{P. D. Schloss}. 2014. The human gut microbiome as a
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-45. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
+46. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
 Ryabin}, \textbf{J. R. Hall}, \textbf{M. Hartmann}, \textbf{E. B.
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-46. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
+47. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
 \textbf{C. Quince}, and \textbf{F. Mahé}. 2016. VSEARCH: A versatile
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-47. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
+48. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
 OptiClust, an Improved Method for Assigning Amplicon-Based Sequence Data
@@
 \hypertarget{ref-rdp_Wang2007}{}
-48. \textbf{Wang, Q.}, \textbf{G. M. Garrity}, \textbf{J. M. Tiedje},
+49. \textbf{Wang, Q.}, \textbf{G. M. Garrity}, \textbf{J. M. Tiedje},
 and \textbf{J. R. Cole}. 2007. Naive bayesian classifier for rapid
@@
 \hypertarget{ref-permanova_Anderson2013}{}
-49. \textbf{Anderson, M. J.}, and \textbf{D. C. I. Walsh}. 2013.
+50. \textbf{Anderson, M. J.}, and \textbf{D. C. I. Walsh}. 2013.
 PERMANOVA, ANOSIM, and the mantel test in the face of heterogeneous
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-50. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
+51. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
 the false discovery rate: A practical and powerful approach to multiple
@@
 \hypertarget{ref-Breiman2001}{}
-51. \textbf{Breiman, L.} 2001. Machine Learning. Springer Nature
+52. \textbf{Breiman, L.} 2001. Machine Learning. Springer Nature
 \textbf{45}:5--32.
@@
 \hypertarget{ref-r_citation_2017}{}
-52. \textbf{R Core Team}. 2017. R: A language and environment for
+53. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-53. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
+54. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
 extension education program evaluation.
@@
 \hypertarget{ref-car_citation_2011}{}
-54. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
+55. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
 applied regressionSecond. Sage, Thousand Oaks CA.
@@
 \hypertarget{ref-lme4_citation_2015}{}
-55. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
+56. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
 \textbf{S. Walker}. 2015. Fitting linear mixed-effects models using
@@
 \hypertarget{ref-epir_citation_2017}{}
-56. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
+57. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
 Heuer}, \textbf{J. Marshall}, \textbf{J. Sanchez}, \textbf{R. Thornton},
@@
 \hypertarget{ref-metafor_citation_2010}{}
-57. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
+58. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
 the metafor package. Journal of Statistical Software \textbf{36}:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-58. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
+59. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
 \textbf{R. Kindt}, \textbf{P. Legendre}, \textbf{D. McGlinn}, \textbf{P.
@@
 \hypertarget{ref-caret_citation_2017}{}
-59. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
+60. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
 Williams}, \textbf{C. Keefer}, \textbf{A. Engelhardt}, \textbf{T.
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-60. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
+61. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-61. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
+62. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
 analysis. Springer-Verlag New York.
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-62. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
+63. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
 ``grid'' graphics.
@@
 
-\textbf{Table 1: Total Individuals in each Study Included in the Stool
+\textbf{Table 1: Total Individuals in each Study Included in the Fecal
 Analysis}
@@
 \textbf{Figure 1: Significant Bacterial Community Metrics for Adenoma or
-Carcinoma in Stool.} A) Adenoma evenness. B) Carcinoma evenness. C)
+Carcinoma in Feces} A) Adenoma evenness. B) Carcinoma evenness. C)
 Carcinoma Shannon diversity. Blue represents controls and red represents
@@
 \textbf{Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial
-Community Metrics in Stool.} A) Community-based odds ratio for adenoma.
+Community Metrics in Feces.} A) Community-based odds ratio for adenoma.
 B) Community-based odds ratio for carcinoma. Colors represent the
@@
 \textbf{Figure 3: The AUC of Indivdiual Significant OR Taxa to classify
-Carcinoma.} A) Stool samples. B) Unmatched tissue samples. The larger
+Carcinoma.} A) Fecal samples. B) Unmatched tissue samples. The larger
 circle represents the median AUC of all studies and the smaller circles
@@
 Carcinoma Models.} A) Common taxa in the top 10 percent for carcinoma
-Random Forest stool-based models. B) Common taxa in the top 10 percent
+Random Forest fecal-based models. B) Common taxa in the top 10 percent
 for carcinoma Random Forest unmatched tissue-based models. Blue
@@
 
-\textbf{Figure 5: Stool Random Forest Model Train AUCs.} A) Adenoma
-random forest model AUCs between all genera, all OTU, and select model
-based on significant OR taxa. B) Carcinoma random forest model AUCs
-between all genera, all OTU, and select model based on significant OR
-taxa. The black line represents the median AUC for the respective group.
-If no values are present in the singificant OR taxa group then there
-were no significant taxa identified and no model was tested.
-
-\textbf{Figure 6: Stool Random Forest Genus-Based Model Test AUCs.} A)
-Test AUCs of adenoma models using all genera across study. B) Test AUCs
-of carcinoma models using all genera or significant OR taxa only. The
-black line represents the AUC at 0.5. The red lines represent the median
-AUC of all test AUCs for a specific study.
+\textbf{Figure 5: Fecal Sample Random Forest Model Train AUCs.} A)
+Adenoma random forest model AUCs between all genera, all OTU, and select
+model based on significant OR taxa. B) Carcinoma random forest model
+AUCs between all genera, all OTU, and select model based on significant
+OR taxa. The black line represents the median AUC for the respective
+group. If no values are present in the singificant OR taxa group then
+there were no significant taxa identified and no model was tested.
+
+\textbf{Figure 6: Fecal Sample Random Forest Genus-Based Model Test
+AUCs.} A) Test AUCs of adenoma models using all genera across study. B)
+Test AUCs of carcinoma models using all genera or significant OR taxa
+only. The black line represents the AUC at 0.5. The red lines represent
+the median AUC of all test AUCs for a specific study.
 
@@
 \textbf{Figure S3: Most Common Taxa Across Carcinoma Full Community
-Stool Study Models.} A) Common taxa in the top 10 percent for carcinoma
-Random Forest all taxa-based models. B) Common taxa in the top 10
-percent for carcinoma Random Forest all OTU-based models. Blue
+Fecal Sample Study Models.} A) Common taxa in the top 10 percent for
+carcinoma Random Forest all taxa-based models. B) Common taxa in the top
+10 percent for carcinoma Random Forest all OTU-based models. Blue
 represents less important and red represents more important to the

### 2018-03-13 submission/supplemental.Rmd
pat updates
all pat's changes except fig adjustment and RF section

@@
 
-**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stool and Tissue**
-
-```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}
-
-temp_t1 <- read_csv("../data/process/tables/beta_perm_adn_stool_summary.csv")
-
-table_s1 <- temp_t1 %>% 
-  mutate(fstat = round(fstat, digits = 2), 
-         r2 = round(r2, digits = 3), 
-         pvalue = round(pvalue, digits = 4), 
-         study = stringr::str_replace(study, 
-                                      c("brim", "zeller", "baxter", "hale"), 
-                                      c("Brim", "Zeller", "Baxter", "Hale")), 
-         comparison = "Adenoma", 
-         sample_type = "Stool") %>% 
-  select(fstat, r2, pvalue, study, comparison, sample_type)
-
-colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
-
-temp_t2 <- read_csv("../data/process/tables/beta_perm_stool_summary.csv")
-
-table_s2 <- temp_t2 %>% 
-  mutate(fstat = round(fstat, digits = 2), 
-         r2 = round(r2, digits = 3), 
-         pvalue = round(pvalue, digits = 4), 
-         study = stringr::str_replace(study, 
-                                      c("wang", "weir", "ahn", "zeller", "baxter", "hale", "flemer"), 
-                                      c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")), 
-         comparison = "Carcinoma", 
-         sample_type = "Stool") %>% 
-  select(fstat, r2, pvalue, study, comparison, sample_type)
-
-colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
-
-combined_table_s1 <- table_s1 %>% bind_rows(table_s2)
-
-temp_t3a <- read_csv("../data/process/tables/beta_perm_adn_unmatched_tissue_summary.csv")
-temp_t3b <- read_csv("../data/process/tables/beta_perm_adn_matched_tissue_summary.csv")
-
-table_s3 <- temp_t3a %>% 
-  bind_rows(temp_t3b) %>% 
-  mutate(fstat = round(fstat, digits = 2), 
-         r2 = round(r2, digits = 3), 
-         pvalue = round(pvalue, digits = 4), 
-         study = c("Lu", "Flemer", "Lu (Matched)"), 
-         comparison = "Adenoma", 
-         sample_type = "Tissue") %>% 
-  select(fstat, r2, pvalue, study, comparison, sample_type)
-
-colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
-
-temp_t4a <- read_csv("../data/process/tables/beta_perm_unmatched_tissue_summary.csv")
-temp_t4b <- read_csv("../data/process/tables/beta_perm_matched_tissue_summary.csv")
-
-table_s4 <- temp_t4a %>% 
-  bind_rows(temp_t4b) %>% 
-  mutate(fstat = round(fstat, digits = 2), 
-         r2 = round(r2, digits = 3), 
-         pvalue = round(pvalue, digits = 4), 
-         study = c("Sanapareddy", "Burns", "Flemer", "Chen", 
-                   "Dejea (Matched)", "Geng (Matched)", "Burns (Matched)"), 
-         comparison = "Carcinoma", 
-         sample_type = "Tissue") %>% 
-  select(fstat, r2, pvalue, study, comparison, sample_type)
-
-colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
-
-combined_table_s2 <- table_s3 %>% bind_rows(table_s4)
-
-combined_table_s1 <- combined_table_s1 %>% bind_rows(combined_table_s2)
-
-kable(combined_table_s1, format = "markdown", align = 'c', padding = 2)
-```
-
-\newpage
-
-**Table S2: Alpha Diversity Metrics OR in Tissue**
+**Table S1: Alpha Diversity Metrics OR in Tissue**
 
@@
 
-**Table S3: OR in Stool of Tumor by Individual Genera with Initial P-Value under 0.05**
+**Table S2: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Feces and Tissue**
+
+```{r supplemental_table1, echo=FALSE, message=FALSE, warning=FALSE}
+
+temp_t1 <- read_csv("../data/process/tables/beta_perm_adn_stool_summary.csv")
+
+table_s1 <- temp_t1 %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = stringr::str_replace(study, 
+                                      c("brim", "zeller", "baxter", "hale"), 
+                                      c("Brim", "Zeller", "Baxter", "Hale")), 
+         comparison = "Adenoma", 
+         sample_type = "Feces") %>% 
+  select(fstat, r2, pvalue, study, comparison, sample_type)
+
+colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
+
+temp_t2 <- read_csv("../data/process/tables/beta_perm_stool_summary.csv")
+
+table_s2 <- temp_t2 %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = stringr::str_replace(study, 
+                                      c("wang", "weir", "ahn", "zeller", "baxter", "hale", "flemer"), 
+                                      c("Wang", "Weir", "Ahn", "Zeller", "Baxter", "Hale", "Flemer")), 
+         comparison = "Carcinoma", 
+         sample_type = "Feces") %>% 
+  select(fstat, r2, pvalue, study, comparison, sample_type)
+
+colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
+
+combined_table_s1 <- table_s1 %>% bind_rows(table_s2)
+
+temp_t3a <- read_csv("../data/process/tables/beta_perm_adn_unmatched_tissue_summary.csv")
+temp_t3b <- read_csv("../data/process/tables/beta_perm_adn_matched_tissue_summary.csv")
+
+table_s3 <- temp_t3a %>% 
+  bind_rows(temp_t3b) %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = c("Lu", "Flemer", "Lu (Matched)"), 
+         comparison = "Adenoma", 
+         sample_type = "Tissue") %>% 
+  select(fstat, r2, pvalue, study, comparison, sample_type)
+
+colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
+
+temp_t4a <- read_csv("../data/process/tables/beta_perm_unmatched_tissue_summary.csv")
+temp_t4b <- read_csv("../data/process/tables/beta_perm_matched_tissue_summary.csv")
+
+table_s4 <- temp_t4a %>% 
+  bind_rows(temp_t4b) %>% 
+  mutate(fstat = round(fstat, digits = 2), 
+         r2 = round(r2, digits = 3), 
+         pvalue = round(pvalue, digits = 4), 
+         study = c("Sanapareddy", "Burns", "Flemer", "Chen", 
+                   "Dejea (Matched)", "Geng (Matched)", "Burns (Matched)"), 
+         comparison = "Carcinoma", 
+         sample_type = "Tissue") %>% 
+  select(fstat, r2, pvalue, study, comparison, sample_type)
+
+colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
+
+combined_table_s2 <- table_s3 %>% bind_rows(table_s4)
+
+combined_table_s1 <- combined_table_s1 %>% bind_rows(combined_table_s2)
+
+kable(combined_table_s1, format = "markdown", align = 'c', padding = 2)
+```
+
+\newpage
+
+**Table S3: OR in Feces or Tumor by Individual Genera with Initial P-Value under 0.05**
 

### 2018-03-14 code/run_random_forest_otu_tissue.R
update to 5 CV

@@ -53,9 +53,8 @@ get_data <- function(i, metadata){
   # i is the study of interest
   
   # grabs subsampled data and assigns rownames from sample names to table
-  shared_data <- read.delim(paste("data/process/", i, "/", i, ".0.03.subsample.shared", 
-                                  sep = ""), header = T, stringsAsFactors = F) %>% 
-    select(-label, -numOtus)
+  shared_data <- as.data.frame(read_tsv(paste("data/process/", i, "/", i, ".0.03.subsample.shared", sep = "")) %>% 
+    select(-label, -numOtus))
   # grabs the meta data and transforms polyp to control (polyp/control vs cancer) 
   study_meta <- metadata %>% filter(study == i)
   
@@ -167,7 +166,7 @@ make_rf_model <- function(run_marker, study, train_data){
     
   fitControl <- trainControl(## 10-fold CV
     method = method_used,
-    number = 2,
+    number = 5,
     p = 0.8,
     classProbs = TRUE, 
     summaryFunction = twoClassSummary, 

### 2018-03-14 code/make_genus_rf_auc_against_study_graph.R
changes to include feces

@@ -123,7 +123,7 @@ adn_tissue_graph <- adn_all_tissue %>%
                         levels = c("flemer", "lu"), 
                         labels = c("Flemer", "Lu\n(Matched)"))) %>% 
   ggplot(aes(model, auc, color = study, group = model_type)) + 
-  geom_point(size = 3.5, show.legend = T) + 
+  geom_point(size = 3.5, position = position_dodge(width = 0.3), show.legend = T) + 
   stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 
                colour = "red", geom = "crossbar", size = 0.5, width = 0.5) + 
   facet_grid(. ~ model_type) + 
@@ -155,7 +155,7 @@ adn_stool_graph <- adn_all_stool %>%
                         levels = c("baxter", "brim", "hale", "zeller"), 
                         labels = c("Baxter", "Brim", "Hale", "Zeller"))) %>% 
   ggplot(aes(model, auc, color = study, group = model_type)) + 
-  geom_jitter(width = 0.2, size = 3.5, show.legend = T) + 
+  geom_point(position = position_dodge(width = 0.5), size = 3.5, show.legend = T) + 
   stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 
                colour = "red", geom = "crossbar", size = 0.5, width = 0.5) + 
   facet_grid(. ~ model_type) + 
@@ -164,7 +164,7 @@ adn_stool_graph <- adn_all_stool %>%
   labs(x = "", y = "AUC") + theme_bw() + ggtitle("A") + 
   scale_color_manual(name = "Study", 
                      values = c('#8968CD', '#34618DFF', '#006400', '#FDE725FF')) + 
-  annotate("text", label = paste("Adenoma (Stool)"), x = 1.0, y = 1.07, size = 2.5) + 
+  annotate("text", label = paste("Adenoma (Feces)"), x = 1.0, y = 1.07, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.09, size = 20), 
         legend.position = "bottom", 
         legend.text = element_text(size = 6),
@@ -187,7 +187,7 @@ crc_unmatched_tissue_graph <- crc_all_unmatched_tissue %>%
                         levels = c("burns", "chen", "flemer", "sana"), 
                         labels = c("Burns", "Chen", "Flemer", "Sanapareddy"))) %>% 
[…]

### 2018-03-14 code/make_ind_genera_auc_graph.R

@@ -65,7 +65,7 @@ stool_graph <- stool_data %>%
   theme_bw() + 
   scale_color_manual(name = "", values = c('#B0C4DE', '#8968CD', '#ED9121', '#006400', '#97D83FFF', '#8B4513', '#FDE725FF')) + 
   labs(x = "AUC", y = "") + ggtitle("A") + 
-  annotate("text", label = paste("Carcinoma (Stool)"), x = 0.15, y = 8.4, size = 2.5) + 
+  annotate("text", label = paste("Carcinoma (Feces)"), x = 0.15, y = 8.4, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 

### 2018-03-14 code/make_rf_auc_full_versus_specific_graph.R

@@ -153,7 +153,7 @@ adn_stool_graph <- adn_stool %>%
   labs(x = "", y = "Model AUC") + theme_bw() + ggtitle("A") + 
   scale_color_manual(values = c('#8968CD', '#34618DFF', '#006400', '#FDE725FF'), 
                      guide = guide_legend(nrow = 1, ncol = 4)) + 
-  annotate("text", label = paste("Adenoma (Stool)"), x = 0.80, y = 1.05, size = 2.5) + 
+  annotate("text", label = paste("Adenoma (Feces)"), x = 0.80, y = 1.05, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
         legend.position = "bottom", 
         legend.text = element_text(size = 6),
@@ -213,7 +213,7 @@ crc_stool_graph <- crc_stool %>%
   scale_color_manual(name = "", 
                      values = c('#B0C4DE', '#8968CD', '#ED9121', '#006400', '#97D83FFF', '#8B4513', '#FDE725FF'), 
                      guide = guide_legend(nrow = 2, ncol = 4)) + 
-  annotate("text", label = paste("Carcinoma (Stool)"), x = 0.8, y = 1.05, size = 2.5) + 
+  annotate("text", label = paste("Carcinoma (Feces)"), x = 0.8, y = 1.05, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.1, size = 20), 
         legend.position = "bottom", 
         legend.text = element_text(size = 6),

### 2018-03-14 code/make_stool_alpha_OR_graph.R

@@ -65,7 +65,7 @@ adn_stool_graph <- adn_all_stool %>%
   facet_grid(. ~ measure) + 
   labs(x = expression(Log["2"]~Odds~Ratio), y = "") + theme_bw() + ggtitle("A") + 
   scale_color_manual(name = "Variable Region", values = c( '#000000', '#FFB90F', '#443A83FF', '#31688EFF')) + 
-  annotate("text", label = paste("Adenoma (Stool)"), x = -2.75, y = 5.5, size = 2.5) + 
+  annotate("text", label = paste("Adenoma (Feces)"), x = -2.75, y = 5.5, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 
@@ -94,7 +94,7 @@ crc_stool_graph <- crc_all_stool %>%
   labs(x = expression(Log["2"]~Odds~Ratio), y = "") + theme_bw() + ggtitle("B") + 
   scale_color_manual(name = "Variable Region", 
                      values = c('#000000', '#FFC1C1', '#FFB90F', '#35B779FF', '#443A83FF')) + 
-  annotate("text", label = paste("Carcinoma (Stool)"), x = -2.70, y = 8.4, size = 2.5) + 
+  annotate("text", label = paste("Carcinoma (Feces)"), x = -2.70, y = 8.4, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
         panel.grid.minor = element_blank(), 

### 2018-03-14 code/make_select_imp_otu_graph.R
update to include pat's reorder comment

@@ -25,6 +25,13 @@ unmatched_tissue_mda <- read_csv("data/process/tables/ALL_genus_unmatched_tissue
   ungroup()
 
 
+stool_vars <- stool_mda %>% group_by(otu) %>% summarise(median_zscore = median(zscore)) %>% 
+  arrange(median_zscore) %>% ungroup()
+
+unmatched_tissue_vars <- unmatched_tissue_mda %>% group_by(otu) %>% summarise(median_zscore = median(zscore)) %>% 
+  arrange(median_zscore) %>% ungroup()
+
+
 ##############################################################################################
 ############################## List of code to make figures ##################################
 ##############################################################################################
@@ -33,7 +40,10 @@ unmatched_tissue_mda <- read_csv("data/process/tables/ALL_genus_unmatched_tissue
 imp_stool_model <- stool_mda %>% 
   mutate(study = factor(study, 
                         levels = c("ahn", "baxter", "flemer", "hale", "wang", "weir", "zeller"), 
-                        labels= c(c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller")))) %>% 
+                        labels= c(c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller"))), 
+         otu = factor(otu, 
+                      levels = stool_vars$otu, 
+                      labels = stool_vars$otu)) %>% 
   ggplot(aes(study, otu, fill = zscore)) + 
   geom_tile(color = "white") + 
   scale_fill_gradient2(name = "Z-Score Median MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
@@ -49,7 +59,10 @@ imp_stool_model <- stool_mda %>%
 imp_unmatched_tissue_model <- unmatched_tissue_mda %>% 
   mutate(study = factor(study, 
                         levels = c("burns", "chen", "flemer", "sana"), 
-                        labels= c(c("Burns", "Chen", "Flemer", "Sanapareddy")))) %>% 
+                        labels= c(c("Burns", "Chen", "Flemer", "Sanapareddy"))), 
+         otu = factor(otu, 
+                      levels = unmatched_tissue_vars$otu, 
+                      labels = unmatched_tissue_vars$otu)) %>% 
   ggplot(aes(study, otu, fill = zscore)) + 
   geom_tile(color = "white") + 
   scale_fill_gradient2(name = "Z-Score Median MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 

### 2018-03-14 submission/manuscript.Rmd
update RF section --ok from josh on understandability

@@
 Because no taxa were identified as having a significant OR associated with adenomas using stool samples or tissue samples, classification models based on OR data were not constructed to classify individuals as having normal colons or adenomas.
-Within the training dataset, 10-fold cross validation (5-fold cross validation for small datasets) was used to build a model that was then evaluated on the testing set.
-For the models constructed using the taxa with significant ORs, the default mtry setting was used to train the model and this model was tested on the other datasets in the meta-analysis.
-The reported AUC values are the AUCs for the application of the model on the test sets.
-For the OTU-based models, the dataset was split into training (80% of samples) and testing (20%) sets and 10-fold cross validation (5-fold cross validation for small datasets) on the training set was used to generate the model for the testing set.
-The original 80/20 split and fitting was repeated 100 times and the average AUC from these 100 repeats was reported.
+For all models the hyperparameters for ntree was set to 500 and the mtry set to the sqaure root of the number of taxa or OTUs within the model ($\sqrt{total~variables}$).
+The $\sqrt{total~variables}$ has been found to approximate the optimum mtry chosen after model tuneing [@Breiman2001].
+All fecal models were built using a 10-fold cross validation (CV) while tissue models were built using 5-fold CV due to study sample size.
+One exception to this rule was the Weir study which was built using a 2-fold CV due to extremely low sample number.
+For significant OR and all taxa data, models were trained on a single study and then tested on the remaining studies with AUCs recorded during both train and testing phases.
+For the OTU data, 100 10-fold CVs were run to generate a range of AUCs that could be reasonably expected to occur and the average AUC from these 100 repeats was reported.
 The Mean Decrease in Accuracy (MDA), a measure of the importance of each taxon to the overall model was used to rank the taxa used in each model.
-For all models, the default setting used was $\sqrt{p}$, where p is all the variables used in the respective model.
-Normally, $\sqrt{p}$ has been found to be what is chosen as the ideal mtry after model tuneing [@Breiman2001].
 

### 2018-03-14 submission/manuscript.tex
update RF section --ok from josh on understandability

@@
 classification models based on OR data were not constructed to classify
-individuals as having normal colons or adenomas. Within the training
-dataset, 10-fold cross validation (5-fold cross validation for small
-datasets) was used to build a model that was then evaluated on the
-testing set. For the models constructed using the taxa with significant
-ORs, the default mtry setting was used to train the model and this model
-was tested on the other datasets in the meta-analysis. The reported AUC
-values are the AUCs for the application of the model on the test sets.
-For the OTU-based models, the dataset was split into training (80\% of
-samples) and testing (20\%) sets and 10-fold cross validation (5-fold
-cross validation for small datasets) on the training set was used to
-generate the model for the testing set. The original 80/20 split and
-fitting was repeated 100 times and the average AUC from these 100
-repeats was reported. The Mean Decrease in Accuracy (MDA), a measure of
-the importance of each taxon to the overall model was used to rank the
-taxa used in each model. For all models, the default setting used was
-\(\sqrt{p}\), where p is all the variables used in the respective model.
-Normally, \(\sqrt{p}\) has been found to be what is chosen as the ideal
-mtry after model tuneing (52).
+individuals as having normal colons or adenomas. For all models the
+hyperparameters for ntree was set to 500 and the mtry set to the sqaure
+root of the number of taxa or OTUs within the model
+(\(\sqrt{total~variables}\)). The \(\sqrt{total~variables}\) has been
+found to approximate the optimum mtry chosen after model tuneing (52).
+All fecal models were built using a 10-fold cross validation (CV) while
+tissue models were built using 5-fold CV due to study sample size. One
+exception to this rule was the Weir study which was built using a 2-fold
+CV due to extremely low sample number. For significant OR and all taxa
+data, models were trained on a single study and then tested on the
+remaining studies with AUCs recorded during both train and testing
+phases. For the OTU data, 100 10-fold CVs were run to generate a range
+of AUCs that could be reasonably expected to occur and the average AUC
+from these 100 repeats was reported. The Mean Decrease in Accuracy
+(MDA), a measure of the importance of each taxon to the overall model
+was used to rank the taxa used in each model.
 

### 2018-03-14 submission/manuscript.Rmd
update with newest references

@@
 Thus, it was not surprising that tissue-based studies did not provide clearer associations between the mucosal microbiota and the presence of tumors.
-Interestingly, *Fusobacterium*, which has received increased attention for its potential role in tumorigenesis [@kostic_fusobacterium_2013] was not consistently identified across the studies in our meta-analysis.
+Interestingly, *Fusobacterium*, which has received increased attention for its potential role in tumorigenesis [@kostic_fusobacterium_2013] was not consistently identified across the studies in our meta-analysis which is consistent with a recent replicability study [@Repass2018].
 This could be due to the relatively small number of individuals in the limited number of studies.

### 2018-03-14 submission/manuscript.tex
update with newest references

@@
 OTU-based models performed similarly to those constructed using the taxa
-with significant ORs (one-tailed paired T-test; P-value=0.966) and those
-using all taxa (one-tailed paired T-test; P-value=0.146) {[}Figure 4{]}.
+with significant ORs (one-tailed paired T-test; P-value=0.979) and those
+using all taxa (one-tailed paired T-test; P-value=0.184) {[}Figure 4{]}.
 Among the OTUs that had the highest mean decrease in accuracy for the
@@
 potential role in tumorigenesis (6) was not consistently identified
-across the studies in our meta-analysis. This could be due to the
-relatively small number of individuals in the limited number of studies.
-The classification models trained using the tissue-based data performed
-well when tested with the training data (Figure S4), but performed
-poorly when tested on the other tissue-associated datasets (Figure S5).
+across the studies in our meta-analysis which is consistent with a
+recent replicability study (36). This could be due to the relatively
+small number of individuals in the limited number of studies. The
+classification models trained using the tissue-based data performed well
+when tested with the training data (Figure S4), but performed poorly
+when tested on the other tissue-associated datasets (Figure S5).
 Disturbingly, taxa that are commonly associated with reagent
@@
 Such contamination is common in studies where there is relatively low
-bacterial biomass (36). The lack of replication among the tissue-based
+bacterial biomass (37). The lack of replication among the tissue-based
 biomarkers may be a product of the relatively small number of studies
@@
 ETBF have been found in tumors in the proximal colon where they tend to
-form biofilms (20, 37). Considering DNA from bacteria that are more
+form biofilms (20, 38). Considering DNA from bacteria that are more
 prevalent in the proximal colon may be degraded by the time it leaves
@@
 been several meta-analyses similar to this study that have sought
-biomarkers for obesity (38--40), inflammatory bowel disease (39), and
+biomarkers for obesity (39--41), inflammatory bowel disease (40), and
 colorectal cancer (28). Considering microbiome research is particularly
-prone to hype and overgeneralization of results (41), these analyses are
+prone to hype and overgeneralization of results (42), these analyses are
 critical. Meta-analyses are difficult to perform because the underlying
@@
 investigating the microbiota in the context of a specific disease (28,
-38--40).
+39--41).
 
@@
 \textbf{\emph{Datasets.}} The studies used for this meta-analysis were
-identified through the review articles written by Keku, et al. (42) and
-Vogtmann, et al. (43). Additional studies, not mentioned in those
+identified through the review articles written by Keku, et al. (43) and
+Vogtmann, et al. (44). Additional studies, not mentioned in those
 reviews were obtained based on the authors' knowledge of the literature.
@@
 Flemer, et al. (17), Geng, et al. (19), Hale, et al. (27), Kostic, et
-al. (44), Lu, et al. (26), Sanapareddy, et al. (25), Wang, et al. (14),
-Weir, et al. (23), and Zeller, et al. (16). The Zackular (45) study was
+al. (45), Lu, et al. (26), Sanapareddy, et al. (25), Wang, et al. (14),
+Weir, et al. (23), and Zeller, et al. (16). The Zackular (46) study was
 excluded because the individuals studied were included within the larger
@@
 (v1.39.3) using the default quality filtering methods for both 454 and
-Illumina sequence data (46). If it was not possible to use the defaults
+Illumina sequence data (47). If it was not possible to use the defaults
 because the trimmed sequences were too short, then the stated quality
 cut-offs from the original study were used. Chimeric sequences were
-identified and removed using VSEARCH (47). The curated sequences were
-assigned to OTUs at 97\% similarity using the OptiClust algorithm (48)
+identified and removed using VSEARCH (48). The curated sequences were
+assigned to OTUs at 97\% similarity using the OptiClust algorithm (49)
 and classified to the deepest taxonomic level that had 80\% support
 using the naïve Bayesian classifier trained on the RDP taxonomy outline
-(version 14, (49)).
+(version 14, (50)).
 
@@
 to be positive. We measured the dissimilarity between individuals by
-calculating the pairwise Bray-Curtis index and used PERMANOVA (50) to
+calculating the pairwise Bray-Curtis index and used PERMANOVA (51) to
 test whether individuals with normal colons were significantly different
@@
 or carcinoma and corrected for multiple comparisons using the
-Benjamini-Hochberg method (51). Again, for each taxon, if the relative
+Benjamini-Hochberg method (52). Again, for each taxon, if the relative
 abundance was greater than the median relative abundance for that taxon
@@
 (\(\sqrt{total~variables}\)). The \(\sqrt{total~variables}\) has been
-found to approximate the optimum mtry chosen after model tuneing (52).
+found to approximate the optimum mtry chosen after model tuneing (53).
 All fecal models were built using a 10-fold cross validation (CV) while
@@
 \textbf{\emph{Statistical Analysis.}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package (53). For
+sequence processing utilized the R (v3.4.3) software package (54). For
 OTU richness, evenness, and Shannon diversity analysis, values were
-power transformed using the rcompanion (v1.11.1) package (54) and
-Z-score normalized using the car (v2.1.6) package (55). Testing for OTU
+power transformed using the rcompanion (v1.11.1) package (55) and
+Z-score normalized using the car (v2.1.6) package (56). Testing for OTU
 richness, evenness, and Shannon diversity differences utilized linear
@@
 (tissue only), and 16S rRNA gene sequence region used using the lme4
-(v1.1.15) package (56). ORs were analyzed using both the epiR (v0.9.93)
-and metafor (v2.0.0) packages (57, 58) by assessing how many individuals
+(v1.1.15) package (57). ORs were analyzed using both the epiR (v0.9.93)
+and metafor (v2.0.0) packages (58, 59) by assessing how many individuals
 with and without disease were above and below the overall median value
@@
 tumor-associated differences in structure with the vegan (v2.4.5)
-package (59). Random Forest models were built using both the caret
-(v6.0.78) and randomForest (v4.6.12) packages (60, 61). All figures were
-created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages (62,
-63).
+package (60). Random Forest models were built using both the caret
+(v6.0.78) and randomForest (v4.6.12) packages (61, 62). All figures were
+created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages (63,
+64).
 
@@
 
+\hypertarget{ref-Repass2018}{}
+36. \textbf{Repass, J.}, \textbf{E. Iorns}, \textbf{A. Denis},
+\textbf{S. R. Williams}, \textbf{N. Perfito}, and \textbf{T. M. E. and}.
+2018. Replication study: Fusobacterium nucleatum infection is prevalent
+in human colorectal carcinoma. eLife. eLife Sciences Organisation, Ltd.
+\textbf{7}.
+
 \hypertarget{ref-Salter_contamination_2014}{}
-36. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
+37. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
 \textbf{S. T. Calus}, \textbf{W. O. Cookson}, \textbf{M. F. Moffatt},
@@
 \hypertarget{ref-Purcell2017}{}
-37. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
+38. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
 \textbf{L. Dixon}, \textbf{F. A. Frizelle}, and \textbf{J. I. Keenan}.
@@
 \hypertarget{ref-Sze2016}{}
-38. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
+39. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
 signal in the noise: Revisiting obesity and the microbiome. mBio.
@@
 \hypertarget{ref-Walters2014}{}
-39. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
+40. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
 2014. Meta-analyses of human gut microbes associated with obesity and
@@
 \hypertarget{ref-Finucane2014}{}
-40. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
+41. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
 Laurent}, and \textbf{K. S. Pollard}. 2014. A taxonomic signature of
@@
 \hypertarget{ref-Hanage2014}{}
-41. \textbf{Hanage, W. P.} 2014. Microbiology: Microbiome science needs
+42. \textbf{Hanage, W. P.} 2014. Microbiology: Microbiome science needs
 a healthy dose of scepticism. Nature. Springer Nature
@@
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-42. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
+43. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
 \textbf{B. Jovov}, and \textbf{X. Han}. 2015. The gastrointestinal
@@
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-43. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
+44. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
 Epidemiologic studies of the human microbiome and cancer. British
@@
 \hypertarget{ref-kostic_genomic_2012}{}
-44. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
+45. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
 Pedamallu}, \textbf{M. Michaud}, \textbf{F. Duke}, \textbf{A. M. Earl},
@@
 \hypertarget{ref-zackular_human_2014}{}
-45. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
+46. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
 Ruffin}, and \textbf{P. D. Schloss}. 2014. The human gut microbiome as a
@@
 \hypertarget{ref-schloss_introducing_2009}{}
-46. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
+47. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
 Ryabin}, \textbf{J. R. Hall}, \textbf{M. Hartmann}, \textbf{E. B.
@@
 \hypertarget{ref-rognes_vsearch_2016}{}
-47. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
+48. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
 \textbf{C. Quince}, and \textbf{F. Mahé}. 2016. VSEARCH: A versatile
@@
 \hypertarget{ref-westcott_opticlust_2017}{}
-48. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
+49. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
 OptiClust, an Improved Method for Assigning Amplicon-Based Sequence Data
@@
 \hypertarget{ref-rdp_Wang2007}{}
-49. \textbf{Wang, Q.}, \textbf{G. M. Garrity}, \textbf{J. M. Tiedje},
+50. \textbf{Wang, Q.}, \textbf{G. M. Garrity}, \textbf{J. M. Tiedje},
 and \textbf{J. R. Cole}. 2007. Naive bayesian classifier for rapid
@@
 \hypertarget{ref-permanova_Anderson2013}{}
-50. \textbf{Anderson, M. J.}, and \textbf{D. C. I. Walsh}. 2013.
+51. \textbf{Anderson, M. J.}, and \textbf{D. C. I. Walsh}. 2013.
 PERMANOVA, ANOSIM, and the mantel test in the face of heterogeneous
@@
 \hypertarget{ref-benjamini_controlling_1995}{}
-51. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
+52. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
 the false discovery rate: A practical and powerful approach to multiple
@@
 \hypertarget{ref-Breiman2001}{}
-52. \textbf{Breiman, L.} 2001. Machine Learning. Springer Nature
+53. \textbf{Breiman, L.} 2001. Machine Learning. Springer Nature
 \textbf{45}:5--32.
@@
 \hypertarget{ref-r_citation_2017}{}
-53. \textbf{R Core Team}. 2017. R: A language and environment for
+54. \textbf{R Core Team}. 2017. R: A language and environment for
 statistical computing. R Foundation for Statistical Computing, Vienna,
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-54. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
+55. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
 extension education program evaluation.
@@
 \hypertarget{ref-car_citation_2011}{}
-55. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
+56. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
 applied regressionSecond. Sage, Thousand Oaks CA.
@@
 \hypertarget{ref-lme4_citation_2015}{}
-56. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
+57. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
 \textbf{S. Walker}. 2015. Fitting linear mixed-effects models using
@@
 \hypertarget{ref-epir_citation_2017}{}
-57. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
+58. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
 Heuer}, \textbf{J. Marshall}, \textbf{J. Sanchez}, \textbf{R. Thornton},
@@
 \hypertarget{ref-metafor_citation_2010}{}
-58. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
+59. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
 the metafor package. Journal of Statistical Software \textbf{36}:1--48.
@@
 \hypertarget{ref-vegan_citation_2017}{}
-59. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
+60. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
 \textbf{R. Kindt}, \textbf{P. Legendre}, \textbf{D. McGlinn}, \textbf{P.
@@
 \hypertarget{ref-caret_citation_2017}{}
-60. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
+61. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
 Williams}, \textbf{C. Keefer}, \textbf{A. Engelhardt}, \textbf{T.
@@
 \hypertarget{ref-randomforest_citation_2002}{}
-61. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
+62. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-62. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
+63. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
 analysis. Springer-Verlag New York.
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-63. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
+64. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
 ``grid'' graphics.

### 2018-03-15 submission/manuscript.Rmd
update to include normal not control fig 1

@@
 
-**Table 1: Total Individuals in each Study Included in the Fecal Analysis**
+**Table 1: Characteristics of the datasets included in the fecal-based analysis**
 
@@
 
-**Table 2: Studies with Tissue Samples Included in the Analysis**
+**Table 2: Characteristics of the datasets included in the tissue-based analyses**
 
@@
 
-**Figure 1: Significant Bacterial Community Metrics for Adenoma or Carcinoma in Feces** A) Adenoma evenness.
-B) Carcinoma evenness.
-C) Carcinoma Shannon diversity.
-Blue represents controls and red represents either adenoma (panel A) or carcinoma (panel B and C).
-The black lines represent the median value for each repsective group.
+**Figure 1: Comparison of alpha diversity indices that were significant between individuals with normal colons, and those with adenomas or carcinomas using data collected from fecal samples** A) Comparison of evenness between individuals with normal colons and adenomas.
+B) Comparison of evenness between individuals with normal colons and carcinomas.
+C) Comparison of Shannon diversity between individuals with normal colons and carcinomas.
+Blue points represent individuals with normal colons and red points represent individuals with either adenomas (panel A) or carcinomas (panel B and C).
+The black lines represent the median value for each group.
 
-**Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Feces.** A) Community-based odds ratio for adenoma.
-B) Community-based odds ratio for carcinoma.
-Colors represent the different variable regions used within the respective study.
+**Figure 2: Comparison of odds ratios calculated using alpha diversity community metrics associated with the presence of adenomas (A) or carcinoma (B) relative to those in individuals with normal colons using data collected from stool samples.**
 
-**Figure 3: The AUC of Indivdiual Significant OR Taxa to classify Carcinoma.** A) Fecal samples.
-B) Unmatched tissue samples.
-The larger circle represents the median AUC of all studies and the smaller circles represent the individual AUC for a particular study.
+**Figure 3: AUC values when classifing individuals as having normal colons or carcinomas using taxa with significant ORs when using stool samples (A) and unmatched tissue samples (B).** We did not identify any taxa as having a significant OR to differentiate individuals with normal colons and adenomas or using matched tissue samples.
+The large black circles represent the median AUC of all studies and the smaller circles represent the individual AUC for a particular study.
 The dotted line denotes an AUC of 0.5.
 
-**Figure 4: Most Important Members in Significant OR Taxa Carcinoma Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest fecal-based models.
-B) Common taxa in the top 10 percent for carcinoma Random Forest unmatched tissue-based models.
-Blue represents less important and red represents more important to the Random Forest Model.
-White means that particular taxa was not in the top 10%.
+**Figure 4: Comparison of Random Forest modeling approaches to classify individuals as having normal colons or adenomas (A) or carcinomas (B) when training the models using the taxa with significant ORs, all taxa in a community, or all OTUs in a community when using stool samples.** No taxa had a significant OR associated with the presence of adenomas using stool samples.
+The black line represents the median AUC for the respective group.
+The dashed gray line indicates an AUC of 0.5.
 
-**Figure 5: Fecal Sample Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
-B) Carcinoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa.
-The black line represents the median AUC for the respective group.
-If no values are present in the singificant OR taxa group then there were no significant taxa identified and no model was tested.
+**Figure 5: Relative importance of taxa with significant ORs in Random Forest models for differentiating between individuals with normal colons and carcinomas using stool samples (A) or unmatched tissue samples (B).** The colors indicate the z-transformed (i.e.
+mean of 0.0 and standard deviation of 1.0) mean decrease in accuracy values calculated from the model for each study.
+The taxa are ranked by their mean z-score-transformed mean decrease in accuracy.
 
-**Figure 6: Fecal Sample Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
-B) Test AUCs of carcinoma models using all genera or significant OR taxa only.
-The black line represents the AUC at 0.5.
-The red lines represent the median AUC of all test AUCs for a specific study.
+**Figure 6: Testing of Random Forest models to classify individuals as having normal colons or adenomas (A) or carcinomas (B) when using sequence data obtained from stool samples.** Models were trained on data from each study (Figure 5) and tested on the other studies.
+The black lines represent the median AUC of all test AUCs for a specific study.
+The dashed gray line represents the AUC at 0.5.
 
@@
 
-**Figure S1: Odds Ratio for Adenoma or Carcinoma based on Bacterial Community Metrics in Tissue.** A) Community-based odds ratio for adenoma.
-B) Community-based odds ratio for carcinoma.
-Colors represent the different variable regions used within the respective study.
+**Figure S1:  Comparison of Odds Ratios associated with normal colons or adenomas (A) or carcinomas (B) calculated using alpha diversity indices with sequence data generated from tissue samples.** The pooled results are from the aggregation of data across all studies.
+The horizontal lines indicate the 95% confidence interval for the OR.
 
-**Figure S2: Tissue Random Forest Model Train AUCs.** A) Adenoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
-B) Carcinoma random forest model AUCs between all genera, all OTU, and select model based on significant OR taxa in unmatched and matched tissue.
+**Figure S2: Relative importance of taxa (A) and OTUs (B) in Random Forest models for differentiating between individuals with normal colons and carcinomas using stool samples.** These taxa and OTUs were among the top 10% most important features in each model.
+The colors indicate the z-transformed (i.e.
+mean of 0.0 and standard deviation of 1.0) mean decrease in accuracy values calculated from the model for each study.
+The taxa are ranked by their mean z-score-transformed mean decrease in accuracy.
+
+**Figure S3: Relative importance of taxa (A, B) and OTUs (C, D) in Random Forest models for differentiating between individuals with normal colons and carcinomas using matched (A, C) and unmatched (B, D) tissue samples.** hese taxa and OTUs were among the top 10% most important features in each model.
+The colors indicate the z-transformed (i.e.
+mean of 0.0 and standard deviation of 1.0) mean decrease in accuracy values calculated from the model for each study.
+The taxa are ranked by their mean z-score-transformed mean decrease in accuracy.
+
+**Figure S4: Comparison of Random Forest modeling approaches to classify individuals as having normal colons or adenomas (A) or carcinomas (B) when training the models using the taxa with significant ORs, all taxa in a community, or all OTUs in a community when using data from tissue samples.** No taxa had a significant OR associated with the presence of adenomas using tissue samples.
 The black line represents the median AUC for the respective group.
-If no values are present in the singificant OR taxa group then there were no significant taxa identified and no model was tested.
+The dashed gray line indicates an AUC of 0.5.
 
-**Figure S3: Most Common Taxa Across Carcinoma Full Community Fecal Sample Study Models.** A) Common taxa in the top 10 percent for carcinoma Random Forest all taxa-based models.
-B) Common taxa in the top 10 percent for carcinoma Random Forest all OTU-based models.
-Blue represents less important and red represents more important to the Random Forest Model.
-White means that particular taxa was not in the top 10%.
-
-**Figure S4: Most Common Genera Across Full Community Tissue Study Models.** A) Common genera in the top 10 percent for matched carcinoma Random Forest all genera-based models.
-B) Common genera in the top 10 percent for unmatched carcinoma Random Forest all genera-based models.
-C) Common genera in the top 10 percent for matched carcinoma Random Forest all OTU-based models.
-D) Common genera in the top 10 percent for unmatched carcinoma Random Forest all OTU-based models.
-Blue represents less important and red represents more important to the Random Forest Model.
-White means that particular taxa was not in the top 10%.
-
-**Figure S5: Tissue Random Forest Genus-Based Model Test AUCs.** A) Test AUCs of adenoma models using all genera across study.
-B) Test AUCs of carcinoma models using all genera for matched tissue studies.
-C) Test AUCs of carcinoma models using all genera or significant OR taxa only for unmatched tissue studies The black line represents the AUC at 0.5.
-The red lines represent the median AUC of all test AUCs for a specific study.
+**Figure S5: Testing of Random Forest models to classify individuals as having normal colons or adenomas (A) or carcinomas (B, C) when using sequence data obtained from tissue samples.** Models were trained on data from each study (Figure S5) and tested on the other studies.
+The black lines represent the median AUC of all test AUCs for a specific study.
+The dashed gray line represents the AUC at 0.5.
 

### 2018-03-15 submission/supplemental.Rmd
update to include normal not control fig 1

--- 2018-02-14 submission/supplemental.Rmd
@@
 
-**Table S1: Bray-Curtis Distance Matrix Analyzed by PERMANOVA for Stooland Tissue**
+**Table S1: Comparison of odds ratios calculated using alpha diversity community metrics associated with the presence of adenomas or carcinoma relative to those in individuals with normal colons using data collected from tissue samples.**
+
+\footnotesize
+
+```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
+
+alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_OR_tissue_composite.csv")) %>% 
+  mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = round(pvalue, digits = 2), 
+         measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
+         comparison = "Adenoma", 
+         tissue_type = "Combined")
+
+colnames(alpha_RR_adn_tissue) <- c("Odds Ratio", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "Measure", 
+                                   "Tumor", "Tissue Group")
+
+alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_unmatched_tissue_composite.csv")) %>%   mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = round(pvalue, digits = 2), 
+         measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
+         comparison = "Carcinoma", 
+         tissue_type = "Unmatched")
+
+colnames(alpha_RR_crc_unmatch_tissue) <- c("Odds Ratio", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "Measure", 
+                                           "Tumor", "Tissue Group")
+
+alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_matched_tissue_composite.csv")) %>% 
+  mutate(rr = round(rr, digits = 2), 
+         ci_lb = round(ci_lb, digits = 2), 
+         ci_ub = round(ci_ub, digits = 2), 
+         pvalue = round(pvalue, digits = 2), 
+         measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
+         comparison = "Carcinoma", 
+         tissue_type = "Matched")
+
+colnames(alpha_RR_crc_match_tissue) <-  c("Odds Ratio", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "Measure", 
+                                          "Tumor", "Tissue Group")
+
+combined_table_S3 <- alpha_RR_adn_tissue %>% bind_rows(alpha_RR_crc_unmatch_tissue, alpha_RR_crc_match_tissue)
+
+kable(combined_table_S3, format = "markdown", align = 'c', padding = 2)
+```
+
+\normalsize
+
+\newpage
+
+**Table S2: Comparison of community dissimilarity between individuals with normal colons and those with adenomas and carcinomas as calculated using Bray-Curtis distance and tested using PERMANOVA.**
 
@@
          comparison = "Adenoma", 
-         sample_type = "Stool") %>% 
-  select(fstat, r2, pvalue, study, comparison, sample_type)
+         sample_type = "Feces") %>% 
+  select(study, comparison, sample_type, r2, pvalue)
 
-colnames(table_s1) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
+colnames(table_s1) <- c("Study", "Tumor", "Sample Type", "R2", "P-value")
 
@@
          comparison = "Carcinoma", 
-         sample_type = "Stool") %>% 
-  select(fstat, r2, pvalue, study, comparison, sample_type)
+         sample_type = "Feces") %>% 
+  select(study, comparison, sample_type, r2, pvalue)
 
-colnames(table_s2) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
+colnames(table_s2) <- c("Study", "Tumor", "Sample Type", "R2", "P-value")
 
@@
          sample_type = "Tissue") %>% 
-  select(fstat, r2, pvalue, study, comparison, sample_type)
+  select(study, comparison, sample_type, r2, pvalue)
 
-colnames(table_s3) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
+colnames(table_s3) <- c("Study", "Tumor", "Sample Type", "R2", "P-value")
 
@@
          sample_type = "Tissue") %>% 
-  select(fstat, r2, pvalue, study, comparison, sample_type)
+  select(study, comparison, sample_type, r2, pvalue)
 
-colnames(table_s4) <- c("F statistic", "R2", "P-value", "Study", "Tumor", "Sample Type")
+colnames(table_s4) <- c("Study", "Tumor", "Sample Type", "R2", "P-value")
 
@@
 
-**Table S2: Alpha Diversity Metrics OR in Tissue**
-
-\footnotesize
-
-```{r supplemental_table3, echo=FALSE, message=FALSE, warning=FALSE}
-
-alpha_RR_adn_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_adn_OR_tissue_composite.csv")) %>% 
-  mutate(rr = round(rr, digits = 2), 
-         ci_lb = round(ci_lb, digits = 2), 
-         ci_ub = round(ci_ub, digits = 2), 
-         pvalue = round(pvalue, digits = 2), 
-         measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
-         comparison = "Adenoma", 
-         tissue_type = "Combined")
-
-colnames(alpha_RR_adn_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure", 
-                                   "Tumor", "Tissue Group")
-
-alpha_RR_crc_unmatch_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_unmatched_tissue_composite.csv")) %>%   mutate(rr = round(rr, digits = 2), 
-         ci_lb = round(ci_lb, digits = 2), 
-         ci_ub = round(ci_ub, digits = 2), 
-         pvalue = round(pvalue, digits = 2), 
-         measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
-         comparison = "Carcinoma", 
-         tissue_type = "Unmatched")
-
-colnames(alpha_RR_crc_unmatch_tissue) <- c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure", 
-                                           "Tumor", "Tissue Group")
-
-alpha_RR_crc_match_tissue <- as.data.frame(read_csv("../data/process/tables/alpha_OR_matched_tissue_composite.csv")) %>% 
-  mutate(rr = round(rr, digits = 2), 
-         ci_lb = round(ci_lb, digits = 2), 
-         ci_ub = round(ci_ub, digits = 2), 
-         pvalue = round(pvalue, digits = 2), 
-         measure = c("OTU Richness", "Shannon Diversity", "Evenness"), 
-         comparison = "Carcinoma", 
-         tissue_type = "Matched")
-
-colnames(alpha_RR_crc_match_tissue) <-  c("Odds Ratio", "Lower Bound", "Upper Bound", "P-value", "Measure", 
-                                          "Tumor", "Tissue Group")
-
-combined_table_S3 <- alpha_RR_adn_tissue %>% bind_rows(alpha_RR_crc_unmatch_tissue, alpha_RR_crc_match_tissue)
-
-kable(combined_table_S3, format = "markdown", align = 'c', padding = 2)
-```
-
-\normalsize
-
-\newpage
-
-**Table S3: OR of Tumor in Stool by Individual Genera with Initial P-Value under 0.05**
+**Table S3: ORs for individual taxa associated with individuals who had a normal colon or adenomas or carcinomas using data collected from stool.** The listed P-values were less than 0.05 prior to using a Benjimani-Hochberg correction for multiple comparisons.
 
@@
          comparison = "Adenoma") %>% 
-  select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison)
+  select(measure, comparison, rr, ci_lb, ci_ub, pvalue, bh)
 
-colnames(temp_8) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Taxa ID", "Tumor")
+colnames(temp_8) <- c("Taxon", "Tumor", "OR", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "BH")
 
@@
          comparison = "Carcinoma") %>% 
-  select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison)
+  select(measure, comparison, rr, ci_lb, ci_ub, pvalue, bh)
 
-colnames(temp_10) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Taxa ID", "Tumor")
+colnames(temp_10) <- c("Taxon", "Tumor", "OR", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "BH")
 
@@
 
-**Table S4: OR of Tumor in Tissue by Individual Genera with Initial P-Value under 0.05**
+\blandscape
+
+**Table S4: ORs for individual taxa associated with individuals who had a normal colon or adenomas or carcinomas using data collected from tissue samples.** The listed P-values were less than 0.05 prior to using a Benjimani-Hochberg correction for multiple comparisons.
 
@@
          tissue_type = "Combined") %>% 
-  select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison, tissue_type)
+  select(measure, comparison, tissue_type, rr, ci_lb, ci_ub, pvalue, bh)
 
-colnames(temp_9) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Taxa ID", 
-                      "Tumor", "Tissue Group")
+colnames(temp_9) <- c("Taxon", "Tumor", "Tissue Group", "OR", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "BH")
 
@@
          tissue_type = "Unmatched") %>% 
-  select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison, tissue_type)
+  select(measure, comparison, tissue_type, rr, ci_lb, ci_ub, pvalue, bh)
 
-colnames(temp_11) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Taxa ID", 
-                      "Tumor", "Tissue Group")
+colnames(temp_11) <- c("Taxon", "Tumor", "Tissue Group", "OR", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "BH")
 
@@
          tissue_type = "Matched") %>% 
-  select(rr, ci_lb, ci_ub, pvalue, bh, measure, comparison, tissue_type)
+  select(measure, comparison, tissue_type, rr, ci_lb, ci_ub, pvalue, bh)
 
-colnames(temp_12) <- c("OR", "Lower Bound", "Upper Bound", "P-value", "BH", "Lowest Taxa ID", 
-                      "Tumor", "Tissue Group")
+colnames(temp_12) <- c("Taxon", "Tumor", "Tissue Group", "OR", "95% CI (Lower Bound)", "95% CI (Upper Bound)", "P-value", "BH")
 
@@
 
+\elandscape
 \newpage

### 2018-03-15 code/make_stool_alpha_OR_graph.R
change colors and control to normal

@@ -29,7 +29,7 @@ crc_all_stool <- read_csv("data/process/tables/alpha_OR_composite.csv") %>%
 
 ### Used the `viridis_pal()(7)` to choose colors to use
 #### zeller = V4 - #FFB90F
-#### hale = V3-V5 - #443A83FF
+#### hale = V3-V5 - #FF00FF
 #### brim = V1-V3 - #31688EFF
 #### baxter = V4 - #FFB90F
 #### weir = V4 - #FFB90F
@@ -37,9 +37,9 @@ crc_all_stool <- read_csv("data/process/tables/alpha_OR_composite.csv") %>%
 ### flemer = V3-V4 - #35B779FF
 ### ahn = V3-V4 - #35B779FF
 ### lu = V3-V4 - #35B779FF
-### dejea = V3-V5 - #443A83FF
-### geng = V1-V2 - #8FD744FF
-### sana = V1-V2 - #8FD744FF
+### dejea = V3-V5 - #FF00FF
+### geng = V1-V2 - #B0171F
+### sana = V1-V2 - #B0171F
 ### burns = V5-V6 - #FDE725FF
 ### chen = V1-V3 - #31688EFF
 
@@ -64,7 +64,7 @@ adn_stool_graph <- adn_all_stool %>%
   geom_point(size = 3, show.legend = T) + 
   facet_grid(. ~ measure) + 
   labs(x = expression(Log["2"]~Odds~Ratio), y = "") + theme_bw() + ggtitle("A") + 
-  scale_color_manual(name = "Variable Region", values = c( '#000000', '#FFB90F', '#443A83FF', '#31688EFF')) + 
+  scale_color_manual(name = "Variable Region", values = c( '#000000', '#FFB90F', '#FF00FF', '#31688EFF')) + 
   annotate("text", label = paste("Adenoma (Feces)"), x = -2.75, y = 5.5, size = 2.5) + 
   theme(plot.title = element_text(face="bold", hjust = -0.07, size = 20), 
         panel.grid.major = element_blank(), 
@@ -93,7 +93,7 @@ crc_stool_graph <- crc_all_stool %>%
   facet_grid(. ~ measure) + 
   labs(x = expression(Log["2"]~Odds~Ratio), y = "") + theme_bw() + ggtitle("B") + 
   scale_color_manual(name = "Variable Region", 
-                     values = c('#000000', '#FFC1C1', '#FFB90F', '#35B779FF', '#443A83FF')) + 
[…]

### 2018-03-15 code/make_stool_alpha_raw_graph.R

@@ -133,7 +133,7 @@ stool_data <- get_combined_table(c(stool_sets, both_sets), "stool") %>%
                         labels = c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller")), 
          disease = factor(disease, 
                           levels = c("control", "cancer"), 
-                          labels = c("Control", "Carcinoma")))
+                          labels = c("Normal", "Carcinoma")))
 
 adn_stool_data <- get_combined_table(c(stool_sets, both_sets), "stool") %>% 
   bind_rows() %>% 
@@ -145,7 +145,7 @@ adn_stool_data <- get_combined_table(c(stool_sets, both_sets), "stool") %>%
                         labels = c("Baxter", "Brim", "Hale", "Zeller")), 
          disease = factor(disease, 
                           levels = c("control", "polyp"), 
-                          labels = c("Control", "Adenoma")))
+                          labels = c("Normal", "Adenoma")))
 
 
 ##############################################################################################

### 2018-03-15 code/make_tissue_alpha_OR_graph.R

@@ -20,14 +20,14 @@ crc_all_tissue <- read_csv("data/process/tables/alpha_OR_tissue_composite.csv") 
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% 
   bind_rows(read_csv("data/process/tables/alpha_OR_ind_tissue_results.csv")) %>% 
-  mutate(region = c(rep("combined", 3), rep("V3-V5", 3), rep("V1-V2", 6), rep("V5-V6", 3), 
-                    rep("V3-V4", 3), rep("V1-V3", 3)))
+  mutate(region = c(rep("combined", 3), rep("V5-V6", 3), rep("V1-V3", 3), rep("V3-V4", 3), 
+                    rep("V1-V2", 3), rep("V3-V5", 3), rep("V1-V2", 3)))
 
 
 
 ### Used the `viridis_pal()(7)` to choose colors to use
 #### zeller = V4 - #440154FF
-#### hale = V3-V5 - #443A83FF
+#### hale = V3-V5 - #FF00FF
 #### brim = V1-V3 - #31688EFF
 #### baxter = V4 - #440154FF
 #### weir = V4 - #440154FF
@@ -35,9 +35,9 @@ crc_all_tissue <- read_csv("data/process/tables/alpha_OR_tissue_composite.csv") 
 ### flemer = V3-V4 - #35B779FF
 ### ahn = V3-V4 - #35B779FF
 ### lu = V3-V4 - #35B779FF
-### dejea = V3-V5 - #443A83FF
-### geng = V1-V2 - #8FD744FF
-### sana = V1-V2 - #8FD744FF
+### dejea = V3-V5 - #FF00FF
+### geng = V1-V2 - #B0171F
+### sana = V1-V2 - #B0171F
 ### burns = V5-V6 - #FDE725FF
 ### chen = V1-V3 - #31688EFF
 
@@ -88,8 +88,8 @@ crc_tissue_graph <- crc_all_tissue %>%
   geom_point(size = 3, show.legend = T) + 
   facet_grid(. ~ measure) + 
   labs(x = expression(Log["2"]~Odds~Ratio), y = "") + theme_bw() + ggtitle("B") + 
-  scale_color_manual(name = "Variable Region", values = c('#000000', '#8FD744FF', '#31688EFF', 
-                                '#35B779FF', '#443A83FF', '#FDE725FF')) + 
[…]

### 2018-03-15 submission/manuscript.Rmd
change fig legend order

@@
 
-**Figure 4: Comparison of Random Forest modeling approaches to classify individuals as having normal colons or adenomas (A) or carcinomas (B) when training the models using the taxa with significant ORs, all taxa in a community, or all OTUs in a community when using stool samples.** No taxa had a significant OR associated with the presence of adenomas using stool samples.
+**Figure 4: Relative importance of taxa with significant ORs in Random Forest models for differentiating between individuals with normal colons and carcinomas using stool samples (A) or unmatched tissue samples (B).** The colors indicate the z-transformed (i.e.
+mean of 0.0 and standard deviation of 1.0) mean decrease in accuracy values calculated from the model for each study.
+The taxa are ranked by their mean z-score-transformed mean decrease in accuracy.
+
+**Figure 5: Comparison of Random Forest modeling approaches to classify individuals as having normal colons or adenomas (A) or carcinomas (B) when training the models using the taxa with significant ORs, all taxa in a community, or all OTUs in a community when using stool samples.** No taxa had a significant OR associated with the presence of adenomas using stool samples.
 The black line represents the median AUC for the respective group.
 The dashed gray line indicates an AUC of 0.5.
-
-**Figure 5: Relative importance of taxa with significant ORs in Random Forest models for differentiating between individuals with normal colons and carcinomas using stool samples (A) or unmatched tissue samples (B).** The colors indicate the z-transformed (i.e.
-mean of 0.0 and standard deviation of 1.0) mean decrease in accuracy values calculated from the model for each study.
-The taxa are ranked by their mean z-score-transformed mean decrease in accuracy.
 

### 2018-03-15 submission/manuscript.tex
change fig legend order

@@
 
+\usepackage{lscape}
+\newcommand{\blandscape}{\begin{landscape}}
+\newcommand{\elandscape}{\end{landscape}}
+
 \begin{document}
@@
 
-\textbf{Table 1: Total Individuals in each Study Included in the Fecal
-Analysis}
+\textbf{Table 1: Characteristics of the datasets included in the
+fecal-based analysis}
 
@@
 
-\textbf{Table 2: Studies with Tissue Samples Included in the Analysis}
+\textbf{Table 2: Characteristics of the datasets included in the
+tissue-based analyses}
 
@@
 
-\textbf{Figure 1: Significant Bacterial Community Metrics for Adenoma or
-Carcinoma in Feces} A) Adenoma evenness. B) Carcinoma evenness. C)
-Carcinoma Shannon diversity. Blue represents controls and red represents
-either adenoma (panel A) or carcinoma (panel B and C). The black lines
-represent the median value for each repsective group.
-
-\textbf{Figure 2: Odds Ratio for Adenoma or Carcinoma based on Bacterial
-Community Metrics in Feces.} A) Community-based odds ratio for adenoma.
-B) Community-based odds ratio for carcinoma. Colors represent the
-different variable regions used within the respective study.
-
-\textbf{Figure 3: The AUC of Indivdiual Significant OR Taxa to classify
-Carcinoma.} A) Fecal samples. B) Unmatched tissue samples. The larger
-circle represents the median AUC of all studies and the smaller circles
-represent the individual AUC for a particular study. The dotted line
-denotes an AUC of 0.5.
-
-\textbf{Figure 4: Most Important Members in Significant OR Taxa
-Carcinoma Models.} A) Common taxa in the top 10 percent for carcinoma
-Random Forest fecal-based models. B) Common taxa in the top 10 percent
-for carcinoma Random Forest unmatched tissue-based models. Blue
-represents less important and red represents more important to the
-Random Forest Model. White means that particular taxa was not in the top
-10\%.
-
-\textbf{Figure 5: Fecal Sample Random Forest Model Train AUCs.} A)
-Adenoma random forest model AUCs between all genera, all OTU, and select
-model based on significant OR taxa. B) Carcinoma random forest model
-AUCs between all genera, all OTU, and select model based on significant
-OR taxa. The black line represents the median AUC for the respective
-group. If no values are present in the singificant OR taxa group then
-there were no significant taxa identified and no model was tested.
-
-\textbf{Figure 6: Fecal Sample Random Forest Genus-Based Model Test
-AUCs.} A) Test AUCs of adenoma models using all genera across study. B)
-Test AUCs of carcinoma models using all genera or significant OR taxa
-only. The black line represents the AUC at 0.5. The red lines represent
-the median AUC of all test AUCs for a specific study.
+\textbf{Figure 1: Comparison of alpha diversity indices that were
+significant between individuals with normal colons, and those with
+adenomas or carcinomas using data collected from fecal samples} A)
+Comparison of evenness between individuals with normal colons and
+adenomas. B) Comparison of evenness between individuals with normal
+colons and carcinomas. C) Comparison of Shannon diversity between
+individuals with normal colons and carcinomas. Blue points represent
+individuals with normal colons and red points represent individuals with
+either adenomas (panel A) or carcinomas (panel B and C). The black lines
+represent the median value for each group.
+
+\textbf{Figure 2: Comparison of odds ratios calculated using alpha
+diversity community metrics associated with the presence of adenomas (A)
+or carcinoma (B) relative to those in individuals with normal colons
+using data collected from stool samples.}
+
+\textbf{Figure 3: AUC values when classifing individuals as having
+normal colons or carcinomas using taxa with significant ORs when using
+stool samples (A) and unmatched tissue samples (B).} We did not identify
+any taxa as having a significant OR to differentiate individuals with
+normal colons and adenomas or using matched tissue samples. The large
+black circles represent the median AUC of all studies and the smaller
+circles represent the individual AUC for a particular study. The dotted
+line denotes an AUC of 0.5.
+
+\textbf{Figure 4: Relative importance of taxa with significant ORs in
+Random Forest models for differentiating between individuals with normal
+colons and carcinomas using stool samples (A) or unmatched tissue
+samples (B).} The colors indicate the z-transformed (i.e.~mean of 0.0
+and standard deviation of 1.0) mean decrease in accuracy values
+calculated from the model for each study. The taxa are ranked by their
+mean z-score-transformed mean decrease in accuracy.
+
+\textbf{Figure 5: Comparison of Random Forest modeling approaches to
+classify individuals as having normal colons or adenomas (A) or
+carcinomas (B) when training the models using the taxa with significant
+ORs, all taxa in a community, or all OTUs in a community when using
+stool samples.} No taxa had a significant OR associated with the
+presence of adenomas using stool samples. The black line represents the
+median AUC for the respective group. The dashed gray line indicates an
+AUC of 0.5.
+
+\textbf{Figure 6: Testing of Random Forest models to classify
+individuals as having normal colons or adenomas (A) or carcinomas (B)
+when using sequence data obtained from stool samples.} Models were
+trained on data from each study (Figure 5) and tested on the other
+studies. The black lines represent the median AUC of all test AUCs for a
+specific study. The dashed gray line represents the AUC at 0.5.
 
@@
 
-\textbf{Figure S1: Odds Ratio for Adenoma or Carcinoma based on
-Bacterial Community Metrics in Tissue.} A) Community-based odds ratio
-for adenoma. B) Community-based odds ratio for carcinoma. Colors
-represent the different variable regions used within the respective
-study.
-
-\textbf{Figure S2: Tissue Random Forest Model Train AUCs.} A) Adenoma
-random forest model AUCs between all genera, all OTU, and select model
-based on significant OR taxa in unmatched and matched tissue. B)
-Carcinoma random forest model AUCs between all genera, all OTU, and
-select model based on significant OR taxa in unmatched and matched
-tissue. The black line represents the median AUC for the respective
-group. If no values are present in the singificant OR taxa group then
-there were no significant taxa identified and no model was tested.
-
-\textbf{Figure S3: Most Common Taxa Across Carcinoma Full Community
-Fecal Sample Study Models.} A) Common taxa in the top 10 percent for
-carcinoma Random Forest all taxa-based models. B) Common taxa in the top
-10 percent for carcinoma Random Forest all OTU-based models. Blue
-represents less important and red represents more important to the
-Random Forest Model. White means that particular taxa was not in the top
-10\%.
-
-\textbf{Figure S4: Most Common Genera Across Full Community Tissue Study
-Models.} A) Common genera in the top 10 percent for matched carcinoma
-Random Forest all genera-based models. B) Common genera in the top 10
-percent for unmatched carcinoma Random Forest all genera-based models.
-C) Common genera in the top 10 percent for matched carcinoma Random
-Forest all OTU-based models. D) Common genera in the top 10 percent for
-unmatched carcinoma Random Forest all OTU-based models. Blue represents
-less important and red represents more important to the Random Forest
-Model. White means that particular taxa was not in the top 10\%.
-
-\textbf{Figure S5: Tissue Random Forest Genus-Based Model Test AUCs.} A)
-Test AUCs of adenoma models using all genera across study. B) Test AUCs
-of carcinoma models using all genera for matched tissue studies. C) Test
-AUCs of carcinoma models using all genera or significant OR taxa only
-for unmatched tissue studies The black line represents the AUC at 0.5.
-The red lines represent the median AUC of all test AUCs for a specific
-study.
+\textbf{Figure S1: Comparison of Odds Ratios associated with normal
+colons or adenomas (A) or carcinomas (B) calculated using alpha
+diversity indices with sequence data generated from tissue samples.} The
+pooled results are from the aggregation of data across all studies. The
+horizontal lines indicate the 95\% confidence interval for the OR.
+
+\textbf{Figure S2: Relative importance of taxa (A) and OTUs (B) in
+Random Forest models for differentiating between individuals with normal
+colons and carcinomas using stool samples.} These taxa and OTUs were
+among the top 10\% most important features in each model. The colors
+indicate the z-transformed (i.e.~mean of 0.0 and standard deviation of
+1.0) mean decrease in accuracy values calculated from the model for each
+study. The taxa are ranked by their mean z-score-transformed mean
+decrease in accuracy.
+
+\textbf{Figure S3: Relative importance of taxa (A, B) and OTUs (C, D) in
+Random Forest models for differentiating between individuals with normal
+colons and carcinomas using matched (A, C) and unmatched (B, D) tissue
+samples.} hese taxa and OTUs were among the top 10\% most important
+features in each model. The colors indicate the z-transformed (i.e.~mean
+of 0.0 and standard deviation of 1.0) mean decrease in accuracy values
+calculated from the model for each study. The taxa are ranked by their
+mean z-score-transformed mean decrease in accuracy.
+
+\textbf{Figure S4: Comparison of Random Forest modeling approaches to
+classify individuals as having normal colons or adenomas (A) or
+carcinomas (B) when training the models using the taxa with significant
+ORs, all taxa in a community, or all OTUs in a community when using data
+from tissue samples.} No taxa had a significant OR associated with the
+presence of adenomas using tissue samples. The black line represents the
+median AUC for the respective group. The dashed gray line indicates an
+AUC of 0.5.
+
+\textbf{Figure S5: Testing of Random Forest models to classify
+individuals as having normal colons or adenomas (A) or carcinomas (B, C)
+when using sequence data obtained from tissue samples.} Models were
+trained on data from each study (Figure S5) and tested on the other
+studies. The black lines represent the median AUC of all test AUCs for a
+specific study. The dashed gray line represents the AUC at 0.5.
 

### 2018-03-15 submission/manuscript.Rmd
update order of supplemental figs

@@
 
-**Figure S2: Relative importance of taxa (A) and OTUs (B) in Random Forest models for differentiating between individuals with normal colons and carcinomas using stool samples.** These taxa and OTUs were among the top 10% most important features in each model.
+**Figure S2: Comparison of Random Forest modeling approaches to classify individuals as having normal colons or adenomas (A) or carcinomas (B) when training the models using the taxa with significant ORs, all taxa in a community, or all OTUs in a community when using data from tissue samples.** No taxa had a significant OR associated with the presence of adenomas using tissue samples.
+The black line represents the median AUC for the respective group.
+The dashed gray line indicates an AUC of 0.5.
+
+**Figure S3: Relative importance of taxa (A) and OTUs (B) in Random Forest models for differentiating between individuals with normal colons and carcinomas using stool samples.** These taxa and OTUs were among the top 10% most important features in each model.
 The colors indicate the z-transformed (i.e.
@@
 
-**Figure S3: Relative importance of taxa (A, B) and OTUs (C, D) in Random Forest models for differentiating between individuals with normal colons and carcinomas using matched (A, C) and unmatched (B, D) tissue samples.** hese taxa and OTUs were among the top 10% most important features in each model.
+**Figure S4: Relative importance of taxa (A, B) and OTUs (C, D) in Random Forest models for differentiating between individuals with normal colons and carcinomas using matched (A, C) and unmatched (B, D) tissue samples.** hese taxa and OTUs were among the top 10% most important features in each model.
 The colors indicate the z-transformed (i.e.
@@
 The taxa are ranked by their mean z-score-transformed mean decrease in accuracy.
-
-**Figure S4: Comparison of Random Forest modeling approaches to classify individuals as having normal colons or adenomas (A) or carcinomas (B) when training the models using the taxa with significant ORs, all taxa in a community, or all OTUs in a community when using data from tissue samples.** No taxa had a significant OR associated with the presence of adenomas using tissue samples.
-The black line represents the median AUC for the respective group.
-The dashed gray line indicates an AUC of 0.5.
 

### 2018-03-15 submission/manuscript.tex
update order of supplemental figs

@@
 
-\textbf{Figure S2: Relative importance of taxa (A) and OTUs (B) in
+\textbf{Figure S2: Comparison of Random Forest modeling approaches to
+classify individuals as having normal colons or adenomas (A) or
+carcinomas (B) when training the models using the taxa with significant
+ORs, all taxa in a community, or all OTUs in a community when using data
+from tissue samples.} No taxa had a significant OR associated with the
+presence of adenomas using tissue samples. The black line represents the
+median AUC for the respective group. The dashed gray line indicates an
+AUC of 0.5.
+
+\textbf{Figure S3: Relative importance of taxa (A) and OTUs (B) in
 Random Forest models for differentiating between individuals with normal
@@
 
-\textbf{Figure S3: Relative importance of taxa (A, B) and OTUs (C, D) in
+\textbf{Figure S4: Relative importance of taxa (A, B) and OTUs (C, D) in
 Random Forest models for differentiating between individuals with normal
@@
 
-\textbf{Figure S4: Comparison of Random Forest modeling approaches to
-classify individuals as having normal colons or adenomas (A) or
-carcinomas (B) when training the models using the taxa with significant
-ORs, all taxa in a community, or all OTUs in a community when using data
-from tissue samples.} No taxa had a significant OR associated with the
-presence of adenomas using tissue samples. The black line represents the
-median AUC for the respective group. The dashed gray line indicates an
-AUC of 0.5.
-
 \textbf{Figure S5: Testing of Random Forest models to classify

### 2018-03-15 code/make_select_imp_otu_graph.R
update with pat's comments

@@ -25,10 +25,10 @@ unmatched_tissue_mda <- read_csv("data/process/tables/ALL_genus_unmatched_tissue
   ungroup()
 
 
-stool_vars <- stool_mda %>% group_by(otu) %>% summarise(median_zscore = median(zscore)) %>% 
+stool_vars <- stool_mda %>% group_by(otu) %>% summarise(median_zscore = mean(zscore)) %>% 
   arrange(median_zscore) %>% ungroup()
 
-unmatched_tissue_vars <- unmatched_tissue_mda %>% group_by(otu) %>% summarise(median_zscore = median(zscore)) %>% 
+unmatched_tissue_vars <- unmatched_tissue_mda %>% group_by(otu) %>% summarise(median_zscore = mean(zscore)) %>% 
   arrange(median_zscore) %>% ungroup()
 
 
@@ -46,7 +46,7 @@ imp_stool_model <- stool_mda %>%
                       labels = stool_vars$otu)) %>% 
   ggplot(aes(study, otu, fill = zscore)) + 
   geom_tile(color = "white") + 
-  scale_fill_gradient2(name = "Z-Score Median MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
+  scale_fill_gradient2(name = "Z-Score MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
   theme_bw() + ggtitle("A") + 
   labs(x = "", y = "") + 
   theme(panel.grid.major = element_blank(), 
@@ -65,7 +65,7 @@ imp_unmatched_tissue_model <- unmatched_tissue_mda %>%
                       labels = unmatched_tissue_vars$otu)) %>% 
   ggplot(aes(study, otu, fill = zscore)) + 
   geom_tile(color = "white") + 
-  scale_fill_gradient2(name = "Z-Score Median MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
+  scale_fill_gradient2(name = "Z-Score MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
   theme_bw() + ggtitle("B") + 
   labs(x = "", y = "") + 
   theme(panel.grid.major = element_blank(), 

### 2018-03-15 code/make_stool_imp_otu_graph.R

@@ -28,6 +28,13 @@ stool_otu_mda <- read_csv("data/process/tables/crc_RF_otu_stool_top10_mda.csv") 
 crc_stool_sets <- 7
 
 
+stool_taxa_labs <- stool_mda %>% group_by(otu) %>% summarise(median_zscore = mean(zscore)) %>% 
+  arrange(median_zscore) %>% ungroup()
+
+stool_otu_labels <- stool_otu_mda %>% group_by(genus) %>% summarise(median_zscore = mean(zscore)) %>% 
+  arrange(median_zscore) %>% ungroup()
+
+
 ##############################################################################################
 ############################## List of code to make figures ##################################
 ##############################################################################################
@@ -36,10 +43,13 @@ crc_stool_sets <- 7
 crc_genera <- stool_mda %>% 
   mutate(study = factor(study, 
                         levels = c("ahn", "baxter", "flemer", "hale", "wang", "weir", "zeller"), 
-                        labels= c(c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller")))) %>% 
+                        labels= c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller")), 
+         otu = factor(otu, 
+                      levels = stool_taxa_labs$otu, 
+                      labels = stool_taxa_labs$otu)) %>% 
   ggplot(aes(study, otu, fill = zscore)) + 
   geom_tile(color = "white") + 
-  scale_fill_gradient2(name = "Z-Score Median MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
+  scale_fill_gradient2(name = "Z-Score MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
   theme_bw() + ggtitle("A") + 
   labs(x = "", y = "") + 
   theme(panel.grid.major = element_blank(), 
@@ -52,7 +62,10 @@ crc_genera <- stool_mda %>%
 crc_otu <- stool_otu_mda %>% 
   mutate(study = factor(study, 
                         levels = c("ahn", "baxter", "flemer", "hale", "wang", "weir", "zeller"), 
-                        labels= c(c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller")))) %>% 
+                        labels= c("Ahn", "Baxter", "Flemer", "Hale", "Wang", "Weir", "Zeller")), 
+         genus = factor(genus, 
[…]

### 2018-03-15 code/make_tissue_imp_otu_graph.R

@@ -28,6 +28,13 @@ crc_matched_genera <- read_csv("data/process/tables/adn_RF_genera_matched_tissue
   ungroup() %>% filter(otu != "NA/")
 
 
+unmatched_genera_labs <- crc_unmatched_genera %>% group_by(otu) %>% summarise(median_zscore = mean(zscore)) %>% 
+  arrange(median_zscore) %>% ungroup()
+
+matched_genera_labs <- crc_matched_genera %>% group_by(otu) %>% summarise(median_zscore = mean(zscore)) %>% 
+  arrange(median_zscore) %>% ungroup()
+
+
 crc_unmatched_otu <- read_csv("data/process/tables/crc_RF_otu_unmatched_tissue_top10_mda.csv") %>% 
   mutate(genus = str_replace_all(genus, "_unclassified", ""), 
          genus = str_replace_all(genus, "\\.", "/"), 
@@ -50,6 +57,12 @@ crc_matched_otu <- read_csv("data/process/tables/adn_RF_otu_matched_tissue_top10
   group_by(study) %>% mutate(zscore = scale(mda_median)) %>% 
   ungroup()
 
+unmatched_otu_labs <- crc_unmatched_otu %>% group_by(genus) %>% summarise(median_zscore = mean(zscore)) %>% 
+  arrange(median_zscore) %>% ungroup()
+
+matched_otu_labs <- crc_matched_otu %>% group_by(genus) %>% summarise(median_zscore = mean(zscore)) %>% 
+  arrange(median_zscore) %>% ungroup()
+
 
 crc_unmatched_sets <- 4
 crc_matched_sets <- 3
@@ -61,10 +74,13 @@ crc_matched_sets <- 3
 crc_matched_genera_graph <- crc_matched_genera %>% 
   mutate(study = factor(study, 
                         levels = c("burns", "dejea", "geng"), 
-                        labels= c("Burns", "Dejea", "Geng"))) %>% 
+                        labels= c("Burns", "Dejea", "Geng")), 
+         otu = factor(otu, 
+                      levels = matched_genera_labs$otu, 
+                      labels = matched_genera_labs$otu)) %>% 
   ggplot(aes(study, otu, fill = zscore)) + 
   geom_tile(color = "white") + 
-  scale_fill_gradient2(name = "Z-Score Median MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
+  scale_fill_gradient2(name = "Z-Score MDA", low = "blue", mid = "white", high = "red", midpoint = 0) + 
[…]

### 2018-03-15 code/make_select_imp_otu_graph.R
add na.rm step

@@ -25,10 +25,10 @@ unmatched_tissue_mda <- read_csv("data/process/tables/ALL_genus_unmatched_tissue
   ungroup()
 
 
-stool_vars <- stool_mda %>% group_by(otu) %>% summarise(median_zscore = mean(zscore)) %>% 
+stool_vars <- stool_mda %>% group_by(otu) %>% summarise(median_zscore = mean(zscore, na.rm = T)) %>% 
   arrange(median_zscore) %>% ungroup()
 
-unmatched_tissue_vars <- unmatched_tissue_mda %>% group_by(otu) %>% summarise(median_zscore = mean(zscore)) %>% 
+unmatched_tissue_vars <- unmatched_tissue_mda %>% group_by(otu) %>% summarise(median_zscore = mean(zscore, na.rm = T)) %>% 
   arrange(median_zscore) %>% ungroup()
 
 

### 2018-03-15 code/make_stool_imp_otu_graph.R

@@ -28,10 +28,10 @@ stool_otu_mda <- read_csv("data/process/tables/crc_RF_otu_stool_top10_mda.csv") 
 crc_stool_sets <- 7
 
 
-stool_taxa_labs <- stool_mda %>% group_by(otu) %>% summarise(median_zscore = mean(zscore)) %>% 
+stool_taxa_labs <- stool_mda %>% group_by(otu) %>% summarise(median_zscore = mean(zscore, na.rm = T)) %>% 
   arrange(median_zscore) %>% ungroup()
 
-stool_otu_labels <- stool_otu_mda %>% group_by(genus) %>% summarise(median_zscore = mean(zscore)) %>% 
+stool_otu_labels <- stool_otu_mda %>% group_by(genus) %>% summarise(median_zscore = mean(zscore, na.rm = T)) %>% 
   arrange(median_zscore) %>% ungroup()
 
 

### 2018-03-15 code/make_tissue_imp_otu_graph.R

@@ -28,10 +28,10 @@ crc_matched_genera <- read_csv("data/process/tables/adn_RF_genera_matched_tissue
   ungroup() %>% filter(otu != "NA/")
 
 
-unmatched_genera_labs <- crc_unmatched_genera %>% group_by(otu) %>% summarise(median_zscore = mean(zscore)) %>% 
+unmatched_genera_labs <- crc_unmatched_genera %>% group_by(otu) %>% summarise(median_zscore = mean(zscore, na.rm = T)) %>% 
   arrange(median_zscore) %>% ungroup()
 
-matched_genera_labs <- crc_matched_genera %>% group_by(otu) %>% summarise(median_zscore = mean(zscore)) %>% 
+matched_genera_labs <- crc_matched_genera %>% group_by(otu) %>% summarise(median_zscore = mean(zscore, na.rm = T)) %>% 
   arrange(median_zscore) %>% ungroup()
 
 
@@ -57,10 +57,10 @@ crc_matched_otu <- read_csv("data/process/tables/adn_RF_otu_matched_tissue_top10
   group_by(study) %>% mutate(zscore = scale(mda_median)) %>% 
   ungroup()
 
-unmatched_otu_labs <- crc_unmatched_otu %>% group_by(genus) %>% summarise(median_zscore = mean(zscore)) %>% 
+unmatched_otu_labs <- crc_unmatched_otu %>% group_by(genus) %>% summarise(median_zscore = mean(zscore, na.rm = T)) %>% 
   arrange(median_zscore) %>% ungroup()
 
-matched_otu_labs <- crc_matched_otu %>% group_by(genus) %>% summarise(median_zscore = mean(zscore)) %>% 
+matched_otu_labs <- crc_matched_otu %>% group_by(genus) %>% summarise(median_zscore = mean(zscore, na.rm = T)) %>% 
   arrange(median_zscore) %>% ungroup()
 
 

### 2018-03-15 code/make_genus_rf_auc_against_study_graph.R
make pat changes for test graphs

@@ -125,9 +125,9 @@ adn_tissue_graph <- adn_all_tissue %>%
   ggplot(aes(model, auc, color = study, group = model_type)) + 
   geom_point(size = 3.5, position = position_dodge(width = 0.3), show.legend = T) + 
   stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 
-               colour = "red", geom = "crossbar", size = 0.5, width = 0.5) + 
+               colour = "black", geom = "crossbar", size = 0.5, width = 0.5) + 
   facet_grid(. ~ model_type) + 
-  geom_hline(yintercept = 0.5, linetype = "dashed") + 
+  geom_hline(color = "gray", yintercept = 0.5, linetype = "dashed") + 
   coord_cartesian(ylim = c(0, 1.05)) + 
   labs(x = "", y = "AUC") + theme_bw() + ggtitle("A") + 
   scale_color_manual(name = "Study", 
@@ -157,9 +157,9 @@ adn_stool_graph <- adn_all_stool %>%
   ggplot(aes(model, auc, color = study, group = model_type)) + 
   geom_point(position = position_dodge(width = 0.5), size = 3.5, show.legend = T) + 
   stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 
-               colour = "red", geom = "crossbar", size = 0.5, width = 0.5) + 
+               colour = "black", geom = "crossbar", size = 0.5, width = 0.5) + 
   facet_grid(. ~ model_type) + 
-  geom_hline(yintercept = 0.5, linetype = "dashed") + 
+  geom_hline(color = "gray", yintercept = 0.5, linetype = "dashed") + 
   coord_cartesian(ylim = c(0, 1.05)) + 
   labs(x = "", y = "AUC") + theme_bw() + ggtitle("A") + 
   scale_color_manual(name = "Study", 
@@ -189,9 +189,9 @@ crc_unmatched_tissue_graph <- crc_all_unmatched_tissue %>%
   ggplot(aes(model, auc, color = study, group = model_type)) + 
   geom_point(position = position_dodge(width = 0.3), size = 3.5, show.legend = T) + 
   stat_summary(fun.y = median, fun.ymin = median, fun.ymax = median, 
-               colour = "red", geom = "crossbar", size = 0.5, width = 0.5) + 
+               colour = "black", geom = "crossbar", size = 0.5, width = 0.5) + 
[…]

### 2018-03-16 submission/manuscript.Rmd
Polishing edits

@@
 We quantified the odds ratios (ORs) for individual bacterial taxa that were associated with an individual having tumors relative to a normal colon.
-Among the fecal samples, there were no taxa that had a significant ORs associated with adenoma and there were `r length(crc_OR_stool$measure)` taxa with significant ORs associated with carcinoma.
+Among the fecal samples, there were no taxa that had significant ORs associated with adenoma and there were `r length(crc_OR_stool$measure)` taxa with significant ORs associated with carcinoma.
 Similarly, among the tissue samples, there were no taxa that had a significant OR associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs associated with carcinoma.
@@
 Numerous case-control studies have characterized the microbiota of individuals with colonic adenomas and carcinomas in an attempt to identify biomarkers of disease progression [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017].
-Because current CRC screening recommendations are poorly adhered to due to person's socioeconomic status, test invasiveness, and frequency of tests, development and validation of microbiota-associated biomarkers for CRC progression could further attempts to develop non-invasive diagnostics [@GimenoGarca2012].
+Because current CRC screening recommendations are poorly adhered to due to a person's socioeconomic status, test invasiveness, and frequency of tests, development and validation of microbiota-associated biomarkers for CRC progression could further attempts to develop non-invasive diagnostics [@GimenoGarca2012].
 
@@
 A recent meta-analysis found that 16S rRNA gene sequences from members of  *Akkermansia*, *Fusobacterium*, and *Parvimonas* were fecal biomarkers for the presence of carcinomas [@shah_leveraging_2017].
-Contrary to previous studies they found sequences similar to members of *Lactobacillus* and *Ruminococcus* to be enriched in patients with adenoma or carcinoma relative to those with normal colons [@baxter_microbiota-based_2016; @burns_virulence_2015; @zeller_potential_2014].
+Contrary to previous studies, they found sequences similar to members of *Lactobacillus* and *Ruminococcus* to be enriched in patients with adenoma or carcinoma relative to those with normal colons [@baxter_microbiota-based_2016; @burns_virulence_2015; @zeller_potential_2014].
 In addition, they found that 16S rRNA gene sequences from members of *Haemophilus*, *Methanosphaera*, *Prevotella*, and *Succinovibrio* were enriched in patients with adenomas and *Pantoea* were enriched in patients with carcinomas.
@@
 richness), the evenness of their abundance, and the overall diversity of the communities were associated with disease stage after controlling for study and variable region differences.
-In fecal samples, both evenness and diversity were significantly lower in successive disease severity categories (evenness P-value=`r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and diversity P-value=`r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`) [Figure 1]; there was no significant difference for richness (P-value=`r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`).
+In fecal samples, both evenness and diversity were significantly lower in successive disease severity categories (P-value=`r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and P-value=`r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1]; there was no significant difference for richness (P-value=`r paste(format(filter(stool_lme, alpha_metric == "sobs")[, "BH"], digits = 2, nsmall = 2))`).
 We next tested whether the lower value of these community metrics translated into significant ORs for having an adenoma or carcinoma.
@@
 ***Individual taxa are associated with significant ORs for carcinomas.***
-Next we identified those taxa that had ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas.
+We next identified those taxa that had ORs that were significantly associated with having a normal colon or the presence of adenomas or carcinomas.
 No taxa had a significant OR for the presence of adenomas when we used data collected from fecal or tissue samples (Table S3 & S4).
@@
 Of these, 4 are commonly associated with the oral cavity: *Fusobacterium* (OR=`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`)), *Parvimonas* (OR=`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Parvimonas")[, "ci_ub"], digits = 2))`)), *Porphyromonas* (OR=`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Porphyromonas")[, "ci_ub"], digits = 2))`)), and *Peptostreptococcus* (OR=`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Peptostreptococcus")[, "ci_ub"], digits = 2))`)) [Table S3].
-The other 4 were *Clostridium XI* (OR=`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_ub"], digits = 2))`)), *Enterobacteriaceae*  (OR=`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_ub"], digits = 2))`)), *Escherichia* (OR = `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_ub"], digits = 2))`)), and *Ruminococcus* (OR=`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_ub"], digits = 2))`)).
+The other 4 were *Clostridium XI* (OR=`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Clostridium_XI")[, "ci_ub"], digits = 2))`)), *Enterobacteriaceae*  (OR=`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Enterobacteriaceae_unclassified")[, "ci_ub"], digits = 2))`)), *Escherichia* (OR=`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Escherichia.Shigella")[, "ci_ub"], digits = 2))`)), and *Ruminococcus* (OR=`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "rr"], digits = 2))` (`r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_OR_stool, measure == "Ruminococcus")[, "ci_ub"], digits = 2))`)).
 Among the data collected from tissue samples, only unmatched carcinoma samples had taxa with a significant OR.
@@
 Mouth-associated genera were not significantly associated with a higher OR for carcinoma in tissue samples [Table S4].
-For example, *Fusobacterium* had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`; however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4].
+For example, *Fusobacterium* had an OR of `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "rr"], digits = 2))` (`r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_lb"], digits = 2))` - `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "ci_ub"], digits = 2))`); however, due to the small number of studies and considerable variation in the data, the Benjimani-Hochberg corrected P-value was `r paste(round(filter(crc_matched_OR_tissue, measure == "Fusobacterium")[, "bh"], digits = 2))` [Table S4].
 It is interesting to note that *Ruminococcus* and members of *Clostridium XI* in fecal samples and *Dorea* and *Blautia* in tissue had ORs that were significantly less than 1.0, which suggests that these populations are protective against the development of carcinomas.
@@
 The taxa that performed the best included *Clostridium XI*, *Ruminococcus*, and *Escherichia*.
-However, these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)` ndicating their limited value as biomarkers when used individually.
+However, these had median AUC values less than `r round(max(filter(ind_crc_combined_stool_auc, taxa %in% c("Escherichia", "Clostridium_XI", "Ruminococcus"))[, "auc"]), digits = 3)` indicating their limited value as biomarkers when used individually.
 Likewise, in unmatched tissue samples the `r length(crc_unmatched_OR_tissue$measure)` taxa with significant OR taxa had AUC values that were marginally better than one would expect by chance [Figure 3B].
@@
 
-We performed a meta-analysis to identify and validate microbiota-base biomarkers that could be used to classify individuals as having normal colons or colonic tumors using fecal or tissue samples.
+We performed a meta-analysis to identify and validate microbiota-based biomarkers that could be used to classify individuals as having normal colons or colonic tumors using fecal or tissue samples.
 To our surprise, Random Forest classification models constructed to differentiate individuals with normal colons from those with carcinomas using a subset of the community performed well relative to models constructed using the full communities.
@@
 Fecal samples represent a non-invasive approach to assess the structure of the gut microbiota and are potentially useful for diagnosing individuals as having colonic tumors.
-However, they do not reflect the structure of the mucosal microbiota [@Flynn_preprint_2017].
+However, they do not reflect the structure of the mucosal microbiota [@Flynn_2018].
 Regardless, the taxa that were the most important in the feces-based models overlapped with those from the models trained using the data from unmatched and matched colon tissue samples [Figure S3].
@@
 Because no taxa were identified as having a significant OR associated with adenomas using stool samples or tissue samples, classification models based on OR data were not constructed to classify individuals as having normal colons or adenomas.
-For all models the hyperparameters for ntree was set to 500 and the mtry set to the sqaure root of the number of taxa or OTUs within the model ($\sqrt{total~variables}$).
-The $\sqrt{total~variables}$ has been found to approximate the optimum mtry chosen after model tuneing [@Breiman2001].
+For all models the value of trees included (i.e.
+ntree) was set to 500 and the number of variables that were randomly tested (i.e.
+mtry) was set to the square root of the number of taxa or OTUs within the model.
+Using the square root of the total number of features as the number of features to test has been found to reliably approximate the optimum value after model tuning [@Breiman2001].
 All fecal models were built using a 10-fold cross validation (CV) while tissue models were built using 5-fold CV due to study sample size.
-One exception to this rule was the Weir study which was built using a 2-fold CV due to extremely low sample number.
-For significant OR and all taxa data, models were trained on a single study and then tested on the remaining studies with AUCs recorded during both train and testing phases.
-For the OTU data, 100 10-fold CVs were run to generate a range of AUCs that could be reasonably expected to occur and the average AUC from these 100 repeats was reported.
-The Mean Decrease in Accuracy (MDA), a measure of the importance of each taxon to the overall model was used to rank the taxa used in each model.
+One exception to this were the models constructed using data from the Weir study, which was built using a 2-fold CV due to the small number of samples sample.
+For models constructed based on the taxa that had a significant OR or using all of the taxa, we trained the models using a single study and then tested on the remaining studies with AUCs recorded during both train and testing phases.
+For the models constructed using OTU data, 100 10-fold CVs were run to generate a range of AUCs that could be reasonably expected to occur.
+The average AUC from these 100 repeats was reported.
+The Mean Decrease in Accuracy (MDA), a measure of the importance of each taxon to the overall model, was used to rank the taxa used in each model.
 
@@
 ***Reproducible Methods.***
-The analysis code can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2017.
+The analysis code can be found at https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018.
 Unless otherwise mentioned, the accession number of raw sequences from the studies used in this analysis can be found directly in the respective batch file in the GitHub repository or in the original manuscript.

### 2018-03-16 submission/manuscript.tex
Polishing edits

@@
 were associated with an individual having tumors relative to a normal
-colon. Among the fecal samples, there were no taxa that had a
-significant ORs associated with adenoma and there were 8 taxa with
-significant ORs associated with carcinoma. Similarly, among the tissue
-samples, there were no taxa that had a significant OR associated with
-adenoma and there were 3 taxa with significant ORs associated with
-carcinoma. Among the significant ORs, the association between individual
-taxa and tumor diagnosis was equal or below 7.11. Because individual
-taxa had limited association with tumor diagnosis, we trained Random
-Forest classification models using only the taxa that had significant
-ORs, using the entire collection of taxa found in each study, and using
+colon. Among the fecal samples, there were no taxa that had significant
+ORs associated with adenoma and there were 8 taxa with significant ORs
+associated with carcinoma. Similarly, among the tissue samples, there
+were no taxa that had a significant OR associated with adenoma and there
+were 3 taxa with significant ORs associated with carcinoma. Among the
+significant ORs, the association between individual taxa and tumor
+diagnosis was equal or below 7.11. Because individual taxa had limited
+association with tumor diagnosis, we trained Random Forest
+classification models using only the taxa that had significant ORs,
+using the entire collection of taxa found in each study, and using
 operational taxonomic units defined based on a 97\% similarity
@@
 (6, 11--17). Because current CRC screening recommendations are poorly
-adhered to due to person's socioeconomic status, test invasiveness, and
-frequency of tests, development and validation of microbiota-associated
-biomarkers for CRC progression could further attempts to develop
-non-invasive diagnostics (18).
+adhered to due to a person's socioeconomic status, test invasiveness,
+and frequency of tests, development and validation of
+microbiota-associated biomarkers for CRC progression could further
+attempts to develop non-invasive diagnostics (18).
 
@@
 fecal biomarkers for the presence of carcinomas (28). Contrary to
-previous studies they found sequences similar to members of
+previous studies, they found sequences similar to members of
 \emph{Lactobacillus} and \emph{Ruminococcus} to be enriched in patients
@@
 both evenness and diversity were significantly lower in successive
-disease severity categories (evenness P-value=0.025 and diversity
-P-value=0.043) {[}Figure 1{]}; there was no significant difference for
+disease severity categories (P-value=0.025 and P-value=0.043,
+respectively) {[}Figure 1{]}; there was no significant difference for
 richness (P-value=0.21). We next tested whether the lower value of these
@@
 \textbf{\emph{Individual taxa are associated with significant ORs for
-carcinomas.}} Next we identified those taxa that had ORs that were
+carcinomas.}} We next identified those taxa that had ORs that were
 significantly associated with having a normal colon or the presence of
@@
 (0.49 - 0.86)), \emph{Enterobacteriaceae} (OR=1.79 (1.33 - 2.41)),
-\emph{Escherichia} (OR = 2.15 (1.57 - 2.95)), and \emph{Ruminococcus}
+\emph{Escherichia} (OR=2.15 (1.57 - 2.95)), and \emph{Ruminococcus}
 (OR=0.63 (0.48 - 0.83)). Among the data collected from tissue samples,
@@
 OR for carcinoma in tissue samples {[}Table S4{]}. For example,
-\emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24; however, due to
+\emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24); however, due to
 the small number of studies and considerable variation in the data, the
@@
 \emph{Escherichia}. However, these had median AUC values less than 0.588
-ndicating their limited value as biomarkers when used individually.
+indicating their limited value as biomarkers when used individually.
 Likewise, in unmatched tissue samples the 3 taxa with significant OR
@@
 
-We performed a meta-analysis to identify and validate microbiota-base
+We performed a meta-analysis to identify and validate microbiota-based
 biomarkers that could be used to classify individuals as having normal
@@
 individuals as having normal colons or adenomas. For all models the
-hyperparameters for ntree was set to 500 and the mtry set to the sqaure
-root of the number of taxa or OTUs within the model
-(\(\sqrt{total~variables}\)). The \(\sqrt{total~variables}\) has been
-found to approximate the optimum mtry chosen after model tuneing (53).
-All fecal models were built using a 10-fold cross validation (CV) while
-tissue models were built using 5-fold CV due to study sample size. One
-exception to this rule was the Weir study which was built using a 2-fold
-CV due to extremely low sample number. For significant OR and all taxa
-data, models were trained on a single study and then tested on the
-remaining studies with AUCs recorded during both train and testing
-phases. For the OTU data, 100 10-fold CVs were run to generate a range
-of AUCs that could be reasonably expected to occur and the average AUC
-from these 100 repeats was reported. The Mean Decrease in Accuracy
-(MDA), a measure of the importance of each taxon to the overall model
-was used to rank the taxa used in each model.
+value of trees included (i.e.~ntree) was set to 500 and the number of
+variables that were randomly tested (i.e.~mtry) was set to the square
+root of the number of taxa or OTUs within the model. Using the square
+root of the total number of features as the number of features to test
+has been found to reliably approximate the optimum value after model
+tuning (53). All fecal models were built using a 10-fold cross
+validation (CV) while tissue models were built using 5-fold CV due to
+study sample size. One exception to this were the models constructed
+using data from the Weir study, which was built using a 2-fold CV due to
+the small number of samples sample. For models constructed based on the
+taxa that had a significant OR or using all of the taxa, we trained the
+models using a single study and then tested on the remaining studies
+with AUCs recorded during both train and testing phases. For the models
+constructed using OTU data, 100 10-fold CVs were run to generate a range
+of AUCs that could be reasonably expected to occur. The average AUC from
+these 100 repeats was reported. The Mean Decrease in Accuracy (MDA), a
+measure of the importance of each taxon to the overall model, was used
+to rank the taxa used in each model.
 
@@
 the Bray-Curtis dissimilarity index and PERMANOVA was used to test for
-tumor-associated differences in structure with the vegan (v2.4.5)
+tumor-associated differences in structure with the vegan (v2.4.4)
 package (60). Random Forest models were built using both the caret
-(v6.0.78) and randomForest (v4.6.12) packages (61, 62). All figures were
-created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages (63,
-64).
+(v6.0.77) and randomForest (v4.6.12) packages (61, 62). All figures were
+created using both ggplot2 (v2.2.1.9000) and gridExtra (v2.3) packages
+(63, 64).
 
 \textbf{\emph{Reproducible Methods.}} The analysis code can be found at
-\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2017}.
+\url{https://github.com/SchlossLab/Sze_CRCMetaAnalysis_mBio_2018}.
 Unless otherwise mentioned, the accession number of raw sequences from
@@
 \hypertarget{ref-siegel_cancer_2016}{}
-1. \textbf{Siegel, R. L.}, \textbf{K. D. Miller}, and \textbf{A. Jemal}.
-2016. Cancer statistics, 2016. CA: a cancer journal for clinicians
+1. \textbf{Siegel RL}, \textbf{Miller KD}, \textbf{Jemal A}. 2016.
+Cancer statistics, 2016. CA: a cancer journal for clinicians
 \textbf{66}:7--30.
+doi:\href{https://doi.org/10.3322/caac.21332}{10.3322/caac.21332}.
 
 \hypertarget{ref-flynn_metabolic_2016}{}
-2. \textbf{Flynn, K. J.}, \textbf{N. T. Baxter}, and \textbf{P. D.
-Schloss}. 2016. Metabolic and Community Synergy of Oral Bacteria in
-Colorectal Cancer. mSphere \textbf{1}.
+2. \textbf{Flynn KJ}, \textbf{Baxter NT}, \textbf{Schloss PD}. 2016.
+Metabolic and Community Synergy of Oral Bacteria in Colorectal Cancer.
+mSphere \textbf{1}.
+doi:\href{https://doi.org/10.1128/mSphere.00102-16}{10.1128/mSphere.00102-16}.
 
 \hypertarget{ref-goodwin_polyamine_2011}{}
-3. \textbf{Goodwin, A. C.}, \textbf{C. E. Destefano Shields}, \textbf{S.
-Wu}, \textbf{D. L. Huso}, \textbf{X. Wu}, \textbf{T. R. Murray-Stewart},
-\textbf{A. Hacker-Prietz}, \textbf{S. Rabizadeh}, \textbf{P. M. Woster},
-\textbf{C. L. Sears}, and \textbf{R. A. Casero}. 2011. Polyamine
-catabolism contributes to enterotoxigenic Bacteroides fragilis-induced
-colon tumorigenesis. Proceedings of the National Academy of Sciences of
-the United States of America \textbf{108}:15354--15359.
+3. \textbf{Goodwin AC}, \textbf{Destefano Shields CE}, \textbf{Wu S},
+\textbf{Huso DL}, \textbf{Wu X}, \textbf{Murray-Stewart TR},
+\textbf{Hacker-Prietz A}, \textbf{Rabizadeh S}, \textbf{Woster PM},
+\textbf{Sears CL}, \textbf{Casero RA}. 2011. Polyamine catabolism
+contributes to enterotoxigenic Bacteroides fragilis-induced colon
+tumorigenesis. Proceedings of the National Academy of Sciences of the
+United States of America \textbf{108}:15354--15359.
+doi:\href{https://doi.org/10.1073/pnas.1010203108}{10.1073/pnas.1010203108}.
 
 \hypertarget{ref-abed_fap2_2016}{}
-4. \textbf{Abed, J.}, \textbf{J. E. M. Emgård}, \textbf{G. Zamir},
-\textbf{M. Faroja}, \textbf{G. Almogy}, \textbf{A. Grenov}, \textbf{A.
-Sol}, \textbf{R. Naor}, \textbf{E. Pikarsky}, \textbf{K. A. Atlan},
-\textbf{A. Mellul}, \textbf{S. Chaushu}, \textbf{A. L. Manson},
-\textbf{A. M. Earl}, \textbf{N. Ou}, \textbf{C. A. Brennan}, \textbf{W.
-S. Garrett}, and \textbf{G. Bachrach}. 2016. Fap2 Mediates Fusobacterium
-nucleatum Colorectal Adenocarcinoma Enrichment by Binding to
-Tumor-Expressed Gal-GalNAc. Cell Host \& Microbe \textbf{20}:215--225.
+4. \textbf{Abed J}, \textbf{Emgård JEM}, \textbf{Zamir G},
+\textbf{Faroja M}, \textbf{Almogy G}, \textbf{Grenov A}, \textbf{Sol A},
+\textbf{Naor R}, \textbf{Pikarsky E}, \textbf{Atlan KA}, \textbf{Mellul
+A}, \textbf{Chaushu S}, \textbf{Manson AL}, \textbf{Earl AM}, \textbf{Ou
+N}, \textbf{Brennan CA}, \textbf{Garrett WS}, \textbf{Bachrach G}. 2016.
+Fap2 Mediates Fusobacterium nucleatum Colorectal Adenocarcinoma
+Enrichment by Binding to Tumor-Expressed Gal-GalNAc. Cell Host \&
+Microbe \textbf{20}:215--225.
+doi:\href{https://doi.org/10.1016/j.chom.2016.07.006}{10.1016/j.chom.2016.07.006}.
 
 \hypertarget{ref-arthur_intestinal_2012}{}
-5. \textbf{Arthur, J. C.}, \textbf{E. Perez-Chanona}, \textbf{M.
-Mühlbauer}, \textbf{S. Tomkovich}, \textbf{J. M. Uronis}, \textbf{T.-J.
-Fan}, \textbf{B. J. Campbell}, \textbf{T. Abujamel}, \textbf{B. Dogan},
-\textbf{A. B. Rogers}, \textbf{J. M. Rhodes}, \textbf{A. Stintzi},
-\textbf{K. W. Simpson}, \textbf{J. J. Hansen}, \textbf{T. O. Keku},
-\textbf{A. A. Fodor}, and \textbf{C. Jobin}. 2012. Intestinal
-inflammation targets cancer-inducing activity of the microbiota. Science
-(New York, N.Y.) \textbf{338}:120--123.
+5. \textbf{Arthur JC}, \textbf{Perez-Chanona E}, \textbf{Mühlbauer M},
+\textbf{Tomkovich S}, \textbf{Uronis JM}, \textbf{Fan T-J},
+\textbf{Campbell BJ}, \textbf{Abujamel T}, \textbf{Dogan B},
+\textbf{Rogers AB}, \textbf{Rhodes JM}, \textbf{Stintzi A},
+\textbf{Simpson KW}, \textbf{Hansen JJ}, \textbf{Keku TO}, \textbf{Fodor
+AA}, \textbf{Jobin C}. 2012. Intestinal inflammation targets
+cancer-inducing activity of the microbiota. Science (New York, NY)
+\textbf{338}:120--123.
+doi:\href{https://doi.org/10.1126/science.1224820}{10.1126/science.1224820}.
 
 \hypertarget{ref-kostic_fusobacterium_2013}{}
-6. \textbf{Kostic, A. D.}, \textbf{E. Chun}, \textbf{L. Robertson},
-\textbf{J. N. Glickman}, \textbf{C. A. Gallini}, \textbf{M. Michaud},
-\textbf{T. E. Clancy}, \textbf{D. C. Chung}, \textbf{P. Lochhead},
-\textbf{G. L. Hold}, \textbf{E. M. El-Omar}, \textbf{D. Brenner},
-\textbf{C. S. Fuchs}, \textbf{M. Meyerson}, and \textbf{W. S. Garrett}.
-2013. Fusobacterium nucleatum potentiates intestinal tumorigenesis and
-modulates the tumor-immune microenvironment. Cell Host \& Microbe
-\textbf{14}:207--215.
+6. \textbf{Kostic AD}, \textbf{Chun E}, \textbf{Robertson L},
+\textbf{Glickman JN}, \textbf{Gallini CA}, \textbf{Michaud M},
+\textbf{Clancy TE}, \textbf{Chung DC}, \textbf{Lochhead P}, \textbf{Hold
+GL}, \textbf{El-Omar EM}, \textbf{Brenner D}, \textbf{Fuchs CS},
+\textbf{Meyerson M}, \textbf{Garrett WS}. 2013. Fusobacterium nucleatum
+potentiates intestinal tumorigenesis and modulates the tumor-immune
+microenvironment. Cell Host \& Microbe \textbf{14}:207--215.
+doi:\href{https://doi.org/10.1016/j.chom.2013.07.007}{10.1016/j.chom.2013.07.007}.
 
 \hypertarget{ref-wu_human_2009}{}
-7. \textbf{Wu, S.}, \textbf{K.-J. Rhee}, \textbf{E. Albesiano},
-\textbf{S. Rabizadeh}, \textbf{X. Wu}, \textbf{H.-R. Yen}, \textbf{D. L.
-Huso}, \textbf{F. L. Brancati}, \textbf{E. Wick}, \textbf{F.
-McAllister}, \textbf{F. Housseau}, \textbf{D. M. Pardoll}, and
-\textbf{C. L. Sears}. 2009. A human colonic commensal promotes colon
-tumorigenesis via activation of T helper type 17 T cell responses.
-Nature Medicine \textbf{15}:1016--1022.
+7. \textbf{Wu S}, \textbf{Rhee K-J}, \textbf{Albesiano E},
+\textbf{Rabizadeh S}, \textbf{Wu X}, \textbf{Yen H-R}, \textbf{Huso DL},
+\textbf{Brancati FL}, \textbf{Wick E}, \textbf{McAllister F},
+\textbf{Housseau F}, \textbf{Pardoll DM}, \textbf{Sears CL}. 2009. A
+human colonic commensal promotes colon tumorigenesis via activation of T
+helper type 17 T cell responses. Nature Medicine \textbf{15}:1016--1022.
+doi:\href{https://doi.org/10.1038/nm.2015}{10.1038/nm.2015}.
 
 \hypertarget{ref-zackular_manipulation_2016}{}
-8. \textbf{Zackular, J. P.}, \textbf{N. T. Baxter}, \textbf{G. Y. Chen},
-and \textbf{P. D. Schloss}. 2016. Manipulation of the Gut Microbiota
-Reveals Role in Colon Tumorigenesis. mSphere \textbf{1}.
+8. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2016. Manipulation of the Gut Microbiota Reveals
+Role in Colon Tumorigenesis. mSphere \textbf{1}.
+doi:\href{https://doi.org/10.1128/mSphere.00001-15}{10.1128/mSphere.00001-15}.
 
 \hypertarget{ref-zackular_gut_2013}{}
-9. \textbf{Zackular, J. P.}, \textbf{N. T. Baxter}, \textbf{K. D.
-Iverson}, \textbf{W. D. Sadler}, \textbf{J. F. Petrosino}, \textbf{G. Y.
-Chen}, and \textbf{P. D. Schloss}. 2013. The gut microbiome modulates
-colon tumorigenesis. mBio \textbf{4}:e00692--00613.
+9. \textbf{Zackular JP}, \textbf{Baxter NT}, \textbf{Iverson KD},
+\textbf{Sadler WD}, \textbf{Petrosino JF}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2013. The gut microbiome modulates colon
+tumorigenesis. mBio \textbf{4}:e00692--00613.
+doi:\href{https://doi.org/10.1128/mBio.00692-13}{10.1128/mBio.00692-13}.
 
 \hypertarget{ref-baxter_structure_2014}{}
-10. \textbf{Baxter, N. T.}, \textbf{J. P. Zackular}, \textbf{G. Y.
-Chen}, and \textbf{P. D. Schloss}. 2014. Structure of the gut microbiome
-following colonization with human feces determines colonic tumor burden.
+10. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
+\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
+colonization with human feces determines colonic tumor burden.
 Microbiome \textbf{2}:20.
+doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.
 
 \hypertarget{ref-ahn_human_2013}{}
-11. \textbf{Ahn, J.}, \textbf{R. Sinha}, \textbf{Z. Pei}, \textbf{C.
-Dominianni}, \textbf{J. Wu}, \textbf{J. Shi}, \textbf{J. J. Goedert},
-\textbf{R. B. Hayes}, and \textbf{L. Yang}. 2013. Human gut microbiome
-and risk for colorectal cancer. Journal of the National Cancer Institute
+11. \textbf{Ahn J}, \textbf{Sinha R}, \textbf{Pei Z}, \textbf{Dominianni
+C}, \textbf{Wu J}, \textbf{Shi J}, \textbf{Goedert JJ}, \textbf{Hayes
+RB}, \textbf{Yang L}. 2013. Human gut microbiome and risk for colorectal
+cancer. Journal of the National Cancer Institute
 \textbf{105}:1907--1911.
+doi:\href{https://doi.org/10.1093/jnci/djt300}{10.1093/jnci/djt300}.
 
 \hypertarget{ref-baxter_microbiota-based_2016}{}
-12. \textbf{Baxter, N. T.}, \textbf{M. T. Ruffin}, \textbf{M. A. M.
-Rogers}, and \textbf{P. D. Schloss}. 2016. Microbiota-based model
-improves the sensitivity of fecal immunochemical test for detecting
-colonic lesions. Genome Medicine \textbf{8}:37.
+12. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
+\textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.
 
 \hypertarget{ref-chen_human_2012}{}
-13. \textbf{Chen, W.}, \textbf{F. Liu}, \textbf{Z. Ling}, \textbf{X.
-Tong}, and \textbf{C. Xiang}. 2012. Human intestinal lumen and
-mucosa-associated microbiota in patients with colorectal cancer. PloS
-One \textbf{7}:e39743.
+13. \textbf{Chen W}, \textbf{Liu F}, \textbf{Ling Z}, \textbf{Tong X},
+\textbf{Xiang C}. 2012. Human intestinal lumen and mucosa-associated
+microbiota in patients with colorectal cancer. PloS One
+\textbf{7}:e39743.
+doi:\href{https://doi.org/10.1371/journal.pone.0039743}{10.1371/journal.pone.0039743}.
 
 \hypertarget{ref-wang_structural_2012}{}
-14. \textbf{Wang, T.}, \textbf{G. Cai}, \textbf{Y. Qiu}, \textbf{N.
-Fei}, \textbf{M. Zhang}, \textbf{X. Pang}, \textbf{W. Jia}, \textbf{S.
-Cai}, and \textbf{L. Zhao}. 2012. Structural segregation of gut
-microbiota between colorectal cancer patients and healthy volunteers.
-The ISME journal \textbf{6}:320--329.
+14. \textbf{Wang T}, \textbf{Cai G}, \textbf{Qiu Y}, \textbf{Fei N},
+\textbf{Zhang M}, \textbf{Pang X}, \textbf{Jia W}, \textbf{Cai S},
+\textbf{Zhao L}. 2012. Structural segregation of gut microbiota between
+colorectal cancer patients and healthy volunteers. The ISME journal
+\textbf{6}:320--329.
+doi:\href{https://doi.org/10.1038/ismej.2011.109}{10.1038/ismej.2011.109}.
 
 \hypertarget{ref-burns_virulence_2015}{}
-15. \textbf{Burns, M. B.}, \textbf{J. Lynch}, \textbf{T. K. Starr},
-\textbf{D. Knights}, and \textbf{R. Blekhman}. 2015. Virulence genes are
-a signature of the microbiome in the colorectal tumor microenvironment.
+15. \textbf{Burns MB}, \textbf{Lynch J}, \textbf{Starr TK},
+\textbf{Knights D}, \textbf{Blekhman R}. 2015. Virulence genes are a
+signature of the microbiome in the colorectal tumor microenvironment.
 Genome Medicine \textbf{7}:55.
+doi:\href{https://doi.org/10.1186/s13073-015-0177-8}{10.1186/s13073-015-0177-8}.
 
 \hypertarget{ref-zeller_potential_2014}{}
-16. \textbf{Zeller, G.}, \textbf{J. Tap}, \textbf{A. Y. Voigt},
-\textbf{S. Sunagawa}, \textbf{J. R. Kultima}, \textbf{P. I. Costea},
-\textbf{A. Amiot}, \textbf{J. Böhm}, \textbf{F. Brunetti}, \textbf{N.
-Habermann}, \textbf{R. Hercog}, \textbf{M. Koch}, \textbf{A. Luciani},
-\textbf{D. R. Mende}, \textbf{M. A. Schneider}, \textbf{P.
-Schrotz-King}, \textbf{C. Tournigand}, \textbf{J. Tran Van Nhieu},
-\textbf{T. Yamada}, \textbf{J. Zimmermann}, \textbf{V. Benes},
-\textbf{M. Kloor}, \textbf{C. M. Ulrich}, \textbf{M. von Knebel
-Doeberitz}, \textbf{I. Sobhani}, and \textbf{P. Bork}. 2014. Potential
-of fecal microbiota for early-stage detection of colorectal cancer.
-Molecular Systems Biology \textbf{10}:766.
+16. \textbf{Zeller G}, \textbf{Tap J}, \textbf{Voigt AY},
+\textbf{Sunagawa S}, \textbf{Kultima JR}, \textbf{Costea PI},
+\textbf{Amiot A}, \textbf{Böhm J}, \textbf{Brunetti F},
+\textbf{Habermann N}, \textbf{Hercog R}, \textbf{Koch M},
+\textbf{Luciani A}, \textbf{Mende DR}, \textbf{Schneider MA},
+\textbf{Schrotz-King P}, \textbf{Tournigand C}, \textbf{Tran Van Nhieu
+J}, \textbf{Yamada T}, \textbf{Zimmermann J}, \textbf{Benes V},
+\textbf{Kloor M}, \textbf{Ulrich CM}, \textbf{Knebel Doeberitz M von},
+\textbf{Sobhani I}, \textbf{Bork P}. 2014. Potential of fecal microbiota
+for early-stage detection of colorectal cancer. Molecular Systems
+Biology \textbf{10}:766.
 
 \hypertarget{ref-flemer_tumour-associated_2017}{}
-17. \textbf{Flemer, B.}, \textbf{D. B. Lynch}, \textbf{J. M. R. Brown},
-\textbf{I. B. Jeffery}, \textbf{F. J. Ryan}, \textbf{M. J. Claesson},
-\textbf{M. O'Riordain}, \textbf{F. Shanahan}, and \textbf{P. W.
-O'Toole}. 2017. Tumour-associated and non-tumour-associated microbiota
-in colorectal cancer. Gut \textbf{66}:633--643.
+17. \textbf{Flemer B}, \textbf{Lynch DB}, \textbf{Brown JMR},
+\textbf{Jeffery IB}, \textbf{Ryan FJ}, \textbf{Claesson MJ},
+\textbf{O'Riordain M}, \textbf{Shanahan F}, \textbf{O'Toole PW}. 2017.
+Tumour-associated and non-tumour-associated microbiota in colorectal
+cancer. Gut \textbf{66}:633--643.
+doi:\href{https://doi.org/10.1136/gutjnl-2015-309595}{10.1136/gutjnl-2015-309595}.
 
 \hypertarget{ref-GimenoGarca2012}{}
-18. \textbf{García, A. Z. G.} 2012. Factors influencing colorectal
-cancer screening participation. Gastroenterology Research and Practice.
-Hindawi Limited \textbf{2012}:1--8.
+18. \textbf{García AZG}. 2012. Factors influencing colorectal cancer
+screening participation. Gastroenterology Research and Practice
+\textbf{2012}:1--8.
+doi:\href{https://doi.org/10.1155/2012/483417}{10.1155/2012/483417}.
 
 \hypertarget{ref-geng_diversified_2013}{}
-19. \textbf{Geng, J.}, \textbf{H. Fan}, \textbf{X. Tang}, \textbf{H.
-Zhai}, and \textbf{Z. Zhang}. 2013. Diversified pattern of the human
-colorectal cancer microbiome. Gut Pathogens \textbf{5}:2.
+19. \textbf{Geng J}, \textbf{Fan H}, \textbf{Tang X}, \textbf{Zhai H},
+\textbf{Zhang Z}. 2013. Diversified pattern of the human colorectal
+cancer microbiome. Gut Pathogens \textbf{5}:2.
+doi:\href{https://doi.org/10.1186/1757-4749-5-2}{10.1186/1757-4749-5-2}.
 
 \hypertarget{ref-dejea_microbiota_2014}{}
-20. \textbf{Dejea, C. M.}, \textbf{E. C. Wick}, \textbf{E. M.
-Hechenbleikner}, \textbf{J. R. White}, \textbf{J. L. Mark Welch},
-\textbf{B. J. Rossetti}, \textbf{S. N. Peterson}, \textbf{E. C.
-Snesrud}, \textbf{G. G. Borisy}, \textbf{M. Lazarev}, \textbf{E. Stein},
-\textbf{J. Vadivelu}, \textbf{A. C. Roslani}, \textbf{A. A. Malik},
-\textbf{J. W. Wanyiri}, \textbf{K. L. Goh}, \textbf{I. Thevambiga},
-\textbf{K. Fu}, \textbf{F. Wan}, \textbf{N. Llosa}, \textbf{F.
-Housseau}, \textbf{K. Romans}, \textbf{X. Wu}, \textbf{F. M.
-McAllister}, \textbf{S. Wu}, \textbf{B. Vogelstein}, \textbf{K. W.
-Kinzler}, \textbf{D. M. Pardoll}, and \textbf{C. L. Sears}. 2014.
+20. \textbf{Dejea CM}, \textbf{Wick EC}, \textbf{Hechenbleikner EM},
+\textbf{White JR}, \textbf{Mark Welch JL}, \textbf{Rossetti BJ},
+\textbf{Peterson SN}, \textbf{Snesrud EC}, \textbf{Borisy GG},
+\textbf{Lazarev M}, \textbf{Stein E}, \textbf{Vadivelu J},
+\textbf{Roslani AC}, \textbf{Malik AA}, \textbf{Wanyiri JW}, \textbf{Goh
+KL}, \textbf{Thevambiga I}, \textbf{Fu K}, \textbf{Wan F}, \textbf{Llosa
+N}, \textbf{Housseau F}, \textbf{Romans K}, \textbf{Wu X},
+\textbf{McAllister FM}, \textbf{Wu S}, \textbf{Vogelstein B},
+\textbf{Kinzler KW}, \textbf{Pardoll DM}, \textbf{Sears CL}. 2014.
 Microbiota organization is a distinct feature of proximal colorectal
@@
 States of America \textbf{111}:18321--18326.
+doi:\href{https://doi.org/10.1073/pnas.1406199111}{10.1073/pnas.1406199111}.
 
 \hypertarget{ref-ecoli_Arthur_2014}{}
-21. \textbf{Arthur, J. C.}, \textbf{R. Z. Gharaibeh}, \textbf{M.
-Mühlbauer}, \textbf{E. Perez-Chanona}, \textbf{J. M. Uronis}, \textbf{J.
-McCafferty}, \textbf{A. A. Fodor}, and \textbf{C. Jobin}. 2014.
-Microbial genomic analysis reveals the essential role of inflammation in
-bacteria-induced colorectal cancer. Nature Communications. Springer
-Nature \textbf{5}:4724.
+21. \textbf{Arthur JC}, \textbf{Gharaibeh RZ}, \textbf{Mühlbauer M},
+\textbf{Perez-Chanona E}, \textbf{Uronis JM}, \textbf{McCafferty J},
+\textbf{Fodor AA}, \textbf{Jobin C}. 2014. Microbial genomic analysis
+reveals the essential role of inflammation in bacteria-induced
+colorectal cancer. Nature Communications \textbf{5}:4724.
+doi:\href{https://doi.org/10.1038/ncomms5724}{10.1038/ncomms5724}.
 
 \hypertarget{ref-strep_Aymeric_2017}{}
-22. \textbf{Aymeric, L.}, \textbf{F. Donnadieu}, \textbf{C. Mulet},
-\textbf{L. du Merle}, \textbf{G. Nigro}, \textbf{A. Saffarian},
-\textbf{M. Bérard}, \textbf{C. Poyart}, \textbf{S. Robine}, \textbf{B.
-Regnault}, \textbf{P. Trieu-Cuot}, \textbf{P. J. Sansonetti}, and
-\textbf{S. Dramsi}. 2017. Colorectal cancer specific conditions
+22. \textbf{Aymeric L}, \textbf{Donnadieu F}, \textbf{Mulet C},
+\textbf{Merle L du}, \textbf{Nigro G}, \textbf{Saffarian A},
+\textbf{Bérard M}, \textbf{Poyart C}, \textbf{Robine S},
+\textbf{Regnault B}, \textbf{Trieu-Cuot P}, \textbf{Sansonetti PJ},
+\textbf{Dramsi S}. 2017. Colorectal cancer specific conditions
 promoteStreptococcus gallolyticusgut colonization. Proceedings of the
-National Academy of Sciences. Proceedings of the National Academy of
-Sciences \textbf{115}:E283--E291.
+National Academy of Sciences \textbf{115}:E283--E291.
+doi:\href{https://doi.org/10.1073/pnas.1715112115}{10.1073/pnas.1715112115}.
 
 \hypertarget{ref-weir_stool_2013}{}
-23. \textbf{Weir, T. L.}, \textbf{D. K. Manter}, \textbf{A. M. Sheflin},
-\textbf{B. A. Barnett}, \textbf{A. L. Heuberger}, and \textbf{E. P.
-Ryan}. 2013. Stool microbiome and metabolome differences between
-colorectal cancer patients and healthy adults. PloS One
-\textbf{8}:e70803.
+23. \textbf{Weir TL}, \textbf{Manter DK}, \textbf{Sheflin AM},
+\textbf{Barnett BA}, \textbf{Heuberger AL}, \textbf{Ryan EP}. 2013.
+Stool microbiome and metabolome differences between colorectal cancer
+patients and healthy adults. PloS One \textbf{8}:e70803.
+doi:\href{https://doi.org/10.1371/journal.pone.0070803}{10.1371/journal.pone.0070803}.
 
 \hypertarget{ref-bfrag_Boleij_2014}{}
-24. \textbf{Boleij, A.}, \textbf{E. M. Hechenbleikner}, \textbf{A. C.
-Goodwin}, \textbf{R. Badani}, \textbf{E. M. Stein}, \textbf{M. G.
-Lazarev}, \textbf{B. Ellis}, \textbf{K. C. Carroll}, \textbf{E.
-Albesiano}, \textbf{E. C. Wick}, \textbf{E. A. Platz}, \textbf{D. M.
-Pardoll}, and \textbf{C. L. Sears}. 2014. The bacteroides fragilis toxin
-gene is prevalent in the colon mucosa of colorectal cancer patients.
-Clinical Infectious Diseases. Oxford University Press (OUP)
+24. \textbf{Boleij A}, \textbf{Hechenbleikner EM}, \textbf{Goodwin AC},
+\textbf{Badani R}, \textbf{Stein EM}, \textbf{Lazarev MG}, \textbf{Ellis
+B}, \textbf{Carroll KC}, \textbf{Albesiano E}, \textbf{Wick EC},
+\textbf{Platz EA}, \textbf{Pardoll DM}, \textbf{Sears CL}. 2014. The
+bacteroides fragilis toxin gene is prevalent in the colon mucosa of
+colorectal cancer patients. Clinical Infectious Diseases
 \textbf{60}:208--215.
+doi:\href{https://doi.org/10.1093/cid/ciu787}{10.1093/cid/ciu787}.
 
 \hypertarget{ref-sanapareddy_increased_2012}{}
-25. \textbf{Sanapareddy, N.}, \textbf{R. M. Legge}, \textbf{B. Jovov},
-\textbf{A. McCoy}, \textbf{L. Burcal}, \textbf{F. Araujo-Perez},
-\textbf{T. A. Randall}, \textbf{J. Galanko}, \textbf{A. Benson},
-\textbf{R. S. Sandler}, \textbf{J. F. Rawls}, \textbf{Z. Abdo},
-\textbf{A. A. Fodor}, and \textbf{T. O. Keku}. 2012. Increased rectal
-microbial richness is associated with the presence of colorectal
-adenomas in humans. The ISME journal \textbf{6}:1858--1868.
+25. \textbf{Sanapareddy N}, \textbf{Legge RM}, \textbf{Jovov B},
+\textbf{McCoy A}, \textbf{Burcal L}, \textbf{Araujo-Perez F},
+\textbf{Randall TA}, \textbf{Galanko J}, \textbf{Benson A},
+\textbf{Sandler RS}, \textbf{Rawls JF}, \textbf{Abdo Z}, \textbf{Fodor
+AA}, \textbf{Keku TO}. 2012. Increased rectal microbial richness is
+associated with the presence of colorectal adenomas in humans. The ISME
+journal \textbf{6}:1858--1868.
+doi:\href{https://doi.org/10.1038/ismej.2012.43}{10.1038/ismej.2012.43}.
 
 \hypertarget{ref-lu_mucosal_2016}{}
-26. \textbf{Lu, Y.}, \textbf{J. Chen}, \textbf{J. Zheng}, \textbf{G.
-Hu}, \textbf{J. Wang}, \textbf{C. Huang}, \textbf{L. Lou}, \textbf{X.
-Wang}, and \textbf{Y. Zeng}. 2016. Mucosal adherent bacterial dysbiosis
-in patients with colorectal adenomas. Scientific Reports
-\textbf{6}:26337.
+26. \textbf{Lu Y}, \textbf{Chen J}, \textbf{Zheng J}, \textbf{Hu G},
+\textbf{Wang J}, \textbf{Huang C}, \textbf{Lou L}, \textbf{Wang X},
+\textbf{Zeng Y}. 2016. Mucosal adherent bacterial dysbiosis in patients
+with colorectal adenomas. Scientific Reports \textbf{6}:26337.
+doi:\href{https://doi.org/10.1038/srep26337}{10.1038/srep26337}.
 
 \hypertarget{ref-hale_shifts_2017}{}
-27. \textbf{Hale, V. L.}, \textbf{J. Chen}, \textbf{S. Johnson},
-\textbf{S. C. Harrington}, \textbf{T. C. Yab}, \textbf{T. C. Smyrk},
-\textbf{H. Nelson}, \textbf{L. A. Boardman}, \textbf{B. R. Druliner},
-\textbf{T. R. Levin}, \textbf{D. K. Rex}, \textbf{D. J. Ahnen},
-\textbf{P. Lance}, \textbf{D. A. Ahlquist}, and \textbf{N. Chia}. 2017.
-Shifts in the Fecal Microbiota Associated with Adenomatous Polyps.
-Cancer Epidemiology, Biomarkers \& Prevention: A Publication of the
-American Association for Cancer Research, Cosponsored by the American
-Society of Preventive Oncology \textbf{26}:85--94.
+27. \textbf{Hale VL}, \textbf{Chen J}, \textbf{Johnson S},
+\textbf{Harrington SC}, \textbf{Yab TC}, \textbf{Smyrk TC},
+\textbf{Nelson H}, \textbf{Boardman LA}, \textbf{Druliner BR},
+\textbf{Levin TR}, \textbf{Rex DK}, \textbf{Ahnen DJ}, \textbf{Lance P},
+\textbf{Ahlquist DA}, \textbf{Chia N}. 2017. Shifts in the Fecal
+Microbiota Associated with Adenomatous Polyps. Cancer Epidemiology,
+Biomarkers \& Prevention: A Publication of the American Association for
+Cancer Research, Cosponsored by the American Society of Preventive
+Oncology \textbf{26}:85--94.
+doi:\href{https://doi.org/10.1158/1055-9965.EPI-16-0337}{10.1158/1055-9965.EPI-16-0337}.
 
 \hypertarget{ref-shah_leveraging_2017}{}
-28. \textbf{Shah, M. S.}, \textbf{T. Z. DeSantis}, \textbf{T.
-Weinmaier}, \textbf{P. J. McMurdie}, \textbf{J. L. Cope}, \textbf{A.
-Altrichter}, \textbf{J.-M. Yamal}, and \textbf{E. B. Hollister}. 2017.
-Leveraging sequence-based faecal microbial community survey data to
-identify a composite biomarker for colorectal cancer. Gut.
+28. \textbf{Shah MS}, \textbf{DeSantis TZ}, \textbf{Weinmaier T},
+\textbf{McMurdie PJ}, \textbf{Cope JL}, \textbf{Altrichter A},
+\textbf{Yamal J-M}, \textbf{Hollister EB}. 2017. Leveraging
+sequence-based faecal microbial community survey data to identify a
+composite biomarker for colorectal cancer. Gut.
+doi:\href{https://doi.org/10.1136/gutjnl-2016-313189}{10.1136/gutjnl-2016-313189}.
 
 \hypertarget{ref-brim_microbiome_2013}{}
-29. \textbf{Brim, H.}, \textbf{S. Yooseph}, \textbf{E. G. Zoetendal},
-\textbf{E. Lee}, \textbf{M. Torralbo}, \textbf{A. O. Laiyemo},
-\textbf{B. Shokrani}, \textbf{K. Nelson}, and \textbf{H. Ashktorab}.
-2013. Microbiome analysis of stool samples from African Americans with
-colon polyps. PloS One \textbf{8}:e81352.
+29. \textbf{Brim H}, \textbf{Yooseph S}, \textbf{Zoetendal EG},
+\textbf{Lee E}, \textbf{Torralbo M}, \textbf{Laiyemo AO},
+\textbf{Shokrani B}, \textbf{Nelson K}, \textbf{Ashktorab H}. 2013.
+Microbiome analysis of stool samples from African Americans with colon
+polyps. PloS One \textbf{8}:e81352.
+doi:\href{https://doi.org/10.1371/journal.pone.0081352}{10.1371/journal.pone.0081352}.
 
 \hypertarget{ref-Sze2017}{}
-30. \textbf{Sze, M. A.}, \textbf{N. T. Baxter}, \textbf{M. T. Ruffin},
-\textbf{M. A. M. Rogers}, and \textbf{P. D. Schloss}. 2017.
-Normalization of the microbiota in patients after treatment for colonic
-lesions. Microbiome. Springer Nature \textbf{5}.
+30. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
+\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
+microbiota in patients after treatment for colonic lesions. Microbiome
+\textbf{5}.
+doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.
 
 \hypertarget{ref-Hannigan2017}{}
-31. \textbf{Hannigan, G. D.}, \textbf{M. B. Duhaime}, \textbf{M. T.
-Ruffin}, \textbf{C. C. Koumpouras}, and \textbf{P. D. Schloss}. 2017.
-Diagnostic potential \& the interactive dynamics of the colorectal
-cancer virome. Cold Spring Harbor Laboratory.
+31. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
+\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
+\& the interactive dynamics of the colorectal cancer virome.
+doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.
 
 \hypertarget{ref-Venkataraman2016}{}
-32. \textbf{Venkataraman, A.}, \textbf{J. R. Sieber}, \textbf{A. W.
-Schmidt}, \textbf{C. Waldron}, \textbf{K. R. Theis}, and \textbf{T. M.
-Schmidt}. 2016. Variable responses of human microbiomes to dietary
-supplementation with resistant starch. Microbiome. Springer Nature
-\textbf{4}.
+32. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
+\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
+Variable responses of human microbiomes to dietary supplementation with
+resistant starch. Microbiome \textbf{4}.
+doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.
 
 \hypertarget{ref-Herrmann2018}{}
-33. \textbf{Herrmann, E.}, \textbf{W. Young}, \textbf{V.
-Reichert-Grimm}, \textbf{S. Weis}, \textbf{C. Riedel}, \textbf{D.
-Rosendale}, \textbf{H. Stoklosinski}, \textbf{M. Hunt}, and \textbf{M.
-Egert}. 2018. In vivo assessment of resistant starch degradation by the
-caecal microbiota of mice using RNA-based stable isotope probingA
-proof-of-principle study. Nutrients. MDPI AG \textbf{10}:179.
+33. \textbf{Herrmann E}, \textbf{Young W}, \textbf{Reichert-Grimm V},
+\textbf{Weis S}, \textbf{Riedel C}, \textbf{Rosendale D},
+\textbf{Stoklosinski H}, \textbf{Hunt M}, \textbf{Egert M}. 2018. In
+vivo assessment of resistant starch degradation by the caecal microbiota
+of mice using RNA-based stable isotope probingA proof-of-principle
+study. Nutrients \textbf{10}:179.
+doi:\href{https://doi.org/10.3390/nu10020179}{10.3390/nu10020179}.
 
 \hypertarget{ref-Reichardt2017}{}
-34. \textbf{Reichardt, N.}, \textbf{M. Vollmer}, \textbf{G. Holtrop},
-\textbf{F. M. Farquharson}, \textbf{D. Wefers}, \textbf{M. Bunzel},
-\textbf{S. H. Duncan}, \textbf{J. E. Drew}, \textbf{L. M. Williams},
-\textbf{G. Milligan}, \textbf{T. Preston}, \textbf{D. Morrison},
-\textbf{H. J. Flint}, and \textbf{P. Louis}. 2017. Specific
-substrate-driven changes in human faecal microbiota composition contrast
-with functional redundancy in short-chain fatty acid production. The
-ISME Journal. Springer Nature \textbf{12}:610--622.
-
-\hypertarget{ref-Flynn_preprint_2017}{}
-35. \textbf{Flynn, K. J.}, \textbf{M. T. Ruffin}, \textbf{D. K.
-Turgeon}, and \textbf{P. D. Schloss}. 2017. Spatial variation of the
-native colon microbiota in healthy adults. Cold Spring Harbor
-Laboratory.
+34. \textbf{Reichardt N}, \textbf{Vollmer M}, \textbf{Holtrop G},
+\textbf{Farquharson FM}, \textbf{Wefers D}, \textbf{Bunzel M},
+\textbf{Duncan SH}, \textbf{Drew JE}, \textbf{Williams LM},
+\textbf{Milligan G}, \textbf{Preston T}, \textbf{Morrison D},
+\textbf{Flint HJ}, \textbf{Louis P}. 2017. Specific substrate-driven
+changes in human faecal microbiota composition contrast with functional
+redundancy in short-chain fatty acid production. The ISME Journal
+\textbf{12}:610--622.
+doi:\href{https://doi.org/10.1038/ismej.2017.196}{10.1038/ismej.2017.196}.
+
+\hypertarget{ref-Flynn_2018}{}
+35. \textbf{Flynn KJ}, \textbf{Ruffin MT}, \textbf{Turgeon DK},
+\textbf{Schloss PD}. 2018. Spatial variation of the native colon
+microbiota in healthy adults. Cancer Prevention and Research \textbf{In
+Press}.
 
 \hypertarget{ref-Repass2018}{}
-36. \textbf{Repass, J.}, \textbf{E. Iorns}, \textbf{A. Denis},
-\textbf{S. R. Williams}, \textbf{N. Perfito}, and \textbf{T. M. E. and}.
-2018. Replication study: Fusobacterium nucleatum infection is prevalent
-in human colorectal carcinoma. eLife. eLife Sciences Organisation, Ltd.
-\textbf{7}.
+36. \textbf{Repass J}, \textbf{Iorns E}, \textbf{Denis A},
+\textbf{Williams SR}, \textbf{Perfito N}, \textbf{and TME}. 2018.
+Replication study: Fusobacterium nucleatum infection is prevalent in
+human colorectal carcinoma. eLife \textbf{7}.
+doi:\href{https://doi.org/10.7554/elife.25801}{10.7554/elife.25801}.
 
 \hypertarget{ref-Salter_contamination_2014}{}
-37. \textbf{Salter, S. J.}, \textbf{M. J. Cox}, \textbf{E. M. Turek},
-\textbf{S. T. Calus}, \textbf{W. O. Cookson}, \textbf{M. F. Moffatt},
-\textbf{P. Turner}, \textbf{J. Parkhill}, \textbf{N. J. Loman}, and
-\textbf{A. W. Walker}. 2014. Reagent and laboratory contamination can
-critically impact sequence-based microbiome analyses. BMC Biology.
-Springer Nature \textbf{12}.
+37. \textbf{Salter SJ}, \textbf{Cox MJ}, \textbf{Turek EM},
+\textbf{Calus ST}, \textbf{Cookson WO}, \textbf{Moffatt MF},
+\textbf{Turner P}, \textbf{Parkhill J}, \textbf{Loman NJ},
+\textbf{Walker AW}. 2014. Reagent and laboratory contamination can
+critically impact sequence-based microbiome analyses. BMC Biology
+\textbf{12}.
+doi:\href{https://doi.org/10.1186/s12915-014-0087-z}{10.1186/s12915-014-0087-z}.
 
 \hypertarget{ref-Purcell2017}{}
-38. \textbf{Purcell, R. V.}, \textbf{J. Pearson}, \textbf{A. Aitchison},
-\textbf{L. Dixon}, \textbf{F. A. Frizelle}, and \textbf{J. I. Keenan}.
-2017. Colonization with enterotoxigenic bacteroides fragilis is
-associated with early-stage colorectal neoplasia. PLOS ONE. Public
-Library of Science (PLoS) \textbf{12}:e0171602.
+38. \textbf{Purcell RV}, \textbf{Pearson J}, \textbf{Aitchison A},
+\textbf{Dixon L}, \textbf{Frizelle FA}, \textbf{Keenan JI}. 2017.
+Colonization with enterotoxigenic bacteroides fragilis is associated
+with early-stage colorectal neoplasia. PLOS ONE \textbf{12}:e0171602.
+doi:\href{https://doi.org/10.1371/journal.pone.0171602}{10.1371/journal.pone.0171602}.
 
 \hypertarget{ref-Sze2016}{}
-39. \textbf{Sze, M. A.}, and \textbf{P. D. Schloss}. 2016. Looking for a
-signal in the noise: Revisiting obesity and the microbiome. mBio.
-American Society for Microbiology \textbf{7}:e01018--16.
+39. \textbf{Sze MA}, \textbf{Schloss PD}. 2016. Looking for a signal in
+the noise: Revisiting obesity and the microbiome. mBio
+\textbf{7}:e01018--16.
+doi:\href{https://doi.org/10.1128/mbio.01018-16}{10.1128/mbio.01018-16}.
 
 \hypertarget{ref-Walters2014}{}
-40. \textbf{Walters, W. A.}, \textbf{Z. Xu}, and \textbf{R. Knight}.
-2014. Meta-analyses of human gut microbes associated with obesity and
-IBD. FEBS Letters. Wiley-Blackwell \textbf{588}:4223--4233.
+40. \textbf{Walters WA}, \textbf{Xu Z}, \textbf{Knight R}. 2014.
+Meta-analyses of human gut microbes associated with obesity and IBD.
+FEBS Letters \textbf{588}:4223--4233.
+doi:\href{https://doi.org/10.1016/j.febslet.2014.09.039}{10.1016/j.febslet.2014.09.039}.
 
 \hypertarget{ref-Finucane2014}{}
-41. \textbf{Finucane, M. M.}, \textbf{T. J. Sharpton}, \textbf{T. J.
-Laurent}, and \textbf{K. S. Pollard}. 2014. A taxonomic signature of
-obesity in the microbiome? Getting to the guts of the matter. PLoS ONE.
-Public Library of Science (PLoS) \textbf{9}:e84689.
+41. \textbf{Finucane MM}, \textbf{Sharpton TJ}, \textbf{Laurent TJ},
+\textbf{Pollard KS}. 2014. A taxonomic signature of obesity in the
+microbiome? Getting to the guts of the matter. PLoS ONE
+\textbf{9}:e84689.
+doi:\href{https://doi.org/10.1371/journal.pone.0084689}{10.1371/journal.pone.0084689}.
 
 \hypertarget{ref-Hanage2014}{}
-42. \textbf{Hanage, W. P.} 2014. Microbiology: Microbiome science needs
-a healthy dose of scepticism. Nature. Springer Nature
-\textbf{512}:247--248.
+42. \textbf{Hanage WP}. 2014. Microbiology: Microbiome science needs a
+healthy dose of scepticism. Nature \textbf{512}:247--248.
+doi:\href{https://doi.org/10.1038/512247a}{10.1038/512247a}.
 
 \hypertarget{ref-keku_gastrointestinal_2015}{}
-43. \textbf{Keku, T. O.}, \textbf{S. Dulal}, \textbf{A. Deveaux},
-\textbf{B. Jovov}, and \textbf{X. Han}. 2015. The gastrointestinal
-microbiota and colorectal cancer. American Journal of Physiology -
-Gastrointestinal and Liver Physiology \textbf{308}:G351--G363.
+43. \textbf{Keku TO}, \textbf{Dulal S}, \textbf{Deveaux A},
+\textbf{Jovov B}, \textbf{Han X}. 2015. The gastrointestinal microbiota
+and colorectal cancer. American Journal of Physiology - Gastrointestinal
+and Liver Physiology \textbf{308}:G351--G363.
+doi:\href{https://doi.org/10.1152/ajpgi.00360.2012}{10.1152/ajpgi.00360.2012}.
 
 \hypertarget{ref-vogtmann_epidemiologic_2016}{}
-44. \textbf{Vogtmann, E.}, and \textbf{J. J. Goedert}. 2016.
-Epidemiologic studies of the human microbiome and cancer. British
-Journal of Cancer \textbf{114}:237--242.
+44. \textbf{Vogtmann E}, \textbf{Goedert JJ}. 2016. Epidemiologic
+studies of the human microbiome and cancer. British Journal of Cancer
+\textbf{114}:237--242.
+doi:\href{https://doi.org/10.1038/bjc.2015.465}{10.1038/bjc.2015.465}.
 
 \hypertarget{ref-kostic_genomic_2012}{}
-45. \textbf{Kostic, A. D.}, \textbf{D. Gevers}, \textbf{C. S.
-Pedamallu}, \textbf{M. Michaud}, \textbf{F. Duke}, \textbf{A. M. Earl},
-\textbf{A. I. Ojesina}, \textbf{J. Jung}, \textbf{A. J. Bass},
-\textbf{J. Tabernero}, \textbf{J. Baselga}, \textbf{C. Liu}, \textbf{R.
-A. Shivdasani}, \textbf{S. Ogino}, \textbf{B. W. Birren}, \textbf{C.
-Huttenhower}, \textbf{W. S. Garrett}, and \textbf{M. Meyerson}. 2012.
-Genomic analysis identifies association of Fusobacterium with colorectal
-carcinoma. Genome Research \textbf{22}:292--298.
+45. \textbf{Kostic AD}, \textbf{Gevers D}, \textbf{Pedamallu CS},
+\textbf{Michaud M}, \textbf{Duke F}, \textbf{Earl AM}, \textbf{Ojesina
+AI}, \textbf{Jung J}, \textbf{Bass AJ}, \textbf{Tabernero J},
+\textbf{Baselga J}, \textbf{Liu C}, \textbf{Shivdasani RA},
+\textbf{Ogino S}, \textbf{Birren BW}, \textbf{Huttenhower C},
+\textbf{Garrett WS}, \textbf{Meyerson M}. 2012. Genomic analysis
+identifies association of Fusobacterium with colorectal carcinoma.
+Genome Research \textbf{22}:292--298.
+doi:\href{https://doi.org/10.1101/gr.126573.111}{10.1101/gr.126573.111}.
 
 \hypertarget{ref-zackular_human_2014}{}
-46. \textbf{Zackular, J. P.}, \textbf{M. A. M. Rogers}, \textbf{M. T.
-Ruffin}, and \textbf{P. D. Schloss}. 2014. The human gut microbiome as a
-screening tool for colorectal cancer. Cancer Prevention Research
-(Philadelphia, Pa.) \textbf{7}:1112--1121.
+46. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
+\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
+for colorectal cancer. Cancer Prevention Research (Philadelphia, Pa)
+\textbf{7}:1112--1121.
+doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
 
 \hypertarget{ref-schloss_introducing_2009}{}
-47. \textbf{Schloss, P. D.}, \textbf{S. L. Westcott}, \textbf{T.
-Ryabin}, \textbf{J. R. Hall}, \textbf{M. Hartmann}, \textbf{E. B.
-Hollister}, \textbf{R. A. Lesniewski}, \textbf{B. B. Oakley}, \textbf{D.
-H. Parks}, \textbf{C. J. Robinson}, \textbf{J. W. Sahl}, \textbf{B.
-Stres}, \textbf{G. G. Thallinger}, \textbf{D. J. Van Horn}, and
-\textbf{C. F. Weber}. 2009. Introducing mothur: Open-Source,
-Platform-Independent, Community-Supported Software for Describing and
-Comparing Microbial Communities. Appl.Environ.Microbiol.
-\textbf{75}:7537--7541.
+47. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
+\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
+\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
+\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
+\textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}. 2009.
+Introducing mothur: Open-Source, Platform-Independent,
+Community-Supported Software for Describing and Comparing Microbial
+Communities. ApplEnvironMicrobiol \textbf{75}:7537--7541.
 
 \hypertarget{ref-rognes_vsearch_2016}{}
-48. \textbf{Rognes, T.}, \textbf{T. Flouri}, \textbf{B. Nichols},
-\textbf{C. Quince}, and \textbf{F. Mahé}. 2016. VSEARCH: A versatile
-open source tool for metagenomics. PeerJ \textbf{4}:e2584.
+48. \textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols B},
+\textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: A versatile open
+source tool for metagenomics. PeerJ \textbf{4}:e2584.
+doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.
 
 \hypertarget{ref-westcott_opticlust_2017}{}
-49. \textbf{Westcott, S. L.}, and \textbf{P. D. Schloss}. 2017.
-OptiClust, an Improved Method for Assigning Amplicon-Based Sequence Data
-to Operational Taxonomic Units. mSphere \textbf{2}.
+49. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
+Improved Method for Assigning Amplicon-Based Sequence Data to
+Operational Taxonomic Units. mSphere \textbf{2}.
+doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.
 
 \hypertarget{ref-rdp_Wang2007}{}
-50. \textbf{Wang, Q.}, \textbf{G. M. Garrity}, \textbf{J. M. Tiedje},
-and \textbf{J. R. Cole}. 2007. Naive bayesian classifier for rapid
-assignment of rRNA sequences into the new bacterial taxonomy. Applied
-and Environmental Microbiology. American Society for Microbiology
-\textbf{73}:5261--5267.
+50. \textbf{Wang Q}, \textbf{Garrity GM}, \textbf{Tiedje JM},
+\textbf{Cole JR}. 2007. Naive bayesian classifier for rapid assignment
+of rRNA sequences into the new bacterial taxonomy. Applied and
+Environmental Microbiology \textbf{73}:5261--5267.
+doi:\href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
 
 \hypertarget{ref-permanova_Anderson2013}{}
-51. \textbf{Anderson, M. J.}, and \textbf{D. C. I. Walsh}. 2013.
-PERMANOVA, ANOSIM, and the mantel test in the face of heterogeneous
-dispersions: What null hypothesis are you testing? Ecological
-Monographs. Wiley-Blackwell \textbf{83}:557--574.
+51. \textbf{Anderson MJ}, \textbf{Walsh DCI}. 2013. PERMANOVA, ANOSIM,
+and the mantel test in the face of heterogeneous dispersions: What null
+hypothesis are you testing? Ecological Monographs \textbf{83}:557--574.
+doi:\href{https://doi.org/10.1890/12-2010.1}{10.1890/12-2010.1}.
 
 \hypertarget{ref-benjamini_controlling_1995}{}
-52. \textbf{Benjamini, Y.}, and \textbf{Y. Hochberg}. 1995. Controlling
-the false discovery rate: A practical and powerful approach to multiple
-testing. Journal of the Royal Statistical Society. Series B
+52. \textbf{Benjamini Y}, \textbf{Hochberg Y}. 1995. Controlling the
+false discovery rate: A practical and powerful approach to multiple
+testing. Journal of the Royal Statistical Society Series B
 (Methodological) \textbf{57}:289--300.
@@
 \hypertarget{ref-Breiman2001}{}
-53. \textbf{Breiman, L.} 2001. Machine Learning. Springer Nature
-\textbf{45}:5--32.
+53. \textbf{Breiman L}. 2001. Machine Learning \textbf{45}:5--32.
+doi:\href{https://doi.org/10.1023/a:1010933404324}{10.1023/a:1010933404324}.
 
@@
 \hypertarget{ref-rcompanion_citation_2017}{}
-55. \textbf{Mangiafico, S.} 2017. Rcompanion: Functions to support
+55. \textbf{Mangiafico S}. 2017. Rcompanion: Functions to support
 extension education program evaluation.
@@
 \hypertarget{ref-car_citation_2011}{}
-56. \textbf{Fox, J.}, and \textbf{S. Weisberg}. 2011. An R companion to
-applied regressionSecond. Sage, Thousand Oaks CA.
+56. \textbf{Fox J}, \textbf{Weisberg S}. 2011. An R companion to applied
+regressionSecond. Sage, Thousand Oaks CA.
 
 \hypertarget{ref-lme4_citation_2015}{}
-57. \textbf{Bates, D.}, \textbf{M. Mächler}, \textbf{B. Bolker}, and
-\textbf{S. Walker}. 2015. Fitting linear mixed-effects models using
-lme4. Journal of Statistical Software \textbf{67}:1--48.
+57. \textbf{Bates D}, \textbf{Mächler M}, \textbf{Bolker B},
+\textbf{Walker S}. 2015. Fitting linear mixed-effects models using lme4.
+Journal of Statistical Software \textbf{67}:1--48.
+doi:\href{https://doi.org/10.18637/jss.v067.i01}{10.18637/jss.v067.i01}.
 
 \hypertarget{ref-epir_citation_2017}{}
-58. \textbf{Telmo Nunes, M. S. with contributions from}, \textbf{C.
-Heuer}, \textbf{J. Marshall}, \textbf{J. Sanchez}, \textbf{R. Thornton},
-\textbf{J. Reiczigel}, \textbf{J. Robison-Cox}, \textbf{P. Sebastiani},
-\textbf{P. Solymos}, \textbf{K. Yoshida}, \textbf{G. Jones}, \textbf{S.
-Pirikahu}, \textbf{S. Firestone}, and \textbf{R. Kyle.} 2017. EpiR:
+58. \textbf{Telmo Nunes MS with contributions from}, \textbf{Heuer C},
+\textbf{Marshall J}, \textbf{Sanchez J}, \textbf{Thornton R},
+\textbf{Reiczigel J}, \textbf{Robison-Cox J}, \textbf{Sebastiani P},
+\textbf{Solymos P}, \textbf{Yoshida K}, \textbf{Jones G},
+\textbf{Pirikahu S}, \textbf{Firestone S}, \textbf{Kyle. R}. 2017. EpiR:
 Tools for the analysis of epidemiological data.
@@
 \hypertarget{ref-metafor_citation_2010}{}
-59. \textbf{Viechtbauer, W.} 2010. Conducting meta-analyses in R with
-the metafor package. Journal of Statistical Software \textbf{36}:1--48.
+59. \textbf{Viechtbauer W}. 2010. Conducting meta-analyses in R with the
+metafor package. Journal of Statistical Software \textbf{36}:1--48.
 
 \hypertarget{ref-vegan_citation_2017}{}
-60. \textbf{Oksanen, J.}, \textbf{F. G. Blanchet}, \textbf{M. Friendly},
-\textbf{R. Kindt}, \textbf{P. Legendre}, \textbf{D. McGlinn}, \textbf{P.
-R. Minchin}, \textbf{R. B. O'Hara}, \textbf{G. L. Simpson}, \textbf{P.
-Solymos}, \textbf{M. H. H. Stevens}, \textbf{E. Szoecs}, and \textbf{H.
-Wagner}. 2017. Vegan: Community ecology package.
+60. \textbf{Oksanen J}, \textbf{Blanchet FG}, \textbf{Friendly M},
+\textbf{Kindt R}, \textbf{Legendre P}, \textbf{McGlinn D},
+\textbf{Minchin PR}, \textbf{O'Hara RB}, \textbf{Simpson GL},
+\textbf{Solymos P}, \textbf{Stevens MHH}, \textbf{Szoecs E},
+\textbf{Wagner H}. 2017. Vegan: Community ecology package.
 
 \hypertarget{ref-caret_citation_2017}{}
-61. \textbf{Jed Wing, M. K. C. from}, \textbf{S. Weston}, \textbf{A.
-Williams}, \textbf{C. Keefer}, \textbf{A. Engelhardt}, \textbf{T.
-Cooper}, \textbf{Z. Mayer}, \textbf{B. Kenkel}, \textbf{the R Core
-Team}, \textbf{M. Benesty}, \textbf{R. Lescarbeau}, \textbf{A. Ziem},
-\textbf{L. Scrucca}, \textbf{Y. Tang}, \textbf{C. Candan}, and
-\textbf{T. Hunt.} 2017. Caret: Classification and regression training.
+61. \textbf{Jed Wing MKC from}, \textbf{Weston S}, \textbf{Williams A},
+\textbf{Keefer C}, \textbf{Engelhardt A}, \textbf{Cooper T},
+\textbf{Mayer Z}, \textbf{Kenkel B}, \textbf{R Core Team},
+\textbf{Benesty M}, \textbf{Lescarbeau R}, \textbf{Ziem A},
+\textbf{Scrucca L}, \textbf{Tang Y}, \textbf{Candan C}, \textbf{Hunt.
+T}. 2017. Caret: Classification and regression training.
 
 \hypertarget{ref-randomforest_citation_2002}{}
-62. \textbf{Liaw, A.}, and \textbf{M. Wiener}. 2002. Classification and
+62. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
 regression by randomForest. R News \textbf{2}:18--22.
@@
 \hypertarget{ref-ggplot2_citation_2009}{}
-63. \textbf{Wickham, H.} 2009. Ggplot2: Elegant graphics for data
+63. \textbf{Wickham H}. 2009. Ggplot2: Elegant graphics for data
 analysis. Springer-Verlag New York.
@@
 \hypertarget{ref-gridextra_citation_2017}{}
-64. \textbf{Auguie, B.} 2017. GridExtra: Miscellaneous functions for
+64. \textbf{Auguie B}. 2017. GridExtra: Miscellaneous functions for
 ``grid'' graphics.

### 2018-03-16 submission/manuscript.Rmd
remove typos and incorporate Pat's edits

@@
 We analyzed previously published 16S rRNA gene sequencing data collected from feces and colon tissue.
-
 We quantified the odds ratios (ORs) for individual bacterial taxa that were associated with an individual having tumors relative to a normal colon.
@@
 To classify individuals as having normal colons or tumors, we built Random Forest classification models for each dataset and comparison using taxa with significant ORs (after multiple comparison correction), all taxa, or OTUs.
-Because no taxa were identified as having a significant OR associated with adenomas using stool samples or tissue samples, classification models based on OR data were not constructed to classify individuals as having normal colons or adenomas.
-For all models the value of trees included (i.e.
+Because no taxa were identified as having a significant OR associated with adenomas using stool or tissue samples, classification models based on OR data were not constructed to classify individuals as having normal colons or adenomas.
+For all models, the value of trees included (i.e.
 ntree) was set to 500 and the number of variables that were randomly tested (i.e.
@@
 All fecal models were built using a 10-fold cross validation (CV) while tissue models were built using 5-fold CV due to study sample size.
-One exception to this were the models constructed using data from the Weir study, which was built using a 2-fold CV due to the small number of samples sample.
+One exception to this were the models constructed using data from the Weir study, which was built using a 2-fold CV due to the small number of samples.
 For models constructed based on the taxa that had a significant OR or using all of the taxa, we trained the models using a single study and then tested on the remaining studies with AUCs recorded during both train and testing phases.

### 2018-03-16 submission/manuscript.tex
remove typos and incorporate Pat's edits

@@
 previously published 16S rRNA gene sequencing data collected from feces
-and colon tissue.
-
-We quantified the odds ratios (ORs) for individual bacterial taxa that
-were associated with an individual having tumors relative to a normal
-colon. Among the fecal samples, there were no taxa that had significant
-ORs associated with adenoma and there were 8 taxa with significant ORs
-associated with carcinoma. Similarly, among the tissue samples, there
-were no taxa that had a significant OR associated with adenoma and there
-were 3 taxa with significant ORs associated with carcinoma. Among the
-significant ORs, the association between individual taxa and tumor
-diagnosis was equal or below 7.11. Because individual taxa had limited
-association with tumor diagnosis, we trained Random Forest
-classification models using only the taxa that had significant ORs,
-using the entire collection of taxa found in each study, and using
+and colon tissue. We quantified the odds ratios (ORs) for individual
+bacterial taxa that were associated with an individual having tumors
+relative to a normal colon. Among the fecal samples, there were no taxa
+that had significant ORs associated with adenoma and there were 8 taxa
+with significant ORs associated with carcinoma. Similarly, among the
+tissue samples, there were no taxa that had a significant OR associated
+with adenoma and there were 3 taxa with significant ORs associated with
+carcinoma. Among the significant ORs, the association between individual
+taxa and tumor diagnosis was equal or below 7.11. Because individual
+taxa had limited association with tumor diagnosis, we trained Random
+Forest classification models using only the taxa that had significant
+ORs, using the entire collection of taxa found in each study, and using
 operational taxonomic units defined based on a 97\% similarity
@@
 OTUs. Because no taxa were identified as having a significant OR
-associated with adenomas using stool samples or tissue samples,
-classification models based on OR data were not constructed to classify
-individuals as having normal colons or adenomas. For all models the
-value of trees included (i.e.~ntree) was set to 500 and the number of
-variables that were randomly tested (i.e.~mtry) was set to the square
-root of the number of taxa or OTUs within the model. Using the square
-root of the total number of features as the number of features to test
-has been found to reliably approximate the optimum value after model
-tuning (53). All fecal models were built using a 10-fold cross
-validation (CV) while tissue models were built using 5-fold CV due to
-study sample size. One exception to this were the models constructed
-using data from the Weir study, which was built using a 2-fold CV due to
-the small number of samples sample. For models constructed based on the
-taxa that had a significant OR or using all of the taxa, we trained the
-models using a single study and then tested on the remaining studies
-with AUCs recorded during both train and testing phases. For the models
-constructed using OTU data, 100 10-fold CVs were run to generate a range
-of AUCs that could be reasonably expected to occur. The average AUC from
-these 100 repeats was reported. The Mean Decrease in Accuracy (MDA), a
-measure of the importance of each taxon to the overall model, was used
-to rank the taxa used in each model.
+associated with adenomas using stool or tissue samples, classification
+models based on OR data were not constructed to classify individuals as
+having normal colons or adenomas. For all models, the value of trees
+included (i.e.~ntree) was set to 500 and the number of variables that
+were randomly tested (i.e.~mtry) was set to the square root of the
+number of taxa or OTUs within the model. Using the square root of the
+total number of features as the number of features to test has been
+found to reliably approximate the optimum value after model tuning (53).
+All fecal models were built using a 10-fold cross validation (CV) while
+tissue models were built using 5-fold CV due to study sample size. One
+exception to this were the models constructed using data from the Weir
+study, which was built using a 2-fold CV due to the small number of
+samples. For models constructed based on the taxa that had a significant
+OR or using all of the taxa, we trained the models using a single study
+and then tested on the remaining studies with AUCs recorded during both
+train and testing phases. For the models constructed using OTU data, 100
+10-fold CVs were run to generate a range of AUCs that could be
+reasonably expected to occur. The average AUC from these 100 repeats was
+reported. The Mean Decrease in Accuracy (MDA), a measure of the
+importance of each taxon to the overall model, was used to rank the taxa
+used in each model.
 
@@
 the Bray-Curtis dissimilarity index and PERMANOVA was used to test for
-tumor-associated differences in structure with the vegan (v2.4.4)
+tumor-associated differences in structure with the vegan (v2.4.5)
 package (60). Random Forest models were built using both the caret
-(v6.0.77) and randomForest (v4.6.12) packages (61, 62). All figures were
-created using both ggplot2 (v2.2.1.9000) and gridExtra (v2.3) packages
-(63, 64).
+(v6.0.78) and randomForest (v4.6.12) packages (61, 62). All figures were
+created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages (63,
+64).
 

### 2018-03-19 submission/manuscript.tex
update title

@@
 
-\section{Making Sense of the Noise: Leveraging Existing 16S rRNA Gene
-Surveys to Identify Key Community Members in Colorectal
-Tumors}\label{making-sense-of-the-noise-leveraging-existing-16s-rrna-gene-surveys-to-identify-key-community-members-in-colorectal-tumors}
+\section{Leveraging Existing 16S rRNA Gene Surveys to Identify
+Reproducible Biomarkers in Individuals with Colorectal
+Tumors}\label{leveraging-existing-16s-rrna-gene-surveys-to-identify-reproducible-biomarkers-in-individuals-with-colorectal-tumors}
 

### 2018-03-19 submission/supplemental.Rmd
Fix Table S4 formatting issues
update title

@@
 
-# Making Sense of the Noise: Leveraging Existing 16S rRNA Gene Surveys to Identify Key Community Members in Colorectal Tumors
+# Leveraging Existing 16S rRNA Gene Surveys to Identify Reproducible Biomarkers in Individuals with Colorectal Tumors
 
@@
 
-\footnotesize
+\scriptsize
 
@@
 
-\blandscape
-
 **Table S4: ORs for individual taxa associated with individuals who had a normal colon or adenomas or carcinomas using data collected from tissue samples.** The listed P-values were less than 0.05 prior to using a Benjimani-Hochberg correction for multiple comparisons.
 
-\footnotesize
+\scriptsize
 
@@
 
-kable(combined_table_S5, format = "markdown", align = 'c', padding = 2)
+kable(combined_table_S5, format = "markdown", align = 'c', padding = 0)
 
 ```
-
-\normalsize
-
-\elandscape
-\newpage

### 2018-05-08 code/Run_render_revision1_paper.R
add typo fix and acknowledgement addition

@@ -1,0 +1,13 @@
+### List of commands to render final pdf
+### renders both main text and supplemental
+
+# Load needed functions
+source('code/functions.R')
+
+# Load needed library
+loadLibs(c("knitr", "rmarkdown"))
+
+# Render the final pdfs
+render('submission/manuscript_R1.Rmd', clean = FALSE)
+
+#render('submission/supplemental_R1.Rmd', clean=FALSE)

### 2018-05-08 submission/manuscript_R1.Rmd
final changes

--- 2018-03-16 submission/manuscript.Rmd
@@
 All training approaches yielded similar classification success as measured using the Area Under the Curve.
-The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from fecal or tissue.
+The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from feces or tissue.
 
@@
 Finally, we would like to thank the members of the Schloss lab for their valuable feedback and proofreading during the formulation of this manuscript.
+Salary support for Marc A.
+Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240.
+Salary support for Patrick D.
+Schloss came from NIH grants P30DK034933 and 1R01CA215574.
 

### 2018-05-08 submission/manuscript_R1.tex
final changes

--- 2018-03-19 submission/manuscript.tex
@@
 classify individuals with carcinomas was considerably better using
-sequences from fecal or tissue.
+sequences from feces or tissue.
 
@@
 \textbf{\emph{Statistical Analysis.}} All statistical analysis after
-sequence processing utilized the R (v3.4.3) software package (54). For
+sequence processing utilized the R (v3.4.4) software package (54). For
 OTU richness, evenness, and Shannon diversity analysis, values were
@@
 of the Schloss lab for their valuable feedback and proofreading during
-the formulation of this manuscript.
+the formulation of this manuscript. Salary support for Marc A. Sze came
+from the Canadian Institute of Health Research and NIH grant
+UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants
+P30DK034933 and 1R01CA215574.
 

### 2018-05-09 submission/supplemental.Rmd
Add supplemental table legends and break up supp tables

@@
     keep_tex: yes
-geometry: margin=1.0in
+geometry: margin=0.75in
 bibliography: null
 ---
-
-# Leveraging Existing 16S rRNA Gene Surveys to Identify Reproducible Biomarkers in Individuals with Colorectal Tumors
-
-\vspace{10mm}
-
-\begin{center}
-Marc A Sze and Patrick D Schloss
-
-\vspace{10mm}
-
-Supplemental
-\end{center}
-
-\newpage
-
+\pagenumbering{gobble}
 ```{r knitr_settings, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}

### 2018-09-17 code/make_genus_rf_auc_against_study_graph.R
update figs rel sana

@@ -55,7 +55,7 @@ adn_all_stool <- make_table(adn_stool_studies, "data/process/tables/",
                             "_pvalue_summary.csv",select_included = F)
 
 # Load needed data (adenoma -- tissue)
-adn_tissue_studies <- c("flemer", "lu")
+adn_tissue_studies <- c("flemer", "lu", "sana")
 
 adn_all_tissue <- make_table(adn_tissue_studies, "data/process/tables/", 
                             "adn_ALL_genus_unmatched_tissue_RF_full_", 
@@ -80,7 +80,7 @@ crc_all_matched_tissue <- make_table(crc_matched_tissue_studies, "data/process/t
                             "_pvalue_summary.csv", select_included = F)
 
 
-crc_unmatched_tissue_studies <- c("burns", "chen", "flemer", "sana")
+crc_unmatched_tissue_studies <- c("burns", "chen", "flemer")
 
 crc_all_unmatched_tissue <- make_table(crc_unmatched_tissue_studies, "data/process/tables/", 
                                      "ALL_genus_unmatched_tissue_RF_", 
@@ -114,14 +114,14 @@ crc_all_unmatched_tissue <- make_table(crc_unmatched_tissue_studies, "data/proce
 adn_tissue_graph <- adn_all_tissue %>% 
   filter(study != model) %>% 
   mutate(model = factor(model, 
-                        levels = c("flemer", "lu"), 
-                        labels = c("Flemer", "Lu\n(Matched)")), 
+                        levels = c("flemer", "sana", "lu"), 
+                        labels = c("Flemer", "Sanapareddy", "Lu\n(Matched)")), 
          model_type = factor(model_type, 
                              levels = c("select", "full"), 
                              labels = c("Significant\nOR Taxa", "All Taxa")), 
          study = factor(study, 
-                        levels = c("flemer", "lu"), 
-                        labels = c("Flemer", "Lu\n(Matched)"))) %>% 
+                        levels = c("flemer", "sana", "lu"), 
+                        labels = c("Flemer", "Sanapareddy", "Lu\n(Matched)"))) %>% 
   ggplot(aes(model, auc, color = study, group = model_type)) + 
   geom_hline(color = "gray", yintercept = 0.5, linetype = "dashed") + 
   geom_point(size = 3.5, position = position_dodge(width = 0.3), show.legend = T) + 
[…]

### 2018-09-17 code/make_rf_auc_full_versus_specific_graph.R

@@ -14,13 +14,13 @@ loadLibs(c("tidyverse", "gridExtra", "viridis"))
 #  mutate(type = "matched")
 adn_tissue <- 
   read_csv("data/process/tables/adn_ALL_genus_unmatched_tissue_RF_fullvsselect_pvalue_summary.csv") %>% 
-  mutate(type = rep(c("unmatched", "matched"), 2), 
+  mutate(type = rep(c("unmatched", "matched", "unmatched"), 3), 
          pvalue = as.numeric(pvalue)) %>% 
   gather(key = model_type, value = AUC, full_model, select_model) %>% 
   mutate(AUC = as.numeric(AUC)) %>% 
   bind_rows(
     read_csv("data/process/tables/adn_tissue_rf_otu_random_comparison_summary.csv") %>% 
-      mutate(type = c("matched", "unmatched"), 
+      mutate(type = c("matched", "unmatched", "unmatched"), 
              model_type = "full_otu", 
              train_model = study, 
              act_mean_auc = as.numeric(act_mean_auc)) %>% 

### 2018-09-17 code/make_tissue_alpha_OR_graph.R

@@ -13,7 +13,7 @@ adn_all_tissue <- read_csv("data/process/tables/alpha_adn_OR_tissue_composite.cs
   rename(est = rr, lower = ci_lb, upper = ci_ub) %>% 
   mutate(study = "composite") %>% 
   bind_rows(read_csv("data/process/tables/alpha_adn_OR_ind_tissue_results.csv")) %>% 
-  mutate(region = c(rep("combined", 3), rep("V3-V4", 6)))
+  mutate(region = c(rep("combined", 3), rep("V3-V4", 3), rep("V1-V2", 3), rep("V3-V4", 3)))
 
 
 crc_all_tissue <- read_csv("data/process/tables/alpha_OR_tissue_composite.csv") %>% 
@@ -21,7 +21,7 @@ crc_all_tissue <- read_csv("data/process/tables/alpha_OR_tissue_composite.csv") 
   mutate(study = "composite") %>% 
   bind_rows(read_csv("data/process/tables/alpha_OR_ind_tissue_results.csv")) %>% 
   mutate(region = c(rep("combined", 3), rep("V5-V6", 3), rep("V1-V3", 3), rep("V3-V4", 3), 
-                    rep("V1-V2", 3), rep("V3-V5", 3), rep("V1-V2", 3)))
+                    rep("V3-V5", 3), rep("V1-V2", 3)))
 
 
 
@@ -48,11 +48,11 @@ crc_all_tissue <- read_csv("data/process/tables/alpha_OR_tissue_composite.csv") 
 
 adn_tissue_graph <- adn_all_tissue %>% 
   mutate(study = factor(study, 
-                        levels = c("composite", "flemer", "lu"), 
-                        labels = c( "Pooled", "Flemer", "Lu")), 
+                        levels = c("composite", "flemer", "lu", "sana"), 
+                        labels = c( "Pooled", "Flemer", "Lu", "Sanapareddy")), 
          region = factor(region, 
-                         levels = c("combined", "V3-V4"), 
-                         labels = c("Combined", "V3-V4")), 
+                         levels = c("combined", "V1-V2", "V3-V4"), 
+                         labels = c("Combined", "V1-V2", "V3-V4")), 
          measure = factor(measure, 
                           levels = c("sobs", "shannoneven", "shannon"), 
                           labels = c("Observed OTUs", "Evenness", "Shannon Diversity")))  %>% 
[…]

### 2018-09-17 code/make_tissue_imp_otu_graph.R

@@ -92,8 +92,8 @@ crc_matched_genera_graph <- crc_matched_genera %>%
 
 crc_unmatched_genera_graph <- crc_unmatched_genera %>% 
   mutate(study = factor(study, 
-                        levels = c("burns", "chen", "flemer", "sana"), 
-                        labels= c("Burns", "Chen", "Flemer", "Sanapareddy")), 
+                        levels = c("burns", "chen", "flemer"), 
+                        labels= c("Burns", "Chen", "Flemer")), 
          otu = factor(otu, 
                       levels = unmatched_genera_labs$otu, 
                       labels = unmatched_genera_labs$otu)) %>% 

### 2018-09-17 submission/manuscript_R1.Rmd
update re sana

@@
   as.data.frame(read_csv("../data/process/tables/adn_ALL_genus_unmatched_tissue_RF_fullvsselect_pvalue_summary.csv") %>%
-  mutate(type = rep(c("unmatched", "matched"), 2),
+                  mutate(type = rep(c("unmatched", "matched", "unmatched"), 3),
                          pvalue = as.numeric(pvalue)) %>%
@@
                     read_csv("../data/process/tables/adn_tissue_rf_otu_random_comparison_summary.csv") %>%
-      mutate(type = c("matched", "unmatched"),
+                      mutate(type = c("matched", "unmatched", "unmatched"),
                              model_type = "full_otu",

### 2018-09-17 submission/manuscript_R1.tex
update re sana

@@
 \setlength{\droptitle}{-2em}
+
   \title{}
@@
 tissue samples, there were no taxa that had a significant OR associated
-with adenoma and there were 3 taxa with significant ORs associated with
+with adenoma and there were 6 taxa with significant ORs associated with
 carcinoma. Among the significant ORs, the association between individual
@@
 Here we provide an updated meta-analysis using 16S rRNA gene sequence
-data from both feces (n=1737) and colon tissue (492 samples from 350
+data from both feces (n=1737) and colon tissue (435 samples from NA
 individuals) from 14 studies (11--17, 19, 20, 23, 25--27, 29) {[}Table 1
@@
 only unmatched carcinoma samples had taxa with a significant OR. Those
-included \emph{Dorea} (OR=0.35 (0.22 - 0.55)), \emph{Blautia} (OR=0.47
-(0.3 - 0.73)), and \emph{Weissella} (OR=5.15 (2.02 - 13.14)).
-Mouth-associated genera were not significantly associated with a higher
-OR for carcinoma in tissue samples {[}Table S4{]}. For example,
-\emph{Fusobacterium} had an OR of 3.98 (1.19 - 13.24); however, due to
-the small number of studies and considerable variation in the data, the
-Benjimani-Hochberg corrected P-value was 0.93 {[}Table S4{]}. It is
-interesting to note that \emph{Ruminococcus} and members of
-\emph{Clostridium XI} in fecal samples and \emph{Dorea} and
-\emph{Blautia} in tissue had ORs that were significantly less than 1.0,
-which suggests that these populations are protective against the
-development of carcinomas. Overall, there was no overlap in the taxa
-with significant OR between fecal and tissue samples.
+included \emph{Dorea} (OR=0.34 (0.21 - 0.57)), \emph{Blautia} (OR=0.41
+(0.25 - 0.67)), and \emph{Weissella} (OR= ( - )). Mouth-associated
+genera were not significantly associated with a higher OR for carcinoma
+in tissue samples {[}Table S4{]}. For example, \emph{Fusobacterium} had
+an OR of 3.98 (1.19 - 13.24); however, due to the small number of
+studies and considerable variation in the data, the Benjimani-Hochberg
+corrected P-value was 0.93 {[}Table S4{]}. It is interesting to note
+that \emph{Ruminococcus} and members of \emph{Clostridium XI} in fecal
+samples and \emph{Dorea} and \emph{Blautia} in tissue had ORs that were
+significantly less than 1.0, which suggests that these populations are
+protective against the development of carcinomas. Overall, there was no
+overlap in the taxa with significant OR between fecal and tissue
+samples.
 
@@
 indicating their limited value as biomarkers when used individually.
-Likewise, in unmatched tissue samples the 3 taxa with significant OR
+Likewise, in unmatched tissue samples the 6 taxa with significant OR
 taxa had AUC values that were marginally better than one would expect by
 chance {[}Figure 3B{]}. The relative abundance of \emph{Dorea} was the
-best predictor of carcinomas and its median AUC was only 0.62. These
+best predictor of carcinomas and its median AUC was only 0.635. These
 results suggest that although these taxa are associated with a
@@
 {[}Figure 4A{]}. Similarly, using data from the unmatched tissue
-samples, the combined model had an AUC of 0.77, which was significantly
+samples, the combined model had an AUC of 0.88, which was significantly
 higher than the AUC values for classifying based on the relative
 abundances of \emph{Blautia} and \emph{Weissella} individually (P-value
-\textless{} 0.037). Both \emph{Dorea} and \emph{Blautia} were the most
+\textless{} 0.13). Both \emph{Dorea} and \emph{Blautia} were the most
 important taxa in the tissue-based models {[}Figure 4B{]}. Pooling the
@@
 samples were only marginally better than 0.5 for any study (median
-AUC=0.549 (range: 0.367 - 0.971)) {[}Figure 5A \& S2A{]}. In contrast,
+AUC=0.555 (range: 0.367 - 0.971)) {[}Figure 5A \& S2A{]}. In contrast,
 the models for classifying subjects as having normal colons or having
@@
 unmatched tissue samples to train classification models, we found that
-the AUC of studies was an average 19.11\% higher when we used all of the
+the AUC of studies was an average 22.86\% higher when we used all of the
 taxa rather than the 3 taxa with significant ORs (one-tailed paired
-T-test; P-value=0.03). For the models trained using data from unmatched
+T-test; P-value=0.053). For the models trained using data from unmatched
 tissue samples, \emph{Lachnospiraceae}, \emph{Bacteroidaceae}, and
@@
 OTU-based models performed similarly to those constructed using the taxa
-with significant ORs (one-tailed paired T-test; P-value=0.979) and those
-using all taxa (one-tailed paired T-test; P-value=0.184) {[}Figure 4{]}.
+with significant ORs (one-tailed paired T-test; P-value=0.945) and those
+using all taxa (one-tailed paired T-test; P-value=0.173) {[}Figure 4{]}.
 Among the OTUs that had the highest mean decrease in accuracy for the
@@
 and tissue samples (Chen and Flemer). After curating the sequences, 1737
-fecal samples and 492 tissue samples remained in the analysis {[}Tables
+fecal samples and 435 tissue samples remained in the analysis {[}Tables
 1 and 2{]}.
@@
 \endhead
-Burns & SRA & V5-6 & 18 & 0 & 16\tabularnewline
+Burns & SRA & V5-6 & 15 & 0 & 13\tabularnewline
 Chen & SRA & V1-3 & 9 & 0 & 9\tabularnewline
-Dejea & SRA & V3-5 & 31 & 0 & 32\tabularnewline
+Dejea & SRA & V3-5 & 15 & 0 & 13\tabularnewline
 Flemer & Author & V3-4 & 103 & 37 & 94\tabularnewline
-Geng & SRA & V1-2 & 16 & 0 & 16\tabularnewline
+Geng & SRA & V1-2 & 8 & 0 & 8\tabularnewline
 Lu & SRA & V3-4 & 20 & 20 & 0\tabularnewline
-Sanapareddy & Author & V1-2 & 38 & 0 & 33\tabularnewline
+Sanapareddy & Author & V1-2 & 38 & 33 & 0\tabularnewline
 \bottomrule

### 2018-09-17 submission/supplemental.Rmd
update re sana re class

@@
          pvalue = round(pvalue, digits = 4),
-         study = c("Lu", "Flemer", "Lu (Matched)"),
+         study = c("Lu", "Sanapareddy", "Flemer", "Lu (Matched)"),
          comparison = "Adenoma",
@@
          pvalue = round(pvalue, digits = 4),
-         study = c("Sanapareddy", "Burns", "Flemer", "Chen",
+         study = c("Burns", "Flemer", "Chen",
                    "Dejea (Matched)", "Geng (Matched)", "Burns (Matched)"),
